```{r}
library(tidyverse)
library(here)
library(ggsci)
library(janitor)
library(readr)
library(explore)
library(maditr)
library(GEOquery)
library(Matrix)
library(circlize)
library(RColorBrewer)
library(celldex)
library(biomaRt)
library(org.Hs.eg.db)
library(plotly)
library(DESeq2)
library(msigdbr)
library(ape)
library(GSVA)
library(sva)
library(clusterProfiler)
library(pheatmap)
library(EnhancedVolcano)
library(ggbeeswarm)
library(tidyverse)
library(corrr)
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(esquisse)
library(corto)
library(factoextra)
library(reshape2)
library(ComplexHeatmap)
library(gghighlight)
library(janitor)
library(paletteer)
library(edgeR)
library(limma)
library(mogene10sttranscriptcluster.db)
```
# Aesthetics
```{r}
library("scales")
pal_npg()
pal_npg("nrc")(10)
show_col(pal_npg("nrc")(10))
```


#Data curation
## BioProject
```{r}
animals_vaccines_bioproject_result <- read_csv("animals_vaccines_bioproject_result.csv", 
    col_names = FALSE)
datacuration_clean = animals_vaccines_bioproject_result %>%
 drop_na() %>% 
  mutate(study = if_else(X1 =="Title", X2, NA)) %>% 
  fill(study, .direction = c("down")) %>% 
  pivot_wider(names_from = X1, values_from = X2) %>% 
  select(-study) %>% 
  clean_names() %>% 
  mutate(title = str_remove(title, "^\\d+\\.\\s"),
         organism = str_remove(organism, "Organism: "),
         organism = str_remove(organism, "Taxonomy ID "),
         bio_project_accession = str_remove(bio_project_accession, "BioProject Accession: "),
         id = str_remove(id, "ID: "))

datacuration_clean

datacuration_clean %>% 
  write_csv(here("tables", "datacuration_step2.csv"))
```


```{r fig.height=5, fig.width=7}
datacuration_clean %>%
  group_by(organism) %>% 
  summarize(n_organism = n()) %>% 
  # filter(!str_detect(organism,"Homo sapiens")) %>% 
  arrange(-n_organism) %>% 
  ggplot() +
  aes(x = n_organism,
      y = organism %>% fct_reorder(n_organism)) +
  geom_col(fill = "black") + 
  theme_light()+
  geom_text(aes(label = n_organism),
            hjust = -0.2,
            size =3) +
  scale_x_continuous(expand = c(0.01,0.01),
                     limits = c(0, 900)) +
  theme(
    legend.position = c(0.95,0.5),
    plot.caption = element_text(hjust = 0, size = 5),
    axis.text.x = element_text(size = 10,
                               color = "black",
                               angle = 0),
    axis.text.y = element_text(size = 10,
                               color = "black",
                               angle = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 15),
    axis.title.y = element_text(color = "black"),
    panel.border = element_blank(),
    panel.grid.major.y = element_line(size = 0, 
                                    linetype = 2,
                                    color = "gray75"),
    panel.grid.major.x = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0 ,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 0.5, color = "black"),
    axis.ticks.y = element_line(size = 0.5, color = "black"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(1, 1, 0, 0, "cm")
  ) +
    labs(x = "Frequency",
       y = "",
       title = "BioProject data curation",
       subtitle = "Immunization and vaccination datasets",
       caption = "Search terms: (('vaccine'[All Fields] OR 'immunization'[All Fields]\nOR 'vaccinated'[All Fields] OR 'vaccination'[All Fields] OR 'vaccines'[All Fields]) AND 'transcriptome gene expression'[Filter]\n AND 'material transcriptome'[Filter] AND 'capture_whole'[Filter] AND 'org mammals'[Filter] AND 'Mus musculus'[Organism]") 
  
```
```{r}
mice_vaccination_mining = datacuration_clean %>% 
  filter(str_detect(organism,"Mus musculus")) %>% 
  filter(!str_detect(str_to_lower(title), "cancer|tumor|autoimmune|drug")) %>% 
  filter(str_detect(str_to_lower(title), "vaccin|immuniz"))


mice_vaccination_mining
```


BioProject metadata
```{r}
library(rentrez)

#Single
bioprojectid = mice_vaccination_mining$id[[1]]
bioproject_meta = entrez_summary(db = "bioproject", id = bioprojectid) %>% 
  as.tibble() %>% 
  select(project_name,
         project_description,
         organism_name,
         project_id,
         project_data_type,
         project_target_material,
         project_target_capture,
         organism_name)


#Loop
bioproject_ids <- mice_vaccination_mining$id
bioproject_metadata_df <- map_dfr(bioproject_ids, ~
  entrez_summary(db = "bioproject", id = .x) %>%
    as.list() %>%
    as_tibble() %>%
    select(
      project_name,
      project_description,
      organism_name,
      project_data_type,
      project_target_material,
      project_target_capture
    ) %>%
    mutate(bioproject_id = .x)
)


bioproject_metadata_df = bioproject_metadata_df %>% 
  filter(!str_detect(project_description, "cancer|neoantigen|tumor|autoimmune|drug|opioid"))

# Visualizar
bioproject_metadata_df %>% 
  write_excel_csv2(here("tables", "mice_vaccination_mining2.csv"))

mice_vaccination_mining2 <- read_delim("tables/mice_vaccination_mining2.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)


bioproject_metadata_df

```


GEO
```{r}
library(glue)
#Single ------
# query NCBI GEO 
geo_ids <- "GSE199750"
terms <- entrez_db_searchable(db = "gds")  # for reference: searchable fields
# first, get additional information about this series, including its db ID.
series <- entrez_search("gds",
                        term = glue("{geo_ids}[ACCN] AND gse[ETYP]"))
# use the ID to fetch details, including the sample table
series_summary = entrez_summary(db = "gds", id = series$ids)
series_summary_df = data_frame(bioproject = series_summary$bioproject,
                               uid = series_summary$uid,
                               gse= series_summary$gse,
                               gds= series_summary$gds,
                               title= series_summary$title,
                               summary= series_summary$summary,
                               gpl= series_summary$gpl,
                               n_samples= series_summary$n_samples,
                               platformtitle= series_summary$platformtitle)

#Multiple -------

# geo_ids = read_delim(here("tables", "gds_result.txt"), delim = "\n", col_names = F) %>% 
#   mutate(X1 = str_c("GSE", X1, sep = "") %>% 
#            str_remove(., "200"))
#   pull()

text = readLines(here("tables", "gds_result (1).txt"))
text_collapsed <- paste(text, collapse = "\n")
gse_ids <- str_extract_all(conteudo, "GSE\\d+\\b") %>% unlist() %>% unique()


geo_metadata_df <- map_dfr(gse_ids, ~ {
  result <- entrez_search("gds", term = glue("{.x}[ACCN] AND gse[ETYP]"))

  if (length(result$ids) == 0) {
    return(tibble(gse = .x, error = "Not found"))
  }

  summary <- entrez_summary(db = "gds", id = result$ids)

  tibble(
    gse = .x,
    uid = summary$uid,
    gds = summary$gds,
    bioproject = summary$bioproject,
    title = summary$title,
    summary_text = summary$summary,
    gpl = summary$gpl,
    n_samples = summary$n_samples,
    platformtitle = summary$platformtitle
  ) %>% 
    mutate(bioproject_id = str_remove(bioproject, "PRJNA"))
})
print(geo_metadata_df)

  
  
```


Merge by Bioproject ID
```{r}
mice_vaccination_mining_ManualAnnotation <- read_csv(here("tables", "mice_vaccination_mining_ManualAnnotation.csv")) %>% 
  mutate(bioproject_id = as.character(bioproject_id),
         project_id = as.character(project_id)) %>% 
  drop_na(project_id)


mice_vaccination_annotated = geo_metadata_df %>% 
  right_join(mice_vaccination_mining_ManualAnnotation,
             by = "bioproject_id") %>% 
  select(title, summary_text,
         project_description, n_samples,
         notes, keywords, timepoints:project_target_material,
         gse, bioproject_id
         ) %>% 
  arrange(-n_samples)

mice_vaccination_annotated %>% 
  write_csv(here("tables", "mice_vaccination_annotated.csv"))

mice_vaccination_annotated
```




# MouseImmuneGO

## Gene sets construction


### Blood transcription modules <https://www.nature.com/articles/ni.2789>

```{r}
#Converter ítens de "Module member genes" em linhas em uma tabela longa
btm_annotation_table_long_immune <- BTM_modules %>%
  separate_rows(`Module member genes`, sep = ",") %>% 
  filter(`Module category` == "immune")

#Salvar
write.csv(btm_annotation_table_long_immune, file = "btm_annotation_table_long_immune.csv")

#Display
head(btm_annotation_table_long_immune)

#Calculate Set size
btm_annotation_nested <- btm_annotation_table_long_immune %>%
  group_by(`Module title`) %>% #Somar e agrupar
  mutate(SetSize = n())

#Convert gene symbol column to aggregated rows
btm_annotation_nested <- btm_annotation_nested %>%
  group_by(`Module title`) %>%
  summarize(Genes = paste(`Module member genes`, collapse = ", "), .groups = "drop") %>% 
  left_join(btm_annotation_nested, by = "Module title") %>% 
  filter(Genes != "NA") %>% 
  relocate(Genes, .after="SetSize")

#Salvar
write.csv(btm_annotation_nested, file = "btm_annotation_genes.csv")

#Convert to unique values table
btm_annotation_unique = btm_annotation_nested %>% 
  dplyr::select(-`Module member genes`)

btm_annotation_unique = btm_annotation_unique %>% 
  distinct() %>% 
  relocate(Genes, .after="SetSize") %>% 
  filter(Genes != "NA")

#salvar
write.csv(btm_annotation_unique, file = "btm_annotation_unique.csv")

#Display
head(btm_annotation_unique)

btm_annotation_table_long_immune
```

### CellMarker

Download database from
"<http://117.50.127.228/CellMarker/CellMarker_download_files/file/Cell_marker_Human.xlsx>"

```{r}
##################### Baixar tabela de marcadores celulares

url <- "http://117.50.127.228/CellMarker/CellMarker_download_files/file/Cell_marker_Human.xlsx" 
f <- tempfile(fileext = ".xlsx")
download.file(url, f)
cell_marker_data <- readxl::read_excel(f, 1)

##################### Sub-agrupar células por sistema imunológico

#Filtrar marcadores e células por tecido (Blood)
cells <- cell_marker_data %>%
    filter(tissue_class == "Blood") %>%
    dplyr::select(cell_name, marker)

##################### Filtrar células
# Acessar todos os tipos de células no dataset de Blood
cellmarker_celltypes <- unique(cells$cell_name)

# Filtrar apenas os tipos que contêm "B cell" e plasmacell
Bcells_no = cellmarker_celltypes[grep("B cell", cellmarker_celltypes)] #44 outputs
Plasmacells_no = cellmarker_celltypes[grep("plasma", cellmarker_celltypes)] #5 outputs

# Filtrar apenas os tipos que contêm "T cell" e seus subtipos "CD4" e "CD8"
Tcells_no = cellmarker_celltypes[grep("T cell", cellmarker_celltypes)] #93 outputs
TCD4_no = cellmarker_celltypes[grep("CD4", cellmarker_celltypes)] #35 outputs
TCD8_no = cellmarker_celltypes[grep("CD8", cellmarker_celltypes)] #34 outputs
Treg_no = cellmarker_celltypes[grep("Treg", cellmarker_celltypes)] #7 outputs
Thelper_no = cellmarker_celltypes[grep("helper", cellmarker_celltypes)] #17 outputs

# Filtrar outros tipos celulares da resposta imune
Neutrophils_no = cellmarker_celltypes[grep("neutrophil", cellmarker_celltypes)] #7 outputs
Monocyte_no = cellmarker_celltypes[grep("monocyte", cellmarker_celltypes)] #14 outputs
DCells_no = cellmarker_celltypes[grep("dendritic", cellmarker_celltypes)] #25 outputs
Macrophages_no = cellmarker_celltypes[grep("macrophage", cellmarker_celltypes)] #7 outputs
NKcells_no = cellmarker_celltypes[grep("killer", cellmarker_celltypes)] # 11 outputs
Neutrophils_no = cellmarker_celltypes[grep("mastcell", cellmarker_celltypes)] #7 outputs

# Criar variáveis para cada célula para enriquecimento

# Filtrar apenas as células da resposta imune adaptativa
Bcells <- filter(cells, str_detect(cell_name, regex("B cell|plasma cell|plasmablast", ignore_case = TRUE)))
TCD4 <- filter(cells, str_detect(cell_name, regex("CD4", ignore_case = TRUE)))
TCD8 <- filter(cells, str_detect(cell_name, regex("CD8", ignore_case = TRUE)))
Treg <- filter(cells, str_detect(cell_name, regex("Treg", ignore_case = TRUE)))
Thelper <- filter(cells, str_detect(cell_name, regex("helper", ignore_case = TRUE)))


# Filtrar células da resposta imune inata
Neutrophils <- filter(cells, str_detect(cell_name, regex("neutrophil", ignore_case = TRUE)))
Monocyte <- filter(cells, str_detect(cell_name, regex("monocyte", ignore_case = TRUE)))
DCells <- filter(cells, str_detect(cell_name, regex("dendritic", ignore_case = TRUE)))
Macrophages <- filter(cells, str_detect(cell_name, regex("macrophage", ignore_case = TRUE)))
Mastcells <- filter(cells, str_detect(cell_name, regex("mast cell", ignore_case = TRUE)))
NKcells <- filter(cells, str_detect(cell_name, regex("killer", ignore_case = TRUE)))
Platelets <- filter(cells, str_detect(cell_name, regex("Platelet", ignore_case = TRUE)))
Eosinophil <- filter(cells, str_detect(cell_name, regex("Eosinophil", ignore_case = TRUE)))
Basophil <- filter(cells, str_detect(cell_name, regex("Basophil", ignore_case = TRUE)))
Granulocyte <- filter(cells, str_detect(cell_name, regex("Granulocyte", ignore_case = TRUE)))

#Agrupar todos os resultados em Immune cells
Immune_adap = rbind(Bcells, TCD4, TCD8, Treg, Thelper)
Immune_adap$Type = "Adaptive"

Immune_innate = rbind(Neutrophils, Monocyte, DCells, Macrophages, Mastcells, NKcells, Platelets, Eosinophil, Basophil, Granulocyte)
Immune_innate$Type = "Innate"

#Unir tabelas
Immune_cells = rbind(Immune_adap, Immune_innate) 

#Salvar
write.csv(Immune_cells, file = "CellMarker_ImmuneCells.csv")

esquisser(Immune_cells_markersbycells)
```

Top 20 cells by #markers

```{r}

Immune_cells_markersbycells = Immune_cells %>% 
  group_by(cell_name, Type) %>% 
  summarise(n_markers = n()) %>% 
  arrange(desc(n_markers))
Immune_cells_markersbycells_top = Immune_cells_markersbycells[1:20, ]

ggImmune_cells_cellsbyimmune = ggplot(Immune_cells_markersbycells_top) +
  aes(x = reorder(cell_name, n_markers), y = n_markers, fill = Type) +
  scale_fill_manual(values = c('Innate' = "#989898",
                               'Adaptive' = "#6DF8DF")) +
  geom_col() +
  labs(
    x = "#Markers",
    y = "Cells",
    title = "Top 20 cells by #markers"
  ) +
  coord_flip() +
  theme_minimal() +
  geom_text(aes(label = n_markers, y = n_markers), 
            position = position_dodge(width = 0.9), 
            vjust = -0.2, 
            hjust = -0.2, 
            size = 3)

ggsave(ggImmune_cells_cellsbyimmune, file= "CellMarker_Top10_cellsbyn_markers.png", width = 10, height = 5)

print(ggImmune_cells_cellsbyimmune)

```

Number of Cells by immune system

```{r}
Immune_cells_cellsbyimmune = Immune_cells %>% 
  group_by(Type, cell_name) %>% 
  summarise(n_cells = n()) %>% 
  group_by(Type) %>% 
  summarise(n_cells = n())

ggImmune_cells_cellsbyimmune = ggplot(Immune_cells_cellsbyimmune) +
  aes(x = reorder(Type, n_cells), y = n_cells, fill = Type) +
  geom_col() +
  scale_fill_manual(values = c('Innate' = "#989898",
                               'Adaptive' = "#6DF8DF")) +
  labs(
    x = "#Cells",
    y = "Immune system",
    title = "#Cells by immune system"
  ) +
  coord_flip() +
  theme_minimal() +
  geom_text(aes(label = n_cells, y = n_cells), 
            position = position_dodge(width = 0.9), 
            vjust = -0.2, 
            hjust = -0.2, 
            size = 3)
ggsave(ggImmune_cells_cellsbyimmune, file= "CellMarker_nCellsbyimmunesystem.png", width = 10, height = 2)

print(ggImmune_cells_cellsbyimmune)
```

### ImmuneGO

```{r}
organism = "org.Mm.eg.db"
# install.packages("GO.db")
BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)
library(clusterProfiler)
library(GO.db)

#HGNC symbol to Entrez
x <- org.Mm.egSYMBOL
mapped_genes <- mappedkeys(x)
symbols_entrez <- as.data.frame(x[mapped_genes])

## Entrez to GO
x <- org.Mm.egGO
mapped_genes <- mappedkeys(x)
entrez_geneontology <- as.data.frame(x[mapped_genes])

## GO BP ancestors
go_bp_ancestor = GOBPANCESTOR %>% 
  as.data.frame()
colnames(go_bp_ancestor)[1] <- "go_id"
colnames(go_bp_ancestor)[2] <- "go_ancestor"

## GO ID to terms
go_bp = AnnotationDbi::select(GO.db, columns=c("GOID","TERM"), keys= "BP", keytype= "ONTOLOGY") %>% 
  clean_names() %>% 
  dplyr::select(go_id = goid, term)

## GO child and ancestor
go_bp_ancestor = go_bp_ancestor %>% 
  inner_join(go_bp %>% rename(go_ancestor = go_id), by = "go_ancestor") %>% 
  rename(term_ancestor = term)

#GO ids to terms
go_bp = AnnotationDbi::select(GO.db, columns=c("GOID","TERM"), keys="BP", keytype="ONTOLOGY") %>% 
  clean_names() %>% 
  rename(go_id = goid) %>% 
  dplyr::select(-ontology) %>% 
  right_join(entrez_geneontology, by = "go_id") %>% 
  right_join(symbols_entrez, by = "gene_id") %>% 
  clean_names() %>% 
  left_join(go_bp_ancestor, by = "go_id") %>% 
  distinct()

#GO immune process

mm_immunego_go.bp = go_bp %>% 
  filter(term_ancestor == "immune system process") %>% 
  rename(go_dataset = term_ancestor,
         process = term,
         genes = symbol) %>% 
  mutate(process = toupper(process),
         immune_set = "Mouse_ImmuneGO")


```

Annotation

There are 943 processes retrieved, but only 18 are directly annotated using Human ImmuneGO. Most general immunological processes are in, like innate immune response,immune response, adaptive immune response, and others.
```{r}
#Merge annotations with Human ImmuneGO

ImmuneGO_BTM_grouped_genesets_specific = read_csv(here("VaxGO", "ImmuneGO_BTM_grouped_genesets_specific.csv"))

ImmuneGO_Annotated_GO_specific_original <- read_csv(here("VaxGO", "ImmuneGO_Annotated_GO_specific_originalGOs.csv"))

## There are 943 processes
mm_immunego_annotation = mm_immunego_go.bp %>% 
  select(process, genes, immune_set) %>% 
  group_by(process) %>% 
  summarize(set_size = n(),
            genes,
            immune_set) %>% 
  summarize(genes = paste0(genes, collapse = ","),
            set_size,
            genes,
            immune_set) %>% 
  distinct() %>% 
  ungroup()  %>% 
  left_join(ImmuneGO_Annotated_GO_specific_original %>% 
              mutate(process_human = str_to_upper(process)) %>% 
              select(process_go_human = process_go, process_human = process, go_term_human = go_term, immune_system:cells),
                by = join_by("process" == "process_go_human"))
#Save
mm_immunego_annotation %>% 
  saveRDS(here("VaxGO", "mm_immunego_annotation.rds"))
mm_immunego_annotation %>% 
  write_excel_csv(here("VaxGO", "mm_immunego_annotation.csv"))
```



### Stats
There are only 18 immune processes matched by name, and 925 unmatched.
```{r}
mm_immunego_annotation = read_csv(here("VaxGO", "mm_immunego_annotation.csv"))

#How many are not in?
## 925
mm_immunego_annotation %>% 
  filter(is.na(immune_system)) 

## 18 are in
mm_immunego_annotation %>% 
  filter(!is.na(immune_system)) %>% 
  arrange(-set_size)
mm_immunego_annotation %>% 
  filter(!is.na(immune_system)) %>% 
  tabyl(immune_system) %>% 
  janitor::adorn_pct_formatting()



```



Annotate by merging genes with the same symbol (using upper case) by specific processes
There are 2,224 genes not annotated in ImmuneGO human

```{r}
mm_immunego_annotation_genes = mm_immunego_annotation %>% 
  separate_rows(genes, sep = ",") %>% 
  mutate(genes_upper = str_to_upper(genes))

ImmuneGO_BTM_grouped_genesets_specific_genes = ImmuneGO_BTM_grouped_genesets_specific %>% 
              mutate(process = str_to_upper(process)) %>% 
              filter(immune_set == "ImmuneGO by Modules") %>% 
              select(process, genes) %>% 
  separate_rows(genes, sep = ",")

BTM_grouped_genesets_specific_genes = ImmuneGO_BTM_grouped_genesets_specific %>% 
              mutate(process = str_to_upper(process)) %>% 
              filter(immune_set == "BTM by Modules") %>% 
              select(process, genes) %>% 
  separate_rows(genes, sep = ",")

#ImmuneGO human
mm_immunego_annotation_genes_human = mm_immunego_annotation_genes %>% 
  select(genes = genes_upper) %>% 
  distinct() %>% 
  left_join(ImmuneGO_BTM_grouped_genesets_specific_genes,
            by = "genes")

#There are 2,224 genes not annotated in ImmuneGO human
mm_immunego_annotation_genes_human %>% 
  filter(is.na(process))

#BTM modules
mm_btm_annotation_genes_human = mm_immunego_annotation_genes %>% 
  select(genes = genes_upper) %>% 
  distinct() %>% 
  left_join(BTM_grouped_genesets_specific_genes,
            by = "genes")


#There are 2,450 genes not annotated in BTM human
mm_btm_annotation_genes_human %>% 
  filter(is.na(process))
```

By general processes, there are 1,940 genes that were not annotated in both BTM and ImmuneGO
```{r}
ImmuneGO_BTM_grouped_genesets <- read_csv("VaxGO/ImmuneGO_BTM_grouped_genesets.csv")
ImmuneGO_BTM_grouped_genes = ImmuneGO_BTM_grouped_genesets %>% 
  separate_rows(genes) %>% 
  select(process, genes) %>% distinct()

#ImmuneGO and BTM modules
mm_ImmuneGO_BTM_annotation_genes_human = mm_immunego_annotation_genes %>% 
  select(genes = genes_upper) %>% 
  distinct() %>% 
  left_join(ImmuneGO_BTM_grouped_genes,
            by = "genes")

#There are 1,940 genes not annotated in ImmuneGO+BTM
mm_ImmuneGO_BTM_annotation_genes_human %>% 
  filter(is.na(process))
```

Which are the most gapping processes by genes?
```{r}
#ImmuneGO and BTM modules
mm_ImmuneGO_BTM_annotation_genes_human = mm_immunego_annotation_genes %>% 
  select(genes = genes_upper, 
         process) %>% 
  distinct() %>% 
  left_join(ImmuneGO_BTM_grouped_genes,
            by = "genes")

#There are 1,940 genes not annotated in ImmuneGO+BTM
mm_ImmuneGO_BTM_annotation_genes_human %>% 
  filter(is.na(process.y)) %>% 
  tabyl(process.x) %>% 
  as_data_frame() %>% 
  arrange(-percent)


```

### Manual annotation

```{r}
mm_immunego_annotated = mm_immunego_go.bp %>% 
  group_by(process) %>%
  summarize(genes = paste0(genes, collapse = ",")) %>%
  left_join(ImmuneGO_BTM_grouped_genesets_specific %>% 
              dplyr::select(-genes) %>% 
              distinct() %>% 
              mutate(process = str_to_upper(process)), by = "process")

```


#### Approach 1

```{r}
ImmuneGO_Mouse_Manual <- read_excel(here("VaxGO", "ImmuneGO_Mouse.xlsx"))
```

#### Approach 2
By percentage: The gene sets were merged exclusively by genes that were in both gene sets. The percentage was calculated by the frequency of each process in human ImmuneGO divided by the mouse set_size.
```{r}
mm_immunego_annotation_genes_match = mm_immunego_annotation_genes %>% 
  select(process_mice = process, genes_mice = genes_upper, set_size) %>% 
  distinct() %>% 
  inner_join(ImmuneGO_BTM_grouped_genesets %>%
            filter(immune_set %in% c("ImmuneGO", "ImmuneGO by Modules")) %>% 
              separate_rows(genes) %>% 
              select(process, genes, immune_system:cells,
                     immune_set,
                     set_size_human = set_size) %>% 
              rename(process_human = process, 
                     genes_human = genes) %>% 
              mutate(immune_system = if_else(is.na(immune_system), "Manual annotation", immune_system),
                     immune_subsystem = if_else(is.na(immune_subsystem), "Manual annotation", immune_subsystem),
                     cells = if_else(is.na(cells), "Manual annotation", cells)),
            by = join_by("genes_mice" == "genes_human")) %>% 
  distinct() %>% 
  group_by(process_mice, process_human) %>% 
  arrange(process_mice, process_human) %>% 
  summarize(set_size_mice = set_size,
            set_size_human, 
            match_human = n(),
            match_perc = match_human/set_size*100,
            immune_system,
            immune_subsystem,
            cells,
            immune_set) %>% 
  distinct() %>% 
  ungroup() %>% 
  group_by(process_mice) %>% 
  arrange(-match_human, -match_perc) %>% 
  ungroup()
mm_immunego_annotation_genes_match
```


Specific processes
```{r}
mm_immunego_annotation_genes_match = mm_immunego_annotation_genes %>% 
  select(process_mice = process, set_size) %>% 
  distinct() %>% 
  inner_join(ImmuneGO_Annotated_GO_specific_original %>%
              select(process, process_go_human = process_go, genes, immune_system:cells,
                     immune_set,
                     set_size_human = set_size) %>% 
              rename(process_human = process, 
                     genes_human = genes),
            by = join_by("process_mice" == "process_go_human")) 
```



General processes
```{r fig.height=5, fig.width=8}
#General processes
mm_immunego_annotation_genes_match_general = mm_immunego_annotation_genes_match %>% 
  group_by(process_mice) %>% 
  filter(immune_set == "ImmuneGO") %>% 
  slice_max(match_perc) %>% 
  slice_max(set_size_human) %>% 
  mutate(immune_system = case_when(process_human == "CELLULAR ADAPTIVE IMMUNE SYSTEM" ~ "Adaptive",				
                                    process_human == "INNATE IMMUNE SYSTEM" ~ "Innate",				
                                    process_human == "COMPLEMENT" ~ "Complement",				
                                    process_human == "LEUKOCYTES" ~ "General",				
                                    process_human == "ADAPTIVE IMMUNE SYSTEM" ~ "Adaptive",				
                                    process_human == "T CELL" ~ "Adaptive",				
                                    process_human == "B CELL" ~ "Adaptive",				
                                    process_human == "HUMORAL ADAPTIVE IMMUNE SYSTEM" ~ "Adaptive",				
                                    process_human == "IMMUNOGLOBULIN MEDIATED RESPONSE" ~ "Adaptive",				
                                    process_human == "ANTIMICROBIAL" ~ "Undefined",
                                    process_human == "ARPP" ~ "Innate",				
                                    process_human == "NEUTROPHIL" ~ "Innate",				
                                    process_human == "ANTIVIRAL & IFN" ~ "Innate",				
                                    process_human == "DENDRITIC CELL" ~ "Innate",				
                                    process_human == "EOSINOPHIL" ~ "Innate",				
                                    process_human == "MAST CELL" ~ "Innate",				
                                    process_human == "MONOCYTE" ~ "Innate",				
                                    process_human == "INFLAMMATION" ~ "Innate",				
                                    process_human == "NK CELL" ~ "Innate",				
                                    process_human == "MACROPHAGE" ~ "Innate",
                                    T ~ immune_system))

mm_immunego_annotation_genes_match_general_plot = mm_immunego_annotation_genes_match_general %>% 
ggplot(.) +
  aes(
    x = match_human,
    y = match_perc,
    size = set_size_mice,
    label = paste("human:", process_human, ", mice:", process_mice)
  ) +
  geom_point(pch = 21, 
             color = "black", 
             fill = NA) +
  geom_point(pch = 21,
             aes(fill = immune_system),
             alpha = 0.3) +
  # geom_text_repel(aes(label = ifelse(match_human > 40,
  #                    paste("mouse:", process_mice, "\nto human:", process_human),
  #                    NA)),
  #                  na.rm = TRUE,
  #                  size = 2,
  #                  hjust = 0,
  #                  fill = NA) +
  scale_x_continuous(expand = c(0.01,0.01),
                     limits = c(0, 200)) +
  scale_y_continuous(expand = c(0.01,0.01),
                     limits = c(0, 100)) +
  theme_minimal() +
  scale_fill_npg() +
  # scale_fill_paletteer_d("ghibli::MarnieLight1") +
  theme(
    legend.position = "right",
    plot.caption = element_text(hjust = 0, size = 5),
    axis.text.x = element_text(size = 10,
                               color = "black",
                               angle = 0),
    axis.text.y = element_text(size = 10,
                               color = "black",
                               angle = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 15),
    axis.title.y = element_text(color = "black"),
    panel.border = element_blank(),
    panel.grid.major.y = element_line(size = 0, 
                                    linetype = 2,
                                    color = "gray75"),
    panel.grid.major.x = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0.5 ,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 0.5, color = "black"),
    axis.ticks.y = element_line(size = 0.5, color = "black"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(1, 1, 0, 0, "cm")
  ) +
  labs(x = "Match to human processes",
       y = "%Match",
       size = "Mouse set size",
       fill = "ImmuneGO Human",
       title = "Mouse GO match to Human ImmuneGO annotation",
       subtitle = "Manual grouped annotation")

mm_immunego_annotation_genes_match_general_plot
  
```


Specific processes
```{r fig.height=5, fig.width=8}
#General processes
mm_immunego_annotation_genes_match_specific = mm_immunego_annotation_genes_match %>% 
  group_by(process_mice) %>% 
  filter(immune_set == "ImmuneGO by Modules") %>% 
  slice_max(match_perc) %>% 
  slice_max(set_size_human)


#Save
mm_immunego_annotation_genes_match_specific %>% 
  write_csv(here("VaxGO", "mm_immunego_annotation_genes_match_specific_semimanual.csv"))
```


```{r fig.height=5, fig.width=8}
mm_immunego_annotation_genes_match_specific_plot = mm_immunego_annotation_genes_match_specific %>% 
ggplot(.) +
  aes(
    x = match_human,
    y = match_perc,
    size = set_size_mice,
    label = paste("human:", process_human, ", mice:", process_mice)
  ) +
  geom_point(pch = 21, 
             color = "black", 
             fill = NA) +
  geom_point(pch = 21,
             aes(fill = immune_system),
             alpha = 0.3) +
  # geom_text_repel(aes(label = ifelse(match_human > 40,
  #                    paste("mouse:", process_mice, "\nto human:", process_human),
  #                    NA)),
  #                  na.rm = TRUE,
  #                  size = 2,
  #                  hjust = 0,
  #                  fill = NA) +
  scale_x_continuous(expand = c(0.01,0.01),
                     limits = c(0, 200)) +
  scale_y_continuous(expand = c(0.01,0.01),
                     limits = c(0, 100)) +
  theme_minimal() +
  scale_fill_npg() +
  # scale_fill_paletteer_d("ghibli::MarnieLight1") +
  theme(
    legend.position = "right",
    plot.caption = element_text(hjust = 0, size = 5),
    axis.text.x = element_text(size = 10,
                               color = "black",
                               angle = 0),
    axis.text.y = element_text(size = 10,
                               color = "black",
                               angle = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 15),
    axis.title.y = element_text(color = "black"),
    panel.border = element_blank(),
    panel.grid.major.y = element_line(size = 0, 
                                    linetype = 2,
                                    color = "gray75"),
    panel.grid.major.x = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0.5 ,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 0.5, color = "black"),
    axis.ticks.y = element_line(size = 0.5, color = "black"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(1, 1, 0, 0, "cm")
  ) +
  labs(x = "Match to human processes",
       y = "%Match",
       size = "Mouse set size",
       fill = "ImmuneGO Human",
       title = "Mouse GO match to Human ImmuneGO annotation",
       subtitle = "Specific processes")

mm_immunego_annotation_genes_match_specific_plot

mm_immunego_annotation_genes_match_general_plot/mm_immunego_annotation_genes_match_specific_plot +
  patchwork::plot_layout(guides = "collect",
                         axis_titles = "collect")
```


Group immune system genes by process and cell

```{r}
ImmuneGO_Annotated_unique = ImmuneGO_Annotated_Genes %>% 
  filter(go_term != "Manual")

#Agrupar genes de todos os processos do sistema imune ----

ImmuneGO_Adaptive <- ImmuneGO_Annotated_unique %>%
  filter(immune_system == "Adaptive") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "ADAPTIVE IMMUNE SYSTEM",
  `Gene set, short` = "Adaptive immune system",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "General",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate <- ImmuneGO_Annotated_unique %>%
  filter(immune_system == "Innate") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "INNATE IMMUNE SYSTEM",
  `Gene set, short` = "Innate immune system",
  `Immune system` = "Innate",
  `Immune sub-system` = "General",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Complement <- ImmuneGO_Annotated_unique %>%
  filter(immune_system == "Complement") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "COMPLEMENT",
  `Gene set, short` = "Complement immune system",
  `Immune system` = "Complement",
  `Immune sub-system` = "General",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Inflammation <- ImmuneGO_Annotated_unique %>%
  filter(immune_sub_system == "Inflammation") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "INFLAMMATION",
  `Gene set, short` = "Inflammation",
  `Immune system` = "Innate",
  `Immune sub-system` = "Inflammation",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Interferon <- ImmuneGO_Annotated_unique %>%
  filter(immune_sub_system %in% c("Antiviral", "Interferon")) %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "ANTIVIRAL & IFN",
  `Gene set, short` = "Antiviral and Interferon",
  `Immune system` = "Innate",
  `Immune sub-system` = "Antiviral and Interferon",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_ARPP = ImmuneGO_Annotated_unique %>%
  filter(immune_sub_system %in% c("PRR", "APC")) %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "ARPP",
  `Gene set, short` = "Antigen receptors, processing and presentation",
  `Immune system` = "Innate",
  `Immune sub-system` = "APC",
  `Immune tissue` = "APC",
  `GO.TERM` = "Manual") %>% 
  clean_names()


ImmuneGO_Innate_DC <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Dendritic cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "DENDRITIC CELL",
  `Gene set, short` = "Dendritic cell",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "APC",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_Macrophage <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Macrophage") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "MACROPHAGE",
  `Gene set, short` = "Macrophage",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "APC",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_Mastcell <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Mast cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "MAST CELL",
  `Gene set, short` = "Mast cell",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "Mast cell",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_Neutrophil <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Neutrophil") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "NEUTROPHIL",
  `Gene set, short` = "Neutrophil",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "Neutrophil",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_Eosinophil <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Eosinophil") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "EOSINOPHIL",
  `Gene set, short` = "Eosinophil",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "Eosinophil",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_NK <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "NK cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "NK CELL",
  `Gene set, short` = "Natural Killer Cell",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "NK cell",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_Monocyte <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Monocyte") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "MONOCYTE",
  `Gene set, short` = "Monocyte",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "APC",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Adap_Cellular <- ImmuneGO_Annotated_unique %>% 
  filter(immune_sub_system == "Cellular") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "CELLULAR ADAPTIVE IMMUNE SYSTEM",
  `Gene set, short` = "Cellular adaptive immune system",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Cellular",
  `Immune tissue` = "Cellular",
  `GO.TERM` = "Manual") %>% 
  clean_names()


ImmuneGO_Adap_Tcell <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "T cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "T CELL",
  `Gene set, short` = "T cell",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Cellular",
  `Immune tissue` = "Cellular",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Adap_Thelper <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "T helper") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "T HELPER CELLS",
  `Gene set, short` = "T helper cell",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Cellular",
  `Immune tissue` = "Cellular",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Adap_Humoral <- ImmuneGO_Annotated_unique %>%
  filter(immune_sub_system == "Humoral") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "HUMORAL ADAPTIVE IMMUNE SYSTEM",
  `Gene set, short` = "Humoral adaptive immune system",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Humoral",
  `Immune tissue` = "Humoral",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Adap_Bcell <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "B cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "B CELL",
  `Gene set, short` = "B cell",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Humoral",
  `Immune tissue` = "B cell",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Adap_Ig <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "B cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "IMMUNOGLOBULIN MEDIATED RESPONSE",
  `Gene set, short` = "Ig-mediated response",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Humoral",
  `Immune tissue` = "B cell",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_other <- ImmuneGO_Annotated_unique %>%
  filter(immune_system == "General" |
         immune_tissue == "Leukocyte") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "GENERAL",
  `Gene set, short` = "General",
  `Immune system` = "General",
  `Immune sub-system` = "General",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()


ImmuneGO_antimicrobial <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Antimicrobial") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "ANTIMICROBIAL",
  `Gene set, short` = "Antimicrobial immune response",
  `Immune system` = "General",
  `Immune sub-system` = "General",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()


# Unite sets -----

ImmuneGO_Annotated_GO = bind_rows(ImmuneGO_Adaptive, 
                            ImmuneGO_Innate, 
                            ImmuneGO_Complement, 
                            ImmuneGO_Inflammation,
                            ImmuneGO_Interferon,
                            ImmuneGO_ARPP,
                            ImmuneGO_Innate_Macrophage,
                            ImmuneGO_Innate_Mastcell,
                            ImmuneGO_Innate_Neutrophil,
                            ImmuneGO_Innate_Eosinophil,
                            ImmuneGO_Innate_NK,
                            ImmuneGO_Innate_Monocyte,
                            ImmuneGO_Adap_Cellular,
                            ImmuneGO_Adap_Tcell,
                            ImmuneGO_Adap_Thelper,
                            ImmuneGO_Adap_Humoral,
                            ImmuneGO_Adap_Bcell,
                            ImmuneGO_Adap_Ig,
                            ImmuneGO_other,
                            ImmuneGO_antimicrobial,
                            ImmuneGO_Innate_DC,
                            ImmuneGO_Annotated_unique) %>% 
  dplyr::select(process:go_term, set_size:genes) %>% 
  distinct()

ImmuneGO_Annotated_Genes = ImmuneGO_Annotated_GO %>% 
   separate_rows(genes, sep = ",")

# Separate TCR and BCR repertoire genes
TCR_BCR_repertoire = ImmuneGO_Annotated_Genes %>% 
  filter(grepl("^TR", genes) & process == "ADAPTIVE IMMUNE SYSTEM" |
           grepl("^IG", genes) & process == "ADAPTIVE IMMUNE SYSTEM" |
           grepl("^TR", genes) & process == "CELLULAR ADAPTIVE IMMUNE SYSTEM" |
         grepl("^IG", genes) & process == "HUMORAL ADAPTIVE IMMUNE SYSTEM") %>% 
  mutate(process = if_else(process == "CELLULAR ADAPTIVE IMMUNE SYSTEM", "TCR REPERTOIRE", 
                           "BCR REPERTOIRE"),
         gene_set_short = if_else(process == "TCR REPERTOIRE", 
                                  "TCR REPERTOIRE", 
                           "BCR REPERTOIRE"),
         process = if_else(grepl("^TR", genes), "TCR REPERTOIRE", "BCR REPERTOIRE"),
         immune_sub_system = if_else(process == "TCR REPERTOIRE", "Cellular", "Humoral"),
         immune_tissue = if_else(process == "TCR REPERTOIRE", "Cellular", "Humoral"))

immunego_manual = ImmuneGO_Annotated_Genes %>%
  filter(go_term == "Manual") %>% 
  anti_join(TCR_BCR_repertoire %>% dplyr::select(genes) %>% distinct(), by = "genes") %>% 
  bind_rows()

#Verify if all genes are annotated. If there are no left genes, you are good to go
a = ImmuneGO_Annotated_Genes %>% filter(go_term == "Manual") %>% dplyr::select(genes) %>% distinct()
ImmuneGO_Annotated_Genes %>% dplyr::select(process, genes, immune_system) %>% anti_join(a, by = "genes")


#Salvar
write.csv(ImmuneGO_Annotated_GO, file = paste0("ImmuneGO_Annotated_GO_", today(), ".csv"), row.names = F)
saveRDS(ImmuneGO_Annotated_GO, file = paste0("ImmuneGO_Annotated_GO_", today(), ".rds"))
saveRDS(ImmuneGO_Annotated_Genes, file = paste0("ImmuneGO_Annotated_Genes_", today(), ".rds"))

#Display
view(ImmuneGO_Annotated_GO)
view(ImmuneGO_Annotated_Genes)

```

Top 10 gene sets
```{r}
#Top 10 gene sets
ImmuneGO_Annotated_unique_top10 = ImmuneGO_Annotated_GO %>% 
  clean_names() %>% 
  arrange(desc(set_size)) %>% 
  slice_head(n = 20) %>% 
  mutate(immune_system = if_else(gene_set_short == "Adaptive Immune response", "Adaptive", immune_system))

#Gráfico de barras
ImmuneGO_top10_barplot = ggplot(ImmuneGO_Annotated_unique_top10) +
 aes(x = reorder(gene_set_short, set_size), 
     weight = set_size, 
     fill = immune_system) +  # Use reorder para reordenar as barras +
  geom_bar()+
  scale_fill_manual(
    values = c(Adaptive = "#6DF8DF",
    Complement = "#e9edc9",
    General = "#bde0fe",
    Innate = "#989898",
    Undefined = "#edede9")) +
 labs(y = "Set size", title = "ImmuneGO", subtitle = "Top 10 GO, by set size") +
 coord_flip() +
 theme_minimal()

#Salvar
ggsave(ImmuneGO_top10_barplot, file="ImmuneGO_top10_barplot.png", width = 10, height = 6)

#Display
head(ImmuneGO_top10_barplot)
```



### ImmuneGO + BTM

```{r}
btm_annotation_genes = readRDS(here("VaxGO gene sets", "btm_annotation_genes.rds"))
ImmuneGO_Annotated_Genes = readRDS(here("VaxGO gene sets", "ImmuneGO_Annotated_Genes_2024-03-26.rds"))
ImmuneGO_Annotated_GO = read_csv(here("VaxGO gene sets", "ImmuneGO_Annotated_GO_2024_03_26.csv"))

#Unite genes
immunego_btm_genes = ImmuneGO_Annotated_Genes %>% 
              dplyr::select(genes) %>% 
              distinct() %>% 
              mutate(set = "ImmuneGO") %>% 
  bind_rows(btm_annotation_genes %>% 
              dplyr::select(genes) %>% 
              distinct() %>% 
              mutate(set = "BTM")) %>% 
  arrange(genes)

immunego_btm_genes_joined = immunego_btm_genes %>% 
  bind_rows(immunego_btm_genes %>% mutate(set = "BTM and ImmuneGO")) %>% 
  distinct()


#Unite subsets
btm_annotation_grouped = btm_annotation_genes %>% 
  group_by(complete_name) %>% 
  rename(process = complete_name) %>% 
  mutate(genes = paste0(genes, collapse = ",")) %>% 
  distinct() %>% 
  mutate(genes = paste0(genes, collapse = ","),
            immune_set = "BTM by Modules") %>% 
    dplyr::select(process,
           immune_system:cells,
           set_size = module_size,
           genes,
           immune_set)

ImmuneGO_annotation_modules = ImmuneGO_Annotated_GO %>% 
  filter(go_term != "Manual") %>% 
    dplyr::select(process = gene_set_short, 
           immune_system:immune_tissue,
           set_size, genes) %>% 
    rename(cells = immune_tissue,
           immune_subsystem = immune_sub_system) %>% 
    mutate(immune_set = "ImmuneGO by Modules")

immunego_btm_specific_grouped = btm_annotation_grouped %>% 
  bind_rows(ImmuneGO_annotation_modules)
  
immunego_btm_specific_grouped %>% 
  write_csv(file = here("VaxGO gene sets", "ImmuneGO_BTM_grouped_genesets_specific.csv"))

  
#By immune system
btm_immunesystem = btm_annotation_genes %>% 
  dplyr::select(immune_system, genes) %>% 
  distinct() %>% 
  rename(process = immune_system) %>% 
  group_by(process) %>% 
  summarize(genes = paste0(genes, collapse = ","),
            set_size = n(),
            process = str_to_upper(process),
            # id = process,
            immune_system = process,
            immune_subsystem = process,
            cells = process,
            complete_name = process,
            annotation_level = "Manual",
            immune_set = "BTM by Immune System") %>% 
  distinct()

#By immune subsystem
btm_immunesubsystem = btm_annotation_genes %>% 
  dplyr::select(immune_subsystem, genes) %>% 
  distinct() %>% 
  rename(process = immune_subsystem) %>% 
  group_by(process) %>% 
  summarize(genes = paste0(genes, collapse = ","),
            process = str_to_upper(process),
            set_size = n(),
            # id = process,
            immune_system = process,
            immune_subsystem = process,
            cells = process,
            complete_name = process,
            annotation_level = "Manual",
            immune_set = "BTM by Immune Sub-System") %>% 
  distinct() 
#By cell
btm_cells = btm_annotation_genes %>% 
  dplyr::select(cells, genes) %>%
  distinct() %>% 
  rename(process = cells) %>% 
  group_by(process) %>% 
  summarize(genes = paste0(genes, collapse = ","),
            set_size = n(),
            process = str_to_upper(process),
            # id = process,
            immune_system = process,
            immune_subsystem = process,
            cells = process,
            annotation_level = "Manual",
            complete_name = process,
            immune_set = "BTM by Immune Cell") %>% 
  distinct() 
#Merge
btm_manual = bind_rows(btm_immunesystem,
          btm_immunesubsystem,
          btm_cells) %>% 
  arrange(complete_name) %>% 
  dplyr::select(complete_name, everything()) %>% 
  distinct()

btm_manual_grouped = bind_rows(btm_manual,
                               btm_annotation_grouped) %>% 
  distinct()


#Merge ImmuneGO and BTM main processes
immunego_btm_manual_grouped = btm_manual %>% 
  dplyr::select(process, genes, immune_set, set_size) %>% 
  mutate(immune_set = "BTM") %>% 
  bind_rows(ImmuneGO_Annotated_GO %>% 
              filter(go_term == "Manual") %>% 
              dplyr::select(process, genes, set_size) %>% 
              mutate(immune_set = "ImmuneGO")) %>% 
  arrange(process) %>% 
  mutate(process = case_when(process == "ADAPTIVE" ~ "ADAPTIVE IMMUNE SYSTEM",
                              process == "ALL" ~ "IMMUNE SYSTEM",
                              process == "GENERAL" ~ "LEUKOCYTES",
                              process == "CELLS" ~ "LEUKOCYTES",
                              process == "DCS" ~ "DENDRITIC CELL",
                              process == "APP" ~ "ARPP",
                              process == "CELLULAR" ~ "CELLULAR ADAPTIVE IMMUNE SYSTEM",
                              process == "DCS, MYELOID" ~ "DENDRITIC CELL",
                              process == "GENERIC" ~ "LEUKOCYTES",
                              process == "HUMORAL" ~ "HUMORAL ADAPTIVE IMMUNE SYSTEM",
                              process == "INNATE" ~ "INNATE IMMUNE SYSTEM",
                              process == "INTERFERON" ~ "ANTIVIRAL & IFN",
                              process == "NK" ~ "NK CELL",
                              process == "B CELL, PLASMA CELL" ~ "B CELL",
                              process == "PLASMA CELL" ~ "B CELL",
                              process == "PRR" ~ "ARPP",
                              process == "SIGNALING" ~ "SIGNALING",
                              process == "TH2" ~ "T CELL",
                              process == "TH1" ~ "T CELL",
                              process == "T HELPER CELLS" ~ "T CELL",
                              TRUE ~ process
                              )) %>% 
  distinct() %>% 
  bind_rows(
    immunego_btm_specific_grouped)

immunego_btm_manual_grouped_genes = immunego_btm_manual_grouped %>%
  separate_rows(genes, sep = ",") %>% 
  distinct()

immunego_btm_manual_grouped_genes = immunego_btm_manual_grouped_genes %>% 
  bind_rows(
    immunego_btm_manual_grouped_genes %>% 
      mutate(immune_set = "BTM and ImmuneGO") %>% 
      distinct()
  )
  
  
immunego_btm_manual_grouped %>% 
  write_csv(file = here("VaxGO gene sets", "ImmuneGO_BTM_grouped_genesets.csv"))

immunego_btm_manual_grouped_genes %>% 
  write_csv(file = here("VaxGO gene sets", "ImmuneGO_BTM_grouped_genes.csv"))
  
```
Plot relative frequency of genes by main process
```{r fig.height=5, fig.width=7}

#Get exclusive genes
immunego_only_genes = immunego_btm_manual_grouped_genes %>% 
  filter(immune_set == "ImmuneGO") %>% 
  dplyr::select(-immune_set, -set_size) %>% 
  anti_join(immunego_btm_manual_grouped_genes %>% 
            filter(immune_set == "BTM"),
            by = "genes") %>% 
  mutate(immune_set = "ImmuneGO only") %>% 
  distinct()

btm_only_genes = immunego_btm_manual_grouped_genes %>% 
  filter(immune_set == "BTM") %>% 
  dplyr::select(-immune_set, -set_size) %>% 
  anti_join(immunego_btm_manual_grouped_genes %>% 
            filter(immune_set == "ImmuneGO"),
            by = "genes") %>% 
  mutate(immune_set = "BTM only") %>% 
  distinct()

btm_immune_joined_stats = immunego_only_genes %>% 
  bind_rows(btm_only_genes) %>% 
  distinct() %>% 
  mutate(immune_set = "BTM and ImmuneGO") %>% 
  dplyr::select(process, genes, immune_set) %>% 
  distinct() %>% 
  group_by(process) %>% 
  distinct() %>% 
  mutate(ntotal=n()) %>% 
  ungroup() %>% 
  dplyr::select(process, ntotal) %>% 
  distinct() %>% 
  inner_join(immunego_only_genes %>% 
               bind_rows(btm_only_genes), by = "process") %>% 
  group_by(process, immune_set) %>% 
  mutate(n = n(),
         genes = paste0(genes, collapse = ",")) %>% 
  dplyr::select(-genes) %>% 
  distinct() %>% 
  ungroup() %>% 
  mutate(n_perc = (n/ntotal) * 100)

btm_immune_joined_stats_plot = btm_immune_joined_stats %>% 
  filter(!process %in% c("LYMPHOCYTE", 
                         "IMMUNOGLOBULIN MEDIATED RESPONSE",
                         "MYELOID CELL", 
                         "SIGNALING", 
                         "IMMUNE SYSTEM")) %>%
  mutate(process = str_to_title(process),
         process = fct_relevel(process, 
                               "Leukocytes",
                               "Adaptive Immune System",
                               "Innate Immune System",
                               "Complement"	,
                               "Cellular Adaptive Immune System",
                               "Humoral Adaptive Immune System",
                               )) %>% 
 ggplot() +
  aes(x = n_perc, y = fct_rev(process), fill = immune_set) +
  geom_bar(stat = "summary", fun = "sum", position = "fill") +
  scale_fill_manual(values = c("BTM only" = "#5e60ce",
                               "ImmuneGO only" = "#00A087FF")) +
  theme_minimal() +
  theme(legend.position = "right",
        strip.text.x = element_text(color = "black", size = 12), 
        axis.text.x = element_text(color = "black", size = 10),
        axis.text.y = element_text(vjust = 0.5, size = 10, color = "black"),
        axis.line = element_line(size = 0.5, colour = "black", linetype=1),
        axis.ticks = element_line(size = 0.5, color = "black")) +
  labs(x = "%Coverage",
       y = "",
       fill = "Immune set",
       subtitle = "Merged data",
       title = "ImmuneGO and BTM main processes genes") +
  scale_x_continuous(expand = c(0,0)) +
  geom_text(aes(label = paste0(round(n_perc, 0))), 
            position = position_fill(vjust = 0.5), 
            size = 3, 
            color = "white",
            fontface = "bold") 

btm_immune_joined_stats_plot %>% 
  ggsave(filename = here("Figures", "ImmuneGO_BTM_Joined_Stats_barplot.png"),
         width = 10,
         height = 7)

btm_immune_joined_stats_plot
```



#Benchmarking models




## Download datasets

Download general data (counts data and metadata). GEO supplementary
files have the counts matrices. Download studies individually. I
recommend downloading large files from the GEO website.


```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
###### GSE120661 ----
#Metadata 
filename = "GSE120661"
GSE120661 <- getGEO("GSE120661", GSEMatrix = TRUE) 
eset <- GSE120661[[1]]

GSE120661_metadata = eset %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    subject_id = "title",
    sex = "gender:ch1", 
    treatment = "treatment:ch1", 
    tissue = "tissue:ch1",
    timepoint = "timepoint:ch1",
    animal_replicate = "animal replicate:ch1",
    strain = "strain/background:ch1") %>% 
  mutate(treatment = fct_relevel(treatment, "saline"),
         timepoint = fct_relevel(timepoint, "0"),
         condition = paste(treatment, timepoint, tissue, sep = "_"))
write_csv(GSE120661_metadata, file = here("tables", "GSE120661_metadata.csv"))

#Counts matrix from Expression set
eset <- exprs(eset)

# Quick check. Normalization needed? Yes
eset_12 <- eset[ ,1:12]
exprs(eset_12) <- exprs(eset_12)[ ,1:12]
boxplot(exprs(eset_12), las=2, cex.axis = 0.7, boxwex=0.6, outline=FALSE)

#Normalize data
exprs_matrix.norm <- normalizeBetweenArrays(exprs(eset))

#Check
exprs_matrix.norm_12 <- exprs_matrix.norm[ ,1:12]
boxplot(exprs_matrix.norm_12, las=2, cex.axis = 0.7, boxwex=0.6, outline=FALSE)
boxplot(exprs(eset_12), las=2, cex.axis = 0.7, boxwex=0.6, outline=FALSE)
boxplot(exprs_matrix.norm_12, las=2, cex.axis = 0.7, boxwex=0.6, outline=FALSE)

#To dataframe
GSE120661_counts_normalized = exprs_matrix.norm %>% 
  as.data.frame() %>% 
  mutate(across(everything(), log2)) #Log transformation

GSE120661_counts_normalized %>% 
  saveRDS(here("tables", "GSE120661_counts_normalized.rds"))
```

Differential expression by limma
```{r}
GSE120661_metadata = read_csv(here("tables", "GSE120661_metadata.csv")) 

GSE120661_metadata = GSE120661_metadata %>% 
  mutate(timepoint = as.integer(timepoint) %>% 
           as.character() %>% 
           as.factor() %>% 
           fct_relevel(., "0"),
         treatment = fct_relevel(treatment, "saline"),
         condition = paste(treatment, timepoint, tissue, sep = "_"))

GSE120661_counts_normalized = read_rds(here("tables", "GSE120661_counts_normalized.rds"))

GSE120661_counts_normalized %>% rownames() %>% length()

#Select and filter
#Filter PBMC samples
GSE120661_samples_pbmc_vector = GSE120661_metadata %>% 
  filter(tissue == "PBMC") %>% 
  select(geo_sample) %>% 
  pull(geo_sample)

GSE120661_samples_pbmc = GSE120661_metadata %>% 
  filter(tissue == "PBMC")

#Select only PBMC samples counts
GSE120661_counts_normalized_matrix = GSE120661_counts_normalized %>% 
  select(c(GSE120661_samples_pbmc_vector)) %>% 
  as.matrix()
```

## Annotate genes
```{r}
library(biomaRt)

## set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## this is the MGI symbol for the gene we're interested in
gene_id <- GSE120661_counts_normalized %>% rownames()

listAttributes(ensembl_mart)

annot <- getBM(
  attributes = c("agilent_sureprint_g3_ge_8x60k", #Agilent microarray data (design ID 028005, retrieved from GEO)
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "agilent_sureprint_g3_ge_8x60k",
  values = gene_id,  # <- Replace with your vector of probe IDs
  mart = ensembl_mart
)
```

## Filter genes with highest median expression

- 39,430 probes in the microarray,
- 39,392 that were annotated in BiomaRt (~99.90%)
- 37,983 probes had a correspondent mgi symbol (96.33% of probes in microarray)
- 31,140 probes had a correspondent entrez id (78.97% of probes in microarray)
- 30,787 probes with entrez id also had a correspondent mgi_symbol (78.08% of probes in microarray)
- 22,435 probes have an entrez gene id. As this is the input in FIT, we used this to filter, not mgi_symbol (56.89% of probes in microarray)
- 21,744 genes were filtered by the highest median (55.14% were kept from the microarray)
- 86 Dupes. The duplication is due to different mgi_symbols. Same Entrez IDs can correspond to different mgi_symbol
- 20849 probes were used (52.87%)

```{r}

#Get annotations for each probe 

GSE120661_counts_normalized_summary = GSE120661_counts_normalized_matrix %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = agilent_sureprint_g3_ge_8x60k), 
            by = "probe") 

# Stats
## 39430 probes in the microarray
GSE120661_counts_normalized_matrix %>% dim()
## 37,983 probes with mgi_symbol
GSE120661_counts_normalized_summary %>% 
  drop_na(mgi_symbol) %>% 
  filter(mgi_symbol != "")
## 31,140 probes with Entrez gene id
GSE120661_counts_normalized_summary %>% 
  drop_na(entrezgene_id)
## 30,787 of these probes with mgi_symbol also had a correspondent entrez gene id
GSE120661_counts_normalized_summary %>% 
  drop_na(mgi_symbol) %>% 
  filter(mgi_symbol != "") %>% 
  drop_na(entrezgene_id)
## 30,787 probes with entrez id also had a correspondent mgi_symbol
GSE120661_counts_normalized_summary %>% 
  drop_na(entrezgene_id) %>% 
  filter(mgi_symbol != "") 

#Get the probes with the highest median across samples for each entrez_id
GSE120661_expression_highest_median_probes = GSE120661_counts_normalized_summary %>% 
  drop_na(entrezgene_id) %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median) %>% 
  ungroup() %>% 
  arrange(entrezgene_id, probe)  %>% 
  select(probe, entrezgene_id, mgi_symbol, median) %>% 
  distinct()

#86 Dupes. The duplication is due to different mgi_symbols. Same Entrez IDs can correspond to different mgi_symbol
GSE120661_expression_highest_median_probes %>% 
  get_dupes(entrezgene_id)

#Modify expression matrix
GSE120661_counts_normalized_matrix = GSE120661_counts_normalized_matrix %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  filter(probe %in% c(GSE120661_expression_highest_median_probes$probe)) %>% 
  column_to_rownames("probe") %>% 
  as.matrix()

GSE120661_counts_normalized_matrix %>% dim()

```



##Differential gene expression
```{r}
dim(GSE120661_counts_normalized_matrix) #20849   279

dim(AnnotatedDataFrame(GSE120661_samples_pbmc %>% column_to_rownames("geo_sample"))) #279
    
eset_filtered <- ExpressionSet(
  assayData = GSE120661_counts_normalized_matrix ,
  phenoData = AnnotatedDataFrame(GSE120661_samples_pbmc %>% column_to_rownames("geo_sample"))
)


design <- model.matrix(~ timepoint * treatment + 0, eset_filtered) # 0 defines the way we compare the samples
colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

#Fit Limma model
fit <- lmFit(eset_filtered, design)
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)
```

```{r}
#Test for Fluad
cont.matrix <- makeContrasts(timepoint4.treatmentFluad - treatmentFluad,
                             timepoint8.treatmentFluad - treatmentFluad,
                             timepoint24.treatmentFluad - treatmentFluad,
                             timepoint48.treatmentFluad - treatmentFluad,
                             timepoint72.treatmentFluad - treatmentFluad,
                             timepoint168.treatmentFluad - treatmentFluad,
                             levels=design)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- treat(fit2, lfc = 0.0)

#Get DEGs
## For one contrast (example)
# all_degs = topTreat(fit2, coef = 1, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE)

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("probe")
    
})

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs
all_degs = all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  mutate(condition = case_when(comparisonA == "Time4.Fluad"	~ "Fluad_4_PBMC",
                               comparisonA == "Time8.Fluad"	~ "Fluad_8_PBMC",
                               comparisonA == "Time24.Fluad"	~ "Fluad_24_PBMC",
                               comparisonA == "Time48.Fluad"	~ "Fluad_48_PBMC",
                               comparisonA == "Time72.Fluad"	~ "Fluad_72_PBMC",
                               comparisonA == "Time168.Fluad"	~ "Fluad_168_PBMC",
                               T~comparisonA)) %>% 
  select(condition, everything())
	
all_degs
```

### Gene annotation
- 27,317 genes in the input matrix
- 25,035 probes from DGE analysis (no lfc threshold)

About FIT genes:
- 16,425 genes in FIT analysis (including human and mouse genes)
- 16,417 unique mouse to human genes (are duplicated)
- 34 genes have either different hgnc symbol to mgi symbol, or the other way.
- 16 human genes are repeated because of different Entrez IDs
- 14,357 mouse genes in FIT are in the limma output. 
- 14,448 unique human genes 
- 223 mouse symbols correspond to the same or another human symbol

```{r}
#Name genes
all_degs_annotated =  all_degs %>% 
  inner_join(GSE120661_expression_highest_median_probes, by = "probe") %>% 
  clean_names()

#Stats
## 20,849 probes
all_degs_annotated %>% 
  select(probe) %>% 
  distinct()

## 21,408 symbols
all_degs_annotated %>% 
  select(mgi_symbol) %>% 
  distinct()

#Get human symbol from FIT genes
library(FIT.mouse2man)

HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez # 16,425 genes

fit_genes = HS_MM_Symbol_Entrez %>% 
  select(mgi_symbol = MGI.symbol,
         human_symbol = HGNC_symbol,
         entrezgene_id_mouse = Mouse.Ortholog,
         entrezgene_id_human = Human.Entrez) %>% 
  rownames_to_column("del") %>% 
  select(!del) 

## 16,417 unique mouse and human genes
fit_genes %>% 
  select(mgi_symbol,
         human_symbol) %>% 
  distinct()

## 34 genes have either different hgnc symbol to mgi symbol, or the other way.
HS_MM_Symbol_Entrez %>% 
  filter(MGI.symbol %in% c(fit_genes %>% 
                             select(human_symbol, mgi_symbol) %>% 
  get_dupes() %>% 
  .$mgi_symbol))

## 16 human genes are repeated because of different Entrez IDs
HS_MM_Symbol_Entrez %>% 
  filter(HGNC_symbol %in% c(fit_genes %>% 
                             select(human_symbol, mgi_symbol) %>% 
  get_dupes() %>% 
  .$human_symbol))

all_degs_annotated_fit = all_degs_annotated %>% 
  select(-mgi_symbol) %>% 
  rename(entrezgene_id_mouse = entrezgene_id) %>% 
  inner_join(fit_genes, by = "entrezgene_id_mouse")

## 15,591 mouse gene symbols from the Limma output are inside FIT genes
all_degs_annotated_fit %>% 
  select(mgi_symbol) %>% 
  distinct()
## 15,613 mouse entrez id from the Limma output are inside FIT genes
all_degs_annotated_fit %>% 
  select(entrezgene_id_mouse) %>% 
  distinct()


## 15,685 human genes from the Limma output are inside FIT genes
all_degs_annotated_fit %>% 
  select(human_symbol) %>% 
  distinct()


all_degs_annotated %>% 
  saveRDS(here("tables", "Fluad_all_degs_annotated.rds"))

all_degs_annotated_fit %>% 
  saveRDS(here("tables", "Fluad_all_degs_annotated_FIT.rds"))

```


```{r}
all_degs_annotated_fit = readRDS(here("tables", "Fluad_all_degs_annotated_FIT.rds"))
all_degs_annotated_fit %>% 
  inner_join(GSE120661_metadata, by = "condition") %>% 
  mutate(timepoint = as.character(timepoint) %>% as.integer()) %>% 
  filter(mgi_symbol == "Tlr8") %>% 
  ggplot() +
  aes(x = timepoint,
      y = log_fc,
      fill = treatment) +
    geom_line(aes(color = treatment)) +
  geom_point(pch = 21,
             color = "black") +
  scale_x_continuous(breaks = c(0, 4, 8, 24, 48, 72, 168),minor_breaks = F) +
    scale_fill_npg() +
  theme_light() +
  theme(
    legend.position = "right",
    plot.caption = element_text(hjust = 0, size = 5),
    axis.text.x = element_text(size = 10,
                               color = "black",
                               angle = 0),
    axis.text.y = element_text(size = 10,
                               color = "black",
                               angle = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 15),
    axis.title.y = element_text(color = "black"),
    panel.border = element_blank(),
    panel.grid.major.y = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    panel.grid.major.x = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0.5 ,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 0.5, color = "black"),
    axis.ticks.y = element_line(size = 0.5, color = "black"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(1, 1, 0, 0, "cm")
  ) +
  labs(x = "Time (h)",
       y = "LogFold-Change")


all_degs_annotated %>% 
  arrange(mgi_symbol)
```




## Found in translation
Ref: Normand, R., Du, W., Briller, M. et al. Found In Translation: a machine learning model for mouse-to-human inference. Nat Methods 15, 1067–1073 (2018). https://doi.org/10.1038/s41592-018-0214-9
Github: https://github.com/shenorrLabTRDF/FIT.mouse2man 
```{r}
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
#Import files
pca_rotations = FIT.mouse2man::pca_rotations
best_models = FIT.mouse2man::best_models
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez
MM_Entrez_symbol_desc = FIT.mouse2man::MM_Entrez_symbol_desc
Psoriasis_classifier_neg = FIT.mouse2man::Psoriasis_classifier_neg
slopes_per_gene_V2.0 = FIT.mouse2man::slopes_per_gene_V2.0
StaphInfection_classifier_pos = FIT.mouse2man::StaphInfection_classifier_pos
AllData_V2.0 = FIT.mouse2man::AllData_V2.0 #Training data
```


```{r}
#Samples
microarray_sample = FIT.mouse2man::microarray_sample #Sample Microarray
microarray_sample %>% 
  rownames_to_column("MM_Entrez_symbol_desc") %>% 
  mutate(MM_Entrez_symbol_desc = as.character(MM_Entrez_symbol_desc)) %>% 
  write_csv(here("FIT_Mouse2man", "microarray_sample.csv"))
RNAseq_sample = FIT.mouse2man::RNAseq_sample #Sample RNAseq
RNAseq_sample %>% 
  write_csv(here("FIT_Mouse2man", "rnaseq_sample.csv"))
```


```{r}
AllData_V2.0 %>% 
  clean_names() %>%
  select(disease:csp_id) %>% 
  distinct()

AllData_V2.0 %>% 
  filter(str_detect(DataType, "Microarray"))

AllData_V2.0 %>% 
  filter(str_detect(DataType, "RNAseq"))
```


```{r}
#Checking applicability of the model to your data
test = FIT.mouse2man::RunClassifier(MouseData = RNAseq_sample, DataType = "rnaseq")
```


```{r}
#Fit
fit_example = FIT.mouse2man::FIT(MouseFile = here("FIT_Mouse2man", "rnaseq_sample.csv"), 
                                 DataType = "rnaseq") # Runs the FIT pipeline which computes prediction of the human effect-size per gene

fit_example %>% 
  saveRDS(here("tables", "fit_example.rds"))

fit_example

fit_example$Mouse.symbol %>% unique() %>% length()
fit_example$Human.symbol %>% unique() %>% length()
?FIT.mouse2man::FIT
```


###Fluad
```{r}

all_degs_fit1 = all_degs_annotated_fit %>% 
  # filter(condition == "Fluad_24_PBMC") %>% 
  arrange(mgi_symbol, entrezgene_id_mouse) %>% 
  arrange(entrezgene_id_mouse) %>% 
  distinct() %>% 
  select(MM.Entrez = entrezgene_id_mouse, 
         row = entrezgene_id_mouse,
         EffectSize = log_fc, 
         condition) 

delete_rows_dupes = all_degs_fit1 %>% 
  filter(condition == "Fluad_4_PBMC") %>% 
  select(row, EffectSize) %>% 
  distinct() %>% 
  drop_na() %>% 
  get_dupes(row) %>% 
  select(row) %>% 
  distinct() %>% 
  pull(row)

all_degs_fit2 = all_degs_fit1 %>% 
  filter(!row %in% c(delete_rows_dupes)) %>% 
  drop_na()

all_degs_fit2 %>% 
  write_csv(here("tables", "Fluad_PBMC_FIT_mouse_allTimePoints.csv"))

```


```{r}
all_degs_fit2 = read_csv(here("tables", "Fluad_PBMC_FIT_mouse_allTimePoints.csv"))

#Fit
delete_dupes = all_degs_fit2 %>% 
  filter(condition == "Fluad_24_PBMC") %>% 
  get_dupes(row) %>% 
  pull(row)

all_degs_fit2 %>% 
  filter(condition == "Fluad_24_PBMC",
         !row %in% c(delete_dupes)) %>% 
  column_to_rownames("row") %>% 
  select(-condition) %>% 
  write_csv(here("tables", "Fluad_24_PBMC_FIT_mouse.csv"))


fit_example = FIT.mouse2man::FIT(MouseFile = here("tables", "Fluad_24_PBMC_FIT_mouse.csv"), 
                                 DataType = "rnaseq") # Runs the FIT pipeline which computes prediction of the human effect-size per gene

fit_example = fit_example %>% 
  mutate(condition = "Fluad_24_PBMC")

fit_example %>% saveRDS(here("FIT_Mouse2man", "Fluad_24_PBMC_FIT_mouse_results.rds"))
fit_example
```


Check gene by gene
14,401 entrez IDs
```{r}
fit_example %>% 
  select(`MM.Entrez` = `Mouse.Entrez`, 
         Mouse.symbol, 
         Human.symbol) %>% 
  distinct()

all_degs_fit2$MM.Entrez %>% unique() %>% length()

```


```{r}
fit_example_geneA = fit_example %>% 
  clean_names() %>% 
  # filter(human_symbol == "TLR8") %>% 
  pivot_longer(cols = c(mouse_effect_size:fit_prediction),
               names_to = "measure", 
               values_to = "value") 


fit_example_geneA = all_degs_annotated_fit %>% 
  rename(mouse_entrez = entrezgene_id_mouse,
         human_entrez = entrezgene_id_human,
         mouse_symbol = mgi_symbol,
         ci_low = ci_l,
         ci_high = ci_r,
         value = log_fc) %>% 
  select(mouse_entrez,
         human_entrez,
         mouse_symbol,
         human_symbol,
         value,
         ci_low,
         ci_high,
         adj_p_val,
         p_value, 
         condition) %>% 
  mutate(measure = "mouse_effect_size",
         mouse_entrez = as.character(mouse_entrez)) %>% 
  bind_rows(fit_example_geneA_mouse)

fit_example_geneA %>% 
  arrange(mouse_entrez)

fit_example_geneA_top10 = fit_example %>% 
  mutate(diff = Mouse_EffectSize - FIT_prediction) %>% 
  arrange(abs(diff)) %>% 
  clean_names() %>% 
           select(mouse_symbol) %>% 
           distinct() %>% 
           slice_head(n = 10) %>% 
           pull(mouse_symbol)
```


```{r}
fit_example_geneA %>% 
  inner_join(GSE120661_metadata %>% 
               select(!c(geo_sample, subject_id, animal_replicate)) %>% 
               distinct() %>% 
               mutate(timepoint = fct_relevel(timepoint, c(GSE120661_metadata$timepoint %>% as.character() %>% as.integer() %>% unique() %>% as.character()))),
             by = "condition") %>% 
  filter(mouse_symbol %in% c(fit_example_geneA_top10)) %>% 
  arrange(mouse_symbol) %>% 
  ggplot() +
  aes(x = human_symbol,
      y = value) +
  geom_point(aes(color = measure,
                 shape = timepoint), position = position_dodge2(width = 0.2)) +
  geom_errorbar(mapping = aes(y))
  theme_light() +
  theme(
    legend.position = "right",
    plot.caption = element_text(hjust = 0, size = 5),
    axis.text.x = element_text(size = 10,
                               color = "black",
                               angle = 45,
                               hjust = 1),
    axis.text.y = element_text(size = 10,
                               color = "black",
                               angle = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 15),
    axis.title.y = element_text(color = "black"),
    panel.border = element_blank(),
    panel.grid.major.y = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    panel.grid.major.x = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0.5 ,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 0.5, color = "black"),
    axis.ticks.y = element_line(size = 0.5, color = "black"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(1, 1, 0, 0, "cm")
  )
  
```


Distribution
```{r fig.height=5, fig.width=5}
distributions_df = fit_example_geneA %>% 
  inner_join(GSE120661_metadata %>% 
               select(!c(geo_sample, subject_id, animal_replicate)) %>% 
               distinct() %>% 
               mutate(timepoint = fct_relevel(timepoint, c(GSE120661_metadata$timepoint %>% as.character() %>% as.integer() %>% unique() %>% as.character()))),
             by = "condition")
 

#Boxplots
distributions_df %>% 
  ggplot() +
  aes(x = timepoint,
         y = value,
         color = measure)+
  geom_boxplot(alpha = 0.1) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.caption = element_text(hjust = 0, size = 5),
    axis.text.x = element_text(size = 10,
                               color = "black",
                               angle = 0),
    axis.text.y = element_text(size = 10,
                               color = "black",
                               angle = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 15),
    axis.title.y = element_text(color = "black"),
    panel.border = element_blank(),
    panel.grid.major.y = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    panel.grid.major.x = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0.5 ,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 0.5, color = "black"),
    axis.ticks.y = element_line(size = 0.5, color = "black"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(1, 1, 0, 0, "cm")
  ) + 
  labs(x = "Timepoint (h)",
       y = "LogFold-Change",
       color = "Measure") +
    scale_color_manual(
  values = c(fit_prediction = "#E64B35FF", 
             mouse_effect_size = "#4DBBD5FF"),
  labels = c(fit_prediction = "FIT Prediction",
             mouse_effect_size = "Mouse Effect Size"))

#Error bars
distributions_df %>% 
 ggplot() +
  aes(x = timepoint,
         y = value,
         color = measure,
      fill = measure) +
  stat_summary(fun.y = mean, geom = "col", position = "dodge",
               color = "black",
               # fill = NA,
               size = 0.3,
               alpha = 0.5) + 
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar",
               color = "black",
               size = 0.5,
               # width = 0.2, 
               position = "dodge") +
  scale_y_continuous(expand = c(0.01,0.01)) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.caption = element_text(hjust = 0, size = 5),
    axis.text.x = element_text(size = 10,
                               color = "black",
                               angle = 0),
    axis.text.y = element_text(size = 10,
                               color = "black",
                               angle = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 15),
    axis.title.y = element_text(color = "black"),
    panel.border = element_blank(),
    panel.grid.major.y = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    panel.grid.major.x = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0.5 ,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 0.5, color = "black"),
    axis.ticks.y = element_line(size = 0.5, color = "black"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(1, 1, 0, 0, "cm")
  ) + 
  labs(x = "Timepoint (h)",
       y = "LogFold-Change",
       color = "Measure") +
  scale_color_manual(
  values = c(fit_prediction = "#E64B35FF", 
             mouse_effect_size = "#4DBBD5FF"),
  labels = c(fit_prediction = "FIT Prediction",
             mouse_effect_size = "Mouse Effect Size")) +
  scale_fill_manual(
  values = c(fit_prediction = "#E64B35FF", 
             mouse_effect_size = "#4DBBD5FF"),
  labels = c(fit_prediction = "FIT Prediction",
             mouse_effect_size = "Mouse Effect Size"))
```



###Loops
```{r}

all_degs_fit2 = read_csv(here("tables", "Fluad_PBMC_FIT_mouse_allTimePoints.csv"))

# Generate csv for every condition
all_conditions <- unique(all_degs_fit2$condition)
purrr::walk(all_conditions, function(cond) {
  
  delete_dupes <- all_degs_fit2 %>%
    filter(condition == cond) %>%
    get_dupes(row) %>%
    pull(row)
  
  filtered_df <- all_degs_fit2 %>%
    filter(condition == cond, !row %in% delete_dupes) %>%
    select(-condition) %>% 
    column_to_rownames("row")
  
  file_name <- paste0(cond, "_FIT_mouse.csv")
  file_path <- here("tables", file_name)
  
  write_csv(filtered_df, file_path)
})

```
## Loop FIT for every timepoint
Obs: Make sure that datasets from FIT package were retrieved

Initial number of genes: 15503
Number of genes for which FIT predictions will be calculated: 14401

```{r}
#Import files
pca_rotations = FIT.mouse2man::pca_rotations
best_models = FIT.mouse2man::best_models
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez
MM_Entrez_symbol_desc = FIT.mouse2man::MM_Entrez_symbol_desc
Psoriasis_classifier_neg = FIT.mouse2man::Psoriasis_classifier_neg
slopes_per_gene_V2.0 = FIT.mouse2man::slopes_per_gene_V2.0
StaphInfection_classifier_pos = FIT.mouse2man::StaphInfection_classifier_pos
AllData_V2.0 = FIT.mouse2man::AllData_V2.0 #Training data


# FIT loop
fit_results_combined <- map2_dfr(all_conditions, all_conditions, function(cond, cond_name) {
  
  # 1. Monta o caminho do CSV
  file_path <- here("tables", paste0(cond, "_FIT_mouse.csv"))
  
  # 2. Roda o FIT
  fit_output <- FIT.mouse2man::FIT(MouseFile = file_path, DataType = "rnaseq")
  
  # 3. Converte para tibble e adiciona a coluna da condição
  fit_output_tbl <- as_tibble(fit_output) %>%
    mutate(condition = cond_name)
  
  return(fit_output_tbl)
})

fit_results_combined

fit_results_combined %>% saveRDS(here("FIT_Mouse2man", "FIT_FLUAD_alltimes.csv"))
```



```{r fig.height=5, fig.width=5}
fit_results_combined = read_rds(here("FIT_Mouse2man", "FIT_FLUAD_alltimes.csv"))
fit_example_results = fit_results_combined %>% 
  clean_names() %>% 
  # filter(human_symbol == "TLR8") %>%
  select(condition, everything()) %>% 
  pivot_longer(cols = c(mouse_effect_size:fit_prediction),
               names_to = "measure", 
               values_to = "value")

fit_example_mouse = all_degs_annotated_fit %>% 
  rename(mouse_entrez = entrezgene_id_mouse,
         human_entrez = entrezgene_id_human,
         mouse_symbol = mgi_symbol,
         ci_low = ci_l,
         ci_high = ci_r,
         value = log_fc) %>% 
  select(mouse_entrez,
         human_entrez,
         mouse_symbol,
         human_symbol,
         value,
         ci_low,
         ci_high,
         adj_p_val,
         p_value, 
         condition) %>% 
  mutate(measure = "mouse_effect_size",
         mouse_entrez = as.character(mouse_entrez),
         up_down = case_when(measure > 0 ~"+",
                             measure <0 ~ "+",
                             T~"-"),
         ci_size = ci_high - ci_low) %>% 
  filter(mouse_entrez %in% c(fit_example_results$mouse_entrez))

fit_example_results_ready = fit_example_mouse %>% 
  bind_rows(fit_example_results) %>% 
  arrange(mouse_symbol) %>% 
  select(condition, everything()) %>% 
  select(!c(adj_p_val, p_value, description)) 

fit_example_results %>% 
  arrange(mouse_entrez)
```


```{r fig.height=5, fig.width=5}
fit_example_geneA_top5 = fit_example_geneA %>% 
           filter(measure == "mouse_effect_size") %>% 
           select(human_symbol, value) %>% 
           distinct() %>% 
           arrange(-value) %>% 
  select(human_symbol) %>% 
  distinct() %>% 
           slice_head(n = 5) %>% 
           pull(human_symbol)
fit_example_geneA_top5
```


```{r fig.height=5, fig.width=5}
fit_example_topgenes_plot = fit_example_results_ready%>% 
  inner_join(GSE120661_metadata %>% 
               mutate(timepoint = as.character(timepoint) %>% 
                        as.integer()) %>% 
               select(!geo_sample:subject_id) %>% 
               select(-animal_replicate) %>% 
               distinct(), by = "condition") %>% 
  filter(human_symbol %in% c(fit_example_geneA_top5)) %>% 
  ggplot() +
  aes(x = timepoint,
      y = value) +
  # geom_ribbon(aes(ymin = ci_low, 
  #                 ymax = ci_high),
  #             fill = "red",
  #             alpha = 0.1) +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high,
                  color = measure)) +
  geom_line(aes(color = measure)) +
  geom_point(aes(color = measure)) +
  scale_x_continuous(breaks = c(0, 4, 8, 24, 48, 72, 168),minor_breaks = F) +
  scale_color_manual(
  values = c(
    fit_prediction = "#E64B35FF", 
    mouse_effect_size = "#4DBBD5FF"
  ),
  labels = c(
    fit_prediction = "FIT Prediction",
    mouse_effect_size = "Mouse Effect Size"
  )
)+
  facet_wrap(~human_symbol, ncol = 1) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.caption = element_text(hjust = 0, size = 5),
    axis.text.x = element_text(size = 10,
                               color = "black",
                               angle = 0),
    axis.text.y = element_text(size = 10,
                               color = "black",
                               angle = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 15),
    axis.title.y = element_text(color = "black"),
    panel.border = element_blank(),
    panel.grid.major.y = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    panel.grid.major.x = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0.5 ,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 0.5, color = "black"),
    axis.ticks.y = element_line(size = 0.5, color = "black"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(0.5, 0.2, 0.2, 0.2, "cm")
  )+
  labs(color = "Measure",
       x = "Time (h)",
       y = "LogFold-Change",
       title = "Prediction - Influenza, Fluad Vaccine")

fit_example_topgenes_plot
```


Compare with human log2fc
```{r}
cleandata_nonorm_flu <- readRDS(here("HumanFluData", "cleandata_nonorm_flu.rds")) 

flu_expFC <- readRDS(here("HumanFluData", "expFC.rds"))
flu_expression_fc = flu_expFC[["exp_fc"]] %>% 
  as.data.frame()

flu_metadata = flu_expFC$cur_phenoData

flu_expression_fc_long = flu_expression_fc %>%
  rownames_to_column("genes") %>% 
  pivot_longer(cols = -genes,
               names_to = "uid",
               values_to = "fold_change") 

flu_metadata %>% 
  select(study_time_collected) %>%
  group_by(study_time_collected) %>% 
  summarize(n = n())

flu_expression_fc_long_metadata = flu_expression_fc_long %>% 
  inner_join(flu_metadata, by = "uid") 

flu_expression_fc_summary = flu_expression_fc_long_metadata %>% 
  mutate(l2fc = sign(fold_change) * log2(abs(fold_change))) %>% 
  group_by(genes, exposure_material_reported, adjuvant, study_time_collected) %>% 
    summarize(mean = mean(l2fc, na.rm = TRUE),
    se = sd(l2fc, na.rm = TRUE) / sqrt(n()),
    lower = mean - 1.96 * se,
    upper = mean + 1.96 * se,
    study_time_collected,
    vaccine_type,
    adjuvant) %>% 
  ungroup() %>% 
  distinct()

flu_expression_fc_summary
```


```{r}
fit_fluad_mouse = fit_example_geneA %>% 
  inner_join(GSE120661_metadata %>% 
               mutate(timepoint = as.character(timepoint) %>% 
                        as.integer()) %>% 
               select(!geo_sample:subject_id) %>% 
               select(-animal_replicate) %>% 
               distinct(), by = "condition") 
   
flu_expression_fc_summary_standard = flu_expression_fc_summary %>% 
  rename(human_symbol = genes,
         value = mean) %>%
  mutate(measure = "human_effect_size") %>% 
  mutate(condition = paste0(exposure_material_reported, "_", study_time_collected, "_", adjuvant)) %>% 
  rename(treatment = exposure_material_reported,
         timepoint = study_time_collected,
         ci_low = lower,
         ci_high = upper) %>% 
  mutate(timepoint = as.numeric(timepoint))

fit_fluad_mouse_human = flu_expression_fc_summary_standard %>% 
  bind_rows(fit_fluad_mouse %>% 
              mutate(timepoint = timepoint/24)) %>% 
  filter(human_symbol %in% c(fit_fluad_mouse$human_symbol))


fit_fluad_mouse %>% 
  filter(human_symbol == "TLR8")

fit_fluad_mouse_human %>% 
  filter(human_symbol == "TLR8")

```



```{r fig.height=5, fig.width=5}
fit_example_human = fit_fluad_mouse_human %>% 
  filter(timepoint %in% c(1, 3, 7),
         treatment %in% c("Fluad", "FLUADD")) %>% 
  select(human_symbol:vaccine_type, condition, measure) %>% 
  mutate(organism = if_else(measure == "mouse_effect_size", "Mouse", "Human"),
         organism = if_else(measure == "fit_prediction", "FIT Prediction", organism))


pairwise_human_fit = fit_example_human %>% 
  drop_na(human_symbol) %>% 
  select(timepoint, organism, value, human_symbol) %>% 
  arrange(human_symbol) %>% 
  pivot_wider(names_from = "organism",
              values_from = "value",
              values_fn = mean) %>% 
  mutate(human_2_mouse = Human - Mouse,
         human_2_fit = Human - `FIT Prediction`) %>% 
  select(human_symbol, timepoint, Human, Mouse, `FIT Prediction`, human_2_mouse, human_2_fit)

pairwise_human_fit$human_symbol %>% unique() %>% length()
```


```{r fig.height=5, fig.width=5}
#Rank genes by differences
meta_human = fit_example_human %>% 
  # filter(human_symbol == "AAK1") %>% 
  filter(organism == "Human") %>% 
  select(human_symbol, 
         timepoint, 
         mean_human = value, 
         se_human = se, 
         ci_low_human = ci_low,
         ci_high_human = ci_high
         ) 


meta_fit = fit_example_human %>% 
  # filter(human_symbol == "AAK1") %>% 
  filter(organism == "FIT Prediction") %>% 
  select(human_symbol, timepoint, 
         mean_fit = value, 
         se_fit = se, 
         ci_low_fit = ci_low,
         ci_high_fit = ci_high
         ) 

meta_mouse = fit_example_human %>% 
  # filter(human_symbol == "AAK1") %>%
  filter(organism == "Mouse") %>% 
  select(human_symbol, timepoint, 
         mean_mouse = value, 
         se_mouse = se, 
         ci_low_mouse = ci_low,
         ci_high_mouse = ci_high
         ) 


pairwise_human_fit_measures = meta_human %>% 
  inner_join(meta_fit, by = join_by(human_symbol, timepoint)) %>% 
  inner_join(meta_mouse, by = join_by(human_symbol, timepoint)) %>% 
  inner_join(pairwise_human_fit, by = join_by(human_symbol, timepoint))

pairwise_human_fit_test = pairwise_human_fit_measures %>%
  mutate(
    # Estimate SE from CIs
    se_fit = (ci_high_fit - ci_low_fit) / (2 * 1.96),
    se_mouse = (ci_high_mouse - ci_low_mouse) / (2 * 1.96),
    # Combine SE
    se_diff_human_fit = sqrt(se_human^2 + se_fit^2),
    se_diff_human_mouse = sqrt(se_human^2 + se_mouse^2),
    
    # Z score
    z_score_human_fit = human_2_fit / se_diff_human_fit,
    z_score_human_mouse = human_2_mouse / se_diff_human_mouse,
    
    # p-value
    p_value_fit = 2 * (1 - pnorm(abs(z_score_human_fit))),
     p_value_mouse = 2 * (1 - pnorm(abs(z_score_human_mouse))),
    
    # FDR (Benjamini-Hochberg)
    p_value_fit_adj = p.adjust(p_value_fit, method = "BH"),
    p_value_mouse_adj = p.adjust(p_value_mouse, method = "BH"),
    
    # Cohen's d
    effect_size_h2fit = human_2_fit / se_diff_human_fit,
    effect_size_h2m = human_2_mouse / se_diff_human_mouse
  ) %>%
  arrange(p_value_fit)


pairwise_human_fit_test_p10 = pairwise_human_fit_test %>% 
  arrange(-p_value_fit_adj) %>% 
  filter(p_value_fit_adj > 0.10)

pairwise_human_fit_test$human_symbol %>% unique() %>% length()

pairwise_human_fit_test_p10$human_symbol %>% unique() %>% length()
  
```

#### FIT vs Human
```{r fig.height=5, fig.width=8}
fit_example_human_plot = 
  fit_example_human %>% 
  filter(human_symbol %in% c(pairwise_human_fit_test_p10 %>% 
                               slice_head(n = 5) %>% 
                               pull(human_symbol))) %>% 
  ggplot() +
  aes(x = timepoint,
      y = value,
      color = measure) +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  "dodge2") +
  scale_x_continuous(breaks = c(0,1, 3, 7),minor_breaks = F) +
    scale_color_manual(
  values = c(
    fit_prediction = "#E64B35FF", 
    mouse_effect_size = "#4DBBD5FF",
    human_effect_size = "black"
  ),
  labels = c(
    fit_prediction = "FIT",
    mouse_effect_size = "Mouse",
    human_effect_size = "Human"
  )
)+
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.caption = element_text(hjust = 0, size = 5),
    axis.text.x = element_text(size = 10,
                               color = "black",
                               angle = 0),
    axis.text.y = element_text(size = 10,
                               color = "black",
                               angle = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 15),
    axis.title.y = element_text(color = "black"),
    panel.border = element_blank(),
    panel.grid.major.y = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    panel.grid.major.x = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0.5 ,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 0.5, color = "black"),
    axis.ticks.y = element_line(size = 0.5, color = "black"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(0.5, 0.2, 0.2, 0.2, "cm")
  )+
  labs(color = "Measure",
       x = "Time (days)",
       y = "LogFold-Change",
       title = "Prediction - Influenza, Fluad Vaccine") +
  facet_wrap(~human_symbol, nrow = 1)

fit_example_human_plot

```
#### Plot mouse vs fit
```{r}

```


#### Plot mouse vs fit vs human
Genes that are not DEGs in mouse but are in humans and are predicted in FIT, and the others that are not predicted in FIT (these should be used to improve model performance)

#### Clustering
```{r fig.height=15, fig.width=30}
df =  pairwise_human_fit_test %>% 
  select(human_symbol, timepoint, p_value_fit_adj, effect_size_h2fit) %>% 
  arrange(-p_value_fit_adj) %>%
  rename(gene = human_symbol,
         day = timepoint,
         p_adj = p_value_fit_adj,
         effect_size = effect_size_h2fit) %>% 
  inner_join(fit_example_human %>% 
               filter(organism == "Human") %>% 
               select(condition, treatment, adjuvant, vaccine_type, day = timepoint),
             by = "day") %>% 
  mutate(day = as.character(day),
         day = day %>% fct_relevel(day, "0", "1", "3", "7")) %>% 
  distinct()

```


```{r fig.height=15, fig.width=30}
row_clustered = df %>% 
  mutate(effect_size = abs(effect_size)) %>% 
  dplyr::select(condition, gene, effect_size) %>% 
  distinct() %>% 
  pivot_wider(names_from = "condition",
              values_from= "effect_size") %>%
  column_to_rownames("gene") %>% 
  as.matrix() %>% 
  replace(is.na(.), 0) %>% 
  dist(.) %>% 
  hclust()


# Reordenar o dataframe original conforme o clustering
df_row_ordered <- df[row_clustered$order, ]
```


```{r fig.height=15, fig.width=5}
row_shared =  df_row_ordered %>%
  filter(gene %in% row_clustered$labels, 
         p_adj > 0.10) %>% 
  mutate(gene = as_factor(gene))

#Plot
dotplot = row_shared %>% 
  filter(p_adj > 0.10) %>% 
  ggplot(aes(x= condition, 
             y = gene, 
             fill = effect_size, 
             # color = effect_size
             )) +
  geom_tile() +
  theme_minimal() + 
  theme(axis.text = element_text(color = "black", size = 12),
        axis.line.x  = element_line(size = 1, color ="black"),
        # axis.line.y  = element_line(size = 1, color ="black"),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 0, angle = 45, vjust = 1, hjust=1),
        axis.ticks = element_line(size = 1, color ="black"),
        # panel.grid = element_line(linetype = 2, color = "gray75"),
        panel.grid = element_blank(),
        legend.direction = "horizontal",
        legend.position = "bottom") +
  labs(x = "",
       y = "") +
  scale_fill_gradient2(
  low = "white",
  mid = "#E64B35FF",
  high = "white",
  midpoint = 0,
  limits = c(-1, 1),
  oob = scales::squish
) +
  scale_y_discrete(position = "right")
dotplot
```


Complex heatmap

By effect size
```{r fig.height=6, fig.width=5}
heatmap_data = df
file = "df_human_fit"

heatmap_data_l2fc= heatmap_data %>% 
  filter(p_adj > 0.10) %>% 
  dplyr::select(genes = gene, condition, effect_size)

matrix_data <- dcast(heatmap_data_l2fc, 
                         genes ~ condition, 
                         value.var = "effect_size", 
                         fun.aggregate = mean) %>% 
  column_to_rownames('genes') %>% 
  mutate_if(is.character, as.numeric) %>% 
  replace(is.na(.), 0)

matrix_data_ready =  matrix_data %>% 
  round(2)

col_fun <- colorRamp2(c(-0.5, 0, 0.5), c("white",
                                     # "#4DBBD5FF", 
                                     "#E64B35FF",
                                     "white"))


mat = matrix_data_ready %>% 
  as.matrix()
width = unit(5, "cm")
height = unit(10, "cm")
width_image = 10
height_image = 20
column_names_gp = gpar(fontsize = 10)
row_names_gp = gpar(fontsize = 0)

fileimageheatmap_ungrouped= here("Figures", paste0(file, sep ="_", "Ungrouped_Complexheatmap.png"))
png(file = fileimageheatmap_ungrouped, width=width_image, height=height_image,units="cm",res=300)
heatmap_plot_all = Heatmap(mat,
                        # top_annotation = ha,
                        # show_row_names = TRUE,
                        show_heatmap_legend = TRUE,
                        heatmap_legend_param = list(direction = "vertical"),
                        name = "Effect size", #Legend
                        col = col_fun, #color
                        column_title = "",
                        # column_title_gp = gpar(fontsize = 15, fontface = "bold"),
                        column_names_rot = 45,
                        column_names_gp = gpar(fontsize = 10),
                        # rect_gp = gpar(col = "gray90", lwd = 1),
                        cluster_rows = TRUE,
                        row_names_gp = row_names_gp,
                        row_split = 5, #Split row clusters
                        row_gap = unit(2, "mm"), row_dend_reorder = T,
                        cluster_columns = T,
                        column_split = 3, #Split group clusters
                        #column_km = 2,
                        column_gap = unit(2, "mm"),
                        show_column_dend = F,
                        show_row_dend = T,
                        row_dend_side = "left",
                        column_dend_side = "top",
                        clustering_distance_rows = "euclidean",
                        border = TRUE,
                        column_dend_reorder = TRUE,
                        width = width, 
                        height = height
)

draw(heatmap_plot_all, annotation_legend_side = "right", heatmap_legend_side = "right")

dev.off()

heatmap_plot_all
```


```{r fig.height=6, fig.width=5}
heatmap_data = df
file = "df_human_fit_pvalue"

heatmap_data_l2fc= heatmap_data %>% 
  filter(p_adj > 0.10) %>% 
  dplyr::select(genes = gene, condition, p_adj)

matrix_data <- dcast(heatmap_data_l2fc, 
                         genes ~ condition, 
                         value.var = "p_adj", 
                         fun.aggregate = mean) %>% 
  column_to_rownames('genes') %>% 
  mutate_if(is.character, as.numeric) %>% 
  replace(is.na(.), 0)

matrix_data_ready =  matrix_data %>% 
  round(2)

col_fun <- colorRamp2(c(0, 1), c("white",
                                     "#E64B35FF"))


mat = matrix_data_ready %>% 
  as.matrix()
width = unit(5, "cm")
height = unit(10, "cm")
width_image = 10
height_image = 20
column_names_gp = gpar(fontsize = 10)
row_names_gp = gpar(fontsize = 0)

fileimageheatmap_ungrouped= here("Figures", paste0(file, sep ="_", "Ungrouped_Complexheatmap.png"))
png(file = fileimageheatmap_ungrouped, width=width_image, height=height_image,units="cm",res=300)
heatmap_plot_all = Heatmap(mat,
                        # top_annotation = ha,
                        # show_row_names = TRUE,
                        show_heatmap_legend = TRUE,
                        heatmap_legend_param = list(direction = "vertical"),
                        name = "Adjusted P-Value", #Legend
                        col = col_fun, #color
                        column_title = "",
                        # column_title_gp = gpar(fontsize = 15, fontface = "bold"),
                        column_names_rot = 45,
                        column_names_gp = gpar(fontsize = 10),
                        # rect_gp = gpar(col = "gray90", lwd = 1),
                        cluster_rows = TRUE,
                        row_names_gp = row_names_gp,
                        row_split = 5, #Split row clusters
                        row_gap = unit(2, "mm"), row_dend_reorder = T,
                        cluster_columns = T,
                        column_split = 3, #Split group clusters
                        #column_km = 2,
                        column_gap = unit(2, "mm"),
                        show_column_dend = F,
                        show_row_dend = T,
                        row_dend_side = "left",
                        column_dend_side = "top",
                        clustering_distance_rows = "euclidean",
                        border = TRUE,
                        column_dend_reorder = TRUE,
                        width = width, 
                        height = height
)

draw(heatmap_plot_all, annotation_legend_side = "right", heatmap_legend_side = "right")

dev.off()

heatmap_plot_all
```



```{r}
df %>% 
  mutate(log_padj = log10(p_adj),
         effect_size_direction = if_else(effect_size > 0, "+", "-")) %>% 
  ggplot() +
  aes(x = effect_size,
         y = p_adj,
         fill = effect_size_direction,
         # shape = day,
      # size = log_padj
      ) +
  geom_jitter(aes(fill = effect_size_direction,
                  alpha = p_adj),
              width = 1,
              pch = 21,
              color = "black") +
  gghighlight(p_adj > 0.10) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.caption = element_text(hjust = 0, size = 5),
    axis.text.x = element_text(size = 10,
                               color = "black",
                               angle = 0),
    axis.text.y = element_text(size = 10,
                               color = "black",
                               angle = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 15),
    axis.title.y = element_text(color = "black"),
    panel.border = element_blank(),
    panel.grid.major.y = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    panel.grid.major.x = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0.5 ,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 0.5, color = "black"),
    axis.ticks.y = element_line(size = 0.5, color = "black"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(0.5, 0.2, 0.2, 0.2, "cm")
  )+
  facet_grid(~day %>% fct_relevel(., "1", "3", "7"), scales = "free_x") 




```
























