```{r}
library(tidyverse)
library(here)
library(ggsci)
library(janitor)
library(readr)
library(explore)
library(maditr)
library(GEOquery)
library(Matrix)
library(circlize)
library(RColorBrewer)
library(celldex)
library(biomaRt)
library(org.Hs.eg.db)
library(plotly)
library(DESeq2)
library(msigdbr)
library(ape)
library(GSVA)
library(sva)
library(clusterProfiler)
library(pheatmap)
library(EnhancedVolcano)
library(ggbeeswarm)
library(tidyverse)
library(corrr)
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(esquisse)
library(corto)
library(factoextra)
library(reshape2)
library(ComplexHeatmap)
library(janitor)
```


#Data curation
## BioProject
```{r}
animals_vaccines_bioproject_result <- read_csv("animals_vaccines_bioproject_result.csv", 
    col_names = FALSE)
datacuration_clean = animals_vaccines_bioproject_result %>%
 drop_na() %>% 
  mutate(study = if_else(X1 =="Title", X2, NA)) %>% 
  fill(study, .direction = c("down")) %>% 
  pivot_wider(names_from = X1, values_from = X2) %>% 
  select(-study) %>% 
  clean_names() %>% 
  mutate(title = str_remove(title, "^\\d+\\.\\s"),
         organism = str_remove(organism, "Organism: "),
         organism = str_remove(organism, "Taxonomy ID "),
         bio_project_accession = str_remove(bio_project_accession, "BioProject Accession: "),
         id = str_remove(id, "ID: "))

datacuration_clean

datacuration_clean %>% 
  write_csv(here("tables", "datacuration_step2.csv"))
```


```{r fig.height=5, fig.width=7}
datacuration_clean %>%
  group_by(organism) %>% 
  summarize(n_organism = n()) %>% 
  # filter(!str_detect(organism,"Homo sapiens")) %>% 
  arrange(-n_organism) %>% 
  ggplot() +
  aes(x = n_organism,
      y = organism %>% fct_reorder(n_organism)) +
  geom_col(fill = "black") + 
  theme_light()+
  geom_text(aes(label = n_organism),
            hjust = -0.2,
            size =3) +
  scale_x_continuous(expand = c(0.01,0.01),
                     limits = c(0, 900)) +
  theme(
    legend.position = c(0.95,0.5),
    plot.caption = element_text(hjust = 0, size = 5),
    axis.text.x = element_text(size = 10,
                               color = "black",
                               angle = 0),
    axis.text.y = element_text(size = 10,
                               color = "black",
                               angle = 0),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(size = 15),
    axis.title.y = element_text(color = "black"),
    panel.border = element_blank(),
    panel.grid.major.y = element_line(size = 0, 
                                    linetype = 2,
                                    color = "gray75"),
    panel.grid.major.x = element_line(size = 0.25, 
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0 ,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 0.5, color = "black"),
    axis.ticks.y = element_line(size = 0.5, color = "black"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(1, 1, 0, 0, "cm")
  ) +
    labs(x = "Frequency",
       y = "",
       title = "BioProject data curation",
       subtitle = "Immunization and vaccination datasets",
       caption = "Search terms: (('vaccine'[All Fields] OR 'immunization'[All Fields]\nOR 'vaccinated'[All Fields] OR 'vaccination'[All Fields] OR 'vaccines'[All Fields]) AND 'transcriptome gene expression'[Filter]\n AND 'material transcriptome'[Filter] AND 'capture_whole'[Filter] AND 'org mammals'[Filter] AND 'Mus musculus'[Organism]") 
  
```
```{r}
mice_vaccination_mining = datacuration_clean %>% 
  filter(str_detect(organism,"Mus musculus")) %>% 
  filter(!str_detect(str_to_lower(title), "cancer|tumor|autoimmune|drug")) %>% 
  filter(str_detect(str_to_lower(title), "vaccin|immuniz"))


mice_vaccination_mining
```


BioProject metadata
```{r}
library(rentrez)

#Single
bioprojectid = mice_vaccination_mining$id[[1]]
bioproject_meta = entrez_summary(db = "bioproject", id = bioprojectid) %>% 
  as.tibble() %>% 
  select(project_name,
         project_description,
         organism_name,
         project_id,
         project_data_type,
         project_target_material,
         project_target_capture,
         organism_name)


#Loop
bioproject_ids <- mice_vaccination_mining$id
bioproject_metadata_df <- map_dfr(bioproject_ids, ~
  entrez_summary(db = "bioproject", id = .x) %>%
    as.list() %>%
    as_tibble() %>%
    select(
      project_name,
      project_description,
      organism_name,
      project_data_type,
      project_target_material,
      project_target_capture
    ) %>%
    mutate(bioproject_id = .x)
)


bioproject_metadata_df = bioproject_metadata_df %>% 
  filter(!str_detect(project_description, "cancer|neoantigen|tumor|autoimmune|drug|opioid"))

# Visualizar
bioproject_metadata_df %>% 
  write_excel_csv2(here("tables", "mice_vaccination_mining2.csv"))

mice_vaccination_mining2 <- read_delim("tables/mice_vaccination_mining2.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)


bioproject_metadata_df

```


GEO
```{r}
library(glue)
#Single ------
# query NCBI GEO 
geo_ids <- "GSE199750"
terms <- entrez_db_searchable(db = "gds")  # for reference: searchable fields
# first, get additional information about this series, including its db ID.
series <- entrez_search("gds",
                        term = glue("{geo_ids}[ACCN] AND gse[ETYP]"))
# use the ID to fetch details, including the sample table
series_summary = entrez_summary(db = "gds", id = series$ids)
series_summary_df = data_frame(bioproject = series_summary$bioproject,
                               uid = series_summary$uid,
                               gse= series_summary$gse,
                               gds= series_summary$gds,
                               title= series_summary$title,
                               summary= series_summary$summary,
                               gpl= series_summary$gpl,
                               n_samples= series_summary$n_samples,
                               platformtitle= series_summary$platformtitle)

#Multiple -------

# geo_ids = read_delim(here("tables", "gds_result.txt"), delim = "\n", col_names = F) %>% 
#   mutate(X1 = str_c("GSE", X1, sep = "") %>% 
#            str_remove(., "200"))
#   pull()

text = readLines(here("tables", "gds_result (1).txt"))
text_collapsed <- paste(text, collapse = "\n")
gse_ids <- str_extract_all(conteudo, "GSE\\d+\\b") %>% unlist() %>% unique()


geo_metadata_df <- map_dfr(gse_ids, ~ {
  result <- entrez_search("gds", term = glue("{.x}[ACCN] AND gse[ETYP]"))

  if (length(result$ids) == 0) {
    return(tibble(gse = .x, error = "Not found"))
  }

  summary <- entrez_summary(db = "gds", id = result$ids)

  tibble(
    gse = .x,
    uid = summary$uid,
    gds = summary$gds,
    bioproject = summary$bioproject,
    title = summary$title,
    summary_text = summary$summary,
    gpl = summary$gpl,
    n_samples = summary$n_samples,
    platformtitle = summary$platformtitle
  ) %>% 
    mutate(bioproject_id = str_remove(bioproject, "PRJNA"))
})
print(geo_metadata_df)

  
  
```


Merge by Bioproject ID
```{r}
mice_vaccination_mining_ManualAnnotation <- read_csv(here("tables", "mice_vaccination_mining_ManualAnnotation.csv")) %>% 
  mutate(bioproject_id = as.character(bioproject_id),
         project_id = as.character(project_id)) %>% 
  drop_na(project_id)


mice_vaccination_annotated = geo_metadata_df %>% 
  right_join(mice_vaccination_mining_ManualAnnotation,
             by = "bioproject_id") %>% 
  select(title, summary_text,
         project_description, n_samples,
         notes, keywords, timepoints:project_target_material,
         gse, bioproject_id
         ) %>% 
  arrange(-n_samples)

mice_vaccination_annotated %>% 
  write_csv(here("tables", "mice_vaccination_annotated.csv"))

mice_vaccination_annotated
```




# MouseImmuneGO

## Gene sets construction


### Blood transcription modules <https://www.nature.com/articles/ni.2789>

```{r}
#Converter ítens de "Module member genes" em linhas em uma tabela longa
btm_annotation_table_long_immune <- BTM_modules %>%
  separate_rows(`Module member genes`, sep = ",") %>% 
  filter(`Module category` == "immune")

#Salvar
write.csv(btm_annotation_table_long_immune, file = "btm_annotation_table_long_immune.csv")

#Display
head(btm_annotation_table_long_immune)

#Calculate Set size
btm_annotation_nested <- btm_annotation_table_long_immune %>%
  group_by(`Module title`) %>% #Somar e agrupar
  mutate(SetSize = n())

#Convert gene symbol column to aggregated rows
btm_annotation_nested <- btm_annotation_nested %>%
  group_by(`Module title`) %>%
  summarize(Genes = paste(`Module member genes`, collapse = ", "), .groups = "drop") %>% 
  left_join(btm_annotation_nested, by = "Module title") %>% 
  filter(Genes != "NA") %>% 
  relocate(Genes, .after="SetSize")

#Salvar
write.csv(btm_annotation_nested, file = "btm_annotation_genes.csv")

#Convert to unique values table
btm_annotation_unique = btm_annotation_nested %>% 
  dplyr::select(-`Module member genes`)

btm_annotation_unique = btm_annotation_unique %>% 
  distinct() %>% 
  relocate(Genes, .after="SetSize") %>% 
  filter(Genes != "NA")

#salvar
write.csv(btm_annotation_unique, file = "btm_annotation_unique.csv")

#Display
head(btm_annotation_unique)

btm_annotation_table_long_immune
```

### CellMarker

Download database from
"<http://117.50.127.228/CellMarker/CellMarker_download_files/file/Cell_marker_Human.xlsx>"

```{r}
##################### Baixar tabela de marcadores celulares

url <- "http://117.50.127.228/CellMarker/CellMarker_download_files/file/Cell_marker_Human.xlsx" 
f <- tempfile(fileext = ".xlsx")
download.file(url, f)
cell_marker_data <- readxl::read_excel(f, 1)

##################### Sub-agrupar células por sistema imunológico

#Filtrar marcadores e células por tecido (Blood)
cells <- cell_marker_data %>%
    filter(tissue_class == "Blood") %>%
    dplyr::select(cell_name, marker)

##################### Filtrar células
# Acessar todos os tipos de células no dataset de Blood
cellmarker_celltypes <- unique(cells$cell_name)

# Filtrar apenas os tipos que contêm "B cell" e plasmacell
Bcells_no = cellmarker_celltypes[grep("B cell", cellmarker_celltypes)] #44 outputs
Plasmacells_no = cellmarker_celltypes[grep("plasma", cellmarker_celltypes)] #5 outputs

# Filtrar apenas os tipos que contêm "T cell" e seus subtipos "CD4" e "CD8"
Tcells_no = cellmarker_celltypes[grep("T cell", cellmarker_celltypes)] #93 outputs
TCD4_no = cellmarker_celltypes[grep("CD4", cellmarker_celltypes)] #35 outputs
TCD8_no = cellmarker_celltypes[grep("CD8", cellmarker_celltypes)] #34 outputs
Treg_no = cellmarker_celltypes[grep("Treg", cellmarker_celltypes)] #7 outputs
Thelper_no = cellmarker_celltypes[grep("helper", cellmarker_celltypes)] #17 outputs

# Filtrar outros tipos celulares da resposta imune
Neutrophils_no = cellmarker_celltypes[grep("neutrophil", cellmarker_celltypes)] #7 outputs
Monocyte_no = cellmarker_celltypes[grep("monocyte", cellmarker_celltypes)] #14 outputs
DCells_no = cellmarker_celltypes[grep("dendritic", cellmarker_celltypes)] #25 outputs
Macrophages_no = cellmarker_celltypes[grep("macrophage", cellmarker_celltypes)] #7 outputs
NKcells_no = cellmarker_celltypes[grep("killer", cellmarker_celltypes)] # 11 outputs
Neutrophils_no = cellmarker_celltypes[grep("mastcell", cellmarker_celltypes)] #7 outputs

# Criar variáveis para cada célula para enriquecimento

# Filtrar apenas as células da resposta imune adaptativa
Bcells <- filter(cells, str_detect(cell_name, regex("B cell|plasma cell|plasmablast", ignore_case = TRUE)))
TCD4 <- filter(cells, str_detect(cell_name, regex("CD4", ignore_case = TRUE)))
TCD8 <- filter(cells, str_detect(cell_name, regex("CD8", ignore_case = TRUE)))
Treg <- filter(cells, str_detect(cell_name, regex("Treg", ignore_case = TRUE)))
Thelper <- filter(cells, str_detect(cell_name, regex("helper", ignore_case = TRUE)))


# Filtrar células da resposta imune inata
Neutrophils <- filter(cells, str_detect(cell_name, regex("neutrophil", ignore_case = TRUE)))
Monocyte <- filter(cells, str_detect(cell_name, regex("monocyte", ignore_case = TRUE)))
DCells <- filter(cells, str_detect(cell_name, regex("dendritic", ignore_case = TRUE)))
Macrophages <- filter(cells, str_detect(cell_name, regex("macrophage", ignore_case = TRUE)))
Mastcells <- filter(cells, str_detect(cell_name, regex("mast cell", ignore_case = TRUE)))
NKcells <- filter(cells, str_detect(cell_name, regex("killer", ignore_case = TRUE)))
Platelets <- filter(cells, str_detect(cell_name, regex("Platelet", ignore_case = TRUE)))
Eosinophil <- filter(cells, str_detect(cell_name, regex("Eosinophil", ignore_case = TRUE)))
Basophil <- filter(cells, str_detect(cell_name, regex("Basophil", ignore_case = TRUE)))
Granulocyte <- filter(cells, str_detect(cell_name, regex("Granulocyte", ignore_case = TRUE)))

#Agrupar todos os resultados em Immune cells
Immune_adap = rbind(Bcells, TCD4, TCD8, Treg, Thelper)
Immune_adap$Type = "Adaptive"

Immune_innate = rbind(Neutrophils, Monocyte, DCells, Macrophages, Mastcells, NKcells, Platelets, Eosinophil, Basophil, Granulocyte)
Immune_innate$Type = "Innate"

#Unir tabelas
Immune_cells = rbind(Immune_adap, Immune_innate) 

#Salvar
write.csv(Immune_cells, file = "CellMarker_ImmuneCells.csv")

esquisser(Immune_cells_markersbycells)
```

Top 20 cells by #markers

```{r}

Immune_cells_markersbycells = Immune_cells %>% 
  group_by(cell_name, Type) %>% 
  summarise(n_markers = n()) %>% 
  arrange(desc(n_markers))
Immune_cells_markersbycells_top = Immune_cells_markersbycells[1:20, ]

ggImmune_cells_cellsbyimmune = ggplot(Immune_cells_markersbycells_top) +
  aes(x = reorder(cell_name, n_markers), y = n_markers, fill = Type) +
  scale_fill_manual(values = c('Innate' = "#989898",
                               'Adaptive' = "#6DF8DF")) +
  geom_col() +
  labs(
    x = "#Markers",
    y = "Cells",
    title = "Top 20 cells by #markers"
  ) +
  coord_flip() +
  theme_minimal() +
  geom_text(aes(label = n_markers, y = n_markers), 
            position = position_dodge(width = 0.9), 
            vjust = -0.2, 
            hjust = -0.2, 
            size = 3)

ggsave(ggImmune_cells_cellsbyimmune, file= "CellMarker_Top10_cellsbyn_markers.png", width = 10, height = 5)

print(ggImmune_cells_cellsbyimmune)

```

Number of Cells by immune system

```{r}
Immune_cells_cellsbyimmune = Immune_cells %>% 
  group_by(Type, cell_name) %>% 
  summarise(n_cells = n()) %>% 
  group_by(Type) %>% 
  summarise(n_cells = n())

ggImmune_cells_cellsbyimmune = ggplot(Immune_cells_cellsbyimmune) +
  aes(x = reorder(Type, n_cells), y = n_cells, fill = Type) +
  geom_col() +
  scale_fill_manual(values = c('Innate' = "#989898",
                               'Adaptive' = "#6DF8DF")) +
  labs(
    x = "#Cells",
    y = "Immune system",
    title = "#Cells by immune system"
  ) +
  coord_flip() +
  theme_minimal() +
  geom_text(aes(label = n_cells, y = n_cells), 
            position = position_dodge(width = 0.9), 
            vjust = -0.2, 
            hjust = -0.2, 
            size = 3)
ggsave(ggImmune_cells_cellsbyimmune, file= "CellMarker_nCellsbyimmunesystem.png", width = 10, height = 2)

print(ggImmune_cells_cellsbyimmune)
```

### ImmuneGO

```{r}
organism = "org.Mm.eg.db"
install.packages("GO.db")
BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)
library(clusterProfiler)
library(GO.db)

#HGNC symbol to Entrez
x <- org.Mm.egSYMBOL
mapped_genes <- mappedkeys(x)
symbols_entrez <- as.data.frame(x[mapped_genes])

## Entrez to GO
x <- org.Mm.egGO
mapped_genes <- mappedkeys(x)
entrez_geneontology <- as.data.frame(x[mapped_genes])

## GO BP ancestors
go_bp_ancestor = GOBPANCESTOR %>% 
  as.data.frame()
colnames(go_bp_ancestor)[1] <- "go_id"
colnames(go_bp_ancestor)[2] <- "go_ancestor"

## GO ID to terms
go_bp = AnnotationDbi::select(GO.db, columns=c("GOID","TERM"), keys= "BP", keytype= "ONTOLOGY") %>% 
  clean_names() %>% 
  dplyr::select(go_id = goid, term)

## GO child and ancestor
go_bp_ancestor = go_bp_ancestor %>% 
  inner_join(go_bp %>% rename(go_ancestor = go_id), by = "go_ancestor") %>% 
  rename(term_ancestor = term)

#GO ids to terms
go_bp = AnnotationDbi::select(GO.db, columns=c("GOID","TERM"), keys="BP", keytype="ONTOLOGY") %>% 
  clean_names() %>% 
  rename(go_id = goid) %>% 
  dplyr::select(-ontology) %>% 
  right_join(entrez_geneontology, by = "go_id") %>% 
  right_join(symbols_entrez, by = "gene_id") %>% 
  clean_names() %>% 
  left_join(go_bp_ancestor, by = "go_id") %>% 
  distinct()

#GO immune process

mm_immunego_go.bp = go_bp %>% 
  filter(term_ancestor == "immune system process") %>% 
  rename(go_dataset = term_ancestor,
         process = term,
         genes = symbol) %>% 
  mutate(process = toupper(process),
         immune_set = "Mouse_ImmuneGO")
```

Annotation

There are 943 processes retrieved, but only 18 are directly annotated using Human ImmuneGO. Most general immunological processes are in, like innate immune response,immune response, adaptive immune response, and others.
```{r}
#Merge annotations with Human ImmuneGO
## There are 943 processes
mm_immunego_annotation = mm_immunego_go.bp %>% 
  select(process, genes, immune_set) %>% 
  group_by(process) %>% 
  summarize(set_size = n(),
            genes,
            immune_set) %>% 
  summarize(genes = paste0(genes, collapse = ","),
            set_size,
            genes,
            immune_set) %>% 
  distinct() %>% 
  ungroup()  %>% 
  left_join(ImmuneGO_BTM_grouped_genesets_specific %>% 
              mutate(process = str_to_upper(process)) %>% 
              filter(immune_set == "ImmuneGO by Modules") %>% 
              select(process:cells),
                by = "process")
#Save
mm_immunego_annotation %>% 
  saveRDS(here("VaxGO", "mm_immunego_annotation.rds"))
mm_immunego_annotation %>% 
  write_excel_csv(here("VaxGO", "mm_immunego_annotation.csv"))
  
#How many are not in?
## 925
mm_immunego_annotation %>% 
  filter(is.na(immune_system)) 

## 18 are in
mm_immunego_annotation %>% 
  filter(!is.na(immune_system)) %>% 
  arrange(-set_size)
mm_immunego_annotation %>% 
  filter(!is.na(immune_system)) %>% 
  tabyl(immune_system) %>% 
  janitor::adorn_pct_formatting()



```
Annotate by merging genes with the same symbol (using upper case) by specific processes
```{r}
mm_immunego_annotation_genes = mm_immunego_annotation %>% 
  separate_rows(genes, sep = ",") %>% 
  mutate(genes_upper = str_to_upper(genes))

ImmuneGO_BTM_grouped_genesets_specific_genes = ImmuneGO_BTM_grouped_genesets_specific %>% 
              mutate(process = str_to_upper(process)) %>% 
              filter(immune_set == "ImmuneGO by Modules") %>% 
              select(process, genes) %>% 
  separate_rows(genes, sep = ",")

BTM_grouped_genesets_specific_genes = ImmuneGO_BTM_grouped_genesets_specific %>% 
              mutate(process = str_to_upper(process)) %>% 
              filter(immune_set == "BTM by Modules") %>% 
              select(process, genes) %>% 
  separate_rows(genes, sep = ",")

#ImmuneGO human
mm_immunego_annotation_genes_human = mm_immunego_annotation_genes %>% 
  select(genes = genes_upper) %>% 
  distinct() %>% 
  left_join(ImmuneGO_BTM_grouped_genesets_specific_genes,
            by = "genes")

#There are 2,224 genes not annotated in ImmuneGO human
mm_immunego_annotation_genes_human %>% 
  filter(is.na(process))

#BTM modules
mm_btm_annotation_genes_human = mm_immunego_annotation_genes %>% 
  select(genes = genes_upper) %>% 
  distinct() %>% 
  left_join(BTM_grouped_genesets_specific_genes,
            by = "genes")



#There are 2,450 genes not annotated in ImmuneGO human
mm_btm_annotation_genes_human %>% 
  filter(is.na(process))
```

By general processes, there are 1,940 genes that were not annotated in both BTM and ImmuneGO
```{r}
ImmuneGO_BTM_grouped_genesets <- read_csv("VaxGO/ImmuneGO_BTM_grouped_genesets.csv")
ImmuneGO_BTM_grouped_genes = ImmuneGO_BTM_grouped_genesets %>% 
  separate_rows(genes) %>% 
  select(process, genes) %>% distinct()

#ImmuneGO and BTM modules
mm_ImmuneGO_BTM_annotation_genes_human = mm_immunego_annotation_genes %>% 
  select(genes = genes_upper) %>% 
  distinct() %>% 
  left_join(ImmuneGO_BTM_grouped_genes,
            by = "genes")

#There are 1,940 genes not annotated in ImmuneGO+BTM
mm_ImmuneGO_BTM_annotation_genes_human %>% 
  filter(is.na(process))
```

Which are the most gapping processes by genes?
```{r}
#ImmuneGO and BTM modules
mm_ImmuneGO_BTM_annotation_genes_human = mm_immunego_annotation_genes %>% 
  select(genes = genes_upper, 
         process) %>% 
  distinct() %>% 
  left_join(ImmuneGO_BTM_grouped_genes,
            by = "genes")

#There are 1,940 genes not annotated in ImmuneGO+BTM
mm_ImmuneGO_BTM_annotation_genes_human %>% 
  filter(is.na(process.y)) %>% 
  tabyl(process.x) %>% 
  as_data_frame() %>% 
  arrange(-percent)


```







```{r}
mm_immunego_annotated = mm_immunego_go.bp %>% 
  group_by(process) %>%
  summarize(genes = paste0(genes, collapse = ",")) %>%
  inner_join(ImmuneGO_BTM_grouped_genesets_specific %>% dplyr::select(-genes) %>% distinct(), by = "process")

# write.csv(ImmuneGO_Annotated_GO_26_2_24, file = "ImmuneGO_Annotated_GO_26_2_24.csv", row.names = F)
# 
# write.csv(ImmuneGO_Annotated_Genes_26_2_24, file = "ImmuneGO_Annotated_Genes_26_2_24.csv", row.names = F)

```

Group immune system genes by process and cell

```{r}
ImmuneGO_Annotated_unique = ImmuneGO_Annotated_Genes %>% 
  filter(go_term != "Manual")

#Agrupar genes de todos os processos do sistema imune ----

ImmuneGO_Adaptive <- ImmuneGO_Annotated_unique %>%
  filter(immune_system == "Adaptive") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "ADAPTIVE IMMUNE SYSTEM",
  `Gene set, short` = "Adaptive immune system",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "General",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate <- ImmuneGO_Annotated_unique %>%
  filter(immune_system == "Innate") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "INNATE IMMUNE SYSTEM",
  `Gene set, short` = "Innate immune system",
  `Immune system` = "Innate",
  `Immune sub-system` = "General",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Complement <- ImmuneGO_Annotated_unique %>%
  filter(immune_system == "Complement") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "COMPLEMENT",
  `Gene set, short` = "Complement immune system",
  `Immune system` = "Complement",
  `Immune sub-system` = "General",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Inflammation <- ImmuneGO_Annotated_unique %>%
  filter(immune_sub_system == "Inflammation") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "INFLAMMATION",
  `Gene set, short` = "Inflammation",
  `Immune system` = "Innate",
  `Immune sub-system` = "Inflammation",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Interferon <- ImmuneGO_Annotated_unique %>%
  filter(immune_sub_system %in% c("Antiviral", "Interferon")) %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "ANTIVIRAL & IFN",
  `Gene set, short` = "Antiviral and Interferon",
  `Immune system` = "Innate",
  `Immune sub-system` = "Antiviral and Interferon",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_ARPP = ImmuneGO_Annotated_unique %>%
  filter(immune_sub_system %in% c("PRR", "APC")) %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "ARPP",
  `Gene set, short` = "Antigen receptors, processing and presentation",
  `Immune system` = "Innate",
  `Immune sub-system` = "APC",
  `Immune tissue` = "APC",
  `GO.TERM` = "Manual") %>% 
  clean_names()


ImmuneGO_Innate_DC <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Dendritic cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "DENDRITIC CELL",
  `Gene set, short` = "Dendritic cell",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "APC",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_Macrophage <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Macrophage") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "MACROPHAGE",
  `Gene set, short` = "Macrophage",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "APC",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_Mastcell <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Mast cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "MAST CELL",
  `Gene set, short` = "Mast cell",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "Mast cell",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_Neutrophil <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Neutrophil") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "NEUTROPHIL",
  `Gene set, short` = "Neutrophil",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "Neutrophil",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_Eosinophil <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Eosinophil") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "EOSINOPHIL",
  `Gene set, short` = "Eosinophil",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "Eosinophil",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_NK <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "NK cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "NK CELL",
  `Gene set, short` = "Natural Killer Cell",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "NK cell",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Innate_Monocyte <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Monocyte") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "MONOCYTE",
  `Gene set, short` = "Monocyte",
  `Immune system` = "Innate",
  `Immune sub-system` = "Cell",
  `Immune tissue` = "APC",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Adap_Cellular <- ImmuneGO_Annotated_unique %>% 
  filter(immune_sub_system == "Cellular") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "CELLULAR ADAPTIVE IMMUNE SYSTEM",
  `Gene set, short` = "Cellular adaptive immune system",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Cellular",
  `Immune tissue` = "Cellular",
  `GO.TERM` = "Manual") %>% 
  clean_names()


ImmuneGO_Adap_Tcell <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "T cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "T CELL",
  `Gene set, short` = "T cell",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Cellular",
  `Immune tissue` = "Cellular",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Adap_Thelper <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "T helper") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "T HELPER CELLS",
  `Gene set, short` = "T helper cell",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Cellular",
  `Immune tissue` = "Cellular",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Adap_Humoral <- ImmuneGO_Annotated_unique %>%
  filter(immune_sub_system == "Humoral") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "HUMORAL ADAPTIVE IMMUNE SYSTEM",
  `Gene set, short` = "Humoral adaptive immune system",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Humoral",
  `Immune tissue` = "Humoral",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Adap_Bcell <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "B cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "B CELL",
  `Gene set, short` = "B cell",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Humoral",
  `Immune tissue` = "B cell",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_Adap_Ig <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "B cell") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "IMMUNOGLOBULIN MEDIATED RESPONSE",
  `Gene set, short` = "Ig-mediated response",
  `Immune system` = "Adaptive",
  `Immune sub-system` = "Humoral",
  `Immune tissue` = "B cell",
  `GO.TERM` = "Manual") %>% 
  clean_names()

ImmuneGO_other <- ImmuneGO_Annotated_unique %>%
  filter(immune_system == "General" |
         immune_tissue == "Leukocyte") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "GENERAL",
  `Gene set, short` = "General",
  `Immune system` = "General",
  `Immune sub-system` = "General",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()


ImmuneGO_antimicrobial <- ImmuneGO_Annotated_unique %>%
  filter(immune_tissue == "Antimicrobial") %>% 
  summarize(set_size = n(), 
            genes = paste(genes, collapse = ",")) %>% 
  mutate(process = "ANTIMICROBIAL",
  `Gene set, short` = "Antimicrobial immune response",
  `Immune system` = "General",
  `Immune sub-system` = "General",
  `Immune tissue` = "General",
  `GO.TERM` = "Manual") %>% 
  clean_names()


# Unite sets -----

ImmuneGO_Annotated_GO = bind_rows(ImmuneGO_Adaptive, 
                            ImmuneGO_Innate, 
                            ImmuneGO_Complement, 
                            ImmuneGO_Inflammation,
                            ImmuneGO_Interferon,
                            ImmuneGO_ARPP,
                            ImmuneGO_Innate_Macrophage,
                            ImmuneGO_Innate_Mastcell,
                            ImmuneGO_Innate_Neutrophil,
                            ImmuneGO_Innate_Eosinophil,
                            ImmuneGO_Innate_NK,
                            ImmuneGO_Innate_Monocyte,
                            ImmuneGO_Adap_Cellular,
                            ImmuneGO_Adap_Tcell,
                            ImmuneGO_Adap_Thelper,
                            ImmuneGO_Adap_Humoral,
                            ImmuneGO_Adap_Bcell,
                            ImmuneGO_Adap_Ig,
                            ImmuneGO_other,
                            ImmuneGO_antimicrobial,
                            ImmuneGO_Innate_DC,
                            ImmuneGO_Annotated_unique) %>% 
  dplyr::select(process:go_term, set_size:genes) %>% 
  distinct()

ImmuneGO_Annotated_Genes = ImmuneGO_Annotated_GO %>% 
   separate_rows(genes, sep = ",")

# Separate TCR and BCR repertoire genes
TCR_BCR_repertoire = ImmuneGO_Annotated_Genes %>% 
  filter(grepl("^TR", genes) & process == "ADAPTIVE IMMUNE SYSTEM" |
           grepl("^IG", genes) & process == "ADAPTIVE IMMUNE SYSTEM" |
           grepl("^TR", genes) & process == "CELLULAR ADAPTIVE IMMUNE SYSTEM" |
         grepl("^IG", genes) & process == "HUMORAL ADAPTIVE IMMUNE SYSTEM") %>% 
  mutate(process = if_else(process == "CELLULAR ADAPTIVE IMMUNE SYSTEM", "TCR REPERTOIRE", 
                           "BCR REPERTOIRE"),
         gene_set_short = if_else(process == "TCR REPERTOIRE", 
                                  "TCR REPERTOIRE", 
                           "BCR REPERTOIRE"),
         process = if_else(grepl("^TR", genes), "TCR REPERTOIRE", "BCR REPERTOIRE"),
         immune_sub_system = if_else(process == "TCR REPERTOIRE", "Cellular", "Humoral"),
         immune_tissue = if_else(process == "TCR REPERTOIRE", "Cellular", "Humoral"))

immunego_manual = ImmuneGO_Annotated_Genes %>%
  filter(go_term == "Manual") %>% 
  anti_join(TCR_BCR_repertoire %>% dplyr::select(genes) %>% distinct(), by = "genes") %>% 
  bind_rows()

#Verify if all genes are annotated. If there are no left genes, you are good to go
a = ImmuneGO_Annotated_Genes %>% filter(go_term == "Manual") %>% dplyr::select(genes) %>% distinct()
ImmuneGO_Annotated_Genes %>% dplyr::select(process, genes, immune_system) %>% anti_join(a, by = "genes")


#Salvar
write.csv(ImmuneGO_Annotated_GO, file = paste0("ImmuneGO_Annotated_GO_", today(), ".csv"), row.names = F)
saveRDS(ImmuneGO_Annotated_GO, file = paste0("ImmuneGO_Annotated_GO_", today(), ".rds"))
saveRDS(ImmuneGO_Annotated_Genes, file = paste0("ImmuneGO_Annotated_Genes_", today(), ".rds"))

#Display
view(ImmuneGO_Annotated_GO)
view(ImmuneGO_Annotated_Genes)

```

Top 10 gene sets
```{r}
#Top 10 gene sets
ImmuneGO_Annotated_unique_top10 = ImmuneGO_Annotated_GO %>% 
  clean_names() %>% 
  arrange(desc(set_size)) %>% 
  slice_head(n = 20) %>% 
  mutate(immune_system = if_else(gene_set_short == "Adaptive Immune response", "Adaptive", immune_system))

#Gráfico de barras
ImmuneGO_top10_barplot = ggplot(ImmuneGO_Annotated_unique_top10) +
 aes(x = reorder(gene_set_short, set_size), 
     weight = set_size, 
     fill = immune_system) +  # Use reorder para reordenar as barras +
  geom_bar()+
  scale_fill_manual(
    values = c(Adaptive = "#6DF8DF",
    Complement = "#e9edc9",
    General = "#bde0fe",
    Innate = "#989898",
    Undefined = "#edede9")) +
 labs(y = "Set size", title = "ImmuneGO", subtitle = "Top 10 GO, by set size") +
 coord_flip() +
 theme_minimal()

#Salvar
ggsave(ImmuneGO_top10_barplot, file="ImmuneGO_top10_barplot.png", width = 10, height = 6)

#Display
head(ImmuneGO_top10_barplot)
```



### ImmuneGO + BTM

```{r}
btm_annotation_genes = readRDS(here("VaxGO gene sets", "btm_annotation_genes.rds"))
ImmuneGO_Annotated_Genes = readRDS(here("VaxGO gene sets", "ImmuneGO_Annotated_Genes_2024-03-26.rds"))
ImmuneGO_Annotated_GO = read_csv(here("VaxGO gene sets", "ImmuneGO_Annotated_GO_2024_03_26.csv"))

#Unite genes
immunego_btm_genes = ImmuneGO_Annotated_Genes %>% 
              dplyr::select(genes) %>% 
              distinct() %>% 
              mutate(set = "ImmuneGO") %>% 
  bind_rows(btm_annotation_genes %>% 
              dplyr::select(genes) %>% 
              distinct() %>% 
              mutate(set = "BTM")) %>% 
  arrange(genes)

immunego_btm_genes_joined = immunego_btm_genes %>% 
  bind_rows(immunego_btm_genes %>% mutate(set = "BTM and ImmuneGO")) %>% 
  distinct()


#Unite subsets
btm_annotation_grouped = btm_annotation_genes %>% 
  group_by(complete_name) %>% 
  rename(process = complete_name) %>% 
  mutate(genes = paste0(genes, collapse = ",")) %>% 
  distinct() %>% 
  mutate(genes = paste0(genes, collapse = ","),
            immune_set = "BTM by Modules") %>% 
    dplyr::select(process,
           immune_system:cells,
           set_size = module_size,
           genes,
           immune_set)

ImmuneGO_annotation_modules = ImmuneGO_Annotated_GO %>% 
  filter(go_term != "Manual") %>% 
    dplyr::select(process = gene_set_short, 
           immune_system:immune_tissue,
           set_size, genes) %>% 
    rename(cells = immune_tissue,
           immune_subsystem = immune_sub_system) %>% 
    mutate(immune_set = "ImmuneGO by Modules")

immunego_btm_specific_grouped = btm_annotation_grouped %>% 
  bind_rows(ImmuneGO_annotation_modules)
  
immunego_btm_specific_grouped %>% 
  write_csv(file = here("VaxGO gene sets", "ImmuneGO_BTM_grouped_genesets_specific.csv"))

  
#By immune system
btm_immunesystem = btm_annotation_genes %>% 
  dplyr::select(immune_system, genes) %>% 
  distinct() %>% 
  rename(process = immune_system) %>% 
  group_by(process) %>% 
  summarize(genes = paste0(genes, collapse = ","),
            set_size = n(),
            process = str_to_upper(process),
            # id = process,
            immune_system = process,
            immune_subsystem = process,
            cells = process,
            complete_name = process,
            annotation_level = "Manual",
            immune_set = "BTM by Immune System") %>% 
  distinct()

#By immune subsystem
btm_immunesubsystem = btm_annotation_genes %>% 
  dplyr::select(immune_subsystem, genes) %>% 
  distinct() %>% 
  rename(process = immune_subsystem) %>% 
  group_by(process) %>% 
  summarize(genes = paste0(genes, collapse = ","),
            process = str_to_upper(process),
            set_size = n(),
            # id = process,
            immune_system = process,
            immune_subsystem = process,
            cells = process,
            complete_name = process,
            annotation_level = "Manual",
            immune_set = "BTM by Immune Sub-System") %>% 
  distinct() 
#By cell
btm_cells = btm_annotation_genes %>% 
  dplyr::select(cells, genes) %>%
  distinct() %>% 
  rename(process = cells) %>% 
  group_by(process) %>% 
  summarize(genes = paste0(genes, collapse = ","),
            set_size = n(),
            process = str_to_upper(process),
            # id = process,
            immune_system = process,
            immune_subsystem = process,
            cells = process,
            annotation_level = "Manual",
            complete_name = process,
            immune_set = "BTM by Immune Cell") %>% 
  distinct() 
#Merge
btm_manual = bind_rows(btm_immunesystem,
          btm_immunesubsystem,
          btm_cells) %>% 
  arrange(complete_name) %>% 
  dplyr::select(complete_name, everything()) %>% 
  distinct()

btm_manual_grouped = bind_rows(btm_manual,
                               btm_annotation_grouped) %>% 
  distinct()


#Merge ImmuneGO and BTM main processes
immunego_btm_manual_grouped = btm_manual %>% 
  dplyr::select(process, genes, immune_set, set_size) %>% 
  mutate(immune_set = "BTM") %>% 
  bind_rows(ImmuneGO_Annotated_GO %>% 
              filter(go_term == "Manual") %>% 
              dplyr::select(process, genes, set_size) %>% 
              mutate(immune_set = "ImmuneGO")) %>% 
  arrange(process) %>% 
  mutate(process = case_when(process == "ADAPTIVE" ~ "ADAPTIVE IMMUNE SYSTEM",
                              process == "ALL" ~ "IMMUNE SYSTEM",
                              process == "GENERAL" ~ "LEUKOCYTES",
                              process == "CELLS" ~ "LEUKOCYTES",
                              process == "DCS" ~ "DENDRITIC CELL",
                              process == "APP" ~ "ARPP",
                              process == "CELLULAR" ~ "CELLULAR ADAPTIVE IMMUNE SYSTEM",
                              process == "DCS, MYELOID" ~ "DENDRITIC CELL",
                              process == "GENERIC" ~ "LEUKOCYTES",
                              process == "HUMORAL" ~ "HUMORAL ADAPTIVE IMMUNE SYSTEM",
                              process == "INNATE" ~ "INNATE IMMUNE SYSTEM",
                              process == "INTERFERON" ~ "ANTIVIRAL & IFN",
                              process == "NK" ~ "NK CELL",
                              process == "B CELL, PLASMA CELL" ~ "B CELL",
                              process == "PLASMA CELL" ~ "B CELL",
                              process == "PRR" ~ "ARPP",
                              process == "SIGNALING" ~ "SIGNALING",
                              process == "TH2" ~ "T CELL",
                              process == "TH1" ~ "T CELL",
                              process == "T HELPER CELLS" ~ "T CELL",
                              TRUE ~ process
                              )) %>% 
  distinct() %>% 
  bind_rows(
    immunego_btm_specific_grouped)

immunego_btm_manual_grouped_genes = immunego_btm_manual_grouped %>%
  separate_rows(genes, sep = ",") %>% 
  distinct()

immunego_btm_manual_grouped_genes = immunego_btm_manual_grouped_genes %>% 
  bind_rows(
    immunego_btm_manual_grouped_genes %>% 
      mutate(immune_set = "BTM and ImmuneGO") %>% 
      distinct()
  )
  
  
immunego_btm_manual_grouped %>% 
  write_csv(file = here("VaxGO gene sets", "ImmuneGO_BTM_grouped_genesets.csv"))

immunego_btm_manual_grouped_genes %>% 
  write_csv(file = here("VaxGO gene sets", "ImmuneGO_BTM_grouped_genes.csv"))
  
```
Plot relative frequency of genes by main process
```{r fig.height=5, fig.width=7}

#Get exclusive genes
immunego_only_genes = immunego_btm_manual_grouped_genes %>% 
  filter(immune_set == "ImmuneGO") %>% 
  dplyr::select(-immune_set, -set_size) %>% 
  anti_join(immunego_btm_manual_grouped_genes %>% 
            filter(immune_set == "BTM"),
            by = "genes") %>% 
  mutate(immune_set = "ImmuneGO only") %>% 
  distinct()

btm_only_genes = immunego_btm_manual_grouped_genes %>% 
  filter(immune_set == "BTM") %>% 
  dplyr::select(-immune_set, -set_size) %>% 
  anti_join(immunego_btm_manual_grouped_genes %>% 
            filter(immune_set == "ImmuneGO"),
            by = "genes") %>% 
  mutate(immune_set = "BTM only") %>% 
  distinct()

btm_immune_joined_stats = immunego_only_genes %>% 
  bind_rows(btm_only_genes) %>% 
  distinct() %>% 
  mutate(immune_set = "BTM and ImmuneGO") %>% 
  dplyr::select(process, genes, immune_set) %>% 
  distinct() %>% 
  group_by(process) %>% 
  distinct() %>% 
  mutate(ntotal=n()) %>% 
  ungroup() %>% 
  dplyr::select(process, ntotal) %>% 
  distinct() %>% 
  inner_join(immunego_only_genes %>% 
               bind_rows(btm_only_genes), by = "process") %>% 
  group_by(process, immune_set) %>% 
  mutate(n = n(),
         genes = paste0(genes, collapse = ",")) %>% 
  dplyr::select(-genes) %>% 
  distinct() %>% 
  ungroup() %>% 
  mutate(n_perc = (n/ntotal) * 100)

btm_immune_joined_stats_plot = btm_immune_joined_stats %>% 
  filter(!process %in% c("LYMPHOCYTE", 
                         "IMMUNOGLOBULIN MEDIATED RESPONSE",
                         "MYELOID CELL", 
                         "SIGNALING", 
                         "IMMUNE SYSTEM")) %>%
  mutate(process = str_to_title(process),
         process = fct_relevel(process, 
                               "Leukocytes",
                               "Adaptive Immune System",
                               "Innate Immune System",
                               "Complement"	,
                               "Cellular Adaptive Immune System",
                               "Humoral Adaptive Immune System",
                               )) %>% 
 ggplot() +
  aes(x = n_perc, y = fct_rev(process), fill = immune_set) +
  geom_bar(stat = "summary", fun = "sum", position = "fill") +
  scale_fill_manual(values = c("BTM only" = "#5e60ce",
                               "ImmuneGO only" = "#00A087FF")) +
  theme_minimal() +
  theme(legend.position = "right",
        strip.text.x = element_text(color = "black", size = 12), 
        axis.text.x = element_text(color = "black", size = 10),
        axis.text.y = element_text(vjust = 0.5, size = 10, color = "black"),
        axis.line = element_line(size = 0.5, colour = "black", linetype=1),
        axis.ticks = element_line(size = 0.5, color = "black")) +
  labs(x = "%Coverage",
       y = "",
       fill = "Immune set",
       subtitle = "Merged data",
       title = "ImmuneGO and BTM main processes genes") +
  scale_x_continuous(expand = c(0,0)) +
  geom_text(aes(label = paste0(round(n_perc, 0))), 
            position = position_fill(vjust = 0.5), 
            size = 3, 
            color = "white",
            fontface = "bold") 

btm_immune_joined_stats_plot %>% 
  ggsave(filename = here("Figures", "ImmuneGO_BTM_Joined_Stats_barplot.png"),
         width = 10,
         height = 7)

btm_immune_joined_stats_plot
```



#Benchmarking models

## Found in translation
Ref: Normand, R., Du, W., Briller, M. et al. Found In Translation: a machine learning model for mouse-to-human inference. Nat Methods 15, 1067–1073 (2018). https://doi.org/10.1038/s41592-018-0214-9
Github: https://github.com/shenorrLabTRDF/FIT.mouse2man 
```{r}
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
#Import files
pca_rotations = FIT.mouse2man::pca_rotations
best_models = FIT.mouse2man::best_models
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez
MM_Entrez_symbol_desc = FIT.mouse2man::MM_Entrez_symbol_desc
Psoriasis_classifier_neg = FIT.mouse2man::Psoriasis_classifier_neg
slopes_per_gene_V2.0 = FIT.mouse2man::slopes_per_gene_V2.0
StaphInfection_classifier_pos = FIT.mouse2man::StaphInfection_classifier_pos
AllData_V2.0 = FIT.mouse2man::AllData_V2.0 #Training data

#Samples
microarray_sample = FIT.mouse2man::microarray_sample #Sample Microarray
microarray_sample %>% 
  rownames_to_column("MM_Entrez_symbol_desc") %>% 
  mutate(MM_Entrez_symbol_desc = as.character(MM_Entrez_symbol_desc)) %>% 
  write_csv(here("FIT_Mouse2man", "microarray_sample.csv"))
RNAseq_sample = FIT.mouse2man::RNAseq_sample #Sample RNAseq
RNAseq_sample %>% 
  write_csv(here("FIT_Mouse2man", "rnaseq_sample.csv"))
```


```{r}
AllData_V2.0 %>% 
  clean_names() %>%
  select(disease:csp_id) %>% 
  distinct()
```


```{r}
#Checking applicability of the model to your data
test = FIT.mouse2man::RunClassifier(MouseData = RNAseq_sample, DataType = "rnaseq")

#Fit
fit_example = FIT.mouse2man::FIT(MouseFile = here("FIT_Mouse2man", "rnaseq_sample.csv"), 
                                 DataType = "rnaseq") # Runs the FIT pipeline which computes prediction of the human effect-size per gene  

```


