---
title: "RunYourAnalysis_MouseToHuman_v2"
output: html_document
---

# Setup

```{r message=FALSE, warning=FALSE}
library(here)
source(here("scripts_notebooks", "required.R"))
library(tidymodels)
```

# Load the trained models

The models were trained on mouse-to-human transfer across infection and injury
conditions. Here we apply them to a new (or the same) mouse experiment.

```{r}
# Best model per task (selected in the modeling script):
#   Shared    = Random Forest
#   Rank      = Random Forest
#   Direction = Random Forest
model_shared    <- readRDS(here("Modelling", "Models", "rf_model_shared.rds"))
model_rank      <- readRDS(here("Modelling", "Models", "rf_model_rank.rds"))
model_direction <- readRDS(here("Modelling", "Models", "rf_model_direction.rds"))
```

# Prepare features table

Please ensure that your input is a table containing the following columns:
pathogen, treatment, timepoint, hgnc_symbol, mean_l2fc, ci_lower, ci_upper,
ave_expr, t, p_value, adj_p_val, b, se, sd, inverse_se, organism, and condition.

```{r}
# Gene expression data (mouse experiment; example: S. aureus, timepoint 12)
dge_input <- readRDS(here("Run your analysis here", "example_dge_result.rds"))

# Protein identity
alignments_human2mouse_all <- readRDS(file = here("tables", "alignments_human2mouse_all.rds"))

# Distance
human_mouse_cds_distance <- readRDS(file = here("tables", "human_mouse_cds_distance.rds"))

# Transcription factors
gene_tf_shared_unique <- readRDS(file = here("tables", "gene_tf_shared_unique.rds"))

# CREs
cres_type_homology_comparison_wide <- readRDS(file = here("tables", "cres_type_homology_comparison_wide.rds"))

# BTMs annotation
btm_annotation_genes <- read_csv(here("tables", "btm_annotation_genes.csv")) %>%
  select(hgnc_symbol = symbol, process, group, subgroup) %>%
  distinct()
```

# Build the feature table

```{r}
# Non-BTM genes (kept as a category)
btm_annotation_genes_others <- bind_rows(
  btm_annotation_genes,
  dge_input %>%
    select(hgnc_symbol) %>%
    anti_join(btm_annotation_genes %>%
                select(hgnc_symbol) %>%
                distinct(),
              by = "hgnc_symbol") %>%
    mutate(process = "Not-BTM",
           group = "Not-BTM",
           subgroup = "Not-BTM")
)

# BTM annotation in wide format (one column per module group)
btm_annotation_genes_wide <- btm_annotation_genes_others %>%
  select(hgnc_symbol, group) %>%
  group_by(hgnc_symbol, group) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = group,
              values_from = n) %>%
  column_to_rownames("hgnc_symbol") %>%
  as.matrix() %>%
  replace(is.na(.), 0) %>%
  as.data.frame() %>%
  rownames_to_column("hgnc_symbol")

# Merge all biological feature layers (wide format)
human_mouse_gene_annotation_wide <- human_mouse_cds_distance %>%
  select(hgnc_symbol, dist_k80) %>%
  inner_join(alignments_human2mouse_all,
             by = join_by("hgnc_symbol" == "gene")) %>%
  inner_join(btm_annotation_genes_wide,
             by = "hgnc_symbol") %>%
  inner_join(gene_tf_shared_unique,
             by = join_by("hgnc_symbol" == "gene")) %>%
  inner_join(cres_type_homology_comparison_wide %>%
               select(hgnc_symbol, 'n_total_cres_gene':'pct_match_ctcf_PLS_CTCF-bound'),
             by = join_by(hgnc_symbol))
```

# GSEA (leading-edge genes)

Rank genes with a composite metric (log2FC x -log(adj_p_val)) and run GSEA
against the BTM gene sets.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping for GSEA
term2gene <- btm_annotation_genes %>%
  select(process, genes = hgnc_symbol)

# Build ranked gene list with composite metric
df_genes2GSEA <- dge_input %>%
  distinct() %>%
  select(condition,
         genes = hgnc_symbol,
         log2fold_change = mean_l2fc,
         adj_p_val) %>%
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
  mutate(log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>%
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA across all conditions
gsea_output <- autoGSEA(df = df_genes2GSEA,
                        TERM2GENE = term2gene,
                        geneset_name = "BTMs")

# Clean GSEA output and merge with experiment metadata
gsea_results <- gsea_output$gsea %>%
  as.data.frame() %>%
  clean_names() %>%
  rename(process = id) %>%
  filter(!is.na(process)) %>%
  rownames_to_column("del") %>%
  select(-del, -description) %>%
  rename(p_adjust_gsea = p_adjust,
         q_value_gsea = qvalue,
         rank_gsea = rank) %>%
  inner_join(dge_input %>%
               select(pathogen, treatment, timepoint, condition, organism) %>%
               distinct(),
             by = "condition")

gsea_results
```

```{r}
# Get leading-edge genes (LEGs)
legs <- gsea_results %>%
  separate_rows(core_enrichment, sep = "/") %>%
  mutate(hgnc_symbol = core_enrichment,
         is_leg = "LEG") %>%
  inner_join(dge_input,
             by = join_by(organism, pathogen, treatment, timepoint, condition, hgnc_symbol))

legs %>%
  dplyr::count(process, is_leg)
```

# Compute the model features

The models expect mouse-derived features. Here we rename the columns and compute
the absolute rank, the BTM-only rank and the direction sign.

```{r}
model_input <- dge_input %>%
  # Rename to the names expected by the models
  rename(mean_log2fc_mouse = mean_l2fc,
         inverse_se_mouse = inverse_se,
         p_adj_mouse = adj_p_val,
         timepoint_comparison = timepoint) %>%
  # Direction of the mouse change
  mutate(sign_mouse = sign(mean_log2fc_mouse)) %>%
  # Absolute rank: percentile of |log2FC| x -log10(adjusted P), per condition
  group_by(treatment, pathogen, timepoint_comparison) %>%
  mutate(rank_mouse = percent_rank(abs(mean_log2fc_mouse) * -log10(pmax(p_adj_mouse, 1e-300))) * 100) %>%
  ungroup() %>%
  # Annotate the biological layers
  inner_join(human_mouse_gene_annotation_wide,
             by = "hgnc_symbol")

# BTM-only absolute rank (rank among genes that belong to a BTM)
btm_gene_set <- btm_annotation_genes %>%
  distinct(hgnc_symbol) %>%
  pull(hgnc_symbol)

model_input <- model_input %>%
  filter(hgnc_symbol %in% btm_gene_set) %>%
  group_by(treatment, pathogen, timepoint_comparison) %>%
  mutate(rank_mouse_btm = percent_rank(abs(mean_log2fc_mouse) * -log10(pmax(p_adj_mouse, 1e-300))) * 100) %>%
  ungroup()

model_input
```

# Predict human rank, direction and shared status

```{r}
# Predictions from the three models (probability for classification, value for regression)
predictions <- model_input %>%
  bind_cols(
    predict(model_rank, new_data = model_input, type = "numeric") %>% rename(score_rank = .pred)
  ) %>%
  bind_cols(
    predict(model_direction, new_data = model_input, type = "prob") %>% select(score_direction = .pred_TRUE)
  ) %>%
  bind_cols(
    predict(model_shared, new_data = model_input, type = "prob") %>% select(score_shared = .pred_TRUE)
  ) %>%
  mutate(score_translational = score_rank * score_direction)

predictions %>%
  select(hgnc_symbol, rank_mouse, rank_mouse_btm, sign_mouse,
         score_shared, score_rank, score_direction, score_translational) %>%
  arrange(desc(score_translational))
```

# Visualise the scores

```{r}
# Spearman rho + BH-adjusted p annotation
cor_label <- function(df, x, y) {
  tibble(rho = cor(df[[x]], df[[y]], method = "spearman", use = "complete.obs"),
         p = cor.test(df[[x]], df[[y]], method = "spearman", exact = FALSE)$p.value) %>%
    mutate(padj = p.adjust(p, method = "BH"),
           label = sprintf("rho=%.2f\npadj=%.1e", rho, padj))
}
```

```{r fig.height=4, fig.width=6}
# score_shared: probability of being a shared (mouse + human) leading-edge gene
label_shared <- cor_label(predictions, "score_shared", "rank_mouse")

predictions %>%
  ggplot() +
  aes(x = score_shared, y = rank_mouse,
      fill = score_shared) +
  geom_point(shape = 21, size = 2, alpha = 0.5, stroke = 0) +
  geom_text(data = label_shared,
            aes(x = Inf, y = Inf, label = label),
            hjust = 1.1,
            vjust = 1.2,
            size = 3,
            inherit.aes = FALSE) +
  theme_vaxgo() +
  scale_fill_gradient2(low = "white", high = "blue") +
  labs(x = "score_shared (probability)",
       y = "Absolute rank (mouse)",
       title = "Shared score")
```

```{r fig.height=4, fig.width=6}
# score_rank: predicted human absolute rank
label_rank <- cor_label(predictions, "score_rank", "rank_mouse")

predictions %>%
  ggplot() +
  aes(x = score_rank, y = rank_mouse) +
  geom_point(shape = 21, size = 2, fill = "#4361ee", alpha = 0.5, stroke = 0) +
  geom_text(data = label_rank,
            aes(x = Inf, y = Inf, label = label),
            hjust = 1.1,
            vjust = 1.2,
            size = 3,
            inherit.aes = FALSE) +
  theme_vaxgo() +
  labs(x = "score_rank (predicted human rank)",
       y = "Absolute rank (mouse)",
       title = "Rank score")
```

```{r fig.height=4, fig.width=6}
# score_direction: probability of concordant direction (same sign in both species)
label_direction <- cor_label(predictions, "score_direction", "rank_mouse")

predictions %>%
  ggplot() +
  aes(x = score_direction, y = rank_mouse) +
  geom_point(shape = 21, size = 2, fill = "gray40", alpha = 0.5, stroke = 0) +
  geom_text(data = label_direction,
            aes(x = Inf, y = Inf, label = label),
            hjust = 1.1,
            vjust = 1.2,
            size = 3,
            inherit.aes = FALSE) +
  theme_vaxgo() +
  labs(x = "score_direction (probability of concordance)",
       y = "Absolute rank (mouse)",
       title = "Direction score")
```

```{r fig.height=4, fig.width=6}
# score_translational: combined score (high rank AND concordant direction)
label_translational <- cor_label(predictions, "score_translational", "rank_mouse")

predictions %>%
  ggplot() +
  aes(x = score_translational, y = rank_mouse) +
  geom_point(shape = 21, size = 2, fill = "#e5383b", alpha = 0.5, stroke = 0) +
  geom_text(data = label_translational,
            aes(x = Inf, y = Inf, label = label),
            hjust = 1.1,
            vjust = 1.2,
            size = 3,
            inherit.aes = FALSE) +
  theme_vaxgo() +
  labs(x = "score_translational",
       y = "Absolute rank (mouse)",
       title = "Translational score")
```

# Select priority genes with a user-defined cutoff

```{r}
# Example: keep genes above a cutoff you choose
score_cutoff <- 0.8

priority_genes <- predictions %>%
  filter(score_translational >= score_cutoff) %>%
  arrange(desc(score_translational)) %>%
  select(hgnc_symbol, score_shared, score_rank, score_direction, score_translational)

priority_genes
```
