---
title: "RunYourAnalysis_MouseToHuman"
output: html_document
---


```{r message=FALSE, warning=FALSE}
install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```

# Prepare features table
Please ensure that your input is a table containing the following columns: pathogen, treatment, timepoint, hgnc_symbol, mean_l2fc, ci_lower, ci_upper, ave_expr, t, p_value, adj_p_val, b, se, sd, inverse_se, organism, and condition, where the condition column represents the combination of organism_pathogen_timepoint.
```{r}
# Gene expression data
# A table containing pathogen, treatment, timepoint, hgnc_symbol, mean_l2fc, ci_lower, ci_upper, ave_expr, t, p_value, adj_p_val, b, se, sd, inverse_se, organism, and condition
dge_input =  readRDS(here("Run your analysis here", 
                          "example_dge_result.rds")) %>% 
  mutate(inverse_se = 1/se)

#Protein identity
alignments_human2mouse_all = readRDS(file = here("tables", "alignments_human2mouse_all.rds"))
#Distance
human_mouse_cds_distance = readRDS(file = here::here("tables", "human_mouse_cds_distance.rds")) 
#Transcription factors
gene_tf_shared_unique = readRDS(file = here("tables", "gene_tf_shared_unique.rds"))
# CREs
cres_type_homology_comparison_wide = 
  readRDS(file = here("tables", "cres_type_homology_comparison_wide.rds"))
#BTMs annotation
btm_annotation_genes <- read_csv(here("tables", "btm_annotation_genes.csv")) %>% 
  select(hgnc_symbol = symbol, process, group, subgroup) %>% 
  distinct()

btm_annotation_genes_others = bind_rows(
  btm_annotation_genes,
  #Add non-BTM genes
  dge_btm_process_genes_input %>% 
    select(hgnc_symbol) %>% 
    anti_join(btm_annotation_genes %>% 
                select(hgnc_symbol) %>% 
                distinct(),
              by = "hgnc_symbol") %>% 
    mutate(process = "Not-BTM",
           group = "Not-BTM",
           subgroup = "Not-BTM")
  )
btm_annotation_genes_wide = btm_annotation_genes_others %>% 
  select(hgnc_symbol, group) %>% 
  group_by(hgnc_symbol, group) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  pivot_wider(names_from = group,
              values_from = n) %>% 
  column_to_rownames("hgnc_symbol") %>% 
  as.matrix() %>% 
  replace(is.na(.), 0) %>% 
  as.data.frame() %>% 
  rownames_to_column("hgnc_symbol")

#Merge features ===========
#Wide format
human_mouse_gene_annotation_wide = 
  #Annotate with CDS distance (Kimura)
  human_mouse_cds_distance %>% select(hgnc_symbol, dist_k80) %>% 
  #Annotate with AA identity %
  inner_join(alignments_human2mouse_all,
             by = join_by("hgnc_symbol" == "gene")) %>% 
  #Annotate with BTMs 
  inner_join(btm_annotation_genes_wide,
             by = "hgnc_symbol") %>% 
  #Annotate with TFs
  inner_join(gene_tf_shared_unique,
             by = join_by("hgnc_symbol" == "gene")) %>% 
  #Annotate with CRES
  inner_join(cres_type_homology_comparison_wide %>% 
               select(hgnc_symbol, 'n_total_cres_gene':'pct_match_ctcf_PLS_CTCF-bound'),
             by = join_by(hgnc_symbol)) 

#Merge features to a long format
human_mouse_gene_annotation_long = 
  #Annotate with CDS distance (Kimura)
  human_mouse_cds_distance %>% select(hgnc_symbol, dist_k80) %>% 
  pivot_longer(cols = dist_k80, names_to = "variable", values_to = "value") %>% 
    mutate(layer = "Kimura distance") %>% 
  #Annotate with AA identity %
  bind_rows(alignments_human2mouse_all %>% 
              rename(hgnc_symbol = gene) %>% 
               pivot_longer(cols = identity_human2mouse, names_to = "variable", values_to = "value")) %>% 
    mutate(layer = "Protein sequence identity") %>% 

  #Annotate with TFs
  bind_rows(gene_tf_shared_unique %>% 
               rename(hgnc_symbol = gene) %>% 
               pivot_longer(cols = pct_tf_unique:n_tf_total, names_to = "variable", values_to = "value") %>% 
              mutate(layer = "Transcription factor")) %>% 
  #Annotate with CRES
  bind_rows(cres_type_homology_comparison_wide %>% 
              mutate(layer = "CREs") %>% 
              pivot_longer(cols = c('n_total_cres_gene':'pct_match_ctcf_PLS_CTCF-bound'), 
                           names_to = "variable",
                           values_to = "value") %>% 
              select(hgnc_symbol, layer, variable, value)
              ) %>% 
  #Annotate with mouse symbol and ensembl ID
  left_join(human_mouse_cds_distance %>% 
              select(hgnc_symbol, mgi_symbol, mouse_ensembl, human_ensembl),
            by = "hgnc_symbol")
  

#Annotate with gene expression ==========
#Wide format
dge_stats_annotated_layers_wide = dge_input %>%   
            inner_join(human_mouse_gene_annotation_wide,
             by = join_by(hgnc_symbol))
 
#Long format
dge_stats_annotated_layers = dge_input %>%   
            inner_join(human_mouse_gene_annotation_long,
             by = join_by(hgnc_symbol)) 


#Save
dge_stats_annotated_layers_wide %>%
  saveRDS(file = here("Run your analysis here", "dge_stats_annotated_layers_wide.rds"))

dge_stats_annotated_layers %>%
  saveRDS(file = here("Run your analysis here", "dge_stats_annotated_layers.rds"))

#Print
dge_stats_annotated_layers
```


# GSEA

Rank genes with a composite metric (log2FC x -log(adj_p_val)) and run
GSEA against the BTM gene sets across all conditions simultaneously.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping for GSEA --------
term2gene <- btm_annotation_genes %>%
  select(process, genes = hgnc_symbol)

# Build ranked gene list with composite metric --------
# Metric: log2FC * -log(adj_p_val) — balances effect size and significance
df_genes2GSEA <- dge_input %>%
  distinct() %>%
  #Rename
  select(condition,
         genes = hgnc_symbol, log2fold_change = mean_l2fc, adj_p_val) %>%
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
  mutate(log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>%
  select(condition, genes, 
         #Rank metric
         rank = log2fold_change_logp)

# Run GSEA across all conditions using autoGSEA (iterates over conditions internally) --------
gsea_output <- autoGSEA(df = df_genes2GSEA,
                         TERM2GENE = term2gene,
                         geneset_name = "BTMs")

# Clean GSEA output and merge with experiment metadata --------
gsea_results <- gsea_output$gsea %>%
  as.data.frame() %>%
  clean_names() %>%
  rename(process = id) %>%
  filter(!is.na(process)) %>%
  rownames_to_column("del") %>% 
  select(-del, -description) %>%
    rename(p_adjust_gsea = p_adjust,
         q_value_gsea = qvalue,
         rank_gsea = rank) %>% 
  #Merge with experiment metadata 
  inner_join(dge_input %>% 
               select(pathogen, treatment, timepoint, condition, organism) %>% 
               distinct(),
             by = "condition") 



#Check
gsea_results


#Save
saveRDS(gsea_results, here("Run your analysis here", "example_gsea_btm_results.rds"))
```



```{r}
#Filter padjusted from GSEA
filter_padjust_gsea = 0.25

#Get LEGs and merge with DGE stats
legs = gsea_results %>%
  separate_rows(core_enrichment, sep = "/") %>%
  mutate(hgnc_symbol = core_enrichment,
         is_leg = "LEG") %>% 
  inner_join(dge_input,
               by = join_by(organism, pathogen, treatment, timepoint, condition, hgnc_symbol))

legs %>% 
  dplyr::count(process, is_leg)

#Complete with not LEGs and group by process
not_legs =  btm_annotation_genes %>% 
  anti_join(dge_gsea_results, 
            by = join_by(hgnc_symbol, process))  %>% 
  #Get values for not-LEGs
  #GSEA results
  inner_join(gsea_results %>%
    select(process:organism, -core_enrichment) %>%
    distinct(),
    by = "process") %>% 
   #Merge with experiment metadata 
  full_join(dge_input,
               by = join_by(organism, pathogen, treatment, timepoint, condition, hgnc_symbol)) %>% 
  mutate(is_leg = "Not LEG",
         process = if_else(is.na(process), "Not-BTM", process),
         group = if_else(is.na(group), "Not-BTM", group),
         subgroup = if_else(is.na(subgroup), "Not-BTM", subgroup))

#Merge
legs_notlegs = bind_rows(legs, not_legs)


legs_notlegs %>% 
  dplyr::count(process, is_leg)

legs_notlegs %>% 
  saveRDS(here("Run your analysis here", "example_dge_gsea_results_legs.rds"))

```



Visualize
```{r}
#Annotate with biomol layers
min_max <- function(x) {
  if (all(is.na(x))) return(x)
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

#Add layers
legs_notlegs_layers = legs_notlegs %>% 
  inner_join(human_mouse_gene_annotation_wide,
             by = join_by(hgnc_symbol))  %>% 
  select(process:subgroup, dist_k80, identity_human2mouse, pct_tf_shared, n_tf_total, n_type_PLS, n_type_dELS, n_type_pELS, `pct_match_ctcf_PLS_CTCF-bound`) 

#Calculate score
legs_notlegs_layers_score = legs_notlegs_layers %>% 
  mutate(
    # --- Tier 1: Expression Metrics (Highest Importance: Weight = 3.0) ---
    w_dge = 3.0 * (min_max(log1p(inverse_se)) + min_max(abs(mean_l2fc))),
    
    # --- Tier 2: Sequence & TF Conservation (High Importance: Weight = 2.0) ---
    w_seq_tf = 2.0 * (
      min_max(identity_human2mouse) + 
      min_max(pct_tf_shared) + 
      min_max(log1p(n_tf_total))
    ),
    
    # --- Tier 3: Cis-Regulatory Elements & CTCF (Moderate Importance: Weight = 1.0) ---
    w_cre = 1.0 * (
      min_max(n_type_PLS) + 
      min_max(n_type_dELS) + 
      min_max(n_type_pELS) + 
      min_max(`pct_match_ctcf_PLS_CTCF-bound`)
    ),
    
    # --- Penalty: Kimura Distance (Weight = -2.5) ---
    w_penalty = -2.5 * min_max(dist_k80),
    
    # Combined raw score
    score_raw = w_dge + w_seq_tf + w_cre + w_penalty,
    
    # Rescale final score from 0 to 100
    score = min_max(score_raw) * 100
  ) %>% 
  select(-starts_with("w_"), -score_raw)

# Select process to get LEGs
select_process = "immune activation - generic cluster"

legs_notlegs_layers %>% 
  filter(process == select_process) %>%
  group_by(organism, treatment, pathogen, process) %>% 
  mutate(rank_log2fc = rank(mean_l2fc * -log10(adj_p_val))) %>% 
  ggplot() +
  aes(x = score,
      y = rank_log2fc,
      fill = is_leg) +
  scale_y_reverse() +
  geom_point(aes(size = score),
             pch = 21) +
  theme_vaxgo() +
  labs(x = "Score",
       y = "Log2FC, ranks (Mouse)",
       title = "Log2FC ranks vs. Score")


```






