---
title: "DataCuration"
output: html_document
---

#Required

```{r}
# install.packages("here") if needed
install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```

# 1. Data curation 

## BioProject

```{r}
#Data curation --------
## BioProject

#Import dataset
animals_vaccines_bioproject_result <- read_csv(here("tables", "animals_vaccines_bioproject_result.csv"), 
    col_names = FALSE)

#Clean table
datacuration_clean = animals_vaccines_bioproject_result %>%
 drop_na() %>% 
  mutate(study = if_else(X1 =="Title", X2, NA)) %>% 
  fill(study, .direction = c("down")) %>% 
  pivot_wider(names_from = X1, values_from = X2) %>% 
  select(-study) %>% 
  clean_names() %>% 
  mutate(title = str_remove(title, "^\\d+\\.\\s"),
         organism = str_remove(organism, "Organism: "),
         organism = str_remove(organism, "Taxonomy ID "),
         bio_project_accession = str_remove(bio_project_accession, "BioProject Accession: "),
         id = str_remove(id, "ID: "))

datacuration_clean

#Save
datacuration_clean %>% 
  write_csv(here("tables", "datacuration_step2.csv"))
```

```{r fig.height=4, fig.width=3}

#Import
datacuration_clean = read_csv(here("tables", "datacuration_step2.csv"))

#Visualize
organisms_datacuration = datacuration_clean %>%
  filter(!str_detect(str_to_lower(title), "cancer|tumor|autoimmune|drug")) %>% 
  filter(str_detect(str_to_lower(title), "vaccin|immuniz")) %>% 
  separate(organism, into = c("genus", "species"), remove = F) %>% 
  group_by(genus) %>% 
  summarize(n_organism = n()) %>% 
  # filter(!str_detect(organism,"Homo sapiens")) %>% 
  arrange(-n_organism) %>% 
  mutate(genus = fct_reorder(genus, -n_organism))

organisms_datacuration_barplot = organisms_datacuration %>% 
  ggplot() +
  aes(x = n_organism,
      y = genus %>% fct_reorder(n_organism)) +
  geom_col(fill = "gray80") + 
  geom_col(data = organisms_datacuration %>% filter(genus %in% c("Homo", "Mus")),
           fill = "#4cc9f0") + 
  theme_light()+
  geom_text(aes(label = n_organism),
            hjust = -0.2,
            size = 2.5) +
  facet_wrap(~genus, ncol = 1, scales = "free_y") +
  scale_x_continuous(expand = c(0.01,0.01),
                     limits = c(0, max(organisms_datacuration$n_organism) + 50)) +
  theme(
    legend.position = c(0.95,0.5), 
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.title = element_markdown(size = 10, face = "bold", margin = margin(0, 0, 0, 2.5)),
    plot.subtitle = element_markdown(size = 8, face = "plain", margin = margin(0, 0, 3, 3)),
    plot.caption = element_textbox_simple(
      size = 5,
      width = unit(1, "npc"),  # faz quebra automática até a largura do gráfico
      halign = 0,              # alinha à esquerda (opcional)
      margin = margin(2, 2, 2, 3)),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    panel.border = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line.x = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_blank(),
    panel.spacing = unit(0, "cm"),
    strip.text = element_text(color = "black", size = 8,face = "italic", hjust = 0, margin = margin(0, 0, 1, 2)),
    plot.margin = margin(0.2, 0, 0, 0, "cm")) +
    labs(x = "",
       y = "",
       title = "BioProject data curation",
       subtitle = "Number of immunization datasets, by genus", 
       caption = "**Search terms:** (('vaccine'[All Fields] OR 'immunization'[All Fields]\nOR 'vaccinated'[All Fields] OR 'vaccination'[All Fields] OR 'vaccines'[All Fields]) AND 'transcriptome gene expression'[Filter]\n AND 'material transcriptome'[Filter] AND 'capture_whole'[Filter] AND 'org mammals'[Filter] AND 'Mus musculus'[Organism].
       **Downstream filters:** In Title (NOT 'cancer|tumor|autoimmune|drug', INCLUDE 'vaccin|immuniz')")
  
organisms_datacuration_barplot

organisms_datacuration_barplot %>% 
  ggsave(filename = here("Figures", "organisms_datacuration_barplot.png"),
         width = 3, height = 4)
```

### BioProject metadata

```{r}

### BioProject metadata -------
#Filter only Mus musculus vaccination datasets
mice_vaccination_mining = datacuration_clean %>% 
  filter(str_detect(organism,"Mus musculus")) %>% 
  filter(!str_detect(str_to_lower(title), "cancer|tumor|autoimmune|drug")) %>% 
  filter(str_detect(str_to_lower(title), "vaccin|immuniz"))

#Get metadata for one dataset
bioprojectid = mice_vaccination_mining$id[[1]]
bioproject_meta = entrez_summary(db = "bioproject", id = bioprojectid) %>% 
  as.tibble() %>% 
  select(project_name,
         project_description,
         organism_name,
         project_id,
         project_data_type,
         project_target_material,
         project_target_capture,
         organism_name)


#Loop for every dataset
bioproject_ids <- mice_vaccination_mining$id
bioproject_metadata_df <- map_dfr(bioproject_ids, ~
  entrez_summary(db = "bioproject", id = .x) %>%
    as.list() %>%
    as_tibble() %>%
    select(
      project_name,
      project_description,
      organism_name,
      project_data_type,
      project_target_material,
      project_target_capture
    ) %>%
    mutate(bioproject_id = .x)
)

#Filter out data
bioproject_metadata_df = bioproject_metadata_df %>% 
  filter(!str_detect(project_description, "cancer|neoantigen|tumor|autoimmune|drug|opioid"))

#Save
bioproject_metadata_df %>% 
  write_excel_csv2(here("tables", "mice_vaccination_mining2.csv"))

#Import
mice_vaccination_mining2 <- read_delim(here("table", "mice_vaccination_mining2.csv"), 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)

#Print
bioproject_metadata_df = mice_vaccination_mining2

```

##GEO
Get metadata from GEO
```{r}
##GEO Metadata -----
#Get metadata 
#Single ------
# query NCBI GEO 
# geo_ids <- "GSE120661"
# terms <- entrez_db_searchable(db = "gds")  # for reference: searchable fields
# # first, get additional information about this series, including its db ID.
# series <- entrez_search(db = "gds",
#                         term = glue("{geo_ids}[ACCN] AND gse[ETYP]"), )
# # use the ID to fetch details, including the sample table
# series_summary = entrez_summary(db = "gds", id = series$ids)
# 
# series_summary_df = data_frame(bioproject = series_summary$bioproject,
#                                uid = series_summary$uid,
#                                gse= series_summary$gse,
#                                gds= series_summary$gds,
#                                title_series= series_summary$title,
#                                summary= series_summary$summary,
#                                gpl= series_summary$gpl,
#                                seriestitle = series_summary$seriestitle,
#                                n_samples= series_summary$n_samples,
#                                platformtitle= series_summary$platformtitle,
#                                platformtaxa = series_summary$platformtaxa,
#                                ptechtype = series_summary$ptechtype) %>% 
#   mutate(gpl = paste0("GPL", gpl))
# 
# 
# #Annotate platform
# platform_human_1_5000 <- read_csv(here("tables", "platform_human_1-5000.csv"))
# platform_human_5000_6821 <- read_csv(here("tables", "platform_human_5000-6821.csv"))
# platform_mouse <- read_csv(here("tables", "platform_mouse.csv"))
# 
# gpl_annotation_human_mouse = platform_mouse %>% 
#   bind_rows(platform_human_1_5000,
#             platform_human_5000_6821) %>% 
#   clean_names() %>% 
#   rename(gpl = accession,
#          title_platform = title)
# 
# #Save
# gpl_annotation_human_mouse %>% 
#   saveRDS(file = here("tables", "gpl_annotation_human_mouse.rds"))
# 
# #Import
# gpl_annotation_human_mouse = 
#   readRDS(file = here("tables", "gpl_annotation_human_mouse.rds"))
# 
# #Join
# series_summary_df %>% 
#   inner_join(gpl_annotation, by = "gpl")

```


```{r}
#Multiple -------
gpl_annotation_human_mouse = 
  readRDS(file = here("tables", "gpl_annotation_human_mouse.rds"))

geo_ids = read_delim(here("tables", "gds_result.txt"), delim = "\n", col_names = F) %>%
  mutate(X1 = str_c("GSE", X1, sep = "") %>%
           str_remove(., "200")) %>% 
  pull()

text = readLines(here("tables", "gds_result (1).txt"))
text_collapsed <- paste(text, collapse = "\n")
gse_ids <- str_extract_all(geo_ids, "GSE\\d+\\b") %>% unlist() %>% unique()


geo_metadata_df <- map_dfr(gse_ids, ~ {
  result <- entrez_search("gds", term = glue("{.x}[ACCN] AND gse[ETYP]"))

  if (length(result$ids) == 0) {
    return(tibble(gse = .x, error = "Not found"))
  }

  summary <- entrez_summary(db = "gds", id = result$ids) %>% as.list()

  tibble(field = names(summary), value = summary) %>%
    pivot_wider(names_from = field, values_from = value) %>%
    mutate(gse = .x)
})



AnimalVax_geo_metadata_df = geo_metadata_df %>% 
  drop_na(uid) %>% 
  rename(title_series = title) %>% 
  mutate(bioproject_id = str_remove(bioproject, "PRJNA"),
           gpl = paste0("GPL", gpl)) %>% 
  select(uid, gse_id = accession, 
         title_series, summary, samples_count,
         gpl, taxon, entrytype, gdstype, pdat, suppfile, pubmedids, bioproject)


#Save
geo_metadata_df %>% 
  saveRDS(file = here("tables", "geo_metadata_df.rds"))

```

## Merge GEO and Bioproject metadata

```{r}
#Merge metadata from GEO with manually annotated
AnimalVax_DataCuration_Step4 <- read_excel(here("DataCuration", "AnimalVax_DataCuration_20-09-25.xlsx"), 
    sheet = "BioProject_Step4_HumanAtlas_Yes")

geo_metadata_df =
  readRDS(file = here("tables", "geo_metadata_df.rds")) %>% 
  select(gse_id, gpl, gdstype, taxon, pdat, suppfile, pubmedids) 



AnimalVax_DataCuration_Annotated = AnimalVax_DataCuration_Step4 %>% 
  #Join annotation
   rename(gse_id = gse) %>% 
  inner_join(geo_metadata_df, by = "gse_id") %>% 
  
  #Arrange columns
  select(
         title,
         project_description,
         gse_id, 
         keywords,
         notes,
         timepoints,
         time_type,
         doses,
         vaccine_type, 
         pathogen_type,
         target_pathogen,
         target_antigen, 
         adjuvanted, 
         viral_vector,
         sample_source, 
         administration_route, 
         sequencing,
         gpl,
         gdstype,
         taxon,
         pdat,
         suppfile,
         pubmedids
         ) %>% 
  
  #Separate rows with multiple entries
  separate_rows(sample_source, sep = "//") %>% 
  separate_rows(administration_route, sep = "//") %>%
  separate_rows(target_antigen, target_pathogen, sep = "//") %>% 
  separate_rows(adjuvanted, sep = "//") %>%
  
  
  #Handling the GPL/RNA assay methods
  #Separate multiple GPL entries
  mutate(gpl = str_replace(gpl, ";", "//")) %>%
  separate_rows(gpl, sequencing, 
                sep = "//") %>%
  #Complete entries without GPLs
  mutate(gpl = ifelse(grepl("GPL", gpl), gpl, paste0("GPL", gpl))) %>% 
  #Check gds_type (Manual vs GEOQuery)
  mutate(sequencing = case_when(gdstype == "Expression profiling by array"~"Microarray",
                                gdstype == "Expression profiling by RT-PCR" ~ "RT-PCR",
                                T ~sequencing)) %>% 
  
  #Separate main sample source from specific sample
  separate(
    sample_source, into = c("main_sample_source", "specific_sample_source"),
    sep = " \\(|\\)",
    remove = TRUE,
    extra = "drop",
    fill = "right"
  ) %>% 
    separate(
    target_antigen, into = c("target_antigen", "specific_antigen"),
    sep = " \\(|\\)",
    remove = TRUE,
    extra = "drop",
    fill = "right"
  ) %>% 
  separate(
    target_pathogen, into = c("target_pathogen", "specific_pathogen"),
    sep = " \\(|\\)",
    remove = TRUE,
    extra = "drop",
    fill = "right"
  ) %>% 
  distinct() 

AnimalVax_DataCuration_Annotated %>% 
  get_dupes()
  
AnimalVax_DataCuration_Annotated %>% 
  View()

AnimalVax_DataCuration_Annotated %>% 
  write_csv(file = here("DataCuration", "AnimalVax_DataCuration_Annotated.csv"))

AnimalVax_DataCuration_Blood = AnimalVax_DataCuration_Annotated %>% 
  filter(str_detect(main_sample_source, "Blood|PBMCs"))
```


### Selected datasets
```{r}

AnimalVax_DataCuration_Annotated = 
  read_csv(file = here("DataCuration", "AnimalVax_DataCuration_Annotated.csv")) %>% 
  mutate(adjuvant = adjuvanted,
         adjuvanted = if_else(adjuvanted != "No", "Yes", adjuvanted))

AnimalVax_DataCuration_SelectedDatasets = read_csv(here("DataCuration", "AnimalVax_DataCuration - DataCuration_Selected.csv")) %>% 
   separate_rows(timepoints, sep = "//") %>% 
  separate_rows(strain, sep = "//")


AnimalVax_DataCuration_SelectedDatasets_long = AnimalVax_DataCuration_SelectedDatasets %>% 
  select(gse_id, n_samples, timepoints, time_type, vaccine_type, target_pathogen, target_antigen:sequencing, strain) %>% 
  mutate(timepoints = as.numeric(timepoints),
         n_samples = as.numeric(n_samples),
         timepoints = case_when(time_type == "Hour" ~ timepoints/24,
                                time_type == "Week" ~ timepoints * 7,
                                T ~ timepoints) %>% round(1),
         time_type = "Hour",
         adjuvant = adjuvanted,
         adjuvanted = if_else(adjuvanted != "No", "Yes", adjuvanted))
```


### Demographics plots

```{r fig.height=2, fig.width=5}
demographics_mice_studies = AnimalVax_DataCuration_SelectedDatasets_long %>% 
  mutate(timepoints = as.character(timepoints),
         target_antigen = if_else(target_antigen == "Quadri 2019-2020", "Quadri 19-20", target_antigen)) %>% 
  ggplot(aes(x = timepoints, y = gse_id)) +
  facet_wrap(~gse_id, scales = "free_y", ncol = 1) +
  
  #Filled
  geom_point(
    aes(fill = target_antigen, 
        color = target_antigen,
        shape = adjuvanted),
    size = 3,
    stroke = 0.2,
    position = position_dodge(width = 0.75)
  ) +
  
  scale_shape_manual(
    values = c("No" = 21, 
               "Yes" = 23),
    name = "Adjuvanted"
  ) +
  scale_fill_manual(values = colors_vaccines,
                    name = "Vaccine") +
  scale_color_manual(values = colors_vaccines,
                    name = "Vaccine") +

  theme_vaxgo() +
  # scale_x_discrete(guide = "prism_bracket")+ 
  # scale_x_continuous(breaks = c(0, -0.1, 0.1, 0.2, 0.3, 1, 3, 7)) +
  labs(x = "Day", 
       y = "") +
  theme_vaxgo() +
  theme(legend.position = "right",
        strip.text.x = element_text(size = 10, hjust = 0, 
                                    margin = margin(0,0,0,0)),
        axis.title.x = element_text(size = 10),
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 10),
        axis.text.x = element_text(angle = 0),
        axis.text.y = element_blank(),
    panel.grid.major.y = element_line(size = 0.5,
                                    linetype = 1,
                                    color = "black"),
    panel.grid.major.x = element_line(size = 0.5,
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 1, color = "black"),
    # axis.ticks.y = element_line(size = 0, color = "black"),
    axis.ticks.y = element_blank(),
    plot.margin = margin(0, 0, 0, 0, unit = "cm"))

demographics_mice_studies

 demographics_mice_studies %>% 
   ggsave(filename = here("DataCuration", "demographics_mice_studies.png"),
          height = 2,
          width = 5)

```


## Visualization

### Sunburst
Visualized on Flourish App
```{r}
#Sunburst
AnimalVax_DataCuration_Annotated_sunburst = AnimalVax_DataCuration_Annotated %>% 
  
#Arrange columns
  select(
         gse_id, 
         title,
         pubmedids,
         # project_description,
         # keywords,
         # notes,
         # timepoints,
         # time_type,
         # doses,
         pathogen_type, 
         target_pathogen,
         target_antigen, 
         vaccine_type, 
         adjuvanted,
         adjuvant,
         administration_route,
         main_sample_source, 
         specific_sample_source,
         sequencing,
         # gpl,
         # gdstype,
         # taxon,
         # pdat,
         # suppfile,
         ) %>% 
  distinct()

AnimalVax_DataCuration_Annotated_sunburst %>% 
  write_csv(file = here("DataCuration", "AnimalVax_DataCuration_Annotated_sunburst.csv"))

```



### Sankey diagram
Visualized on Flourish App

```{r}

#Create sankey diagram
AnimalVax_DataCuration_Annotated_blood = AnimalVax_DataCuration_Annotated %>% 
  filter(main_sample_source == "Blood")

##### Layer 1
AnimalVax_DataCuration_Annotated_sankey_level1 = AnimalVax_DataCuration_Annotated_blood %>% 
  select(source = pathogen_type, target = target_pathogen, gse_id) %>%  
  group_by(source, target) %>% 
  summarize(value = n(),
            step_from = 0, step_to = 1,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "pathogen_type to target_pathogen") 

##### Layer 2
AnimalVax_DataCuration_Annotated_sankey_level2 = AnimalVax_DataCuration_Annotated_blood %>% 
  select(source = target_pathogen, target = target_antigen, gse_id) %>%
  group_by(source, target) %>% 
  summarize(value = n(),
            step_from = 1, step_to = 2,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "target_pathogen to target_antigen") 

##### Layer 3
AnimalVax_DataCuration_Annotated_sankey_level3 = AnimalVax_DataCuration_Annotated_blood %>% 
  select(source = target_antigen, target = vaccine_type, gse_id) %>%
  group_by(source, target) %>% 
  summarize(value = n(),
            step_from = 2, step_to = 3,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "target_antigen to vaccine_type")

AnimalVax_DataCuration_Annotated_sankey_level4 = AnimalVax_DataCuration_Annotated_blood %>% 
  select(source = vaccine_type, target = adjuvanted, gse_id) %>%
  group_by(source, target) %>% 
  summarize(value = n(),
            step_from = 3, step_to = 4,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "vaccine_type to adjuvanted") %>% 
  mutate(source = if_else(is.na(source), "No", source),
         target = if_else(is.na(target), "No", target))

AnimalVax_DataCuration_Annotated_sankey_level5 = AnimalVax_DataCuration_Annotated_blood %>% 
  select(source = adjuvanted, target = administration_route, gse_id) %>%
  group_by(source, target) %>% 
  summarize(value = n(),
            step_from = 4, step_to = 5,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "adjuvanted to administration_route") %>% 
  mutate(source = if_else(is.na(source), "No", source),
         target = if_else(is.na(target), "No information", target))

AnimalVax_DataCuration_Annotated_sankey_level6 = AnimalVax_DataCuration_Annotated_blood %>% 
  select(source = administration_route, target = main_sample_source, gse_id) %>%
  group_by(source, target) %>% 
  summarize(value = n(),
            step_from = 5, step_to = 6,
            gse_ids = paste0(gse_id, collapse = ";"),
            layer = "administration_route to main_sample_source")


#Bind tables
AnimalVax_DataCuration_Annotated_sankey = AnimalVax_DataCuration_Annotated_sankey_level1 %>% 
  bind_rows(AnimalVax_DataCuration_Annotated_sankey_level2,
            AnimalVax_DataCuration_Annotated_sankey_level3,
            AnimalVax_DataCuration_Annotated_sankey_level4,
            AnimalVax_DataCuration_Annotated_sankey_level5,
            AnimalVax_DataCuration_Annotated_sankey_level6) %>% 
  distinct()
  

AnimalVax_DataCuration_Annotated_sankey %>% 
  View()
  
#Save
AnimalVax_DataCuration_Annotated_sankey %>% 
  write_csv(file = here("DataCuration", "AnimalVax_DataCuration_Annotated_sankey_blood.csv"))
  
  
```


Sankey example
```{r fig.height=10, fig.width=10}
# install.packages("remotes")
# library(remotes)
# install_github("davidsjoberg/ggsankey")
library(ggsankey)

sankey_df <- AnimalVax_DataCuration_Annotated_blood %>%
  make_long(pathogen_type, 
         target_pathogen,
         vaccine_type, 
         target_antigen, 
         adjuvanted,
         # administration_route,
         # main_sample_source, 
         specific_sample_source,
         sequencing,  
         gse_id) %>% 
  drop_na()

datacuration_blood_sankey_ggplot = ggplot(sankey_df, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = factor(node),
               label = node
               )) +
  geom_alluvial(flow.alpha = .6, 
              space = 1) +
  geom_alluvial_label(size = 3.5,space = 1, 
                     color = "black", 
                     fill = "white", 
                     alpha = 0.5) +
  scale_fill_viridis_d(drop = FALSE) +
  theme_alluvial(base_size = 12) +
  theme(legend.position = "none")

# ggplotly(sankey_ggplot)

```

### Heatmap
#### All datasets
```{r fig.height=21, fig.width=10}
heatmap_df = AnimalVax_DataCuration_Annotated %>%
  select(pathogen_type, 
         target_pathogen,
         target_antigen,
         vaccine_type, 
         adjuvanted,
         administration_route,
         main_sample_source,
         specific_sample_source,
         sequencing,  
         gse_id) %>% 
  #Handling duplicate gse ids
   group_by(gse_id) %>%
   mutate(gse_id_unique = paste0(gse_id, "_", row_number()))  %>%
  ungroup(gse_id) %>% 
  select(-gse_id) %>% 
   pivot_longer(cols = -gse_id_unique,
                names_to = "variable",
                values_to = "value") %>% 
  mutate(variable = fct_relevel(variable, c("pathogen_type",
                                            "target_pathogen",
                                            "target_antigen",
                                            "vaccine_type",
                                            "adjuvanted",
                                            "administration_route",
                                            "main_sample_source",
                                            "specific_sample_source",
                                            "sequencing")))


#Color categories
colors_all <- c("#90A4AEFF", "#4DBBD5FF", "#4361ee")
paleta <- colorRampPalette(colors_all)
cores_15 <- paleta(100)
levels_heatmap <- unique(na.omit(heatmap_df$value))
levels_heatmap <- levels_heatmap[!levels_heatmap %in% c("NA", "No")]

# sorteia cores para eles
set.seed(123) # opcional, pra reprodutibilidade
colors_random <- sample(cores_15, length(levels_heatmap), replace = TRUE)

# monta a paleta completa
palette_final <- c(setNames(colors_random, levels_heatmap))


#Clustering
heatmap_wide <- heatmap_df %>%
  select(gse_id_unique, variable, value) %>%
  pivot_wider(names_from = variable, values_from = value) %>% 
  mutate(gse_id_unique = as.character(gse_id_unique))


#Clusterize by pathogen type

#Virus
heatmap_numeric <- heatmap_wide %>%
  filter(pathogen_type == "Virus") %>% 
  mutate(across(-gse_id_unique, ~as.integer(as.factor(.))))
dist_matrix <- dist(heatmap_numeric %>% select(-gse_id_unique), 
                    method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
ordered_gse_virus <- heatmap_numeric$gse_id_unique[hc$order]

#Bacteria
heatmap_numeric <- heatmap_wide %>%
  filter(pathogen_type == "Bacteria") %>% 
  mutate(across(-gse_id_unique, ~as.integer(as.factor(.))))
dist_matrix <- dist(heatmap_numeric %>% select(-gse_id_unique), 
                    method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
ordered_gse_bacteria <- heatmap_numeric$gse_id_unique[hc$order]

#Fungi
heatmap_numeric <- heatmap_wide %>%
  filter(pathogen_type == "Fungi") %>% 
  mutate(across(-gse_id_unique, ~as.integer(as.factor(.))))
dist_matrix <- dist(heatmap_numeric %>% select(-gse_id_unique), 
                    method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
ordered_gse_fungi <- heatmap_numeric$gse_id_unique[hc$order]

#Protozoa
heatmap_numeric <- heatmap_wide %>%
  filter(pathogen_type == "Protozoa") %>% 
  mutate(across(-gse_id_unique, ~as.integer(as.factor(.))))
dist_matrix <- dist(heatmap_numeric %>% select(-gse_id_unique), 
                    method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")
ordered_gse_protozoa <- heatmap_numeric$gse_id_unique[hc$order]

# reorder
heatmap_df <- heatmap_df %>%
  mutate(gse_id_unique = fct_relevel(gse_id_unique, c(ordered_gse_virus, ordered_gse_bacteria,ordered_gse_fungi ,ordered_gse_protozoa)) %>% fct_rev())

#Plot
heatmap_plot = heatmap_df %>% 
  ggplot(aes(x = variable, y = gse_id_unique,
             label = value)) +
  
  #Layer
  geom_tile(data = ~filter(.x, variable == "pathogen_type"),
            aes(fill = value)) +
  scale_fill_manual(values = c(palette_final),
                    na.value = "transparent",
                    name = "pathogen_type") +
  new_scale_fill() +
  
   #Layer
  geom_tile(data = ~filter(.x, variable == "target_pathogen"),
            aes(fill = value)) +
  scale_fill_manual(values = c(palette_final),
                    na.value = "transparent",
                    name = "target_pathogen") +
  new_scale_fill() +
  
  #Layer
  geom_tile(data = ~filter(.x, variable == "target_antigen"),
            aes(fill = value)) +
  scale_fill_manual(values = c(palette_final),
                    na.value = "transparent",
                    name = "target_antigen") +
   new_scale_fill() +
  #Layer
  geom_tile(data = ~filter(.x, variable == "vaccine_type"),
            aes(fill = value)) +
  scale_fill_manual(values = c(palette_final),
                    na.value = "transparent",
                    name = "vaccine_type") +
   new_scale_fill() +
  #Layer
  geom_tile(data = ~filter(.x, variable == "adjuvanted"),
            aes(fill = value),
            na.rm = T) +
  scale_fill_manual(values = c(palette_final),
                    na.value = "transparent",
                    name = "adjuvanted") +
   new_scale_fill() +
  #Layer
  geom_tile(data = ~filter(.x, variable == "administration_route"),
            aes(fill = value)) +
  scale_fill_manual(values = c(palette_final),
                    na.value = "transparent",
                    name = "administration_route") +
   new_scale_fill() +
  #Layer
  geom_tile(data = ~filter(.x, variable == "main_sample_source"),
            aes(fill = value)) +
  scale_fill_manual(values = c(palette_final),
                    na.value = "transparent",
                    name = "main_sample_source") +
   new_scale_fill() +
  #Layer
  geom_tile(data = ~filter(.x, variable == "specific_sample_source"),
            aes(fill = value)) +
  scale_fill_manual(values = c(palette_final),
                     na.value = "transparent",
                    name = "specific_sample_source") +
   new_scale_fill() +
   #Layer
  geom_tile(data = ~filter(.x, variable == "sequencing"),
            aes(fill = value)) +
  scale_fill_manual(values = c(palette_final),
                    na.value = "transparent",
                    name = "sequencing") +
  theme_vaxgo() +
  # facet_wrap(~variable, scales = "free_x", nrow = 1) +
  geom_text(size = 2, color = "white", fontface = "bold") +
  #Do not visualize the legends here
  theme(strip.text = element_blank(), 
        # axis.text.y = element_text(size = 5),
        panel.spacing = unit(0, "lines"),
    legend.position = "none", 
    # legend.direction = "vertical", 
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8)
  )  
  
heatmap_plot
  
```


#### Blood only
```{r fig.height=21, fig.width=10}
heatmap_df = AnimalVax_DataCuration_Blood %>%
  select(pathogen_type, 
         target_pathogen,
         target_antigen,
         vaccine_type, 
         adjuvanted,
         administration_route,
         main_sample_source,
         specific_sample_source,
         sequencing,  
         gse_id) %>% 
  #Handling duplicate gse ids
   group_by(gse_id) %>%
   mutate(gse_id_unique = paste0(gse_id, "_", row_number()))  %>%
  ungroup(gse_id) %>% 
  select(-gse_id) %>% 
   pivot_longer(cols = -gse_id_unique,
                names_to = "variable",
                values_to = "value") %>% 
  mutate(variable = fct_relevel(variable, c("pathogen_type",
                                            "target_pathogen",
                                            "target_antigen",
                                            "vaccine_type",
                                            "adjuvanted",
                                            "administration_route",
                                            "main_sample_source",
                                            "specific_sample_source",
                                            "sequencing")))


#Color categories
colors_all <- c("#90A4AEFF", "#4DBBD5FF", "#4361ee", "#669bbc", "#a2d2ff", "#219ebc")
paleta <- colorRampPalette(colors_all)
cores_15 <- paleta(3)
levels_heatmap <- unique(na.omit(heatmap_df$value))
levels_heatmap <- levels_heatmap[!levels_heatmap %in% c("NA", "No")]

# sorteia cores para eles
set.seed(123) # opcional, pra reprodutibilidade
colors_random <- sample(cores_15, length(levels_heatmap), replace = TRUE)

# monta a paleta completa
palette_final <- c(setNames(colors_random, levels_heatmap))


#Clustering
heatmap_wide <- heatmap_df %>%
  select(gse_id_unique, variable, value) %>%
  pivot_wider(names_from = variable, values_from = value) %>% 
  mutate(gse_id_unique = as.character(gse_id_unique))


#Clusterize by pathogen type

#Virus
heatmap_wide_arranged = heatmap_wide %>% 
  arrange(pathogen_type, target_pathogen, target_antigen)

# reorder
heatmap_df <- heatmap_df %>% 
  mutate(gse_id_unique = fct_relevel(gse_id_unique, c(heatmap_wide_arranged$gse_id_unique)) %>% fct_rev())
```


```{r fig.height=5, fig.width=8}
library("paletteer")

#Plot
heatmap_plot_blood = heatmap_df %>% 
  
  ggplot(aes(x = variable, y = gse_id_unique,
             label = value)) +
  
  #Layer
  geom_tile(data = ~filter(.x, variable == "pathogen_type"),
            aes(fill = value)) +
  scale_fill_paletteer_d("dichromat::LightBluetoDarkBlue_10") +
  new_scale_fill() +
  
   #Layer
  geom_tile(data = ~filter(.x, variable == "target_pathogen"),
            aes(fill = value)) +
  scale_fill_paletteer_d("dichromat::LightBluetoDarkBlue_10") +
  new_scale_fill() +
  
  #Layer
  geom_tile(data = ~filter(.x, variable == "target_antigen"),
            aes(fill = value)) +
  scale_fill_paletteer_d("dichromat::LightBluetoDarkBlue_10") +
   new_scale_fill() +
  #Layer
  geom_tile(data = ~filter(.x, variable == "vaccine_type"),
            aes(fill = value)) +
  scale_fill_paletteer_d("dichromat::LightBluetoDarkBlue_10") +
   new_scale_fill() +
  #Layer
  geom_tile(data = ~filter(.x, variable == "adjuvanted"),
            aes(fill = value),
            na.rm = T) +
  scale_fill_paletteer_d("dichromat::LightBluetoDarkBlue_10") +
   new_scale_fill() +
  #Layer
  geom_tile(data = ~filter(.x, variable == "administration_route"),
            aes(fill = value)) +
  scale_fill_paletteer_d("dichromat::LightBluetoDarkBlue_10") +
   new_scale_fill() +
  #Layer
  geom_tile(data = ~filter(.x, variable == "main_sample_source"),
            aes(fill = value)) +
  scale_fill_paletteer_d("dichromat::LightBluetoDarkBlue_10") +
   new_scale_fill() +
  #Layer
  geom_tile(data = ~filter(.x, variable == "specific_sample_source"),
            aes(fill = value)) +
  scale_fill_paletteer_d("dichromat::LightBluetoDarkBlue_10") +
   new_scale_fill() +
   #Layer
  geom_tile(data = ~filter(.x, variable == "sequencing"),
            aes(fill = value)) +
  scale_fill_paletteer_d("dichromat::LightBluetoDarkBlue_10") +
  # scale_fill_manual(values = c(palette_final),
  #                   na.value = "transparent",
  #                   name = "sequencing") +
  theme_vaxgo() +
  # facet_wrap(~variable, scales = "free_x", nrow = 1) +
  geom_text(size = 2, color = "black", fontface = "bold") +
  #Do not visualize the legends here
  theme(strip.text = element_blank(), 
        # axis.text.y = element_text(size = 5),
        panel.spacing = unit(0, "lines"),
    legend.position = "none", 
    # legend.direction = "vertical", 
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8)
  )  +
  labs(x = "",
       y = "")
  
heatmap_plot_blood

# heatmap_plot_blood %>% 
#   ggsave(filename = here("Figures", "AnimalVax_DataCuration_Blood_heatmap.png"),
#          width = 10,
#          height = 5)
  
```


```{r}
#Save
mice_vaccination_annotated %>% 
  write_csv(here("tables", "mice_vaccination_annotated.csv"))

#Print
mice_vaccination_annotated
```




# 2. Demographics
## Fluad and Engerix B

```{r}
#Import datasets
GSE124689_expression_not_normalized_log2_raw_symbol = read_rds(file = here("tables", "influenza_GSE124689_exprs_not_normalized.rds"))
GSE124689_metadata_clean = readRDS(file = here("tables", "influenza_fluad_human_metadata_clean.rds"))
GSE120661_counts_not_normalized_log2_annotations = read_rds(here("tables", "GSE120661_counts_not_normalized_log2_annotations_not_normalized_log2.rds"))
GSE120661_samples_pbmc = read_rds(here("tables", "GSE120661_samples_pbmc.rds"))

#Bind mouse metadata to human data
human_mouse_fluad_metadata = GSE120661_samples_pbmc %>% 
  filter(treatment == "Fluad") %>% 
  select(sample_id = geo_sample, 
         participant_id = subject_id, 
         vaccine = treatment,
         sex,
         strain,
         sample_type = tissue,
         timepoint,
         condition) %>% 
  mutate(organism = "Mouse",
         type = "Inactivated",
         adjuvant = "MF59",
         age = 7, #6–8 weeks
         timepoint = as.character(timepoint) %>% as.numeric(.) / 24,
         timepoint = if_else(timepoint < 1 & timepoint != 0,
                             as.character(round(timepoint, 1)),
                             as.character(as.integer(round(timepoint, 0)))),
         race = strain, 
         array = "Agilent-028005 SurePrint G3 Mouse GE 8x60K Microarray") %>% 
  
  #Bind human data
  bind_rows(GSE124689_metadata_clean %>% 
               select(sample_id,
                      participant_id,
                      condition, timepoint,
                      sex, age, race, vaccine, 
                      sample_type, array, 
                      type, adjuvant) %>% 
              mutate(organism = "Human",
                     vaccine = "Fluad",
                     timepoint = as.character(timepoint)))

human_mouse_fluad_metadata

human_mouse_fluad_metadata %>% 
  saveRDS(file = here("tables", "influenza_fluad_human_mouse_metadata.rds"))
```

##### By organism
```{r fig.height=5, fig.width=5}
#Visualize samples ------
demography_samples = human_mouse_fluad_metadata %>% 
  group_by(organism, timepoint, race) %>% 
  summarize(n = n()) %>% 
  mutate(timepoint = as.character(timepoint)) 

demography_samples_plot = demography_samples %>% 
  ggplot() +
  aes(x = timepoint, 
      y = organism) +
  geom_point(aes(size = n,
                 fill = organism),
             pch = 22) +
  scale_size_binned(breaks = c(1, 10, 50),
                    range = c(2, 10)) +
  scale_fill_manual(values = c(
    "Human"  = "#90A4AEFF",
    "Mouse" = "#4DBBD5FF",
    "FIT"   = "#4361ee"
  )) +
    scale_color_manual(values = c(
    "Human"  = "#90A4AEFF",
    "Mouse" = "#4DBBD5FF",
    "FIT"   = "#4361ee"
  )) +
  theme_vaxgo() +
  theme(legend.position = "right",
        axis.title.x = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.text.x = element_text(angle = 0),
    panel.grid.major.y = element_line(size = 1,
                                    linetype = 1,
                                    color = "black"),
    panel.grid.major.x = element_line(size = 0.5,
                                    linetype = 2,
                                    color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 1, color = "black"),
    axis.ticks.y = element_line(size = 0, color = "black"),
    plot.margin = margin(0, 0, 0, 0, unit = "cm")) +
  labs(x = "Days",
       y = "",
       size = "Samples",
       fill = "Organism")  +
  geom_label(aes(label = n),
             fill = "white",
             label.size = NA,
            size = 3,
            nudge_x = 0,
            nudge_y = 0.1,
            data = demography_samples %>% 
              filter(organism == "Human")) +
  geom_label(aes(label = n),
             fill = "white",
             label.size = NA,
            size = 3,
            nudge_x = 0,
            nudge_y = 0.1,
            data = demography_samples %>% 
              filter(organism == "Mouse")) +
  theme(legend.position = "none")

#Print
demography_samples_plot

#Save
demography_samples_plot %>% 
  ggsave(filename = here("Figures", "influenza_fluad_human_mouse_demography.png"),
         width = 5,
         height = 5)

```



### By sex
```{r fig.height=5, fig.width=2}
#Visualize samples by sex ------
demography_samples_sex = human_mouse_fluad_metadata %>% 
  select(participant_id, sex, organism) %>% 
  distinct() %>% 
  group_by(organism, sex) %>%
  summarize(n = n())

demography_samples_sex %>% 
ggplot() +
  aes(
    x = organism,
    y = n,
    fill = sex,
    weight = n,
    label = n
  ) +
  geom_col(color = "black") +
  coord_polar(theta = "y") +
  theme_void() +
  labs(fill = "Sex") +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        plot.title = element_text(size = 15, hjust = 0.5),
        strip.text = element_text(size = 12),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = margin(1, 0, 0, 0,unit = "cm"),
        legend.position = "top") +
  facet_wrap(vars(fct_rev(organism)), scales = "free", ncol = 1) +
  scale_fill_manual(values = c("Male" = "#4DBBD5FF",
          "Female" = "#4361ee"))
```

By age
```{r fig.height=5, fig.width=2}
#Visualize samples by age ------
demography_samples_age = human_mouse_fluad_metadata %>% 
  select(participant_id, age, organism) %>% 
  distinct() %>% 
  filter(organism == "Human")

demography_samples_age %>% 
  ggplot() +
  aes(x = organism,
      y = age) +
  geom_violin(aes(fill = organism)) +
  geom_boxplot(fill = "white", 
               outliers = F, width = 0.1) +
  geom_quasirandom(color = "black", 
              fill = NA, 
              position = position_dodge2(width = 0.2),
              pch = 21,
              alpha = 0.5) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 50))+
  scale_fill_manual(values = c(
    "Human"  = "#90A4AEFF",
    "Mouse" = "#4DBBD5FF")) +
  theme_vaxgo() +
  theme(legend.position = "none",
        legend.key = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0),
        strip.text.x = element_text(color = "white", size = 10, face = "bold"),
        strip.text.y = element_text(color = "black", size = 10, hjust = 0, angle = 0),
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.x = element_rect(fill = "black"),
        )  +
  labs(y = "Age",
       x = "Human")
```


## All
```{r}
all_human_mouse_metadata = readRDS(here("tables", "all_human_mouse_metadata.rds")) %>% 
  mutate(pathogen = as.factor(pathogen) %>% fct_relevel("Influenza", "Hepatitis B", "Staphylococcus aureus"),
         timepoint = if_else(str_detect(time_unit, "Hour|hours"), timepoint/24, timepoint) %>% round(2),
         timepoint = fct_reorder(as.character(timepoint), timepoint),
         sex = str_to_title(sex))
all_human_mouse_metadata

demography_samples = all_human_mouse_metadata %>% 
  group_by(pathogen, organism, timepoint) %>% 
  summarise(n = n(),
            time_unit) %>% 
  ungroup() %>% 
  distinct()
demography_samples
```


```{r fig.height=5, fig.width=5}
demography_samples_plot = demography_samples %>%
  ggplot() +
  aes(x = timepoint, 
      y = organism) +
  facet_wrap(~pathogen, ncol = 1) +
  geom_point(aes(size = n,
                 fill = organism),
             pch = 22) +
  scale_size_binned( range = c(1, 5)) +
  scale_fill_manual(values = c(
    "Human"  = "#90A4AEFF",
    "Mouse" = "#4DBBD5FF"
  )) +
  theme_vaxgo() +
  theme(legend.position = "right",
        axis.title.x = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.text.x = element_text(angle = 0),
    panel.grid.major.y = element_line(size = 1,
                                    linetype = 1,
                                    color = "black"),
    # panel.grid.major.x = element_line(size = 0.5,
    #                                 linetype = 2,
    #                                 color = "gray75"),
    axis.line.x = element_line(size = 0.5,
                               colour = "black",
                               linetype = 1),
    axis.line.y = element_line(size = 0,
                               colour = "black",
                               linetype = 1),
    axis.ticks.x = element_line(size = 1, color = "black"),
    axis.ticks.y = element_line(size = 1, color = "black"),
    plot.margin = margin(0, 0, 0, 0, unit = "cm")) +
  labs(x = "Days",
       y = "",
       size = "Samples",
       fill = "Organism")  +
  geom_text(aes(label = n),
             label.size = NA,
            size = 3,
            hjust = 0.5,
            vjust = -1,
            data = demography_samples %>% 
              filter(organism == "Human")) +
  geom_text(aes(label = n),
             label.size = NA,
            size = 3,
            hjust = 0.5,
            vjust = -0.8,
            data = demography_samples %>% 
              filter(organism == "Mouse")) +
  theme(legend.position = "none")

#Print
demography_samples_plot

demography_samples_plot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_demography.png"),
         width = 5,
         height = 5)

```

```{r fig.height=5, fig.width=10}
demography_samples_sex = all_human_mouse_metadata %>%
  select(pathogen, participant_id, sex, organism) %>%
  distinct() %>%
  dplyr::count(pathogen, organism, sex, name = "n") %>%
  group_by(pathogen, organism) %>%
  mutate(
    total = sum(n),
    perc = (100 * n / total) %>% round(0)
  ) %>%
  ungroup()
  

demography_samples_sex %>% 
ggplot() +
  aes(
    x = organism,
    y = perc,
    fill = sex,
    weight = perc,
    label = perc
  ) +
  geom_col(color = "black") +
  geom_text(aes(label = perc), position = position_stack(vjust = .5), color = "white")+
  coord_polar(theta = "y") +
  theme_void() +
  labs(fill = "Sex") +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        plot.title = element_text(size = 15, hjust = 0.5),
        strip.text.y = element_text(size = 12, hjust = 0),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = margin(0, 0, 0, 0,unit = "cm"),
        legend.position = "top") +
  facet_grid(~pathogen~organism, scales = "free") +
  scale_fill_manual(values = c("Male" = "#4DBBD5FF",
          "Female" = "#4361ee"))
```

