---
title: "Preprocessing"
output: html_document
---

# Required
```{r}
# install.packages("here") if needed
install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```


# Preprocessing


# 1. Fluad + MF59, Vaccine

## Mouse
### Download dataset
```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Mouse-------
#Download datasets

#GSE120661 ----

#Metadata 
filename = "GSE120661"
GSE120661 <- getGEO("GSE120661", GSEMatrix = TRUE) 
eset <- GSE120661[[1]]

GSE120661_metadata = eset %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    subject_id = "title",
    sex = "gender:ch1", 
    treatment = "treatment:ch1", 
    tissue = "tissue:ch1",
    timepoint = "timepoint:ch1",
    animal_replicate = "animal replicate:ch1",
    strain = "strain/background:ch1") %>% 
  mutate(treatment = fct_relevel(treatment, "saline"),
         timepoint = fct_relevel(timepoint, "0"),
         condition = paste(treatment, timepoint, tissue, sep = "_"))

#Save
write_csv(GSE120661_metadata, file = here("tables", "influenza_hepatitisb_GSE120661_metadata.csv"))

#Counts matrix from Expression set
eset <- exprs(eset)

eset %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "influenza_hepatitisb_GSE120661_exprs.rds"))

```


```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Import
GSE120661_exprs = readRDS(here("tables", "influenza_hepatitisb_GSE120661_exprs.rds"))
GSE120661_counts_not_normalized_log2 = readRDS(here("tables","influenza_hepatitisb_GSE120661_exprs.rds"))

#Check distribution
GSE120661_exprs_distribution = GSE120661_exprs %>% 
  rownames_to_column("probes") %>% 
  pivot_longer(cols = -probes,
               names_to = "sample_id",
               values_to = "value") 

#Plot
GSE120661_exprs_distribution %>% 
  filter(sample_id == "GSM3407136") %>% 
  ggplot() +
  aes(x = value) +
  geom_histogram()
```


Get metadata
```{r}
#Import
GSE120661_metadata = read_csv(here("tables", "influenza_hepatitisb_GSE120661_metadata.csv")) 

#Clean
GSE120661_metadata = GSE120661_metadata %>% 
  mutate(timepoint = as.integer(timepoint) %>% 
           as.character() %>% 
           as.factor() %>% 
           fct_relevel(., "0"),
         treatment = fct_relevel(treatment, "saline"),
         condition = paste(treatment, timepoint, tissue, sep = "_"))

#Save
GSE120661_metadata %>% saveRDS(here("tables", "influenza_fluad_mouse_metadata.rds"))
```

Import
```{r}
#Import
GSE120661_metadata = readRDS(here("tables", "influenza_fluad_mouse_metadata.rds"))
GSE120661_exprs = read_rds(here("tables", "influenza_hepatitisb_GSE120661_exprs.rds"))

#Check length
GSE120661_exprs %>% rownames() %>% length()

mouse_fluad_metadata = GSE120661_metadata %>% 
  filter(treatment == "Fluad",
         tissue == "PBMC") %>% 
  mutate(timepoint = as.character(timepoint) %>% as.numeric(),
         timepoint = (timepoint / 24) %>% round(1),
         participant_id = paste0("TriFLU-MF59_BL", animal_replicate)) %>% 
  filter(timepoint %in% c(0, 1, 3, 7))

#Save
mouse_fluad_metadata %>% 
  saveRDS(here("tables", "influenza_fluad_mouse_metadata.rds"))

#Select only PBMC samples counts
GSE120661_exprs = GSE120661_exprs %>% 
  as.data.frame() %>% 
  select(c(mouse_fluad_metadata$geo_sample)) %>% 
  as.matrix()

GSE120661_exprs %>% 
  as.data.frame() %>% 
  saveRDS(file = here("tables", "influenza_fluad_mouse_exprs.rds"))
```

### Annotate genes
```{r}
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- GSE120661_counts_not_normalized_log2 %>% rownames()

#Get other annotations
listAttributes(ensembl_mart)

#Annotate
annot <- getBM(
  attributes = c("agilent_sureprint_g3_ge_8x60k", #Agilent microarray data (design ID 028005, retrieved from GEO)
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "agilent_sureprint_g3_ge_8x60k",
  values = gene_id,
  mart = ensembl_mart)

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_fluad_log2summary = GSE120661_exprs %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = agilent_sureprint_g3_ge_8x60k), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = mouse_fluad_log2summary %>% 
  #filter ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

mouse_fluad_log2summary = mouse_fluad_log2summary %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
mouse_fluad_log2_filtered_probes = GSE120661_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(mouse_fluad_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol,
         human_entrez,
         mouse_entrez,
         c(mouse_fluad_metadata$geo_sample))
#14395
mouse_fluad_log2_filtered_probes %>% dim()

mouse_fluad_log2_hgnc_symbol = mouse_fluad_log2_filtered_probes 

mouse_fluad_log2_hgnc_symbol %>% get_dupes(human_symbol)

#Save 
mouse_fluad_log2_hgnc_symbol %>% 
  saveRDS(here("tables", "influenza_fluad_mouse_exprs_hgnc_symbol.rds"))
```

Create expression set
```{r}
#Import
mouse_fluad_log2_hgnc_symbol = readRDS(here("tables", "influenza_fluad_mouse_exprs_hgnc_symbol.rds"))
mouse_fluad_metadata = readRDS(here("tables", "influenza_fluad_mouse_metadata.rds")) 


#Annotate
mouse_fluad_log2fc_filtered_sample_long = mouse_fluad_log2_hgnc_symbol %>%
  # rownames_to_column("human_symbol") %>% 
  pivot_longer(cols = -human_symbol,
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(mouse_fluad_metadata %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>%
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")

#Save
mouse_fluad_log2fc_filtered_sample_long %>% 
  saveRDS(here("tables", "influenza_fluad_mouse_log2fc_sample_clean_long.rds"))


#Pivot wider
mouse_fluad_log2fc_filtered_sample_wide = mouse_fluad_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")

beep()

```


### Differential gene expression
```{r}
#Create ExpressionSet --------------
#Import
mouse_fluad_log2_hgnc_symbol = readRDS(here("tables", "influenza_fluad_mouse_exprs_hgnc_symbol.rds"))
mouse_fluad_metadata = readRDS(here("tables", "influenza_fluad_mouse_metadata.rds")) 


GSE120661_exprs_matrix_fluad = mouse_fluad_log2_hgnc_symbol %>% 
  as.data.frame() %>% 
  select(human_symbol, c(mouse_fluad_metadata$geo_sample)) %>% 
  column_to_rownames("human_symbol") %>% 
  as.matrix()

#Normalize quantiles
GSE120661_exprs_matrix_fluad_normalized <- normalizeBetweenArrays(GSE120661_exprs_matrix_fluad, 
                                                                   method = "quantile")

#Diagnostics
dim(GSE120661_exprs_matrix_fluad_normalized) #14357   20

dim(AnnotatedDataFrame(mouse_fluad_metadata %>% column_to_rownames("geo_sample"))) #35
```


```{r}
#Create new Expression set with the filtered data
eset_filtered <- ExpressionSet(
  assayData = GSE120661_exprs_matrix_fluad_normalized,
  phenoData = AnnotatedDataFrame(mouse_fluad_metadata %>%
                                   mutate(timepoint = as.character(timepoint) %>% as.factor()) %>% 
                                   column_to_rownames("geo_sample")))

#Save
eset_filtered %>% 
  saveRDS(file = here("tables", "influenza_fluad_mouse_eset.rds"))
```







#### Limma

```{r}
#Create design matrix -------
eset_mouse_fluad = readRDS(file = here("tables", "influenza_fluad_mouse_eset.rds"))

design <- model.matrix(~ timepoint + 0, eset_mouse_fluad) # 0 defines the way we compare the samples
colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

# Estimate correlation between doses on the same individual
corfit <- duplicateCorrelation(eset_mouse_fluad, 
                               design = design, 
                               block = eset_mouse_fluad$participant_id)
corfit$consensus #0.024

# Use the correlation into the linear model
#Fit Limma model
fit <- lmFit(eset_mouse_fluad, design, 
             block = eset_mouse_fluad$participant_id, 
             correlation = corfit$consensus)

#Contrasts
#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

cont.matrix <- makeContrasts(timepoint1  - timepoint0 ,
                             timepoint3  - timepoint0 ,
                             timepoint7  - timepoint0 ,
                             levels=design)

fit2 <- contrasts.fit(fit, cont.matrix)

# eBayes
fit2 <- eBayes(fit2)

#Get DEGs ------
## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
    
})

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs
all_degs_mouse = all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  mutate(condition = case_when(comparisonA == "Time1"	~ "Fluad_1_PBMC",
                               comparisonA == "Time3"	~ "Fluad_3_PBMC",
                               comparisonA == "Time7"	~ "Fluad_7_PBMC",
                               T~comparisonA),
         timepoint = str_remove(comparisonA, "Time"),
         organism = "Mouse") %>% 
  select(condition, everything()) %>% 
  clean_names() %>% 
  select(condition, symbol, mean = log_fc, lower = ci_l, upper = ci_r, ave_expr,
         everything())

all_degs_mouse %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarize(n_degs = n())
	
#Save
all_degs_mouse %>% saveRDS(file = here("tables", "influenza_fluad_mouse_dge_limma_degs.rds"))
```


#### Simple t-Test
```{r}
#Import
mouse_fluad_log2fc_filtered_sample_long = 
  readRDS(here("tables", "influenza_fluad_mouse_log2fc_sample_clean_long.rds"))
#Import
mouse_fluad_log2_hgnc_symbol = readRDS(here("tables", "influenza_fluad_mouse_exprs_hgnc_symbol.rds")) %>% 
  distinct()
```


```{r}
#Compute summary statistics
fluad_log2fc_summary = mouse_fluad_log2fc_filtered_sample_long %>% 
  group_by(human_symbol, timepoint) %>%
  summarize(
    avg_log2fc = mean(log2fc, na.rm = TRUE),  # mean log2 fold change
    n_subjects = n(),                         # number of samples
    sd = sd(value, na.rm = TRUE),             # standard deviation
    se = sd / sqrt(n_subjects),               # standard error
    effect_size = avg_log2fc / sd)            # Cohen’s d-like effect size

mouse_fluad_log2_hgnc_symbol_wide = mouse_fluad_log2_hgnc_symbol %>%
  # rownames_to_column("human_symbol") %>% 
  pivot_longer(cols = -human_symbol,
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(mouse_fluad_metadata %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") 

# Paired t-test
mouse_fluad_log2fc_paired_ttest <- mouse_fluad_log2_hgnc_symbol_wide %>%
  group_by(human_symbol) %>%
  
  summarise(
    pval_day1 = tryCatch(t.test(day_1, day_0, paired = TRUE, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_day3 = tryCatch(t.test(day_3, day_0, paired = TRUE, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_day7 = tryCatch(t.test(day_7, day_0, paired = TRUE, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    .groups = 'drop'
  ) %>%
  
  mutate(
    padj_day1 = p.adjust(pval_day1, method = "fdr"),
    padj_day3 = p.adjust(pval_day3, method = "fdr"),
    padj_day7 = p.adjust(pval_day7, method = "fdr")
  ) %>% 
  pivot_longer(
    cols = starts_with(c("pval_", "padj_")),
    names_to = c(".value", "timepoint"),
    names_pattern = "(pval|padj)_(day\\d+)"
  ) %>% 
  mutate(condition = case_when(timepoint == "day1"	~ "Fluad_24_PBMC",
                               timepoint == "day3"	~ "Fluad_72_PBMC",
                               timepoint == "day7"	~ "Fluad_168_PBMC",
                               T~timepoint),
         timepoint = str_remove(timepoint, "day"),
         test = "t-Test",
         paired = "Paired")

#Unpaired
mouse_fluad_log2fc_unpaired_ttest <- mouse_fluad_log2_hgnc_symbol_wide %>%
  group_by(human_symbol) %>%
  
  summarise(
    pval_day1 = tryCatch(t.test(day_1, day_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_day3 = tryCatch(t.test(day_3, day_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_day7 = tryCatch(t.test(day_7, day_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    .groups = 'drop'
  ) %>%
  
  mutate(
    padj_day1 = p.adjust(pval_day1, method = "fdr"),
    padj_day3 = p.adjust(pval_day3, method = "fdr"),
    padj_day7 = p.adjust(pval_day7, method = "fdr")
  ) %>% 
  pivot_longer(
    cols = starts_with(c("pval_", "padj_")),
    names_to = c(".value", "timepoint"),
    names_pattern = "(pval|padj)_(day\\d+)"
  ) %>% 
  mutate(condition = case_when(timepoint == "day1"	~ "Fluad_24_PBMC",
                               timepoint == "day3"	~ "Fluad_72_PBMC",
                               timepoint == "day7"	~ "Fluad_168_PBMC",
                               T~timepoint),
         timepoint = str_remove(timepoint, "day"),
         test = "t-Test",
         paired = "Unpaired")

mouse_fluad_log2fc_paired_unpaired_ttest = mouse_fluad_log2fc_paired_ttest %>% 
  bind_rows(mouse_fluad_log2fc_unpaired_ttest) 

mouse_fluad_log2fc_paired_unpaired_ttest %>% 
  saveRDS(file = here("tables", "influenza_fluad_mouse_log2fc_paired_unpaired_ttest.rds"))

```


## Human

### Download dataset
Data was quantile normalized signal intensity using limma 3.38.3, read.maimages source = agilent.median, backgroundCorrect method= normexp, normalizeBetweenArrays method= quantile. Ref: GEO.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
### Download dataset --------
# Data was quantile normalized signal intensity using limma 3.38.3, read.maimages source = agilent.median, backgroundCorrect method= normexp, normalizeBetweenArrays method= quantile. Ref: GEO

###### GSE124689 ----
#Download dataset
GSE124689 <- getGEO("GSE124689", GSEMatrix = TRUE) 
#Get ExpressionSet
eset <- GSE124689[[1]]
#Expression
GSE124689_exprs <- eset %>% exprs()
#Metadata
GSE124689_metadata = eset %>% pData()
#Save
GSE124689_exprs %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "influenza_human_GSE124689_exprs.rds"))
```

Import data
```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Import data
GSE124689_exprs = readRDS(here("tables", "influenza_human_GSE124689_exprs.rds"))

#Get metadata annotated by Amnah
# cleandata_nonorm_flu <- readRDS(here("HumanFluData", "cleandata_nonorm_flu.rds")) 
flu_expFC <- readRDS(here("HumanFluData", "expFC.rds"))
# flu_expression_fc = flu_expFC[["exp_fc"]] %>% as.data.frame()
flu_metadata = flu_expFC$cur_phenoData
flu_metadata %>% filter(exposure_material_reported == "FLUADD")

#Filter samples by vaccine
human_samples_pbmc_fluad = flu_metadata %>% 
  filter(exposure_material_reported == "FLUADD")

#Check distribution
GSE124689_exprs %>% 
  as.data.frame() %>% 
  select(value = GSM3542845) %>% 
  rownames_to_column("probe") %>% 
  ggplot() +
  aes(x = value) +
  geom_histogram()
```




### Annotate features
```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
# Annotate genes
probes_annotation <- read.delim(here("tables", "GPL21272-58809.txt"), skip = 19, header = TRUE, sep = "\t", quote = "") %>% 
  select(gene_id = ID, entrez_id = ENTREZ_GENE_ID, ensembl_id = ENSEMBL_ID, gene_symbol = GENE_SYMBOL) %>% 
  mutate(gene_id = as.character(gene_id))

GSE124689_counts <- GSE124689_exprs %>%
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>% 
  inner_join(probes_annotation, by = "gene_id") %>% 
  select(gene_id, entrez_id, ensembl_id, gene_symbol, c(human_samples_pbmc_fluad$biosample_accession)) %>% 
  mutate(entrez_id = as.character(entrez_id)) 

#Get probes and entrez IDs with the highest median by ensembl_id
GSE124689_counts_genes_selected = GSE124689_counts %>% 
  pivot_longer(cols = -c(gene_id, entrez_id, ensembl_id, gene_symbol),
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(entrez_id, gene_id) %>% 
  drop_na(entrez_id) %>% 
  summarize(median = median(expression),
            median = round(median, 10)) %>% 
  slice_max(median, n = 1) %>% 
  ungroup()

#Merge
GSE124689_counts_raw = GSE124689_counts %>% 
  inner_join(GSE124689_counts_genes_selected %>% 
               select(gene_id, entrez_id),
             by = join_by("gene_id", "entrez_id"))

#Delete duplicates by sampling
dupes_entrez = GSE124689_counts_genes_selected %>% 
  get_dupes(entrez_id) %>% 
  group_by(entrez_id) %>% 
  slice_sample(n = 1)

#Merge
human_fluadd_not_normalized = GSE124689_counts_raw %>%
  select(-gene_symbol) %>% 
  filter(!gene_id %in% c(dupes_entrez$gene_id)) %>% 
  distinct() %>% 
  select(c(entrez_id, human_samples_pbmc_fluad$uid)) %>% 
  column_to_rownames("entrez_id") 

#Save
human_fluadd_not_normalized %>% saveRDS(file = here("tables", "influenza_GSE124689_exprs.rds"))

#Get symbols only
GSE124689_counts_genes_selected_symbol = GSE124689_counts %>% 
  pivot_longer(cols = -c(gene_id, entrez_id, ensembl_id, gene_symbol),
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(gene_symbol, gene_id) %>% 
  filter(gene_symbol != "") %>% 
  drop_na(gene_symbol) %>% 
  summarize(median = median(expression),
            median = round(median, 10)) %>% 
  slice_max(median, n = 1) %>% 
  ungroup()

GSE124689_counts_raw_symbol = GSE124689_counts %>% 
  inner_join(GSE124689_counts_genes_selected_symbol %>% 
               select(gene_id, gene_symbol),
             by = join_by("gene_id", "gene_symbol"))

dupes_symbols = GSE124689_counts_genes_selected_symbol %>% 
  get_dupes(gene_symbol) %>% 
  group_by(gene_symbol) %>% 
  slice_sample(n = 1)

human_fluadd_not_normalized_symbol = GSE124689_counts_raw_symbol %>%
  filter(!gene_symbol %in% c(dupes_symbols$gene_symbol)) %>% 
  distinct() %>% 
  select(c(gene_symbol, human_samples_pbmc_fluad$uid)) %>% 
  column_to_rownames("gene_symbol")

human_fluadd_symbol_entrez = GSE124689_counts_raw_symbol %>%
  filter(!gene_symbol %in% c(dupes_symbols$gene_symbol)) %>% 
  distinct() %>% 
  select(symbol = gene_symbol, entrez_id)

#Save
human_fluadd_not_normalized_symbol %>% saveRDS(file = here("tables", "influenza_GSE124689_exprs_not_normalized.rds"))
```


```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Counts matrix from Expression set
flu_metadata_clean = flu_metadata %>% 
  arrange(uid) %>% 
  separate(participant_id, into = c("del", "participant_id")) %>% 
  select(-del) %>% 
  select(sample_id = uid, participant_id, timepoint = study_time_collected, age = age_imputed, sex = gender, race, vaccine = exposure_material_reported, sample_type = cell_type, array = featureSetName2, type = vaccine_type, adjuvant) %>% 
  mutate(age = as.numeric(age),
         sex = as.factor(sex), 
         race = as.factor(race),
         condition = paste(vaccine, "Day", timepoint)) %>%
  filter(sample_id %in% c(GSE124689_counts %>% colnames())) %>% 
  arrange(sample_id) %>% 
  select(condition, everything()) 

#Save
saveRDS(flu_metadata_clean, file = here("tables", "influenza_fluad_human_metadata_clean.rds"))
```

### Unite and normalize with human
```{r}
# Unite tables by "humanized" symbols ---------
#Filter out duplicated symbols
filter_out_mgi = GSE120661_counts_not_normalized_log2_annotations %>% 
  select(mgi_symbol) %>% 
  get_dupes()

GSE120661_counts_not_normalized_log2_symbol = GSE120661_counts_not_normalized_log2_annotations %>% 
  select(-probe, -entrezgene_id) %>% #21,856 rows
  distinct() %>%  #21,666 rows
  filter(mgi_symbol != c(filter_out_mgi$mgi_symbol)) %>% #21,557 rows
  mutate(mgi_symbol = str_to_upper(mgi_symbol)) %>% 
  rename(symbol = mgi_symbol)

#Unite and normalize
human_mouse_fluad_matrix = GSE124689_expression_not_normalized_log2_raw_symbol %>% 
  rownames_to_column("symbol") %>% 
  inner_join(GSE120661_counts_not_normalized_log2_symbol, 
             by = "symbol")  %>% 
  column_to_rownames("symbol") %>% 
  select(human_mouse_fluad_metadata %>% 
           # filter(timepoint %in% c("0", "1", "3", "7")) %>% 
           pull(sample_id)) %>% 
  as.matrix() %>% 
  normalizeBetweenArrays(method = "quantile")

human_mouse_fluad_df = human_mouse_fluad_matrix %>% 
  as.data.frame()


#Create ExpressionSet
eset_human_mouse <- ExpressionSet(
  assayData = human_mouse_fluad_matrix,
  phenoData = AnnotatedDataFrame(human_mouse_fluad_metadata %>% 
                                   filter(sample_id %in% c(human_mouse_fluad_matrix %>% colnames())) %>% 
                                   column_to_rownames("sample_id"))
)

eset_human_mouse %>% saveRDS(file = here("tables", "influenza_fluad_human_mouse_eset.rds"))

```


### Differential gene expression
```{r}
#Import
eset_human_mouse = readRDS(file = here("tables", "influenza_fluad_human_mouse_eset.rds"))
human_mouse_fluad_matrix = eset_human_mouse %>% exprs()
human_mouse_fluad_metadata = readRDS(file = here("tables", "influenza_fluad_human_mouse_metadata.rds.png"))

human_mouse_fluad_matrix_human = human_mouse_fluad_matrix %>% 
  as.data.frame() %>% 
  select(c(human_mouse_fluad_metadata %>% 
           filter(organism == "Human") %>% 
             filter(!participant_id %in% c("CRC305DGENT1483036",
                                             "CRC305DGENT0351013",
                                             "CRC305DGENT1471032")) %>% 
             pull(sample_id))) %>% 
  as.matrix()

human_mouse_fluad_metadata_pData = human_mouse_fluad_metadata %>% 
                                   column_to_rownames("sample_id") %>% 
                                   filter(organism == "Human") %>% 
                                   filter(!participant_id %in% c("CRC305DGENT1483036",
                                             "CRC305DGENT0351013",
                                             "CRC305DGENT1471032")) %>% 
  mutate(timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         age = scale(as.numeric(age)),
         race = as.factor(race))
```

#### Limma
```{r}
# DGE -------

#Create ExpressionSet
eset_human <- ExpressionSet(
  assayData = human_mouse_fluad_matrix_human,
  phenoData = AnnotatedDataFrame(human_mouse_fluad_metadata_pData))

#Save
eset_human %>% saveRDS(file = here("tables", "fluad_eset_human.rds"))
```


```{r}
##### Run analysis in the cluster -------
eset_human = readRDS(file = here("tables", "fluad_eset_human.rds"))
#Create design matrix
design <- model.matrix(~ condition + sex + age + race + 0, eset_human) 
#Avoid blank spaces
colnames(design) <- make.names(colnames(design))

#Get intra-individual correlation
corfit <- duplicateCorrelation(eset_human, 
                               design = design, 
                               block = eset_human$participant_id)
corfit$consensus #0.11

# Fit Limma model
fluad_human_limma_fit <- lmFit(eset_human, 
                             design, 
                             block = eset_human$participant_id, 
                             correlation = corfit$consensus)

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

cont.matrix <- makeContrasts(conditionFLUADD.Day.1 - conditionFLUADD.Day.0,
                             conditionFLUADD.Day.3 - conditionFLUADD.Day.0,
                             conditionFLUADD.Day.7 - conditionFLUADD.Day.0,
                             levels=design)

fit2 <- contrasts.fit(fluad_human_limma_fit, cont.matrix)

# eBayes
fit2 <- eBayes(fit2)


fit2 %>% saveRDS(here("tables","fluad_human_limma_fit.rds"))


# Import results from cluster -------
fit2 = readRDS(here("tables","fluad_human_limma_fit.rds"))

#Get DEGs ------
## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs
fluad_human_degs = all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace_all(comparison, "\\.", " "),
        comparison = str_replace_all(comparison, "condition", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  select(condition = comparisonA, everything()) %>% 
  select(-comparisonRef) %>% 
  clean_names() %>% 
  select(condition, symbol, mean = log_fc, lower = ci_l, upper = ci_r, ave_expr,
         everything()) %>% 
  inner_join(human_mouse_fluad_metadata_pData %>% 
               select(condition, timepoint, vaccine, sample_type) %>% 
               distinct(), by = "condition") %>% 
  mutate(organism = "Human") 

#Save
fluad_human_degs %>% saveRDS(file = here("tables", "influenza_fluad_human_dge_limma_degs.rds"))

#Print results

fluad_human_degs %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarize(n_degs = n())
```


#### Simple t-Test
```{r}
eset_human_mouse = readRDS(file = here("tables", "influenza_fluad_human_mouse_eset.rds"))
human_mouse_fluad_matrix = eset_human_mouse %>% exprs()
human_mouse_fluad_metadata = readRDS(file = here("tables", "influenza_fluad_human_mouse_metadata.rds.png"))


human_mouse_fluad_matrix_human = human_mouse_fluad_matrix %>% 
  as.data.frame() %>% 
  select(c(human_mouse_fluad_metadata %>% 
           filter(organism == "Human") %>% 
             filter(!participant_id %in% c("CRC305DGENT1483036",
                                             "CRC305DGENT0351013",
                                             "CRC305DGENT1471032")) %>% 
             pull(sample_id))) %>% 
  as.matrix()

human_mouse_fluad_metadata_pData = human_mouse_fluad_metadata %>% 
                                   column_to_rownames("sample_id") %>% 
                                   filter(organism == "Human") %>% 
                                   filter(!participant_id %in% c("CRC305DGENT1483036",
                                             "CRC305DGENT0351013",
                                             "CRC305DGENT1471032")) %>% 
  mutate(timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         age = scale(as.numeric(age)),
         race = as.factor(race))

#Annotate
human_fluad_log2fc_filtered_sample_long = human_mouse_fluad_matrix_human %>%
  as.data.frame() %>% 
  rownames_to_column("human_symbol") %>% 
  pivot_longer(cols = -human_symbol,
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(human_mouse_fluad_metadata_pData %>% 
               rownames_to_column("geo_sample") %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>%
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")

#Save
human_fluad_log2fc_filtered_sample_long %>% 
  saveRDS(here("tables", "influenza_fluad_human_log2fc_sample_long.rds"))


#Pivot wider
human_fluad_log2fc_filtered_sample_wide = human_fluad_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")

beep()
```


```{r}
#Import
human_fluad_log2fc_filtered_sample_long = 
  readRDS(here("tables", "influenza_fluad_human_log2fc_sample_long.rds"))


#Compute summary statistics
human_fluad_log2fc_summary = human_fluad_log2fc_filtered_sample_long %>% 
  group_by(human_symbol, timepoint) %>%
  summarize(
    avg_log2fc = mean(log2fc, na.rm = TRUE),  # mean log2 fold change
    n_subjects = n(),                         # number of samples
    sd = sd(value, na.rm = TRUE),             # standard deviation
    se = sd / sqrt(n_subjects),               # standard error
    effect_size = avg_log2fc / sd)            # Cohen’s d-like effect size

human_fluad_log2_hgnc_symbol_wide = human_mouse_fluad_matrix_human %>%
  as.data.frame() %>% 
  rownames_to_column("human_symbol") %>% 
  pivot_longer(cols = -human_symbol,
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(human_mouse_fluad_metadata_pData %>% 
               rownames_to_column("geo_sample") %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_")

# Paired t-test
human_fluad_log2fc_paired_ttest <- human_fluad_log2_hgnc_symbol_wide %>%
  group_by(human_symbol) %>%
  
  summarise(
    pval_day1 = tryCatch(t.test(day_1, day_0, paired = TRUE, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_day3 = tryCatch(t.test(day_3, day_0, paired = TRUE, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_day7 = tryCatch(t.test(day_7, day_0, paired = TRUE, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    .groups = 'drop'
  ) %>%
  
  mutate(
    padj_day1 = p.adjust(pval_day1, method = "fdr"),
    padj_day3 = p.adjust(pval_day3, method = "fdr"),
    padj_day7 = p.adjust(pval_day7, method = "fdr")
  ) %>% 
  pivot_longer(
    cols = starts_with(c("pval_", "padj_")),
    names_to = c(".value", "timepoint"),
    names_pattern = "(pval|padj)_(day\\d+)"
  ) %>% 
  mutate(condition = case_when(timepoint == "day1"	~ "Fluad_24_PBMC",
                               timepoint == "day3"	~ "Fluad_72_PBMC",
                               timepoint == "day7"	~ "Fluad_168_PBMC",
                               T~timepoint),
         timepoint = str_remove(timepoint, "day"),
         test = "t-Test",
         paired = "Paired")

# Unpaired t-test ----------
human_fluad_log2fc_unpaired_ttest <- human_fluad_log2_hgnc_symbol_wide %>%
  group_by(human_symbol) %>%
  
  summarise(
    pval_day1 = tryCatch(
      t.test(day_1, day_0, paired = FALSE, var.equal = FALSE, alternative = "two.sided")$p.value,
      error = function(e) NA_real_
    ),
    pval_day3 = tryCatch(
      t.test(day_3, day_0, paired = FALSE, var.equal = FALSE, alternative = "two.sided")$p.value,
      error = function(e) NA_real_
    ),
    pval_day7 = tryCatch(
      t.test(day_7, day_0, paired = FALSE, var.equal = FALSE, alternative = "two.sided")$p.value,
      error = function(e) NA_real_
    ),
    .groups = 'drop'
  ) %>%
  
  mutate(
    padj_day1 = p.adjust(pval_day1, method = "fdr"),
    padj_day3 = p.adjust(pval_day3, method = "fdr"),
    padj_day7 = p.adjust(pval_day7, method = "fdr")
  ) %>% 
  pivot_longer(
    cols = starts_with(c("pval_", "padj_")),
    names_to = c(".value", "timepoint"),
    names_pattern = "(pval|padj)_(day\\d+)"
  ) %>% 
  mutate(
    condition = case_when(
      timepoint == "day1" ~ "Fluad_24_PBMC",
      timepoint == "day3" ~ "Fluad_72_PBMC",
      timepoint == "day7" ~ "Fluad_168_PBMC",
      TRUE ~ timepoint),
    timepoint = str_remove(timepoint, "day"),
    test = "t-Test",
    paired = "Unpaired")




human_fluad_log2fc_paired_unpaired_ttest = human_fluad_log2fc_paired_ttest %>% 
  bind_rows(human_fluad_log2fc_unpaired_ttest) 

human_fluad_log2fc_paired_unpaired_ttest %>% 
  saveRDS(file = here("tables", "influenza_fluad_human_dge_paired_unpaired_ttest.rds"))

```






## Integrate datasets

```{r}
#Sample log2fc
human_fluad_log2fc_filtered_sample_long = readRDS(here("tables", "influenza_fluad_human_log2fc_sample_long.rds"))
mouse_fluad_log2fc_filtered_sample_long = readRDS(here("tables", "influenza_fluad_mouse_log2fc_sample_clean_long.rds"))

human_mouse_log2fc_samples = mouse_fluad_log2fc_filtered_sample_long %>% 
  mutate(organism = "Mouse") %>% 
  bind_rows(human_fluad_log2fc_filtered_sample_long %>% 
              mutate(organism = "Human"))

human_mouse_log2fc_samples %>% 
  saveRDS(file = here("tables", "influenza_fluad_human_mouse_log2fc_samples.rds"))

#Unite metadata
fluad_human_metadata = readRDS(file = here("tables", "influenza_fluad_human_metadata_clean.rds"))


#Unite Mouse and Human datasets. Samples.
human_mouse_log2fc_samples

#Unite DEGs
all_degs_human = readRDS(file = here("tables", "influenza_fluad_human_dge_limma_degs.rds")) 
all_degs_mouse = readRDS(file = here("tables", "influenza_fluad_mouse_dge_limma_degs.rds")) 

all_degs_human_mouse = all_degs_mouse %>% 
  bind_rows(all_degs_human) %>% 
  select(condition:ave_expr, t:organism) %>% 
  rename(human_symbol = symbol)

all_degs_human_mouse %>% saveRDS(here("tables", "influenza_fluad_human_mouse_dge_limma_degs.rds"))
```

## FIT
Ref: Normand, R., Du, W., Briller, M. et al. Found In Translation: a machine learning model for mouse-to-human inference. Nat Methods 15, 1067–1073 (2018). https://doi.org/10.1038/s41592-018-0214-9
Github: https://github.com/shenorrLabTRDF/FIT.mouse2man 

### Run prediction
Import package
```{r}
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
#Import files
pca_rotations = FIT.mouse2man::pca_rotations
best_models = FIT.mouse2man::best_models
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez
MM_Entrez_symbol_desc = FIT.mouse2man::MM_Entrez_symbol_desc
Psoriasis_classifier_neg = FIT.mouse2man::Psoriasis_classifier_neg
slopes_per_gene_V2.0 = FIT.mouse2man::slopes_per_gene_V2.0
StaphInfection_classifier_pos = FIT.mouse2man::StaphInfection_classifier_pos
AllData_V2.0 = FIT.mouse2man::AllData_V2.0 #Training data
```

Obs.: Make sure that datasets from FIT package were retrieved
Initial number of genes: 15503
Number of genes for which FIT predictions will be calculated: 14401


```{r}
all_degs_mouse = readRDS(file = here("tables", "influenza_fluad_mouse_dge_limma_degs.rds")) 

mouse_fluad_log2_hgnc_symbol = readRDS(here("tables", "influenza_fluad_mouse_exprs_hgnc_symbol.rds"))
### Loop for every time point -----
all_degs_fit2 = all_degs_mouse %>% 
  select(symbol, EffectSize = logFC, condition) %>%
  inner_join(mouse_fluad_log2_hgnc_symbol %>% 
               select(symbol = human_symbol, MM.Entrez = mouse_entrez),
             by = "symbol") %>% 
  mutate(row = MM.Entrez)

# Generate csv for every condition
all_conditions <- unique(all_degs_fit2$condition)
purrr::walk(all_conditions, function(cond) {
  
  delete_dupes <- all_degs_fit2 %>%
    filter(condition == cond) %>%
    get_dupes(row) %>%
    pull(row)
  
  filtered_df <- all_degs_fit2 %>%
    filter(condition == cond, !row %in% delete_dupes) %>%
    select(-condition) %>% 
    column_to_rownames("row") %>% 
    select(MM.Entrez, EffectSize)
  
  file_name <- paste0(cond, "_FIT_mouse.csv")
  file_path <- here("tables", file_name)
  
  write_csv(filtered_df, file_path)
})

# FIT loop
fit_results_combined <- map2_dfr(all_conditions, all_conditions, function(cond, cond_name) {
  
  file_path <- here("tables", paste0(cond, "_FIT_mouse.csv"))
  
  fit_output <- FIT.mouse2man::FIT(MouseFile = file_path, DataType = "rnaseq")

  fit_output_tbl <- as_tibble(fit_output) %>%
    mutate(condition = cond_name)
  
  return(fit_output_tbl)
})
beep()
#Print
fit_results_combined

#Save
fit_results_combined %>% saveRDS(here("tables", "influenza_fluad_mouse_fit_prediction_alltimes.rds"))


#Visualize distribution
fit_results_combined %>% 
  ggplot() +
  aes(x = FIT_prediction,
      fill = condition,
      color = condition) +
  geom_histogram() +
  theme_vaxgo()
```

Import
```{r fig.height=5, fig.width=5}
#Import
fit_results_combined = read_rds(here("tables", "influenza_fluad_mouse_fit_prediction_alltimes.rds"))

#Pivot longer
fit_example_results = fit_results_combined %>% 
  clean_names() %>%  
  pivot_longer(cols = c(mouse_effect_size, fit_prediction),
               names_to = "measure", 
               values_to = "value")


fit_example_results_ready = fit_example_results %>% 
  arrange(mouse_symbol) %>% 
  select(condition, everything())

#Save
fit_example_results_ready %>% 
  saveRDS(file = here("tables", "influenza_fluad_mouse_fit_prediction_example_results_ready.rds"))
```


### Integrate Mouse, FIT, and Human data
```{r}
#Import datasets
fit_example_results_ready = readRDS(file = here("tables", "influenza_fluad_mouse_fit_prediction_example_results_ready.rds"))
fluad_human_degs = read_rds(file = here("tables", "influenza_fluad_human_dge_limma_degs.rds"))
all_degs_mouse = read_rds(file = here("tables", "influenza_fluad_mouse_dge_limma_degs.rds"))
GSE120661_metadata = read_csv(here("tables", "influenza_hepatitisb_GSE120661_metadata.csv")) 
human_mouse_fluad_metadata = readRDS(file = here("tables", "influenza_fluad_human_mouse_metadata.rds"))

#Clean tables
human_mouse_fluad_metadata_datasets = human_mouse_fluad_metadata %>% 
  select(vaccine, sample_type, timepoint, condition, organism, type, array) %>% 
  distinct() %>% 
  mutate(timepoint = as.numeric(timepoint))


# Annotate genes by entrez_id ------
#Get gene annotation
genes_ann = fit_example_results_ready %>% 
  select(mouse_entrez, human_entrez,human_symbol) %>% 
  distinct() %>% 
  mutate(human_entrez = as.character(human_entrez))

#Human
fluad_human = fluad_human_degs %>% 
  select(condition,
         human_symbol = symbol, 
         log2fc = mean,
         ci_low = lower,
         ci_high = upper,
         p_value,
         adj_p_val,
         organism,
         t
         ) %>% 
  mutate(ci_size = ci_high - ci_low,
         organism = "Human") %>% 
  inner_join(genes_ann, by = join_by("human_symbol")) %>% 
  rename(symbol = human_symbol)

#Mouse
fluad_mouse = all_degs_mouse %>%
  select(condition,
         symbol, 
         log2fc = logFC,
         ci_low = CI.L,
         ci_high = CI.R,
         p_value = P.Value,
         adj_p_val = adj.P.Val,
         t,
         ) %>% 
  mutate(ci_size = ci_high - ci_low,
         organism = "Mouse")

#FIT
fluad_fit = fit_example_results_ready %>% 
  filter(measure == "fit_prediction") %>% 
  select(condition, 
         mouse_entrez, 
         human_entrez,
         mouse_symbol,
         symbol = human_symbol, 
         log2fc = value,
         ci_low,
         ci_high,
         ) %>% 
  mutate(organism = "FIT",
         ci_size = ci_high - ci_low) %>% 
  #Get only the human entrez id used
  filter(human_entrez %in% c(fluad_human$human_entrez)) %>% 
  select(!c(mouse_entrez, human_entrez, mouse_symbol)) %>% 
  #Calculate pvalues
   mutate(
  se = (ci_high - ci_low) / (2 * 1.96),
  z = log2fc / se,
  p_value = 2 * pnorm(-abs(z)),
  adj_p_val = p.adjust(p_value, method = "BH")) %>% 
  select(-se)

#Combine tables
fit_fluad_mouse_human = fluad_human %>% 
  bind_rows(fluad_mouse, 
            fluad_fit) %>% 
  inner_join(human_mouse_fluad_metadata_datasets %>% 
               select(condition, vaccine, timepoint, type),
             by = "condition") %>% 
  arrange(symbol) %>% 
  filter(timepoint %in% c(1, 3, 7)) %>% 
  select(!c(human_entrez, mouse_entrez))

#Save
fit_fluad_mouse_human %>% 
  saveRDS(here("tables", "influenza_fluad_human_mouse_fit_dge_degs.rds"))

#Save long dataset
fit_fluad_mouse_human %>% 
  mutate(condition =  paste(organism, "Day", timepoint)) %>% 
  saveRDS(file = here("tables", "influenza_fluad_human_mouse_fit_dge_degs_long.rds"))

#Print
fit_fluad_mouse_human

```


6373 human genes were analyzed

Filtered:
- 5563 genes with padj > 0.10
- 3673 genes with padj > 0.50
- 1001 genes with padj > 0.90

```{r fig.height=5, fig.width=5}
#Make the table wider manually -----

#FIT
meta_fit = fit_fluad_mouse_human %>% 
  filter(organism == "FIT") %>% 
  select(symbol, 
         timepoint, 
         z_fit = z,
         mean_fit = log2fc, 
         ci_low_fit = ci_low,
         ci_high_fit = ci_high,
         ci_width_fit = ci_size,
         pval_adj_fit = adj_p_val,
         pval_fit = p_value
         )

#Mouse
meta_mouse = fit_fluad_mouse_human %>% 
  filter(organism == "Mouse") %>% 
  select(symbol, 
         timepoint, 
         t_mouse = t,
         mean_mouse = log2fc, 
         ci_low_mouse = ci_low,
         ci_high_mouse = ci_high,
         ci_width_mouse = ci_size,
         pval_adj_mouse = adj_p_val,
         pval_mouse = p_value
         ) %>% 
  #Filter out repeated values
  arrange(symbol, timepoint) %>% 
  mutate(mean_mouse_round =  mean_mouse %>% round(3),
           value = (ci_high_mouse + ci_low_mouse) / 2,
         value = value %>% round(3)) %>% 
  mutate(test = if_else(value == mean_mouse_round, "T", "no")) %>% 
  filter(test == "T") %>% 
  select(-c(test, value, mean_mouse_round))

#Human
meta_human = fit_fluad_mouse_human %>% 
  filter(organism == "Human") %>% 
  select(symbol, 
         timepoint, 
         t_human = t,
         mean_human = log2fc, 
         ci_low_human = ci_low,
         ci_high_human = ci_high,
         ci_width_human = ci_size,
         pval_adj_human = adj_p_val,
         pval_human = p_value)

#Unite and calculate summary statistics
pairwise_human_fit_measures = meta_human %>% 
  inner_join(meta_fit, by = join_by(symbol, timepoint)) %>% 
  inner_join(meta_mouse, by = join_by(symbol, timepoint)) %>% 
  mutate(human_2_mouse = mean_human - mean_mouse,
         human_2_fit = mean_human - mean_fit) %>% 
  group_by(timepoint) %>%
  ungroup() %>% 
  
    mutate(
    # Estimate SE from CIs
    se_human = (ci_high_human - ci_low_human) / (2 * 1.96),
    se_fit = (ci_high_fit - ci_low_fit) / (2 * 1.96),
    se_mouse = (ci_high_mouse - ci_low_mouse) / (2 * 1.96),
    
    # Combine SE
    se_diff_h2fit = sqrt(se_human^2 + se_fit^2),
    se_diff_h2m = sqrt(se_human^2 + se_mouse^2),
    
    # Z score
    z_score_h2fit = human_2_fit / se_diff_h2fit,
    z_score_h2m = human_2_mouse / se_diff_h2m,
    
    # p-value
    p_value_h2fit = 2 * (1 - pnorm(abs(z_score_h2fit))),
    p_value_h2m = 2 * (1 - pnorm(abs(z_score_h2m))),
    
    # FDR (Benjamini-Hochberg)
    p_value_h2fit_adj = p.adjust(p_value_h2fit, method = "BH"),
    p_value_h2m_adj = p.adjust(p_value_h2m, method = "BH"),
    
    # Cohen's d
    effect_size_h2fit = human_2_fit / se_diff_h2fit,
    effect_size_h2m = human_2_mouse / se_diff_h2m)  %>% 
  
  
  #Reorder cols
  select(
    timepoint, symbol,
    #Human
    mean_human, t_human,
    ci_low_human, ci_high_human, ci_width_human,
    se_human, pval_human, pval_adj_human,
    #FIT
    mean_fit, z_fit,
    ci_low_fit, ci_high_fit, ci_width_fit,
    se_fit, pval_fit, pval_adj_fit,
    #Mouse
    mean_mouse, t_mouse,
    ci_low_mouse, ci_high_mouse, ci_width_mouse,
    se_mouse, pval_mouse, pval_adj_mouse,
    #Comparison
    human_2_mouse,
    human_2_fit,
    se_diff_h2fit, se_diff_h2m,
    z_score_h2fit, z_score_h2m,
    p_value_h2fit, p_value_h2fit_adj,
    p_value_h2m, p_value_h2m_adj,
    effect_size_h2fit, effect_size_h2m
  )

#Save wide dataset
pairwise_human_fit_measures %>% 
  saveRDS(here("tables", "influenza_fluad_human_mouse_fit_dge_pairwise2human.rds"))

#Print
pairwise_human_fit_measures
```


# 2. Engerix B, Vaccine

## Mouse

### Prepare data
```{r}
#Import
GSE120661_counts_not_normalized_log2 = readRDS(here("tables", "GSE120661_counts_not_normalized_log2.rds")) %>% 
  as.data.frame()
GSE120661_metadata = read_csv(file = here("tables", "influenza_hepatitisb_GSE120661_metadata.csv"))

#Filter data
mouse_engerixb_metadata = GSE120661_metadata %>% 
  filter(treatment == "Engerix B",
         tissue == "PBMC") %>% 
  mutate(timepoint = (timepoint / 24) %>% round(1),
         participant_id = paste0("ENG_BL_", animal_replicate)) %>% 
  filter(timepoint %in% c(0, 1, 3, 7))

mouse_engerixb_metadata %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))

#Get log2 values
mouse_engerixb_log2 = GSE120661_counts_not_normalized_log2 %>% 
  select(mouse_engerixb_metadata %>% pull(geo_sample))

```

Annotate genes

```{r}
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- mouse_engerixb_log2 %>% rownames()

#Get other annotations
listAttributes(ensembl_mart)

#Annotate
annot <- getBM(
  attributes = c("agilent_sureprint_g3_ge_8x60k", #Agilent microarray data (design ID 028005, retrieved from GEO)
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "agilent_sureprint_g3_ge_8x60k",
  values = gene_id,
  mart = ensembl_mart)

beep()

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_engerixb_log2summary = mouse_engerixb_log2 %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = agilent_sureprint_g3_ge_8x60k), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = mouse_engerixb_log2summary %>% 
  #filter ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

mouse_engerixb_log2summary = mouse_engerixb_log2summary %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
mouse_engerixb_log2_filtered_probes = mouse_engerixb_log2 %>% 
  rownames_to_column("probe") %>% 
  inner_join(mouse_engerixb_log2summary, by = "probe") %>%   #214,396
  select(probe, mouse_entrez, human_symbol = hgnc_symbol, c(mouse_engerixb_metadata %>% pull(geo_sample)))
#14395
mouse_engerixb_log2_filtered %>% dim()



mouse_engerixb_log2_hgnc_symbol = mouse_engerixb_log2_filtered_probes %>% 
  select(human_symbol, c(mouse_engerixb_metadata$geo_sample)) %>% 
  column_to_rownames("human_symbol")


#Save 
mouse_engerixb_log2_hgnc_symbol %>% 
  saveRDS(here("tables", "hepatitisb_engerixb_mouse_exprs.rds"))
```

Create expression set

```{r}
#Import
mouse_engerixb_log2_hgnc_symbol = readRDS(here("tables", "hepatitisb_engerixb_mouse_exprs.rds"))
mouse_engerixb_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))

#Create Expression Set
eset_mouse_engerixb <- ExpressionSet(
  assayData = mouse_engerixb_log2_hgnc_symbol %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(mouse_engerixb_metadata %>% 
                                   column_to_rownames("geo_sample")))

#Save
eset_mouse_engerixb %>% saveRDS(file = here("tables", "hepatitisb_engerixb_mouse_eset.rds"))
```


```{r}
#Annotate
mouse_engerixb_log2fc_filtered_sample_long = mouse_engerixb_log2_hgnc_symbol %>%
  rownames_to_column("human_symbol") %>% 
  pivot_longer(cols = -human_symbol,
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(mouse_engerixb_metadata %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>%
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")

#Save
mouse_engerixb_log2fc_filtered_sample_long %>% 
  saveRDS(here("tables", "mouse_engerixb_log2fc_filtered_sample_long.rds"))


#Pivot wider
mouse_engerixb_log2fc_filtered_sample_wide = mouse_engerixb_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")

beep()
```


#### Differential expression (Limma)
Paired samples design
```{r}

#Import
mouse_engerixb_log2_hgnc_symbol = readRDS(here("tables", "hepatitisb_engerixb_mouse_exprs.rds"))
mouse_engerixb_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))
eset_mouse = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_eset.rds"))

#Create model matrix
design <- model.matrix(~ condition + 0, # 0 defines the way we compare the samples
                       eset_mouse) 
colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

# Estimate the correlation between measurements made on the same subject (paired-samples)
corfit <- duplicateCorrelation(eset_mouse, 
                               design = design, 
                               block = eset_mouse$participant_id)
corfit$consensus #-0.02688727

#Fit Limma model
fit <- lmFit(eset_mouse, 
             design, 
             block = eset_mouse$participant_id, 
             correlation = corfit$consensus) #Use inter-subject correlation as input into the linear model fit

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(conditionEngerix.B_24_PBMC	 - conditionEngerix.B_0_PBMC,
                             conditionEngerix.B_72_PBMC	 - conditionEngerix.B_0_PBMC,
                             conditionEngerix.B_168_PBMC	 - conditionEngerix.B_0_PBMC,
                             levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_mouse_engerixb = all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  mutate(condition = case_when(comparisonA == "conditionEngerix.B_24_PBMC"	~ "Engerix_24_PBMC",
                               comparisonA == "conditionEngerix.B_72_PBMC"	~ "Engerix_72_PBMC",
                               comparisonA == "conditionEngerix.B_168_PBMC"	~ "Engerix_168_PBMC",
                               T~comparisonA),
         organism = "Mouse") %>% 
  select(condition, everything()) %>% 
  clean_names() %>% 
  rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs_mouse_engerixb %>% saveRDS(file = here("tables", "hepatitisb_engerixb_mouse_degs_mouse.rds"))

all_degs_mouse_engerixb %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarize(n_degs = n())

```



#### Correlation plots, pairwise

Plot correlations between mouse samples across all timepoints using
pairwise scatterplots and density plots from the GGally package.

To do this, define custom plotting functions for densities on the
diagonal and points on the lower panels, then apply ggpairs to the
filtered log2 fold change expression matrix. Reorder the timepoints as
factors, color and fill by timepoint, and apply a custom theme along
with manual color scales.

The result is a matrix of correlation plots titled “All days,
HepB/EngerixB, GSE120661, C57BL/6J”.

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(mouse_engerixb_log2fc_filtered_sample_wide %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, HepB/EngerixB, GSE120661, C57BL/6J", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) 

mouse_correlation_ggpairs_alldays

# Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "hepatitisb_engerixb_mouse_samples_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)
beep()
```

#### PCA

Perform a PCA on the **mouse** EngerixB dataset using log2 fold change
values from each **sample**.

Merge metadata with expression values, reshape the data into a wide
matrix with genes as rows and samples as columns, and replace missing or
non-finite values with zeros.

Run PCA on the transposed matrix to capture sample-level variation, then
extract the first two principal components and merge them back with
sample annotations.

Finally, plot PC1 versus PC2, coloring and filling by timepoint and
labeling by participant, with custom color scales and a themed layout.

```{r fig.height=4, fig.width=5}

#Import
mouse_engerixb_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))
mouse_engerixb_log2fc_filtered_sample_long = readRDS(here("tables", "mouse_engerixb_log2fc_filtered_sample_long.rds"))

#Input
samples_metadata_pca_input = mouse_engerixb_metadata
samples_log2fc_values_pca_input = mouse_engerixb_log2fc_filtered_sample_long


#Import metadata
annotation_samples_pca = samples_metadata_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = samples_log2fc_values_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, human_symbol, log2fc) %>% 
  filter(!human_symbol == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("human_symbol") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Mouse, HepB, EngerixB, GSE120661",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

beep()

```







#### BTMs

Annotate mouse genes with Blood Transcription Module (BTM) functional
groups and use them to assess correlations across timepoints. Load
pre-curated BTM annotations and clean the table of functional groups,
then merge it with an external list of immune-related genes. Map mouse
log2 fold change values to BTM processes by joining gene symbols with
their corresponding modules. For each participant and timepoint, compute
the mean log2 fold change per process, reshape the data into a wide
format, and attach module group information.

```{r}
#Import files
btm_annotation_genes =
  read_csv(file = here("tables", "btm_annotation_genes.csv"))
mouse_engerixb_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))
mouse_engerixb_log2fc_filtered_sample_long = readRDS(here("tables", "mouse_engerixb_log2fc_filtered_sample_long.rds"))

#Annotate genes by BTM
samples_genes_btm = mouse_engerixb_log2fc_filtered_sample_long %>% 
  ungroup() %>% 
  select(symbol = human_symbol, participant_id, timepoint, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = samples_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process

dge_btm_mean_process_mouse = dge_btm_mean_process

dge_btm_mean_process_mouse %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_mouse_gsea_btm_mean.rds"))
```



##### Using GGPairs

Using this summarized BTM-level expression, generate pairwise
correlation plots with ggpairs, showing scatterplots and density plots
across modules, colored by timepoint. Add reference lines at zero, apply
custom color scales, and use a themed layout.

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

mouse_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple"))

mouse_correlation_ggpairs_btms_gsea_alldays

mouse_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "hepatitisb_engerixb_mouse_gsea_samples_btm_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)

```

##### Using ggplot

Compute pairwise correlations of BTM module expression for five mouse
participants at a selected timepoint. Create unique participant pairs,
calculate Pearson correlations and R², and plot scatterplots with
regression lines, colored by module group, faceted by participant pairs,
and labeled with correlation values.

```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(3:7) %>% 
  colnames()

#Inputs
timepoint_selected = 7


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == timepoint_selected) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == timepoint_selected) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-1, 1) +
  xlim(-1, 1) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste0("Mouse, Hepatitis B, EngerixB Mouse, Day", timepoint_selected))
```

## Human

Download directly from GEO. The data is too large.

#### Prepare data

```{r}
#EngerixB: GSE124533 -----
filename = "GSE124533"
GSE124533 <- getGEO(filename = here("data_studies", "GSE124533_series_matrix.txt"), GSEMatrix = F)

#Metadata ------
GSE124533_metadata = GSE124533@phenoData %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    participant_id = "participant:ch1",
    sex = "gender:ch1",
    treatment = "treatment:ch1",
    ethnicity = "ethnicity:ch1",
    race = "race:ch1",
    age = "age:ch1",
    timepoint = "day:ch1") %>% 
  mutate(condition = treatment,
         organism = "Human",
         time_unit = "Day")
#Save
GSE124533_metadata %>% saveRDS(file = here("tables", "hepatitisb_engerixb_human_GSE124533_metadata.rds"))

#Import
GSE124533_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_human_GSE124533_metadata.rds"))


#Filter metadata
GSE124533_metadata_filtered = GSE124533_metadata %>%
  filter(treatment %in% c("ENGERIXB1"),
         timepoint %in% c("0", "1" , "3", "7"))

#Save
GSE124533_metadata_filtered %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_metadata.rds"))


#Counts matrix
GSE124533_exprs = GSE124533@assayData[["exprs"]] %>% as.data.frame()

GSE124533_exprs_engerixb = GSE124533_exprs %>% 
  select(c(GSE124533_metadata_filtered$geo_sample))
#Save
GSE124533_exprs_engerixb %>% saveRDS(file = here("tables", "GSE124533_exprs_engerixb.rds"))
```



Import
```{r}
#Import
human_exprs_engerixb = readRDS(file = here("tables", "GSE124533_exprs_engerixb.rds"))
human_engerixb_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_human_metadata.rds"))

########### Filter and annotate probes -----
probes_annotation <- read.delim(here("tables", "GPL21272-58809.txt"), skip = 19, header = TRUE, sep = "\t", quote = "") %>% 
  select(gene_id = ID, entrez_id = ENTREZ_GENE_ID, ensembl_id = ENSEMBL_ID, gene_symbol = GENE_SYMBOL) %>% 
  mutate(gene_id = as.character(gene_id))


#Get annotations for each probe 
### Filter probes with highest median expression
human_engerix_log2summary = human_exprs_engerixb %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(probes_annotation %>% 
              rename(probe = gene_id,
                     entrezgene_id = entrez_id,
                     human_symbol = gene_symbol), 
            by = "probe") %>% 
  #Filter ensembl id
  group_by(ensembl_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() 

#Filter out genes with multiple hgnc_symbols
hgnc_symbol_dupes = human_engerix_log2summary %>% 
  #filter ensembl id with multiple hgnc_symbols
  get_dupes(human_symbol)

human_engerix_log2summary = human_engerix_log2summary %>% 
  filter(!human_symbol %in% c(hgnc_symbol_dupes$human_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
human_engerix_log2_filtered_probes = human_exprs_engerixb %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(human_engerix_log2summary, by = "probe") %>%   #214,396
  select(human_symbol,
         c(human_engerixb_metadata$geo_sample))
#14395
human_engerix_log2_filtered_probes %>% dim()

human_engerix_log2_hgnc_symbol = human_engerix_log2_filtered_probes 

human_engerix_log2_hgnc_symbol %>% 
  get_dupes(human_symbol)

#Save 
human_engerix_log2_hgnc_symbol %>% 
  saveRDS(here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol.rds"))
```




```{r}

#Import
human_engerix_log2_hgnc_symbol = readRDS(here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol.rds"))
human_metadata_filtered = readRDS(file = here("tables", "hepatitisb_engerixb_human_metadata.rds"))

#Get value from time 0 and calculate Log2FC
human_engerixb_log2fc_filtered_sample_long = human_engerix_log2_hgnc_symbol %>%
  ungroup() %>% 
  pivot_longer(cols = -(human_symbol),
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(human_metadata_filtered %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) 

# #Get over duplicated symbols
# overduplicated = human_engerixb_log2fc_filtered_sample_long %>% 
#   select(human_symbol) %>% 
#   group_by(human_symbol) %>% 
#   dplyr::count() %>% 
#   arrange(-n) 

#Calculate log2fc
human_engerixb_log2fc_filtered_sample_long = human_engerixb_log2fc_filtered_sample_long %>%
  # filter(!human_symbol %in% c(overduplicated %>% 
  #                               ungroup() %>% 
  #                               filter(n > median(n)) %>% 
  #                               pull(human_symbol))) %>% 
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>% 
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")


#Annotate sample and pivot longer
human_engerixb_log2fc_filtered_sample_long_annotated = human_engerixb_log2fc_filtered_sample_long %>% 
  select(-value, -day0) %>% 
  inner_join(GSE124533_metadata_filtered %>% 
               select(sample_id = geo_sample,
                      participant_id, timepoint), by = join_by("participant_id", "timepoint"))

#Save
human_engerixb_log2fc_filtered_sample_long_annotated %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol_long.rds"))

beep()

#Pivot wider
human_engerixb_log2fc_filtered_sample_wide = human_engerixb_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")
  
```


Create expression set
```{r}
#Import
human_engerix_log2_hgnc_symbol = readRDS(here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol.rds"))
human_metadata_filtered = readRDS(file = here("tables", "hepatitisb_engerixb_human_metadata.rds"))

human_metadata_filtered = human_metadata_filtered %>% 
  mutate(age = str_remove(age, "y") %>% as.numeric() %>% scale(),
         timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         race = as.factor(race))

#Create Expression Set
eset_human <- ExpressionSet(
  assayData = human_engerix_log2_hgnc_symbol %>% 
    column_to_rownames("human_symbol") %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(human_metadata_filtered))

#Save
eset_human %>% saveRDS(file = here("tables", "hepatitisb_engerixb_human_eset.rds"))
```

### Differential expression


##### Limma
```{r}
#Import
human_engerix_log2_hgnc_symbol = readRDS(here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol.rds"))
human_metadata_filtered = readRDS(file = here("tables", "hepatitisb_engerixb_human_metadata.rds"))
eset_human = readRDS(file = here("tables", "hepatitisb_engerixb_human_eset.rds"))

#Create design matrix
design <- model.matrix(~ timepoint + sex + age + race + 0, eset_human) 

#Avoid blank spaces
colnames(design) <- make.names(colnames(design))

#Get intra-individual correlation
corfit <- duplicateCorrelation(eset_human, 
                               design = design, 
                               block = eset_human$participant_id)
corfit$consensus #0.22

# Fit
engerix_human_limma_fit <- lmFit(eset_human, 
                                 design, 
                                 block = eset_human$participant_id, 
                                 correlation = corfit$consensus)

# Contrasts
cont.matrix <- makeContrasts(
  timepoint1 - timepoint0,
  timepoint3 - timepoint0,
  timepoint7 - timepoint0,
  levels = design
)

fit2 <- contrasts.fit(engerix_human_limma_fit, cont.matrix)

# eBayes
fit2 <- eBayes(fit2)

# Extract
results_list <- lapply(1:ncol(fit2), function(i) {
  topTable(fit2, coef = i, adjust.method = "fdr", number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>%
    rownames_to_column("symbol")
})


# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs
all_degs_human = all_contrasts_degs %>% 
  #Clean table
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  mutate(condition = case_when(comparisonA == "Time1"	~ "EngerixB_Day1",
                               comparisonA == "Time3"	~ "EngerixB_Day3",
                               comparisonA == "Time7"	~ "EngerixB_Day7",
                               T~comparisonA),
         organism = "Human") %>% 
  select(condition, everything()) %>% 
  clean_names() %>% 
  rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)

all_degs_human %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarize(n_degs = n())
	
#Save
all_degs_human %>% saveRDS(file = here("tables", "hepatitisb_engerixb_human_dge_limma_degs.rds"))
```




#### Correlation plots, pairwise

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Sample 5 participants
set.seed(3141)
sample_participants_5 = GSE124533_metadata_filtered %>% 
  select(participant_id) %>% 
  distinct() %>% 
  slice_sample(n = 5) %>% 
  pull()

#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(human_engerixb_log2fc_filtered_sample_wide %>% 
                                              select(human_symbol, timepoint, c(sample_participants_5)) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, HepB/EngerixB, GSE124533, Human", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) 

mouse_correlation_ggpairs_alldays
# Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "hepatitisb_engerixb_human_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)
beep()
```

#### PCA

```{r fig.height=4, fig.width=5}

GSE124533_counts_log2_summary_annotated_probes =
  readRDS(file = here("tables", "GSE124533_counts_log2_summary_annotated_probes.rds"))

#Input
samples_metadata_pca_input = GSE124533_metadata_filtered
samples_log2fc_values_pca_input = human_engerixb_log2fc_filtered_sample_long_annotated


#Import metadata
annotation_samples_pca = samples_metadata_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = samples_log2fc_values_pca_input %>% 
  ungroup() %>% 
  select(human_symbol, participant_id, timepoint, log2fc) %>%
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, human_symbol, log2fc) %>% 
  filter(!human_symbol == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("human_symbol") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Human, HepB, EngerixB, GSE124533",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

beep()

```

#### BTMs

```{r}

#Import files
btm_annotation_genes =
  read_csv(file = here("tables", "btm_annotation_genes.csv"))
human_engerixb_log2fc_filtered_sample_long_annotated = 
  readRDS(file = here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol_long.rds"))
GSE124533_metadata_filtered =
  readRDS(file = here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol.rds"))


#Input
samples_metadata_pca_input = GSE124533_metadata_filtered
samples_log2fc_values_pca_input = human_engerixb_log2fc_filtered_sample_long_annotated

#Annotate genes by BTM
samples_genes_btm = samples_log2fc_values_pca_input %>% 
  ungroup() %>% 
  select(symbol = human_symbol, participant_id, timepoint, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = samples_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process_human = dge_btm_mean_process

dge_btm_mean_process_human %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_gsea_btm_mean.rds"))
```

##### GGPairs

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

human_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = sample_participants_5,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days, BTM"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple"))

human_correlation_ggpairs_btms_gsea_alldays

human_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "hepatitisb_engerixb_human_gsea_btm_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)

```

##### Ggplot manual

```{r fig.height=8, fig.width=10}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#Input
timepoint_selected = 7

#Plot table
plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == timepoint_selected) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == timepoint_selected) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove identical pairs
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))

#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-1, 1) +
  xlim(-1, 1) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste("Human, Hepatitis B, EngerixB Human, Day", timepoint_selected))
```


## Combine datasets
```{r}
#Unite Mouse and Human datasets. Samples.
human_mouse_log2fc_samples = 
#Mouse
mouse_engerixb_log2fc_filtered_sample_long %>%
  ungroup() %>% 
  select(human_symbol, log2fc, participant_id,
         timepoint, value, day0) %>% 
  mutate(organism = "Mouse") %>% 
  
  #Human
  bind_rows(human_engerixb_log2fc_filtered_sample_long %>%
  ungroup() %>% 
  select(human_symbol, log2fc, participant_id,
         timepoint,  value, day0) %>% 
  mutate(organism = "Human"))

#Unite metadata
human_mouse_engerix_metadata = 
  #Human
  GSE124533_metadata_filtered %>% 
  select(geo_sample, participant_id, sex, organism, timepoint) %>% 
  mutate(timepoint = as.numeric(timepoint),
         sex = str_to_title(sex)) %>% 
  #Mouse
  bind_rows(
    mouse_engerixb_metadata %>%
      select(geo_sample, participant_id, sex, timepoint) %>%
      mutate(organism = "Mouse"))

human_mouse_log2fc_samples %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_exprs_samples_log2fc.rds"))

human_mouse_engerix_metadata %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_metadata.rds"))


all_degs_human = readRDS(file = here("tables", "hepatitisb_engerixb_human_dge_limma_degs.rds")) %>% 
  mutate(timepoint = str_remove(comparison_a, "Time"))
all_degs_mouse = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_degs_mouse.rds")) %>% 
  mutate(timepoint = str_remove(condition, "Engerix_") %>% str_remove("_PBMC"))
all_degs_human_mouse = all_degs_mouse %>% 
  bind_rows(all_degs_human) %>% 
  select(condition, timepoint, human_symbol, mean_l2fc:organism)
all_degs_human_mouse %>% saveRDS(here("tables", "hepatitisb_engerixb_human_mouse_dge_limma_degs.rds"))
```

## FIT

### Run prediction

```{r}
#Get log2fc mean
GSE19668_input2FIT = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
    filter(timepoint == 2) %>% 
  select(participant_id, MM.Entrez = entrezgene_id, EffectSize = log2fc) %>% 
  group_by(MM.Entrez) %>% 
  summarize(mean_log2fc = mean(EffectSize))

#Save
GSE19668_input2FIT %>% 
  write_csv(file = here("FIT_Mouse2man", "GSE19668_day2_FIT_mouse.csv"))

#Run FIT
GSE19668_input2FIT_day2 = FIT.mouse2man::FIT(MouseFile = here("FIT_Mouse2man", "GSE19668_day2_FIT_mouse.csv"), DataType = "rnaseq")
```

### Unite log2FC values from mouse, human, and FIT

```{r fig.height=5, fig.width=8}
#Bind rows
mouse_human_fit_log2fc = bind_rows(GSE19668_input2FIT_day2 %>%
                                     #Get DEGs for FIT
                                     filter((CI_low != 0 | CI_high != 0) & !is.na(CI_low) & !is.na(CI_high)) %>%  
                                   select(human_symbol = Human.symbol,
                                          log2fc = FIT_prediction) %>%
                                     mutate(organism = "FIT"),
                                   
                                   #Get DEGs for Human
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_human <= 0.05) %>% 
                                     select(human_symbol, log2fc = Human) %>%
                                     mutate(organism = "Human"),
                                   
                                   #Get DEGs for Mouse
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_mouse <= 0.05) %>% 
                                     select(human_symbol, log2fc = Mouse) %>%
                                     mutate(organism = "Mouse"))


#Pivot wider
mouse_human_fit_log2fc_wide = GSE19668_input2FIT_day2 %>%
            select(human_symbol = Human.symbol,
                   FIT_prediction) %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Human),
            by = "human_symbol") %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Mouse),
            by = "human_symbol")

#Save
mouse_human_fit_log2fc %>% 
  saveRDS(file = here("tables", "mouse_human_fit_log2fc_long_SAureus.rds"))

mouse_human_fit_log2fc_wide %>% 
  saveRDS(file = here("tables", "mouse_human_fit_log2fc_wide_SAureus.rds"))
```

### Distribution

Import tables

```{r fig.height=5, fig.width=8}
#Import
mouse_human_fit_log2fc = readRDS(file = here("tables", "mouse_human_fit_log2fc_long_SAureus.rds"))
mouse_human_fit_log2fc_wide = readRDS(file = here("tables", "mouse_human_fit_log2fc_wide_SAureus.rds"))

GSE33341_metadata = readRDS(file = here("tables", "GSE33341_metadata.rds"))
GSE19668_metadata = readRDS(file = here("tables", "GSE19668_metadata.rds"))
```

Plot histogram

```{r fig.height=5, fig.width=8}
#Plot histogram
mouse_human_fit_log2fc_histogram = mouse_human_fit_log2fc %>% 
  ggplot() +
  aes(x = log2fc,
      fill = organism) +
  geom_density(aes(y = after_stat(count),
                   color = organism),
                  position = position_identity(), 
               alpha = 0.1,
               size = 1,
              show.legend = T) +
  scale_fill_manual(values = c(FIT = "#4361ee",
                               Mouse = "#4DBBD5FF",
                               Human = "#546E7AFF"))+
  scale_color_manual(values = c(FIT = "#4361ee",
                                Mouse = "#4DBBD5FF",
                                Human = "#546E7AFF"))+
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        legend.key = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0),
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(limits = c(-4, 4)) +
  labs(color = "",
       fill = "")

mouse_human_fit_log2fc_histogram

#Faceted
mouse_human_fit_log2fc_histogram_faceted = mouse_human_fit_log2fc_histogram +
  facet_wrap(~organism,
             scales = "free_y")


#Plot scatter plot
mouse_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = Mouse,
      y = Human) +
#All DEGs
geom_point(color = "#4DBBD5FF",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "Mouse L2FC",
         y = "Human L2FC")

mouse_human_scatterplot


fit_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = FIT_prediction,
      y = Human) +
#All DEGs
geom_point(color = "#4361ee",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "FIT L2FC",
         y = "Human L2FC")

#Histograms
(((mouse_human_fit_log2fc_histogram + mouse_human_fit_log2fc_histogram_faceted) + 
    plot_layout(widths = c(0.5, 1), guides = "collect") +
    plot_annotation(theme = theme(legend.position = "top")))  /
    #Scatter plots
    (mouse_human_scatterplot + fit_human_scatterplot)) +
  plot_layout(heights = c(0.5, 1))

```


# 3. S. aureus
## Mouse

### Retrieve dataset
Retrieve Exprs and Metadata from GEO
```{r}
#S aureus: GSE19668 -----
filename = "saureus_infection_GSE19668"
GSE19668 <- getGEO("GSE19668", GSEMatrix = TRUE) 
eset <- GSE19668[[1]]

#Metadata ------
GSE19668_metadata = eset %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    subject_id = "title",
    # sex = "gender:ch1", 
    treatment = "infection:ch1", 
    # tissue = "tissue:ch1",
    timepoint = "time point:ch1",
    # animal_replicate = "animal replicate:ch1",
    strain = "strain:ch1") %>% 
  mutate(treatment = fct_relevel(treatment, "none"),
         timepoint = fct_relevel(timepoint, "0"),
         condition = paste(treatment, strain, timepoint, sep = "_"))

#Save
GSE19668_metadata %>% saveRDS(file = here("tables", paste0(filename, "_metadata.rds")))


#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% as.data.frame() %>% saveRDS(here("tables", paste0(filename, "_exprs.rds")))
```

Clean metadata
```{r}
#Import
GSE19668_metadata = readRDS(file = here("tables", "saureus_infection_GSE19668_metadata.rds"))

GSE19668_metadata %>% 
  select(strain)

#Filter metadata
GSE19668_metadata_filtered = GSE19668_metadata %>%
  filter(strain == "C57BL/6J") %>% 
  separate(subject_id, into = c("tissue", "strain1", "timepoint1", "participant_id"), sep = " ") %>% 
  select(!c(tissue:timepoint1)) %>% 
  arrange(participant_id, timepoint) %>% 
  separate(timepoint, into = c("timepoint_number", "time_unit")) %>% 
  mutate(time_unit = "hours",
         timepoint = timepoint_number %>% as.numeric(),
         organism = "Mouse") %>% 
  arrange(timepoint) %>% 
  select(-timepoint_number) %>% 
  select(geo_sample, participant_id, condition, timepoint, time_unit, everything())

#Save
GSE19668_metadata_filtered %>% saveRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

```

## Annotate dataset
```{r}
#Import
saureus_mouse_exprs = readRDS(here("tables","saureus_infection_GSE19668_exprs.rds"))
saureus_mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

#Inputs
filename = "saureus_infection_mouse"
mouse_exprs = saureus_mouse_exprs 
mouse_metadata = saureus_mouse_metadata
```


```{r}

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
mouse_exprs_metadata  = mouse_exprs %>% 
  select(c(mouse_metadata$geo_sample)) %>%
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = c(mouse_metadata$geo_sample),
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(mouse_metadata, 
             by = "geo_sample") 


#Get value from day 0 and calculate Log2FC
mouse_exprs_log2fc_sample = mouse_exprs_metadata %>% 
  filter(timepoint == 0) %>% 
  select(probe_id, participant_id, day0_value_raw = value_raw) %>% 
  #Filter batch
  inner_join(mouse_exprs_metadata %>% 
               filter(timepoint != "0"),
             by = join_by("probe_id", "participant_id")) %>% 
  mutate(log2fc = value_raw - day0_value_raw) %>% 
  distinct() 

#Check duplicates
mouse_exprs_log2fc_sample %>% 
  filter(participant_id == "rep1",
         probe_id == "1415670_at")


mouse_exprs_log2fc_sample %>% 
  saveRDS(here("tables", "saureus_infection_mouse_log2fc_sample_clean_long.rds"))
```


```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pull(probe_id) %>% unique()

# listAttributes(ensembl_mart)
#Annotate
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2", #Probes
  values = gene_id,
  mart = ensembl_mart)

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_log2summary = mouse_exprs %>% #20,845 probes
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_mouse430_2), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = mouse_log2summary %>% 
  #filter ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

#Filter out probes with multiple mgi and hgnc symbols
probe_dupes = mouse_log2summary %>% get_dupes(probe)

mouse_log2summary = mouse_log2summary %>% 
  select(probe, hgnc_symbol) %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique()),
         !probe %in% c(probe_dupes$probe %>% unique())) 

#Modify expression matrix

#Filtered probes
mouse_log2_filtered_probes = mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
  inner_join(mouse_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol,
         probe,
         c(mouse_metadata %>% 
             pull(geo_sample)))
#Check
mouse_log2_filtered_probes %>% dim()
mouse_log2_hgnc_symbol = mouse_log2_filtered_probes 
mouse_log2_hgnc_symbol %>% get_dupes(probe)

#Save
mouse_log2_hgnc_symbol %>% saveRDS(file = here("tables","saureus_infection_mouse_exprs_hgnc_symbol.rds"))
```

Pivot data and create expression set
```{r}
#Import
mouse_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_mouse_exprs_hgnc_symbol.rds"))
mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

#Pivot to long and wide format, and annotate
mouse_log2_samples_log2fc_long = mouse_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(mouse_metadata,
             by = "geo_sample")


mouse_log2_samples_log2fc_wide = mouse_log2_samples_log2fc_long %>% 
  select(human_symbol, probe, participant_id, timepoint, log2fc) %>%
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc") %>% 
  mutate(timepoint = as.character(timepoint) %>% fct_reorder(timepoint %>% as.numeric()))


#Create Expression Set
eset_mouse <- ExpressionSet(
  assayData = mouse_log2_hgnc_symbol %>% 
    select(-probe) %>% 
    column_to_rownames("human_symbol") %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(mouse_metadata))

#Save
eset_mouse %>% saveRDS(file = here("tables", "saureus_infection_mouse_eset.rds"))

```


### Differential expression 

#### Limma
Paired samples design
```{r}

#Import
mouse_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_mouse_exprs_hgnc_symbol.rds"))
mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))
eset_mouse = readRDS(file = here("tables", "saureus_infection_mouse_eset.rds"))

#Create model matrix
design <- model.matrix(~ condition + 0, # 0 defines the way we compare the samples
                       eset_mouse) 

colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

# Estimate the correlation between measurements made on the same subject (paired-samples)
corfit <- duplicateCorrelation(eset_mouse, 
                               design = design, 
                               block = eset_mouse$participant_id)
corfit$consensus #0.001357108

#Fit Limma model
fit <- lmFit(eset_mouse, 
             design, 
             block = eset_mouse$participant_id, 
             correlation = corfit$consensus) #Use inter-subject correlation as input into the linear model fit

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(conditionStaphylococcus.aureus_C57BL.6J_2.hours - conditionnone_C57BL.6J_0,
                             conditionStaphylococcus.aureus_C57BL.6J_4.hours - conditionnone_C57BL.6J_0,
                             conditionStaphylococcus.aureus_C57BL.6J_6.hours - conditionnone_C57BL.6J_0,
                             conditionStaphylococcus.aureus_C57BL.6J_12.hours - conditionnone_C57BL.6J_0,
                             levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs = all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "conditionnone_C57BL.6J_0", "0")) %>% 
  mutate(condition = case_when(comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_2.hours"	~ "Saureus_2h",
                               comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_4.hours"	~ "Saureus_4h",
                               comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_6.hours"	~ "Saureus_6h",
                               comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_12.hours"	~ "Saureus_12h",
                               T~comparisonA),
         organism = "Mouse",
         timepoint = str_remove(condition, "Saureus_") %>% str_remove("h")) %>% 
  select(condition, everything()) %>% 
  clean_names() %>% 
  rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs %>% saveRDS(file = here("tables", "saureus_infection_mouse_dge_limma_degs.rds"))

all_degs %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarize(n_degs = n())

```



##Import


```{r}
mouse_degs = readRDS(file = here("tables", "saureus_infection_mouse_dge_limma_degs.rds"))
```



#### Correlation plots, pairwise
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, GSE19668, C57BL/6J", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_alldays
#Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)
beep()



#Timepoint 2
timepoint_select = 2
mouse_correlation_ggpairs_day2 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == 2) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day2
#Save
mouse_correlation_ggpairs_day2 %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_day2.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 6
timepoint_select = 6
mouse_correlation_ggpairs_day6 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day6
#Save
mouse_correlation_ggpairs_day6 %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_day6.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 12
timepoint_select = 12
mouse_correlation_ggpairs_day12 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day12
#Save
mouse_correlation_ggpairs_day12 %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_day12.png"),
         height = 10,
         width = 10)
beep()

```




#### PCA

```{r fig.height=4, fig.width=5}
#Import metadata
annotation_samples_pca = GSE19668_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = GSE19668_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, entrezgene_id, log2fc) %>% 
  filter(!entrezgene_id == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("entrezgene_id") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Mouse, S. aureus, GSE19668",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

```

#### BTMs
```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)

#Annotate genes by BTM
GSE19668_genes_btm = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  rename(mm_entrez = entrezgene_id) %>% 
  select(mgi_symbol, symbol = human_symbol, participant_id, geo_sample, timepoint, treatment, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = GSE19668_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process
```



##### GGPairs
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

mouse_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) 

mouse_correlation_ggpairs_btms_gsea_alldays

mouse_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_gsea_btm_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)


#All days, colored by group
mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor = ggpairs_data_dge_btm_mean_process %>%
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = group,
                                 fill = group),
          title = paste("Day", timepoint_selected),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors)

mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor

mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_gsea_btm_correlation_ggpairs_alldays_groupcolor.png"),
         height = 10,
         width = 10)

beep()

```




##### Ggplot manual
```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(3:7) %>% 
  colnames()


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == 2) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == 2) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = "GSE19668, Mouse, SAureus infeciton, random sample")
```













## Human

### Prepare data
```{r}
#S. aureus: GSE33341 -----
filename = "GSE33341"
GSE <- getGEO("GSE33341", GSEMatrix = TRUE) 
eset <- GSE[[2]] #The matrix #1 the mouse annotated

#Metadata ------
GSE_metadata = eset %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    participant_id = "title",
    sex = "host gender:ch1",
    condition = "source_name_ch1",
    pathogen = "pathogen:ch1",
    description = "source_name_ch1", 
    ethnicity = "ethnicity:ch1",
    tissue_infection = "anatomic site of infection:ch1",
    batch_experimental = "experimental batch:ch1",
    age = "host age (yr):ch1") %>% 
  mutate(condition = pathogen,
         organism = "Human",
         timepoint = 1,
         time_unit = "Day",
         tissue = "Blood",
         participant_id = str_extract(participant_id, "\\[.*?\\]") %>% 
           str_remove("\\[") %>% 
           str_remove("\\]"),
         condition = case_when(description == "blood patient with E. coli bacteremia prior antibiotic treatment (Emergency Department)" ~ "E. coli prior antibiotic",
                               description == "blood patient with S. aureus bacteremia prior antibiotic treatment (Emergency Department)" ~ "S aureus prior antibiotic",
                               description == "blood human NOT infected" ~ "No infection",
                               T ~ description),
         pathogen = if_else(pathogen == "-", "Healthy", pathogen))
#Save
GSE_metadata %>% saveRDS(file = here("tables", "saureus_infection_human_metadata.rds"))

#Filter metadata
GSE_metadata_filtered = GSE_metadata %>%
  filter(pathogen %in% c("Staphylococcus aureus", "Healthy")) %>% 
  select(geo_sample, participant_id, condition, timepoint, everything()) %>% 
  mutate(ethnicity = str_to_title(ethnicity)) %>% 
  filter(batch_experimental != "2739",
         ethnicity != "Unknown",
         age != "unknown")

GSE_metadata_filtered %>% 
  saveRDS(file = here("tables", "saureus_infection_human_metadata.rds"))



#Expression data ------------
GSE_exprs <- exprs(eset) %>% as.data.frame()

#Save
GSE_exprs %>% saveRDS(file = here("tables", "saureus_infection_human_exprs.rds"))
```

Prepare data
```{r}
#Import
saureus_human_exprs = readRDS(file = here("tables", "saureus_infection_human_exprs.rds"))
saureus_human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))


#Inputs
filename = "saureus_infection_human"
human_exprs = saureus_human_exprs 
human_metadata = saureus_human_metadata
```


```{r}
#Annotate genes ------
#Human -----
#Annotate data and calculate log2fc
human_exprs_metadata  = human_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = -probe_id,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(human_metadata, 
             by = "geo_sample") 

#Get value from healthy and calculate Log2FC
human_exprs_log2fc_sample = human_exprs_metadata %>% 
  filter(pathogen == "Healthy") %>% 
  select(probe_id, value_raw_ref = value_raw) %>% 
  group_by(probe_id) %>% 
  summarize(value_raw_ref = mean(value_raw_ref)) %>% 
  inner_join(human_exprs_metadata %>% 
               filter(pathogen != "Healthy"),
             by = join_by("probe_id")) %>% 
  filter(condition != "No infection") %>% 
  mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

#Check duplicates
human_exprs_log2fc_sample %>% 
  filter(participant_id == "CSD4620.CEL",
         probe_id == "1007_s_at")

human_exprs_log2fc_sample %>% 
  saveRDS(here("tables", "saureus_infection_human_log2fc_sample_clean_long.rds"))
```


```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
# listDatasets(ensembl_mart)
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- human_exprs_log2fc_sample$probe_id %>% unique()

#Annotate
annot <- getBM(
  attributes = c("affy_hg_u133a_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133a_2",
  values = gene_id,
  mart = ensembl_mart) %>% 
  #Standardize
  rename(probe = affy_hg_u133a_2,
                     entrezgene_id = entrezgene_id,
                     human_symbol = hgnc_symbol,
                     ensembl_id = ensembl_gene_id)

#Filter genes with highest median expression
human_log2summary = human_exprs %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot, 
            by = "probe") %>% 
  #Filter ensembl id
  group_by(ensembl_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup()  


#Filter out genes with multiple hgnc_symbols
hgnc_symbol_dupes = human_log2summary %>% 
  #filter ensembl id with multiple hgnc_symbols
  get_dupes(human_symbol)

human_log2summary = human_log2summary %>% 
  filter(!human_symbol %in% c(hgnc_symbol_dupes$human_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
human_log2_filtered_probes = human_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(human_log2summary, by = "probe") %>%   #214,396
  select(human_symbol, probe,
         c(human_metadata$geo_sample))
#14395
human_log2_filtered_probes %>% dim()

human_log2_filtered_probes = human_log2_filtered_probes 

human_log2_filtered_probes %>% 
  get_dupes(human_symbol)

#Save 
human_log2_filtered_probes %>% 
  saveRDS(here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
```


```{r}
#Import
human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))

#Pivot to long and wide format, and annotate
human_log2_samples_log2fc_long = human_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(human_metadata,
             by = "geo_sample")


human_log2_samples_log2fc_wide = human_log2_samples_log2fc_long %>% 
  select(human_symbol, probe, participant_id, timepoint, log2fc) %>%
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc") %>% 
  mutate(timepoint = as.character(timepoint) %>% fct_reorder(timepoint %>% as.numeric()))


#Create Expression Set
human_metadata_eset = human_metadata %>% 
  mutate(age = as.numeric(age) %>% scale(),
         timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         pathogen = fct_relevel(pathogen, "Healthy"),
         batch_experimental = as.factor(batch_experimental),
         ethnicity = as.factor(ethnicity)) 

eset_human <- ExpressionSet(
  assayData = human_log2_hgnc_symbol %>% 
    select(-probe) %>% 
    column_to_rownames("human_symbol") %>% 
    select(c(human_metadata_eset$geo_sample)) %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(human_metadata_eset))

#Save
eset_human %>% saveRDS(file = here("tables", "saureus_infection_human_eset.rds"))

```






### Differential expression

#### Limma

```{r}
#Import
human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))
eset_human = readRDS(file = here("tables", "saureus_infection_human_eset.rds"))

#Create model matrix
design <- model.matrix(~ age + sex + ethnicity + batch_experimental + pathogen,
                       data = pData(eset_human))

colnames(design) <- make.names(colnames(design))  
rownames(design) <- rownames(pData(eset_human))

#Fit Limma model
fit <- lmFit(eset_human, 
             design) 

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(pathogenStaphylococcus.aureus, levels = design)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- limma::eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_human = all_contrasts_degs %>% 
  dplyr::select(comparison, everything()) %>% 
  dplyr::mutate(condition = case_when(comparison == "pathogenStaphylococcus.aureus"	~ "Saureus_Day1",
                               T~comparison),
         organism = "Human",
         timepoint = "1") %>% 
  dplyr::select(condition, everything()) %>% 
  clean_names() %>% 
  dplyr::rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs_human %>% saveRDS(file = here("tables", "saureus_infection_human_dge_limma_degs.rds"))

all_degs_human %>% 
  dplyr::group_by(condition) %>%
  dplyr::filter(adj_p_val <= 0.05) %>% 
  dplyr::summarize(n_degs = n())

```


#### t-Test
```{r}
#Import ================
human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))

#Wrangle ==============
#Pivot to long and wide format, and annotate
human_log2_samples_log2fc_long = human_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(human_metadata,
             by = "geo_sample")

human_log2_wide = human_log2_hgnc_symbol %>%
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(human_metadata %>% 
               select(geo_sample, participant_id, pathogen),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_")

human_exprs_log2_infected_healthy = human_exprs_metadata %>% 
  filter(pathogen == "Healthy") %>% 
  select(probe_id, value_raw_ref = value_raw) %>% 
  group_by(probe_id) %>% 
  summarize(value_raw_ref = mean(value_raw_ref)) %>% 
  inner_join(human_exprs_metadata %>% 
               filter(pathogen != "Healthy"),
             by = join_by("probe_id")) %>% 
  filter(pathogen != "Healthy") %>% 
  inner_join(human_log2_samples_log2fc_long %>% 
               select(probe_id = probe, human_symbol),
             by = join_by("probe_id")) %>% 
  mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

# Unpaired t-test ----------
human_log2fc_unpaired_ttest <- human_exprs_log2_infected_healthy %>%
  group_by(human_symbol) %>%
  
  summarise(
    pval = tryCatch(
      t.test(value_raw, value_raw_ref,
             paired = FALSE, var.equal = FALSE, 
             alternative = "two.sided")$p.value,
      error = function(e) NA_real_
    ),
    .groups = 'drop'
  ) %>%
  mutate(
    padj = p.adjust(pval, method = "fdr")
    ) %>% 
  mutate(
    condition = "Saureus infection",
    timepoint = "1",
    test = "t-Test",
    paired = "Unpaired")


human_log2fc_unpaired_ttest %>% saveRDS(file = here("tables", "saureus_infection_human_dge_unpaired_ttest.rds"))

```







#### Import
```{r}
#Import
GSE33341_counts_log2_summary_annotated_probes = 
  readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol_long.rds"))

human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))



#Pivot to long and wide format, and annotate
human_log2_samples_log2fc_long = human_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(human_metadata,
             by = "geo_sample")


human_log2_samples_log2fc_wide = human_log2_samples_log2fc_long %>% 
  select(human_symbol, probe, participant_id, timepoint, log2fc) %>%
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc") %>% 
  mutate(timepoint = as.character(timepoint) %>% fct_reorder(timepoint %>% as.numeric()))

#Pivot wider
GSE33341_log2fc_sample_wide = GSE33341_counts_log2_summary_annotated_probes %>%
  select(human_symbol, participant_id, timepoint, log2fc) %>% 
  pivot_wider(names_from = participant_id,
              values_from = log2fc) %>% 
  ungroup()
```





## Combine datasets

```{r}
#Import results =======
all_degs_mouse = read_rds(here("tables", "saureus_infection_mouse_dge_limma_degs.rds"))
all_degs_human = read_rds(here("tables", "saureus_infection_human_dge_limma_degs.rds"))
mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))
mouse_log2fc_sample = readRDS(here("tables", "saureus_infection_mouse_log2fc_sample_clean_long.rds")) 
mouse_exprs = readRDS(file = here("tables", "saureus_infection_mouse_exprs_hgnc_symbol_long.rds"))
human_log2fc_sample = readRDS(here("tables", "saureus_infection_human_log2fc_sample_clean_long.rds")) 
human_exprs = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))



#DGE results ===========
all_degs_human_mouse = all_degs_mouse %>% 
  bind_rows(all_degs_human) %>% 
  select(condition, timepoint, human_symbol, mean_l2fc:organism)
all_degs_human_mouse %>% saveRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds"))


#Metadata ===========
human_mouse_metadata = mouse_metadata %>% bind_rows(human_metadata)
human_mouse_metadata %>% saveRDS(file = here("tables", "saureus_infection_human_mouse_metadata.rds"))

#Merge sample log2fc =============
mouse_log2fc_sample_symbol = mouse_log2fc_sample %>% 
  inner_join(mouse_exprs %>% 
               ungroup() %>% 
               select(human_symbol, probe_id) %>% 
               distinct(),
             by = "probe_id") %>% 
  mutate(organism = "Mouse",
         subject_id = participant_id)

human_symbol_dupes = mouse_log2fc_sample_symbol %>% 
select(organism, timepoint, participant_id, human_symbol, log2fc) %>% 
distinct() |>
dplyr::summarise(n = dplyr::n(), .by = c(human_symbol, participant_id, timepoint)) |>
dplyr::filter(n > 1L)

mouse_log2fc_sample_symbol = mouse_log2fc_sample_symbol %>% 
  filter(!human_symbol %in% c(human_symbol_dupes$human_symbol))

human_log2fc_sample_symbol = human_log2fc_sample %>% 
  inner_join(human_exprs %>% 
               select(human_symbol, probe_id = probe) %>% 
               distinct(),
             by = "probe_id") %>% 
  mutate(organism = "Human",
         timepoint = 24,
         subject_id = participant_id) 

human_mouse_log2fc_samples = bind_rows(mouse_log2fc_sample_symbol, human_log2fc_sample_symbol) %>% 
  ungroup()

#Check duplicates and remove
human_symbol_dupes = human_mouse_log2fc_samples %>% 
select(organism, timepoint, participant_id, geo_sample, human_symbol, log2fc) %>% 
distinct() |>
dplyr::summarise(n = dplyr::n(), .by = c(human_symbol, participant_id, timepoint)) |>
dplyr::filter(n > 1L)

human_mouse_log2fc_samples$organism %>% unique()

human_mouse_log2fc_samples %>% 
  saveRDS(here("tables", "saureus_infection_human_mouse_samples_log2fc.rds"))


```




## FIT

```{r}
#Get log2fc mean
GSE19668_input2FIT = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
    filter(timepoint == 2) %>% 
  select(participant_id, MM.Entrez = entrezgene_id, EffectSize = log2fc) %>% 
  group_by(MM.Entrez) %>% 
  summarize(mean_log2fc = mean(EffectSize))

#Save
GSE19668_input2FIT %>% 
  write_csv(file = here("FIT_Mouse2man", "saureus_infection_mouse_fit_prediction_dge_degs_day2.csv"))

#Run FIT
GSE19668_input2FIT_day2 = FIT.mouse2man::FIT(MouseFile = here("FIT_Mouse2man", "saureus_infection_mouse_fit_prediction_dge_degs_day2.csv"), DataType = "rnaseq")
```

```{r fig.height=5, fig.width=8}
#Bind rows
mouse_human_fit_log2fc = bind_rows(GSE19668_input2FIT_day2 %>%
                                     #Get DEGs for FIT
                                     filter((CI_low != 0 | CI_high != 0) & !is.na(CI_low) & !is.na(CI_high)) %>%  
                                   select(human_symbol = Human.symbol,
                                          log2fc = FIT_prediction) %>%
                                     mutate(organism = "FIT"),
                                   
                                   #Get DEGs for Human
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_human <= 0.05) %>% 
                                     select(human_symbol, log2fc = Human) %>%
                                     mutate(organism = "Human"),
                                   
                                   #Get DEGs for Mouse
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_mouse <= 0.05) %>% 
                                     select(human_symbol, log2fc = Mouse) %>%
                                     mutate(organism = "Mouse"))


#Pivot wider
mouse_human_fit_log2fc_wide = GSE19668_input2FIT_day2 %>%
            select(human_symbol = Human.symbol,
                   FIT_prediction) %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Human),
            by = "human_symbol") %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Mouse),
            by = "human_symbol")

#Save
mouse_human_fit_log2fc %>% 
  saveRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))

mouse_human_fit_log2fc_wide %>% 
  saveRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))
```

### Distribution

```{r fig.height=5, fig.width=8}
#Import
mouse_human_fit_log2fc = readRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))
mouse_human_fit_log2fc_wide = readRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))

GSE33341_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))
GSE19668_metadata = readRDS(file = here("tables", "saureus_infection_GSE19668_metadata.rds"))
```

```{r fig.height=5, fig.width=8}
#Plot histogram
mouse_human_fit_log2fc_histogram = mouse_human_fit_log2fc %>% 
  ggplot() +
  aes(x = log2fc,
      fill = organism) +
  geom_density(aes(y = after_stat(count),
                   color = organism),
                  position = position_identity(), 
               alpha = 0.1,
               size = 1,
              show.legend = T) +
  scale_fill_manual(values = c(FIT = "#4361ee",
                               Mouse = "#4DBBD5FF",
                               Human = "#546E7AFF"))+
  scale_color_manual(values = c(FIT = "#4361ee",
                                Mouse = "#4DBBD5FF",
                                Human = "#546E7AFF"))+
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        legend.key = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0),
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(limits = c(-4, 4)) +
  labs(color = "",
       fill = "")

mouse_human_fit_log2fc_histogram

#Faceted
mouse_human_fit_log2fc_histogram_faceted = mouse_human_fit_log2fc_histogram +
  facet_wrap(~organism,
             scales = "free_y")


#Plot scatter plot
mouse_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = Mouse,
      y = Human) +
#All DEGs
geom_point(color = "#4DBBD5FF",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "Mouse L2FC",
         y = "Human L2FC")

mouse_human_scatterplot


fit_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = FIT_prediction,
      y = Human) +
#All DEGs
geom_point(color = "#4361ee",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "FIT L2FC",
         y = "Human L2FC")

#Histograms
(((mouse_human_fit_log2fc_histogram + mouse_human_fit_log2fc_histogram_faceted) + 
    plot_layout(widths = c(0.5, 1), guides = "collect") +
    plot_annotation(theme = theme(legend.position = "top")))  /
    #Scatter plots
    (mouse_human_scatterplot + fit_human_scatterplot)) +
  plot_layout(heights = c(0.5, 1))

```



### BTMs
Import
```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)
```

```{r}
#Annotate genes by BTM
dge_btm_mean_process = mouse_human_fit_log2fc %>%
  filter(organism != "Mouse") %>% 
  rename(symbol = human_symbol) %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  group_by(organism, process) %>% 
  summarize(mean_log2fc = mean(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism", values_from = "mean_log2fc") 

dge_btm_mean_process
```

```{r fig.height=5, fig.width=7}

#Calculate correlation
cor_labels <- dge_btm_mean_process %>%
  summarise(cor_value = cor(Human, FIT, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
dge_btm_mean_process %>% 
ggplot(aes(x = FIT, y = Human)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  stat_smooth(aes(x = FIT, y = Human),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "FIT Log2FC", 
       y = "Human Log2FC", 
       color = "",
       title = "Mouse vs. Human")

```












# 4. E. coli infection

### Human and Mouse, same dataset

```{r}
#E. coli: GSE33341 -----
filename = "GSE33341"
GSE33341 <- getGEO("GSE33341", GSEMatrix = TRUE) 
eset_mouse <- GSE33341[[1]]
eset_human <- GSE33341[[2]]

#Metadata ------
#Mouse
ecoli_mouse_metadata = eset_mouse %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    #Host
    subject_title = "title",
    subject_num = "replicate number:ch1",
    batch = "experimental batch:ch1",
    host_strain = "host strain:ch1",
    organism = "organism_ch1",
    age = "host age (yr):ch1",
    sex = "host gender:ch1", 
    #Pathogen
    pathogen = "pathogen:ch1",
    status = "infection status:ch1",
    site_infection = "anatomic site of infection:ch1",
    route_infection = "inoculum route:ch1",
    dosage = "pathogen   dosage cfu/g:ch1",
    timepoint = "time after infection (h):ch1",
    bac_strain = "bacterial strain:ch1"
    ) %>% 
  filter(pathogen %in% c("Escherichia  coli", "-")) %>%
  mutate(pathogen = if_else(pathogen == "Escherichia  coli", "escherichia coli", "none"),
         subject_id = paste0("Mouse_", host_strain, "_Animal", subject_num, "_Batch", batch),
         status = str_replace(status, "-", "Healthy"),
         condition = paste("Mouse", host_strain, timepoint, sep = "_"),
         comparison = paste0("Mouse_", host_strain, "_Batch", batch)) %>% 
  select(geo_sample, comparison, condition, subject_id, everything())


#Human
ecoli_human_metadata = eset_human %>% 
  pData() %>% 
  filter(`pathogen:ch1` %in% c("Escherichia coli", "-")) %>%
  dplyr::select( 
    geo_sample = "geo_accession", 
    #Host
    subject_title = "title",
    # subject_num = "replicate number:ch1",
    batch = "experimental batch:ch1",
    # host_strain = "host strain:ch1",
    organism = "organism_ch1",
    age = "host age (yr):ch1",
    sex = "host gender:ch1", 
    ethnicity = "ethnicity:ch1",
    #Pathogen
    pathogen = "pathogen:ch1",
    # status = "infection status:ch1",
    site_infection = "anatomic site of infection:ch1"
    # route_infection = "inoculum route:ch1",
    # dosage = "pathogen   dosage cfu/g:ch1",
    # timepoint = "time after infection (h):ch1",
    # bac_strain = "bacterial strain:ch1"
    ) %>% 
  mutate(site_infection = str_to_lower(site_infection),
         organism = "Human",
         pathogen = if_else(pathogen == "-", "Healthy", pathogen),
         status = if_else(pathogen == "Healthy", pathogen, "infected"),
         timepoint = "0",
         subject_id = if_else(pathogen == "Escherichia coli", 
                              paste0("Ecoli_human_", site_infection,"_",geo_sample), 
                              paste0("Healthy_human_", geo_sample)),
         site_infection = if_else(site_infection == "-", "none", site_infection),
         condition = subject_id) %>% 
  select(geo_sample, condition, subject_id, everything()) %>% 
  mutate(timepoint = if_else(pathogen == "Escherichia coli", "24", timepoint))

ecoli_human_metadata

#Unite metadata
ecoli_mouse_human_metadata = ecoli_human_metadata %>% 
  bind_rows(ecoli_mouse_metadata)

ecoli_mouse_human_metadata %>% 
  saveRDS(file = here("tables", "ecoli_infection_human_mouse_metadata.rds"))
```

```{r}
#Import
ecoli_mouse_human_metadata =
  readRDS(file = here("tables", "ecoli_infection_human_mouse_metadata.rds"))


ecoli_mouse_human_metadata %>% 
  select(strain, treatment, tissue) %>% 
  distinct()

#Get expression data -----
#Mouse
mouse_exprs <- exprs(eset_mouse) %>% 
  as.data.frame() %>% 
  select(GSE33341_ecoli_mouse_human_metadata %>% 
           filter(organism == "Mus musculus") %>% 
           pull(geo_sample)) %>% 
  rownames_to_column("probe_id") 

mouse_exprs %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "ecoli_infection_mouse_exprs.rds"))

#Human
human_exprs <- exprs(eset_human) %>% 
  as.data.frame() %>% 
  select(GSE33341_ecoli_mouse_human_metadata %>% 
           filter(organism == "Human") %>% 
           pull(geo_sample)) %>% 
  rownames_to_column("probe_id")

human_exprs %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "ecoli_infection_human_exprs.rds"))

```

#### Mouse

##### Annotate genes

This procedure was done using only one batch (8)

```{r}
#Import
ecoli_human_exprs = readRDS(here("tables", "ecoli_infection_human_exprs.rds"))
ecoli_mouse_exprs = readRDS(here("tables", "ecoli_infection_mouse_exprs.rds"))
ecoli_mouse_human_metadata = readRDS(file = here("tables", "ecoli_infection_human_mouse_metadata.rds"))

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
mouse_expression_metadata  = ecoli_mouse_exprs %>% 
  pivot_longer(cols = 2:125,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(ecoli_mouse_human_metadata, 
             by = "geo_sample") 

mouse_metadata =  ecoli_mouse_human_metadata %>% 
               filter(host_strain == "C57BL/6J",
                      batch == "8")

mouse_metadata %>% 
  saveRDS(file = here("tables", "ecoli_infection_mouse_metadata.rds"))

#Get value from day 0 and calculate Log2FC
log2fc_sample = mouse_expression_metadata %>% 
  #Filter batch
  filter(organism == "Mus musculus",
         status == "Healthy",
         host_strain == "C57BL/6J",
         batch == "8") %>% 
  select(probe_id, batch, subject_num, day0_value_raw = value_raw) %>% 
  #Get the mean expression value for healthy
  group_by(probe_id, batch) %>% 
  summarize(n = n(),
            #Mean of healthy
            day0_value_raw = sum(day0_value_raw)/n,
            batch) %>% 
  select(probe_id, day0_value_raw) %>% 
  #Filter batch
  inner_join(GSE33341_mouse_expression_metadata %>% 
               filter(timepoint != "0",
                      host_strain == "C57BL/6J",
                      batch == "8"),
             by = join_by("probe_id", "batch")) %>% 
  mutate(log2fc = value_raw - day0_value_raw) %>% 
  distinct() 

#Check duplicates
log2fc_sample %>% 
  filter(subject_id == "Mouse_C57BL/6J_Animal1_Batch8",
         probe_id == "1415670_at")


#Save
log2fc_sample %>% 
  saveRDS(here("tables", "ecoli_infection_mouse_log2fc_sample_clean_long.rds"))
```

```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- log2fc_sample$probe_id %>% unique()
# listAttributes(ensembl_mart)
#Annotate
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2",
  values = gene_id,
  mart = ensembl_mart)

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_log2summary = ecoli_mouse_exprs %>% #20,845 probes
  as.data.frame() %>% 
  rename(probe = probe_id) %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_mouse430_2), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = mouse_log2summary %>% 
  #filter ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

mouse_log2summary = mouse_log2summary %>% 
  select(probe, hgnc_symbol) %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
mouse_log2_filtered_probes = ecoli_mouse_exprs %>% 
  as.data.frame() %>% 
  rename(probe = probe_id) %>% 
  inner_join(mouse_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol, 
         probe,
         c(mouse_metadata %>% 
             pull(geo_sample)))
#14395
mouse_log2_filtered_probes %>% dim()

mouse_log2_hgnc_symbol = mouse_log2_filtered_probes 

mouse_log2_hgnc_symbol %>% 
  get_dupes(human_symbol)

#Save
mouse_log2_hgnc_symbol %>% 
  saveRDS(file = here("tables", "ecoli_infection_mouse_exprs_hgnc_symbol_long.rds"))
```

Import

```{r}
#Import
mouse_counts_log2_summary_annotated_genes = 
  readRDS(file = here("tables", "ecoli_infection_mouse_exprs_hgnc_symbol_long.rds"))

#Pivot wider
log2fc_sample_wide = mouse_counts_log2_summary_annotated_probes %>% 
  select(probe = probe_id, entrezgene_id, batch:human_symbol) %>% 
  distinct() %>% 
  select(human_symbol, subject_id, timepoint, log2fc) %>%
  group_by(human_symbol, timepoint) %>% 
  pivot_wider(names_from = subject_id,
              values_from = log2fc) %>% 
  ungroup()  %>% 
  mutate(timepoint = fct_reorder(timepoint, timepoint %>% as.numeric()))
  
#Print
log2fc_sample_wide
```


##### Differential expression


```{r}
#Import
ecoli_mouse_log2_hgnc_symbol = readRDS(file = here("tables", "ecoli_infection_mouse_exprs_hgnc_symbol_long.rds"))
ecoli_mouse_metadata = readRDS(file = here("tables", "ecoli_infection_mouse_metadata.rds"))

#Create Expression Set
eset_mouse_ecoli <- ExpressionSet(
  assayData = ecoli_mouse_log2_hgnc_symbol %>% 
    column_to_rownames("human_symbol") %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(ecoli_mouse_metadata))

#Save
eset_mouse_ecoli %>% saveRDS(file = here("tables", "ecoli_infection_mouse_eset.rds"))
```



###### Limma
Paired samples design
```{r}

#Import
ecoli_mouse_log2_hgnc_symbol = readRDS(file = here("tables", "ecoli_infection_mouse_exprs_hgnc_symbol_long.rds"))
ecoli_mouse_metadata = readRDS(file = here("tables", "ecoli_infection_mouse_metadata.rds"))
eset_mouse_ecoli = readRDS(file = here("tables", "ecoli_infection_mouse_eset.rds"))

#Create model matrix
design <- model.matrix(~ condition + 0, # 0 defines the way we compare the samples
                       eset_mouse_ecoli) 

colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

# Estimate the correlation between measurements made on the same subject (paired-samples)
corfit <- duplicateCorrelation(eset_mouse_ecoli, 
                               design = design, 
                               block = eset_mouse_ecoli$subject_id)
corfit$consensus #-0.06

#Fit Limma model
fit <- lmFit(eset_mouse_ecoli, 
             design, 
             block = eset_mouse_ecoli$subject_id, 
             correlation = corfit$consensus) #Use inter-subject correlation as input into the linear model fit

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(conditionMouse_C57BL.6J_2	 - conditionMouse_C57BL.6J_0,
                             conditionMouse_C57BL.6J_6	 - conditionMouse_C57BL.6J_0,
                             conditionMouse_C57BL.6J_12	 - conditionMouse_C57BL.6J_0,
                             conditionMouse_C57BL.6J_24	 - conditionMouse_C57BL.6J_0,
                             levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_mouse_ecoli= all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  mutate(condition = case_when(comparisonA == "conditionMouse_C57BL.6J_2"	~ "Ecoli_2h",
                               comparisonA == "conditionMouse_C57BL.6J_6"	~ "Ecoli_6h",
                               comparisonA == "conditionMouse_C57BL.6J_12"	~ "Ecoli_12h",
                               comparisonA == "conditionMouse_C57BL.6J_24"	~ "Ecoli_24h",
                               T~comparisonA),
         organism = "Mouse",
         timepoint = str_remove(condition, "Ecoli_") %>% str_remove("h")) %>% 
  select(condition, everything()) %>% 
  clean_names() %>% 
  rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs_mouse_ecoli %>% saveRDS(file = here("tables", "ecoli_infection_mouse_dge_limma_degs.rds"))

all_degs_mouse_ecoli %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarize(n_degs = n())

```



###### T test
```{r}
#Import
mouse_exprs = readRDS(file = here("tables", "ecoli_infection_mouse_exprs_hgnc_symbol_long.rds"))
mouse_metadata = readRDS(file = here("tables", "ecoli_infection_mouse_metadata.rds"))
log2fc_sample = readRDS(here("tables", "ecoli_infection_mouse_log2fc_sample_clean_long.rds"))
log2fc_sample_symbol = log2fc_sample %>% 
  inner_join(mouse_exprs %>% 
               select(human_symbol, probe_id = probe),
             by = "probe_id")

```


```{r}
#Compute summary statistics
log2fc_summary = log2fc_sample_symbol %>% 
  group_by(human_symbol, timepoint) %>%
  summarize(
    avg_log2fc = mean(log2fc, na.rm = TRUE),  # mean log2 fold change
    n_subjects = n(),                         # number of samples
    sd = sd(log2fc, na.rm = TRUE),             # standard deviation
    se = sd / sqrt(n_subjects),               # standard error
    effect_size = avg_log2fc / sd)            # Cohen’s d-like effect size

log2_hgnc_symbol_wide = mouse_exprs %>%
  select(-probe) %>% 
  pivot_longer(cols = -human_symbol,
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(mouse_metadata %>% 
               select(geo_sample, subject_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, subject_id, timepoint, log2) %>% 
  distinct() %>% 
  select(subject_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "hour_") 

# Paired t-test
log2fc_paired_ttest <- log2_hgnc_symbol_wide %>%
  group_by(human_symbol) %>%
  
  summarise(
    pval_hour2 = tryCatch(t.test(hour_2, hour_0, paired = TRUE, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_hour6 = tryCatch(t.test(hour_6, hour_0, paired = TRUE, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_hour12 = tryCatch(t.test(hour_12, hour_0, paired = TRUE, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_hour24 = tryCatch(t.test(hour_24, hour_0, paired = TRUE, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    .groups = 'drop'
  ) %>%
  
  mutate(
    padj_hour2 = p.adjust(pval_hour2, method = "fdr"),
    padj_hour6 = p.adjust(pval_hour6, method = "fdr"),
    padj_hour12 = p.adjust(pval_hour12, method = "fdr"),
    padj_hour24 = p.adjust(pval_hour24, method = "fdr"),
  ) %>% 
  pivot_longer(
    cols = starts_with(c("pval_", "padj_")),
    names_to = c(".value", "timepoint"),
    names_pattern = "(pval|padj)_(hour\\d+)"
  ) %>% 
  mutate(timepoint = str_remove(timepoint, "hour"),
         test = "t-Test",
         paired = "Paired")

#Unpaired
log2fc_unpaired_ttest = log2_hgnc_symbol_wide %>%
  group_by(human_symbol) %>%
  
  summarise(
    pval_hour2 = tryCatch(t.test(hour_2, hour_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_hour6 = tryCatch(t.test(hour_6, hour_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_hour12 = tryCatch(t.test(hour_12, hour_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_hour24 = tryCatch(t.test(hour_24, hour_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    .groups = 'drop'
  ) %>%
  
  mutate(
    padj_hour2 = p.adjust(pval_hour2, method = "fdr"),
    padj_hour6 = p.adjust(pval_hour6, method = "fdr"),
    padj_hour12 = p.adjust(pval_hour12, method = "fdr"),
    padj_hour24 = p.adjust(pval_hour24, method = "fdr"),
  ) %>% 
  pivot_longer(
    cols = starts_with(c("pval_", "padj_")),
    names_to = c(".value", "timepoint"),
    names_pattern = "(pval|padj)_(hour\\d+)"
  ) %>% 
  mutate(timepoint = str_remove(timepoint, "hour"),
         test = "t-Test",
         paired = "Unpaired") 

log2fc_paired_unpaired_ttest = log2fc_unpaired_ttest %>% 
  bind_rows(log2fc_paired_ttest) 

log2fc_paired_unpaired_ttest %>% 
  saveRDS(file = here("tables", "ecoli_infection_mouse_log2fc_paired_unpaired_ttest.rds"))

```


###### Compare DGE methods

```{r}
#Import ------
#t-Test
mouse_log2fc_paired_unpaired_ttest =
  readRDS(file = here("tables", "ecoli_infection_mouse_log2fc_paired_unpaired_ttest.rds")) %>% 
  mutate(organism = "Mouse")

#Limma
mouse_degs = readRDS(file = here("tables", "ecoli_infection_mouse_dge_limma_degs.rds")) %>% 
  clean_names() %>% 
  rename(padj = adj_p_val) %>% 
  mutate(organism = "Mouse",
         test = "Limma",
         paired = "Paired") 


#Merge datasets
degs_ttest_limma =
  bind_rows(
    mouse_log2fc_paired_unpaired_ttest,
    mouse_degs
    )

#Summarize  
degs_ttest_limma_summary = degs_ttest_limma %>% 
  group_by(organism, timepoint, test, paired) %>% 
  summarise(
    n_degs = sum(padj <= 0.05, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) 

degs_ttest_limma_summary

```


```{r fig.height=5, fig.width=5}
# Plot number of DEGs per test and significance level
degs_ttest_limma_summary_plot_gg <- degs_ttest_limma_summary %>%
  ggplot(aes(
    x = timepoint,
    y = n_degs,
    fill = test,
    alpha = paired
  )) +
  geom_col(width = 0.8, position = position_dodge2()) +
  
  # Manual colors per test
  scale_fill_manual(values = c(
    "t-Test" = "gray25",
    "Limma" = "#4DBBD5FF"
  )) +
  
  # Alpha: paired = 1, unpaired = 0.5
  scale_alpha_manual(values = c("Paired" = 1, "Unpaired" = 0.2)) +
  
  # Continuous x-axis for numeric timepoints
  scale_x_discrete(
    breaks = sort(unique(degs_ttest_limma_summary$timepoint)),
    expand = c(0.01, 0.01)
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  
  theme_vaxgo() +
  labs(
    title = "Total DEGs by test and timepoint",
    subtitle = "Comparison across tests (FDR ≤ 0.05)",
    x = "Day",
    y = "Number of DEGs",
    fill = "Test",
    alpha = "Pairing"
  ) +
  facet_wrap(~organism, scales = "free_y") +
  theme(
    legend.position = "top",
    axis.line.y = element_blank(),
    panel.grid.major.y = element_line(color = "gray75", linetype = 3),
    axis.ticks.y = element_line(color = "gray75")
  )

# Render plot
degs_ttest_limma_summary_plot_gg

degs_ttest_limma_summary_plot_gg %>% 
  ggsave(filename = here("Figures", "ecoli_infection_mouse_dge_ttest_limma_summary.png"),
         width = 5,
         height = 5)
```








#### Human

##### Annotate genes

There are batches

```{r}

ecoli_human_exprs = readRDS(here("tables", "ecoli_infection_human_exprs.rds"))
ecoli_mouse_exprs = readRDS(here("tables", "ecoli_infection_mouse_exprs.rds"))
ecoli_mouse_human_metadata = readRDS(file = here("tables", "ecoli_infection_human_mouse_metadata.rds"))


#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
human_expression_metadata  = ecoli_human_exprs %>% 
  pivot_longer(cols = 2:63,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(ecoli_mouse_human_metadata, 
             by = "geo_sample") 

human_expression_metadata %>% 
  saveRDS(file = here("tables", "ecoli_infection_human_exprs_metadata.rds"))


human_expression_metadata =
  readRDS(file = here("tables", "ecoli_infection_human_exprs_metadata.rds"))

human_metadata =  human_expression_metadata %>% 
  filter(organism == "Human") %>% 
  select(-value_raw, - probe_id) %>% distinct()

human_metadata %>% saveRDS(file = here("tables", "ecoli_infection_human_metadata.rds"))

#Get value from day 0 and calculate Log2FC
human_log2fc_sample = human_expression_metadata %>% 
  #Filter batch
  filter(organism == "Human",
         status == "Healthy") %>% 
  select(probe_id, participant_id = subject_id, status, day0_value_raw = value_raw) %>% 
  group_by(probe_id, status) %>% 
  summarize(n = n(),
            #Mean of healthy
            day0_value_raw = sum(day0_value_raw)/n) %>% 
  select(probe_id, day0_value_raw) %>% 
  #Join by probe id to infected participants
  inner_join(human_expression_metadata %>% 
               filter(site_infection == "urinary tract"), #1 lung, 19 urinary tract),
             by = join_by("probe_id")) %>% 
  mutate(log2fc = value_raw - day0_value_raw) %>% 
  select(probe_id:subject_id, batch:status, log2fc)

#Check duplicates
human_log2fc_sample %>% 
  filter(subject_id == "Ecoli_human_urinary tract_GSM824707",
         probe_id == "1007_s_at") #OK

human_log2fc_sample %>% 
  saveRDS(here("tables", "ecoli_infection_human_log2fc_sample_clean_long.rds"))

```

```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- human_log2fc_sample$probe_id %>% unique()
# listAttributes(ensembl_mart)%>% View()
#Annotate
#[HG-U133A_2] Affymetrix Human Genome U133A 2.0 Array
annot <- getBM(
  attributes = c("affy_hg_u133a_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133a_2",
  values = gene_id,
  mart = ensembl_mart)

#Get annotations for each probe 
### Filter probes with highest median expression
#Get annotations for each probe 
### Filter probes with highest median expression
human_ecoli_log2summary = human_exprs %>% 
  rename(probe = probe_id) %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_hg_u133a_2), 
            by = "probe") %>% 
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(human_symbol = hgnc_symbol) %>% 
  filter(human_symbol != "") %>%
  group_by(entrezgene_id) %>%
  slice_sample(n = 1) %>%  # one line per entrezgene_id
  ungroup() %>%
  distinct(probe, .keep_all = TRUE)  # remove dup probes

human_ecoli_log2summary %>% 
  get_dupes(probe)



#Filtered probes
human_ecoli_log2_filtered_probes = human_exprs %>%
  rename(probe = probe_id) %>% 
  inner_join(human_ecoli_log2summary, by = "probe") %>%   #214,396
  select(human_symbol, probe,
         c(human_metadata$geo_sample))
#14395
human_ecoli_log2_filtered_probes %>% dim()

human_ecoli_log2_filtered_symbol = human_ecoli_log2_filtered_probes 

human_ecoli_log2_filtered_symbol %>% 
  get_dupes(human_symbol)

#Save 
human_ecoli_log2_filtered_symbol %>% 
  saveRDS(here("tables", "ecoli_infection_human_exprs_hgnc_symbol.rds"))
```




##### Differential expression

###### Create ExpressionSet

```{r}
#Import
human_ecoli_log2_filtered_symbol = readRDS(file = here("tables", "ecoli_infection_human_exprs_hgnc_symbol.rds"))

dim(human_ecoli_log2_filtered_symbol)
  
ecoli_human_metadata = readRDS(file = here("tables", "ecoli_infection_human_metadata.rds")) %>% 
  mutate(timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         age = scale(as.numeric(age)),
         ethnicity = as.factor(ethnicity),
         status = as.factor(status),
         status = fct_relevel(status, "Healthy", "infected"),
         site_infection = as.factor(site_infection)) %>% 
  dplyr::filter(!is.na(age))

dim(ecoli_human_metadata)

#Create Expression Set
eset_human_ecoli <- ExpressionSet(
  assayData = human_ecoli_log2_filtered_symbol %>% 
    column_to_rownames("human_symbol") %>% 
    dplyr::select(ecoli_human_metadata$geo_sample) %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(ecoli_human_metadata)) 

#metadata = 62 rows
#exprs data = 62 cols

#Save
eset_human_ecoli %>% saveRDS(file = here("tables", "ecoli_infection_human_eset.rds"))


```



###### Limma
Paired samples design
```{r}

#Import
human_ecoli_log2_filtered_symbol = readRDS(file = here("tables", "ecoli_infection_human_exprs_hgnc_symbol.rds"))
ecoli_human_metadata = readRDS(file = here("tables", "ecoli_infection_human_metadata.rds"))
eset_human_ecoli = readRDS(file = here("tables", "ecoli_infection_human_eset.rds"))


#Create model matrix
design <- model.matrix(~ age + sex + ethnicity + status,
                       data = pData(eset_human_ecoli))

colnames(design) <- make.names(colnames(design))  # Avoids blank spaces
rownames(design) <- rownames(pData(eset_human_ecoli))

#Fit Limma model
fit <- lmFit(eset_human_ecoli, 
             design) 

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(statusinfected, levels = design)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- limma::eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_human_ecoli= all_contrasts_degs %>% 
  dplyr::select(comparison, everything()) %>% 
  dplyr::mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  dplyr::mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  dplyr::mutate(condition = case_when(comparisonA == "statusinfected"	~ "Ecoli_24h",
                               T~comparisonA),
         organism = "Human",
        timepoint = str_remove(condition, "Ecoli_") %>% str_remove("h")) %>% 
  dplyr::select(condition, everything()) %>% 
  clean_names() %>% 
  dplyr::rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs_human_ecoli %>% saveRDS(file = here("tables", "ecoli_infection_human_dge_limma_degs.rds"))

all_degs_human_ecoli %>% 
  dplyr::group_by(condition) %>%
  dplyr::filter(adj_p_val <= 0.05) %>% 
  dplyr::summarize(n_degs = n())

```




###### T test
```{r}
#Import
human_exprs = readRDS(file = here("tables", "ecoli_infection_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "ecoli_infection_human_metadata.rds"))
log2fc_sample = readRDS(here("tables", "ecoli_infection_human_log2fc_sample_clean_long.rds"))
log2fc_sample_symbol = log2fc_sample %>% 
  inner_join(human_exprs %>% 
               select(human_symbol, probe_id = probe),
             by = "probe_id")


#Compute summary statistics
log2fc_summary = log2fc_sample_symbol %>% 
  group_by(human_symbol) %>%
  summarize(
    avg_log2fc = mean(log2fc, na.rm = TRUE),  # mean log2 fold change
    n_subjects = n(),                         # number of samples
    sd = sd(log2fc, na.rm = TRUE),             # standard deviation
    se = sd / sqrt(n_subjects),               # standard error
    effect_size = avg_log2fc / sd)            # Cohen’s d-like effect size

log2_hgnc_symbol_wide = log2fc_sample_symbol %>%
  ungroup() %>% 
  inner_join(human_metadata %>% 
               select(geo_sample, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, subject_id, timepoint, day0_value_raw, value_raw) %>% 
  distinct() %>% 
  select(subject_id, human_symbol, timepoint, hour_0 = day0_value_raw, value = value_raw) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "hour_")
```


```{r}
#Unpaired
log2fc_unpaired_ttest = log2_hgnc_symbol_wide %>%
  group_by(human_symbol) %>%
  
  summarise(pval_hour24 = tryCatch(t.test(hour_24, hour_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    .groups = 'drop'
  ) %>%
  
  mutate(padj_hour24 = p.adjust(pval_hour24, method = "fdr")) %>% 
  pivot_longer(
    cols = starts_with(c("pval_", "padj_")),
    names_to = c(".value", "timepoint"),
    names_pattern = "(pval|padj)_(hour\\d+)"
  ) %>% 
  mutate(timepoint = str_remove(timepoint, "hour"),
         test = "t-Test",
         paired = "Unpaired") 


log2fc_unpaired_ttest %>% 
  saveRDS(file = here("tables", "ecoli_infection_human_log2fc_unpaired_ttest.rds"))

```



###### Compare DGE methods

```{r}
#Import ------
#t-Test
human_log2fc_unpaired_ttest =
  readRDS(file = here("tables", "ecoli_infection_human_log2fc_unpaired_ttest.rds")) %>% 
  mutate(organism = "Human")

#Limma
human_degs = readRDS(file = here("tables", "ecoli_infection_human_dge_limma_degs.rds")) %>% 
  clean_names() %>% 
  rename(padj = adj_p_val) %>% 
  mutate(organism = "Human",
         test = "Limma",
         paired = "Unpaired") 


#Merge datasets
degs_ttest_limma =
  bind_rows(
    human_log2fc_unpaired_ttest,
    human_degs
    )

#Summarize  
degs_ttest_limma_summary = degs_ttest_limma %>% 
  group_by(organism, timepoint, test, paired) %>% 
  summarise(
    n_degs = sum(padj <= 0.05, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) 

degs_ttest_limma_summary

```


```{r fig.height=5, fig.width=5}
# Plot number of DEGs per test and significance level
degs_ttest_limma_summary_plot_gg <- degs_ttest_limma_summary %>%
  ggplot(aes(
    x = timepoint,
    y = n_degs,
    fill = test,
    alpha = paired
  )) +
  geom_col(width = 0.8, position = position_dodge2()) +
  
  # Manual colors per test
  scale_fill_manual(values = c(
    "t-Test" = "gray25",
    "Limma" = "#4DBBD5FF"
  )) +
  
  # Alpha: paired = 1, unpaired = 0.5
  scale_alpha_manual(values = c("Unpaired" = 0.5)) +
  
  # Continuous x-axis for numeric timepoints
  scale_x_discrete(
    breaks = sort(unique(degs_ttest_limma_summary$timepoint)),
    expand = c(0.01, 0.01)
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  
  theme_vaxgo() +
  labs(
    title = "Total DEGs by test and timepoint",
    subtitle = "Comparison across tests (FDR ≤ 0.05)",
    x = "Day",
    y = "Number of DEGs",
    fill = "Test",
    alpha = "Pairing"
  ) +
  facet_wrap(~organism, scales = "free_y") +
  theme(
    legend.position = "top",
    axis.line.y = element_blank(),
    panel.grid.major.y = element_line(color = "gray75", linetype = 3),
    axis.ticks.y = element_line(color = "gray75")
  )

# Render plot
degs_ttest_limma_summary_plot_gg

degs_ttest_limma_summary_plot_gg %>% 
  ggsave(filename = here("Figures", "ecoli_infection_human_dge_ttest_limma_summary.png"),
         width = 5,
         height = 5)
```





## Combine datasets

```{r}

#Merge DGE results
all_degs_mouse = read_rds(here("tables", "ecoli_infection_mouse_dge_limma_degs.rds"))
all_degs_human = read_rds(here("tables", "ecoli_infection_human_dge_limma_degs.rds"))
mouse_log2fc_sample = readRDS(here("tables", "ecoli_infection_mouse_log2fc_sample_clean_long.rds")) 
mouse_exprs = readRDS(file = here("tables", "ecoli_infection_mouse_exprs_hgnc_symbol_long.rds"))
human_log2fc_sample = readRDS(here("tables", "ecoli_infection_human_log2fc_sample_clean_long.rds")) 
human_exprs = readRDS(file = here("tables", "ecoli_infection_human_exprs_hgnc_symbol.rds"))

#Merge DGE results ===========
all_degs_human_mouse = all_degs_mouse %>% 
  bind_rows(all_degs_human) %>% 
  select(condition, timepoint, human_symbol, mean_l2fc:organism)

#Save
all_degs_human_mouse %>% saveRDS(here("tables", "ecoli_infection_human_mouse_dge_limma_degs.rds"))

#Merge sample log2fc =============
#Mouse

mouse_log2fc_sample_symbol = mouse_log2fc_sample %>% 
  inner_join(mouse_exprs %>% 
               select(human_symbol, probe_id = probe),
             by = "probe_id") %>% 
  mutate(organism = "Mouse")

human_log2fc_sample_symbol = human_log2fc_sample %>% 
  inner_join(human_exprs %>% 
               select(human_symbol, probe_id = probe),
             by = "probe_id") %>% 
  mutate(organism = "Human",
         timepoint = "24")

human_mouse_ecoli_log2fc_samples = bind_rows(mouse_log2fc_sample_symbol, human_log2fc_sample_symbol) %>% 
  ungroup()

human_mouse_ecoli_log2fc_samples %>% 
  saveRDS(here("tables", "ecoli_infection_human_mouse_samples_log2fc.rds"))
```




