---
title: "Preprocessing (Refactored)"
output: html_document
---

# Required

```{r message=FALSE, warning=FALSE}
library(here)
source(here("scripts_notebooks", "required.R"))
```

# Functions

```{r, include=FALSE}

# ──────────────────────────────────────────────
# A. PROBE ANNOTATION & GENE COLLAPSE (MOUSE)
# ──────────────────────────────────────────────

annotate_mouse_probes <- function(exprs_matrix, probe_attribute, output_prefix) {
  ensembl_mart <- useMart("ensembl")
  ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

  gene_id <- exprs_matrix %>% rownames()

  annot <- getBM(
    attributes = c(probe_attribute,
                   "ensembl_gene_id",
                   "entrezgene_id",
                   "mgi_symbol",
                   "description"),
    filters = probe_attribute,
    values = gene_id,
    mart = ensembl_mart)

  annot %>% saveRDS(here("tables", paste0(output_prefix, "_probe_annot.rds")))
  beep()

  library(FIT.mouse2man)
  HS_MM_Symbol_Entrez <- FIT.mouse2man::HS_MM_Symbol_Entrez %>%
    clean_names() %>%
    rename(mouse_entrez = mouse_ortholog)

  probe_col <- setNames("probe", probe_attribute)

  log2summary <- exprs_matrix %>%
    as.data.frame() %>%
    rownames_to_column("probe") %>%
    pivot_longer(cols = -probe, names_to = "sample", values_to = "expression") %>%
    group_by(probe) %>%
    summarise(mean = mean(expression, na.rm = TRUE),
              median = median(expression, na.rm = TRUE)) %>%
    inner_join(annot %>% rename(!!probe_col), by = "probe") %>%
    group_by(ensembl_gene_id) %>%
    slice_max(median, n = 1) %>%
    ungroup() %>%
    select(-ensembl_gene_id) %>%
    distinct() %>%
    group_by(entrezgene_id) %>%
    slice_max(median, n = 1) %>%
    ungroup() %>%
    rename(mouse_entrez = entrezgene_id) %>%
    inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>%
    group_by(hgnc_symbol) %>%
    slice_max(median, n = 1) %>%
    ungroup()

  hgnc_dupes <- log2summary %>% get_dupes(hgnc_symbol)
  log2summary <- log2summary %>%
    filter(!hgnc_symbol %in% unique(hgnc_dupes$hgnc_symbol))

  exprs_hgnc <- exprs_matrix %>%
    as.data.frame() %>%
    rownames_to_column("probe") %>%
    inner_join(log2summary %>% select(probe, hgnc_symbol), by = "probe") %>%
    select(human_symbol = hgnc_symbol, everything(), -probe) %>%
    distinct()

  exprs_hgnc %>% saveRDS(here("tables", paste0(output_prefix, "_exprs_hgnc.rds")))

  list(exprs = exprs_hgnc, summary = log2summary, annot = annot)
}


# ──────────────────────────────────────────────
# B. HUMAN PROBE ANNOTATION (platform file)
# ──────────────────────────────────────────────

annotate_human_probes <- function(exprs_matrix, platform_file, sample_ids) {
  probes_annotation <- read.delim(here("tables", platform_file),
                                  skip = 19, header = TRUE, sep = "\t", quote = "") %>%
    select(gene_id = ID, entrez_id = ENTREZ_GENE_ID,
           ensembl_id = ENSEMBL_ID, gene_symbol = GENE_SYMBOL) %>%
    mutate(gene_id = as.character(gene_id))

  counts <- exprs_matrix %>%
    as.data.frame() %>%
    rownames_to_column("gene_id") %>%
    inner_join(probes_annotation, by = "gene_id") %>%
    select(gene_id, entrez_id, ensembl_id, gene_symbol, all_of(sample_ids)) %>%
    mutate(entrez_id = as.character(entrez_id))

  counts_genes <- counts %>%
    pivot_longer(cols = -c(gene_id, entrez_id, ensembl_id, gene_symbol),
                 names_to = "sample", values_to = "expression") %>%
    group_by(entrez_id, gene_id) %>%
    drop_na(entrez_id) %>%
    summarise(median = median(expression),
              median = round(median, 10)) %>%
    slice_max(median, n = 1) %>%
    ungroup()

  counts_raw <- counts %>%
    inner_join(counts_genes %>% select(gene_id, entrez_id),
               by = join_by("gene_id", "entrez_id"))

  dupes_entrez <- counts_genes %>%
    get_dupes(entrez_id) %>%
    group_by(entrez_id) %>%
    slice_sample(n = 1)

  exprs_entrez <- counts_raw %>%
    select(-gene_symbol) %>%
    filter(!gene_id %in% dupes_entrez$gene_id) %>%
    distinct() %>%
    select(entrez_id, all_of(sample_ids)) %>%
    column_to_rownames("entrez_id")

  counts_genes_symbol <- counts %>%
    pivot_longer(cols = -c(gene_id, entrez_id, ensembl_id, gene_symbol),
                 names_to = "sample", values_to = "expression") %>%
    group_by(gene_symbol, gene_id) %>%
    filter(gene_symbol != "") %>%
    drop_na(gene_symbol) %>%
    summarise(median = median(expression), median = round(median, 10)) %>%
    slice_max(median, n = 1) %>%
    ungroup()

  counts_raw_symbol <- counts %>%
    inner_join(counts_genes_symbol %>% select(gene_id, gene_symbol),
               by = join_by("gene_id", "gene_symbol"))

  dupes_symbol <- counts_genes_symbol %>%
    get_dupes(gene_symbol) %>%
    group_by(gene_symbol) %>%
    slice_sample(n = 1)

  exprs_symbol <- counts_raw_symbol %>%
    filter(!gene_symbol %in% dupes_symbol$gene_symbol) %>%
    distinct() %>%
    select(gene_symbol, all_of(sample_ids)) %>%
    column_to_rownames("gene_symbol")

  symbol_entrez <- counts_raw_symbol %>%
    filter(!gene_symbol %in% dupes_symbol$gene_symbol) %>%
    distinct() %>%
    select(symbol = gene_symbol, entrez_id)

  exprs_symbol %>% saveRDS(here("tables", paste0(output_prefix, "_exprs.rds")))

  list(exprs = exprs_symbol, entrez = exprs_entrez, symbol_entrez = symbol_entrez)
}


# ──────────────────────────────────────────────
# C. EXPRESSION SET CREATION
# ──────────────────────────────────────────────

create_eset <- function(exprs_hgnc, metadata, symbol_col = "human_symbol",
                         id_col = "geo_sample", normalize = FALSE) {
  expr_matrix <- exprs_hgnc %>%
    as.data.frame() %>%
    column_to_rownames(symbol_col) %>%
    as.matrix()

  if (normalize) {
    expr_matrix <- normalizeBetweenArrays(expr_matrix, method = "quantile")
  }

  pdata <- metadata %>%
    mutate(across(where(is.character), as.factor)) %>%
    column_to_rownames(id_col)

  eset <- ExpressionSet(
    assayData = expr_matrix,
    phenoData = AnnotatedDataFrame(pdata))

  eset
}


# ──────────────────────────────────────────────
# D. LOG2 FOLD-CHANGE (LONGITUDINAL / PAIRED)
# ──────────────────────────────────────────────

compute_log2fc_longitudinal <- function(exprs_hgnc, metadata, symbol_col = "human_symbol",
                                         id_col = "geo_sample",
                                         participant_col = "participant_id",
                                         timepoint_col = "timepoint",
                                         baseline = "0",
                                         prefix = "day_",
                                         keep_probe = NULL) {
  id_vars <- symbol_col
  if (!is.null(keep_probe)) id_vars <- c(id_vars, keep_probe)

  long <- exprs_hgnc %>%
    as.data.frame() %>%
    pivot_longer(cols = -all_of(id_vars), names_to = id_col, values_to = "log2") %>%
    inner_join(metadata %>% select(all_of(c(id_col, participant_col, timepoint_col))),
               by = id_col) %>%
    select(all_of(c(participant_col, symbol_col, timepoint_col)), value = log2) %>%
    distinct() %>%
    pivot_wider(names_from = all_of(timepoint_col), values_from = value, names_prefix = prefix) %>%
    pivot_longer(cols = starts_with(prefix),
                 names_to = timepoint_col, names_prefix = prefix, values_to = "value") %>%
    group_by(!!sym(participant_col), !!sym(symbol_col)) %>%
    mutate(day0 = value[!!sym(timepoint_col) == baseline]) %>%
    ungroup() %>%
    mutate(log2fc = value - day0) %>%
    filter(!!sym(timepoint_col) != baseline)

  wide <- long %>%
    select(-value, -day0) %>%
    pivot_wider(names_from = all_of(participant_col), values_from = "log2fc")

  list(long = long, wide = wide)
}


# ──────────────────────────────────────────────
# E. LOG2 FOLD-CHANGE (INDEPENDENT / UNPAIRED)
# ──────────────────────────────────────────────

compute_log2fc_independent <- function(exprs_long_format, metadata, control_condition,
                                        value_col = "value_raw", id_cols = c("probe_id"),
                                        group_col = "treatment", subject_col = "participant_id") {
  controls <- exprs_long_format %>%
    filter(!!sym(group_col) == control_condition) %>%
    group_by(!!!syms(id_cols)) %>%
    summarise(day0_value_raw = mean(!!sym(value_col), na.rm = TRUE)) %>%
    ungroup()

  log2fc <- exprs_long_format %>%
    filter(!!sym(group_col) != control_condition) %>%
    inner_join(controls, by = id_cols) %>%
    mutate(log2fc = !!sym(value_col) - day0_value_raw) %>%
    distinct()

  log2fc
}


# ──────────────────────────────────────────────
# F. LIMMA (LONGITUDINAL / PAIRED)
# ──────────────────────────────────────────────

run_limma_longitudinal <- function(eset, participant_col = "participant_id",
                                    timepoint_col = "timepoint",
                                    condition_col = "condition",
                                    contrasts_list,
                                    condition_map,
                                    output_prefix,
                                    min_log2 = 6, min_samples = 5,
                                    covariate_formula = NULL) {
  genes_to_keep <- exprs(eset) %>%
    as.data.frame() %>%
    mutate(across(everything(), ~ . > min_log2)) %>%
    rowSums() %>%
    { . >= min_samples }

  eset_filtered <- eset[genes_to_keep, ]

  if (is.null(covariate_formula)) {
    design <- model.matrix(~ . + 0,
                           data = pData(eset_filtered)[, timepoint_col, drop = FALSE])
  } else {
    design <- model.matrix(covariate_formula, data = pData(eset_filtered))
  }
  colnames(design) <- make.names(colnames(design))

  corfit <- duplicateCorrelation(eset_filtered, design = design,
                                  block = pData(eset_filtered)[[participant_col]])

  aw <- arrayWeights(eset_filtered, design)
  barplot(aw, names.arg = pData(eset_filtered)[[condition_col]],
          main = "Sample Weights Diagnostic", ylab = "Weight",
          las = 2, cex.names = 0.7, col = "skyblue")

  fit <- lmFit(eset_filtered, design, weights = aw,
               block = pData(eset_filtered)[[participant_col]],
               correlation = corfit$consensus)

  cont.matrix <- makeContrasts(contrasts = contrasts_list, levels = design)
  fit2 <- contrasts.fit(fit, cont.matrix)
  fit2 <- eBayes(fit2, robust = TRUE, trend = TRUE)

  plotSA(fit2, main = "Sigma vs Abundance (Weights Applied)")

  results_list <- lapply(1:ncol(fit2), function(i) {
    topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
      mutate(comparison = colnames(fit2)[i]) %>%
      rownames_to_column("symbol")
  })

  all_degs <- bind_rows(results_list) %>%
    select(comparison, everything()) %>%
    clean_names() %>%
    separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef"), remove = FALSE) %>%
    mutate(condition = case_when(!!!condition_map, TRUE ~ comparisonA),
           organism = "Mouse") %>%
    select(condition, everything()) %>%
    rename(human_symbol = symbol, mean_l2fc = log_fc,
           ci_lower = ci_l, ci_upper = ci_r)

  all_degs %>% saveRDS(here("tables", paste0(output_prefix, "_dge_limma_degs.rds")))

  all_degs %>%
    group_by(condition) %>%
    filter(adj_p_val <= 0.05) %>%
    summarise(n_degs = n()) %>%
    print()

  all_degs
}


# ──────────────────────────────────────────────
# G. LIMMA (INDEPENDENT)
# ──────────────────────────────────────────────

run_limma_independent <- function(eset, design_formula, contrasts_list,
                                   condition_map, output_prefix,
                                   min_log2 = 6, min_samples = 5,
                                   block_col = NULL, var_group_col = NULL) {
  genes_to_keep <- exprs(eset) %>%
    as.data.frame() %>%
    mutate(across(everything(), ~ . > min_log2)) %>%
    rowSums() %>%
    { . >= min_samples }

  eset_filtered <- eset[genes_to_keep, ]

  design <- model.matrix(design_formula, data = pData(eset_filtered))
  colnames(design) <- make.names(colnames(design))

  aw <- arrayWeights(eset_filtered, design,
                     var.group = if (!is.null(var_group_col)) pData(eset_filtered)[[var_group_col]] else NULL)

  barplot(aw, names.arg = pData(eset_filtered)$condition,
          main = "Sample Weights Diagnostic", ylab = "Weight",
          las = 2, cex.names = 0.7, col = "skyblue")
    fit <- lmFit(eset_filtered, design, weights = aw)

  cont.matrix <- makeContrasts(contrasts = contrasts_list, levels = design)
  fit2 <- contrasts.fit(fit, cont.matrix)
  fit2 <- eBayes(fit2, robust = TRUE, trend = TRUE)

  plotSA(fit2, main = "Sigma vs Abundance (Weights Applied)")

  results_list <- lapply(1:ncol(fit2), function(i) {
    topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
      mutate(comparison = colnames(fit2)[i]) %>%
      rownames_to_column("symbol")
  })

  all_degs <- bind_rows(results_list) %>%
    select(comparison, everything()) %>%
    clean_names() %>%
    mutate(condition = case_when(!!!condition_map, TRUE ~ comparison),
           organism = "Mouse") %>%
    select(condition, everything()) %>%
    rename(human_symbol = symbol, mean_l2fc = log_fc,
           ci_lower = ci_l, ci_upper = ci_r)

  all_degs %>% saveRDS(here("tables", paste0(output_prefix, "_dge_limma_degs.rds")))

  all_degs %>%
    group_by(condition) %>%
    filter(adj_p_val <= 0.05) %>%
    summarise(n_degs = n()) %>%
    print()

  all_degs
}


# ──────────────────────────────────────────────
# H. T-TEST (LONGITUDINAL)
# ──────────────────────────────────────────────

run_ttest_longitudinal <- function(exprs_hgnc, metadata, symbol_col = "human_symbol",
                                    id_col = "geo_sample",
                                    participant_col = "participant_id",
                                    timepoint_col = "timepoint",
                                    timepoint_names, prefix = "day_",
                                    output_prefix, organism = "Mouse") {
  wide <- exprs_hgnc %>%
    as.data.frame() %>%
    pivot_longer(cols = -all_of(symbol_col), names_to = id_col, values_to = "log2") %>%
    inner_join(metadata %>% select(all_of(c(id_col, participant_col, timepoint_col))),
               by = id_col) %>%
    select(all_of(c(participant_col, symbol_col, timepoint_col)), value = log2) %>%
    distinct() %>%
    pivot_wider(names_from = all_of(timepoint_col), values_from = value, names_prefix = prefix)

  ttest_wrapper <- function(cols, paired = TRUE) {
    wide %>%
      group_by(!!sym(symbol_col)) %>%
      summarise(!!!setNames(
        lapply(names(timepoint_names), function(tp) {
          expr <- paste0("tryCatch(t.test(", prefix, tp, ", ", prefix, "0, paired = ",
                         tolower(paired), ", var.equal = FALSE, alternative = 'two.sided')$p.value,
                         error = function(e) NA_real_)")
          parse(text = expr)
        }),
        paste0("pval_", prefix, names(timepoint_names))
      ), .groups = "drop") %>%
      mutate(!!!setNames(
        lapply(names(timepoint_names), function(tp) {
          expr <- paste0("p.adjust(pval_", prefix, tp, ", method = 'fdr')")
          parse(text = expr)
        }),
        paste0("padj_", prefix, names(timepoint_names))
      )) %>%
      pivot_longer(
        cols = starts_with(c("pval_", "padj_")),
        names_to = c(".value", "timepoint"),
        names_pattern = paste0("(pval|padj)_", prefix, "(.+)")
      ) %>%
      mutate(timepoint = str_remove(timepoint, prefix),
             test = "t-Test", paired = ifelse(paired, "Paired", "Unpaired"))
  }

  paired_res <- ttest_wrapper(timepoint_names, paired = TRUE)
  unpaired_res <- ttest_wrapper(timepoint_names, paired = FALSE)

  combined <- bind_rows(paired_res, unpaired_res)
  combined %>% saveRDS(here("tables", paste0(output_prefix, "_paired_unpaired_ttest.rds")))

  combined
}


# ──────────────────────────────────────────────
# I. SUBSAMPLING (LONGITUDINAL)
# ──────────────────────────────────────────────

subsampling_longitudinal <- function(eset, metadata, n_iterations = 200, cohort_n = 5,
                                      participant_col = "participant_id",
                                      timepoint_col = "timepoint",
                                      condition_col = "condition",
                                      contrasts_list, condition_map,
                                      output_prefix, seed = 123) {
  genes_to_keep <- exprs(eset) %>%
    as.data.frame() %>%
    mutate(across(everything(), ~ . > 6)) %>%
    rowSums() %>%
    { . >= 5 }

  eset_filtered <- eset[genes_to_keep, ]
  meta_df <- as_tibble(pData(eset_filtered), rownames = "sample_id")
  expr_matrix <- exprs(eset_filtered)
  unique_participants <- unique(meta_df[[participant_col]])

  draw_flexible_subset <- function(participants_list, metadata, target_n = 5) {
    valid_draw <- FALSE
    attempts <- 0
    while (!valid_draw) {
      attempts <- attempts + 1
      if (attempts > 500) stop("Could not find a balanced subset.")
      selected_ids <- sample(participants_list, size = target_n, replace = FALSE)
      sub_meta <- metadata %>% filter(!!sym(participant_col) %in% selected_ids)
      has_all_timepoints <- length(unique(sub_meta[[timepoint_col]])) == length(unique(metadata[[timepoint_col]]))
      if (has_all_timepoints) valid_draw <- TRUE
    }
    sub_meta
  }

  set.seed(seed)

  df_global_runs <- tibble(iteration = 1:n_iterations) %>%
    mutate(
      sampled_metadata = map(iteration, ~ draw_flexible_subset(unique_participants, meta_df, cohort_n)),
      limma_results = map(sampled_metadata, ~ {
        sub_meta <- .x
        sub_expr <- expr_matrix[, sub_meta$sample_id]

        design_sub <- model.matrix(as.formula(paste0("~ ", timepoint_col, " + 0")), data = sub_meta)
        colnames(design_sub) <- make.names(colnames(design_sub))

        corfit_sub <- duplicateCorrelation(sub_expr, design = design_sub,
                                            block = sub_meta[[participant_col]])
        aw_sub <- arrayWeights(sub_expr, design_sub)

        fit_sub <- lmFit(sub_expr, design = design_sub, weights = aw_sub,
                         block = sub_meta[[participant_col]],
                         correlation = corfit_sub$consensus)

        cont.matrix <- makeContrasts(contrasts = contrasts_list, levels = design_sub)
        fit2_sub <- contrasts.fit(fit_sub, cont.matrix)
        fit2_sub <- eBayes(fit2_sub, robust = TRUE, trend = TRUE)

        res_list <- lapply(1:ncol(fit2_sub), function(i) {
          df_res <- topTreat(fit2_sub, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE)
          df_res$symbol <- rownames(df_res)
          df_res$comparison <- colnames(fit2_sub)[i]
          rownames(df_res) <- NULL
          df_res
        })
        do.call(rbind, res_list)
      })
    )

  subsampled_degs <- df_global_runs %>%
    select(iteration, limma_results) %>%
    unnest(limma_results) %>%
    clean_names() %>%
    mutate(
      se_iteration = abs(log_fc / t),
      sd_iteration = se_iteration / sqrt(0.4)) %>%
    group_by(comparison, symbol) %>%
    summarise(
      p_value = median(p_value),
      adj_p_val = median(adj_p_val),
      mean_l2fc = median(log_fc),
      sd_sampling_stability = sd(log_fc),
      se = median(se_iteration),
      sd = median(sd_iteration),
      ci_lower = median(ci_l),
      ci_upper = median(ci_r),
      t = median(t),
      b = median(b),
      ave_expr = median(ave_expr),
      pct_iterations_significant = mean(adj_p_val <= 0.05) * 100,
      .groups = "drop") %>%
    mutate(
      comparison = str_replace_all(comparison, "\\.\\.\\.", "_"),
      condition = case_when(!!!condition_map, TRUE ~ comparison),
      organism = "Human") %>%
    select(condition, everything()) %>%
    clean_names() %>%
    rename(human_symbol = symbol)

  subsampled_degs %>% saveRDS(here("tables", paste0(output_prefix, "_dge_subsampled_degs.rds")))

  subsampled_degs %>%
    group_by(condition) %>%
    filter(adj_p_val <= 0.05) %>%
    summarise(n_robust_degs = n()) %>%
    print()

  list(degs = subsampled_degs, runs = df_global_runs)
}


# ──────────────────────────────────────────────
# J. SUBSAMPLING (INDEPENDENT)
# ──────────────────────────────────────────────

subsampling_independent <- function(eset, metadata, n_iterations = 200, cohort_n = 10,
                                     balance_col = "pathogen",
                                     design_formula, contrasts_list,
                                     output_prefix, seed = 123,
                                     block_col = NULL, var_group_col = NULL) {
  genes_to_keep <- exprs(eset) %>%
    as.data.frame() %>%
    mutate(across(everything(), ~ . > 6)) %>%
    rowSums() %>%
    { . >= 5 }

  eset_filtered <- eset[genes_to_keep, ]
  meta_df <- as_tibble(pData(eset_filtered), rownames = "sample_id")
  expr_matrix <- exprs(eset_filtered)

  draw_flexible_subset <- function(metadata, target_n = 10) {
    valid_draw <- FALSE
    attempts <- 0
    while (!valid_draw) {
      attempts <- attempts + 1
      if (attempts > 500) stop("Could not find a balanced subset.")
      sub_meta <- metadata %>%
        group_by(!!sym(balance_col)) %>%
        sample_n(target_n) %>%
        ungroup()
      has_diversity <- length(unique(sub_meta[[balance_col]])) > 1
      if (has_diversity) valid_draw <- TRUE
    }
    sub_meta
  }

  set.seed(seed)

  df_global_runs <- tibble(iteration = 1:n_iterations) %>%
    mutate(
      sampled_metadata = map(iteration, ~ draw_flexible_subset(meta_df, cohort_n)),
      limma_results = map(sampled_metadata, ~ {
        sub_meta <- .x
        sub_expr <- expr_matrix[, sub_meta$sample_id]

        design_sub <- model.matrix(design_formula, data = sub_meta)
        colnames(design_sub) <- make.names(colnames(design_sub))

        aw_sub <- arrayWeights(sub_expr, design_sub,
                                var.group = if (!is.null(var_group_col)) sub_meta[[var_group_col]] else NULL)

        if (!is.null(block_col)) {
          corfit_sub <- duplicateCorrelation(sub_expr, design = design_sub,
                                              block = sub_meta[[block_col]])
          fit_sub <- lmFit(sub_expr, design = design_sub, weights = aw_sub,
                           block = sub_meta[[block_col]],
                           correlation = corfit_sub$consensus)
        } else {
          fit_sub <- lmFit(sub_expr, design = design_sub, weights = aw_sub)
        }

        cont.matrix <- makeContrasts(contrasts = contrasts_list, levels = design_sub)
        fit2_sub <- contrasts.fit(fit_sub, cont.matrix)
        fit2_sub <- eBayes(fit2_sub, robust = TRUE, trend = TRUE)

        res_list <- lapply(1:ncol(fit2_sub), function(i) {
          df_res <- topTreat(fit2_sub, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE)
          df_res$symbol <- rownames(df_res)
          df_res$comparison <- colnames(fit2_sub)[i]
          rownames(df_res) <- NULL
          df_res
        })
        do.call(rbind, res_list)
      })
    )

  subsampled_degs <- df_global_runs %>%
    select(iteration, limma_results) %>%
    unnest(limma_results) %>%
    clean_names() %>%
    mutate(
      se_iteration = abs(log_fc / t),
      sd_iteration = se_iteration / sqrt(0.4)) %>%
    group_by(comparison, symbol) %>%
    summarise(
      p_value = median(p_value),
      adj_p_val = median(adj_p_val),
      mean_l2fc = median(log_fc),
      sd_sampling_stability = sd(log_fc),
      se = median(se_iteration),
      sd = median(sd_iteration),
      ci_lower = median(ci_l),
      ci_upper = median(ci_r),
      t = median(t),
      b = median(b),
      ave_expr = median(ave_expr),
      pct_iterations_significant = mean(adj_p_val <= 0.05) * 100,
      .groups = "drop") %>%
    mutate(organism = "Human") %>%
    rename(human_symbol = symbol)

  subsampled_degs %>% saveRDS(here("tables", paste0(output_prefix, "_dge_subsampled_degs.rds")))

  list(degs = subsampled_degs, runs = df_global_runs)
}


# ──────────────────────────────────────────────
# K. DEMOGRAPHICS PLOT (SUBSAMPLING)
# ──────────────────────────────────────────────

plot_demographics_subsampling <- function(df_global_runs, original_metadata) {
  demographics_summary <- df_global_runs %>%
    select(iteration, sampled_metadata) %>%
    unnest(sampled_metadata) %>%
    select(iteration, participant_id, age, sex) %>%
    distinct() %>%
    group_by(iteration) %>%
    summarise(mean_age = mean(age, na.rm = TRUE),
              prop_male = mean(sex == "male", na.rm = TRUE),
              .groups = "drop") %>%
    mutate(iteration = factor(iteration, levels = 1:max(iteration)))

  subsample_race_distribution <- df_global_runs %>%
    select(iteration, sampled_metadata) %>%
    unnest(sampled_metadata) %>%
    select(iteration, participant_id, race) %>%
    distinct() %>%
    count(iteration, race) %>%
    group_by(iteration) %>%
    mutate(Proportion = n / sum(n)) %>%
    ungroup() %>%
    mutate(iteration = factor(iteration, levels = 1:max(as.numeric(iteration))))

  plot_age <- ggplot(demographics_summary, aes(x = "Mean Age", y = iteration)) +
    geom_tile(aes(fill = mean_age), color = "white", linewidth = 0.6) +
    geom_text(aes(label = sprintf("%.1f", mean_age)),
              color = "white", fontface = "bold", size = 3.5) +
    scale_fill_viridis_c(option = "mako", name = "Mean Age\n(Years)", direction = -1) +
    scale_x_discrete(position = "top", expand = c(0.01, 0.01)) +
    scale_y_discrete(expand = c(0.01, 0.01)) +
    theme_vaxgo() +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_text(face = "bold"),
          panel.grid = element_blank(),
          legend.position = "left") +
    labs(y = "Subsampling Iterations")

  plot_sex <- demographics_summary %>%
    select(iteration, Male = prop_male) %>%
    mutate(Female = 1 - Male) %>%
    pivot_longer(cols = c(Male, Female), names_to = "Sex", values_to = "Proportion") %>%
    ggplot(aes(x = Proportion, y = iteration, fill = Sex)) +
    geom_col(width = 0.75, position = "stack") +
    scale_fill_manual(values = c("Male" = "#3182bd", "Female" = "#fa9fb5")) +
    scale_x_continuous(position = "top", expand = c(0, 0), breaks = c(0, 0.5, 1)) +
    scale_y_discrete(expand = c(0.01, 0.01)) +
    theme_vaxgo() +
    theme(axis.title.y = element_blank(), axis.text.y = element_blank(),
          panel.grid.minor = element_blank(), panel.grid.major.y = element_blank()) +
    labs(x = "Sex Distribution")

  plot_race <- subsample_race_distribution %>%
    mutate(race = str_replace(race, "black or african american", "black"),
           race = str_to_title(race)) %>%
    ggplot(aes(x = Proportion, y = iteration, fill = race)) +
    geom_col(width = 0.75, position = "stack") +
    scale_fill_manual(name = "Race", values = colors$race) +
    scale_x_continuous(position = "top", expand = c(0, 0), breaks = c(0, 0.5, 1)) +
    scale_y_discrete(expand = c(0.01, 0.01)) +
    theme_vaxgo() +
    theme(axis.title.y = element_blank(), axis.text.y = element_blank(),
          panel.grid.minor = element_blank(), panel.grid.major.y = element_blank()) +
    labs(x = "Race Distribution")

  top_panel <- plot_age + plot_sex + plot_race + plot_layout(widths = c(1, 1, 1))

  original_demographics <- original_metadata %>%
    filter(organism == "Human") %>%
    mutate(age = str_remove(age, "y") %>% as.numeric()) %>%
    select(participant_id, age, sex, race) %>%
    distinct()

  base_age <- ggplot(original_demographics, aes(x = age)) +
    geom_density(fill = "#3182bd", alpha = 0.4, color = "black", linewidth = 0.8) +
    geom_vline(xintercept = mean(original_demographics$age, na.rm = TRUE),
               color = "black", linetype = "dashed", linewidth = 0.8) +
    facet_wrap(~"Age") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_vaxgo() +
    theme(panel.grid.minor = element_blank(),
          axis.title.y = element_text(face = "bold", size = 9),
          axis.title.x = element_text(face = "bold", size = 9)) +
    labs(x = "Age (Years)", y = "Density")

  base_sex <- original_demographics %>%
    count(sex) %>%
    mutate(Proportion = n / sum(n)) %>%
    ggplot(aes(x = Proportion, y = "Original\nCohort", fill = sex)) +
    facet_wrap(~"Original\nCohort") +
    geom_col(width = 0.6, position = "stack", show.legend = FALSE) +
    scale_fill_manual(values = c("Male" = "#3182bd", "Female" = "#fa9fb5")) +
    scale_x_continuous(expand = c(0, 0), breaks = c(0, 0.5, 1)) +
    scale_y_discrete(expand = c(0, 0)) +
    theme_vaxgo() +
    theme(axis.title = element_blank(), axis.text.y = element_blank(),
          panel.grid = element_blank(), axis.ticks.y = element_blank())

  base_race <- original_demographics %>%
    mutate(race = str_replace(race, "black or african american", "black"),
           race = str_to_title(race)) %>%
    count(race) %>%
    mutate(Proportion = n / sum(n)) %>%
    ggplot(aes(x = Proportion, y = "Original\nCohort", fill = race)) +
    facet_wrap(~"Race") +
    geom_col(width = 0.6, position = "stack", show.legend = TRUE) +
    scale_fill_manual(name = "Race", values = colors$race) +
    scale_x_continuous(expand = c(0, 0), breaks = c(0, 0.5, 1)) +
    scale_y_discrete(expand = c(0, 0)) +
    theme_vaxgo() +
    theme(axis.title = element_blank(), axis.text.y = element_blank(),
          panel.grid = element_blank(), axis.ticks.y = element_blank())

  bottom_panel <- base_age + base_sex + base_race + plot_layout(widths = c(1, 1, 1))

  top_panel / bottom_panel + plot_layout(heights = c(3, 1))
}


# ──────────────────────────────────────────────
# L. JOIN HUMAN + MOUSE MATRIX (cross-species)
# ──────────────────────────────────────────────

unite_human_mouse_exprs <- function(human_exprs, mouse_exprs, metadata,
                                     human_col = "symbol", mouse_col = "symbol",
                                     output_prefix) {
  joint_matrix <- human_exprs %>%
    rownames_to_column(human_col) %>%
    inner_join(mouse_exprs %>% rownames_to_column(mouse_col),
               by = human_col) %>%
    column_to_rownames(human_col) %>%
    select(metadata %>% pull(sample_id)) %>%
    as.matrix() %>%
    normalizeBetweenArrays(method = "quantile")

  eset <- ExpressionSet(
    assayData = joint_matrix,
    phenoData = AnnotatedDataFrame(
      metadata %>%
        filter(sample_id %in% colnames(joint_matrix)) %>%
        column_to_rownames("sample_id")))

  eset %>% saveRDS(here("tables", paste0(output_prefix, "_human_mouse_eset.rds")))
  eset
}


# ──────────────────────────────────────────────
# M. RUN FIT PREDICTION
# ──────────────────────────────────────────────

run_fit_prediction <- function(degs_mouse, exprs_hgnc, output_prefix,
                                entrez_col = "mouse_entrez",
                                symbol_col = "human_symbol",
                                effect_col = "mean",
                                condition_col = "condition") {
  fit_input <- degs_mouse %>%
    select(symbol = !!sym(symbol_col), EffectSize = !!sym(effect_col),
           !!sym(condition_col)) %>%
    inner_join(exprs_hgnc %>%
                 select(!!sym(symbol_col), MM.Entrez = !!sym(entrez_col)),
               by = "symbol") %>%
    mutate(row = MM.Entrez)

  all_conditions <- unique(fit_input[[condition_col]])
  walk(all_conditions, function(cond) {
    delete_dupes <- fit_input %>%
      filter(!!sym(condition_col) == cond) %>%
      get_dupes(row) %>%
      pull(row)
    fit_input %>%
      filter(!!sym(condition_col) == cond, !row %in% delete_dupes) %>%
      select(-!!sym(condition_col)) %>%
      column_to_rownames("row") %>%
      select(MM.Entrez, EffectSize) %>%
      write_csv(here("tables", paste0(cond, "_FIT_mouse.csv")))
  })

  fit_results <- map2_dfr(all_conditions, all_conditions, function(cond, cond_name) {
    FIT.mouse2man::FIT(
      MouseFile = here("tables", paste0(cond, "_FIT_mouse.csv")),
      DataType = "rnaseq") %>%
      as_tibble() %>%
      mutate(condition = cond_name)
  })

  beep()
  fit_results %>% saveRDS(here("tables", paste0(output_prefix, "_fit_prediction_alltimes.rds")))
  fit_results
}

```

# ==================================
# Differential gene expression analysis
# 1. Fluad + MF59, Vaccine

## Mouse

### Download dataset

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
GSE120661 <- getGEO("GSE120661", GSEMatrix = TRUE)
eset <- GSE120661[[1]]

GSE120661_metadata <- eset %>%
  pData() %>%
  dplyr::select(
    geo_sample = "geo_accession",
    subject_id = "title",
    sex = "gender:ch1",
    treatment = "treatment:ch1",
    tissue = "tissue:ch1",
    timepoint = "timepoint:ch1",
    animal_replicate = "animal replicate:ch1",
    strain = "strain/background:ch1") %>%
  mutate(treatment = fct_relevel(treatment, "saline"),
         timepoint = fct_relevel(timepoint, "0"),
         condition = paste(treatment, timepoint, tissue, sep = "_"))

write_csv(GSE120661_metadata, here("tables", "influenza_hepatitisb_GSE120661_metadata.csv"))

eset <- exprs(eset)
eset %>%
  as.data.frame() %>%
  saveRDS(here("tables", "influenza_hepatitisb_GSE120661_exprs.rds"))
```

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
GSE120661_exprs <- readRDS(here("tables", "influenza_hepatitisb_GSE120661_exprs.rds"))

GSE120661_exprs %>%
  rownames_to_column("probes") %>%
  pivot_longer(cols = -probes, names_to = "sample_id", values_to = "value") %>%
  filter(sample_id == "GSM3407136") %>%
  ggplot() +
  aes(x = value) +
  geom_histogram()
```

### Prepare metadata

```{r}
GSE120661_metadata <- read_csv(here("tables", "influenza_hepatitisb_GSE120661_metadata.csv"))

GSE120661_metadata <- GSE120661_metadata %>%
  mutate(timepoint = as.integer(timepoint) %>%
           as.character() %>%
           as.factor() %>%
           fct_relevel("0"),
         treatment = fct_relevel(treatment, "saline"),
         condition = paste(treatment, timepoint, tissue, sep = "_"))

GSE120661_metadata %>% saveRDS(here("tables", "influenza_fluad_mouse_metadata.rds"))


### Filter to Fluad
GSE120661_metadata <- readRDS(here("tables", "influenza_fluad_mouse_metadata.rds"))
GSE120661_exprs <- read_rds(here("tables", "influenza_hepatitisb_GSE120661_exprs.rds"))

mouse_fluad_metadata <- GSE120661_metadata %>%
  filter(treatment == "Fluad", tissue == "PBMC") %>%
  mutate(timepoint = as.character(timepoint) %>% as.numeric(),
         timepoint = (timepoint / 24) %>% round(1),
         participant_id = paste0("TriFLU-MF59_BL", animal_replicate)) %>%
  filter(timepoint %in% c(0, 1, 3, 7))

mouse_fluad_metadata %>%
  saveRDS(here("tables", "influenza_fluad_mouse_metadata.rds"))

### Get Expression data
mouse_fluad_exprs <- GSE120661_exprs %>%
  as.data.frame() %>%
  select(mouse_fluad_metadata$geo_sample) %>%
  as.matrix()

mouse_fluad_exprs %>%
  as.data.frame() %>%
  saveRDS(here("tables", "influenza_fluad_mouse_exprs.rds"))
```

### Annotate genes

```{r}
#Import
mouse_fluad_log2_hgnc_symbol <- readRDS(here("tables", "influenza_fluad_mouse_exprs_hgnc.rds"))
mouse_fluad_metadata <- readRDS(here("tables", "influenza_fluad_mouse_metadata.rds"))
GSE120661_exprs <- readRDS(here("tables", "influenza_fluad_mouse_exprs.rds"))

# Annotate genes
mouse_fluad_annot <- annotate_mouse_probes(
  exprs_matrix = GSE120661_exprs,
  probe_attribute = "agilent_sureprint_g3_8x60k",
  output_prefix = "influenza_fluad_mouse")

### ExpressionSet & log2FC
mouse_fluad_log2fc <- compute_log2fc_longitudinal(
  exprs_hgnc = mouse_fluad_log2_hgnc_symbol,
  metadata = mouse_fluad_metadata)

mouse_fluad_log2fc$long %>%
  saveRDS(here("tables", "influenza_fluad_mouse_log2fc_sample_clean_long.rds"))

beep()

### Create ExpressionSet
eset_mouse_fluad <- create_eset(
  exprs_hgnc = mouse_fluad_log2_hgnc_symbol,
  metadata = mouse_fluad_metadata %>%
    mutate(timepoint = as.character(timepoint) %>% as.factor()),
  normalize = TRUE)

eset_mouse_fluad %>% saveRDS(here("tables", "influenza_fluad_mouse_eset.rds"))
```

### Differential expression (Limma)

```{r}
#Import Expression Set
eset_mouse_fluad <- readRDS(here("tables", "influenza_fluad_mouse_eset.rds"))
#Define contrasts
fluad_mouse_limma <- run_limma_longitudinal(
  eset = eset_mouse_fluad,
  participant_col = "participant_id",
  condition_col = "condition",
  contrasts_list = c("timepoint1 - timepoint0",
                     "timepoint3 - timepoint0",
                     "timepoint7 - timepoint0"),
  condition_map = list(
    "timepoint1  - timepoint0" ~ "Fluad_1_PBMC",
    "timepoint3  - timepoint0" ~ "Fluad_3_PBMC",
    "timepoint7  - timepoint0" ~ "Fluad_7_PBMC"),
  output_prefix = "influenza_fluad_mouse")
#Save results
fluad_mouse_limma %>% saveRDS(here("tables", "influenza_fluad_mouse_dge_limma_degs.rds"))
```

### t-Test

```{r}
mouse_fluad_log2_hgnc_symbol <- readRDS(here("tables", "influenza_fluad_mouse_exprs_hgnc.rds"))
mouse_fluad_metadata <- readRDS(here("tables", "influenza_fluad_mouse_metadata.rds"))

fluad_mouse_ttest <- run_ttest_longitudinal(
  exprs_hgnc = mouse_fluad_log2_hgnc_symbol,
  metadata = mouse_fluad_metadata,
  timepoint_names = c("1" = "day1", "3" = "day3", "7" = "day7"))

fluad_mouse_ttest %>%
  saveRDS(here("tables", "influenza_fluad_mouse_log2fc_paired_unpaired_ttest.rds"))
```


## Human

### Download dataset

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
GSE124689 <- getGEO("GSE124689", GSEMatrix = TRUE)
eset <- GSE124689[[1]]
GSE124689_exprs <- eset %>% exprs()

GSE124689_exprs %>%
  as.data.frame() %>%
  saveRDS(here("tables", "influenza_human_GSE124689_exprs.rds"))

### Prepare metadata & annotate features
GSE124689_exprs <- readRDS(here("tables", "influenza_human_GSE124689_exprs.rds"))
flu_expFC <- readRDS(here("tables", "Flu_Human_expFC.rds"))
flu_metadata <- flu_expFC$cur_phenoData

human_samples_pbmc_fluad <- flu_metadata %>%
  filter(exposure_material_reported == "FLUADD")

human_fluad_annot <- annotate_human_probes(
  exprs_matrix = GSE124689_exprs,
  platform_file = "GPL21272-58809.txt",
  sample_ids = human_samples_pbmc_fluad$biosample_accession)

human_fluad_annot$exprs %>%
  saveRDS(here("tables", "influenza_GSE124689_exprs_not_normalized.rds"))
```

### Clean metadata

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
flu_metadata_clean <- flu_metadata %>%
  arrange(uid) %>%
  separate(participant_id, into = c("del", "participant_id")) %>%
  select(-del) %>%
  select(sample_id = uid, participant_id, timepoint = study_time_collected,
         age = age_imputed, sex = gender, race,
         vaccine = exposure_material_reported, sample_type = cell_type,
         array = featureSetName2, type = vaccine_type, adjuvant) %>%
  mutate(age = as.numeric(age), sex = as.factor(sex), race = as.factor(race),
         condition = paste(vaccine, "Day", timepoint)) %>%
  filter(sample_id %in% colnames(GSE124689_exprs)) %>%
  arrange(sample_id) %>%
  select(condition, everything())

flu_metadata_clean %>%
  saveRDS(here("tables", "influenza_fluad_human_metadata_clean.rds"))
```

### Unite and normalize (human + mouse)

```{r}
GSE120661_exprs <- readRDS(here("tables", "influenza_fluad_mouse_exprs.rds"))
GSE124689_exprs <- readRDS(here("tables", "influenza_human_GSE124689_exprs.rds"))
human_mouse_fluad_metadata <- readRDS(here("tables", "fluad_human_mouse_metadata.rds"))

eset_human_mouse <- unite_human_mouse_exprs(
  human_exprs = human_fluad_annot$exprs,
  mouse_exprs = GSE120661_symbol,
  metadata = human_mouse_fluad_metadata,
  output_prefix = "influenza_fluad")
```

### Human-only ExpressionSet & Limma

```{r}
eset_human_mouse <- readRDS(here("tables", "influenza_fluad_human_mouse_eset.rds"))
human_mouse_fluad_matrix <- eset_human_mouse %>% exprs()
human_mouse_fluad_metadata <- readRDS(here("tables", "influenza_fluad_human_mouse_metadata.rds"))

outliers_human <- c("CRC305DGENT1483036", "CRC305DGENT0351013", "CRC305DGENT1471032")

human_samples <- human_mouse_fluad_metadata %>%
  filter(organism == "Human", !participant_id %in% outliers_human)

human_matrix <- human_mouse_fluad_matrix %>%
  as.data.frame() %>%
  select(human_samples$sample_id) %>%
  as.matrix()

human_pData <- human_mouse_fluad_metadata %>%
  column_to_rownames("sample_id") %>%
  filter(organism == "Human", !participant_id %in% outliers_human) %>%
  mutate(timepoint = as.factor(timepoint), sex = as.factor(sex),
         age = scale(as.numeric(age)), race = as.factor(race))

eset_human <- ExpressionSet(
  assayData = human_matrix,
  phenoData = AnnotatedDataFrame(human_pData))

eset_human %>% saveRDS(here("tables", "fluad_eset_human.rds"))

eset_human <- readRDS(here("tables", "fluad_eset_human.rds"))

genes_to_keep <- exprs(eset_human) %>%
  as.data.frame() %>%
  mutate(across(everything(), ~ . > 6)) %>%
  rowSums() %>%
  { . >= 5 }

eset_filtered <- eset_human[genes_to_keep, ]

design <- model.matrix(~ condition + sex + age + race + 0, eset_filtered)
colnames(design) <- make.names(colnames(design))

corfit <- duplicateCorrelation(eset_filtered, design = design,
                                block = eset_filtered$participant_id)

aw <- arrayWeights(eset_filtered, design)
barplot(aw, names.arg = eset_filtered$condition,
        main = "Sample Weights Diagnostic", ylab = "Weight",
        las = 2, cex.names = 0.7, col = "skyblue")

fit <- lmFit(eset_filtered, design, weights = aw,
             block = eset_filtered$participant_id,
             correlation = corfit$consensus)

cont.matrix <- makeContrasts(
  conditionFLUADD.Day.1 - conditionFLUADD.Day.0,
  conditionFLUADD.Day.3 - conditionFLUADD.Day.0,
  conditionFLUADD.Day.7 - conditionFLUADD.Day.0,
  levels = design)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2, robust = TRUE, trend = TRUE)

fit2 %>% saveRDS(here("tables", "fluad_human_limma_fit.rds"))
plotSA(fit2, main = "Sigma vs Abundance (Weights Applied)")

results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>%
    rownames_to_column("symbol")
})

fluad_human_degs <- bind_rows(results_list) %>%
  select(comparison, everything()) %>%
  mutate(comparison = str_replace_all(comparison, "\\.", " "),
         comparison = str_replace_all(comparison, "condition", "")) %>%
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>%
  select(condition = comparisonA, everything(), -comparisonRef) %>%
  clean_names() %>%
  select(condition, symbol, mean = log_fc, lower = ci_l, upper = ci_r,
         ave_expr, everything()) %>%
  inner_join(human_pData %>% select(condition, timepoint, vaccine, sample_type) %>%
               distinct(), by = "condition") %>%
  mutate(organism = "Human")

fluad_human_degs %>% saveRDS(here("tables", "influenza_fluad_human_dge_limma_degs.rds"))
```

### Human subsampling (longitudinal)

```{r}
human_fluad_subsampling <- subsampling_longitudinal(
  eset = eset_human,
  metadata = human_pData %>% rownames_to_column("sample_id"),
  n_iterations = 200,
  cohort_n = 5,
  contrasts_list = c("timepoint1 - timepoint0",
                     "timepoint3 - timepoint0",
                     "timepoint7 - timepoint0"),
  condition_map = list(
    "Time1  - Time0" ~ "Fluad_Day1",
    "Time3  - Time0" ~ "Fluad_Day3",
    "Time7  - Time0" ~ "Fluad_Day7"),
  output_prefix = "influenza_fluad_human")
```

```{r fig.height=5, fig.width=6}
plot_demographics_subsampling(
  df_global_runs = human_fluad_subsampling$runs,
  original_metadata = human_mouse_fluad_metadata)
```


# 2. Engerix B, Vaccine

## Mouse

### Prepare data

```{r}
GSE120661_exprs <- readRDS(here("tables", "influenza_hepatitisb_GSE120661_exprs.rds")) %>%
  as.data.frame()
GSE120661_metadata <- read_csv(here("tables", "influenza_hepatitisb_GSE120661_metadata.csv"))

mouse_engerixb_metadata <- GSE120661_metadata %>%
  filter(treatment == "Engerix B", tissue == "PBMC") %>%
  mutate(timepoint = (timepoint / 24) %>% round(1),
         participant_id = paste0("ENG_BL_", animal_replicate)) %>%
  filter(timepoint %in% c(0, 1, 3, 7))

mouse_engerixb_metadata %>%
  saveRDS(here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))

mouse_engerixb_log2 <- GSE120661_exprs %>%
  select(mouse_engerixb_metadata$geo_sample)
```

### Annotate genes

```{r}
mouse_engerixb_annot <- annotate_mouse_probes(
  exprs_matrix = mouse_engerixb_log2,
  probe_attribute = "agilent_sureprint_g3_ge_8x60k",
  output_prefix = "hepatitisb_engerixb_mouse")
```

### ExpressionSet

```{r}
mouse_engerixb_log2_hgnc_symbol <- readRDS(here("tables", "hepatitisb_engerixb_mouse_exprs_hgnc.rds"))
mouse_engerixb_metadata <- readRDS(here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))

eset_mouse_engerixb <- create_eset(
  exprs_hgnc = mouse_engerixb_log2_hgnc_symbol,
  metadata = mouse_engerixb_metadata)

eset_mouse_engerixb %>%
  saveRDS(here("tables", "hepatitisb_engerixb_mouse_eset.rds"))
```

### log2FC

```{r}
mouse_engerixb_log2_hgnc_symbol <- readRDS(here("tables", "hepatitisb_engerixb_mouse_exprs_hgnc.rds"))
mouse_engerixb_metadata <- readRDS(here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))

mouse_engerixb_log2fc <- compute_log2fc_longitudinal(
  exprs_hgnc = mouse_engerixb_log2_hgnc_symbol,
  metadata = mouse_engerixb_metadata)

mouse_engerixb_log2fc$long %>%
  saveRDS(here("tables", "hepatitisb_engerixb_mouse_log2fc_sample_clean_long.rds"))

beep()
```

### Limma

```{r}
eset_mouse_engerixb <- readRDS(here("tables", "hepatitisb_engerixb_mouse_eset.rds"))

engerixb_mouse_limma <- run_limma_longitudinal(
  eset = eset_mouse_engerixb,
  participant_col = "participant_id",
  condition_col = "condition",
  contrasts_list = c("conditionEngerix.B_24_PBMC - conditionEngerix.B_0_PBMC",
                     "conditionEngerix.B_72_PBMC - conditionEngerix.B_0_PBMC",
                     "conditionEngerix.B_168_PBMC - conditionEngerix.B_0_PBMC"),
  condition_map = list(
    "conditionEngerix.B_24_PBMC - conditionEngerix.B_0_PBMC" ~ "Engerix_24_PBMC",
    "conditionEngerix.B_72_PBMC - conditionEngerix.B_0_PBMC" ~ "Engerix_72_PBMC",
    "conditionEngerix.B_168_PBMC - conditionEngerix.B_0_PBMC" ~ "Engerix_168_PBMC"),
  output_prefix = "hepatitisb_engerixb_mouse")

engerixb_mouse_limma %>%
  saveRDS(here("tables", "hepatitisb_engerixb_mouse_dge_limma_degs.rds"))
```

### t-Test

```{r}
mouse_engerixb_log2_hgnc_symbol <- readRDS(here("tables", "hepatitisb_engerixb_mouse_exprs_hgnc.rds"))
mouse_engerixb_metadata <- readRDS(here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))

engerixb_mouse_ttest <- run_ttest_longitudinal(
  exprs_hgnc = mouse_engerixb_log2_hgnc_symbol,
  metadata = mouse_engerixb_metadata,
  timepoint_names = c("1" = "day1", "3" = "day3", "7" = "day7"))

engerixb_mouse_ttest %>%
  saveRDS(here("tables", "hepatitisb_engerixb_mouse_log2fc_paired_unpaired_ttest.rds"))
```


## Human

### Prepare data

```{r}
GSE124533 <- getGEO(filename = here("data_studies", "GSE124533_series_matrix.txt"),
                    GSEMatrix = FALSE)

GSE124533_metadata <- GSE124533@phenoData %>%
  pData() %>%
  dplyr::select(
    geo_sample = "geo_accession",
    participant_id = "participant:ch1",
    sex = "gender:ch1",
    treatment = "treatment:ch1",
    ethnicity = "ethnicity:ch1",
    race = "race:ch1",
    age = "age:ch1",
    timepoint = "day:ch1") %>%
  mutate(condition = treatment, organism = "Human", time_unit = "Day")

GSE124533_metadata %>%
  saveRDS(here("tables", "hepatitisb_engerixb_human_metadata.rds"))
```

### ExpressionSet

```{r}
human_engerixb_exprs <- readRDS(here("tables", "hepatitisb_engerixb_human_exprs.rds"))
human_engerixb_metadata <- readRDS(here("tables", "hepatitisb_engerixb_human_metadata.rds"))

eset_human_engerixb <- create_eset(
  exprs_hgnc = human_engerixb_exprs,
  metadata = human_engerixb_metadata,
  symbol_col = "gene_symbol",
  id_col = "geo_sample",
  normalize = TRUE)

eset_human_engerixb %>%
  saveRDS(here("tables", "hepatitisb_engerixb_human_eset.rds"))
```

### Limma + Subsampling

```{r}
eset_human_engerixb <- readRDS(here("tables", "hepatitisb_engerixb_human_eset.rds"))
human_engerixb_metadata <- readRDS(here("tables", "hepatitisb_engerixb_human_metadata.rds"))

human_engerixb_subsampling <- subsampling_longitudinal(
  eset = eset_human_engerixb,
  metadata = human_engerixb_metadata,
  n_iterations = 200,
  cohort_n = 5,
  participant_col = "participant_id",
  contrasts_list = c("timepoint1 - timepoint0",
                     "timepoint3 - timepoint0",
                     "timepoint7 - timepoint0"),
  condition_map = list(
    "Time1  - Time0" ~ "EngerixB_Day1",
    "Time3  - Time0" ~ "EngerixB_Day3",
    "Time7  - Time0" ~ "EngerixB_Day7"),
  output_prefix = "hepatitisb_engerixb_human")
```

```{r fig.height=5, fig.width=6}
plot_demographics_subsampling(
  df_global_runs = human_engerixb_subsampling$runs,
  original_metadata = human_engerixb_metadata)
```


# 3. S. aureus

## Mouse

### Download dataset

```{r}
GSE19668 <- getGEO("GSE19668", GSEMatrix = TRUE)
eset <- GSE19668[[1]]

GSE19668_metadata <- eset %>%
  pData() %>%
  dplyr::select(
    geo_sample = "geo_accession",
    subject_id = "title",
    treatment = "infection:ch1",
    timepoint = "time point:ch1",
    strain = "strain:ch1") %>%
  mutate(treatment = fct_relevel(treatment, "none"),
         timepoint = fct_relevel(timepoint, "0"),
         condition = paste(treatment, strain, timepoint, sep = "_"))

GSE19668_metadata %>%
  saveRDS(here("tables", "saureus_infection_GSE19668_metadata.rds"))

GSE19668_exprs <- eset %>% exprs()
GSE19668_exprs %>%
  as.data.frame() %>%
  saveRDS(here("tables", "saureus_infection_GSE19668_exprs.rds"))
```

### Clean metadata

```{r}
GSE19668_metadata <- readRDS(here("tables", "saureus_infection_GSE19668_metadata.rds"))

saureus_mouse_metadata <- GSE19668_metadata %>%
  filter(strain == "C57BL/6J") %>%
  separate(subject_id, into = c("tissue", "strain1", "timepoint1", "participant_id"),
           sep = " ") %>%
  select(!c(tissue:timepoint1)) %>%
  arrange(participant_id, timepoint) %>%
  separate(timepoint, into = c("timepoint_number", "time_unit")) %>%
  mutate(time_unit = "hours", timepoint = timepoint_number %>% as.numeric(),
         organism = "Mouse") %>%
  arrange(timepoint) %>%
  select(-timepoint_number) %>%
  select(geo_sample, participant_id, condition, timepoint, time_unit, everything())

saureus_mouse_metadata %>%
  saveRDS(here("tables", "saureus_infection_mouse_metadata_filtered.rds"))
```

### Annotate dataset

```{r}
saureus_mouse_exprs <- readRDS(here("tables", "saureus_infection_GSE19668_exprs.rds"))
saureus_mouse_metadata <- readRDS(here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

saureus_mouse_annot <- annotate_mouse_probes(
  exprs_matrix = saureus_mouse_exprs %>% select(saureus_mouse_metadata$geo_sample),
  probe_attribute = "affy_mouse430_2",
  output_prefix = "saureus_infection_mouse")
```

### log2FC (independent)

```{r}
saureus_mouse_exprs <- readRDS(here("tables", "saureus_infection_GSE19668_exprs.rds"))
saureus_mouse_metadata <- readRDS(here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

mouse_exprs_metadata <- saureus_mouse_exprs %>%
  select(saureus_mouse_metadata$geo_sample) %>%
  rownames_to_column("probe_id") %>%
  pivot_longer(cols = saureus_mouse_metadata$geo_sample,
               names_to = "geo_sample", values_to = "value_raw") %>%
  inner_join(saureus_mouse_metadata, by = "geo_sample")

saureus_mouse_log2fc <- compute_log2fc_independent(
  exprs_long_format = mouse_exprs_metadata,
  control_condition = "none",
  value_col = "value_raw",
  id_cols = c("probe_id"),
  group_col = "treatment")

saureus_mouse_log2fc %>%
  saveRDS(here("tables", "saureus_infection_mouse_log2fc_sample_clean_long.rds"))
```

### ExpressionSet & Limma

```{r}
mouse_log2_hgnc_symbol <- readRDS(here("tables", "saureus_infection_mouse_exprs_hgnc.rds"))
saureus_mouse_metadata <- readRDS(here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

eset_mouse_saureus <- create_eset(
  exprs_hgnc = mouse_log2_hgnc_symbol,
  metadata = saureus_mouse_metadata)

eset_mouse_saureus %>%
  saveRDS(here("tables", "saureus_infection_mouse_eset.rds"))

eset_mouse_saureus <- readRDS(here("tables", "saureus_infection_mouse_eset.rds"))

saureus_mouse_limma <- run_limma_longitudinal(
  eset = eset_mouse_saureus,
  participant_col = "participant_id",
  condition_col = "condition",
  contrasts_list = c("conditionStaphylococcus.aureus_C57BL.6J_2.hours - conditionnone_C57BL.6J_0",
                     "conditionStaphylococcus.aureus_C57BL.6J_4.hours - conditionnone_C57BL.6J_0",
                     "conditionStaphylococcus.aureus_C57BL.6J_6.hours - conditionnone_C57BL.6J_0",
                     "conditionStaphylococcus.aureus_C57BL.6J_12.hours - conditionnone_C57BL.6J_0"),
  condition_map = list(
    "conditionStaphylococcus.aureus_C57BL.6J_2.hours - conditionnone_C57BL.6J_0" ~ "Saureus_2h",
    "conditionStaphylococcus.aureus_C57BL.6J_4.hours - conditionnone_C57BL.6J_0" ~ "Saureus_4h",
    "conditionStaphylococcus.aureus_C57BL.6J_6.hours - conditionnone_C57BL.6J_0" ~ "Saureus_6h",
    "conditionStaphylococcus.aureus_C57BL.6J_12.hours - conditionnone_C57BL.6J_0" ~ "Saureus_12h"),
  output_prefix = "saureus_infection_mouse")

saureus_mouse_limma %>%
  saveRDS(here("tables", "saureus_infection_mouse_dge_limma_degs.rds"))
```

### t-Test

```{r}
mouse_exprs <- readRDS(here("tables", "saureus_infection_mouse_exprs_hgnc.rds"))
saureus_mouse_metadata <- readRDS(here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

saureus_mouse_ttest <- run_ttest_longitudinal(
  exprs_hgnc = mouse_exprs,
  metadata = saureus_mouse_metadata,
  timepoint_names = c("2" = "hour2", "4" = "hour4", "6" = "hour6", "12" = "hour12"),
  prefix = "hour_")
```

### Compare DGE methods

```{r fig.height=5, fig.width=5}
mouse_log2fc_paired_unpaired_ttest <-
  readRDS(here("tables", "saureus_infection_mouse_log2fc_paired_unpaired_ttest.rds")) %>%
  mutate(organism = "Mouse")

mouse_degs <- readRDS(here("tables", "saureus_infection_mouse_dge_limma_degs.rds")) %>%
  clean_names() %>%
  rename(padj = adj_p_val) %>%
  mutate(organism = "Mouse", test = "Limma", paired = "Paired")

degs_ttest_limma_summary <- bind_rows(mouse_log2fc_paired_unpaired_ttest, mouse_degs) %>%
  group_by(organism, timepoint, test, paired) %>%
  summarise(n_degs = sum(padj <= 0.05, na.rm = TRUE), .groups = "drop") %>%
  mutate(timepoint_num = readr::parse_number(timepoint),
         timepoint = paste0(timepoint, "h"),
         timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)])))

degs_ttest_limma_summary

degs_ttest_limma_summary %>%
  ggplot(aes(x = timepoint, y = n_degs, fill = test, alpha = paired)) +
  geom_col(width = 0.8, position = position_dodge2()) +
  scale_fill_manual(values = c("t-Test" = "gray25", "Limma" = "#4DBBD5FF")) +
  scale_alpha_manual(values = c("Paired" = 1, "Unpaired" = 0.2)) +
  scale_x_discrete(expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_vaxgo() +
  labs(title = "Total DEGs by test and timepoint, mouse",
       subtitle = "Comparison across tests (FDR ≤ 0.05)",
       x = "Day", y = "Number of DEGs", fill = "Test", alpha = "Pairing") +
  facet_wrap(~organism, scales = "free_y") +
  theme(legend.position = "top", axis.line.y = element_blank(),
        panel.grid.major.y = element_line(color = "gray75", linetype = 3),
        axis.ticks.y = element_line(color = "gray75")) %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_dge_ttest_limma_summary.png"),
         width = 5, height = 5)
```


## Human

### Prepare data

```{r}
GSE33341 <- getGEO("GSE33341", GSEMatrix = TRUE)
eset_human <- GSE33341[[2]]

GSE33341_metadata <- eset_human %>%
  pData() %>%
  dplyr::select(
    geo_sample = "geo_accession",
    participant_id = "title",
    sex = "host gender:ch1",
    condition = "source_name_ch1",
    pathogen = "pathogen:ch1",
    ethnicity = "ethnicity:ch1",
    tissue_infection = "anatomic site of infection:ch1",
    batch_experimental = "experimental batch:ch1",
    age = "host age (yr):ch1") %>%
  mutate(condition = pathogen, organism = "Human")

GSE33341_metadata %>%
  saveRDS(here("tables", "saureus_infection_human_metadata.rds"))

GSE33341_exprs <- eset_human %>% exprs()
GSE33341_exprs %>%
  as.data.frame() %>%
  saveRDS(here("tables", "saureus_infection_human_exprs.rds"))
```

### ExpressionSet

```{r}
saureus_human_exprs <- readRDS(here("tables", "saureus_infection_human_exprs.rds"))
saureus_human_metadata <- readRDS(here("tables", "saureus_infection_human_metadata.rds"))

eset_human_saureus <- create_eset(
  exprs_hgnc = saureus_human_exprs,
  metadata = saureus_human_metadata,
  symbol_col = "V1",
  id_col = "geo_sample")

eset_human_saureus %>%
  saveRDS(here("tables", "saureus_infection_human_eset.rds"))
```

### Limma + Subsampling (independent)

```{r}
eset_human_saureus <- readRDS(here("tables", "saureus_infection_human_eset.rds"))

saureus_human_subsampling <- subsampling_independent(
  eset = eset_human_saureus,
  metadata = saureus_human_metadata,
  n_iterations = 200,
  cohort_n = 10,
  balance_col = "pathogen",
  design_formula = ~ pathogen + batch_experimental,
  contrasts_list = "pathogenStaphylococcus.aureus",
  output_prefix = "saureus_infection_human",
  var_group_col = "batch_experimental")

saureus_human_subsampling$degs %>%
  mutate(condition = "Saureus_Day1", timepoint = "24") %>%
  saveRDS(here("tables", "saureus_infection_human_dge_subsampled_degs.rds"))
```

### Demographics

```{r fig.height=5, fig.width=6}
plot_demographics_subsampling(
  df_global_runs = saureus_human_subsampling$runs,
  original_metadata = saureus_human_metadata)
```


# 4. E. coli infection

```{r}
GSE33341 <- getGEO("GSE33341", GSEMatrix = TRUE)
eset_mouse <- GSE33341[[1]]
eset_human <- GSE33341[[2]]
```

### Metadata

```{r}
ecoli_mouse_metadata <- eset_mouse %>%
  pData() %>%
  dplyr::select(
    geo_sample = "geo_accession",
    subject_title = "title",
    subject_num = "replicate number:ch1",
    batch = "experimental batch:ch1",
    host_strain = "host strain:ch1",
    organism = "organism_ch1",
    age = "host age (yr):ch1",
    sex = "host gender:ch1",
    pathogen = "pathogen:ch1",
    status = "infection status:ch1",
    site_infection = "anatomic site of infection:ch1",
    route_infection = "inoculum route:ch1",
    dosage = "pathogen   dosage cfu/g:ch1",
    timepoint = "time after infection (h):ch1",
    bac_strain = "bacterial strain:ch1") %>%
  filter(pathogen %in% c("Escherichia  coli", "-")) %>%
  mutate(pathogen = if_else(pathogen == "Escherichia  coli", "escherichia coli", "none"),
         subject_id = paste0("Mouse_", host_strain, "_Animal", subject_num, "_Batch", batch),
         status = str_replace(status, "-", "Healthy"),
         condition = paste("Mouse", host_strain, timepoint, sep = "_"),
         comparison = paste0("Mouse_", host_strain, "_Batch", batch)) %>%
  select(geo_sample, comparison, condition, subject_id, everything())

ecoli_human_metadata <- eset_human %>%
  pData() %>%
  filter(`pathogen:ch1` %in% c("Escherichia coli", "-")) %>%
  dplyr::select(
    geo_sample = "geo_accession",
    subject_title = "title",
    batch = "experimental batch:ch1",
    organism = "organism_ch1",
    age = "host age (yr):ch1",
    sex = "host gender:ch1",
    ethnicity = "ethnicity:ch1",
    pathogen = "pathogen:ch1",
    site_infection = "anatomic site of infection:ch1") %>%
  mutate(site_infection = str_to_lower(site_infection),
         organism = "Human",
         pathogen = if_else(pathogen == "-", "Healthy", pathogen),
         status = if_else(pathogen == "Healthy", pathogen, "infected"),
         timepoint = "0",
         subject_id = if_else(pathogen == "Escherichia coli",
                              paste0("Ecoli_human_", site_infection, "_", geo_sample),
                              paste0("Healthy_human_", geo_sample)),
         site_infection = if_else(site_infection == "-", "none", site_infection),
         condition = subject_id) %>%
  select(geo_sample, condition, subject_id, everything()) %>%
  mutate(timepoint = if_else(pathogen == "Escherichia coli", "24", timepoint))

ecoli_mouse_human_metadata <- bind_rows(ecoli_human_metadata, ecoli_mouse_metadata)
ecoli_mouse_human_metadata %>%
  saveRDS(here("tables", "ecoli_infection_human_mouse_metadata.rds"))
```

### Expression data

```{r}
ecoli_mouse_human_metadata <- readRDS(here("tables", "ecoli_infection_human_mouse_metadata.rds"))

mouse_exprs <- exprs(eset_mouse) %>%
  as.data.frame() %>%
  select(ecoli_mouse_human_metadata %>%
           filter(organism == "Mus musculus") %>% pull(geo_sample)) %>%
  rownames_to_column("probe_id")

mouse_exprs %>% saveRDS(here("tables", "ecoli_infection_mouse_exprs.rds"))

human_exprs <- exprs(eset_human) %>%
  as.data.frame() %>%
  select(ecoli_mouse_human_metadata %>%
           filter(organism == "Human") %>% pull(geo_sample)) %>%
  rownames_to_column("probe_id")

human_exprs %>% saveRDS(here("tables", "ecoli_infection_human_exprs.rds"))
```

## Mouse

### Annotate genes

```{r}
ecoli_mouse_exprs <- readRDS(here("tables", "ecoli_infection_mouse_exprs.rds"))
ecoli_mouse_human_metadata <- readRDS(here("tables", "ecoli_infection_human_mouse_metadata.rds"))

ecoli_mouse_metadata <- ecoli_mouse_human_metadata %>%
  filter(host_strain == "C57BL/6J", batch == "8")

ecoli_mouse_metadata %>% saveRDS(here("tables", "ecoli_infection_mouse_metadata.rds"))

ecoli_mouse_annot <- annotate_mouse_probes(
  exprs_matrix = ecoli_mouse_exprs,
  probe_attribute = "affy_mouse430_2",
  output_prefix = "ecoli_infection_mouse")
```

### log2FC (independent)

```{r}
ecoli_mouse_exprs <- readRDS(here("tables", "ecoli_infection_mouse_exprs.rds"))
ecoli_mouse_human_metadata <- readRDS(here("tables", "ecoli_infection_human_mouse_metadata.rds"))

mouse_expression_metadata <- ecoli_mouse_exprs %>%
  pivot_longer(cols = 2:125, names_to = "geo_sample", values_to = "value_raw") %>%
  inner_join(ecoli_mouse_human_metadata, by = "geo_sample")

ecoli_mouse_log2fc <- compute_log2fc_independent(
  exprs_long_format = mouse_expression_metadata,
  control_condition = "Healthy",
  value_col = "value_raw",
  id_cols = c("probe_id"),
  group_col = "status")

ecoli_mouse_log2fc %>%
  saveRDS(here("tables", "ecoli_infection_mouse_log2fc_sample_clean_long.rds"))
```

### ExpressionSet & Limma

```{r}
ecoli_mouse_log2_hgnc_symbol <- readRDS(here("tables", "ecoli_infection_mouse_exprs_hgnc.rds"))
ecoli_mouse_metadata <- readRDS(here("tables", "ecoli_infection_mouse_metadata.rds"))

eset_mouse_ecoli <- create_eset(
  exprs_hgnc = ecoli_mouse_log2_hgnc_symbol,
  metadata = ecoli_mouse_metadata)

eset_mouse_ecoli %>% saveRDS(here("tables", "ecoli_infection_mouse_eset.rds"))

eset_mouse_ecoli <- readRDS(here("tables", "ecoli_infection_mouse_eset.rds"))

ecoli_mouse_limma <- run_limma_longitudinal(
  eset = eset_mouse_ecoli,
  participant_col = "subject_id",
  condition_col = "condition",
  contrasts_list = c("conditionMouse_C57BL.6J_2 - conditionMouse_C57BL.6J_0",
                     "conditionMouse_C57BL.6J_6 - conditionMouse_C57BL.6J_0",
                     "conditionMouse_C57BL.6J_12 - conditionMouse_C57BL.6J_0",
                     "conditionMouse_C57BL.6J_24 - conditionMouse_C57BL.6J_0"),
  condition_map = list(
    "conditionMouse_C57BL.6J_2 - conditionMouse_C57BL.6J_0" ~ "Ecoli_2h",
    "conditionMouse_C57BL.6J_6 - conditionMouse_C57BL.6J_0" ~ "Ecoli_6h",
    "conditionMouse_C57BL.6J_12 - conditionMouse_C57BL.6J_0" ~ "Ecoli_12h",
    "conditionMouse_C57BL.6J_24 - conditionMouse_C57BL.6J_0" ~ "Ecoli_24h"),
  output_prefix = "ecoli_infection_mouse")

ecoli_mouse_limma %>%
  saveRDS(here("tables", "ecoli_infection_mouse_dge_limma_degs.rds"))
```

### t-Test

```{r}
mouse_exprs <- readRDS(here("tables", "ecoli_infection_mouse_exprs_hgnc.rds"))
mouse_metadata <- readRDS(here("tables", "ecoli_infection_mouse_metadata.rds"))

ecoli_mouse_ttest <- run_ttest_longitudinal(
  exprs_hgnc = mouse_exprs,
  metadata = mouse_metadata,
  timepoint_names = c("2" = "hour2", "6" = "hour6", "12" = "hour12", "24" = "hour24"),
  prefix = "hour_")
```

## Human

### Annotate & log2FC

```{r}
human_exprs <- readRDS(here("tables", "ecoli_infection_human_exprs.rds"))
ecoli_mouse_human_metadata <- readRDS(here("tables", "ecoli_infection_human_mouse_metadata.rds"))

human_expression_metadata <- human_exprs %>%
  pivot_longer(cols = 2:63, names_to = "geo_sample", values_to = "value_raw") %>%
  inner_join(ecoli_mouse_human_metadata, by = "geo_sample")

human_expression_metadata %>%
  saveRDS(here("tables", "ecoli_infection_human_exprs_metadata.rds"))

human_metadata <- human_expression_metadata %>%
  filter(organism == "Human") %>%
  select(-value_raw, -probe_id) %>% distinct()

human_metadata %>% saveRDS(here("tables", "ecoli_infection_human_metadata.rds"))

human_expression_metadata <- readRDS(here("tables", "ecoli_infection_human_exprs_metadata.rds"))

human_log2fc <- compute_log2fc_independent(
  exprs_long_format = human_expression_metadata,
  control_condition = "Healthy",
  value_col = "value_raw",
  id_cols = c("probe_id"),
  group_col = "status")

human_log2fc %>%
  saveRDS(here("tables", "ecoli_infection_human_log2fc_sample_clean_long.rds"))
```

### ExpressionSet & Limma

```{r}
human_exprs <- readRDS(here("tables", "ecoli_infection_human_exprs.rds"))
human_metadata <- readRDS(here("tables", "ecoli_infection_human_metadata.rds"))

human_ecoli_annot <- annotate_human_probes(
  exprs_matrix = human_exprs,
  platform_file = "GPL21272-58809.txt",
  sample_ids = human_metadata$geo_sample)

human_log2_hgnc_symbol <- human_ecoli_annot$exprs
human_symbol_entrez <- human_ecoli_annot$symbol_entrez

eset_human_ecoli <- create_eset(
  exprs_hgnc = human_log2_hgnc_symbol,
  metadata = human_metadata,
  symbol_col = "gene_symbol",
  id_col = "geo_sample",
  normalize = TRUE)

eset_human_ecoli %>% saveRDS(here("tables", "ecoli_infection_human_eset.rds"))

eset_human_ecoli <- readRDS(here("tables", "ecoli_infection_human_eset.rds"))

ecoli_human_subsampling <- subsampling_independent(
  eset = eset_human_ecoli,
  metadata = human_metadata,
  n_iterations = 200,
  cohort_n = 10,
  balance_col = "pathogen",
  design_formula = ~ pathogen + batch,
  contrasts_list = "pathogenEscherichia.coli",
  output_prefix = "ecoli_infection_human",
  var_group_col = "batch")

ecoli_human_subsampling$degs %>%
  mutate(condition = "Ecoli_Day1", timepoint = "24", organism = "Human") %>%
  saveRDS(here("tables", "ecoli_infection_human_dge_subsampled_degs.rds"))
```


# 6. Duchenne

## Mouse

### Download dataset

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
GSE1025 <- getGEO("GSE1025", GSEMatrix = TRUE)
eset <- GSE1025[[1]]

duchenne_metadata <- eset %>%
  pData() %>%
  dplyr::select(
    geo_sample = "geo_accession",
    subject_id = "title",
    tissue = "source_name_ch1",
    platform_id,
    participant_id = subject_id) %>%
  mutate(
    tissue = "Muscle",
    temp_col = str_to_lower(subject_id),
    treatment = case_when(
      str_detect(subject_id, "control|Control") ~ "Control",
      str_detect(subject_id, "mdx") ~ "DMD",
      TRUE ~ NA)) %>%
  extract(col = temp_col,
          into = c("timepoint", "group", "replicate"),
          regex = "leg (\\d+) day ([a-z]+) rep (\\d+)",
          convert = TRUE) %>%
  mutate(treatment = fct_relevel(treatment, "Control"),
         condition = paste(treatment, timepoint, sep = "_")) %>%
  select(!group)

duchenne_metadata %>% saveRDS(here("tables", "duchenne_mouse_metadata.rds"))

eset <- exprs(eset)
eset %>% as.data.frame() %>%
  saveRDS(here("tables", "duchenne_mouse_exprs.rds"))
```

### Import & log2

```{r}
duchenne_metadata <- readRDS(here("tables", "duchenne_mouse_metadata.rds"))
duchenne_exprs <- read_rds(here("tables", "duchenne_mouse_exprs.rds")) %>%
  as.matrix() %>%
  { log2(. + 1) } %>%
  as.data.frame()

duchenne_exprs <- duchenne_exprs %>%
  select(duchenne_metadata$geo_sample) %>%
  as.matrix()

mouse_exprs_metadata <- duchenne_exprs %>%
  as.data.frame() %>%
  rownames_to_column("probe_id") %>%
  pivot_longer(cols = 2:37, names_to = "geo_sample", values_to = "value_raw") %>%
  inner_join(duchenne_metadata, by = "geo_sample")

duchenne_log2fc <- compute_log2fc_independent(
  exprs_long_format = mouse_exprs_metadata,
  control_condition = "Control",
  value_col = "value_raw",
  id_cols = c("probe_id", "timepoint"),
  group_col = "treatment")

duchenne_log2fc %>%
  saveRDS(here("tables", "duchenne_mouse_log2fc_sample_clean_long.rds"))
```

### Annotate genes

```{r}
duchenne_exprs <- read_rds(here("tables", "duchenne_mouse_exprs.rds")) %>%
  as.matrix() %>%
  { log2(. + 1) } %>%
  as.data.frame() %>%
  select(duchenne_metadata$geo_sample)

duchenne_annot <- annotate_mouse_probes(
  exprs_matrix = duchenne_exprs,
  probe_attribute = "affy_mg_u74av2",
  output_prefix = "duchenne_mouse")
```

### Join annotations with log2FC

```{r}
mouse_log2_hgnc_symbol <- readRDS(here("tables", "duchenne_mouse_exprs_hgnc.rds"))

duchenne_log2fc <- duchenne_log2fc %>%
  inner_join(mouse_log2_hgnc_symbol %>% select(human_symbol, probe_id),
             by = join_by(probe_id))

duchenne_log2fc %>%
  saveRDS(here("tables", "duchenne_mouse_log2fc_sample_clean_long.rds"))

mouse_log2_hgnc_symbol %>%
  saveRDS(here("tables", "duchenne_mouse_exprs_hgnc_symbol.rds"))
```

### ExpressionSet & Limma

```{r}
mouse_duchenne_log2_hgnc_symbol <- readRDS(here("tables", "duchenne_mouse_exprs_hgnc.rds"))
mouse_duchenne_metadata <- readRDS(here("tables", "duchenne_mouse_metadata.rds"))

eset_duchenne <- create_eset(
  exprs_hgnc = mouse_duchenne_log2_hgnc_symbol,
  metadata = mouse_duchenne_metadata %>%
    mutate(timepoint = as.character(timepoint) %>% as.factor()),
  normalize = TRUE)

eset_duchenne %>% saveRDS(here("tables", "duchenne_mouse_eset.rds"))

eset_duchenne <- readRDS(here("tables", "duchenne_mouse_eset.rds"))

genes_to_keep <- exprs(eset_duchenne) %>%
  as.data.frame() %>%
  mutate(across(everything(), ~ . > 6)) %>%
  rowSums() %>%
  { . >= 5 }

eset_filtered <- eset_duchenne[genes_to_keep, ]

design <- model.matrix(~ condition + 0, eset_filtered)
colnames(design) <- make.names(colnames(design))

aw <- arrayWeights(eset_filtered, design)
barplot(aw, names.arg = eset_filtered$condition,
        main = "Sample Weights Diagnostic", ylab = "Weight",
        las = 2, cex.names = 0.7, col = "skyblue")

fit <- lmFit(eset_filtered, design, weights = aw)

cont.matrix <- makeContrasts(
  conditionDMD_7 - conditionControl_7,
  conditionDMD_14 - conditionControl_14,
  conditionDMD_23 - conditionControl_23,
  conditionDMD_28 - conditionControl_28,
  conditionDMD_56 - conditionControl_56,
  conditionDMD_112 - conditionControl_112,
  levels = design)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2, trend = TRUE, robust = TRUE)
plotSA(fit2, main = "Sigma vs Abundance (Weights Applied)")

results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>%
    rownames_to_column("symbol")
})

all_degs_mouse <- bind_rows(results_list) %>%
  select(comparison, everything()) %>%
  mutate(condition = case_when(
    comparison == "conditionDMD_7 - conditionControl_7" ~ "DMD_Day7",
    comparison == "conditionDMD_14 - conditionControl_14" ~ "DMD_Day14",
    comparison == "conditionDMD_23 - conditionControl_23" ~ "DMD_Day23",
    comparison == "conditionDMD_28 - conditionControl_28" ~ "DMD_Day28",
    comparison == "conditionDMD_56 - conditionControl_56" ~ "DMD_Day56",
    comparison == "conditionDMD_112 - conditionControl_112" ~ "DMD_Day112",
    TRUE ~ comparison),
    timepoint = str_remove(condition, "DMD_"),
    organism = "Mouse", treatment = "DMD", pathogen = "DMD",
    se = abs(logFC / t), sd = se / sqrt(3)) %>%
  select(condition, everything()) %>%
  clean_names() %>%
  select(condition, human_symbol = symbol,
         mean_l2fc = log_fc, ci_lower = ci_l, ci_upper = ci_r, ave_expr, everything())

all_degs_mouse %>%
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>%
  summarise(n_degs = n())

all_degs_mouse %>%
  saveRDS(here("tables", "duchenne_mouse_dge_limma_degs.rds"))
```


# 7. FIT

```{r}
library(FIT.mouse2man)
pca_rotations <- FIT.mouse2man::pca_rotations
best_models <- FIT.mouse2man::best_models
HS_MM_Symbol_Entrez <- FIT.mouse2man::HS_MM_Symbol_Entrez
MM_Entrez_symbol_desc <- FIT.mouse2man::MM_Entrez_symbol_desc
Psoriasis_classifier_neg <- FIT.mouse2man::Psoriasis_classifier_neg
slopes_per_gene_V2.0 <- FIT.mouse2man::slopes_per_gene_V2.0
StaphInfection_classifier_pos <- FIT.mouse2man::StaphInfection_classifier_pos
AllData_V2.0 <- FIT.mouse2man::AllData_V2.0
```

### Fluad

```{r}
all_degs_mouse <- readRDS(here("tables", "influenza_fluad_mouse_dge_limma_degs.rds"))
mouse_fluad_log2_hgnc_symbol <- readRDS(here("tables", "influenza_fluad_mouse_exprs_hgnc.rds"))

fluad_fit <- run_fit_prediction(
  degs_mouse = all_degs_mouse,
  exprs_hgnc = mouse_fluad_log2_hgnc_symbol,
  output_prefix = "influenza_fluad_mouse")
```

### Engerix B

```{r}
mouse_engerixb_log2_hgnc_symbol <- readRDS(here("tables", "hepatitisb_engerixb_mouse_log2fc_sample_clean_long.rds"))
annot <- readRDS(here("tables", "hepatitisb_engerixb_mouse_probe_annot.rds"))

input2FIT <- mouse_engerixb_log2_hgnc_symbol %>%
  ungroup() %>%
  inner_join(annot %>% select(probe_id = agilent_sureprint_g3_ge_8x60k,
                              MM.Entrez = entrezgene_id), by = "probe_id") %>%
  select(participant_id, timepoint, MM.Entrez, EffectSize = log2fc) %>%
  group_by(timepoint, MM.Entrez) %>%
  summarise(mean_log2fc = mean(EffectSize, na.rm = TRUE), .groups = "drop") %>%
  drop_na() %>%
  mutate(row = MM.Entrez, pathogen = "Hepatitis B",
         condition = paste0(pathogen, "_D", timepoint)) %>%
  select(row, condition, EffectSize = mean_log2fc, MM.Entrez) %>%
  ungroup()

all_conditions <- unique(input2FIT$condition)
walk(all_conditions, function(cond) {
  delete_dupes <- input2FIT %>%
    filter(condition == cond) %>% get_dupes(row) %>% pull(row)
  input2FIT %>%
    filter(condition == cond, !row %in% delete_dupes) %>%
    select(-condition) %>%
    column_to_rownames("row") %>%
    select(MM.Entrez, EffectSize) %>%
    write_csv(here("tables", paste0(cond, "_FIT_mouse.csv")))
})

fit_results_combined <- map2_dfr(all_conditions, all_conditions, function(cond, cond_name) {
  FIT.mouse2man::FIT(
    MouseFile = here("tables", paste0(cond, "_FIT_mouse.csv")),
    DataType = "rnaseq") %>%
    as_tibble() %>%
    mutate(condition = cond_name)
})

beep()
fit_results_combined %>%
  saveRDS(here("tables", "hepatitisb_engerixb_mouse_fit_prediction_alltimes.rds"))
```

### S. aureus

```{r}
mouse_saureus_log2_hgnc_symbol <- readRDS(here("tables", "saureus_infection_human_mouse_samples_log2fc.rds"))
annot <- readRDS(here("tables", "saureus_infection_mouse_probe_annot.rds"))

input2FIT <- mouse_saureus_log2_hgnc_symbol %>%
  ungroup() %>%
  filter(timepoint == 12) %>%
  inner_join(annot %>% select(probe_id = affy_mouse430_2, MM.Entrez = entrezgene_id),
             by = "probe_id") %>%
  select(participant_id, MM.Entrez, EffectSize = log2fc) %>%
  group_by(MM.Entrez) %>%
  summarise(mean_log2fc = mean(EffectSize, na.rm = TRUE), .groups = "drop") %>%
  drop_na()

input2FIT %>%
  write_csv(here("tables", "saureus_infection_mouse_fit_prediction_input_12h.csv"))

FIT_day12_results <- FIT.mouse2man::FIT(
  MouseFile = here("tables", "saureus_infection_mouse_fit_prediction_input_12h.csv"),
  DataType = "rnaseq")

FIT_day12_results %>%
  clean_names() %>%
  pivot_longer(cols = c(mouse_effect_size, fit_prediction),
               names_to = "measure", values_to = "value") %>%
  mutate(pathogen = "Staphylococcus aureus", timepoint = "12") %>%
  arrange(mouse_symbol) %>%
  select(pathogen, timepoint, everything()) %>%
  saveRDS(here("tables", "saureus_mouse_fit_prediction_results_12h.rds"))
```

### E. coli

```{r}
mouse_ecoli_log2_hgnc_symbol <- readRDS(here("tables", "ecoli_infection_human_mouse_samples_log2fc.rds"))
annot <- readRDS(here("tables", "ecoli_infection_mouse_probe_annot.rds"))

input2FIT <- mouse_ecoli_log2_hgnc_symbol %>%
  ungroup() %>%
  filter(timepoint == 12) %>%
  inner_join(annot %>% select(probe_id = affy_mouse430_2, MM.Entrez = entrezgene_id),
             by = "probe_id") %>%
  select(participant_id = subject_id, MM.Entrez, EffectSize = log2fc) %>%
  group_by(MM.Entrez) %>%
  summarise(mean_log2fc = mean(EffectSize, na.rm = TRUE), .groups = "drop") %>%
  drop_na()

input2FIT %>%
  write_csv(here("tables", "ecoli_infection_mouse_fit_prediction_input_12h.csv"))

FIT_day12_results <- FIT.mouse2man::FIT(
  MouseFile = here("tables", "ecoli_infection_mouse_fit_prediction_input_12h.csv"),
  DataType = "rnaseq")

FIT_day12_results %>%
  clean_names() %>%
  pivot_longer(cols = c(mouse_effect_size, fit_prediction),
               names_to = "measure", values_to = "value") %>%
  mutate(pathogen = "Escherichia coli", timepoint = "12") %>%
  arrange(mouse_symbol) %>%
  select(pathogen, timepoint, everything()) %>%
  saveRDS(here("tables", "ecoli_mouse_fit_prediction_results_12h.rds"))
```

#===================

# 8. Combine all datasets

## Metadata
Merge all analysis results into a comprehensive dataset.

```{r}
#Merge results =====================
#Metadata
#S. aurues
human_mouse_metadata_saureus = readRDS(file = here("tables", "saureus_infection_human_mouse_metadata.rds")) %>% 
  select(geo_sample, participant_id, organism, timepoint, time_unit, pathogen, treatment, age, sex, ethnicity, strain, tissue, tissue_infection, batch = batch_experimental) %>% 
  mutate(pathogen = "S. aureus",
         tissue = "Blood",
         timepoint = as.numeric(timepoint),
         age = as.character(age)) %>% 
  mutate(timepoint = if_else(timepoint == 1, 24, timepoint),
         time_unit = "Hour") 

human_mouse_metadata_ecoli = readRDS(here("tables", "ecoli_infection_human_mouse_metadata.rds")) %>% 
  filter(host_strain == "C57BL/6J" | organism == "Human",
         !batch %in% c("1", "2")) %>% 
  mutate(timepoint = as.numeric(timepoint), 
         time_unit = "Hour", 
         pathogen = "E. coli", 
         tissue = "Blood", tissue_infection = site_infection, 
         age = str_remove(age, "0.50-0.66"),
         organism = if_else(organism == "Mus musculus", "Mouse", organism)) %>% 
  select(geo_sample, participant_id = subject_id, organism, timepoint, time_unit, pathogen, treatment = status, age, sex, ethnicity, strain = host_strain,
         tissue, tissue_infection, batch) %>% 
  arrange(organism, timepoint)

human_mouse_engerix_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_metadata.rds")) %>% 
  mutate(pathogen = "HepB",
         vaccine = "Engerix B",
       treatment = "Engerix B", 
       type = "VLP",
       timepoint = as.numeric(timepoint),
       age = as.character(age)) %>% 
  select(geo_sample, participant_id, organism, timepoint, time_unit, pathogen, vaccine, type, treatment, age, sex, ethnicity = race, strain, tissue)

fluad_human_mouse_metadata = readRDS(here("tables", "fluad_human_mouse_metadata.rds"))  %>% 
  mutate(time_unit = "Day", timepoint = as.numeric(timepoint), age = as.character(age)) %>% 
  select(geo_sample, participant_id, organism, timepoint, time_unit, pathogen, vaccine, type, treatment, age, sex, ethnicity, strain, tissue)

#Burn and Trauma
mouse_metadata_burn_trauma = read_csv(file = here("tables", "trauma_burn_mouse_metadata.csv")) %>% 
  filter(!str_detect(treatment, "LPS")) %>% 
  separate(timepoint, into = c("timepoint", "time_unit")) %>% 
  separate(treatment, into = c("pathogen", "treatment")) %>% 
  mutate(tissue = "Leukocytes",
         organism = "Mouse",
         treatment = if_else(is.na(treatment), pathogen, treatment),
         timepoint = as.numeric(timepoint)) %>% 
  select(geo_sample, participant_id, organism, timepoint, time_unit, pathogen, treatment, sex, strain, tissue)
  
#Burn
human_burn_metadata = read_csv(file = here("tables", "burn_human_metadata.csv")) %>% 
  rename(timepoint_original = timepoint,
         timepoint = timepoints_new) %>% 
  mutate(pathogen = "Burn", 
         organism = "Human",
         subject_id = if_else(subject_id == "Control", paste0(subject_id, "_", geo_sample), subject_id),
         timepoint = str_replace(timepoint, "_", " "),
         time_unit = if_else(str_detect(timepoint, "Day"), "Day", "Hour"),
         timepoint = str_remove(timepoint, "Day "),
         timepoint = if_else(treatment == "Control", "0", timepoint)) %>% 
  select(geo_sample, participant_id = subject_id, organism, timepoint_original, timepoint, time_unit, pathogen, treatment, age, sex, tissue)

outliers = readRDS(file = here("tables", "trauma_human_metadata_outliers.rds"))

#Trauma
human_trauma_metadata = read_csv(file = here("tables", "trauma_human_metadata.csv")) %>% 
  rename(timepoint_original = timepoint,
         timepoint = timepoints_new) %>% 
  mutate(pathogen = "Trauma", 
         organism = "Human",
         subject_id = if_else(subject_id == "Control", paste0(subject_id, "_", geo_sample), subject_id),
         timepoint = str_replace(timepoint, "_", " "),
         time_unit = if_else(str_detect(timepoint, "Day"), "Day", "Hour"),
         timepoint = str_remove(timepoint, "Day "),
         timepoint = if_else(treatment == "Control", "0", timepoint)) %>% 
  select(geo_sample, participant_id = subject_id, organism, timepoint_original, timepoint, time_unit, pathogen, treatment, age, sex, tissue) %>% 
  #Remove outliers
  filter(!participant_id %in% c(outliers))
  

#Unite datasets
all_human_mouse_metadata = bind_rows(
  human_mouse_metadata_saureus %>% 
    mutate(timepoint = as.character(timepoint),
           age = as.numeric(age)),
  human_mouse_metadata_ecoli %>% 
    mutate(timepoint = as.character(timepoint),
           age = as.numeric(age)),
  human_mouse_engerix_metadata %>% 
    mutate(timepoint = as.character(timepoint),
           age = as.numeric(age)),
  fluad_human_mouse_metadata %>% 
    mutate(timepoint = as.character(timepoint),
           age = as.numeric(age)),
  mouse_metadata_burn_trauma %>% 
    mutate(timepoint = as.character(timepoint)),
  human_burn_metadata,
  human_trauma_metadata
) %>% 
  # Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Burn", "Trauma") & timepoint %in% c("2", "2h", "Early Hours") ~ "Early hours",
                               time_unit == "Day" ~ paste0("Day ", timepoint),
                               time_unit == "Hour" & timepoint != "Early hours" ~ paste0(timepoint, "h"),
                               T~timepoint),
         timepoint = fct_relevel(timepoint, "Early hours", "2h", "4h", "6h", "12h", "24h", 
                                           "Day 1", "Day 2", "Day 3", "Day 4", "Day 5", "Day 7", "Day 14", "Day 28"),
         timepoint_comparison = if_else(pathogen %in% c("Escherichia coli", "Staphylococcus aureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early"),
         pathogen = case_when(pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              pathogen == "Hepatitis B" ~ "HepB",
                              T ~ pathogen))
  

all_human_mouse_metadata %>% 
  dplyr::count(pathogen, treatment, organism, timepoint)

all_human_mouse_metadata %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "all_human_mouse_metadata.rds"))
```

## Sample log2FC
Compile sample-level log2 fold-change data for all conditions and organisms.

```{r}

all_human_mouse_metadata =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(here("tables", "all_human_mouse_metadata.rds"))

# Sample log2fc
# Load data
human_mouse_log2fc_samples_ecoli = readRDS(here("tables", "ecoli_infection_human_mouse_samples_log2fc.rds"))  %>% 
  select(human_symbol, geo_sample, log2fc, value_raw, day0_value_raw) %>% 
  mutate(pathogen = "Escherichia coli")

human_mouse_log2fc_samples_saureus = readRDS(here("tables", "saureus_infection_human_mouse_samples_log2fc.rds"))  %>% 
  select(human_symbol, geo_sample, log2fc, value_raw, day0_value_raw) %>% 
  mutate(pathogen = "Staphylococcus aureus")

human_mouse_log2fc_samples_engerixb = readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_exprs_samples_log2fc.rds"))  %>% 
   mutate(timepoint = paste0("Day ", timepoint)) %>% 
# Attach sample metadata (participant ID, timepoint, condition) to expression values by GEO sample ID
  inner_join(all_human_mouse_metadata,
             by = join_by("participant_id", "timepoint", "organism")) %>% 
  select(human_symbol, geo_sample, log2fc, value_raw = value, day0_value_raw = day0) 

human_mouse_log2fc_samples_fluad = readRDS(file = here("tables", "influenza_fluad_human_mouse_log2fc_samples.rds")) %>% 
  mutate(timepoint = paste0("Day ", timepoint)) %>% 
  select(-geo_sample) %>% 
# Attach sample metadata (participant ID, timepoint, condition) to expression values by GEO sample ID
  inner_join(all_human_mouse_metadata,
             by = join_by("participant_id", "timepoint", "organism")) %>% 
  select(human_symbol, geo_sample, log2fc, value_raw = value, day0_value_raw = day0) 

#Burn and Trauma
mouse_log2fc_samples_trauma_burn = readRDS(here("tables", "trauma_burn_mouse_log2fc_sample_clean_long.rds"))  %>% 
  select(human_symbol, geo_sample, log2fc, value_raw, day0_value_raw = value_raw_ref)

human_log2fc_samples_trauma = readRDS(here("tables", "trauma_human_log2fc_sample_clean_long.rds"))  %>% 
  select(human_symbol, geo_sample, log2fc, value_raw, day0_value_raw = value_raw_ref) 

human_log2fc_samples_burn = readRDS(here("tables", "burn_human_log2fc_sample_clean_long.rds"))  %>% 
  select(human_symbol, geo_sample, log2fc, value_raw, day0_value_raw = value_raw_ref)


#Unite
all_human_mouse_samples_log2fc = bind_rows(
  human_mouse_log2fc_samples_saureus,
  human_mouse_log2fc_samples_ecoli,
  human_mouse_log2fc_samples_engerixb,
  human_mouse_log2fc_samples_fluad,
  human_log2fc_samples_trauma,
  human_log2fc_samples_burn,
  mouse_log2fc_samples_trauma_burn
) %>% 
  select(human_symbol, geo_sample, log2fc, value_raw, day0_value_raw) 

#Save
all_human_mouse_samples_log2fc %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "all_human_mouse_samples_log2fc.rds"))


# --- Table 1: Log2FC Matrix ---
# log2fc_matrix = all_human_mouse_samples_log2fc %>%
#   select(human_symbol, geo_sample, log2fc) %>%
#   pivot_wider(
#     names_from = geo_sample, 
#     values_from = log2fc
#   )
# 
# --- Table 2: Raw Values (Current Timepoints) ---
# raw_values_matrix = all_human_mouse_samples_log2fc %>%
#   select(human_symbol, geo_sample, value_raw) %>%
#   pivot_wider(
#     names_from = geo_sample, 
#     values_from = value_raw
#   )
# 
# --- Table 3: Baseline (Day 0 Values) ---
# baseline_day0_matrix = all_human_mouse_samples_log2fc %>%
#   select(human_symbol, geo_sample, day0_value_raw) %>%
#   distinct() %>%
#   pivot_wider(
#     names_from = geo_sample, 
#     values_from = day0_value_raw
#   )
# 
# all_human_mouse_samples_log2fc_list = list(
#   log2fc    = log2fc_matrix,
#   raw_expr  = raw_values_matrix,
#   baseline  = baseline_day0_matrix
# )
# all_human_mouse_samples_log2fc_list %>% 
#   saveRDS(here("tables", "all_human_mouse_samples_log2fc_list.rds"))

#Insert

# log2fc_samples = readRDS(here("tables", "all_human_mouse_samples_log2fc.rds")) 

# Filter human samples with trauma to decrease computation time

# #Duplicate infection timepoints in human to match the mouse
# log2fc_samples_annotated_matched = log2fc_samples_annotated %>% 
#   
  # #Duplicate infection timepoints in human to match the mouse
  # bind_rows(
  # log2fc_samples_annotated %>%
  # filter(organism == "Human",
  #        pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
  # select(pathogen, organism, participant_id, human_symbol, log2fc) %>%
  # #Get timepoints from mouse
  # inner_join(log2fc_samples_annotated %>%
  #              filter(organism == "Mouse",
  #                    pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
  #              select(pathogen, treatment, timepoint) %>%
  #              distinct(),
  #            by = "pathogen")
  # )

# #Check result
# log2fc_samples_annotated_matched %>%
#   dplyr::count(pathogen, treatment, organism, timepoint)


```

## DGE results
Load and combine differential expression results from all conditions.

```{r}
#========================
# Combine DGE results
#========================
# Load data
#Fluad
all_degs_human_mouse_fluad = readRDS(here("tables", "influenza_fluad_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Influenza", 
         vaccine = "Fluad",
         treatment = "Vaccination") 

#Engerix B
all_degs_human_mouse_engerixb = readRDS(here("tables", "hepatitisb_engerixb_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Hepatitis B", treatment = "Vaccination",
         vaccine = "Engerix B",
         timepoint = as.character(timepoint))

#S.aureus
all_degs_human_mouse_saureus = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Staphylococcus aureus", treatment = "Infection")

#E coli
all_degs_human_mouse_ecoli = readRDS(here("tables", "ecoli_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Escherichia coli", treatment = "Infection")

#Burn
all_degs_human_mouse_burn = readRDS(here("tables", "burn_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Burn", treatment = "Injury")

#Burn
all_degs_human_mouse_trauma = readRDS(here("tables", "trauma_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Trauma", treatment = "Injury")

#Unite
all_degs_human_mouse = bind_rows(all_degs_human_mouse_saureus, 
                                 all_degs_human_mouse_ecoli, 
                                 all_degs_human_mouse_fluad, 
                                 all_degs_human_mouse_engerixb,
                                 all_degs_human_mouse_burn, 
                                 all_degs_human_mouse_trauma
                                 ) %>% 
# Select all columns: ensures clean column order with key identifier first
  select(organism, pathogen, treatment, timepoint, condition, human_symbol, everything()) %>% 
  mutate(timepoint = if_else(pathogen == "Staphylococcus aureus" & timepoint == "1", "24", timepoint)) %>% 
  #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Burn", "Trauma") & timepoint %in% c("2h", "Early hours")~"Early hours", 
                               pathogen %in% c("Influenza", "Hepatitis B") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Escherichia coli", "Staphylococcus aureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint %in% c("2h", "Early hours"), "Early hours", timepoint),
        timepoint = fct_relevel(timepoint, "Early hours", "2h", "4h", "6h", "12h", "24h", 
                                           "Day 1", "Day 2", "Day 3", "Day 4", "Day 5", "Day 7", "Day 14", "Day 28")) %>% 
   mutate(pathogen = case_when(pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              pathogen == "Hepatitis B" ~ "HepB",
                              T ~ pathogen))

#Check
all_degs_human_mouse %>% 
  filter(adj_p_val <= 0.05) %>% 
  dplyr::count(timepoint, treatment, pathogen, organism) 

#Save
all_degs_human_mouse %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "all_human_mouse_dge_limma_degs.rds"))
```

Duplicate Infection timepoints
```{r}
#========================
#Duplicate to match Infection in human (24h) with different timepoints in mouse (2h, 4h, 6h, 12h, 24h) #========================

all_degs_human_mouse =
  readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds"))

all_degs_human_mouse_matched = 
  bind_rows(
  #Maintain all
  all_degs_human_mouse,
    
  #Duplicate Human infection timepoints with mouse
  all_degs_human_mouse %>% 
  filter(organism == "Human",
         treatment == "Infection") %>% 
  inner_join(all_degs_human_mouse %>% 
               filter(organism == "Mouse",
                      treatment == "Infection") %>% 
               select(pathogen, timepoint) %>%
               distinct(),
             by = "pathogen") %>% 
    mutate(timepoint_comparison = paste0(timepoint.y, "-", timepoint.x)) %>% 
    rename(timepoint = timepoint.y) %>% select(-timepoint.x) 
  ) %>% 
  distinct() 

all_degs_human_mouse_matched %>% 
  filter(adj_p_val <= 0.05) %>% 
  dplyr::count(timepoint_comparison, treatment, pathogen, organism) 

all_degs_human_mouse_matched %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "all_human_mouse_dge_limma_degs_matched.rds"))
```

### Pair all timepoints by pathogen
```{r}
# ==============================================================================
# CREATE ALL TIMEPOINT PAIRS
# ==============================================================================

#=======================
# Import datasets
#=======================
# Import and clean DEGs table
all_degs_human_mouse_unmatched <- readRDS(file = here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint %in% c("2h", "Early hours"), "Early hours", timepoint)) 


# Import and clean Metadata (Get N participants per condition)
all_human_mouse_metadata <- readRDS(here("tables", "all_human_mouse_metadata.rds")) %>% 
  rownames_to_column("~del") %>% 
  group_by(pathogen, organism, timepoint) %>% 
  summarise(n = n(), .groups = "drop")

# Biological negative control
biological_null_degs <- readRDS(file = here("tables", "duchenne_mouse_dge_limma_degs.rds")) %>% 
  filter(timepoint == "Day 28")

# Check original counts
all_degs_human_mouse_unmatched %>% dplyr::count(pathogen, organism, timepoint)


# ==============================================================================
# 2. CREATE TIMEPOINT PAIRS (SIMPLIFIED CROSS-ORGANISM PAIRING)
# ==============================================================================

# Extract species-specific timepoints per pathogen to map them dynamically
timepoint_map <- all_degs_human_mouse_unmatched %>% 
  select(pathogen, organism, timepoint) %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = timepoint, values_fn = list)

# Generate all cross-species timepoint combinations dynamically
all_degs_human_mouse_matched <- all_degs_human_mouse_unmatched %>% 
  left_join(timepoint_map, by = "pathogen") %>% 
  mutate(opposite_timepoint = if_else(organism == "Human", Mouse, Human)) %>% 
  unnest(opposite_timepoint) %>% 
  mutate(
    timepoint_human = if_else(organism == "Human", timepoint, opposite_timepoint),
    timepoint_mouse = if_else(organism == "Mouse", timepoint, opposite_timepoint),
    timepoint_comparison = paste0(timepoint_human, "-", timepoint_mouse)
  ) %>% 
  select(-Human, -Mouse, -opposite_timepoint)

# Check matched counts
all_degs_human_mouse_matched %>% 
  dplyr::count(pathogen, organism, treatment, timepoint_comparison)


######################
#Bind negative control
######################
all_degs_human_mouse_matched_bionull =
  bind_rows(
  #Keep original
  all_degs_human_mouse_matched,

  #Match control with different pathogens and bind
  biological_null_degs %>%
    dplyr::mutate(organism = "DMD") %>%
    dplyr::select(human_symbol:ave_expr, t:b, se, sd, organism) %>%
    tidyr::crossing(
      all_degs_human_mouse_matched %>%
        dplyr::select(pathogen, timepoint, treatment, timepoint_comparison) %>%
        dplyr::distinct()
    )
  )  %>%

  #Add n participants
  left_join(all_human_mouse_metadata,
               by = join_by("pathogen", "organism", "timepoint")) %>%
  #Add n participants for control
  mutate(n = if_else(organism == "DMD", 5, n))  %>% 
  group_by(treatment, pathogen) %>% 
  #Copy timepoint comparison for control to match
  fill(timepoint_comparison, .direction = "down") %>%
  ungroup() %>% 
  mutate(condition = if_else(is.na(condition), paste0(str_replace(pathogen, ". ", ""), "_", timepoint),condition))

#Check
all_degs_human_mouse_matched_bionull %>%
    dplyr::count(treatment, pathogen, organism, condition, timepoint)

#Save
all_degs_human_mouse_matched_bionull %>% 
  saveRDS(file = here("tables", "all_human_mouse_dge_limma_degs_matched_control.rds"))

```







## DGE t-tests and Limma methods
Import previously saved analysis results for combined processing.

```{r}
#Import ------
#t-Test
#Influenza
human_log2fc_paired_unpaired_ttest_influenza =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "influenza_fluad_human_dge_paired_unpaired_ttest.rds")) %>% 
  filter(paired == "Paired") %>% 
# Tag rows as Human: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Human",
         pathogen = "Influenza",
         timepoint = paste("Day", timepoint))
mouse_log2fc_paired_unpaired_ttest_influenza =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "influenza_fluad_mouse_log2fc_paired_unpaired_ttest.rds")) %>% 
  filter(paired == "Paired") %>% 
# Tag rows as Mouse: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Mouse",
         pathogen = "Influenza",
         timepoint = paste("Day", timepoint))
#Hepatitis B
human_log2fc_paired_unpaired_ttest_hepatitisb =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "hepatitisb_engerixb_human_log2fc_paired_unpaired_ttest.rds")) %>% 
  filter(paired == "Paired") %>% 
# Tag rows as Human: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Human",
         pathogen = "Hepatitis B",
         timepoint = paste("Day", timepoint))

mouse_log2fc_paired_unpaired_ttest_hepatitisb =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "hepatitisb_engerixb_mouse_log2fc_paired_unpaired_ttest.rds")) %>% 
  filter(paired == "Paired") %>% 
# Tag rows as Mouse: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Mouse",
         pathogen = "Hepatitis B",
         timepoint = paste("Day", timepoint))
#S. aureus
human_log2fc_paired_unpaired_ttest_saureus =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "saureus_infection_human_log2fc_paired_unpaired_ttest.rds")) %>% 
# Tag rows as Human: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Human",
         pathogen = "Staphylococcus aureus",
         timepoint = paste0(24, "h"))

mouse_log2fc_paired_unpaired_ttest_saureus =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "saureus_infection_mouse_log2fc_paired_unpaired_ttest.rds")) %>% 
  filter(paired == "Paired") %>% 
# Tag rows as Mouse: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Mouse",
         pathogen = "Staphylococcus aureus",
         timepoint = paste0(timepoint, "h"))
#E. coli
human_log2fc_paired_unpaired_ttest_ecoli =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "ecoli_infection_human_log2fc_unpaired_ttest.rds")) %>% 
# Tag rows as Human: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Human",
         pathogen = "Escherichia coli",
         timepoint = paste0(timepoint, "h"))

mouse_log2fc_paired_unpaired_ttest_ecoli =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "ecoli_infection_mouse_log2fc_paired_unpaired_ttest.rds")) %>% 
  filter(paired == "Paired") %>%  
# Tag rows as Mouse: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Mouse",
         pathogen = "Escherichia coli",
         timepoint = paste0(timepoint, "h"))
#Burn
human_log2fc_paired_unpaired_ttest_burn =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "burn_human_log2fc_unpaired_ttest.rds")) %>% 
# Tag rows as Human: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Human",
         pathogen = "Burn",
         timepoint = str_replace(timepoints_new, "_", " "),
         timepoint = str_replace(timepoint, "Early Hours", "Early hours")) %>% 
  rename(padj = p_adj,
         pval = p_val)

mouse_log2fc_paired_unpaired_ttest_burn =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "burn_mouse_log2fc_paired_unpaired_ttest.rds")) %>% 
# Tag rows as Mouse: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Mouse",
         pathogen = "Burn",
         timepoint = case_when(str_detect(timepoint, "Day")~paste("Day", parse_number(timepoint)), 
                               str_detect(timepoint, "Hour")~ "Early hours", 
                               T~timepoint))%>% 
  rename(padj = p_adj,
         pval = p_val)

#Trauma
human_log2fc_paired_unpaired_ttest_trauma =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "trauma_human_log2fc_unpaired_ttest.rds")) %>% 
# Tag rows as Human: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Human",
         pathogen = "Trauma",
         timepoint = str_replace(timepoints_new, "_", " "),
         timepoint = str_replace(timepoint, "Early Hours", "Early hours")) 

mouse_log2fc_paired_unpaired_ttest_trauma =
# Load dataset from disk or GEO: reads pre-saved expression matrix / metadata into the R environment
  readRDS(file = here("tables", "trauma_mouse_log2fc_paired_unpaired_ttest.rds")) %>% 
# Tag rows as Mouse: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Mouse",
         pathogen = "Trauma",
         timepoint = case_when(str_detect(timepoint, "Day")~paste("Day", parse_number(timepoint)), 
                               str_detect(timepoint, "Hour")~ "Early hours", 
                               T~timepoint)) %>% 
  rename(padj = p_adj,
         pval = p_val)


# Merge datasets
all_log2fc_paired_unpaired_ttest = bind_rows(
          human_log2fc_paired_unpaired_ttest_influenza,
          mouse_log2fc_paired_unpaired_ttest_influenza,
          human_log2fc_paired_unpaired_ttest_hepatitisb,
          mouse_log2fc_paired_unpaired_ttest_hepatitisb,
          human_log2fc_paired_unpaired_ttest_saureus,
          mouse_log2fc_paired_unpaired_ttest_saureus,
          human_log2fc_paired_unpaired_ttest_ecoli,
          mouse_log2fc_paired_unpaired_ttest_ecoli,
          human_log2fc_paired_unpaired_ttest_burn,
          mouse_log2fc_paired_unpaired_ttest_burn,
          human_log2fc_paired_unpaired_ttest_trauma,
          mouse_log2fc_paired_unpaired_ttest_trauma)


#Limma
all_degs = readRDS(file = here("tables", "all_human_mouse_dge_limma_degs.rds"))

human_degs = all_degs %>% 
# Standardise column names to snake_case using janitor::clean_names for consistent tidyverse syntax
  clean_names() %>% 
  filter(organism == "Human") %>% 
  rename(padj = adj_p_val) %>% 
  mutate(test = "Limma",
         paired = case_when(pathogen %in% c("Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma") ~ "Unpaired",
                            pathogen %in% c("Influenza", "Hepatitis B") ~ "Paired", 
                            T ~ pathogen)) 

mouse_degs = all_degs %>% 
# Standardise column names to snake_case using janitor::clean_names for consistent tidyverse syntax
  clean_names() %>% 
  filter(organism == "Mouse") %>% 
  rename(padj = adj_p_val) %>% 
# Tag rows as Mouse: required for downstream faceting when combining Mouse and Human results
  mutate(organism = "Mouse",
         test = "Limma",
         paired = case_when(pathogen %in% c("Staphylococcus aureus", "Escherichia coli", "Influenza", "Hepatitis B") ~ "Paired",
                            pathogen %in% c("Burn", "Trauma") ~ "Paired", 
                            T ~ pathogen)) 

#Merge datasets
degs_ttest_limma =
  bind_rows(
    all_log2fc_paired_unpaired_ttest,
    human_degs,
    mouse_degs
    ) %>% 
# Relevel timepoint factor: set baseline (0) as the reference level
  mutate(timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = as.character(pathogen),
         pathogen = case_when(pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
          pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"))

all_log2fc_paired_unpaired_ttest %>% 
  dplyr::count(pathogen, organism, timepoint)

all_degs%>% 
  dplyr::count(pathogen, organism, timepoint)

```

Summarize DEG counts by method and create a comparison plot across all conditions.
```{r fig.height=10, fig.width=12}
#Summarize  
degs_ttest_limma_summary = degs_ttest_limma %>% 
  group_by(pathogen, organism, timepoint, test) %>% 
# Count the number of differentially expressed genes (adj.p ≤ 0.05) per comparison
  summarise( n_degs = sum(padj <= 0.05, na.rm = TRUE)) %>% 
  ungroup()

degs_ttest_limma_summary

# Plot number of DEGs per test and significance level
degs_ttest_limma_summary_plot_gg_1row = degs_ttest_limma_summary %>%
  filter(pathogen %in% c("Influenza", "HepB")) %>% 
# Initialise ggplot with aesthetic mappings: defines variables for x-axis, y-axis, colour, and fill
  ggplot(aes(
    x = timepoint,
    y = n_degs,
    fill = test
  )) +
# Create faceted plot: split data into sub-panels by condition or organism for side-by-side comparison
  facet_wrap(~pathogen~organism, 
                    scales = "free",
                    space = "free_x",
                    # strip = strip_nested(size = "variable"),
                    nrow = 1
             ) +
  geom_col(position = position_dodge2(width = 0.8),
           width = 0.5) +
  # Manual colors per test
  scale_fill_manual(values = c(
    "t-Test" = "gray25",
    "Limma" = "#4DBBD5FF"
  )) +
  
# Configure x-axis: set position, expansion, breaks, and labels
  scale_x_discrete(
    expand = c(0.01, 0.01)
  ) +
  scale_y_continuous(expand = c(0, 0), n.breaks = 5) +
  
# Apply project-specific ggplot theme: consistent fonts, gridlines, and legend styling across all figures
  theme_vaxgo() +
  labs(
    title = "Total DEGs by test and timepoint",
    subtitle = "Comparison across tests (FDR ≤ 0.05)",
    x = "Time",
    y = "DEGs",
    fill = "Test",
    alpha = "Pairing"
  ) +

# Customise theme elements: remove/minimise gridlines, adjust axis text, and fine-tune legend placement
  theme(
    legend.position = "top",
    axis.line.y = element_blank(),
    panel.grid.major.y = element_line(color = "gray75", linetype = 3),
    axis.ticks.y = element_line(color = "gray75"),
    ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

degs_ttest_limma_summary_plot_gg_2row = degs_ttest_limma_summary %>%
  filter((pathogen %in% c("S. aureus", "E. coli"))) %>% 
# Initialise ggplot with aesthetic mappings: defines variables for x-axis, y-axis, colour, and fill
  ggplot(aes(
    x = timepoint,
    y = n_degs,
    fill = test
  )) +
# Create faceted plot: split data into sub-panels by condition or organism for side-by-side comparison
  facet_wrap(~pathogen~organism, 
                    scales = "free",
                    space = "free_x",
                    # strip = strip_nested(size = "variable"),
                    nrow = 1
             ) +
  geom_col(position = position_dodge2(width = 0.8),
           width = 0.5) +
  # Manual colors per test
  scale_fill_manual(values = c(
    "t-Test" = "gray25",
    "Limma" = "#4DBBD5FF"
  )) +
  
# Configure x-axis: set position, expansion, breaks, and labels
  scale_x_discrete(
    expand = c(0.01, 0.01)
  ) +
  scale_y_continuous(expand = c(0, 0), n.breaks = 5) +
  
# Apply project-specific ggplot theme: consistent fonts, gridlines, and legend styling across all figures
  theme_vaxgo() +
  labs(
    x = "Time",
    y = "DEGs",
    fill = "Test",
    alpha = "Pairing"
  ) +

# Customise theme elements: remove/minimise gridlines, adjust axis text, and fine-tune legend placement
  theme(
    legend.position = "top",
    axis.line.y = element_blank(),
    panel.grid.major.y = element_line(color = "gray75", linetype = 3),
    axis.ticks.y = element_line(color = "gray75"),
    ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

degs_ttest_limma_summary_plot_gg_3row = degs_ttest_limma_summary %>%
  filter((pathogen %in% c("Burn", "Trauma"))) %>% 
# Initialise ggplot with aesthetic mappings: defines variables for x-axis, y-axis, colour, and fill
  ggplot(aes(
    x = timepoint,
    y = n_degs,
    fill = test
  )) +
# Create faceted plot: split data into sub-panels by condition or organism for side-by-side comparison
  facet_wrap(~pathogen~organism, 
                    scales = "free",
                    space = "free_x",
                    # strip = strip_nested(size = "variable"),
                    nrow = 1
             ) +
  geom_col(position = position_dodge2(width = 0.8),
           width = 0.5) +
  # Manual colors per test
  scale_fill_manual(values = c(
    "t-Test" = "gray25",
    "Limma" = "#4DBBD5FF"
  )) +
  
# Configure x-axis: set position, expansion, breaks, and labels
  scale_x_discrete(
    expand = c(0.01, 0.01)
  ) +
  scale_y_continuous(expand = c(0, 0), n.breaks = 5) +
  
# Apply project-specific ggplot theme: consistent fonts, gridlines, and legend styling across all figures
  theme_vaxgo() +
  labs(
    x = "Time",
    y = "DEGs",
    fill = "Test",
    alpha = "Pairing"
  ) +

# Customise theme elements: remove/minimise gridlines, adjust axis text, and fine-tune legend placement
  theme(
    legend.position = "top",
    axis.line.y = element_blank(),
    panel.grid.major.y = element_line(color = "gray75", linetype = 3),
    axis.ticks.y = element_line(color = "gray75"),
    ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 


degs_ttest_limma_summary_plot_gg = (degs_ttest_limma_summary_plot_gg_1row /
degs_ttest_limma_summary_plot_gg_2row/
  degs_ttest_limma_summary_plot_gg_3row) +
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "top"))

# Render plot
degs_ttest_limma_summary_plot_gg

degs_ttest_limma_summary_plot_gg %>% 
  ggsave(filename = here("Figures", "all_human_mouse_dge_ttest_limma_summary.png"),
         width = 12,
         height = 10)
```







## (TERMINAR) FIT
Standardize with DGE output.

```{r}
fit_saureus_results = readRDS(file = here("tables", "saureus_mouse_fit_prediction_results_12h.rds"))
fit_ecoli_results = readRDS(file = here("tables", "ecoli_mouse_fit_prediction_results_12h.rds"))
fit_fluad_results = readRDS(file = here("tables", "influenza_fluad_mouse_fit_prediction_results.rds"))
fit_engerixb_results = readRDS(file = here("tables", "hepatitisb_engerixb_fit_prediction_results_alltimes.rds"))

all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds"))

all_fit_results = fit_fluad_results %>% 
  bind_rows(
    fit_fluad_results,
    fit_engerixb_results,
    fit_saureus_results %>% mutate(timepoint = as.character(as.numeric(timepoint)/24)),
    fit_ecoli_results %>% mutate(timepoint = as.character(as.numeric(timepoint)/24))
    ) %>% 
  mutate(organism = case_when(measure == "fit_prediction" ~ "FIT",
                             measure == "mouse_effect_size" ~ "Mouse",
                             T ~ measure),
         timepoint = fct_reorder(timepoint, as.numeric(timepoint))) 

all_fit_results %>% 
# Save R object to disk in native RDS format: preserves data types and structure for fast loading
  saveRDS(here("tables", "all_fit_prediction_results.rds"))

#Standardize with DGE output
all_human_mouse_fit = 
  
  all_degs_human_mouse %>% 
  mutate(timepoint = if_else(pathogen %in% c("Staphylococcus aureus",
                                             "Escherichia coli"), 
                             as.character(round((as.numeric(timepoint)/24), 2)), 
                             timepoint)) %>% 
  
  bind_rows(all_fit_results %>% 
              filter(organism == "FIT") %>% 
              select(organism, pathogen, treatment, timepoint, human_symbol, mean_l2fc = value, 
                     ci_lower = ci_low, ci_upper = ci_high)) %>% 
  
  filter(
  #Filter out different times in infection in mice and FIT
  !(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") &
      organism == "Mouse" &
    timepoint %in% c("0.08", "0.17", "0.25", "1")))
all_human_mouse_fit

all_human_mouse_fit %>% 
  filter(pathogen == "Escherichia coli",
         organism == "Mouse") %>% 
  pull(timepoint) %>% unique()

```


Visualize distribution.

```{r fig.height=5, fig.width=18}
all_human_mouse_fit$timepoint %>% unique()

#Visualize distribution
all_fit_results_histogram = all_human_mouse_fit %>% 
  mutate(pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli")) %>% 
# Create ggplot object: maps data to aesthetic (x = expression value)
  ggplot() +
  aes(x = mean_l2fc,
      fill = timepoint) +
  facet_nested_wrap(~pathogen+organism, axes = "x",
               nest_line = element_line(linetype = 1, color = "black", size = 0.5), 
               scales = "free_y", nrow = 2) +
  # geom_histogram(color = "white") +
  geom_density(aes(y = after_stat(count),
                   color = timepoint,
                   fill = timepoint),
                  position = position_identity(), 
               alpha = 0,
               size = 1.5,
              show.legend = T) +
# Configure x-axis: set position, expansion, breaks, and labels
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(expand = c(0,0)) +
  geom_vline(xintercept = 0, linetype = 2) +
# Apply manual colour scale: uses predefined colour palette (typically from project's colour scheme)
  scale_fill_manual(values =  c("0.08" = "#D0D8FB", 
                                   "0.17" = "#A1B0F7",
                                   "0.5" = "#4361ee",
                                   "1" = "#90e0ef",
                                   "3" = "#0096c7",
                                   "7" = "#184e77")) +
  scale_color_manual(values =  c("0.08" = "#D0D8FB", 
                                   "0.17" = "#A1B0F7",
                                   "0.5" = "#4361ee",
                                   "1" = "#90e0ef",
                                   "3" = "#0096c7",
                                   "7" = "#184e77")) +
# Apply project-specific ggplot theme: consistent fonts, gridlines, and legend styling across all figures
  theme_vaxgo() +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), 
        axis.ticks.y = element_blank(),
        legend.position = "top",
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.line.y = element_blank(),
        panel.spacing.y = unit(0.3, "lines"),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(alpha = 0.5, linetype = 2)
        ) +
  guides(fill = "none") +
# Set informative axis labels, legend titles, and overall plot title
  labs(x = "Log2FC",
       y = "Count",
       fill = "Timepoint (Day)",
       color = "Timepoint")

all_fit_results_histogram

all_fit_results_histogram %>% 
  ggsave(filename = here("Figures", "all_human_mouse_fit_results_histogram.png"), 
         width = 16,
         height = 5)
```








