---
title: "3.1_Comparing_Human_Mouse_ByCondition"
output: html_document
---
# Required

```{r message=FALSE, warning=FALSE}
# install.packages("here") if needed
install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```

#=============================

# 1. Fluad, Vaccine

##Functional analysis

### Condition-wise

BTMs correlation, intra- and inter-species

#### Immune

##### Mean Log2FC and GSEA (Immune)

Load the pre-computed differential expression results and the Blood Transcription Module annotation table. Join them on gene symbol to assign each gene to its BTM module, then collapse genes to module-level mean log2FC for each organism and timepoint. For each module, test whether its mean log2FC differs from zero using a one-sample t-test with Benjamini-Hochberg correction. Save the resulting module-level summary for downstream plotting.

```{r}
# Load pre-computed DGE results from limma output, annotate pathogen/vaccine identity and cast timepoint to numeric --------
# Annotate pathogen/vaccine/timepoint for downstream grouping
all_degs_human_mouse = readRDS(here("tables", "influenza_fluad_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Influenza", vaccine = "Fluad",
         timepoint = as.numeric(timepoint)) 

# Import BTM gene-to-module mapping (one row per gene, columns: symbol, process, group, subgroup) --------
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "influenza_fluad_human_mouse"

# Extract unique condition labels from DGE table for downstream filtering of annotation_conditions --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Join DEGs with BTM annotation --------
# Rename to generic "symbol" for the join, then attach BTM metadata
# Keep only Human and Mouse (exclude any reference or FIT entries)
dge_btm_process_genes = all_degs_human_mouse %>% 
# Rename human_symbol to "symbol" to match the BTM annotation key --------
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
# Attach BTM module membership (process, group, subgroup) for each gene --------
  inner_join(btm_annotation_genes, by = "symbol") %>% 
# Exclude FIT or reference species — keep only Human and Mouse for cross-species comparison --------
  filter(organism %in% c("Human", "Mouse")) 

# Save intermediate gene-level BTM table for reuse in plotting without re-joining --------
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Collapse gene-level log2FC values to BTM-level means per organism and timepoint --------
# For each organism x pathogen x timepoint x process:
#   - mean, median, sd, var of the constituent genes' log2FC
#   - n_genes: how many genes from the module were detected
#   - group/subgroup: preserved from BTM annotation (constant per process)
dge_btm_mean_process = dge_btm_process_genes %>% 
# Group by organism, pathogen, timepoint, and BTM: each group is a set of genes belonging to that module for a given condition --------
  group_by(organism, pathogen, timepoint, process) %>% 
# Compute summary statistics for each module: mean and median as centrality measures, sd and var as dispersion measures --------
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group),
            subgroup = dplyr::first(subgroup)) %>%
# Remove duplicate module-by-organism entries if the join produced any --------
  distinct() %>% 
  ungroup() %>% 
# Build a human-readable condition identifier combining organism, pathogen, and timepoint for downstream merging --------
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint)) 

# Test whether each BTM's mean log2FC differs significantly from zero --------
# One-sample t-test per module (mean_l2fc ~ 1, mu = 0): tests if the module is up- or down-regulated as a whole
# BH correction across all modules within each organism x timepoint grouping
# Modules with fewer than 2 detected genes are excluded (cannot compute variance)
pvals_vs_zero <- dge_btm_process_genes %>%
# Group by the same keys as the mean summarisation for per-module testing --------
  group_by(organism, pathogen, timepoint, process) %>%
# Exclude modules that have only one gene in the dataset (t-test requires at least 2 observations) --------
  filter(n() >= 2) %>% 
# Run one-sample t-test: tests whether the mean log2FC of genes in this module differs from zero --------
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%    
# Apply Benjamini-Hochberg FDR correction across all modules within each group --------
  adjust_pvalue(method = "BH") %>%  
# Add significance stars based on adjusted p-value thresholds --------
  add_significance()

# Diagnostic: print modules with fewer than 2 detected genes (excluded from the t-test) --------
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join module-level mean log2FC with their per-module t-test p-values --------
# Left join preserves all modules from the mean table, even those excluded from the t-test (NA p-values)
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
# Standardise column names to snake_case via janitor --------
  clean_names()

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))

```



##### GSEA

Prepare a rank-ordered gene list from the DGE results using the combined metric `log2FC * -log(adj_p_val)` and run Gene Set Enrichment Analysis against the BTM gene sets. Merge the GSEA output (Normalized Enrichment Score, q-value) with the module-level mean log2FC table for integrated visualization of both magnitude and enrichment direction.

```{r message=FALSE}
# Prepare term-to-gene mapping for GSEA from the BTM annotation table --------
term2gene = btm_annotation_genes %>% 
# Keep only the two columns needed by GSEA: process (gene set name) and genes (gene symbols) --------
  select(process, genes = symbol)

# Build a ranked gene list for GSEA using a composite metric --------
# Metric: log2FC * -log(adj_p_val) — balances effect size and statistical significance
# Positive rank = up-regulated and highly significant; negative rank = down-regulated and highly significant
df_genes2GSEA <- all_degs_human_mouse %>% 
# Remove duplicate gene entries that may arise from multiple probes --------
  distinct() %>% 
# Select only the columns needed for GSEA: rename to match the autoGSEA helper's expected column names --------
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
# Drop genes with missing symbol or fold-change (cannot rank them) --------
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
# Create condition identifier and compute composite rank metric --------
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
# Keep only columns required by autoGSEA: condition label, gene symbol, and rank metric --------
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA across all conditions using the autoGSEA helper --------
# autoGSEA iterates over each unique condition internally, performing fgsea for each
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTMs")

# Merge GSEA enrichment results with module-level means from the mean-Log2FC analysis --------
# Full join keeps: (a) modules with GSEA NES but no mean (rare) and (b) modules with mean but no GSEA result
# p_adjust_mean distinguishes the t-test p-value from the GSEA q-value
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
# Standardise column names --------
  clean_names() %>% 
# Rename 'id' (fgsea's gene-set identifier) to 'process' for consistent merging --------
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
# Merge with the pre-computed module means and t-test results --------
  full_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), 
            by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


#### MSigDB Hallmarks

##### Mean Log2FC

Repeat the module-level analysis using MSigDB Hallmarks instead of BTMs. Load the DGE results and the Hallmark gene sets, join them on gene symbol, collapse to Hallmark-level mean log2FC per organism and timepoint, and test each Hallmark against zero with a one-sample t-test and BH correction. Save the Hallmark-level summary.

```{r fig.width=10, fig.height= 8}
# Load DGE results --------
# Same approach as the BTM analysis: annotate pathogen and cast timepoint to numeric
all_degs_human_mouse = readRDS(here("tables", "influenza_fluad_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Influenza", vaccine = "Fluad",
         timepoint = as.numeric(timepoint)) 

filename = "influenza_fluad_human_mouse"

# Extract unique condition labels for downstream filtering --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Load MSigDB Hallmark gene set annotation with process-group mapping --------
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Join DEGs with Hallmark annotation by gene symbol --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
# Rename to match the Hallmark annotation key --------
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
# Attach Hallmark process and group membership for each gene --------
  inner_join(msigdbr_grouped, by = "symbol")

# Save intermediate Hallmark-gene table for reuse in plotting without re-joining --------
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse gene-level log2FC values to Hallmark-level means per organism and timepoint --------
# Unlike BTMs, Hallmarks lack subgroup columns, so only mean, median, sd, var, and n_genes are computed
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
# Group by organism, timepoint, Hallmark process, and pathogen: each group is a set of genes in that Hallmark for a given condition --------
  group_by(organism, timepoint, process, pathogen) %>% 
# Compute per-Hallmark summary statistics (mean, median, sd, var) as descriptors of Hallmark activity --------
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>% 
  ungroup() %>% 
# Create condition identifier for downstream merging --------
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero, with BH correction --------
# Tests whether the mean log2FC of genes in this Hallmark differs from zero (up- or down-regulated as a group)
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
# Exclude Hallmarks with only one gene (cannot compute t-test variance) --------
  filter(n() >= 2) %>% 
# Run one-sample t-test for each Hallmark group --------
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
# Apply Benjamini-Hochberg FDR correction within each grouping --------
  rstatix::adjust_pvalue(method = "BH") %>%
# Add significance indicator stars --------
  rstatix::add_significance()

# Diagnostic: print Hallmarks with fewer than 2 genes (excluded from the t-test) --------
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join Hallmark-level mean log2FC with their per-Hallmark t-test p-values --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
# Standardise column names --------
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

##### GSEA

Run Gene Set Enrichment Analysis against the MSigDB Hallmarks using the same composite rank metric as the BTMs analysis. Merge the GSEA output (NES, q-value) with the Hallmark-level mean log2FC table so that both magnitude and enrichment can be compared side by side.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping from the MSigDB annotation table --------
term2gene = msigdbr_grouped  %>% 
# Keep only the columns required by autoGSEA: process name and gene symbol --------
  select(process, genes = symbol)

# Build ranked gene list with the same composite metric as the BTM GSEA --------
# Metric: log2FC * -log(adj_p_val) — balances effect size and significance for ranking
df_genes2GSEA <- all_degs_human_mouse %>% 
# Remove duplicate gene entries from multi-probe mappings --------
  distinct() %>% 
# Select and rename columns to match autoGSEA input format --------
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
# Exclude genes with missing symbol or fold-change (cannot compute rank) --------
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
# Create condition label and compute the composite ranking metric --------
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
# Keep only the three columns that autoGSEA expects --------
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks using the autoGSEA helper (iterates over conditions internally) --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level mean log2FC --------
# The full_join keeps Hallmarks that have mean log2FC but no GSEA result (q-value NA) and vice versa
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
# Standardise column names --------
  clean_names() %>% 
# Rename 'id' (fgsea output) to 'process' for merging with the mean table --------
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
# Merge GSEA enrichment (NES, q-value) with the pre-computed Hallmark means and t-test results --------
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```

Plot distributions

Load the Fluad Hallmark GSEA and expression results, pivot to wide format, and arrange Hallmarks by Human log2FC at day 1. The plotting code below is commented out (gated) and can be activated when needed.

```{r fig.height=12, fig.width=10}
# Load the GSEA results, module-level means, and gene-level data --------
gsea_hallmarks_results = readRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
dge_hallmarks_mean_process = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
dge_hallmarks_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Assign handle for matched timepoint analysis (no matching needed here) --------
dge_hallmarks_process_genes_matched = dge_hallmarks_process_genes
gsea_hallmarks_results_matched = gsea_hallmarks_results 

# Pivot GSEA enrichment and mean log2FC to wide format (one column per organism) --------
# Creates a side-by-side view: Human_mean_log2fc | Mouse_mean_log2fc | Human_NES | Mouse_NES etc.
dge_hallmarks_gsea_mean_wide = gsea_hallmarks_results_matched %>% 
# Keep only the columns needed for the wide pivot: organism-specific metrics, timepoint, and process identifier --------
  select(organism, timepoint, process, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
# Group by timepoint and Hallmark: each group has at most one Human and one Mouse row --------
  group_by(timepoint, process,  set_size) %>%
# Widen so that organism becomes a suffix: Human values and Mouse values sit side by side in the same row --------
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

# Order Hallmarks by Human mean log2FC at the first timepoint for consistent plotting --------
process_order_log2fc = dge_hallmarks_mean_process %>% 
# Keep only the variables needed for ordering: process, timepoint, organism, and mean log2FC --------
  select(process, timepoint, organism, mean_log2fc) %>%
# Filter to Human, day 1 (the first timepoint) to establish a single ordering --------
  filter(timepoint == 1,
         organism == "Human") %>% 
  group_by(organism, timepoint) %>% 
# Sort descending by mean log2FC so the most up-regulated Hallmarks appear first --------
  arrange(-mean_log2fc) %>%
  pull(process) %>% 
  unique() 

dge_hallmarks_process_ordered = dge_hallmarks_mean_process 

# Plot code (gated) — activate by removing comment prefix --------
# dge_hallmarks_process_distribution = dge_hallmarks_process_genes_matched %>% 
#   inner_join(msigdbr_grouped %>% select(process, group), by = "process") %>% 
#     mutate(timepoint_num = timepoint,
#     timepoint = paste0("Day ", timepoint_num),
#     timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)])))  %>% 
#   ggplot() +
#   aes(y = process,
#       x = mean_l2fc,
#       color = organism) +
#   facet_grid(~timepoint~group, scales = "free", space = "free") +
#   geom_vline(xintercept = 0, color = "black") +
#   geom_point(pch = "|", alpha = 1, 
#              show.legend = F,
#              size = 3) +
#   stat_summary(
#   aes(color = organism),
#   fun.data = mean_se,
#   geom = "errorbar",
#   width = 0.5,
#   size = 1) +
#   
#   stat_summary(
#     aes(fill = organism), 
#     fun = mean,             
#     color = "black",
#     geom = "point",             
#     size = 2,    
#     pch = 21,  
#     alpha = 1,
#     show.legend = T 
#   ) +
#   
#   theme_vaxgo() +
#   theme(strip.text.y = element_text(angle = 0, hjust = 0), 
#         axis.ticks.y = element_line(color = "gray", alpha = 0.5),
#         legend.position = "top",
#         axis.text.x = element_text(size = 8),
#         axis.text.y = element_text(size = 8),
#         axis.line.y = element_blank(),
#         panel.spacing.y = unit(0.3, "lines"),
#         panel.grid.major.y = element_line(alpha = 0.5),
#         panel.grid.major.x = element_line(alpha = 0.5)) +
#   scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
#                                "Human" = "#90A4AEFF")) +
#   scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
#                                "Human" = "#90A4AEFF")) +
#     labs(color = "Organism",
#        y = "",
#        x = "Log2FC",
#        fill = "Organism",
#        title = "Mouse vs Human, MSigDB genes, mean Log2FC") 
# 
# dge_hallmarks_process_distribution 
# 
# dge_hallmarks_process_distribution %>% 
#   ggsave(filename = here("Figures", paste0(filename, "_dge_hallmarks_process_distribution.png")),
#          height = 12, 
#          width = 10)
```

#=============================

# 2. Engerix B, Vaccine

### Condition-wise

BTMs correlation, intra- and inter-species

#### BTMs

##### Mean Log2FC

Repeat the same module-level analysis for Engerix B (Hepatitis B vaccine). Load the DGE results for this vaccine, join with the BTM annotation, collapse to module-level mean log2FC per organism and timepoint, test each module against zero, and save the summary.

```{r}
# Load Engerix B DGE results from the combined human-mouse limma output --------
all_degs_human_mouse = readRDS(here("tables", "hepatitisb_engerixb_human_mouse_dge_limma_degs.rds")) %>% 
# Annotate pathogen/vaccine identities and cast timepoint to numeric for consistent downstream handling --------
  mutate(pathogen = "Hepatitis B", vaccine = "EngerixB",
         timepoint = as.numeric(timepoint))

# Import BTM gene-to-module mapping (reused across all pathogens) --------
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "hepatitisb_engerixb_human_mouse"

# Extract unique condition labels for downstream filtering --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Join DEGs with BTM module annotation by gene symbol --------
dge_btm_process_genes = all_degs_human_mouse %>% 
# Rename to generic "symbol" for the join, then attach BTM metadata --------
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
# Attach BTM module membership (process, group, subgroup) for each gene --------
  inner_join(btm_annotation_genes, by = "symbol") %>% 
# Keep only Human and Mouse — exclude any reference or FIT entries --------
  filter(organism %in% c("Human", "Mouse")) 

# Save intermediate gene-level BTM table for reuse in plotting --------
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Collapse gene-level log2FC to BTM-level means per organism and timepoint --------
dge_btm_mean_process = dge_btm_process_genes %>% 
# Group by organism x pathogen x timepoint x process: one group per BTM per condition --------
  group_by(organism, pathogen, timepoint, process) %>% 
# Compute per-module summary statistics: mean and median as centrality, sd and var as dispersion, n_genes as coverage --------
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group), 
            subgroup = dplyr::first(subgroup)) %>%
# Remove any duplicate entries from the join --------
  distinct() %>% 
  ungroup() %>% 
# Create a human-readable condition identifier for downstream merging --------
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint)) 

# One-sample t-test per BTM vs zero, with BH correction --------
# Tests whether the module as a whole is up- or down-regulated
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
# Exclude modules with only one detected gene (cannot compute t-test) --------
  filter(n() >= 2) %>% 
# Run one-sample t-test testing whether mean_log2fc differs from zero --------
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
# Apply Benjamini-Hochberg FDR correction per grouping --------
  adjust_pvalue(method = "BH") %>%
# Add significance indicator stars --------
  add_significance()

# Diagnostic: print modules with fewer than 2 detected genes (excluded from the test) --------
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join module-level mean log2FC with their per-module t-test p-values --------
# Left join preserves all modules, even those excluded from the t-test (NA p-values)
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

##### GSEA

Run Gene Set Enrichment Analysis on the Engerix B DGE results using the BTM gene sets with the same composite rank metric as the Fluad analysis. Merge the GSEA output with the module-level mean log2FC table.

```{r}
# Prepare BTM term-to-gene mapping for GSEA --------
term2gene = btm_annotation_genes %>% 
# Keep only the columns required by autoGSEA: process (gene-set name) and genes (gene symbols) --------
  select(process, genes = symbol)

# Build a ranked gene list for GSEA using the composite metric (log2FC * -log(adj_p_val)) --------
# This metric balances effect size and statistical significance for ranking
df_genes2GSEA <- all_degs_human_mouse %>% 
# Remove duplicate gene entries from multi-probe mappings --------
  distinct() %>% 
# Select and rename columns to match autoGSEA input format --------
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
# Drop genes with missing symbol or fold-change (cannot rank) --------
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
# Create condition label and compute the composite ranking metric --------
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
# Keep only the three columns autoGSEA expects: condition, genes, and rank --------
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA across all conditions using the autoGSEA helper --------
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTMs")

# Clean GSEA output and merge with BTM-level means from the mean-Log2FC analysis --------
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
# Standardise column names --------
  clean_names() %>% 
# Rename 'id' (fgsea default) to 'process' for merging --------
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
# Merge GSEA enrichment (NES, q-value) with pre-computed module means and t-test p-values --------
  full_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```

##### Plot distributions

Visualize the distribution of gene-level log2FC within each BTM for both organisms across timepoints. Individual genes are shown as tick marks, the module mean as a filled circle with standard error bars. BTMs are ordered by Human log2FC at the latest timepoint and grouped by immune process category.

```{r fig.height=12, fig.width=10}
# Load GSEA results, module-level means, and gene-level data from saved RDS --------
gsea_btm_results = readRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
dge_btm_mean_process = readRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
dge_btm_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Assign handle for matched timepoint analysis (no matching needed here because all conditions share the same timepoints) --------
dge_btm_process_genes_matched = dge_btm_process_genes

# Pivot GSEA enrichment and mean log2FC to wide format (Human and Mouse side by side) --------
# Creates parallel columns: Human_mean_log2fc | Mouse_mean_log2fc, Human_NES | Mouse_NES, etc.
dge_btm_gsea_mean_wide = gsea_btm_results %>% 
# Parse organism from the condition label (stored as "Human_Influenza_1" or "Mouse_Influenza_1") --------
  mutate(organism = case_when(str_detect(condition, "Mouse") ~ "Mouse",
                              str_detect(condition, "Human") ~ "Human",
                              T~condition)) %>% 
# Keep only the columns needed for the wide pivot --------
  select(organism, timepoint, process, group, subgroup, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
# Group by all non-organism columns: each group has at most one Human and one Mouse row --------
  group_by(timepoint, process, subgroup, group, set_size) %>%
# Widen so that organism becomes a suffix — Human and Mouse values sit side by side in the same row --------
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

# Order BTMs by Human mean log2FC at the latest timepoint, within each immune-process group --------
process_order_log2fc = dge_btm_mean_process %>% 
  select(process, timepoint, organism, mean_log2fc, group) %>%
# Use day 7 (latest timepoint) and Human only as the reference for ordering --------
  filter(timepoint == 7,
         organism == "Human") %>% 
  group_by(organism, timepoint, group) %>% 
# Sort descending so the most up-regulated BTMs appear first within their group --------
  arrange(-mean_log2fc) %>%
  pull(process) %>% 
  unique() 

dge_btm_process_ordered = dge_btm_mean_process %>% 
  ungroup() %>% 
# Relevel group factor to match the immune-process display order, and reverse the process order for plotting (bottom = highest log2FC) --------
    mutate(group = fct_relevel(as.factor(group), c(immune_order)),
           process = fct_relevel(as.factor(process), c(process_order_log2fc)) %>% fct_rev()) 

# Create label styling for BTMs significant in both species (adj.p ≤ 0.10 in Human AND Mouse) --------
# Significant BTMs get bold face and blue colour; non-significant get plain face and grey
labels_process <- dge_btm_gsea_mean_wide %>%
  ungroup() %>% 
  distinct(process, p_adjust_Human, p_adjust_Mouse) %>%
  mutate(label = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
                         "bold", "plain"),
         label_color = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
                         "#4DBBD5FF", "gray50"))  %>% 
  ungroup() %>% 
  distinct() %>% 
  mutate(process = fct_relevel(process, process_order_log2fc)) %>% 
  filter(!is.na(p_adjust_Human))

# Data preparation for the distribution plot: recode numeric timepoint to labelled factor --------
dge_btm_process_distribution = dge_btm_process_genes_matched %>% 
    mutate(timepoint_num = timepoint,
    timepoint = paste0("Day ", timepoint_num),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
# Initialise ggplot with aesthetic mappings: y = BTM process name, x = mean log2FC, colour = organism --------
  ggplot() +
  aes(y = process,
      x = mean_l2fc,
      color = organism) +
# Facet by immune-process group (rows) and timepoint (columns), allowing free scales per panel --------
  facet_grid(~group~timepoint, scales = "free", space = "free") +
# Geom layers: reference line at zero, individual gene points as tick marks, module mean with SE bars --------
  geom_vline(xintercept = 0, color = "black") +
  geom_point(pch = "|", alpha = 1, 
             show.legend = F,
             size = 3) +
  stat_summary(
  aes(color = organism),
  fun.data = mean_se,
  geom = "errorbar",
  width = 0.5,
  size = 1) +
# Module mean as a filled circle with black outline (overlaid on the gene points) --------
  stat_summary(
    aes(fill = organism), 
    fun = mean,             
    color = "black",
    geom = "point",             
    size = 2,    
    pch = 21,  
    alpha = 1,
    show.legend = T 
  ) +
# Theme customization: apply vaxgo theme, adjust facet labels, legend position, and axis text --------
  theme_vaxgo() +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), 
        axis.ticks.y = element_line(color = "gray", alpha = 0.5),
        legend.position = "top",
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8, face = c(labels_process$label), 
                                   color = c(labels_process$label_color)),
        axis.line.y = element_blank(),
        panel.spacing.y = unit(0.3, "lines"),
        panel.grid.major.y = element_line(alpha = 0.5),
        panel.grid.major.x = element_line(alpha = 0.5)) +
# Colour scales: use project-defined colours for Human and Mouse --------
  scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
                                "Human" = "#90A4AEFF")) +
  scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
                                "Human" = "#90A4AEFF")) +
# Labels: set informative axis titles and plot title --------
    labs(color = "Organism",
       y = "",
       x = "Log2FC",
       fill = "Organism",
       title = "Mouse vs Human, BTMs genes, mean Log2FC") 

# Print the plot to the RMarkdown output --------
dge_btm_process_distribution 

#Save
dge_btm_process_distribution %>% 
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_process_distribution.png")),
         height = 12, 
         width = 10)
```

#### MSigDB Hallmarks

##### Mean Log2FC

Repeat the Hallmark-level analysis for Engerix B. Load the all-pathogen DGE file filtered to Hepatitis B, join with the MSigDB Hallmark gene sets, collapse to Hallmark-level mean log2FC per organism and timepoint, test each Hallmark against zero, and save the summary.




```{r fig.width=10, fig.height= 8}
# Load Engerix B DGE results from the combined all-pathogen limma table, filtered to Hepatitis B --------
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Hepatitis B")

filename = "hepatitisb_engerixb_human_mouse"

# Extract unique condition labels for downstream filtering --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Load MSigDB Hallmark gene set annotation with process-group mapping --------
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Join DEGs with Hallmark annotation by gene symbol --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
# Rename to match the Hallmark annotation key --------
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
# Attach Hallmark process and group membership for each gene --------
  inner_join(msigdbr_grouped, by = "symbol")

# Save intermediate Hallmark-gene table for reuse in plotting --------
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse gene-level log2FC to Hallmark-level means per organism and timepoint --------
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
# Group by organism, pathogen, timepoint, and Hallmark process: one group per Hallmark per condition --------
  group_by(organism, pathogen, timepoint, process) %>% 
# Compute per-Hallmark summary statistics (mean, median, sd, var) as descriptors of Hallmark activity --------
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
# Remove duplicate entries from the join --------
  distinct() %>% 
  ungroup() %>% 
# Create condition identifier for downstream merging --------
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero, with BH correction --------
# Tests whether the Hallmark as a whole is up- or down-regulated
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
# Exclude Hallmarks with only one detected gene --------
  filter(n() >= 2) %>% 
# Run one-sample t-test for each Hallmark --------
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
# Apply Benjamini-Hochberg FDR correction --------
  rstatix::adjust_pvalue(method = "BH") %>%
# Add significance stars --------
  rstatix::add_significance()

# Diagnostic: print Hallmarks with fewer than 2 genes (excluded from the t-test) --------
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join Hallmark means with their per-Hallmark t-test p-values --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
# Standardise column names --------
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

##### GSEA

Run Gene Set Enrichment Analysis on the Engerix B DGE results using the MSigDB Hallmark gene sets with the same composite rank metric. Merge the GSEA output with the Hallmark-level mean log2FC table.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping for GSEA from the MSigDB annotation --------
term2gene = msigdbr_grouped  %>% 
# Keep only the columns required by autoGSEA: process name and gene symbol --------
  select(process, genes = symbol)

# Build ranked gene list with the same composite metric as the BTM analysis --------
# Metric: log2FC * -log(adj_p_val) — balances effect size and significance
df_genes2GSEA <- all_degs_human_mouse %>% 
# Remove duplicate gene entries from multi-probe mappings --------
  distinct() %>% 
# Select and rename columns to match autoGSEA input format --------
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
# Drop genes with missing symbol or fold-change (cannot compute rank) --------
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
# Create condition label and compute composite ranking metric --------
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
# Keep only the three columns autoGSEA expects --------
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks using the autoGSEA helper --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level means from the mean-Log2FC analysis --------
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
# Standardise column names --------
  clean_names() %>% 
# Rename 'id' (fgsea output) to 'process' for merging --------
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
# Merge GSEA enrichment (NES, q-value) with pre-computed Hallmark means and t-test results --------
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```

Plot distributions

Visualise the distribution of gene-level log2FC within each Hallmark for both organisms across timepoints. The plot shows individual gene values as points, the module mean as a filled circle, and standard error bars. Hallmarks are ordered by Human log2FC at the latest timepoint.

```{r fig.height=12, fig.width=10}
# Load GSEA results, module-level means, and gene-level data from saved RDS --------
gsea_hallmarks_results = readRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
dge_hallmarks_mean_process = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
dge_hallmarks_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Assign handles for matched timepoint analysis (no matching needed here) --------
dge_hallmarks_process_genes_matched = dge_hallmarks_process_genes
gsea_hallmarks_results_matched = gsea_hallmarks_results 

# Pivot GSEA enrichment and mean log2FC to wide format (Human and Mouse side by side) --------
dge_hallmarks_gsea_mean_wide = gsea_hallmarks_results_matched %>% 
# Parse organism from the condition label --------
  mutate(organism = case_when(str_detect(condition, "Mouse") ~ "Mouse",
                              str_detect(condition, "Human") ~ "Human",
                              T~condition)) %>% 
# Keep only the columns needed for the wide pivot --------
  select(organism, timepoint, process, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
# Group by all non-organism columns: each group has one Human and one Mouse row --------
  group_by(timepoint, process,  set_size) %>%
# Widen so that organism becomes a suffix: Human and Mouse values sit side by side --------
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

# Order Hallmarks by Human mean log2FC at the latest timepoint for consistent plotting --------
process_order_log2fc = dge_hallmarks_mean_process %>% 
  select(process, timepoint, organism, mean_log2fc) %>%
# Use day 7 and Human only as the reference for ordering --------
  filter(timepoint == 7,
         organism == "Human") %>% 
  group_by(organism, timepoint) %>% 
# Sort descending so the most up-regulated Hallmarks appear first --------
  arrange(-mean_log2fc) %>%
  pull(process) %>% 
  unique() 

# Data preparation for the distribution plot: recode numeric timepoint to labelled factor --------
dge_hallmarks_process_distribution = dge_hallmarks_process_genes_matched %>% 
    mutate(timepoint_num = timepoint,
    timepoint = paste0("Day ", timepoint_num),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)])))  %>% 
# Initialise ggplot with aesthetic mappings --------
  ggplot() +
  aes(y = process,
      x = mean_l2fc,
      color = organism) +
# Facetting: one row per immune-process group, one column per timepoint --------
  facet_grid(~group~timepoint, scales = "free", space = "free") +
# Geom layers: reference line at zero, individual gene tick marks, module mean with SE bars --------
  geom_vline(xintercept = 0, color = "black") +
  geom_point(pch = "|", alpha = 1, 
             show.legend = F,
             size = 3) +
  stat_summary(
  aes(color = organism),
  fun.data = mean_se,
  geom = "errorbar",
  width = 0.5,
  size = 1) +
# Module mean as a filled circle overlaid on the gene-level points --------
  stat_summary(
    aes(fill = organism), 
    fun = mean,             
    color = "black",
    geom = "point",             
    size = 2,    
    pch = 21,  
    alpha = 1,
    show.legend = T 
  ) +
# Theme customization: apply vaxgo theme, adjust facet labels and legend --------
  theme_vaxgo() +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), 
        axis.ticks.y = element_line(color = "gray", alpha = 0.5),
        legend.position = "top",
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.line.y = element_blank(),
        panel.spacing.y = unit(0.3, "lines"),
        panel.grid.major.y = element_line(alpha = 0.5),
        panel.grid.major.x = element_line(alpha = 0.5)) +
# Colour scales: use project-defined colours for Human and Mouse --------
  scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
                                "Human" = "#90A4AEFF")) +
  scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
                                "Human" = "#90A4AEFF")) +
# Labels: set informative axis titles and plot title --------
    labs(color = "Organism",
       y = "",
       x = "Log2FC",
       fill = "Organism",
       title = "Mouse vs Human, MSigDB genes, mean Log2FC") 

# Print and save the plot
dge_hallmarks_process_distribution 

dge_hallmarks_process_distribution %>% 
  ggsave(filename = here("Figures", paste0(filename, "_dge_hallmarks_process_distribution.png")),
         height = 12, 
         width = 10)

```


#=============================

# 3. S. aureus

## Functional analysis

### Immune

BTMs correlation, intra- and inter-species

#### Mean Log2FC

Load the S. aureus DGE results, annotate each gene with its BTM membership, collapse to a module-level mean log2FC per organism and timepoint, test each module against zero with a one-sample t-test, and save the outputs.

```{r}
# Load S. aureus DGE output and BTM annotation --------
all_degs_human_mouse = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Saureus")
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "saureus_infection_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

# Annotate DEGs with BTM membership and save gene-level table --------
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

#Save
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Collapse gene-level log2FC to BTM-level means per organism and timepoint --------
dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc), 
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group), 
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per BTM vs zero with BH correction --------
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Identify BTMs with fewer than 2 genes (will fail t-test)
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the S. aureus DGE results using the BTM gene sets. Build the composite rank metric, pass it to autoGSEA, and merge the enrichment results with the BTM-level mean log2FC table.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping --------
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against BTMs --------
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")

# Clean GSEA output and merge with BTM-level means --------
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


### MSigDB Hallmarks

#### Mean Log2FC

Repeat the module-level analysis for S. aureus using MSigDB Hallmarks instead of BTMs. Load the Hallmark gene sets, join with the DGE results, collapse to Hallmark-level mean log2FC per organism and timepoint, test each Hallmark against zero, and save.

```{r fig.width=10, fig.height= 8}
# Load S. aureus DGE output --------
all_degs_human_mouse = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Saureus")

filename = "saureus_infection_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Load Hallmark gene-set annotation --------
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Annotate DEGs with Hallmark membership and save gene-level table --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

#Save
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse gene-level log2FC to Hallmark-level means --------
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero with BH correction --------
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  rstatix::adjust_pvalue(method = "BH") %>%     
  rstatix::add_significance()

# Identify Hallmarks with fewer than 2 genes (will fail t-test)
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the S. aureus DGE results using the MSigDB Hallmark gene sets with the composite rank metric. Merge the GSEA output with the Hallmark-level mean log2FC table.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping --------
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, 
         log2fold_change = mean_l2fc, 
         adj_p_val)  %>% 
  filter(!is.na(genes), 
         !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level means --------
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```


#===========================


# 4. E. coli

## Functional analysis

### Condition-wise analysis

#### Immune

BTMs correlation, intra- and inter-species Compare means and NES

Load the E. coli DGE results, annotate each gene with its BTM membership, collapse to a module-level mean log2FC per organism and timepoint, test each module against zero with a one-sample t-test, and save the outputs.

```{r}
# Load E. coli DGE output and BTM annotation --------
all_degs_human_mouse = readRDS(here("tables", "ecoli_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Ecoli")
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "ecoli_infection_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Annotate DEGs with BTM membership and save gene-level table --------
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

#Save
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Collapse gene-level log2FC to BTM-level means per organism and timepoint --------
dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group), 
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per BTM vs zero with BH correction --------
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Identify BTMs with fewer than 2 genes (will fail t-test)
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

##### GSEA

Run Gene Set Enrichment Analysis on the E. coli DGE results using the BTM gene sets. Build the composite rank metric, pass it to autoGSEA, and merge with the BTM-level means.

```{r}
# Prepare BTM term-to-gene mapping --------
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against BTMs --------
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")

# Clean GSEA output and merge with BTM-level means --------
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


#### MSigDB Hallmarks

##### Mean Log2FC

Repeat the module-level analysis for E. coli using MSigDB Hallmarks instead of BTMs. Load the Hallmark gene sets, join with the DGE results, collapse to Hallmark-level mean log2FC per organism and timepoint, test each Hallmark against zero, and save.

```{r fig.width=10, fig.height= 8}
# Load E. coli DGE output --------
all_degs_human_mouse = readRDS(here("tables", "ecoli_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Ecoli")

filename = "ecoli_infection_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

# Load Hallmark gene-set annotation --------
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Annotate DEGs with Hallmark membership and save gene-level table --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

#Save
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse gene-level log2FC to Hallmark-level means --------
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero with BH correction --------
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance()

# Identify Hallmarks with fewer than 2 genes (will fail t-test)
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

##### GSEA

Run Gene Set Enrichment Analysis on the E. coli DGE results using the MSigDB Hallmark gene sets with the composite rank metric. Merge the GSEA output with the Hallmark-level mean log2FC table.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping --------
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level means --------
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```



#===========================
#5. Burn

## Functional analysis

### Immune

BTMs correlation, intra- and inter-species

#### Mean Log2FC

Load the Burn DGE results, annotate each gene with its BTM membership, collapse to a module-level mean log2FC per organism and timepoint, test each module against zero, and save.

```{r}
# Load Burn DGE output and BTM annotation --------
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Burn")
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "burn_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

# Annotate DEGs with BTM membership and save gene-level table --------
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

#Save
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Collapse gene-level log2FC to BTM-level means per organism and timepoint --------
dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc), 
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group), 
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per BTM vs zero with BH correction --------
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  adjust_pvalue(method = "BH") %>%     
  add_significance()

# Identify BTMs with fewer than 2 genes (will fail t-test)
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the Burn DGE results using the BTM gene sets. Build the composite rank metric, pass it to autoGSEA, and merge with the BTM-level means.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping --------
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against BTMs --------
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")

# Clean GSEA output and merge with BTM-level means --------
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```

### MSigDB Hallmarks

#### Mean Log2FC

Repeat the module-level analysis for Burn using MSigDB Hallmarks. Load the Hallmark gene sets, join with the DGE results, collapse to Hallmark-level mean log2FC per organism and timepoint, test each Hallmark against zero, and save.

```{r fig.width=10, fig.height= 8}
# Load Burn DGE output --------
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Burn")

filename = "burn_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Load Hallmark gene-set annotation --------
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Annotate DEGs with Hallmark membership and save gene-level table --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

#Save
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse gene-level log2FC to Hallmark-level means --------
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero with BH correction --------
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  rstatix::adjust_pvalue(method = "BH") %>%     
  rstatix::add_significance()

# Identify Hallmarks with fewer than 2 genes (will fail t-test)
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the Burn DGE results using the MSigDB Hallmark gene sets with the composite rank metric. Merge the GSEA output with the Hallmark-level mean log2FC table.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping --------
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level means --------
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```


#===========================
#6. Trauma

## Functional analysis

### Immune

BTMs correlation, intra- and inter-species

#### Mean Log2FC

Load the Trauma DGE results, annotate each gene with its BTM membership, collapse to a module-level mean log2FC per organism and timepoint, test each module against zero, and save.

```{r}
# Load Trauma DGE output and BTM annotation --------
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Trauma")
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "trauma_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

# Annotate DEGs with BTM membership and save gene-level table --------
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

#Save
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Collapse gene-level log2FC to BTM-level means per organism and timepoint --------
dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc), 
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group), 
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per BTM vs zero with BH correction --------
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  adjust_pvalue(method = "BH") %>%     
  add_significance()

# Identify BTMs with fewer than 2 genes (will fail t-test)
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the Trauma DGE results using the BTM gene sets. Build the composite rank metric, pass it to autoGSEA, and merge with the BTM-level means.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping --------
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against BTMs --------
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")

# Clean GSEA output and merge with BTM-level means --------
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


### MSigDB Hallmarks

#### Mean Log2FC

Repeat the module-level analysis for Trauma using MSigDB Hallmarks. Load the Hallmark gene sets, join with the DGE results, collapse to Hallmark-level mean log2FC per organism and timepoint, test each Hallmark against zero, and save.

```{r fig.width=10, fig.height= 8}
# Load Trauma DGE output --------
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Trauma")

filename = "trauma_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Load Hallmark gene-set annotation --------
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Annotate DEGs with Hallmark membership and save gene-level table --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

#Save
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse gene-level log2FC to Hallmark-level means --------
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero with BH correction --------
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  rstatix::adjust_pvalue(method = "BH") %>%     
  rstatix::add_significance()

# Identify Hallmarks with fewer than 2 genes (will fail t-test)
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the Trauma DGE results using the MSigDB Hallmark gene sets with the composite rank metric. Merge the GSEA output with the Hallmark-level mean log2FC table.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping --------
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, 
         log2fold_change = mean_l2fc, 
         adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level means --------
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```


#===========================

#7. Unite output datasets

```{r}
#Import datasets ------
#Import DGE output
dge_long_genes_matched = readRDS(here("tables", "all_human_mouse_dge_limma_degs_matched.rds")) 
#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "all_human_mouse"

#Clean metadata 
annotation_conditions = dge_long_genes_matched %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

#GSEA BTMs

#Import ========================

all_human_mouse_metadata = readRDS(here("tables", "all_human_mouse_metadata.rds"))
#Unite BTM means and NES
gsea_btm_results_saureus = readRDS(here("tables", paste0("saureus_infection_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_ecoli = readRDS(here("tables", paste0("ecoli_infection_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_fluad = readRDS(here("tables", paste0("influenza_fluad_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_engerix = readRDS(here("tables", paste0("hepatitisb_engerixb_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_burn = readRDS(here("tables", paste0("burn_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_trauma = readRDS(here("tables", paste0("trauma_human_mouse_gsea_btm_results.rds")))

gsea_btm_results_all = bind_rows(
  gsea_btm_results_saureus,
  gsea_btm_results_ecoli,
  gsea_btm_results_fluad %>% mutate(timepoint = as.character(timepoint)),
  gsea_btm_results_engerix %>% mutate(timepoint = as.character(timepoint)),
  gsea_btm_results_trauma,
  gsea_btm_results_burn
)%>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint == "2h", "Early hours", timepoint)) %>% 
  #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Influenza", "Hepatitis B") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Ecoli", "Saureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = if_else(timepoint == "1h" & pathogen == "Saureus", "24h", timepoint),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_comparison = if_else(pathogen %in% c("Ecoli", "Saureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early"),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Saureus", "Ecoli", "Burn", "Trauma"))

gsea_btm_results_all %>% 
  dplyr::count(pathogen, timepoint, organism)

#Unite BTM means
dge_btm_mean_process_saureus = readRDS(file = here("tables", paste0("saureus_infection_human_mouse_dge_btm_mean_process.rds")))
dge_btm_mean_process_ecoli = readRDS(file = here("tables", paste0("ecoli_infection_human_mouse_dge_btm_mean_process.rds")))
dge_btm_mean_process_fluad = readRDS(file = here("tables", paste0("influenza_fluad_human_mouse_dge_btm_mean_process.rds")))
dge_btm_mean_process_engerix  = readRDS(file = here("tables", paste0("hepatitisb_engerixb_human_mouse_dge_btm_mean_process.rds")))
dge_btm_mean_process_trauma  = readRDS(file = here("tables", paste0("trauma_human_mouse_dge_btm_mean_process.rds")))
dge_btm_mean_process_burn  = readRDS(file = here("tables", paste0("burn_human_mouse_dge_btm_mean_process.rds")))

dge_btm_mean_process_all = bind_rows(
  dge_btm_mean_process_saureus,
  dge_btm_mean_process_ecoli,
  dge_btm_mean_process_fluad %>% mutate(timepoint = as.character(timepoint)),
  dge_btm_mean_process_engerix %>% mutate(timepoint = as.character(timepoint)),
  dge_btm_mean_process_burn,
  dge_btm_mean_process_trauma
) %>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint == "2h", "Early hours", timepoint)) %>% 
  #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Influenza", "Hepatitis B") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Ecoli", "Saureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = if_else(timepoint == "1h" & pathogen == "Saureus", "24h", timepoint),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_comparison = if_else(pathogen %in% c("Ecoli", "Saureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early"),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Saureus", "Ecoli", "Burn", "Trauma"))
dge_btm_mean_process_all %>% 
  dplyr::count(pathogen, timepoint, organism)


#Unite DGE results annotated with BTMs
dge_btm_process_genes_saureus = readRDS(file = here("tables", paste0("saureus_infection_human_mouse_dge_btm_process_genes.rds")))
dge_btm_process_genes_ecoli = readRDS(file = here("tables", paste0("ecoli_infection_human_mouse_dge_btm_process_genes.rds")))
dge_btm_process_genes_fluad = readRDS(file = here("tables", paste0("influenza_fluad_human_mouse_dge_btm_process_genes.rds")))
dge_btm_process_genes_engerix = readRDS(file = here("tables", paste0("hepatitisb_engerixb_human_mouse_dge_btm_process_genes.rds")))
dge_btm_process_genes_burn = readRDS(file = here("tables", paste0("burn_human_mouse_dge_btm_process_genes.rds")))
dge_btm_process_genes_trauma = readRDS(file = here("tables", paste0("trauma_human_mouse_dge_btm_process_genes.rds")))

dge_btm_process_genes_all = bind_rows(
  dge_btm_process_genes_saureus,
  dge_btm_process_genes_ecoli,
  dge_btm_process_genes_fluad %>% mutate(timepoint = as.character(timepoint)),
  dge_btm_process_genes_engerix %>% mutate(timepoint = as.character(timepoint)),
  dge_btm_process_genes_burn,
  dge_btm_process_genes_trauma
) %>% 
   mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint == "2h", "Early hours", timepoint)) %>% 
  #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Influenza", "Hepatitis B") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Ecoli", "Saureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = if_else(timepoint == "1h" & pathogen == "Saureus", "24h", timepoint),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_comparison = if_else(pathogen %in% c("Ecoli", "Saureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early"),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Saureus", "Ecoli", "Burn", "Trauma"))
dge_btm_process_genes_all %>% 
  dplyr::count(pathogen, timepoint, organism)


#Import results -------

filename = "all_human_mouse"

#Duplicate reference, if timepoints do not match =================
gsea_btm_results_all_matched = gsea_btm_results_all %>% 
  select(pathogen, organism, process, group, timepoint, nes, p_adjust, p_adjust_mean, mean_log2fc, core_enrichment, set_size) %>% 
  bind_rows(
  #Duplicate Human
  gsea_btm_results_all %>% 
  filter(organism == "Human",
         pathogen %in% c("Saureus", "Ecoli")) %>% 
  select(pathogen, organism, process, group, nes, p_adjust, p_adjust_mean, mean_log2fc, core_enrichment, set_size) %>% 
  inner_join(gsea_btm_results_all %>% 
               filter(organism == "Mouse",
                      pathogen %in% c("Saureus", "Ecoli")) %>% 
               select(pathogen, timepoint) %>% 
               distinct(),
             by = "pathogen")
  ) %>% 
  distinct()

dge_btm_process_genes_matched = dge_btm_process_genes_all %>% 
  select(pathogen, organism, symbol, process, group, timepoint, adj_p_val, mean_l2fc) %>% 
  bind_rows(
  #Duplicate Human
  dge_btm_process_genes_all %>% 
  filter(organism == "Human",
         pathogen %in% c("Saureus", "Ecoli")) %>% 
  select(pathogen, symbol, organism, process, group, adj_p_val,mean_l2fc) %>% 
  inner_join(dge_btm_process_genes_all %>% 
               filter(organism == "Mouse",
                      pathogen %in% c("Saureus", "Ecoli")) %>% 
               select(pathogen, timepoint) %>% 
               distinct(),
             by = "pathogen")
  ) %>% 
  distinct()


#Pivot to wide GSEA + mean
dge_btm_gsea_mean_wide = gsea_btm_results_all_matched %>% 
  select(organism, pathogen, timepoint, process, group, set_size, nes, mean_log2fc, p_adjust, p_adjust_mean) %>% 
  group_by(pathogen, process, group, set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, p_adjust_mean, set_size),
    names_sep = "_") 

dge_btm_gsea_mean_wide  %>% 
  ungroup() %>% 
  dplyr::count(pathogen, timepoint)

gsea_btm_results_all_matched %>% 
  dplyr::count(pathogen, organism, timepoint)

#Save
dge_btm_gsea_mean_wide %>% 
  saveRDS(file = here("tables", "dge_btm_gsea_mean_wide.rds"))
 
gsea_btm_results_all_matched %>% 
  saveRDS(file = here("tables", "gsea_btm_results_all_matched.rds"))

```


#===========================


