---
title: "2_Comparing_Human_Mouse"
output: html_document
---

# Required

```{r message=FALSE, warning=FALSE}
# install.packages("here") if needed
install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```


# Unified analyses
## 1. Differential gene expression analysis
### Volcano plot
```{r fig.height=16, fig.width=16}
all_degs_human_mouse = readRDS(file = here("tables", "all_human_mouse_dge_limma_degs.rds"))

#Volcano plot -------
volcano_data <- all_degs_human_mouse %>% 
  mutate(logp = -log10(adj_p_val),
         sig = case_when(mean_l2fc > 0 & logp > 1.3 ~ "Up",
                         mean_l2fc < 0 & logp > 1.3 ~ "Down",
                         TRUE ~ "NS"))

volcano_plot = volcano_data %>% 
  mutate(pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma")) %>% 
ggplot() +
  aes(x = mean_l2fc, y = logp, fill = sig) +
  facet_nested_wrap(~pathogen+timepoint+organism, 
               scales = "free", 
               axes = "all",
               nest_line = element_line(linetype = 1, color = "black")) +
  geom_point(aes(fill = sig,
                 color = sig),
             pch = 21, 
             size = 1, 
             alpha = 1) +
  geom_vline(xintercept = 0, color = "black", linetype = 1, size = 0.5) +
  geom_hline(yintercept = 1.3, color = "black", linetype = 2, size = 0.5) +
  scale_y_continuous(n.breaks = 5,
                     # limits = c(0, 70),
                     expand = c(0.01, 0.01)) +
  guides(color = F) +
  scale_color_manual(
    values = c(
      "Up" = "#165289", 
      "Down" = "#45efc6",
      "NS" = "gray80")) +
  scale_fill_manual(
    values = c(
      "Up" = "#165289", 
      "Down" = "#45efc6",
      "NS" = "gray80")) +
  labs(x = "Log2-Fold Change",
       y = "-log(FDR)",
       fill = "Direction") +
  theme_vaxgo() +
  theme(legend.position = "top",
        legend.direction = "horizontal",
        legend.title = element_blank(),
        legend.key = element_blank(),
        strip.text = element_text(size = 10),
        axis.text = element_text(size = 10),
        axis.text.x = element_text(angle = 0))  +
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = "none")

#Save
volcano_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_qc_volcano_plot_raw.png"),
         width = 16,
         height = 16)

#Print
volcano_plot
```




### Overlapping DEGs


```{r fig.width=10, fig.height=4}
#Import datasets
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs_matched.rds")) 

btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

dge_long_genes_matched = all_degs_human_mouse %>% 
  mutate(pathogen = case_when(pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              pathogen == "Hepatitits B" ~ "HepB",
                              T ~ pathogen))

dge_long_genes_matched %>% 
  dplyr::count(pathogen, organism, timepoint_comparison) 

#Get total genes by each condition
degs_mouse = dge_long_genes_matched %>% 
  filter(adj_p_val <= 0.05,
         organism == "Mouse")
degs_human = dge_long_genes_matched %>% 
  filter(adj_p_val <= 0.05,
         organism == "Human")

degs_mouse_summarize = dge_long_genes_matched %>% 
  filter(adj_p_val <= 0.05) %>% 
  group_by(organism, pathogen, timepoint_comparison) %>% 
  summarise(n_genes = n()) 

#Print
degs_mouse_summarize

#Get overlap -----
#Get symbols
degs_mouse_genes = degs_mouse %>% 
  select(human_symbol, timepoint_comparison, pathogen) %>% 
  distinct()

degs_human_genes = degs_human %>% 
  select(human_symbol, timepoint_comparison, pathogen) %>% 
  distinct()

#Inner join, Mouse and Human ------

#Mouse
degs_human_and_mouse_genes = degs_human_genes %>% 
  inner_join(degs_mouse_genes, 
             by = join_by("pathogen", "human_symbol", "timepoint_comparison")) %>% 
  mutate(condition = "Mouse AND Human",
         comparison = "Mouse vs Human",
         organism = "Mouse",
         group = "Common")

#Duplicate rows for human
degs_human_and_mouse_genes_human = degs_human_and_mouse_genes %>% 
  mutate(organism = "Human")


#Anti join, Human vs Mouse -------
#Human
#Human vs Mouse
degs_human_Not_mouse_genes = degs_human_genes %>% 
  anti_join(degs_mouse_genes, 
             by = join_by("pathogen", "human_symbol", "timepoint_comparison")) %>% 
  mutate(condition = "Human NOT Mouse", 
         comparison = "Mouse vs Human",
         organism = "Human",
         group = "Not common")

#Models
#Mouse vs Human
degs_mouse_Not_human_genes = degs_mouse_genes %>% 
  anti_join(degs_human_genes, 
             by = join_by("pathogen", "human_symbol", "timepoint_comparison")) %>% 
  mutate(condition = "Mouse NOT Human",
         organism = "Mouse",
         comparison = "Mouse vs Human",
         group = "Not common")


#Merge data ------
degs_human_mouse_overlap_genes = 
  bind_rows(degs_human_and_mouse_genes,
            degs_human_and_mouse_genes_human,
            degs_human_Not_mouse_genes,
            degs_mouse_Not_human_genes 
            )

#Classify by same and exclusive DEGs -----
degs_human_mouse_overlap_genes_summary = degs_human_mouse_overlap_genes %>% 
  group_by(pathogen, organism, condition, timepoint_comparison) %>% 
  summarise(n = n()) %>% 
  distinct() %>%
  ungroup() %>% 
  separate(timepoint_comparison, into = c("timepoint_mouse", "timepoint_human"), sep = "-", remove = F) %>% 
  filter(!c(pathogen == "Staphylococcus aureus" & timepoint_mouse == "24h")) %>% 
    mutate(condition_new = case_when(condition == "Human NOT Mouse" ~ "Human only",
                                   condition == "Mouse NOT Human" ~ "Mouse only",
                                   condition == "Mouse AND Human" ~ "Shared",
                                   T~condition),
         condition_new = fct_relevel(condition_new,  "Human only",  "Mouse only", "Shared")) %>% 
  #Factor reordering
    mutate(pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma"),
       timepoint_comparison = paste0(timepoint_mouse,  sep = "-", timepoint_human)) %>% 
  filter(!c(pathogen == "Staphylococcus aureus" & timepoint_mouse == "24h")) %>%
  mutate(timepoint_mouse = as.character(timepoint_mouse),
         timepoint_mouse = fct_relevel(timepoint_mouse, c("Early", "Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_human = as.character(timepoint_human),
         timepoint_human = fct_relevel(timepoint_human, c("Early", "Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         treatment = case_when(pathogen %in% c("Hepatitis B", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")) %>% 
  arrange(timepoint_mouse)
  

ranges_y = degs_human_mouse_overlap_genes_summary %>%
  group_by(pathogen) %>%
  summarise(max = max(n))

degs_human_mouse_overlap_genes_summary 
ranges_y
```

Plot overlap (bars)

```{r fig.height=12, fig.width=5}
# #Plot overlap
# degs_human_mouse_overlap_common_plot = degs_human_mouse_overlap_genes_summary %>%
#   ggplot() +
#   aes(x = organism,
#       y = n,
#       fill = condition_new) + 
#   geom_col(alpha = 1) +
#   facet_nested_wrap(~treatment+pathogen+timepoint, 
#                     axes = "all",
#                     ncol = 6, 
#                     scales = "free_y", 
#                     # remove_labels = "all",
#                nest_line = element_line(linetype = 1, color = "black")) +
# 
#   #Manual scales
#   scale_y_continuous(n.breaks = 5,
#                      expand = c(0.01, 0.01)) +
#   facetted_pos_scales(y = list(
#     #Influenza
#     scale_y_continuous(limits = c(0, 6000), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),  
#     scale_y_continuous(limits = c(0, 6000), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 6000), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     #Hepatitis B
#     scale_y_continuous(limits = c(0, 6020), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 6020), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 6020), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     #Staphylococcus aureus
#     scale_y_continuous(limits = c(0, 3650), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 3650), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 3650), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 3650), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     #E. coli
#     scale_y_continuous(limits = c(0, 4200), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 4200), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 4200), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 4200), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     #Burn
#     scale_y_continuous(limits = c(0, 10100), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 10100), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 10100), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 10100), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     #Trauma
#     scale_y_continuous(limits = c(0, 12000), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 12000), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 12000), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale())),
#     scale_y_continuous(limits = c(0, 12000), expand = c(0, 0),
#                        labels = label_number(scale_cut = cut_short_scale()))
#     
#   )) +
#   # scale_fill_manual(values = c("Common" = "#175289",
#   #                              "Not common" = "gray90"))    +
#   scale_fill_manual(values = colors$comparison)    +
#   theme_vaxgo() +
#   theme(
#     strip.text = element_text(size = 10, margin = margin(1, 0, 1, 0)),
#     strip.text.y = element_text(hjust = 0.5, size = 10),
#     axis.text.x = element_text(angle = 0),
#     axis.text.y = element_text(size = 8, hjust = 1),
#     panel.grid.major.x = element_blank(),
#     panel.grid.major.y = element_blank(),
#     panel.spacing.y = unit(0, "cm"), 
#     panel.spacing.x = unit(0, "cm"), 
#     plot.margin = margin(0, 0.1, 0, 0, unit = "cm"), 
#     legend.margin = margin(0,0,0,0),
#     legend.box.margin = margin(0, 0, -10, 0), 
#     legend.position = "top",
#     legend.direction = "horizontal", 
#     legend.key.width = unit(0.4, 'cm'),
#     legend.key.height = unit(0.4, 'cm'),
#     legend.title = element_blank(),
#     legend.text = element_text(size = 10, vjust = 0.5, hjust = -0.2)
#   ) +
#   labs(
#     x = "",
#     y = "Total"
#     ) 
# 
# degs_human_mouse_overlap_common_plot

#Save
# degs_human_mouse_overlap_common_plot %>%
#   ggsave(filename = here("Figures", "all_human_mouse_dge_degs_overlap_barplot.png"),
#          width = 5,
#          height = 12)

# degs_human_mouse_overlap_common_plot
```

### Lineplot of overlapped genes
```{r fig.height=12, fig.width=3}
#Plot overlap
degs_human_mouse_overlap_lineplot = degs_human_mouse_overlap_genes_summary %>%
  select(-organism) %>% 
  distinct() %>% 
  ggplot() +
  
  aes(x = timepoint_mouse,
      y = n,
      fill = condition_new,
      color = condition_new,
      group = condition_new) + 
  
  facet_nested_wrap(~treatment+pathogen, 
                    axes = "all",
                    ncol = 1, 
                    scales = "free", 
                    # remove_labels = "all",
               nest_line = element_line(linetype = 1, color = "black")) +
  geom_line(width = 1) +
  
  geom_point(size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n), 
                  size = 2, 
                  # direction = "y", 
                  segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.05, 0.05),
                     labels = label_number(scale_cut = cut_short_scale())) +
  theme_vaxgo() +
  scale_fill_manual(values = colors$comparison)    +
  scale_color_manual(values = colors$comparison)    +
  labs(x = "Time",
       y = "Total",
       fill = "",
       color = "") +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          axis.title.x = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 2, 0, 1, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  guides(color = "none")



degs_human_mouse_overlap_lineplot

#Save
degs_human_mouse_overlap_lineplot %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_degs_overlap_lineplot.png"),
         width = 3.5,
         height = 12)
```

## 2. Gene expression correlation analysis

### 2.1. Correlation: Genes, means, Overall

#### Pivot DEGs table
```{r fig.height=20, fig.width=20}
#Import datasets ==============
#DEGs (matched)
all_degs_human_mouse =
  readRDS(here("tables", "all_human_mouse_dge_limma_degs_matched.rds"))
#DEGs (unmatched)
all_degs_human_mouse_unmatched =
  readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds"))

#Check
all_degs_human_mouse %>% 
  dplyr::count(timepoint, organism, treatment, pathogen) 

#Functional modules
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))


#####################################
# Different timepoints, same species
#####################################

human_mouse_log2fc_avg_wide_all = 

  #Condition 1
  all_degs_human_mouse_unmatched %>% 
  select(organism, pathogen:human_symbol, mean_l2fc:sd, sd_sampling_stability) %>% 
  rename(mean_log2fc = mean_l2fc, p_adj = adj_p_val, 
    tstatistic = t, bstatistic = b) %>% 
  rename_with(~ paste0(.x, "_1"), all_of(c("organism", "condition", "timepoint", "mean_log2fc", "ci_lower", "ci_upper", "ave_expr", "tstatistic", "p_value", "p_adj", "bstatistic", "se", "sd", "sd_sampling_stability"))) %>% 
  inner_join(
  #Condition 2
   all_degs_human_mouse_unmatched %>% 
  select(organism, pathogen:human_symbol, mean_l2fc:sd, sd_sampling_stability) %>% 
  rename(mean_log2fc = mean_l2fc, p_adj = adj_p_val, 
    tstatistic = t, bstatistic = b) %>% 
  rename_with(~ paste0(.x, "_2"), all_of(c("organism", "condition", "timepoint", "mean_log2fc", "ci_lower", "ci_upper", "ave_expr", "tstatistic", "p_value", "p_adj", "bstatistic", "se", "sd", "sd_sampling_stability"))),
  
  by = join_by("pathogen", "treatment", "human_symbol")
  ) %>% 
  
  
  # Filter for strict 1:1 ortholog pairs with complete expression profiles
  drop_na(mean_log2fc_1, mean_log2fc_2) %>% 
  
  # Compute expression divergence and volatility metrics
  mutate(
    # Cross-species effect size delta (Directional macroevolutionary divergence)
    mean_log2fc_diff = mean_log2fc_1 - mean_log2fc_2,
    
    # Coefficient of Variation - CV
    # --------------------------------------------------------------------------
    cv_1 = sd_1 / abs(mean_log2fc_1),
    cv_2 = sd_2 / abs(mean_log2fc_2),
    cv_mean  = (cv_1 + cv_2) / 2, # Combined cross-species relative noise
    
    # Statistical Weights for Downstream Weighted Correlation
    # Joint geometric weight based on downsampling iteration variance (SD^2)
    weight_sd_inverse = 1 / (sd_1^2 + sd_2^2),
    # Joint geometric weight based on native linear model standard error (SE^2)
    weight_se_inverse = 1 / (se_1^2 + se_2^2),
    
    se_combined = sqrt(se_1^2 + se_2^2)
  ) %>% 
  
  
  #Factor reordering
    mutate(pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma"),
       timepoint_comparison = paste0(timepoint_1,  sep = "-", timepoint_2)) %>% 
  # filter(!c(pathogen == "Staphylococcus aureus" & timepoint_mouse == "24h")) %>%
  mutate(timepoint_2 = as.character(timepoint_2),
         timepoint_2 = fct_relevel(timepoint_2, c("Early", "Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_1 = as.character(timepoint_1),
         timepoint_1 = fct_relevel(timepoint_1, c("Early", "Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         treatment = case_when(pathogen %in% c("Hepatitis B", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "Staphylococcus aureus", "Escherichia coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")) %>% 
  arrange(timepoint_1) %>%
  #Annotate Inter and intra-species comparisons
  mutate(comparison_set = case_when(organism_1 != organism_2 ~ "Inter-species",
                                    organism_1 == organism_2 ~ "Intra-species",
                                    T ~ NA)) 



```







#### Create Biological null distribution
```{r}
biological_null_degs = readRDS(file = here("tables", "duchenne_mouse_dge_limma_degs.rds")) %>% 
  select(-comparison) %>% 
  rename(mean_log2fc = mean_l2fc, 
         p_adj = adj_p_val, 
         tstatistic = t, 
         bstatistic = b) %>% 
  rename_with(~ paste0(.x, "_2"), all_of(c( "pathogen", "organism", "treatment", "condition", "timepoint", "mean_log2fc", "ci_lower", "ci_upper", "ave_expr", "tstatistic", "p_value", "p_adj", "bstatistic", "se", "sd")))

# Create Biological null distribution =====


# 1. Get the target data using the arguments
target_df <- human_mouse_log2fc_avg_wide_all %>%
  filter(organism_1 == "Human", organism_2 == "Mouse") %>%
  mutate(divergence = mean_log2fc_1 - mean_log2fc_2, model_type = "Observed Target") %>% 
  rename(pathogen_1 = pathogen, treatment_1 = treatment) %>% 
  mutate(pathogen_2 = pathogen_1, treament_2 = treatment_1)

# 2. Get the biological null data and join
null_df <- biological_null_degs %>%
  filter(condition_2 == "DMD_Day28") %>%
  inner_join(target_df %>% select(!ends_with("_2")), by = "human_symbol", suffix = c("_2", "_1")) %>%
  mutate(divergence = mean_log2fc_1 - mean_log2fc_2,
         model_type = "Biological Null") %>% 
  # Compute expression divergence and volatility metrics
  mutate(
    # Cross-species effect size delta (Directional macroevolutionary divergence)
    mean_log2fc_diff = mean_log2fc_1 - mean_log2fc_2,
    
    # Coefficient of Variation - CV
    # --------------------------------------------------------------------------
    cv_1 = sd_1 / abs(mean_log2fc_1),
    cv_2 = sd_2 / abs(mean_log2fc_2),
    cv_mean  = (cv_1 + cv_2) / 2, # Combined cross-species relative noise
    
    # Statistical Weights for Downstream Weighted Correlation
    # Joint geometric weight based on downsampling iteration variance (SD^2)
    weight_sd_inverse = 1 / (sd_1^2 + sd_2^2),
    # Joint geometric weight based on native linear model standard error (SE^2)
    weight_se_inverse = 1 / (se_1^2 + se_2^2),
    
    se_combined = sqrt(se_1^2 + se_2^2)
  ) 

# 3. Return combined rows
human_mouse_log2fc_avg_wide_interspecies_bionull = bind_rows(target_df, null_df)


#Get immune and non immune genes
human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune = 
  
  human_mouse_log2fc_avg_wide_interspecies_bionull %>%
  mutate(gene_set = "All genes") %>% 
  
  bind_rows(
  #Same DEGs
  human_mouse_log2fc_avg_wide_interspecies_bionull %>%
  filter(p_adj_1 <= 0.05 & p_adj_2 <= 0.05) %>% 
  mutate(gene_set = "DEGs"), 
  
  #Classify as immune or non_immune 
  human_mouse_log2fc_avg_wide_interspecies_bionull %>% 
  mutate(gene_set = if_else(human_symbol %in% c(btm_annotation_genes %>%
                               filter(module_category == "immune") %>% 
                               pull(symbol)),
                             "Immune",
                             "Not-immune")) %>% 
  filter(p_adj_1 <= 0.05 & p_adj_2 <= 0.05))


human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>% 
  dplyr::count(pathogen_1, pathogen_2, comparison_set)

# #Save
# human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>% 
#   saveRDS(here("tables", "human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune.rds"))

```

#### Compute correlation
```{r fig.height=20, fig.width=20}
#####################
#Correlation (Weighted)
#####################

# Inter-species ==========
correlation_filtered_inter_weighted =      
  human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>%
  #In the inter-species set, keep the Human as a reference (Organism_1)
  filter(!(comparison_set == "Inter-species" & organism_1 == "Mouse")) %>% 
  filter(comparison_set == "Inter-species") %>% 
  group_by(comparison_set, treatment_1, treatment_2, pathogen_1, pathogen_2, 
           timepoint_1, timepoint_2,
           gene_set) %>%
  summarise({
    # Compute weighted correlation
    res <- wtd.cor(x = rank(mean_log2fc_1), 
                   y = rank(mean_log2fc_2), 
                   weight = weight_se_inverse)
    
    # Extract statistics
    data.frame(
      rho = res[1, "correlation"],
      se  = res[1, "std.err"],
      t   = res[1, "t.value"],
      p   = res[1, "p.value"]
    )
  }, .groups = 'drop')  %>% 
  #Rename 1 and 2 as human and mouse, respectively
     rename_with(
    ~ str_replace(.x, "_1$", "_human"), 
    ends_with("_1")
  ) %>% 
  rename_with(
    ~ str_replace(.x, "_2$", "_mouse"), 
    ends_with("_2")
  ) %>% 
  mutate(p_adj = p.adjust(p, method = "BH"),
         comparison = "Human vs. Mouse",
         timepoint_comparison = paste0(timepoint_human,  sep = "-", timepoint_mouse),
         pathogen_comparison = paste0(pathogen_human, "-", pathogen_mouse)) %>% 
  
  #Standardize factors
     mutate(
         timepoint_human = str_replace(timepoint_human,"Early hours", "Early\nhours"),
         timepoint_mouse = str_replace(timepoint_mouse,"Early hours", "Early\nhours"),
         timepoint_mouse = str_replace(timepoint_mouse,"Day ", "Day\n"),
         timepoint_human = fct_relevel(timepoint_human, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_mouse = fct_relevel(timepoint_mouse, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         pathogen_human = case_when(pathogen_human == "Hepatitis B" ~ "HepB",
                              pathogen_human == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen_human == "Escherichia coli" ~ "E. coli",
                              T ~pathogen_human),
          pathogen_human = fct_relevel(pathogen_human, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         pathogen_mouse= case_when(pathogen_mouse == "Hepatitis B" ~ "HepB",
                              pathogen_mouse == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen_mouse == "Escherichia coli" ~ "E. coli",
                              T ~pathogen_mouse),
          pathogen_mouse = fct_relevel(pathogen_mouse, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment_human = case_when(pathogen_human %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen_human %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen_human %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen_human),
         treatment_human = fct_relevel(treatment_human, "Vaccination", "Infection", "Injury"),
         treatment_mouse = case_when(pathogen_mouse %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen_mouse %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen_mouse %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen_mouse),
         treatment_mouse = fct_relevel(treatment_mouse, "Vaccination", "Infection", "Injury")) 

#Summary
degs_immune_nonimmune_summary = human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>%
  filter(comparison_set == "Inter-species") %>%  
  group_by(pathogen_1, pathogen_2, timepoint_comparison, gene_set) %>% 
  summarise(n_genes = n(), .groups = "drop")





```








#### Matrix plot


##### Interspecies


```{r fig.height=10, fig.width=11}

# Plot Matrices
# All
correlation_heatmap_all = correlation_filtered_inter_weighted %>% 
  filter(pathogen_mouse != "DMD") %>% 
  ggplot() +
  aes(x = timepoint_mouse,
      y = timepoint_human) +
  #Facet
    facet_nested_wrap(~gene_set+treatment_mouse+pathogen_mouse,
                    scales = "free",
                    nrow = 4, 
                    axes = "all",
                    remove_labels = "none",
                    strip = strip_nested(size = "variable")) +
  labs(x = "Mouse",
       y = "Human") +
  geom_tile(aes(fill = rho)) +
  geom_text(aes(label = round(rho, 2)), size = 3, 
                  color = "black") +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(-1, 1, length.out = 10)),
        limits = c(-1, 1),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 1,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.5, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

correlation_heatmap_all 

correlation_heatmap_all %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_interspecies_matrixplot_weighted.png"),
         width = 11, height = 8)

```

```{r fig.height=6, fig.width=12}
#Except immune
correlation_heatmap_all_degs = correlation_filtered_inter_weighted %>% 
  filter(pathogen_mouse != "DMD") %>% 
  filter(gene_set != "Immune") %>% 
  ggplot() +
  aes(x = timepoint_mouse,
      y = timepoint_human) +
  #Facet
  facet_nested_wrap(~gene_set+treatment_mouse+pathogen_mouse,
                    scales = "free",
                    ncol = 6, 
                    axes = "all",
                    remove_labels = "none",
                    strip = strip_nested(size = "variable")) +
  labs(x = "Mouse",
       y = "Human") +
  geom_tile(aes(fill = rho)) +
  geom_text(aes(label = round(rho, 2)), size = 3, 
                  color = "black") +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(-1, 1, length.out = 10)),
        limits = c(-1, 1),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 1,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.5, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

correlation_heatmap_all_degs 

correlation_heatmap_all_degs %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_interspecies_all_degs_matrixplot.png"),
         width = 12, height = 6)
```


```{r fig.height=3, fig.width=12}
#Only DEGs
correlation_heatmap_degs = correlation_filtered_inter_weighted %>% 
  filter(gene_set == "DEGs") %>% 
  ggplot() +
  aes(x = timepoint_mouse,
      y = timepoint_human) +
  #Facet
  facet_nested_wrap(~gene_set+treatment+pathogen,
                    scales = "free",
                    nrow = 1, 
                    axes = "all",
                    remove_labels = "none",
                    strip = strip_nested(size = "variable")) +
  labs(x = "Mouse",
       y = "Human") +
  geom_tile(aes(fill = rho),
            color = "white",
            size = 0.5) +
  geom_text(aes(label = round(rho, 2)), size = 3, 
                  color = "black") +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(-1, 1, length.out = 10)),
        limits = c(-1, 1),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 0.5,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.5, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

correlation_heatmap_degs 

correlation_heatmap_degs %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_interspecies_matrixplot_degs_weighted.png"),
         width = 12, height = 3)



```

#####Intraspecies

All
```{r fig.height=12, fig.width=18}
# Intra-species ==========
correlation_filtered_intra_weighted <- human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>%
  filter(comparison_set == "Intra-species") %>% 
  group_by(
    comparison_set, organism_1, treatment, pathogen,
    timepoint_1, timepoint_2, gene_set
  ) %>%
  summarise({
    # Capture the data for the current group
    d <- cur_data()
    
    # Clean: filter out NAs and invalid weights
    valid_data <- d %>% 
      filter(!is.na(mean_log2fc_1), 
             !is.na(mean_log2fc_2), 
             weight_se_inverse > 0)
    
    # Check if we have enough observations (e.g., at least 3)
    if (nrow(valid_data) >= 3) {
      res <- wtd.cor(
        x = rank(valid_data$mean_log2fc_1), 
        y = rank(valid_data$mean_log2fc_2), 
        weight = valid_data$weight_se_inverse
      )
      
      data.frame(
        rho = res[1, "correlation"],
        se  = res[1, "std.err"],
        t   = res[1, "t.value"],
        p   = res[1, "p.value"]
      )
    } else {
      # Return NAs if the group is too small or invalid
      data.frame(rho = NA, se = NA, t = NA, p = NA)
    }
  }, .groups = 'drop') %>% 
  mutate(comparison = paste0(organism_1, " vs. ", organism_1),
         timepoint_comparison = paste0(timepoint_1, "-", timepoint_2)) %>%

  #Standardize factors
  mutate(timepoint_label = if_else(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"),
                                   timepoint_2, timepoint_1),
         timepoint_label = as.character(timepoint_label),
         timepoint_label = str_replace(timepoint_label,"Early hours", "Early\nhours"),
         timepoint_1 = str_replace(timepoint_1,"Early hours", "Early\nhours"),
         timepoint_2 = str_replace(timepoint_2,"Early hours", "Early\nhours"),
         timepoint_2 = str_replace(timepoint_2,"Day ", "Day\n"),
         timepoint = fct_relevel(timepoint_label, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_1 = fct_relevel(timepoint_1, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_2 = fct_relevel(timepoint_2, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         pathogen = as.character(pathogen),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~pathogen),
          pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"))
```


```{r fig.height=12, fig.width=18}
#Matrix plot for intraspecies correlation =======
#Human vs Human
correlation_heatmap_intra_human_all = correlation_filtered_intra_weighted %>% 
  # filter(!(pathogen %in% c("S. aureus", "E. coli") & organism_1 == "Human")) %>%
  filter(organism_1 == "Human",
         # gene_set != "Immune"
         ) %>% 
  ggplot() +
  aes(x = timepoint_2,
      y = timepoint_1) +
  labs(x = "Timepoint",
       y = "Timepoint") +
  #Facet
  facet_nested_wrap(~organism_1+gene_set+treatment+pathogen,
                    scales = "free", trim_blank = F, ncol = 6,
                    strip = strip_nested(size = "variable")) +
  geom_tile(aes(fill = rho),
            color = "white",
            size = 0.5) +
    geom_text(aes(label = round(rho, 2), 
              color = ifelse(abs(rho) > 0.5, "white", "black")), 
          size = 4, 
          fontface = "bold") +
  scale_color_identity() +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        na.value = "white",
        values = scales::rescale(seq(-0.999999, 0.999999, length.out = 10)),
        limits = c(-0.999999, 0.99999),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 0.5,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.2, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 
#Print
correlation_heatmap_intra_human_all

#Save
correlation_heatmap_intra_human_all %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_intraspecies_matrixplot_human_weighted.png"),
         width = 18, height = 10)


#Mouse vs. Mouse
correlation_heatmap_intra_mouse_all = correlation_filtered_intra_weighted %>% 
  filter(!(pathogen %in% c("S. aureus", "E. coli") & organism_1 == "Human")) %>% 
  filter(organism_1 == "Mouse",
          # gene_set != "Immune"
         ) %>% 
  # View()
  ggplot() +
  aes(x = timepoint_2,
      y = timepoint_1) +
  labs(x = "Timepoint",
       y = "Timepoint") +
  #Facet
  facet_nested_wrap(~organism_1+gene_set+treatment+pathogen,
                    scales = "free",
                    trim_blank = F,
                    ncol = 6,
                    strip = strip_nested(size = "variable")) +
  geom_tile(aes(fill = rho),
            color = "white",
            size = 0.5) +
  geom_text(aes(label = round(rho, 2), 
              color = ifelse(abs(rho) > 0.5, "white", "black")), 
          size = 4, 
          fontface = "bold") +
  scale_color_identity() +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        na.value = "white",
        values = scales::rescale(seq(-0.999999, 0.999999, length.out = 10)),
        limits = c(-0.999999, 0.99999),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 0.5,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.2, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 
#Print
correlation_heatmap_intra_mouse_all

#Save
correlation_heatmap_intra_mouse_all %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_intraspecies_matrixplot_mouse_weighted.png"),
         width = 18, height = 10)


correlation_filtered_intra_weighted %>% 
  filter(!(pathogen %in% c("S. aureus", "E. coli") & organism_1 == "Human")) %>% 
  filter(organism_1 == "Mouse",
          gene_set != "Immune") 
```


```{r fig.height=10, fig.width=18}
######### Wrap plots
design <- 
"
AAAABBBB
AAAABBBB
AAAA    
"
correlation_heatmap_intra_human_mouse_all = wrap_plots(A = correlation_heatmap_intra_mouse_all, 
           B = correlation_heatmap_intra_human_all, 
           design = design) +
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "top"))
  
correlation_heatmap_intra_human_mouse_all

#Save
correlation_heatmap_intra_human_mouse_all %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_intraspecies_matrixplot_all_weighted.png"),
         width = 18, height = 10)


```



DEGs only
```{r fig.height=6, fig.width=12}
#Matrix plot for intraspecies correlation =======
#Human vs Human
correlation_heatmap_intra_human = correlation_filtered_intra %>% 
  filter(!(pathogen %in% c("S. aureus", "E. coli") & organism1 == "Human")) %>% 
  filter(organism1 == "Human") %>% 
  filter(gene_set == "DEGs") %>% 
  # View()
  ggplot() +
  aes(x = timepoint2,
      y = timepoint1) +
  labs(x = "Timepoint",
       y = "Timepoint") +
  #Facet
  facet_nested_wrap(~organism1+treatment+pathogen,
                    scales = "free",
                    nrow = 1, trim_blank = F,
                    strip = strip_nested(size = "variable")) +
  geom_tile(aes(fill = rho),
            color = "white",
            size = 0.5) +
    geom_text(aes(label = round(rho, 2), 
              color = ifelse(abs(rho) > 0.2, "white", "black")), 
          size = 3, 
          fontface = "bold") +
  scale_color_identity() +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(-1, 1, length.out = 10)),
        limits = c(-1, 1),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 0.5,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.2, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

# correlation_heatmap_intra_human

#Mouse vs. Mouse
correlation_heatmap_intra_mouse = correlation_filtered_intra %>% 
  filter(!(pathogen %in% c("S. aureus", "E. coli") & organism1 == "Human")) %>% 
  filter(organism1 == "Mouse") %>% 
  filter(gene_set == "DEGs") %>% 
  # View()
  ggplot() +
  aes(x = timepoint2,
      y = timepoint1) +
  labs(x = "Timepoint",
       y = "Timepoint") +
  #Facet
  facet_nested_wrap(~organism1+treatment+pathogen,
                    scales = "free",
                    nrow = 1, trim_blank = F,
                    strip = strip_nested(size = "variable")) +
  geom_tile(aes(fill = rho),
            color = "white",
            size = 0.5) +
    geom_text(aes(label = round(rho, 2), 
              color = ifelse(abs(rho) > 0.2, "white", "black")), 
          size = 3, 
          fontface = "bold") +
  scale_color_identity() +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(-1, 1, length.out = 10)),
        limits = c(-1, 1),
        name = "Spearman's rho",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 0.5,
          nbin = 10,
          raster = FALSE
        )
      ) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        panel.spacing.y = unit(0, "lines"),
        panel.spacing.x = unit(0.2, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

# correlation_heatmap_intra_mouse

######### Wrap plots
design <- "
  AAAAAA
  BBBB##
"
correlation_heatmap_intra_human_mouse = wrap_plots(A = correlation_heatmap_intra_mouse, 
           B = correlation_heatmap_intra_human, 
           design = design) +
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "top"))
  
correlation_heatmap_intra_human_mouse

#Save
correlation_heatmap_intra_human_mouse %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_correlation_intraspecies_matrixplot_degs.png"),
         width = 12, height = 6)


```


#### Scatter plots



```{r fig.height=16, fig.width=16}
#Plot weighted correlation scatterplot
human_mouse_log2fc_avg_wide %>%
  mutate(diff_log2FC_cat = case_when(
    mean_log2fc_diff < -2 ~ "<-2",
    mean_log2fc_diff > 2  ~ ">2",
    TRUE        ~ as.character(round(mean_log2fc_diff, 0))),
  diff_log2FC_cat = fct_relevel(diff_log2FC_cat, "<-2", "-2", "-1", "0", "1", "2", ">2")) %>% 
  ggplot() +
  aes(x = mean_log2fc_human, y = mean_log2fc_mouse) +
  facet_wrap(~treatment+pathogen+timepoint_comparison, scales = "free") +
  geom_point(aes(size = weight_se_inverse,
                 fill = diff_log2FC_cat),
             alpha = 0.5,
             stroke = 0,
             pch = 21) +
  # Weighted Regression Line
  geom_smooth(aes(weight = weight_se_inverse), 
              method = "lm", color = "red", linetype = "dashed", se = T) +
  
  # Unweighted Regression Line
  geom_smooth(method = "lm", color = "blue", linetype = "solid", se = T) +
  
  scale_fill_manual(
    values = c("<-2" = "darkred", "-2" = "red", "-1" = "orange", 
               "0" = "gray", "1" = "orange", "2" = "red", ">2" = "darkred")
  ) +
  labs(
    x = "Human Median Log2FC",
    y = "Mouse Median Log2FC",
    title = "Comparison: Weighted vs Unweighted Regression",
    subtitle = "Red Dashed: Weighted | Blue Solid: Unweighted"
  ) +
  theme_vaxgo()+
  theme(legend.position = "top")
```

##### Log2FC vs Log2fc

```{r}
#Correlation ------

#Log2fc vs Log2fc
cor_data <- human_mouse_log2fc_avg_wide_degs_immune_nonimmune %>%
  group_by(pathogen, timepoint_comparison, gene_set) %>%
  summarise(
    cor_test_spearman = list(safe_cor_test(mean_log2fc_mouse, mean_log2fc_human, method = "spearman")),
    cor_test_pearson = list(safe_cor_test(mean_log2fc_mouse, mean_log2fc_human, method = "pearson")),,
    .groups = "drop"
  ) %>%
  inner_join(degs_immune_nonimmune_summary, by = join_by("pathogen", "timepoint_comparison", "gene_set")) %>% 
  mutate(
    #Spearman
    cor_value_spearman = map_dbl(cor_test_spearman, ~ if (is.null(.x)) NA_real_ else .x$estimate),
    p_spearman   = map_dbl(cor_test_spearman, ~ if (is.null(.x)) NA_real_ else .x$p.value),
    p_label_spearman = case_when(
      is.na(p_spearman)      ~ "ns",
      p_spearman <= 0.01     ~ "***",
      p_spearman <= 0.05     ~ "**",
      p_spearman <= 0.10     ~ "*",
      TRUE                ~ "ns"
    ),
    label_spearman = paste0("rho=", round(cor_value_spearman, 2), ", n=", n_genes, ", ", p_label_spearman),
    #Pearson
    cor_value_pearson = map_dbl(cor_test_pearson, ~ if (is.null(.x)) NA_real_ else .x$estimate),
    p_pearson   = map_dbl(cor_test_pearson, ~ if (is.null(.x)) NA_real_ else .x$p.value),
    p_label_pearson = case_when(
      is.na(p_pearson)      ~ "ns",
      p_pearson <= 0.01     ~ "***",
      p_pearson <= 0.05     ~ "**",
      p_pearson <= 0.10     ~ "*",
      TRUE                ~ "ns"
    ),
    label_pearson = paste0("r=", round(cor_value_pearson, 2), ", n=", n_genes, ", ", p_label_pearson)
  ) %>% 
  drop_na(cor_value_spearman,
          cor_value_pearson) %>% 
  mutate(
    gene_set = factor(gene_set, levels = c("All genes", "DEGs", "Not-immune", "Immune")),
    facet_label = paste0(gene_set, ": ", label_spearman)) %>% 
    #Reorder timepoints
  separate(timepoint_comparison, into = c("timepoint_mouse", "timepoint_human"), remove = F) %>% 
  mutate(timepoint_mouse = fct_relevel(timepoint_mouse, c("2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7"))) %>% 
  arrange(timepoint_mouse) %>%
  mutate(timepoint_comparison = fct_inorder(timepoint_comparison))

cor_data_facetted = cor_data %>% 
  arrange(gene_set) %>% 
   group_by(pathogen, timepoint_comparison) %>%
  summarise(
    facet_label = paste(facet_label, collapse = "\n"),
    .groups = "drop"
  )  %>% 
  mutate(facet_label = paste0(timepoint_comparison, "\n", facet_label)) %>% 
  arrange(timepoint_comparison) %>% 
  mutate(facet_label = fct_inorder(facet_label))

cor_data
cor_data_facetted


human_mouse_log2fc_avg_wide_labels = human_mouse_log2fc_avg_wide %>% 
  inner_join(cor_data_facetted, 
             by = join_by("pathogen","timepoint_comparison")) %>% 
  mutate(pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma")) %>% 
  distinct() 

#Unite gene data and correlation/facets
human_mouse_log2fc_avg_wide_labels_filters = 
  bind_rows(
    #All genes
    human_mouse_log2fc_avg_wide_labels %>% 
  mutate(color_group = "All genes"),

    #All DEGs
   human_mouse_log2fc_avg_wide_labels %>% 
  filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>% 
  mutate(color_group = "All DEGs"),

   #Shared DEGs
   human_mouse_log2fc_avg_wide_labels %>% 
  filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05) %>% 
  mutate(color_group = "Shared DEGs"),

  #Immune DEGs
    human_mouse_log2fc_avg_wide_labels %>% 
  filter(human_symbol %in% c(btm_annotation_genes %>% 
                               filter(module_category == "immune") %>% 
                               pull(symbol)),
         p_adj_human <= 0.05 & p_adj_mouse <= 0.05) %>% 
  mutate(color_group = "Immune")
  ) %>% 
  mutate(color_group = fct_relevel(color_group, "All genes", "All DEGs", "Shared DEGs", "Immune"))



```


```{r fig.height=12, fig.width=16}
# Log2FC vs Log2fc
human_mouse_log2fc_log2fc_avg_wide_scatterplot = human_mouse_log2fc_avg_wide_labels_filters %>% 
  
  #Plot
  ggplot() +
  aes(x = mean_log2fc_mouse,
      y = mean_log2fc_human) +
  #Facet
  facet_nested_wrap(vars(pathogen, facet_label),
                    scales = "free",
                    nrow = 4,
                    strip = strip_nested(size = "variable")) +
  
  geom_point(aes(fill = color_group),
             alpha = 1,
             color = "gray95",
             pch = 21,
             size = 1.5,
             data = . %>% filter(color_group == "All genes")) +
  
  geom_point(aes(fill = color_group),
             alpha = 1,
             color = "gray85",
             pch = 21,
             size = 1.5,
             data = . %>% filter(color_group == "All DEGs")) +
  
  geom_point(aes(fill = color_group,
                 color = color_group),
             alpha = 1,
             pch = 21,
             size = 1.5,
             data = . %>% filter(!color_group %in% c("All genes", "All DEGs"))) +
  
  #Regression line: Linear model, Pearson
  stat_smooth(method = "lm", geom = "line", alpha = 1, se = F, size = 1,
              mapping = aes(group = color_group,
                        color = color_group)) +
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  scale_fill_manual(values = c("All genes" = "gray95",
                               "All DEGs" = "gray85",
                               "Shared DEGs" = "#175289",
                               "Immune" ="#46eec6")) +
  scale_color_manual(values = c("All genes" = "black",
                               "All DEGs" = "#4361ee",
                               "Shared DEGs" = "#175289",
                               "Immune" ="#46eec6")) +
  #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "top",
        legend.direction = "horizontal",
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       fill = "Gene type",
       color = "Linear regression",
       title = "Correlation plot,\nHuman vs Mouse")


human_mouse_log2fc_log2fc_avg_wide_scatterplot 

#Save
# human_mouse_log2fc_log2fc_avg_wide_scatterplot %>%
#   ggsave(filename = here("Figures", "all_human_mouse_dge_degs_scatterplot.png"),
#          width = 16, height = 12)

```


##### Ranks vs Ranks

```{r fig.height=12, fig.width=16}
# 
# #Create ranks
# human_mouse_log2fc_avg_wide_labels_ranked <-
#   human_mouse_log2fc_avg_wide_labels %>%
#   group_by(pathogen, facet_label) %>%
#   mutate(
#     rank_log2fc_mouse = dense_rank(mean_log2fc_mouse),
#     rank_log2fc_human = dense_rank(mean_log2fc_human)
#   ) %>%
#   ungroup()
# 
# 
# human_mouse_rank_rank_avg_wide_scatterplot = human_mouse_log2fc_avg_wide_labels_ranked %>% 
#   
#   #Plot
#   ggplot() +
#   aes(x = rank_log2fc_mouse,
#       y = rank_log2fc_human) +
#   #Facet
#   facet_nested_wrap(vars(pathogen, facet_label),
#                     scales = "free",
#                     nrow = 4,
#                     strip = strip_nested(size = "variable")) +
#   
#   geom_point(aes(color = color_group,
#                  fill = color_group),
#              alpha = 1,
#              size = 1.5,
#              data = human_mouse_log2fc_avg_wide_labels_ranked %>% 
#                mutate(color_group = "All genes")) +
#   
#   #All DEGs
#   geom_point(aes(color = color_group,
#                  fill = color_group), 
#              alpha = 1,
#              size = 1.5,
#              data = human_mouse_log2fc_avg_wide_labels_ranked %>% 
#                filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>% 
#                mutate(color_group = "All DEGs")) +
#   
#   #DEGs in both organisms
#   geom_point(aes(fill = color_group,
#                  color = color_group),
#              alpha = 1,
#              size = 1.5,
#              pch = 21,
#              data = human_mouse_log2fc_avg_wide_labels_ranked %>%
#                #Filter DEGs in both organisms, same sign
#                filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05) %>% 
#              mutate(color_group = "Shared DEGs")) +
# 
#   #Immune DEGs
#   geom_point(aes(fill = color_group,
#                  color = color_group),
#              alpha = 1,
#              size = 1.5,
#              pch = 21,
#              data = human_mouse_log2fc_avg_wide_labels_ranked %>%
#                filter(human_symbol %in% c(btm_annotation_genes %>%                                                
#                                             filter(module_category == "immune") %>%
#                                             pull(symbol)),
#                       p_adj_human <= 0.05 & p_adj_mouse <= 0.05) %>% 
#                mutate(color_group = "Immune")) +
#   guides(color = "none",
#          fill = guide_legend(override.aes = list(size = 4))) +
#   scale_fill_manual(values = colors$shared) +
#   scale_color_manual(values = colors$shared) +
#   #Add lines
#   geom_abline(alpha = 0.2, linetype = 2) +
#   #Theme
#   theme_vaxgo() +
#   theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
#         axis.text.y = element_text(size = 10),
#         strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
#         legend.position = "top",
#         legend.direction = "horizontal",
#         ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) +
#   labs(x = "Mouse L2FC",
#        y = "Human L2FC",
#        fill = "Gene type",
#        title = "Correlation plot,\nHuman vs Mouse")
# 
# 
# human_mouse_rank_rank_avg_wide_scatterplot 
# human_mouse_rank_rank_avg_wide_scatterplot %>%
#   ggsave(filename = here("Figures", "all_human_mouse_dge_degs_scatterplot_ranks.png"),
#          width = 16, height = 12)

```




#### Other exploration analysis

##### Compare Diff Log2FC distributions
```{r fig.height=4, fig.width=6}
#Compare distributions =============

#Plot
log2fc_diff_density_bionull_plot =   human_mouse_log2fc_avg_wide_bionull %>% 
  filter(p_adj_human < 0.05) %>% 
    ggplot() +
  # Eixo X agora usa nossa coluna unificada 'divergence'
  aes(x = abs(divergence)) +
  
  # 2. Camada de frente: Suas linhas reais e longitudinais mapeadas pelo Mako
  geom_density(
    data = filter(human_mouse_log2fc_avg_wide_bionull, model_type == "Observed Target"),
    aes(color = timepoint_comparison,
        # linetype = model_type
        ),
    alpha = 1, size = 1
  ) +
  
  # 1. Camada de fundo: A distribuição nula simulada (Sombra cinza estável)
  geom_density(
    data = filter(human_mouse_log2fc_avg_wide_bionull, model_type == "Biological Null"),
    aes(y = ..density..),
    color = "red",size = 0.5, linetype = 1,  alpha = 1
  ) +
  
  geom_vline(xintercept = 0, color = "red", linetype = 2) +
  
  # Facetamento controlado por patógeno
  facet_wrap(~pathogen, scales = "free") +
  
  # Manter sua paleta profissional
  scale_color_viridis_d(option = "mako", direction = -1) +
    
  # scale_x_continuous(limits = c(-2, 2)) +
  
  labs(
    x = "Relative Expression Divergence (Log2FC Human - Log2FC Mouse)", # Atualizado de 'Absolute' para 'Relative'
    y = "Density",
    title = "Empirical vs. Biological Null Distribution of Cross-Species Divergence",
    subtitle = "Red curve sets the independent baseline drift utilizing an unmatched vaccine control (Hepatitis B Day 1)",
    color = "Infection Timeline"
  ) +
  theme_vaxgo() +
  theme(legend.position = "top")
  
log2fc_diff_density_bionull_plot

```


```{r fig.height=8, fig.width=12}
#Distribution of Cross-Species Divergence Magnitude
ggplot(human_mouse_log2fc_avg_wide) +
  aes(x = cv_mean,
      y = mean_log2fc_diff, 
      color = weight_se_inverse) +
  geom_point(alpha = 1, size = 1) +
  facet_wrap(~pathogen+timepoint_mouse, scales = "free") +
  scale_color_viridis_c(option = "mako", direction = -1) +
  # scale_fill_manual(values = colors$timepoint) +
  theme_vaxgo()



ggplot(human_mouse_log2fc_avg_wide) +
  aes(x = abs(mean_log2fc_diff),
      y = se_mouse, 
      color = weight_se_inverse) +
  geom_point(alpha = 1, size = 1) +
  facet_wrap(~pathogen+timepoint_mouse, scales = "free") +
  scale_color_viridis_c(option = "mako", direction = -1) +
  # scale_fill_manual(values = colors$timepoint) +
  theme_vaxgo()

ggplot(human_mouse_log2fc_avg_wide) +
  aes(x = mean_log2fc_human,
      y = se_human, 
      color = weight_se_inverse) +
  geom_point(alpha = 1, size = 1) +
  facet_wrap(~pathogen+timepoint_mouse, scales = "free") +
  scale_color_viridis_c(option = "mako", direction = -1) +
  # scale_fill_manual(values = colors$timepoint) +
  theme_vaxgo()



human_mouse_se = bind_rows(
human_mouse_log2fc_avg_wide %>% 
  group_by(pathogen, timepoint_mouse) %>% 
  summarise(median_se = median(se_human, na.rm = T),
            median_cv = median(cv_human, na.rm = T),
            median_diff = median(mean_log2fc_diff, na.rm = T)) %>% 
  mutate(organism = "Human") %>% 
  ungroup(),

human_mouse_log2fc_avg_wide %>% 
  group_by(pathogen, timepoint_mouse) %>% 
  summarise(median_se = median(se_mouse, na.rm = T),
            median_cv = median(cv_mouse, na.rm = T),
            median_diff = median(mean_log2fc_diff, na.rm = T)) %>% 
  mutate(organism = "Mouse") %>% 
  ungroup(),

human_mouse_log2fc_avg_wide %>% 
  group_by(pathogen, timepoint_mouse) %>% 
  summarise(median_se = median(se_combined, na.rm = T),
            median_cv = median(cv_mean, na.rm = T),
            median_diff = median(mean_log2fc_diff, na.rm = T)) %>% 
  mutate(organism = "Mouse-Human") %>% 
  ungroup())


human_mouse_se %>% 
  ggplot() +
  aes(x = timepoint_mouse,
      y = median_se,
      color = organism,
      group = organism) +
  geom_line() +
  geom_point() +
  geom_line(aes(x = timepoint_mouse,
      y = abs(median_diff),
      group = organism),
      color = "black") +
  geom_point(aes(x = timepoint_mouse,
      y = abs(median_diff),
      group = organism),
      color = "black") +
  facet_wrap(~pathogen, scales = "free_x")

human_mouse_se %>% 
  ggplot() +
  aes(x = timepoint_mouse,
      y = abs(median_diff),
      color = organism,
      group = organism) +
  geom_line() +
  geom_point() +
  facet_wrap(~pathogen, scales = "free")
```


```{r fig.height=3, fig.width=9}
human_mouse_se %>% 
  ggplot() +
  aes(y = abs(median_diff),
      x = median_se,
      color = timepoint_mouse,
      shape = pathogen) +
  geom_point() +
  facet_wrap(~organism, scales = "free") +
  theme_vaxgo()+
  theme(legend.position = "top")

```













### 2.3. Correlation: RRHO, means

```{r}
# 1. Identify all unique pathogen and timepoint combinations
combinations <- human_mouse_log2fc_avg_wide %>%
  distinct(pathogen, timepoint_comparison)

# 2. Initialize a list to store results from each iteration
all_rrho_results <- list()

for(i in 1:nrow(combinations)) {
  
  curr_pathogen <- combinations$pathogen[i]
  curr_timepoint <- combinations$timepoint_comparison[i]
  
  # A. Filter and Prepare Differential Expression (DDE) Scores
  rrho_input_data <- human_mouse_log2fc_avg_wide %>% 
    filter(pathogen == curr_pathogen, timepoint_comparison == curr_timepoint) %>%
    filter(!is.na(p_adj_human), !is.na(p_adj_mouse), 
           !is.na(mean_log2fc_human), !is.na(mean_log2fc_mouse)) %>%
    mutate(
      # Handle p-values of 0 to avoid infinite logs
      p_adj_human = ifelse(p_adj_human == 0, .Machine$double.xmin, p_adj_human),
      p_adj_mouse = ifelse(p_adj_mouse == 0, .Machine$double.xmin, p_adj_mouse),
      # Calculate DDE score: -log10(p-value) * sign of Fold Change
      score_human = -log10(p_adj_human) * sign(mean_log2fc_human),
      score_mouse = -log10(p_adj_mouse) * sign(mean_log2fc_mouse)
    )

  # Skip if the gene list is too small (RRHO2 requires a minimum set)
  if(nrow(rrho_input_data) < 100) next 

  # B. Prepare Sorted Gene Lists for RRHO2
  # Sorting from highest to lowest DDE score is mandatory for RRHO analysis
  gene_list1 <- rrho_input_data %>% 
    arrange(desc(score_human)) %>%
    select(Genes = human_symbol, DDE = score_human) %>% 
    as.data.frame()
    
  gene_list2 <- rrho_input_data %>% 
    arrange(desc(score_mouse)) %>%
    select(Genes = human_symbol, DDE = score_mouse) %>% 
    as.data.frame()

  # Identify the Biological Split Point (where DDE score crosses zero)
  # This marks the boundary between Up-regulated and Down-regulated genes
  split_h_val <- sum(gene_list1$DDE > 0)
  split_m_val <- sum(gene_list2$DDE > 0)

  # C. Execute RRHO2
  # Using 'try' to skip iterations that might encounter mathematical errors
  rrho_obj <- try(RRHO2_initialize(gene_list2, gene_list1, 
                                   labels = c("Mouse", "Human"), 
                                   boundary = 0.01,
                                   log10.ind = TRUE, 
                                   method = "hyper"), silent = TRUE)
  
  if(inherits(rrho_obj, "try-error")) next

  # Transform the hypermat (matrix) into a long format dataframe for ggplot2
  temp_df <- as.data.frame(rrho_obj$hypermat) %>%
    mutate(Mouse_Rank = row_number()) %>%
    pivot_longer(cols = -Mouse_Rank, names_to = "Human_Rank", values_to = "LogP") %>%
    mutate(
      Human_Rank = as.numeric(gsub("V", "", Human_Rank)),
      pathogen = curr_pathogen,       
      timepoint = curr_timepoint,
      # Store the split points to draw exact quadrant boundaries later
      split_h = split_h_val,
      split_m = split_m_val
    )

  all_rrho_results[[i]] <- temp_df
}

# Combine all results into a single final dataframe
final_rrho_df <- bind_rows(all_rrho_results)

# Save the processed data for visualization
final_rrho_df %>% 
  saveRDS(file = here("tables", "all_human_mouse_rrho_df.rds"))
 

```

Extract RRHO2 results and gene lists by quadrant
```{r}
library(dplyr)
library(tidyr)
library(RRHO2) # Ensure the library is loaded

# 1. Identify all unique pathogen and timepoint combinations
combinations <- human_mouse_log2fc_avg_wide %>%
  distinct(pathogen, timepoint_comparison)

# 2. Initialize lists to store results
plot_data_list <- list()
gene_extraction_list <- list() # New list for the extracted genes
raw_objects_list <- list()

# --- MAIN LOOP ---
for(i in 1:nrow(combinations)) {
  
  curr_pathogen <- combinations$pathogen[i]
  curr_timepoint <- combinations$timepoint_comparison[i]
  # Create a unique name for each comparison
  iter_name <- paste(curr_pathogen, curr_timepoint, sep = "_")
  
  # --- A. Prepare Data ---
  rrho_input_data <- human_mouse_log2fc_avg_wide %>% 
    filter(pathogen == curr_pathogen, timepoint_comparison == curr_timepoint) %>%
    filter(!is.na(p_adj_human), !is.na(p_adj_mouse), 
           !is.na(mean_log2fc_human), !is.na(mean_log2fc_mouse)) %>%
    mutate(
      p_adj_human = ifelse(p_adj_human == 0, .Machine$double.xmin, p_adj_human),
      p_adj_mouse = ifelse(p_adj_mouse == 0, .Machine$double.xmin, p_adj_mouse),
      score_human = -log10(p_adj_human) * sign(mean_log2fc_human),
      score_mouse = -log10(p_adj_mouse) * sign(mean_log2fc_mouse)
    )

  if(nrow(rrho_input_data) < 100) next 

  gene_list_human <- rrho_input_data %>% 
    arrange(desc(score_human)) %>%
    select(Genes = human_symbol, DDE = score_human) %>% as.data.frame()
    
  gene_list_mouse <- rrho_input_data %>% 
    arrange(desc(score_mouse)) %>%
    select(Genes = human_symbol, DDE = score_mouse) %>% as.data.frame()

  split_human <- sum(gene_list_human$DDE > 0)
  split_mouse <- sum(gene_list_mouse$DDE > 0)

  # --- B. Execute RRHO2 ---
  rrho_obj <- try(RRHO2_initialize(gene_list_mouse, gene_list_human, 
                                   labels = c("Mouse", "Human"), 
                                   boundary = 0.01, log10.ind = TRUE, method = "hyper"), 
                  silent = TRUE)
  
  if(inherits(rrho_obj, "try-error")) next

  # Store the raw object
  raw_objects_list[[iter_name]] <- rrho_obj

  # --- C. Save Plot Data ---
  temp_df <- as.data.frame(rrho_obj$hypermat) %>%
    mutate(Mouse_Rank = row_number()) %>%
    pivot_longer(cols = -Mouse_Rank, names_to = "Human_Rank", values_to = "LogP") %>%
    mutate(
      Human_Rank = as.numeric(gsub("V", "", Human_Rank)),
      pathogen = curr_pathogen,        
      timepoint = curr_timepoint,
      split_h = split_human,
      split_m = split_mouse
    )
  plot_data_list[[iter_name]] <- temp_df
}

# --- 3. CONSOLIDATE INTO final_results ---
final_results <- list(
  plot_data = bind_rows(plot_data_list),
  objects = raw_objects_list
)


#Extract gene lists
# 1. Define the quadrants we want to extract
quadrants <- c("uu", "dd", "ud", "du")

# 2. Use map_dfr to iterate through each comparison and quadrant
all_extracted_genes <- imap_dfr(final_results$objects, function(obj, condition_name) {
  
  # For each object (comparison), iterate through the 4 quadrants
  map_dfr(quadrants, function(q) {
    
    # Construct the internal names used by RRHO2 (e.g., genelist_uu and gene_list1_uu)
    genelist_slot <- paste0("genelist_", q)
    gene_item_slot <- paste0("gene_list1_", q)
    
    # Extract the list of genes
    genes <- obj[[genelist_slot]][[gene_item_slot]]
    
    # Only create a dataframe if there are genes in that quadrant
    if (!is.null(genes) && length(genes) > 0) {
      data.frame(
        genes = as.character(genes),
        condition = condition_name,
        comparison = genelist_slot,
        stringsAsFactors = FALSE
      )
    } else {
      NULL
    }
  })
}) %>% 
  separate(condition, into = c("pathogen", "timepoint_comparison"), sep = "_") %>% 
  separate(comparison, into = c("del", "direction"), sep = "_") %>%
  select(-del) %>% 
  mutate(direction = case_when(direction == "uu" ~ "Up-Up",
                               direction == "ud" ~ "Up-Down",
                               direction == "dd" ~ "Down-Down",
                               direction == "du" ~ "Down-Up",
                               T~direction))


final_results_final = list(
  plotdata = final_results$plot_data, 
  rrho2_obj = final_results$objects, 
     genes = all_extracted_genes)


final_results_final %>% saveRDS(file = here("tables", "all_human_mouse_rrho_df.rds"))

```


#### Plot RRHO Heatmap
```{r fig.height=8, fig.width=12}
library(tidyverse)
library(ggh4x)

final_rrho_list = readRDS(file = here("tables", "all_human_mouse_rrho_df.rds"))

final_rrho_df_plot = final_results_final$plotdata

final_rrho_df_plot <- final_rrho_df %>%
  mutate(pathogen = as.character(pathogen),
    pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma")) %>%
    #Reorder timepoints
  separate(timepoint, into = c("timepoint_mouse", "timepoint_human"), remove = F) %>% 
  mutate(timepoint_mouse = fct_relevel(timepoint_mouse, c("2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7"))) %>% 
  arrange(timepoint_mouse) %>%
  mutate(timepoint = fct_inorder(timepoint)) %>% 
  mutate(
    LogP_Final = ifelse(LogP > 50, 50, LogP)
  )
    

# --- Step 2: Generate the Visualization ---
rrho_plot <- ggplot() +
    geom_raster(data = final_rrho_df_plot, 
              aes(x = Human_Rank, y = Mouse_Rank, fill = LogP_Final),
              interpolate = T) +
  
  # C. Faceting: Using ggh4x for nested labels
  facet_nested_wrap(~pathogen + timepoint,
                    scales = "free", 
                    nrow = 4) +
  
  # D. Scale Adjustments
  scale_x_reverse(expand = c(0,0)) + 
  scale_y_reverse(expand = c(0,0)) +

 scale_fill_gradientn(
    colors = c("white", "#4361ee", "#46eec6"),
    values = c(0, 0.2, 1, 1.3), 
    limits = c(0, 50),
    name = "-log10(p-value)",
    breaks = c(0, 25, 50),
    labels = c("0", "25", "≥50"),
    na.value = "black"
  ) +
  # F. Labels and Styling
  labs(
    title = "Comparative RRHO2 Analysis",
    x = "Human Gene Rank (Up-to-Down)",
    y = "Mouse Gene Rank (Up-to-Down)"
  ) +
  theme_vaxgo() + 
  theme(aspect.ratio = 1,
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.key.width = unit(0.5, "cm"),
        legend.key.height = unit(0.2, "cm"), 
        legend.text = element_text(size = 10, 
                                   # hjust = 1, vjust = 1, 
                                   angle = 0),
        legend.title = element_text(size = 10),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1))

# Display and Save
print(rrho_plot)

ggsave(rrho_plot, 
       filename = here("Figures", "all_human_mouse_rrho_heatmap.png"),
       width = 12, height = 8)
```



#### Plot raw ranks
```{r fig.height=8, fig.width=12}
# Prepare data
rank_data <- human_mouse_log2fc_avg_wide %>%
  filter(!is.na(mean_log2fc_human) | !is.na(mean_log2fc_mouse),
         !is.na(p_adj_human), !is.na(p_adj_mouse)) %>%
  group_by(pathogen, timepoint_comparison) %>%
  mutate(
    score_human = -log10(p_adj_human) * sign(mean_log2fc_human),
    score_mouse = -log10(p_adj_mouse) * sign(mean_log2fc_mouse),
    Human_Rank_Val = rank(score_human),
    Mouse_Rank_Val = rank(score_mouse),
  ) %>%
  ungroup() %>% 
  mutate(pathogen = as.character(pathogen),
    pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma"))

# Calcular os limites por grupo
thresholds <- rank_data %>%
  group_by(pathogen, timepoint_comparison) %>%
  summarise(
    # Encontra o maior rank entre os genes com log2fc negativo
    v_line = max(Human_Rank_Val[mean_log2fc_human < 0], na.rm = TRUE),
    h_line = max(Mouse_Rank_Val[mean_log2fc_mouse < 0], na.rm = TRUE),
    .groups = "drop"
  )

# 2. Plotar como Heatmap de Densidade (Raster)
rank_density_heatmap = rank_data %>% 
  ggplot() + 
  aes(x = Human_Rank_Val, y = Mouse_Rank_Val) +
  facet_nested_wrap(~pathogen + timepoint_comparison,
                    scales = "free", 
                    nrow = 4,
                    strip = strip_nested(size = "variable"), 
                    axes = "all") +
 stat_density_2d(aes(fill = after_stat(density / sum(density))), 
                  geom = "raster", 
                  contour = F, 
                  n = 50) +    
  scale_fill_gradientn(
  colors = c("white", "#4361ee", "#4DBBD5FF", "#46eec6"),
  na.value = "white", 
  labels = scales::label_percent(), 
  name = "Density (%)",
  guide = guide_colorbar(
    barwidth = 1, 
    barheight = 8, 
    title.position = "top", 
    title.hjust = 0.5,
    ticks.colour = "black",
    frame.colour = "black"
  )
) +
  scale_x_reverse(expand = c(0,0)) +
  scale_y_reverse(expand = c(0,0)) +
  # Divide by threshold
  geom_vline(data = thresholds, aes(xintercept = v_line), 
             linetype = "solid", color = "white", alpha = 1, linewidth = 2) +
  geom_hline(data = thresholds, aes(yintercept = h_line), 
             linetype = "solid", color = "white", alpha = 1, linewidth = 2) +
  labs(
    title = "Rank-Rank Density Heatmap",
    subtitle = "Normalized density per condition",
    x = "Human Gene Rank",
    y = "Mouse Gene Rank"
  ) +
  theme_vaxgo() + 
  theme(aspect.ratio = 1,
        axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "right",
        legend.direction = "vertical",
        legend.key.width = unit(1, "cm"),
        legend.text = element_text(size = 8, 
                                   # hjust = 1, vjust = 1, 
                                   angle = 0),
        legend.title = element_text(size = 8),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1))


rank_density_heatmap


rank_density_heatmap %>% 
  ggsave(filename = here("Figures", "all_human_mouse_allgenes_rankrank_heatmap.png"),
         width = 12,
         height = 8)
```







#### Gene list analysis


```{r fig.height=10, fig.width=10}
final_rrho_list = readRDS(file = here("tables", "all_human_mouse_rrho_df.rds"))

gene_list_df = final_rrho_list$genes %>% 
  dplyr::count(pathogen, timepoint_comparison, direction) %>% 
  separate(timepoint_comparison, into = c("timepoint_mouse", "timepoint_human"), remove = F, sep = "-") %>% 
  mutate(timepoint_mouse = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_human = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7"))) %>% 
  arrange(timepoint_mouse) %>%
  mutate(timepoint_comparison = fct_inorder(timepoint_comparison),
         direction = fct_relevel(direction, "Up-Up", "Up-Down", "Down-Down", "Down-Up")) %>% 
  #Calculate percentage
  inner_join(
    human_mouse_log2fc_avg_wide  %>% 
      select(pathogen, timepoint_comparison) %>%
      dplyr::count(pathogen, timepoint_comparison) %>%
      rename(n_total = n),
    by = join_by(pathogen, timepoint_comparison)
  ) %>% 
  mutate(pct = ((n/n_total)*100) %>% round(0)) %>% 
  mutate(pathogen = as.character(pathogen),
         pathogen = case_when(pathogen == "Staphylococcus aureus" ~ "S. aureus",
                             pathogen == "Escherichia coli" ~ "E. coli", 
                             T~pathogen),
    pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "S. aureus", "E. coli", "Burn", "Trauma"))


```

##### Line plot
```{r fig.height=3, fig.width=12}
gene_list_lineplot = gene_list_df %>% 
  ggplot() +
  aes(x = timepoint_mouse, y = pct) +
  labs(x = "Timepoint Mouse",
         y = "Shared genes (%)",
         fill = "Direction",
         color = "Direction",
        linetype = "Direction") +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(0, 100)
                     ) +
  facet_grid(~pathogen, space = "free_x", scales = "free_x", axes = "all", axis.labels = "margins") +
  geom_line(aes(color = direction, 
                group = direction,
                linetype = direction),
            size = 1) +
  geom_point(aes(fill = direction),
             alpha = 1,
             color = "black",
             stroke = 0.5,
             pch = 21,
             size = 3) +
  geom_text(aes(label = n), color = "black", size = 2, vjust = -0.8) +
  
  scale_color_manual(values = colors$direction) +
  scale_fill_manual(values = colors$direction) +
  scale_linetype_manual(values = c("Up-Up" = 1, 
                                "Up-Down" = 2, 
                                "Down-Down" = 1, 
                                "Down-Up" = 2)) +
  
    theme_vaxgo() + 
  theme(# axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.key.width = unit(1, "cm"),
        legend.key.height = unit(1, "cm"), 
        legend.text = element_text(size = 10, 
                                   # hjust = 1, vjust = 1, 
                                   angle = 0),
        legend.title = element_text(size = 10),
        legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1))  +

  guides(fill = "none",
         linetype = guide_legend(override.aes = list(linewidth = 1)),
         color = guide_legend(override.aes = list(size = 0)))

gene_list_lineplot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_rrho_lineplot.png"),
         width = 12,
         height = 3)

gene_list_lineplot

  
```




##### Heatmap
```{r fig.height=10, fig.width=10}
gene_list_df_heatmap = final_rrho_list$genes %>% 
  separate(timepoint_comparison, into = c("timepoint_mouse", "timepoint_human"), remove = F, sep = "-") %>% 
  mutate(timepoint_mouse = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_human = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7"))) %>% 
  arrange(timepoint_mouse) %>%
  mutate(timepoint_comparison = fct_inorder(timepoint_comparison),
         direction = fct_relevel(direction, "Up-Up", "Up-Down", "Down-Down", "Down-Up")) %>% 
  mutate(pathogen = case_when(pathogen == "Staphylococcus aureus" ~ "S. aureus",
                             pathogen == "Escherichia coli" ~ "E. coli", 
                             pathogen == "Hepatitis B" ~  "HepB",
                             T~pathogen),
    pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(timepoint_direction = paste0(direction, " (", timepoint_mouse,")")) %>% 
  arrange(timepoint_mouse, direction ) %>% 
  mutate(timepoint_direction = fct_inorder(timepoint_direction))


# Clustering
df_to_cluster <- gene_list_df_heatmap %>%
  mutate(val = case_when(
    direction == "Up-Up"     ~  2,
    direction == "Up-Down"   ~  1,
    direction == "Down-Up"   ~ -1,
    direction == "Down-Down" ~ -2,
    TRUE ~ 0
  )) %>%
  select(genes, timepoint_direction, pathogen, val) %>%
  mutate(cond = paste(pathogen, timepoint_direction, sep = "_")) %>%
  select(genes, cond, val) %>%
  distinct() %>%
  pivot_wider(names_from = cond, values_from = val, values_fill = 0) %>%
  column_to_rownames("genes")

gene_dist <- dist(as.matrix(df_to_cluster)) 
gene_hclust <- hclust(gene_dist, method = "ward.D2")

# 3. Extrair a ordem dos genes clusterizados
gene_order <- rownames(df_to_cluster)[gene_hclust$order]

# 4. Aplicar a ordem de volta ao dataframe original do ggplot
gene_list_df_heatmap_clustered <- gene_list_df_heatmap %>%
  mutate(genes = factor(genes, levels = gene_order))

```


```{r fig.height=7, fig.width=12}
#Heatmap
gene_list_df_heatmap_plot = gene_list_df_heatmap %>% 
  mutate(direction_short = as.character(direction),
         direction_short = case_when(direction_short == "Up-Up" ~ "UU",
                               direction_short == "Up-Down" ~ "UD",
                               direction_short == "Down-Down" ~ "DD",
                               direction_short == "Down-Up" ~ "DU",
                               T~direction_short),
         direction_short = fct_relevel(direction_short, "UU", "UD", "DD", "DU"),
        timepoint_direction = paste0(timepoint_mouse, " (", direction,")")) %>% 
  arrange(timepoint_mouse, direction ) %>% 
  mutate(timepoint_mouse = fct_rev(timepoint_mouse)) %>% 
  # filter(pathogen == "Influenza") %>% 
  ggplot() +
  aes(x = genes, 
      y = timepoint_mouse,
       label = genes,
      fill = direction) +
  labs(x = "Genes",
         y = "Timepoint",
         fill = "Genes (%)") +
  facet_grid(~pathogen~direction, 
             scales = "free", 
             axes = "all", 
             space = "free",
             axis.labels = "margins",
             switch = "y") +
  geom_tile() +
  scale_fill_manual(values = colors$direction) +
  theme_vaxgo() + 
  theme(axis.text.y = element_text(hjust = 1, size =8),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text.y = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
        panel.spacing = unit(0, "lines"),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.key.width = unit(1, "cm"),
        legend.text = element_text(size = 8, 
                                   # hjust = 1, vjust = 1, 
                                   angle = 0),
        legend.title = element_text(size = 8)) 


gene_list_df_heatmap_plot

gene_list_df_heatmap_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_rrho_heatmap_genes_bydirection.png"),
         width = 12,
         height = 7)

```



##### ORA: Hallmarks
```{r}

msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

#Discretize by quartile
ora_df_input = gene_list_df_heatmap %>% 
  mutate(condition = paste0(pathogen, "_", timepoint_comparison, "_", direction)) %>% 
  select(genes, condition)

term2gene_df_input = msigdbr_grouped %>% 
  select(process, genes = symbol)

#Run ORA Hallmarks
ora_all_diff_hallmark <- enricher_condition(
  df = ora_df_input,
  term2gene_df = term2gene_df_input) %>% 
  mutate(log_p_adjust = -log(p_adjust),
         process = str_replace_all(description, "_", " ")) %>% 
  separate(condition, into = c("pathogen", "timepoint_comparison", "direction"), sep = "_")  %>% 
  separate(timepoint_comparison, sep = "-", into = c("timepoint_mouse", "timepoint_human")) %>% 
  inner_join(msigdbr_grouped %>% 
               select(process, group) %>% 
               distinct(),
             by = join_by(id == "process")) %>% 
  mutate(timepoint_mouse = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_human = fct_relevel(timepoint_mouse, c("Early", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         timepoint_direction = paste0(timepoint_mouse, " (", direction, ")")) %>% 
  arrange(fct_rev(direction), timepoint_mouse) %>% 
  mutate(timepoint_direction = fct_inorder(timepoint_direction)) 


ora_all_diff_hallmark2$timepoint_direction %>% levels()

#Save
ora_all_diff_hallmark %>% 
  saveRDS(file = here("tables", "all_human_mouse_rrho_ora_hallmark.rds"))

#Print
ora_all_diff_hallmark
  
```


```{r}
# Clustering
df_to_cluster <- ora_all_diff_hallmark %>%
  mutate(val = case_when(
    direction == "Up-Up"     ~  2,
    direction == "Up-Down"   ~  1,
    direction == "Down-Up"   ~ -1,
    direction == "Down-Down" ~ -2,
    TRUE ~ 0
  )) %>%
  select(description, timepoint_direction, pathogen, val) %>%
  mutate(cond = paste(pathogen, timepoint_direction, sep = "_")) %>%
  select(description, cond, val) %>%
  distinct() %>%
  pivot_wider(names_from = cond, values_from = val, values_fill = 0) %>%
  column_to_rownames("description")

gene_dist <- dist(as.matrix(df_to_cluster)) 
gene_hclust <- hclust(gene_dist, method = "ward.D2")

# 3. Extrair a ordem dos description clusterizados
gene_order <- rownames(df_to_cluster)[gene_hclust$order]

# 4. Aplicar a ordem de volta ao dataframe original do ggplot
ora_all_diff_hallmark_clustered <- ora_all_diff_hallmark %>%
  mutate(description = factor(description, levels = gene_order))
```




```{r fig.height=8, fig.width=20}
ora_all_diff_hallmark_heatmap = ora_all_diff_hallmark %>% 
  mutate(direction = fct_relevel(direction, "Up-Up", "Down-Down", "Up-Down", "Down-Up")) %>% 
  filter(log_p_adjust > 1) %>% 
  mutate(log_p_adjust = if_else(log_p_adjust >10, 10, log_p_adjust)) %>% 
  ggplot() +
  aes(x = timepoint_mouse ,
      y = description) +
  # facet_grid(~direction+pathogen, space = "free_x", scales = "free_x") +
    facet_nested_wrap(~direction+pathogen, scales = "free_x", nrow = 1) +
  geom_tile(aes(fill = log_p_adjust)) +
  theme_vaxgo() + 
  labs(x = "",
       y = "") +
  theme(axis.text.y = element_text(hjust = 1, size =8),
        axis.text.x= element_text(size =8, angle = 90, hjust = 1, vjust = 0.5),
        # axis.ticks.x = element_blank(),
        strip.text.x = element_text(hjust = 0.5,size = 10, color = "white", face = "bold"),
        strip.text.y = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
        panel.spacing = unit(0.1, "lines"),
        strip.background.x = element_rect(fill = "black", color = "white"),
        panel.grid.major.x = element_line(linetype = 1, color = "gray90"), 
        panel.grid.major.y = element_line(linetype = 1, color = "gray90"), 
        plot.margin = margin(0, 0.3, 0, 0.3, unit = "cm"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.key.width = unit(1, "cm"),
        axis.line.y.right = element_line(linetype = 1, size = 0.5, color = "black"),
        legend.text = element_text(size = 10, 
                                   # hjust = 1, vjust = 1, 
                                   angle = 0),
        legend.title = element_text(size = 10),
        ggh4x.facet.nestline = element_line(colour = "white", linetype = 1))  +
   scale_fill_gradient(
    low = "#4361ee", 
    high = "#46eec6",
    limits = c(0, 10),
    breaks = c(0, 1, 1.3, 2, 10), 
    labels = c("0", "1", "1.3", "2", "≥10"), 
    na.value = "white"
  )

ora_all_diff_hallmark_heatmap

ora_all_diff_hallmark_heatmap %>%
   ggsave(filename = here("Figures", "all_human_mouse_rrho_heatmap_process_hallmarks.png"),
         width = 20,
         height = 8)

```















### 2.4. Correlation: Genes, samples

```{r}
#Import
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))
all_human_mouse_metadata = readRDS(here("tables", "all_human_mouse_metadata.rds"))


#Filter conditions with too many human samples to reduce computation time
set.seed(420)
all_human_mouse_metadata_filtered =
    all_human_mouse_metadata %>% 
   filter(!(pathogen %in% c("Trauma", "Influenza", "Burn") & organism == "Human")) %>% 
  
   bind_rows(
     #Sample trauma patients
     all_human_mouse_metadata %>%
      filter(pathogen == "Trauma", treatment == "Trauma", organism == "Human") %>% 
      group_by(timepoint) %>% 
      slice_sample(n = 30, replace = F),
     #Sample Burn patients
     all_human_mouse_metadata %>%
      filter(pathogen == "Burn", treatment == "Burn", organism == "Human") %>% 
      group_by(timepoint) %>% 
      slice_sample(n = 30, replace = F),
     #Sample Fluad patients
     all_human_mouse_metadata %>% 
       filter(participant_id %in% c(all_human_mouse_metadata %>%
                                      filter(pathogen == "Influenza", organism == "Human") %>% 
                                      pull(participant_id) %>% 
                                      unique() %>% 
                                      sample(size = 30, replace = FALSE)))
  )

#Import samples and annotate with metadata
log2fc_samples = readRDS(here("tables", "all_human_mouse_samples_log2fc.rds")) 

log2fc_samples_annotated = log2fc_samples %>% 
  inner_join(all_human_mouse_metadata_filtered, 
             by = join_by("geo_sample"))   

#Check
log2fc_samples_annotated %>% 
  select(pathogen, organism, timepoint, participant_id) %>% 
  distinct() %>% 
  dplyr::count(pathogen, organism, timepoint) 


#DEGs
all_degs_human_mouse = 
  readRDS(file = here("tables", "all_human_mouse_dge_limma_degs_matched.rds"))

all_degs_human_mouse %>% 
  dplyr::count(pathogen, organism, timepoint)


#Get DEGs from different organisms
mouse_degs = all_degs_human_mouse %>%
  group_by(pathogen, timepoint) %>%
  filter(adj_p_val <= 0.05, organism == "Mouse") %>%
  select(pathogen, human_symbol, timepoint) %>% 
  ungroup()

human_degs = all_degs_human_mouse %>%
  group_by(pathogen, timepoint) %>%
  filter(adj_p_val <= 0.05, organism == "Human") %>%
  select(pathogen, human_symbol, timepoint) %>% 
  ungroup()

#Get shared DEGs 
human_mouse = mouse_degs %>%
  full_join(human_degs %>%
               ungroup() %>%
               distinct(),
             by = join_by("human_symbol", "timepoint", "pathogen")) %>% 
  distinct()

human_mouse %>% 
  dplyr::count(pathogen, timepoint)

```

```{r}
# #Input
# filename = "all_human_mouse"
# human_mouse_log2fc_samples = log2fc_samples_annotated

#Input
filename = "all_human_mouse_DEGs"

human_mouse_log2fc_samples =
  log2fc_samples_annotated %>%
  inner_join(human_mouse %>%
               mutate(timepoint = as.character(timepoint)),
             by = join_by(timepoint, human_symbol, pathogen))
  

```



By pathogen, all vs all.
```{r}
library(tidyverse)
library(Hmisc)
library(here)
library(janitor)

# 1. Define Inputs and Gene Subsets -----------------------------------------
# We create a list to automate the loop over both sets
gene_sets_list <- list(
  "AllGenes" = human_mouse_log2fc_samples,
  "Immune"   = human_mouse_log2fc_samples %>% 
                 filter(human_symbol %in% (btm_annotation_genes %>% 
                                          filter(module_category == "immune") %>% 
                                          pull(symbol) %>% unique()))
)


# 2. Main Processing Loop ----------------------------------------------------
# map_dfr iterates through "AllGenes" and "Immune", stacking the results
final_correlations <- map_dfr(names(gene_sets_list), function(set_name) {
  
  message(paste("Processing gene set:", set_name))
  current_data <- gene_sets_list[[set_name]]
  
  # Data Preparation: Long format to Wide Matrix
  genes_samples_long <- current_data %>%
    select(pathogen, organism, human_symbol, participant_id, timepoint, log2fc) %>%
    distinct() %>%
    mutate(condition = paste0(pathogen, "_", participant_id, "_", timepoint))

  samples_annotation <- genes_samples_long %>%
    select(pathogen, organism, participant_id, timepoint, condition) %>%
    distinct() 

  genes_samples_wide <- genes_samples_long %>%
    select(human_symbol, condition, log2fc) %>% 
    distinct() %>%
    pivot_wider(names_from = "condition", values_from = "log2fc", values_fn = mean) %>%
    column_to_rownames("human_symbol")

  # 3. Pathogen-Specific Correlation Calculation -----------------------------
  pathogen_list <- unique(samples_annotation$pathogen)
  
  map_dfr(pathogen_list, function(path) {
    
    pathogen_samples <- samples_annotation %>% 
      filter(pathogen == path) %>% 
      pull(condition) 
    
    # Subset matrix and drop NAs
    matrix_subset <- genes_samples_wide %>%
      select(all_of(intersect(pathogen_samples, colnames(.)))) %>%
      # drop_na() %>%
      as.matrix()
    
    if(nrow(matrix_subset) <= 4 | ncol(matrix_subset) < 2) {
      warning(paste("Skipping pathogen:", path, "- only", nrow(matrix_subset), "shared genes found."))
      return(NULL)
    }
    
    # Calculate Spearman Correlation
    cor_obj <- rcorr(matrix_subset, type = "spearman")
    
    # Tidy extraction of Rho and P-values
    r_long <- as.data.frame(cor_obj$r) %>%
      rownames_to_column("condition1") %>%
      pivot_longer(-condition1, names_to = "condition2", values_to = "rho")
    
    p_long <- as.data.frame(cor_obj$P) %>%
      rownames_to_column("condition1") %>%
      pivot_longer(-condition1, names_to = "condition2", values_to = "p_val")
    
    # Join with Metadata and Filter Duplicates
    r_long %>%
      left_join(p_long, by = c("condition1", "condition2")) %>%
      filter(condition1 < condition2) %>%
      left_join(samples_annotation %>% select(condition, organism1 = organism, timepoint1 = timepoint), 
                by = c("condition1" = "condition")) %>%
      left_join(samples_annotation %>% select(condition, organism2 = organism, timepoint2 = timepoint), 
                by = c("condition2" = "condition")) %>%
      mutate(pathogen = path, gene_set = set_name)
  })
})

# 4. Statistical Summary per Comparison and Timepoint -----------------------
correlation_summary <- final_correlations %>% 
  # Keep Human as reference for cross-species comparisons
  filter(!(organism1 == "Mouse" & organism2 == "Human")) %>% 
  mutate(comparison = paste0(organism1, " vs. ", organism2),
         timepoint_comparison = paste0(timepoint1, "-", timepoint2)) %>% 
  group_by(gene_set, pathogen, comparison, timepoint1, timepoint2) %>% 
  summarise(
    mean_rho = mean(rho, na.rm = TRUE),
    sd_rho   = sd(rho, na.rm = TRUE),
    n        = n(),
    se       = sd_rho / sqrt(n),
    ci_low   = mean_rho - (qt(0.975, df = n - 1) * se),
    ci_high  = mean_rho + (qt(0.975, df = n - 1) * se),
    .groups = "drop"
  )

# 5. Final Filtering (Matched Timepoints and 24h Reference) -----------------
correlation_summary_filtered <- bind_rows(
  # Matched timepoints (e.g., 2h vs 2h)
  correlation_summary %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & comparison == "Human vs. Mouse"),
           timepoint1 == timepoint2),
  
  # 24h reference for acute infections
  correlation_summary %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"),
           comparison == "Human vs. Mouse",
           timepoint1 == "24h")
) %>%
  mutate(timepoint_label = if_else(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"), 
                                   timepoint2, timepoint1),
         timepoint_label = as.character(timepoint_label),
         timepoint_label = str_replace(timepoint_label,"Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint_label, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
          pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma"),
         gene_set = str_replace(gene_set, "AllGenes", "All genes")) 

correlation_summary_filtered
# Save Output
saveRDS(correlation_summary_filtered, file = here("tables", paste0(filename, "_gene_sets_correlation_summary.rds")))
```


Horizontal
```{r fig.height=3, fig.width=16}
#Import
correlation_summary_filtered_DEGs = readRDS(file = here("tables","all_human_mouse_DEGs_gene_sets_correlation_summary.rds")) %>% 
  mutate(filter = "DEGs")
correlation_summary_filtered_all = readRDS(file = here("tables","all_human_mouse_gene_sets_correlation_summary.rds")) %>% 
  mutate(filter = "No filter")

correlation_summaries =  
  bind_rows(correlation_summary_filtered_DEGs,
            correlation_summary_filtered_all) %>% 
  mutate(timepoint = str_replace(timepoint, "Day ", "Day\n"),
         timepoint = as.character(timepoint),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         filter = fct_relevel(filter, c("No filter", "DEGs")),
         pathogen = as.character(pathogen),
         pathogen = str_replace(pathogen, "Staphylococcus aureus", "S. aureus") %>% 
           str_replace("Escherichia coli", "E. coli"),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "S. aureus", "E. coli", "Burn", "Trauma"))

#Plot
correlation_long_all_mouse_human_timepoints_summary_plot = correlation_summaries %>% 
  ggplot() +
  aes(x = timepoint,
      y = mean_rho,
      group = interaction(comparison, gene_set), 
      color = comparison,
      fill = comparison) +
  
  facet_nested_wrap(~filter + pathogen, scales = "free_x", nrow = 1) +
  geom_line(aes(linetype = gene_set,
                alpha = gene_set),
            linewidth = 0.5) +
    geom_errorbar(aes(ymin = ci_low, 
                    ymax = ci_high,
                    alpha = gene_set), 
                width = 0.2,
                size = 1,
                linetype = "solid") + 
   geom_point(aes(alpha = gene_set,
                  fill = comparison),
             size = 2, 
             pch = 21,
             color = "black",
             stroke = 0.5) + 

  scale_alpha_manual(values = c("Immune" = 1, "All genes" = 0.5)) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(-0.1, 1)
                     ) +
  theme_vaxgo() +
  scale_color_manual(values = colors$comparison_versus) +
  scale_fill_manual(values = colors$comparison_versus) +
  labs(
    # title = gene_set_name,
       x = "Time",
       y = "Correlation",
       shape = "Type",
       linetype = "Gene category (Line)",
       alpha = "Gene category (Opacity)",
       color = "Comparison") +
  geom_hline(yintercept = 0, linetype = 2, color = "black") +
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  guides(fill = "none",
         linetype = guide_legend(order = 1, linewidth = 5), 
        alpha = guide_legend(order = 2, linewidth = 3),
        color = guide_legend(override.aes = list(linewidth = 3)))

correlation_long_all_mouse_human_timepoints_summary_plot

correlation_long_all_mouse_human_timepoints_summary_plot %>%
  ggsave(filename = here("Figures", paste0("all_human_mouse_all_DEGs_correlation_timepoints_summary_plot_horizontal.png")),
         width = 16,
         height = 3)

```

Vertical
```{r fig.height=6, fig.width=10}
#Import
correlation_summary_filtered_DEGs = readRDS(file = here("tables","all_human_mouse_DEGs_gene_sets_correlation_summary.rds")) %>% 
  mutate(filter = "DEGs")
correlation_summary_filtered_all = readRDS(file = here("tables","all_human_mouse_gene_sets_correlation_summary.rds")) %>% 
  mutate(filter = "No filter")

correlation_summaries =  
  bind_rows(correlation_summary_filtered_DEGs,
            correlation_summary_filtered_all) %>% 
  mutate(timepoint = str_replace(timepoint, "Day ", "Day\n"),
         timepoint = as.character(timepoint),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         filter = fct_relevel(filter, c("No filter", "DEGs")),
         pathogen = as.character(pathogen),
         pathogen = str_replace(pathogen, "Staphylococcus aureus", "S. aureus") %>% 
           str_replace("Escherichia coli", "E. coli"),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "S. aureus", "E. coli", "Burn", "Trauma"))

#Plot
correlation_long_all_mouse_human_timepoints_summary_plot = correlation_summaries %>% 
  ggplot() +
  aes(x = timepoint,
      y = mean_rho,
      group = interaction(comparison, gene_set), 
      color = comparison,
      fill = comparison) +
  
  facet_nested_wrap(~filter + pathogen, scales = "free_x", nrow = 2) +
  geom_line(aes(linetype = gene_set,
                alpha = gene_set),
            linewidth = 0.5) +
    geom_errorbar(aes(ymin = ci_low, 
                    ymax = ci_high,
                    alpha = gene_set), 
                width = 0.2,
                size = 1,
                linetype = "solid") + 
   geom_point(aes(alpha = gene_set,
                  fill = comparison),
             size = 2, 
             pch = 21,
             color = "black",
             stroke = 0.5) + 

  scale_alpha_manual(values = c("Immune" = 1, "All genes" = 0.5)) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(-0.1, 1)
                     ) +
  theme_vaxgo() +
  scale_color_manual(values = colors$comparison_versus) +
  scale_fill_manual(values = colors$comparison_versus) +
  labs(
    # title = gene_set_name,
       x = "Time",
       y = "Correlation",
       shape = "Type",
       linetype = "Gene category (Line)",
       alpha = "Gene category (Opacity)",
       color = "Comparison") +
  geom_hline(yintercept = 0, linetype = 2, color = "black") +
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          panel.spacing.y = unit(0, "lines"),
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 10, hjust = 0),
          legend.spacing.y = unit(0, "cm"),
          legend.title = element_text(size = 10),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  guides(fill = "none",
         linetype = guide_legend(order = 1, linewidth = 5, nrow = 2, byrow = TRUE), 
        alpha = guide_legend(order = 2, linewidth = 3, nrow = 2, byrow = TRUE,),
        color = guide_legend(override.aes = list(linewidth = 3), nrow = 3, byrow = TRUE,))

correlation_long_all_mouse_human_timepoints_summary_plot

correlation_long_all_mouse_human_timepoints_summary_plot %>%
  ggsave(filename = here("Figures", paste0("all_human_mouse_all_DEGs_correlation_timepoints_summary_plot_vertical.png")),
         width = 10,
         height = 6)

```






