---
title: "3.1_Comparing_Human_Mouse_ByCondition"
output: html_document
---
# Required

```{r message=FALSE, warning=FALSE}
# install.packages("here") if needed
install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```

#=============================

# 1. Fluad, Vaccine

##Functional analysis

### Condition-wise

BTMs correlation, intra- and inter-species

#### Immune

##### Mean Log2FC and GSEA (Immune)

Compare means and NES Import

```{r}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "influenza_fluad_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Influenza", vaccine = "Fluad",
         timepoint = as.numeric(timepoint)) %>% 
  rename(mean_l2fc = mean,
         ci_lower = lower,
         ci_upper = upper)

#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "influenza_fluad_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Gene set enrichment analysis (GSEA) --------
#Get calculate means -----
#Annotate genes by BTM 

dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen,
            group, subgroup) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%    
  adjust_pvalue(method = "BH") %>%  
  add_significance()

dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Print
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))

```



##### GSEA

```{r message=FALSE}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

#Run multiple GSEA and unite means
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTMs")


gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  full_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


#### MSigDB Hallmarks

##### Mean Log2FC

```{r fig.width=10, fig.height= 8}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "influenza_fluad_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Influenza", vaccine = "Fluad",
         timepoint = as.numeric(timepoint)) %>% 
   rename(mean_l2fc = mean,
         ci_lower = lower,
         ci_upper = upper)

filename = "influenza_fluad_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Import Hallmarks -------

#Get annotated processes
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Gene set enrichment analysis (GSEA) --------
#Get calculate means -----
#Annotate genes by BTM 

dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     # testa média contra zero
  rstatix::adjust_pvalue(method = "BH") %>%      # corrige FDR
  rstatix::add_significance()

dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Print
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

##### GSEA

```{r message=FALSE, warning=TRUE}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

#Run multiple GSEA and unite means
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")


gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```

Plot distributions

```{r fig.height=12, fig.width=10}
#Import ========================
gsea_hallmarks_results = readRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
dge_hallmarks_mean_process = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
dge_hallmarks_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))
#Get annotated processes
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

#If timepoints match between ref and target, use this
dge_hallmarks_process_genes_matched = dge_hallmarks_process_genes

gsea_hallmarks_results_matched = gsea_hallmarks_results 

#Pivot to wide GSEA + mean
dge_hallmarks_gsea_mean_wide = gsea_hallmarks_results_matched %>% 
  select(organism, timepoint, process, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process,  set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

#Arrange process by log2fc
process_order_log2fc = dge_hallmarks_mean_process %>% 
  select(process, timepoint, organism, mean_log2fc) %>%
  filter(timepoint == 1,
         organism == "Human") %>% 
  group_by(organism, timepoint) %>% 
  arrange(-mean_log2fc) %>%
  pull(process) %>% 
  unique() 

dge_hallmarks_process_ordered = dge_hallmarks_mean_process 
  

# #Plot
# dge_hallmarks_process_distribution = dge_hallmarks_process_genes_matched %>% 
#   inner_join(msigdbr_grouped %>% select(process, group), by = "process") %>% 
#     mutate(timepoint_num = timepoint,
#     timepoint = paste0("Day ", timepoint_num),
#     timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)])))  %>% 
#   ggplot() +
#   aes(y = process,
#       x = mean_l2fc,
#       color = organism) +
#   facet_grid(~timepoint~group, scales = "free", space = "free") +
#   geom_vline(xintercept = 0, color = "black") +
#   geom_point(pch = "|", alpha = 1, 
#              show.legend = F,
#              size = 3) +
#   stat_summary(
#   aes(color = organism),
#   fun.data = mean_se,
#   geom = "errorbar",
#   width = 0.5,
#   size = 1) +
#   
#   stat_summary(
#     aes(fill = organism), 
#     fun = mean,             
#     color = "black",
#     geom = "point",             
#     size = 2,    
#     pch = 21,  
#     alpha = 1,
#     show.legend = T 
#   ) +
#   
#   # xlim(-2, 2) +
#   theme_vaxgo() +
#   theme(strip.text.y = element_text(angle = 0, hjust = 0), 
#         axis.ticks.y = element_line(color = "gray", alpha = 0.5),
#         legend.position = "top",
#         axis.text.x = element_text(size = 8),
#         axis.text.y = element_text(size = 8),
#         axis.line.y = element_blank(),
#         panel.spacing.y = unit(0.3, "lines"),
#         panel.grid.major.y = element_line(alpha = 0.5),
#         panel.grid.major.x = element_line(alpha = 0.5)) +
#   scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
#                                "Human" = "#90A4AEFF")) +
#   scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
#                                "Human" = "#90A4AEFF")) +
#     labs(color = "Organism",
#        y = "",
#        x = "Log2FC",
#        fill = "Organism",
#        title = "Mouse vs Human, MSigDB genes, mean Log2FC") 
# 
# dge_hallmarks_process_distribution 
# 
# dge_hallmarks_process_distribution %>% 
#   ggsave(filename = here("Figures", paste0(filename, "_dge_hallmarks_process_distribution.png")),
#          height = 12, 
#          width = 10)

```



#=============================

# 2. Engerix B, Vaccine

### Condition-wise

BTMs correlation, intra- and inter-species

#### BTMs

##### Mean Log2FC

Compare means and NES Import

```{r}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "hepatitisb_engerixb_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Hepatitis B", vaccine = "EngerixB",
         timepoint = as.numeric(timepoint))

#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "hepatitisb_engerixb_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

#Get calculate means -----
#Annotate genes by BTM 

dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen,
            group, subgroup) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     # testa média contra zero
  adjust_pvalue(method = "BH") %>%      # corrige FDR
  add_significance()

dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  
  

#Print
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

##### GSEA

```{r}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

#Run multiple GSEA and unite means
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTMs")


gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  full_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```

#####Plot distributions

```{r fig.height=12, fig.width=10}
#Import ========================
gsea_btm_results = readRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
dge_btm_mean_process = readRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
dge_btm_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))


#If timepoints match between ref and target, use this
dge_btm_process_genes_matched = dge_btm_process_genes

#Pivot to wide GSEA + mean
dge_btm_gsea_mean_wide = gsea_btm_results %>% 
  mutate(organism = case_when(str_detect(condition, "Mouse") ~ "Mouse",
                              str_detect(condition, "Human") ~ "Human",
                              T~condition)) %>% 
  select(organism, timepoint, process, group, subgroup, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process, subgroup, group, set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

#Arrange process by log2fc
process_order_log2fc = dge_btm_mean_process %>% 
  select(process, timepoint, organism, mean_log2fc, group) %>%
  filter(timepoint == 7,
         organism == "Human") %>% 
  group_by(organism, timepoint, group) %>% 
  arrange(-mean_log2fc) %>%
  pull(process) %>% 
  unique() 

dge_btm_process_ordered = dge_btm_mean_process %>% 
  ungroup() %>% 
    mutate(group = fct_relevel(as.factor(group), c(immune_order)),
           process = fct_relevel(as.factor(process), c(process_order_log2fc)) %>% fct_rev()) 

labels_process <- dge_btm_gsea_mean_wide %>%
  ungroup() %>% 
  distinct(process, p_adjust_Human, p_adjust_Mouse) %>%
  mutate(label = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
                         "bold", "plain"),
         label_color = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
                         "#4DBBD5FF", "gray50"))  %>% 
  ungroup() %>% 
  distinct() %>% 
  mutate(process = fct_relevel(process, process_order_log2fc)) %>% 
  filter(!is.na(p_adjust_Human))
  

#Plot
dge_btm_process_distribution = dge_btm_process_genes_matched %>% 
    mutate(timepoint_num = timepoint,
    timepoint = paste0("Day ", timepoint_num),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(y = process,
      x = mean_l2fc,
      color = organism) +
  facet_grid(~group~timepoint, scales = "free", space = "free") +
  geom_vline(xintercept = 0, color = "black") +
  geom_point(pch = "|", alpha = 1, 
             show.legend = F,
             size = 3) +
  stat_summary(
  aes(color = organism),
  fun.data = mean_se,
  geom = "errorbar",
  width = 0.5,
  size = 1) +
  
  stat_summary(
    aes(fill = organism), 
    fun = mean,             
    color = "black",
    geom = "point",             
    size = 2,    
    pch = 21,  
    alpha = 1,
    show.legend = T 
  ) +
  
  # xlim(-2, 2) +
  theme_vaxgo() +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), 
        axis.ticks.y = element_line(color = "gray", alpha = 0.5),
        legend.position = "top",
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8, face = c(labels_process$label), 
                                   color = c(labels_process$label_color)),
        axis.line.y = element_blank(),
        panel.spacing.y = unit(0.3, "lines"),
        panel.grid.major.y = element_line(alpha = 0.5),
        panel.grid.major.x = element_line(alpha = 0.5)) +
  scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
  scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
    labs(color = "Organism",
       y = "",
       x = "Log2FC",
       fill = "Organism",
       title = "Mouse vs Human, BTMs genes, mean Log2FC") 

dge_btm_process_distribution 

dge_btm_process_distribution %>% 
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_process_distribution.png")),
         height = 12, 
         width = 10)


```


#### MSigDB Hallmarks

##### Mean Log2FC

```{r fig.width=10, fig.height= 8}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Hepatitis B")

filename = "hepatitisb_engerixb_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Import Hallmarks -------

#Get annotated processes
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Gene set enrichment analysis (GSEA) --------
#Get calculate means -----
#Annotate genes by BTM 

dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     # testa média contra zero
  rstatix::adjust_pvalue(method = "BH") %>%      # corrige FDR
  rstatix::add_significance()

dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Print
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

##### GSEA

```{r message=FALSE, warning=TRUE}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

#Run multiple GSEA and unite means
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")


gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```

Plot distributions

```{r fig.height=12, fig.width=10}
#Import ========================
gsea_hallmarks_results = readRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
dge_hallmarks_mean_process = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
dge_hallmarks_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))
#Get annotated processes
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

#If timepoints match between ref and target, use this
dge_hallmarks_process_genes_matched = dge_hallmarks_process_genes

gsea_hallmarks_results_matched = gsea_hallmarks_results 

#Pivot to wide GSEA + mean
dge_hallmarks_gsea_mean_wide = gsea_hallmarks_results_matched %>% 
  mutate(organism = case_when(str_detect(condition, "Mouse") ~ "Mouse",
                              str_detect(condition, "Human") ~ "Human",
                              T~condition)) %>% 
  select(organism, timepoint, process, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process,  set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

#Arrange process by log2fc
process_order_log2fc = dge_hallmarks_mean_process %>% 
  select(process, timepoint, organism, mean_log2fc) %>%
  filter(timepoint == 7,
         organism == "Human") %>% 
  group_by(organism, timepoint) %>% 
  arrange(-mean_log2fc) %>%
  pull(process) %>% 
  unique() 

dge_hallmarks_process_ordered = dge_hallmarks_mean_process 
  

#Plot
dge_hallmarks_process_distribution = dge_hallmarks_process_genes_matched %>% 
    mutate(timepoint_num = timepoint,
    timepoint = paste0("Day ", timepoint_num),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)])))  %>% 
  ggplot() +
  aes(y = process,
      x = mean_l2fc,
      color = organism) +
  facet_grid(~group~timepoint, scales = "free", space = "free") +
  geom_vline(xintercept = 0, color = "black") +
  geom_point(pch = "|", alpha = 1, 
             show.legend = F,
             size = 3) +
  stat_summary(
  aes(color = organism),
  fun.data = mean_se,
  geom = "errorbar",
  width = 0.5,
  size = 1) +
  
  stat_summary(
    aes(fill = organism), 
    fun = mean,             
    color = "black",
    geom = "point",             
    size = 2,    
    pch = 21,  
    alpha = 1,
    show.legend = T 
  ) +
  
  # xlim(-2, 2) +
  theme_vaxgo() +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), 
        axis.ticks.y = element_line(color = "gray", alpha = 0.5),
        legend.position = "top",
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.line.y = element_blank(),
        panel.spacing.y = unit(0.3, "lines"),
        panel.grid.major.y = element_line(alpha = 0.5),
        panel.grid.major.x = element_line(alpha = 0.5)) +
  scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
  scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
    labs(color = "Organism",
       y = "",
       x = "Log2FC",
       fill = "Organism",
       title = "Mouse vs Human, MSigDB genes, mean Log2FC") 

dge_hallmarks_process_distribution 

dge_hallmarks_process_distribution %>% 
  ggsave(filename = here("Figures", paste0(filename, "_dge_hallmarks_process_distribution.png")),
         height = 12, 
         width = 10)

```


#=============================

# 3. S. aureus

## Functional analysis

### Immune

BTMs correlation, intra- and inter-species

#### Mean Log2FC

Compare means and NES Import

```{r}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Saureus")
#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "saureus_infection_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

#Get calculate means -----
#Annotate genes by BTM 

dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc), 
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen,
            group, subgroup) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     # testa média contra zero
  adjust_pvalue(method = "BH") %>%      # corrige FDR
  add_significance()

dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Print
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

#### GSEA

```{r message=FALSE, warning=FALSE}
# Gene set enrichment analysis (GSEA) --------
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)


#Run multiple GSEA and unite means
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")


gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


### MSigDB Hallmarks

#### Mean Log2FC

```{r fig.width=10, fig.height= 8}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Saureus")

filename = "saureus_infection_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Import Hallmarks -------

msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Gene set enrichment analysis (GSEA) --------
#Get calculate means -----
#Annotate genes by BTM 

dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  rstatix::adjust_pvalue(method = "BH") %>%     
  rstatix::add_significance()

dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Print
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

#### GSEA

```{r message=FALSE, warning=TRUE}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

#Run multiple GSEA and unite means
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")


gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```


#=============================

# 4. E. coli

## Functional analysis

### Condition-wise analysis

#### Immune

BTMs correlation, intra- and inter-species Compare means and NES

```{r}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "ecoli_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Ecoli")
#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "ecoli_infection_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Gene set enrichment analysis (GSEA) --------
#Get calculate means -----
#Annotate genes by BTM 

dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen,
            group, subgroup) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     # testa média contra zero
  adjust_pvalue(method = "BH") %>%      # corrige FDR
  add_significance()

dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Print
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

##### GSEA

```{r}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

#Run multiple GSEA and unite means
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")


gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


#### MSigDB Hallmarks


##### Mean Log2FC
```{r fig.width=10, fig.height= 8}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "ecoli_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Ecoli")

filename = "ecoli_infection_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()
```

```{r fig.width=10, fig.height= 8}
# Import Hallmarks -------

#Get annotated processes
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Gene set enrichment analysis (GSEA) --------
#Get calculate means -----
#Annotate genes by BTM 

dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     # testa média contra zero
  rstatix::adjust_pvalue(method = "BH") %>%      # corrige FDR
  rstatix::add_significance()

dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Print
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

##### GSEA

```{r message=FALSE, warning=TRUE}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

#Run multiple GSEA and unite means
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")


gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```



#===========================
#5. Burn

## Functional analysis

### Immune

BTMs correlation, intra- and inter-species

#### Mean Log2FC

Compare means and NES Import

```{r}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Burn")

#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "burn_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

#Get calculate means -----
#Annotate genes by BTM 
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc), 
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen,
            group, subgroup) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  adjust_pvalue(method = "BH") %>%     
  add_significance()

dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Print
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

#### GSEA

```{r message=FALSE, warning=FALSE}
# Gene set enrichment analysis (GSEA) --------
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)


#Run multiple GSEA and unite means
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")


gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


### MSigDB Hallmarks

#### Mean Log2FC

```{r fig.width=10, fig.height= 8}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Burn")

#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "burn_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Import Hallmarks -------

msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Gene set enrichment analysis (GSEA) --------
#Get calculate means -----
#Annotate genes by BTM 

dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  rstatix::adjust_pvalue(method = "BH") %>%     
  rstatix::add_significance()

dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Print
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

#### GSEA

```{r message=FALSE, warning=TRUE}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

#Run multiple GSEA and unite means
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")


gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```


#===========================
#6. Trauma

## Functional analysis

### Immune

BTMs correlation, intra- and inter-species

#### Mean Log2FC

Compare means and NES Import

```{r}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Trauma")

#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "trauma_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

#Get calculate means -----
#Annotate genes by BTM 
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc), 
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen,
            group, subgroup) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  adjust_pvalue(method = "BH") %>%     
  add_significance()

dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

dge_btm_mean_process

#Print
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

#### GSEA

```{r message=FALSE, warning=FALSE}
# Gene set enrichment analysis (GSEA) --------
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)


#Run multiple GSEA and unite means
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")


gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


### MSigDB Hallmarks

#### Mean Log2FC

```{r fig.width=10, fig.height= 8}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Trauma")

#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "trauma_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Import Hallmarks -------

msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Gene set enrichment analysis (GSEA) --------
#Get calculate means -----
#Annotate genes by BTM 

dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  rstatix::adjust_pvalue(method = "BH") %>%     
  rstatix::add_significance()

dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Print
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

#### GSEA

```{r message=FALSE, warning=TRUE}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

#Run multiple GSEA and unite means
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")


gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))
  
  
#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```


#===========================