---
title: "3.1_Comparing_Human_Mouse_ByCondition"
output: html_document
---
# Required

```{r message=FALSE, warning=FALSE}
# install.packages("here") if needed
install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```

#=============================

# 1. Fluad, Vaccine

##Functional analysis

### Condition-wise

BTMs correlation, intra- and inter-species

#### Immune

##### Mean Log2FC and GSEA (Immune)

Load the pre-computed differential expression results and the Blood Transcription Module annotation table. Join them on gene symbol to assign each gene to its BTM module, then collapse genes to module-level mean log2FC for each organism and timepoint. For each module, test whether its mean log2FC differs from zero using a one-sample t-test with Benjamini-Hochberg correction. Save the resulting module-level summary for downstream plotting.

```{r}
# Load data --------
# Import pre-computed DGE results (limma output: mean_l2fc, adj_p_val per gene)
# Annotate pathogen/vaccine/timepoint for downstream grouping
all_degs_human_mouse = readRDS(here("tables", "influenza_fluad_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Influenza", vaccine = "Fluad",
         timepoint = as.numeric(timepoint)) 

# Import BTM gene-to-module mapping (one row per gene, columns: symbol, process, group, subgroup)
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "influenza_fluad_human_mouse"

# Extract unique condition labels for downstream filtering
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Annotate DEGs with BTM modules --------
# Rename to generic "symbol" for the join, then attach BTM metadata
# Keep only Human and Mouse (exclude any reference or FIT entries)
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

# Save intermediate for reuse in plotting without re-joining
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Summarise gene-level log2FC to module-level means --------
# For each organism x pathogen x timepoint x process:
#   - mean, median, sd, var of the constituent genes' log2FC
#   - n_genes: how many genes from the module were detected
#   - group/subgroup: preserved from BTM annotation (constant per process)
dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group),
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint)) 

# Test if module mean log2FC differs from zero --------
# One-sample t-test per module (mean_l2fc ~ 1, mu = 0)
# BH correction across all modules within each group
# Modules with <2 genes are skipped (filter(n() >= 2))
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%    
  adjust_pvalue(method = "BH") %>%  
  add_significance()

# Diagnostic: print modules with <2 genes (excluded from the t-test)
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join module means with their significance tests --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))

```



##### GSEA

Prepare a rank-ordered gene list from the DGE results using the combined metric `log2FC * -log(adj_p_val)` and run Gene Set Enrichment Analysis against the BTM gene sets. Merge the GSEA output (Normalized Enrichment Score, q-value) with the module-level mean log2FC table for integrated visualization of both magnitude and enrichment direction.

```{r message=FALSE}
# Prepare term-to-gene mapping for GSEA --------
# BTM annotation already has process and symbol columns
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

# Build ranked gene list --------
# Use a composite metric: log2FC * -log(p-value) to rank genes
# This balances effect size and statistical significance
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA across all conditions using the autoGSEA helper --------
# autoGSEA iterates over each condition internally
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTMs")

# Merge GSEA results with module-level means --------
# This produces a single table with both NES (from GSEA) and mean_log2fc (from summarise)
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), 
            by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


#### MSigDB Hallmarks

##### Mean Log2FC

Repeat the module-level analysis using MSigDB Hallmarks instead of BTMs. Load the DGE results and the Hallmark gene sets, join them on gene symbol, collapse to Hallmark-level mean log2FC per organism and timepoint, and test each Hallmark against zero with a one-sample t-test and BH correction. Save the Hallmark-level summary.

```{r fig.width=10, fig.height= 8}
# Load DGE results --------
# Same approach as the BTM analysis: annotate pathogen and cast timepoint to numeric
all_degs_human_mouse = readRDS(here("tables", "influenza_fluad_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Influenza", vaccine = "Fluad",
         timepoint = as.numeric(timepoint)) 

filename = "influenza_fluad_human_mouse"

# Extract unique conditions for downstream use
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Load MSigDB Hallmark gene set annotation
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Join DEGs with Hallmark annotation --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

# Save intermediate for re-use in plotting
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse genes to Hallmark-level means --------
# Note: Hallmarks lack group/subgroup columns, so only mean_log2fc and n_genes are computed
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, timepoint, process, pathogen) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero, with BH correction
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance()

# Print Hallmarks with fewer than 2 genes (excluded from the test)
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join means with significance tests and save
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

##### GSEA

Run Gene Set Enrichment Analysis against the MSigDB Hallmarks using the same composite rank metric as the BTMs analysis. Merge the GSEA output (NES, q-value) with the Hallmark-level mean log2FC table so that both magnitude and enrichment can be compared side by side.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping --------
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

# Build ranked gene list with the same composite metric as the BTM GSEA --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with module-level mean log2FC --------
# The full_join keeps modules that have means but no GSEA result and vice versa
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```

Plot distributions

Load the Fluad Hallmark GSEA and expression results, pivot to wide format, and arrange Hallmarks by Human log2FC at day 1. The plotting code below is commented out (gated) and can be activated when needed.

```{r fig.height=12, fig.width=10}
# Load the GSEA results, module-level means, and gene-level data --------
gsea_hallmarks_results = readRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
dge_hallmarks_mean_process = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
dge_hallmarks_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Assign handle for matched timepoint analysis (no matching needed here)
dge_hallmarks_process_genes_matched = dge_hallmarks_process_genes
gsea_hallmarks_results_matched = gsea_hallmarks_results 

# Pivot GSEA + mean to wide format (Human and Mouse side by side) --------
dge_hallmarks_gsea_mean_wide = gsea_hallmarks_results_matched %>% 
  select(organism, timepoint, process, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process,  set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

# Order Hallmarks by Human mean log2FC at the first timepoint --------
process_order_log2fc = dge_hallmarks_mean_process %>% 
  select(process, timepoint, organism, mean_log2fc) %>%
  filter(timepoint == 1,
         organism == "Human") %>% 
  group_by(organism, timepoint) %>% 
  arrange(-mean_log2fc) %>%
  pull(process) %>% 
  unique() 

dge_hallmarks_process_ordered = dge_hallmarks_mean_process 

# Plot code (gated) — activate by removing comment prefix --------
# dge_hallmarks_process_distribution = dge_hallmarks_process_genes_matched %>% 
#   inner_join(msigdbr_grouped %>% select(process, group), by = "process") %>% 
#     mutate(timepoint_num = timepoint,
#     timepoint = paste0("Day ", timepoint_num),
#     timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)])))  %>% 
#   ggplot() +
#   aes(y = process,
#       x = mean_l2fc,
#       color = organism) +
#   facet_grid(~timepoint~group, scales = "free", space = "free") +
#   geom_vline(xintercept = 0, color = "black") +
#   geom_point(pch = "|", alpha = 1, 
#              show.legend = F,
#              size = 3) +
#   stat_summary(
#   aes(color = organism),
#   fun.data = mean_se,
#   geom = "errorbar",
#   width = 0.5,
#   size = 1) +
#   
#   stat_summary(
#     aes(fill = organism), 
#     fun = mean,             
#     color = "black",
#     geom = "point",             
#     size = 2,    
#     pch = 21,  
#     alpha = 1,
#     show.legend = T 
#   ) +
#   
#   theme_vaxgo() +
#   theme(strip.text.y = element_text(angle = 0, hjust = 0), 
#         axis.ticks.y = element_line(color = "gray", alpha = 0.5),
#         legend.position = "top",
#         axis.text.x = element_text(size = 8),
#         axis.text.y = element_text(size = 8),
#         axis.line.y = element_blank(),
#         panel.spacing.y = unit(0.3, "lines"),
#         panel.grid.major.y = element_line(alpha = 0.5),
#         panel.grid.major.x = element_line(alpha = 0.5)) +
#   scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
#                                "Human" = "#90A4AEFF")) +
#   scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
#                                "Human" = "#90A4AEFF")) +
#     labs(color = "Organism",
#        y = "",
#        x = "Log2FC",
#        fill = "Organism",
#        title = "Mouse vs Human, MSigDB genes, mean Log2FC") 
# 
# dge_hallmarks_process_distribution 
# 
# dge_hallmarks_process_distribution %>% 
#   ggsave(filename = here("Figures", paste0(filename, "_dge_hallmarks_process_distribution.png")),
#          height = 12, 
#          width = 10)



#=============================

# 2. Engerix B, Vaccine

### Condition-wise

BTMs correlation, intra- and inter-species

#### BTMs

##### Mean Log2FC

Repeat the same module-level analysis for Engerix B (Hepatitis B vaccine). Load the DGE results for this vaccine, join with the BTM annotation, collapse to module-level mean log2FC per organism and timepoint, test each module against zero, and save the summary.

```{r}
# Load data --------
# Import Engerix B DGE results and annotate pathogen/vaccine
all_degs_human_mouse = readRDS(here("tables", "hepatitisb_engerixb_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Hepatitis B", vaccine = "EngerixB",
         timepoint = as.numeric(timepoint))

# Import BTM annotation (same table used for all pathogens)
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "hepatitisb_engerixb_human_mouse"

# Extract unique condition labels
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Annotate DEGs with BTM modules --------
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

#Save
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Collapse genes to module-level means per organism and timepoint --------
dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group), 
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint)) 

# One-sample t-test per module vs zero, with BH correction --------
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Diagnostic: modules with fewer than 2 genes (excluded from the test)
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join module means with their significance tests --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

##### GSEA

Run Gene Set Enrichment Analysis on the Engerix B DGE results using the BTM gene sets with the same composite rank metric as the Fluad analysis. Merge the GSEA output with the module-level mean log2FC table.

```{r}
# Prepare term-to-gene mapping for GSEA --------
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

# Build ranked gene list --------
# Use the same composite metric: log2FC * -log(adj_p_val)
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA across all conditions --------
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTMs")

# Clean GSEA output and merge with module-level means --------
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```

##### Plot distributions

Visualize the distribution of gene-level log2FC within each BTM for both organisms across timepoints. Individual genes are shown as tick marks, the module mean as a filled circle with standard error bars. BTMs are ordered by Human log2FC at the latest timepoint and grouped by immune process category.

```{r fig.height=12, fig.width=10}
# Load GSEA, module means, and gene-level data --------
gsea_btm_results = readRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
dge_btm_mean_process = readRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
dge_btm_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Assign handle for matched timepoint analysis (no matching needed here)
dge_btm_process_genes_matched = dge_btm_process_genes

# Pivot GSEA + mean to wide format with Human and Mouse side by side --------
dge_btm_gsea_mean_wide = gsea_btm_results %>% 
  mutate(organism = case_when(str_detect(condition, "Mouse") ~ "Mouse",
                              str_detect(condition, "Human") ~ "Human",
                              T~condition)) %>% 
  select(organism, timepoint, process, group, subgroup, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process, subgroup, group, set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

# Order BTMs by Human mean log2FC at the latest timepoint, within each group --------
process_order_log2fc = dge_btm_mean_process %>% 
  select(process, timepoint, organism, mean_log2fc, group) %>%
  filter(timepoint == 7,
         organism == "Human") %>% 
  group_by(organism, timepoint, group) %>% 
  arrange(-mean_log2fc) %>%
  pull(process) %>% 
  unique() 

dge_btm_process_ordered = dge_btm_mean_process %>% 
  ungroup() %>% 
    mutate(group = fct_relevel(as.factor(group), c(immune_order)),
           process = fct_relevel(as.factor(process), c(process_order_log2fc)) %>% fct_rev()) 

# Create label styling for BTMs significant in both species --------
labels_process <- dge_btm_gsea_mean_wide %>%
  ungroup() %>% 
  distinct(process, p_adjust_Human, p_adjust_Mouse) %>%
  mutate(label = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
                         "bold", "plain"),
         label_color = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
                         "#4DBBD5FF", "gray50"))  %>% 
  ungroup() %>% 
  distinct() %>% 
  mutate(process = fct_relevel(process, process_order_log2fc)) %>% 
  filter(!is.na(p_adjust_Human))

# Build the distribution plot --------
dge_btm_process_distribution = dge_btm_process_genes_matched %>% 
    mutate(timepoint_num = timepoint,
    timepoint = paste0("Day ", timepoint_num),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(y = process,
      x = mean_l2fc,
      color = organism) +
  facet_grid(~group~timepoint, scales = "free", space = "free") +
  geom_vline(xintercept = 0, color = "black") +
  geom_point(pch = "|", alpha = 1, 
             show.legend = F,
             size = 3) +
  stat_summary(
  aes(color = organism),
  fun.data = mean_se,
  geom = "errorbar",
  width = 0.5,
  size = 1) +
  
  stat_summary(
    aes(fill = organism), 
    fun = mean,             
    color = "black",
    geom = "point",             
    size = 2,    
    pch = 21,  
    alpha = 1,
    show.legend = T 
  ) +
  
  theme_vaxgo() +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), 
        axis.ticks.y = element_line(color = "gray", alpha = 0.5),
        legend.position = "top",
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8, face = c(labels_process$label), 
                                   color = c(labels_process$label_color)),
        axis.line.y = element_blank(),
        panel.spacing.y = unit(0.3, "lines"),
        panel.grid.major.y = element_line(alpha = 0.5),
        panel.grid.major.x = element_line(alpha = 0.5)) +
  scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
  scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
    labs(color = "Organism",
       y = "",
       x = "Log2FC",
       fill = "Organism",
       title = "Mouse vs Human, BTMs genes, mean Log2FC") 

# Print
dge_btm_process_distribution 

#Save
dge_btm_process_distribution %>% 
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_process_distribution.png")),
         height = 12, 
         width = 10)


#### MSigDB Hallmarks

##### Mean Log2FC

Repeat the Hallmark-level analysis for Engerix B. Load the all-pathogen DGE file filtered to Hepatitis B, join with the MSigDB Hallmark gene sets, collapse to Hallmark-level mean log2FC per organism and timepoint, test each Hallmark against zero, and save the summary.

```{r fig.width=10, fig.height= 8}
# Load DGE results for Engerix B --------
# Uses the combined all-pathogen table, filtered to Hepatitis B
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Hepatitis B")

filename = "hepatitisb_engerixb_human_mouse"

# Extract unique condition labels
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Load MSigDB Hallmark gene set annotation
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Join DEGs with Hallmark annotation --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

#Save
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse genes to Hallmark-level means per organism and timepoint --------
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero, with BH correction --------
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance()

# Diagnostic: Hallmarks with fewer than 2 genes (excluded from the test)
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join Hallmark means with significance tests --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

##### GSEA

Run Gene Set Enrichment Analysis on the Engerix B DGE results using the MSigDB Hallmark gene sets with the same composite rank metric. Merge the GSEA output with the Hallmark-level mean log2FC table.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping --------
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

# Build ranked gene list with the same composite metric as the BTM analysis --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level means --------
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```

Plot distributions

Visualise the distribution of gene-level log2FC within each Hallmark for both organisms across timepoints. The plot shows individual gene values as points, the module mean as a filled circle, and standard error bars. Hallmarks are ordered by Human log2FC at the latest timepoint.

```{r fig.height=12, fig.width=10}
# Load the GSEA results, module-level means, and gene-level data --------
gsea_hallmarks_results = readRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
dge_hallmarks_mean_process = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
dge_hallmarks_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Assign handle for matched timepoint analysis (no matching needed here)
dge_hallmarks_process_genes_matched = dge_hallmarks_process_genes
gsea_hallmarks_results_matched = gsea_hallmarks_results 

# Pivot GSEA + mean to wide format (Human and Mouse side by side) --------
dge_hallmarks_gsea_mean_wide = gsea_hallmarks_results_matched %>% 
  mutate(organism = case_when(str_detect(condition, "Mouse") ~ "Mouse",
                              str_detect(condition, "Human") ~ "Human",
                              T~condition)) %>% 
  select(organism, timepoint, process, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process,  set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

# Order Hallmarks by Human mean log2FC at the latest timepoint --------
process_order_log2fc = dge_hallmarks_mean_process %>% 
  select(process, timepoint, organism, mean_log2fc) %>%
  filter(timepoint == 7,
         organism == "Human") %>% 
  group_by(organism, timepoint) %>% 
  arrange(-mean_log2fc) %>%
  pull(process) %>% 
  unique() 

# Build the distribution plot --------
# geom_point shows each gene, stat_summary shows the module mean and SE
dge_hallmarks_process_distribution = dge_hallmarks_process_genes_matched %>% 
    mutate(timepoint_num = timepoint,
    timepoint = paste0("Day ", timepoint_num),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)])))  %>% 
  ggplot() +
  aes(y = process,
      x = mean_l2fc,
      color = organism) +
  facet_grid(~group~timepoint, scales = "free", space = "free") +
  geom_vline(xintercept = 0, color = "black") +
  geom_point(pch = "|", alpha = 1, 
             show.legend = F,
             size = 3) +
  stat_summary(
  aes(color = organism),
  fun.data = mean_se,
  geom = "errorbar",
  width = 0.5,
  size = 1) +
  
  stat_summary(
    aes(fill = organism), 
    fun = mean,             
    color = "black",
    geom = "point",             
    size = 2,    
    pch = 21,  
    alpha = 1,
    show.legend = T 
  ) +
  
  theme_vaxgo() +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), 
        axis.ticks.y = element_line(color = "gray", alpha = 0.5),
        legend.position = "top",
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.line.y = element_blank(),
        panel.spacing.y = unit(0.3, "lines"),
        panel.grid.major.y = element_line(alpha = 0.5),
        panel.grid.major.x = element_line(alpha = 0.5)) +
  scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
  scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
    labs(color = "Organism",
       y = "",
       x = "Log2FC",
       fill = "Organism",
       title = "Mouse vs Human, MSigDB genes, mean Log2FC") 

# Print and save the plot
dge_hallmarks_process_distribution 

dge_hallmarks_process_distribution %>% 
  ggsave(filename = here("Figures", paste0(filename, "_dge_hallmarks_process_distribution.png")),
         height = 12, 
         width = 10)

```


#=============================

# 3. S. aureus

## Functional analysis

### Immune

BTMs correlation, intra- and inter-species

#### Mean Log2FC

Load the S. aureus DGE results, annotate each gene with its BTM membership, collapse to a module-level mean log2FC per organism and timepoint, test each module against zero with a one-sample t-test, and save the outputs.

```{r}
# Load S. aureus DGE output and BTM annotation --------
all_degs_human_mouse = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Saureus")
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "saureus_infection_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

# Annotate DEGs with BTM membership and save gene-level table --------
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

#Save
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Collapse gene-level log2FC to BTM-level means per organism and timepoint --------
dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc), 
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group), 
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per BTM vs zero with BH correction --------
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Identify BTMs with fewer than 2 genes (will fail t-test)
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the S. aureus DGE results using the BTM gene sets. Build the composite rank metric, pass it to autoGSEA, and merge the enrichment results with the BTM-level mean log2FC table.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping --------
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against BTMs --------
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")

# Clean GSEA output and merge with BTM-level means --------
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


### MSigDB Hallmarks

#### Mean Log2FC

Repeat the module-level analysis for S. aureus using MSigDB Hallmarks instead of BTMs. Load the Hallmark gene sets, join with the DGE results, collapse to Hallmark-level mean log2FC per organism and timepoint, test each Hallmark against zero, and save.

```{r fig.width=10, fig.height= 8}
# Load S. aureus DGE output --------
all_degs_human_mouse = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Saureus")

filename = "saureus_infection_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Load Hallmark gene-set annotation --------
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Annotate DEGs with Hallmark membership and save gene-level table --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

#Save
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse gene-level log2FC to Hallmark-level means --------
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero with BH correction --------
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  rstatix::adjust_pvalue(method = "BH") %>%     
  rstatix::add_significance()

# Identify Hallmarks with fewer than 2 genes (will fail t-test)
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the S. aureus DGE results using the MSigDB Hallmark gene sets with the composite rank metric. Merge the GSEA output with the Hallmark-level mean log2FC table.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping --------
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, 
         log2fold_change = mean_l2fc, 
         adj_p_val)  %>% 
  filter(!is.na(genes), 
         !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level means --------
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```


#===========================


# 4. E. coli

## Functional analysis

### Condition-wise analysis

#### Immune

BTMs correlation, intra- and inter-species Compare means and NES

Load the E. coli DGE results, annotate each gene with its BTM membership, collapse to a module-level mean log2FC per organism and timepoint, test each module against zero with a one-sample t-test, and save the outputs.

```{r}
# Load E. coli DGE output and BTM annotation --------
all_degs_human_mouse = readRDS(here("tables", "ecoli_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Ecoli")
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "ecoli_infection_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Annotate DEGs with BTM membership and save gene-level table --------
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

#Save
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Collapse gene-level log2FC to BTM-level means per organism and timepoint --------
dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group), 
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per BTM vs zero with BH correction --------
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Identify BTMs with fewer than 2 genes (will fail t-test)
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

##### GSEA

Run Gene Set Enrichment Analysis on the E. coli DGE results using the BTM gene sets. Build the composite rank metric, pass it to autoGSEA, and merge with the BTM-level means.

```{r}
# Prepare BTM term-to-gene mapping --------
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against BTMs --------
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")

# Clean GSEA output and merge with BTM-level means --------
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


#### MSigDB Hallmarks

##### Mean Log2FC

Repeat the module-level analysis for E. coli using MSigDB Hallmarks instead of BTMs. Load the Hallmark gene sets, join with the DGE results, collapse to Hallmark-level mean log2FC per organism and timepoint, test each Hallmark against zero, and save.

```{r fig.width=10, fig.height= 8}
# Load E. coli DGE output --------
all_degs_human_mouse = readRDS(here("tables", "ecoli_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Ecoli")

filename = "ecoli_infection_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

# Load Hallmark gene-set annotation --------
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Annotate DEGs with Hallmark membership and save gene-level table --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

#Save
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse gene-level log2FC to Hallmark-level means --------
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero with BH correction --------
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance()

# Identify Hallmarks with fewer than 2 genes (will fail t-test)
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

##### GSEA

Run Gene Set Enrichment Analysis on the E. coli DGE results using the MSigDB Hallmark gene sets with the composite rank metric. Merge the GSEA output with the Hallmark-level mean log2FC table.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping --------
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level means --------
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```



#===========================
#5. Burn

## Functional analysis

### Immune

BTMs correlation, intra- and inter-species

#### Mean Log2FC

Load the Burn DGE results, annotate each gene with its BTM membership, collapse to a module-level mean log2FC per organism and timepoint, test each module against zero, and save.

```{r}
# Load Burn DGE output and BTM annotation --------
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Burn")
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "burn_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

# Annotate DEGs with BTM membership and save gene-level table --------
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

#Save
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Collapse gene-level log2FC to BTM-level means per organism and timepoint --------
dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc), 
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group), 
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per BTM vs zero with BH correction --------
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  adjust_pvalue(method = "BH") %>%     
  add_significance()

# Identify BTMs with fewer than 2 genes (will fail t-test)
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the Burn DGE results using the BTM gene sets. Build the composite rank metric, pass it to autoGSEA, and merge with the BTM-level means.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping --------
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against BTMs --------
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")

# Clean GSEA output and merge with BTM-level means --------
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```

### MSigDB Hallmarks

#### Mean Log2FC

Repeat the module-level analysis for Burn using MSigDB Hallmarks. Load the Hallmark gene sets, join with the DGE results, collapse to Hallmark-level mean log2FC per organism and timepoint, test each Hallmark against zero, and save.

```{r fig.width=10, fig.height= 8}
# Load Burn DGE output --------
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Burn")

filename = "burn_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Load Hallmark gene-set annotation --------
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Annotate DEGs with Hallmark membership and save gene-level table --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

#Save
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse gene-level log2FC to Hallmark-level means --------
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero with BH correction --------
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  rstatix::adjust_pvalue(method = "BH") %>%     
  rstatix::add_significance()

# Identify Hallmarks with fewer than 2 genes (will fail t-test)
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the Burn DGE results using the MSigDB Hallmark gene sets with the composite rank metric. Merge the GSEA output with the Hallmark-level mean log2FC table.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping --------
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level means --------
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```


#===========================
#6. Trauma

## Functional analysis

### Immune

BTMs correlation, intra- and inter-species

#### Mean Log2FC

Load the Trauma DGE results, annotate each gene with its BTM membership, collapse to a module-level mean log2FC per organism and timepoint, test each module against zero, and save.

```{r}
# Load Trauma DGE output and BTM annotation --------
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Trauma")
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "trauma_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

# Annotate DEGs with BTM membership and save gene-level table --------
dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(organism %in% c("Human", "Mouse")) 

#Save
dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

# Collapse gene-level log2FC to BTM-level means per organism and timepoint --------
dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc), 
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group), 
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per BTM vs zero with BH correction --------
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  adjust_pvalue(method = "BH") %>%     
  add_significance()

# Identify BTMs with fewer than 2 genes (will fail t-test)
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  )  

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the Trauma DGE results using the BTM gene sets. Build the composite rank metric, pass it to autoGSEA, and merge with the BTM-level means.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping --------
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against BTMs --------
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")

# Clean GSEA output and merge with BTM-level means --------
gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  inner_join(dge_btm_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```


### MSigDB Hallmarks

#### Mean Log2FC

Repeat the module-level analysis for Trauma using MSigDB Hallmarks. Load the Hallmark gene sets, join with the DGE results, collapse to Hallmark-level mean log2FC per organism and timepoint, test each Hallmark against zero, and save.

```{r fig.width=10, fig.height= 8}
# Load Trauma DGE output --------
all_degs_human_mouse = readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds")) %>% 
  filter(pathogen == "Trauma")

filename = "trauma_human_mouse"

# Extract condition-to-timepoint mapping --------
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Load Hallmark gene-set annotation --------
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Annotate DEGs with Hallmark membership and save gene-level table --------
dge_hallmarks_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(msigdbr_grouped, by = "symbol")

#Save
dge_hallmarks_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_process_genes.rds")))

# Collapse gene-level log2FC to Hallmark-level means --------
dge_hallmarks_mean_process = dge_hallmarks_process_genes %>% 
  group_by(organism, pathogen, timepoint, process) %>% 
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

# One-sample t-test per Hallmark vs zero with BH correction --------
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>% 
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%     
  rstatix::adjust_pvalue(method = "BH") %>%     
  rstatix::add_significance()

# Identify Hallmarks with fewer than 2 genes (will fail t-test)
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join p-values back to the mean table --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>% 
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>% 
  clean_names()

#Save
dge_hallmarks_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_hallmarks_mean_process.rds")))
```

#### GSEA

Run Gene Set Enrichment Analysis on the Trauma DGE results using the MSigDB Hallmark gene sets with the composite rank metric. Merge the GSEA output with the Hallmark-level mean log2FC table.

```{r message=FALSE, warning=TRUE}
# Prepare Hallmark term-to-gene mapping --------
term2gene = msigdbr_grouped  %>% 
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, 
         log2fold_change = mean_l2fc, 
         adj_p_val)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint),
          log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>% 
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks --------
gsea_hallmarks_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level means --------
gsea_hallmarks_results = gsea_hallmarks_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  filter(!is.na(process)) %>% 
  full_join(dge_hallmarks_mean_process %>% 
              rename(p_adjust_mean = p_adjust), by = join_by("condition", "process"))

#Save
gsea_hallmarks_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_hallmarks_results.rds")))
```


#===========================