---
title: "Comparing Human vs Mouse — Functional analysis"
author: Wasim Aluísio Prates-Syed
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Setup

```{r}
library(tidyverse)
library(here)
library(janitor)
library(rstatix)
library(glue)

source(here("scripts_notebooks", "required.R"))
```

# Functional enrichment

Unified pipeline for all 6 conditions (Fluad, Engerix B, S. aureus, E.
coli,Burn, Trauma). All conditions are loaded from the pre-merged DGE
file, then each analysis (BTM means, BTM GSEA, Hallmarks means,
Hallmarks GSEA) runs once on the combined data and produces a single
output file.

## Import all conditions

Load the pre-merged DGE file that already contains all 6 conditions with
consistent column names and pathogen labels.

```{r}
#Import gene sets
btm_annotation_genes <- read_csv(here("tables", "btm_annotation_genes.csv"))
msigdbr_grouped <- read_csv(here("tables", "msigdb_hallmarks_grouped_genes.csv"))

#Import DEGs table
#Save
all_human_mouse_dge_limma_degs_matched_control = readRDS(file = here("tables", "all_human_mouse_dge_limma_degs_matched_control.rds"))

#Filter out matched Control (DMD) and add only one timepoint
all_degs = all_human_mouse_dge_limma_degs_matched_control %>% 
  filter(organism != "DMD") %>%
  bind_rows(all_human_mouse_dge_limma_degs_matched_control %>%
              filter(organism == "DMD",
                     pathogen == "Burn",
                     timepoint == "Day 1") %>%
              mutate(timepoint = "Day 28",
                     pathogen = "DMD",
                     organism = "Mouse",
                     treatment = "Control",
                     condition = "DMD_Day 28")) %>%
  select(organism:human_symbol, mean_l2fc:sd, n) %>% 
  distinct()

# Extract unique condition labels for downstream filtering --------
annotation_conditions <- all_degs %>%
  select(condition, timepoint,  organism) %>%
  distinct()
```

## Blood Transcriptional Modules

### t-Test

Join DEGs with BTM annotation, collapse to module-level means per
organism and timepoint, test each module against zero, and save the
unified output.

```{r}
# Annotate each genes with its BTM membership by joining on gene symbol --------
dge_btm_process_genes <- all_degs %>%
  rename(symbol = human_symbol) %>%
  ungroup() %>%
  inner_join(btm_annotation_genes, by = "symbol") %>%
  filter(organism %in% c("Human", "Mouse"))

# Pivot the gene-level BTM table wide
#Create timepoint pairs
timepoint_map <- dge_btm_process_genes %>%
  filter(pathogen != "DMD") %>%
  select(pathogen, organism, timepoint) %>%
  distinct() %>%
  pivot_wider(
    names_from = organism,
    values_from = timepoint,
    values_fn = list
  )

#Pivot and bind control
dge_btm_process_genes_wide <- dge_btm_process_genes %>%
  filter(pathogen != "DMD") %>%
  select(symbol, mean_l2fc, adj_p_val, t, se, n_participants = n, pathogen, organism, timepoint, process, group) %>%
  distinct() %>%
  
  # Map and unnest cross-species timepoint combinations
  left_join(timepoint_map, by = "pathogen") %>%
  mutate(opposite_timepoint = if_else(organism == "Human", Mouse, Human)) %>%
  unnest(opposite_timepoint) %>%
  mutate(
    timepoint_human = if_else(organism == "Human", timepoint, opposite_timepoint),
    timepoint_mouse = if_else(organism == "Mouse", timepoint, opposite_timepoint),
    timepoint_comparison = stringr::str_c(timepoint_human, "-", timepoint_mouse)
  ) %>%
  select(-Human, -Mouse, -opposite_timepoint, -timepoint_human, -timepoint_mouse, -timepoint) %>%
  
  # Pivot the table wide
  pivot_wider(
    names_from = organism,
    values_from = c(mean_l2fc, adj_p_val, t, se, n_participants),
    values_fn = median
  ) %>%
  mutate(group = fct_relevel(group, immune_order)) %>%
  drop_na(mean_l2fc_Human, mean_l2fc_Mouse) %>% 
  inner_join(dge_btm_process_genes %>% 
                ungroup() %>%
              filter(pathogen == "DMD") %>%
              mutate(organism = "Control") %>% 
              select(symbol, mean_l2fc, adj_p_val, t, se, n_participants = n, organism, process, group) %>%
              distinct() %>%
              group_by(organism, process) %>%
              pivot_wider(names_from = organism,
                          values_from = c(mean_l2fc, adj_p_val, t, se, n_participants), values_fn = median) %>%
              ungroup() %>%
              mutate(group = fct_relevel(group, c(immune_order))),
             by = join_by(symbol, process, group)
  )

#Check
dge_btm_process_genes_wide

#Save
saveRDS(dge_btm_process_genes_wide, here("tables", "all_human_mouse_dge_btm_process_genes_wide.rds"))

# Collapse gene-level log2FC to module-level
dge_btm_mean_process <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  summarise(mean_log2fc = mean(mean_l2fc, na.rm = T),
            median_log2fc = median(mean_l2fc, na.rm = T),
            sd_log2fc = sd(mean_l2fc, na.rm = T),
            var_log2fc = var(mean_l2fc, na.rm = T),
            n_genes = n(),
            n_participants = dplyr::first(n),
            group = dplyr::first(group),
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>%
  ungroup() %>%
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint))

# One-sample t-test per BTM vs zero with BH correction --------
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>%
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Check
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint)

# Join module-level means with their t-test p-values --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>%
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>%
  clean_names()

#Pivot wide
dge_btm_mean_wide <- dge_btm_mean_process %>%
  select(organism, pathogen, timepoint, process, group,
         mean_log2fc, median_log2fc, sd_log2fc, var_log2fc, n_genes, p, p_adjust) %>%
  group_by(pathogen, timepoint, process) %>%
  pivot_wider(names_from = organism,
              values_from = c(mean_log2fc, median_log2fc, sd_log2fc, var_log2fc, n_genes, p, p_adjust),
              names_sep = "_") %>% 
  ungroup() 



dge_btm_mean_process %>% 
  dplyr::count(organism, pathogen, timepoint, n_participants) %>% 
  arrange(pathogen, timepoint)

#Save
saveRDS(dge_btm_mean_process, here("tables", "all_human_mouse_dge_btm_mean_process.rds"))
saveRDS(dge_btm_process_genes, here("tables", "all_human_mouse_dge_btm_process_genes.rds"))
```

### BTM GSEA

Rank genes with a composite metric (log2FC x -log(adj_p_val)) and run
GSEA against the BTM gene sets across all conditions simultaneously.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping for GSEA --------
term2gene <- btm_annotation_genes %>%
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
# Metric: log2FC * -log(adj_p_val) — balances effect size and significance
df_genes2GSEA <- all_degs %>%
  distinct() %>%
  select(organism, pathogen, timepoint,
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val) %>%
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint),
         log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>%
  select(condition, genes, 
         #Rank metric
         rank = log2fold_change_logp)

# Run GSEA across all conditions using autoGSEA (iterates over conditions internally) --------
gsea_output <- autoGSEA(df = df_genes2GSEA,
                         TERM2GENE = term2gene,
                         geneset_name = "BTMs")

# Clean GSEA output and merge with module-level means --------
gsea_results <- gsea_output$gsea %>%
  as.data.frame() %>%
  clean_names() %>%
  rename(process = id) %>%
  filter(!is.na(process)) %>%
  rownames_to_column("del") %>% 
  select(-del, -description) %>% 
  full_join(dge_btm_mean_process %>% 
               rename(p_adjust_mean = p_adjust),
            by = join_by("condition", "process"))

#Check
gsea_results %>% dplyr::count(pathogen, organism, timepoint)


#Save
saveRDS(gsea_results, here("tables", "all_human_mouse_gsea_btm_results.rds"))
```


```{r message=FALSE, warning=FALSE}
#Compare humans to mice
gsea_results_human = gsea_results %>% 
  filter(organism == "Human",
         pathogen != "DMD") %>% 
  select(-organism) %>% 
  distinct() %>% 
  rename_with( .fn = ~ stringr::str_c(.x, "_Human"),
               cols = c(enrichment_score:condition, timepoint, mean_log2fc:p_adjust_mean)) %>% 
  rename(process = process_Human, gsea_enrichment = gsea_enrichment_Human,
         pathogen = pathogen_Human, group = group_Human, subgroup = subgroup_Human)

gsea_results_mouse = gsea_results %>% 
  filter(organism == "Mouse",
         pathogen != "DMD") %>% 
  select(-organism) %>% 
  distinct() %>% 
  rename_with( .fn = ~ stringr::str_c(.x, "_Mouse"),
               cols = c(enrichment_score:condition, timepoint, mean_log2fc:p_adjust_mean)) %>% 
  rename(process = process_Mouse, gsea_enrichment = gsea_enrichment_Mouse,
         pathogen = pathogen_Mouse, group = group_Mouse, subgroup = subgroup_Mouse)

gsea_results_control = gsea_results %>% 
  filter(organism == "Mouse",
         pathogen == "DMD") %>% 
  mutate(organism = "Control",
         pathogen = "Control") %>% 
  select(-organism, -timepoint, -pathogen,) %>% 
  distinct() %>% 
  rename_with( .fn = ~ stringr::str_c(.x, "_Control"),
               cols = c(enrichment_score:condition, timepoint, mean_log2fc:p_adjust_mean)) %>% 
  rename(process = process_Control, gsea_enrichment = gsea_enrichment_Control, group = group_Control, subgroup = subgroup_Control)

dge_btm_gsea_mean_wide = gsea_results_human %>% 
  full_join(gsea_results_mouse,
            by = join_by(process, gsea_enrichment, pathogen, group, subgroup)) %>% 
  full_join(gsea_results_control,
            by = join_by(process, gsea_enrichment, group, subgroup)) %>% 
  drop_na(timepoint_Human, timepoint_Mouse) %>% 
  mutate(timepoint_comparison = paste0(timepoint_Human, "-", timepoint_Mouse)) %>% 
#Calculate statistics
  mutate(
    # Cross-species effect size delta 
    mean_log2fc_diff = mean_log2fc_Human - mean_log2fc_Mouse,
    
    # Joint FDR significance weighting (GSEA)
    weight_human_gsea_fdr = -log10(pmax(p_adjust_Human, 1e-10)),
    weight_mouse_gsea_fdr = -log10(pmax(p_adjust_Mouse, 1e-10)),
    weight_joint_gsea_fdr = sqrt(weight_human_gsea_fdr * weight_mouse_gsea_fdr),
    
    # Statistical Weights for Weighted Correlation
    se_sq_Human = var_log2fc_Human / n_genes_Human,
    se_sq_Mouse = var_log2fc_Mouse / n_genes_Mouse,
    
    # Combined variance of the cross-species mean difference
    var_combined = se_sq_Human + se_sq_Mouse,
    
    # Correct inverse variance weight
    weight_var_inverse = 1 / var_combined,
  )

#Save
saveRDS(dge_btm_gsea_mean_wide, here("tables", "all_human_mouse_dge_btm_gsea_mean_wide.rds"))


#Pivot longer
gsea_btm_results_all_matched = dge_btm_gsea_mean_wide %>% 
  tidyr::pivot_longer(
    # Select all organism-suffixed columns
    cols = dplyr::ends_with(c("_Human", "_Mouse", "_Control")),
    
    # Split column names into the metric name (.value) and the organism level
    names_to = c(".value", "organism"),
    names_sep = "_(?=[^_]+$)" # Regex splits strictly at the final underscore
  )

#Save
saveRDS(gsea_btm_results_all_matched, here("tables", "all_human_mouse_gsea_btm_results_matched.rds"))



```



### BTM distribution plots (one per condition)

Gene-level distribution plot for each condition: individual gene log2FC
values as points, module mean as a filled circle, standard error bars,
faceted by immune group and timepoint. BTMs with adj.p \<= 0.10 in both
species are highlighted in bold blue on the Y-axis.

```{r fig.height=12, fig.width=10}
walk(seq_len(nrow(conditions)), \(i) {
  row <- conditions[i, ]

  # Filter unified data to this condition --------
  cond_btm_genes <- dge_btm_process_genes %>%
    filter(pathogen == row$label)

  cond_btm_mean <- dge_btm_mean_process %>%
    filter(pathogen == row$label)

  cond_gsea <- gsea_results %>%
    semi_join(cond_btm_mean %>% distinct(condition), by = "condition")

  # Parse organism from condition label (stored as "Human_Influenza_1") --------
  cond_gsea <- cond_gsea %>%
    mutate(organism = case_when(
      str_detect(condition, "Mouse") ~ "Mouse",
      str_detect(condition, "Human") ~ "Human",
      TRUE ~ condition
    ))

  # Pivot GSEA results to wide format (Human and Mouse side by side) --------
  cond_gsea_wide <- cond_gsea %>%
    select(organism, timepoint, process, group, subgroup, set_size,
           nes, mean_log2fc, p_adjust, qvalue) %>%
    group_by(timepoint, process, subgroup, group, set_size) %>%
    pivot_wider(names_from = organism,
                values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
                names_sep = "_")

  # Determine ordering timepoint: last timepoint for this condition --------
  if (is.numeric(cond_btm_mean$timepoint)) {
    order_tp <- max(cond_btm_mean$timepoint, na.rm = TRUE)
  } else {
    tp_levels <- c("2h", "4h", "6h", "12h", "24h", "Early hours")
    available <- intersect(tp_levels, unique(cond_btm_mean$timepoint))
    order_tp <- if (length(available) > 0) dplyr::last(available) else unique(cond_btm_mean$timepoint)[1]
  }

  # Order BTMs by Human mean log2FC at the ordering timepoint, per immune group --------
  process_order <- cond_btm_mean %>%
    filter(timepoint == order_tp, organism == "Human") %>%
    group_by(organism, timepoint, group) %>%
    arrange(-mean_log2fc) %>%
    pull(process) %>%
    unique()

  # Create label styling: BTMs significant in both species get bold + blue --------
  labels_process <- cond_gsea_wide %>%
    ungroup() %>%
    distinct(process, p_adjust_Human, p_adjust_Mouse) %>%
    mutate(
      label = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10,
                     "bold", "plain"),
      label_color = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10,
                           "#4DBBD5FF", "gray50")
    ) %>%
    ungroup() %>%
    distinct() %>%
    mutate(process = fct_relevel(process, process_order)) %>%
    filter(!is.na(p_adjust_Human))

  # Factor releveling for display order --------
  dge_btm_process_ordered <- cond_btm_mean %>%
    ungroup() %>%
    mutate(
      group = fct_relevel(as.factor(group), immune_order),
      process = fct_relevel(as.factor(process), process_order) %>% fct_rev()
    )

  # Format timepoint labels for the plot --------
  if (is.numeric(cond_btm_genes$timepoint)) {
    dge_plot <- cond_btm_genes %>%
      mutate(
        timepoint_num = timepoint,
        timepoint = paste0("Day ", timepoint_num),
        timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))
      )
  } else {
    dge_plot <- cond_btm_genes %>%
      mutate(
        timepoint = factor(timepoint, levels = unique(timepoint))
      )
  }

  # Data preparation: set up ggplot with aesthetic mappings --------
  p <- dge_plot %>%
    ggplot() +
    aes(y = process, x = mean_l2fc, color = organism) +
    # Facetting: rows = immune group, columns = timepoint --------
    facet_grid(~group ~ timepoint, scales = "free", space = "free") +
    # Geom layers: reference line, individual gene points, module mean with SE --------
    geom_vline(xintercept = 0, color = "black") +
    geom_point(pch = "|", alpha = 1, show.legend = FALSE, size = 3) +
    stat_summary(
      aes(color = organism),
      fun.data = mean_se,
      geom = "errorbar",
      width = 0.5,
      size = 1
    ) +
    stat_summary(
      aes(fill = organism),
      fun = mean,
      color = "black",
      geom = "point",
      size = 2,
      pch = 21,
      alpha = 1,
      show.legend = TRUE
    ) +
    # Theme customization: vaxgo theme with significance-styled axis text --------
    theme_vaxgo() +
    theme(
      strip.text.y = element_text(angle = 0, hjust = 0),
      axis.ticks.y = element_line(color = "gray", alpha = 0.5),
      legend.position = "top",
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(
        size = 8,
        face = c(labels_process$label),
        color = c(labels_process$label_color)
      ),
      axis.line.y = element_blank(),
      panel.spacing.y = unit(0.3, "lines"),
      panel.grid.major.y = element_line(alpha = 0.5),
      panel.grid.major.x = element_line(alpha = 0.5)
    ) +
    # Colour scales: project colours for Human and Mouse --------
    scale_fill_manual(values = c("Mouse" = "#4DBBD5FF", "Human" = "#90A4AEFF")) +
    scale_color_manual(values = c("Mouse" = "#4DBBD5FF", "Human" = "#90A4AEFF")) +
    # Labels --------
    labs(
      color = "Organism",
      y = "",
      x = "Log2FC",
      fill = "Organism",
      title = glue("Mouse vs Human, BTMs genes, {row$name}")
    )

  print(p)

  #Save
  p %>%
    ggsave(
      filename = here("Figures", paste0(row$filename_prefix, "_dge_btm_process_distribution.png")),
      height = 12,
      width = 10
    )
})
```

## MSigDB Hallmarks

### t-Test

Join DEGs with MSigDB Hallmark annotation, collapse to Hallmark-level
means, test each Hallmark against zero, and save the unified output.

```{r}
# Annotate each DEG with its Hallmark membership --------
dge_hallmarks_process_genes <- all_degs %>%
  rename(symbol = human_symbol) %>%
  ungroup() %>%
  inner_join(msigdbr_grouped, by = "symbol")

# Collapse gene-level log2FC to BTM-level means per organism, pathogen, and timepoint --------
# Group by organism, pathogen, timepoint, and BTM process: each group is a set of genes in that module
dge_hallmarks_mean_process <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group)) %>%
  distinct() %>%
  ungroup() %>%
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint))

# One-sample t-test per BTM vs zero with BH correction --------
# Tests whether the mean log2FC of genes in this module differs from zero
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>%
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Check
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint)

# Join module-level means with their t-test p-values --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>%
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>%
  clean_names()


dge_hallmarks_mean_process %>% 
  dplyr::count(organism, pathogen, timepoint) %>% 
  arrange(pathogen, timepoint)

#Save
saveRDS(dge_hallmarks_mean_process, here("tables", "all_human_mouse_dge_hallmarks_mean_process.rds"))
saveRDS(dge_hallmarks_process_genes, here("tables", "all_human_mouse_dge_hallmarks_process_genes.rds"))
```

###  GSEA

Rank genes with a composite metric (log2FC x -log(adj_p_val)) and run
GSEA against the BTM gene sets across all conditions simultaneously.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping for GSEA --------
term2gene <- msigdbr_grouped %>%
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
# Metric: log2FC * -log(adj_p_val) — balances effect size and significance
df_genes2GSEA <- all_degs %>%
  distinct() %>%
  select(organism, pathogen, timepoint,
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val) %>%
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint),
         log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>%
  select(condition, genes, 
         #Rank metric
         rank = log2fold_change_logp)

# Run GSEA across all conditions using autoGSEA (iterates over conditions internally) --------
gsea_output <- autoGSEA(df = df_genes2GSEA,
                         TERM2GENE = term2gene,
                         geneset_name = "Hallmarks")

# Clean GSEA output and merge with module-level means --------
gsea_results <- gsea_output$gsea %>%
  as.data.frame() %>%
  clean_names() %>%
  rename(process = id) %>%
  filter(!is.na(process)) %>%
  inner_join(dge_hallmarks_mean_process %>% 
               rename(p_adjust_mean = p_adjust),
            by = join_by("condition", "process"))

#Check
gsea_results %>% dplyr::count(pathogen, organism, timepoint)

#Save
saveRDS(gsea_results, here("tables", "all_human_mouse_gsea_hallmarks_results.rds"))
```

BTM matching and wide pivot

Duplicate Human timepoints for S. aureus and E. coli (where Human and
Mouse timepoints differ) so cross-species comparisons align. Pivot GSEA
and means to wide format (one column per organism).

```{r}
# Extract species-specific timepoints per pathogen to map them dynamically
timepoint_map <- gsea_results %>% 
  dplyr::filter(pathogen != "DMD") %>%
  select(pathogen, organism, timepoint) %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = timepoint, values_fn = list)


# Generate matched timepoints for active organisms (Human and Mouse)
active_matched_timepoints <- gsea_results %>% 
  #Filter out control
  dplyr::filter(!(pathogen == "DMD")) %>%
  #Filter only Human and Mouse
  dplyr::filter(organism %in% c("Human", "Mouse")) %>% 
  dplyr::left_join(timepoint_map, by = "pathogen") %>% 
  dplyr::mutate(opposite_timepoint = dplyr::if_else(organism == "Human", Mouse, Human)) %>% 
  tidyr::unnest(opposite_timepoint) %>% 
  dplyr::mutate(
    timepoint_human = dplyr::if_else(organism == "Human", timepoint, opposite_timepoint),
    timepoint_mouse = dplyr::if_else(organism == "Mouse", timepoint, opposite_timepoint),
    timepoint_comparison = stringr::str_c(timepoint_human, "-", timepoint_mouse)
  ) %>% 
  dplyr::select(-Human, -Mouse, -opposite_timepoint)

# Check
 active_matched_timepoints %>%
  dplyr::count(organism, pathogen, timepoint_comparison)


# Match DMD control across all comparisons
gsea_hallmarks_results_all_matched <- dplyr::bind_rows(
  
  # 1. Retain the dynamically paired active experimental rows
  active_matched_timepoints,
  
  # 2. Extract DMD, strip original identifiers, rename the organism, and map universally
  gsea_results %>%
    dplyr::filter(organism == "Mouse" & pathogen == "DMD") %>%
    dplyr::mutate(organism = "DMD") %>%
    # Remove original pathogen/treatment/timepoint so they don't conflict during crossing
    dplyr::select(-pathogen,  -timepoint) %>% 
    tidyr::crossing(
      active_matched_timepoints %>%
        dplyr::select(pathogen, timepoint_comparison) %>%
        dplyr::distinct()
    )
)

# Pivot to wide format: one row per module per timepoint with Human and Mouse side by side --------
dge_hallmarks_gsea_mean_wide <- gsea_hallmarks_results_all_matched %>%
  select(organism, pathogen, timepoint_comparison, process, group, set_size,
         nes, mean_log2fc, p_adjust, p_adjust_mean) %>%
  group_by(pathogen, process, group, set_size) %>%
  pivot_wider(names_from = organism,
              values_from = c(mean_log2fc, nes, p_adjust, p_adjust_mean, set_size),
              names_sep = "_") %>% 
  ungroup()

dge_hallmarks_gsea_mean_wide %>% 
  dplyr::count(pathogen, timepoint_comparison)

#Save
saveRDS(dge_hallmarks_gsea_mean_wide, here("tables", "all_human_mouse_dge_hallmarks_gsea_mean_wide.rds"))
saveRDS(gsea_hallmarks_results_all_matched, here("tables", "all_human_mouse_gsea_hallmarks_results_all_matched.rds"))
```

### Distribution plots (one per condition)

Gene-level distribution plot for each condition: individual gene log2FC
values as points, module mean as a filled circle, standard error bars,
faceted by immune group and timepoint. BTMs with adj.p \<= 0.10 in both
species are highlighted in bold blue on the Y-axis.

```{r fig.height=12, fig.width=10}
walk(seq_len(nrow(conditions)), \(i) {
  row <- conditions[i, ]

  # Filter unified data to this condition --------
  cond_hallmarks_genes <- dge_hallmarks_process_genes %>%
    filter(pathogen == row$label)

  cond_hallmarks_mean <- dge_hallmarks_mean_process %>%
    filter(pathogen == row$label)

  cond_gsea <- gsea_results %>%
    semi_join(cond_hallmarks_mean %>% distinct(condition), by = "condition")

  # Parse organism from condition label (stored as "Human_Influenza_1") --------
  cond_gsea <- cond_gsea %>%
    mutate(organism = case_when(
      str_detect(condition, "Mouse") ~ "Mouse",
      str_detect(condition, "Human") ~ "Human",
      TRUE ~ condition
    ))

  # Pivot GSEA results to wide format (Human and Mouse side by side) --------
  cond_gsea_wide <- cond_gsea %>%
    select(organism, timepoint, process, group, subgroup, set_size,
           nes, mean_log2fc, p_adjust, qvalue) %>%
    group_by(timepoint, process, subgroup, group, set_size) %>%
    pivot_wider(names_from = organism,
                values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
                names_sep = "_")

  # Determine ordering timepoint: last timepoint for this condition --------
  if (is.numeric(cond_hallmarks_mean$timepoint)) {
    order_tp <- max(cond_hallmarks_mean$timepoint, na.rm = TRUE)
  } else {
    tp_levels <- c("2h", "4h", "6h", "12h", "24h", "Early hours")
    available <- intersect(tp_levels, unique(cond_hallmarks_mean$timepoint))
    order_tp <- if (length(available) > 0) dplyr::last(available) else unique(cond_hallmarks_mean$timepoint)[1]
  }

  # Order BTMs by Human mean log2FC at the ordering timepoint, per immune group --------
  process_order <- cond_hallmarks_mean %>%
    filter(timepoint == order_tp, organism == "Human") %>%
    group_by(organism, timepoint, group) %>%
    arrange(-mean_log2fc) %>%
    pull(process) %>%
    unique()

  # Create label styling: BTMs significant in both species get bold + blue --------
  labels_process <- cond_gsea_wide %>%
    ungroup() %>%
    distinct(process, p_adjust_Human, p_adjust_Mouse) %>%
    mutate(
      label = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10,
                     "bold", "plain"),
      label_color = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10,
                           "#4DBBD5FF", "gray50")
    ) %>%
    ungroup() %>%
    distinct() %>%
    mutate(process = fct_relevel(process, process_order)) %>%
    filter(!is.na(p_adjust_Human))

  # Factor releveling for display order --------
  dge_hallmarks_process_ordered <- cond_hallmarks_mean %>%
    ungroup() %>%
    mutate(
      group = fct_relevel(as.factor(group), immune_order),
      process = fct_relevel(as.factor(process), process_order) %>% fct_rev()
    )

  # Format timepoint labels for the plot --------
  if (is.numeric(cond_hallmarks_genes$timepoint)) {
    dge_plot <- cond_hallmarks_genes %>%
      mutate(
        timepoint_num = timepoint,
        timepoint = paste0("Day ", timepoint_num),
        timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))
      )
  } else {
    dge_plot <- cond_hallmarks_genes %>%
      mutate(
        timepoint = factor(timepoint, levels = unique(timepoint))
      )
  }

  # Data preparation: set up ggplot with aesthetic mappings --------
  p <- dge_plot %>%
    ggplot() +
    aes(y = process, x = mean_l2fc, color = organism) +
    # Facetting: rows = immune group, columns = timepoint --------
    facet_grid(~group ~ timepoint, scales = "free", space = "free") +
    # Geom layers: reference line, individual gene points, module mean with SE --------
    geom_vline(xintercept = 0, color = "black") +
    geom_point(pch = "|", alpha = 1, show.legend = FALSE, size = 3) +
    stat_summary(
      aes(color = organism),
      fun.data = mean_se,
      geom = "errorbar",
      width = 0.5,
      size = 1
    ) +
    stat_summary(
      aes(fill = organism),
      fun = mean,
      color = "black",
      geom = "point",
      size = 2,
      pch = 21,
      alpha = 1,
      show.legend = TRUE
    ) +
    # Theme customization: vaxgo theme with significance-styled axis text --------
    theme_vaxgo() +
    theme(
      strip.text.y = element_text(angle = 0, hjust = 0),
      axis.ticks.y = element_line(color = "gray", alpha = 0.5),
      legend.position = "top",
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(
        size = 8,
        face = c(labels_process$label),
        color = c(labels_process$label_color)
      ),
      axis.line.y = element_blank(),
      panel.spacing.y = unit(0.3, "lines"),
      panel.grid.major.y = element_line(alpha = 0.5),
      panel.grid.major.x = element_line(alpha = 0.5)
    ) +
    # Colour scales: project colours for Human and Mouse --------
    scale_fill_manual(values = c("Mouse" = "#4DBBD5FF", "Human" = "#90A4AEFF")) +
    scale_color_manual(values = c("Mouse" = "#4DBBD5FF", "Human" = "#90A4AEFF")) +
    # Labels --------
    labs(
      color = "Organism",
      y = "",
      x = "Log2FC",
      fill = "Organism",
      title = glue("Mouse vs Human, BTMs genes, {row$name}")
    )

  print(p)

  #Save
  p %>%
    ggsave(
      filename = here("Figures", paste0(row$filename_prefix, "_dge_hallmarks_process_distribution.png")),
      height = 12,
      width = 10
    )
})
```