---
title: "Comparing Human vs Mouse — Functional analysis"
author: Wasim Aluísio Prates-Syed
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Setup

```{r}
library(tidyverse)
library(here)
library(janitor)
library(rstatix)
library(glue)

source(here("scripts_notebooks", "required.R"))
```

# Functional enrichment

Unified pipeline for all 6 conditions (Fluad, Engerix B, S. aureus, E.
coli,Burn, Trauma). All conditions are loaded from the pre-merged DGE
file, then each analysis (BTM means, BTM GSEA, Hallmarks means,
Hallmarks GSEA) runs once on the combined data and produces a single
output file.

## Condition configuration

```{r}
conditions <- tribble(
  ~name,        ~label,        ~hallmarks_plot, ~filename_prefix,
  "Fluad",      "Influenza",   FALSE,           "influenza_fluad_human_mouse",
  "Engerix B",  "Hepatitis B", TRUE,            "hepatitisb_engerixb_human_mouse",
  "S. aureus",  "Saureus",     FALSE,           "saureus_infection_human_mouse",
  "E. coli",    "Ecoli",       FALSE,           "ecoli_infection_human_mouse",
  "Burn",       "Burn",        FALSE,           "burn_human_mouse",
  "Trauma",     "Trauma",      FALSE,           "trauma_human_mouse"
)
```

## Import all conditions

Load the pre-merged DGE file that already contains all 6 conditions with
consistent column names and pathogen labels.

```{r}
btm_annotation_genes <- read_csv(here("tables", "btm_annotation_genes.csv"))
msigdbr_grouped <- read_csv(here("tables", "msigdb_hallmarks_grouped_genes.csv"))

#Import DEGs table
all_degs <- readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds"))

#Biological negative control
biological_null_degs = readRDS(file = here("tables", "duchenne_mouse_dge_limma_degs.rds")) %>% 
  filter(timepoint == "Day 28")

#Bind with biological negative control
all_degs = all_degs %>% 
  bind_rows(biological_null_degs)

# Extract unique condition labels for downstream filtering --------
annotation_conditions <- all_degs %>%
  select(condition, timepoint, model = organism) %>%
  distinct()
```

## Blood Transcriptional Modules

### t-Test

Join DEGs with BTM annotation, collapse to module-level means per
organism and timepoint, test each module against zero, and save the
unified output.

```{r}
# Annotate each DEG with its BTM membership by joining on gene symbol --------
# Rename human_symbol to symbol to match the BTM annotation key
# Keep only Human and Mouse (exclude reference organisms)
dge_btm_process_genes <- all_degs %>%
  rename(symbol = human_symbol) %>%
  ungroup() %>%
  inner_join(btm_annotation_genes, by = "symbol") %>%
  filter(organism %in% c("Human", "Mouse"))

# Collapse gene-level log2FC to BTM-level means per organism, pathogen, and timepoint --------
# Group by organism, pathogen, timepoint, and BTM process: each group is a set of genes in that module
dge_btm_mean_process <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            group = dplyr::first(group),
            subgroup = dplyr::first(subgroup)) %>%
  distinct() %>%
  ungroup() %>%
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint))

# One-sample t-test per BTM vs zero with BH correction --------
# Tests whether the mean log2FC of genes in this module differs from zero
pvals_vs_zero <- dge_btm_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>%
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Diagnostic: print modules with fewer than 2 genes (excluded from t-test) --------
dge_btm_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join module-level means with their t-test p-values --------
dge_btm_mean_process <- dge_btm_mean_process %>%
  left_join(
    pvals_vs_zero %>%
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>%
  clean_names()

#Save
saveRDS(dge_btm_mean_process, here("tables", "all_human_mouse_dge_btm_mean_process.rds"))
saveRDS(dge_btm_process_genes, here("tables", "all_human_mouse_dge_btm_process_genes.rds"))
```

### BTM GSEA

Rank genes with a composite metric (log2FC x -log(adj_p_val)) and run
GSEA against the BTM gene sets across all conditions simultaneously.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping for GSEA --------
term2gene <- btm_annotation_genes %>%
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
# Metric: log2FC * -log(adj_p_val) — balances effect size and significance
df_genes2GSEA <- all_degs %>%
  distinct() %>%
  select(organism, pathogen, timepoint,
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val) %>%
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint),
         log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>%
  select(condition, genes, 
         #Rank metric
         rank = log2fold_change_logp)

# Run GSEA across all conditions using autoGSEA (iterates over conditions internally) --------
gsea_output <- autoGSEA(df = df_genes2GSEA,
                         TERM2GENE = term2gene,
                         geneset_name = "BTMs")

# Clean GSEA output and merge with module-level means --------
gsea_results <- gsea_output$gsea %>%
  as.data.frame() %>%
  clean_names() %>%
  rename(process = id) %>%
  filter(!is.na(process)) %>%
  inner_join(dge_btm_mean_process %>% 
               rename(p_adjust_mean = p_adjust),
            by = join_by("condition", "process"))

#Check
gsea_results %>% dplyr::count(pathogen, organism, timepoint)

#Save
saveRDS(gsea_results, here("tables", "all_human_mouse_gsea_btm_results.rds"))
```

BTM matching and wide pivot

Duplicate Human timepoints for S. aureus and E. coli (where Human and
Mouse timepoints differ) so cross-species comparisons align. Pivot GSEA
and means to wide format (one column per organism).

```{r}
# Duplicate Human rows at Mouse timepoints for infections with mismatched timepoints --------
# S. aureus and E. coli have Human at 2/4/6/12/24h but Mouse only at 24h;
# this creates Human entries at 24h for every Mouse timepoint
gsea_btm_results_all_matched <- gsea_results %>%
  select(pathogen, organism, process, group, timepoint, nes,
         p_adjust, p_adjust_mean, mean_log2fc, core_enrichment, set_size) %>%
  bind_rows(
    gsea_results %>%
      filter(organism == "Human",
             pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
      select(pathogen, organism, process, group, nes,
             p_adjust, p_adjust_mean, mean_log2fc, core_enrichment, set_size) %>%
      inner_join(gsea_results %>%
                   filter(organism == "Mouse",
                          pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
                   select(pathogen, timepoint) %>%
                   distinct(),
                 by = "pathogen")
  ) %>%
  distinct()

#Check
gsea_btm_results_all_matched %>% 
  dplyr::count(organism, pathogen, timepoint)

# Pivot to wide format: one row per module per timepoint with Human and Mouse side by side --------
dge_btm_gsea_mean_wide <- gsea_btm_results_all_matched %>%
  select(organism, pathogen, timepoint, process, group, set_size,
         nes, mean_log2fc, p_adjust, p_adjust_mean) %>%
  group_by(pathogen, process, group, set_size) %>%
  pivot_wider(names_from = organism,
              values_from = c(mean_log2fc, nes, p_adjust, p_adjust_mean, set_size),
              names_sep = "_")

#Save
saveRDS(dge_btm_gsea_mean_wide, here("tables", "all_human_mouse_dge_btm_gsea_mean_wide.rds"))
saveRDS(gsea_btm_results_all_matched, here("tables", "all_human_mouse_gsea_btm_results_all_matched.rds"))
```

### BTM distribution plots (one per condition)

Gene-level distribution plot for each condition: individual gene log2FC
values as points, module mean as a filled circle, standard error bars,
faceted by immune group and timepoint. BTMs with adj.p \<= 0.10 in both
species are highlighted in bold blue on the Y-axis.

```{r fig.height=12, fig.width=10}
walk(seq_len(nrow(conditions)), \(i) {
  row <- conditions[i, ]

  # Filter unified data to this condition --------
  cond_btm_genes <- dge_btm_process_genes %>%
    filter(pathogen == row$label)

  cond_btm_mean <- dge_btm_mean_process %>%
    filter(pathogen == row$label)

  cond_gsea <- gsea_results %>%
    semi_join(cond_btm_mean %>% distinct(condition), by = "condition")

  # Parse organism from condition label (stored as "Human_Influenza_1") --------
  cond_gsea <- cond_gsea %>%
    mutate(organism = case_when(
      str_detect(condition, "Mouse") ~ "Mouse",
      str_detect(condition, "Human") ~ "Human",
      TRUE ~ condition
    ))

  # Pivot GSEA results to wide format (Human and Mouse side by side) --------
  cond_gsea_wide <- cond_gsea %>%
    select(organism, timepoint, process, group, subgroup, set_size,
           nes, mean_log2fc, p_adjust, qvalue) %>%
    group_by(timepoint, process, subgroup, group, set_size) %>%
    pivot_wider(names_from = organism,
                values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
                names_sep = "_")

  # Determine ordering timepoint: last timepoint for this condition --------
  if (is.numeric(cond_btm_mean$timepoint)) {
    order_tp <- max(cond_btm_mean$timepoint, na.rm = TRUE)
  } else {
    tp_levels <- c("2h", "4h", "6h", "12h", "24h", "Early hours")
    available <- intersect(tp_levels, unique(cond_btm_mean$timepoint))
    order_tp <- if (length(available) > 0) dplyr::last(available) else unique(cond_btm_mean$timepoint)[1]
  }

  # Order BTMs by Human mean log2FC at the ordering timepoint, per immune group --------
  process_order <- cond_btm_mean %>%
    filter(timepoint == order_tp, organism == "Human") %>%
    group_by(organism, timepoint, group) %>%
    arrange(-mean_log2fc) %>%
    pull(process) %>%
    unique()

  # Create label styling: BTMs significant in both species get bold + blue --------
  labels_process <- cond_gsea_wide %>%
    ungroup() %>%
    distinct(process, p_adjust_Human, p_adjust_Mouse) %>%
    mutate(
      label = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10,
                     "bold", "plain"),
      label_color = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10,
                           "#4DBBD5FF", "gray50")
    ) %>%
    ungroup() %>%
    distinct() %>%
    mutate(process = fct_relevel(process, process_order)) %>%
    filter(!is.na(p_adjust_Human))

  # Factor releveling for display order --------
  dge_btm_process_ordered <- cond_btm_mean %>%
    ungroup() %>%
    mutate(
      group = fct_relevel(as.factor(group), immune_order),
      process = fct_relevel(as.factor(process), process_order) %>% fct_rev()
    )

  # Format timepoint labels for the plot --------
  if (is.numeric(cond_btm_genes$timepoint)) {
    dge_plot <- cond_btm_genes %>%
      mutate(
        timepoint_num = timepoint,
        timepoint = paste0("Day ", timepoint_num),
        timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))
      )
  } else {
    dge_plot <- cond_btm_genes %>%
      mutate(
        timepoint = factor(timepoint, levels = unique(timepoint))
      )
  }

  # Data preparation: set up ggplot with aesthetic mappings --------
  p <- dge_plot %>%
    ggplot() +
    aes(y = process, x = mean_l2fc, color = organism) +
    # Facetting: rows = immune group, columns = timepoint --------
    facet_grid(~group ~ timepoint, scales = "free", space = "free") +
    # Geom layers: reference line, individual gene points, module mean with SE --------
    geom_vline(xintercept = 0, color = "black") +
    geom_point(pch = "|", alpha = 1, show.legend = FALSE, size = 3) +
    stat_summary(
      aes(color = organism),
      fun.data = mean_se,
      geom = "errorbar",
      width = 0.5,
      size = 1
    ) +
    stat_summary(
      aes(fill = organism),
      fun = mean,
      color = "black",
      geom = "point",
      size = 2,
      pch = 21,
      alpha = 1,
      show.legend = TRUE
    ) +
    # Theme customization: vaxgo theme with significance-styled axis text --------
    theme_vaxgo() +
    theme(
      strip.text.y = element_text(angle = 0, hjust = 0),
      axis.ticks.y = element_line(color = "gray", alpha = 0.5),
      legend.position = "top",
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(
        size = 8,
        face = c(labels_process$label),
        color = c(labels_process$label_color)
      ),
      axis.line.y = element_blank(),
      panel.spacing.y = unit(0.3, "lines"),
      panel.grid.major.y = element_line(alpha = 0.5),
      panel.grid.major.x = element_line(alpha = 0.5)
    ) +
    # Colour scales: project colours for Human and Mouse --------
    scale_fill_manual(values = c("Mouse" = "#4DBBD5FF", "Human" = "#90A4AEFF")) +
    scale_color_manual(values = c("Mouse" = "#4DBBD5FF", "Human" = "#90A4AEFF")) +
    # Labels --------
    labs(
      color = "Organism",
      y = "",
      x = "Log2FC",
      fill = "Organism",
      title = glue("Mouse vs Human, BTMs genes, {row$name}")
    )

  print(p)

  #Save
  p %>%
    ggsave(
      filename = here("Figures", paste0(row$filename_prefix, "_dge_btm_process_distribution.png")),
      height = 12,
      width = 10
    )
})
```

## MSigDB Hallmarks

### t-Test

Join DEGs with MSigDB Hallmark annotation, collapse to Hallmark-level
means, test each Hallmark against zero, and save the unified output.

```{r}
# Annotate each DEG with its Hallmark membership --------
dge_hallmarks_process_genes <- all_degs %>%
  rename(symbol = human_symbol) %>%
  ungroup() %>%
  inner_join(msigdbr_grouped, by = "symbol")

# Collapse gene-level log2FC to Hallmark-level means per organism and timepoint --------
dge_hallmarks_mean_process <- dge_hallmarks_process_genes %>%
  group_by(organism, timepoint, process, pathogen) %>%
  summarise(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n()) %>%
  distinct() %>%
  ungroup() %>%
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint))

# One-sample t-test per Hallmark vs zero with BH correction --------
pvals_vs_zero <- dge_hallmarks_process_genes %>%
  group_by(organism, pathogen, timepoint, process) %>%
  filter(n() >= 2) %>%
  rstatix::t_test(mean_l2fc ~ 1, mu = 0) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Diagnostic: print Hallmarks with fewer than 2 genes --------
dge_hallmarks_process_genes %>%
  dplyr::count(organism, pathogen, timepoint, process) %>%
  filter(n < 2)

# Join Hallmark-level means with their t-test p-values --------
dge_hallmarks_mean_process <- dge_hallmarks_mean_process %>%
  left_join(
    pvals_vs_zero %>%
      select(organism, pathogen, timepoint, process, p, p_adjust = p.adj),
    by = c("organism", "pathogen", "timepoint", "process")
  ) %>%
  clean_names()

#Save
saveRDS(dge_hallmarks_mean_process, here("tables", "all_human_mouse_dge_hallmarks_mean_process.rds"))
saveRDS(dge_hallmarks_process_genes, here("tables", "all_human_mouse_dge_hallmarks_process_genes.rds"))
```

### Hallmarks GSEA

Run GSEA against MSigDB Hallmarks across all conditions simultaneously.

```{r}
# Prepare Hallmark term-to-gene mapping for GSEA --------
term2gene <- msigdbr_grouped %>%
  select(process, genes = symbol)

# Build ranked gene list with composite metric --------
df_genes2GSEA <- all_degs %>%
  distinct() %>%
  select(organism, pathogen, timepoint,
         genes = human_symbol, log2fold_change = mean_l2fc, adj_p_val) %>%
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
  mutate(condition = paste0(organism, "_", pathogen, "_", timepoint),
         log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>%
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA against Hallmarks using autoGSEA --------
gsea_output <- autoGSEA(df = df_genes2GSEA,
                         TERM2GENE = term2gene,
                         geneset_name = "MSigDB Hallmarks")

# Clean GSEA output and merge with Hallmark-level means --------
gsea_results_hallmarks <- gsea_output$gsea %>%
  as.data.frame() %>%
  clean_names() %>%
  rename(process = id) %>%
  filter(!is.na(process)) %>%
  full_join(dge_hallmarks_mean_process %>% rename(p_adjust_mean = p_adjust),
            by = join_by("condition", "process"))

#Save
saveRDS(gsea_results_hallmarks, here("tables", "all_human_mouse_gsea_hallmarks_results.rds"))
```

### Hallmarks distribution plots (per condition)

Gene-level distribution plot for conditions where
`hallmarks_plot = TRUE` (currently Engerix B only). Shows individual
gene points, module mean, and standard error bars, faceted by timepoint
and Hallmark group.

```{r fig.height=12, fig.width=10}
walk(seq_len(nrow(conditions)), \(i) {
  row <- conditions[i, ]
  if (!row$hallmarks_plot) return()

  # Filter unified data to this condition --------
  cond_hall_genes <- dge_hallmarks_process_genes %>%
    filter(pathogen == row$label)

  cond_hall_mean <- dge_hallmarks_mean_process %>%
    filter(pathogen == row$label)

  cond_gsea <- gsea_results_hallmarks %>%
    semi_join(cond_hall_mean %>% distinct(condition), by = "condition")

  # Pivot GSEA to wide format --------
  cond_gsea_wide <- cond_gsea %>%
    select(organism, timepoint, process, set_size, nes, mean_log2fc, p_adjust, qvalue) %>%
    group_by(timepoint, process, set_size) %>%
    pivot_wider(names_from = organism,
                values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
                names_sep = "_")

  # Determine ordering timepoint --------
  if (is.numeric(cond_hall_mean$timepoint)) {
    order_tp <- max(cond_hall_mean$timepoint, na.rm = TRUE)
  } else {
    tp_levels <- c("2h", "4h", "6h", "12h", "24h", "Early hours")
    available <- intersect(tp_levels, unique(cond_hall_mean$timepoint))
    order_tp <- if (length(available) > 0) dplyr::last(available) else unique(cond_hall_mean$timepoint)[1]
  }

  # Order Hallmarks by Human mean log2FC --------
  process_order <- cond_hall_mean %>%
    filter(timepoint == order_tp, organism == "Human") %>%
    group_by(organism, timepoint) %>%
    arrange(-mean_log2fc) %>%
    pull(process) %>%
    unique()

  # Format timepoint labels for the plot --------
  if (is.numeric(cond_hall_genes$timepoint)) {
    dge_plot <- cond_hall_genes %>%
      inner_join(msigdbr_grouped %>% select(process, group), by = "process") %>%
      mutate(
        timepoint_num = timepoint,
        timepoint = paste0("Day ", timepoint_num),
        timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))
      )
  } else {
    dge_plot <- cond_hall_genes %>%
      inner_join(msigdbr_grouped %>% select(process, group), by = "process") %>%
      mutate(
        timepoint = factor(timepoint, levels = unique(timepoint))
      )
  }

  # Data preparation: set up ggplot with aesthetic mappings --------
  p <- dge_plot %>%
    ggplot() +
    aes(y = process, x = mean_l2fc, color = organism) +
    # Facetting: columns = timepoint, rows = Hallmark group --------
    facet_grid(~timepoint ~ group, scales = "free", space = "free") +
    # Geom layers --------
    geom_vline(xintercept = 0, color = "black") +
    geom_point(pch = "|", alpha = 1, show.legend = FALSE, size = 3) +
    stat_summary(
      aes(color = organism),
      fun.data = mean_se,
      geom = "errorbar",
      width = 0.5,
      size = 1
    ) +
    stat_summary(
      aes(fill = organism),
      fun = mean,
      color = "black",
      geom = "point",
      size = 2,
      pch = 21,
      alpha = 1,
      show.legend = TRUE
    ) +
    # Theme customization --------
    theme_vaxgo() +
    theme(
      strip.text.y = element_text(angle = 0, hjust = 0),
      axis.ticks.y = element_line(color = "gray", alpha = 0.5),
      legend.position = "top",
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8),
      axis.line.y = element_blank(),
      panel.spacing.y = unit(0.3, "lines"),
      panel.grid.major.y = element_line(alpha = 0.5),
      panel.grid.major.x = element_line(alpha = 0.5)
    ) +
    # Colour scales --------
    scale_fill_manual(values = c("Mouse" = "#4DBBD5FF", "Human" = "#90A4AEFF")) +
    scale_color_manual(values = c("Mouse" = "#4DBBD5FF", "Human" = "#90A4AEFF")) +
    # Labels --------
    labs(
      color = "Organism",
      y = "",
      x = "Log2FC",
      fill = "Organism",
      title = glue("Mouse vs Human, MSigDB genes, mean Log2FC — {row$name}")
    )

  print(p)

  #Save
  p %>%
    ggsave(
      filename = here("Figures", paste0(row$filename_prefix, "_dge_hallmarks_process_distribution.png")),
      height = 12,
      width = 10
    )
})
```

