---
title: "Comparing Human vs Mouse — Functional analysis"
author: Wasim Aluísio Prates-Syed
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Setup

```{r message=FALSE, warning=FALSE}
# Load core packages for the analysis session
library(here)
# Source project-wide functions, themes, and colour palettes
source(here("scripts_notebooks", "required.R"))


```

# Functional enrichment

Unified pipeline for all 6 conditions (Fluad, Engerix B, S. aureus, E.
coli,Burn, Trauma). All conditions are loaded from the pre-merged DGE
file, then each analysis (BTM means, BTM GSEA, Hallmarks means,
Hallmarks GSEA) runs once on the combined data and produces a single
output file.

## Import all conditions

Load the pre-merged DGE file that already contains all 6 conditions with
consistent column names and pathogen labels.

```{r}
# Import annotation tables for BTMs and Hallmark gene sets
btm_annotation_genes <- read_csv(here("tables", "btm_annotation_genes.csv"))
msigdbr_grouped <- read_csv(here("tables", "msigdb_hallmarks_grouped_genes.csv"))

#Load DGE results
all_degs <- readRDS(here("tables", "all_human_mouse_dge_limma_degs.rds"))

# Extract unique condition labels 
annotation_conditions <- all_degs %>%
  select(condition, timepoint, model = organism) %>%
  distinct()

#Filter timepoints combinations
conditions_filtered <- readRDS(file = here("tables", "all_human_mouse_correlation_filtered_inter_weighted_filtered.rds")) %>% 
  mutate(pathogen_human = case_when(pathogen_human == "Hepatitis B" ~ "HepB",
                              pathogen_human == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen_human == "Escherichia coli" ~ "E. coli",
                              T ~pathogen_human),
          pathogen_human = fct_relevel(pathogen_human, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  select(treatment_human, treatment_mouse, pathogen_human, pathogen_mouse,
         pathogen_comparison, timepoint_comparison,
         timepoint_human, timepoint_mouse) %>% 
  distinct() %>% 
  #Define control
  select(pathogen = pathogen_human, timepoint_comparison) %>% 
  distinct()

# ── Load BTM (Blood Transcriptional Modules) results
# Load mean log2FC per BTM process, collapsed across genes within each module
dge_btm_mean_process <- readRDS(here("tables", "all_human_mouse_dge_btm_mean_process.rds")) 

# Load gene-level BTM membership
dge_btm_process_genes <- readRDS(here("tables", "all_human_mouse_dge_btm_process_genes.rds"))

dge_btm_process_genes_wide = readRDS(here("tables", "all_human_mouse_dge_btm_process_genes_wide.rds")) %>% 
  inner_join(conditions_filtered,
             by = join_by(pathogen, timepoint_comparison))

# Load pre-computed GSEA results
gsea_results <- readRDS(here("tables", "all_human_mouse_gsea_btm_results.rds"))

# Load BTM GSEA + matched results
dge_btm_gsea_mean_wide <- readRDS(here("tables", "all_human_mouse_dge_btm_gsea_mean_wide.rds")) %>% 
  inner_join(conditions_filtered,
             by = join_by(pathogen, timepoint_comparison))

# Matched GSEA results where human and mouse timepoints have been aligned for direct comparison
gsea_btm_results_all_matched <- readRDS(here("tables", "all_human_mouse_gsea_btm_results_matched.rds")) %>% 
  inner_join(conditions_filtered,
             by = join_by(pathogen, timepoint_comparison))



```

# =========================

# Functional analysis

## Condition-wise analysis

### BTMs

BTMs correlation, intra- and inter-species Compare means and NES

#### Module-level, BTMs

Prepare the wide-format table of BTM mean log2FC and NES values for
human vs mouse comparison, then compute Pearson correlations between
organisms under three significance filters: all modules, modules
significant in mouse only, and modules significant in both. Assemble a
combined data frame with facet-ready labels for downstream plotting.



##### Mean Log2-FC

###### Correlation

####### Condition vs. Condition
```{r fig.height=5, fig.width=7}
# Correlation between modules ==================

# Conditions (No-control)
# Compute correlation between human and mouse mean log2FC across all modules (no filtering)
cor_labels =      
  dge_btm_gsea_mean_wide %>%
  calculate_weighted_correlation_pearson(
    x_col = mean_log2fc_Human,
    y_col = mean_log2fc_Mouse,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  )

# Compute correlation restricted to modules significant in mouse (p_adjust <= 0.25)
cor_labels_Mouse <- dge_btm_gsea_mean_wide %>%
  filter(p_adjust_mean_Mouse <= 0.25) %>%
  calculate_weighted_correlation_pearson(
    x_col = mean_log2fc_Human,
    y_col = mean_log2fc_Mouse,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  ) 


# Compute correlation restricted to modules significant in both human and mouse
cor_labels_Mouse_Human <- dge_btm_gsea_mean_wide %>%
  filter(p_adjust_mean_Mouse <= 0.25, p_adjust_mean_Human <= 0.25) %>%
  calculate_weighted_correlation_pearson(
    x_col = mean_log2fc_Human,
    y_col = mean_log2fc_Mouse,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  )

#Bind to one df
cor_labels_mean_long = bind_rows(cor_labels %>% mutate(filter = "All"), 
                                cor_labels_Mouse %>% mutate(filter = "Mouse"), 
                                cor_labels_Mouse_Human %>% mutate(filter = "Mouse and Human")) %>% 
  #Refactor
  mutate(pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c("S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c("Burn", "Trauma") ~ "Injury",
                               TRUE ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"))

# Join GSEA table with all three correlation summaries
dge_btm_mean_group_correlation_df <- dge_btm_gsea_mean_wide %>%
  ungroup() %>%
  
  # Reorder module groups and wrap long process names for display
  mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>%
  drop_na(mean_log2fc_Mouse) %>%
  # Attach the three sets of correlation labels by (pathogen, timepoint_comparison
  inner_join(cor_labels, by = join_by(pathogen, timepoint_comparison)) %>%
  inner_join(cor_labels_Mouse %>% 
                dplyr::rename_with(
                .fn = ~ stringr::str_c(.x, "_filter_Mouse"),
                .cols = cor_value:cor_label
              ), by = join_by(pathogen, timepoint_comparison)) %>%
  inner_join(cor_labels_Mouse_Human %>% 
                dplyr::rename_with(
                .fn = ~ stringr::str_c(.x, "_filter_Mouse_Human"),
                .cols = cor_value:cor_label
              ), by = join_by(pathogen, timepoint_comparison)) %>%
  
  
  # Combine the three correlation labels into a single facet annotation
  mutate(cor_labels_all_filtered = paste0("All: ", cor_label, ",\nM: ", cor_label_filter_Mouse, ",\nM&H: ", cor_label_filter_Mouse_Human),
         facet_name = paste0(timepoint_comparison, "\n", cor_labels_all_filtered),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("Hep B", "Influenza") ~ "Vaccination",
                               pathogen %in% c("S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c("Burn", "Trauma") ~ "Injury",
                               TRUE ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"))

# Check
dge_btm_mean_group_correlation_df %>%
  dplyr::count(pathogen, timepoint_comparison)

```

Conditions vs. Control

```{r}

# Conditions vs. Control =======
# Compute correlation between human and mouse mean log2FC across all modules (no filtering)
cor_labels_control =      
  dge_btm_gsea_mean_wide %>%
  calculate_weighted_correlation_pearson(
    x_col = mean_log2fc_Human,
    y_col = mean_log2fc_Control,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  )

# Compute correlation restricted to modules significant in mouse (p_adjust <= 0.25)
cor_labels_Mouse_control <- dge_btm_gsea_mean_wide %>%
  filter(p_adjust_mean_Mouse <= 0.25) %>%
  calculate_weighted_correlation_pearson(
    x_col = mean_log2fc_Human,
    y_col = mean_log2fc_Control,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  ) 


# Compute correlation restricted to modules significant in both human and mouse
cor_labels_Mouse_Human_control <- dge_btm_gsea_mean_wide %>%
  filter(p_adjust_mean_Mouse <= 0.25, p_adjust_mean_Human <= 0.25) %>%
  calculate_weighted_correlation_pearson(
    x_col = mean_log2fc_Human,
    y_col = mean_log2fc_Control,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  )

#Bind to one df
cor_labels_mean_long_control = bind_rows(cor_labels_control %>% mutate(filter = "All"), 
                                cor_labels_Mouse_control %>% mutate(filter = "Mouse"), 
                                cor_labels_Mouse_Human_control %>% mutate(filter = "Mouse and Human")) %>% 
  #Refactor
  mutate(pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c("S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c("Burn", "Trauma") ~ "Injury",
                               TRUE ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"))



# Join GSEA table with all three correlation summaries
dge_btm_mean_group_correlation_df_control <- dge_btm_gsea_mean_wide %>%
  ungroup() %>%
  
  # Reorder module groups and wrap long process names for display
  mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>%
  drop_na(mean_log2fc_Control) %>%
  # Attach the three sets of correlation labels by (pathogen, timepoint_comparison
  inner_join(cor_labels_control, by = join_by(pathogen, timepoint_comparison)) %>%
  inner_join(cor_labels_Mouse_control %>% 
                dplyr::rename_with(
                .fn = ~ stringr::str_c(.x, "_filter_Mouse"),
                .cols = cor_value:cor_label
              ), by = join_by(pathogen, timepoint_comparison)) %>%
  inner_join(cor_labels_Mouse_Human_control %>% 
                dplyr::rename_with(
                .fn = ~ stringr::str_c(.x, "_filter_Mouse_Human"),
                .cols = cor_value:cor_label
              ), by = join_by(pathogen, timepoint_comparison)) %>%
  
  
  # Combine the three correlation labels into a single facet annotation
  mutate(cor_labels_all_filtered = paste0("All: ", cor_label, ",\nM: ", cor_label_filter_Mouse, ",\nM&H: ", cor_label_filter_Mouse_Human),
         facet_name = paste0(timepoint_comparison, "\n", cor_labels_all_filtered),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("Hep B", "Influenza") ~ "Vaccination",
                               pathogen %in% c("S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c("Burn", "Trauma") ~ "Injury",
                               TRUE ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"))

# Check
dge_btm_mean_group_correlation_df_control %>%
  dplyr::count(pathogen, timepoint_comparison)
```



###### Heatmap of summary results
```{r fig.height=3, fig.width=14}

#Merge to one DF
cor_labels_mean_long_condition_control = bind_rows(cor_labels_mean_long,
                                                  cor_labels_mean_long_control %>% 
                                                    mutate(timepoint_comparison = "Control"))


# Data preparation
correlation_long_all_btm_mouse_human_timepoints_summary_heatmap_mean = cor_labels_mean_long_condition_control %>%
  mutate(filter = str_replace(filter, "Mouse and Human", "Mouse\nand\nHuman"),
         timepoint_comparison = fct_relevel(timepoint_comparison, "Control")) %>% 
  # Aesthetics
  ggplot() +
  aes(x = timepoint_comparison,
      y = filter %>% fct_rev(),
      fill = cor_value) +
  # Facets
  facet_nested_wrap(~treatment+pathogen,
                    scales = "free_x",
                    nrow = 1) +
  # Geoms
  geom_point(aes(size = n_obs),
             pch = 21) +
  geom_text(aes(label = paste0("r=",round(cor_value,2), ", R²=", round(r_squared,2))),
            color = "black",
            vjust = -2.5, fontface = "plain",
            size = 3) +
  geom_text(aes(label = paste0("n=", n_obs)),
            color = "black",
            vjust = 0.5, fontface = "plain",
            hjust = -0.5,
            size = 3)+
  # Scales
  scale_size_continuous(range = c(2, 10)) +
    scale_fill_gradientn(
    colors = c("#4DBBD5FF", "white", "#4361ee"),
    values = scales::rescale(c(-1, 0, 1)),
    limits = c(-1, 1),
    name = "Correlation",
    guide = ggplot2::guide_colorbar(
      barwidth = 10,
      barheight = 1,
      nbin = 10,
      raster = FALSE
    )
  ) +
  # Labels
  labs(
    x = "Time",
    y = "Filter",
    color = "Correlation",
    size = "N genes") +
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0, size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.y = element_blank(),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        panel.spacing = unit(1, "lines"),
        plot.margin = margin(0, 0.1, 0, 0.1, unit = "cm"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, -10, 0),
        legend.text = element_text(size = 10, hjust = 0.5),
        legend.title = element_text(size = 10),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

# Render plot
correlation_long_all_btm_mouse_human_timepoints_summary_heatmap_mean 
   
 

# Save
correlation_long_all_btm_mouse_human_timepoints_summary_heatmap_mean %>%
  ggsave(filename = here("Figures",  "all_human_mouse_mean_btm_correlation_all_timepoints_summary_heatmap_filtered.png"),
         width = 14,
         height = 4)
```

####### Scatter plot 
```{r fig.height=4, fig.width=14, message=FALSE, warning=FALSE}
# Data preparation
dge_btm_mean_group_correlation_df <-
  bind_rows(
    # All modules (no filtering)
    dge_btm_mean_group_correlation_df %>%
      mutate(filter = "All"),
    # Modules significant in mouse only (padj <= 0.25 in mouse, > 0.25 in human)
    dge_btm_mean_group_correlation_df %>%
      filter(p_adjust_mean_Mouse <= 0.25, p_adjust_mean_Human > 0.25) %>%
      mutate(filter = "Mouse"),
    # Modules significant in both organisms
    dge_btm_mean_group_correlation_df %>%
      filter(p_adjust_mean_Mouse <= 0.25 & p_adjust_mean_Human <= 0.25) %>%
      mutate(filter = "Mouse and Human"),
  )

# Scatterplot
dge_btm_mean_group_correlation_plot <- dge_btm_mean_group_correlation_df %>%
  ggplot() +
  aes(x = mean_log2fc_Mouse,
      y = mean_log2fc_Human) +
  facet_nested_wrap(vars(pathogen, facet_name),
                    scales = "free",
                    nrow = 1,
                    strip = strip_nested(size = "variable")) +

  # Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +

  # Geom layer: all modules shown with blur for background density
  with_blur(
    geom_point(aes(fill = group, color = group, shape = filter),
               data = . %>% filter(filter == "All"),
               size = 2, alpha = 0.2, stroke = 0.1, sigma = 2)
  ) +

  # Geom layer: mouse-only significant modules drawn with full opacity
  geom_point(aes(fill = group, shape = filter, color = group),
             data = . %>% filter(filter == "Mouse"),
             size = 2, stroke = 0.1, alpha = 1) +

  # Geom layer: both-significant modules drawn with black outline for emphasis
  geom_point(aes(fill = group, shape = filter),
             color = "black",
             data = . %>% filter(filter == "Mouse and Human"),
             size = 2, stroke = 0.1, alpha = 1) +

  # Scales
  scale_shape_manual(values = c(21, 22, 24)) +
  scale_fill_manual(values = colors_genesets$immune_colors) +
  scale_color_manual(values = colors_genesets$immune_colors) +
  guides(color = FALSE,
         fill = guide_legend(byrow = TRUE, nrow = 3,
                             keywidth = unit(0.3, "cm"),
                             keyheight = unit(0.3, "cm"),
                             override.aes = list(size = 3, shape = 21, color = "black"),
                             default.unit = "cm"),
         shape = guide_legend(byrow = TRUE, nrow = 3,
                              keywidth = unit(0.3, "cm"),
                              keyheight = unit(0.3, "cm"),
                              override.aes = list(size = 3),
                              default.unit = "cm")) +

  # Labels
  labs(fill = "Group",
       shape = "Significance",
       y = "Human (Mean Log2FC)",
       x = "Mouse (Mean Log2FC)") +

  # Theme customization
    theme_vaxgo() +
  theme(
        strip.text.x = element_text(hjust = 0),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.text = element_text(hjust = 0),
        panel.spacing = unit(0, "lines"))

# Render plot
dge_btm_mean_group_correlation_plot

# Save
dge_btm_mean_group_correlation_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_btm_mean_group_correlation_plot_filtered.png"),
         width = 14, height = 4)
```






###### Conditions vs. Control

####### Scatter plot 
```{r fig.height=4, fig.width=14, message=FALSE, warning=FALSE}
# Data preparation
dge_btm_mean_group_correlation_df_control_plot <-
  bind_rows(
    # All modules (no filtering)
    dge_btm_mean_group_correlation_df_control %>%
      mutate(filter = "All"),
    # Modules significant in mouse only (padj <= 0.25 in mouse, > 0.25 in human)
    dge_btm_mean_group_correlation_df_control %>%
      filter(p_adjust_mean_Mouse <= 0.25, p_adjust_mean_Human > 0.25) %>%
      mutate(filter = "Mouse"),
    # Modules significant in both organisms
    dge_btm_mean_group_correlation_df_control %>%
      filter(p_adjust_mean_Mouse <= 0.25 & p_adjust_mean_Human <= 0.25) %>%
      mutate(filter = "Mouse and Human"),
  )

# Scatterplot
dge_btm_mean_group_correlation_plot_control <- dge_btm_mean_group_correlation_df_control_plot %>%
  ggplot() +
  aes(x = mean_log2fc_Mouse,
      y = mean_log2fc_Control) +
  facet_nested_wrap(vars(pathogen, facet_name),
                    scales = "free",
                    nrow = 1,
                    strip = strip_nested(size = "variable")) +

  # Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +

  # Geom layer: all modules shown with blur for background density
  with_blur(
    geom_point(aes(fill = group, color = group, shape = filter),
               data = . %>% filter(filter == "All"),
               size = 2, alpha = 0.2, stroke = 0.1, sigma = 2)
  ) +

  # Geom layer: mouse-only significant modules drawn with full opacity
  geom_point(aes(fill = group, shape = filter, color = group),
             data = . %>% filter(filter == "Mouse"),
             size = 2, stroke = 0.1, alpha = 1) +

  # Geom layer: both-significant modules drawn with black outline for emphasis
  geom_point(aes(fill = group, shape = filter),
             color = "black",
             data = . %>% filter(filter == "Mouse and Human"),
             size = 2, stroke = 0.1, alpha = 1) +

  # Scales
  scale_shape_manual(values = c(21, 22, 24)) +
  scale_fill_manual(values = colors_genesets$immune_colors) +
  scale_color_manual(values = colors_genesets$immune_colors) +
  guides(color = FALSE,
         fill = guide_legend(byrow = TRUE, nrow = 3,
                             keywidth = unit(0.3, "cm"),
                             keyheight = unit(0.3, "cm"),
                             override.aes = list(size = 3, shape = 21, color = "black"),
                             default.unit = "cm"),
         shape = guide_legend(byrow = TRUE, nrow = 3,
                              keywidth = unit(0.3, "cm"),
                              keyheight = unit(0.3, "cm"),
                              override.aes = list(size = 3),
                              default.unit = "cm")) +

  # Labels
  labs(fill = "Group",
       shape = "Significance",
       y = "Human (Mean Log2FC)",
       x = "Control (Mean Log2FC)") +

  # Theme customization
    theme_vaxgo() +
  theme(
        strip.text.x = element_text(hjust = 0),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.text = element_text(hjust = 0),
        panel.spacing = unit(0, "lines"))

# Render plot
dge_btm_mean_group_correlation_plot_control

# Save
dge_btm_mean_group_correlation_plot_control %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_btm_mean_group_correlation_plot_filtered_Control.png"),
         width = 14, height = 4)
```










##### Plot NES

Compute Pearson correlations between Human and Mouse NES values across
BTMs, stratified by pathogen and timepoint, for all genes and for
significance-filtered subsets.

Condition vs. Condition
```{r fig.height=5, fig.width=7}
# Correlation for all modules
cor_labels_nes <- dge_btm_gsea_mean_wide %>%
  calculate_weighted_correlation_pearson(
    x_col = nes_Mouse,
    y_col = nes_Human,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  ) 

# Correlation for Mouse-significant modules
cor_labels_nes_Mouse <- dge_btm_gsea_mean_wide %>%
  filter(p_adjust_Mouse <= 0.25) %>% 
  calculate_weighted_correlation_pearson(
    x_col = nes_Mouse,
    y_col = nes_Human,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  )

# Correlation for both Mouse and Human significant modules 
cor_labels_nes_Mouse_Human <- dge_btm_gsea_mean_wide %>%
  filter(p_adjust_Mouse <= 0.25, p_adjust_Human <= 0.25) %>% 
  calculate_weighted_correlation_pearson(
    x_col = nes_Mouse,
    y_col = nes_Human,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  ) 

#Bind to one df
cor_labels_nes_long = bind_rows(cor_labels_nes %>% mutate(filter = "All"), 
                                cor_labels_nes_Mouse %>% mutate(filter = "Mouse"), 
                                cor_labels_nes_Mouse_Human %>% mutate(filter = "Mouse and Human")) %>% 
  #Refactor
  mutate(pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c("S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c("Burn", "Trauma") ~ "Injury",
                               TRUE ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"))



# Build master correlation data frame with scores and annotations
dge_btm_nes_group_correlation_df = dge_btm_gsea_mean_wide %>%
  ungroup() %>%
  # Reorder module groups and wrap long process names for display
  mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>%
  drop_na(mean_log2fc_Mouse) %>%
  
  # Attach the three sets of correlation labels by (pathogen, timepoint_comparison
  inner_join(cor_labels_nes, by = join_by(pathogen, timepoint_comparison)) %>%
  inner_join(cor_labels_nes_Mouse %>% 
              dplyr::rename_with(
                .fn = ~ stringr::str_c(.x, "_filter_Mouse"),
                .cols = cor_value:cor_label
              ), by = join_by(pathogen, timepoint_comparison)) %>%
  inner_join(cor_labels_nes_Mouse_Human %>% 
              dplyr::rename_with(
                .fn = ~ stringr::str_c(.x, "_filter_Mouse_Human"),
                .cols = cor_value:cor_label
              ), by = join_by(pathogen, timepoint_comparison)) %>%
  
  
  # Combine the three correlation labels into a single facet annotation
  mutate(cor_labels_all_filtered = paste0("All: ", cor_label, ",\nM: ", cor_label_filter_Mouse, ",\nM&H: ", cor_label_filter_Mouse_Human),
         facet_name = paste0(timepoint_comparison, "\n", cor_labels_all_filtered),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c("S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c("Burn", "Trauma") ~ "Injury",
                               TRUE ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"))

dge_btm_nes_group_correlation_df %>% dplyr::count(pathogen, timepoint_comparison)

```


Condition vs. Control
```{r fig.height=5, fig.width=7}
# Correlation for all modules
cor_labels_nes_control <- dge_btm_gsea_mean_wide %>%
  calculate_weighted_correlation_pearson(
    x_col = nes_Control,
    y_col = nes_Human,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  ) 

# Correlation for Mouse-significant modules
cor_labels_nes_Mouse_control <- dge_btm_gsea_mean_wide %>%
  filter(p_adjust_Control <= 0.25) %>% 
  calculate_weighted_correlation_pearson(
    x_col = nes_Control,
    y_col = nes_Human,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  )

# Correlation for both Mouse and Human significant modules 
cor_labels_nes_Mouse_Human_control <- dge_btm_gsea_mean_wide %>%
  filter(p_adjust_Control <= 0.25, p_adjust_Human <= 0.25) %>% 
  calculate_weighted_correlation_pearson(
    x_col = nes_Control,
    y_col = nes_Human,
    weight_col = weight_joint_gsea_fdr,
    group_cols = c("pathogen", "timepoint_comparison")
  ) 

#Bind to one df
cor_labels_nes_long_control = bind_rows(cor_labels_nes_control %>% mutate(filter = "All"), 
                                cor_labels_nes_Mouse_control %>% mutate(filter = "Mouse"), 
                                cor_labels_nes_Mouse_Human_control %>% mutate(filter = "Mouse and Human")) %>% 
  #Refactor
  mutate(pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c("S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c("Burn", "Trauma") ~ "Injury",
                               TRUE ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"))


# Build master correlation data frame with scores and annotations
dge_btm_nes_group_correlation_df_control = dge_btm_gsea_mean_wide %>%
  ungroup() %>%
  # Reorder module groups and wrap long process names for display
  mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>%
  drop_na(mean_log2fc_Mouse) %>%
  
  # Attach the three sets of correlation labels by (pathogen, timepoint_comparison
  inner_join(cor_labels_nes_control, by = join_by(pathogen, timepoint_comparison)) %>%
  inner_join(cor_labels_nes_Mouse_control %>% 
              dplyr::rename_with(
                .fn = ~ stringr::str_c(.x, "_filter_Mouse"),
                .cols = cor_value:cor_label
              ), by = join_by(pathogen, timepoint_comparison)) %>%
  inner_join(cor_labels_nes_Mouse_Human_control %>% 
              dplyr::rename_with(
                .fn = ~ stringr::str_c(.x, "_filter_Mouse_Human"),
                .cols = cor_value:cor_label
              ), by = join_by(pathogen, timepoint_comparison)) %>%
  
  
  # Combine the three correlation labels into a single facet annotation
  mutate(cor_labels_all_filtered = paste0("All: ", cor_label, ",\nM: ", cor_label_filter_Mouse, ",\nM&H: ", cor_label_filter_Mouse_Human),
         facet_name = paste0(timepoint_comparison, "\n", cor_labels_all_filtered),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c("S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c("Burn", "Trauma") ~ "Injury",
                               TRUE ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"))

dge_btm_nes_group_correlation_df_control %>% dplyr::count(pathogen, timepoint_comparison)

```


###### Heatmap of summary results

```{r fig.height=3, fig.width=12}

#Merge to one DF
cor_labels_nes_long_condition_control = bind_rows(cor_labels_nes_long,
                                                  cor_labels_nes_long_control %>% 
                                                    mutate(timepoint_comparison = "Control")) %>% 
  filter(treatment %in% c("Infection", "Injury"))


# Data preparation
correlation_long_all_btm_mouse_human_timepoints_summary_heatmap = cor_labels_nes_long_condition_control %>%
  mutate(filter = str_replace(filter, "Mouse and Human", "Mouse\nand\nHuman"),
         timepoint_comparison = fct_relevel(timepoint_comparison, "Control")) %>% 
  # Aesthetics
  ggplot() +
  aes(x = timepoint_comparison,
      y = filter %>% fct_rev(),
      fill = cor_value) +
  # Facets
  facet_nested_wrap(~treatment+pathogen,
                    scales = "free_x",
                    nrow = 1) +
  # Geoms
  geom_point(aes(size = n_obs),
             pch = 21) +
  geom_text(aes(label = paste0("r=",round(cor_value,2), ", R²=", round(r_squared,2))),
            color = "black",
            vjust = -2.5, fontface = "plain",
            size = 3) +
  geom_text(aes(label = paste0("n=", n_obs)),
            color = "black",
            vjust = 0.5, fontface = "plain",
            hjust = -0.5,
            size = 3)+
  # Scales
  scale_size_continuous(range = c(2, 10)) +
    scale_fill_gradientn(
    colors = c("#4DBBD5FF", "white", "#4361ee"),
    values = scales::rescale(c(-1, 0, 1)),
    limits = c(-1, 1),
    name = "Correlation",
    guide = ggplot2::guide_colorbar(
      barwidth = 10,
      barheight = 1,
      nbin = 10,
      raster = FALSE
    )
  ) +
  # Labels
  labs(
    x = "Time",
    y = "Filter",
    color = "Correlation",
    size = "N genes") +
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0, size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.y = element_blank(),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        panel.spacing = unit(1, "lines"),
        plot.margin = margin(0, 0.1, 0, 0.1, unit = "cm"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, -10, 0),
        legend.text = element_text(size = 10, hjust = 0.5),
        legend.title = element_text(size = 10),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

# Render plot
correlation_long_all_btm_mouse_human_timepoints_summary_heatmap 
   
 

# Save
correlation_long_all_btm_mouse_human_timepoints_summary_heatmap %>%
  ggsave(filename = here("Figures",  "all_human_mouse_NES_btm_correlation_all_timepoints_summary_heatmap_filtered.png"),
         width = 10,
         height = 3)
```



###### Barplot of summary results

```{r fig.height=3, fig.width=6}
# Data preparation
correlation_long_all_btm_mouse_human_timepoints_summary_barplot = dge_btm_nes_group_correlation_df %>%
  # Reshape 
  pivot_longer(cols = c(cor_value, cor_value_filter_Mouse, cor_value_filter_Mouse_Human),
               names_to = "correlation_filter",
               values_to = "correlation") %>%
  # Rename
  mutate(correlation_filter = case_when(correlation_filter == "cor_value" ~ "All",
                                        correlation_filter == "cor_value_filter_Mouse" ~ "Mouse",
                                        correlation_filter == "cor_value_filter_Mouse_Human" ~ "Mouse and Human",
                                        TRUE ~ correlation_filter)) %>%
  # Aesthetics
  ggplot() +
  aes(x = timepoint_comparison,
      y = correlation,
      group = correlation_filter,
      fill = correlation_filter) +
  # Geoms
  geom_line(size = 1) +
  # Geoms
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  # Facets
  facet_nested_wrap(~treatment+pathogen,
                    scales = "free_x",
                    nrow = 1) +
  # Scales
  # scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0, 0), 
                     limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
  scale_fill_manual(values = c("All" = "gray90",
                               "Mouse" = "gray50",
                               "Mouse and Human" = "#4DBBD5FF")) +
  # Labels
  labs(
    x = "Time",
    y = "Correlation",
    color = "Comparison") +
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0, size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        panel.spacing = unit(1, "lines"),
        plot.margin = margin(0, 0.5, 0, 0.5, unit = "cm"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, -10, 0),
        legend.text = element_text(size = 10, hjust = 0),
        legend.title = element_text(size = 10),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) 

# Render plot
correlation_long_all_btm_mouse_human_timepoints_summary_barplot

#Save
correlation_long_all_btm_mouse_human_timepoints_summary_barplot %>%
  ggsave(filename = here("Figures",  "all_human_mouse_NES_btm_correlation_all_timepoints_summary_barplot_filtered.png"),
         width = 6,
         height = 3)
```


######### Line plot of summary results
```{r fig.height=3, fig.width=10}

# Data preparation
correlation_long_all_btm_mouse_human_timepoints_summary_lineplot = dge_btm_nes_group_correlation_df %>%
  # Reshape 
  pivot_longer(cols = c(cor_value, cor_value_filter_Mouse, cor_value_filter_Mouse_Human),
               names_to = "correlation_filter",
               values_to = "correlation") %>%
  # Rename
  mutate(correlation_filter = case_when(correlation_filter == "cor_value" ~ "All",
                                        correlation_filter == "cor_value_filter_Mouse" ~ "Mouse",
                                        correlation_filter == "cor_value_filter_Mouse_Human" ~ "Mouse and Human",
                                        TRUE ~ correlation_filter)) %>%
  # Aesthetics
  ggplot() +
  aes(x = timepoint_comparison,
      y = correlation,
      group = correlation_filter,
      color = correlation_filter,
      fill = correlation_filter) +
  # Geoms
  geom_line(size = 1) +
  geom_point(pch = 21, fill = "white", size = 1.5, stroke = 1) +
  geom_hline(yintercept = 0, linetype = 2, color = "black") +
  # Facets
  facet_nested_wrap(~treatment+pathogen,
                    scales = "free_x",
                    nrow = 1) +
  # Scales
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.05, 0.05)) +
  scale_color_manual(values = c("All" = "gray90",
                                "Mouse" = "gray50",
                                "Mouse and Human" = "#4DBBD5FF")) +
  scale_fill_manual(values = c("All" = "gray90",
                                "Mouse" = "gray50",
                                "Mouse and Human" = "#4DBBD5FF")) +
  # Labels
  labs(
       x = "Time",
       y = "Correlation",
       color = "Comparison") +
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0, size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        panel.spacing = unit(2, "lines"),
        plot.margin = margin(0, 0.5, 0, 0.5, unit = "cm"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, -10, 0),
        legend.text = element_text(size = 10, hjust = 0),
        legend.title = element_text(size = 10),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) +
  guides(fill = "none")

# Render plot
correlation_long_all_btm_mouse_human_timepoints_summary_lineplot

# correlation_long_all_btm_mouse_human_timepoints_summary_lineplot %>% 
#   ggsave(filename = here("Figures", paste0(filename, "_NES_btm_correlation_all_timepoints_summary_lineplot.png")),
#          width = 10,
#          height = 3)
```

###### Scatter plot

```{r fig.height=4, fig.width=10}

# Data preparation: stack filter levels
dge_btm_nes_group_correlation_plot =
  bind_rows(
    # All modules
    dge_btm_nes_group_correlation_df %>%
      mutate(filter = "All"),
    # Only Mouse-significant modules
    dge_btm_nes_group_correlation_df %>%
      filter(p_adjust_Mouse <= 0.25, p_adjust_Human > 0.25) %>%
      mutate(filter = "Mouse"),
    # Both-significant modules
    dge_btm_nes_group_correlation_df %>%
      filter(p_adjust_Mouse <= 0.25 & p_adjust_Human <= 0.25) %>%
      mutate(filter = "Mouse and Human"),
  ) %>%
  # Aesthetics
  ggplot() +
  aes(x = nes_Mouse,
      y = nes_Human,
      labels = process) +
  # Facets
  facet_nested_wrap(vars(pathogen, facet_name),
                    scales = "free",
                    nrow = 1,
                    strip = strip_nested(size = "variable")) +
  # Guide lines: diagonal, anti-diagonal, axes
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +

# All points with blur for density visualisation
  with_blur(
    geom_point(aes(fill = group,
                 color = group,
                 shape = filter),
                 data = . %>% filter(filter == c("All")),
             size = 1.5,
             alpha = 0.2,
    sigma = 2)
    ) +
   # Mouse-only significant points
  geom_point(aes(fill = group,
                 shape = filter,
                 color = group),
             data = . %>% filter(filter == c("Mouse")),
             size = 2,
             stroke = 0.1,
             alpha = 1
             ) +
  # Both-significant points
  geom_point(aes(fill = group,
                 shape = filter),
             color = "black",
             data = . %>% filter(filter == c("Mouse and Human")),
             size = 2,
             stroke = 0.1,
             alpha = 1
             ) +
  # Scales
  scale_shape_manual(values = c(21, 22, 24)) +
  scale_fill_manual(values = colors_genesets$immune_colors) +
  scale_color_manual(values = colors_genesets$immune_colors) +
  # Guides
  guides(color = FALSE,
         fill = guide_legend(byrow = TRUE, nrow = 3,
                           keywidth = unit(0.3, "cm"),
                           keyheight = unit(0.3, "cm"),
                           override.aes = list(size = 3, shape = 21, color = "black"),
                           default.unit = "cm"),
         shape = guide_legend(byrow = TRUE,
                              nrow = 3,
                             keywidth = unit(0.3, "cm"),
                           keyheight = unit(0.3, "cm"),
                          override.aes = list(size = 3),
                           default.unit = "cm")) +
  # Labels
  labs(fill = "Group",
       shape = "Significance",
       y = "Human (NES)",
       x = "Mouse (NES)") +
  # Theme
  theme_vaxgo() +
  theme(
        strip.text.x = element_text(hjust = 0),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.text = element_text(hjust = 0),
        panel.spacing = unit(0, "lines"))

# Render plot
dge_btm_nes_group_correlation_plot



# Save plot
dge_btm_nes_group_correlation_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_btm_nes_group_correlation_plot_filtered.png"),
         width = 10,
         height = 4)
```


###### Conditions vs. Control
```{r fig.height=4, fig.width=10}

# Data preparation: stack filter levels
dge_btm_nes_group_correlation_control_plot =
  bind_rows(
    # All modules
    dge_btm_nes_group_correlation_df %>%
      mutate(filter = "All"),
    # Only Mouse-significant modules
    dge_btm_nes_group_correlation_df %>%
      filter(p_adjust_Control <= 0.25, p_adjust_Human > 0.25) %>%
      mutate(filter = "Control"),
    # Both-significant modules
    dge_btm_nes_group_correlation_df %>%
      filter(p_adjust_Control <= 0.25 & p_adjust_Human <= 0.25) %>%
      mutate(filter = "Control and Human"),
  ) %>%
  # Aesthetics
  ggplot() +
  aes(x = nes_Control,
      y = nes_Human,
      labels = process) +
  # Facets
  facet_nested_wrap(vars(pathogen, facet_name),
                    scales = "free",
                    nrow = 1,
                    strip = strip_nested(size = "variable")) +
  # Guide lines: diagonal, anti-diagonal, axes
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +

# All points with blur for density visualisation
  with_blur(
    geom_point(aes(fill = group,
                 color = group,
                 shape = filter),
                 data = . %>% filter(filter == c("All")),
             size = 1.5,
             alpha = 0.2,
    sigma = 2)
    ) +
   # Mouse-only significant points
  geom_point(aes(fill = group,
                 shape = filter,
                 color = group),
             data = . %>% filter(filter == c("Control")),
             size = 2,
             stroke = 0.1,
             alpha = 1
             ) +
  # Both-significant points
  geom_point(aes(fill = group,
                 shape = filter),
             color = "black",
             data = . %>% filter(filter == c("Control and Human")),
             size = 2,
             stroke = 0.1,
             alpha = 1
             ) +
  # Scales
  scale_shape_manual(values = c(21, 22, 24)) +
  scale_fill_manual(values = colors_genesets$immune_colors) +
  scale_color_manual(values = colors_genesets$immune_colors) +
  # Guides
  guides(color = FALSE,
         fill = guide_legend(byrow = TRUE, nrow = 3,
                           keywidth = unit(0.3, "cm"),
                           keyheight = unit(0.3, "cm"),
                           override.aes = list(size = 3, shape = 21, color = "black"),
                           default.unit = "cm"),
         shape = guide_legend(byrow = TRUE,
                              nrow = 3,
                             keywidth = unit(0.3, "cm"),
                           keyheight = unit(0.3, "cm"),
                          override.aes = list(size = 3),
                           default.unit = "cm")) +
  # Labels
  labs(fill = "Group",
       shape = "Significance",
       y = "Human (NES)",
       x = "Control (NES)") +
  # Theme
  theme_vaxgo() +
  theme(
        strip.text.x = element_text(hjust = 0),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.text = element_text(hjust = 0),
        panel.spacing = unit(0, "lines"))

# Render plot
dge_btm_nes_group_correlation_control_plot



#Save plot
dge_btm_nes_group_correlation_control_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_btm_nes_group_correlation_plot_Control_filtered.png"),
         width = 10,
         height = 4)
```



###### Do modules with higher activation in the model show greater concordance in Human?

```{r fig.height=3, fig.width=10}
# Precision analysis by Mouse NES quartiles

# Compute precision per quartile
results_quartiles = dge_btm_nes_group_correlation_df %>%
  # Retain only all-module filter; exclude pathogens with few timepoints
  # filter(!pathogen %in% c("Influenza", "Hepatitis B")) %>%
  # Stratify by pathogen and timepoint
  group_by(pathogen, timepoint_comparison) %>%
  # Restrict to positively enriched modules only
  filter(nes_Mouse > 0,
         nes_Human > 0) %>%
  # Divide modules into quartiles by Mouse NES
  mutate(quartile = ntile(nes_Mouse, 4)) %>%
  # Compute precision per quartile
  group_by(pathogen, timepoint_comparison, quartile) %>%
  summarise(
    total_in_quartile = n(),
    true_positives = sum(p_adjust_Human <= 0.05),
    precision_pct = (true_positives / total_in_quartile) * 100,
    .groups = "drop"
  )

# Visualisation: precision ladder plot
ggplot(results_quartiles) +
  aes(x = precision_pct, y = as.character(quartile), fill = as.character(quartile)) +
  # Geoms
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  # Reference line for random chance (5% = 0.05 alpha)
  geom_vline(xintercept = 5, linetype = "dashed", color = "red") +
  # Facets
  facet_wrap(~pathogen, scales = "free_y", nrow = 1) +
  # Scales
  scale_fill_viridis_d(name = "Mouse Quartile", option = "mako", direction = -1) +
  scale_x_continuous(limits = c(0, 100), labels = function(x) paste0(x, "%")) +
  # Theme
  theme_vaxgo() +
  theme(legend.position = "top") +
  # Labels
  labs(title = "Precision by Mouse NES Quartile",
       subtitle = "Higher Quartiles (4) should show higher % of Human validation",
       x = "Precision (% of BTMs Significant in Human)",
       y = "Timepoint")



```






##----------------

### Gene-level, BTMs

#### Intra-module correlation 

```{r}
# Define a safe correlation function robust to low n and zero variance
safe_cor_test <- function(x, y, method) {
  # Remove non-finite values
  ok <- is.finite(x) & is.finite(y)
  x <- x[ok]
  y <- y[ok]
  # Require at least 3 observations and non-zero variance
  if (length(x) < 3 || sd(x) == 0 || sd(y) == 0) {
    return(list(estimate = NA_real_, p.value = NA_real_))
  } else {
    tryCatch({
      ct <- cor.test(x, y, method = method)
      return(list(estimate = unname(ct$estimate), p.value = ct$p.value))
    }, error = function(e) {
      return(list(estimate = NA_real_, p.value = NA_real_))
    })
  }
}

# Compute per-BTM Spearman correlation
dge_btm_process_genes_correlation <- dge_btm_process_genes_wide %>%
  group_by(pathogen, timepoint_comparison, process) %>%
  
  # Compute correlation and count genes
  summarise(
    n_genes = n(),
    test_result = list(safe_cor_test(x = mean_l2fc_Mouse, y = mean_l2fc_Human, method = "spearman")),
    .groups = "drop"
  ) %>%
  
  # Extract correlation estimate and p-value
  mutate(
    cor_value = map_dbl(test_result, ~ .x$estimate %||% NA_real_),
    p_value   = map_dbl(test_result, ~ .x$p.value %||% NA_real_)
  ) %>%
  
  # Compute R-squared and significance label
  mutate(
    r_squared = cor_value^2,
    p_label = case_when(
      p_value <= 0.01 ~ "***",
      p_value <= 0.05 ~ "**",
      p_value <= 0.10 ~ "*",
      TRUE            ~ "ns"
    ),
    cor_label = sprintf("r=%.2f, R²=%.2f, %s", cor_value, r_squared, p_label)
  ) %>%
  
  # Remove nested column and organise
  select(-test_result) %>%
  arrange(timepoint_comparison) %>%
  distinct()

# Clean and rank processes by absolute correlation strength
correlation_processes_bypathogen_time_df <- dge_btm_process_genes_correlation %>%
  # Remove missing correlations
  drop_na(cor_value) %>%
  # Group by pathogen and timepoint
  group_by(pathogen, timepoint_comparison) %>%
  # Rank within each stratum by descending absolute correlation
  arrange(process, -abs(cor_value)) %>%
  distinct() %>%
  # Remove grouping
  ungroup() %>%
  # Attach functional group annotation
  inner_join(
    btm_annotation_genes %>% select(process, group),
    by = "process"
  ) %>%
  # Deduplicate after join
  distinct()

# Check
correlation_processes_bypathogen_time_df %>%
  dplyr::count(pathogen, timepoint_comparison)

#Save
correlation_processes_bypathogen_time_df %>%
  saveRDS(file = here("tables", "all_human_mouse_dge_btm_process_genes_correlation.rds"))

# # Compute per-gene Spearman correlation across conditions
# dge_btm_process_genes_correlation_bygene <- dge_btm_process_genes_wide %>%
#   # Select relevant columns; exclude Influenza and Hepatitis B
#   select(symbol, pathogen, timepoint_comparison, Mouse = mean_l2fc_Mouse, Human = mean_l2fc_Human) %>%
#   filter(!pathogen %in% c("Influenza", "Hepatitis B")) %>%
#   distinct() %>%
#   # Group by gene symbol
#   group_by(symbol) %>%
#   summarise(
#     n_pathogens = n(),
#       test_result = list(safe_cor_test(x = Human, y = Mouse, method = "spearman")),
#     .groups = "drop"
#   ) %>%
#   # Extract correlation estimate and p-value
#   mutate(
#     cor_value = map_dbl(test_result, ~ .x$estimate %||% NA_real_),
#     p_value   = map_dbl(test_result, ~ .x$p.value   %||% NA_real_)
#   ) %>%
#   # Compute R-squared and significance label
#   mutate(
#     r_squared = cor_value^2,
#     p_label = case_when(
#       p_value <= 0.01 ~ "***",
#       p_value <= 0.05 ~ "**",
#       p_value <= 0.10 ~ "*",
#       TRUE            ~ "ns"
#     ),
#     cor_label = sprintf("r=%.2f, R²=%.2f, %s", cor_value, r_squared, p_label)
#   ) %>%
#   # Remove nested column and sort by correlation strength
#   select(-test_result) %>%
#   arrange(cor_value) %>%
#   distinct()
# 
# # Inspect and save per-gene correlation
# dge_btm_process_genes_correlation_bygene
# 
# dge_btm_process_genes_correlation_bygene %>%
#   saveRDS(file = here("tables", "all_human_mouse_dge_btm_process_genes_correlation_bygene.rds"))
```

###### Violinplot: Distribution of Correlation values

Distribution of correlations across timepoints and pathogens with a violin-boxplot

```{r fig.height = 5, fig.width=12}
# Select extreme processes for labelling
selected_processes_df = bind_rows(
    # Top positive correlations
    correlation_processes_bypathogen_time_df %>%
      filter(p_value <= 0.05,
             cor_value > 0) %>%
      group_by(pathogen, timepoint_comparison) %>%
      filter(n_genes > 3) %>%
      arrange(-cor_value, -n_genes) %>%
      slice_head(n = 2),
    # Top negative correlations
    correlation_processes_bypathogen_time_df %>%
      filter(p_value <= 0.05,
             cor_value < 0) %>%
      group_by(pathogen, timepoint_comparison) %>%
      filter(n_genes > 3) %>%
      arrange(cor_value, -n_genes) %>%
      slice_head(n = 2),
    # Bottom positive correlations
    correlation_processes_bypathogen_time_df %>%
      filter(p_value <= 0.05,
             cor_value > 0) %>%
      group_by(pathogen, timepoint_comparison) %>%
      filter(n_genes > 3) %>%
      arrange(cor_value, -n_genes) %>%
      slice_head(n = 2),
    # Bottom negative correlations
    correlation_processes_bypathogen_time_df %>%
      filter(p_value <= 0.05,
             cor_value < 0) %>%
      group_by(pathogen, timepoint_comparison) %>%
      filter(n_genes > 3) %>%
      arrange(cor_value, -n_genes) %>%
      slice_head(n = 2)
  )

# Violin-boxplot of correlation distributions
correlation_processes_bypathogen_time = correlation_processes_bypathogen_time_df %>%
  # Aesthetics
  ggplot() +
  aes(x = timepoint_comparison,
      y = cor_value,
      fill = pathogen,
      color = pathogen) +
  # Facets
  facet_nested_wrap(~pathogen, scales = "free_x", nrow = 1) +
  # Geoms: background violin for all data
  geom_violin(color = NA,
              fill = "black",
              alpha = 0.1) +
  # Violin for significant positive correlations
  geom_violin(color = NA,
              data = . %>%
                filter(p_value <= 0.05),
              alpha = 0.3) +
  # Boxplot overlay
  geom_boxplot(color = "black",
               width = 0.1,
               fill = NA,
               outliers = FALSE) +
  # Highlight extreme processes
  geom_point(data = selected_processes_df,
             color = "black",
             pch = 21,
             fill = "white") +
  # Repel labels for extreme processes
  geom_text_repel(aes(label = process),
                  max.overlaps = 10000,
                  size = 3,
                  nudge_x = 0.2,
                  hjust = 0,
                  color = "black",
                  fill = "white", lineheight = 0.6,
                  box.padding = unit(0, "cm"),
                  label.padding = unit(0.1, "cm"),
                  data = selected_processes_df %>%
                   mutate(process = str_wrap(process, width = 10))
  ) +
  # Median annotation
  stat_summary(
    fun = median,
    geom = "label",
    aes(label = after_stat(round(y, 1))),
    size = 3,
    hjust = -0.2,
    color = "black",
    fill = "white"
  ) +
  # Scales
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(-1, 1)) +
  scale_fill_manual(values = colors$pathogen) +
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        panel.spacing = unit(1.1, "lines"),
        plot.margin = margin(0.3, 0.3, 0, 0, unit = "cm"),
        legend.position = "none") +
  # Labels
  labs(x = "Time",
       y = "Correlation, Spearman",
       color = "",
       fill = "",
       title = "Correlation of biological processes by gene expression, stratified by pathogen and time.",
       subtitle = "Gray distribution: All observed values (unfiltered).\nColored distribution: Significant positive correlations (r > 0, p<=0.05)")

# Render plot
correlation_processes_bypathogen_time

# Save plot
correlation_processes_bypathogen_time %>%
  ggsave(filename = here("Figures", "all_human_mouse_gsea_btm_intramodule_correlation_violinplot_labeled.png"),
         width = 12,
         height = 5)
```

###### Heatmap: Distribution of Correlation values

```{r fig.height=20, fig.width=10}
# Prepare heatmap data
btm_correlation_process_heatmap_df =
  correlation_processes_bypathogen_time_df %>%
  # Retain only significant correlations
  # filter(p_value <= 0.05) %>%
  mutate(
    # Reorder processes by mean correlation
    process = fct_reorder(process, cor_value, .fun = mean),
    timepoint = as.factor(timepoint_comparison),
    group = fct_relevel(group, immune_order)
  )

# Build heatmap
btm_correlation_process_heatmap = btm_correlation_process_heatmap_df %>%
  # Aesthetics
  ggplot(aes(x = timepoint_comparison, y = process, fill = cor_value)) +
  # Facets by functional group and pathogen
  facet_grid(
    rows = vars(group),
    cols = vars(pathogen),
    scales = "free",
    space = "free"
  ) +
  # Geoms
  geom_tile(color = "white", linewidth = 0.2) +
  # Scales
  scale_fill_gradient2(
    low = "#06d6a0",
    mid = "white",
    high = "#669bbc",
    midpoint = 0,
    name = "Correlation"
  ) +
  # Labels
  labs(
    x = "",
    y = "",
    title = paste0("BTM immune – intramodule correlation")
  ) +
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.text = element_text(size = 8),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1))

# Render heatmap
btm_correlation_process_heatmap
```

#### Leading-edge genes analysis

Compare the leading-edge genes (LEGs): classify each gene as shared, Human-only, or Mouse-only.

```{r}
#Import


# Split core enrichment genes into separate rows
gsea_btm_results_core = gsea_btm_results_all_matched %>%
  filter(p_adjust <= 0.25) %>%
  separate_rows(core_enrichment, sep = "/")

# Join Human and Mouse core enrichment genes
gsea_btm_results_core_joined =
  # Extract Human core genes
  gsea_btm_results_core %>%
  filter(organism == "Human") %>%
  select(pathogen, process, timepoint_comparison,
         gene = core_enrichment,
         set_size_human = set_size,
         p_adjust_human_gsea = p_adjust) %>%
  mutate(human_core = TRUE) %>%
  
  # Full join with Mouse core genes
  full_join(
    gsea_btm_results_core %>%
      filter(organism == "Mouse") %>%
      select(pathogen, process, timepoint_comparison,
             gene = core_enrichment,
             set_size_mouse = set_size,
             p_adjust_mouse_gsea = p_adjust) %>%
      mutate(mouse_core = TRUE),
    by = c("pathogen", "process", "timepoint_comparison", "gene")
  ) %>% 
  # Classify each gene's conservation status
  mutate(
    status = case_when(
      human_core == TRUE & mouse_core == TRUE ~ "Shared",
      human_core == TRUE & is.na(mouse_core) ~ "Human only",
      mouse_core == TRUE & is.na(human_core) ~ "Mouse only",
      TRUE ~ "Not core"
    )
  ) %>%
  # Arrange and attach functional group
  arrange(pathogen, timepoint_comparison, process, status) %>%
  inner_join(btm_annotation_genes %>% 
               select(process, group),
             by = "process") %>%
  distinct()

# Inspect and save
gsea_btm_results_core_joined %>%
  dplyr::count(pathogen, timepoint_comparison, status)

#Save
gsea_btm_results_core_joined %>%
  saveRDS(file = here("tables", "all_human_mouse_gsea_btm_results_core_joined.rds"))
```

Compute the percentage of leading-edge genes shared between organisms,
stratified by status, process, pathogen, and timepoint.

```{r fig.height = 12, fig.width=4}

#Filter genes analysed in both species
matching_genes = dge_btm_process_genes_wide %>%
  select(pathogen, human_symbol = symbol) %>% 
  distinct() 


# Compute percentage of LEGs per status and stratum
legs_perc <- gsea_btm_results_core_joined %>%
  # filter(process == "immune activation - generic cluster") %>%
  inner_join(matching_genes, by = join_by(pathogen, gene == human_symbol)) %>% 
  dplyr::distinct() %>%
  # Count genes per status within each process stratum
  dplyr::group_by(pathogen, timepoint_comparison, process, status) %>%
  dplyr::summarise(n = dplyr::n(), 
                   .groups = "drop_last",
                   group = dplyr::first(group)) %>%
  # Calculate overall metrics and Mouse-specific denominator
  dplyr::mutate(
    #All in the denominator
    percent       = (n / sum(n)) * 100,
    ntotal        = sum(n),
    percent       = round(percent, 0),
    #Mouse total LEGs in the denominator
    n_mouse_total = sum(n[status %in% c("Mouse only", "Shared")]),
    percent_mouse = dplyr::case_when(
      status %in% c("Mouse only", "Shared") & n_mouse_total > 0 ~ (n / n_mouse_total) * 100,
      TRUE                                                     ~ NA_real_
    ),
    percent_mouse = round(percent_mouse, 0)
  ) %>%
  dplyr::ungroup() %>%
  # Refactor treatment categories and set explicit factor order
  dplyr::mutate(
    treatment = dplyr::case_when(
      pathogen %in% c("HepB", "Influenza")     ~ "Vaccination",
      pathogen %in% c("S. aureus", "E. coli")  ~ "Infection",
      pathogen %in% c("Burn", "Trauma")        ~ "Injury",
      TRUE                                     ~ pathogen
    ),
    treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
  ) %>%
  dplyr::arrange(timepoint_comparison)

# # Save summary table
legs_perc %>%
  saveRDS(file = here("tables", "all_human_mouse_gsea_btm_results_core_summary.rds"))

# Report mean and median shared percentage per pathogen
legs_perc %>%
  filter(status == "Shared") %>%
  # group_by(pathogen) %>%
  summarise(
            median_perc = median(percent, na.rm = T),
            mean_perc = mean(percent, na.rm = T),
            n = n(),
            se_perc = sd(percent, na.rm = T),
            median_perc_mouse = median(percent_mouse, na.rm = T),
            mean_perc_mouse = mean(percent_mouse, na.rm = T))



```

Visualise the distribution of shared leading-edge gene percentages
across processes, pathogens, and timepoints with violin-boxplots,
annotating extreme processes.

```{r fig.height = 3, fig.width=10}
#Import
legs_perc = readRDS(file = here("tables", "all_human_mouse_gsea_btm_results_core_summary.rds"))

# Prepare plot data
shared_perc_plot_df = legs_perc %>%
  # Keep only shared status
  filter(status == "Shared") %>%
  mutate(
    # Classify treatment type
    treatment = case_when(
      pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
      pathogen %in% c("S. aureus", "E. coli") ~ "Infection",
      pathogen %in% c("Burn", "Trauma") ~ "Injury",
      TRUE ~ pathogen
    ),
    treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"),
    # timepoint_original = timepoint,
    # timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7"))
  )

#Labels
top_bottom_shared_modules <- dplyr::bind_rows(
  # Extract top 5 process modules per pathogen and timepoint
  shared_perc_plot_df %>%
    dplyr::group_by(pathogen, timepoint_comparison) %>%
    dplyr::arrange(-percent_mouse, -ntotal) %>%
    dplyr::slice_head(n = 2),
  # Extract bottom 5 process modules per pathogen and timepoint
  shared_perc_plot_df %>%
    dplyr::group_by(pathogen, timepoint_comparison) %>%
    dplyr::arrange(percent_mouse, -ntotal) %>%
    dplyr::slice_head(n = 2)
) %>%
  # Apply text wrapping and assign randomized direction (-1 = left, 1 = right)
  dplyr::mutate(
    process = stringr::str_wrap(process, width = 10),
    side    = sample(c(-1, 1), size = dplyr::n(), replace = TRUE),
    nudge_x = side * 0.2,
    hjust   = dplyr::if_else(side == 1, 0, 1)
  )


# Violin-boxplot of shared percentage
shared_processes_bypathogen_time = shared_perc_plot_df %>%
  # Aesthetics
  ggplot() +
  aes(x = timepoint_comparison,
      y = percent_mouse,
      fill = pathogen,
      color = pathogen) +
  # Facets
  facet_nested_wrap(~treatment + pathogen,
             scales = "free_x", nrow = 1) +
  # Geoms
  geom_violin(color = NA,
              alpha = 0.5,
              fill = "gray90") +
  geom_boxplot(
    color = "black",
    width = 0.1,
    fill = "white",
    outliers = FALSE
  ) +
  geom_point(
    mapping = aes(fill = group,
                  group = group, 
      label = process),
    color = "black",
    pch = 21,
    size = 2,
    position = position_dodge2(width = 0.3)
  ) +
  # Median annotation
  stat_summary(
    fun = median,
    geom = "label",
    aes(label = after_stat(round(y, 1))),
    size = 3,
    hjust = -0.2,
    color = "black",
    fill = "white"
  ) +
  # Scales
  scale_y_continuous(expand = c(0.05, 0.05)) +
  scale_fill_manual(values = btm_immune_groups$group) +
  guides(
    fill = guide_legend(
      byrow = TRUE,
      ncol = 1, 
      override.aes = list(shape = 22, size = 3)
    )
  ) +
  theme(
    legend.position = "right",
    legend.key.spacing.y = unit(1, "pt"), 
    legend.margin = margin(t = 0, b = 10)
  ) +
  # Theme
  theme_vaxgo() +
  theme(panel.spacing = unit(0.2, "lines"),
        plot.margin = margin(0.3, 0.3, 0, 0, unit = "cm"),
        plot.title = element_text(hjust = 0.5, face = "bold"), plot.title.position = "panel",
        legend.position = "right",
        legend.key.spacing.y = unit(1, "pt"), 
        legend.margin = margin(t = 0, b = 10),
        legend.key.size = unit(2, "pt"),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) +
  # Labels
  labs(x = "Time",
       y = "%Shared LEGs",
       color = "",
       fill = "BTM Group",
       title = "Distribution of %Shared LEGs, modules with p_adj <= 0.25") 

# Render plot
shared_processes_bypathogen_time 

# Save plot
shared_processes_bypathogen_time %>%
  ggsave(filename = here("Figures", "all_human_mouse_gsea_btm_intramodule_shared_violinplot_percentage_labeled_horizontal.png"),
         width = 10,
         height = 3)

```



##### Bubble plot of shared LEG percentage across processes

```{r fig.height=15, fig.width=10}
# Build shared percentage bubble plot
shared_perc_plot = shared_perc_plot_df %>%
  # Align process order with the correlation heatmap
  mutate(process = fct_relevel(process, btm_correlation_process_heatmap$data %>%
                                  arrange(process) %>%
                                  mutate(process = as.character(process)) %>%
                                  pull(process) %>%
                                  unique()),
         group = fct_relevel(group, immune_order),
         timepoint_comparison = str_replace_all(timepoint_comparison, "-", "\n")) %>%
  # Aesthetics
  ggplot() +
  aes(x = timepoint_comparison,
      y = process,
      fill = percent_mouse,
      size = n) +
  # Facets
  facet_grid(
    rows = vars(group),
    cols = vars(pathogen),
    scales = "free",
    space = "free_y"
  ) +
  # Geoms
  geom_point(pch = 21) +
  # Scales
  scale_fill_stepsn(
    colours = c("white", "#06d6a0", "#669bbc", "black"),
    breaks = c(0, 25, 50, 75, 100),
    limits = c(0, 100),
    name = "Shared %"
  ) +
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.text = element_text(size = 10, hjust = 0),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) +
  # Labels
  labs(y = "", x = "Timepoint", size = "N genes")

# Render bubble plot
shared_perc_plot

```

Combine the correlation heatmap and the shared-percentage bubble plot
into a single composite figure.

```{r fig.height=20, fig.width=20}
# Combine heatmap and bubble plot
btm_correlation_shared_perc_plots = (btm_correlation_process_heatmap + theme(plot.title = element_blank())) | (shared_perc_plot) +
  plot_layout(axes = "collect")

# Save and render combined plot
btm_correlation_shared_perc_plots %>%
  ggsave(filename = here("Figures", "all_human_mouse_gsea_btm_results_correlation_and_leadingedge_shared_perc_heatmap.png"),
         width = 20,
         height = 20)

btm_correlation_shared_perc_plots
```

##### Compare the log2FC and ranks of core LEGs 

```{r}
# Compare log2FC and ranks of Leading Edge genes across species

# 1. Define target process
correlation_processes_bypathogen_time_df = readRDS(file = here("tables", "all_human_mouse_dge_btm_process_genes_correlation.rds"))
dge_btm_process_genes_correlation = correlation_processes_bypathogen_time_df

# Unite all data sources into a master data frame by sequential joins
df_legs <- dge_btm_process_genes_wide %>%
  
  # 2. Join with GSEA core enrichment (Leading Edge) genes
  full_join(
    gsea_btm_results_core_joined,
    # filter(process == process_selected)
    by = join_by(group, process, pathogen, timepoint_comparison, symbol == gene)
  ) %>% 
  drop_na(status) %>% 
  # Drop genes not measured in both organisms
  drop_na(mean_l2fc_Mouse, mean_l2fc_Human) %>% 
  
  # 3. Integrate shared percentage metrics
  inner_join(
    legs_perc %>%
      # filter(status == "Shared") %>%
      rename(n_status_perc = percent_mouse),
    by = join_by(pathogen, group, process, timepoint_comparison, status)
  ) %>%
  
  # 4. Integrate Normalized Enrichment Scores (NES)
  inner_join(
    dge_btm_gsea_mean_wide %>%
      ungroup() %>%
      # filter(process == process_selected)
      select(process, group, pathogen, timepoint_comparison, starts_with("nes_")),
    by = join_by(pathogen, group, process, timepoint_comparison)
  ) %>%

  # 5. Integrate correlation statistics
  inner_join(
    dge_btm_process_genes_correlation %>%
      select(timepoint_comparison, pathogen, process, cor_value, r_squared, p_label, cor_label),
    by = join_by(pathogen, process, timepoint_comparison)
  ) %>%

  # 6. Data cleaning and dynamic label generation for faceting
  mutate(
    across(starts_with("nes_"), ~ round(.x, 2)),
    status = if_else(is.na(status), "Not core", status),
    facet_label = sprintf(
      "%s\nShared = %s%%\nNES Mouse = %.2f\nNES Human = %.2f",
      timepoint_comparison, n_status_perc, nes_Mouse, nes_Human
    )
  ) %>%
  arrange(timepoint_comparison) %>%
  mutate(facet_label = fct_inorder(facet_label)) %>%
  distinct() %>%
  
  # Rank genes by log2FC within each condition group
  group_by(pathogen, process, group, timepoint_comparison) %>%
  
  
  mutate(Mouse_rank = rank(mean_l2fc_Mouse),
         Mouse_rank_p = rank(mean_l2fc_Mouse * -log(p_adjust_mouse_gsea)),
         Human_rank_p = rank(mean_l2fc_Human * -log(p_adjust_human_gsea)),
         Human_rank = rank(mean_l2fc_Human)
         ) %>%
  ungroup()

#Fix facet label to match only Shared LEGs
df_legs = df_legs %>% 
  filter(status == "Shared") %>% 
  select(facet_label, pathogen, timepoint_comparison, process, group) %>% 
  distinct() %>% 
  inner_join(df_legs %>% 
               select(!facet_label),
             by = join_by(pathogen, timepoint_comparison, process, group))

#Save
df_legs %>% 
  saveRDS(file = here("tables", "all_human_mouse_gsea_btm_legs.rds"))

# Check distinct pathogen-timepoint combinations
dge_btm_gsea_mean_wide %>%
  ungroup() %>%
  select(pathogen, timepoint_comparison) %>%
  distinct()

# Reshape rank variables into long format
df_legs_ranked = df_legs %>%
  filter(process == process_selected) %>%
  select(symbol:timepoint_comparison, group,
         status, n_status_perc:facet_label,
         Mouse_rank:Human_rank) %>%
  pivot_longer(
    names_to = "variable",
    values_to = "value",
    cols = c("Mouse_rank", "Human_rank",
             "Mouse_rank_p", "Human_rank_p"
             )
  ) %>%
  mutate(organism = if_else(str_detect(variable, "Mouse"), "M", "H")) %>%
  
  # Refine labels and categorical grouping (Vaccination, Infection, Injury)
  mutate(
    # facet_label = sprintf(
    #   "%s\n%s\nShared = %s%%\nNES M=%.2f, H=%.2f",
    #   timepoint_comparison, cor_label, n_status_perc, nes_Mouse, nes_Human
    # ),
    treatment = case_when(
      pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
      pathogen %in% c("S. aureus", "E. coli")         ~ "Infection",
      pathogen %in% c("Burn", "Trauma")           ~ "Injury",
      TRUE                                        ~ pathogen
    ),
    treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
  ) %>%
  arrange(timepoint_comparison) %>%
  mutate(facet_label = fct_inorder(facet_label))

df_legs_ranked
```


##### %Shared LEGs: Barplot 

```{r fig.height=2.5, fig.width=8}
#Import
df_legs = readRDS(file = here("tables", "all_human_mouse_gsea_btm_legs.rds"))
legs_perc = readRDS(file = here("tables", "all_human_mouse_gsea_btm_results_core_summary.rds"))

#Define process
# process_selected <- "cell cycle and transcription"
process_selected = "immune activation - generic cluster"
```


```{r fig.height=2.5, fig.width=8}
# Plot percent of shared Leading Edge genes per condition
# 1. Prepare data 
legs_perc_barplot = legs_perc %>%
  
  # Filter to target process and exclude Vaccination and Trauma
  filter(process %in% c(process_selected),
         !treatment == "Vaccination"
         ) %>%
  
  # Data aesthetics
  ggplot() +
  aes(x = percent,
      y = fct_rev(timepoint_comparison),
      fill = status) +
  # Facets
  facet_nested_wrap(~process + treatment + pathogen + timepoint_comparison,
             scales = "free",
             nrow = 1,
             strip = strip_nested(size = "variable")) +
  
  # Geoms
  geom_col(position = "fill") +
  
  # Add percentage labels on bars
  geom_text(aes(label = paste0(percent, "%\n(", n, ")")),
            lineheight = 0.8,
            position = "fill",
            color = "white",
            fontface = "bold",
            hjust = 1,
            size = 3) +
  # Scales and Labels
  scale_fill_manual(values = colors$comparison) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(fill = "Core Status",
       y = "Timepoint",
       x = "Percent") +
  
  # Theme and Legend styling
  theme_vaxgo() +
  theme(
    axis.text = element_text(size = 10),
    axis.text.y = element_blank(),
    strip.text = element_text(size = 10, hjust = 0),
    panel.spacing = unit(1, "cm"),
    legend.position = "top", legend.text = element_text(size = 10),
    legend.margin = margin(0, 0, -10, 0),
    ggh4x.facet.nestline = element_line(colour = "black")
  ) 

legs_perc_barplot

# Save and render barplot
legs_perc_barplot %>%
ggsave(filename = here("Figures", paste0("all_human_mouse_gsea_btm_results_leadingedge_core_barplot_selected_process_",process_selected, ".png")),
         width = 8,
         height = 2.5)

```


##### Shared LEGs, Ranks and Effect size: Scatterplot

Vertical
```{r fig.height=8, fig.width=4}
# Scatterplot of significance-weighted ranks
legs_scatterplot_rank_significance = df_legs %>%
  mutate(facet_label_vertical = paste0(pathogen, "\n", facet_label)) %>% 
  arrange(treatment) %>% 
  mutate(facet_label_vertical= fct_inorder(facet_label_vertical)) %>% 
  filter(process == process_selected) %>%
  # Data aesthetics
  ggplot() +
  aes(x = Mouse_rank_p,
      y = Human_rank_p) +
  
  labs(
       x = "Mouse Rank (Log2FC x -logP)",
       y = "Human Rank (Log2FC x -logP)",
       fill = "Status") +
 
  # Facets
  facet_grid(facet_label_vertical~.,
             axes = "all") +

  # Geoms
  geom_point(aes(fill = status), pch = 21, size = 2, alpha = 0.3, data = . %>% filter(status != "Shared")) +
  geom_point(aes(fill = status), pch = 21, size = 2, data = . %>% filter(status == "Shared")) +
  # Scales
  scale_fill_manual(values = colors$comparison) +
  scale_x_continuous(transform = "reverse") +
  scale_y_continuous(transform = "reverse") +

  # Theme and Legend styling
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0, size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        panel.spacing = unit(1, "lines"),
        plot.margin = margin(0, 0.1, 0, 0.1, unit = "cm"),
        legend.position = "top",
        legend.location = "plot",
        legend.direction = "horizontal",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, -10, 0),
        legend.text = element_text(size = 10, hjust = 0.5),
        legend.title = element_text(size = 10),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1))

legs_scatterplot_rank_significance

#Save
legs_scatterplot_rank_significance %>% 
  ggsave(filename = here("Figures", paste0("all_human_mouse_gsea_btm_results_leadingedge_core_ranks_scatterplot_",process_selected, ".png")),
         width = 4, height =8 )

#Labelled genes
legs_scatterplot_rank_significance_label = legs_scatterplot_rank_significance +
  geom_text_repel(aes(label = symbol),
                  data = . %>% filter(status == "Shared"),
                  size = 2, max.overlaps = 10000)  


legs_scatterplot_rank_significance_label

legs_scatterplot_rank_significance_label %>% 
  ggsave(filename =  here("Figures", paste0("all_human_mouse_gsea_btm_results_leadingedge_core_ranks_scatterplot_LabelledGenes_",process_selected, ".png")),
         width = 10, height = 4)
```

Horizontal
```{r fig.height=3.5, fig.width=8}
legs_scatterplot_rank_significance_horizontal = df_legs %>%
  mutate(facet_label_vertical = paste0(pathogen, "\n", facet_label)) %>% 
  arrange(treatment) %>% 
  mutate(facet_label_vertical= fct_inorder(facet_label_vertical)) %>% 
  filter(process == process_selected,
         ) %>%
  # Data aesthetics
  ggplot() +
  aes(x = Mouse_rank_p,
      y = Human_rank_p) +
  
  labs(
       x = "Mouse Rank (Log2FC x -logP)",
       y = "Human Rank (Log2FC x -logP)",
       fill = "Status") +
 
  # Facets
  facet_nested_wrap(process+treatment+facet_label_vertical~.,
              scales = "free",
             nrow = 1,
             strip = strip_nested(size = "variable")) +

  # Geoms
  geom_point(aes(fill = status), pch = 21, size = 2, alpha = 0.3, data = . %>% filter(status != "Shared")) +
  geom_point(aes(fill = status), pch = 21, size = 2, data = . %>% filter(status == "Shared")) +
  # Scales
  scale_fill_manual(values = colors$comparison) +
  scale_x_continuous(transform = "reverse") +
  scale_y_continuous(transform = "reverse") +

  # Theme and Legend styling
  # Theme
  theme_vaxgo() +
  theme(legend.position = "top", legend.direction = "horizontal",
        strip.text.x           = element_text(hjust = 0, margin = margin(5, 0, 5, 0))) 

legs_scatterplot_rank_significance_horizontal

# Render both scatterplots
legs_scatterplot_rank_significance_horizontal %>%
  ggsave(filename = here("Figures", paste0("all_human_mouse_gsea_btm_results_leadingedge_core_ranks_scatterplot_horizontal_",process_selected, ".png")),
         width = 8, height = 3.5 )


```




##### Comparing gene ranks: Slope plot

```{r fig.height= 10, fig.width=4}
# Plot gene positions across species using ranks
df_legs_ranked_significance_plot = df_legs_ranked %>%
  # mutate(facet_label = str_replace_all(facet_label, "\n", " ")) %>% 
  # filter(status == "Shared") %>% 
  # Select only significance-weighted rank variables
  filter(variable %in% c("Mouse_rank_p", "Human_rank_p")) %>%
  # Data aesthetics
  ggplot() +
  aes(x = organism,
      y = value,
      group = symbol,
      color = status) +
  # Facets
  facet_nested_wrap(
    ~pathogen + facet_label,
    scales = "free",
    ncol = 1,
    strip = strip_nested(size = "variable")
  ) +
  # Geoms: layered lines - Shared genes drawn last (on top) with full opacity
  geom_line(size = 0.5, alpha = 0.3, data = . %>% filter(status %in% c("Human only", "Mouse only"))) +
  geom_line(size = 0.7, alpha = 1.0, data = . %>% filter(status == "Shared")) +
  # Scales and Labels
  scale_color_manual(values = colors$comparison) +
  scale_x_discrete(expand = c(0.01, 0.01)) +
  scale_y_continuous(
    # Commented out: labels = label_number(suffix = "k", scale = 1e-3)
    transform = "reverse",
    expand = c(0.01, 0.01)
  ) +
  labs(color = "Core Status",
       y = "Rank (Descending log2FC)",
       x = NULL,
       title = process_selected) +
  # Theme and Legend styling
  theme_vaxgo() +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  theme(
    axis.text = element_text(size = 10),
    strip.text = element_text(size = 10, hjust = 0),
    panel.spacing = unit(0.2, "lines"),
    
    legend.position = "top",
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    legend.margin = margin(0, 0, -10, 0),
    ggh4x.facet.nestline = element_line(colour = "black")
  ) 

df_legs_ranked_significance_plot +
  theme(plot.margin = margin(0, 1, 0, 1, unit = "cm"))
```


```{r fig.height= 10, fig.width=5}
# # ADD  LABELS
# df_legs_ranked_plot +
#   geom_text_repel(
#     data = df_legs_ranked %>%
#       # filter(pathogen == "Saureus") %>%
#       filter(status == "Shared") %>%
#       group_by(pathogen, process, timepoint) %>%
#       arrange(rank, .by_group = TRUE) %>%
#   filter(
#     (nes_Mouse > 0 & row_number() <= 5) |
#     (nes_Mouse < 0 & row_number() > (n() - 5))
#   ) %>%
#   ungroup(),
#     aes(label = symbol,
#         x = "M"),
#     size = 3,
#     color = "black",
#     direction = "y",
#     hjust = 0,
#   max.iter = 1e4, max.time = 1,
#     nudge_x = 0.1,
#     # segment.size = .7,
#     # segment.alpha = .5,
#   force        = 0.5,
#     segment.linetype = "solid",
#     box.padding = .4,
#   point.padding = 0.2,
#   segment.square    = T,
#     segment.inflect   = TRUE
#     # segment.curvature = -1,
#     # segment.ncp = 3,
#     # segment.angle = 20
#   ) +
#   coord_cartesian(clip = "off")

# Save significance-weighted slope plot
df_legs_ranked_significance_plot %>%
  ggsave(filename = here("Figures", paste0("all_human_mouse_gsea_btm_results_leadingedge_core_ranks_lineplot_", process_selected, "2.png")),
         width = 12,
         height = 5)

# Render both slope plots
df_legs_ranked_significance_plot
df_legs_ranked_plot
```




Combine all leading edge analysis plots into composite figures using
patchwork. Two versions are assembled: a full composite with
scatterplot, log2FC comparison, and slope plot, and a shortened version
with only the scatterplot and slope plot.

```{r fig.height=16, fig.width=20}
# Combine all leading edge plots into composite figures
# Full composite: scatter + spacer + log2FC + spacer + slope
leading_edges_plots = (
  (df_test_scatter + theme(legend.position = "top")) /
  plot_spacer() /
  (df_test_log2fc + theme(legend.position = "none")) /
  plot_spacer() /
  (df_legs_ranked_plot + theme(legend.position = "none"))
) +
plot_layout(
  guides = "collect",
  heights = c(0.8, 0, 1, 0, 1)
) +
plot_annotation(
  theme = theme(
    legend.position = "top",
    plot.margin = margin(1, 1, 1, 1, "pt")
  )
) &
theme(
  panel.spacing = unit(0, "lines"),
  plot.margin = margin(0, 2, 0, 2, "pt")
)

# Render full composite
leading_edges_plots

# Save full composite
leading_edges_plots %>%
  ggsave(filename = here("Figures", paste0("all_human_mouse_gsea_btm_results_leadingedge_core_", process_selected, ".png")),
         width = 20,
         height = 16)

# Shortened version: scatter + spacer + slope only
leading_edges_plots_short = (
  (df_test_scatter + theme(legend.position = "top")) /
  plot_spacer() /
  (df_legs_ranked_plot + theme(legend.position = "none"))
) +
plot_layout(
  guides = "collect",
  heights = c(1, -0.2, 1)
) +
plot_annotation(
  theme = theme(
    legend.position = "top",
    plot.margin = margin(1, 1, 1, 1, "pt")
  )
) &
theme(
  panel.spacing = unit(0, "lines"),
  # Commented out: plot.margin = margin(0, 2, 0, 2, "pt")
  legend.margin = margin(0, 0, 0, 0),
  legend.box.margin = margin(1, 0, -10, 0),
  plot.margin = margin(0.5, 0.5, 0.5, 0.5, unit = "cm")
)

# Render shortened composite
leading_edges_plots_short

# Save shortened composite
leading_edges_plots_short %>%
  ggsave(filename = here("Figures", paste0("all_human_mouse_gsea_btm_results_leadingedge_core_short_", process_selected, ".png")),
         width = 20,
         height = 12)
```




##### Unite LEGs with DGE for future analyses
```{r}
# Gene expression
dge_btm_process_genes_diff_bygene = readRDS(here("tables", "human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune.rds"))

dge_btm_process_genes_diff_bygene_clean = dge_btm_process_genes_diff_bygene %>% 
  filter(gene_set == "All genes", comparison == "Human vs. Mouse") %>% 
  filter(pathogen_mouse != "DMD") %>% 
  select(treatment = treatment_human, pathogen = pathogen_human, timepoint_comparison, hgnc_symbol = human_symbol, mean_log2fc_mouse,
         mean_log2fc_diff, se_mouse) %>% 
  mutate(inverse_se_mouse = 1/se_mouse,
         abs_log2fc_diff = abs(mean_log2fc_diff)) %>% 
  select(-se_mouse, -mean_log2fc_diff)
  

#LEGs
all_human_mouse_gsea_btm_results_core_joined = readRDS(file = here("tables", "all_human_mouse_gsea_btm_results_core_joined.rds"))
gsea_btm_legs_clean =
  all_human_mouse_gsea_btm_results_core_joined %>%
  distinct() %>% 
  select(pathogen, timepoint_comparison, hgnc_symbol = gene, status) %>% 
  distinct() %>% 
  mutate(status_original = status) %>% 
  mutate(status = if_else(status == "Shared", "Shared", "Not-shared"))


#Filter timepoints

selected_timepoints = gsea_btm_legs_clean %>% 
  select(pathogen, timepoint_comparison) %>% 
  distinct()

dge_btm_process_genes_diff_bygene_clean_filtered = 
  bind_rows(
    
  #LEGs categorized
  dge_btm_process_genes_diff_bygene_clean %>% 
  inner_join(selected_timepoints,
               by = join_by(pathogen, timepoint_comparison)) %>% 
  inner_join(
    gsea_btm_legs_clean %>%
      select(hgnc_symbol, pathogen, status, timepoint_comparison) %>% distinct(),
    by = join_by(hgnc_symbol, pathogen, timepoint_comparison)
  ) %>% 
  mutate(status = if_else(is.na(status), "Not LEG", status)),
  
  #Bind all genes
  dge_btm_process_genes_diff_bygene_clean %>% 
    inner_join(selected_timepoints,
               by = join_by(pathogen, timepoint_comparison)) %>% 
  left_join(
    gsea_btm_legs_clean %>%
      select(hgnc_symbol, pathogen, status, timepoint_comparison) %>% distinct(),
    by = join_by(hgnc_symbol, pathogen, timepoint_comparison)
  ) %>% 
  mutate(status = "All genes") 
  )

dge_btm_process_genes_diff_bygene_clean_filtered %>% dplyr::count(pathogen, timepoint_comparison, status)

#Save
dge_btm_process_genes_diff_bygene_clean_filtered %>% 
  saveRDS(file = here("tables", "dge_btm_process_genes_diff_bygene_clean_filtered.rds"))



```


##### Are LEGs shared and not-shared between species different gene expression?


```{r fig.height=2.5, fig.width=8}

dge_btm_process_genes_diff_bygene_clean_filtered =
  readRDS(file = here("tables", "dge_btm_process_genes_diff_bygene_clean_filtered.rds"))

# Gene expression
legs_divergence_plot = dge_btm_process_genes_diff_bygene_clean_filtered %>% 
  filter(treatment %in% c("Infection", "Injury"),
         status %in% c("Shared", "Not-shared"),) %>%
  bind_rows(dge_btm_process_genes_diff_bygene_clean_filtered %>% 
  filter(treatment %in% c("Infection", "Injury"),
         status %in% c("Shared", "Not-shared"),) %>% 
    mutate(treatment = "Pooled",
           pathogen = "")) %>% 
  ggplot() +
  aes(x = status,
      y = abs_log2fc_diff,
      color = status,
      fill =  status,
      ) +
  facet_nested_wrap(~treatment+pathogen,
                    nrow = 1, axes = "all", scales = "free") +
  geom_violin(width = 1) +
  geom_boxplot(width = 0.3,
               alpha = 0,
               color = "black",
               outliers = F) +
  scale_y_continuous(limits = c(0, 3), expand = c(0.01, 0.01)) +
    # STATISTICAL TEST LAYER
  stat_compare_means(
    comparisons = list(c("Shared", "Not-shared")),
    method = "wilcox.test",
    p.adjust.method = "BH",
    label = "p.adj",
    label.y.np = "top", 
    label.y = 2.2,
    vjust = 1.5,
    hide.ns = T, step.increase = 0,
    size = 2.5, tip.length = 0,
  ) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(x = "N",
       y = "|ΔMean Log2FC|",
       fill = "",
       color = "") +
  theme_vaxgo() +
  theme(legend.position = "top",
        axis.title.x = element_blank(), 
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) 

#Print
legs_divergence_plot

#Save plot
legs_divergence_plot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_legs_divergence_plot.png"),
         height = 2.5, width = 8)

```




```{r fig.height=3, fig.width=8, message=FALSE, warning=FALSE}

##### Difference in means vs. status
stat_results <- dge_btm_process_genes_diff_bygene_clean_filtered %>%
  filter(status %in% c("Shared", "Not-shared"),
         treatment %in% c("Infection", "Injury")) %>%
  group_by(treatment, pathogen) %>%
  summarise(
    n_shared     = sum(status == "Shared" & !is.na(abs_log2fc_diff)),
    n_not_shared = sum(status == "Not-shared" & !is.na(abs_log2fc_diff)),
    
    median_shared     = if (n_shared > 0) median(abs_log2fc_diff[status == "Shared"], na.rm = TRUE) else NA_real_,
    median_not_shared = if (n_not_shared > 0) median(abs_log2fc_diff[status == "Not-shared"], na.rm = TRUE) else NA_real_,
    
    wilcox_p = if (n_shared >= 1 && n_not_shared >= 1) {
      wilcox.test(
        abs_log2fc_diff[status == "Shared"],
        abs_log2fc_diff[status == "Not-shared"]
      )$p.value
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  mutate(
    fdr_wilcox = p.adjust(wilcox_p, method = "BH"),
    fdr_wilcox_sci = sprintf("%.2e", fdr_wilcox),
    
    total_n       = n_shared + n_not_shared,
    z_score       = abs(qnorm(wilcox_p / 2)),
    effect_size_r = z_score / sqrt(total_n),
    
    label_facet   = sprintf("padj=%s, r=%.2f", fdr_wilcox_sci, effect_size_r)
  )

#Plot
difflog2fc_distribution_legs_plot = dge_btm_process_genes_diff_bygene_clean_filtered %>% 
  filter(status %in% c("Shared", "Not-shared"),
         treatment %in% c("Infection", "Injury")) %>%
  inner_join(stat_results, by = c("treatment", "pathogen")) %>% 
  ggplot() +
  aes(
    x     = abs_log2fc_diff,
    color = status,
    fill  = status
  ) +
  # Facetamento agrupado com o p-ajustado no título
  facet_nested_wrap(
    ~ treatment + pathogen + label_facet,
    nrow = 1
  ) +
  # Duas curvas de densidade (uma para cada status)
  geom_density(
    linewidth = 1,
    data = . %>% filter(status == "Shared")
  ) +
  geom_density(
    linewidth = 1,
    data = . %>% filter(status == "Not-shared")
  ) +
  # Boxplot alinhado na base do eixo Y
  geom_boxplot(
    aes(y = -0.1),
    width = 0.2, 
    outliers = FALSE,
    color = "black", show.legend = F
  ) +
      scale_y_continuous(expand = c(0.01, 0.01)) +
    scale_x_continuous(expand = c(0.01, 0.01)) +
   # Set manual color palette for model types
  scale_color_manual(values = c("Not-shared" = "black", "Shared" = "#4DBBD5FF")) +
  scale_fill_manual(values = c("Not-shared" = NA, "Shared" = "#4DBBD5FF")) +
  theme_vaxgo() +
  theme(
    legend.position = "top",
    panel.spacing   = grid::unit(0.7, "cm"),
    plot.margin     = ggplot2::margin(0.1, 0.5, 0.1, 0.1, "cm")
  ) +
  labs(
    x     = "|ΔMean Log2FC|",
    y     = "Density",
    fill  = "Status",
    color = "Status"
  )

difflog2fc_distribution_legs_plot

# Save to file
difflog2fc_distribution_legs_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_btms_legs_meanlog2fc.png"),
         width = 8,
         height = 3)
```



```{r fig.height=2, fig.width=8}
# Gene expression

legs_error_plot = dge_btm_process_genes_diff_bygene_clean_filtered %>% 
  filter(treatment %in% c("Infection", "Injury"),
         status %in% c("Shared", "Not-shared"),) %>%
  bind_rows(dge_btm_process_genes_diff_bygene_clean_filtered %>% 
  filter(treatment %in% c("Infection", "Injury"),
         status %in% c("Shared", "Not-shared"),) %>% 
    mutate(treatment = "Pooled",
           pathogen = "")) %>% 
  ggplot() +
  aes(x = status,
      y = inverse_se_mouse,
      color = status,
      fill =  status,
      ) +
  facet_nested_wrap(~treatment+pathogen,
                    nrow = 1, axes = "all",
                    # scales = "free_y"
                    ) +
  
   geom_violin(alpha = 1, 
              width = 1,
              color = NA) +
  
  # Boxplot centered on each status
  geom_boxplot(
    width = 0.3,
    alpha = 1,
    outliers = FALSE
  ) +
  # scale_y_continuous(expand = c(0.02, 0.02)) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison_color) +
  labs(x = "N",
       y = "1/SE (Mouse)",
       fill = "",
       color = "") +
  theme_vaxgo() +
  theme(legend.position = "top",
        axis.title.x = element_blank(), 
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
    # STATISTICAL TEST LAYER
  stat_compare_means(
    comparisons = list(c("Shared", "Not-shared")),
    method = "wilcox.test",
    p.adjust.method = "BH",
    label = "p.adj",
    label.y.np = "top", 
    vjust = 1.5,
    hide.ns = T, step.increase = 0,
    size = 2.5, tip.length = 0,
  ) 

#Plot
legs_error_plot

#Save
legs_error_plot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_legs_error_plot.png"),
         height = 2.5, width = 8)

```



##### Do LEGS interact with Error and divergence?

By treatment and pathogen
```{r fig.height=2.5, fig.width=8}
# Gene expression
legs_error_divergence_plot = dge_btm_process_genes_diff_bygene_clean_filtered %>% 
  filter(treatment %in% c("Infection", "Injury")) %>%
  bind_rows(dge_btm_process_genes_diff_bygene_clean_filtered %>% 
  filter(treatment %in% c("Infection", "Injury")) %>% 
    mutate(treatment = "Pooled",
           pathogen = "")) %>% 
  ggplot() +
  aes(x = inverse_se_mouse,
      y = abs_log2fc_diff,
      color = status,
      fill =  status,
      ) +
  facet_nested_wrap(~treatment + pathogen, nrow = 1, scales = "free_x") +

  # Regression
  geom_smooth(
    method = "lm",
    mapping = aes(color = status, fill = status),
    se = TRUE
  ) +
  
  #Marginal Boxplot
  # TOP  
  geom_violin(
    mapping = aes(
      x = inverse_se_mouse,         
      y = 3.5,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0.5,
    width = 0.75,
    color = NA,
    show.legend = FALSE
  ) +
  geom_boxplot(
    mapping = aes(
      x = inverse_se_mouse,         
      y = 3.5,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0, linewidth = 0.2,
    color = "black",
    width = 0.75,
    show.legend = FALSE
  ) +
  # RIGHT
  geom_violin(
    mapping = aes(
      x = 18,         
      y = abs_log2fc_diff,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0.5,
    width = 4,
    color = NA,
    show.legend = FALSE
  ) +
  geom_boxplot(
    mapping = aes(
      x = 18,         
      y = abs_log2fc_diff,
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0, 
    linewidth = 0.2,
    color = "black",
    width = 4,
    show.legend = FALSE
  ) +
  scale_y_continuous(limits = c(0, 4)) +
  scale_x_continuous(limits = c(0, 20))+
  
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(
    x = "1/SE (Mouse)",
    y = "|ΔMean Log2FC|",
    fill = "LEG status"
  ) +
  theme_vaxgo() +
  theme(
    legend.position = "top",
    # panel.spacing = grid::unit(0.7, "cm")
  ) +
  guides(
    color = "none", 
    fill = guide_legend(
      title.position = "left", 
      byrow = TRUE, 
      nrow = 1,
      override.aes = list(shape = 22, alpha = 1, size = 3, linetype = 0)
    )
  )

legs_error_divergence_plot 

#Save
legs_error_divergence_plot %>%
  ggsave(filename = here("Figures", "all_human_mouse_legs_error_divergence_plot_pooled.png"),
         height = 3, width = 16)


```





##### Is divergence in Log2FC associated with groups of processes?

```{r fig.height=3, fig.width=6}
##### Is divergence in Log2FC associated with groups of processes?
dge_btm_process_genes_diff_bygene_process = dge_btm_process_genes_diff_bygene_clean_filtered %>% 
  full_join(btm_annotation_genes %>% 
  select(process, hgnc_symbol = symbol, group, subgroup),
  by = "hgnc_symbol") %>% 
  mutate(process = if_else(is.na(process), "Not BTM", process),
         group = if_else(is.na(group), "Not BTM", group))



# 1. Prepare and filter data
plot_data <- dge_btm_process_genes_diff_bygene_process %>% 
  filter(
    status %in% c("All genes"),
    treatment %in% c("Infection", "Injury")
  ) %>%
  mutate(group = fct_relevel(group, c("Not BTM", immune_order)) %>% fct_rev()) %>% 
  select(-process) %>% 
  distinct()

# 2. Perform Wilcoxon test against "Not BTM" with BH adjustment
stat_results <- plot_data %>% 
  rstatix::wilcox_test(
    abs_log2fc_diff ~ group, 
    ref.group = "Not BTM", 
    p.adjust.method = "BH"
  ) 

# 3. Generate Plot
dge_btm_process_genes_diff_bygene_process_pooled_plot <- ggplot(
  plot_data, 
  aes(x = group, y = abs_log2fc_diff, color = group, fill = group)
) +
  geom_violin(alpha = 1, width = 1, color = NA) +
  geom_boxplot(width = 0.5, alpha = 0, color = "black", outliers = FALSE) +
  
  # STATISTICAL TEST LAYER
  stat_pvalue_manual(
    stat_results,
    x = "group2",             
    y.position = 3.5,           
    label = "p.adj",     
    remove.bracket = TRUE,    
    coord.flip = TRUE,
    size = 4, hide.ns = F,
    color = "black", hjust = 0
  ) +
  
  coord_flip() +
  scale_y_continuous(limits = c(0, 4.5), expand = c(0, 0)) +
  scale_fill_manual(values = btm_immune_groups$group) +
  scale_color_manual(values = btm_immune_groups$group) +
  labs(
    x = "",
    y = "|ΔMean Log2FC|",
    fill = "",
    color = ""
  ) +
  theme_vaxgo() +
  theme(legend.position = "none")

#Display plot
dge_btm_process_genes_diff_bygene_process_pooled_plot

#Save
dge_btm_process_genes_diff_bygene_process_pooled_plot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_dge_btm_process_genes_diff_bygene_process_pooled.png"),
         height = 4, width = 6)

```


