---
title: "3_Performance"
output: html_document
---
#Required

```{r}
# install.packages("here") if needed
library(here)
source(here("scripts_notebooks", "required.R"))
```

# Performance tests 

# 1. All conditions
## PR Curve: Genes

```{r}
#Import ============
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

#Metadata 
all_human_mouse_metadata = readRDS(here("tables", "all_human_mouse_metadata.rds")) 

all_human_mouse_metadata_matched =     #Duplicate infection timepoints in human to match the mouse
  bind_rows(
    all_human_mouse_metadata %>% 
      filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human")) %>% 
      select(pathogen, organism, timepoint),
    
  #Duplicate Human infection timepoints with mouse
  all_human_mouse_metadata %>% 
  filter((pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human")) %>% 
  select(pathogen, organism) %>% 
  inner_join(all_human_mouse_metadata %>% 
               filter(organism == "Mouse",
                      pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
               select(pathogen, timepoint) %>% 
               distinct(),
             by = "pathogen")
  ) %>% 
  group_by(pathogen, organism, timepoint) %>%
  summarise(n = n())

all_degs_human_mouse = 
  readRDS(file = here("tables", "all_human_mouse_dge_limma_degs_matched.rds")) %>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint %in% c("2h", "Early hours"), "Early hours", timepoint))

all_degs_human_mouse_matched = all_degs_human_mouse %>% 
       inner_join(all_human_mouse_metadata_matched,
               by = join_by("pathogen", "organism", "timepoint")) 

all_degs_human_mouse_matched 

#DEGs

#Functional modules
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

#Input ======================
dge_long_genes = all_degs_human_mouse_matched %>% 
  select(pathogen, treatment, organism, human_symbol, timepoint, mean_l2fc, adj_p_val, everything())

dge_long_genes %>% dplyr::count(pathogen,organism,  timepoint)

#All genes
genes_pr_allgenes = dge_long_genes
#Filter immune genes
genes_pr_allgenes_immunegenes = dge_long_genes %>% 
  filter(human_symbol %in% c(btm_annotation_genes %>% 
                         filter(module_category == "immune") %>% 
                         pull(symbol) %>% 
                         unique()))

#Only DEGs
degs_genes_human = dge_long_genes %>%
  filter(adj_p_val <= 0.05,
         organism == "Human") %>%
  select(human_symbol, pathogen, organism, timepoint)

#Only immune DEGs
genes_pr_DEGs_all = dge_long_genes %>%
  inner_join(degs_genes_human %>% 
               select(-organism),
             by = join_by("human_symbol", "pathogen", "timepoint"))

#Only immune DEGs
genes_pr_DEGs_immune = dge_long_genes %>%
  inner_join(degs_genes_human %>% 
               select(-organism),
             by = join_by("human_symbol", "pathogen", "timepoint")) %>%
  #Filter immune genes only
  filter(human_symbol %in% c(btm_annotation_genes$symbol %>% unique()))


degs_genes_human %>% 
  dplyr::count(pathogen, organism, timepoint)

genes_pr_DEGs_all %>% 
  dplyr::count(pathogen, organism, timepoint)

genes_pr_DEGs_immune %>% 
  dplyr::count(pathogen, timepoint)

```

Calculate PR curve

Inputs
Double check the code
```{r fig.height=4, fig.width=10}
set.seed(123)
# ######## Input -----
filename = "all_human_mouse"
gene_set_filename = "All_OnlyImmune"
genes_df_input = bind_rows(
  genes_pr_allgenes %>% mutate(model_type = "Mouse"),
  genes_pr_allgenes_immunegenes %>% mutate(model_type = "Immune")
)

# filename = "all_human_mouse"
# gene_set_filename = "HumanDEGs_OnlyImmune"
# genes_df_input = bind_rows(
#   genes_pr_DEGs_all %>% mutate(model_type = "Mouse"),
#   genes_pr_DEGs_immune %>% mutate(model_type = "Immune")
# )

```

Compute PR
```{r fig.height=6, fig.width=10}
#=====================================================
# Compute probabilities
#=====================================================

dge_long_genes_up = genes_df_input %>% 
  mutate(
    df = n - 1,
    pval_one_sided = pt(t, df, lower.tail = FALSE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t > 0, "up", "not_up"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  select(-adj_p_val) %>%
  rename(adj_p_val = pval_one_sided_adj)

dge_long_genes_down = genes_df_input %>% 
  mutate(
    df = n - 1,
    pval_one_sided = pt(t, df, lower.tail = TRUE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t < 0, "down", "not_down"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  select(-adj_p_val) %>%
  rename(adj_p_val = pval_one_sided_adj)

#=====================================================
# Assign human truth
#=====================================================
truth_human_up = dge_long_genes_up %>% 
  select(human_symbol, pathogen, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")


truth_human_down = dge_long_genes_down %>% 
  select(human_symbol, pathogen, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#=====================================================
#Generate reference distributions
#=====================================================
perfect_prediction_up = truth_human_up %>%
  mutate(prob = if_else(truth == "up", 1, 0),
         model = "Human") %>%
  rename(pred = truth)

permutation_prediction_up = perfect_prediction_up %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), organism = "Permutation",
         model = "Permutation")

perfect_prediction_down = truth_human_down %>%
  mutate(prob = if_else(truth == "down", 1, 0),
          model = "Human") %>%
  rename(pred = truth)

permutation_prediction_down = perfect_prediction_down %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), 
         organism = "Permutation",
         model = "Permutation")

#=====================================================
# Combine models (Mouse + Immune + Permutation + Perfect)
#=====================================================
models_up = dge_long_genes_up %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         organism,
         model = if_else(model_type == "Immune", "Immune", organism)) %>%
  bind_rows(perfect_prediction_up, permutation_prediction_up)

models_down = dge_long_genes_down %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = if_else(model_type == "Immune", "Immune", organism)) %>%
  bind_rows(perfect_prediction_down, permutation_prediction_down)

#=====================================================
# Classify probabilities 
#=====================================================
classified_df_probs_up = truth_human_up %>% 
  select(human_symbol, pathogen, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(human_symbol, organism, pathogen, timepoint, pred, prob, model),
             by = join_by(human_symbol, timepoint, pathogen)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct()

classified_df_probs_down = truth_human_down %>% 
  select(human_symbol, pathogen, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(human_symbol, organism, pathogen, timepoint, pred, prob, model),
             by = join_by(human_symbol, timepoint, pathogen)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct() 


classified_df_probs_up_DEGs = truth_human_up %>% 
  select(human_symbol, pathogen, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(human_symbol, organism, pathogen, timepoint, pred, prob, model),
             by = join_by(human_symbol, timepoint, pathogen)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct() %>% 
    inner_join(degs_genes_human %>% 
               select(-organism), 
             by = join_by(human_symbol, pathogen, timepoint))



classified_df_probs_down_DEGs = truth_human_down %>% 
  select(human_symbol, pathogen, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(human_symbol, organism, pathogen, timepoint, pred, prob, model),
             by = join_by(human_symbol, timepoint, pathogen)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct()  %>% 
    inner_join(degs_genes_human %>% 
               select(-organism), 
             by = join_by(human_symbol, pathogen, timepoint))

classified_df_probs_up %>% 
  dplyr::count(pathogen, model, timepoint)
classified_df_probs_up_DEGs %>% 
  dplyr::count(pathogen, model, timepoint)

classified_df_probs_down %>% 
  dplyr::count(pathogen, model, timepoint)
classified_df_probs_down_DEGs %>% 
  dplyr::count(pathogen, model, timepoint)
```


```{r fig.height=6, fig.width=10}
#=====================================================
#Compute PR 
#=====================================================

pr_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Up")

pr_curve_up_notup_DEGs = classified_df_probs_up_DEGs %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Up")

pr_curve_up_notup %>% 
  dplyr::count(pathogen, model, timepoint)

pr_curve_up_notup_DEGs %>% 
  dplyr::count(pathogen, model, timepoint)

pr_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Down")

pr_curve_down_notdown_DEGs = classified_df_probs_down_DEGs %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Down")

pr_curve_down_notdown %>% 
  dplyr::count(pathogen, model, timepoint)

pr_curve_down_notdown_DEGs %>% 
  dplyr::count(pathogen, model, timepoint)

```


```{r fig.height=6, fig.width=10}
#=====================================================
#Compute PR-AUC 
#=====================================================
pr_auc_results_up = classified_df_probs_up %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, pr_auc, comparison)

pr_pr_auc_results_up_DEGs = classified_df_probs_up_DEGs %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, pr_auc,  comparison)

pr_auc_results_down = classified_df_probs_down %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, pr_auc,  comparison)

pr_pr_auc_results_down_DEGs = classified_df_probs_down_DEGs %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, pr_auc,  comparison)


##### Unite all results
pr_curve_auc_df = bind_rows(pr_auc_results_up,
                             pr_auc_results_down) %>% 
  inner_join(
    pr_curve_up_notup %>% 
      bind_rows(pr_curve_down_notdown %>% 
                  filter(model != "Human")),
    by = join_by(pathogen, model, timepoint, comparison)) %>% 
    mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         # timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(pr_auc = round(pr_auc, 2) %>% as.character())

pr_curve_auc_df_DEGs = bind_rows(pr_pr_auc_results_up_DEGs,pr_pr_auc_results_down_DEGs) %>% 
  inner_join(
    pr_curve_up_notup_DEGs %>% 
      bind_rows(pr_curve_down_notdown_DEGs %>% 
                  filter(model != "Human")),
    by = join_by(pathogen, model, timepoint, comparison)) %>% 
    mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         # timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(pr_auc = round(pr_auc, 2) %>% as.character())

pr_curve_auc_df %>% 
  select(pathogen, model, comparison, timepoint, ) %>% 
  filter(model == "Immune") %>% 
              distinct()

pr_curve_auc_df_DEGs %>% 
  select(pathogen, model, comparison, timepoint, ) %>% 
  filter(model == "Immune") %>% 
              distinct()
```

Plot PR curve
```{r fig.height=6, fig.width=10}
#Plot curves ------------------
#PLOT PR Curve
pr_curve_plot = pr_curve_auc_df %>% 
  # mutate(comparison = as.character(comparison),
  #        comparison = if_else(comparison == "Up", "U", "D")) %>% 
  arrange(pathogen) %>% 
  ungroup() %>% 
  ggplot(aes(x = recall, 
             y = precision, 
             color = model,
             linetype = comparison)) +
  geom_path(size = 0.75, alpha = 1) +
  theme_vaxgo() +
  facet_nested_wrap(~pathogen+timepoint, nrow = 3, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Mouse
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint, pr_auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.4),
            size = 2,
            hjust = 1) +
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint,pr_auc ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.3),
            size = 2,
            hjust = 1) +
  #Immune
  geom_text(data = . %>% 
              filter(model == "Immune", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint, pr_auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.2),
            size = 2,
            hjust = 1) +
   geom_text(data = . %>% 
              filter(model == "Immune", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint,pr_auc ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.1),
            size = 2,
            hjust = 1) +
  scale_linetype_manual(values = c("Up" = 1,
                                   "Down" = 2)) +
  # coord_equal() +
  scale_color_manual(values = colors$organism) +
  labs(
  x = "Recall",
  y = "Precision",
  color = "Model",
  linetype = "Comparison") + 
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
          axis.title.x = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 8),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          legend.spacing.x = unit(0, 'cm'),
          legend.spacing.y = unit(0, 'cm'),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing.y = unit(0, "lines"), 
          panel.spacing.x = unit(1, "lines"), 
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          
          ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 0.5, size = 1),
                                 order = 2), 
         fill = "none",
         color = guide_legend(override.aes = list(linewidth = 3)))

pr_curve_plot


# pr_curve_plot %>%
#   ggsave(filename = here("Figures", paste0(filename, "_performance_pr_", gene_set_filename, ".png")),
#          width = 8,
#          height = 5)
```



PR-AUC as lineplots
```{r fig.height=12, fig.width=3}

#Get n genes
ngenes = bind_rows(
  classified_df_probs_up %>%
    filter(pred == "up") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Up") %>% 
    ungroup(),
  classified_df_probs_down %>%
    filter(pred == "down") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Down") %>% 
    ungroup(),
)


# Visualize AUC
pr_pr_auc_lineplot_df <- bind_rows(pr_auc_results_up,
                      pr_auc_results_down) %>%
  inner_join(ngenes, by = join_by(comparison, pathogen, timepoint, model)) %>% 
  mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 

pr_auc_lineplot = pr_pr_auc_lineplot_df %>% 
  ggplot() +
   aes(x = timepoint,
      y = pr_auc,
      fill = model,
      color = model,
      group = model) + 
  
  facet_nested_wrap(~treatment+pathogen+comparison,
                    scales = "free",
                    ncol = 2,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_line(width = 1,
            aes(linetype = model)) +
  
  geom_point(size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n_genes), size = 2) +
  # Add a reference line for random prediction (AUC = 0.5)
    # geom_text_repel(aes(label = auc), 
  #                 size = 2.5, 
  #                 # direction = "y", 
  #                 segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  # scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_linetype_manual(values = c(
    "Mouse" = 1,
    "Immune" = 1,
    "Human" = 1,
    "Permutation" = 2
    )) +
  scale_fill_manual(values = colors$organism) +
  scale_color_manual(values = colors$organism) +
  labs(
    x = "Time",
    y = "PR-AUC",
    fill = "Model",
    shape = "Model",
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(0, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) +
  guides(color = "none",
         shape = "none",
         linetype = "none",
         fill = guide_legend(nrow = 2, by.row = T))

pr_auc_lineplot
# pr_auc_lineplot %>%
#   ggsave(filename = here("Figures", "all_human_mouse_performance_pr_pr_auc_lineplot_All_OnlyImmune.png"),
#          width = 3,
#          height = 12)

```


```{r fig.height=12, fig.width=3}
#Get n genes
ngenes_DEGs = bind_rows(
  classified_df_probs_up_DEGs  %>%
    filter(pred == "up") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Up") %>% 
    ungroup(),
  classified_df_probs_down_DEGs %>%
    filter(pred == "down") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Down") %>% 
    ungroup(),
)

classified_df_probs_up %>% 
  dplyr::count(pathogen, model, timepoint, pred)
classified_df_probs_up_DEGs %>% 
  dplyr::count(pathogen, model, timepoint, pred)
```


```{r fig.height=12, fig.width=3}
# Visualize AUC
pr_pr_pr_auc_lineplot_df_DEGs <- bind_rows(pr_pr_auc_results_up_DEGs,
                      pr_pr_auc_results_down_DEGs) %>%
  inner_join(ngenes_DEGs, by = join_by(comparison, pathogen, timepoint, model)) %>% 
  mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 

pr_pr_auc_lineplot_DEGs = pr_pr_pr_auc_lineplot_df_DEGs %>% 
  ggplot() +
   aes(x = timepoint,
      y = pr_auc,
      fill = model,
      color = model,
      group = model) + 
  
  facet_nested_wrap(~treatment+pathogen+comparison,
                    scales = "free",
                    ncol = 2,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_line(width = 1,
            aes(linetype = model)) +
  
  geom_point(size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n_genes), size = 2) +
  # Add a reference line for random prediction (AUC = 0.5)
    # geom_text_repel(aes(label = auc), 
  #                 size = 2.5, 
  #                 # direction = "y", 
  #                 segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(0, 1)) +
  theme_vaxgo() +
  scale_linetype_manual(values = c(
    "Mouse" = 1,
    "Immune" = 1,
    "Human" = 1,
    "Permutation" = 2
    )) +
  scale_fill_manual(values = colors$organism) +
  scale_color_manual(values = colors$organism) +
  labs(
    x = "Time",
    y = "PR-AUC",
    fill = "Model",
    shape = "Model",
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(0, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) +
  guides(color = "none",
         shape = "none",
         linetype = "none",
         fill = guide_legend(nrow = 2, by.row = T))

pr_pr_auc_lineplot_DEGs
# pr_pr_auc_lineplot_DEGs %>%
#   ggsave(filename = here("Figures", "all_human_mouse_performance_pr_pr_pr_auc_lineplot_DEGs_OnlyImmune.png"),
#          width = 3,
#          height = 12)

```




## PR Curve: BTMs

### Means

```{r}
#Import ========================
#Unite BTM means
dge_btm_mean_process_saureus = readRDS(file = here("tables", paste0("saureus_infection_human_mouse_dge_btm_mean_process.rds")))
dge_btm_mean_process_ecoli = readRDS(file = here("tables", paste0("ecoli_infection_human_mouse_dge_btm_mean_process.rds")))
dge_btm_mean_process_fluad = readRDS(file = here("tables", paste0("influenza_fluad_human_mouse_dge_btm_mean_process.rds")))
dge_btm_mean_process_engerix  = readRDS(file = here("tables", paste0("hepatitisb_engerixb_human_mouse_dge_btm_mean_process.rds")))
dge_btm_mean_process_trauma  = readRDS(file = here("tables", paste0("trauma_human_mouse_dge_btm_mean_process.rds")))
dge_btm_mean_process_burn  = readRDS(file = here("tables", paste0("burn_human_mouse_dge_btm_mean_process.rds")))

dge_btm_mean_process_all = bind_rows(
  dge_btm_mean_process_saureus,
  dge_btm_mean_process_ecoli,
  dge_btm_mean_process_fluad %>% mutate(timepoint = as.character(timepoint)),
  dge_btm_mean_process_engerix %>% mutate(timepoint = as.character(timepoint)),
  dge_btm_mean_process_burn,
  dge_btm_mean_process_trauma
) %>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint == "2h", "Early hours", timepoint),
         pathogen = case_when(pathogen == "Saureus" ~ "Staphylococcus aureus",
                              pathogen == "Ecoli" ~ "Escherichia coli",
                              T~pathogen)) %>% 
  #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Influenza", "Hepatitis B") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Escherichia coli", "Staphylococcus aureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_comparison = if_else(pathogen %in% c("Escherichia coli", "Staphylococcus aureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early")) %>% 
  mutate(timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")))

dge_btm_mean_process_all %>% 
  dplyr::count(pathogen, organism, timepoint)


#Import ============
#Import DEGs table
dge_btm_mean_process = dge_btm_mean_process_all

#Import metadata
#Metadata 
all_human_mouse_metadata = readRDS(here("tables", "all_human_mouse_metadata.rds")) %>% 
  filter(!(timepoint %in% c("0h", "Day 0")))

all_human_mouse_metadata_matched = bind_rows(
  # 1. Keep everything for OTHER pathogens (neither S. aureus nor E. coli)
  all_human_mouse_metadata %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))) %>% 
    select(pathogen, organism, timepoint),
    
  # 2. Keep the original MOUSE timepoints for S. aureus and E. coli
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse") %>% 
    select(pathogen, organism, timepoint),
    
  # 3. Create the HUMAN timepoints mapped to Mouse timepoints for these pathogens
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(pathogen, organism) %>% # Keep 'Human' label, drop original timepoint
    inner_join(
      all_human_mouse_metadata %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
) %>% 
  group_by(pathogen, organism, timepoint) %>%
  summarise(n = n(), .groups = "drop") %>% 
  ungroup()



#Pre-process data =============

all_btms_human_mouse_matched = bind_rows(
  # 1. Keep everything that is NOT S. aureus or E. coli
  dge_btm_mean_process %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))),
  
  # 2. Keep the original MOUSE data for S. aureus and E. coli
  dge_btm_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse"),
    
  # 3. Add the DUPLICATED HUMAN data mapped to Mouse timepoints
  dge_btm_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(-timepoint) %>% 
    inner_join(
      all_human_mouse_metadata_matched %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
)

all_btms_human_mouse_matched %>% select(pathogen, organism, timepoint, ) %>% distinct()

all_btms_human_mouse = all_btms_human_mouse_matched %>% 
  inner_join(all_human_mouse_metadata_matched,
               by = join_by("pathogen", "organism", "timepoint")) %>% 
  distinct()
#Check
all_btms_human_mouse %>% select(pathogen, organism, timepoint, ) %>% distinct()

#Input ======================
set.seed(123)
filename = "all_human_mouse"
gene_set_filename = "BTMs"

btms_df_input = all_btms_human_mouse_matched %>% 
  rename(symbol = process)
```


Calculate PR curve


```{r fig.height=8, fig.width=10}
#=====================================================
# Compute probabilities
#=====================================================

dge_long_genes_up = btms_df_input %>% 
  mutate(
    df = n_genes - 1,
    t = mean_log2fc/(sd_log2fc/sqrt(n_genes)), 
    pval_one_sided = pt(t, df, lower.tail = FALSE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t > 0, "up", "not_up"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  rename(adj_p_val = pval_one_sided_adj)

dge_long_genes_up %>% 
  filter(pathogen == "Staphylococcus aureus",
         organism == "Human")

dge_long_genes_down = btms_df_input %>% 
  mutate(
    df = n_genes - 1,
    t = mean_log2fc/(sd_log2fc/sqrt(n_genes)), 
    pval_one_sided = pt(t, df, lower.tail = TRUE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t < 0, "down", "not_down"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  rename(adj_p_val = pval_one_sided_adj) %>% 
  drop_na(adj_p_val)

dge_long_genes_down %>% 
  filter(pathogen == "Staphylococcus aureus",
         organism == "Human")

#=====================================================
# Assign human truth
#=====================================================
truth_human_up = dge_long_genes_up %>% 
  select(pathogen, symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human") 

truth_human_down = dge_long_genes_down %>% 
  select(pathogen, symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

truth_human_up %>% 
  dplyr::count(pathogen, organism, timepoint)

#=====================================================
#Generate reference distributions
#=====================================================
perfect_prediction_up = truth_human_up %>%
  mutate(prob = if_else(truth == "up", 1, 0),
         model = "Human") %>%
  rename(pred = truth) 

permutation_prediction_up = perfect_prediction_up %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), organism = "Permutation",
         model = "Permutation")%>% 
  drop_na(pred)

perfect_prediction_down = truth_human_down %>%
  mutate(prob = if_else(truth == "down", 1, 0),
          model = "Human") %>%
  rename(pred = truth) %>% 
  drop_na(pred)

permutation_prediction_down = perfect_prediction_down %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), organism = "Permutation",
         model = "Permutation") %>% 
  drop_na(prob)


perfect_prediction_down %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
# Combine models (Mouse + Immune + Permutation + Perfect)
#=====================================================
models_up = dge_long_genes_up %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_up, permutation_prediction_up) %>% 
  drop_na(prob)

models_down = dge_long_genes_down %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_down, permutation_prediction_down) %>% 
  drop_na(prob)

models_down %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
# Classify probabilities 
#=====================================================
classified_df_probs_up = truth_human_up %>% 
  select(pathogen, symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(pathogen, symbol, timepoint, pred, prob, model),
             by = join_by(pathogen, symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct()

classified_df_probs_down = truth_human_down %>% 
  select(pathogen, symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(pathogen, symbol, timepoint, pred, prob, model),
             by = join_by(pathogen, symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct()

classified_df_probs_up %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
#Compute PR 
#=====================================================

pr_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  drop_na() %>% 
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Up")

pr_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  drop_na() %>% 
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Down")

pr_curve_up_notup %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
#Compute PR-AUC 
#=====================================================
pr_auc_results_up = classified_df_probs_up %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, , comparison)

pr_auc_results_down = classified_df_probs_down %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, , comparison)

pr_auc_results_down %>% 
  dplyr::count(pathogen, model, timepoint)


#=====================================================
#Unite all results
#=====================================================
pr_curve_auc_df = bind_rows(pr_auc_results_up,pr_auc_results_down) %>% 
  inner_join(
    pr_curve_up_notup %>% 
      bind_rows(pr_curve_down_notdown %>% 
                  filter(model != "Human")),
    by = join_by(pathogen, model, timepoint, comparison)) %>% 
    mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         # timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(pr_auc = round(pr_auc, 2) %>% as.character())

pr_curve_auc_df %>% 
  select(pathogen, model, comparison, timepoint, ) %>% 
  filter(model == "Immune") %>% 
              distinct()
```


```{r fig.height=13, fig.width=4}
#Plot curves ------------------
#PLOT PR Curve
pr_curve_plot = pr_curve_auc_df %>% 
  # mutate(comparison = as.character(comparison),
  #        comparison = if_else(comparison == "Up", "U", "D")) %>% 
  arrange(pathogen) %>% 
  ungroup() %>% 
  ggplot(aes(x = recall, 
             y = precision, 
             color = model,
             linetype = comparison)) +
  geom_path(size = 0.75, alpha = 1) +
  theme_vaxgo() +
  facet_nested_wrap(~pathogen+timepoint, ncol = 3, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Mouse
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint, ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.2),
            size = 2,
            hjust = 1) +
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint, ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.05),
            size = 2,
            hjust = 1) +

  scale_linetype_manual(values = c("Up" = 1,
                                   "Down" = 2)) +
  # coord_equal() +
  scale_color_manual(values = colors$organism) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model",
  linetype = "Comparison") + 
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
          axis.title.x = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 8),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          legend.spacing.x = unit(0, 'cm'),
          legend.spacing.y = unit(0, 'cm'),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing.y = unit(0, "lines"), 
          panel.spacing.x = unit(1, "lines"), 
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          
          ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 0.5,
                                                     size = 1),
                                 order = 2,
                                 nrow = 2), 
         fill = "none",
         color = guide_legend(override.aes = list(linewidth = 3),
                              nrow = 2))

pr_curve_plot


pr_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_pr_", gene_set_filename, ".png")),
         width = 5,
         height = 13)
```


```{r fig.height=12, fig.width=6}
#Unite plots
roc_auc_curve_plot = ((pr_curve_plot) /
  (auc_plot + theme(legend.position = "none")
   )
  ) +
  plot_layout(
  guides = "collect",
  heights = c(0.8, 1) 
) +
plot_annotation(
  theme = theme(
    legend.position = "top",
    plot.margin = margin(0.5, 0.5,0.5,0.5, "pt")
  )
) & 
theme(
  panel.spacing = unit(0, "lines"),
  # plot.margin = margin(0, 2, 0, 2, "pt"),
   legend.margin=margin(0,0,0,0),
  legend.title = element_text(size = 8),
  legend.text = element_text(size = 8),
   legend.box.margin=margin(1,0,-10,0),
   plot.margin = margin(0.1, 0.1, 0.1, 0.1, unit = "cm")
)
roc_auc_curve_plot
roc_auc_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_pr_", gene_set_filename, ".png")),
         width = 5,
         height = 12)

```


PR-AUC as lineplots
```{r fig.height=4, fig.width=12}

#Get n genes
ngenes = bind_rows(
  classified_df_probs_up %>%
    filter(pred == "up") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Up") %>% 
    ungroup(),
  classified_df_probs_down %>%
    filter(pred == "down") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Down") %>% 
    ungroup(),
)


# Visualize AUC
pr_pr_auc_lineplot_df <- bind_rows(pr_auc_results_up,
                      pr_auc_results_down) %>%
  inner_join(ngenes, by = join_by(comparison, pathogen, timepoint, model)) %>% 
  mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 

pr_auc_lineplot = pr_pr_auc_lineplot_df %>% 
  ggplot() +
   aes(x = timepoint,
      y = pr_auc,
      fill = model,
      color = model,
      group = model) + 
  
  facet_nested_wrap(~treatment+pathogen+comparison,
                    scales = "free",
                    nrow = 2,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_line(width = 1,
            aes(linetype = model)) +
  
  geom_point(size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n_genes), 
                  color = "black",
                  size = 3) +
  # Add a reference line for random prediction (AUC = 0.5)
    # geom_text_repel(aes(label = auc), 
  #                 size = 2.5, 
  #                 # direction = "y", 
  #                 segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  # scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_linetype_manual(values = c(
    "Mouse" = 1,
    "Immune" = 1,
    "Human" = 1,
    "Permutation" = 2
    )) +
  scale_fill_manual(values = colors$organism) +
  scale_color_manual(values = colors$organism) +
  labs(
    x = "Time",
    y = "PR-AUC",
    fill = "Model",
    shape = "Model",
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(0, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) +
  guides(color = "none",
         shape = "none",
         linetype = "none",
         fill = guide_legend(nrow = 1, by.row = T))

pr_auc_lineplot
pr_auc_lineplot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_pr_pr_auc_lineplot_", gene_set_filename, ".png")),
         width = 12,
         height = 4)

```






### NES
```{r}
#Unite BTM means and NES
gsea_btm_results_saureus = readRDS(here("tables", paste0("saureus_infection_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_ecoli = readRDS(here("tables", paste0("ecoli_infection_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_fluad = readRDS(here("tables", paste0("influenza_fluad_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_engerix = readRDS(here("tables", paste0("hepatitisb_engerixb_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_burn = readRDS(here("tables", paste0("burn_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_trauma = readRDS(here("tables", paste0("trauma_human_mouse_gsea_btm_results.rds")))

gsea_btm_results_all = bind_rows(
  gsea_btm_results_saureus,
  gsea_btm_results_ecoli,
  gsea_btm_results_fluad %>% mutate(timepoint = as.character(timepoint)),
  gsea_btm_results_engerix %>% mutate(timepoint = as.character(timepoint)),
  gsea_btm_results_trauma,
  gsea_btm_results_burn
)%>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint == "2h", "Early hours", timepoint),
         pathogen = case_when(pathogen == "Saureus" ~ "Staphylococcus aureus",
                              pathogen == "Ecoli" ~ "Escherichia coli",
                              T~pathogen)) %>% 
  #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Influenza", "Hepatitis B") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Escherichia coli", "Staphylococcus aureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_comparison = if_else(pathogen %in% c("Escherichia coli", "Staphylococcus aureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early")) %>% 
  mutate(timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")))


#Import ============
#Import DEGs table
dge_btm_mean_process = gsea_btm_results_all

#Import metadata
#Metadata 
all_human_mouse_metadata = readRDS(here("tables", "all_human_mouse_metadata.rds")) %>% 
  filter(!(timepoint %in% c("0h", "Day 0")))

all_human_mouse_metadata_matched = bind_rows(
  # 1. Keep everything for OTHER pathogens (neither S. aureus nor E. coli)
  all_human_mouse_metadata %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))) %>% 
    select(pathogen, organism, timepoint),
    
  # 2. Keep the original MOUSE timepoints for S. aureus and E. coli
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse") %>% 
    select(pathogen, organism, timepoint),
    
  # 3. Create the HUMAN timepoints mapped to Mouse timepoints for these pathogens
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(pathogen, organism) %>% # Keep 'Human' label, drop original timepoint
    inner_join(
      all_human_mouse_metadata %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
) %>% 
  group_by(pathogen, organism, timepoint) %>%
  summarise(n = n(), .groups = "drop") %>% 
  ungroup()



#Pre-process data =============

all_btms_human_mouse_matched = bind_rows(
  # 1. Keep everything that is NOT S. aureus or E. coli
  dge_btm_mean_process %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))),
  
  # 2. Keep the original MOUSE data for S. aureus and E. coli
  dge_btm_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse"),
    
  # 3. Add the DUPLICATED HUMAN data mapped to Mouse timepoints
  dge_btm_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(-timepoint) %>% 
    inner_join(
      all_human_mouse_metadata_matched %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
)

all_btms_human_mouse_matched %>% select(pathogen, organism, timepoint, ) %>% distinct()

all_btms_human_mouse = all_btms_human_mouse_matched %>% 
  inner_join(all_human_mouse_metadata_matched,
               by = join_by("pathogen", "organism", "timepoint")) %>% 
  distinct()
#Check
all_btms_human_mouse %>% select(pathogen, organism, timepoint, ) %>% distinct()

#Input ======================
set.seed(123)
filename = "all_human_mouse"
gene_set_filename = "BTMs_NES"

btms_df_input = all_btms_human_mouse_matched %>% 
  rename(symbol = process)
```
#### Significant Up/Down vs Not-significant Up/Down

Calculate PR curve
```{r fig.height=8, fig.width=10}
#=====================================================
# Compute probabilities
#=====================================================

dge_long_genes_up = btms_df_input %>% 
  mutate(
    df = n_genes - 1,
    t = nes/(sd_log2fc/sqrt(n_genes)), 
    pval_one_sided = pt(t, df, lower.tail = FALSE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t > 0, "up", "not_up"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  rename(adj_p_val = pval_one_sided_adj)

dge_long_genes_down = btms_df_input %>% 
  mutate(
    df = n_genes - 1,
    t = nes/(sd_log2fc/sqrt(n_genes)), 
    pval_one_sided = pt(t, df, lower.tail = TRUE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t < 0, "down", "not_down"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  rename(adj_p_val = pval_one_sided_adj) %>% 
  drop_na(adj_p_val)


#=====================================================
# Assign human truth
#=====================================================
truth_human_up = dge_long_genes_up %>% 
  select(pathogen, symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human") 

truth_human_down = dge_long_genes_down %>% 
  select(pathogen, symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

truth_human_up %>% 
  dplyr::count(pathogen, organism, timepoint)

#=====================================================
#Generate reference distributions
#=====================================================
perfect_prediction_up = truth_human_up %>%
  mutate(prob = if_else(truth == "up", 1, 0),
         model = "Human") %>%
  rename(pred = truth) 

permutation_prediction_up = perfect_prediction_up %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), organism = "Permutation",
         model = "Permutation")%>% 
  drop_na(pred)

perfect_prediction_down = truth_human_down %>%
  mutate(prob = if_else(truth == "down", 1, 0),
          model = "Human") %>%
  rename(pred = truth) %>% 
  drop_na(pred)

permutation_prediction_down = perfect_prediction_down %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), organism = "Permutation",
         model = "Permutation") %>% 
  drop_na(prob)


perfect_prediction_down %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
# Combine models (Mouse + Immune + Permutation + Perfect)
#=====================================================
models_up = dge_long_genes_up %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_up, permutation_prediction_up) %>% 
  drop_na(prob)

models_down = dge_long_genes_down %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_down, permutation_prediction_down) %>% 
  drop_na(prob)

models_down %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
# Classify probabilities 
#=====================================================
classified_df_probs_up = truth_human_up %>% 
  select(pathogen, symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(pathogen, symbol, timepoint, pred, prob, model),
             by = join_by(pathogen, symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct()

classified_df_probs_down = truth_human_down %>% 
  select(pathogen, symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(pathogen, symbol, timepoint, pred, prob, model),
             by = join_by(pathogen, symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct()

classified_df_probs_up %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
#Compute PR 
#=====================================================

pr_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  drop_na() %>% 
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Up")

pr_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  drop_na() %>% 
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Down")

pr_curve_up_notup %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
#Compute PR-AUC 
#=====================================================
pr_auc_results_up = classified_df_probs_up %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, , comparison)

pr_auc_results_down = classified_df_probs_down %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, , comparison)

pr_auc_results_down %>% 
  dplyr::count(pathogen, model, timepoint)


#=====================================================
#Unite all results
#=====================================================
pr_curve_auc_df = bind_rows(pr_auc_results_up,pr_auc_results_down) %>% 
  inner_join(
    pr_curve_up_notup %>% 
      bind_rows(pr_curve_down_notdown %>% 
                  filter(model != "Human")),
    by = join_by(pathogen, model, timepoint, comparison)) %>% 
    mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         # timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(pr_auc = round(pr_auc, 2) %>% as.character())


```


```{r fig.height=13, fig.width=4}
#Plot curves ------------------
#PLOT PR Curve
pr_curve_plot = pr_curve_auc_df %>% 
  # mutate(comparison = as.character(comparison),
  #        comparison = if_else(comparison == "Up", "U", "D")) %>% 
  arrange(pathogen) %>% 
  ungroup() %>% 
  ggplot(aes(x = recall, 
             y = precision, 
             color = model,
             linetype = comparison)) +
  geom_path(size = 0.75, alpha = 1) +
  theme_vaxgo() +
  facet_nested_wrap(~pathogen+timepoint, ncol = 3, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Mouse
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint, ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.2),
            size = 2,
            hjust = 1) +
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint, ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.05),
            size = 2,
            hjust = 1) +

  scale_linetype_manual(values = c("Up" = 1,
                                   "Down" = 2)) +
  # coord_equal() +
  scale_color_manual(values = colors$organism) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model",
  linetype = "Comparison") + 
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
          axis.title.x = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 8),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          legend.spacing.x = unit(0, 'cm'),
          legend.spacing.y = unit(0, 'cm'),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing.y = unit(0, "lines"), 
          panel.spacing.x = unit(1, "lines"), 
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          
          ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 0.5,
                                                     size = 1),
                                 order = 2,
                                 nrow = 2), 
         fill = "none",
         color = guide_legend(override.aes = list(linewidth = 3),
                              nrow = 2))

pr_curve_plot


pr_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_pr_", gene_set_filename, ".png")),
         width = 5,
         height = 13)
```



PR-AUC as lineplots
```{r fig.height=12, fig.width=4}

#Get n genes
ngenes = bind_rows(
  classified_df_probs_up %>%
    filter(pred == "up") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Up") %>% 
    ungroup(),
  classified_df_probs_down %>%
    filter(pred == "down") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Down") %>% 
    ungroup(),
)


# Visualize AUC
pr_pr_auc_lineplot_df <- bind_rows(pr_auc_results_up,
                      pr_auc_results_down) %>%
  inner_join(ngenes, by = join_by(comparison, pathogen, timepoint, model)) %>% 
  mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 

pr_auc_lineplot = pr_pr_auc_lineplot_df %>% 
  ggplot() +
   aes(x = timepoint,
      y = pr_auc,
      fill = model,
      color = model,
      group = model) + 
  
  facet_nested_wrap(~treatment+pathogen+comparison,
                    scales = "free",
                    ncol = 2,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_line(width = 1,
            aes(linetype = model)) +
  
  geom_point(size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n_genes), 
                  color = "black",
                  size = 3) +
  # Add a reference line for random prediction (AUC = 0.5)
    # geom_text_repel(aes(label = auc), 
  #                 size = 2.5, 
  #                 # direction = "y", 
  #                 segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  # scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_linetype_manual(values = c(
    "Mouse" = 1,
    "Immune" = 1,
    "Human" = 1,
    "Permutation" = 2
    )) +
  scale_fill_manual(values = colors$organism) +
  scale_color_manual(values = colors$organism) +
  labs(
    x = "Time",
    y = "PR-AUC",
    fill = "Model",
    shape = "Model",
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(0, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) +
  guides(color = "none",
         shape = "none",
         linetype = "none",
         fill = guide_legend(nrow = 1, by.row = T))

pr_auc_lineplot
# pr_auc_lineplot %>%
#   ggsave(filename = here("Figures", paste0(filename, "_performance_pr_pr_auc_lineplot_", gene_set_filename, ".png")),
#          width = 12,
#          height = 4)

pr_auc_lineplot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_pr_pr_auc_lineplot_", gene_set_filename, "_vertical.png")),
         width = 4,
         height = 12)

```


#####Classify by ranks

```{r fig.height=5, fig.width=10}
# Classificar em quartis para ver a acurácia por faixa
ranking_validation = classified_df_probs_up %>%
  group_by(pathogen, model, timepoint) %>%
  arrange(prob) %>% 
  mutate(quartile = ntile(prob, 2)) %>% # Divide em 2 partes
  group_by(pathogen, model, timepoint, quartile) %>%
  # filter(quartile == 3) %>% # Foca no topo (os 25% maiores NES/Prob)
  summarise(accuracy_in_top_quartile = sum(truth == "up") / n()) %>% 
  
  #Clean
  mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         # comparison = factor(comparison, levels = c("Up", "Down")),
         timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 


# Plotar para comparar os modelos
  ggplot(ranking_validation) +
    aes(
      x = timepoint,
      y = accuracy_in_top_quartile,
      color = model,
      shape = as.character(quartile)
    ) +
  geom_line(aes(group = interaction(model, quartile))) +
  geom_point() +
  # ylim(0, 1) +
  facet_wrap(~treatment+pathogen, 
             nrow= 1,
             scales = "free_x") +
  geom_hline(yintercept = 0.5,
             linetype = 2) +
  theme_vaxgo() +
    scale_color_manual(values = c("Human" = "black",
                                  "Mouse" = "blue", 
                                  "Permutation" = "gray")) +
    # scale_color_viridis_d(option = "mako", direction = -1) +
    theme(legend.position = "top") +
  labs(title = "Accuracy by quartile")

```







## PR Curve: Hallmarks

### Means Hallmarks

```{r}
#Import ========================
#Unite BTM means
dge_hallmarks_mean_process_saureus = readRDS(file = here("tables", paste0("saureus_infection_human_mouse_dge_hallmarks_mean_process.rds")))
dge_hallmarks_mean_process_ecoli = readRDS(file = here("tables", paste0("ecoli_infection_human_mouse_dge_hallmarks_mean_process.rds")))
dge_hallmarks_mean_process_fluad = readRDS(file = here("tables", paste0("influenza_fluad_human_mouse_dge_hallmarks_mean_process.rds")))
dge_hallmarks_mean_process_engerix  = readRDS(file = here("tables", paste0("hepatitisb_engerixb_human_mouse_dge_hallmarks_mean_process.rds")))
dge_hallmarks_mean_process_trauma  = readRDS(file = here("tables", paste0("trauma_human_mouse_dge_hallmarks_mean_process.rds")))
dge_hallmarks_mean_process_burn  = readRDS(file = here("tables", paste0("burn_human_mouse_dge_hallmarks_mean_process.rds")))

dge_hallmarks_mean_process_all = bind_rows(
  dge_hallmarks_mean_process_saureus,
  dge_hallmarks_mean_process_ecoli,
  dge_hallmarks_mean_process_fluad %>% mutate(timepoint = as.character(timepoint)),
  dge_hallmarks_mean_process_engerix %>% mutate(timepoint = as.character(timepoint)),
  dge_hallmarks_mean_process_burn,
  dge_hallmarks_mean_process_trauma
) %>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint == "2h", "Early hours", timepoint),
         pathogen = case_when(pathogen == "Saureus" ~ "Staphylococcus aureus",
                              pathogen == "Ecoli" ~ "Escherichia coli",
                              T~pathogen)) %>% 
  #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Influenza", "Hepatitis B") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Escherichia coli", "Staphylococcus aureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_comparison = if_else(pathogen %in% c("Escherichia coli", "Staphylococcus aureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early")) %>% 
  mutate(timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")))

dge_hallmarks_mean_process_all %>% 
  dplyr::count(pathogen, organism, timepoint)


#Import ============
#Import DEGs table
dge_hallmarks_mean_process = dge_hallmarks_mean_process_all

#Import metadata
#Metadata 
all_human_mouse_metadata = readRDS(here("tables", "all_human_mouse_metadata.rds")) %>% 
  filter(!(timepoint %in% c("0h", "Day 0")))

all_human_mouse_metadata_matched = bind_rows(
  # 1. Keep everything for OTHER pathogens (neither S. aureus nor E. coli)
  all_human_mouse_metadata %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))) %>% 
    select(pathogen, organism, timepoint),
    
  # 2. Keep the original MOUSE timepoints for S. aureus and E. coli
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse") %>% 
    select(pathogen, organism, timepoint),
    
  # 3. Create the HUMAN timepoints mapped to Mouse timepoints for these pathogens
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(pathogen, organism) %>% # Keep 'Human' label, drop original timepoint
    inner_join(
      all_human_mouse_metadata %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
) %>% 
  group_by(pathogen, organism, timepoint) %>%
  summarise(n = n(), .groups = "drop") %>% 
  ungroup()



#Pre-process data =============

all_hallmarks_human_mouse_matched = bind_rows(
  # 1. Keep everything that is NOT S. aureus or E. coli
  dge_hallmarks_mean_process %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))),
  
  # 2. Keep the original MOUSE data for S. aureus and E. coli
  dge_hallmarks_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse"),
    
  # 3. Add the DUPLICATED HUMAN data mapped to Mouse timepoints
  dge_hallmarks_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(-timepoint) %>% 
    inner_join(
      all_human_mouse_metadata_matched %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
)

all_hallmarks_human_mouse_matched %>% select(pathogen, organism, timepoint, ) %>% distinct()

all_hallmarks_human_mouse = all_hallmarks_human_mouse_matched %>% 
  inner_join(all_human_mouse_metadata_matched,
               by = join_by("pathogen", "organism", "timepoint")) %>% 
  distinct()
#Check
all_hallmarks_human_mouse %>% select(pathogen, organism, timepoint, ) %>% distinct()

#Input ======================
set.seed(123)
filename = "all_human_mouse"
gene_set_filename = "Hallmarks"

hallmarks_df_input = all_hallmarks_human_mouse_matched %>% 
  rename(symbol = process)
```


Calculate PR curve


```{r fig.height=8, fig.width=10}
#=====================================================
# Compute probabilities
#=====================================================

dge_long_genes_up = hallmarks_df_input %>% 
  mutate(
    df = n_genes - 1,
    t = mean_log2fc/(sd_log2fc/sqrt(n_genes)), 
    pval_one_sided = pt(t, df, lower.tail = FALSE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t > 0, "up", "not_up"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  rename(adj_p_val = pval_one_sided_adj)

dge_long_genes_up %>% 
  filter(pathogen == "Staphylococcus aureus",
         organism == "Human")

dge_long_genes_down = hallmarks_df_input %>% 
  mutate(
    df = n_genes - 1,
    t = mean_log2fc/(sd_log2fc/sqrt(n_genes)), 
    pval_one_sided = pt(t, df, lower.tail = TRUE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t < 0, "down", "not_down"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  rename(adj_p_val = pval_one_sided_adj) %>% 
  drop_na(adj_p_val)

dge_long_genes_down %>% 
  filter(pathogen == "Staphylococcus aureus",
         organism == "Human")

#=====================================================
# Assign human truth
#=====================================================
truth_human_up = dge_long_genes_up %>% 
  select(pathogen, symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human") 

truth_human_down = dge_long_genes_down %>% 
  select(pathogen, symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

truth_human_up %>% 
  dplyr::count(pathogen, organism, timepoint)

#=====================================================
#Generate reference distributions
#=====================================================
perfect_prediction_up = truth_human_up %>%
  mutate(prob = if_else(truth == "up", 1, 0),
         model = "Human") %>%
  rename(pred = truth) 

permutation_prediction_up = perfect_prediction_up %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), organism = "Permutation",
         model = "Permutation")%>% 
  drop_na(pred)

perfect_prediction_down = truth_human_down %>%
  mutate(prob = if_else(truth == "down", 1, 0),
          model = "Human") %>%
  rename(pred = truth) %>% 
  drop_na(pred)

permutation_prediction_down = perfect_prediction_down %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), organism = "Permutation",
         model = "Permutation") %>% 
  drop_na(prob)


perfect_prediction_down %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
# Combine models (Mouse + Immune + Permutation + Perfect)
#=====================================================
models_up = dge_long_genes_up %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_up, permutation_prediction_up) %>% 
  drop_na(prob)

models_down = dge_long_genes_down %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_down, permutation_prediction_down) %>% 
  drop_na(prob)

models_down %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
# Classify probabilities 
#=====================================================
classified_df_probs_up = truth_human_up %>% 
  select(pathogen, symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(pathogen, symbol, timepoint, pred, prob, model),
             by = join_by(pathogen, symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct()

classified_df_probs_down = truth_human_down %>% 
  select(pathogen, symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(pathogen, symbol, timepoint, pred, prob, model),
             by = join_by(pathogen, symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct()

classified_df_probs_up %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
#Compute PR 
#=====================================================

pr_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  drop_na() %>% 
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Up")

pr_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  drop_na() %>% 
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Down")

pr_curve_up_notup %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
#Compute PR-AUC 
#=====================================================
pr_auc_results_up = classified_df_probs_up %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, , comparison)

pr_auc_results_down = classified_df_probs_down %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, , comparison)

pr_auc_results_down %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
#Unite all results
#=====================================================
pr_curve_auc_df = bind_rows(pr_auc_results_up,pr_auc_results_down) %>% 
  inner_join(
    pr_curve_up_notup %>% 
      bind_rows(pr_curve_down_notdown %>% 
                  filter(model != "Human")),
    by = join_by(pathogen, model, timepoint, comparison)) %>% 
    mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         # timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(pr_auc = round(pr_auc, 2) %>% as.character())


```


```{r fig.height=13, fig.width=4}
#Plot curves ------------------
#PLOT PR Curve
pr_curve_plot = pr_curve_auc_df %>% 
  # mutate(comparison = as.character(comparison),
  #        comparison = if_else(comparison == "Up", "U", "D")) %>% 
  arrange(pathogen) %>% 
  ungroup() %>% 
  ggplot(aes(x = recall, 
             y = precision, 
             color = model,
             linetype = comparison)) +
  geom_path(size = 0.75, alpha = 1) +
  theme_vaxgo() +
  facet_nested_wrap(~pathogen+timepoint, ncol = 3, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Mouse
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint, ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.2),
            size = 2,
            hjust = 1) +
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint, ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.05),
            size = 2,
            hjust = 1) +

  scale_linetype_manual(values = c("Up" = 1,
                                   "Down" = 2)) +
  # coord_equal() +
  scale_color_manual(values = colors$organism) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model",
  linetype = "Comparison") + 
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
          axis.title.x = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 8),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          legend.spacing.x = unit(0, 'cm'),
          legend.spacing.y = unit(0, 'cm'),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing.y = unit(0, "lines"), 
          panel.spacing.x = unit(1, "lines"), 
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          
          ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 0.5,
                                                     size = 1),
                                 order = 2,
                                 nrow = 2), 
         fill = "none",
         color = guide_legend(override.aes = list(linewidth = 3),
                              nrow = 2))

pr_curve_plot


pr_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_pr_", gene_set_filename, ".png")),
         width = 5,
         height = 13)
```

PR-AUC as lineplots
```{r fig.height=4, fig.width=12}

#Get n genes
ngenes = bind_rows(
  classified_df_probs_up %>%
    filter(pred == "up") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Up") %>% 
    ungroup(),
  classified_df_probs_down %>%
    filter(pred == "down") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Down") %>% 
    ungroup(),
)


# Visualize AUC
pr_pr_auc_lineplot_df <- bind_rows(pr_auc_results_up,
                      pr_auc_results_down) %>%
  inner_join(ngenes, by = join_by(comparison, pathogen, timepoint, model)) %>% 
  mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 

pr_auc_lineplot = pr_pr_auc_lineplot_df %>% 
  ggplot() +
   aes(x = timepoint,
      y = pr_auc,
      fill = model,
      color = model,
      group = model) + 
  
  facet_nested_wrap(~treatment+pathogen+comparison,
                    scales = "free",
                    nrow = 2,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_line(width = 1,
            aes(linetype = model)) +
  
  geom_point(size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n_genes), 
                  color = "black",
                  size = 3) +
  # Add a reference line for random prediction (AUC = 0.5)
    # geom_text_repel(aes(label = auc), 
  #                 size = 2.5, 
  #                 # direction = "y", 
  #                 segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  # scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_linetype_manual(values = c(
    "Mouse" = 1,
    "Immune" = 1,
    "Human" = 1,
    "Permutation" = 2
    )) +
  scale_fill_manual(values = colors$organism) +
  scale_color_manual(values = colors$organism) +
  labs(
    x = "Time",
    y = "PR-AUC",
    fill = "Model",
    shape = "Model",
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(0, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) +
  guides(color = "none",
         shape = "none",
         linetype = "none",
         fill = guide_legend(nrow = 1, by.row = T))

pr_auc_lineplot
pr_auc_lineplot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_pr_pr_auc_lineplot_", gene_set_filename, ".png")),
         width = 12,
         height = 4)

```





### NES

```{r}
#Unite Hallmarks means and NES
gsea_hallmarks_results_saureus = readRDS(here("tables", paste0("saureus_infection_human_mouse_gsea_hallmarks_results.rds")))
gsea_hallmarks_results_ecoli = readRDS(here("tables", paste0("ecoli_infection_human_mouse_gsea_hallmarks_results.rds")))
gsea_hallmarks_results_fluad = readRDS(here("tables", paste0("influenza_fluad_human_mouse_gsea_hallmarks_results.rds")))
gsea_hallmarks_results_engerix = readRDS(here("tables", paste0("hepatitisb_engerixb_human_mouse_gsea_hallmarks_results.rds")))
gsea_hallmarks_results_burn = readRDS(here("tables", paste0("burn_human_mouse_gsea_hallmarks_results.rds")))
gsea_hallmarks_results_trauma = readRDS(here("tables", paste0("trauma_human_mouse_gsea_hallmarks_results.rds")))

gsea_hallmarks_results_all = bind_rows(
  gsea_hallmarks_results_saureus,
  gsea_hallmarks_results_ecoli,
  gsea_hallmarks_results_fluad %>% mutate(timepoint = as.character(timepoint)),
  gsea_hallmarks_results_engerix %>% mutate(timepoint = as.character(timepoint)),
  gsea_hallmarks_results_trauma,
  gsea_hallmarks_results_burn
)%>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint == "2h", "Early hours", timepoint),
         pathogen = case_when(pathogen == "Saureus" ~ "Staphylococcus aureus",
                              pathogen == "Ecoli" ~ "Escherichia coli",
                              T~pathogen)) %>% 
  #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Influenza", "Hepatitis B") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Escherichia coli", "Staphylococcus aureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_comparison = if_else(pathogen %in% c("Escherichia coli", "Staphylococcus aureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early")) %>% 
  mutate(timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")))


#Import ============
#Import DEGs table
dge_hallmarks_mean_process = gsea_hallmarks_results_all

#Import metadata
#Metadata 
all_human_mouse_metadata = readRDS(here("tables", "all_human_mouse_metadata.rds")) %>% 
  filter(!(timepoint %in% c("0h", "Day 0")))

all_human_mouse_metadata_matched = bind_rows(
  # 1. Keep everything for OTHER pathogens (neither S. aureus nor E. coli)
  all_human_mouse_metadata %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))) %>% 
    select(pathogen, organism, timepoint),
    
  # 2. Keep the original MOUSE timepoints for S. aureus and E. coli
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse") %>% 
    select(pathogen, organism, timepoint),
    
  # 3. Create the HUMAN timepoints mapped to Mouse timepoints for these pathogens
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(pathogen, organism) %>% # Keep 'Human' label, drop original timepoint
    inner_join(
      all_human_mouse_metadata %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
) %>% 
  group_by(pathogen, organism, timepoint) %>%
  summarise(n = n(), .groups = "drop") %>% 
  ungroup()



#Pre-process data =============

all_hallmarks_human_mouse_matched = bind_rows(
  # 1. Keep everything that is NOT S. aureus or E. coli
  dge_hallmarks_mean_process %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))),
  
  # 2. Keep the original MOUSE data for S. aureus and E. coli
  dge_hallmarks_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse"),
    
  # 3. Add the DUPLICATED HUMAN data mapped to Mouse timepoints
  dge_hallmarks_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(-timepoint) %>% 
    inner_join(
      all_human_mouse_metadata_matched %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
)

all_hallmarks_human_mouse_matched %>% select(pathogen, organism, timepoint, ) %>% distinct()

all_hallmarks_human_mouse = all_hallmarks_human_mouse_matched %>% 
  inner_join(all_human_mouse_metadata_matched,
               by = join_by("pathogen", "organism", "timepoint")) %>% 
  distinct()
#Check
all_hallmarks_human_mouse %>% select(pathogen, organism, timepoint, ) %>% distinct()

#Input ======================
set.seed(123)
filename = "all_human_mouse"
gene_set_filename = "Hallmarks_NES"

hallmarks_df_input = all_hallmarks_human_mouse_matched %>% 
  rename(symbol = process)
```


Calculate PR curve


```{r fig.height=8, fig.width=10}
#=====================================================
# Compute probabilities
#=====================================================
# The script derives a t-score by scaling the NES by the standard error of the log_2FC values within each hallmark. It then computes one-sided p-values using the Student's t-distribution, applies a Benjamini-Hochberg (BH) correction for multiple testing, and classifies pathways as significantly "up-regulated" or "down-regulated" based on a 5% False Discovery Rate (FDR) threshold.

dge_long_genes_up <- hallmarks_df_input %>% 
  mutate(
    # 1. Degrees of freedom
    df = n_genes - 1,
    
    # 2. Calculate t-statistic (NES / Standard Error)
    t = nes / (sd_log2fc / sqrt(n_genes)), 
    
    # 3. Compute one-sided p-value (upper tail)
    pval_one_sided = pt(t, df, lower.tail = FALSE),
    
    # 4. Multiple testing correction (Benjamini-Hochberg)
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    
    # 5. Classify significant up-regulation
    direction = if_else(pval_one_sided_adj <= 0.05 & t > 0, "up", "not_up"),
    
    # 6. Probability score (Complement of adjusted p-value)
    prob = 1 - pval_one_sided_adj
  ) %>%
  # Rename for consistency with downstream pipelines
  rename(adj_p_val = pval_one_sided_adj)


dge_long_genes_down <- hallmarks_df_input %>% 
  mutate(
    df = n_genes - 1,
    t = nes / (sd_log2fc / sqrt(n_genes)), 
    # Compute one-sided p-value (lower tail for down-regulation)
    pval_one_sided = pt(t, df, lower.tail = TRUE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    
    # Classify significant down-regulation
    direction = if_else(pval_one_sided_adj <= 0.05 & t < 0, "down", "not_down"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  # 7. Cleanup and final renaming
  rename(adj_p_val = pval_one_sided_adj) %>% 
  drop_na(adj_p_val)


#=====================================================
# Assign human truth
#=====================================================
truth_human_up = dge_long_genes_up %>% 
  select(pathogen, symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human") 

truth_human_down = dge_long_genes_down %>% 
  select(pathogen, symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

truth_human_up %>% 
  dplyr::count(pathogen, organism, timepoint)

#=====================================================
#Generate reference distributions
#=====================================================
perfect_prediction_up = truth_human_up %>%
  mutate(prob = if_else(truth == "up", 1, 0),
         model = "Human") %>%
  rename(pred = truth) 

permutation_prediction_up = perfect_prediction_up %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), organism = "Permutation",
         model = "Permutation")%>% 
  drop_na(pred)

perfect_prediction_down = truth_human_down %>%
  mutate(prob = if_else(truth == "down", 1, 0),
          model = "Human") %>%
  rename(pred = truth) %>% 
  drop_na(pred)

permutation_prediction_down = perfect_prediction_down %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), organism = "Permutation",
         model = "Permutation") %>% 
  drop_na(prob)


perfect_prediction_down %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
# Combine models (Mouse + Immune + Permutation + Perfect)
#=====================================================
models_up = dge_long_genes_up %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_up, permutation_prediction_up) %>% 
  drop_na(prob)

models_down = dge_long_genes_down %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_down, permutation_prediction_down) %>% 
  drop_na(prob)

models_down %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
# Classify probabilities 
#=====================================================
classified_df_probs_up = truth_human_up %>% 
  select(pathogen, symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(pathogen, symbol, timepoint, pred, prob, model),
             by = join_by(pathogen, symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct()

classified_df_probs_down = truth_human_down %>% 
  select(pathogen, symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(pathogen, symbol, timepoint, pred, prob, model),
             by = join_by(pathogen, symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct()

classified_df_probs_up %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
#Compute PR 
#=====================================================

pr_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  drop_na() %>% 
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Up")

pr_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  drop_na() %>% 
  group_by(pathogen, model, timepoint) %>%
  pr_curve(truth, prob) %>%
  mutate(comparison = "Down")

pr_curve_up_notup %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
#Compute PR-AUC 
#=====================================================
pr_auc_results_up = classified_df_probs_up %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, , comparison)

pr_auc_results_down = classified_df_probs_down %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  pr_auc(truth, prob) %>%
  rename(pr_auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, , comparison)

pr_auc_results_down %>% 
  dplyr::count(pathogen, model, timepoint)
#=====================================================
#Unite all results
#=====================================================
pr_curve_auc_df = bind_rows(pr_auc_results_up,pr_auc_results_down) %>% 
  inner_join(
    pr_curve_up_notup %>% 
      bind_rows(pr_curve_down_notdown %>% 
                  filter(model != "Human")),
    by = join_by(pathogen, model, timepoint, comparison)) %>% 
    mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         # timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(pr_auc = round(pr_auc, 2) %>% as.character())


```


```{r fig.height=13, fig.width=4}
#Plot curves ------------------
#PLOT PR Curve
pr_curve_plot = pr_curve_auc_df %>% 
  # mutate(comparison = as.character(comparison),
  #        comparison = if_else(comparison == "Up", "U", "D")) %>% 
  arrange(pathogen) %>% 
  ungroup() %>% 
  ggplot(aes(x = recall, 
             y = precision, 
             color = model,
             linetype = comparison)) +
  geom_path(size = 0.75, alpha = 1) +
  theme_vaxgo() +
  facet_nested_wrap(~pathogen+timepoint, ncol = 3, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Mouse
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint, ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.2),
            size = 2,
            hjust = 1) +
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint, ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 1,
                y = 0.05),
            size = 2,
            hjust = 1) +

  scale_linetype_manual(values = c("Up" = 1,
                                   "Down" = 2)) +
  # coord_equal() +
  scale_color_manual(values = colors$organism) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model",
  linetype = "Comparison") + 
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
          axis.title.x = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 8),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          legend.spacing.x = unit(0, 'cm'),
          legend.spacing.y = unit(0, 'cm'),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing.y = unit(0, "lines"), 
          panel.spacing.x = unit(1, "lines"), 
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          
          ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 0.5,
                                                     size = 1),
                                 order = 2,
                                 nrow = 2), 
         fill = "none",
         color = guide_legend(override.aes = list(linewidth = 3),
                              nrow = 2))

pr_curve_plot


pr_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_pr_", gene_set_filename, ".png")),
         width = 5,
         height = 13)
```



PR-AUC as lineplots
```{r fig.height=4, fig.width=12}

#Get n genes
ngenes = bind_rows(
  classified_df_probs_up %>%
    filter(pred == "up") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Up") %>% 
    ungroup(),
  classified_df_probs_down %>%
    filter(pred == "down") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Down") %>% 
    ungroup(),
)


# Visualize AUC
pr_pr_auc_lineplot_df <- bind_rows(pr_auc_results_up,
                      pr_auc_results_down) %>%
  inner_join(ngenes, by = join_by(comparison, pathogen, timepoint, model)) %>% 
  mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 

pr_auc_lineplot = pr_pr_auc_lineplot_df %>% 
  ggplot() +
   aes(x = timepoint,
      y = pr_auc,
      fill = model,
      color = model,
      group = model) + 
  
  facet_nested_wrap(~treatment+pathogen+comparison,
                    scales = "free",
                    nrow = 2,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_line(width = 1,
            aes(linetype = model)) +
  
  geom_point(size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n_genes), 
                  color = "black",
                  size = 3) +
  # Add a reference line for random prediction (AUC = 0.5)
    # geom_text_repel(aes(label = auc), 
  #                 size = 2.5, 
  #                 # direction = "y", 
  #                 segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  # scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_linetype_manual(values = c(
    "Mouse" = 1,
    "Immune" = 1,
    "Human" = 1,
    "Permutation" = 2
    )) +
  scale_fill_manual(values = colors$organism) +
  scale_color_manual(values = colors$organism) +
  labs(
    x = "Time",
    y = "PR-AUC",
    fill = "Model",
    shape = "Model",
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(0, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) +
  guides(color = "none",
         shape = "none",
         linetype = "none",
         fill = guide_legend(nrow = 1, by.row = T))

pr_auc_lineplot
pr_auc_lineplot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_pr_pr_auc_lineplot_", gene_set_filename, ".png")),
         width = 12,
         height = 4)

```










#===================================
# (ADAPT) Fluad, Vaccine

## Regression model
### Compute overall metrics for all genes
```{r}
#Import
pairwise_human_fit_test = readRDS(here("FIT_Mouse2man", "pairwise_human_fit_measures.rds"))
pairwise_human_fit_measures = pairwise_human_fit_test
pairwise_human_fit_test$symbol %>% unique() %>% length() #6373

# Calculate metrics --------
#FIT------
#Define ranges of values for combinatorial analysis
filter_conditions_fit <- expand_grid(
  padj_fit = c(0.01, 0.05, seq(0.1, 1, by = 0.1)),
  expr_cutoff = c(0.01, seq(0.05, 0.5, by = 0.05), 0.75, 1),
  ci_percentile = c(0.3, 1.0),
  timepoint = c(1, 3, 7)
)

#Run analysis
filter_conditions_results_fit_degs <- filter_conditions_fit %>%
  rowwise() %>%
  do({
    fc <- .
    
    # Filter timepoint
    data_timepoint <- pairwise_human_fit_test %>%
      filter(timepoint == fc$timepoint)

    # Calculate the cutoff for each percentile
    ci_cutoff <- quantile(data_timepoint$ci_width_fit, probs = fc$ci_percentile, na.rm = TRUE)

    # Apply cutoffs
    filtered_data <- data_timepoint %>%
      filter(
        pval_adj_fit < fc$padj_fit,
        abs(mean_fit) > fc$expr_cutoff,
        ci_width_fit <= ci_cutoff
      )

    if (nrow(filtered_data) >= 2) {
      metrics <- filtered_data %>%
        summarise(
          n_genes = n(),
          RMSE = rmse_vec(mean_human, mean_fit),
          MAE = mae_vec(mean_human, mean_fit),
          R2 = rsq_vec(mean_human, mean_fit),
          Spearman = cor(mean_human, mean_fit, method = "spearman"),
          Pearson = cor(mean_human, mean_fit, method = "pearson")
        )
    } else {
      metrics <- tibble(
        n_genes = 0,
        RMSE = NA, MAE = NA, R2 = NA,
        Spearman = NA, Pearson = NA
      )
    }

    bind_cols(fc, tibble(ci_width_cutoff = ci_cutoff), metrics)
  }) %>%
  ungroup() %>%
  mutate(organism = "FIT") %>%
  rename(padj_cutoff = padj_fit) %>% 
  mutate(ci_percentile = paste0(ci_percentile * 100, "%"))
#Print
filter_conditions_results_fit_degs

#Mouse --------
filter_conditions_mouse <- expand_grid(
  padj_mouse = c(0.01, 0.05, seq(0.1, 1, by = 0.1)),
  expr_cutoff = c(0.01, seq(0.05, 0.5, by = 0.05), 0.75, 1),
  ci_percentile = c(0.3, 1.0),
  timepoint = c(1, 3, 7)
)

filter_conditions_results_mouse_degs <- filter_conditions_mouse %>%
  rowwise() %>%
  do({
    fc <- .
    
    # Filtrar apenas dados do timepoint atual
    data_timepoint <- pairwise_human_fit_test %>%
      filter(timepoint == fc$timepoint)

    # Calcular o cutoff do percentil escolhido
    ci_cutoff <- quantile(data_timepoint$ci_width_mouse, probs = fc$ci_percentile, na.rm = TRUE)

    # Aplicar filtros
    filtered_data <- data_timepoint %>%
      filter(
        pval_adj_mouse < fc$padj_mouse,
        abs(mean_mouse) > fc$expr_cutoff,
        ci_width_mouse <= ci_cutoff
      )

    if (nrow(filtered_data) >= 2) {
      metrics <- filtered_data %>%
        summarise(
          n_genes = n(),
          RMSE = rmse_vec(mean_human, mean_mouse),
          MAE = mae_vec(mean_human, mean_mouse),
          R2 = rsq_vec(mean_human, mean_mouse),
          Spearman = cor(mean_human, mean_mouse, method = "spearman"),
          Pearson = cor(mean_human, mean_mouse, method = "pearson")
        )
    } else {
      metrics <- tibble(
        n_genes = 0,
        RMSE = NA, MAE = NA, R2 = NA,
        Spearman = NA, Pearson = NA
      )
    }

    bind_cols(fc, tibble(ci_width_cutoff = ci_cutoff), metrics)
  }) %>%
  ungroup() %>%
  mutate(organism = "Mouse") %>%
  rename(padj_cutoff = padj_mouse) %>% 
  mutate(ci_percentile = paste0(ci_percentile * 100, "%"))
#Print
filter_conditions_results_mouse_degs  

#Merge
metrics_FIT_M_vs_FIT_filtered_DEGs = filter_conditions_results_mouse_degs %>% 
  bind_rows(filter_conditions_results_fit_degs) %>% 
  mutate(filter = if_else(organism == "FIT", paste0("FIT CI ", ci_percentile), NA),
         filter = if_else(organism == "Mouse", paste0("Mouse CI ", ci_percentile), filter)) %>% 
  select(organism,filter, everything()) %>% 
  filter(expr_cutoff <= 0.5)

#Save
metrics_FIT_M_vs_FIT_filtered_DEGs %>% 
  saveRDS(here("FIT_Mouse2man", "metrics_FIT_M_vs_FIT_filtered_DEGs.rds"))

#Print
metrics_FIT_M_vs_FIT_filtered_DEGs
```

#### Import metrics
```{r}
#### Import metrics ------
metrics_FIT_M_vs_FIT_filtered_DEGs =  readRDS(here("FIT_Mouse2man", "metrics_FIT_M_vs_FIT_filtered_DEGs.rds"))

metrics_FIT_M_vs_FIT_filtered_DEGs$filter %>% unique()

#Get best combinations
metrics_FIT_M_vs_FIT_filtered_DEGs %>% 
  group_by(organism) %>% 
  slice_min(RMSE, n = 1) %>% 
  slice_min(padj_cutoff, n = 1)

metrics_FIT_M_vs_FIT_filtered_DEGs %>% 
  group_by(organism) %>% 
  slice_max(R2, n = 5) %>% 
  slice_min(padj_cutoff, n = 5) 
```

#### Plot regression metrics


#### RMSE
```{r fig.height=12, fig.width=12}
metrics_FIT_M_vs_FIT_filtered_RMSE_plot = metrics_FIT_M_vs_FIT_filtered_DEGs %>%
  mutate(across(ci_width_cutoff:timepoint, as.character)) %>%
  mutate(filter = fct_relevel(filter, 
                              "Mouse CI 10%", 
                              "Mouse CI 30%", 
                              "Mouse CI 50%",
                              "Mouse CI 100%",
                              "FIT CI 10%",
                              "FIT CI 30%", 
                              "FIT CI 50%", 
                              "FIT CI 100%")) %>% 
  filter(filter %in% c("Mouse CI 30%",
                        "Mouse CI 100%",
                        "FIT CI 30%",
                        "FIT CI 100%")) %>% 
  filter(padj_cutoff <= 0.25) %>% 
  ggplot(aes(x = expr_cutoff, y = RMSE)) +
  geom_line(aes(group = filter,
                color = filter)) +
    geom_point(aes(fill = filter,
                 # size = n_genes
                 ),
             alpha = 0.8,
             pch = 21,
             # fill = NA,
             width = 0.1,
             # size = 0.5
             ) +
  ggh4x::facet_grid2(
    rows = vars(timepoint),
    cols = vars(padj_cutoff),
    switch = "both",
    labeller = labeller(
      timepoint = as_labeller(c("1" = "Day 1", 
                                "3" = "Day 3", 
                                "7" = "Day 7")),
      padj_cutoff = function(x) paste("padj:", x)  
    )
  ) +
  scale_x_continuous(n.breaks = 4) +
  scale_y_continuous(n.breaks = 4) +
  scale_size(range = c(0.1, 3), breaks = c(10, 100, 500, 1000)) +
  theme_vaxgo() +
  theme(
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    strip.background = element_blank(),
    legend.key = element_blank(),
    axis.text.x = element_text(size = 8, angle = 0),
    axis.text.y = element_text(size = 8),
    panel.spacing.x = unit(1, "lines")
  ) +
  labs(
    title = "RMSE",
    x = "L2FC Cutoff",
    y = "RMSE",
    color = "",
    size = "Number of genes"
  ) +
    scale_color_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#4361ee",
    "FIT CI 100%"   = "#dec0f1"
  )) +
    scale_fill_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#4361ee",
    "FIT CI 100%"   = "#dec0f1"
  )) 

metrics_FIT_M_vs_FIT_filtered_RMSE_plot
```


#### Spearman
```{r fig.height=12, fig.width=12}
metrics_FIT_M_vs_FIT_filtered_Spearman_plot = metrics_FIT_M_vs_FIT_filtered_DEGs %>%
  mutate(across(ci_width_cutoff:timepoint, as.character)) %>%
  mutate(filter = fct_relevel(filter, 
                              "Mouse CI 10%", 
                              "Mouse CI 30%", 
                              "Mouse CI 50%",
                              "Mouse CI 100%",
                              "FIT CI 10%",
                              "FIT CI 30%", 
                              "FIT CI 50%", 
                              "FIT CI 100%")) %>% 
  filter(filter %in% c("Mouse CI 30%",
                        "Mouse CI 100%",
                        "FIT CI 30%",
                        "FIT CI 100%")) %>% 
  filter(padj_cutoff <= 0.25) %>% 
  ggplot(aes(x = expr_cutoff, y = Spearman)) +
  geom_line(aes(group = filter,
                color = filter)) +
  geom_point(aes(fill = filter,
                 # size = n_genes
                 ),
             alpha = 0.8,
             pch = 21,
             # fill = NA,
             # size = 0.5
             ) +
  ggh4x::facet_grid2(
    rows = vars(timepoint),
    cols = vars(padj_cutoff),
    switch = "both",
    labeller = labeller(
      timepoint = as_labeller(c("1" = "Day 1", 
                                "3" = "Day 3", 
                                "7" = "Day 7")),
      padj_cutoff = function(x) paste("padj:", x)  
    )
  ) +
  scale_size(range = c(0.1, 3), breaks = c(10, 100, 500, 1000)) +
  scale_x_continuous(n.breaks = 4) +
  scale_y_continuous(n.breaks = 4) +
  theme_vaxgo() +
  theme(
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    strip.background = element_blank(),
    legend.key = element_blank(),
    axis.text.x = element_text(size = 8, angle = 0),
    axis.text.y = element_text(size = 8),
    panel.spacing.x = unit(1, "lines")
  ) +
  labs(
    title = "Spearman correlation",
    x = "L2FC Cutoff",
    y = "Rho",
    color = "",
    size = "Number of genes"
  ) +
    scale_color_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#4361ee",
    "FIT CI 100%"   = "#dec0f1"
  )) +
    scale_fill_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#4361ee",
    "FIT CI 100%"   = "#dec0f1"
  )) 
metrics_FIT_M_vs_FIT_filtered_Spearman_plot
```

#### Merge plots

```{r fig.height=4, fig.width=12}

rmse_plot = metrics_FIT_M_vs_FIT_filtered_RMSE_plot + guides(color = "none") + labs(fill = "Model")
spearman_plot = metrics_FIT_M_vs_FIT_filtered_Spearman_plot + guides(color = "none") + labs(fill = "Model")


metrics_FIT_tuning_params_plots_2 = rmse_plot + spearman_plot + plot_layout(axes = "collect", 
                                                                            guides = "collect")

metrics_FIT_tuning_params_plots_2
metrics_FIT_tuning_params_plots_2 %>%
  ggsave(filename = here("Figures", "metrics_FIT_tuning_params_plots_2.png"), height = 4, width = 12)
```





#### Plot correlation

Mouse vs Human (cutoffs for mouse)

```{r}
corr_plots_data = pairwise_human_fit_test %>%
  select(symbol, timepoint, mean_mouse, mean_human, mean_fit,
         pval_adj_human, pval_adj_mouse, pval_adj_fit) %>%
  pivot_longer(cols = c(pval_adj_human, pval_adj_mouse, pval_adj_fit),
               names_to = "measure",
               values_to = "value") %>%
  mutate(
    pval_cuts = case_when(
      value <= 0.01 ~ "0.01",
      value <= 0.05 ~ "0.05",
      TRUE ~ "1"
    )
  ) %>%
  pivot_wider(
    names_from = measure,
    values_from = pval_cuts,
    values_fn = list
  ) %>%
  unnest(cols = c(pval_adj_human, pval_adj_mouse, pval_adj_fit)) %>% 
  fill(pval_adj_human, .direction = "down") %>% 
  fill(pval_adj_mouse, .direction = "up") %>% 
  fill(pval_adj_fit, .direction = "up") %>% 
  mutate(pval_cuts_mouse_human = if_else(pval_adj_human == pval_adj_mouse, pval_adj_human, "Not equal"),
         pval_cuts_fit_human = if_else(pval_adj_human == pval_adj_fit, pval_adj_fit, "Not equal"),
         pval_cuts_mouse_fit = if_else(pval_adj_fit == pval_adj_mouse, pval_adj_fit, "Not equal")) 

corr_plots_data
```


Plot correlation with different padj cutoffs
```{r fig.height=12, fig.width=6}
human_vs_mouse_corr = corr_plots_data %>% 
  filter(pval_cuts_mouse_human != "Not equal") %>% 
  ggplot() +
  aes(x = mean_mouse,
      y = mean_human) +
  geom_point(alpha = 0.5,
             fill = "#4DBBD5FF",
             pch = 21,
             size = 1,
             show.legend = F,
             color = "#4DBBD5FF") +
  geom_smooth(method = 'lm', se = TRUE, color = "black") +
  geom_vline(xintercept = 0, linetype = 2, size = 0.5, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = 2, size = 0.5, alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, size = 0.5, alpha = 0.5) +
  facet_grid(~paste("Day", timepoint)~pval_cuts_mouse_human, scales = "free_x") +
  scale_x_continuous(n.breaks = 4, limits = c(-4,4)) +
  scale_y_continuous(n.breaks = 4, limits =  c(-2, 4)) +
  theme_vaxgo() +
  labs(x = "Mouse L2FC",
       y = "Human L2FC") +
    theme(
      legend.position = "top",
    strip.text = element_text(size = 10),
    strip.text.y = element_text(size = 10, angle = 0),
    strip.placement = "outside",
    strip.background = element_blank(),
    axis.text.x = element_text(size = 8, angle = 0),
    axis.text.y = element_text(size = 8),
    panel.spacing.x = unit(1, "lines"),
  )

fit_vs_human_corr = corr_plots_data %>% 
  filter(pval_cuts_fit_human != "Not equal") %>% 
  ggplot() +
  aes(x = mean_fit,
      y = mean_human) +
  geom_point(alpha = 0.5,
             fill = "#4361ee",
             pch = 21,
             size = 1,
             show.legend = F,
             color = "#4361ee") +
  geom_smooth(method = 'lm', se = TRUE, color = "black") +
  geom_vline(xintercept = 0, linetype = 2, size = 0.5, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = 2, size = 0.5, alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, size = 0.5, alpha = 0.5) +
  scale_x_continuous(n.breaks = 4, limits = c(-4,4)) +
  scale_y_continuous(n.breaks = 4, limits =  c(-2, 4)) +
  theme_vaxgo() +
  facet_grid(~paste("Day", timepoint)~pval_cuts_fit_human, scales = "free_x") +
  labs(x = "FIT L2FC",
       y = "Human L2FC") +
    theme(
      legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    strip.text.y = element_text(size = 10, angle = 0),
    strip.background = element_blank(),
    axis.text.x = element_text(size = 8, angle = 0),
    axis.text.y = element_text(size = 8),
    panel.spacing.x = unit(1, "lines"),
  )


fit_vs_mouse_corr = corr_plots_data %>% 
  filter(pval_cuts_mouse_fit != "Not equal") %>% 
  ggplot() +
  aes(x = mean_mouse,
      y = mean_fit) +
  geom_point(alpha = 0.5,
             fill = "#90A4AEFF",
             pch = 21,
             size = 1,
             show.legend = F,
             color = "#90A4AEFF") +
  geom_vline(xintercept = 0, linetype = 2, size = 0.5, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = 2, size = 0.5, alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, size = 0.5, alpha = 0.5) +
  geom_smooth(method = 'lm', se = TRUE, color = "black") +
  facet_grid(~paste("Day", timepoint)~pval_cuts_mouse_fit, scales = "free_x") +
  scale_x_continuous(n.breaks = 4, limits = c(-4,4)) +
  scale_y_continuous(n.breaks = 4, limits =  c(-2, 4)) +
  theme_vaxgo() +
  labs(y = "FIT L2FC",
       x = "Mouse L2FC") +
    theme(
      legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    strip.text.y = element_text(size = 10, angle = 0),
    strip.background = element_blank(),
    axis.text.x = element_text(size = 8, angle = 0),
    axis.text.y = element_text(size = 8),
    panel.spacing.x = unit(1, "lines"),
  )

pairwise_human_fit_test_corr_plots = human_vs_mouse_corr / 
  fit_vs_human_corr / 
  fit_vs_mouse_corr +
  plot_layout(guides =  "collect") +
  plot_annotation(theme = theme(legend.position = "top"))

pairwise_human_fit_test_corr_plots
pairwise_human_fit_test_corr_plots %>% 
  ggsave(filename = here("Figures", "pairwise_human_fit_test_corr_plots.png"), width = 6, height = 12)


```


```{r fig.height=4, fig.width = 12}
pairwise_human_fit_test_corr_plots_2 = human_vs_mouse_corr + 
  fit_vs_human_corr + 
  fit_vs_mouse_corr +
  # plot_layout(guides =  "collect", axes = "collect") +
  plot_annotation(theme = theme(legend.position = "top"))

pairwise_human_fit_test_corr_plots_2
pairwise_human_fit_test_corr_plots_2 %>% 
  ggsave(filename = here("Figures", "pairwise_human_fit_test_corr_plots_2.png"), width = 12, height = 5)
```



## Classification model


### Classify by grid of cutoffs
```{r}
# Define cutoffs
padj_cutoffs <- c(0.05)
l2fc_cutoffs <- c(0, 0.5)
ci_percentiles <- c(0.3, 1.0)

# Expand grid
cutoff_grid <- expand_grid(
  padj_cutoff = padj_cutoffs,
  l2fc_cutoff = l2fc_cutoffs,
  ci_percentile = ci_percentiles
)

# Classify by every combination
base_df <- pairwise_human_fit_genes_wide

classified_df <- pmap_dfr(
  cutoff_grid,
  function(padj_cutoff, l2fc_cutoff, ci_percentile) {
    
    ci_cutoff <- quantile(base_df$ci_width, probs = ci_percentile, na.rm = TRUE)
    
    base_df %>%
      filter(ci_width <= ci_cutoff) %>%
      mutate(
        padj_cutoff = padj_cutoff,
        l2fc_cutoff = l2fc_cutoff,
        ci_percentile = ci_percentile,
        classification = case_when(
          pval_adj < padj_cutoff & l2fc >= l2fc_cutoff  ~ "up",
          pval_adj < padj_cutoff & l2fc < -l2fc_cutoff ~ "down",
          TRUE                                          ~ "neutral"
    ))})

#Save
classified_df %>% 
  saveRDS(here("FIT_Mouse2man", "genes_l2fc_fit_mouse_human_classified_cutoffs.rds"))
```

###Import

```{r}
#Import
classified_df = read_rds(here("FIT_Mouse2man", "genes_l2fc_fit_mouse_human_classified_cutoffs.rds")) %>% 
  rename(padj_deg = pval_adj)

#Get only DEGs in humans
degs_human = classified_df %>% 
  filter(organism == "Human",
         padj_deg < 0.05) %>% 
  pull(symbol) %>% 
  unique()

length(degs_human)

#Summarize genes
classified_df_countgenes_degs_direction = classified_df %>% 
  filter(symbol %in% c(degs_human)) %>% 
  group_by(organism, timepoint, padj_cutoff, l2fc_cutoff, ci_percentile, classification) %>% 
  summarise(n_genes = n()) %>% 
  ungroup()  %>% 
  mutate(across(c(timepoint, padj_cutoff, l2fc_cutoff), as.character))

#Print
classified_df_countgenes_degs_direction

```


### Number of DEGs
Lineplot
```{r fig.width=5, fig.height=5}
#Number of DEGs: Lineplot
classification_degs_ngenes_plot = classified_df_countgenes_degs_direction %>%
  filter(classification != "neutral") %>% #Filter out neutral
  filter(padj_cutoff <= 0.1, 
         ci_percentile == 1.0, 
         l2fc_cutoff %in% c("0", "0.5")) %>% 
  mutate(row_labels = paste0("padj: ", padj_cutoff),
         timepoint = paste0("Day ", timepoint),
         l2fc_cutoff = as.numeric(l2fc_cutoff)) %>%   # cria label combinada
  ggplot() +
  aes(x = l2fc_cutoff,
      y = n_genes) +
    geom_point(aes(color = organism,
                 shape = classification,
                 # fill = organism
                 ),
             size = 2,
             fill = NA,
             alpha = 1
             ) +
  geom_line(aes(group = interaction(classification, organism),
                color = organism)) +
    scale_shape_manual(values=c("up" = 24, 
                                "down" = 25, 
                                "neutral" = 1))+
  facet_grid(~timepoint~row_labels)+ 
  scale_y_continuous(n.breaks = 4,
                     # limits = c(0, 3000),
                     # oob = scales::squish
                     ) +
  theme_vaxgo() +
  theme(
    strip.text = element_text(size = 10),
    strip.text.y = element_text(angle = 0),
    strip.placement = "outside",
    legend.position = "top",
    legend.direction = "horizontal",
    legend.key = element_blank(),
     # strip.background.x = element_rect(fill = "gray95", color = NA), 
    axis.text.x = element_text(angle = 0),
    panel.spacing.x = unit(0.4, "lines"),
  ) +
  labs(
    x = "L2FC Cutoff",
    y = "Total",
    color = "Organism",
    fill = "Organism",
    shape = "Classification"
  ) +
  scale_color_manual(values = c(
    "Human"  = "#90A4AEFF",
    "Mouse" = "#4DBBD5FF",
    "FIT"   = "#4361ee"
  )) +
  scale_fill_manual(values = c(
    "Human"  = "#90A4AEFF",
    "Mouse" = "#4DBBD5FF",
    "FIT"   = "#4361ee"
  )) 

#Print
classification_degs_ngenes_plot

```


```{r}
#Pivot wider
classified_wide <- classified_df %>%
  select(symbol, timepoint, padj_cutoff, l2fc_cutoff, organism, classification) %>%
  filter(l2fc_cutoff == 0) %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism", values_from = "classification") %>%
  filter(!is.na(Human) & !is.na(FIT) & !is.na(Mouse)) %>% 
  mutate(
    fit_correct = FIT == Human,
    mouse_correct = Mouse == Human
  )

classified_wide %>% glimpse()

classified_eval = classified_wide

my_metrics <- metric_set(accuracy, precision, recall, f_meas, spec, sens)

### FIT
# "up" vs all
acc_up_vs_all_fit <- classified_eval %>%
  mutate(
    truth = factor(ifelse(Human == "up", "up", "not_up"), levels = c("up", "not_up")),
    estimate = factor(ifelse(FIT == "up", "up", "not_up"), levels = c("up", "not_up"))
  ) %>%
  group_by(timepoint, padj_cutoff, l2fc_cutoff) %>%
  my_metrics(truth = truth, estimate = estimate) %>%
  mutate(model = "FIT", target = "up")
# "down" vs all
acc_down_vs_all_fit <- classified_eval %>%
  mutate(
    truth = factor(ifelse(Human == "down", "down", "not_down"), levels = c("down", "not_down")),
    estimate = factor(ifelse(FIT == "down", "down", "not_down"), levels = c("down", "not_down"))
  ) %>%
  group_by(timepoint, padj_cutoff, l2fc_cutoff) %>%
  my_metrics(truth = truth, estimate = estimate) %>%
  mutate(model = "FIT", target = "down")

acc_fit_results <- bind_rows(acc_up_vs_all_fit, acc_down_vs_all_fit)

fit_classification_degs_ngenes <- classified_eval %>%
  mutate(
    truth = factor(Human, levels = c("up", "down", "neutral")),
    estimate = factor(FIT, levels = c("up", "down", "neutral")),
    correct = truth == estimate,
    incorrect = truth != estimate
  ) %>% 
  group_by(timepoint, padj_cutoff, l2fc_cutoff, truth) %>%
  summarise(
    num_correct_tp = sum(correct, na.rm = T),
    num_incorrect_fp = sum(incorrect, na.rm = T),
    .groups = "drop"
  ) %>% 
  mutate(model = "FIT")


fit_classification_degs_ngenes = bind_rows(fit_classification_degs_ngenes)

fit_classification_degs = acc_fit_results %>% 
  rename(truth = target) %>% 
  inner_join(fit_classification_degs_ngenes,
             by = join_by("model", "timepoint", "padj_cutoff", "l2fc_cutoff", "truth"))

fit_classification_degs %>% 
  arrange(-.estimate) %>% 
  filter(num_correct_tp > 0)
```


```{r}
### MOUSE
# "up" vs all
acc_up_vs_all_mouse <- classified_eval %>%
  mutate(
    truth = factor(ifelse(Human == "up", "up", "not_up"), levels = c("up", "not_up")),
    estimate = factor(ifelse(Mouse == "up", "up", "not_up"), levels = c("up", "not_up"))
  ) %>%
  group_by(timepoint, padj_cutoff, l2fc_cutoff) %>%
  my_metrics(truth = truth, estimate = estimate) %>%
  mutate(model = "Mouse", target = "up")

# "down" vs all
acc_down_vs_all_mouse <- classified_eval %>%
  mutate(
    truth = factor(ifelse(Human == "down", "down", "not_down"), levels = c("down", "not_down")),
    estimate = factor(ifelse(Mouse == "down", "down", "not_down"), levels = c("down", "not_down"))
  ) %>%
  group_by(timepoint, padj_cutoff, l2fc_cutoff) %>%
  my_metrics(truth = truth, estimate = estimate) %>%
  mutate(model = "Mouse", target = "down")

acc_mouse_results <- bind_rows(acc_down_vs_all_mouse, acc_up_vs_all_mouse)

mouse_classification_degs_ngenes <- classified_eval %>%
  mutate(
    truth = factor(Human, levels = c("up", "down", "neutral")),
    estimate = factor(FIT, levels = c("up", "down", "neutral")),
    correct = truth == estimate,
    incorrect = truth != estimate
  ) %>% 
  group_by(timepoint, padj_cutoff, l2fc_cutoff, truth) %>%
  summarise(
    num_correct_tp = sum(correct, na.rm = T),
    num_incorrect_fp = sum(incorrect, na.rm = T),
    .groups = "drop"
  ) %>% 
  mutate(model = "Mouse")

#Bind tables
mouse_classification_degs_ngenes = bind_rows(mouse_classification_degs_ngenes)


mouse_classification_degs = acc_mouse_results %>% 
  rename(truth = target) %>% 
  inner_join(mouse_classification_degs_ngenes,
             by = join_by("model", "timepoint", "padj_cutoff", "l2fc_cutoff", "truth"))

classification_degs_comparison = fit_classification_degs %>% 
  bind_rows(mouse_classification_degs)

classification_degs_comparison %>% 
  write_csv(file = here("FIT_Mouse2man", "classification_degs_comparison.csv"))

classification_degs_comparison

```

Table
```{r}

library(gtable)
library(gtExtras)


summary_table_accuracy = 
  classification_degs_comparison %>% 
  filter(
    .metric == "accuracy")  %>% 
  arrange(timepoint, -.estimate, -num_correct_tp) %>% 
  select(timepoint:l2fc_cutoff, .estimate,truth, model:num_incorrect_fp) %>% 
  gt() %>% 
  gt_theme_pff() %>% 
  tab_style(
    style = list(
      cell_fill(color = "lightblue"),
      cell_text(weight = "bold", color = "black")
    ),
    locations = cells_row_groups(groups = everything())
  ) %>% 
  tab_header(title = md("**Accuracy**"))

# summary_table_accuracy %>% 
#   gtsave(filename = here("Tables" , "summary_table_accuracy.docx"))
  
summary_table_accuracy
  

```

### Classification metrics 
Import
```{r fig.height=8, fig.width=15, message=FALSE, warning=FALSE}
classification_degs_comparison = read_csv(file = here("FIT_Mouse2man", "classification_degs_comparison.csv"))
```

#### Accuracy
```{r fig.height=8, fig.width=15, message=FALSE, warning=FALSE}
classification_degs_accuracy_plot = classification_degs_comparison %>% 
  mutate(l2fc_cutoff = as.character(l2fc_cutoff),
         timepoint = paste0("Day ", timepoint),
         padj_cutoff = paste0("padj: ", padj_cutoff)) %>% 
  filter(.metric %in% c("accuracy")) %>%
  ggplot() +
  aes(x = l2fc_cutoff,
      y = .estimate,
      color = model) +
  geom_point(aes(shape = truth,
                  # fill = model,
                   size = num_correct_tp),
             alpha = 1,
             fill = NA
             ) +
    scale_size(range = c(0.1, 4),
               # breaks = c(10, 100, 500, 1000)
               ) +
      scale_shape_manual(values=c("up" = 24, 
                                "down" = 25, 
                                "neutral" = 1))+
  scale_shape(solid = FALSE) +
  geom_line(aes(group = interaction(model, truth))) +
  facet_grid(~timepoint~padj_cutoff, scales = "free_y")+
  scale_y_continuous(n.breaks = 6, expand = c(0.05, 0.05)) +
    scale_color_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#45efc6",
    "FIT CI 100%"   = "#4361ee"
  )) +
    scale_fill_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#45efc6",
    "FIT CI 100%"   = "#4361ee"
  )) +
  theme_vaxgo() +
  theme(
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0),
    axis.text.x = element_text(size = 8, angle = 0),
    axis.text.y = element_text(size = 8),
    panel.spacing.x = unit(0.5, "lines"),
  ) +
  labs(
    title = "Accuracy",
    x = "L2FC Cutoff",
    y = "Accuracy",
    color = "Model",
    fill = "Model",
    size = "Correct"
  )

classification_degs_accuracy_plot
```


#### Recall
```{r fig.height=8, fig.width=15, message=FALSE, warning=FALSE}
classification_degs_recall_plot = classification_degs_comparison %>% 
  mutate(l2fc_cutoff = as.character(l2fc_cutoff),
         timepoint = paste0("Day ", timepoint),
         padj_cutoff = paste0("padj: ", padj_cutoff)) %>% 
  filter(.metric %in% c("recall")) %>%
  ggplot() +
  aes(x = l2fc_cutoff,
      y = .estimate,
      color = model,
      fill = model) +
  geom_line(aes(group = interaction(model, truth))) +
   geom_point(aes(shape = truth,
                  # fill = model,
                   size = num_correct_tp),
             alpha = 1,
             fill = NA
             ) +
    scale_size(range = c(0.1, 4)) +
      scale_shape_manual(values=c("up" = 24, 
                                "down" = 25, 
                                "neutral" = 1))+
  scale_shape(solid = FALSE) +
  facet_grid(~timepoint~padj_cutoff, scales = "free_y")+
  scale_y_continuous(n.breaks = 4, expand = c(0.05, 0.05)) +
    scale_color_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#4361ee",
    "FIT CI 100%"   = "#dec0f1"
  )) +
    scale_fill_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#4361ee",
    "FIT CI 100%"   = "#dec0f1"
  )) +
  theme_vaxgo() +
  theme(
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0),
    axis.text.x = element_text(size = 8, angle = 0),
    axis.text.y = element_text(size = 8),
    panel.spacing.x = unit(0.5, "lines"),
  ) +
  labs(
    title = "Recall",
    x = "L2FC Cutoff",
    y = "Recall",
    color = "Model",
    fill = "Model",
    size = "Correct"
  )

classification_degs_recall_plot
```


#### Precision
```{r fig.height=8, fig.width=15, message=FALSE, warning=FALSE}
classification_degs_precision_plot = classification_degs_comparison %>% 
  mutate(l2fc_cutoff = as.character(l2fc_cutoff),
         timepoint = paste0("Day ", timepoint),
         padj_cutoff = paste0("padj: ", padj_cutoff)) %>% 
  filter(.metric %in% c("precision")) %>%
  ggplot() +
  aes(x = l2fc_cutoff,
      y = .estimate,
      color = model) +
 geom_line(aes(group = interaction(model, truth))) +
   geom_point(aes(shape = truth,
                  fill = model,
                   size = num_correct_tp),
              fill = NA,
             alpha = 1, 
             ) +
    scale_size(range = c(0.1, 4),
               # breaks = c(10, 100, 500, 1000)
               ) +
      scale_shape_manual(values=c("up" = 24, 
                                "down" = 25, 
                                "neutral" = 1))+
  # scale_shape(solid = FALSE) +
  facet_grid(~timepoint~padj_cutoff, scales = "free_y")+
  scale_y_continuous(n.breaks = 4, expand = c(0.05, 0.05)) +
    scale_color_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#4361ee",
    "FIT CI 100%"   = "#dec0f1"
  )) +
    scale_fill_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#4361ee",
    "FIT CI 100%"   = "#dec0f1"
  )) +
  theme_vaxgo() +
  theme(
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0),
    axis.text.x = element_text(size = 8, angle = 0),
    axis.text.y = element_text(size = 8),
    panel.spacing.x = unit(0.5, "lines"),
  ) +
  labs(
    title = "Precision",
    x = "L2FC Cutoff",
    y = "Precision",
    color = "Model",
    fill = "Model",
    size = "Correct"
  )

classification_degs_ngenes_plot +
classification_degs_accuracy_plot +
classification_degs_recall_plot+
  classification_degs_precision_plot +
  plot_layout(axes = "collect", ncol = 2,
              guides = "collect", 
              axis_titles = "collect")

classification_degs_precision_plot
```


#### Precision recall
```{r fig.height=4, fig.width=15}
precision_recall_plot = classification_degs_comparison %>%
  filter(.metric %in% c("recall", "precision")) %>%
  mutate(timepoint = paste0("Day ", timepoint)) %>% 
  pivot_wider(names_from = .metric,
              values_from = .estimate) %>% 
  ggplot(aes(x = recall, y = precision, shape = truth, color = model, size = num_correct_tp)) +
  geom_point(aes(shape = truth)) +
  scale_size(range = c(0.1, 4),
               # breaks = c(10, 100, 500, 1000)
               ) +
  scale_shape_manual(values=c("up" = 24, 
                              "down" = 25))+
  scale_y_continuous(n.breaks = 4, expand = c(0.05, 0.05)) +
    scale_color_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#4361ee",
    "FIT CI 100%"   = "#dec0f1"
  )) +
    scale_fill_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#4361ee",
    "FIT CI 100%"   = "#dec0f1"
  )) +
  facet_wrap(~timepoint) +
  theme_vaxgo() +
  theme(
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0),
    axis.text.x = element_text(size = 8, angle = 0),
    axis.text.y = element_text(size = 8),
    panel.spacing.x = unit(0.5, "lines"),
  ) +
  labs(
    title = "Precision-Recall",
    x = "Recall",
    y = "Precision",
    shape = "Direction",
    color = "Model",
    fill = "Model",
    size = "Correct"
  )

precision_recall_plot
```





#### Classification of DEGs, distributions
```{r}
# Suponha um exemplo
log2fc <- 1.2
ci_low <- 0.8
ci_high <- 1.6
se <- (ci_high - ci_low) / (2 * 1.96)

# Geração de x para a curva
x_vals <- seq(log2fc - 4*se, log2fc + 4*se, length.out = 500)
y_vals <- dnorm(x_vals, mean = log2fc, sd = se)

# Criar o data.frame
df_curve <- data.frame(x = x_vals, y = y_vals)

# Plot
ggplot(df_curve, aes(x = x, y = y)) +
  geom_line(color = "blue") +
  geom_vline(xintercept = c(ci_low, ci_high), linetype = "dashed", color = "red") +
  geom_vline(xintercept = log2fc, linetype = "solid", color = "black") +
  theme_vaxgo() +
  labs(title = "Normal distribution from log2FC ± CI",
       x = "log2 Fold Change",
       y = "Density")
```


```{r}
#Manual example
dge_long_genes_up_example = dge_long_genes_down%>% 
  filter(organism == "Human",
         symbol == "TLR4") %>% 
  slice(3)

# Parameters
condition = dge_long_genes_up_example$condition
log2fc <- dge_long_genes_up_example$log2fc
# log2fc <- -0.15
ci_low <- dge_long_genes_up_example$ci_low
ci_high <- dge_long_genes_up_example$ci_high
z = dge_long_genes_up_example$z
pval_one_sided = pnorm(z, lower.tail = T) #one sided, higher
se <- dge_long_genes_up_example$se
direction = dge_long_genes_up_example$direction
df <- 120  #samples

# Null parameters
se_null <- se # Same SE
ci_low_null <- 0 - qt(0.975, df) * se_null
ci_high_null <- 0 + qt(0.975, df) * se_null

# x values (scaled for t-distribution)
x_vals <- seq(log2fc - 4*se, log2fc + 4*se, length.out = 500)
x_vals_null <- seq(-4*se, 4*se, length.out = 500)

# t-distribution densities
y_vals <- dt((x_vals - log2fc)/se, df = df) / se
y_vals_null <- dt(x_vals_null / se_null, df = df) / se_null

# Data frames
df_curve_gene <- data.frame(x = x_vals, y = y_vals, distribution = "Observed")
df_curve_null <- data.frame(x = x_vals_null, y = y_vals_null, distribution = "Null")
actual_null_dist <- bind_rows(df_curve_gene, df_curve_null)

# Plot

# Destaque da cauda inferior da distribuição nula
shade_df_superior <- df_curve_null %>%
  filter(x >= log2fc) 

shade_df_inferior <- df_curve_gene %>%
  filter(x <= log2fc) 

ggplot(actual_null_dist, aes(x = x, y = y, color = distribution)) +
  geom_line(size = 1.2) +
  
  # Área sombreada da cauda superior
  geom_area(data = shade_df_superior, aes(x = x, y = y), fill = "black", alpha = 0.3, inherit.aes = FALSE) +
  
  # Área sombreada da cauda inferior
  geom_area(data = shade_df_inferior, aes(x = x, y = y), fill = "lightblue", alpha = 0.3, inherit.aes = FALSE) +
  
  # Linhas de referência
  geom_vline(xintercept = c(ci_high_null), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(ci_low), linetype = "dashed", color = "lightblue") +
  geom_vline(xintercept = log2fc, linetype = "solid", color = "lightblue") +
  geom_vline(xintercept = 0, linetype = "solid", color = "black") +
  
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  labs(title = paste0(condition, ", ", "adj_pval = ", as.character(pval_one_sided), ", ", direction),
       subtitle = condition,
       x = "log2 Fold Change",
       y = "Density") +
  scale_color_manual(values = c("Observed" = "lightblue", "Null" = "black"))

```


Automated
Generate distributions for multiple genes
```{r}

#Plot T distributions
plot_t_distribution <- function(df, df_samples = 10, n_points = 500) {
  # df: data frame containing condition, log2fc, ci_low, ci_high, se columns
  # df_samples: degrees of freedom for t-distribution (default 120)
  # n_points: number of points for the curves
  
  # Function to create a plot for one row (one condition)
  create_plot <- function(row) {
    condition <- row$condition
    log2fc <- row$log2fc
    ci_low <- row$ci_low
    ci_high <- row$ci_high
    se <- row$se
    adj_p_val = row$adj_p_val
    direction = row$direction
    df <- df_samples
    
    
    # Null parameters (mean=0, same SE)
    se_null <- se
    ci_low_null <- 0 - qt(0.975, df) * se_null
    ci_high_null <- 0 + qt(0.975, df) * se_null
    
    # x ranges for observed and null
    x_vals <- seq(log2fc - 4*se, log2fc + 4*se, length.out = n_points)
    x_vals_null <- seq(-4*se_null, 4*se_null, length.out = n_points)
    
    # t-distribution densities
    y_vals <- dt((x_vals - log2fc) / se, df = df) / se
    y_vals_null <- dt(x_vals_null / se_null, df = df) / se_null
    
    # Data frames for plotting
    df_curve_gene <- data.frame(x = x_vals, y = y_vals, distribution = "Observed")
    df_curve_null <- data.frame(x = x_vals_null, y = y_vals_null, distribution = "Null")
    actual_null_dist <- bind_rows(df_curve_gene, df_curve_null)
    
    # Plot
    p <- ggplot(actual_null_dist, aes(x = x, y = y, color = distribution)) +
      geom_line(size = 1.2) +
      geom_vline(xintercept = c(ci_high_null), linetype = "dashed", color = "black") +
      geom_vline(xintercept = c(ci_low), linetype = "dashed", color = "lightblue") +
      geom_vline(xintercept = log2fc, linetype = "solid", color = "lightblue") +
      geom_vline(xintercept = 0, linetype = "solid", color = "black") +
      theme_vaxgo() +
      theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
            legend.position = "top") +
      labs(title = paste0(condition, ", ", "adj_pval = ", round(adj_p_val, 4), ", ", direction),
           x = "log2 Fold Change",
           y = "Density") +
      scale_color_manual(values = c("Observed" = "lightblue", "Null" = "black"))
    
    return(p)
  }
  
  # Apply the plotting function to each row
  plots <- lapply(seq_len(nrow(df)), function(i) create_plot(df[i, ]))
  
  # Optionally, name the list by conditions
  names(plots) <- df$condition
  
  return(plots)
}
```

Run
```{r fig.height=8, fig.width=12}

df_example <- dge_long_genes_up %>%
  filter(symbol == "TLR8") %>%
  distinct(condition, .keep_all = TRUE)  # One row per condition

plots_list <- plot_t_distribution(df_example, df_samples = 100)

wrap_plots(plots_list, guides = "collect") +
  plot_annotation(title = df_example$symbol, theme = theme(legend.position = "top"))
```









#### Confusion matrix
```{r}
library(tidyr)
library(purrr) # Make sure this line is run before using reduce

##### Function
#Get summaries
get_confmat_summary <- function(data, model_selected, timepoint_selected) {
  data %>%
    filter(model == model_selected, timepoint == timepoint_selected) %>%
    conf_mat(truth, predicted) %>% # Assumes 'truth' and 'predicted' columns exist
    summary() %>% 
    mutate(model = model_selected,
           timepoint = timepoint_selected)
}


# Plot heatmap for every combination
plot_xconfmat_heatmap <- function(data, 
                                 model_selected, 
                                 timepoint_selected) {
  
  data %>%
    filter(model == model_selected, timepoint == timepoint_selected) %>%
    conf_mat(truth, predicted) %>%
    autoplot(type = "heatmap") +
    aes(
      fill = Freq,
      x = fct_rev(str_to_upper(Prediction)),
      y = str_to_upper(Truth),
      label = Freq
    ) +
    geom_tile(color = "black") +
    geom_text() +
    scale_x_discrete(position = "top", expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) +
    labs(
      x = "Predicted",
      y = "Truth",
      title = paste(model_selected, "Day", timepoint_selected)
    ) +
    theme_vaxgo() +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 12, hjust = 0.5),
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      axis.line.x = element_blank(),
      axis.line.y = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1)
    ) +
    scale_fill_gradient2(
      low = "white",
      high = "#80a1d4",
      n.breaks = 3
    )
}
```


```{r fig.height=5, fig.width=10}
filtered_df_confmat = classified_df_probs 


plot_params <- expand_grid(model = unique(filtered_df_confmat$model),
                           timepoint = unique(filtered_df_confmat$timepoint))


#Get performance
confmat_summaries <- map2_dfr(plot_params$model, plot_params$timepoint,
                               .f = ~ get_confmat_summary(data = filtered_df_confmat, # Pass your full data here
                                                           model_selected = .x,
                                                           timepoint_selected = .y))

confmat_summaries_plot = confmat_summaries %>% 
  filter(.metric %in% c("accuracy", "sens", wa)) %>%
  ggplot() +
  aes(y = .estimate, x = model, fill = model, label = round(.estimate, 2)) +
  geom_col() +
  geom_text(hjust = -0.2, vjust = 0.5, angle = 90, size = 3) +
  scale_y_continuous(limits = c(0, 1.10)) +
  facet_grid(~paste("Day", timepoint) ~.metric) +
  theme_vaxgo() +
  theme(strip.text.y = element_text(angle = 0),
        axis.text.x = element_blank()) +
  scale_fill_manual(values = c(
    "Mouse CI 30%" = "#4DBBD5FF",
    "Mouse CI 100%" = "#80a1d4",
    "FIT CI 30%"   = "#4361ee",
    "FIT CI 100%"   = "#dec0f1"
  ))

confmat_summaries_plot

confmat_summaries_plot %>% 
  ggsave(filename = here("Figures", "Conf_Matrix_FIT_Mouse_Summaries.png"),
       width = 10,
       height = 5)
```


```{r fig.height=10, fig.width=10}
#Plot heatmaps
conf_matrix_heatmap_list <- list()
heatmap_list <- map2(plot_params$model, plot_params$timepoint,
                     .f = ~ plot_confmat_heatmap(data = filtered_df_confmat, # Use filtered_df_confmat here
                                                 model_selected = .x,
                                                 timepoint_selected = .y))


combined_heatmaps <- purrr::reduce(heatmap_list, `+`) + plot_layout( ncol = 3)

combined_heatmaps

combined_heatmaps %>% 
  ggsave(filename = here("Figures", "Conf_Matrix_FIT_Mouse.png"),
       width = 10,
       height = 10)

```



































































































































































































































































