---
title: "4_Performance_Biocontrol"
output: html_document
---
#Required

```{r message=FALSE, warning=FALSE}
# install.packages("here") if needed
library(here)
source(here("scripts_notebooks", "required.R"))
```

# Performance tests 



# Functions

```{r}


```


## Classification: Genes

```{r}
#=======================
# Import datasets
#=======================
# Import BTM annotation genes
btm_annotation_genes <- read_csv(file = here("tables", "btm_annotation_genes.csv"))

# Import and clean DEGs table with control
all_degs_human_mouse_matched_bionull = 
  readRDS(file = here("tables", "all_human_mouse_dge_limma_degs_matched_control.rds")) %>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint %in% c("2h", "Early hours"), "Early hours", timepoint)) 

# Check original counts
all_degs_human_mouse_matched_bionull %>% dplyr::count(pathogen, organism, timepoint)


# ==============================================================================
# FILTER COMPARISONS BY CORRELATION
# ==============================================================================

conditions_filtered <- readRDS(file = here("tables", "all_human_mouse_correlation_filtered_inter_weighted_filtered.rds")) %>% 
  mutate(pathogen_human = case_when(pathogen_human == "Hepatitis B" ~ "HepB",
                              pathogen_human == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen_human == "Escherichia coli" ~ "E. coli",
                              T ~pathogen_human),
          pathogen_human = fct_relevel(pathogen_human, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  select(treatment_human, treatment_mouse, pathogen_human, pathogen_mouse,
         pathogen_comparison, timepoint_comparison,
         timepoint_human, timepoint_mouse) %>% 
  distinct() %>% 
  pivot_longer(
    cols = contains("_human") | contains("_mouse"), 
    names_to = c(".value", "organism"),             
    names_sep = "_",                                
    values_drop_na = TRUE
  ) %>% 
  mutate(
    organism = str_to_title(organism),
    timepoint = str_replace_all(timepoint, "\n", " ")
  )

# Filter paired data
all_degs_human_mouse_matched_filtered <- all_degs_human_mouse_matched_bionull %>% 
  select(organism:human_symbol, mean_l2fc:sd, timepoint_comparison, n) %>% 
  inner_join(conditions_filtered %>% 
               select(-timepoint, -organism), by = c("pathogen", "treatment", "timepoint_comparison")) %>%
  
  #Define control
  mutate(organism = if_else(organism == "DMD", "Control", organism))

# Check filtered counts
all_degs_human_mouse_matched_filtered %>% 
  dplyr::count(treatment, pathogen, organism, timepoint_comparison)


##If the control is only permuted values
# permutation_prediction_up = perfect_prediction_up %>%
#   group_by(pathogen, timepoint_comparison) %>% 
#   mutate(prob = runif(n()), organism = "Permutation",
#          model = "Permutation")

# permutation_prediction_down = perfect_prediction_down %>%
#   group_by(pathogen, timepoint_comparison) %>% 
#   mutate(prob = runif(n()), 
#          organism = "Permutation",
#          model = "Permutation")

#############################
# Input
#############################

#All genes
genes_roc_allgenes = all_degs_human_mouse_matched_filtered %>%
  select(pathogen, treatment, organism, human_symbol, timepoint_comparison, mean_l2fc, adj_p_val, everything())

genes_roc_allgenes %>% dplyr::count(pathogen, organism,  timepoint_comparison)
genes_roc_allgenes

#Filter immune genes
genes_roc_allgenes_immunegenes = genes_roc_allgenes %>%
  filter(human_symbol %in% c(btm_annotation_genes %>%
                         filter(module_category == "immune") %>%
                         pull(symbol) %>%
                         unique()))

#Check
genes_roc_allgenes %>%
  dplyr::count(pathogen, organism, timepoint_comparison)
```


```{r}
set.seed(123)
######## Input =======
filename = "all_human_mouse_alternative_unicaudal"
genes_df_input = bind_rows(
  genes_roc_allgenes,
  genes_roc_allgenes_immunegenes %>% mutate(organism = "Immune")
)

######## Run analysis ======
classification_performance_results = classification_performance_compute_genes_degs(genes_df_input)

#Extract results
roc_auc_df = classification_performance_results$roc_auc_df
roc_auc_df_DEGs = classification_performance_results$roc_auc_df_DEGs
roc_curve_auc_df = classification_performance_results$roc_curve_auc_df
roc_curve_auc_df_DEGs = classification_performance_results$roc_curve_auc_df_DEGs
pr_auc_df = classification_performance_results$pr_auc_df
pr_auc_DEGs_df = classification_performance_results$pr_auc_DEGs_df
pr_curve_auc_df = classification_performance_results$pr_curve_auc_df
pr_curve_auc_df_DEGs = classification_performance_results$pr_curve_auc_df_DEGs

#Refactor
roc_curve_auc_df = roc_curve_auc_df %>% 
    mutate(model = fct_relevel(model, "Human", "Mouse", "Control", "Immune", "Permutation"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(auc = round(auc, 2) %>% as.character())

roc_curve_auc_df_DEGs = roc_curve_auc_df_DEGs %>% 
    mutate(model = fct_relevel(model, "Human", "Mouse", "Control", "Immune", "Permutation"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(auc = round(auc, 2) %>% as.character())

roc_auc_df = roc_auc_df%>% 
  mutate(model = fct_relevel(model, "Human", "Mouse", "Immune", "Permutation", "Control", "FIT"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 

roc_auc_df_DEGs = roc_auc_df_DEGs %>% 
  mutate(model = fct_relevel(model, "Human", "Mouse", "Immune", "Permutation", "Control", "FIT"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 


pr_curve_auc_df = pr_curve_auc_df %>% 
    mutate(model = fct_relevel(model, "Mouse", "Immune", "Permutation", "Human", "FIT", "Control"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(pr_auc = round(pr_auc, 2) %>% as.character())

pr_curve_auc_df_DEGs = pr_curve_auc_df_DEGs  %>% 
    mutate(model = fct_relevel(model, "Mouse", "Immune", "Permutation", "Human", "FIT", "Control"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(pr_auc = round(pr_auc, 2) %>% as.character())

pr_auc_df = pr_auc_df %>% 
  mutate(model = fct_relevel(model, "Human", "Mouse", "Immune", "Permutation", "Control", "FIT"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 

pr_auc_DEGs_df = pr_auc_DEGs_df %>% 
  mutate(model = fct_relevel(model, "Human", "Mouse", "Immune", "Permutation", "Control", "FIT"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 

```



### Plot ROC curve

#### All genes
```{r fig.height=6, fig.width=8}
#Plot curves ------------------
#PLOT ROC Curve
roc_curve_plot = roc_curve_auc_df %>% 
  arrange(pathogen) %>% 
  ungroup() %>% 
  ggplot(aes(x = 1 - specificity, 
             y = sensitivity, 
             color = model,
             linetype = comparison)) +
  geom_path(size = 0.75, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_nested_wrap(~pathogen+timepoint_comparison, nrow = 2, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Mouse
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint_comparison, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.4),
            size = 3,
            hjust = 1) +
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint_comparison, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.3),
            size = 3,
            hjust = 1) +
  #Immune
  geom_text(data = . %>% 
              filter(model == "Immune", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint_comparison, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.2),
            size = 3,
            hjust = 1) +
   geom_text(data = . %>% 
              filter(model == "Immune", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint_comparison, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.1),
            size = 3,
            hjust = 1) +
  scale_linetype_manual(values = c("Up" = 1,
                                   "Down" = 2)) +
  # coord_equal() +
  scale_color_manual(values = colors$organism) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model",
  title = "All genes",
  linetype = "Comparison") + 
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
          axis.title.x = element_text(size = 10),
          axis.text.y = element_text(size = 10),
          axis.title.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 10),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          legend.spacing.x = unit(0, 'cm'),
          legend.spacing.y = unit(0, 'cm'),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing.y = unit(0, "lines"), 
          panel.spacing.x = unit(1, "lines"), 
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          
          ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 0.5, size = 1),
                                 order = 2), 
         fill = "none",
         color = guide_legend(override.aes = list(linewidth = 3)))

roc_curve_plot

roc_curve_auc_df %>% 
  dplyr::count(pathogen, timepoint_comparison, comparison, model)

#Save
roc_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_All_OnlyImmune.png")),
         width = 8,
         height = 6)
```


#### DEGs
```{r fig.height=6, fig.width=8}
#Plot curves ------------------
#PLOT ROC Curve
roc_curve_degs_plot = roc_curve_auc_df_DEGs %>% 
  arrange(pathogen) %>% 
  ungroup() %>% 
  ggplot(aes(x = 1 - specificity, 
             y = sensitivity, 
             color = model,
             linetype = comparison)) +
  geom_path(size = 0.75, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_nested_wrap(~pathogen+timepoint_comparison, nrow = 2, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Mouse
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint_comparison, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.4),
            size = 3,
            hjust = 1) +
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint_comparison, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.3),
            size = 3,
            hjust = 1) +
  #Immune
  geom_text(data = . %>% 
              filter(model == "Immune", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint_comparison, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.2),
            size = 3,
            hjust = 1) +
   geom_text(data = . %>% 
              filter(model == "Immune", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint_comparison, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.1),
            size = 3,
            hjust = 1) +
  scale_linetype_manual(values = c("Up" = 1,
                                   "Down" = 2)) +
  # coord_equal() +
  scale_color_manual(values = colors$organism) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model",
  title = "DEGs",
  linetype = "Comparison") + 
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
          axis.title.x = element_text(size = 10),
          axis.text.y = element_text(size = 10),
          axis.title.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 10),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          legend.spacing.x = unit(0, 'cm'),
          legend.spacing.y = unit(0, 'cm'),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing.y = unit(0, "lines"), 
          panel.spacing.x = unit(1, "lines"), 
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          
          ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 0.5, size = 1),
                                 order = 2), 
         fill = "none",
         color = guide_legend(override.aes = list(linewidth = 3)))

roc_curve_degs_plot

roc_curve_degs_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_DEGs_OnlyImmune.png")),
         width = 8,
         height = 6)
```



### Plot AUC 
#### All Genes: Heatmap
```{r fig.height=3, fig.width=8}
roc_auc_heatmap = roc_auc_df %>% 
  ggplot() +
   aes(x = comparison,
      y = model %>% fct_rev(),
      fill = auc,
      color = model %>% fct_rev(),
      group = model %>% fct_rev()) + 
  
  facet_nested_wrap(~treatment+pathogen+timepoint_comparison,
                    # scales = "free",
                    nrow = 1,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_tile(color = "black",
             stroke = 0.5) + 
  
  #Add labels
  geom_text(aes(label = round(auc, 2)), size = 3,
             color = "black",
             fontface = "bold",
             vjust = 0.25,
            data = . %>% filter(auc <= 0.7)) +
  geom_text(aes(label = round(auc, 2)), size = 3,
             color = "white",
             fontface = "bold",
             vjust = 0.25,
            data = . %>% filter(auc > 0.7)) +
   geom_text(aes(label = paste0("n=", n_genes)), size = 2.5,
             color = "black",
             vjust = 2,
             data = . %>% filter(auc <= 0.7)) +
  
   geom_text(aes(label = paste0("n=", n_genes)), size = 2.5,
             color = "white",
             vjust = 2,
             data = . %>% filter(auc > 0.7)) +
 

  scale_x_discrete(expand = c(0.01, 0.01)) +
  scale_y_discrete(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
    scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(0, 1, length.out = 10)),
        limits = c(0, 1), 
        name = "ROC AUC",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 1,
          nbin = 10,
          raster = FALSE
        )
      ) +
  labs(
    x = "Comparison",
    y = "Model",
    fill = "ROC-AUC",
    title = "All genes"
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(1, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) 

roc_auc_heatmap
  

roc_auc_heatmap %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_auc_heatmap_All_OnlyImmune.png")),
         width = 8,
         height = 3)
```
#### DEGs: Heatmap
```{r fig.height=3, fig.width=8}
#### DEGs: Heatmap
roc_auc_DEGs_heatmap = roc_auc_df_DEGs %>% 
  ggplot() +
   aes(x = comparison,
      y = model %>% fct_rev(),
      fill = auc,
      color = model %>% fct_rev(),
      group = model %>% fct_rev()) + 
  
  facet_nested_wrap(~treatment+pathogen+timepoint_comparison,
                    # scales = "free",
                    nrow = 1,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_tile(color = "black",
             stroke = 0.5) + 
  
  #Add labels
  geom_text(aes(label = round(auc, 2)), size = 3,
             color = "black",
             fontface = "bold",
             vjust = 0.25,
            data = . %>% filter(auc <= 0.9)) +
  geom_text(aes(label = round(auc, 2)), size = 3,
             color = "white",
             fontface = "bold",
             vjust = 0.25,
            data = . %>% filter(auc > 0.9)) +
   geom_text(aes(label = paste0("n=", n_genes)), size = 2.5,
             color = "black",
             vjust = 2,
             data = . %>% filter(auc <= 0.9)) +
  
   geom_text(aes(label = paste0("n=", n_genes)), size = 2.5,
             color = "white",
             vjust = 2,
             data = . %>% filter(auc > 0.9)) +
 

  scale_x_discrete(expand = c(0.01, 0.01)) +
  scale_y_discrete(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
    scale_fill_gradientn(
        colors = c("#4DBBD5FF", "white","#4361ee"),
        values = scales::rescale(seq(0, 1, length.out = 10)),
        limits = c(0, 1), 
        name = "ROC AUC",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 1,
          nbin = 10,
          raster = FALSE
        )
      ) +
  labs(
    x = "Comparison",
    y = "Model",
    fill = "ROC-AUC",
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(1, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) 

roc_auc_DEGs_heatmap
  
#Save
roc_auc_DEGs_heatmap %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_auc_heatmap_DEGs_OnlyImmune.png")),
         width = 8,
         height = 3)
```


#### All Genes: Lineplots

```{r fig.height=12, fig.width=3}
# 
# auc_lineplot = auc_lineplot_df %>% 
#   ggplot() +
#    aes(x = timepoint_comparison,
#       y = auc,
#       fill = model,
#       color = model,
#       group = model) + 
#   
#   facet_nested_wrap(~treatment+pathogen+comparison,
#                     scales = "free",
#                     ncol = 2,
#                     strip = strip_nested(size = "variable")
#                     ) +
#   
#   geom_line(width = 1,
#             aes(linetype = model)) +
#   
#   geom_point(size = 2,
#              pch = 21,
#              color = "black",
#              stroke = 0.5) + 
#   geom_text_repel(aes(label = n_genes), size = 2) +
#   # Add a reference line for random prediction (AUC = 0.5)
#   geom_hline(yintercept = 0.5, linetype = "dashed", color = "black", linewidth = 0.2) +
#   # geom_text_repel(aes(label = auc), 
#   #                 size = 2.5, 
#   #                 # direction = "y", 
#   #                 segment.alpha = 1) +
# 
#   scale_x_discrete(expand = c(0.05, 0.05)) +
#   # scale_y_continuous(expand = c(0.01, 0.01)) +
#   theme_vaxgo() +
#   scale_linetype_manual(values = c(
#     "Mouse" = 1,
#     "Immune" = 1,
#     "Human" = 1,
#     "Permutation" = 2
#     )) +
#   scale_fill_manual(values = colors$organism) +
#   scale_color_manual(values = colors$organism) +
#   labs(
#     x = "Time",
#     y = "AUC",
#     fill = "Model",
#     shape = "Model",
#   ) +
#     theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
#           axis.text.y = element_text(size = 8, hjust = 1),
#           axis.title.x = element_text(size = 8),
#           axis.title.y = element_text(size = 8),
#           strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
#           strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
#           strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
#           panel.spacing = unit(0, "lines"),
#           plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
#           legend.position = "top",
#           legend.direction = "horizontal",
#           legend.margin=margin(0,0,0,0),
#           legend.box.margin=margin(0,0,-10,0),
#           legend.text = element_text(size = 8, hjust = 0),
#           legend.title = element_text(size = 8),
#           ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
#     ) +
#   guides(color = "none",
#          shape = "none",
#          linetype = "none",
#          # fill = guide_legend(nrow = 2, by.row = T)
#          )
# 
# auc_lineplot
# auc_lineplot %>%
#   ggsave(filename = here("Figures", paste0(filename, "_performance_roc_auc_lineplot_All_OnlyImmune.png"),
#          width = 3,
#          height = 12)

```





#### DEGs: Lineplots
```{r fig.height=12, fig.width=3}
# 
# auc_DEGs_lineplot = auc_lineplot_df_DEGs %>% 
#   ggplot() +
#    aes(x = timepoint_comparison,
#       y = auc,
#       fill = model,
#       color = model,
#       group = model) + 
#   
#   facet_nested_wrap(~treatment+pathogen+comparison,
#                     scales = "free",
#                     ncol = 2,
#                     strip = strip_nested(size = "variable")
#                     ) +
#   
#   geom_line(width = 1,
#             aes(linetype = model)) +
#   
#   geom_point(size = 2,
#              pch = 21,
#              color = "black",
#              stroke = 0.5) + 
#   geom_text_repel(aes(label = n_genes), size = 2) +
#   # Add a reference line for random prediction (AUC = 0.5)
#   geom_hline(yintercept = 0.5, linetype = "dashed", color = "black", linewidth = 0.2) +
#   # geom_text_repel(aes(label = auc), 
#   #                 size = 2.5, 
#   #                 # direction = "y", 
#   #                 segment.alpha = 1) +
# 
#   scale_x_discrete(expand = c(0.05, 0.05)) +
#   # scale_y_continuous(expand = c(0.01, 0.01)) +
#   theme_vaxgo() +
#   scale_linetype_manual(values = c(
#     "Mouse" = 1,
#     "Immune" = 1,
#     "Human" = 1,
#     "Permutation" = 2
#     )) +
#   scale_fill_manual(values = colors$organism) +
#   scale_color_manual(values = colors$organism) +
#   labs(
#     x = "Time",
#     y = "AUC",
#     fill = "Model",
#     shape = "Model",
#   ) +
#     theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
#           axis.text.y = element_text(size = 8, hjust = 1),
#           axis.title.x = element_text(size = 8),
#           axis.title.y = element_text(size = 8),
#           strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
#           strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
#           strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
#           panel.spacing = unit(0, "lines"),
#           plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
#           legend.position = "top",
#           legend.direction = "horizontal",
#           legend.margin=margin(0,0,0,0),
#           legend.box.margin=margin(0,0,-10,0),
#           legend.text = element_text(size = 8, hjust = 0),
#           legend.title = element_text(size = 8),
#           ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
#     ) +
#   guides(color = "none",
#          shape = "none",
#          linetype = "none",
#          fill = guide_legend(nrow = 2, by.row = T))
# 
# auc_DEGs_lineplot
# 
# auc_DEGs_lineplot %>%
#   ggsave(filename = here("Figures", paste0(filename, "_performance_roc_auc_lineplot_DEGs_OnlyImmune.png"),
#          width = 3,
#          height = 12)

```












###Plot PR curve


#### All genes
```{r fig.height=6, fig.width=8}
#Plot curves ------------------
#PLOT PR Curve
pr_curve_plot = pr_curve_auc_df %>% 
  arrange(pathogen) %>% 
  ungroup() %>% 
  ggplot(aes(x = recall, 
             y = precision, 
             color = model,
             linetype = comparison)) +
  geom_path(size = 0.75, alpha = 1) +
  theme_vaxgo() +
  facet_nested_wrap(~pathogen+timepoint_comparison, nrow = 2, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Mouse
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint_comparison, pr_auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 0.95,
                y = 0.9),
            size = 3,
            hjust = 1) +
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint_comparison,pr_auc ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 0.95,
                y = 0.8),
            size = 3,
            hjust = 1) +
  #Immune
  geom_text(data = . %>% 
              filter(model == "Immune", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint_comparison, pr_auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 0.95,
                y = 0.7),
            size = 3,
            hjust = 1) +
   geom_text(data = . %>% 
              filter(model == "Immune", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint_comparison,pr_auc ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 0.95,
                y = 0.6),
            size = 3,
            hjust = 1) +
  scale_linetype_manual(values = c("Up" = 1,
                                   "Down" = 2)) +
  # coord_equal() +
  scale_color_manual(values = colors$organism) +
  labs(
  x = "Recall",
  y = "Precision",
  title = "All genes",
  color = "Model",
  linetype = "Comparison") + 
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
          axis.title.x = element_text(size = 10),
          axis.text.y = element_text(size = 10),
          axis.title.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 10),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          legend.spacing.x = unit(0, 'cm'),
          legend.spacing.y = unit(0, 'cm'),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing.y = unit(0, "lines"), 
          panel.spacing.x = unit(1, "lines"), 
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          
          ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 0.5, size = 1),
                                 order = 2), 
         fill = "none",
         color = guide_legend(override.aes = list(linewidth = 3)))

pr_curve_plot
pr_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_pr_All_OnlyImmune.png")),
         width = 8,
         height = 6)
```



#### DEGs

```{r fig.height=6, fig.width=10}
#Plot curves ------------------
#PLOT PR Curve
pr_curve_DEGs_plot = pr_curve_auc_df_DEGs %>% 
  arrange(pathogen) %>% 
  ungroup() %>% 
  ggplot(aes(x = recall, 
             y = precision, 
             color = model,
             linetype = comparison)) +
  geom_path(size = 0.75, alpha = 1) +
  theme_vaxgo() +
  facet_nested_wrap(~pathogen+timepoint_comparison, nrow = 2, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Mouse
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint_comparison, pr_auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 0.95,
                y = 0.9),
            size = 3,
            hjust = 1) +
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint_comparison,pr_auc ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 0.95,
                y = 0.8),
            size = 3,
            hjust = 1) +
  #Immune
  geom_text(data = . %>% 
              filter(model == "Immune", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint_comparison, pr_auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 0.95,
                y = 0.7),
            size = 3,
            hjust = 1) +
   geom_text(data = . %>% 
              filter(model == "Immune", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint_comparison,pr_auc ) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, pr_auc),
                color = model,
                x = 0.95,
                y = 0.6),
            size = 3,
            hjust = 1) +
  scale_linetype_manual(values = c("Up" = 1,
                                   "Down" = 2)) +
  # coord_equal() +
  scale_color_manual(values = colors$organism) +
  labs(
  x = "Recall",
  y = "Precision",
  title = "DEGs",
  color = "Model",
  linetype = "Comparison") + 
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
          axis.title.x = element_text(size = 10),
          axis.text.y = element_text(size = 10),
          axis.title.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 10),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          legend.spacing.x = unit(0, 'cm'),
          legend.spacing.y = unit(0, 'cm'),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing.y = unit(0, "lines"), 
          panel.spacing.x = unit(1, "lines"), 
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          
          ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 0.5, size = 1),
                                 order = 2), 
         fill = "none",
         color = guide_legend(override.aes = list(linewidth = 3)))

pr_curve_DEGs_plot

pr_curve_DEGs_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_pr_DEGS_OnlyImmune.png")),
         width = 8,
         height = 6)
```




### Plot AUC
#### All genes: Heatmap
PR-AUC as heatmap
```{r fig.height=3, fig.width=8}
#PR-AUC as heatmap
pr_auc_heatmap = pr_auc_df %>% 
  ggplot() +
   aes(x = comparison,
      y = model %>% fct_rev(),
      fill = pr_auc,
      color = model %>% fct_rev(),
      group = model %>% fct_rev()) + 
  
  facet_nested_wrap(~treatment+pathogen+timepoint_comparison,
                    # scales = "free",
                    nrow = 1,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_tile(color = "black",
             stroke = 0.5) + 
  
  #Add labels
  geom_text(aes(label = round(pr_auc, 2)), size = 3,
             color = "black",
             fontface = "bold",
             vjust = 0.25,
            data = . %>% filter(pr_auc <= 0.7)) +
  geom_text(aes(label = round(pr_auc, 2)), size = 3,
             color = "white",
             fontface = "bold",
             vjust = 0.25,
            data = . %>% filter(pr_auc > 0.7)) +
   geom_text(aes(label = paste0("n=", n_genes)), size = 2.5,
             color = "black",
             vjust = 2,
             data = . %>% filter(pr_auc <= 0.7)) +
  
   geom_text(aes(label = paste0("n=", n_genes)), size = 2,
             color = "white",
             vjust = 2,
             data = . %>% filter(pr_auc > 0.7)) +
 

    scale_x_discrete(expand = c(0.01, 0.01)) +
  scale_y_discrete(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
    scale_fill_gradientn(
        colors = c("white", "#4DBBD5FF","#4361ee"),
        values = scales::rescale(seq(0, 1, length.out = 10)),
        limits = c(0, 1),
        name = "PR_AUC",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 1,
          nbin = 10,
          raster = FALSE
        )
      ) +
  labs(
    x = "Comparison",
    y = "Model",
    title = "All genes",
    fill = "PR-AUC",
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(1, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) 

pr_auc_heatmap
  
#Save
pr_auc_heatmap %>%
  ggsave(filename = here("Figures", 
                         paste0(filename, 
                               "_performance_pr_auc_heatmap_All_OnlyImmune.png")),
         width = 8,
         height = 3)

```


#### DEGs: Heatmap
PR-AUC as heatmap
```{r fig.height=3, fig.width=8}
#PR-AUC as heatmap
pr_auc_DEGs_heatmap = pr_auc_DEGs_df %>% 
  ggplot() +
   aes(x = comparison,
      y = model %>% fct_rev(),
      fill = pr_auc,
      color = model %>% fct_rev(),
      group = model %>% fct_rev()) + 
  
  facet_nested_wrap(~treatment+pathogen+timepoint_comparison,
                    # scales = "free",
                    nrow = 1,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_tile(color = "black",
             stroke = 0.5) + 
  
  #Add labels
  geom_text(aes(label = round(pr_auc, 2)), size = 3,
             color = "black",
             fontface = "bold",
             vjust = 0.25,
            data = . %>% filter(pr_auc <= 0.9)) +
  geom_text(aes(label = round(pr_auc, 2)), size = 3,
             color = "white",
             fontface = "bold",
             vjust = 0.25,
            data = . %>% filter(pr_auc > 0.9)) +
   geom_text(aes(label = paste0("n=", n_genes)), size = 2.5,
             color = "black",
             vjust = 2,
             data = . %>% filter(pr_auc <= 0.9)) +
  
   geom_text(aes(label = paste0("n=", n_genes)), size = 2,
             color = "white",
             vjust = 2,
             data = . %>% filter(pr_auc > 0.9)) +
 

    scale_x_discrete(expand = c(0.01, 0.01)) +
  scale_y_discrete(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
    scale_fill_gradientn(
        colors = c("white", "#4DBBD5FF","#4361ee"),,
        values = scales::rescale(seq(0, 1, length.out = 10)),
        limits = c(0, 1),
        name = "PR_AUC",
        guide = guide_colorbar(
          barwidth = 15,
          barheight = 1,
          nbin = 10,
          raster = FALSE
        )
      ) +
  labs(
    x = "Comparison",
    y = "Model",
    title = "DEGs",
    fill = "PR-AUC",
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(1, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) 

pr_auc_DEGs_heatmap
  
#Save
pr_auc_DEGs_heatmap %>%
  ggsave(filename = here("Figures", 
                         paste0(filename, "_performance_pr_auc_heatmap_DEGs_OnlyImmune.png")),
         width = 8,
         height = 3)

```

####All genes: Lineplots
PR-AUC as lineplots
```{r fig.height=12, fig.width=3}
# pr_auc_lineplot = pr_auc_df %>% 
#   ggplot() +
#    aes(x = timepoint_comparison,
#       y = pr_auc,
#       fill = model,
#       color = model,
#       group = model) + 
#   
#   facet_nested_wrap(~treatment+pathogen+comparison,
#                     scales = "free",
#                     ncol = 2,
#                     strip = strip_nested(size = "variable")
#                     ) +
#   
#   geom_line(width = 1,
#             aes(linetype = model)) +
#   
#   geom_point(size = 2,
#              pch = 21,
#              color = "black",
#              stroke = 0.5) + 
#   geom_text_repel(aes(label = n_genes), size = 2) +
#   # Add a reference line for random prediction (AUC = 0.5)
#     # geom_text_repel(aes(label = auc), 
#   #                 size = 2.5, 
#   #                 # direction = "y", 
#   #                 segment.alpha = 1) +
# 
#   scale_x_discrete(expand = c(0.05, 0.05)) +
#   # scale_y_continuous(expand = c(0.01, 0.01)) +
#   theme_vaxgo() +
#   scale_linetype_manual(values = c(
#     "Mouse" = 1,
#     "Immune" = 1,
#     "Human" = 1,
#     "Permutation" = 2
#     )) +
#   scale_fill_manual(values = colors$organism) +
#   scale_color_manual(values = colors$organism) +
#   labs(
#     x = "Time",
#     y = "PR-AUC",
#     fill = "Model",
#     shape = "Model",
#   ) +
#     theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
#           axis.text.y = element_text(size = 8, hjust = 1),
#           axis.title.x = element_text(size = 8),
#           axis.title.y = element_text(size = 8),
#           strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
#           strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
#           strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
#           panel.spacing = unit(0, "lines"),
#           plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
#           legend.position = "top",
#           legend.direction = "horizontal",
#           legend.margin=margin(0,0,0,0),
#           legend.box.margin=margin(0,0,-10,0),
#           legend.text = element_text(size = 8, hjust = 0),
#           legend.title = element_text(size = 8),
#           ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
#     ) +
#   guides(color = "none",
#          shape = "none",
#          linetype = "none",
#          fill = guide_legend(nrow = 2, by.row = T))
# 
# pr_auc_lineplot
# 
# pr_auc_lineplot %>%
#   ggsave(filename = here("Figures",
#                          paste0(filename, "_performance_pr_auc_lineplot_All_OnlyImmune.png")),
#          width = 3,
#          height = 12)

```





#### DEGs: Lineplots
PR-AUC as lineplots
```{r fig.height=12, fig.width=3}


# pr_auc_lineplot = pr_auc_df %>% 
#   ggplot() +
#    aes(x = timepoint_comparison,
#       y = pr_auc,
#       fill = model,
#       color = model,
#       group = model) + 
#   
#   facet_nested_wrap(~treatment+pathogen+comparison,
#                     scales = "free",
#                     ncol = 2,
#                     strip = strip_nested(size = "variable")
#                     ) +
#   
#   geom_line(width = 1,
#             aes(linetype = model)) +
#   
#   geom_point(size = 2,
#              pch = 21,
#              color = "black",
#              stroke = 0.5) + 
#   geom_text_repel(aes(label = n_genes), size = 2) +
#   # Add a reference line for random prediction (AUC = 0.5)
#     # geom_text_repel(aes(label = auc), 
#   #                 size = 2.5, 
#   #                 # direction = "y", 
#   #                 segment.alpha = 1) +
# 
#   scale_x_discrete(expand = c(0.05, 0.05)) +
#   # scale_y_continuous(expand = c(0.01, 0.01)) +
#   theme_vaxgo() +
#   scale_linetype_manual(values = c(
#     "Mouse" = 1,
#     "Immune" = 1,
#     "Human" = 1,
#     "Permutation" = 2
#     )) +
#   scale_fill_manual(values = colors$organism) +
#   scale_color_manual(values = colors$organism) +
#   labs(
#     x = "Time",
#     y = "PR-AUC",
#     fill = "Model",
#     shape = "Model",
#   ) +
#     theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
#           axis.text.y = element_text(size = 8, hjust = 1),
#           axis.title.x = element_text(size = 8),
#           axis.title.y = element_text(size = 8),
#           strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
#           strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
#           strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
#           panel.spacing = unit(0, "lines"),
#           plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
#           legend.position = "top",
#           legend.direction = "horizontal",
#           legend.margin=margin(0,0,0,0),
#           legend.box.margin=margin(0,0,-10,0),
#           legend.text = element_text(size = 8, hjust = 0),
#           legend.title = element_text(size = 8),
#           ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
#     ) +
#   guides(color = "none",
#          shape = "none",
#          linetype = "none",
#          fill = guide_legend(nrow = 2, by.row = T))
# 
# pr_auc_lineplot
# 
# pr_auc_lineplot %>%
#   ggsave(filename = here("Figures",
#                          paste0(filename, "_performance_pr_auc_lineplot_DEGs_OnlyImmune.png")),
#          width = 3,
#          height = 12)

```








#====================


# BTMs

### Means

```{r}
#Import ========================
#Unite BTM means
dge_btm_mean_process_all = readRDS(dge_btm_mean_process, here("tables", "all_human_mouse_dge_btm_mean_process.rds"))
dge_btm_process_genes_all = readRDS(dge_btm_process_genes, here("tables", "all_human_mouse_dge_btm_process_genes.rds"))


#Import ============
#Import DEGs table
dge_btm_mean_process = dge_btm_mean_process_all

#Import metadata
#Metadata 
all_human_mouse_metadata = readRDS(here("tables", "all_human_mouse_metadata.rds")) %>% 
  filter(!(timepoint %in% c("0h", "Day 0")))

all_human_mouse_metadata_matched = bind_rows(
  # 1. Keep everything for OTHER pathogens (neither S. aureus nor E. coli)
  all_human_mouse_metadata %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))) %>% 
    select(pathogen, organism, timepoint),
    
  # 2. Keep the original MOUSE timepoints for S. aureus and E. coli
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse") %>% 
    select(pathogen, organism, timepoint),
    
  # 3. Create the HUMAN timepoints mapped to Mouse timepoints for these pathogens
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(pathogen, organism) %>% # Keep 'Human' label, drop original timepoint
    inner_join(
      all_human_mouse_metadata %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
) %>% 
  group_by(pathogen, organism, timepoint) %>%
  summarise(n = n(), .groups = "drop") %>% 
  ungroup()



#Pre-process data =============

all_btms_human_mouse_matched = bind_rows(
  # 1. Keep everything that is NOT S. aureus or E. coli
  dge_btm_mean_process %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))),
  
  # 2. Keep the original MOUSE data for S. aureus and E. coli
  dge_btm_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse"),
    
  # 3. Add the DUPLICATED HUMAN data mapped to Mouse timepoints
  dge_btm_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(-timepoint) %>% 
    inner_join(
      all_human_mouse_metadata_matched %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
)

all_btms_human_mouse_matched %>% select(pathogen, organism, timepoint, ) %>% distinct()

all_btms_human_mouse = all_btms_human_mouse_matched %>% 
  inner_join(all_human_mouse_metadata_matched,
               by = join_by("pathogen", "organism", "timepoint")) %>% 
  distinct()
#Check
all_btms_human_mouse %>% select(pathogen, organism, timepoint, ) %>% distinct()
```


```{r}
#Input ======================
set.seed(123)
filename = "all_human_mouse"
gene_set_filename = "BTMs"

btms_df_input = all_btms_human_mouse_matched %>% 
  rename(symbol = process)

######## Run analysis ======
classification_performance_genes_modules_results = classification_performance_compute_genes_modules(btms_df_input)

#Extract results
roc_auc_df = classification_performance_genes_modules_results$roc_auc_df
roc_curve_auc_df = classification_performance_genes_modules_results$roc_curve_auc_df
pr_auc_df = classification_performance_genes_modules_results$pr_auc_df
pr_curve_auc_df = classification_performance_genes_modules_results$pr_curve_auc_df

#Refactor
roc_curve_auc_df = roc_curve_auc_df %>% 
    mutate(model = fct_relevel(model, "Human", "Mouse", "Control", "Immune", "Permutation"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(auc = round(auc, 2) %>% as.character())

roc_auc_df = roc_auc_df%>% 
  mutate(model = fct_relevel(model, "Human", "Mouse", "Immune", "Permutation", "Control", "FIT"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 


pr_curve_auc_df = pr_curve_auc_df %>% 
    mutate(model = fct_relevel(model, "Mouse", "Immune", "Permutation", "Human", "FIT", "Control"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(pr_auc = round(pr_auc, 2) %>% as.character())

pr_auc_df = pr_auc_df %>% 
  mutate(model = fct_relevel(model, "Human", "Mouse", "Immune", "Permutation", "Control", "FIT"),
         comparison = fct_relevel(comparison, "Up", "Down"),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         )

```


Plot ROC curve
```{r fig.height=13, fig.width=4}
#Plot curves ------------------
#PLOT ROC Curve
roc_curve_plot = roc_curve_auc_df %>% 
  # mutate(comparison = as.character(comparison),
  #        comparison = if_else(comparison == "Up", "U", "D")) %>% 
  arrange(pathogen) %>% 
  ungroup() %>% 
  ggplot(aes(x = 1 - specificity, 
             y = sensitivity, 
             color = model,
             linetype = comparison)) +
  geom_path(size = 0.75, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_nested_wrap(~pathogen+timepoint, ncol = 3, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Mouse
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.2),
            size = 2,
            hjust = 1) +
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.05),
            size = 2,
            hjust = 1) +

  scale_linetype_manual(values = c("Up" = 1,
                                   "Down" = 2)) +
  # coord_equal() +
  scale_color_manual(values = colors$organism) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model",
  linetype = "Comparison") + 
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
          axis.title.x = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 8),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          legend.spacing.x = unit(0, 'cm'),
          legend.spacing.y = unit(0, 'cm'),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing.y = unit(0, "lines"), 
          panel.spacing.x = unit(1, "lines"), 
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          
          ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 0.5,
                                                     size = 1),
                                 order = 2,
                                 nrow = 2), 
         fill = "none",
         color = guide_legend(override.aes = list(linewidth = 3),
                              nrow = 2))

roc_curve_plot


roc_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_", gene_set_filename, ".png")),
         width = 5,
         height = 13)
```


```{r fig.height=12, fig.width=6}
#Unite plots
roc_auc_curve_plot = ((roc_curve_plot) /
  (auc_plot + theme(legend.position = "none")
   )
  ) +
  plot_layout(
  guides = "collect",
  heights = c(0.8, 1) 
) +
plot_annotation(
  theme = theme(
    legend.position = "top",
    plot.margin = margin(0.5, 0.5,0.5,0.5, "pt")
  )
) & 
theme(
  panel.spacing = unit(0, "lines"),
  # plot.margin = margin(0, 2, 0, 2, "pt"),
   legend.margin=margin(0,0,0,0),
  legend.title = element_text(size = 8),
  legend.text = element_text(size = 8),
   legend.box.margin=margin(1,0,-10,0),
   plot.margin = margin(0.1, 0.1, 0.1, 0.1, unit = "cm")
)
roc_auc_curve_plot
roc_auc_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_", gene_set_filename, ".png")),
         width = 5,
         height = 12)

```


AUC as lineplots
```{r fig.height=4, fig.width=12}

#Get n genes
ngenes = bind_rows(
  classified_df_probs_up %>%
    filter(pred == "up") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Up") %>% 
    ungroup(),
  classified_df_probs_down %>%
    filter(pred == "down") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Down") %>% 
    ungroup(),
)


# Visualize AUC
auc_lineplot_df <- bind_rows(auc_results_up,
                      auc_results_down) %>%
  inner_join(ngenes, by = join_by(comparison, pathogen, timepoint, model)) %>% 
  mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 

auc_lineplot = auc_lineplot_df %>% 
  ggplot() +
   aes(x = timepoint,
      y = auc,
      fill = model,
      color = model,
      group = model) + 
  
  facet_nested_wrap(~treatment+pathogen+comparison,
                    scales = "free",
                    nrow = 2,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_line(width = 1,
            aes(linetype = model)) +
  
  geom_point(size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n_genes), 
                  color = "black",
                  size = 3) +
  # Add a reference line for random prediction (AUC = 0.5)
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black", linewidth = 0.2) +
  # geom_text_repel(aes(label = auc), 
  #                 size = 2.5, 
  #                 # direction = "y", 
  #                 segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  # scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_linetype_manual(values = c(
    "Mouse" = 1,
    "Immune" = 1,
    "Human" = 1,
    "Permutation" = 2
    )) +
  scale_fill_manual(values = colors$organism) +
  scale_color_manual(values = colors$organism) +
  labs(
    x = "Time",
    y = "AUC",
    fill = "Model",
    shape = "Model",
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(0, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) +
  guides(color = "none",
         shape = "none",
         linetype = "none",
         fill = guide_legend(nrow = 1, by.row = T))

auc_lineplot
auc_lineplot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_auc_lineplot_", gene_set_filename, ".png")),
         width = 12,
         height = 4)

```






### NES
```{r}
#Unite BTM means and NES
gsea_btm_results_saureus = readRDS(here("tables", paste0("saureus_infection_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_ecoli = readRDS(here("tables", paste0("ecoli_infection_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_fluad = readRDS(here("tables", paste0("influenza_fluad_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_engerix = readRDS(here("tables", paste0("hepatitisb_engerixb_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_burn = readRDS(here("tables", paste0("burn_human_mouse_gsea_btm_results.rds")))
gsea_btm_results_trauma = readRDS(here("tables", paste0("trauma_human_mouse_gsea_btm_results.rds")))

gsea_btm_results_all = bind_rows(
  gsea_btm_results_saureus,
  gsea_btm_results_ecoli,
  gsea_btm_results_fluad %>% mutate(timepoint = as.character(timepoint)),
  gsea_btm_results_engerix %>% mutate(timepoint = as.character(timepoint)),
  gsea_btm_results_trauma,
  gsea_btm_results_burn
)%>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint == "2h", "Early hours", timepoint),
         pathogen = case_when(pathogen == "Saureus" ~ "Staphylococcus aureus",
                              pathogen == "Ecoli" ~ "Escherichia coli",
                              T~pathogen)) %>% 
  #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Influenza", "Hepatitis B") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Escherichia coli", "Staphylococcus aureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_comparison = if_else(pathogen %in% c("Escherichia coli", "Staphylococcus aureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early")) %>% 
  mutate(timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")))


#Import ============
#Import DEGs table
dge_btm_mean_process = gsea_btm_results_all

#Import metadata
#Metadata 
all_human_mouse_metadata = readRDS(here("tables", "all_human_mouse_metadata.rds")) %>% 
  filter(!(timepoint %in% c("0h", "Day 0")))

all_human_mouse_metadata_matched = bind_rows(
  # 1. Keep everything for OTHER pathogens (neither S. aureus nor E. coli)
  all_human_mouse_metadata %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))) %>% 
    select(pathogen, organism, timepoint),
    
  # 2. Keep the original MOUSE timepoints for S. aureus and E. coli
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse") %>% 
    select(pathogen, organism, timepoint),
    
  # 3. Create the HUMAN timepoints mapped to Mouse timepoints for these pathogens
  all_human_mouse_metadata %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(pathogen, organism) %>% # Keep 'Human' label, drop original timepoint
    inner_join(
      all_human_mouse_metadata %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
) %>% 
  group_by(pathogen, organism, timepoint) %>%
  summarise(n = n(), .groups = "drop") %>% 
  ungroup()



#Pre-process data =============

all_btms_human_mouse_matched = bind_rows(
  # 1. Keep everything that is NOT S. aureus or E. coli
  dge_btm_mean_process %>% 
    filter(!(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))),
  
  # 2. Keep the original MOUSE data for S. aureus and E. coli
  dge_btm_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Mouse"),
    
  # 3. Add the DUPLICATED HUMAN data mapped to Mouse timepoints
  dge_btm_mean_process %>% 
    filter(pathogen %in% c("Staphylococcus aureus", "Escherichia coli") & organism == "Human") %>% 
    select(-timepoint) %>% 
    inner_join(
      all_human_mouse_metadata_matched %>% 
        filter(organism == "Mouse", 
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>% 
        distinct(),
      by = "pathogen"
    )
)

all_btms_human_mouse_matched %>% select(pathogen, organism, timepoint, ) %>% distinct()

all_btms_human_mouse = all_btms_human_mouse_matched %>% 
  inner_join(all_human_mouse_metadata_matched,
               by = join_by("pathogen", "organism", "timepoint")) %>% 
  distinct()
#Check
all_btms_human_mouse %>% select(pathogen, organism, timepoint, ) %>% distinct()

#Input ======================
set.seed(123)
filename = "all_human_mouse"
gene_set_filename = "BTMs_NES"

btms_df_input = all_btms_human_mouse_matched %>% 
  rename(symbol = process)
```
#### Significant Up/Down vs Not-significant Up/Down

Calculate ROC curve
```{r fig.height=8, fig.width=10}
#=====================================================
# Compute probabilities
#=====================================================

dge_long_genes_up = btms_df_input %>% 
  mutate(
    df = n_genes - 1,
    t = nes/(sd_log2fc/sqrt(n_genes)), 
    pval_one_sided = pt(t, df, lower.tail = FALSE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t > 0, "up", "not_up"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  rename(adj_p_val = pval_one_sided_adj)

dge_long_genes_down = btms_df_input %>% 
  mutate(
    df = n_genes - 1,
    t = nes/(sd_log2fc/sqrt(n_genes)), 
    pval_one_sided = pt(t, df, lower.tail = TRUE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t < 0, "down", "not_down"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  rename(adj_p_val = pval_one_sided_adj) %>% 
  drop_na(adj_p_val)


#=====================================================
# Assign human truth
#=====================================================
truth_human_up = dge_long_genes_up %>% 
  select(pathogen, symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human") 

truth_human_down = dge_long_genes_down %>% 
  select(pathogen, symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

truth_human_up %>% 
  dplyr::count(pathogen, organism, timepoint)

#=====================================================
#Generate reference distributions
#=====================================================
perfect_prediction_up = truth_human_up %>%
  mutate(prob = if_else(truth == "up", 1, 0),
         model = "Human") %>%
  rename(pred = truth) 

permutation_prediction_up = perfect_prediction_up %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), organism = "Permutation",
         model = "Permutation")%>% 
  drop_na(pred)

perfect_prediction_down = truth_human_down %>%
  mutate(prob = if_else(truth == "down", 1, 0),
          model = "Human") %>%
  rename(pred = truth) %>% 
  drop_na(pred)

permutation_prediction_down = perfect_prediction_down %>%
  group_by(pathogen, timepoint) %>% 
  mutate(prob = runif(n()), organism = "Permutation",
         model = "Permutation") %>% 
  drop_na(prob)


perfect_prediction_down %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
# Combine models (Mouse + Immune + Permutation + Perfect)
#=====================================================
models_up = dge_long_genes_up %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_up, permutation_prediction_up) %>% 
  drop_na(prob)

models_down = dge_long_genes_down %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_down, permutation_prediction_down) %>% 
  drop_na(prob)

models_down %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
# Classify probabilities 
#=====================================================
classified_df_probs_up = truth_human_up %>% 
  select(pathogen, symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(pathogen, symbol, timepoint, pred, prob, model),
             by = join_by(pathogen, symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct()

classified_df_probs_down = truth_human_down %>% 
  select(pathogen, symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(pathogen, symbol, timepoint, pred, prob, model),
             by = join_by(pathogen, symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct()

classified_df_probs_up %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
#Compute ROC 
#=====================================================

roc_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  drop_na() %>% 
  group_by(pathogen, model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Up")

roc_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human")) %>%
  drop_na() %>% 
  group_by(pathogen, model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Down")

roc_curve_up_notup %>% 
  dplyr::count(pathogen, model, timepoint)

#=====================================================
#Compute AUC 
#=====================================================
auc_results_up = classified_df_probs_up %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, auc, comparison)

auc_results_down = classified_df_probs_down %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Permutation", "Human", "FIT")) %>%
  group_by(pathogen, model, timepoint) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(pathogen, model, timepoint, auc, comparison)

auc_results_down %>% 
  dplyr::count(pathogen, model, timepoint)


#=====================================================
#Unite all results
#=====================================================
roc_curve_auc_df = bind_rows(auc_results_up,auc_results_down) %>% 
  inner_join(
    roc_curve_up_notup %>% 
      bind_rows(roc_curve_down_notdown %>% 
                  filter(model != "Human")),
    by = join_by(pathogen, model, timepoint, comparison)) %>% 
    mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         # timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma")) %>% 
  mutate(auc = round(auc, 2) %>% as.character())


```


```{r fig.height=13, fig.width=4}
#Plot curves ------------------
#PLOT ROC Curve
roc_curve_plot = roc_curve_auc_df %>% 
  # mutate(comparison = as.character(comparison),
  #        comparison = if_else(comparison == "Up", "U", "D")) %>% 
  arrange(pathogen) %>% 
  ungroup() %>% 
  ggplot(aes(x = 1 - specificity, 
             y = sensitivity, 
             color = model,
             linetype = comparison)) +
  geom_path(size = 0.75, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_nested_wrap(~pathogen+timepoint, ncol = 3, axes = "x",
                    strip = strip_nested(size = "variable")) +
  #Mouse
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Up") %>% 
              select(pathogen, model, comparison, timepoint, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.2),
            size = 2,
            hjust = 1) +
  geom_text(data = . %>% 
              filter(model == "Mouse", comparison == "Down") %>% 
              select(pathogen, model, comparison, timepoint, auc) %>% 
              distinct(),
            mapping = aes(label = paste(comparison, auc),
                color = model,
                x = 1,
                y = 0.05),
            size = 2,
            hjust = 1) +

  scale_linetype_manual(values = c("Up" = 1,
                                   "Down" = 2)) +
  # coord_equal() +
  scale_color_manual(values = colors$organism) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model",
  linetype = "Comparison") + 
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
          axis.title.x = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 8),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          legend.spacing.x = unit(0, 'cm'),
          legend.spacing.y = unit(0, 'cm'),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing.y = unit(0, "lines"), 
          panel.spacing.x = unit(1, "lines"), 
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          
          ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3) +
  guides(linetype = guide_legend(override.aes = list(linewidth = 0.5,
                                                     size = 1),
                                 order = 2,
                                 nrow = 2), 
         fill = "none",
         color = guide_legend(override.aes = list(linewidth = 3),
                              nrow = 2))

roc_curve_plot


roc_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_", gene_set_filename, ".png")),
         width = 5,
         height = 13)
```



AUC as lineplots
```{r fig.height=12, fig.width=4}

#Get n genes
ngenes = bind_rows(
  classified_df_probs_up %>%
    filter(pred == "up") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Up") %>% 
    ungroup(),
  classified_df_probs_down %>%
    filter(pred == "down") %>% 
  group_by(pathogen, timepoint, model) %>% 
  summarise(n_genes = n(),
            comparison = "Down") %>% 
    ungroup(),
)


# Visualize AUC
auc_lineplot_df <- bind_rows(auc_results_up,
                      auc_results_down) %>%
  inner_join(ngenes, by = join_by(comparison, pathogen, timepoint, model)) %>% 
  mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         comparison = factor(comparison, levels = c("Up", "Down")),
         timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 

auc_lineplot = auc_lineplot_df %>% 
  ggplot() +
   aes(x = timepoint,
      y = auc,
      fill = model,
      color = model,
      group = model) + 
  
  facet_nested_wrap(~treatment+pathogen+comparison,
                    scales = "free",
                    ncol = 2,
                    strip = strip_nested(size = "variable")
                    ) +
  
  geom_line(width = 1,
            aes(linetype = model)) +
  
  geom_point(size = 2,
             pch = 21,
             color = "black",
             stroke = 0.5) + 
  geom_text_repel(aes(label = n_genes), 
                  color = "black",
                  size = 3) +
  # Add a reference line for random prediction (AUC = 0.5)
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black", linewidth = 0.2) +
  # geom_text_repel(aes(label = auc), 
  #                 size = 2.5, 
  #                 # direction = "y", 
  #                 segment.alpha = 1) +

  scale_x_discrete(expand = c(0.05, 0.05)) +
  # scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_linetype_manual(values = c(
    "Mouse" = 1,
    "Immune" = 1,
    "Human" = 1,
    "Permutation" = 2
    )) +
  scale_fill_manual(values = colors$organism) +
  scale_color_manual(values = colors$organism) +
  labs(
    x = "Time",
    y = "AUC",
    fill = "Model",
    shape = "Model",
  ) +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8, hjust = 1),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          strip.text = element_text(size = 8, margin = margin(1, 0, 1, 0)),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
          panel.spacing = unit(0, "lines"),
          plot.margin = margin(0, 0.5, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1, size = 0.5)
    ) +
  guides(color = "none",
         shape = "none",
         linetype = "none",
         fill = guide_legend(nrow = 1, by.row = T))

auc_lineplot
# auc_lineplot %>%
#   ggsave(filename = here("Figures", paste0(filename, "_performance_roc_auc_lineplot_", gene_set_filename, ".png")),
#          width = 12,
#          height = 4)

auc_lineplot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_auc_lineplot_", gene_set_filename, "_vertical.png")),
         width = 4,
         height = 12)

```


#####Classify by ranks

```{r fig.height=5, fig.width=10}
# Classificar em quartis para ver a acurácia por faixa
ranking_validation = classified_df_probs_up %>%
  group_by(pathogen, model, timepoint) %>%
  arrange(prob) %>% 
  mutate(quartile = ntile(prob, 2)) %>% # Divide em 2 partes
  group_by(pathogen, model, timepoint, quartile) %>%
  # filter(quartile == 3) %>% # Foca no topo (os 25% maiores NES/Prob)
  summarise(accuracy_in_top_quartile = sum(truth == "up") / n()) %>% 
  
  #Clean
  mutate(model = factor(model, levels = c("Human", "Mouse", "Immune", "Permutation")),
         # comparison = factor(comparison, levels = c("Up", "Down")),
         timepoint = str_replace(timepoint, "Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         pathogen = case_when(pathogen == "Hepatitis B" ~ "HepB",
                              pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              T ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "HepB", "S. aureus", "E. coli", "Burn", "Trauma"),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")
         ) 


# Plotar para comparar os modelos
  ggplot(ranking_validation) +
    aes(
      x = timepoint,
      y = accuracy_in_top_quartile,
      color = model,
      shape = as.character(quartile)
    ) +
  geom_line(aes(group = interaction(model, quartile))) +
  geom_point() +
  # ylim(0, 1) +
  facet_wrap(~treatment+pathogen, 
             nrow= 1,
             scales = "free_x") +
  geom_hline(yintercept = 0.5,
             linetype = 2) +
  theme_vaxgo() +
    scale_color_manual(values = c("Human" = "black",
                                  "Mouse" = "blue", 
                                  "Permutation" = "gray")) +
    # scale_color_viridis_d(option = "mako", direction = -1) +
    theme(legend.position = "top") +
  labs(title = "Accuracy by quartile")

```









