---
title: "5_EvolutionaryAnalysis"
output: html_document
---


# Evolutionary analysis
# Setup

```{r message=FALSE, warning=FALSE}
# Load core packages for the analysis session
library(tidyverse)
library(here)
library(janitor)
library(rstatix)
library(glue)

# Source project-wide functions, themes, and colour palettes
source(here("scripts_notebooks", "required.R"))


```
# Regulation

## Import data

```{r fig.height=5, fig.width=12}
# Import BTM core enrichment results
gsea_btm_results_core_joined =
  readRDS(file = here("tables", "all_human_mouse_gsea_btm_results_core_joined.rds")) %>%
  distinct()

#Import modules and LEGs
dge_btm_process_module_correlation_and_legs = 
  readRDS(file = here("tables", "all_human_mouse_gsea_btm_legs.rds"))

# Import BTM gene-level correlation data
human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune =
  readRDS(here("tables", "human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune.rds"))

# Import homolog alignment data
all_alignments = read_csv(here("tables", "human_to_mouse_alignment_mart_export.txt"))

# Import BTM gene annotations
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

# Reference: all high-confidence orthologs in the database (unfiltered)
alignments_human2mouse_all = all_alignments %>%
  clean_names() %>%
  filter(mouse_orthology_confidence_0_low_1_high == 1) %>%
  drop_na(mouse_gene_name) %>%
  select(gene = mouse_gene_name, mouse_homology_type,
         identity_human2mouse = percent_id_query_gene_identical_to_target_mouse_gene,
         identity_mouse2human = percent_id_target_mouse_gene_identical_to_query_gene) %>%
  mutate(gene = str_to_upper(gene)) %>%
  # Resolve multiple identity scores by taking the median per gene
  group_by(gene) %>%
  summarise(identity_human2mouse = median(identity_human2mouse)) %>%
  ungroup()

# Filter to only BTM core enrichment genes and append process membership
alignments_human2mouse = alignments_human2mouse_all %>% 
  inner_join(dge_btm_process_module_correlation_and_legs,
             by = join_by(gene == symbol))  %>% 
  mutate(status = if_else(status == "Shared", status, "Not-shared"))

#Group by process
protein_identity_process = alignments_human2mouse %>% 
  group_by(treatment, pathogen, timepoint_comparison, process, status) %>% 
  summarise(facet_label = dplyr::first(facet_label), group = dplyr::first(group),
            mean_identity = mean(identity_human2mouse, na.rm = T))

# Import pre-computed human gene information and join with alignment data
gene_info_human =
  readRDS(file = here("tables", "gene_info_human.rds")) %>%
  inner_join(alignments_human2mouse_all, by = "gene")
```


## Transcription factors
Annotate with gene regulation - TF Identify conserved Transcription
Factors (TFs) regulating specific genes in both Humans and Mice

```{r}
# Clean and parse the ChEA3 transcription factor–target gene matrix
gene_attribute_chea <- read_tsv(here("tables", "gene_attribute_matrix.txt"))

# Transform the wide ChEA matrix into long format, keeping only positive scores
gene_attribute_chea_clean <- gene_attribute_chea %>%
  rename(gene = Gene) %>%
  pivot_longer(-gene, names_to = "experiment", values_to = "score") %>%
  filter(score > 0) %>%
  arrange(desc(score)) %>%
  # Split the experiment metadata into TF name, PMID, and method/tissue/species
  separate(experiment,
           into = c("tf", "pmid", "method_tissue_species"),
           sep = "-",
           extra = "merge",
           fill = "right") %>%
  # Detect species from the metadata string (HUMAN or MOUSE)
  mutate(across(where(is.character), str_trim),
         species = case_when(str_detect(method_tissue_species, "HUMAN") ~ "Human",
                              str_detect(method_tissue_species, "MOUSE") ~ "Mouse",
                              TRUE ~ method_tissue_species)) %>%
  filter(species %in% c("Human", "Mouse"))

# Count evidence per (gene, TF) pair across species
# For each TF–target pair, count how many PMIDs support it in Human and in Mouse
gene_attribute_tf_count <- gene_attribute_chea_clean %>%
  ungroup() %>%
  select(gene, tf, pmid, species) %>%
  group_by(gene, tf) %>%
  pivot_wider(names_from = "species",
              values_from = "pmid") %>%
  mutate(
    Human_pmids = map_chr(Human, ~ paste(.x, collapse = ", ")),
    Mouse_pmids = map_chr(Mouse, ~ paste(.x, collapse = ", ")),
    Human_pmids_n = map_int(Human, length),
    Mouse_pmids_n = map_int(Mouse, length)
  ) %>%
  select(!c(Human, Mouse)) %>%
  mutate(across(ends_with("pmids"), ~ na_if(.x, ""))) %>%
  # Flag whether this TF regulates this gene in both organisms
  mutate(both = if_else(!is.na(Human_pmids) & !is.na(Mouse_pmids), "TRUE", "FALSE"))

# Save the cross-species TF-target count for downstream analysis
saveRDS(gene_attribute_tf_count,
        file = here("tables", "regulation_gene_tf_count.rds"))


```

```{r}
# Reload the TF-count data and gene alignment quadrants
gene_attribute_tf_count <- readRDS(file = here("tables", "regulation_gene_tf_count.rds"))
alignments_human2mouse_genes_filtered <- read_rds(file = here("tables", "alignments_human2mouse_genes_filtered.rds"))
alignments_human2mouse_all_global_df_nshared
dge_btm_process_genes_wide

# Build a contingency table of cross-species TF regulation for a few selected genes
# Join with the filtered gene quadrants to show how many TFs are shared vs species-specific
gene_attribute_tf_count %>%
  filter(gene %in% c("XPO1", "CD2", "CDK14", "NDUFB10")) %>%
  inner_join(alignments_human2mouse_genes_filtered %>%
               group_by(status, filter) %>%
               slice_head(n = 5) %>%
               filter(status == "Shared"),
             by = "gene") %>%
  mutate(gene = paste(gene, filter, cor_label)) %>%
  arrange(desc(both)) %>%
  tabyl(gene, both) %>%
  adorn_totals(where = "col") %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns() %>%
  as.data.frame()


```

#### Calculate Shared TFs

```{r fig.height=3, fig.width=5}
# Join the TF regulation table with the gene alignment quadrants
gene_attribute_tf_count_filter_df <- gene_attribute_tf_count %>%
  inner_join(alignments_human2mouse_all_global_df_nshared,
             by = "gene")

# Summarise TF counts per gene, split by cross-species status (both vs species-specific)
gene_attribute_tf_count_filter_df_summary <- gene_attribute_tf_count_filter_df %>%
  arrange(desc(both)) %>%
  group_by(gene, both) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(gene) %>%
  mutate(total = sum(n),
         percentage = round(n / total * 100, 1))

# Compute unique TF counts per gene for each species
gene_attribute_tf_count_test <- gene_attribute_tf_count %>%
  select(gene, tf, Human_pmids_n, Mouse_pmids_n) %>%
  mutate(Mouse_TF = if_else(Mouse_pmids_n > 0, TRUE, FALSE),
         Human_TF = if_else(Human_pmids_n > 0, TRUE, FALSE)) %>%
  group_by(gene) %>%
  summarise(unique_tfs_human = n_distinct(tf[Human_TF == TRUE]),
            unique_tfs_mouse = n_distinct(tf[Mouse_TF == TRUE]))

# gene_attribute_tf_count_filter_df_summary %>% 
#   filter(both == "T") %>% 
#   ggplot() +
#   aes(y = filter,
#       x = percentage,
#       fill = filter) +
#   geom_boxplot() +
#   geom_point(alpha = 0.2) +
#     stat_summary(geom = "pointrange",
#                fun.data = "mean_cl_boot",
#                position = "dodge",
#                color = "red") +
#   theme_vaxgo() +
#   theme(legend.position = "none") +
#   scale_fill_manual(values = colors_all)

gene_attribute_tf_count_filter_df_summary_annotated = gene_attribute_tf_count_filter_df_summary %>%
  arrange(gene) %>% 
  distinct() %>%
    select(-percentage) %>% 
  pivot_wider(values_from = "n",
              names_from = "both") %>% 
  distinct() %>%
  rename(not_shared_tf = "F",
         shared_tf = "T") %>% 
  mutate(shared_tf = if_else(is.na(shared_tf), 0, shared_tf)) %>% 
  inner_join(gene_attribute_tf_count_filter_df %>% 
               ungroup() %>% 
               filter(class %in% c("Shared", "Not-shared")) %>% 
               select(gene, class:cor_label) %>% 
               distinct(),
             by = join_by(gene)) %>% 
  ungroup() %>% 
  mutate(shared_tf_pct = (shared_tf/total) %>% round(2)) %>% 
    mutate(logp = -log(p_value),
         cor_value_abs = abs(cor_value)) %>% 
  #Highlight genes
  mutate(classification = "All") %>% 
  inner_join(gene_attribute_tf_count_test,
             by = "gene")
  
#Highlight genes
shared_genes = gene_attribute_tf_count_filter_df_summary_annotated %>% 
                    filter(gene %in% c(top_shared_notshared_genes %>% 
                                 filter(class == "Shared") %>% 
                                   pull(gene) %>% unique())) %>% 
              mutate(classification = "Shared") 

gene_attribute_tf_count_filter_df_summary_annotated = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  bind_rows(shared_genes) 

gene_attribute_tf_count_filter_df_summary_annotated %>% 
  dplyr::count(classification, class)
  
```

#### Plot Total of TFs by species

Scatterplot of total unique TFs per gene in Human vs Mouse. Points are
coloured by classification (Shared vs All). A diagonal line indicates
equal TF counts between species.

```{r fig.height=5, fig.width=7}
# Compare total unique TF counts per gene between human and mouse
regulation_tfs_total_by_species_plot <- gene_attribute_tf_count_filter_df_summary_annotated %>%
  drop_na(cor_value) %>%
  ggscatter(x = "unique_tfs_mouse",
            y = "unique_tfs_human",
            fill = "classification",
            color = "classification",
            point = FALSE,
            add = "reg.line",
            conf.int = TRUE) +
    scale_x_continuous(limits = c(0, 80)) +
  scale_y_continuous(limits = c(0, 80)) +
  geom_abline(linetype = 2,
              alpha = 0.5) +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "Human (Total TFs)",
       y = "Mouse (Total TFs)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value,
                 label = gene),
             alpha = 0.8,
             pch = 21,
             stroke = 0,
             size = 2) +
    geom_point(aes(fill = cor_value),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(80, 75), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_total_by_species_plot
  
```

```{r fig.height=3, fig.width=5}
#Identity
regulation_tfs_total_identity_plot =  
  gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  ggscatter(x = "total",
            y = "identity_human2mouse", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line", 
            conf.int = T
              ) +
  scale_color_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "Total TFs",
       y = "%Identity (a.a.)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value),
             alpha = 0.2,
             pch = 21,
             stroke = 0) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(120, 115), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_total_identity_plot 

#Identity
regulation_tfs_shared_pct_identity_plot =  
  gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  ggscatter(x = "shared_tf_pct",
            y = "identity_human2mouse", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line", 
            conf.int = T
              ) +
  scale_color_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "%Shared TFs",
       y = "%Identity (a.a.)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value),
             alpha = 0.2,
             pch = 21,
             stroke = 0) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(120, 115), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_shared_pct_identity_plot 

regulation_tfs_identity_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  ggscatter(x = "shared_tf",
            y = "identity_human2mouse", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line", 
            conf.int = T
              ) +
  scale_color_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "Total Shared TFs",
       y = "%Identity (a.a.)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value),
             alpha = 0.2,
             pch = 21,
             stroke = 0) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(120, 115), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_total_identity_plot 
regulation_tfs_identity_plot
```

```{r fig.height=5, fig.width=5}
#Correlation
regulation_tfs_corvalue_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  mutate(quadrant = if_else(cor_value>0, "Positive correlation", "Negative correlation")) %>% 
  drop_na(quadrant) %>% 
  mutate(logp = -log(p_value),
         cor_value_abs = abs(cor_value)) %>% 
  ggscatter(x = "cor_value_abs",
            y = "total", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line",
            # conf.int = TRUE
              ) +
    scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "|Correlation|",
       y = "Total TFs",
       color = "",
       fill = "") +
  geom_point(aes(fill = logp),
             alpha = 1,
             pch = 21, 
             color = "black",
             stroke = 0,
             show.legend = T) +
  scale_fill_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
    guide = guide_coloursteps(
      # show.limits = TRUE,
      # barwidth = unit(4, "cm"), 
      # barheight = unit(0.2, "cm")    
    )
  ) +
  
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(750, 800), 
           size = 4, show.legend = F) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white", 
                  show.legend = F) +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_corvalue_plot

#Correlation
regulation_tfs_corvalue_shared_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  mutate(quadrant = if_else(cor_value>0, "Positive correlation", "Negative correlation")) %>% 
  drop_na(quadrant) %>% 
  ggscatter(x = "cor_value_abs",
            y = "shared_tf", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line",
            # conf.int = TRUE
              ) +
    scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "|Correlation|",
       y = "Shared TFs",
       color = "",
       fill = "") +
  geom_point(aes(fill = logp),
             alpha = 1,
             pch = 21, 
             color = "black",
             stroke = 0,
             show.legend = T) +
  ylim(0, 100) +
  scale_fill_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
    # guide = guide_coloursteps(
    #   show.limits = TRUE,
    #   barwidth = unit(4, "cm"), 
    #   barheight = unit(0.2, "cm")    
    # )
  ) +
  
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(100, 90), 
           size = 4, show.legend = F) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white", 
                  show.legend = F) +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_corvalue_shared_plot

#Correlation
regulation_tfs_corvalue_shared_perc_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  mutate(logp = -log(p_value)) %>% 
  ggscatter(x = "cor_value_abs",
            y = "shared_tf_pct", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line",
            # conf.int = TRUE
              ) +
    scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "|Correlation|",
       y = "%Shared TFs",
       color = "",
       fill = "") +
  geom_point(aes(fill = logp),
             alpha = 1,
             pch = 21, 
             color = "black",
             stroke = 0,
             show.legend = T) +
  # ylim(0, 100) +
  scale_fill_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
    # guide = guide_coloursteps(
    #   show.limits = TRUE,
    #   barwidth = unit(4, "cm"), 
    #   barheight = unit(0.2, "cm")    
    # )
  ) +
  
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(0.20, 0.18), 
           size = 4, show.legend = F) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white", 
                  show.legend = F) +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_corvalue_shared_perc_plot
```

```{r fig.height=6, fig.width=12}
# ── Patchwork: combine all 6 TF regulation panels into two rows (identity and correlation) and save
regulation_plots <- (regulation_tfs_total_identity_plot | regulation_tfs_identity_plot | regulation_tfs_shared_pct_identity_plot) /
  (regulation_tfs_corvalue_plot | regulation_tfs_corvalue_shared_plot | regulation_tfs_corvalue_shared_perc_plot) +
  plot_annotation(title = "Transcription factors vs. Aminoacid identity and Expression correlation",
                  subtitle = "Shared LEGs with p < 0.10 and rsq > 0.2 are highlighted in blue")

regulation_plots

# Save complete regulation figure
ggsave(regulation_plots,
       filename = here("Figures", "all_human_mouse_gsea_leganalysis_regulation_plots.png"),
       width = 12, height = 6)
```

```{r fig.height=4, fig.width=20}
# ── Alternative layout: single-row patchwork with 4 panels (identity + correlation) and save
regulation_plots <- (regulation_tfs_total_identity_plot | regulation_tfs_shared_pct_identity_plot | regulation_tfs_corvalue_plot | regulation_tfs_corvalue_shared_perc_plot) +
  plot_annotation(title = "Transcription factors vs. Aminoacid identity and Expression correlation",
                  subtitle = "Shared LEGs with p < 0.10 and rsq > 0.2 are highlighted in blue")

#Print
regulation_plots

# Save compact regulation figure
ggsave(regulation_plots,
       filename = here("Figures", "all_human_mouse_gsea_leganalysis_regulation_plots.png"),
       width = 20, height = 4)
```

#### By correlation

```{r}
gene_attribute_tf_count_filter_df = gene_attribute_tf_count %>% 
  ungroup() %>% 
  # Join with the filtered gene quadrants (e.g., High Corr/Low Identity)
  inner_join(alignments_human2mouse_genes_filtered,
             by = "gene")

gene_attribute_tf_count_filter_df_summary = gene_attribute_tf_count_filter_df %>% 
  # Create a descriptive label for the final table
  arrange(desc(both)) %>% 
  group_by(
          # filter, 
           gene, both) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  group_by(gene) %>% 
  mutate(total = sum(n),
         percentage = (n/total*100) %>% round(1))  



gene_attribute_tf_count_filter_df_summary_annotated = gene_attribute_tf_count_filter_df_summary %>%
  arrange(gene) %>% 
  distinct() %>%
    select(-percentage) %>% 
  pivot_wider(values_from = "n",
              names_from = "both") %>% 
  distinct() %>%
  rename(not_shared_tf = "F",
         shared_tf = "T") %>% 
  mutate(shared_tf = if_else(is.na(shared_tf), 0, shared_tf)) %>% 
  inner_join(gene_attribute_tf_count_filter_df %>% 
               ungroup() %>% 
               filter(class %in% c("Immune\nBTMs", "Non-Immune\nBTMs")) %>%
               select(gene, filter, identity_human2mouse, class:cor_label) %>% 
               distinct(),
             by = join_by(gene)) %>% 
  ungroup() %>% 
  mutate(shared_tf_pct = (shared_tf/total) %>% round(2))



#Correlation
gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  # mutate(quadrant = if_else(cor_value>0, "Positive correlation", "Negative correlation")) %>% 
  # drop_na(quadrant) %>% 
  ggplot() +
  aes(x = total,
      y = cor_value,
      label = gene) +
  # facet_wrap(~quadrant,
  #            scales = "free") +
  geom_point(alpha = 0.1) +
  geom_point(data = . %>% 
                filter(p_value < 0.10,
                       r_squared > 0.2),
             color = "black",
             alpha = 0.8) +
  geom_point(color = "black",
             size = 3,
             pch = 21,
             fill = "#4cc9f0",
             data =  gene_attribute_tf_count_filter_df_summary_annotated %>% 
                    inner_join(top_shared_notshared_genes %>% 
                                 filter(class == "All") %>% 
                                 filter(p_value < 0.10,
                                 r_squared > 0.2) %>% 
                                 select(gene) %>% 
                                 distinct(),
                               by = "gene")) +
  geom_label_repel(color = "black",
                   data =  gene_attribute_tf_count_filter_df_summary_annotated %>% 
                    inner_join(top_shared_notshared_genes %>% 
                                 filter(class == "All") %>%
                                 filter(p_value < 0.10,
                                 r_squared > 0.2) %>% 
                                 select(gene) %>% 
                                 distinct(),
                               by = "gene"),
                  size = 2.5,
                  fill = "white") +
  # geom_smooth(method = "lm") +
  # geom_smooth(method = "lm",
  #             data = . %>% 
  #               filter(p_value < 0.10,
  #                      r_squared > 0.2),
  #            color = "blue") +
  theme_vaxgo()



#Identity
gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  ggplot() +
  aes(x = total,
      y = identity_human2mouse,
      # label = gene
      ) +
  geom_point(alpha = 0.1) +
  geom_point(
    # data = . %>% 
    #             filter(p_value < 0.10,
    #                    r_squared > 0.2),
             color = "black",
             pch = 21,
             alpha = 0.8,
             mapping = aes(fill = filter)) +

  # geom_smooth(method = "lm") +
  theme_vaxgo()



gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  ggplot() +
  aes(x = shared_tf,
      y = identity_human2mouse,
      # label = gene
      ) +
  geom_point(alpha = 0.1) +
  geom_point(
    # data = . %>% 
    #             filter(p_value < 0.10,
    #                    r_squared > 0.2),
             color = "black",
             pch = 21,
             alpha = 0.8,
             mapping = aes(fill = filter)) +

  # geom_smooth(method = "lm") +
  theme_vaxgo()

```















#------------------------

## Cis-regulatory elements

```{r packages}
source(here::here("scripts_notebooks", "required.R"))

# library(BSgenome.Hsapiens.UCSC.hg38)
# library(BSgenome.Mmusculus.UCSC.mm10)
# library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(GenomeInfoDb)
library(GenomicRanges)
library(IRanges)
library(rtracklayer)
library(org.Hs.eg.db)
library(pwalign)
library(biomaRt)
```

### Import and merge data

```{r import-data}
#Import Human and Mouse CREs ===========
#In contrast to the Human dataset, there are no CTCF-bound CREs in the main, full CRE mouse dataset. So, we imported CREs separately and merged them.

# Human CREs (GRCh38) =============

#PLS
cres_human_pls <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_PLS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_original_human = type_human,
         subtype_human = "not-CTCF") 

#pELS-dELS
cres_human_pelsdels <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_pELS-dELS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_original_human = type_human,
         subtype_human = "not-CTCF") 

#CTCF BOUND
cres_human_ctcfbound <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_CTCF-bound.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type_original"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_human = str_remove(type_original_human, ",CTCF-bound"),
         subtype_human = "CTCF")

#Check overlaps and remove CTCF-bound cCREs that are not annotated accordingly =====
#PLS
cres_human_ctcfbound_pls_overlap = cres_human_ctcfbound %>% 
  inner_join(cres_human_pls,
             by = "id_ccre_human") %>% 
  select(id_ccre_human) %>% 
  distinct() %>% 
  arrange(id_ccre_human) #36,430 IDs

#47,532 -> 11,102 IDs
cres_human_pls_clean = cres_human_pls %>% 
  filter(!id_ccre_human %in% cres_human_ctcfbound_pls_overlap$id_ccre_human) 

#pELS-dELS
cres_human_ctcfbound_pelsdels_overlap = cres_human_ctcfbound %>% 
  inner_join(cres_human_pelsdels,
             by = "id_ccre_human")%>% 
  select(id_ccre_human) %>% 
  distinct() #720,306 IDs

#1,718,669 -> 998,363
cres_human_pelsdels_clean = cres_human_pelsdels %>% 
  filter(!id_ccre_human %in% cres_human_ctcfbound_pelsdels_overlap$id_ccre_human)

#Merge
cres_human = bind_rows(cres_human_pls_clean,
          cres_human_pelsdels_clean,
          cres_human_ctcfbound) 


# Mouse CREs (mm10) ============
#PLS
cres_mouse_pls <- read_tsv(
  here("tables", "Genomic", "genomic_mouse_cCREs_mm10_PLS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_original_mouse = type_mouse,
         subtype_mouse = "not-CTCF") 

#pELS-dELS
cres_mouse_pelsdels <- read_tsv(
  here("tables", "Genomic", "genomic_mouse_cCREs_mm10_pELS-dELS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_original_mouse = type_mouse,
         subtype_mouse = "not-CTCF")

#CTCF BOUND
cres_mouse_ctcfbound <- read_tsv(
  here("tables", "Genomic", "genomic_mouse_cCREs_mm10_CTCF-bound.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type_original"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_mouse = str_remove(type_original_mouse, ",CTCF-bound"),
         subtype_mouse = "CTCF")

#Check overlaps and remove CTCF-bound cCREs that are not annotated accordingly =====
#PLS
cres_mouse_ctcfbound_pls_overlap = cres_mouse_ctcfbound %>% 
  inner_join(cres_mouse_pls,
             by = "id_ccre_mouse") %>% 
  select(id_ccre_mouse) %>% 
  distinct() %>% 
  arrange(id_ccre_mouse) #11.884 IDs

#27.332 -> 15,448 IDs
cres_mouse_pls_clean = cres_mouse_pls %>% 
  filter(!id_ccre_mouse %in% cres_mouse_ctcfbound_pls_overlap$id_ccre_mouse) 

#pELS-dELS
cres_mouse_ctcfbound_pelsdels_overlap = cres_mouse_ctcfbound %>% 
  inner_join(cres_mouse_pelsdels,
             by = "id_ccre_mouse")%>% 
  select(id_ccre_mouse) %>% 
  distinct() #66.186 IDs

#512.001 -> 445,815 
cres_mouse_pelsdels_clean = cres_mouse_pelsdels %>% 
  filter(!id_ccre_mouse %in% cres_mouse_ctcfbound_pelsdels_overlap$id_ccre_mouse)

#Merge
cres_mouse = bind_rows(cres_mouse_pls_clean,
          cres_mouse_pelsdels_clean,
          cres_mouse_ctcfbound)


# Import homologous CREs ====================
# Homologous CRE pairs from ENCODE (hg38–mm10) Nearest gene, protein coding only
homologous_cres <- read_tsv(
  here("tables", "Genomic", "genomic_human_mouse_cCREs_homologous_GRCh38_Mm10.tsv"),
  col_names = c("human_enhancer", "mouse_enhancer"),
  show_col_types = FALSE
) #537,822

# Load ENCODE comprehensive gene-CRE association table
genes_cres_human <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_Closest-Genes-PC.tsv"),
  col_names = c("chrom_cre", "start_cre", "end_cre", "id_dhs", "id_ccre", "type_cre",
                "chrom_gene", "start_gene", "end_gene", "transcript_id", "score_gene", "strand_gene", "gene_id"),
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) #2,651,787

########### GENEHANCER

#Gene interactions with enhancers
genehancer_interactions_doubleelite = read_tsv(here("tables", "Genomic", "GeneHancer_Chr1-22_ChrX_Interactions_DoubleElite_hgTables.txt")) %>% 
  separate(name, into = c("hgnc_symbol", "genehancer_id")) %>%
  filter(chromEnd >= chromStart) %>% 
  rename(geneScore = score,
         geneValue = value,)
                                               
#Enhancers                                              
genehancer_enhancers_doubleelite = read_tsv(here("tables", "Genomic", "GeneHancer_Chr1-22_ChrX_Enhancers_DoubleElite_hgTables.txt")) %>%
  filter(chromEnd >= chromStart)


########## Get overlap between ENCODE4 and Genehancer
library(GenomicRanges)
#Convert to GRanges
genehancer_enhancers_doubleelite_ranges = makeGRangesFromDataFrame(genehancer_enhancers_doubleelite, 
                                  keep.extra.columns = TRUE,
                                  seqnames.field = "chrom",
                                  start.field = "chromStart",
                                  end.field = "chromEnd", 
                                  na.rm = T)

#Convert to GRanges
cres_human_ranges = makeGRangesFromDataFrame(cres_human, 
                                  keep.extra.columns = TRUE,
                                  seqnames.field = "chrom_human",
                                  start.field = "start_human",
                                  end.field = "end_human", 
                                  na.rm = T)
# Find overlaps
distances_ranges <- distanceToNearest(genehancer_enhancers_doubleelite_ranges, 
                     cres_human_ranges) %>% 
  as.data.frame()

#Annotate Genehancer ids (ELITE) with ENCODE4 ids
cres_human_encode_genehancer <- data.frame(
  genehancer_enhancers_doubleelite_ranges[distances_ranges$queryHits], 
  cres_human_ranges[distances_ranges$subjectHits],
  distance = distances_ranges$distance
) %>% 
  rename_with(
    ~ str_replace(., "\\.1$", "_encode4"), 
    ends_with(".1")                        
  ) %>% 
  rename(chrom = seqnames,
         chromStart = start,
         chromEnd = end,
         genehancer_id = name) %>%
  
  #Merge gene interaction scores
  inner_join(genehancer_interactions_doubleelite,
             by = join_by("genehancer_id", "chrom", "chromStart", "chromEnd")
             )


cres_human_encode_genehancer %>% 
  glimpse()
```




### Summarize data 
Count how many mouse CREs each human CRE maps to (and vice versa)

```{r classify-match-type}

#Summarize classes
cres_human_encode_genehancer %>% 
  dplyr::count(type_human, subtype_human) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_human, desc(n))

cres_mouse %>% 
  dplyr::count(type_mouse, subtype_mouse) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_mouse, desc(n))

# Count how many mouse CREs each human CRE maps to (and vice versa)
human_counts <- homologous_cres %>%
  dplyr::count(human_enhancer, name = "n_mouse_per_human")

mouse_counts <- homologous_cres %>%
  dplyr::count(mouse_enhancer, name = "n_human_per_mouse")

# Full join: preserves orphan CREs on both sides
cres_annotated <- homologous_cres %>%
  left_join(human_counts, by = "human_enhancer") %>%
  left_join(mouse_counts, by = "mouse_enhancer") %>%
  full_join(cres_human_encode_genehancer, by = join_by(human_enhancer == id_ccre_human)) %>%
  full_join(cres_mouse, by = join_by(mouse_enhancer == id_ccre_mouse)) %>%
  mutate(
    is_homologous = !is.na(human_enhancer) & !is.na(mouse_enhancer),
    match_type = case_when(
      is.na(human_enhancer)                            ~ "mouse_only",
      is.na(mouse_enhancer)                            ~ "human_only",
      n_mouse_per_human == 1 & n_human_per_mouse == 1  ~ "1:1",
      n_mouse_per_human >  1 & n_human_per_mouse == 1  ~ "N:1",
      n_mouse_per_human == 1 & n_human_per_mouse >  1  ~ "1:N",
      TRUE                                             ~ "N:M"
    )
  ) %>% 
  rename(chrom_human = chrom,
         start_human = chromStart,
         end_human = chromEnd,
         )

#Filter 1 to 1
cres_homologous_1to1 <- cres_annotated %>% #856,415
  dplyr::filter(match_type == "1:1") %>% #277,898
  #Filter enhancers with complete annotation in both species
  dplyr::filter(!is.na(chrom_human), !is.na(chrom_mouse))#1,567

# Separate tables
cres_homologous <- cres_annotated %>% dplyr::filter(is_homologous)
cres_human_only <- cres_annotated %>% dplyr::filter(match_type == "human_only")
cres_mouse_only <- cres_annotated %>% dplyr::filter(match_type == "mouse_only")

#Summarize
cres_homologous_1to1$type_human %>% unique()
#Compare classes of CREs between species
cres_homologous_1to1_comparison = cres_homologous_1to1 %>% 
  
  #Filter PLS, pELS, dELS
    filter(type_human %in% c("pELS", "dELS", "PLS"),
         type_mouse %in% c("pELS", "dELS", "PLS")) %>% 
  
  #Summarize
  group_by(type_human, subtype_human) %>% 
  dplyr::count(type_human, subtype_human, type_mouse, subtype_mouse) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(desc(n)) %>%
  mutate(type_human = fct_relevel(type_human, "PLS", "pELS", "dELS"),
         type_mouse = fct_relevel(type_mouse, "PLS", "pELS", "dELS")
         )

#Print
cres_homologous_1to1_comparison
```

Visualize

```{r classify-match-type, fig.height=5, fig.width=5}
#Heatmap

#Elite CREs
cres_homologous_1to1_comparison %>% 
    dplyr::filter(!is.na(hgnc_symbol)) %>% #1,706
  #Filter ELITE enhancers
  dplyr::filter(eliteness == "Elite") %>%  ##1,567
  #Plot
  ggplot() +
    aes(x = subtype_mouse %>% str_replace(., "-", "\n"),
         y = subtype_human %>% str_replace(., "-", "\n")) +
  facet_grid(~type_human~type_mouse) +
  geom_tile(aes(fill = pct), color = "black", linewidth = 0.4) +
  geom_text(aes(label = paste0(pct, "%")),
            size = 3) +
  theme_vaxgo() +
  scale_fill_gradient2(low = "white",
                       high = "#4361ee",
                       guide = guide_colorbar(barwidth = unit(2, "cm"))) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  theme(axis.text.x = element_text(angle = 0,
                                   # vjust = 0,
                                   hjust = 0.5),
        legend.position = "top") +
  labs(x = "Mouse",
       y = "Human",
       fill = "%Shared, from human")


# All CREs
cres_homologous %>% 
  #Plot
  ggplot() +
    aes(x = subtype_mouse %>% str_replace(., "-", "\n"),
         y = subtype_human %>% str_replace(., "-", "\n")) +
  facet_grid(~type_human~type_mouse) +
  geom_tile(aes(fill = pct), color = "black", linewidth = 0.4) +
  geom_text(aes(label = paste0(pct, "%")),
            size = 3) +
  theme_vaxgo() +
  scale_fill_gradient2(low = "white",
                       high = "#4361ee",
                       guide = guide_colorbar(barwidth = unit(2, "cm"))) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  theme(axis.text.x = element_text(angle = 0,
                                   # vjust = 0,
                                   hjust = 0.5),
        legend.position = "top") +
  labs(x = "Mouse",
       y = "Human",
       fill = "%Shared, from human")


```

------------------------------------------------------------------------

### Gene annotation

```{r import-gene-annotation}

###############
# ADAPT
###############

# Import gene expression correlation
# Mock
expression_correlation <- tibble(
  hgnc_symbol = c("ENTPD1", "ASGR1", "MARCKS", "IL7R"),
  mgi_symbol  = c("Entpd1", "Asgr1", "Marcks", "Il7r"),
  rho         = c(0.85, 0.79, -0.42, -0.61)
)

all_human_mouse_dge_btm_process_genes_correlation_bygene <- readRDS( here("tables", "all_human_mouse_dge_btm_process_genes_correlation_bygene.rds"))

############## 
# Query biomaRt to map human Ensembl IDs to HGNC and MGI symbols
human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

hgnc_mgi_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = unique(cres_human_encode_genehancer$gene_id_clean),
  mart       = human_mart
)

# Join HGNC symbols to gene table
genes_cres_human <- genes_cres_human %>%
  left_join(hgnc_mgi_map, by = c("gene_id_clean" = "ensembl_gene_id"))


# Compute multi-CRE metrics per homologous gene pair
gene_level_cre_metrics <- genes_cres_human %>%
  group_by(hgnc_symbol, type_cre) %>%
  summarise(
    n_cres_human       = n(),
    n_cres_orthologous = sum(id_ccre %in% cres_homologous_1to1$human_enhancer),
    pct_cres_conserved = n_cres_orthologous / n_cres_human,
    .groups = "drop"
  ) 
  # inner_join(all_human_mouse_dge_btm_process_genes_correlation_bygene, by = c("hgnc_symbol" = "hgnc_symbol"))


# Annotate cres_annotated with human gene info (one gene per CRE)
cres_annotated <- cres_annotated %>%
  left_join(
    genes_cres_human %>%
      dplyr::select(id_ccre, gene_id, hgnc_symbol,                     transcript_id, chrom_gene, start_gene, end_gene, strand_gene) %>%
      distinct(id_ccre, .keep_all = TRUE),
    by = c("human_enhancer" = "id_ccre")
  )

```



(DOES NOT WORK) Get evolutionary distance

```{r}
# library(biomaRt)
# 
# # 1. Definir o banco de dados antigo usando o host do archive desejado (ex: jan2020)
# # Como você quer partir do camundongo para buscar os homólogos humanos, usamos "mmusculus_gene_ensembl"
# listEnsemblArchives()
# listMarts()
# 
# options(timeout = 600)
# 
# # ensembl_human_old = useEnsembl(biomart = "ensembl",
# #                        dataset="hsapiens_gene_ensembl",
# #                        version="77")
# 
# ensembl_human_old = useMart(biomart = "ENSEMBL_MART_ENSEMBL",
#                              dataset="hsapiens_gene_ensembl",
#                             host = "https://oct2014.archive.ensembl.org")
# 
# listFilters(ensembl_human_old) %>% View()
# listAttributes(ensembl_human_old) %>% View()
# 
# # 2. Carregar sua lista de IDs de genes de camundongo (ex: ENSMUSG0000XXXXXXXX)
# human_genes_list <- all_human_mouse_dge_btm_process_genes_correlation_bygene$symbol
# 
# # 3. Fazer a busca trazendo o ID do camundongo e os dados do homólogo humano (hsapiens)
# options(timeout = 600)
# human_mouse_orthologs_old <- getBM(attributes = c(
#                                                   "ensembl_gene_id",
#                                                   "external_gene_name"
#                                           # "mmusculus_homolog_dn", 
#                                           # "mmusculus_homolog_ds"
#                                           ), 
#                             filters = 'external_gene_name', 
#                             values = "CD6", 
                            # mart = ensembl_human_old)

```






How many CREs per gene? And how many genes per enhancer?
```{r fig.height=5, fig.width=10}
#Distribution
gene_level_cre_metrics %>% 
  filter(hgnc_symbol != "") %>% 
  ggplot() +
  aes(x = n_cres_human,
      y = "Human",
      label = hgnc_symbol) +
  geom_violin(fill = "gray90",
              color = NA) +
  geom_boxplot(width = 0.2,
               color = "#4DBBD5FF",
               alpha = 0) +
  scale_x_continuous(transform = "log10") +
  geom_quasirandom(groupOnX = FALSE, # Permite a distribuição vertical em um X contínuo
                 varwidth = TRUE,  # Ajusta a largura de acordo com a densidade
                 pch = 21,
                 mapping = aes(fill = pct_cres_conserved)) +
  scale_color_continuous() +
  geom_label_repel(size = 2.5,
                  data = . %>% slice_max(n = 5, 
                                         order_by = n_cres_human)) +
  theme_vaxgo()

```


```{r fig.height=4, fig.width=10}

library(ggpubr)
#Is there any correlation between CREs metrics and types of CREs?
gene_level_cre_metrics %>% 
  inner_join(all_human_mouse_dge_btm_process_genes_correlation_bygene %>% 
               rename(hgnc_symbol = symbol), 
             by = join_by(hgnc_symbol)) %>% 
  filter(hgnc_symbol != "",
         # p_value <= 0.10,
         n_pathogens > 10,
         type_cre %in% c("PLS", "pELS", "dELS", "TF")) %>% 
  ggplot() +
  aes(
      x = n_cres_orthologous,
      # x = n_cres_human,
      # x = pct_cres_conserved,
      y = cor_value, 
      label = hgnc_symbol,
      color = type_cre,
      # size = n_pathogens
      ) +
  theme_vaxgo() +
  facet_wrap(~type_cre, nrow = 1, 
             scales = "free_x")+
  geom_point(alpha = 0.2,
             ) +
  geom_smooth(method = "lm", color = "black", se = TRUE, size = 0.8) +
  
  stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top", color = "black") +
  theme(legend.position = "top") 
  
  # geom_label_repel(color = "black",
  #                 size = 2.5,
  #                 data = . %>% 
  #                   slice_max(pct_cres_conserved, n = 10))


```



------------------------------------------------------------------------

###  Sequence extraction

The HOXD55 matrix (originally derived by Chiaromonte et al., 2002, and deeply analyzed in Blanchette et al., 2004) defines nucleotide matches and mismatches using large integers to maximize sensitivity when calculating neutral substitution rates in mammalian alignments:Match: $+91$Mismatch: $-114$Because a single mismatch penalizes the score by $-114$, the gap penalties must be scaled proportionally. If you used a standard gap opening penalty of $-10$, the alignment algorithm would erroneously introduce hundreds of micro-gaps instead of accepting a single nucleotide mismatch, completely ruining the biological alignment of your cis-regulatory elements (CREs).

To maintain perfect parity with the alignments you see on the UCSC Genome Browser for human (hg38) and mouse (mm10), you use their default parameters:Gap Opening Penalty ($O$): $-400$Gap Extension Penalty ($E$): $-30$

Chiaromonte, F., et al. (2002). Scoring pairwise genomic alignments. Pacific Symposium on Biocomputing. This paper established the ancestral mammalian repeat substitution matrices (including the HOXD matrices). Blanchette, M., et al. (2004). Blastz: A tool for aligning a large number of long genomic sequences. Genome Research. This paper formally documents the open-source BLASTZ/LASTZ defaults ($O = 400, E = 30$, match $= 91$, mismatch $= 114$) used by UCSC. Kent, W. J., et al. (2003). Evolution's Cauldron: Duplication, Deletion, and Rearrangement in the Mouse and Human Genomes. PNAS. Documents how UCSC constructs the chain and net alignments utilizing these exact parameters.

Align homolog enhancer sequences

```{r extract-sequences}
#Filter homologs with mapped coordinates
cres_to_align_homologous <- cres_homologous_1to1 %>%
  filter(human_enhancer %in% genes_cres_human$id_ccre) %>%
  filter(!is.na(chrom_human), !is.na(chrom_mouse)) %>%
  #Annotate gene symbol
  inner_join(cres_annotated %>% 
            select(human_enhancer, mouse_enhancer, hgnc_symbol),
        by = join_by(human_enhancer, mouse_enhancer))

#Convert to GRanges objects
gr_human <- GRanges(seqnames = cres_to_align_homologous$chrom_human, ranges = IRanges(start = cres_to_align_homologous$start_human, end = cres_to_align_homologous$end_human))
gr_mouse <- GRanges(seqnames = cres_to_align_homologous$chrom_mouse, ranges = IRanges(start = cres_to_align_homologous$start_mouse, end = cres_to_align_homologous$end_mouse))

#Get sequences
cres_to_align_homologous$sequence_human <- as.character(getSeq(BSgenome.Hsapiens.UCSC.hg38, gr_human))
cres_to_align_homologous$sequence_mouse <- as.character(getSeq(BSgenome.Mmusculus.UCSC.mm10, gr_mouse))

#Clean sequences from undefined characters
cres_to_align_homologous_clean = cres_to_align_homologous %>% 
  mutate(sequence_human = str_remove_all(sequence_human, "N"),
         sequence_mouse = str_remove_all(sequence_mouse, "N"))

# Parameterize a nucleotide substitution matrix mirroring the standard HOXD55 values
matriz_hoxd55 <- nucleotideSubstitutionMatrix(match = 91, mismatch = -114, baseOnly = TRUE)

#Align sequences
aln_cres <- pairwiseAlignment(
  DNAStringSet(cres_to_align_homologous_clean$sequence_human),
  DNAStringSet(cres_to_align_homologous_clean$sequence_mouse),
  type         = "local",
  substitutionMatrix = matriz_hoxd55,
  gapOpening   = -400, # Score scale adjusted for standard UCSC genomic alignment scoring
  gapExtension = -30
)

#Calculate percent identity
cres_to_align_homologous_identity <- cres_to_align_homologous_clean %>%
  mutate(
    cre_identity_pct = pid(aln_cres, type = "PID1"),
    cre_align_score  = score(aln_cres),
    cre_align_length = nchar(alignedPattern(aln_cres)),
    cre_coverage_human = cre_align_length / nchar(sequence_human)
  )



#Save results
cres_to_align_homologous_identity %>% 
  saveRDS(file = here("tables", "cres_homologous_1to1_annotated_aligned.rds"))

#Summarize alignment results
summary(cres_to_align$cre_identity_pct)

cat("\n% identity > 80%:", round(mean(cres_to_align$cre_identity_pct > 80) * 100, 1), "%\n")
cat("% identity > 70%:", round(mean(cres_to_align$cre_identity_pct > 70) * 100, 1), "%\n")
cat("% identity > 60%:", round(mean(cres_to_align$cre_identity_pct > 60) * 100, 1), "%\n")

#Which genes have the most conserved enhancers?
cres_to_align %>% 
  select(hgnc_symbol, human_enhancer, mouse_enhancer, type_human, type_mouse, cre_identity_pct:cre_coverage_human) %>% 
  group_by(hgnc_symbol, type_human, type_mouse) %>% 
  summarise(median_pct_id = median(cre_identity_pct)) %>%
  arrange(-median_pct_id) 

```

###  Alignment of proximal TSS windows

```{r}
tx_human <- transcripts(TxDb.Hsapiens.UCSC.hg38.knownGene)

# Multi-scale biological window definitions centered on the TSS
windows_definition <- list(
  w3000 = list(up = 3000, down = 300),  # Full proximal regulatory landscape (pELS/dELS capture)
  w1500 = list(up = 1500, down = 300),  # Core promoter + immediate proximal enhancers
  w500  = list(up = 500,  down = 100),  # Core promoter locus only
  w200  = list(up = 200,  down = 50)    # Minimal transcription initiation footprint
)

# Exemplary extraction and global alignment (Needleman-Wunsch) loop initialization for w3000
tss_human_w3000 <- promoters(TxDb.Hsapiens.UCSC.hg38.knownGene, upstream = 3000, downstream = 300)
tss_mouse_w3000 <- promoters(TxDb.Mmusculus.UCSC.mm10.knownGene, upstream = 3000, downstream = 300)

# Implementation Note: Filter your GRanges objects by matched homologous pairs prior to running global alignment:
aln_global <- pairwiseAlignment(seq_human_gene, seq_mouse_gene, type = "global", substitutionMatrix = matriz_hoxd55)
```

###  Annotate conservation and TEs

```{r}
ah <- AnnotationHub()

# 1. Fetch the most recent RepeatMasker records for hg38
rmsk_query <- query(ah, c("RepeatMasker", "Homo sapiens", "hg38"))
rmsk_hg38  <- ah[[names(rmsk_query)[1]]]

# 2. Compute the exact fraction of the CRE masked by repetitive components
gr_cres_human_all <- GRanges(
  seqnames = cres_to_align$chrom_human, 
  ranges = IRanges(start = cres_to_align$start_human, end = cres_to_align$end_human)
)

hits_rmsk <- findOverlaps(gr_cres_human_all, rmsk_hg38)

# Calculate coverage to map base-pair level masking within the ranges
repeat_cov <- GenomicRanges::coverage(rmsk_hg38)
seqlevels(gr_cres_human_all, pruning.mode = "coercion") <- names(repeat_cov)
masked_views <- Views(repeat_cov, ranges(gr_cres_human_all))

cres_to_align$repeat_masked_pct <- sapply(masked_views, function(x) sum(as.numeric(x) > 0)) / width(gr_cres_human_all) * 100
```

### Integrative Statistical Modeling (Linear Mixed-Effects Models)

Because individual homologous genes map to multiple independent flanking CREs throughout the genome, we employ a Linear Mixed-Effects Model (LMM). Treating the gene pair as a random effect handles the nested data dependency and protects against degrees-of-freedom inflation.

```{r}
# Consolidate feature matrix across all annotation scales
df_integrated <- cres_to_align %>%
  select(human_enhancer, mouse_enhancer, cre_identity_pct, repeat_masked_pct) %>%
  left_join(genes_cres_human, by = join_by(human_enhancer == id_ccre)) %>%
  left_join(gene_level_cre_metrics, by = "hgnc_symbol") %>%
  mutate(gene_pair_id = paste0(hgnc_symbol, "_", mmusculus_homolog_associated_gene_name)) %>%
  filter(!is.na(rho))

# Fit Linear Mixed-Effects Model
lmm_fit <- lmer(
  rho ~ cre_identity_pct + pct_cres_conserved + repeat_masked_pct + (1 | gene_pair_id),
  data = df_integrated
)

# Extract clean model coefficients and variance components
tidy(lmm_fit)
```

Export

```{r export}
# Merge alignment results back into cres_annotated
cres_annotated <- cres_annotated %>%
  left_join(
    cres_to_align %>%
      dplyr::select(human_enhancer, mouse_enhancer,
                    sequence_human, sequence_mouse,
                    cre_identity_pct, cre_align_score, cre_align_length),
    by = c("human_enhancer", "mouse_enhancer")
  )

# Update split tables
# cres_homologous <- cres_annotated %>% filter(is_homologous)
# cres_human_only <- cres_annotated %>% filter(match_type == "human_only")
# cres_mouse_only <- cres_annotated %>% filter(match_type == "mouse_only")

# Save
saveRDS(cres_annotated,  here("tables", "cres_annotated_full.rds"))
# saveRDS(cres_homologous, here("tables", "cres_homologous_annotated.rds"))
# saveRDS(cres_human_only, here("tables", "cres_human_only_annotated.rds"))
# saveRDS(cres_mouse_only, here("tables", "cres_mouse_only_annotated.rds"))

message("All outputs saved to tables/")
```

