---
title: "5_EvolutionaryAnalysis"
output: html_document
---


# Evolutionary analysis
# Setup

```{r message=FALSE, warning=FALSE}
# Load core packages for the analysis session
library(here)
source(here("scripts_notebooks", "required.R"))
```
# Regulation

## Import data

```{r fig.height=5, fig.width=12}
# Import BTM core enrichment results
gsea_btm_results_core_joined =
  readRDS(file = here("tables", "all_human_mouse_gsea_btm_results_core_joined.rds")) %>%
  distinct()

#Import modules and LEGs
dge_btm_process_module_diff_and_legs = 
  readRDS(file = here("tables", "all_human_mouse_gsea_btm_legs.rds"))

# Import BTM gene-level correlation data
dge_btm_process_genes_diff_bygene =
  readRDS(here("tables", "human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune.rds"))

# Import homolog alignment data
all_alignments = read_csv(here("tables", "human_to_mouse_alignment_mart_export.txt"))

# Import BTM gene annotations
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

# Annotate genes with sequence identity and BTM membership
# Build alignment-annotated dataset for core enrichment gene

# Reference: all high-confidence orthologs in the database (unfiltered)
alignments_human2mouse_all = all_alignments %>%
  clean_names() %>%
  filter(mouse_orthology_confidence_0_low_1_high == 1) %>%
  drop_na(mouse_gene_name) %>%
  select(gene = mouse_gene_name, mouse_homology_type,
         identity_human2mouse = percent_id_query_gene_identical_to_target_mouse_gene,
         identity_mouse2human = percent_id_target_mouse_gene_identical_to_query_gene) %>%
  mutate(gene = str_to_upper(gene)) %>%
  # Resolve multiple identity scores by taking the median per gene
  group_by(gene) %>%
  summarise(identity_human2mouse = median(identity_human2mouse)) %>%
  ungroup()

# Filter to only BTM core enrichment genes and append process membership
alignments_human2mouse = alignments_human2mouse_all %>% 
  inner_join(dge_btm_process_module_diff_and_legs,
             by = join_by(gene == symbol))  %>% 
  mutate(status = if_else(status == "Shared", status, "Not-shared"))

#Group by process
protein_identity_process = alignments_human2mouse %>% 
  group_by(treatment, pathogen, timepoint_comparison, process, status) %>% 
  summarise(facet_label = dplyr::first(facet_label), group = dplyr::first(group),
            mean_identity = mean(identity_human2mouse, na.rm = T))


# Commented out: biomaRt gene info retrieval (pre-computed and imported below)
# human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
# gene_info_human = getBM(
#   attributes = c("hgnc_symbol", "gene_biotype", "go_id", "namespace_1003", "name_1006"),
#   mart = human_mart,
#   values = alignments_human2mouse_all$gene %>% unique(),
#   filters = "hgnc_symbol"
# ) %>%
#   rename(go_name = name_1006, go_domain = namespace_1003, gene = hgnc_symbol)
# gene_info_human %>% saveRDS(file = here("tables", "gene_info_human.rds"))

# Import pre-computed human gene information and join with alignment data
gene_info_human_log2fc_stats =
  readRDS(file = here("tables", "gene_info_human.rds")) %>%
  select(-hgnc_symbol) %>% 
  inner_join(alignments_human2mouse_all, by = "gene") %>% 
  full_join(alignments_human2mouse %>% 
              select(gene, process, pathogen, treatment, timepoint_comparison, group, status) %>% 
              distinct(),
          by = "gene") %>%
  
  #Add difference in log2FC
  inner_join(dge_btm_process_genes_diff_bygene %>% 
               filter(gene_set == "All genes") %>% 
               filter(pathogen_mouse != "DMD") %>% 
               select(pathogen = pathogen_human, timepoint_comparison, 
                      treatment = treatment_human, 
                      gene = human_symbol, mean_log2fc_human, 
                      mean_log2fc_mouse, mean_log2fc_diff, 
                      weight_se_inverse_combined) %>% 
               distinct(),
             by = join_by(gene, pathogen, timepoint_comparison, treatment)) 
#Save
gene_info_human_log2fc_stats %>% 
  saveRDS(file = here("tables", "gene_info_human_log2fc_stats.rds"))
```

#------------------------

## Cis-regulatory elements

```{r packages}
# library(BSgenome.Hsapiens.UCSC.hg38)
# library(BSgenome.Mmusculus.UCSC.mm10)
# library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(GenomeInfoDb)
library(GenomicRanges)
library(IRanges)
library(rtracklayer)
library(org.Hs.eg.db)
library(pwalign)
library(biomaRt)
```

### Import and merge data

```{r import-data}

#Import gene info with log2fc stats
gene_info_human_log2fc_stats = readRDS(file = here("tables", "gene_info_human_log2fc_stats.rds"))
#ENCODE ===============
#In contrast to the Human dataset, there are no CTCF-bound CREs in the main, full CRE mouse dataset. So, we imported CREs separately and merged them.

# Human CREs (GRCh38) =============

#PLS
cres_human_pls <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_PLS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_original_human = type_human,
         subtype_human = "not-CTCF") 

#pELS-dELS
cres_human_pelsdels <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_pELS-dELS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_original_human = type_human,
         subtype_human = "not-CTCF") 

#CTCF BOUND
cres_human_ctcfbound <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_CTCF-bound.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type_original"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_human = str_remove(type_original_human, ",CTCF-bound"),
         subtype_human = "CTCF")

# Load ENCODE comprehensive gene-CRE association table
genes_cres_human_original <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_Closest-Genes-PC.tsv"),
  col_names = c("chrom_cre", "start_cre", "end_cre", "id_dhs", "id_ccre", "type_cre",
                "chrom_gene", "start_gene", "end_gene", "transcript_id", "score_gene", "strand_gene", "gene_id"),
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) #2,651,787


#Check overlaps and remove CTCF-bound cCREs that are not annotated accordingly =====
#PLS
cres_human_ctcfbound_pls_overlap = cres_human_ctcfbound %>% 
  inner_join(cres_human_pls,
             by = "id_ccre_human") %>% 
  select(id_ccre_human) %>% 
  distinct() %>% 
  arrange(id_ccre_human) #36,430 IDs

#47,532 -> 11,102 IDs
cres_human_pls_clean = cres_human_pls %>% 
  filter(!id_ccre_human %in% cres_human_ctcfbound_pls_overlap$id_ccre_human) 

#pELS-dELS
cres_human_ctcfbound_pelsdels_overlap = cres_human_ctcfbound %>% 
  inner_join(cres_human_pelsdels,
             by = "id_ccre_human")%>% 
  select(id_ccre_human) %>% 
  distinct() #720,306 IDs

#1,718,669 -> 998,363
cres_human_pelsdels_clean = cres_human_pelsdels %>% 
  filter(!id_ccre_human %in% cres_human_ctcfbound_pelsdels_overlap$id_ccre_human)

#Merge
cres_human = bind_rows(cres_human_pls_clean,
          cres_human_pelsdels_clean,
          cres_human_ctcfbound)


#Annotate Human symbol ===========

library(biomaRt)
human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
listAttributes(human_mart) %>% view()

hgnc_symbol <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = unique(genes_cres_human_original$gene_id_clean),
  mart       = human_mart
)

# Join HGNC symbols to gene table
genes_cres_human <- genes_cres_human_original %>%
  left_join(hgnc_symbol, by = c("gene_id_clean" = "ensembl_gene_id")) %>% 
  rename(chrom_human = chrom_cre, start_human = start_cre, end_human = end_cre,
         id_ccre_human = id_ccre, id_dhs_human = id_dhs) %>% 
  #Add CRE subtype
  inner_join(cres_human %>% #1,958,107
               select(chrom_human, start_human, end_human, id_dhs_human, id_ccre_human, type_human, subtype_human),
             by = join_by(chrom_human, start_human, end_human, id_dhs_human, id_ccre_human)) %>% #2,214,457
  select(-type_cre)

genes_cres_human
```

Import Mouse CREs
```{r import-data}
# Mouse CREs (mm10) ============
#PLS
cres_mouse_pls <- read_tsv(
  here("tables", "Genomic", "genomic_mouse_cCREs_mm10_PLS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_original_mouse = type_mouse,
         subtype_mouse = "not-CTCF") 

#pELS-dELS
cres_mouse_pelsdels <- read_tsv(
  here("tables", "Genomic", "genomic_mouse_cCREs_mm10_pELS-dELS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_original_mouse = type_mouse,
         subtype_mouse = "not-CTCF")

#CTCF BOUND
cres_mouse_ctcfbound <- read_tsv(
  here("tables", "Genomic", "genomic_mouse_cCREs_mm10_CTCF-bound.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type_original"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_mouse = str_remove(type_original_mouse, ",CTCF-bound"),
         subtype_mouse = "CTCF")


#Check overlaps and remove CTCF-bound cCREs that are not annotated accordingly =====
#PLS
cres_mouse_ctcfbound_pls_overlap = cres_mouse_ctcfbound %>% 
  inner_join(cres_mouse_pls,
             by = "id_ccre_mouse") %>% 
  select(id_ccre_mouse) %>% 
  distinct() %>% 
  arrange(id_ccre_mouse) #11.884 IDs

#27.332 -> 15,448 IDs
cres_mouse_pls_clean = cres_mouse_pls %>% 
  filter(!id_ccre_mouse %in% cres_mouse_ctcfbound_pls_overlap$id_ccre_mouse) 

#pELS-dELS
cres_mouse_ctcfbound_pelsdels_overlap = cres_mouse_ctcfbound %>% 
  inner_join(cres_mouse_pelsdels,
             by = "id_ccre_mouse")%>% 
  select(id_ccre_mouse) %>% 
  distinct() #66.186 IDs

#512.001 -> 445,815 
cres_mouse_pelsdels_clean = cres_mouse_pelsdels %>% 
  filter(!id_ccre_mouse %in% cres_mouse_ctcfbound_pelsdels_overlap$id_ccre_mouse)

#Merge
cres_mouse = bind_rows(cres_mouse_pls_clean,
          cres_mouse_pelsdels_clean,
          cres_mouse_ctcfbound)

# Import homologous CREs ====================
# Homologous CRE pairs from ENCODE (hg38–mm10) Nearest gene, protein coding only
homologous_cres <- read_tsv(
  here("tables", "Genomic", "genomic_human_mouse_cCREs_homologous_GRCh38_Mm10.tsv"),
  col_names = c("human_enhancer", "mouse_enhancer"),
  show_col_types = FALSE
) #537,822

```


```{r import-data}
# GENEHANCER ===================

#Gene interactions with enhancers (ELITE)
genehancer_interactions_doubleelite = read_tsv(here("tables", "Genomic", "GeneHancer_Chr1-22_ChrX_Interactions_DoubleElite_hgTables.txt")) %>% 
  separate(name, into = c("hgnc_symbol", "genehancer_id")) %>%
  filter(chromEnd >= chromStart) %>% 
  rename(geneScore = score,
         geneValue = value,)
                                               
#Enhancers                                              
genehancer_enhancers_doubleelite = read_tsv(here("tables", "Genomic", "GeneHancer_Chr1-22_ChrX_Enhancers_DoubleElite_hgTables.txt")) %>%
  filter(chromEnd >= chromStart)


# Get overlap between ENCODE4 and Genehancer =========
library(GenomicRanges)
#Convert to GRanges
genehancer_enhancers_doubleelite_ranges = makeGRangesFromDataFrame(genehancer_enhancers_doubleelite, 
                                  keep.extra.columns = TRUE,
                                  seqnames.field = "chrom",
                                  start.field = "chromStart",
                                  end.field = "chromEnd", 
                                  na.rm = T)

#Convert to GRanges
cres_human_ranges = makeGRangesFromDataFrame(genes_cres_human, 
                                  keep.extra.columns = TRUE,
                                  seqnames.field = "chrom_human",
                                  start.field = "start_human",
                                  end.field = "end_human", 
                                  na.rm = T)
# Find overlaps
distances_ranges <- distanceToNearest(genehancer_enhancers_doubleelite_ranges, 
                     cres_human_ranges) %>% 
  as.data.frame()

#Annotate Genehancer ids (ELITE) with ENCODE4 ids
cres_human_encode_genehancer <- data.frame(
  genehancer_enhancers_doubleelite_ranges[distances_ranges$queryHits], 
  cres_human_ranges[distances_ranges$subjectHits],
  distance = distances_ranges$distance
) %>% 
  rename_with(
    ~ str_replace(., "\\.1$", "_encode4"), 
    ends_with(".1")                        
  ) %>% 
  rename(chrom = seqnames,
         chromStart = start,
         chromEnd = end,
         genehancer_id = name) %>%
  
  #Merge gene interaction scores
  full_join(genehancer_interactions_doubleelite,
             by = join_by("genehancer_id", "hgnc_symbol", "chrom", "chromStart", "chromEnd")
             )

#Check
cres_human_encode_genehancer %>% 
  glimpse()

#Save ===========
cres_human_encode_genehancer %>% 
  saveRDS(file = here("tables", "cres_human_encode_genehancer.rds"))

# 
# #Add more metadata (eliteness) =========
# cres_human_encode_genehancer_elite_notelite = genes_cres_human %>% 
#   full_join(cres_human_encode_genehancer %>% 
#              select(id_dhs_human, id_ccre_human, genehancer_id, evidenceSources, eliteness)  ,
#             join_by(id_dhs_human, id_ccre_human))
# #Save
# cres_human_encode_genehancer_elite_notelite %>% 
#   saveRDS(file = here("tables", "cres_human_encode_genehancer_elite_notelite.rds"))
```




### Summarize data 
Count how many mouse CREs each human CRE maps to (and vice versa)

```{r classify-match-type}
#Summarize classes
#Only elite enhancers
#Per gene
genes_cres_human %>% 
  filter(type_human %in% c("pELS", "dELS", "PLS")) %>% 
  group_by(hgnc_symbol, type_human, subtype_human) %>% 
  dplyr::summarise(n = n()) %>%
  arrange(hgnc_symbol) %>% 
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_human, desc(n))

cres_mouse %>% 
  dplyr::count(type_mouse, subtype_mouse) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_mouse, desc(n))


cres_human_encode_genehancer %>% 
  dplyr::count(type_human, subtype_human) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_human, desc(n))

cres_mouse %>% 
  dplyr::count(type_mouse, subtype_mouse) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_mouse, desc(n))

# Count how many mouse CREs each human CRE maps to (and vice versa)
human_counts <- homologous_cres %>%
  dplyr::count(human_enhancer, name = "n_mouse_per_human")

mouse_counts <- homologous_cres %>%
  dplyr::count(mouse_enhancer, name = "n_human_per_mouse")

# Full join: preserves orphan CREs on both sides
cres_annotated <- homologous_cres %>%
  left_join(human_counts, by = "human_enhancer") %>%
  left_join(mouse_counts, by = "mouse_enhancer") %>%
  full_join(genes_cres_human, by = join_by(human_enhancer == id_ccre_human)) %>%
  full_join(cres_mouse, by = join_by(mouse_enhancer == id_ccre_mouse)) %>%
  mutate(
    is_homologous = !is.na(human_enhancer) & !is.na(mouse_enhancer),
    match_homology = case_when(
      is.na(human_enhancer)                            ~ "mouse_only",
      is.na(mouse_enhancer)                            ~ "human_only",
      n_mouse_per_human == 1 & n_human_per_mouse == 1  ~ "1:1",
      n_mouse_per_human >  1 & n_human_per_mouse == 1  ~ "N:1",
      n_mouse_per_human == 1 & n_human_per_mouse >  1  ~ "1:N",
      TRUE                                             ~ "N:M"
    ),
    match_cretype = if_else(type_human == type_mouse, "Same", "Different"),
    match_cresubtype = if_else(subtype_human == subtype_mouse, "Same", "Different")
  ) 

#Save
cres_annotated %>% saveRDS(file = here("tables", "cres_homologous_annotated.rds"))


#Filter 1 to 1
cres_homologous_1to1 <- cres_annotated %>% #856,415
  dplyr::filter(match_type == "1:1") %>% #277,898
  #Filter enhancers with complete annotation in both species
  dplyr::filter(!is.na(chrom_human), !is.na(chrom_mouse))#1,567

# Separate tables
cres_homologous <- cres_annotated %>% dplyr::filter(is_homologous)
cres_human_only <- cres_annotated %>% dplyr::filter(match_type == "human_only")
cres_mouse_only <- cres_annotated %>% dplyr::filter(match_type == "mouse_only")

#Summarize
cres_homologous_1to1$type_human %>% unique()
#Compare classes of CREs between species
cres_homologous_1to1_comparison = cres_homologous_1to1 %>% 
  
  #Filter ELITE PLS, pELS, dELS
    filter(type_human %in% c("pELS", "dELS", "PLS"),
         type_mouse %in% c("pELS", "dELS", "PLS")) %>% 
  
  #Summarize
  group_by(type_human, subtype_human) %>% 
  dplyr::count(type_human, subtype_human, type_mouse, subtype_mouse) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(desc(n)) %>%
  mutate(type_human = fct_relevel(type_human, "PLS", "pELS", "dELS"),
         type_mouse = fct_relevel(type_mouse, "PLS", "pELS", "dELS")
         )

#Print
cres_homologous_1to1_comparison

#Save
cres_homologous_1to1_comparison %>% saveRDS(file = here("tables", "cres_homologous_1to1_annotated_aligned.rds"))


```



### Questions
#### Human: How many CREs per gene? And how many genes per enhancer?
```{r fig.height=5, fig.width=10}
#Import annotated CREs
cres_annotated = readRDS(file = here("tables", "cres_homologous_annotated.rds"))

#=======================================
#Compare classes of CREs between species
#======================================

# Calculate cCRE type, subtype, and homology percentages per locus

cres_type_homology_comparison <- cres_annotated %>%
  # Filter target cCRE types and valid gene symbols
  dplyr::filter(
    type_human %in% c("pELS", "dELS", "PLS"),
    type_mouse %in% c("pELS", "dELS", "PLS"),
    !is.na(hgnc_symbol) & hgnc_symbol != ""
  ) %>%
  # Collapse granular match_homology into broad simplified categories
  dplyr::mutate(
    match_homology_group = dplyr::case_when(
      match_homology == "1:1" ~ "1:1 Match",
      match_homology %in% c("1:N", "N:1", "N:M") ~ "Complex Homology",
      match_homology %in% c("human_only", "mouse_only") ~ "Species Specific",
      TRUE ~ "Other"
    ),
    match_homology_group = factor(
      match_homology_group, 
      levels = c("1:1 Match", "Complex Homology", "Species Specific")
    )
  ) %>%
  select(-match_homology) %>% 
  distinct() %>% 
  # Step 1: Count occurrences per gene locus and annotation/homology stratum
  dplyr::group_by(
    hgnc_symbol, type_human, subtype_human, type_mouse, subtype_mouse,
    match_homology_group, match_cretype, match_cresubtype
  ) %>%
  dplyr::summarise(n_cres = dplyr::n(), .groups = "drop") %>%
  
  # Step 2: Complete missing combinations with 0 to explicitly retain n = 0
  tidyr::complete(
    hgnc_symbol,
    type_human = c("pELS", "dELS", "PLS"),
    subtype_human,
    fill = list(n_cres = 0)
  ) %>%
  
  # Step 3: Compute relative percentage for each stratum over total locus cCREs
  dplyr::group_by(hgnc_symbol) %>%
  dplyr::mutate(
    n_total_cres_gene = sum(n_cres),
    pct_cres_gene     = dplyr::if_else(
      n_total_cres_gene > 0, 
      round(100 * n_cres / n_total_cres_gene, 2), 
      0
    )
  ) %>%
  
  # Step 4: Compute percentage relative to same Human Type
  dplyr::group_by(hgnc_symbol, type_human) %>%
  dplyr::mutate(
    n_total_cre_type_human = sum(n_cres),
    pct_same_cre_type      = dplyr::if_else(
      n_total_cre_type_human > 0,
      round(
        100 * sum(n_cres[match_cretype == "Same" | match_cretype == TRUE], na.rm = TRUE) / n_total_cre_type_human, 
        2
      ),
      0
    )
  ) %>%
  
  # Step 5: Compute percentage relative to same Human Type & Subtype
  dplyr::group_by(hgnc_symbol, type_human, subtype_human) %>%
  dplyr::mutate(
    n_total_cre_subtype_human = sum(n_cres),
    pct_same_cre_subtype      = dplyr::if_else(
      n_total_cre_subtype_human > 0,
      round(
        100 * sum(n_cres[match_cresubtype == "Same" | match_cresubtype == TRUE], na.rm = TRUE) / n_total_cre_subtype_human, 
        2
      ),
      0
    )
  ) %>%
  
  # Step 6: Compute percentage relative to Homology Class per Gene
  dplyr::group_by(hgnc_symbol, type_human, subtype_human, match_homology_group) %>%
  dplyr::mutate(
    
    n_total_homology = sum(n_cres),
    pct_homology     = dplyr::if_else(
      n_total_cres_gene > 0,
      round(100 * n_total_homology / n_total_cres_gene, 2),
      0
    )
  ) %>%
  
  # Step 7: Categorize cCRE counts into unified frequency bins (n_cat)
  dplyr::group_by(hgnc_symbol, type_human, subtype_human) %>%
  dplyr::mutate(
    n_cat = dplyr::case_when(
      n_cres == 0 ~ "0",
      n_cres == 1 ~ "1",
      n_cres == 2 ~ "2",
      n_cres == 3 ~ "3",
      n_cres == 4 ~ "4",
      n_cres >= 5 ~ ">5",
      TRUE        ~ NA_character_
    ),
    n_cat = factor(n_cat, levels = c("0", "1", "2", "3", "4", ">5"))
  ) %>%
  dplyr::ungroup() %>%
  
  # Step 8: Sort by locus and count, then set factor levels for plotting
  dplyr::arrange(hgnc_symbol, dplyr::desc(n_cres)) %>%
  dplyr::mutate(
    type_human = factor(type_human, levels = c("PLS", "pELS", "dELS")),
    type_mouse = factor(type_mouse, levels = c("PLS", "pELS", "dELS"))
  )

#Print
cres_type_homology_comparison

#Save 
cres_type_homology_comparison %>% 
  saveRDS(file = here("tables", "cres_type_homology_comparison.rds"))


# Inner join with cross-species differential expression statistics
cres_type_homology_comparison_dge_legs_stats <- cres_type_homology_comparison %>%
    dplyr::inner_join(
    gene_info_human_log2fc_stats %>%
      dplyr::select(
        treatment, pathogen, status,
        hgnc_symbol = gene, identity_human2mouse,
        mean_log2fc_human:weight_se_inverse_combined
      ) %>%
      dplyr::distinct() %>%
      dplyr::arrange(hgnc_symbol),
    by = dplyr::join_by(hgnc_symbol)
  )

#Print
cres_type_homology_comparison_dge_legs_stats

#Save
cres_type_homology_comparison_dge_legs_stats %>% 
  saveRDS(file = here("tables", "cres_type_homology_comparison_dge_legs_stats.rds"))

```

#### Have CREs diverged their functions?

```{r classify-match-type, fig.height=6, fig.width=8}


# Calculate shared cCRE type and subtype proportions across species ======
cres_type_matrix_pct <- cres_type_homology_comparison %>%
  
  # Step 1: Aggregate total cCRE counts per Human & Mouse type/subtype combination
  dplyr::group_by(type_human, subtype_human, type_mouse, subtype_mouse) %>%
  dplyr::summarise(n_pair_cres = sum(n_cres), .groups = "drop") %>%
  
  # Step 2: Compute relative percentage relative to total human cCREs in that stratum
  dplyr::group_by(type_human, subtype_human) %>%
  dplyr::mutate(
    n_human_stratum = sum(n_pair_cres),
    pct             = dplyr::if_else(
      n_human_stratum > 0, 
      round(100 * n_pair_cres / n_human_stratum, 1), 
      0
    )
  ) %>%
  dplyr::ungroup() %>%
  
  # Step 3: Order factor levels for heatmaps
  dplyr::mutate(
    type_human = factor(type_human, levels = c("PLS", "pELS", "dELS")),
    type_mouse = factor(type_mouse, levels = c("PLS", "pELS", "dELS"))
  )

#Heatmap
cres_type_matrix_pct_mouse_human_plot = cres_type_matrix_pct %>% 
  #Filter out CREs that do not exist in mice
  drop_na(type_mouse) %>% 
  #Plot
  ggplot() +
    aes(x = subtype_mouse %>% str_replace(., "-", "\n") %>% fct_rev(),
         y = subtype_human %>% str_replace(., "-", "\n")%>% fct_rev()) +
  facet_grid(~paste0("Mouse ", type_mouse) %>% fct_rev()~paste0("Human ", type_human) %>% fct_rev()) +
  # geom_tile(aes(fill = pct), color = "black", linewidth = 0.4) +
  geom_point(aes(fill = pct, size = n_pair_cres), 
             color = "black",
             pch = 21) +
  geom_text(aes(label = paste0(pct, "%")),
            size = 3.5,
            nudge_y = 0.4) +

  scale_fill_gradient2(low = "white",
                       high = "#4DBBD5FF",
                       guide = guide_colorbar(barwidth = unit(2, "cm"))) +
  scale_size_continuous(range = c(4, 12)) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0,
                                   # vjust = 0,
                                   hjust = 0.5),
        legend.position = "right"
        ) +
  labs(x = "Mouse",
       y = "Human",
       fill = "%Shared, from human",
       size = "CRE pairs") 
  # guides(size = guide_legend(
  #     nrow           = 3,
  #     byrow          = TRUE)) 

cres_type_matrix_pct_mouse_human_plot

cres_type_matrix_pct_mouse_human_plot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_regulation_cres_type_matrix_pct_plot.png"),
         width = 8, height = 6)

```




#### Distribution of CREs by gene, by LEGs
https://pubmed.ncbi.nlm.nih.gov/29180706/#&gid=article-figures&pid=figure-2-uid-1

```{r fig.height=3, fig.width=12}
#Plot
genes_cres_human_core_stats %>% 
  # filter(eliteness == "Elite") %>% 
  # filter(type_human == "PLS") %>% 
  ggplot() +
  aes(x = n_cat,
      y = abs(mean_log2fc_diff),
      color = status,
      fill =  status,
      ) +
  facet_nested_wrap(~fct_rev(type_human), 
                    nrow = 1, axes = "all", scales = "free") +
  geom_point(alpha = 0.2,
             size = 0.5,
            position = position_jitterdodge(jitter.width = 0.15, 
                                    dodge.width = 0.9),) +
  geom_boxplot(width = 0.7,
               alpha = 0.5,
               outliers = F) +
  # STATISTICAL TEST LAYER
  stat_compare_means(
    aes(group = status),
    method = "wilcox.test",
    label = "p.signif",
    symnum.args = list(
      cutpoints = c(0, 0.001, 0.01, 0.05, 1),
      symbols = c("***", "**", "*", "ns")
    ),
    vjust = 1.2,
    label.y.np = "top",
    hide.ns = T
  ) +
#   # Linear regression
#   stat_smooth(
#   aes(x = as.numeric(forcats::fct_drop(n_cat)), group = status),
#   method = "loess",
#   se = TRUE,
#   linewidth = 1.2,
#   alpha = 0.25
# ) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(x = "N",
       y = "|Log2FC|",
       color = "",
       fill = "") +
  theme_vaxgo() +
  theme(legend.position = "top")

```

#### Distribution of diffLog2FC, by CREs, by gene, by LEGs
```{r fig.height=8, fig.width=12}
#Plot
cres_type_homology_comparison_dge_legs_stats %>% 
  # filter(type_human == "PLS") %>%
  ggplot() +
  aes(x = n_cat,
      y = mean_log2fc_human,
      # label = hgnc_symbol,
      fill =  status,
      color = status
      ) +
  facet_nested_wrap(~fct_rev(type_human)+ 
                      # subtype_human + 
                      treatment+ pathogen, nrow = 3,
                    scales = "free") +
  geom_hline(yintercept = 0, color = "black", linetype = 2) +
    geom_point(position = position_jitterdodge(jitter.width = 0.2, 
                                    dodge.width = 0.9),
               mapping = aes(color = status),
               size = 0.5,
               alpha = 0.5) +
  geom_violin(alpha = 0.1, ) +
    # geom_boxplot(width = 0.7,
    #            alpha = 0.5,
    #            outliers = F) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(x = "N",
       y = "Human Log2FC",
       fill = "",
       color = "") +
  theme_vaxgo() +
  theme(legend.position = "top")

```

What is the distribution of absolute diffLog2FC, by CREs, by gene, by LEGs
https://pubmed.ncbi.nlm.nih.gov/29180706/#&gid=article-figures&pid=figure-2-uid-1
```{r fig.height=8, fig.width=12}
#Plot
cres_type_homology_comparison_dge_legs_stats %>% 
  # filter(type_human == "PLS") %>% 
  ggplot() +
  aes(x = n_cat,
      y = abs(mean_log2fc_diff),
      color = status,
      fill =  status,
      ) +
  facet_nested_wrap(~fct_rev(type_human)+treatment+pathogen, 
                    nrow = 3, axes = "all", scales = "free") +
  geom_point(alpha = 0.2,
             size = 0.5,
            position = position_jitterdodge(jitter.width = 0.15, 
                                    dodge.width = 0.5),) +
  geom_boxplot(width = 0.7,
               alpha = 0.5,
               outliers = F) +
  # STATISTICAL TEST LAYER
  stat_compare_means(
    aes(group = status),
    method = "wilcox.test",
    label = "p.signif",
    symnum.args = list(
      cutpoints = c(0, 0.001, 0.01, 0.05, 1),
      symbols = c("***", "**", "*", "ns")
    ),
    label.y.np = "top",
    vjust = 1.2,        
    hide.ns = T
  ) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(x = "N",
       y = expression("|" * Delta * "Log"[2] * "FC|"),
       fill = "",
       color = "") +
  theme_vaxgo() +
  theme(legend.position = "top")

```


#### nCREs vs. Identity
```{r fig.height=6, fig.width=8}
cres_type_homology_comparison_dge_legs_stats %>% 
  ggplot(aes(
    x = abs(mean_log2fc_diff),
    y = identity_human2mouse,
    color = status,
    fill = status
  )) +
  # Sub-divide facets by both cCRE class and n_cat bin
  facet_nested_wrap(
    ~ fct_rev(type_human) + n_cat, 
    axes = "all", 
    scales = "free_x",
    ncol = 6
  ) +
  geom_point(
    alpha = 0.25,
    size = 0.6
  ) +
  # Linear regression line calculated per status inside each n_cat bin
  stat_smooth(
    method = "lm",
    se = TRUE,          # Shading for 95% Confidence Interval
    linewidth = 1,
    alpha = 0.2
  ) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(
    x = expression("|" * Delta * "Log"[2] * "FC|"),
    y = "% Protein Identity",
    fill = "",
    color = ""
  ) +
  theme_vaxgo() +
  theme(legend.position = "top")
```






#### Regression
```{r}
# Load required core libraries
# --------
library(tidyverse)
library(tidymodels)
library(here)

# --------
# Prepare master wide features dataset per gene locus
# --------
genes_features_master <- cres_type_homology_comparison_dge_legs_stats %>%
  # Filter missing identities to ensure clean modeling data
  drop_na(identity_human2mouse, status) %>%
  select(
         hgnc_symbol, treatment, pathogen, status, identity_human2mouse, mean_log2fc_human, mean_log2fc_mouse, 
         type_human, subtype_human, match_homology_group, 
         n_cres, n_cat, pct_same_cre_type, pct_same_cre_subtype, pct_homology) %>% 
  distinct() %>% 
  
  # Step 1: Pivot all required features (counts and percentages) to wide format
  group_by(hgnc_symbol, treatment, pathogen, status, identity_human2mouse) %>% 
  tidyr::pivot_wider(
    id_cols     = c(hgnc_symbol, treatment, pathogen, status, identity_human2mouse, mean_log2fc_human, mean_log2fc_mouse),
    names_from  = c(type_human, subtype_human, match_homology),
    values_from = c(n_cres, n_cat, pct_same_cre_type, pct_same_cre_subtype, pct_homology),
    names_sep   = "_",
    # Corrected values_fill syntax passing explicit default scalar types
    # values_fill = 0
    ) %>%
  janitor::clean_names() %>%
  
  # Step 2: Define Target Variables for Logistic and Linear Regression
  dplyr::mutate(
    # Target for Logistic Regression (1 = Shared, 0 = Not Shared)
    status_bin = factor(dplyr::if_else(status == "Shared", "Shared", "Not_Shared"), levels = c("Shared", "Not_Shared")),
    
    # Target for Linear Regression (Absolute Log2FC divergence magnitude)
    abs_log2fc_diff = abs(mean_log2fc_human - mean_log2fc_mouse)
  )




cres_type_homology_comparison_dge_legs_stats %>%
  # Filter missing identities to ensure clean modeling data
  drop_na(identity_human2mouse, status) %>%
  filter(hgnc_symbol == "GNAI2") %>% 
  select(
         hgnc_symbol, treatment, pathogen, status, 
         type_human, subtype_human, match_homology_group, identity_human2mouse, mean_log2fc_human, mean_log2fc_mouse, mean_log2fc_diff,
         n_cres, n_cat, pct_same_cre_type, pct_same_cre_subtype, pct_homology) %>% 
  distinct() %>% 
  
  # Step 1: Pivot all required features (counts and percentages) to wide format
  tidyr::pivot_wider(
    id_cols     = c(hgnc_symbol, treatment, pathogen, status, identity_human2mouse, mean_log2fc_human, mean_log2fc_mouse, mean_log2fc_diff),
    names_from  = c(type_human, subtype_human, match_homology_group),
    values_from = c(n_cres, 
                    # n_cat, 
                    pct_same_cre_type, pct_same_cre_subtype, pct_homology),
    names_sep   = "_", 
    values_fn = median
    ) %>% 
  # Step 2: Define Target Variables for Logistic and Linear Regression
  dplyr::mutate(
    # Target for Logistic Regression (1 = Shared, 0 = Not Shared)
    status_bin = factor(dplyr::if_else(status == "Shared", "Shared", "Not_Shared"), levels = c("Shared", "Not_Shared")),
    
    # Target for Linear Regression (Absolute Log2FC divergence magnitude)
    abs_log2fc_diff = abs(mean_log2fc_diff)
  )

```


```{r}
# ==============================================================================
# PIPELINE 1: LOGISTIC REGRESSION (Classify Shared vs Not-Shared LEGs)
# ==============================================================================

# --------
# Configure logistic regression specification and recipe
# --------
logistic_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

logistic_recipe <- recipes::recipe(status_bin ~ ., data = genes_features_master) %>%
  # Preserve metadata columns by assigning them to an "id" role (they won't be used as predictors)
  recipes::update_role(hgnc_symbol, treatment, pathogen, status, mean_log2fc_human, mean_log2fc_mouse, abs_log2fc_diff, new_role = "id") %>%
  # Convert n_cat categorical variables into dummy binaries for glm
  recipes::step_dummy(recipes::all_nominal_predictors()) %>%
  # Remove zero-variance predictors
  recipes::step_zv(recipes::all_predictors()) %>%
  # Scale and center all numeric predictor metrics (percentages and raw counts)
  recipes::step_normalize(recipes::all_numeric_predictors())

# --------
# Fit logistic workflow and extract engine
# --------
logistic_fit <- tidymodels::workflow() %>%
  tidymodels::add_recipe(logistic_recipe) %>%
  tidymodels::add_model(logistic_spec) %>%
  tidymodels::fit(data = genes_features_master)

# Extract native GLM object for direct diagnostics
logistic_glm_engine <- tidymodels::extract_fit_engine(logistic_fit)

# -> Example direct analysis via pROC
genes_features_master$prob_shared <- predict(logistic_glm_engine, type = "response")
roc_obj <- pROC::roc(genes_features_master$status_bin, genes_features_master$prob_shared, levels = c("Not_Shared", "Shared"))
cat("Logistic AUC:", round(pROC::auc(roc_obj), 3), "\n")

# -> Example direct analysis via broom
logistic_or_table <- broom::tidy(logistic_glm_engine, exponentiate = TRUE, conf.int = TRUE)
```


```{r}
# ==============================================================================
# PIPELINE 2: LINEAR REGRESSION (Predict Log2FC Divergence Magnitude)
# ==============================================================================

# --------
# Configure linear regression specification and recipe
# --------
linear_spec <- tidymodels::linear_reg() %>%
  tidymodels::set_engine("lm") %>%
  tidymodels::set_mode("regression")

linear_recipe <- recipes::recipe(abs_log2fc_diff ~ ., data = genes_features_master) %>%
  recipes::update_role(hgnc_symbol, treatment, pathogen, status, status_bin, prob_shared, mean_log2fc_human, mean_log2fc_mouse, new_role = "id") %>%
  recipes::step_dummy(recipes::all_nominal_predictors()) %>%
  recipes::step_zv(recipes::all_predictors()) %>%
  recipes::step_normalize(recipes::all_numeric_predictors())

# --------
# Fit linear workflow and extract engine
# --------
linear_fit <- tidymodels::workflow() %>%
  tidymodels::add_recipe(linear_recipe) %>%
  tidymodels::add_model(linear_spec) %>%
  tidymodels::fit(data = genes_features_master)

# Extract native LM object for direct diagnostics
linear_lm_engine <- tidymodels::extract_fit_engine(linear_fit)

# -> Example direct analysis via broom
linear_coefficients <- broom::tidy(linear_lm_engine, conf.int = TRUE)


# --------
# Save model workflows and extracted engines to disk
# --------
# Save
saveRDS(logistic_glm_engine, file = here::here("tables", "logistic_glm_engine.rds"))
saveRDS(linear_lm_engine, file = here::here("tables", "linear_lm_engine.rds"))
```
















------------------------------------------------------------------------

### Gene annotation

```{r import-gene-annotation}

###############
# ADAPT
###############

# Import gene expression correlation
gene_info_human_log2fc_stats = readRDS(file = here("tables", "gene_info_human_log2fc_stats.rds"))
############## 
# Query biomaRt to map human Ensembl IDs to HGNC and MGI symbols
library(biomaRt)
human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

hgnc_mgi_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = unique(cres_human_encode_genehancer$gene_id_clean),
  mart       = human_mart
)

# Join HGNC symbols to gene table
genes_cres_human <- genes_cres_human %>%
  left_join(hgnc_mgi_map, by = c("gene_id_clean" = "ensembl_gene_id"))


# Compute multi-CRE metrics per homologous gene pair
gene_level_cre_metrics <- genes_cres_human %>%
  group_by(hgnc_symbol, type_cre) %>%
  summarise(
    n_cres_human       = n(),
    n_cres_orthologous = sum(id_ccre %in% cres_homologous_1to1$human_enhancer),
    pct_cres_conserved = n_cres_orthologous / n_cres_human,
    .groups = "drop"
  ) 
  # inner_join(all_human_mouse_dge_btm_process_genes_correlation_bygene, by = c("hgnc_symbol" = "hgnc_symbol"))


# Annotate cres_annotated with human gene info (one gene per CRE)
cres_annotated <- cres_annotated %>%
  left_join(
    genes_cres_human %>%
      dplyr::select(id_ccre, gene_id, hgnc_symbol,                     transcript_id, chrom_gene, start_gene, end_gene, strand_gene) %>%
      distinct(id_ccre, .keep_all = TRUE),
    by = c("human_enhancer" = "id_ccre")
  )

```



(DOES NOT WORK) Get evolutionary distance

```{r}
# library(biomaRt)
# 
# # 1. Definir o banco de dados antigo usando o host do archive desejado (ex: jan2020)
# # Como você quer partir do camundongo para buscar os homólogos humanos, usamos "mmusculus_gene_ensembl"
# listEnsemblArchives()
# listMarts()
# 
# options(timeout = 600)
# 
# # ensembl_human_old = useEnsembl(biomart = "ensembl",
# #                        dataset="hsapiens_gene_ensembl",
# #                        version="77")
# 
# ensembl_human_old = useMart(biomart = "ENSEMBL_MART_ENSEMBL",
#                              dataset="hsapiens_gene_ensembl",
#                             host = "https://oct2014.archive.ensembl.org")
# 
# listFilters(ensembl_human_old) %>% View()
# listAttributes(ensembl_human_old) %>% View()
# 
# # 2. Carregar sua lista de IDs de genes de camundongo (ex: ENSMUSG0000XXXXXXXX)
# human_genes_list <- all_human_mouse_dge_btm_process_genes_correlation_bygene$symbol
# 
# # 3. Fazer a busca trazendo o ID do camundongo e os dados do homólogo humano (hsapiens)
# options(timeout = 600)
# human_mouse_orthologs_old <- getBM(attributes = c(
#                                                   "ensembl_gene_id",
#                                                   "external_gene_name"
#                                           # "mmusculus_homolog_dn", 
#                                           # "mmusculus_homolog_ds"
#                                           ), 
#                             filters = 'external_gene_name', 
#                             values = "CD6", 
                            # mart = ensembl_human_old)

```










------------------------------------------------------------------------

###  Sequence extraction

The HOXD55 matrix (originally derived by Chiaromonte et al., 2002, and deeply analyzed in Blanchette et al., 2004) defines nucleotide matches and mismatches using large integers to maximize sensitivity when calculating neutral substitution rates in mammalian alignments:Match: $+91$Mismatch: $-114$Because a single mismatch penalizes the score by $-114$, the gap penalties must be scaled proportionally. If you used a standard gap opening penalty of $-10$, the alignment algorithm would erroneously introduce hundreds of micro-gaps instead of accepting a single nucleotide mismatch, completely ruining the biological alignment of your cis-regulatory elements (CREs).

To maintain perfect parity with the alignments you see on the UCSC Genome Browser for human (hg38) and mouse (mm10), you use their default parameters:Gap Opening Penalty ($O$): $-400$Gap Extension Penalty ($E$): $-30$

Chiaromonte, F., et al. (2002). Scoring pairwise genomic alignments. Pacific Symposium on Biocomputing. This paper established the ancestral mammalian repeat substitution matrices (including the HOXD matrices). Blanchette, M., et al. (2004). Blastz: A tool for aligning a large number of long genomic sequences. Genome Research. This paper formally documents the open-source BLASTZ/LASTZ defaults ($O = 400, E = 30$, match $= 91$, mismatch $= 114$) used by UCSC. Kent, W. J., et al. (2003). Evolution's Cauldron: Duplication, Deletion, and Rearrangement in the Mouse and Human Genomes. PNAS. Documents how UCSC constructs the chain and net alignments utilizing these exact parameters.

Align homolog enhancer sequences

```{r extract-sequences}
#Filter homologs with mapped coordinates
cres_to_align_homologous <- cres_homologous_1to1 %>%
  filter(human_enhancer %in% genes_cres_human$id_ccre) %>%
  filter(!is.na(chrom_human), !is.na(chrom_mouse)) %>%
  #Annotate gene symbol
  inner_join(cres_annotated %>% 
            select(human_enhancer, mouse_enhancer, hgnc_symbol),
        by = join_by(human_enhancer, mouse_enhancer))

#Convert to GRanges objects
gr_human <- GRanges(seqnames = cres_to_align_homologous$chrom_human, ranges = IRanges(start = cres_to_align_homologous$start_human, end = cres_to_align_homologous$end_human))
gr_mouse <- GRanges(seqnames = cres_to_align_homologous$chrom_mouse, ranges = IRanges(start = cres_to_align_homologous$start_mouse, end = cres_to_align_homologous$end_mouse))

#Get sequences
cres_to_align_homologous$sequence_human <- as.character(getSeq(BSgenome.Hsapiens.UCSC.hg38, gr_human))
cres_to_align_homologous$sequence_mouse <- as.character(getSeq(BSgenome.Mmusculus.UCSC.mm10, gr_mouse))

#Clean sequences from undefined characters
cres_to_align_homologous_clean = cres_to_align_homologous %>% 
  mutate(sequence_human = str_remove_all(sequence_human, "N"),
         sequence_mouse = str_remove_all(sequence_mouse, "N"))

# Parameterize a nucleotide substitution matrix mirroring the standard HOXD55 values
matriz_hoxd55 <- nucleotideSubstitutionMatrix(match = 91, mismatch = -114, baseOnly = TRUE)

#Align sequences
aln_cres <- pairwiseAlignment(
  DNAStringSet(cres_to_align_homologous_clean$sequence_human),
  DNAStringSet(cres_to_align_homologous_clean$sequence_mouse),
  type         = "local",
  substitutionMatrix = matriz_hoxd55,
  gapOpening   = -400, # Score scale adjusted for standard UCSC genomic alignment scoring
  gapExtension = -30
)

#Calculate percent identity
cres_to_align_homologous_identity <- cres_to_align_homologous_clean %>%
  mutate(
    cre_identity_pct = pid(aln_cres, type = "PID1"),
    cre_align_score  = score(aln_cres),
    cre_align_length = nchar(alignedPattern(aln_cres)),
    cre_coverage_human = cre_align_length / nchar(sequence_human)
  )



#Save results
cres_to_align_homologous_identity %>% 
  saveRDS(file = here("tables", "cres_homologous_1to1_annotated_aligned.rds"))

#Summarize alignment results
summary(cres_to_align$cre_identity_pct)

cat("\n% identity > 80%:", round(mean(cres_to_align$cre_identity_pct > 80) * 100, 1), "%\n")
cat("% identity > 70%:", round(mean(cres_to_align$cre_identity_pct > 70) * 100, 1), "%\n")
cat("% identity > 60%:", round(mean(cres_to_align$cre_identity_pct > 60) * 100, 1), "%\n")

#Which genes have the most conserved enhancers?
cres_to_align %>% 
  select(hgnc_symbol, human_enhancer, mouse_enhancer, type_human, type_mouse, cre_identity_pct:cre_coverage_human) %>% 
  group_by(hgnc_symbol, type_human, type_mouse) %>% 
  summarise(median_pct_id = median(cre_identity_pct)) %>%
  arrange(-median_pct_id) 

```

###  Alignment of proximal TSS windows

```{r}
tx_human <- transcripts(TxDb.Hsapiens.UCSC.hg38.knownGene)

# Multi-scale biological window definitions centered on the TSS
windows_definition <- list(
  w3000 = list(up = 3000, down = 300),  # Full proximal regulatory landscape (pELS/dELS capture)
  w1500 = list(up = 1500, down = 300),  # Core promoter + immediate proximal enhancers
  w500  = list(up = 500,  down = 100),  # Core promoter locus only
  w200  = list(up = 200,  down = 50)    # Minimal transcription initiation footprint
)

# Exemplary extraction and global alignment (Needleman-Wunsch) loop initialization for w3000
tss_human_w3000 <- promoters(TxDb.Hsapiens.UCSC.hg38.knownGene, upstream = 3000, downstream = 300)
tss_mouse_w3000 <- promoters(TxDb.Mmusculus.UCSC.mm10.knownGene, upstream = 3000, downstream = 300)

# Implementation Note: Filter your GRanges objects by matched homologous pairs prior to running global alignment:
aln_global <- pairwiseAlignment(seq_human_gene, seq_mouse_gene, type = "global", substitutionMatrix = matriz_hoxd55)
```

###  Annotate conservation and TEs

```{r}
ah <- AnnotationHub()

# 1. Fetch the most recent RepeatMasker records for hg38
rmsk_query <- query(ah, c("RepeatMasker", "Homo sapiens", "hg38"))
rmsk_hg38  <- ah[[names(rmsk_query)[1]]]

# 2. Compute the exact fraction of the CRE masked by repetitive components
gr_cres_human_all <- GRanges(
  seqnames = cres_to_align$chrom_human, 
  ranges = IRanges(start = cres_to_align$start_human, end = cres_to_align$end_human)
)

hits_rmsk <- findOverlaps(gr_cres_human_all, rmsk_hg38)

# Calculate coverage to map base-pair level masking within the ranges
repeat_cov <- GenomicRanges::coverage(rmsk_hg38)
seqlevels(gr_cres_human_all, pruning.mode = "coercion") <- names(repeat_cov)
masked_views <- Views(repeat_cov, ranges(gr_cres_human_all))

cres_to_align$repeat_masked_pct <- sapply(masked_views, function(x) sum(as.numeric(x) > 0)) / width(gr_cres_human_all) * 100
```

### Integrative Statistical Modeling (Linear Mixed-Effects Models)

Because individual homologous genes map to multiple independent flanking CREs throughout the genome, we employ a Linear Mixed-Effects Model (LMM). Treating the gene pair as a random effect handles the nested data dependency and protects against degrees-of-freedom inflation.

```{r}
# Consolidate feature matrix across all annotation scales
df_integrated <- cres_to_align %>%
  select(human_enhancer, mouse_enhancer, cre_identity_pct, repeat_masked_pct) %>%
  left_join(genes_cres_human, by = join_by(human_enhancer == id_ccre)) %>%
  left_join(gene_level_cre_metrics, by = "hgnc_symbol") %>%
  mutate(gene_pair_id = paste0(hgnc_symbol, "_", mmusculus_homolog_associated_gene_name)) %>%
  filter(!is.na(rho))

# Fit Linear Mixed-Effects Model
lmm_fit <- lmer(
  rho ~ cre_identity_pct + pct_cres_conserved + repeat_masked_pct + (1 | gene_pair_id),
  data = df_integrated
)

# Extract clean model coefficients and variance components
tidy(lmm_fit)
```

Export

```{r export}
# Merge alignment results back into cres_annotated
cres_annotated <- cres_annotated %>%
  left_join(
    cres_to_align %>%
      dplyr::select(human_enhancer, mouse_enhancer,
                    sequence_human, sequence_mouse,
                    cre_identity_pct, cre_align_score, cre_align_length),
    by = c("human_enhancer", "mouse_enhancer")
  )

# Update split tables
# cres_homologous <- cres_annotated %>% filter(is_homologous)
# cres_human_only <- cres_annotated %>% filter(match_type == "human_only")
# cres_mouse_only <- cres_annotated %>% filter(match_type == "mouse_only")

# Save
saveRDS(cres_annotated,  here("tables", "cres_annotated_full.rds"))
# saveRDS(cres_homologous, here("tables", "cres_homologous_annotated.rds"))
# saveRDS(cres_human_only, here("tables", "cres_human_only_annotated.rds"))
# saveRDS(cres_mouse_only, here("tables", "cres_mouse_only_annotated.rds"))

message("All outputs saved to tables/")
```











## Transcription factors
Annotate with gene regulation - TF Identify conserved Transcription
Factors (TFs) regulating specific genes in both Humans and Mice

```{r}
# Clean and parse the ChEA3 transcription factor–target gene matrix
gene_attribute_chea <- read_tsv(here("tables", "gene_attribute_matrix.txt"))

# Transform the wide ChEA matrix into long format, keeping only positive scores
gene_attribute_chea_clean <- gene_attribute_chea %>%
  rename(gene = Gene) %>%
  pivot_longer(-gene, names_to = "experiment", values_to = "score") %>%
  filter(score > 0) %>%
  arrange(desc(score)) %>%
  # Split the experiment metadata into TF name, PMID, and method/tissue/species
  separate(experiment,
           into = c("tf", "pmid", "method_tissue_species"),
           sep = "-",
           extra = "merge",
           fill = "right") %>%
  # Detect species from the metadata string (HUMAN or MOUSE)
  mutate(across(where(is.character), str_trim),
         species = case_when(str_detect(method_tissue_species, "HUMAN") ~ "Human",
                              str_detect(method_tissue_species, "MOUSE") ~ "Mouse",
                              TRUE ~ method_tissue_species)) %>%
  filter(species %in% c("Human", "Mouse"))

# Count evidence per (gene, TF) pair across species
# For each TF–target pair, count how many PMIDs support it in Human and in Mouse
gene_attribute_tf_count <- gene_attribute_chea_clean %>%
  ungroup() %>%
  select(gene, tf, pmid, species) %>%
  group_by(gene, tf) %>%
  pivot_wider(names_from = "species",
              values_from = "pmid") %>%
  mutate(
    Human_pmids = map_chr(Human, ~ paste(.x, collapse = ", ")),
    Mouse_pmids = map_chr(Mouse, ~ paste(.x, collapse = ", ")),
    Human_pmids_n = map_int(Human, length),
    Mouse_pmids_n = map_int(Mouse, length)
  ) %>%
  select(!c(Human, Mouse)) %>%
  mutate(across(ends_with("pmids"), ~ na_if(.x, ""))) %>%
  # Flag whether this TF regulates this gene in both organisms
  mutate(both = if_else(!is.na(Human_pmids) & !is.na(Mouse_pmids), "TRUE", "FALSE"))

# Save the cross-species TF-target count for downstream analysis
saveRDS(gene_attribute_tf_count,
        file = here("tables", "regulation_gene_tf_count.rds"))


gene_attribute_tf_count

```

```{r}
# Reload the TF-count data and gene DGE stats
gene_attribute_tf_count <- readRDS(file = here("tables", "regulation_gene_tf_count.rds"))
gene_info_human_log2fc_stats = readRDS(file = here("tables", "gene_info_human_log2fc_stats.rds"))

# Build a contingency table of cross-species TF regulation for a few selected genes
# Join with the filtered gene quadrants to show how many TFs are shared vs species-specific
gene_attribute_tf_count %>%
  inner_join(gene_info_human_log2fc_stats,
             by = join_by(gene)) %>% 
  filter(gene %in% c("XPO1", "CD2", "CDK14", "NDUFB10")) %>%
  mutate(gene_info = paste(gene, status, mean_log2fc_diff)) %>%
  arrange(desc(both)) %>%
  select(gene, pathogen, treatment, both) %>% 
  distinct() %>% 
  tabyl(gene, both) %>%
  adorn_totals(where = "col") %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns() %>%
  as.data.frame()


```

#### Calculate Shared TFs

```{r fig.height=3, fig.width=5}
# Join the TF regulation table with the gene alignment quadrants
gene_attribute_tf_count_log2fc_stats <- gene_attribute_tf_count %>%
  inner_join(gene_info_human_log2fc_stats,
             by = join_by(gene))

# Summarise TF counts per gene, split by cross-species status (both vs species-specific)
gene_attribute_tf_count_df_summary = gene_attribute_tf_count_log2fc_stats %>%
  select(gene, both, pathogen, treatment) %>%
  distinct() %>% 
  arrange(desc(both)) %>%
  group_by(treatment, pathogen, gene, both) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(treatment, pathogen, gene) %>%
  mutate(total = sum(n),
         percentage = round(n / total * 100, 1))

# Compute unique TF counts per gene for each species
gene_attribute_tf_count_test <- gene_attribute_tf_count %>%
  select(gene, tf, Human_pmids_n, Mouse_pmids_n) %>%
  mutate(Mouse_TF = if_else(Mouse_pmids_n > 0, TRUE, FALSE),
         Human_TF = if_else(Human_pmids_n > 0, TRUE, FALSE)) %>%
  group_by(gene) %>%
  summarise(unique_tfs_human = n_distinct(tf[Human_TF == TRUE]),
            unique_tfs_mouse = n_distinct(tf[Mouse_TF == TRUE]))

gene_attribute_tf_count_df_summary %>%
  filter(both == "TRUE") %>%
  ggplot() +
  aes(y = percentage,
      x = pathogen,
      fill = pathogen) +
  geom_boxplot() +
  geom_point(alpha = 0.2) +
    stat_summary(geom = "pointrange",
               fun.data = "mean_cl_boot",
               position = "dodge",
               color = "red") +
  theme_vaxgo() +
  theme(legend.position = "none") +
  scale_fill_manual(values = colors$pathogen)

gene_attribute_tf_count_filter_df_summary_annotated = gene_attribute_tf_count_df_summary %>%
  arrange(gene) %>% 
  distinct() %>%
    select(-percentage) %>% 
  pivot_wider(values_from = "n",
              names_from = "both") %>% 
  distinct() %>%
  rename(not_shared_tf = "F",
         shared_tf = "T") %>% 
  mutate(shared_tf = if_else(is.na(shared_tf), 0, shared_tf)) %>% 
  inner_join(gene_attribute_tf_count_log2fc_stats %>% 
               ungroup() %>% 
               filter(class %in% c("Shared", "Not-shared")) %>% 
               select(gene, class:cor_label) %>% 
               distinct(),
             by = join_by(gene)) %>% 
  ungroup() %>% 
  mutate(shared_tf_pct = (shared_tf/total) %>% round(2)) %>% 
    mutate(logp = -log(p_value),
         cor_value_abs = abs(cor_value)) %>% 
  #Highlight genes
  mutate(classification = "All") %>% 
  inner_join(gene_attribute_tf_count_test,
             by = "gene")
  
#Highlight genes
shared_genes = gene_attribute_tf_count_filter_df_summary_annotated %>% 
                    filter(gene %in% c(top_shared_notshared_genes %>% 
                                 filter(class == "Shared") %>% 
                                   pull(gene) %>% unique())) %>% 
              mutate(classification = "Shared") 

gene_attribute_tf_count_filter_df_summary_annotated = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  bind_rows(shared_genes) 

gene_attribute_tf_count_filter_df_summary_annotated %>% 
  dplyr::count(classification, class)
  
```

#### Plot Total of TFs by species

Scatterplot of total unique TFs per gene in Human vs Mouse. Points are
coloured by classification (Shared vs All). A diagonal line indicates
equal TF counts between species.

```{r fig.height=5, fig.width=7}
# Compare total unique TF counts per gene between human and mouse
regulation_tfs_total_by_species_plot <- gene_attribute_tf_count_filter_df_summary_annotated %>%
  drop_na(cor_value) %>%
  ggscatter(x = "unique_tfs_mouse",
            y = "unique_tfs_human",
            fill = "classification",
            color = "classification",
            point = FALSE,
            add = "reg.line",
            conf.int = TRUE) +
    scale_x_continuous(limits = c(0, 80)) +
  scale_y_continuous(limits = c(0, 80)) +
  geom_abline(linetype = 2,
              alpha = 0.5) +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "Human (Total TFs)",
       y = "Mouse (Total TFs)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value,
                 label = gene),
             alpha = 0.8,
             pch = 21,
             stroke = 0,
             size = 2) +
    geom_point(aes(fill = cor_value),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(80, 75), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_total_by_species_plot
  
```

```{r fig.height=3, fig.width=5}
#Identity
regulation_tfs_total_identity_plot =  
  gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  ggscatter(x = "total",
            y = "identity_human2mouse", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line", 
            conf.int = T
              ) +
  scale_color_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "Total TFs",
       y = "%Identity (a.a.)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value),
             alpha = 0.2,
             pch = 21,
             stroke = 0) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(120, 115), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_total_identity_plot 

#Identity
regulation_tfs_shared_pct_identity_plot =  
  gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  ggscatter(x = "shared_tf_pct",
            y = "identity_human2mouse", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line", 
            conf.int = T
              ) +
  scale_color_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "%Shared TFs",
       y = "%Identity (a.a.)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value),
             alpha = 0.2,
             pch = 21,
             stroke = 0) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(120, 115), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_shared_pct_identity_plot 

regulation_tfs_identity_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  ggscatter(x = "shared_tf",
            y = "identity_human2mouse", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line", 
            conf.int = T
              ) +
  scale_color_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "Total Shared TFs",
       y = "%Identity (a.a.)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value),
             alpha = 0.2,
             pch = 21,
             stroke = 0) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(120, 115), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_total_identity_plot 
regulation_tfs_identity_plot
```

```{r fig.height=5, fig.width=5}
#Correlation
regulation_tfs_corvalue_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  mutate(quadrant = if_else(cor_value>0, "Positive correlation", "Negative correlation")) %>% 
  drop_na(quadrant) %>% 
  mutate(logp = -log(p_value),
         cor_value_abs = abs(cor_value)) %>% 
  ggscatter(x = "cor_value_abs",
            y = "total", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line",
            # conf.int = TRUE
              ) +
    scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "|Correlation|",
       y = "Total TFs",
       color = "",
       fill = "") +
  geom_point(aes(fill = logp),
             alpha = 1,
             pch = 21, 
             color = "black",
             stroke = 0,
             show.legend = T) +
  scale_fill_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
    guide = guide_coloursteps(
      # show.limits = TRUE,
      # barwidth = unit(4, "cm"), 
      # barheight = unit(0.2, "cm")    
    )
  ) +
  
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(750, 800), 
           size = 4, show.legend = F) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white", 
                  show.legend = F) +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_corvalue_plot

#Correlation
regulation_tfs_corvalue_shared_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  mutate(quadrant = if_else(cor_value>0, "Positive correlation", "Negative correlation")) %>% 
  drop_na(quadrant) %>% 
  ggscatter(x = "cor_value_abs",
            y = "shared_tf", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line",
            # conf.int = TRUE
              ) +
    scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "|Correlation|",
       y = "Shared TFs",
       color = "",
       fill = "") +
  geom_point(aes(fill = logp),
             alpha = 1,
             pch = 21, 
             color = "black",
             stroke = 0,
             show.legend = T) +
  ylim(0, 100) +
  scale_fill_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
    # guide = guide_coloursteps(
    #   show.limits = TRUE,
    #   barwidth = unit(4, "cm"), 
    #   barheight = unit(0.2, "cm")    
    # )
  ) +
  
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(100, 90), 
           size = 4, show.legend = F) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white", 
                  show.legend = F) +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_corvalue_shared_plot

#Correlation
regulation_tfs_corvalue_shared_perc_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  mutate(logp = -log(p_value)) %>% 
  ggscatter(x = "cor_value_abs",
            y = "shared_tf_pct", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line",
            # conf.int = TRUE
              ) +
    scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "|Correlation|",
       y = "%Shared TFs",
       color = "",
       fill = "") +
  geom_point(aes(fill = logp),
             alpha = 1,
             pch = 21, 
             color = "black",
             stroke = 0,
             show.legend = T) +
  # ylim(0, 100) +
  scale_fill_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
    # guide = guide_coloursteps(
    #   show.limits = TRUE,
    #   barwidth = unit(4, "cm"), 
    #   barheight = unit(0.2, "cm")    
    # )
  ) +
  
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(0.20, 0.18), 
           size = 4, show.legend = F) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white", 
                  show.legend = F) +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_corvalue_shared_perc_plot
```

```{r fig.height=6, fig.width=12}
# ── Patchwork: combine all 6 TF regulation panels into two rows (identity and correlation) and save
regulation_plots <- (regulation_tfs_total_identity_plot | regulation_tfs_identity_plot | regulation_tfs_shared_pct_identity_plot) /
  (regulation_tfs_corvalue_plot | regulation_tfs_corvalue_shared_plot | regulation_tfs_corvalue_shared_perc_plot) +
  plot_annotation(title = "Transcription factors vs. Aminoacid identity and Expression correlation",
                  subtitle = "Shared LEGs with p < 0.10 and rsq > 0.2 are highlighted in blue")

regulation_plots

# Save complete regulation figure
ggsave(regulation_plots,
       filename = here("Figures", "all_human_mouse_gsea_leganalysis_regulation_plots.png"),
       width = 12, height = 6)
```

```{r fig.height=4, fig.width=20}
# ── Alternative layout: single-row patchwork with 4 panels (identity + correlation) and save
regulation_plots <- (regulation_tfs_total_identity_plot | regulation_tfs_shared_pct_identity_plot | regulation_tfs_corvalue_plot | regulation_tfs_corvalue_shared_perc_plot) +
  plot_annotation(title = "Transcription factors vs. Aminoacid identity and Expression correlation",
                  subtitle = "Shared LEGs with p < 0.10 and rsq > 0.2 are highlighted in blue")

#Print
regulation_plots

# Save compact regulation figure
ggsave(regulation_plots,
       filename = here("Figures", "all_human_mouse_gsea_leganalysis_regulation_plots.png"),
       width = 20, height = 4)
```

#### By correlation

```{r}
gene_attribute_tf_count_filter_df = gene_attribute_tf_count %>% 
  ungroup() %>% 
  # Join with the filtered gene quadrants (e.g., High Corr/Low Identity)
  inner_join(alignments_human2mouse_genes_filtered,
             by = "gene")

gene_attribute_tf_count_filter_df_summary = gene_attribute_tf_count_filter_df %>% 
  # Create a descriptive label for the final table
  arrange(desc(both)) %>% 
  group_by(
          # filter, 
           gene, both) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  group_by(gene) %>% 
  mutate(total = sum(n),
         percentage = (n/total*100) %>% round(1))  



gene_attribute_tf_count_filter_df_summary_annotated = gene_attribute_tf_count_filter_df_summary %>%
  arrange(gene) %>% 
  distinct() %>%
    select(-percentage) %>% 
  pivot_wider(values_from = "n",
              names_from = "both") %>% 
  distinct() %>%
  rename(not_shared_tf = "F",
         shared_tf = "T") %>% 
  mutate(shared_tf = if_else(is.na(shared_tf), 0, shared_tf)) %>% 
  inner_join(gene_attribute_tf_count_filter_df %>% 
               ungroup() %>% 
               filter(class %in% c("Immune\nBTMs", "Non-Immune\nBTMs")) %>%
               select(gene, filter, identity_human2mouse, class:cor_label) %>% 
               distinct(),
             by = join_by(gene)) %>% 
  ungroup() %>% 
  mutate(shared_tf_pct = (shared_tf/total) %>% round(2))



#Correlation
gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  # mutate(quadrant = if_else(cor_value>0, "Positive correlation", "Negative correlation")) %>% 
  # drop_na(quadrant) %>% 
  ggplot() +
  aes(x = total,
      y = cor_value,
      label = gene) +
  # facet_wrap(~quadrant,
  #            scales = "free") +
  geom_point(alpha = 0.1) +
  geom_point(data = . %>% 
                filter(p_value < 0.10,
                       r_squared > 0.2),
             color = "black",
             alpha = 0.8) +
  geom_point(color = "black",
             size = 3,
             pch = 21,
             fill = "#4cc9f0",
             data =  gene_attribute_tf_count_filter_df_summary_annotated %>% 
                    inner_join(top_shared_notshared_genes %>% 
                                 filter(class == "All") %>% 
                                 filter(p_value < 0.10,
                                 r_squared > 0.2) %>% 
                                 select(gene) %>% 
                                 distinct(),
                               by = "gene")) +
  geom_label_repel(color = "black",
                   data =  gene_attribute_tf_count_filter_df_summary_annotated %>% 
                    inner_join(top_shared_notshared_genes %>% 
                                 filter(class == "All") %>%
                                 filter(p_value < 0.10,
                                 r_squared > 0.2) %>% 
                                 select(gene) %>% 
                                 distinct(),
                               by = "gene"),
                  size = 2.5,
                  fill = "white") +
  # geom_smooth(method = "lm") +
  # geom_smooth(method = "lm",
  #             data = . %>% 
  #               filter(p_value < 0.10,
  #                      r_squared > 0.2),
  #            color = "blue") +
  theme_vaxgo()



#Identity
gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  ggplot() +
  aes(x = total,
      y = identity_human2mouse,
      # label = gene
      ) +
  geom_point(alpha = 0.1) +
  geom_point(
    # data = . %>% 
    #             filter(p_value < 0.10,
    #                    r_squared > 0.2),
             color = "black",
             pch = 21,
             alpha = 0.8,
             mapping = aes(fill = filter)) +

  # geom_smooth(method = "lm") +
  theme_vaxgo()



gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  ggplot() +
  aes(x = shared_tf,
      y = identity_human2mouse,
      # label = gene
      ) +
  geom_point(alpha = 0.1) +
  geom_point(
    # data = . %>% 
    #             filter(p_value < 0.10,
    #                    r_squared > 0.2),
             color = "black",
             pch = 21,
             alpha = 0.8,
             mapping = aes(fill = filter)) +

  # geom_smooth(method = "lm") +
  theme_vaxgo()

```














