---
title: "5_EvolutionaryAnalysis"
output: html_document
---


# Evolutionary analysis
# Setup

```{r message=FALSE, warning=FALSE}
# Load core packages for the analysis session
library(here)
source(here("scripts_notebooks", "required.R"))


library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(GenomeInfoDb)
library(GenomicRanges)
library(IRanges)
library(rtracklayer)
library(org.Hs.eg.db)
library(pwalign)
library(biomaRt)
```
# Regulation

## Import data

```{r fig.height=5, fig.width=12}
# Import BTM core enrichment results
gsea_btm_results_core_joined =
  readRDS(file = here("tables", "all_human_mouse_gsea_btm_results_core_joined.rds")) %>%
  distinct()

#Import modules and LEGs
dge_btm_process_module_diff_and_legs = 
  readRDS(file = here("tables", "all_human_mouse_gsea_btm_legs.rds"))

# Import BTM gene-level correlation data
dge_btm_process_genes_diff_bygene =
  readRDS(here("tables", "human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune.rds"))

# Import homolog alignment data
all_alignments = read_csv(here("tables", "human_to_mouse_alignment_mart_export.txt"))

# Import BTM gene annotations
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

# Annotate genes with sequence identity and BTM membership
# Build alignment-annotated dataset for core enrichment gene

# Reference: all high-confidence orthologs in the database (unfiltered)
alignments_human2mouse_all = all_alignments %>%
  clean_names() %>%
  filter(mouse_orthology_confidence_0_low_1_high == 1) %>%
  drop_na(mouse_gene_name) %>%
  select(gene = mouse_gene_name, mouse_homology_type,
         identity_human2mouse = percent_id_query_gene_identical_to_target_mouse_gene,
         identity_mouse2human = percent_id_target_mouse_gene_identical_to_query_gene) %>%
  mutate(gene = str_to_upper(gene)) %>%
  # Resolve multiple identity scores by taking the median per gene
  group_by(gene) %>%
  summarise(identity_human2mouse = median(identity_human2mouse)) %>%
  ungroup()

# Filter to only BTM core enrichment genes and append process membership
alignments_human2mouse = alignments_human2mouse_all %>% 
  inner_join(dge_btm_process_module_diff_and_legs,
             by = join_by(gene == symbol))  %>% 
  mutate(status = if_else(status == "Shared", status, "Not-shared"))

#Group by process
protein_identity_process = alignments_human2mouse %>% 
  group_by(treatment, pathogen, timepoint_comparison, process, status) %>% 
  summarise(facet_label = dplyr::first(facet_label), group = dplyr::first(group),
            mean_identity = mean(identity_human2mouse, na.rm = T))

# Import pre-computed human gene information and join with alignment data
gene_info_human_log2fc_stats =
  readRDS(file = here("tables", "gene_info_human.rds")) %>%
  select(-hgnc_symbol) %>% 
  inner_join(alignments_human2mouse_all, by = "gene") %>% 
  full_join(alignments_human2mouse %>% 
              select(gene, process, pathogen, treatment, timepoint_comparison, group, status) %>% 
              distinct(),
          by = "gene") %>%
  
  #Add difference in log2FC
  right_join(dge_btm_process_genes_diff_bygene %>% 
               filter(gene_set == "All genes") %>% 
               filter(pathogen_mouse != "DMD") %>% 
               select(pathogen = pathogen_human, timepoint_comparison, 
                      treatment = treatment_human, 
                      gene = human_symbol, mean_log2fc_human, 
                      mean_log2fc_mouse, mean_log2fc_diff, 
                      weight_se_inverse_combined) %>% 
               # mutate(status = "not LEG") %>% 
               distinct(),
             by = join_by(gene, pathogen, timepoint_comparison, treatment)) %>% 
  mutate(status = if_else(is.na(status), "Not core", status))
#Save
gene_info_human_log2fc_stats %>% 
  saveRDS(file = here("tables", "gene_info_human_log2fc_stats.rds"))


gene_info_human_log2fc_stats$status %>% unique()
```

#------------------------

## Cis-regulatory elements



### Import and merge data

```{r import-data}

#Import gene info with log2fc stats
gene_info_human_log2fc_stats = readRDS(file = here("tables", "gene_info_human_log2fc_stats.rds"))
#ENCODE ===============
#In contrast to the Human dataset, there are no CTCF-bound CREs in the main, full CRE mouse dataset. So, we imported CREs separately and merged them.

# Human CREs (GRCh38) =============

#PLS
cres_human_pls <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_PLS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_original_human = type_human,
         ctcf_human = "not-CTCF") 

#pELS-dELS
cres_human_pelsdels <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_pELS-dELS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_original_human = type_human,
         ctcf_human = "not-CTCF") 

#CTCF BOUND
cres_human_ctcfbound <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_CTCF-bound.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type_original"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_human = str_remove(type_original_human, ",CTCF-bound"),
         ctcf_human = "CTCF-bound")

# Load ENCODE comprehensive gene-CRE association table
genes_cres_human_original <- read_tsv(
  here("tables", "Genomic", "genomic_human_cCREs_GRCh38_Closest-Genes-PC.tsv"),
  col_names = c("chrom_cre", "start_cre", "end_cre", "id_dhs", "id_ccre", "type_cre",
                "chrom_gene", "start_gene", "end_gene", "transcript_id", "score_gene", "strand_gene", "gene_id"),
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) #2,651,787


#Check overlaps and remove CTCF-bound cCREs that are not annotated accordingly =====
#PLS
cres_human_ctcfbound_pls_overlap = cres_human_ctcfbound %>% 
  inner_join(cres_human_pls,
             by = "id_ccre_human") %>% 
  select(id_ccre_human) %>% 
  distinct() %>% 
  arrange(id_ccre_human) #36,430 IDs

#47,532 -> 11,102 IDs
cres_human_pls_clean = cres_human_pls %>% 
  filter(!id_ccre_human %in% cres_human_ctcfbound_pls_overlap$id_ccre_human) 

#pELS-dELS
cres_human_ctcfbound_pelsdels_overlap = cres_human_ctcfbound %>% 
  inner_join(cres_human_pelsdels,
             by = "id_ccre_human")%>% 
  select(id_ccre_human) %>% 
  distinct() #720,306 IDs

#1,718,669 -> 998,363
cres_human_pelsdels_clean = cres_human_pelsdels %>% 
  filter(!id_ccre_human %in% cres_human_ctcfbound_pelsdels_overlap$id_ccre_human)

#Merge
cres_human = bind_rows(cres_human_pls_clean,
          cres_human_pelsdels_clean,
          cres_human_ctcfbound)


#Annotate Human symbol ===========
library(org.Hs.eg.db)

hgnc_symbol <- AnnotationDbi::mapIds(
  x         = org.Hs.eg.db,
  keys      = unique(genes_cres_human_original$gene_id_clean),
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
) %>%
  tibble::enframe(name = "ensembl_gene_id", value = "hgnc_symbol") %>%
  dplyr::filter(!is.na(hgnc_symbol) & hgnc_symbol != "")


# Join HGNC symbols to gene table
genes_cres_human <- genes_cres_human_original %>%
  left_join(hgnc_symbol, by = c("gene_id_clean" = "ensembl_gene_id")) %>% 
  rename(chrom_human = chrom_cre, start_human = start_cre, end_human = end_cre,
         id_ccre_human = id_ccre, id_dhs_human = id_dhs) %>% 
  #Add CRE ctcf
  inner_join(cres_human %>% #1,958,107
               select(chrom_human, start_human, end_human, id_dhs_human, id_ccre_human, type_human, ctcf_human),
             by = join_by(chrom_human, start_human, end_human, id_dhs_human, id_ccre_human)) %>% #2,214,457
  select(-type_cre)

genes_cres_human

# Mouse CREs (mm10) ============
#PLS
cres_mouse_pls <- read_tsv(
  here("tables", "Genomic", "genomic_mouse_cCREs_mm10_PLS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_original_mouse = type_mouse,
         ctcf_mouse = "not-CTCF") 

#pELS-dELS
cres_mouse_pelsdels <- read_tsv(
  here("tables", "Genomic", "genomic_mouse_cCREs_mm10_pELS-dELS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_original_mouse = type_mouse,
         ctcf_mouse = "not-CTCF")

#CTCF BOUND
cres_mouse_ctcfbound <- read_tsv(
  here("tables", "Genomic", "genomic_mouse_cCREs_mm10_CTCF-bound.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type_original"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_mouse = str_remove(type_original_mouse, ",CTCF-bound"),
         ctcf_mouse = "CTCF-bound")


#Check overlaps and remove CTCF-bound cCREs that are not annotated accordingly =====
#PLS
cres_mouse_ctcfbound_pls_overlap = cres_mouse_ctcfbound %>% 
  inner_join(cres_mouse_pls,
             by = "id_ccre_mouse") %>% 
  select(id_ccre_mouse) %>% 
  distinct() %>% 
  arrange(id_ccre_mouse) #11.884 IDs

#27.332 -> 15,448 IDs
cres_mouse_pls_clean = cres_mouse_pls %>% 
  filter(!id_ccre_mouse %in% cres_mouse_ctcfbound_pls_overlap$id_ccre_mouse) 

#pELS-dELS
cres_mouse_ctcfbound_pelsdels_overlap = cres_mouse_ctcfbound %>% 
  inner_join(cres_mouse_pelsdels,
             by = "id_ccre_mouse")%>% 
  select(id_ccre_mouse) %>% 
  distinct() #66.186 IDs

#512.001 -> 445,815 
cres_mouse_pelsdels_clean = cres_mouse_pelsdels %>% 
  filter(!id_ccre_mouse %in% cres_mouse_ctcfbound_pelsdels_overlap$id_ccre_mouse)

#Merge
cres_mouse = bind_rows(cres_mouse_pls_clean,
          cres_mouse_pelsdels_clean,
          cres_mouse_ctcfbound)

# Import homologous CREs ====================
# Homologous CRE pairs from ENCODE (hg38–mm10) Nearest gene, protein coding only
homologous_cres <- read_tsv(
  here("tables", "Genomic", "genomic_human_mouse_cCREs_homologous_GRCh38_Mm10.tsv"),
  col_names = c("human_enhancer", "mouse_enhancer"),
  show_col_types = FALSE
) #537,822

```

### Summarize data 
Count how many mouse CREs each human CRE maps to (and vice versa)

```{r classify-match-type}
#Summarize classes
#Only elite enhancers
#Per gene
genes_cres_human %>% 
  filter(type_human %in% c("pELS", "dELS", "PLS")) %>% 
  group_by(hgnc_symbol, type_human, ctcf_human) %>% 
  dplyr::summarise(n = n()) %>%
  arrange(hgnc_symbol) %>% 
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_human, desc(n))

cres_mouse %>% 
  dplyr::count(type_mouse, ctcf_mouse) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_mouse, desc(n))

# Count how many mouse CREs each human CRE maps to (and vice versa)
human_counts <- homologous_cres %>%
  dplyr::count(human_enhancer, name = "n_mouse_per_human")

mouse_counts <- homologous_cres %>%
  dplyr::count(mouse_enhancer, name = "n_human_per_mouse")

# Full join: preserves orphan CREs on both sides
cres_annotated <- homologous_cres %>%
  left_join(human_counts, by = "human_enhancer") %>%
  left_join(mouse_counts, by = "mouse_enhancer") %>%
  #Annotate with human gene
  full_join(genes_cres_human, by = join_by(human_enhancer == id_ccre_human)) %>%
  #Annotate mouse enhancer with other info
  full_join(cres_mouse, by = join_by(mouse_enhancer == id_ccre_mouse)) %>%
  mutate(type_mouse = case_when(!type_mouse %in% c("PLS", "pELS", "dELS") ~ "Other",
                                T ~ type_mouse)) %>% 
  #Filter out enhancers without mouse annotation
  drop_na(chrom_mouse, hgnc_symbol) %>% 
  mutate(
    is_homologous = !is.na(human_enhancer) & !is.na(mouse_enhancer),
    match_homology = case_when(
      is.na(human_enhancer)                            ~ "mouse_only",
      is.na(mouse_enhancer)                            ~ "human_only",
      n_mouse_per_human == 1 & n_human_per_mouse == 1  ~ "1:1",
      n_mouse_per_human >  1 & n_human_per_mouse == 1  ~ "N:1",
      n_mouse_per_human == 1 & n_human_per_mouse >  1  ~ "1:N",
      TRUE                                             ~ "N:M"
    ),
    match_cretype = if_else(type_human == type_mouse, "Same", "Different"),
    match_ctcf = if_else(ctcf_human == ctcf_mouse, "Same", "Different")
  ) 
#Print
cres_annotated

#Save
cres_annotated %>% saveRDS(file = here("tables", "cres_homologous_annotated.rds"))


```

#### How many CREs per gene?
```{r fig.height=5, fig.width=10}
#Import annotated CREs
cres_annotated = readRDS(file = here("tables", "cres_homologous_annotated.rds"))

#=======================================
#Compare classes of CREs between species
#======================================

# Calculate cCRE type, ctcf, and homology percentages per locus

cres_type_homology_comparison <- cres_annotated %>% #453,679
  
  # Step 1: Filter target human cCRE types and valid gene symbols
  # Note: We do not filter type_mouse here so we don't lose 'Species Specific' cCREs
  dplyr::filter(
    type_human %in% c("pELS", "dELS", "PLS"),
    !is.na(hgnc_symbol) & hgnc_symbol != ""
  ) %>% #442,995
  
  # Step 2: Handle NAs BEFORE grouping and calculations
  dplyr::mutate(
    type_original_mouse       = dplyr::if_else(is.na(type_original_mouse), "Unknown", as.character(type_original_mouse)),
    type_mouse       = dplyr::if_else(is.na(type_mouse), "Unknown", as.character(type_mouse)),
    ctcf_mouse    = dplyr::if_else(is.na(ctcf_mouse), "Unknown", as.character(ctcf_mouse)),
    match_cretype    = dplyr::if_else(is.na(match_cretype), "Unknown", as.character(match_cretype)),
    match_ctcf = dplyr::if_else(is.na(match_ctcf), "Unknown", as.character(match_ctcf)),
    
    # Collapse granular match_homology into broad simplified categories
    match_homology_group = dplyr::case_when(
      match_homology %in% c("1:1") ~ "Homologous",
      match_homology %in% c("1:N", "N:1", "N:M") ~ "Complex homology",
      match_homology %in% c("human_only", "mouse_only") ~ "Species Specific",
      TRUE ~ "Other"
    ),
    match_homology_group = forcats::fct_relevel(
      match_homology_group, "Homologous", "Complex homology", "Species Specific"
    )
  ) %>%
  tidyr::drop_na(match_homology_group) %>%
  
  # Step 3: Count occurrences per gene locus and annotation/homology stratum
  dplyr::group_by(
    hgnc_symbol, type_human, ctcf_human, type_mouse, ctcf_mouse,
    match_cretype, match_ctcf, match_homology_group
  ) %>%
  dplyr::summarise(n_cres = dplyr::n(), .groups = "drop") %>%
  
  # # Step 4: Complete missing combinations with 0 to explicitly retain n = 0 per stratum
  # tidyr::complete(
  #   hgnc_symbol,
  #   type_human = c("pELS", "dELS", "PLS"),
  #   ctcf_human,
  #   match_homology_group,
  #   fill = list(n_cres = 0)
  # ) %>%
  # 
  # Step 5: Compute total locus cCREs
  dplyr::group_by(hgnc_symbol) %>%
  dplyr::mutate(
    n_total_cres_gene = sum(n_cres, na.rm = TRUE)
  ) %>%
  
  # Step 6: Compute counts and percentage matching relative to the same Human Type
  dplyr::group_by(hgnc_symbol, type_human) %>%
  dplyr::mutate(
    n_type         = sum(n_cres, na.rm = TRUE),
    # Using %in% safely ignores NAs introduced by tidyr::complete
    n_match_type   = sum(n_cres[match_cretype %in% c("Same", "TRUE", TRUE)], na.rm = TRUE),
    pct_match_type = dplyr::if_else(n_type > 0, round(100 * n_match_type / n_type, 2), 0)
  ) %>%
  
  # Step 7: Compute counts and percentage matching relative to Human Type & ctcf
  dplyr::group_by(hgnc_symbol, type_human, ctcf_human) %>%
  dplyr::mutate(
    n_ctcf         = sum(n_cres, na.rm = TRUE),
    n_match_ctcf   = sum(n_cres[match_ctcf %in% c("Same", "TRUE", TRUE)], na.rm = TRUE),
    pct_match_ctcf = dplyr::if_else(n_ctcf > 0, round(100 * n_match_ctcf / n_ctcf, 2), 0)
  ) %>%
  
  # Step 8: Compute homology counts and percentage WITHIN each Human Type & ctcf stratum
  dplyr::group_by(hgnc_symbol, type_human, ctcf_human, match_homology_group) %>%
  dplyr::mutate(
    n_homology   = sum(n_cres, na.rm = TRUE),
    pct_homology = dplyr::if_else(n_ctcf > 0, round(100 * n_homology / n_ctcf, 2), 0)
  ) %>%
  dplyr::ungroup() %>%
  
  # Step 9: Categorize cCRE counts into unified frequency bins (n_type_cat)
  dplyr::mutate(
    n_type_cat = dplyr::case_when(
      type_human == "PLS" & n_type >= 5 ~ ">=5",
      type_human %in% c("pELS", "dELS") & n_type >= 20 ~ ">=20",
      TRUE ~ as.character(n_type)
    ),
      n_type_cat = factor(
      n_type_cat, 
      levels = c(
        as.character(0:4), ">=5",      
        as.character(5:19), ">=20"  
      )
    )
  ) %>%
  
  # Step 10: Set factor levels for plotting and sort by locus and count
  dplyr::mutate(
    type_mouse       = dplyr::if_else(is.na(type_mouse), "Unknown", as.character(type_mouse)),
    ctcf_mouse    = dplyr::if_else(is.na(ctcf_mouse), "Unknown", as.character(ctcf_mouse)),
    match_cretype    = dplyr::if_else(is.na(match_cretype), "Unknown", as.character(match_cretype)),
    match_ctcf = dplyr::if_else(is.na(match_ctcf), "Unknown", as.character(match_ctcf)),
    type_human = fct_relevel(type_human, "PLS", "pELS", "dELS"),
    type_mouse = fct_relevel(type_mouse, "PLS", "pELS", "dELS")
  ) %>% 
  dplyr::arrange(hgnc_symbol, dplyr::desc(n_type))

 
  
#Print
cres_type_homology_comparison 

#Save 
cres_type_homology_comparison %>% 
  saveRDS(file = here("tables", "cres_type_homology_comparison.rds"))



```

#### Have CREs diverged their functions?

```{r classify-match-type, fig.height=6, fig.width=8}
#Import
cres_type_homology_comparison_dge_legs_stats =  
  readRDS(file = here("tables", "cres_type_homology_comparison_dge_legs_stats.rds"))
cres_type_homology_comparison =  
  readRDS(file = here("tables", "cres_type_homology_comparison.rds"))

# Calculate shared cCRE type and ctcf proportions across species ======
cres_type_matrix_pct <- cres_type_homology_comparison %>%
  
  # Step 1: Aggregate total cCRE counts per Human & Mouse type/ctcf combination
  dplyr::group_by(type_human, ctcf_human, type_mouse, ctcf_mouse) %>%
  dplyr::summarise(n_pair_cres = sum(n_cres, na.rm = T), .groups = "drop") %>%
  
  # Step 2: Compute relative percentage relative to total human cCREs in that stratum
  dplyr::group_by(type_human, ctcf_human) %>%
  dplyr::mutate(
    n_human_stratum = sum(n_pair_cres, na.rm = T),
    pct             = dplyr::if_else(
      n_human_stratum > 0, 
      round(100 * n_pair_cres / n_human_stratum, 1), 
      0
    )
  ) %>%
  dplyr::ungroup() %>%
  
  # Step 3: Order factor levels for heatmaps
  dplyr::mutate(
    type_human = fct_relevel(type_human, "PLS", "pELS", "dELS"),
    type_mouse = fct_relevel(type_mouse, "PLS", "pELS", "dELS", "CA-H3K4", "CA-CTCF", "TF"),
    facet_human = paste0("Human ", type_human),
    facet_mouse = paste0("Mouse ", type_mouse),
  ) %>% 
  arrange(type_mouse) %>% 
  mutate(facet_mouse = fct_inorder(facet_mouse)) %>% 
  arrange(type_human) %>% 
  mutate(facet_human = fct_inorder(facet_human))

#Heatmap
cres_type_matrix_pct_mouse_human_plot = cres_type_matrix_pct %>% 
  #Filter out CREs that do not exist in mice
  drop_na(type_mouse) %>%
  #Plot
  ggplot() +
    aes(x = ctcf_mouse %>% str_replace(., "-", "\n") %>% fct_rev(),
         y = ctcf_human %>% str_replace(., "-", "\n")%>% fct_rev()) +
  facet_grid(~facet_mouse~facet_human) +
  # geom_tile(aes(fill = pct), color = "black", linewidth = 0.4) +
  geom_point(aes(fill = pct, size = n_pair_cres), 
             color = "black",
             pch = 21) +
  geom_text(aes(label = paste0(pct, "%")),
            size = 3.5,
            nudge_y = 0.4) +

  scale_fill_gradient2(low = "white",
                       high = "#4DBBD5FF",
                       guide = guide_colorbar(barwidth = unit(1, "cm"))) +
  scale_size_continuous(range = c(4, 12)) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0,
                                   # vjust = 0,
                                   hjust = 0.5),
        legend.position = "right",
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)
        ) +
  labs(x = "Mouse",
       y = "Human",
       fill = "%Shared,\nfrom human",
       size = "CRE pairs") 

#Print
cres_type_matrix_pct_mouse_human_plot

#Save
cres_type_matrix_pct_mouse_human_plot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_regulation_cres_type_matrix_pct_plot.png"),
         width = 8, height = 6)

```


#### Annotate with gene expression and LEGs

```{r}
#### Annotate with gene expression and LEGs

# Gene expression
dge_btm_process_genes_diff_bygene_clean_filtered =  readRDS(file = here("tables", "dge_btm_process_genes_diff_bygene_clean_filtered.rds"))

  
  
  # Inner join with cross-species differential expression statistics =============
cres_type_homology_comparison_dge_legs_stats <- 
  
  bind_rows( cres_type_homology_comparison %>%
   #Annotate with gene expression
  inner_join(dge_btm_process_genes_diff_bygene_clean_filtered,
             by = join_by(hgnc_symbol)) %>% 
  #Annotate with LEG classification
  left_join(gsea_btm_legs_clean, 
             by = join_by(hgnc_symbol, pathogen, timepoint_comparison)) %>% 
  mutate(status = case_when(is.na(status)~"Not LEG", 
                            status %in% c("Human only", "Mouse only") ~ "Not-shared",
                            T~status)),
  #All genes
  cres_type_homology_comparison %>%
  inner_join(dge_btm_process_genes_diff_bygene_clean_filtered,
             by = join_by(hgnc_symbol)) %>% 
  mutate(status = "All genes")
    )

  
#Print
cres_type_homology_comparison_dge_legs_stats


cres_type_homology_comparison_dge_legs_stats %>% dplyr::count(status)

#Save
cres_type_homology_comparison_dge_legs_stats %>% 
  saveRDS(file = here("tables", "cres_type_homology_comparison_dge_legs_stats.rds"))
```


#### Distribution of CREs by gene vs Divergence in Log2FC

```{r fig.height=3, fig.width=8}
#Plot
cres_type_divergence_pooled_plot = cres_type_homology_comparison_dge_legs_stats %>% 
  distinct() %>% 
  filter(status != "Not LEG") %>% 
  ggplot() +
  aes(x = n_type,
      y = abs_log2fc_diff,
      color = status,
      fill = status) +
  
  facet_wrap(~type_human, scales = "free_x") +
  #By LEG
  geom_smooth(method = "lm",
              mapping = aes(color = status,
                            fill = status),
              se = T) +
  scale_x_continuous(expand = c(0.01, 0.01), limits = c(0, NA)) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(
       x = "N CREs",
       y = "|ΔMean Log2FC|",
       fill = "LEG status") +
  theme_vaxgo() +
  theme(legend.position = "top",
        panel.spacing = grid::unit(0.7, "cm"),
        plot.margin          = ggplot2::margin(0.1, 0.5, 0.1, 0.1, "cm")) +
  guides(
    color = "none",
    fill  = guide_legend(title.position = "left", 
                         byrow = T, nrow = 1,
                         override.aes = list(shape = 22, alpha = 1, size = 3,
                                             linetype = 0)))

cres_type_inverseSE_pooled_plot

#Print
cres_type_divergence_pooled_marginal = cres_type_divergence_pooled_plot +
  # 1. TOP SIDE: Boxplot horizontal no limite superior do eixo Y
  geom_boxplot(
    data = . %>% 
      group_by(type_human) %>% 
      mutate(
        max_y = max(abs_log2fc_diff, na.rm = TRUE),
      ),
    mapping = aes(
      x = n_type,         # Variável contínua no eixo X
      y = 3,             # Posição fixa no topo do eixo Y
      fill = status,
      group = interaction(status, type_human),
      width = box_height           # Controla a espessura no eio Y
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0.5,
    width = 0.5,
    show.legend = FALSE
  ) +
  #Right side
  geom_boxplot(
    data = . %>%
      group_by(type_human) %>%
      mutate(
          max_x = max(n_type, na.rm = TRUE),
        box_width = max_x * 0.15 
      ),
    mapping = aes(
      x = max_x,
      y = abs_log2fc_diff,
      fill = status,
      group = interaction(status, type_human),
      width = box_width
    ),
    outliers = F,
    outlier.shape = NA,
    alpha = 0.5,
    show.legend = FALSE
  ) 


# Display composite plot
print(cres_type_divergence_pooled_marginal)

#Save
cres_type_divergence_pooled_marginal %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_degs_cretype_divergence_pooled.png"),
         width = 8,
         height = 3)

```



```{r fig.height=3, fig.width=12}
#Plot
cres_type_ctcf_divergence_pooled_plot = cres_type_homology_comparison_dge_legs_stats %>% 
  distinct() %>% 
  filter(status != "Not LEG") %>% 
  ggplot() +
  aes(x = n_type,
      y = abs_log2fc_diff,
      color = status,
      fill = status) +
  
  facet_nested_wrap(~type_human+ctcf_human, scales = "free_x", nrow = 1) +
  #By LEG
  geom_smooth(method = "lm",
              mapping = aes(color = status,
                            fill = status),
              se = T) +
  scale_x_continuous(expand = c(0.01, 0.01), limits = c(0, NA)) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(
       x = "N CREs",
       y = "|ΔMean Log2FC|",
       fill = "LEG status") +
  theme_vaxgo() +
  theme(legend.position = "top",
        panel.spacing = grid::unit(0.7, "cm"),
        # plot.margin          = ggplot2::margin(0.1, 0.5, 0.1, 0.1, "cm")
        ) +
  guides(
    color = "none",
    fill  = guide_legend(title.position = "left", 
                         byrow = T, nrow = 1,
                         override.aes = list(shape = 22, alpha = 1, size = 3,
                                             linetype = 0)))

cres_type_ctcf_divergence_pooled_plot

#Print
cres_type_ctcf_divergence_pooled_plot_marginal = cres_type_ctcf_divergence_pooled_plot +
  # 1. TOP SIDE: Boxplot horizontal no limite superior do eixo Y
  geom_boxplot(
    data = . %>% 
      group_by(type_human, ctcf_human) %>% 
      mutate(
        max_y = max(abs_log2fc_diff, na.rm = TRUE),
      ),
    mapping = aes(
      x = n_type,         # Variável contínua no eixo X
      y = 3,             # Posição fixa no topo do eixo Y
      fill = status,
      group = interaction(status, type_human, ctcf_human),
      width = box_height           # Controla a espessura no eio Y
    ),
    outliers = FALSE,
    outlier.shape = NA,
    alpha = 0.5,
    width = 0.5,
    show.legend = FALSE
  ) +
  #Right side
  geom_boxplot(
    data = . %>%
      group_by(type_human, ctcf_human) %>%
      mutate(
          max_x = max(n_type, na.rm = TRUE),
        box_width = max_x * 0.15 
      ),
    mapping = aes(
      x = max_x,
      y = abs_log2fc_diff,
      fill = status,
      group = interaction(status, type_human, ctcf_human),
      width = box_width
    ),
    outliers = F,
    outlier.shape = NA,
    alpha = 0.5,
    show.legend = FALSE
  ) 


# Display composite plot
print(cres_type_ctcf_divergence_pooled_plot_marginal)

#Save
cres_type_ctcf_divergence_pooled_plot_marginal %>%
  ggsave(filename = here("Figures", "all_human_mouse_dge_degs_cretype_ctcf_human_pooled.png"),
         width = 8,
         height = 2.5)

```












#### Distribution of CREs by gene, by LEGs
https://pubmed.ncbi.nlm.nih.gov/29180706/#&gid=article-figures&pid=figure-2-uid-1

```{r fig.height=3, fig.width=16}
#Plot
cres_type_categorization_pooled_plot = cres_type_homology_comparison_dge_legs_stats %>% 
  filter(n_type_cat != "0") %>%
  filter(status %in% c("Shared", "Not-shared")) %>%
  ggplot() +
  aes(x = n_type_cat,
      y = abs_log2fc_diff,
      color = status,
      fill =  status,
      ) +
  facet_wrap(~type_human, space = "free_x",
                    nrow = 1, axes = "all", scales = "free") +
    geom_point(alpha = 0.2,
             size = 0.5,
            position = position_jitterdodge(jitter.width = 0.3, 
                                    dodge.width = 0.75),) +
  geom_boxplot(width = 0.7,
               alpha = 0.5,
               outliers = F,
               color = "black")+
  
  # STATISTICAL TEST LAYER
  stat_compare_means(
    aes(group = status),
    method = "wilcox.test",
    label = "p.signif",
    symnum.args = list(
      cutpoints = c(0, 0.001, 0.01, 0.05, 1),
      symbols = c("***", "**", "*", "ns")
    ),
    vjust = 1.2,
    label.y.np = "top",
    hide.ns = T
  ) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(x = "N",
       y = expression("|" * Delta * "Log"[2] * "FC|"),
       color = "LEGs",
       fill = "LEGs",
       title = "Pooled gene expression conditions") +
  theme_vaxgo() +
  theme(legend.position = "top")

#Print
cres_type_categorization_pooled_plot

#Save
cres_type_categorization_pooled_plot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_cres_type_categorization_pooled_plot.png"),
         height = 3, width = 16)

```


```{r fig.height=3, fig.width=8}

cres_stat_results <- cres_type_homology_comparison_dge_legs_stats %>%
  dplyr::filter(
    n_type_cat != "0",
    status %in% c("Shared", "Not-shared")
  ) %>%
  dplyr::group_by(type_human) %>%
  dplyr::summarise(
    n_shared     = sum(status == "Shared", na.rm = TRUE),
    n_not_shared = sum(status == "Not-shared", na.rm = TRUE),
    
    # Calculate Wilcoxon test p-value per cCRE type
    p_val        = if (n_shared >= 1 && n_not_shared >= 1) {
      stats::wilcox.test(
        n_total_cre_type_human[status == "Shared"],
        n_total_cre_type_human[status == "Not-shared"]
      )$p.value
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    # Format p-values into adjusted FDR scientific notation strings
    fdr_p_val      = stats::p.adjust(p_val, method = "BH"),
    fdr_wilcox_sci = sprintf("%.2e", fdr_p_val),
    
    # Construct clean label for facet headers
    facet_label    = stringr::str_c(type_human, "\n(p = ", fdr_wilcox_sci, ")")
  )


#Plot
cres_type_comparison_plot <- cres_type_homology_comparison_dge_legs_stats %>%
  dplyr::filter(
    n_type_cat != "0",
    status %in% c("Shared", "Not-shared")
  ) %>%
  # Attach pre-calculated p-value labels
  dplyr::inner_join(
    cres_stat_results %>% dplyr::select(type_human, facet_label),
    by = "type_human"
  ) %>%
  dplyr::mutate(
    status      = factor(status, levels = c("Shared", "Not-shared")),
    facet_label = forcats::fct_reorder(facet_label, as.numeric(factor(type_human, levels = c("PLS", "pELS", "dELS"))))
  ) %>%
  ggplot2::ggplot() +
  ggplot2::aes(
    x     = status,
    y     = n_total_cre_type_human,
    color = status,
    fill  = status
  ) +
  # Create nested wrap facets incorporating the formatted p-value string
  ggh4x::facet_nested_wrap(
    ~ facet_label,
    nrow   = 1,
    scales = "free_y"
  ) +
  # Render boxplot layer without outliers
  ggplot2::geom_boxplot(
    width    = 0.3,
    alpha    = 1,
    outliers = FALSE,
    color    = "black"
  ) +
  # Apply project custom palettes and labels
  ggplot2::scale_fill_manual(values = colors$comparison) +
  ggplot2::scale_color_manual(values = colors$comparison) +
  ggplot2::labs(
    x     = "Status",
    y     = "Total cCREs per Gene",
    color = "",
    fill  = ""
  ) +
  # Set Y-axis view boundaries
  ggplot2::coord_cartesian(ylim = c(0, NA)) +
  theme_vaxgo() +
  ggplot2::theme(
    legend.position = "top",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )


cres_type_comparison_plot
```

CRE ctcf
```{r fig.height=3, fig.width=12}
cres_ctcf_stat_results <- cres_type_homology_comparison_dge_legs_stats %>%
  dplyr::filter(
    n_type_cat != "0",
    status %in% c("Shared", "Not-shared")
  ) %>%
  dplyr::group_by(type_human, ctcf_human) %>%
  dplyr::summarise(
    n_shared     = sum(status == "Shared", na.rm = TRUE),
    n_not_shared = sum(status == "Not-shared", na.rm = TRUE),
    
    # Calculate Wilcoxon test p-value per cCRE type
    p_val        = if (n_shared >= 1 && n_not_shared >= 1) {
      stats::wilcox.test(
        n_total_cre_type_human[status == "Shared"],
        n_total_cre_type_human[status == "Not-shared"]
      )$p.value
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    # Format p-values into adjusted FDR scientific notation strings
    fdr_p_val      = stats::p.adjust(p_val, method = "BH"),
    fdr_wilcox_sci = sprintf("%.2e", fdr_p_val),
    
    # Construct clean label for facet headers
    facet_label    = stringr::str_c(type_human, ", ", ctcf_human, "\n(p = ", fdr_wilcox_sci, ")")
  )
#Plot
cres_ctcf_comparison_plot <- cres_type_homology_comparison_dge_legs_stats %>%
  dplyr::filter(
    n_type_cat != "0",
    status %in% c("Shared", "Not-shared")
  ) %>%
  # Attach pre-calculated p-value labels
  dplyr::inner_join(
    cres_ctcf_stat_results %>% dplyr::select(type_human, ctcf_human, facet_label),
    by = join_by(type_human, ctcf_human)
  ) %>%
  dplyr::mutate(
    status      = factor(status, levels = c("Shared", "Not-shared")),
    facet_label = forcats::fct_reorder(facet_label, as.numeric(factor(type_human, levels = c("PLS", "pELS", "dELS"))))
  ) %>%
  ggplot2::ggplot() +
  ggplot2::aes(
    x     = status,
    y     = n_total_cre_type_human,
    color = status,
    fill  = status
  ) +
  # Create nested wrap facets incorporating the formatted p-value string
  ggh4x::facet_nested_wrap(
    ~ facet_label,
    nrow   = 1,
    scales = "free_y"
  ) +
  # Render boxplot layer without outliers
  ggplot2::geom_boxplot(
    width    = 0.3,
    alpha    = 1,
    outliers = FALSE,
    color    = "black"
  ) +
  # Apply project custom palettes and labels
  ggplot2::scale_fill_manual(values = colors$comparison) +
  ggplot2::scale_color_manual(values = colors$comparison) +
  ggplot2::labs(
    x     = "Status",
    y     = "Total cCREs per Gene",
    color = "",
    fill  = ""
  ) +
  # Set Y-axis view boundaries
  ggplot2::coord_cartesian(ylim = c(0, NA)) +
  theme_vaxgo() +
  ggplot2::theme(
    legend.position = "top",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )


cres_ctcf_comparison_plot

```

CRE type per pathogen
```{r fig.height=8, fig.width=8}
# Step 1: Pre-calculate statistical p-values per pathogen and human cCRE type
cres_stat_results_pathogen <- cres_type_homology_comparison_dge_legs_stats %>%
  dplyr::filter(
    n_type_cat != "0",
    status %in% c("Shared", "Not-shared")
  ) %>%
  dplyr::group_by(pathogen, type_human) %>%
  dplyr::summarise(
    n_shared     = sum(status == "Shared", na.rm = TRUE),
    n_not_shared = sum(status == "Not-shared", na.rm = TRUE),
    
    # Calculate Wilcoxon test p-value per pathogen and cCRE type
    p_val        = if (n_shared >= 1 && n_not_shared >= 1) {
      stats::wilcox.test(
        n_total_cre_type_human[status == "Shared"],
        n_total_cre_type_human[status == "Not-shared"]
      )$p.value
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    # Format p-values into adjusted FDR scientific notation strings
    fdr_p_val      = stats::p.adjust(p_val, method = "BH"),
    fdr_wilcox_sci = sprintf("%.2e", fdr_p_val),
    
    # Construct clean label for facet headers
    facet_label    = stringr::str_c(type_human, "\n(p = ", fdr_wilcox_sci, ")")
  )

# --------
# Step 2: Render boxplot comparison plot with p-values in nested facet titles
# --------
cres_type_comparison_plot_pathogen <- cres_type_homology_comparison_dge_legs_stats %>%
  dplyr::filter(
    n_type_cat != "0",
    status %in% c("Shared", "Not-shared")
  ) %>%
  # Attach pre-calculated p-value labels
  dplyr::inner_join(
    cres_stat_results_pathogen %>% dplyr::select(pathogen, type_human, facet_label),
    by = c("pathogen", "type_human")
  ) %>%
  dplyr::mutate(
    status      = factor(status, levels = c("Shared", "Not-shared")),
    facet_label = forcats::fct_reorder(facet_label, as.numeric(factor(type_human, levels = c("PLS", "pELS", "dELS"))))
  ) %>%
  ggplot2::ggplot() +
  ggplot2::aes(
    x     = status,
    y     = n_total_cre_type_human,
    color = status,
    fill  = status
  ) +
  ggh4x::facet_nested_wrap(
    ~ pathogen + facet_label,
    ncol   = 3,
    scales = "free_y", axes = "all"
  ) +
  ggplot2::geom_boxplot(
    width    = 0.3,
    alpha    = 1,
    outliers = FALSE,
    color    = "black"
  ) +
  ggplot2::scale_fill_manual(values = colors$comparison) +
  ggplot2::scale_color_manual(values = colors$comparison) +
  ggplot2::labs(
    x     = "Status",
    y     = "Total cCREs per Gene",
    color = "",
    fill  = ""
  ) +
  theme_vaxgo() +
  ggplot2::theme(
    legend.position = "top",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

# Render plot output
print(cres_type_comparison_plot_pathogen)

```


CRE ctcftype per pathogen
```{r fig.height=8, fig.width=8}
# Step 1: Pre-calculate statistical p-values per pathogen and human cCRE type
cres_ctcf_stat_results_pathogen <- cres_type_homology_comparison_dge_legs_stats %>%
  dplyr::filter(
    n_type_cat != "0",
    status %in% c("Shared", "Not-shared")
  ) %>%
  dplyr::group_by(pathogen, type_human, ctcf_human) %>%
  dplyr::summarise(
    n_shared     = sum(status == "Shared", na.rm = TRUE),
    n_not_shared = sum(status == "Not-shared", na.rm = TRUE),
    
    # Calculate Wilcoxon test p-value per pathogen and cCRE type
    p_val        = if (n_shared >= 1 && n_not_shared >= 1) {
      stats::wilcox.test(
        n_total_cre_type_human[status == "Shared"],
        n_total_cre_type_human[status == "Not-shared"]
      )$p.value
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    # Format p-values into adjusted FDR scientific notation strings
    fdr_p_val      = stats::p.adjust(p_val, method = "BH"),
    fdr_wilcox_sci = sprintf("%.2e", fdr_p_val),
    
    # Construct clean label for facet headers
    facet_label    = stringr::str_c("p = ", fdr_wilcox_sci)
  )

# Step 2: Render boxplot comparison plot with p-values in nested facet titles
cres_type_comparison_ctcf_plot_pathogen <- cres_type_homology_comparison_dge_legs_stats %>%
  dplyr::filter(
    n_type_cat != "0",
    status %in% c("Shared", "Not-shared")
  ) %>%
  # Attach pre-calculated p-value labels
  dplyr::inner_join(
    cres_ctcf_stat_results_pathogen %>% dplyr::select(pathogen, type_human, ctcf_human, facet_label),
    by = c("pathogen", "type_human", "ctcf_human")
  ) %>%
  dplyr::mutate(
    status      = factor(status, levels = c("Shared", "Not-shared")),
    type_human = fct_relevel(type_human, "PLS", "pELS", "dELS")
  ) %>%
  ggplot2::ggplot() +
  ggplot2::aes(
    x     = status,
    y     = n_total_cre_type_human,
    color = status,
    fill  = status
  ) +
  ggh4x::facet_nested_wrap(
    ~ treatment+ pathogen + type_human + ctcf_human + facet_label,
    nrow   = 4,
    scales = "free_y",
    axes = "all"
  ) +
  ggplot2::geom_boxplot(
    width    = 0.3,
    alpha    = 1,
    outliers = FALSE,
    color    = "black"
  ) +
  ggplot2::scale_fill_manual(values = colors$comparison) +
  ggplot2::scale_color_manual(values = colors$comparison) +
  ggplot2::labs(
    x     = "Status",
    y     = "Total cCREs per Gene",
    color = "",
    fill  = ""
  ) +
  theme_vaxgo() +
  ggplot2::theme(
    legend.position = "top",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

# Render plot output
print(cres_type_comparison_ctcf_plot_pathogen)

```



Distribution of diffLog2FC, by CREs, by gene, by LEGs
```{r fig.height=8, fig.width=12}
# #Plot
# cres_type_homology_comparison_dge_legs_stats %>%
#   filter(n_type_cat != "0") %>%
# filter(status %in% c("Shared", "Not-shared")) %>%
#   ggplot() +
#   aes(x = n_type_cat,
#       y = mean_log2fc_human,
#       # label = hgnc_symbol,
#       fill =  status,
#       color = status
#       ) +
#   facet_nested_wrap(~type_human+ 
#                       # ctcf_human + 
#                       treatment+ pathogen, nrow = 3,
#                     scales = "free") +
#   geom_hline(yintercept = 0, color = "black", linetype = 2) +
#     geom_point(position = position_jitterdodge(jitter.width = 0.2, 
#                                     dodge.width = 0.9),
#                mapping = aes(color = status),
#                size = 0.5,
#                alpha = 0.5) +
  # geom_violin(alpha = 0.1, ) +
    # geom_boxplot(width = 0.7,
    #            alpha = 0.5,
    #            outliers = F) +
#   scale_fill_manual(values = colors$comparison) +
#   scale_color_manual(values = colors$comparison) +
#   labs(x = "N",
#        y = "Human Log2FC",
#        fill = "",
#        color = "") +
#   theme_vaxgo() +
#   theme(legend.position = "top")

```

What is the distribution of absolute diffLog2FC, by CREs, by gene, by LEGs
https://pubmed.ncbi.nlm.nih.gov/29180706/#&gid=article-figures&pid=figure-2-uid-1
```{r fig.height=8, fig.width=12}
#Plot
cres_type_homology_comparison_dge_legs_stats %>% 
  filter(n_type_cat != "0") %>%
  filter(status %in% c("Shared", "Not-shared")) %>%
  ggplot() +
  aes(x = n_type_cat,
      y = abs(mean_log2fc_diff),
      color = status,
      fill =  status,
      ) +
  facet_nested_wrap(~type_human+treatment+pathogen, 
                    nrow = 3, axes = "all", scales = "free") +
  geom_point(alpha = 0.2,
             size = 0.5,
            position = position_jitterdodge(jitter.width = 0.3, 
                                    dodge.width = 0.75),) +
  geom_boxplot(width = 0.7,
               alpha = 0.5,
               outliers = F) +
  # STATISTICAL TEST LAYER
  stat_compare_means(
    aes(group = status),
    method = "wilcox.test",
    label = "p.signif",
    symnum.args = list(
      cutpoints = c(0, 0.001, 0.01, 0.05, 1),
      symbols = c("***", "**", "*", "ns")
    ),
    label.y.np = "top",
    vjust = 1.2,        
    hide.ns = T
  ) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(x = "N",
       y = expression("|" * Delta * "Log"[2] * "FC|"),
       fill = "",
       color = "") +
  theme_vaxgo() +
  theme(legend.position = "top")

```


```{r fig.height=2.5, fig.width=12}
#Plot
#Plot
cres_type_homology_comparison_dge_legs_stats %>% 
  filter(n_type_cat != "0") %>%
  filter(status %in% c("Shared", "Not-shared")) %>%
  # Remove os 10% maiores valores do eixo Y
  filter(n_total_cre_type_human <= quantile(n_total_cre_type_human, 0.9, na.rm = TRUE)) %>%
  ggplot() +
  aes(x = n_total_cre_type_human,
      y = abs(mean_log2fc_diff),
      color = status,
      fill =  status,
      ) +
  facet_nested_wrap(~type_human, 
                    nrow = 1, axes = "all", scales = "free") +
  geom_point(alpha = 0.2,
             size = 0.5,
            position = position_jitterdodge(jitter.width = 0.3, 
                                    dodge.width = 0.75),) +
    geom_smooth(method = "lm", se = T, 
              data = cres_type_homology_comparison_dge_legs_stats %>% 
                # Remove os 10% maiores valores do eixo Y
                filter(n_total_cre_type_human <= quantile(n_total_cre_type_human, 0.95, na.rm = TRUE)) %>%
                mutate(status = "All") %>% 
                filter(n_type_cat != "0")) +
  geom_smooth(method = "lm", se = T) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  scale_x_continuous(expand = c(0.01, 0.01), 
                     # transform = "log10"
                     ) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  labs(x = "N",
       y = expression("|" * Delta * "Log"[2] * "FC|"),
       fill = "",
       color = "") +
  theme_vaxgo() +
  theme(legend.position = "top")


```




```{r fig.height=3, fig.width=12}
#Plot
cres_type_homology_comparison_dge_legs_stats %>% 
  filter(n_type_cat != "0") %>%
  filter(status %in% c("Shared", "Not-shared")) %>%
  # Remove os 10% maiores valores do eixo Y
  # filter(n_total_cre_ctcf_human <= quantile(n_total_cre_ctcf_human, 0.90, na.rm = TRUE)) %>%
  ggplot() +
  aes(x = n_total_cre_ctcf_human,
      y = abs(mean_log2fc_diff),
      color = status,
      fill =  status,
      ) +
  facet_nested_wrap(~type_human+ctcf_human, 
                    nrow = 1, axes = "all", scales = "free") +
  geom_point(alpha = 0.2,
             size = 0.5,
            position = position_jitterdodge(jitter.width = 0.3, 
                                    dodge.width = 0.75),) +
    geom_smooth(method = "lm", se = T, 
              inherit.aes = F,
              mapping = aes(x = n_total_cre_ctcf_human, 
                            y = abs(mean_log2fc_diff)),
              data = cres_type_homology_comparison_dge_legs_stats %>% 
                # Remove os 10% maiores valores do eixo Y
  # filter(n_total_cre_ctcf_human <= quantile(n_total_cre_ctcf_human, 0.90, na.rm = TRUE)) %>%
                mutate(status = "All") %>% 
                filter(n_type_cat != "0"),
      color = "black",
      fill = "black") +
  geom_smooth(method = "lm", se = T) +

  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  scale_x_continuous(expand = c(0.01, 0.01), 
                       limits = c(0, 20)
                       # transform = "log10"
                       ) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  labs(x = "N",
       y = expression("|" * Delta * "Log"[2] * "FC|"),
       fill = "",
       color = "") +
  theme_vaxgo() +
  theme(legend.position = "top")


```
```{r fig.height=4, fig.width=12}
#Plot
cres_type_homology_comparison_dge_legs_stats %>% 
  filter(n_type_cat != "0") %>%
  filter(status %in% c("Shared", "Not-shared"),
         treatment != "Vaccination") %>%
  ggplot() +
  aes(x = n_total_cre_type_human,
      y = abs(mean_log2fc_diff),
      color = status,
      fill =  status,
      ) +
  facet_nested_wrap(~treatment + pathogen+type_human, 
                    nrow = 2, axes = "all", scales = "free") +
    geom_smooth(method = "lm", se = T, 
              inherit.aes = F,
              mapping = aes(x = n_total_cre_type_human, 
                            y = abs(mean_log2fc_diff)),
              data = cres_type_homology_comparison_dge_legs_stats %>% 
                filter(treatment != "Vaccination") %>% 
                mutate(status = "All") %>% 
                filter(n_type_cat != "0"),
      color = "black",
      fill = "black") +
  geom_smooth(method = "lm", se = T) +

  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  scale_x_continuous(expand = c(0.01, 0.01), 
                       limits = c(0, 20)
                       # transform = "log10"
                       ) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(0, NA)) +
  labs(x = "N",
       y = expression("|" * Delta * "Log"[2] * "FC|"),
       fill = "",
       color = "") +
  theme_vaxgo() +
  theme(legend.position = "top")


```


```{r fig.height=8, fig.width=10}
#Plot
cres_type_homology_comparison_dge_legs_stats %>% 
  filter(n_type_cat != "0") %>%
  filter(status %in% c("Shared", "Not-shared"),
         treatment != "Vaccination") %>%
  # Remove os 10% maiores valores do eixo Y
  # filter(n_total_cre_ctcf_human <= quantile(n_total_cre_ctcf_human, 0.90, na.rm = TRUE)) %>%
  ggplot() +
  aes(x = n_total_cre_ctcf_human,
      y = abs(mean_log2fc_diff),
      color = status,
      fill =  status,
      ) +
  facet_nested_wrap(~treatment + pathogen+type_human+ctcf_human, 
                    nrow = 4, axes = "all", scales = "free") +
    geom_smooth(method = "lm", se = T, 
              inherit.aes = F,
              mapping = aes(x = n_total_cre_ctcf_human, 
                            y = abs(mean_log2fc_diff)),
              data = cres_type_homology_comparison_dge_legs_stats %>% 
                filter(treatment != "Vaccination") %>% 
                mutate(status = "All") %>% 
                filter(n_type_cat != "0"),
      color = "black",
      fill = "black") +
  geom_smooth(method = "lm", se = T) +

  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  scale_x_continuous(expand = c(0.01, 0.01), 
                       limits = c(0, 20)
                       # transform = "log10"
                       ) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(0, NA)) +
  labs(x = "N",
       y = expression("|" * Delta * "Log"[2] * "FC|"),
       fill = "",
       color = "") +
  theme_vaxgo() +
  theme(legend.position = "top")


```








------------------------------------------------------------------------

### Gene annotation

```{r import-gene-annotation}

###############
# ADAPT
###############

# Import gene expression correlation
gene_info_human_log2fc_stats = readRDS(file = here("tables", "gene_info_human_log2fc_stats.rds"))
############## 
# Query biomaRt to map human Ensembl IDs to HGNC and MGI symbols
library(biomaRt)
human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

hgnc_mgi_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = unique(gene_info_human_log2fc_stats$gene_id_clean),
  mart       = human_mart
)

# Join HGNC symbols to gene table
genes_cres_human <- genes_cres_human %>%
  left_join(hgnc_mgi_map, by = c("gene_id_clean" = "ensembl_gene_id"))


# Compute multi-CRE metrics per homologous gene pair
gene_level_cre_metrics <- genes_cres_human %>%
  group_by(hgnc_symbol, type_cre) %>%
  summarise(
    n_cres_human       = n(),
    n_cres_orthologous = sum(id_ccre %in% cres_homologous_1to1$human_enhancer),
    pct_cres_conserved = n_cres_orthologous / n_cres_human,
    .groups = "drop"
  ) 
  # inner_join(all_human_mouse_dge_btm_process_genes_correlation_bygene, by = c("hgnc_symbol" = "hgnc_symbol"))


# Annotate cres_annotated with human gene info (one gene per CRE)
cres_annotated <- cres_annotated %>%
  left_join(
    genes_cres_human %>%
      dplyr::select(id_ccre, gene_id, hgnc_symbol,                     transcript_id, chrom_gene, start_gene, end_gene, strand_gene) %>%
      distinct(id_ccre, .keep_all = TRUE),
    by = c("human_enhancer" = "id_ccre")
  )

```



(DOES NOT WORK) Get evolutionary distance

```{r}
# library(biomaRt)
# 
# # 1. Definir o banco de dados antigo usando o host do archive desejado (ex: jan2020)
# # Como você quer partir do camundongo para buscar os homólogos humanos, usamos "mmusculus_gene_ensembl"
# listEnsemblArchives()
# listMarts()
# 
# options(timeout = 600)
# 
# # ensembl_human_old = useEnsembl(biomart = "ensembl",
# #                        dataset="hsapiens_gene_ensembl",
# #                        version="77")
# 
# ensembl_human_old = useMart(biomart = "ENSEMBL_MART_ENSEMBL",
#                              dataset="hsapiens_gene_ensembl",
#                             host = "https://oct2014.archive.ensembl.org")
# 
# listFilters(ensembl_human_old) %>% View()
# listAttributes(ensembl_human_old) %>% View()
# 
# # 2. Carregar sua lista de IDs de genes de camundongo (ex: ENSMUSG0000XXXXXXXX)
# human_genes_list <- all_human_mouse_dge_btm_process_genes_correlation_bygene$symbol
# 
# # 3. Fazer a busca trazendo o ID do camundongo e os dados do homólogo humano (hsapiens)
# options(timeout = 600)
# human_mouse_orthologs_old <- getBM(attributes = c(
#                                                   "ensembl_gene_id",
#                                                   "external_gene_name"
#                                           # "mmusculus_homolog_dn", 
#                                           # "mmusculus_homolog_ds"
#                                           ), 
#                             filters = 'external_gene_name', 
#                             values = "CD6", 
                            # mart = ensembl_human_old)

```










------------------------------------------------------------------------

####  Sequence extraction

The HOXD55 matrix (originally derived by Chiaromonte et al., 2002, and deeply analyzed in Blanchette et al., 2004) defines nucleotide matches and mismatches using large integers to maximize sensitivity when calculating neutral substitution rates in mammalian alignments:Match: $+91$Mismatch: $-114$Because a single mismatch penalizes the score by $-114$, the gap penalties must be scaled proportionally. If you used a standard gap opening penalty of $-10$, the alignment algorithm would erroneously introduce hundreds of micro-gaps instead of accepting a single nucleotide mismatch, completely ruining the biological alignment of your cis-regulatory elements (CREs).

To maintain perfect parity with the alignments you see on the UCSC Genome Browser for human (hg38) and mouse (mm10), you use their default parameters:Gap Opening Penalty ($O$): $-400$Gap Extension Penalty ($E$): $-30$

Chiaromonte, F., et al. (2002). Scoring pairwise genomic alignments. Pacific Symposium on Biocomputing. This paper established the ancestral mammalian repeat substitution matrices (including the HOXD matrices). Blanchette, M., et al. (2004). Blastz: A tool for aligning a large number of long genomic sequences. Genome Research. This paper formally documents the open-source BLASTZ/LASTZ defaults ($O = 400, E = 30$, match $= 91$, mismatch $= 114$) used by UCSC. Kent, W. J., et al. (2003). Evolution's Cauldron: Duplication, Deletion, and Rearrangement in the Mouse and Human Genomes. PNAS. Documents how UCSC constructs the chain and net alignments utilizing these exact parameters.

Align homolog enhancer sequences

```{r extract-sequences}
#Filter homologs with mapped coordinates
cres_to_align_homologous <- cres_homologous_1to1 %>%
  filter(human_enhancer %in% genes_cres_human$id_ccre) %>%
  filter(!is.na(chrom_human), !is.na(chrom_mouse)) %>%
  #Annotate gene symbol
  inner_join(cres_annotated %>% 
            select(human_enhancer, mouse_enhancer, hgnc_symbol),
        by = join_by(human_enhancer, mouse_enhancer))

#Convert to GRanges objects
gr_human <- GRanges(seqnames = cres_to_align_homologous$chrom_human, ranges = IRanges(start = cres_to_align_homologous$start_human, end = cres_to_align_homologous$end_human))
gr_mouse <- GRanges(seqnames = cres_to_align_homologous$chrom_mouse, ranges = IRanges(start = cres_to_align_homologous$start_mouse, end = cres_to_align_homologous$end_mouse))

#Get sequences
cres_to_align_homologous$sequence_human <- as.character(getSeq(BSgenome.Hsapiens.UCSC.hg38, gr_human))
cres_to_align_homologous$sequence_mouse <- as.character(getSeq(BSgenome.Mmusculus.UCSC.mm10, gr_mouse))

#Clean sequences from undefined characters
cres_to_align_homologous_clean = cres_to_align_homologous %>% 
  mutate(sequence_human = str_remove_all(sequence_human, "N"),
         sequence_mouse = str_remove_all(sequence_mouse, "N"))

# Parameterize a nucleotide substitution matrix mirroring the standard HOXD55 values
matriz_hoxd55 <- nucleotideSubstitutionMatrix(match = 91, mismatch = -114, baseOnly = TRUE)

#Align sequences
aln_cres <- pairwiseAlignment(
  DNAStringSet(cres_to_align_homologous_clean$sequence_human),
  DNAStringSet(cres_to_align_homologous_clean$sequence_mouse),
  type         = "local",
  substitutionMatrix = matriz_hoxd55,
  gapOpening   = -400, # Score scale adjusted for standard UCSC genomic alignment scoring
  gapExtension = -30
)

#Calculate percent identity
cres_to_align_homologous_identity <- cres_to_align_homologous_clean %>%
  mutate(
    cre_identity_pct = pid(aln_cres, type = "PID1"),
    cre_align_score  = score(aln_cres),
    cre_align_length = nchar(alignedPattern(aln_cres)),
    cre_coverage_human = cre_align_length / nchar(sequence_human)
  )



#Save results
cres_to_align_homologous_identity %>% 
  saveRDS(file = here("tables", "cres_homologous_1to1_annotated_aligned.rds"))

#Summarize alignment results
summary(cres_to_align$cre_identity_pct)

cat("\n% identity > 80%:", round(mean(cres_to_align$cre_identity_pct > 80) * 100, 1), "%\n")
cat("% identity > 70%:", round(mean(cres_to_align$cre_identity_pct > 70) * 100, 1), "%\n")
cat("% identity > 60%:", round(mean(cres_to_align$cre_identity_pct > 60) * 100, 1), "%\n")

#Which genes have the most conserved enhancers?
cres_to_align %>% 
  select(hgnc_symbol, human_enhancer, mouse_enhancer, type_human, type_mouse, cre_identity_pct:cre_coverage_human) %>% 
  group_by(hgnc_symbol, type_human, type_mouse) %>% 
  summarise(median_pct_id = median(cre_identity_pct)) %>%
  arrange(-median_pct_id) 

```

####  Alignment of proximal TSS windows

```{r}
tx_human <- transcripts(TxDb.Hsapiens.UCSC.hg38.knownGene)

# Multi-scale biological window definitions centered on the TSS
windows_definition <- list(
  w3000 = list(up = 3000, down = 300),  # Full proximal regulatory landscape (pELS/dELS capture)
  w1500 = list(up = 1500, down = 300),  # Core promoter + immediate proximal enhancers
  w500  = list(up = 500,  down = 100),  # Core promoter locus only
  w200  = list(up = 200,  down = 50)    # Minimal transcription initiation footprint
)

# Exemplary extraction and global alignment (Needleman-Wunsch) loop initialization for w3000
tss_human_w3000 <- promoters(TxDb.Hsapiens.UCSC.hg38.knownGene, upstream = 3000, downstream = 300)
tss_mouse_w3000 <- promoters(TxDb.Mmusculus.UCSC.mm10.knownGene, upstream = 3000, downstream = 300)

# Implementation Note: Filter your GRanges objects by matched homologous pairs prior to running global alignment:
aln_global <- pairwiseAlignment(seq_human_gene, seq_mouse_gene, type = "global", substitutionMatrix = matriz_hoxd55)
```


```{r}
ah <- AnnotationHub()

rmsk_query <- query(ah, c("RepeatMasker", "Homo sapiens", "hg38"))
rmsk_hg38  <- ah[[names(rmsk_query)[1]]]

# 2. Compute the exact fraction of the CRE masked by repetitive components
gr_cres_human_all <- GRanges(
  seqnames = cres_to_align$chrom_human, 
  ranges = IRanges(start = cres_to_align$start_human, end = cres_to_align$end_human)
)

hits_rmsk <- findOverlaps(gr_cres_human_all, rmsk_hg38)

# Calculate coverage to map base-pair level masking within the ranges
repeat_cov <- GenomicRanges::coverage(rmsk_hg38)
seqlevels(gr_cres_human_all, pruning.mode = "coercion") <- names(repeat_cov)
masked_views <- Views(repeat_cov, ranges(gr_cres_human_all))

cres_to_align$repeat_masked_pct <- sapply(masked_views, function(x) sum(as.numeric(x) > 0)) / width(gr_cres_human_all) * 100
```


## Transcription factors
### Import, merge, and summarise data 
Annotate with gene regulation - TF Identify conserved Transcription
Factors (TFs) regulating specific genes in both Humans and Mice

```{r}
#Import ChEA3 ===============================

# Clean and parse the ChEA3 transcription factor–target gene matrix
gene_attribute_chea <- read_tsv(here("tables", "gene_attribute_matrix.txt"))

# Transform the wide ChEA matrix into long format, keeping only positive scores
gene_attribute_chea_clean <- gene_attribute_chea %>%
  rename(gene = Gene) %>%
  pivot_longer(-gene, names_to = "experiment", values_to = "score") %>%
  filter(score > 0) %>%
  arrange(desc(score)) %>%
  # Split the experiment metadata into TF name, PMID, and method/tissue/species
  separate(experiment,
           into = c("tf", "pmid", "method_tissue_species"),
           sep = "-",
           extra = "merge",
           fill = "right") %>%
  # Detect species from the metadata string (HUMAN or MOUSE)
  mutate(across(where(is.character), str_trim),
         species = case_when(str_detect(method_tissue_species, "HUMAN") ~ "Human",
                              str_detect(method_tissue_species, "MOUSE") ~ "Mouse",
                              TRUE ~ method_tissue_species)) %>%
  filter(species %in% c("Human", "Mouse"))

# Count evidence per (gene, TF) pair across species
# For each TF–target pair, count how many PMIDs support it in Human and in Mouse
gene_attribute_tf_count <- gene_attribute_chea_clean %>%
  ungroup() %>%
  select(gene, tf, pmid, species) %>%
  group_by(gene, tf) %>%
  pivot_wider(names_from = "species",
              values_from = "pmid") %>%
  mutate(
    Human_pmids = map_chr(Human, ~ paste(.x, collapse = ", ")),
    Mouse_pmids = map_chr(Mouse, ~ paste(.x, collapse = ", ")),
    Human_pmids_n = map_int(Human, length),
    Mouse_pmids_n = map_int(Mouse, length)
  ) %>%
  select(!c(Human, Mouse)) %>%
  mutate(across(ends_with("pmids"), ~ na_if(.x, ""))) %>%
  # Flag whether this TF regulates this gene in both organisms
  mutate(both = if_else(!is.na(Human_pmids) & !is.na(Mouse_pmids), "TRUE", "FALSE"))

# Save the cross-species TF-target count for downstream analysis
saveRDS(gene_attribute_tf_count,
        file = here("tables", "regulation_gene_tf_count.rds"))


# Summarise table =====================
# Reload the TF-count data and gene DGE stats
gene_attribute_tf_count <- readRDS(file = here("tables", "regulation_gene_tf_count.rds"))

#Get percentages
tf_percentages <- gene_attribute_tf_count %>%
  dplyr::arrange(dplyr::desc(both)) %>%
  dplyr::select(gene, both) %>% 
  dplyr::distinct() %>% 
  janitor::tabyl(gene, both) %>%
  janitor::adorn_percentages("row") %>%
  as.data.frame() %>%
  dplyr::rename(
    pct_tf_unique = `FALSE`,
    pct_tf_shared = `TRUE`
  ) %>% 
  dplyr::mutate(
    pct_tf_unique = round(pct_tf_unique * 100, 2),
    pct_tf_shared = round(pct_tf_shared * 100, 2)
  )

# Compute raw counts table
tf_counts <- gene_attribute_tf_count %>%
  dplyr::arrange(dplyr::desc(both)) %>%
  dplyr::select(gene, both) %>% 
  dplyr::distinct() %>% 
  janitor::tabyl(gene, both) %>%
  as.data.frame() %>%
  dplyr::rename(
    n_tf_unique = `FALSE`,
    n_tf_shared = `TRUE`
  )

# Merge 
gene_tf_shared_unique <- tf_percentages %>%
  dplyr::left_join(tf_counts, by = "gene") %>% 
  mutate(n_tf_total = (n_tf_unique + n_tf_shared))

# Print
print(gene_tf_shared_unique)

#Save
gene_tf_shared_unique %>% 
  saveRDS(file = here("tables", "gene_tf_shared_unique.rds"))

# Join the TF regulation table with the gene alignment quadrants
gene_attribute_tf_count_log2fc_stats <- gene_tf_shared_unique %>%
  inner_join(gene_info_human_log2fc_stats ,
             by = join_by(gene))

```

### Plot Total of TFs by species

Scatterplot of total unique TFs per gene in Human vs Mouse. Points are
coloured by classification (Shared vs All). A diagonal line indicates
equal TF counts between species.

```{r fig.height=3, fig.width=6}
#Import
#### Annotate with gene expression and LEGs
#TFs
gene_attribute_tf_count <- readRDS(file = here("tables", "regulation_gene_tf_count.rds"))
gene_tf_shared_unique = readRDS(file = here("tables", "gene_tf_shared_unique.rds"))
dge_btm_process_genes_diff_bygene_clean_filtered =  readRDS(file = here("tables", "dge_btm_process_genes_diff_bygene_clean_filtered.rds"))
  
  
  # Inner join with cross-species differential expression statistics =============
gene_attribute_tf_count_log2fc_stats <- 
  
  bind_rows( 
    
  gene_tf_shared_unique %>%
               rename(hgnc_symbol = gene) %>% 
   #Annotate with gene expression
  inner_join(dge_btm_process_genes_diff_bygene_clean_filtered,
             by = join_by(hgnc_symbol)) %>% 
  #Annotate with LEG classification
  left_join(gsea_btm_legs_clean, 
             by = join_by(hgnc_symbol, pathogen, timepoint_comparison)) %>% 
  mutate(status = case_when(is.na(status)~"Not LEG", 
                            status %in% c("Human only", "Mouse only") ~ "Not-shared",
                            T~status)),
  
  #All genes
  gene_tf_shared_unique %>%
    rename(hgnc_symbol = gene) %>% 
  inner_join(dge_btm_process_genes_diff_bygene_clean_filtered,
             by = join_by(hgnc_symbol)) %>% 
  mutate(status = "All genes")
    )
```


```{r fig.height=3, fig.width=8}
# Compare total unique TF counts per gene between human and mouse
#Plot
# Custom labels for the facet headers
variable_names <- c(
  "pct_tf_shared" = "Shared TFs (%)",
  "n_tf_shared"   = "N of Shared TFs",
  "n_tf_total"    = "Total TFs"
)

gene_attribute_tf_divergence_pooled_plot = gene_attribute_tf_count_log2fc_stats %>% 
  select(hgnc_symbol, pathogen, treatment, status, abs_log2fc_diff, pct_tf_shared, n_tf_shared, n_tf_total) %>% 
  pivot_longer(
    cols = c(pct_tf_shared, n_tf_shared, n_tf_total),
    names_to = "variable", 
    values_to = "value"
  ) %>% 
  mutate(variable = fct_relevel(variable, "n_tf_total", "n_tf_shared", "pct_tf_shared")) %>% 
  # filter(status %in% c("Shared", "Not-shared")) %>%
  
  ggplot(aes(x = value, y = abs_log2fc_diff, 
             color = status, fill = status)) +
  
  # Relabel facets cleanly
  facet_wrap(
    ~variable, 
    scales = "free", 
    nrow = 1,
    labeller = labeller(variable = variable_names)
  ) +
  geom_smooth(method = "lm") +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  scale_y_continuous(limits = c(0, NA), expand = c(0.01, 0.01)) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  
  labs(
    x = NULL,
    y = "|ΔMean Log2FC|",
    color = "LEGs",
    fill = "LEGs",
    title = "Pooled gene expression conditions"
  ) +
  theme_vaxgo() +
  theme(
    legend.position = "top",
    strip.text = element_text(size = 10) 
  )

#Print
gene_attribute_tf_divergence_pooled_plot_marginal = gene_attribute_tf_divergence_pooled_plot +
      # 1. TOP SIDE: Boxplot horizontal no limite superior do eixo Y
  geom_boxplot(
    data = . %>% 
      group_by(variable) %>% 
      mutate(
        max_y = max(abs_log2fc_diff, na.rm = TRUE),
      ),
    mapping = aes(
      x = value,         # Variável contínua no eixo X
      y = 3,             # Posição fixa no topo do eixo Y
      fill = status,
      group = interaction(status, variable),
      width = box_height           # Controla a espessura no eio Y
    ),
    outliers = FALSE, 
    outlier.shape = NA,
    alpha = 0.7,
    linewidth = 0.3,
    width = 0.5,
    color = "black",
    show.legend = FALSE
  ) +
  #Right side
  geom_boxplot(
    data = . %>%
      group_by(variable) %>%
      mutate(
          max_x = max(value, na.rm = TRUE),
        box_width = max_x * 0.2 
      ),
    mapping = aes(
      x = max_x,
      y = abs_log2fc_diff,
      fill = status,
      group = interaction(status, variable),
      width = box_width
    ),
    outliers = F,
    color = "black",
    outlier.shape = NA,
        alpha = 0.7,
    linewidth = 0.3,
    show.legend = FALSE
  ) +
    guides(
    color = "none",
    fill  = guide_legend(title.position = "left", 
                         byrow = T, nrow = 1,
                         override.aes = list(shape = 22, alpha = 1, size = 3,
                                             linetype = 0)))


# Save complete regulation figure
ggsave(gene_attribute_tf_divergence_pooled_plot_marginal,
       filename = here("Figures", "all_human_mouse_regulation_tfs_divergence_plots.png"),
       width = 8, height = 3)
  
```











```{r fig.height=3, fig.width=8}
#Plot
# Custom labels for the facet headers
variable_names <- c(
  "pct_tf_shared" = "Shared TFs (%)",
  "n_tf_shared"   = "N of Shared TFs",
  "n_tf_total"    = "Total TFs"
)

gene_attribute_tf_count_legs_pooled_plot = gene_attribute_tf_count_log2fc_stats %>% 
  select(hgnc_symbol, pathogen, treatment, status, abs_log2fc_diff, pct_tf_shared, n_tf_shared, n_tf_total) %>% 
  pivot_longer(
    cols = c(pct_tf_shared, n_tf_shared, n_tf_total),
    names_to = "variable", 
    values_to = "value"
  ) %>% 
  mutate(variable = fct_relevel(variable, "n_tf_total", "n_tf_shared", "pct_tf_shared")) %>% 
  filter(status %in% c("Shared", "Not-shared", "All genes")) %>%
  
  ggplot(aes(x = status, y = value, color = status, fill = status)) +
  
  # Relabel facets cleanly
  facet_wrap(
    ~variable, 
    scales = "free_y", 
    nrow = 1,
    labeller = labeller(variable = variable_names)
  ) +
  
  geom_violin(alpha = 1, 
              width = 1,
              color = NA) +
  
  # Boxplot centered on each status
  geom_boxplot(
    width = 0.3,
    alpha = 1,
    outliers = FALSE
  ) +
  
  ggpubr::stat_compare_means(
    comparisons = list(c("Shared", "Not-shared"),
                       c("Shared", "All genes")),
    method = "wilcox.test",
    label = "p.signif",
    symnum.args = list(
      cutpoints = c(0, 0.001, 0.01, 0.05, 1),
      symbols = c("***", "**", "*", "ns")
    ),
    size = 3,
    vjust = 0.3,
    hide.ns = F, 
    tip.length = 0.01, 
  ) +

  
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison_color) +
  
  labs(
    x = "Status",
    y = NULL, 
    color = "LEGs",
    fill = "LEGs",
    title = "Pooled gene expression conditions"
  ) +
  theme_vaxgo() +
  theme(
    legend.position = "top", axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    strip.text = element_text(size = 10) 
  )

#Print
gene_attribute_tf_count_legs_pooled_plot
    

#Save
gene_attribute_tf_count_legs_pooled_plot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_gene_attribute_tf_count_legs_pooled_plot.png"),
         height = 3, width = 8)
```














































