---
title: "5_EvolutionaryAnalysis"
output: html_document
---


## Evolutionary analysis
### Annotate with sequence identity

Import human-to-mouse sequence identity data from Ensembl BioMart and
annotate BTM core enrichment genes with homology information. This
annotation enables correlation of sequence conservation with
cross-species transcriptional concordance.

```{r fig.height=5, fig.width=12}
# Download sequence identity data from Ensembl BioMart
# http://www.ensembl.org/biomart/martview/cdbaeb7d08ea2dbc5bae9a1b0d6ad871?VIRTUALSCHEMANAME=default&ATTRIBUTES=hsapiens_gene_ensembl.default.homologs.ensembl_gene_id|hsapiens_gene_ensembl.default.homologs.ensembl_gene_id_version|hsapiens_gene_ensembl.default.homologs.ensembl_transcript_id|hsapiens_gene_ensembl.default.homologs.ensembl_transcript_id_version|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_perc_id_r1|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_perc_id|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_associated_gene_name|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_orthology_type|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_canonical_transcript_protein|hsapiens_gene_ensembl.default.homologs.mmusculus_homolog_ensembl_gene&FILTERS=&VISIBLEPANEL=attributepanel
# Import and annotate

# Import BTM core enrichment results
gsea_btm_results_core_joined =
  readRDS(file = here("tables", "all_human_mouse_gsea_btm_results_core_joined.rds")) %>%
  distinct()

# Import BTM gene-level correlation data
dge_btm_process_genes_correlation_bygene =
  readRDS(file = here("tables", "all_human_mouse_dge_btm_process_genes_correlation_bygene.rds"))

# Import homolog alignment data
all_alignments = read_csv(here("tables", "human_to_mouse_alignment_mart_export.txt"))

# Import BTM gene annotations
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

# Annotate genes with sequence identity and BTM membership
# Build alignment-annotated dataset for core enrichment genes
alignments_human2mouse = all_alignments %>%
  # Clean column names
  clean_names() %>%
  # Keep only high-confidence orthologs
  filter(mouse_orthology_confidence_0_low_1_high == 1) %>%
  # Remove rows without mouse gene names
  drop_na(mouse_gene_name) %>%
  # Select identity, homology type, and gene name columns
  select(gene = mouse_gene_name, mouse_homology_type,
         identity_human2mouse = percent_id_query_gene_identical_to_target_mouse_gene,
         identity_mouse2human = percent_id_target_mouse_gene_identical_to_query_gene) %>%
  # Standardize gene names to uppercase
  mutate(gene = str_to_upper(gene)) %>%
  # Resolve multiple identity scores by taking the median per gene
  group_by(gene) %>%
  summarise(identity_human2mouse = median(identity_human2mouse)) %>%
  ungroup() %>%
  # Filter to only BTM core enrichment genes and append process membership
  inner_join(gsea_btm_results_core_joined,
             by = join_by(gene))

# Reference: all high-confidence orthologs in the database (unfiltered)
alignments_human2mouse_all = all_alignments %>%
  clean_names() %>%
  filter(mouse_orthology_confidence_0_low_1_high == 1) %>%
  drop_na(mouse_gene_name) %>%
  select(gene = mouse_gene_name, mouse_homology_type,
         identity_human2mouse = percent_id_query_gene_identical_to_target_mouse_gene,
         identity_mouse2human = percent_id_target_mouse_gene_identical_to_query_gene) %>%
  mutate(gene = str_to_upper(gene)) %>%
  # Resolve multiple identity scores by taking the median per gene
  group_by(gene) %>%
  summarise(identity_human2mouse = median(identity_human2mouse)) %>%
  ungroup()

# Commented out: biomaRt gene info retrieval (pre-computed and imported below)
# human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
# gene_info_human = getBM(
#   attributes = c("hgnc_symbol", "gene_biotype", "go_id", "namespace_1003", "name_1006"),
#   mart = human_mart,
#   values = alignments_human2mouse_all$gene %>% unique(),
#   filters = "hgnc_symbol"
# ) %>%
#   rename(go_name = name_1006, go_domain = namespace_1003, gene = hgnc_symbol)
# gene_info_human %>% saveRDS(file = here("tables", "gene_info_human.rds"))

# Import pre-computed human gene information and join with alignment data
gene_info_human =
  readRDS(file = here("tables", "gene_info_human.rds")) %>%
  inner_join(alignments_human2mouse_all, by = "gene")
```

Plot correlation distribution by gene

```{r}
dge_btm_process_genes_correlation_bygene %>% 
  ggplot() +
  aes(x = cor_value) +
  geom_density() 
```

Integrate sequence identity with cross-species gene correlation,
organized by functional hierarchy. This prepares data for comparing
sequence conservation with gene-level expression concordance.

```{r fig.height=5, fig.width=12}
# Establish process order based on median sequence identity
process_order_median = alignments_human2mouse %>%
  select(process, group, gene, identity_human2mouse) %>%
  group_by(process) %>%
  summarise(median = median(identity_human2mouse)) %>%
  arrange(median)

# Prepare the detailed process-level dataframe with correlation stats
alignments_human2mouse_process_cor = alignments_human2mouse %>%
  select(process, group, gene, identity_human2mouse) %>%
  mutate(process = fct_relevel(process, process_order_median$process),
         group = fct_relevel(group, group_order_mean$group) %>% fct_rev()) %>%
  distinct() %>%
  # Join with gene-level Spearman correlation results
  left_join(dge_btm_process_genes_correlation_bygene,
             by = join_by(gene == symbol)) %>%
  drop_na(cor_value)

# Create a group-level summary dataset (duplicating group as process)
alignments_human2mouse_group_btm_summary_process = alignments_human2mouse %>%
  select(process, group, gene, identity_human2mouse) %>%
  mutate(group = fct_relevel(group, group_order_mean$group) %>% fct_rev()) %>%
  distinct() %>%
  left_join(dge_btm_process_genes_correlation_bygene,
             by = join_by(gene == symbol)) %>%
  mutate(process = group,
         process = fct_relevel(process, alignments_human2mouse$group %>% unique())) %>%
  drop_na(cor_value)

# Bind both levels into a single dataframe for integrated plotting
alignments_human2mouse_process_cor_withsummary = alignments_human2mouse_process_cor %>%
  bind_rows(alignments_human2mouse_group_btm_summary_process)
```

Compare sequence identity distributions across high-level functional
groups (BTM groups). A violin-boxplot-pointrange composite visualizes
the density, quartiles, and bootstrapped mean of human-to-mouse sequence
identity per group.

```{r fig.height=3, fig.width=4}
# Calculate the mean identity per group to define the plot order
group_order_mean = alignments_human2mouse %>%
  select(process, group, gene, identity_human2mouse) %>%
  group_by(group) %>%
  summarise(mean = mean(identity_human2mouse)) %>%
  arrange(mean)

# Create the distribution plot (Violin + Boxplot + Pointrange)
alignments_human2mouse_pointrange_group_btm = alignments_human2mouse %>%
  select(process, group, gene, identity_human2mouse) %>%
  # Reorder groups based on the previously calculated means
  mutate(group = fct_relevel(group, group_order_mean$group)) %>%
  distinct() %>%
  # Data aesthetics
  ggplot() +
  aes(x = identity_human2mouse,
      y = group,
      fill = group,
      color = group) +
  # Geoms
  geom_violin() +
  geom_boxplot(alpha = 0.1,
               width = 0.1,
               fill = NA,
               color = "white") +
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               size = 1,
               fatten = 0.5,
               color = "black", show.legend = F) +
  # Scales
  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        panel.spacing = unit(0, "lines"),
        plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
        legend.position = "none",
        legend.direction = "horizontal",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, -10, 0),
        legend.text = element_text(size = 10, hjust = 0),
        legend.title = element_text(size = 10),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
  ) +
  # Labels
  labs(x = "% Identity (Human-to-Mouse)",
       y = NULL,
       color = "Group",
       fill = "Group")

# Render plot
alignments_human2mouse_pointrange_group_btm

# Commented out: save to file
# alignments_human2mouse_pointrange_group_btm %>%
#   ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_violinplot_group.png"),
#          width = 12,
#          height = 4)
```

Visualize sequence identity distributions for individual processes
within their functional groups. This per-process view reveals which
specific biological modules have the highest sequence conservation
between human and mouse.

```{r fig.height=18, fig.width=10}
# Generate the detailed per-process sequence identity plot
alignments_human2mouse_pointrange_process_btm = alignments_human2mouse_process_cor_withsummary %>%
  # Data aesthetics
  ggplot() +
  aes(x = identity_human2mouse,
      y = process,
      fill = group,
      color = group) +
  # Facets
  facet_grid(group ~ .,
             scales = "free_y",
             space = "free_y") +
  # Geoms: scatter points and bootstrapped mean CI
  geom_point(color = "gray75",
             alpha = 0.2,
             size = 0.5) +
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               size = 1,
               fatten = 0.5,
               mapping = aes(color = group),
               show.legend = F) +
  # Scales
  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        axis.line.y.right = element_line(color = "black", linetype = 1),
        panel.spacing = unit(0, "lines"),
        plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
        legend.position = "none",
        legend.direction = "horizontal",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, -10, 0),
        legend.text = element_text(size = 8, hjust = 0),
        legend.title = element_text(size = 8),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
  ) +
  # Labels
  labs(x = "% Identity (Human-to-Mouse)",
       y = NULL,
       color = "Group",
       fill = "Group") +
  # Legend layout control
  guides(color = guide_legend(nrow = 2, byrow = T))

# Render plot
alignments_human2mouse_pointrange_process_btm

# Commented out: save to file
# alignments_human2mouse_pointrange_process_btm %>%
#   ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_lineplot_processes.png"),
#          width = 10,
#          height = 18)
```

Correlation of cross-species gene expression concordance (Spearman rho)
across functional groups. A violin-boxplot-pointrange composite
visualizes the distribution of gene-level correlations between human and
mouse within each BTM group.

```{r fig.height=4, fig.width=5}
# Calculate the mean correlation per group to define the plot order
group_order_mean = alignments_human2mouse_process_cor_withsummary %>%
  select(process, group, gene, cor_value) %>%
  group_by(group) %>%
  summarise(mean = mean(cor_value)) %>%
  arrange(mean)

# Create the distribution plot (Violin + Boxplot + Pointrange)
alignments_human2mouse_pointrange_group_btm_correlation = alignments_human2mouse_process_cor_withsummary %>%
  select(process, group, gene, cor_value) %>%
  # Reorder groups based on the previously calculated means
  mutate(group = fct_relevel(group, as.character(group_order_mean$group))) %>%
  distinct() %>%
  # Data aesthetics
  ggplot() +
  aes(x = cor_value,
      y = group,
      fill = group,
      color = group) +
  # Geoms
  geom_violin() +
  geom_boxplot(alpha = 0.1,
               width = 0.1,
               fill = NA,
               color = "white") +
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               size = 1,
               fatten = 0.5,
               color = "black", show.legend = F) +
  geom_vline(xintercept = 0) +
  # Scales
  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  # Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        panel.spacing = unit(0, "lines"),
        plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
        legend.position = "none",
        legend.direction = "horizontal",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, -10, 0),
        legend.text = element_text(size = 10, hjust = 0),
        legend.title = element_text(size = 10),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
  ) +
  # Labels
  labs(x = "Correlation (Human-to-Mouse)",
       y = NULL,
       color = "Group",
       fill = "Group")

# Render plot
alignments_human2mouse_pointrange_group_btm_correlation

# Commented out: save to file
# alignments_human2mouse_pointrange_group_btm %>%
#   ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_violinplot_group.png"),
#          width = 12,
#          height = 4)
```

Visualize sequence identity distributions for individual processes
within their functional groups.

```{r fig.height=18, fig.width=10}
# Get correlation of processes, by correlating their genes values
# Generate the detailed per-process sequence identity plot
alignments_human2mouse_pointrange_process_btm_correlation = alignments_human2mouse_process_cor_withsummary %>% 
    # Reorder groups based on the previously calculated means
  mutate(group = fct_relevel(group, as.character(group_order_mean$group)) %>% fct_rev()) %>% 
  ggplot() +
  aes(x = cor_value,
      y = process,
      fill = group,
      color = group) +
  
  # Vertical faceting by group with proportional spacing
  facet_grid(group~., 
             scales = "free_y", 
             space = "free_y") +
  
  # Boxplot for distribution and Pointrange for mean CI
  # geom_boxplot(alpha = 0) +
    # stat_summary(fun.data = "median_hilow",
    #            geom = "pointrange",
    #            size = 1,
    #            fatten = 0.1, 
    #            color = "gray90",
    #            # mapping = aes(color = group),
    #            show.legend = F) +
  geom_point(color = "gray75",
             alpha = 0.2,
             size = 0.5) +
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               size = 1,
               fatten = 0.5, 
               # color = "black", 
               mapping = aes(color = group),
               show.legend = F) +
  
  # Scales and Colors
  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  
  # theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        axis.line.y.right = element_line(color = "black", linetype = 1),
        panel.spacing = unit(0, "lines"),
        plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,-10,0),
        legend.text = element_text(size = 8, hjust = 0),
        legend.title = element_text(size = 8),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
  ) +
  labs(x = "%Identity (Human-to-Mouse)",
       y = "",
       color = "Group",
       fill = "Group") +
  
  # Legend layout control
  guides(color = guide_legend(nrow = 2, byrow = T)) +
  geom_vline(xintercept = 0)

# Display 
alignments_human2mouse_pointrange_process_btm_correlation

# alignments_human2mouse_pointrange_process_btm %>%
# ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_lineplot_processes.png"),
#          width = 10,
#          height = 18)

```

By group, get important processes

```{r}
correlation_processes_bypathogen_time_df %>% 
  group_by(process) %>% 
  summarise(mean_cor_value = mean(cor_value, ))


alignments_human2mouse_process_cor %>% 
  group_by(group, process) %>% 
  summarise(identity_human2mouse = median(identity_human2mouse, na.rm = T)) %>% 
  inner_join(,
             by = "process")

#Histogram, violin, filtered genes
alignments_human2mouse_processes_filtered = bind_rows(
  
                    #Genes with high positive correlation, high frequency, high identity
                    alignments_human2mouse_all_global_df_nshared %>%
                      filter(p_value <= 0.05) %>% 
                      group_by(status) %>%
                      filter(n > 3) %>% 
                      filter(cor_value >= 0) %>% 
                      filter(identity_human2mouse > 90) %>%
                      arrange(-cor_value, -n) %>%
                      slice_head(n = 5) %>% 
                      mutate(filter = "High correlation,\nHigh %Identity"),
                    
                    
                    alignments_human2mouse_all_global_df_nshared %>%
                      filter(status == "Immune BTMs") %>%
                      slice_max(identity_human2mouse, n = 5,  with_ties = F) %>% 
                      mutate(filter = "High %Identity"),
                    
                    alignments_human2mouse_all_global_df_nshared %>%
                      filter(status == "Non-Immune BTMs") %>%
                      slice_max(identity_human2mouse, n = 5,  with_ties = F) %>% 
                      mutate(filter = "High %Identity"),
                    
                    alignments_human2mouse_all_global_df_nshared %>%
                      filter(status == "All") %>%
                      slice_max(identity_human2mouse, n = 5,  with_ties = F) %>% 
                      mutate(filter = "High %Identity"),
                    
                    #BOTTOM
                    #Genes with high correlation, high frequency, low identity
                    alignments_human2mouse_all_global_df_nshared %>%
                      filter(n > 3) %>% 
                      filter(p_value <= 0.05) %>% 
                      arrange(-cor_value, identity_human2mouse) %>% 
                      group_by(status) %>% 
                      slice_min(identity_human2mouse, n = 5)  %>% 
                      mutate(filter = "High correlation,\nLow %Identity"),
                    
                    
                    alignments_human2mouse_all_global_df_nshared %>%
                      filter(status == "Immune BTMs") %>%
                      slice_min(identity_human2mouse, n = 5,  with_ties = F) %>% 
                      mutate(filter = "Low %Identity"),
                    
                    alignments_human2mouse_all_global_df_nshared %>%
                      filter(status == "Non-Immune BTMs") %>%
                      slice_min(identity_human2mouse, n = 5,  with_ties = F) %>% 
                      mutate(filter = "Low %Identity"),
                    
                    alignments_human2mouse_all_global_df_nshared %>%
                      filter(status == "All") %>%
                      slice_min(identity_human2mouse, n = 5, with_ties = F) %>% 
                      mutate(filter = "Low %Identity"),

                    #Genes with low positive correlation, high frequency, high identity
                    alignments_human2mouse_all_global_df_nshared %>%
                      filter(p_value <= 0.05) %>% 
                      filter(cor_value >= 0) %>% 
                      filter(identity_human2mouse > 90) %>%
                      group_by(status) %>%
                      arrange(cor_value, -n) %>%
                      slice_head(n = 5) %>% 
                      ungroup() %>% 
                      mutate(filter = "Low correlation,\nHigh %Identity"),
                    #Genes with high positive correlation, high frequency, low identity
                    alignments_human2mouse_all_global_df_nshared %>%
                      filter(p_value <= 0.05) %>% 
                      filter(cor_value >= 0) %>% 
                      filter(identity_human2mouse < 90) %>%
                      group_by(status) %>%
                      arrange(-cor_value, -n) %>%
                      slice_head(n = 5) %>% 
                      ungroup() %>% 
                      mutate(filter = "Low correlation,\nLow %Identity")
) %>% 
    mutate(
    status = str_replace(status, " ", "\n"),
    status = fct_relevel(
    status,
    "Human\nonly",
    "Mouse\nonly",
    "Shared",
    "All",
    "Non-Immune\nBTMs",
    "Immune\nBTMs"
  )) 

#Histogram and violin
alignments_human2mouse_processes_histogram_all_global = alignments_human2mouse_all_global_df_filtered %>%
  mutate(analysis = "Distribution of conservation") %>%
  mutate(status = str_replace(status, " ", "\n"),
         status = fct_relevel(
           status,
           "Human\nonly",
           "Mouse\nonly",
           "Shared",
           "All",
           "Non-Immune\nBTMs",
           "Immune\nBTMs"
         )) %>% 
  ggplot() +
  aes(x = status,
      y = identity_human2mouse,
      fill = status,
      color = status,
      group = status) +
  facet_wrap(~analysis) +
  geom_violin(color = "black",
              alpha = 0.7) +
  geom_boxplot(
              # color = "white",
               mapping = aes(color = status),
               width = 0.1,
               fill = NA,
               outliers = F) +
  scale_fill_manual(name = "Status",
                    values = c(colors$comparison,
                               "All" = "white")) +
  scale_color_manual(
    name = "Status",
    values = c(
    "Mouse\nonly" = "white",
    "Human\nonly" = "white",
    "Shared" = "white",
    "Immune\nBTMs" = "white",
    "Non-Immune\nBTMs" = "white",
    "All" = "black")) +
  new_scale_color() + 
  
  scale_color_manual(
     name = "Filter",
     values = c(
    "High %Identity" = "#4361ee",
    "High correlation,\nHigh %Identity" = "#4DBBD5FF",
    "Low correlation,\nHigh %Identity" = "#7209b7",
    
    "Low %Identity" = "red",
    "High correlation,\nLow %Identity" = "#CA6702",
    "Low correlation,\nLow %Identity" = "#AE2012"
  )) +
  geom_point(data = alignments_human2mouse_genes_filtered,
             aes(color = filter),
             pch = 21,
             fill = "white") +
  stat_summary(
    fun = median, 
    geom = "label", 
    aes(label = after_stat(round(y, 1))),
    size = 3,
    hjust = -0.2,   
    color = "black",
    fill = "white",
  ) +
  #Get top and bottom genes
  geom_label_repel(aes(label = gene,
                       color = filter), 
                  max.overlaps = 10000,
                  size = 2, 
                  direction = c("y"),
                  nudge_x = -0.4,
                  hjust = 0,
                  fill = "white", 
                  box.padding = unit(0, "cm"), 
                  label.padding = unit(0.1, "cm"), 
                  data = alignments_human2mouse_genes_filtered) +
  
  scale_x_discrete(expand = c(0.1, 0.1)) +
  scale_y_continuous(expand = c(0.01, 0.01), 
                     limits = c(50, 105)
                     ) +
  theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 10),
          axis.text.y = element_text(size = 10),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0.3, 0.3, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "vertical",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  labs(x = "Status",
       y = "%Identity (Human-to-Mouse)",
       color = "",
       fill = "") 


#Box plot, independent, by process 
alignments_human2mouse_pointrange_process_btm =  alignments_human2mouse_process_cor_withsummary %>% 
  ggplot() +
  aes(x = identity_human2mouse,
      y = process,
      fill = group,
      color = group) +
  facet_grid(group~., 
             scales = "free_y",
             space = "free_y") +
  # geom_violin(alpha = 0.2) +
  geom_boxplot(alpha = 0)+
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               size = 1,
               fatten = 0.5, 
               # mapping = aes(color = group),
               color = "black", show.legend = F)+
  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          axis.line.y.right = element_line(color = "black", linetype = 1),
          panel.spacing = unit(0, "lines"),
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  labs(x = "%Identity (Human-to-Mouse)",
       y = "",
       color = "Group",
       fill = "Group") +
  guides(color = guide_legend(nrow = 2, byrow = T))

alignments_human2mouse_pointrange_process_btm
```

Construct the global comparison dataframe integrating sequence identity,
BTM membership, and functional correlation across all genes. This
dataset classifies genes into Human-only, Mouse-only, Shared, Immune
BTM, Non-Immune BTM, and All categories for downstream conservation
analysis.

```{r fig.height=5, fig.width=6}
# Build the unified dataframe of all gene conservation values across categories
# 1. Construct the Global comparison dataframe (All genes vs BTMs vs Orthologs)
alignments_human2mouse_all_global_df = alignments_human2mouse %>% 
  select(gene:pathogen, timepoint, status) %>% 
  distinct() %>% 
  mutate(gene_set = "Human vs. Mouse") %>% 
  # Bind with specialized BTM categories
  bind_rows(
    alignments_human2mouse %>% 
      mutate(gene_set = "BTMs",
             # Aggregate BTM groups into functional "Immune" and "Non-Immune" blocks
             status = case_when(
               group %in% c("ENERGY METABOLISM", "ECM AND MIGRATION", "CELL CYCLE", 
                            "SIGNAL TRANSDUCTION", "PLATELETS") ~ "Non-Immune BTMs",
               group %in% c("NEUTROPHILS", "T CELLS", "MONOCYTES", "B CELLS", 
                            "INNATE RESPONSE", "NK CELLS", "IFN") ~ "Immune BTMs",
               TRUE ~ status)
      ) %>% 
      select(gene, status, identity_human2mouse, gene_set) %>% 
      distinct(),
    # Bind with the total ortholog reference set
    gene_info_human %>% 
      distinct() %>% 
      mutate(status = "All", gene_set = "All") %>% 
      select(gene, status, identity_human2mouse, gene_set) %>% 
      distinct()
  ) %>% 
  # 2. Relevel status for logical visualization flow
  mutate(status = fct_relevel(status, "Human only", "Mouse only", "Shared", 
                              "Immune BTMs", "ENERGY METABOLISM", "ECM AND MIGRATION", 
                              "CELL CYCLE", "SIGNAL TRANSDUCTION", "PLATELETS", "All")) %>% 
  # 3. Integrate with functional correlation results
  left_join(dge_btm_process_genes_correlation_bygene,
            by = join_by(gene == symbol)) %>% 
  filter(gene_set == "Human vs. Mouse" | !is.na(cor_value))

# 4. Calculate gene frequency (How many conditions/pathogens each gene appears in)
alignments_human2mouse_all_global_df_nshared = alignments_human2mouse_all_global_df %>%
  # Filter to only status categories relevant for sharing analysis
  filter(status %in% c("Mouse only", "Human only", "Shared")) %>% 
  arrange(gene) %>% 
  select(gene, status) %>%
  # Count how many conditions each gene appears in per status category
  group_by(status, gene) %>% 
  summarise(n_conditions_status = n()) %>% 
  ungroup() %>% 
  arrange(gene) %>% 
  # Compute total conditions per gene and derive sharing proportion and ratio
  group_by(gene) %>% 
  mutate(ntotal = sum(n_conditions_status),
         nstatus_to_total_pct = (n_conditions_status/ntotal) %>% round(2),
         nstatus_to_total = paste0(n_conditions_status, "-", ntotal)) %>% 
  # Classify each gene as Shared or Not-shared across conditions
  mutate(is_shared = if_else(status == "Shared", "Shared", "Not-shared")) %>%
  group_by(gene, is_shared) %>%
  mutate(n_group = sum(n_conditions_status)) %>%
  ungroup() %>%
  group_by(gene) %>%
  mutate(
    n_shared = sum(n_conditions_status[status == "Shared"]),
    n_notshared = ntotal - n_shared,
    ratio_shared_notshared_pct = round(n_shared / n_notshared, 2),
    ratio_shared_notshared_pct = if_else(is.infinite(ratio_shared_notshared_pct), NA_real_, ratio_shared_notshared_pct),
    # Classify based on whether shared proportion exceeds not-shared proportion
    class = if_else(ratio_shared_notshared_pct >= 1, "Shared", "Not-shared")
  ) %>%
  ungroup() %>% 
  # Join back the identity and correlation values from the full dataset
  inner_join(alignments_human2mouse_all_global_df %>% 
               select(gene, identity_human2mouse, cor_value:cor_label) %>% 
               distinct(),
             by = "gene") %>% 
  # Append the All, Immune BTM, and Non-Immune BTM categories without sharing classification
  bind_rows(alignments_human2mouse_all_global_df %>%
              filter(status %in% c("All", "Immune BTMs", "Non-Immune BTMs")) %>% 
               select(status,gene, identity_human2mouse, cor_value:cor_label) %>% 
               distinct() %>% 
              mutate(is_shared = status,
                     class = status))

# Save the combined sharing and conservation table to disk
alignments_human2mouse_all_global_df_nshared %>% 
  saveRDS(file = here("tables", "alignments_human2mouse_all_global_df_nshared.rds"))

# 5. Final cleaned reference for identity distribution
alignments_human2mouse_all_global_df_filtered = alignments_human2mouse_all_global_df %>% 
  select(gene, status, identity_human2mouse) %>% 
  distinct()

# Create a filtered dataset of exemplar genes based on identity and correlation
top_shared_notshared_genes = bind_rows(
    # Select top 50 shared genes with highest sharing ratio
    alignments_human2mouse_all_global_df_nshared %>%
    select(gene, 
           class, n_conditions_status, n_shared, ntotal, cor_value, p_value, r_squared, cor_label, identity_human2mouse,  ratio_shared_notshared_pct) %>%
    arrange(-ratio_shared_notshared_pct, -n_shared) %>% 
    filter(class == "Shared",
           n_shared > 3) %>%
      arrange(-ratio_shared_notshared_pct) %>% 
    slice_head(n = 50) %>% 
      select(-n_conditions_status) %>% 
    distinct() %>% 
      arrange(-identity_human2mouse, -cor_value),
    
    # Select top 50 not-shared genes with lowest sharing ratio
    alignments_human2mouse_all_global_df_nshared %>%
    drop_na(cor_value) %>% 
    select(gene, 
           n_shared, class,  n_notshared, identity_human2mouse, cor_value, p_value, r_squared, cor_label, ntotal,ratio_shared_notshared_pct) %>%
    arrange(ratio_shared_notshared_pct, -n_notshared) %>% 
    filter(class == "Not-shared",
           n_notshared > 3) %>%
    arrange(ratio_shared_notshared_pct) %>% 
      slice_head(n = 50) %>% 
    distinct(),
    
    # Include all Immune BTM, Non-Immune BTM, and All category genes
    alignments_human2mouse_all_global_df_nshared %>%
      filter(status %in% c("Immune BTMs", "Non-Immune BTMs", "All")) %>% 
    drop_na(cor_value) %>% 
    select(gene, ntotal, 
           n_shared, class, n_notshared, identity_human2mouse, cor_value, p_value, r_squared, cor_label, ntotal,ratio_shared_notshared_pct) %>%
    distinct()
    ) %>% 
  mutate(n_conditions_is_shared = ntotal)

# Save the top exemplar gene table to disk
top_shared_notshared_genes %>% 
  saveRDS(file = here("tables", "top_shared_notshared_genes.rds"))
```

##### Scatterplots of top shared LEGs

Scatterplots comparing mouse versus human log2 fold changes for the top
shared Leading Edge Genes (LEGs). Each panel shows a single gene across
conditions, with points colored by pathogen and a linear regression line
indicating cross-species concordance.

```{r fig.height=8, fig.width=4}
# Scatterplots of top shared LEGs ----
# Log2FC cross-species comparison for top shared LEGs ----
top_shared_notshared_genes %>%
                 ungroup() %>%
                 filter(class == "Shared") %>% 
  # Join species-specific log2FC values from the wide-format GSEA table
  inner_join(dge_btm_process_genes_wide %>% 
               select(gene = symbol, pathogen, timepoint, Mouse:adj_p_val_Human) %>% 
               distinct(),
             by = "gene") %>%
  # Retain only genes with strong cross-species correlation
  filter(r_squared > 0.2) %>% 
    # Exclude vaccination conditions to focus on infection-only comparisons
    filter(!pathogen %in% c("Influenza", "Hepatitis B")) %>% 
  ggplot() +
    aes(x = Mouse,
      y = Human) +
    # One panel per gene with its correlation label
    facet_wrap(~gene+cor_label, 
             scales = "free",
             ncol = 2) +
  geom_point(aes(fill = pathogen),
             size = 3,
             pch = 21,
             color = "black") +
  # Linear regression line with confidence band
  stat_smooth(method = "lm",
              color = "black",
              fill = "gray",
              alpha = 0.2) +
  # Reference lines at zero
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = 0, linetype = 2) +
  scale_fill_manual(values = colors$pathogen) +
  theme_vaxgo() +
  theme(legend.position = "top") +
    labs(title = "Shared LEGs, R2 > 0.2",
       subtitle = "Without vaccination",
       fill = "Pathogen")





```

Four-panel scatterplot matrix exploring relationships between gene
sharing frequency, sequence identity, and cross-species correlation.
These plots identify which shared LEGs have the strongest conservation
signals across conditions.

```{r fig.height=4, fig.width=20}
library(viridis)

# Plot shared vs not-shared gene count scatter
# 1. Shared vs Not-shared count scatterplot
nshared_shared_plot = (alignments_human2mouse_all_global_df_nshared %>% 
  ungroup() %>% 
  filter(status == "Shared") %>% 
  select(gene, ntotal, ratio_shared_notshared_pct, n_shared, n_notshared, class, identity_human2mouse:cor_label) %>% 
  distinct() %>% 
  arrange(-ratio_shared_notshared_pct) %>% 
  ggplot() +
  aes(x = n_notshared,
      y = n_shared,
      fill = class,
      color = class,
      label = gene) +
  geom_point() +
  geom_point(color = "black",
              pch = 21,
              data = alignments_human2mouse_all_global_df_nshared %>% 
                filter(class == "Shared")) +
  geom_text_repel(data = alignments_human2mouse_all_global_df_nshared %>%
                 ungroup() %>%
                 filter(status == "Shared",
                        class == "Shared"),
             color = "black",
             size = 3) +
  scale_fill_manual(values = c("Not-shared" = "gray90",
                                "Shared" = "#4DBBD5FF")) +
      scale_color_manual(values = c("Not-shared" = "gray90",
                                "Shared" = "#4DBBD5FF")) +
  theme_vaxgo() +
  theme(legend.position = "right") +
  labs(title = "Shared vs. Not-shared",
       x = "Not-Shared",
       y = "Shared",
       fill = "",
       color = "")) 

# Plot the ratio of shared-to-not-shared vs sequence identity
# 2. Sharing ratio vs %Identity scatterplot
ratio_identity_plot = (alignments_human2mouse_all_global_df_nshared %>% 
  ungroup() %>% 
  filter(status == "Shared") %>% 
  select(gene, ntotal, ratio_shared_notshared_pct, n_shared, n_notshared, class, identity_human2mouse:cor_label) %>% 
  distinct() %>% 
  arrange(-ratio_shared_notshared_pct) %>% 
  ggplot() +
  aes(x = ratio_shared_notshared_pct,
      y = identity_human2mouse,
      fill = class,
      color = class,
      label = gene) +
  geom_point() +
  geom_point(color = "black",
              pch = 21,
              data = alignments_human2mouse_all_global_df_nshared %>% 
                filter(class == "Shared")) +
  geom_text_repel(data = alignments_human2mouse_all_global_df_nshared %>%
                 ungroup() %>%
                 filter(status == "Shared",
                        class == "Shared"),
             color = "black",
             size = 3)  +
  scale_fill_manual(values = c("Not-shared" = "gray90",
                                "Shared" = "#4DBBD5FF")) +
      scale_color_manual(values = c("Not-shared" = "gray90",
                                "Shared" = "#4DBBD5FF")) +
  theme_vaxgo() +
    theme(legend.position = "right") +
  labs(title = "Ratio of Shared to Not-shared vs Identity",
       x = "Ratio Shared-to-Not-Shared",
       y = "%Identity",
       fill = "",
       color = ""))  

# 3. Sharing ratio vs cross-species correlation scatterplot
ratio_correlation_plot = (alignments_human2mouse_all_global_df_nshared %>% 
  ungroup() %>% 
  filter(status == "Shared") %>% 
  select(gene, ntotal, ratio_shared_notshared_pct, n_shared, n_notshared, class, identity_human2mouse:cor_label) %>% 
  distinct() %>% 
  arrange(-ratio_shared_notshared_pct) %>% 
  ggplot() +
  aes(x = cor_value,
      y = ratio_shared_notshared_pct,
      fill = -log(p_value),
      color = -log(p_value),
      label = gene) +
  geom_point() +
  geom_point(color = "black",
              pch = 21,
              data = alignments_human2mouse_all_global_df_nshared %>% 
                filter(class == "Shared")) +
  geom_label_repel(data = alignments_human2mouse_all_global_df_nshared %>%
                 ungroup() %>%
                 filter(status == "Shared",
                        class == "Shared"),
             color = "black",
             fill = "white",
             size = 3)  +
  scale_fill_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
  ) +
      scale_color_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
  ) +
    geom_vline(xintercept = 0, linetype = 2) +
  theme_vaxgo() +
   theme(legend.position = "right") +
  labs(title = "Ratio of Shared to Not-shared vs Correlation",
       x = "Correlation",
       y = "Ratio Shared-to-Not-Shared"))

# 4. Sequence identity vs cross-species correlation scatterplot
identity_correlation_plot = (alignments_human2mouse_all_global_df_nshared %>% 
  ungroup() %>% 
  filter(status == "Shared") %>% 
  select(gene, ntotal, ratio_shared_notshared_pct, n_shared, n_notshared, class, identity_human2mouse:cor_label) %>% 
  distinct() %>% 
  arrange(-ratio_shared_notshared_pct) %>% 
  ggplot() +
  aes(x = cor_value,
      y = identity_human2mouse,
      fill = -log(p_value),
      color = -log(p_value),
      label = gene) +
  geom_point() +
  geom_point(color = "black",
              pch = 21,
              data = alignments_human2mouse_all_global_df_nshared %>% 
                filter(class == "Shared")) +
  geom_label_repel(data = alignments_human2mouse_all_global_df_nshared %>%
                 ungroup() %>%
                 filter(status == "Shared",
                        class == "Shared"),
             color = "black",
             fill = "white",
             size = 3)  +
  scale_fill_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
  ) +
      scale_color_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
  ) +
     geom_vline(xintercept = 0, linetype = 2) +
  theme_vaxgo() +
   theme(legend.position = "right") +
  labs(title = "Identity vs Correlation, Shared",
       x = "Correlation",
       y = "%Identity"))

# Combine all four panels into a single composite figure and save
nshared_identity_correlation_plots = (nshared_shared_plot | ratio_identity_plot | identity_correlation_plot | ratio_correlation_plot)

nshared_identity_correlation_plots %>% 
  ggsave(filename = here("plot.png"),
         width = 20,
         height = 4)

```

Create a filtered dataset of exemplar genes based on identity and
correlation

```{r fig.height=5, fig.width=6}
# Build filtered exemplar gene dataset stratified by correlation direction and sequence identity
alignments_human2mouse_genes_filtered = bind_rows(

  # SHARED vs NOT-SHARED CATEGORY
  # Positive correlation, high identity (>90%)
  top_shared_notshared_genes %>%
    filter(p_value <= 0.10,
           cor_value >= 0, 
            r_squared > 0.2,
           ntotal > 3,
           identity_human2mouse > 90) %>% 
    group_by(class) %>%
    arrange(-cor_value, 
            -ntotal) %>%
    slice_head(n = 100) %>% 
    mutate(filter = "Positive correlation,\nHigh %Identity"),
  
  # Positive correlation, low identity (<90%)
  top_shared_notshared_genes %>%
    filter(ntotal > 3, 
           cor_value > 0,
           p_value <= 0.10,
            r_squared > 0.2,
           identity_human2mouse < 90) %>% 
    arrange(-cor_value, identity_human2mouse) %>% 
    group_by(class) %>% 
    slice_head(n = 100)  %>% 
    mutate(filter = "Positive correlation,\nLow %Identity"),

  # Negative correlation, high identity (>90%)
  top_shared_notshared_genes %>%
    filter(ntotal > 3,
           p_value <= 0.10, 
            r_squared > 0.2,
           cor_value < 0, 
           identity_human2mouse > 90) %>% 
    group_by(class) %>%
    arrange(cor_value, -ntotal) %>%
    slice_head(n = 100) %>% 
    ungroup() %>% 
    mutate(filter = "Negative correlation,\nHigh %Identity"),

  # Negative correlation, low identity (<90%)
  top_shared_notshared_genes %>%
    filter(p_value <= 0.10, 
           cor_value < 0, 
           r_squared > 0.2,
           ntotal > 3,
           identity_human2mouse < 90) %>% 
    group_by(class) %>%
    arrange(cor_value, -ntotal) %>%
    slice_head(n = 100) %>% 
    ungroup() %>% 
    mutate(filter = "Negative correlation,\nLow %Identity"),
  
  # BTM CATEGORY (Immune, Non-Immune, All)
  # Positive correlation, high identity (>90%) for BTM sets
  top_shared_notshared_genes %>%
    filter(class %in% c("Immune BTMs", "Non-Immune BTMs", "All"),
           p_value < 0.10,
            r_squared > 0.2,
           identity_human2mouse > 90,
           cor_value > 0,
           ) %>%
    group_by(class) %>%
    arrange(-cor_value, -ntotal) %>% 
    slice_head(n = 100) %>% 
    mutate(filter = "Positive correlation,\nHigh %Identity"),
  
  # Negative correlation, low identity (<90%) for BTM sets
  top_shared_notshared_genes %>%
    filter(class %in% c("Immune BTMs", "Non-Immune BTMs", "All"),
           p_value < 0.10,
           cor_value < 0,
            r_squared > 0.2,
           identity_human2mouse <90,
           ) %>%
    group_by(class) %>%
    arrange(-cor_value, -ntotal) %>% 
    slice_head(n = 100) %>% 
    mutate(filter = "Negative correlation,\nLow %Identity"),

) %>% 
  # Clean labels and reorder factors for consistent plotting
  mutate(
    class = str_replace(class, " ", "\n"),
    class = fct_relevel(class, "Not-shared", "Shared", 
                         "All", "Non-Immune\nBTMs", "Immune\nBTMs"),
    filter = fct_relevel(filter, 
                         "Positive correlation,\nHigh %Identity",
                         "Negative correlation,\nHigh %Identity", 
                         "Positive correlation,\nLow %Identity",
                         "Negative correlation,\nLow %Identity")
    ) 


```

Correlation and identity, shared vs not-shared, filtered by importance
(n and ratio)

```{r fig.height=4, fig.width=16}
# Correlation by shared vs not-shared, filtered by importance
# Build violin-half plot of cross-species correlation across gene categories
correlation_shared_notshared_plot = alignments_human2mouse_all_global_df_nshared %>% 
  ungroup() %>% 
  select(gene, ntotal, ratio_shared_notshared_pct, is_shared, n_shared, n_notshared, class, identity_human2mouse:cor_label) %>% 
  distinct() %>% 
  arrange(-ratio_shared_notshared_pct) %>% 
  mutate(class = str_replace(class, " ", "\n"),
         class = fct_relevel(class, "Not-shared", "Shared", 
                         "All", "Non-Immune\nBTMs", "Immune\nBTMs")) %>% 
  ggplot() +
  aes(x = class,
      y = cor_value,
      label = gene) +
  # Half-violin plot showing distribution density by class
  see::geom_violinhalf(flip = TRUE,
                  alpha = T,
                  mapping = aes(fill = class))+
  geom_point(color = "black",
              alpha = 0.2) +
   # First color scale: category status
  scale_fill_manual(name = "Status", values = c(colors$comparison, "All" = "white")) +
  new_scale_color() + 
  new_scale_fill() + 
  # Second color scale: filter category (correlation x identity)
  scale_color_manual(
     name = "Filter",
     values = c(
                "Positive correlation,\nHigh %Identity" = "#4DBBD5FF",
                "Negative correlation,\nHigh %Identity" = "#4361ee", 
                "Positive correlation,\nLow %Identity" = "#CA6702",
                "Negative correlation,\nLow %Identity" = "#AE2012")
     ) +
  scale_fill_manual(
     name = "Filter",
     values = c(
                "Positive correlation,\nHigh %Identity" = "#4DBBD5FF",
                "Negative correlation,\nHigh %Identity" = "#4361ee", 
                "Positive correlation,\nLow %Identity" = "#CA6702",
                "Negative correlation,\nLow %Identity" = "#AE2012")) +
  # Overlay top exemplar genes colored by their filter category
  geom_point(aes(color = filter,
                 fill = filter),
             alpha = 1,
              data = alignments_human2mouse_genes_filtered %>% 
  group_by(class, filter) %>% 
  slice_head(n = 10)) +
  
  geom_hline(yintercept = 0) +
  
  # Label top exemplar genes with repel to avoid overlap
  geom_label_repel(aes(color = filter),
                   data = alignments_human2mouse_genes_filtered %>% 
  group_by(class, filter) %>% 
  slice_head(n = 10), 
    nudge_x = 0.3,
    fill = "white",
    hjust = 0, 
    direction = "y",
    size = 2.5)  +
  theme_vaxgo() +
   theme(legend.position = "right") +
  labs(title = "Ratio of Shared to Not-shared vs Correlation",
       x = "Group",
       y = "Correlation")

correlation_shared_notshared_plot

# Identity by shared vs not-shared, filtered by importance
# Build violin-half plot of sequence identity across gene categories
identity_shared_not_shared_plot = alignments_human2mouse_all_global_df_nshared %>% 
  ungroup() %>% 
  select(gene, ntotal, ratio_shared_notshared_pct, is_shared, n_shared, n_notshared, class, identity_human2mouse:cor_label) %>% 
  distinct() %>% 
  arrange(-ratio_shared_notshared_pct) %>% 
    mutate(class = str_replace(class, " ", "\n"),
         class = fct_relevel(class, "Not-shared", "Shared", 
                         "All", "Non-Immune\nBTMs", "Immune\nBTMs")) %>% 
  ggplot() +
  aes(x = class,
      y = identity_human2mouse,
      label = gene) +
  see::geom_violinhalf(flip = TRUE,
                  alpha = T,
                  mapping = aes(fill = class))+
  geom_point(color = "black",
              alpha = 0.2) +
   # First color scale: category status
  scale_fill_manual(name = "Status", values = c(colors$comparison, "All" = "white")) +
  new_scale_color() + 
  new_scale_fill() + 
  # Second color scale: filter category (correlation x identity)
   scale_color_manual(
     name = "Filter",
     values = c(
                "Positive correlation,\nHigh %Identity" = "#4DBBD5FF",
                "Negative correlation,\nHigh %Identity" = "#4361ee", 
                "Positive correlation,\nLow %Identity" = "#CA6702",
                "Negative correlation,\nLow %Identity" = "#AE2012")
     ) +
  scale_fill_manual(
     name = "Filter",
     values = c(
                "Positive correlation,\nHigh %Identity" = "#4DBBD5FF",
                "Negative correlation,\nHigh %Identity" = "#4361ee", 
                "Positive correlation,\nLow %Identity" = "#CA6702",
                "Negative correlation,\nLow %Identity" = "#AE2012")) +
  geom_point(aes(color = filter,
                 fill = filter),
             alpha = 1,
              data = alignments_human2mouse_genes_filtered %>% 
  group_by(class, filter) %>% 
  slice_head(n = 10)) +
  
  # Label top exemplar genes with repel to avoid overlap
  geom_label_repel(aes(color = filter),
                   data = alignments_human2mouse_genes_filtered %>% 
  group_by(class, filter) %>% 
  slice_head(n = 10), 
    nudge_x = 0.3,
    fill = "white",
    hjust = 0, 
    direction = "y",
    size = 2.5)  +
  theme_vaxgo() +
   theme(legend.position = "right") +
  labs(title = "Ratio of Shared to Not-shared vs Identity",
       y = "%Identity",
       x = "Group")

identity_shared_not_shared_plot

# Combine both violin-half plots into a single composite figure
correlation_shared_notshared_plot + identity_shared_not_shared_plot +
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "right"))

```

Correlation between variables?

```{r}
# Scatterplot of %Identity vs cross-species Rho (positive correlations only)
alignments_human2mouse_all_global_df %>%
  select(-pathogen, -timepoint) %>%
  distinct() %>% 
  filter(cor_value > 0) %>%
  ggplot() +
  aes(x = identity_human2mouse,
      y = cor_value,
      color = -log(p_value),
      ) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_vaxgo() +
  labs(x = "%Identity",
       y = "Rho")

# Boxplot of correlation by number of pathogens where the gene is shared
alignments_human2mouse_all_global_df %>%
  select(-pathogen, -timepoint) %>% 
  distinct() %>% 
  drop_na(n_pathogens) %>% 
  mutate(n_pathogens_f = fct_reorder(as.character(n_pathogens), n_pathogens)) %>% 
 drop_na(cor_value) %>% 
  filter(p_value <= 0.05) %>%
  filter(cor_value > 0) %>%
  ggplot() +
  aes(x = n_pathogens_f,
      y = cor_value
      ) +
  geom_boxplot(aes(fill = class)) +
  geom_point(aes(color = class))

# Boxplot of %Identity by number of pathogens where the gene is shared
alignments_human2mouse_all_global_df %>%
  select(-pathogen, -timepoint) %>% 
  distinct() %>% 
  drop_na(n_pathogens) %>% 
  mutate(n_pathogens_f = fct_reorder(as.character(n_pathogens), n_pathogens)) %>% 
  drop_na(cor_value) %>% 
  filter(p_value <= 0.05) %>%
  filter(cor_value > 0) %>%
  ggplot() +
  aes(x = n_pathogens_f,
      y = identity_human2mouse,
      color = cor_value
      ) +
  geom_boxplot() +
  geom_point()
```

### Gene biotype

Examine whether gene biotype categories (miRNA, protein-coding,
pseudogenes, lncRNA, snoRNA) differ in their sequence identity to mouse
orthologs. This tests whether conservation level relates to functional
gene class.

```{r fig.height=3, fig.width=4}
# Test whether gene biotype is associated with %Identity
gene_info_human %>% 
  select(hgnc_symbol, gene_biotype, identity_human2mouse) %>% 
  distinct() %>% 
  group_by(gene_biotype) %>% 
  mutate(median_identity = median(identity_human2mouse)) %>% 
  ungroup() %>%
  filter(gene_biotype %in% c("miRNA", "protein_coding",
                             "processed_pseudogene",
                             "snoRNA",
                             "lncRNA",
                             "unprocessed_pseudogene")) %>% 
  arrange(median_identity) %>%
  mutate(gene_biotype = fct_inorder(gene_biotype)) %>% 
  # Violin with overlaid boxplot for each biotype
  ggplot() +
  aes(x = identity_human2mouse,
      y = gene_biotype,) +
  geom_violin(fill = "#4DBBD5FF",
              alpha = 0.2,
              color = NA, width = 2) +
  geom_boxplot(width = 0.3,
               fill = NA) +
  theme_vaxgo() +
  labs(x = "%Identity",
       y = "",
       title = "Biotype and %Identity of mouse orthologs")
```

### Cellular component ontology

Evaluate sequence identity of Leading Edge Genes (LEGs) and other
categories across major subcellular compartments.

```{r fig.height=3, fig.width=12}
# How LEGs are associated with identity by gene cellular component ontology
# Point-range plot of %Identity across gene categories, faceted by subcellular compartment

gene_info_human %>%
  # 1. Integrate GO annotations with global identity and status data
  inner_join(
    alignments_human2mouse_all_global_df_filtered %>% 
      select(-identity_human2mouse),
    by = join_by("hgnc_symbol" == gene)
  ) %>% 
  
  # 2. Filter for specific Cellular Component domains
  filter(
    go_domain == "cellular_component",
    go_name %in% c("nucleus", "cytoplasm", "membrane", "extracellular region")
  ) %>% 
  
  # 3. Clean and reformat status labels for plotting
  select(hgnc_symbol, go_name, identity_human2mouse, status) %>% 
  distinct() %>% 
  mutate(status = str_replace(status, " ", "\n")) %>%
  mutate(status = fct_relevel(
    status,
    "Human\nonly", "Mouse\nonly", "Shared", "All", 
    "Non-Immune\nBTMs", "Immune\nBTMs"
  )) %>% 
  
  # 4. Visualization: Compare identity across GO categories using facets
  ggplot() +
  aes(x = status,
      y = identity_human2mouse,
      fill = status,
      color = status) +
  facet_wrap(~str_to_title(go_name),
             nrow = 1,
             axes = "all") +
  
  # 5. Statistical Summary: Mean and 95% CI via bootstrapping
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange") +
  
  # 6. Styling, Scales, and Legends
  theme_vaxgo() +
  theme(legend.position = "top",
        axis.text.x = element_blank()) +
  scale_color_manual(values = colors$comparison) +
  scale_fill_manual(values = colors$comparison) +
  labs(color = "Status",
       y = "%Identity",
       x = "") +
  guides(color = guide_legend(nrow = 1, byrow = T),
         fill = "none")

# Distribution, boxplot and violin
gene_info_human %>%
  # 1. Join with global filtered identity data
  inner_join(
    alignments_human2mouse_all_global_df_filtered %>% 
      select(-identity_human2mouse),
    by = join_by("hgnc_symbol" == gene)
  ) %>% 
  
  # 2. Filter for major subcellular domains (Gene Ontology)
  filter(
    go_domain == "cellular_component",
    go_name %in% c("nucleus", "cytoplasm", "membrane", "extracellular region")
  ) %>% 
  
  # 3. Clean and reorder categories for consistent visualization
  select(hgnc_symbol, go_name, identity_human2mouse, status) %>% 
  distinct() %>% 
  arrange(identity_human2mouse) %>% 
  mutate(status = str_replace(status, " ", "\n")) %>% 
  mutate(status = fct_relevel(
    status,
    "Human\nonly", "Mouse\nonly", "Shared", "All", 
    "Non-Immune\nBTMs", "Immune\nBTMs"
  )) %>% 
  
  # 4. Visualization: Layered distribution per compartment
  ggplot() +
  aes(x = status,
      y = identity_human2mouse,
      fill = status,
      color = status) +
  facet_wrap(~str_to_title(go_name),
             scales = "free",
             nrow = 1,
             axes = "all") +
  
  # 5. Geoms: Violin for density + Boxplot for summary stats
  geom_violin(alpha = 0.2, color = NA) +
  geom_boxplot(width = 0.2) +
  
  # 6. Styling and Custom Color Scales
  theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        legend.position = "top",
        axis.title.x = element_blank()) +
  scale_color_manual(values = c(colors$comparison_color, "All" = "black")) +
  scale_fill_manual(values = c(colors$comparison, "All" = "black")) +
  
  # 7. Legend and Labels
  guides(color = "none",
         fill = guide_legend(nrow = 1, byrow = T)) +
  labs(fill = "Status",
       y = "%Identity",
       x = "")

```

Visualize the global distribution of sequence identity while
highlighting specific "exemplar" genes of interest

```{r fig.height=8, fig.width=12}
# Global Statistical Summary: Mean Conservation (Bootstrap)
alignments_human2mouse_pointrange_all_global = alignments_human2mouse_all_global_df_filtered %>%
  # 1. Clean labels and establish the logical factor order for the comparison
  mutate(
    status = str_replace(status, " ", "\n"),
    status = fct_relevel(
      status,
      "Human\nonly", "Mouse\nonly", "Shared", "All", 
      "Non-Immune\nBTMs", "Immune\nBTMs"
    )
  ) %>% 
  mutate(analysis = "Mean Conservation, Bootstrap") %>%
  
  # 2. Setup the visualization: Aesthetics and Faceting
  ggplot() +
  aes(x = status,
      y = identity_human2mouse,
      fill = status,
      color = status,
      group = status) +
  facet_wrap(~analysis) +
  
  # 3. Layer 1: Pointrange with bootstrapped Mean and CI
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               size = 1,
               fatten = 2,
               stroke = 0.5,
               pch = 21,
               mapping = aes(color = status),
               position = position_dodge2(0)) +
  
  # 4. Layer 2: Text labels for the precise mean values
  stat_summary(fun.data = "mean_cl_boot", 
               geom = "text", 
               aes(label = after_stat(round(y, 1))), 
               hjust = -0.2, 
               size = 3,
               color = "black") +
  
  # 5. Custom scales for fill and color
  scale_fill_manual(values = c(colors$comparison,
                               "BTMs" = "#4361ee",
                               "All" = "white")) +
  scale_color_manual(values = c(colors$comparison,
                                "BTMs" = "#4361ee",
                                "All" = "black")) +
  
  # 6. Adjust scales and coordinate system
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  
  # 7. Final Theme and Legend Cleanup
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
        panel.spacing = unit(1.1, "lines"),
        plot.margin = margin(0.3, 0.3, 0, 0, unit = "cm"),
        legend.position = "none",
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) +
  labs(x = "Status",
       y = "%Identity (Human-to-Mouse)",
       color = "",
       fill = "")
# Display
alignments_human2mouse_pointrange_all_global
```

Plot boxplot with violin plot and unite with point-range plot

```{r fig.height=8, fig.width=12}
# Global Distribution of Conservation with Highlighted Exemplar Genes
# Violin + boxplot with highlighted exemplar genes overlaid
alignments_human2mouse_histogram_all_global = alignments_human2mouse_all_global_df_filtered %>%
  # 1. Prepare categories and factor levels
  mutate(analysis = "Distribution of conservation",
         status = str_replace(status, " ", "\n"),
         status = fct_relevel(status, "Human\nonly", "Mouse\nonly", "Shared", 
                              "All", "Non-Immune\nBTMs", "Immune\nBTMs")) %>% 
  
  ggplot() +
  aes(x = status, y = identity_human2mouse, fill = status, color = status, group = status) +
  facet_wrap(~analysis) +
  
  # 2. Main Layers: Violin for density and transparent Boxplot
  geom_violin(color = "black", alpha = 0.7) +
  geom_boxplot(mapping = aes(color = status), width = 0.1, fill = NA, outliers = F) +
  
  # 3. First Color Scale: Category status
  scale_fill_manual(name = "Status", values = c(colors$comparison, "All" = "white")) +
  scale_color_manual(name = "Status", 
                     values = c("Mouse\nonly" = "white", "Human\nonly" = "white", "Shared" = "white", 
                                "Immune\nBTMs" = "white", "Non-Immune\nBTMs" = "white", "All" = "black")) +
  
  # 4. Secondary Color Scale: Custom filters for highlighted genes
  new_scale_color() + 
  scale_color_manual(
     name = "Filter",
     values = c("High %Identity" = "#4361ee", "Positive correlation,\nHigh %Identity" = "#4DBBD5FF",
                "Negative correlation,\nHigh %Identity" = "#7209b7", "Low %Identity" = "red",
                "Positive correlation,\nLow %Identity" = "#CA6702", "Negative correlation,\nLow %Identity" = "#AE2012")) +
  
  # 5. Adding highlighted genes as points
  geom_point(data = alignments_human2mouse_genes_filtered %>% 
               group_by(status, filter) %>% 
               slice_head(n = 5), aes(color = filter), pch = 21, fill = "white") +
  
  # 6. Statistical annotation: Median identity labels
  stat_summary(fun = median, geom = "label", aes(label = after_stat(round(y, 1))),
               size = 3, hjust = -0.2, color = "black", fill = "white") +
  
  # 7. Gene Labeling: Repel prevents overlapping names
  geom_label_repel(data = alignments_human2mouse_genes_filtered %>% 
               group_by(status, filter) %>% 
               slice_head(n = 5),
                   aes(label = gene, color = filter), 
                   max.overlaps = 10000, size = 2, direction = "y", nudge_x = -0.4,
                   hjust = 0, fill = "white", box.padding = unit(0, "cm"), label.padding = unit(0.1, "cm")) +
  
  # 8. Scales, Theme, and Layout
  scale_x_discrete(expand = c(0.1, 0.1)) +
  scale_y_continuous(expand = c(0.01, 0.01), limits = c(50, 105)) +
  theme_vaxgo() +
  theme(axis.text = element_text(size = 10),
        legend.position = "top", legend.direction = "vertical",
        legend.margin = margin(0,0,0,0), legend.box.margin = margin(0,0,-10,0),
        plot.margin = margin(0.3, 0.3, 0, 0, unit = "cm")) +
  labs(x = "Status", y = "%Identity (Human-to-Mouse)", color = "", fill = "")

# Display
alignments_human2mouse_histogram_all_global

# Unite summary and distribution plots using patchwork
alignments_human2mouse_pointrange_histogram = (alignments_human2mouse_pointrange_all_global /
alignments_human2mouse_histogram_all_global) +
  plot_layout(heights = c(0.25, 0.75),
              guides = "collect",
              axis_titles = "collect"
              ) +
  plot_annotation(theme = theme(legend.position = "right",
                                legend.justification = "top"))
alignments_human2mouse_pointrange_histogram

# Save composite figure
alignments_human2mouse_pointrange_histogram %>%
ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_violinplot_all_global.png"),
         width = 8,
         height = 8)
```

Line plot of %Identity across all genes, faceted by treatment type and
pathogen. This shows how sequence conservation evolves over time for
each gene category.

```{r fig.height=3, fig.width=10}
# Prepare data: clean timepoints, assign treatment groups, filter outliers
alignments_human2mouse_pointrange_all_df = alignments_human2mouse %>% 
  mutate(timepoint = str_replace(timepoint," ", "\n"),
         timepoint = fct_relevel(timepoint, c("Ref", "Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         treatment = case_when(pathogen %in% c("Hepatitis B", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "Saureus", "Ecoli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")) %>% 
  filter(!(pathogen == "Saureus" & timepoint == "24h")) %>% 
  select(!c(process, group, set_size_human:mouse_core, group)) %>% 
  distinct()

# Compute reference identity means for Immune BTM, Non-Immune BTM, and All
mean_ref_identity = alignments_human2mouse_all_global_df_filtered %>% 
  filter(status %in% c("Immune BTMs", "Non-Immune BTMs", "All")) %>% 
  group_by(status) %>% 
  reframe(mean_cl_boot(identity_human2mouse))

# Build line plot with pointrange for each gene category across timepoints
alignments_human2mouse_pointrange_all_plot = alignments_human2mouse_pointrange_all_df %>% 
  ggplot() +
  aes(x = timepoint,
      y = identity_human2mouse,
      fill = status,
      color = status,
      group = status) +
  facet_grid(~treatment+pathogen, 
             scales = "free_x",
             space = "free_x") +
  # Line connecting mean values
  stat_summary(fun = mean,
               geom = "line",
               size = 0.5) +
  # Pointrange with bootstrapped CI
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               size = 1,
               fatten = 1,
               mapping = aes(color = status))+
  # Label with number of genes per timepoint
  geom_text_repel(
    data = alignments_human2mouse_pointrange_all_df %>%
      group_by(treatment, pathogen, timepoint, status) %>%
      summarise(n_genes = n(),
                identity_human2mouse = mean(identity_human2mouse),
                .groups = "drop"),
    aes(x = timepoint,
        y = identity_human2mouse,
        label = n_genes,
        color = status),
    inherit.aes = FALSE,
    size = 2, segment.size = 0,
    box.padding = 0.5
  ) +
  # Reference pointrange for overall category mean at baseline
  geom_pointrange(mapping = aes(x = "Ref", y = y, ymin = ymin, ymax = ymax),
                  size = 1,
                  fatten = 1,
                  data = mean_ref_identity) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = c(colors$comparison, 
                     "All" = "black")
                     ) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 10),
          axis.text.y = element_text(size = 10),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  labs(x = "Time",
       y = "%Identity (Human-to-Mouse)",
       color = "",
       fill = "") +
  guides(fill = "none")

# Print
alignments_human2mouse_pointrange_all_plot

# Save
alignments_human2mouse_pointrange_all_plot %>%
ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_lineplot_all.png"),
         width = 10,
         height = 3)
```

Point-range lineplot of %Identity across time, faceted by BTM functional
group and treatment. This provides a high-resolution view of
conservation dynamics within each immunological and metabolic module.

```{r fig.height=12, fig.width=12}
# Lineplot faceted by BTM group and treatment across timepoints
alignments_human2mouse_lineplot_all = 
  # Append an "All" category alongside individual BTM groups
  bind_rows(alignments_human2mouse %>% 
              mutate(group = "All"),
            alignments_human2mouse
            ) %>% 
  # Clean timepoint labels and assign treatment and factor orders
  mutate(timepoint = str_replace(timepoint,"Early hours", "Early\nhours"),
         timepoint = fct_relevel(timepoint, c("Early\nhours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         group = fct_relevel(group, "All", immune_order),
         treatment = case_when(pathogen %in% c("HepB", "Influenza") ~ "Vaccination",
                               pathogen %in% c( "Saureus", "Ecoli") ~ "Infection",
                               pathogen %in% c( "Burn", "Trauma") ~ "Injury",
                               T ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury")) %>% 
  
  ggplot() +
  aes(x = timepoint,
      y = identity_human2mouse,
      fill = status,
      color = status,
      group = status) +
  # One row per BTM group, one column per treatment:pathogen
  facet_grid(group ~ treatment+pathogen, 
             scales = "free",
             space = "free_x") +
  # Line connecting mean identity values
  stat_summary(fun = mean, 
               geom = "line", 
               size = 0.8) +
  # Pointrange with bootstrapped 95% CI
  stat_summary(fun.data = "mean_cl_boot",
               geom = "pointrange",
               mapping = aes(color = status),
               size = 0.2
               ) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, vjust = 1, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          panel.spacing = unit(1.1, "lines"),
          plot.margin = margin(0, 0.3, 0, 0, unit = "cm"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.margin=margin(0,0,0,0),
          legend.box.margin=margin(0,0,-10,0),
          legend.text = element_text(size = 10, hjust = 0),
          legend.title = element_text(size = 10),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
    ) +
  labs(x = "Time",
       y = "%Identity (Human-to-Mouse)",
       color = "",
       fill = "")

alignments_human2mouse_lineplot_all

# alignments_human2mouse_lineplot_all %>% 
#   ggsave(filename = here("Figures", "all_human_mouse_gsea_LEGs_alignment_btms_human2mouse_lineplot_bygroup.png"),
#          width = 12,
#          height = 12)

```

### Align most conserved genes

Load the shared LEG data and prepare for gene annotation and alignment.
The top exemplar genes are extracted from the filtered dataset for
downstream biomaRt queries and protein sequence alignment.

```{r}
# Import GSEA core results and BTM annotation table
gsea_btm_results_core_joined =
  readRDS(file = here("tables", "all_human_mouse_gsea_btm_results_core_joined.rds"))
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

# Extract top exemplar genes from the filtered dataset
shared_genes_df = alignments_human2mouse_genes_filtered %>% 
  select(filter, status, everything()) %>% 
  group_by(filter, status) %>% 
  slice_head(n  = 1) %>% 
  ungroup()

shared_genes = shared_genes_df %>% 
  pull(gene) %>% 
  unique()

# 1. Setup biomaRt for gene annotation
## Gene annotation
# human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
# mouse_mart <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")

# 2. Retrieve Pfam domains, GO terms, and Identity
# We include homology identity to correlate directly
domains_human <- getBM(
  attributes = c(
    "hgnc_symbol",
    "interpro",
    "interpro_start",
    "interpro_end",
    "interpro_short_description",
    # "tmhmm_start",
    # "tmhmm_end",
    "phobius",
    "phobius_start",
    "phobius_end"
  ),
  filters = "hgnc_symbol",
  values = shared_genes,
  # values = "ABCE1",
  mart = human_mart
)

sequences_human = getBM(
  attributes = c(
    "hgnc_symbol",
    "peptide",
    "uniprotswissprot"
  ),
  filters = "hgnc_symbol",
  values = shared_genes,
  # values = "ABCE1",
  mart = human_mart
) %>% 
  #Filter out non-canonical/annotated sequences
  filter(uniprotswissprot != "")

identity_human <- getBM(
  attributes = c("external_gene_name", 
                 "mmusculus_homolog_perc_id_r1",
                 "mmusculus_homolog_goc_score",
                 "mmusculus_homolog_orthology_confidence"),
  filters = "hgnc_symbol",
  # values = "ABCE1",
  values = shared_genes,
  mart = human_mart
) %>% 
  rename(hgnc_symbol = external_gene_name,
         identity = mmusculus_homolog_perc_id_r1)

#REGULATION
regulation_enhancer_human = getBM(
  attributes = c(
    "external_gene_name",
    "transcription_start_site"
  ),
  filters = "external_gene_name",
  values = "MARCKS",
  mart = human_mart
)
```

```{r}


# Inspect available BioMart attributes for human — used to identify which annotation columns can be retrieved
listAttributes(human_mart) %>% View()
```

```{r}
# Join the tables
domains_identity_sequence_human = 
  sequences_human %>% 
  inner_join(domains_human, 
             by = "hgnc_symbol") %>% 
  left_join(identity_human, 
            domains_human,
            by = "hgnc_symbol") %>% 
  group_by(hgnc_symbol, interpro) %>% 
  mutate(interpro_length = interpro_end - interpro_start) %>% 
  slice_max(interpro_length, n = 1)


#Mouse
genes_mouse = getBM(
  attributes = c("external_gene_name", 
                 "hsapiens_homolog_associated_gene_name"),
  filters = "external_gene_name",
  values = str_to_title(shared_genes),
  mart = mouse_mart
) %>% 
  rename(mgi_symbol = external_gene_name,
         hgnc_symbol = hsapiens_homolog_associated_gene_name)

domains_mouse <- getBM(
  attributes = c(
    "external_gene_name",
    "interpro",
    "interpro_start",
    "interpro_end",
    "interpro_short_description",
    # "tmhmm_start",
    # "tmhmm_end",
    "phobius",
    "phobius_start",
    "phobius_end"
    # "signalp_start",
    # "signalp_end"
  ), 
  filters = "external_gene_name",
  values = genes_mouse$mgi_symbol,
  mart = mouse_mart
)

sequences_mouse = getBM(
  attributes = c(
    "external_gene_name",
    "peptide",
    "uniprotswissprot"
  ),
  filters = "external_gene_name",
  values = genes_mouse$mgi_symbol,
  # values = "Abce1",
  mart = mouse_mart
) %>% 
  #Filter out non-canonical/annotated sequences
  filter(uniprotswissprot != "")

# enhancer_mouse = getBM(
#   attributes = c(
#     "external_gene_name",
#     "peptide",
#     "uniprotswissprot"
#   ),
#   filters = "external_gene_name",
#   values = genes_mouse$mgi_symbol,
#   # values = "Abce1",
#   mart = mouse_mart
# ) %>% 
#   #Filter out non-canonical/annotated sequences
#   filter(uniprotswissprot != "")


#Merge
domains_identity_sequence_mouse = 
  sequences_mouse %>% 
  inner_join(domains_mouse, 
             by = "external_gene_name") %>% 
  rename(mgi_symbol = external_gene_name) %>% 
  left_join(genes_mouse, 
            by = "mgi_symbol") %>% 
  arrange(interpro_start) %>% 
  group_by(mgi_symbol, interpro) %>% 
  mutate(interpro_length = interpro_end - interpro_start) %>% 
  slice_max(interpro_length, n = 1)


#Merge Human and Mouse
genes_domain_mouse_human <- domains_identity_sequence_mouse %>% 
  inner_join(domains_identity_sequence_human, 
             by = join_by(hgnc_symbol, interpro), 
             suffix = c("_mouse", "_human")) %>% 
  drop_na(peptide_mouse, peptide_human) %>% 
  inner_join(shared_genes_df %>% 
               select(filter, status, hgnc_symbol = gene, n, cor_value, p_value, r_squared, cor_label) %>% 
               distinct(),
             by = "hgnc_symbol")

glimpse(genes_domain_mouse_human)

genes_domain_mouse_human %>% 
  saveRDS(file = here("tables", "all_human_mouse_gsea_leganalysis_biomart_annotation.rds"))


# listAttributes(mouse_mart) %>% View()
# listAttributes(human_mart) %>% View()

```

Is there an association between domains and correlation?

```{r}
# Quick inspection: select distinct domain-level metrics per gene to check whether domain composition correlates with expression conservation
genes_domain_mouse_human %>%
  select(hgnc_symbol,
         interpro,
         interpro_short_description_human,
         identity, interpro_length_human,
         filter:cor_label) %>%
  distinct() %>%
  drop_na(cor_value)
```

Alignment

```{r}
genes_domain_mouse_human =
  readRDS(file = here("tables", "all_human_mouse_gsea_leganalysis_biomart_annotation.rds")) %>% 
  distinct()

selected_gene = "CD2"

# #Align domains
# sequences_align_domains = genes_domain_mouse_human %>% 
# mutate(
#     # Cut sequence: peptide[start : end]
#     domain_seq_human = stringr::str_sub(peptide_human, interpro_start_human, interpro_end_human),
#     domain_seq_mouse = stringr::str_sub(peptide_mouse, interpro_start_mouse, interpro_end_mouse)
#   )
# 
# #Align full sequence
# sequences_align_gene <- genes_domain_mouse_human %>%
#   filter(status == "Shared") %>% 
#   ungroup() %>% 
#   filter(hgnc_symbol == selected_gene) %>% 
#   select(hgnc_symbol, peptide_human, peptide_mouse, status, cor_label) %>%
#   distinct()
# 
# #Alignment
# gene_data <- sequences_align_gene
# 
# # Create AAStringSet to run on msa
# input_seqs <- AAStringSet(c(gene_data$peptide_human, gene_data$peptide_mouse))
# names(input_seqs) <- c("Human", "Mouse")
# 
# # Run MSA
# aln <- msa(input_seqs, method = "Muscle", type = "protein")
# 
# # Extract sequences
# # Convert to text
# seqs_alinhadas <- as.character(unmasked(aln))

# Pairwise sequence alignment
library(pwalign)
#Compare two original sequences
alignment <- pairwiseAlignment(
  pattern = gene_data$peptide_human, 
  subject = gene_data$peptide_mouse, 
  substitutionMatrix = "BLOSUM62",
  type = "global", # Garante alinhamento de ponta a ponta
  gapOpening = 10,  # Penalidade padrão
  gapExtension = 0.5
)

# Get identity score
identity_global <- pid(alignment, type = "PID1")

#Convert to a df
alignment_results_pairwise <- tibble(
  hgnc_symbol       = gene_data$hgnc_symbol,
  seq_human_orig    = gene_data$peptide_human,
  seq_mouse_orig    = gene_data$peptide_mouse,
  seq_human_aligned = as.character(pattern(alignment)),
  seq_mouse_aligned = as.character(subject(alignment)),
  identity          = identity_global
) %>% 
  inner_join(genes_domain_mouse_human %>% 
  ungroup() %>% 
  select(mgi_symbol, hgnc_symbol, identity_biomart = identity, peptide_mouse, peptide_human, uniprotswissprot_human,
         filter, status, n, cor_value, p_value, r_squared, cor_label) %>% 
  distinct(),
  by = "hgnc_symbol")



```

### Protein structure

#### Prepare protein data for CD2

Fetch UniProt annotations (topology, domains, PTMs) and compute pairwise
sequence differences between human and mouse CD2 from the alignment.
These will feed into a multi-track protein annotation plot and 3D
visualisation.

```{r fig.height=4, fig.width=8}
# Load tidyverse for data processing
library(tidyverse)

# Prepare sequence and annotation data for the selected gene
# Extract the row corresponding to selected_gene and keep the longest isoforms
# to ensure we have full-length sequences for comparison
gene_info <- alignment_results_pairwise %>%
  filter(hgnc_symbol == selected_gene) %>%
  mutate(
    len_mouse = nchar(seq_mouse_orig),
    len_human = nchar(seq_human_orig)
  ) %>%
  slice_max(len_human, n = 1) %>%
  slice_max(len_mouse, n = 1)

# Create a "Full Protein" annotation track covering the entire sequence
# This provides the background rectangle spanning the whole protein length
full_protein <- gene_info %>%
  select(hgnc_symbol, len_mouse, len_human, status, cor_label,
         seq_human_orig, seq_mouse_orig, filter, identity, identity_biomart,
         seq_human_aligned, seq_mouse_aligned) %>%
  distinct() %>%
  pivot_longer(cols = starts_with("len_"), names_to = "species", values_to = "end") %>%
  group_by(hgnc_symbol, species) %>%
  slice_max(end, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    species = str_remove(species, "len_"),
    start = 1,
    interpro_short_description_human = "Full Protein",
    interpro = "Full",
    species_end = species
  ) %>%
  distinct()

# Retrieve UniProt feature annotations for CD2 (human and mouse orthologs)
# This fetches transmembrane domains, topology, and other protein features
prot_data <- drawProteins::get_features("P06729 P08920") %>%
  drawProteins::feature_to_dataframe() %>%
  mutate(description = case_when(type == "TRANSMEM" ~ "TM",
                                  type == "TOPO_DOM" & description == "Extracellular" ~ "EC",
                                  type == "TOPO_DOM" & description == "Cytoplasmic" ~ "CP",
                                  description == "NONE" ~ type,
                                  description == "" ~ type,
                                  TRUE ~ description)) %>%
  separate(entryName, sep = "_", into = c("gene", "organism"), remove = FALSE) %>%
  mutate(organism = str_to_title(organism))

# Identify sequence differences between human and mouse from the pairwise alignment
vec_h <- strsplit(unique(full_protein$seq_human_aligned), "")[[1]]
vec_m <- strsplit(unique(full_protein$seq_mouse_aligned), "")[[1]]

# Find positions where the aligned residues differ
diff_idx <- which(vec_h != vec_m)

# Map alignment positions back to real (ungapped) residue positions per species
mutations_base <- tibble(
  align_pos = diff_idx,
  res_h = vec_h[diff_idx],
  res_m = vec_m[diff_idx]
) %>%
  mutate(
    pos_real_human = map_int(align_pos, ~ sum(vec_h[1:.x] != "-")),
    pos_real_mouse = map_int(align_pos, ~ sum(vec_m[1:.x] != "-"))
  )

# Separate by species to track mutations in each organism's coordinate system
mutations_human <- mutations_base %>%
  filter(res_h != "-") %>%
  mutate(species = "human", x_pos = pos_real_human)

mutations_mouse <- mutations_base %>%
  filter(res_m != "-") %>%
  mutate(species = "mouse", x_pos = pos_real_mouse)

# Combine mutations into a single table formatted for ggplot annotation
mutations_plot <- bind_rows(mutations_human, mutations_mouse) %>%
  rename(organism = species) %>%
  mutate(gene = unique(gene_info$hgnc_symbol),
         organism = str_to_title(organism),
         type = "Alignment",
         description = "Mutations")

```

#### Visualize structures

draw proteins

```{r fig.height=4, fig.width= 10, paged.print=TRUE}
# Get a human-readable protein name for the y-axis label
prot_name <- unique(prot_data$entryName)

# Split UniProt features into thematic tracks for multi-panel visualisation
# Track 1: Topology — transmembrane and subcellular localisation
df_topology <- prot_data %>%
  filter(type %in% c("TOPO_DOM", "TRANSMEM", "SIGNAL")) %>%
  mutate(y_pos = 1)

# Track 2: Domains — functional regions (Ig-like, receptors, etc.)
df_features <- prot_data %>%
  filter(type %in% c("DOMAIN", "REGION", "MOTIF", "RECEPT_DOM")) %>%
  mutate(y_pos = 1.5)

# Track 3: Modifications — PTMs and natural variants
df_modifications <- prot_data %>%
  filter(type %in% c("MOD_RES", "PHOSPHO", "VARIANT")) %>%
  mutate(y_pos = 1.61)

# Alignments-derived mutation data frame
df_mutations <- mutations_plot

# Create a summary dataframe to define the background line length for each facet
# This ensures the grey backbone scales to the correct length per organism
background_lines <- prot_data %>%
  group_by(organism, gene) %>%
  summarise(max_end = max(end), .groups = 'drop') %>%
  tidyr::crossing(y_val = c(1, 1.5))

# Initialise the multi-layer protein annotation ggplot
drawprotein_uniprot_ggplot <- ggplot() +

  # Track background: grey line spanning the full protein length per facet
  geom_segment(data = background_lines,
               aes(x = 1, xend = max_end, y = y_val, yend = y_val),
               color = "grey80", size = 2) +

  # Track 1: Topology rectangles — extracellular, cytoplasmic, transmembrane
  geom_rect(data = df_topology,
            aes(xmin = begin, xmax = end,
                ymin = y_pos - 0.1, ymax = y_pos + 0.1,
                fill = description),
            color = "black", size = 0.2) +

  # Facet by organism and gene for side-by-side species comparison
  facet_wrap(~ paste(organism, gene), ncol = 1) +

  # Custom fill scale for topology track
  scale_fill_manual(values = c("EC" = "#4cc9f0",
                                "CP" = "#3a86ff",
                                "TM" = "#90A4AEFF",
                                "SIGNAL" = "#f72585"),
                    name = "Topology") +

  # Reset fill scale for the next aesthetic layer
  ggnewscale::new_scale_fill() +

  # Track 2: Functional domains — semi-transparent rectangles with independent fill
  geom_rect(data = df_features,
            aes(xmin = begin, xmax = end,
                ymin = y_pos - 0.1, ymax = y_pos + 0.1,
                fill = description),
            color = "black", size = 0.2, alpha = 0.5) +
  scale_fill_viridis_d(option = "mako", name = "Domains") +

  # Track 3: Modification points — diamonds for phosphorylation / variants
  geom_point(data = df_modifications,
             aes(x = begin, y = y_pos + 0.02, color = type),
             size = 3, shape = 18) +
  scale_color_manual(values = c("PHOSPHO" = "yellow", "VARIANT" = "red"),
                     name = "PTMs") +

  # Reset color for mutation layer
  new_scale_color() +

  # Mutation layer: violin + points to show alignment divergence density
  geom_violin(data = mutations_plot,
              aes(x = x_pos, y = 1.5, color = "Mutations"),
              fill = NA, size = 0.4, width = 0.2, scale = "count", adjust = 0.05) +
  geom_point(data = mutations_plot,
             aes(x = x_pos, y = 1.5, color = "Mutations",
                 label = paste(align_pos, res_h, res_m)),
             shape = 124, size = 1, alpha = 0.6) +
  scale_color_manual(values = c("Mutations" = "red"), name = "Alignment") +

  # Axis formatting: map y_pos to track labels
  scale_y_continuous(breaks = c(unique(df_topology$y_pos), unique(df_features$y_pos)),
                     labels = c("Topology", "Domains")) +
  scale_x_continuous(expand = c(0.01, 0.01)) +

  # Annotate domain start and end coordinates on the plot
  geom_text(aes(label = begin, y = y_pos, x = begin),
            data = df_features,
            nudge_y = 0.15, hjust = 0.5, size = 3) +
  geom_text(aes(label = end, y = y_pos, x = end),
            data = df_features,
            nudge_y = -0.15, hjust = 0.5, size = 3) +
  geom_text(aes(label = begin, y = y_pos, x = begin),
            data = df_topology,
            nudge_y = 0.15, hjust = 0.5, size = 3) +
  geom_text(aes(label = end, y = y_pos, x = end),
            data = df_topology,
            nudge_y = -0.15, hjust = 0.5, size = 3) +

  # Custom lab theme
  theme_vaxgo() +
  theme(axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0),
        panel.grid.minor = element_blank(),
        legend.position = "right",
        legend.justification = "top",
        legend.spacing.y = unit(0, "cm"),
        plot.margin = margin(1, 1, 2, 1)) +

  # Labels: include identity and correlation metrics in the title
  labs(x = "Residue position",
       y = "Features",
       title = paste0("Gene: ", unique(prot_data$gene),
                      ", UNIPROT: ", unique(prot_data$accession),
                      ". Status: ", unique(gene_info$status),
                      ". Filter: ", str_replace(unique(gene_info$filter), "\n", " "),
                      ", Correlation: ", unique(gene_info$cor_label),
                      "\nIdentity (MUSCLE, BLOSUM62): ", round(unique(gene_info$identity), 1),
                      ". Identity (BioMart): ", round(unique(gene_info$identity_biomart), 1)))

# Display the plot
drawprotein_uniprot_ggplot

```

VIsualize 3d

```{r}
# Define hex colour codes for topology tracks — each type gets a distinct colour for the 3D viewer
topo_colors <- list(
  "EC" = "#4cc9f0",
  "CP" = "#3a86ff",
  "TM" = "black",
  "SIGNAL" = "#FF1493"
)

# Prepare topology data for NGLVieweR: assign colours and create residue-range selections
df_topology_3d <- df_topology %>%
  mutate(
    color_3d = as.character(topo_colors[description]),
    selection = paste0(begin, "-", end)
  )

# Prepare domain feature data for NGLVieweR: map mako colours and create selections
df_features_3d <- df_features %>%
  mutate(
    selection = paste0(begin, "-", end),
    color_hex = as.character(mako(n())[as.numeric(as.factor(description))])
  )

# Initialise NGLVieweR with the CD2 human AlphaFold structure
view_initial <- NGLVieweR(here::here("Structures", "CD2_HumanAF-P06729-F1-model_v6.pdb")) %>%
  stageParameters(backgroundColor = "white") %>%
  addRepresentation("ribbon",
                    param = list(color = "#D3D3D3", opacity = 0.2))

# Apply a small rotation for a better initial view angle
view_initial %>%
  setRotation(10, 0, 0)

view_initial
```

Topology

```{r}
# Add topology cartoon representations on top of the initial structure
# Each colour-coded segment (EC, CP, TM, SIGNAL) maps to a UniProt-annotated region
view_topology <- view_initial
for (i in seq_len(nrow(df_topology_3d))) {
  current_color <- df_topology_3d$color_3d[i]
  if (!is.na(current_color)) {
    view_topology <- view_topology %>%
      addRepresentation("cartoon", param = list(
        sele = df_topology_3d$selection[i],
        color = current_color,
        opacity = 0.5
      ))
  }
}
view_topology
```

domains

```{r}
library(NGLVieweR)
library(dplyr)

# Define distinct colours for each domain feature to ensure clear visual separation in the 3D structure
domain_colors <- colors_all

# Prepare the domain feature table with colour assignment and residue range selection
df_features_3d <- df_features %>%
  mutate(
    color_3d = domain_colors[1:n()],
    selection = paste0(begin, "-", end)
  )

# Initialise NGLVieweR with CD2 human AlphaFold structure
# Set orthographic camera, high sample level for antialiasing, and a transparent ribbon backbone
view_domain <- NGLVieweR(here::here("Structures", "CD2_HumanAF-P06729-F1-model_v6.pdb")) %>%
  stageParameters(
    backgroundColor = "white",
    cameraType = "orthographic",
    clipDist = 0,
    sampleLevel = 4,
    preserveDrawingBuffer = TRUE
  ) %>%
  addRepresentation("cartoon", param = list(color = "#D3D3D3", opacity = 0.2))

# Overlay each domain with its assigned colour as a cartoon representation
for (i in seq_len(nrow(df_features_3d))) {
  current_color <- as.character(df_features_3d$color_3d[i])
  current_sele  <- as.character(df_features_3d$selection[i])
  if (!is.na(current_color)) {
    view_domain <- view_domain %>%
      addRepresentation("cartoon", param = list(
        sele = current_sele,
        color = current_color,
        opacity = 1
      ))
  }
}

# Render and rotate for optimal viewing angle
view_domain
view_domain <- view_domain %>%
  NGLVieweR::setRotation(c(0, 1.57, 0)) %>%
  NGLVieweR::setSelection("*")
view_domain

# Annotate chains in the PDB object: assign a single-letter code per domain for downstream visualisation
for (i in seq_len(nrow(df_features_3d))) {
  desc_char <- toupper(substr(as.character(df_features_3d$description[i]), 1, 1))
  res_range <- as.numeric(unlist(strsplit(as.character(df_features_3d$selection[i]), "-")))
  sele <- atom.select(pdb, resno = res_range[1]:res_range[2])
  if (length(sele$atom) > 0) {
    pdb$atom$chain[sele$atom] <- desc_char
  }
}

# Save the chain-annotated PDB for later use
write.pdb(pdb, file = here("Structures", "cd2_com_chains.pdb"))
```

```{r}
# Render mutations from the pairwise alignment on the 3D structure
# Each divergence position between human and mouse CD2 is shown as a solid red cartoon segment
if (nrow(mutations_plot) > 0) {
  mut_residues <- unique(as.character(mutations_plot$x_pos))
  mut_sele <- paste(mut_residues, collapse = " OR ")
  view_mutations <- view_topology %>%
    addRepresentation("cartoon", param = list(
      sele = mut_sele,
      color = "#FF0000",
      opacity = 1
    ))
}
view_mutations
```

Display alignment with ggmsa

```{r fig.height=3, fig.width=16}
# Set the gene for visualisation
selected_gene <- "CD2"

# Extract the most complete row for CD2 from the alignment results
data_row <- plot_domains %>%
  select(hgnc_symbol, identity, identity_biomart,
          seq_human_orig, seq_mouse_orig,
          seq_mouse_aligned, seq_human_aligned) %>%
  distinct()

# Create an AAStringSet object from the human and mouse aligned sequences
seqs_to_plot <- AAStringSet(c(data_row$seq_human_aligned, data_row$seq_mouse_aligned))
names(seqs_to_plot) <- c("Human", "Mouse")

# Render the multiple sequence alignment with ggmsa
# Colour by amino acid chemistry, use Human as reference, and show consensus tracks
ggmsa(seqs_to_plot,
      ref = "Human",
      start = 1,
      end = 352,
      color = "Chemistry_AA",
      font = "DroidSansMono",
      char_width = 0.8,
      border = "white",
      seq_name = TRUE,
      consensus_views = TRUE,
      disagreement = FALSE,
      use_dot = FALSE,
      by_conservation = FALSE,
      ignore_gaps = FALSE) +
  facet_msa(field = 100) +
  theme(panel.spacing.y = unit(0, "lines"),
        legend.position = "right",
        axis.text.y = element_text(color = "black", size = 10),
        axis.text.x = element_text(color = "black", size = 10)) +
  labs(title = paste0(data_row$hgnc_symbol, ", Identity: ", round(data_row$identity, 1), "%")) 


```

## Regulation

### Transcription factors
Annotate with gene regulation - TF Identify conserved Transcription
Factors (TFs) regulating specific genes in both Humans and Mice

```{r}
# Clean and parse the ChEA3 transcription factor–target gene matrix
gene_attribute_chea <- read_tsv(here("tables", "gene_attribute_matrix.txt"))

# Transform the wide ChEA matrix into long format, keeping only positive scores
gene_attribute_chea_clean <- gene_attribute_chea %>%
  rename(gene = Gene) %>%
  pivot_longer(-gene, names_to = "experiment", values_to = "score") %>%
  filter(score > 0) %>%
  arrange(desc(score)) %>%
  # Split the experiment metadata into TF name, PMID, and method/tissue/species
  separate(experiment,
           into = c("tf", "pmid", "method_tissue_species"),
           sep = "-",
           extra = "merge",
           fill = "right") %>%
  # Detect species from the metadata string (HUMAN or MOUSE)
  mutate(across(where(is.character), str_trim),
         species = case_when(str_detect(method_tissue_species, "HUMAN") ~ "Human",
                              str_detect(method_tissue_species, "MOUSE") ~ "Mouse",
                              TRUE ~ method_tissue_species)) %>%
  filter(species %in% c("Human", "Mouse"))

# Count evidence per (gene, TF) pair across species
# For each TF–target pair, count how many PMIDs support it in Human and in Mouse
gene_attribute_tf_count <- gene_attribute_chea_clean %>%
  ungroup() %>%
  select(gene, tf, pmid, species) %>%
  group_by(gene, tf) %>%
  pivot_wider(names_from = "species",
              values_from = "pmid") %>%
  mutate(
    Human_pmids = map_chr(Human, ~ paste(.x, collapse = ", ")),
    Mouse_pmids = map_chr(Mouse, ~ paste(.x, collapse = ", ")),
    Human_pmids_n = map_int(Human, length),
    Mouse_pmids_n = map_int(Mouse, length)
  ) %>%
  select(!c(Human, Mouse)) %>%
  mutate(across(ends_with("pmids"), ~ na_if(.x, ""))) %>%
  # Flag whether this TF regulates this gene in both organisms
  mutate(both = if_else(!is.na(Human_pmids) & !is.na(Mouse_pmids), "TRUE", "FALSE"))

# Save the cross-species TF-target count for downstream analysis
saveRDS(gene_attribute_tf_count,
        file = here("tables", "regulation_gene_tf_count.rds"))


```

```{r}
# Reload the TF-count data and gene alignment quadrants
gene_attribute_tf_count <- readRDS(file = here("tables", "regulation_gene_tf_count.rds"))
alignments_human2mouse_genes_filtered <- read_rds(file = here("tables", "alignments_human2mouse_genes_filtered.rds"))
alignments_human2mouse_all_global_df_nshared
dge_btm_process_genes_wide

# Build a contingency table of cross-species TF regulation for a few selected genes
# Join with the filtered gene quadrants to show how many TFs are shared vs species-specific
gene_attribute_tf_count %>%
  filter(gene %in% c("XPO1", "CD2", "CDK14", "NDUFB10")) %>%
  inner_join(alignments_human2mouse_genes_filtered %>%
               group_by(status, filter) %>%
               slice_head(n = 5) %>%
               filter(status == "Shared"),
             by = "gene") %>%
  mutate(gene = paste(gene, filter, cor_label)) %>%
  arrange(desc(both)) %>%
  tabyl(gene, both) %>%
  adorn_totals(where = "col") %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns() %>%
  as.data.frame()


```

#### Calculate Shared TFs

```{r fig.height=3, fig.width=5}
# Join the TF regulation table with the gene alignment quadrants
gene_attribute_tf_count_filter_df <- gene_attribute_tf_count %>%
  inner_join(alignments_human2mouse_all_global_df_nshared,
             by = "gene")

# Summarise TF counts per gene, split by cross-species status (both vs species-specific)
gene_attribute_tf_count_filter_df_summary <- gene_attribute_tf_count_filter_df %>%
  arrange(desc(both)) %>%
  group_by(gene, both) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(gene) %>%
  mutate(total = sum(n),
         percentage = round(n / total * 100, 1))

# Compute unique TF counts per gene for each species
gene_attribute_tf_count_test <- gene_attribute_tf_count %>%
  select(gene, tf, Human_pmids_n, Mouse_pmids_n) %>%
  mutate(Mouse_TF = if_else(Mouse_pmids_n > 0, TRUE, FALSE),
         Human_TF = if_else(Human_pmids_n > 0, TRUE, FALSE)) %>%
  group_by(gene) %>%
  summarise(unique_tfs_human = n_distinct(tf[Human_TF == TRUE]),
            unique_tfs_mouse = n_distinct(tf[Mouse_TF == TRUE]))

# gene_attribute_tf_count_filter_df_summary %>% 
#   filter(both == "T") %>% 
#   ggplot() +
#   aes(y = filter,
#       x = percentage,
#       fill = filter) +
#   geom_boxplot() +
#   geom_point(alpha = 0.2) +
#     stat_summary(geom = "pointrange",
#                fun.data = "mean_cl_boot",
#                position = "dodge",
#                color = "red") +
#   theme_vaxgo() +
#   theme(legend.position = "none") +
#   scale_fill_manual(values = colors_all)

gene_attribute_tf_count_filter_df_summary_annotated = gene_attribute_tf_count_filter_df_summary %>%
  arrange(gene) %>% 
  distinct() %>%
    select(-percentage) %>% 
  pivot_wider(values_from = "n",
              names_from = "both") %>% 
  distinct() %>%
  rename(not_shared_tf = "F",
         shared_tf = "T") %>% 
  mutate(shared_tf = if_else(is.na(shared_tf), 0, shared_tf)) %>% 
  inner_join(gene_attribute_tf_count_filter_df %>% 
               ungroup() %>% 
               filter(class %in% c("Shared", "Not-shared")) %>% 
               select(gene, class:cor_label) %>% 
               distinct(),
             by = join_by(gene)) %>% 
  ungroup() %>% 
  mutate(shared_tf_pct = (shared_tf/total) %>% round(2)) %>% 
    mutate(logp = -log(p_value),
         cor_value_abs = abs(cor_value)) %>% 
  #Highlight genes
  mutate(classification = "All") %>% 
  inner_join(gene_attribute_tf_count_test,
             by = "gene")
  
#Highlight genes
shared_genes = gene_attribute_tf_count_filter_df_summary_annotated %>% 
                    filter(gene %in% c(top_shared_notshared_genes %>% 
                                 filter(class == "Shared") %>% 
                                   pull(gene) %>% unique())) %>% 
              mutate(classification = "Shared") 

gene_attribute_tf_count_filter_df_summary_annotated = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  bind_rows(shared_genes) 

gene_attribute_tf_count_filter_df_summary_annotated %>% 
  dplyr::count(classification, class)
  
```

#### Plot Total of TFs by species

Scatterplot of total unique TFs per gene in Human vs Mouse. Points are
coloured by classification (Shared vs All). A diagonal line indicates
equal TF counts between species.

```{r fig.height=5, fig.width=7}
# Compare total unique TF counts per gene between human and mouse
regulation_tfs_total_by_species_plot <- gene_attribute_tf_count_filter_df_summary_annotated %>%
  drop_na(cor_value) %>%
  ggscatter(x = "unique_tfs_mouse",
            y = "unique_tfs_human",
            fill = "classification",
            color = "classification",
            point = FALSE,
            add = "reg.line",
            conf.int = TRUE) +
    scale_x_continuous(limits = c(0, 80)) +
  scale_y_continuous(limits = c(0, 80)) +
  geom_abline(linetype = 2,
              alpha = 0.5) +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "Human (Total TFs)",
       y = "Mouse (Total TFs)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value,
                 label = gene),
             alpha = 0.8,
             pch = 21,
             stroke = 0,
             size = 2) +
    geom_point(aes(fill = cor_value),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(80, 75), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_total_by_species_plot
  
```

```{r fig.height=3, fig.width=5}
#Identity
regulation_tfs_total_identity_plot =  
  gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  ggscatter(x = "total",
            y = "identity_human2mouse", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line", 
            conf.int = T
              ) +
  scale_color_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "Total TFs",
       y = "%Identity (a.a.)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value),
             alpha = 0.2,
             pch = 21,
             stroke = 0) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(120, 115), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_total_identity_plot 

#Identity
regulation_tfs_shared_pct_identity_plot =  
  gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  ggscatter(x = "shared_tf_pct",
            y = "identity_human2mouse", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line", 
            conf.int = T
              ) +
  scale_color_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "%Shared TFs",
       y = "%Identity (a.a.)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value),
             alpha = 0.2,
             pch = 21,
             stroke = 0) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(120, 115), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_shared_pct_identity_plot 

regulation_tfs_identity_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  ggscatter(x = "shared_tf",
            y = "identity_human2mouse", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line", 
            conf.int = T
              ) +
  scale_color_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "Total Shared TFs",
       y = "%Identity (a.a.)",
       color = "",
       fill = "") +
  geom_point(aes(fill = cor_value),
             alpha = 0.2,
             pch = 21,
             stroke = 0) +
  scale_fill_gradient2(
    name = "r",
    low = "#4361ee",  
    mid = "white",  
    high = "#4cc9f0", 
    midpoint = 0) +
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(120, 115), 
           size = 4) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white") +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_total_identity_plot 
regulation_tfs_identity_plot
```

```{r fig.height=5, fig.width=5}
#Correlation
regulation_tfs_corvalue_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  mutate(quadrant = if_else(cor_value>0, "Positive correlation", "Negative correlation")) %>% 
  drop_na(quadrant) %>% 
  mutate(logp = -log(p_value),
         cor_value_abs = abs(cor_value)) %>% 
  ggscatter(x = "cor_value_abs",
            y = "total", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line",
            # conf.int = TRUE
              ) +
    scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "|Correlation|",
       y = "Total TFs",
       color = "",
       fill = "") +
  geom_point(aes(fill = logp),
             alpha = 1,
             pch = 21, 
             color = "black",
             stroke = 0,
             show.legend = T) +
  scale_fill_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
    guide = guide_coloursteps(
      # show.limits = TRUE,
      # barwidth = unit(4, "cm"), 
      # barheight = unit(0.2, "cm")    
    )
  ) +
  
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(750, 800), 
           size = 4, show.legend = F) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white", 
                  show.legend = F) +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_corvalue_plot

#Correlation
regulation_tfs_corvalue_shared_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  mutate(quadrant = if_else(cor_value>0, "Positive correlation", "Negative correlation")) %>% 
  drop_na(quadrant) %>% 
  ggscatter(x = "cor_value_abs",
            y = "shared_tf", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line",
            # conf.int = TRUE
              ) +
    scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "|Correlation|",
       y = "Shared TFs",
       color = "",
       fill = "") +
  geom_point(aes(fill = logp),
             alpha = 1,
             pch = 21, 
             color = "black",
             stroke = 0,
             show.legend = T) +
  ylim(0, 100) +
  scale_fill_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
    # guide = guide_coloursteps(
    #   show.limits = TRUE,
    #   barwidth = unit(4, "cm"), 
    #   barheight = unit(0.2, "cm")    
    # )
  ) +
  
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(100, 90), 
           size = 4, show.legend = F) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white", 
                  show.legend = F) +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_corvalue_shared_plot

#Correlation
regulation_tfs_corvalue_shared_perc_plot = gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  drop_na(cor_label) %>% 
  mutate(logp = -log(p_value)) %>% 
  ggscatter(x = "cor_value_abs",
            y = "shared_tf_pct", 
            fill = "classification",
            color = "classification",
            point = F,
            add = "reg.line",
            # conf.int = TRUE
              ) +
    scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  new_scale_color() +
  new_scale_fill() +
  labs(x = "|Correlation|",
       y = "%Shared TFs",
       color = "",
       fill = "") +
  geom_point(aes(fill = logp),
             alpha = 1,
             pch = 21, 
             color = "black",
             stroke = 0,
             show.legend = T) +
  # ylim(0, 100) +
  scale_fill_binned(
    name = "p-value",
    breaks = c(1, 1.3, 2), 
    type = "viridis", option = "mako",   
    direction = 1,
    labels = c("0.1", "0.05", "0.01"),
    # guide = guide_coloursteps(
    #   show.limits = TRUE,
    #   barwidth = unit(4, "cm"), 
    #   barheight = unit(0.2, "cm")    
    # )
  ) +
  
  new_scale_fill() +
  
  geom_point(aes(fill = classification),
             color = "black",
             size = 3,
             pch = 21,
             data =  . %>% 
                    filter(classification == "Shared"),
             show.legend = F) +
  stat_cor(mapping = aes(color = classification,
                         fill = classification),
           method = "spearman", 
           label.y = c(0.20, 0.18), 
           size = 4, show.legend = F) +
  geom_label_repel(aes(label = gene),
                   color = "black",
                   data =  . %>% 
                     filter(classification == "Shared"),
                  size = 2.5,
                  fill = "white", 
                  show.legend = F) +
  theme_vaxgo() +
  theme(legend.position = "right") +
  scale_color_manual(name = "Class",
                     values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  scale_fill_manual(name = "Class",
                    values = c("Shared" = "#4DBBD5FF",
                                "All" = "black")) +
  guides(color = "none")

regulation_tfs_corvalue_shared_perc_plot
```

```{r fig.height=6, fig.width=12}
# ── Patchwork: combine all 6 TF regulation panels into two rows (identity and correlation) and save
regulation_plots <- (regulation_tfs_total_identity_plot | regulation_tfs_identity_plot | regulation_tfs_shared_pct_identity_plot) /
  (regulation_tfs_corvalue_plot | regulation_tfs_corvalue_shared_plot | regulation_tfs_corvalue_shared_perc_plot) +
  plot_annotation(title = "Transcription factors vs. Aminoacid identity and Expression correlation",
                  subtitle = "Shared LEGs with p < 0.10 and rsq > 0.2 are highlighted in blue")

regulation_plots

# Save complete regulation figure
ggsave(regulation_plots,
       filename = here("Figures", "all_human_mouse_gsea_leganalysis_regulation_plots.png"),
       width = 12, height = 6)
```

```{r fig.height=4, fig.width=20}
# ── Alternative layout: single-row patchwork with 4 panels (identity + correlation) and save
regulation_plots <- (regulation_tfs_total_identity_plot | regulation_tfs_shared_pct_identity_plot | regulation_tfs_corvalue_plot | regulation_tfs_corvalue_shared_perc_plot) +
  plot_annotation(title = "Transcription factors vs. Aminoacid identity and Expression correlation",
                  subtitle = "Shared LEGs with p < 0.10 and rsq > 0.2 are highlighted in blue")

#Print
regulation_plots

# Save compact regulation figure
ggsave(regulation_plots,
       filename = here("Figures", "all_human_mouse_gsea_leganalysis_regulation_plots.png"),
       width = 20, height = 4)
```

#### By correlation

```{r}
gene_attribute_tf_count_filter_df = gene_attribute_tf_count %>% 
  ungroup() %>% 
  # Join with the filtered gene quadrants (e.g., High Corr/Low Identity)
  inner_join(alignments_human2mouse_genes_filtered,
             by = "gene")

gene_attribute_tf_count_filter_df_summary = gene_attribute_tf_count_filter_df %>% 
  # Create a descriptive label for the final table
  arrange(desc(both)) %>% 
  group_by(
          # filter, 
           gene, both) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  group_by(gene) %>% 
  mutate(total = sum(n),
         percentage = (n/total*100) %>% round(1))  



gene_attribute_tf_count_filter_df_summary_annotated = gene_attribute_tf_count_filter_df_summary %>%
  arrange(gene) %>% 
  distinct() %>%
    select(-percentage) %>% 
  pivot_wider(values_from = "n",
              names_from = "both") %>% 
  distinct() %>%
  rename(not_shared_tf = "F",
         shared_tf = "T") %>% 
  mutate(shared_tf = if_else(is.na(shared_tf), 0, shared_tf)) %>% 
  inner_join(gene_attribute_tf_count_filter_df %>% 
               ungroup() %>% 
               filter(class %in% c("Immune\nBTMs", "Non-Immune\nBTMs")) %>%
               select(gene, filter, identity_human2mouse, class:cor_label) %>% 
               distinct(),
             by = join_by(gene)) %>% 
  ungroup() %>% 
  mutate(shared_tf_pct = (shared_tf/total) %>% round(2))



#Correlation
gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  # mutate(quadrant = if_else(cor_value>0, "Positive correlation", "Negative correlation")) %>% 
  # drop_na(quadrant) %>% 
  ggplot() +
  aes(x = total,
      y = cor_value,
      label = gene) +
  # facet_wrap(~quadrant,
  #            scales = "free") +
  geom_point(alpha = 0.1) +
  geom_point(data = . %>% 
                filter(p_value < 0.10,
                       r_squared > 0.2),
             color = "black",
             alpha = 0.8) +
  geom_point(color = "black",
             size = 3,
             pch = 21,
             fill = "#4cc9f0",
             data =  gene_attribute_tf_count_filter_df_summary_annotated %>% 
                    inner_join(top_shared_notshared_genes %>% 
                                 filter(class == "All") %>% 
                                 filter(p_value < 0.10,
                                 r_squared > 0.2) %>% 
                                 select(gene) %>% 
                                 distinct(),
                               by = "gene")) +
  geom_label_repel(color = "black",
                   data =  gene_attribute_tf_count_filter_df_summary_annotated %>% 
                    inner_join(top_shared_notshared_genes %>% 
                                 filter(class == "All") %>%
                                 filter(p_value < 0.10,
                                 r_squared > 0.2) %>% 
                                 select(gene) %>% 
                                 distinct(),
                               by = "gene"),
                  size = 2.5,
                  fill = "white") +
  # geom_smooth(method = "lm") +
  # geom_smooth(method = "lm",
  #             data = . %>% 
  #               filter(p_value < 0.10,
  #                      r_squared > 0.2),
  #            color = "blue") +
  theme_vaxgo()



#Identity
gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  ggplot() +
  aes(x = total,
      y = identity_human2mouse,
      # label = gene
      ) +
  geom_point(alpha = 0.1) +
  geom_point(
    # data = . %>% 
    #             filter(p_value < 0.10,
    #                    r_squared > 0.2),
             color = "black",
             pch = 21,
             alpha = 0.8,
             mapping = aes(fill = filter)) +

  # geom_smooth(method = "lm") +
  theme_vaxgo()



gene_attribute_tf_count_filter_df_summary_annotated %>% 
  arrange(-shared_tf) %>% 
  ggplot() +
  aes(x = shared_tf,
      y = identity_human2mouse,
      # label = gene
      ) +
  geom_point(alpha = 0.1) +
  geom_point(
    # data = . %>% 
    #             filter(p_value < 0.10,
    #                    r_squared > 0.2),
             color = "black",
             pch = 21,
             alpha = 0.8,
             mapping = aes(fill = filter)) +

  # geom_smooth(method = "lm") +
  theme_vaxgo()

```















###------------------------

### Cis-regulatory elements

```{r packages}
source(here::here("scripts_notebooks", "required.R"))

library(BSgenome.Hsapiens.UCSC.hg38)
library(BSgenome.Mmusculus.UCSC.mm10)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(ChIPseeker)
library(GenomeInfoDb)
library(GenomicRanges)
library(IRanges)
library(rtracklayer)
library(org.Hs.eg.db)
library(pwalign)
library(biomaRt)
```


```{r import-data}
#Import Human and Mouse CREs ===========
#In contrast to the Human dataset, there are no CTCF-bound CREs in the main, full CRE mouse dataset. So, we imported CREs separately and merged them.

# Human CREs (GRCh38) =============

#PLS
cres_human_pls <- read_tsv(
  here("Genomic", "genomic_human_cCREs_GRCh38_PLS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_original_human = type_human,
         subtype_human = "not-CTCF") 

#pELS-dELS
cres_human_pelsdels <- read_tsv(
  here("Genomic", "genomic_human_cCREs_GRCh38_pELS-dELS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_original_human = type_human,
         subtype_human = "not-CTCF") 

#CTCF BOUND
cres_human_ctcfbound <- read_tsv(
  here("Genomic", "genomic_human_cCREs_GRCh38_CTCF-bound.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type_original"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_human = str_remove(type_original_human, ",CTCF-bound"),
         subtype_human = "CTCF")

#Check overlaps and remove CTCF-bound cCREs that are not annotated accordingly =====
#PLS
cres_human_ctcfbound_pls_overlap = cres_human_ctcfbound %>% 
  inner_join(cres_human_pls,
             by = "id_ccre_human") %>% 
  select(id_ccre_human) %>% 
  distinct() %>% 
  arrange(id_ccre_human) #36,430 IDs

#47,532 -> 11,102 IDs
cres_human_pls_clean = cres_human_pls %>% 
  filter(!id_ccre_human %in% cres_human_ctcfbound_pls_overlap$id_ccre_human) 

#pELS-dELS
cres_human_ctcfbound_pelsdels_overlap = cres_human_ctcfbound %>% 
  inner_join(cres_human_pelsdels,
             by = "id_ccre_human")%>% 
  select(id_ccre_human) %>% 
  distinct() #720,306 IDs

#1,718,669 -> 998,363
cres_human_pelsdels_clean = cres_human_pelsdels %>% 
  filter(!id_ccre_human %in% cres_human_ctcfbound_pelsdels_overlap$id_ccre_human)

#Merge
cres_human = bind_rows(cres_human_pls_clean,
          cres_human_pelsdels_clean,
          cres_human_ctcfbound) S

```

```{r import-data}
# Mouse CREs (mm10) ============
#PLS
cres_mouse_pls <- read_tsv(
  here("Genomic", "genomic_mouse_cCREs_mm10_PLS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_original_mouse = type_mouse,
         subtype_mouse = "not-CTCF") 

#pELS-dELS
cres_mouse_pelsdels <- read_tsv(
  here("Genomic", "genomic_mouse_cCREs_mm10_pELS-dELS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_original_mouse = type_mouse,
         subtype_mouse = "not-CTCF")

#CTCF BOUND
cres_mouse_ctcfbound <- read_tsv(
  here("Genomic", "genomic_mouse_cCREs_mm10_CTCF-bound.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type_original"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_mouse = str_remove(type_original_mouse, ",CTCF-bound"),
         subtype_mouse = "CTCF")

#Check overlaps and remove CTCF-bound cCREs that are not annotated accordingly =====
#PLS
cres_mouse_ctcfbound_pls_overlap = cres_mouse_ctcfbound %>% 
  inner_join(cres_mouse_pls,
             by = "id_ccre_mouse") %>% 
  select(id_ccre_mouse) %>% 
  distinct() %>% 
  arrange(id_ccre_mouse) #11.884 IDs

#27.332 -> 15,448 IDs
cres_mouse_pls_clean = cres_mouse_pls %>% 
  filter(!id_ccre_mouse %in% cres_mouse_ctcfbound_pls_overlap$id_ccre_mouse) 

#pELS-dELS
cres_mouse_ctcfbound_pelsdels_overlap = cres_mouse_ctcfbound %>% 
  inner_join(cres_mouse_pelsdels,
             by = "id_ccre_mouse")%>% 
  select(id_ccre_mouse) %>% 
  distinct() #66.186 IDs

#512.001 -> 445,815 
cres_mouse_pelsdels_clean = cres_mouse_pelsdels %>% 
  filter(!id_ccre_mouse %in% cres_mouse_ctcfbound_pelsdels_overlap$id_ccre_mouse)

#Merge
cres_mouse = bind_rows(cres_mouse_pls_clean,
          cres_mouse_pelsdels_clean,
          cres_mouse_ctcfbound)

```

Import homologous CREs

```{r}
# Homologous CRE pairs from ENCODE (hg38–mm10) Nearest gene, protein coding only
homologous_cres <- read_tsv(
  here("Genomic", "genomic_human_mouse_cCREs_homologous_GRCh38_Mm10.tsv"),
  col_names = c("human_enhancer", "mouse_enhancer"),
  show_col_types = FALSE
) #537,822

# Load ENCODE comprehensive gene-CRE association table
genes_cres_human <- read_tsv(
  here("Genomic", "genomic_human_cCREs_GRCh38_Closest-Genes-PC.tsv"),
  col_names = c("chrom_cre", "start_cre", "end_cre", "id_dhs", "id_ccre", "type_cre",
                "chrom_gene", "start_gene", "end_gene", "transcript_id", "score_gene", "strand_gene", "gene_id"),
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) #2,651,787
```
Annotate with GeneHancer
```{r}
########### GENEHANCER

#Gene interactions with enhancers
genehancer_interactions_doubleelite = read_tsv(here("Genomic", "GeneHancer_Chr1-22_ChrX_Interactions_DoubleElite_hgTables.txt")) %>% 
  separate(name, into = c("hgnc_symbol", "genehancer_id")) %>%
  filter(chromEnd >= chromStart) %>% 
  rename(geneScore = score,
         geneValue = value,)
                                               
#Enhancers                                              
genehancer_enhancers_doubleelite = read_tsv(here("Genomic", "GeneHancer_Chr1-22_ChrX_Enhancers_DoubleElite_hgTables.txt")) %>%
  filter(chromEnd >= chromStart)


########## Get overlap between ENCODE4 and Genehancer
library(GenomicRanges)
#Convert to GRanges
genehancer_enhancers_doubleelite_ranges = makeGRangesFromDataFrame(genehancer_enhancers_doubleelite, 
                                  keep.extra.columns = TRUE,
                                  seqnames.field = "chrom",
                                  start.field = "chromStart",
                                  end.field = "chromEnd", 
                                  na.rm = T)

#Convert to GRanges
cres_human_ranges = makeGRangesFromDataFrame(cres_human, 
                                  keep.extra.columns = TRUE,
                                  seqnames.field = "chrom_human",
                                  start.field = "start_human",
                                  end.field = "end_human", 
                                  na.rm = T)
# Find overlaps
distances_ranges <- distanceToNearest(genehancer_enhancers_doubleelite_ranges, 
                     cres_human_ranges) %>% 
  as.data.frame()

#Annotate Genehancer ids (ELITE) with ENCODE4 ids
cres_human_encode_genehancer <- data.frame(
  genehancer_enhancers_doubleelite_ranges[distances_ranges$queryHits], 
  cres_human_ranges[distances_ranges$subjectHits],
  distance = distances_ranges$distance
) %>% 
  rename_with(
    ~ str_replace(., "\\.1$", "_encode4"), 
    ends_with(".1")                        
  ) %>% 
  rename(chrom = seqnames,
         chromStart = start,
         chromEnd = end,
         genehancer_id = name) %>%
  
  #Merge gene interaction scores
  inner_join(genehancer_interactions_doubleelite,
             by = join_by("genehancer_id", "chrom", "chromStart", "chromEnd")
             )


cres_human_encode_genehancer %>% 
  glimpse()
```




Summarize data Count how many mouse CREs each human CRE maps to (and vice versa)

```{r classify-match-type}

#Summarize classes
cres_human_encode_genehancer %>% 
  dplyr::count(type_human, subtype_human) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_human, desc(n))

cres_mouse %>% 
  dplyr::count(type_mouse, subtype_mouse) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_mouse, desc(n))

# Count how many mouse CREs each human CRE maps to (and vice versa)
human_counts <- homologous_cres %>%
  dplyr::count(human_enhancer, name = "n_mouse_per_human")

mouse_counts <- homologous_cres %>%
  dplyr::count(mouse_enhancer, name = "n_human_per_mouse")

# Full join: preserves orphan CREs on both sides
cres_annotated <- homologous_cres %>%
  left_join(human_counts, by = "human_enhancer") %>%
  left_join(mouse_counts, by = "mouse_enhancer") %>%
  full_join(cres_human_encode_genehancer, by = join_by(human_enhancer == id_ccre_human)) %>%
  full_join(cres_mouse, by = join_by(mouse_enhancer == id_ccre_mouse)) %>%
  mutate(
    is_homologous = !is.na(human_enhancer) & !is.na(mouse_enhancer),
    match_type = case_when(
      is.na(human_enhancer)                            ~ "mouse_only",
      is.na(mouse_enhancer)                            ~ "human_only",
      n_mouse_per_human == 1 & n_human_per_mouse == 1  ~ "1:1",
      n_mouse_per_human >  1 & n_human_per_mouse == 1  ~ "N:1",
      n_mouse_per_human == 1 & n_human_per_mouse >  1  ~ "1:N",
      TRUE                                             ~ "N:M"
    )
  ) %>% 
  rename(chrom_human = chrom,
         start_human = chromStart,
         end_human = chromEnd,
         )

#Filter 1 to 1
cres_homologous_1to1 <- cres_annotated %>% #856,415
  dplyr::filter(match_type == "1:1") %>% #277,898
  #Filter enhancers with complete annotation in both species
  dplyr::filter(!is.na(chrom_human), !is.na(chrom_mouse))#1,567

# Separate tables
cres_homologous <- cres_annotated %>% dplyr::filter(is_homologous)
cres_human_only <- cres_annotated %>% dplyr::filter(match_type == "human_only")
cres_mouse_only <- cres_annotated %>% dplyr::filter(match_type == "mouse_only")
```

Visualize

```{r classify-match-type, fig.height=5, fig.width=6}
#Summarize
cres_homologous_1to1$type_human %>% unique()
#Compare classes of CREs between species
cres_homologous_1to1_comparison = cres_homologous_1to1 %>% 
  
  #Filter PLS, pELS, dELS
    filter(type_human %in% c("pELS", "dELS", "PLS"),
         type_mouse %in% c("pELS", "dELS", "PLS")) %>% 
  
  #Summarize
  group_by(type_human, subtype_human) %>% 
  dplyr::count(type_human, subtype_human, type_mouse, subtype_mouse) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(desc(n)) %>%
  mutate(type_human = fct_relevel(type_human, "PLS", "pELS", "dELS"),
         type_mouse = fct_relevel(type_mouse, "PLS", "pELS", "dELS")
         )
```

```{r classify-match-type, fig.height=5, fig.width=5}
#Heatmap

#Elite CREs
cres_homologous %>% 
    dplyr::filter(!is.na(hgnc_symbol)) %>% #1,706
  #Filter ELITE enhancers
  dplyr::filter(eliteness == "Elite") %>%  ##1,567
  #Plot
  ggplot() +
    aes(x = subtype_mouse %>% str_replace(., "-", "\n"),
         y = subtype_human %>% str_replace(., "-", "\n")) +
  facet_grid(~type_human~type_mouse) +
  geom_tile(aes(fill = pct), color = "black", linewidth = 0.4) +
  geom_text(aes(label = paste0(pct, "%")),
            size = 3) +
  theme_vaxgo() +
  scale_fill_gradient2(low = "white",
                       high = "#4361ee",
                       guide = guide_colorbar(barwidth = unit(2, "cm"))) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  theme(axis.text.x = element_text(angle = 0,
                                   # vjust = 0,
                                   hjust = 0.5),
        legend.position = "top") +
  labs(x = "Mouse",
       y = "Human",
       fill = "%Shared, from human")


# All CREs
cres_homologous %>% 
  #Plot
  ggplot() +
    aes(x = subtype_mouse %>% str_replace(., "-", "\n"),
         y = subtype_human %>% str_replace(., "-", "\n")) +
  facet_grid(~type_human~type_mouse) +
  geom_tile(aes(fill = pct), color = "black", linewidth = 0.4) +
  geom_text(aes(label = paste0(pct, "%")),
            size = 3) +
  theme_vaxgo() +
  scale_fill_gradient2(low = "white",
                       high = "#4361ee",
                       guide = guide_colorbar(barwidth = unit(2, "cm"))) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  theme(axis.text.x = element_text(angle = 0,
                                   # vjust = 0,
                                   hjust = 0.5),
        legend.position = "top") +
  labs(x = "Mouse",
       y = "Human",
       fill = "%Shared, from human")


```

------------------------------------------------------------------------

#### Gene annotation

```{r import-gene-annotation}

###############
# ADAPT
###############

# Import gene expression correlation
# Mock
expression_correlation <- tibble(
  hgnc_symbol = c("ENTPD1", "ASGR1", "MARCKS", "IL7R"),
  mgi_symbol  = c("Entpd1", "Asgr1", "Marcks", "Il7r"),
  rho         = c(0.85, 0.79, -0.42, -0.61)
)

all_human_mouse_dge_btm_process_genes_correlation_bygene <- readRDS( here("tables", "all_human_mouse_dge_btm_process_genes_correlation_bygene.rds"))

############## 
# Query biomaRt to map human Ensembl IDs to HGNC and MGI symbols
human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

hgnc_mgi_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = unique(cres_human_encode_genehancer$gene_id_clean),
  mart       = human_mart
)

# Join HGNC symbols to gene table
genes_cres_human <- genes_cres_human %>%
  left_join(hgnc_mgi_map, by = c("gene_id_clean" = "ensembl_gene_id"))


# Compute multi-CRE metrics per homologous gene pair
gene_level_cre_metrics <- genes_cres_human %>%
  group_by(hgnc_symbol, type_cre) %>%
  summarise(
    n_cres_human       = n(),
    n_cres_orthologous = sum(id_ccre %in% cres_homologous_1to1$human_enhancer),
    pct_cres_conserved = n_cres_orthologous / n_cres_human,
    .groups = "drop"
  ) 
  # inner_join(all_human_mouse_dge_btm_process_genes_correlation_bygene, by = c("hgnc_symbol" = "hgnc_symbol"))


# Annotate cres_annotated with human gene info (one gene per CRE)
cres_annotated <- cres_annotated %>%
  left_join(
    genes_cres_human %>%
      dplyr::select(id_ccre, gene_id, hgnc_symbol,                     transcript_id, chrom_gene, start_gene, end_gene, strand_gene) %>%
      distinct(id_ccre, .keep_all = TRUE),
    by = c("human_enhancer" = "id_ccre")
  )

```



(DOES NOT WORK) Get evolutionary distance

```{r}
# library(biomaRt)
# 
# # 1. Definir o banco de dados antigo usando o host do archive desejado (ex: jan2020)
# # Como você quer partir do camundongo para buscar os homólogos humanos, usamos "mmusculus_gene_ensembl"
# listEnsemblArchives()
# listMarts()
# 
# options(timeout = 600)
# 
# # ensembl_human_old = useEnsembl(biomart = "ensembl",
# #                        dataset="hsapiens_gene_ensembl",
# #                        version="77")
# 
# ensembl_human_old = useMart(biomart = "ENSEMBL_MART_ENSEMBL",
#                              dataset="hsapiens_gene_ensembl",
#                             host = "https://oct2014.archive.ensembl.org")
# 
# listFilters(ensembl_human_old) %>% View()
# listAttributes(ensembl_human_old) %>% View()
# 
# # 2. Carregar sua lista de IDs de genes de camundongo (ex: ENSMUSG0000XXXXXXXX)
# human_genes_list <- all_human_mouse_dge_btm_process_genes_correlation_bygene$symbol
# 
# # 3. Fazer a busca trazendo o ID do camundongo e os dados do homólogo humano (hsapiens)
# options(timeout = 600)
# human_mouse_orthologs_old <- getBM(attributes = c(
#                                                   "ensembl_gene_id",
#                                                   "external_gene_name"
#                                           # "mmusculus_homolog_dn", 
#                                           # "mmusculus_homolog_ds"
#                                           ), 
#                             filters = 'external_gene_name', 
#                             values = "CD6", 
                            # mart = ensembl_human_old)

```






How many CREs per gene? And how many genes per enhancer?
```{r fig.height=5, fig.width=10}
#Distribution
gene_level_cre_metrics %>% 
  filter(hgnc_symbol != "") %>% 
  ggplot() +
  aes(x = n_cres_human,
      y = "Human",
      label = hgnc_symbol) +
  geom_violin(fill = "gray90",
              color = NA) +
  geom_boxplot(width = 0.2,
               color = "#4DBBD5FF",
               alpha = 0) +
  scale_x_continuous(transform = "log10") +
  geom_quasirandom(groupOnX = FALSE, # Permite a distribuição vertical em um X contínuo
                 varwidth = TRUE,  # Ajusta a largura de acordo com a densidade
                 pch = 21,
                 mapping = aes(fill = pct_cres_conserved)) +
  scale_color_continuous() +
  geom_label_repel(size = 2.5,
                  data = . %>% slice_max(n = 5, 
                                         order_by = n_cres_human)) +
  theme_vaxgo()

```


```{r fig.height=4, fig.width=10}

library(ggpubr)
#Is there any correlation between CREs metrics and types of CREs?
gene_level_cre_metrics %>% 
  inner_join(all_human_mouse_dge_btm_process_genes_correlation_bygene %>% 
               rename(hgnc_symbol = symbol), 
             by = join_by(hgnc_symbol)) %>% 
  filter(hgnc_symbol != "",
         # p_value <= 0.10,
         n_pathogens > 10,
         type_cre %in% c("PLS", "pELS", "dELS", "TF")) %>% 
  ggplot() +
  aes(
      x = n_cres_orthologous,
      # x = n_cres_human,
      # x = pct_cres_conserved,
      y = cor_value, 
      label = hgnc_symbol,
      color = type_cre,
      # size = n_pathogens
      ) +
  theme_vaxgo() +
  facet_wrap(~type_cre, nrow = 1, 
             scales = "free_x")+
  geom_point(alpha = 0.2,
             ) +
  geom_smooth(method = "lm", color = "black", se = TRUE, size = 0.8) +
  
  stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top", color = "black") +
  theme(legend.position = "top") 
  
  # geom_label_repel(color = "black",
  #                 size = 2.5,
  #                 data = . %>% 
  #                   slice_max(pct_cres_conserved, n = 10))


```



------------------------------------------------------------------------

####  Sequence extraction

The HOXD55 matrix (originally derived by Chiaromonte et al., 2002, and deeply analyzed in Blanchette et al., 2004) defines nucleotide matches and mismatches using large integers to maximize sensitivity when calculating neutral substitution rates in mammalian alignments:Match: $+91$Mismatch: $-114$Because a single mismatch penalizes the score by $-114$, the gap penalties must be scaled proportionally. If you used a standard gap opening penalty of $-10$, the alignment algorithm would erroneously introduce hundreds of micro-gaps instead of accepting a single nucleotide mismatch, completely ruining the biological alignment of your cis-regulatory elements (CREs).

To maintain perfect parity with the alignments you see on the UCSC Genome Browser for human (hg38) and mouse (mm10), you use their default parameters:Gap Opening Penalty ($O$): $-400$Gap Extension Penalty ($E$): $-30$

Chiaromonte, F., et al. (2002). Scoring pairwise genomic alignments. Pacific Symposium on Biocomputing. This paper established the ancestral mammalian repeat substitution matrices (including the HOXD matrices). Blanchette, M., et al. (2004). Blastz: A tool for aligning a large number of long genomic sequences. Genome Research. This paper formally documents the open-source BLASTZ/LASTZ defaults ($O = 400, E = 30$, match $= 91$, mismatch $= 114$) used by UCSC. Kent, W. J., et al. (2003). Evolution's Cauldron: Duplication, Deletion, and Rearrangement in the Mouse and Human Genomes. PNAS. Documents how UCSC constructs the chain and net alignments utilizing these exact parameters.

Align homolog enhancer sequences

```{r extract-sequences}
#Filter homologs with mapped coordinates
cres_to_align_homologous <- cres_homologous_1to1 %>%
  filter(human_enhancer %in% genes_cres_human$id_ccre) %>%
  filter(!is.na(chrom_human), !is.na(chrom_mouse)) %>%
  #Annotate gene symbol
  inner_join(cres_annotated %>% 
            select(human_enhancer, mouse_enhancer, hgnc_symbol),
        by = join_by(human_enhancer, mouse_enhancer))

#Convert to GRanges objects
gr_human <- GRanges(seqnames = cres_to_align_homologous$chrom_human, ranges = IRanges(start = cres_to_align_homologous$start_human, end = cres_to_align_homologous$end_human))
gr_mouse <- GRanges(seqnames = cres_to_align_homologous$chrom_mouse, ranges = IRanges(start = cres_to_align_homologous$start_mouse, end = cres_to_align_homologous$end_mouse))

#Get sequences
cres_to_align_homologous$sequence_human <- as.character(getSeq(BSgenome.Hsapiens.UCSC.hg38, gr_human))
cres_to_align_homologous$sequence_mouse <- as.character(getSeq(BSgenome.Mmusculus.UCSC.mm10, gr_mouse))

#Clean sequences from undefined characters
cres_to_align_homologous_clean = cres_to_align_homologous %>% 
  mutate(sequence_human = str_remove_all(sequence_human, "N"),
         sequence_mouse = str_remove_all(sequence_mouse, "N"))

# Parameterize a nucleotide substitution matrix mirroring the standard HOXD55 values
matriz_hoxd55 <- nucleotideSubstitutionMatrix(match = 91, mismatch = -114, baseOnly = TRUE)

#Align sequences
aln_cres <- pairwiseAlignment(
  DNAStringSet(cres_to_align_homologous_clean$sequence_human),
  DNAStringSet(cres_to_align_homologous_clean$sequence_mouse),
  type         = "local",
  substitutionMatrix = matriz_hoxd55,
  gapOpening   = -400, # Score scale adjusted for standard UCSC genomic alignment scoring
  gapExtension = -30
)

#Calculate percent identity
cres_to_align_homologous_identity <- cres_to_align_homologous_clean %>%
  mutate(
    cre_identity_pct = pid(aln_cres, type = "PID1"),
    cre_align_score  = score(aln_cres),
    cre_align_length = nchar(alignedPattern(aln_cres)),
    cre_coverage_human = cre_align_length / nchar(sequence_human)
  )



#Save results
cres_to_align_homologous_identity %>% 
  saveRDS(file = here("tables", "cres_homologous_1to1_annotated_aligned.rds"))

#Summarize alignment results
summary(cres_to_align$cre_identity_pct)

cat("\n% identity > 80%:", round(mean(cres_to_align$cre_identity_pct > 80) * 100, 1), "%\n")
cat("% identity > 70%:", round(mean(cres_to_align$cre_identity_pct > 70) * 100, 1), "%\n")
cat("% identity > 60%:", round(mean(cres_to_align$cre_identity_pct > 60) * 100, 1), "%\n")

#Which genes have the most conserved enhancers?
cres_to_align %>% 
  select(hgnc_symbol, human_enhancer, mouse_enhancer, type_human, type_mouse, cre_identity_pct:cre_coverage_human) %>% 
  group_by(hgnc_symbol, type_human, type_mouse) %>% 
  summarise(median_pct_id = median(cre_identity_pct)) %>%
  arrange(-median_pct_id) 

```

####  Alignment of proximal TSS windows

```{r}
tx_human <- transcripts(TxDb.Hsapiens.UCSC.hg38.knownGene)

# Multi-scale biological window definitions centered on the TSS
windows_definition <- list(
  w3000 = list(up = 3000, down = 300),  # Full proximal regulatory landscape (pELS/dELS capture)
  w1500 = list(up = 1500, down = 300),  # Core promoter + immediate proximal enhancers
  w500  = list(up = 500,  down = 100),  # Core promoter locus only
  w200  = list(up = 200,  down = 50)    # Minimal transcription initiation footprint
)

# Exemplary extraction and global alignment (Needleman-Wunsch) loop initialization for w3000
tss_human_w3000 <- promoters(TxDb.Hsapiens.UCSC.hg38.knownGene, upstream = 3000, downstream = 300)
tss_mouse_w3000 <- promoters(TxDb.Mmusculus.UCSC.mm10.knownGene, upstream = 3000, downstream = 300)

# Implementation Note: Filter your GRanges objects by matched homologous pairs prior to running global alignment:
aln_global <- pairwiseAlignment(seq_human_gene, seq_mouse_gene, type = "global", substitutionMatrix = matriz_hoxd55)
```

####  Annotate conservation and TEs

```{r}
ah <- AnnotationHub()

# 1. Fetch the most recent RepeatMasker records for hg38
rmsk_query <- query(ah, c("RepeatMasker", "Homo sapiens", "hg38"))
rmsk_hg38  <- ah[[names(rmsk_query)[1]]]

# 2. Compute the exact fraction of the CRE masked by repetitive components
gr_cres_human_all <- GRanges(
  seqnames = cres_to_align$chrom_human, 
  ranges = IRanges(start = cres_to_align$start_human, end = cres_to_align$end_human)
)

hits_rmsk <- findOverlaps(gr_cres_human_all, rmsk_hg38)

# Calculate coverage to map base-pair level masking within the ranges
repeat_cov <- GenomicRanges::coverage(rmsk_hg38)
seqlevels(gr_cres_human_all, pruning.mode = "coercion") <- names(repeat_cov)
masked_views <- Views(repeat_cov, ranges(gr_cres_human_all))

cres_to_align$repeat_masked_pct <- sapply(masked_views, function(x) sum(as.numeric(x) > 0)) / width(gr_cres_human_all) * 100
```

#### Integrative Statistical Modeling (Linear Mixed-Effects Models)

Because individual homologous genes map to multiple independent flanking CREs throughout the genome, we employ a Linear Mixed-Effects Model (LMM). Treating the gene pair as a random effect handles the nested data dependency and protects against degrees-of-freedom inflation.

```{r}
# Consolidate feature matrix across all annotation scales
df_integrated <- cres_to_align %>%
  select(human_enhancer, mouse_enhancer, cre_identity_pct, repeat_masked_pct) %>%
  left_join(genes_cres_human, by = join_by(human_enhancer == id_ccre)) %>%
  left_join(gene_level_cre_metrics, by = "hgnc_symbol") %>%
  mutate(gene_pair_id = paste0(hgnc_symbol, "_", mmusculus_homolog_associated_gene_name)) %>%
  filter(!is.na(rho))

# Fit Linear Mixed-Effects Model
lmm_fit <- lmer(
  rho ~ cre_identity_pct + pct_cres_conserved + repeat_masked_pct + (1 | gene_pair_id),
  data = df_integrated
)

# Extract clean model coefficients and variance components
tidy(lmm_fit)
```

Export

```{r export}
# Merge alignment results back into cres_annotated
cres_annotated <- cres_annotated %>%
  left_join(
    cres_to_align %>%
      dplyr::select(human_enhancer, mouse_enhancer,
                    sequence_human, sequence_mouse,
                    cre_identity_pct, cre_align_score, cre_align_length),
    by = c("human_enhancer", "mouse_enhancer")
  )

# Update split tables
# cres_homologous <- cres_annotated %>% filter(is_homologous)
# cres_human_only <- cres_annotated %>% filter(match_type == "human_only")
# cres_mouse_only <- cres_annotated %>% filter(match_type == "mouse_only")

# Save
saveRDS(cres_annotated,  here("tables", "cres_annotated_full.rds"))
# saveRDS(cres_homologous, here("tables", "cres_homologous_annotated.rds"))
# saveRDS(cres_human_only, here("tables", "cres_human_only_annotated.rds"))
# saveRDS(cres_mouse_only, here("tables", "cres_mouse_only_annotated.rds"))

message("All outputs saved to tables/")
```

