---
title: "6 — Genomic CRE Annotation"
subtitle: "Comparative annotation of human and mouse cis-regulatory elements"
author: "Wasim Aluísio Prates-Syed"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

# Setup

```{r packages}
source(here::here("scripts_notebooks", "required.R"))

library(BSgenome.Hsapiens.UCSC.hg38)
library(BSgenome.Mmusculus.UCSC.mm10)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(ChIPseeker)
library(GenomeInfoDb)
library(GenomicRanges)
library(IRanges)
library(rtracklayer)
library(org.Hs.eg.db)
library(pwalign)
library(biomaRt)
```

---

# Import

```{r import-data}
# Homologous CRE pairs from ENCODE (hg38–mm10)
homologous_cres <- read_tsv(
  here("Genomic", "hg38-mm10-Homologous.tsv"),
  col_names = c("human_enhancer", "mouse_enhancer"),
  show_col_types = FALSE
)

# Human CREs (GRCh38)
cres_human <- read_tsv(
  here("Genomic", "GRCh38-cCREs.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human"))

# Mouse CREs (mm10)
cres_mouse <- read_tsv(
  here("Genomic", "mm10-cCREs.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse"))


# Load ENCODE comprehensive gene-CRE association table
genes_cres_human <- read_tsv(
  here("Genomic", "GRCh38-Closest-Genes-PC.tsv"),
  col_names = c("chrom_cre", "start_cre", "end_cre", "id_dhs", "id_ccre", "type_cre",
                "chrom_gene", "start_gene", "end_gene", "transcript_id", "score_gene", "strand_gene", "gene_id"),
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))
```

```{r classify-match-type}
# Count how many mouse CREs each human CRE maps to (and vice versa)
human_counts <- homologous_cres %>%
  count(human_enhancer, name = "n_mouse_per_human")

mouse_counts <- homologous_cres %>%
  count(mouse_enhancer, name = "n_human_per_mouse")

# Full join: preserves orphan CREs on both sides
cres_annotated <- homologous_cres %>%
  left_join(human_counts, by = "human_enhancer") %>%
  left_join(mouse_counts, by = "mouse_enhancer") %>%
  full_join(cres_human, by = join_by(human_enhancer == id_ccre_human)) %>%
  full_join(cres_mouse, by = join_by(mouse_enhancer == id_ccre_mouse)) %>%
  mutate(
    is_homologous = !is.na(human_enhancer) & !is.na(mouse_enhancer),
    match_type = case_when(
      is.na(human_enhancer)                            ~ "mouse_only",
      is.na(mouse_enhancer)                            ~ "human_only",
      n_mouse_per_human == 1 & n_human_per_mouse == 1  ~ "1:1",
      n_mouse_per_human >  1 & n_human_per_mouse == 1  ~ "N:1",
      n_mouse_per_human == 1 & n_human_per_mouse >  1  ~ "1:N",
      TRUE                                             ~ "N:M"
    )
  )

#Filter 1 to 1
cres_homologous_1to1 <- cres_annotated %>% 
  filter(match_type == "1:1") %>% 
  #Filter enhancers with complete annotation in both species
  filter(!is.na(chrom_human), !is.na(chrom_mouse))

# Separate tables
cres_homologous <- cres_annotated %>% filter(is_homologous)
cres_human_only <- cres_annotated %>% filter(match_type == "human_only")
cres_mouse_only <- cres_annotated %>% filter(match_type == "mouse_only")


#Summarize
cres_annotated %>%
  count(match_type) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n))
```

---

# Gene annotation

```{r import-gene-annotation}

###############
# ADAPT
###############

# Import gene expression correlation
# Mock
expression_correlation <- tibble(
  hgnc_symbol = c("ENTPD1", "ASGR1", "MARCKS", "IL7R"),
  mgi_symbol  = c("Entpd1", "Asgr1", "Marcks", "Il7r"),
  rho         = c(0.85, 0.79, -0.42, -0.61)
)


############## 
# Query biomaRt to map human Ensembl IDs to HGNC and MGI symbols
human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

hgnc_mgi_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol", "mmusculus_homolog_associated_gene_name"),
  filters    = "ensembl_gene_id",
  values     = unique(genes_cres_human$gene_id_clean),
  mart       = mart_human
)

# Join HGNC symbols to gene table
genes_cres_human <- genes_cres_human %>%
  left_join(hgnc_mgi_map, by = c("gene_id_clean" = "ensembl_gene_id")) %>%
  # Restrict to classic, highly interpretable signatures
  filter(type_cre %in% c("PLS", "pELS", "dELS")) 


# Compute multi-CRE metrics per homologous gene pair
gene_level_cre_metrics <- genes_cres_human %>%
  group_by(hgnc_symbol, mmusculus_homolog_associated_gene_name) %>%
  summarise(
    n_cres_human       = n(),
    n_cres_orthologous = sum(id_ccre %in% cres_1to1$human_enhancer),
    pct_cres_conserved = n_cres_orthologous / n_cres_human,
    .groups = "drop"
  ) %>%
  inner_join(expression_correlation, by = c("hgnc_symbol" = "hgnc_symbol"))


# Annotate cres_annotated with human gene info (one gene per CRE)
cres_annotated <- cres_annotated %>%
  left_join(
    genes_cres_human %>%
      dplyr::select(id_ccre, gene_id, hgnc_symbol, gene_biotype,
                    transcript_id, chrom_gene, start_gene, end_gene, strand_gene) %>%
      distinct(id_ccre, .keep_all = TRUE),
    by = c("human_enhancer" = "id_ccre")
  )

```

---

# Sequence extraction
The HOXD55 matrix (originally derived by Chiaromonte et al., 2002, and deeply analyzed in Blanchette et al., 2004) defines nucleotide matches and mismatches using large integers to maximize sensitivity when calculating neutral substitution rates in mammalian alignments:Match: $+91$Mismatch: $-114$Because a single mismatch penalizes the score by $-114$, the gap penalties must be scaled proportionally. If you used a standard gap opening penalty of $-10$, the alignment algorithm would erroneously introduce hundreds of micro-gaps instead of accepting a single nucleotide mismatch, completely ruining the biological alignment of your cis-regulatory elements (CREs).

To maintain perfect parity with the alignments you see on the UCSC Genome Browser for human (hg38) and mouse (mm10), you use their default parameters:Gap Opening Penalty ($O$): $-400$Gap Extension Penalty ($E$): $-30$

Chiaromonte, F., et al. (2002). Scoring pairwise genomic alignments. Pacific Symposium on Biocomputing. This paper established the ancestral mammalian repeat substitution matrices (including the HOXD matrices).
Blanchette, M., et al. (2004). Blastz: A tool for aligning a large number of long genomic sequences. Genome Research. This paper formally documents the open-source BLASTZ/LASTZ defaults ($O = 400, E = 30$, match $= 91$, mismatch $= 114$) used by UCSC.
Kent, W. J., et al. (2003). Evolution's Cauldron: Duplication, Deletion, and Rearrangement in the Mouse and Human Genomes. PNAS. Documents how UCSC constructs the chain and net alignments utilizing these exact parameters.

```{r extract-sequences}
cres_to_align <- cres_1to1 %>%
  filter(human_enhancer %in% genes_cres_human$id_ccre) %>%
  filter(!is.na(chrom_human), !is.na(chrom_mouse))

gr_human <- GRanges(seqnames = cres_to_align$chrom_human, ranges = IRanges(start = cres_to_align$start_human, end = cres_to_align$end_human))
gr_mouse <- GRanges(seqnames = cres_to_align$chrom_mouse, ranges = IRanges(start = cres_to_align$start_mouse, end = cres_to_align$end_mouse))

cres_to_align$sequence_human <- as.character(getSeq(BSgenome.Hsapiens.UCSC.hg38, gr_human))
cres_to_align$sequence_mouse <- as.character(getSeq(BSgenome.Mmusculus.UCSC.mm10, gr_mouse))

# Parameterize a nucleotide substitution matrix mirroring the standard HOXD55 values
matriz_hoxd55 <- nucleotideSubstitutionMatrix(match = 91, mismatch = -114, baseOnly = TRUE)

aln_cres <- pairwiseAlignment(
  DNAStringSet(cres_to_align$sequence_human),
  DNAStringSet(cres_to_align$sequence_mouse),
  type         = "local",
  substitutionMatrix = matriz_hoxd55,
  gapOpening   = -400, # Score scale adjusted for standard UCSC genomic alignment scoring
  gapExtension = -30
)

cres_to_align <- cres_to_align %>%
  mutate(
    cre_identity_pct = pid(aln_cres, type = "PID1"),
    cre_align_score  = score(aln_cres),
    cre_align_length = nchar(alignedPattern(aln_cres)),
    cre_coverage_human = cre_align_length / nchar(sequence_human)
  )

```




```{r alignment-summary}
summary(cres_to_align$cre_identity_pct)

cat("\n% identity > 80%:", round(mean(cres_to_align$cre_identity_pct > 80) * 100, 1), "%\n")
cat("% identity > 70%:", round(mean(cres_to_align$cre_identity_pct > 70) * 100, 1), "%\n")
cat("% identity > 60%:", round(mean(cres_to_align$cre_identity_pct > 60) * 100, 1), "%\n")
```


# Multiscale alignment of proximal TSS windows
```{r}
tx_human <- transcripts(TxDb.Hsapiens.UCSC.hg38.knownGene)

# Multi-scale biological window definitions centered on the TSS
windows_definition <- list(
  w3000 = list(up = 3000, down = 300),  # Full proximal regulatory landscape (pELS/dELS capture)
  w1500 = list(up = 1500, down = 300),  # Core promoter + immediate proximal enhancers
  w500  = list(up = 500,  down = 100),  # Core promoter locus only
  w200  = list(up = 200,  down = 50)    # Minimal transcription initiation footprint
)

# Exemplary extraction and global alignment (Needleman-Wunsch) loop initialization for w3000
tss_human_w3000 <- promoters(TxDb.Hsapiens.UCSC.hg38.knownGene, upstream = 3000, downstream = 300)
tss_mouse_w3000 <- promoters(TxDb.Mmusculus.UCSC.mm10.knownGene, upstream = 3000, downstream = 300)

# Implementation Note: Filter your GRanges objects by matched homologous pairs prior to running global alignment:
# aln_global <- pairwiseAlignment(seq_human_gene, seq_mouse_gene, type = "global", substitutionMatrix = matriz_hoxd55)
```



# Annotate conservation and TEs

```{r}
ah <- AnnotationHub()

# 1. Fetch the most recent RepeatMasker records for hg38
rmsk_query <- query(ah, c("RepeatMasker", "Homo sapiens", "hg38"))
rmsk_hg38  <- ah[[names(rmsk_query)[1]]]

# 2. Compute the exact fraction of the CRE masked by repetitive components
gr_cres_human_all <- GRanges(
  seqnames = cres_to_align$chrom_human, 
  ranges = IRanges(start = cres_to_align$start_human, end = cres_to_align$end_human)
)

hits_rmsk <- findOverlaps(gr_cres_human_all, rmsk_hg38)

# Calculate coverage to map base-pair level masking within the ranges
repeat_cov <- GenomicRanges::coverage(rmsk_hg38)
seqlevels(gr_cres_human_all, pruning.mode = "coercion") <- names(repeat_cov)
masked_views <- Views(repeat_cov, ranges(gr_cres_human_all))

cres_to_align$repeat_masked_pct <- sapply(masked_views, function(x) sum(as.numeric(x) > 0)) / width(gr_cres_human_all) * 100
```



#Integrative Statistical Modeling (Linear Mixed-Effects Models)

Because individual homologous genes map to multiple independent flanking CREs throughout the genome, we employ a Linear Mixed-Effects Model (LMM). Treating the gene pair as a random effect handles the nested data dependency and protects against degrees-of-freedom inflation.

```{r}
# Consolidate feature matrix across all annotation scales
df_integrated <- cres_to_align %>%
  select(human_enhancer, mouse_enhancer, cre_identity_pct, repeat_masked_pct) %>%
  left_join(genes_cres_human, by = join_by(human_enhancer == id_ccre)) %>%
  left_join(gene_level_cre_metrics, by = "hgnc_symbol") %>%
  mutate(gene_pair_id = paste0(hgnc_symbol, "_", mmusculus_homolog_associated_gene_name)) %>%
  filter(!is.na(rho))

# Fit Linear Mixed-Effects Model
lmm_fit <- lmer(
  rho ~ cre_identity_pct + pct_cres_conserved + repeat_masked_pct + (1 | gene_pair_id),
  data = df_integrated
)

# Extract clean model coefficients and variance components
tidy(lmm_fit)
```



# Export

```{r export}
# Merge alignment results back into cres_annotated
cres_annotated <- cres_annotated %>%
  left_join(
    cres_to_align %>%
      dplyr::select(human_enhancer, mouse_enhancer,
                    sequence_human, sequence_mouse,
                    cre_identity_pct, cre_align_score, cre_align_length),
    by = c("human_enhancer", "mouse_enhancer")
  )

# Update split tables
# cres_homologous <- cres_annotated %>% filter(is_homologous)
# cres_human_only <- cres_annotated %>% filter(match_type == "human_only")
# cres_mouse_only <- cres_annotated %>% filter(match_type == "mouse_only")

# Save
saveRDS(cres_annotated,  here("tables", "cres_annotated_full.rds"))
# saveRDS(cres_homologous, here("tables", "cres_homologous_annotated.rds"))
# saveRDS(cres_human_only, here("tables", "cres_human_only_annotated.rds"))
# saveRDS(cres_mouse_only, here("tables", "cres_mouse_only_annotated.rds"))

message("All outputs saved to tables/")
```
