---
title: "6_Statistical_Modelling"
output: html_document
---
# Setup

```{r message=FALSE, warning=FALSE}
# Load core packages for the analysis session
library(here)
source(here("scripts_notebooks", "required.R"))

library(tidyverse)
library(tidymodels)
```


# Prepare features table
```{r}
# Gene expression
dge_btm_process_genes_diff_bygene_clean_filtered =  readRDS(file = here("tables", "dge_btm_process_genes_diff_bygene_clean_filtered.rds")) 


#Protein identity
alignments_human2mouse_all = readRDS(file = here("tables", "alignments_human2mouse_all.rds"))

#Distance
human_mouse_cds_distance = readRDS(file = here::here("tables", "human_mouse_cds_distance.rds"))

#Transcription factors
gene_tf_shared_unique = readRDS(file = here("tables", "gene_tf_shared_unique.rds"))

#BTMs annotation
btm_annotation_genes <- read_csv(here("tables", "btm_annotation_genes.csv")) %>% 
  select(hgnc_symbol = symbol, process, group, subgroup) %>% 
  distinct()

btm_annotation_genes_others = bind_rows(
  btm_annotation_genes,
  #Add non-BTM genes
  dge_btm_process_genes_diff_bygene_clean_filtered %>% 
    select(hgnc_symbol) %>% 
    anti_join(btm_annotation_genes %>% 
                select(hgnc_symbol) %>% 
                distinct(),
              by = "hgnc_symbol") %>% 
    mutate(process = "Not-BTM",
           group = "Not-BTM",
           subgroup = "Not-BTM")
  )


btm_annotation_genes_wide = btm_annotation_genes_others %>% 
  select(hgnc_symbol, group) %>% 
  group_by(hgnc_symbol, group) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  pivot_wider(names_from = group,
              values_from = n) %>% 
  column_to_rownames("hgnc_symbol") %>% 
  as.matrix() %>% 
  replace(is.na(.), 0) %>% 
  as.data.frame() %>% 
  rownames_to_column("hgnc_symbol")


# CREs ===============
cres_type_homology_comparison_wide = 
  readRDS(file = here("tables", "cres_type_homology_comparison_wide.rds"))

#Merge features ==========
#By genes only
gene_annotated_layers = 
  #Annotate with CDS distance (Kimura)
  human_mouse_cds_distance %>% select(hgnc_symbol, dist_k80) %>% 
  #Annotate with AA identity %
  inner_join(alignments_human2mouse_all,
             by = join_by("hgnc_symbol" == "gene")) %>% 
  #Annotate with TFs
  inner_join(gene_tf_shared_unique,
             by = join_by("hgnc_symbol" == "gene")) %>% 
  #Annotate with CRES
  inner_join(cres_type_homology_comparison_wide,
             by = join_by(hgnc_symbol)) %>% 
     #Annotate with BTMs 
  inner_join(btm_annotation_genes_wide,
             by = "hgnc_symbol")

#Save
# gene_annotated_layers %>% 
#   saveRDS(file = here("tables", "human_mouse_statsmodelling_gene_annotated_layers.rds"))

#All parameters
human_mouse_parameters_values = 
  gene_annotated_layers %>% 
  #Annotate with gene expression
  inner_join(
    dge_btm_process_genes_diff_bygene_clean_filtered %>% 
               select(!c(p_adj_mouse, status_original, process, p_adjust_mouse_gsea, mean_log2fc_human, group, status_original, nes_human, nes_mouse)) %>% 
               distinct(),
             by = join_by(hgnc_symbol)) 

human_mouse_parameters_values

#Save
human_mouse_parameters_values %>%
  saveRDS(file = here("tables", "human_mouse_statsmodelling_parameters_values.rds"))

human_mouse_parameters_values %>% 
  glimpse()



# #Create features for regression and classification
# human_mouse_features_regression = human_mouse_parameters_values %>%
#   select(-status) %>%
#   distinct() %>%
#   select(treatment:abs_log2fc_diff,inverse_se_mouse, hgnc_symbol, everything())
# human_mouse_features_regression
# 
# human_mouse_features_classification_legs = human_mouse_parameters_values %>%
#   drop_na(status) %>%
#   filter(status %in% c("Shared", "Not-shared")) %>%
#   mutate(status = fct_relevel(status, "Not-shared")) %>%
#   select(treatment:inverse_se_mouse, inverse_se_mouse, hgnc_symbol, status, everything()) %>%
#   select(!c(abs_log2fc_diff))
# human_mouse_features_classification_legs

```



# Statistical modelling

## Features dictionary
```{r}
#Import
human_mouse_parameters_values =
  readRDS(file = here("tables", "human_mouse_statsmodelling_parameters_values.rds"))

feature_dictionary <- tibble(
  feature_name_original = names(human_mouse_parameters_values)
) %>%
  filter(!feature_name_original %in% c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison", "status")) %>%
  
  # 1. Regex parsing for cCRE feature sets
  extract(
    col     = feature_name_original,
    into    = c("cre_metric", "cre_type", "cre_ctcf", "cre_homology"),
    regex   = "^(n_type|pct_match_type|n_ctcf|pct_match_ctcf|n_homology|pct_homology)_(PLS|pELS|dELS)(?:_(CTCF-bound|not-CTCF))?(?:_([A-Za-z0-9_]+))?$",
    remove  = FALSE
  ) %>%
  mutate(cre_ctcf = str_replace(cre_ctcf, "-bound", "")) %>% 
  
  # 2. Consolidated classification and label formatting
  mutate(
    # --- Category Assignment ---
    feature_type = case_when(
      feature_name_original %in% c("inverse_se_mouse", "abs_log2fc_diff", 
                                   "mean_log2fc_mouse", "abs_rank_log2fc_mouse") ~ "Gene expression",
      feature_name_original %in% c("dist_k80", "identity_human2mouse") ~ "Sequence evolution",
      str_detect(feature_name_original, "^(n_tf|pct_tf)") ~ "Transcription Factor",
      feature_name_original %in% c(
        "Not-BTM", "INNATE RESPONSE", "B CELLS", "T CELLS", "PLATELETS", 
        "ECM AND MIGRATION", "ENERGY METABOLISM", "MONOCYTES", 
        "CELL CYCLE", "IFN", "SIGNAL TRANSDUCTION", "NK CELLS", "NEUTROPHILS"
      ) ~ "BTMs",
      feature_name_original == "n_total_cres_gene" ~ "CREs | Summary",
      
      !is.na(cre_type) & !is.na(cre_ctcf) ~ str_c(cre_type, " ", cre_ctcf),
      !is.na(cre_type)                    ~ cre_type,
      
      str_detect(feature_name_original, "PLS")  ~ "PLS",
      str_detect(feature_name_original, "pELS") ~ "pELS",
      str_detect(feature_name_original, "dELS") ~ "dELS",
      TRUE ~ "Other"
    ),
    
    # Clean whitespace and trailing commas
    feature_type = str_trim(feature_type),
    
    # --- Formatted Clean Feature Labels ---
    feature_name = case_when(
      feature_name_original == "mean_log2fc_mouse"      ~ "Log2FC (Mouse)",
      feature_name_original == "abs_rank_log2fc_mouse"      ~ "|Rank (Mouse)|",
      feature_name_original == "abs_log2fc_diff"        ~ "|ΔLog2FC|",
      feature_name_original == "inverse_se_mouse"        ~ "1/SE",
      feature_name_original == "identity_human2mouse"   ~ "%Protein AA identity",
      feature_name_original == "dist_k80"                ~ "Kimura distance",
      
      feature_name_original == "n_tf_unique"             ~ "nUnique TFs",
      feature_name_original == "pct_tf_unique"           ~ "%Unique TFs",
      feature_name_original == "n_tf_shared"             ~ "nShared TFs",
      feature_name_original == "pct_tf_shared"           ~ "%Shared TFs",
      feature_name_original == "n_tf_total"              ~ "nTotal TFs",
      
      feature_name_original == "n_total_cres_gene"      ~ "nTotal cCREs",
      
      cre_metric == "n_type"                             ~ "nCREs",
      cre_metric == "pct_match_type"                    ~ "%Match Type",
      cre_metric == "n_ctcf"                             ~ "nCTCF",
      cre_metric == "pct_match_ctcf"                    ~ "%Match CTCF",
      
      cre_metric == "n_homology"                         ~ str_c("n", str_replace_all(cre_homology, "_", " ")),
      cre_metric == "pct_homology"                       ~ str_c("%", str_replace_all(cre_homology, "_", " ")),
      
      TRUE ~ feature_name_original
    )
  ) %>%
  
  # 3. Final selection, factor releveling, and ordering
  select(feature_name_original, feature_type, feature_name) %>% 
  mutate(
    feature_type = fct_relevel(
      feature_type, 
      "Gene expression", 
      "Sequence evolution", 
      "Transcription Factor", 
      "CREs | Summary",
      "PLS",              
      "PLS CTCF", 
      "PLS not-CTCF",
      "pELS",             
      "pELS CTCF", 
      "pELS not-CTCF",
      "dELS",             
      "dELS CTCF", 
      "dELS not-CTCF",
      "Other",
      "BTMs"
    )
  ) %>% 
  arrange(feature_type)

feature_dictionary

```


```{r}
# ==============================================================================
# 1. BUILDING BLOCKS (CATEGORY-SPECIFIC FEATURE DEFINITIONS)
# ==============================================================================

# Experimental metadata and target outcome
feat_base_experiment <- c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol")
feat_base_outcome    <- c("inverse_se_mouse", "abs_log2fc_diff", "status", "mean_log2fc_mouse", "abs_rank_log2fc_mouse")

# Biological Modules
feat_sequence_evolution <- c("dist_k80", "identity_human2mouse")

feat_btms <- c(
  "Not-BTM", "INNATE RESPONSE", "B CELLS", "T CELLS", "PLATELETS", 
  "ECM AND MIGRATION", "ENERGY METABOLISM", "MONOCYTES", "CELL CYCLE", 
  "IFN", "SIGNAL TRANSDUCTION", "NK CELLS", "NEUTROPHILS"
)

feat_tfs_selected    <- c("n_tf_total", "pct_tf_shared")


# 2. BIOLOGICAL ABLATION CONTROLS (NEGATIVE CONTROLS)

# [Ablation 1] Experimental metadata and DGE statistical precision only
features_ablation_dge_only <- c(feat_base_experiment, feat_base_outcome)

# [Ablation 2] DGE + Protein sequence conservation
features_ablation_dge_seq <- c(features_ablation_dge_only, feat_sequence_evolution)


# [Ablation 3] DGE + Sequence + Transcriptional Networks (TFs & BTMs) - NO epigenomics/cCREs
features_ablation_no_ccre <- c(features_ablation_dge_seq, feat_tfs_selected)


# ==============================================================================
# 3. REGULATORY RESOLUTION SETS (cCREs / ENHANCERS / PROMOTERS)
# ==============================================================================
features_ccre_baseline_minimal <- c(
  features_ablation_no_ccre,
  "n_total_cres_gene", 
  "n_type_PLS", "n_type_pELS", "n_type_dELS"
)


# [Resolution 4] High: Includes subtype matching rates (% match)
features_ccre_ctcf_match <- c(
  features_ablation_no_ccre,
  "n_total_cres_gene",
  ## PLS
  "n_type_PLS",                               # Count of PLS
  "pct_match_type_PLS",                       # % Matching PLS subtype
  "pct_match_ctcf_PLS_CTCF-bound",            # % Matching subtype
  ## pELS
  "n_type_pELS",                               # Count of pELS
  "pct_match_type_pELS",                       # % Matching pELS subtype
  "pct_match_ctcf_pELS_CTCF-bound",            # % Matching subtype
  ## dELS
  "n_type_dELS",                               # Count of dELS
  "pct_match_type_dELS",                       # % Matching dELS subtype
  "pct_match_ctcf_dELS_CTCF-bound"            # % Matching subtype
)

# [Resolution 5] Ultra-High: Adds complete homology and complex homology metrics
features_ccre_full_homology <- c(
  features_ablation_no_ccre,
  "n_total_cres_gene",
  
  # --- cCREs: Promoter-like Sequences (PLS) ---
  ## General PLS
  "n_type_PLS",                               # Count of PLS
  "pct_match_type_PLS",                       # % Matching PLS subtype
  "pct_match_ctcf_PLS_CTCF-bound",            # % Matching subtype
  "pct_homology_PLS_CTCF-bound_Homologous",   # % Homologous

  # --- cCREs: Proximal Enhancer-like Sequences (pELS) ---
  ## General pELS
  "n_type_pELS",                               # Count of pELS
  "pct_match_type_pELS",                       # % Matching pELS subtype
  "pct_match_ctcf_pELS_CTCF-bound",            # % Matching subtype
  "pct_homology_pELS_CTCF-bound_Homologous",   # % Homologous

  # --- cCREs: Distal Enhancer-like Sequences (dELS) ---
  ## General dELS
  "n_type_dELS",                               # Count of dELS
  "pct_match_type_dELS",                       # % Matching dELS subtype
  "pct_match_ctcf_dELS_CTCF-bound",            # % Matching subtype
  "pct_homology_dELS_CTCF-bound_Homologous"   # % Homologous
)



features_dge_seq_btms <- c(
  features_ccre_ctcf_match,
  feat_btms
)

# ==============================================================================
# 4. MASTER LIST FOR PIPELINE AND ROC EVALUATION
# ==============================================================================

feature_sets_list <- list(
  # --- Biological Ablation Controls ---
  "DGE Baseline"      = features_ablation_dge_only,
  "+ Sequence"        = features_ablation_dge_seq,
  "+ Sequence + TFs"             = features_ablation_no_ccre,
  
  # --- Regulatory Models (cCRE Resolution Gradient) ---
  "+ Sequence + TFs + CRE type"      = features_ccre_baseline_minimal,
  "+ Sequence + TFs + CRE type + CTCF"        = features_ccre_ctcf_match,
  # "+ Sequence + TFs + CRE type + CTCF + Homology"        = features_ccre_full_homology,
  # --- Functional Models ---
  "Full + BTM"            = features_dge_seq_btms
)

features_names_vector = c(
      "DGE Baseline",
      "+ Sequence",
      "+ Sequence + TFs",
      "+ Sequence + TFs + CRE type",
      "+ Sequence + TFs + CRE type + CTCF",
      "Full + BTM" 
    )

```





# Modelling

## Classify Shared vs Not-Shared LEGs

### Iterate over all features

```{r fig.height=12, fig.width=10}
# Create a directory to store the lightweight RDS files if it doesn't exist
dir.create(here::here("Models"), showWarnings = FALSE)

# Execute the iterative classification pipeline
all_classification_results <- imap(feature_sets_list, function(current_selection, set_name) {
  
  cat("\n=======================================================\n")
  cat(" FITTING CLASSIFICATION MODELS:", set_name, "\n")
  cat("=======================================================\n")
  
  clean_name <- str_replace_all(set_name, "[ +]+", "_") %>% 
    str_replace_all("^_|_$", "")
  
  # ==============================================================================
  # 1. PREPARE DATASET
  # ==============================================================================
  df_classification <- human_mouse_parameters_values %>%
    select(all_of(current_selection)) %>%
    drop_na(status) %>%
    filter(status %in% c("Shared", "Not-shared")) %>%
    mutate(status = factor(status, levels = c("Not-shared", "Shared"))) %>%
    select(!any_of("abs_log2fc_diff"))
  
  # ==============================================================================
  # 2. RECIPE DEFINITION
  # ==============================================================================
  cls_recipe <- recipe(status ~ ., data = df_classification) %>%
    update_role(any_of(c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison")), new_role = "id") %>%
    step_dummy(all_nominal_predictors()) %>%
    step_nzv(all_predictors()) %>%
    step_corr(all_numeric_predictors(), threshold = 0.9) %>%
    step_normalize(all_numeric_predictors())
  
  prep_rec <- prep(cls_recipe, training = df_classification)
  baked_predictors <- bake(prep_rec, new_data = NULL) %>% 
    select(-status, -has_role("id"))
  
  num_cols <- ncol(baked_predictors)
  
  # ==============================================================================
  # 3. MODEL FITTING
  # ==============================================================================
  
  # --- Logistic Regression (Native GLM) ---
  logistic_spec <- logistic_reg() %>% 
    set_engine("glm")
  
  fit_logistic <- workflow() %>%
    add_recipe(cls_recipe) %>%
    add_model(logistic_spec) %>%
    fit(data = df_classification)
  
  # --- Random Forest ---
  rf_mtry <- min(8, max(1, num_cols))
  
  rf_spec <- rand_forest(trees = 1000, mtry = rf_mtry, min_n = 5) %>% 
    set_engine("ranger", importance = "impurity") %>% 
    set_mode("classification")
  
  fit_rf <- workflow() %>%
    add_recipe(cls_recipe) %>%
    add_model(rf_spec) %>%
    fit(data = df_classification)
  
  # ==============================================================================
  # 4. EXTRACT COEFFICIENTS & IMPORTANCE
  # ==============================================================================
  
  log_coefs <- tidy(extract_fit_engine(fit_logistic), conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    rename(feature_name_original = term) %>%
    mutate(
      feature_name_original = str_replace_all(feature_name_original, "`", ""),
      feature_set = set_name
    ) %>% 
    inner_join(feature_dictionary, by = join_by("feature_name_original"))
  
  rf_engine <- extract_fit_engine(fit_rf)
  rf_imp <- tibble(
    feature_name_original = names(rf_engine$variable.importance),
    importance            = as.numeric(rf_engine$variable.importance),
    feature_set           = set_name
  ) %>%
    inner_join(feature_dictionary, by = join_by("feature_name_original"))
  
  # ==============================================================================
  # 5. EVALUATION METRICS (ROC & PR-AUC)
  # ==============================================================================
  
  pred_logistic <- stats::predict(fit_logistic, new_data = df_classification, type = "prob")$.pred_Shared
  pred_rf        <- stats::predict(fit_rf, new_data = df_classification, type = "prob")$.pred_Shared
  
  # ROC Curves & AUC
  roc_logistic  <- roc(response = df_classification$status, predictor = pred_logistic, levels = c("Not-shared", "Shared"))
  roc_rf        <- roc(response = df_classification$status, predictor = pred_rf, levels = c("Not-shared", "Shared"))
  
  auc_log <- round(auc(roc_logistic), 3)
  auc_rf  <- round(auc(roc_rf), 3)
  
  # PR-AUC Calculation (target event = "Shared")
  eval_df_log <- tibble(truth = df_classification$status, .pred_Shared = pred_logistic)
  eval_df_rf  <- tibble(truth = df_classification$status, .pred_Shared = pred_rf)
  
  prauc_log <- round(pr_auc(eval_df_log, truth = truth, .pred_Shared, event_level = "second")$.estimate, 3)
  prauc_rf  <- round(pr_auc(eval_df_rf, truth = truth, .pred_Shared, event_level = "second")$.estimate, 3)
  
  metrics_table <- tibble(
    feature_set = set_name,
    model = c("Logistic Regression", "Random Forest"),
    auc = c(auc_log, auc_rf),
    pr_auc = c(prauc_log, prauc_rf)
  )
  
  roc_data <- bind_rows(
    ggroc(roc_logistic)$data %>% mutate(model = "Logistic Regression", auc = auc_log, pr_auc = prauc_log),
    ggroc(roc_rf)$data %>% mutate(model = "Random Forest", auc = auc_rf, pr_auc = prauc_rf)
  ) %>% 
    mutate(
      feature_set = set_name,
      curve_label = paste0(model, " - ", set_name, " (ROC=", auc, ", PR=", pr_auc, ")")
    )
  
  # ==============================================================================
  # 6. COMPILE AND SAVE LIGHTWEIGHT RDS
  # ==============================================================================
  
  iteration_results <- list(
    set_name     = set_name,
    metrics      = metrics_table,
    coefficients = log_coefs,
    importance   = rf_imp,
    roc_data     = roc_data
  )
  
  rds_path <- here::here("Models", paste0("classification_extracted_data_", clean_name, ".rds"))
  saveRDS(iteration_results, file = rds_path)
  
  return(iteration_results)
})


# ==============================================================================
# GENERATE MASTER TABLES
# ==============================================================================

master_class_metrics <- map_dfr(all_classification_results, "metrics") %>% 
  mutate(
    feature_set = fct_relevel(feature_set, features_names_vector)
  )
saveRDS(master_class_metrics, file = here::here("Models", "master_class_metrics.rds"))

master_class_coefs <- map_dfr(all_classification_results, "coefficients") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))
saveRDS(master_class_coefs, file = here::here("Models", "master_class_coefs.rds"))

master_class_var_imp <- map_dfr(all_classification_results, "importance")  %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

saveRDS(master_class_var_imp, file = here::here("Models", "master_class_var_imp.rds"))

master_class_roc_data <- map_dfr(all_classification_results, "roc_data")  %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

saveRDS(master_class_roc_data, file = here::here("Models", "master_class_roc_data.rds"))

# Generate wide summary table of AUC and PR-AUC values across models and feature sets
classification_auc_table <- master_class_metrics %>%
  pivot_wider(names_from = model, values_from = c(auc, pr_auc))  %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector)) %>% 
  arrange(feature_set)

classification_auc_table


```


### Visualization
```{r fig.height=12, fig.width=10}
# ==============================================================================
# 5. VISUALIZATIONS
# ==============================================================================

master_class_metrics = readRDS(file = here::here("Models", "master_class_metrics.rds"))
master_class_coefs = readRDS(file = here::here("Models", "master_class_coefs.rds"))
master_class_var_imp = readRDS(file = here::here("Models", "master_class_var_imp.rds"))
master_class_roc_data = readRDS(file = here::here("Models", "master_class_roc_data.rds"))
# Logistic Regression Coefficients Plot================
log_coefs_plot_df = master_class_coefs %>%
  inner_join(
    master_class_metrics %>% filter(model == "Logistic Regression"),
    by = "feature_set"
  ) %>%
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  mutate(
    feature_set_label = paste0(feature_set, "\n(AUC = ", auc, ")"),
    feature_set       = fct_relevel(feature_set, features_names_vector),
    feature_name = if_else(
      str_detect(feature_type, "PLS|pELS|dELS"),
      paste(feature_type, feature_name),
      feature_name
    )
  ) %>%
  arrange(feature_set) %>%
  mutate(feature_name = fct_rev(fct_inorder(feature_name)))


log_coefs_plot = log_coefs_plot_df %>% 
  ggplot() +
  aes(x = estimate, 
      y = feature_set %>% fct_rev(), 
      xmin = conf.low, 
      xmax = conf.high, 
      fill = feature_set, 
      group = feature_set) +
  facet_grid(feature_name %>% fct_rev() ~., scales="free_y", space="free", switch = "y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.6) +
  geom_errorbarh(height = 0.2, linewidth = 0.6, width = 0,
                 alpha = 0.5,
                 position = position_dodge(width = 1), 
                 show.legend = FALSE) +
  geom_point(size = 2.2, pch = 21, position = position_dodge(width = 1)) +
  scale_fill_manual(values = colors$feature_sets) +
  labs(
    x = "Log-Odds",
    y = "Feature",
    fill = "Feature Set"
  ) +
  theme_vaxgo() +
  theme(legend.position = "top", 
    legend.location = "plot",
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1, vjust = 0.5),
    axis.text.y = element_blank()) +
  guides(fill = guide_legend(override.aes = list(pch = 22),
                             ncol = 1))

# 2. Random Forest Importance Plot================
# Classification RF Importance Data Preparation
rf_imp_plot_class_df <- master_class_var_imp %>%
  inner_join(
    master_class_metrics %>% filter(model == "Random Forest"),
    by = "feature_set"
  ) %>%
  mutate(
    feature_set_label = paste0(feature_set, "\n(AUC = ", auc, ")"),
    feature_set       = fct_relevel(feature_set, features_names_vector),
    feature_name = if_else(
      str_detect(feature_type, "PLS|pELS|dELS"),
      paste(feature_type, feature_name),
      feature_name
    )
  ) %>%
  arrange(feature_set) %>%
  mutate(feature_name = fct_rev(fct_inorder(feature_name)))

# Classification RF Importance Visualization
rf_imp_plot_class <- rf_imp_plot_class_df %>%
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  ggplot() +
  aes(
    x    = importance + 1, 
    y    = feature_set %>% fct_rev(), 
    fill = feature_set
  ) +
  facet_grid(feature_name %>% fct_rev() ~ ., scales = "free_y", space = "free", switch = "y") +
  geom_col(
    width    = 0.6, 
    position = position_dodge(1)
  ) +
  scale_fill_manual(values = colors$feature_sets) +
  scale_x_log10(
    breaks = c(1, 10, 50, 100, 250, 1000), 
    labels = c("0", "10", "50", "100", "250", "1000"), 
    expand = c(0, 0)
  ) +
  labs(
    x    = "Gini",
    y    = "Feature",
    fill = "Feature Set"
  ) +
  theme_vaxgo() +
  theme(
    legend.position   = "top", 
    legend.location   = "plot",
    strip.placement   = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1, vjust = 0.5),
    axis.text.y       = element_blank()
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(pch = 22),
      byrow        = TRUE,
      ncol         = 1
    )
  )

rf_imp_plot_class

# AUC ROC Bar Chart Comparison ========================
p_auc_bars = master_class_metrics  %>%
  pivot_longer(cols = c("pr_auc", "auc"),
               names_to = "method",
               values_to = "auc") %>%
  mutate(method = if_else(method == "auc", "ROC", "PR")) %>% 
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector)) %>% 
  ggplot() +
  aes(x = auc, y = method, fill = model) +
  facet_wrap(.~feature_set,
             scales = "free_y",
             ncol = 1) +
# Draw side-by-side bars
  geom_col(position = position_dodge(width = 0.8), width = 0.7, color = "black", linewidth = 0.3) +
  # Add exact AUC values on top of the bars
  geom_text(
    aes(label = sprintf("%.3f", auc)), 
    position = position_dodge(width = 0.8), 
    vjust = 0.5,
    hjust = 1.5,
    size = 3,
    color = "black",
    fontface = "bold",
  ) +
  # Match the color scheme used in previous algorithm comparisons
  scale_fill_manual(values = c(
    "Random Forest" = "#4DBBD5FF", 
    "Logistic Regression" = "gray70" 
  )) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    x = "AUC ROC",
    y = "",
    fill = "Algorithm"
  ) +
  theme_vaxgo() +
  theme(
    legend.position = "top", 
    legend.title.position = "top",
    # axis.text.y = element_blank(),
    plot.title.position = "plot",
    strip.text.x = element_text(hjust = 0)
  ) +
   guides(fill = guide_legend(byrow = T, nrow = 2))

# Merge plots
class_patch_plot_1 <- (
  (log_coefs_plot + theme(strip.text.y = element_blank(), legend.position = "none")) |
  (rf_imp_plot_class + guides(fill = guide_legend(byrow = TRUE, nrow = 2)) + 
     theme(legend.title.position = "top", legend.location = "plot", legend.justification = "left",
           strip.text.y.left = element_blank()))
) +
  plot_layout(guides = "collect", axes = "collect") +
  plot_annotation(theme = theme(legend.position = "top", legend.location = "plot", legend.justification = "left"))

# Final merge
class_patch_plot <- (class_patch_plot_1 |p_auc_bars) 

#Display
class_patch_plot

#Save
class_patch_plot %>% 
ggsave(filename = here("Figures", "statistical_modelling_classification_coefficients_rocauc_allmodels.png"),
         width = 10,
         height = 12)
```








## Predict Log2FC Divergence Magnitude


### Iterate over multiple models

```{r fig.height=12, fig.width=10}
# Create a directory to store the lightweight RDS files if it doesn't exist
dir.create(here::here("Models"), showWarnings = FALSE)

# Execute the iterative regression pipeline and store extracted data in a list
all_regression_results <- imap(feature_sets_list, function(current_selection, set_name) {
  
  cat("\n=======================================================\n")
  cat(" FITTING REGRESSION MODELS:", set_name, "\n")
  cat("=======================================================\n")
  
  # Clean the set name for file saving 
  clean_name <- str_replace_all(set_name, "[ +]+", "_") %>% 
    str_replace_all("^_|_$", "")
  
  # ==============================================================================
  # 1. PREPARE DATASET 
  # ==============================================================================
  reg_cols <- setdiff(current_selection, "status")
  if (!"abs_log2fc_diff" %in% reg_cols) {
    reg_cols <- c(reg_cols, "abs_log2fc_diff")
  }
  
  df_regression <- human_mouse_parameters_values %>%
    select(all_of(reg_cols)) %>%
    drop_na(abs_log2fc_diff) %>%
    distinct()
  
  # ==============================================================================
  # 2. RECIPE DEFINITION 
  # ==============================================================================
  reg_recipe <- recipe(abs_log2fc_diff ~ ., data = df_regression) %>%
    update_role(any_of(c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison")), new_role = "id") %>%
    step_dummy(all_nominal_predictors()) %>%
    step_zv(all_predictors()) %>%
    step_normalize(all_numeric_predictors())
  
  # Determine number of valid predictors
  prep_rec <- prep(reg_recipe, training = df_regression)
  baked_pred <- bake(prep_rec, new_data = NULL) %>% 
    select(-abs_log2fc_diff, -has_role("id"))
  
  num_cols <- ncol(baked_pred)
  
  # ==============================================================================
  # 3. MODEL FITTING 
  # ==============================================================================
  
  # --- Linear Regression (LM) ---
  lm_spec <- parsnip::linear_reg() %>%
    parsnip::set_engine("lm") %>%
    parsnip::set_mode("regression")
  
  lm_fit <- workflow() %>%
    add_recipe(reg_recipe) %>%
    add_model(lm_spec) %>%
    fit(data = df_regression)
  
  # --- Random Forest (RF) ---
  rf_mtry <- min(8, max(1, num_cols))
  
  rf_spec <- parsnip::rand_forest(trees = 1000, mtry = rf_mtry, min_n = 5) %>%
    parsnip::set_engine("ranger", importance = "impurity") %>%
    parsnip::set_mode("regression")
  
  rf_fit <- workflow() %>%
    add_recipe(reg_recipe) %>%
    add_model(rf_spec) %>%
    fit(data = df_regression)
  
  # ==============================================================================
  # 4. EXTRACT COEFFICIENTS & IMPORTANCE
  # ==============================================================================
  
  lm_coefs <- tidy(extract_fit_engine(lm_fit), conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    rename(feature_name_original = term) %>%
    mutate(
      feature_name_original = str_replace_all(feature_name_original, "`", ""),
      feature_set = set_name # Add identifier for the master table
    ) %>% 
    inner_join(feature_dictionary, by = join_by("feature_name_original"))
  
  rf_engine <- extract_fit_engine(rf_fit)
  rf_imp <- tibble(
    feature_name_original = names(rf_engine$variable.importance),
    importance            = as.numeric(rf_engine$variable.importance),
    feature_set           = set_name # Add identifier for the master table
  ) %>%
    inner_join(feature_dictionary, by = join_by("feature_name_original"))
  
  # ==============================================================================
  # 5. EVALUATION METRICS & PREDICTIONS
  # ==============================================================================
  
  eval_table <- df_regression %>%
    select(abs_log2fc_diff) %>%
    bind_cols(
      stats::predict(lm_fit, new_data = df_regression) %>% rename(pred_linear = .pred),
      stats::predict(rf_fit, new_data = df_regression) %>% rename(pred_rf = .pred)
    ) %>%
    mutate(feature_set = set_name)
  
  metrics_lm <- metric_set(rmse, rsq)(eval_table, truth = abs_log2fc_diff, estimate = pred_linear) %>% 
    mutate(model = "Linear (LM)", feature_set = set_name)
  
  metrics_rf <- metric_set(rmse, rsq)(eval_table, truth = abs_log2fc_diff, estimate = pred_rf) %>% 
    mutate(model = "Random Forest", feature_set = set_name)
  
  current_metrics <- bind_rows(metrics_lm, metrics_rf)
  
  # ==============================================================================
  # 6. COMPILE AND SAVE LIGHTWEIGHT RDS
  # ==============================================================================
  
  # Bundle ONLY the extracted data frames (excluding the heavy model fits)
  iteration_results <- list(
    set_name     = set_name,
    metrics      = current_metrics,
    coefficients = lm_coefs,
    importance   = rf_imp,
    predictions  = eval_table
  )
  
  # Save the lightweight bundle as an RDS file
  rds_path <- here::here("Models", paste0("regression_extracted_data_", clean_name, ".rds"))
  saveRDS(iteration_results, file = rds_path)
  
  cat("  -> Saved extracted data to:", rds_path, "\n")
  
  return(iteration_results)
})

# ==============================================================================
# GENERATE MASTER TABLES
# ==============================================================================

# 1. Master Metrics Table (RMSE, R2 for all models and sets)
master_regression_metrics <- map_dfr(all_regression_results, "metrics") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

master_regression_metrics %>% 
  saveRDS(file = here("Models", "master_regression_metrics.rds"))

# 2. Master Coefficients Table (Linear Model terms across all sets)
master_regression_coefs <- map_dfr(all_regression_results, "coefficients") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

master_regression_coefs %>% 
  saveRDS(file = here("Models", "master_regression_coefs.rds"))

# 3. Master Variable Importance Table (Random Forest Gini across all sets)
master_regression_var_imp <- map_dfr(all_regression_results, "importance") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

master_regression_var_imp %>% 
  saveRDS(file = here("Models", "master_regression_var_imp.rds"))

# 4. Master Predictions Table (Observed vs Predicted data across all sets)
master_regression_predictions <- map_dfr(all_regression_results, "predictions") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

master_regression_predictions %>% 
  saveRDS(file = here("Models", "master_regression_predictions.rds"))

# Display the summary metrics
print(master_regression_metrics %>% arrange(model, feature_set))

master_regression_metrics %>% 
  filter(.metric == "rsq") %>% 
  arrange(model)

```


```{r fig.height=12, fig.width=10}
# ==============================================================================
# 5. VISUALIZATIONS
# ==============================================================================
master_regression_metrics = 
  readRDS(file = here("Models", "master_regression_metrics.rds"))
master_regression_coefs =
  readRDS(file = here("Models", "master_regression_coefs.rds"))
master_regression_var_imp = 
  readRDS(file = here("Models", "master_regression_var_imp.rds"))
master_regression_predictions =
  readRDS(file = here("Models", "master_regression_predictions.rds"))

# 1. Generate wide summary table of R² values across models and feature sets
regression_rsq_table <- master_regression_metrics %>%
  filter(.metric == "rsq") %>%
  mutate(.estimate = round(.estimate, 3)) %>%
  select(feature_set, model, .estimate) %>%
  pivot_wider(
    names_from = model,
    values_from = .estimate
  ) %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector)) %>% 
  arrange(feature_set)

# Display the summary table
print(regression_rsq_table)

# 1. LM Forest Plot Data Preparation
linear_coefficients_df <- master_regression_coefs %>%
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  inner_join(
    master_regression_metrics %>% 
      filter(model == "Linear (LM)", .metric == "rsq") %>% 
      select(model, .estimate, feature_set),
    by = "feature_set"
  ) %>%
  mutate(
    feature_set_label = paste0(feature_set, "\n(R² = ", round(.estimate, 2), ")"),
    feature_set       = fct_relevel(feature_set, features_names_vector),
    feature_name = if_else(
      str_detect(feature_type, "PLS|pELS|dELS"),
      paste(feature_type, feature_name),
      feature_name
    )
  ) %>%
  arrange(feature_set) %>%
  mutate(feature_name = fct_rev(fct_inorder(feature_name)))

# 2. LM Forest Plot Visualization
linear_coefficients_forest_plot <- linear_coefficients_df %>%
  ggplot() +
  aes(
    x     = estimate, 
    y     = feature_set %>% fct_rev(), 
    xmin  = conf.low, 
    xmax  = conf.high, 
    fill  = feature_set, 
    group = feature_set
  ) +
  facet_grid(feature_name %>% fct_rev() ~ ., scales = "free_y", space = "free", switch = "y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.6) +
  geom_errorbarh(
    height      = 0.2, 
    linewidth   = 0.6, 
    width       = 0,
    alpha       = 0.5,
    position    = position_dodge(width = 1), 
    show.legend = FALSE
  ) +
  geom_point(size = 2.2, pch = 21, position = position_dodge(width = 1)) +
  scale_fill_manual(values = colors$feature_sets) +
  labs(
    x    = "Std. Effect Size",
    y    = "Feature",
    fill = "Feature Set"
  ) +
  theme_vaxgo() +
  theme(
    legend.title.position = "top", 
    legend.position   = "top", 
    legend.location   = "plot",
    strip.placement   = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1, vjust = 0.5),
    axis.text.y       = element_blank()
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(pch = 22),
      ncol         = 1
    )
  )

linear_coefficients_forest_plot

# 3. Random Forest Gini Importance Plot
# 1. RF Importance Data Preparation
rf_importance_df <- master_regression_var_imp %>%
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  inner_join(
    master_regression_metrics %>% 
      filter(model == "Random Forest", .metric == "rsq") %>% 
      select(model, .estimate, feature_set),
    by = "feature_set"
  ) %>%
  mutate(
    feature_set_label = paste0(feature_set, "\n(R² = ", round(.estimate, 2), ")"),
    feature_set       = fct_relevel(feature_set, features_names_vector),
    feature_name = if_else(
      str_detect(feature_type, "PLS|pELS|dELS"),
      paste(feature_type, feature_name),
      feature_name
    )
  ) %>%
  arrange(feature_set) %>%
  mutate(feature_name = fct_rev(fct_inorder(feature_name)))

# 2. RF Importance Visualization
p_rf_importance <- rf_importance_df %>% 
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  ggplot() +
  aes(
    x    = importance + 1, 
    y    = feature_set %>% fct_rev(), 
    fill = feature_set
  ) +
  facet_grid(feature_name %>% fct_rev() ~ ., scales = "free_y", space = "free", switch = "y") +
  geom_col(
    width    = 0.6,
    position = position_dodge(1)
  ) +
  scale_fill_manual(values = colors$feature_sets) +
  scale_x_log10(
    breaks = c(1, 10, 50, 100, 250, 1000),
    labels = c("0", "10", "50", "100", "250", "1000"),
    expand = c(0, 0)
  ) +
  labs(
    x    = "Importance",
    y    = "Feature",
    fill = "Feature Set"
  ) +
  theme_vaxgo() +
  theme(
    legend.title.position = "top", 
    legend.position   = "top", 
    legend.location   = "plot",
    legend.justification = "left",
    strip.placement   = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1, vjust = 0.5),
    axis.text.y       = element_blank()
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(pch = 22),
      nrow        = 2
    )
  )

p_rf_importance

# 4. Merge plots (LM vs RF)
coeffiecients_patch_plot = (linear_coefficients_forest_plot + 
                              theme(strip.text.y = element_blank(),
                                    legend.position = "none")) +
  (p_rf_importance + theme(strip.text.y.left = element_blank()))+
  plot_layout(widths = c(0.5, 0.5), 
              guides = "collect", axes = "collect") +
  plot_annotation(theme = theme(legend.position = "top", legend.location = "plot"))

coeffiecients_patch_plot

#### Plot Observed vs Predicted values across models ============

pobs_vs_pred_df = master_regression_predictions %>% 
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  inner_join(regression_rsq_table,
             by = "feature_set") %>% 
  mutate(feature_set_label = paste0(feature_set, "\nLinear model: R² = ", `Linear (LM)`, ",\nRandom forest: R² = ", `Random Forest`)) %>% 
  select(feature_set, feature_set_label, abs_log2fc_diff, pred_linear, pred_rf) %>% 
  pivot_longer(cols = c(pred_linear, pred_rf),
               names_to = "algorithm",
               values_to = "predicted") %>% 
  mutate(algorithm = case_when(algorithm == "pred_linear" ~ "Linear model",
                               algorithm == "pred_rf" ~ "Random forest",
                               T ~ algorithm),
         feature_set = fct_relevel(feature_set,features_names_vector)
    ) %>% 
  arrange(feature_set) %>% 
  mutate(feature_set_label = fct_inorder(feature_set_label))

p_obs_vs_pred = pobs_vs_pred_df %>%
  ggplot(aes(x = abs_log2fc_diff, 
             y = predicted,
             color = algorithm)) +
  facet_wrap(~ feature_set_label, ncol = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.6) +
  geom_point(alpha = 0.1, size = 1, pch = 21, stroke = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1, show.legend = F) +
  scale_color_manual(values = c("Random forest" = "#4DBBD5FF",
                                "Linear model" = "gray70")) +
  guides(
    color = guide_legend(override.aes = list(alpha = 1,
                                             size = 4,
                                             pch = 21,
                                             fill = c("gray70", "#4DBBD5FF")),
                         nrow = 2))  +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  
  labs(
    x = "Observed |ΔLog2FC|",
    y = "Predicted |ΔLog2FC|",
    fill = "",
    color = ""
  ) +
  theme_vaxgo() +
  theme(
    legend.position = "top",
    plot.title.position = "plot", 
    legend.justification = "left",
    legend.title = element_blank(),
  )

#Merge plots
coeffiecients_pobsvspred_patch_plot = (coeffiecients_patch_plot | p_obs_vs_pred ) +
  plot_layout(widths = c(0.6, 0.4)) 

coeffiecients_pobsvspred_patch_plot

coeffiecients_pobsvspred_patch_plot %>% 
  ggsave(filename = here("Figures", "statistical_modelling_regression_coefficients_predicted_vs_observed_patch_plot_allmodes.png"),
         width = 10,
         height = 12)
```

## Combine plots for figures

```{r fig.height=12, fig.width=6}
#Combine performance plots
performance_plots = p_auc_bars + p_obs_vs_pred

performance_plots

#Save plots
performance_plots %>% 
      ggsave(filename = here("Figures", "statistical_modelling_classification_regression_aucroc_obs_pred.png"),
         width = 6,
         height = 12)
```


```{r fig.height=12, fig.width=12}

#Combine features plots
features_plots = wrap_plots(
  class_patch_plot_1 + theme(legend.position = "bottom", legend.direction = "vertical"),
  coeffiecients_patch_plot + theme(legend.position = "bottom"),
  ncol = 2,
  widths = c(1, 1)
) + 
  plot_layout(
    guides = "collect",
    axes = "collect",
    axis_titles = "collect"
  ) &
  plot_annotation(theme = theme(legend.position = "bottom"),
                  title = "Features")

features_plots %>% 
  ggsave(filename = here("Figures", "statistical_modelling_classification_regression_features.png"),
         width = 12,
         height = 12)

features_plots

```










# Create Scores

```{r}
#Import
#Multilayer table
gene_annotated_layers = 
  readRDS(file = here("tables", "human_mouse_statsmodelling_gene_annotated_layers.rds"))

#Gene expression + BTMs results
dge_btm_process_genes_diff_bygene_clean_filtered =  readRDS(file = here("tables", "dge_btm_process_genes_diff_bygene_clean_filtered.rds"))

#Classification
logit_coefs = readRDS(file = here::here("Models", "master_class_coefs.rds"))#Logistic coefficients
rf_importance_class <- readRDS(file = here::here("Models", "master_class_var_imp.rds"))#Random forest importance

#Regression 
lm_coefs = readRDS(file = here("Models", "master_regression_coefs.rds")) #Logistic coefficients
rf_importance_reg = readRDS(file = here("Models", "master_regression_var_imp.rds")) #Random forest importance

#Annotate with status by organism ==========
legs_notlegs_layers = dge_btm_process_genes_diff_bygene_clean_filtered %>% 
  filter(treatment %in% c("Infection", "Injury")) %>% 
  #Annotate with layers
  inner_join(gene_annotated_layers,
             by = "hgnc_symbol")  %>%
  filter(status != "All genes")

#Select model to use corresponding weights =========
select_model = "+ Sequence + TFs + CRE type + CTCF"

# Extract classification weights (Shared vs. Not-Shared)
weight_vec_class <- rf_importance_class %>% 
  filter(feature_set == select_model) %>% 
  select(feature_name, importance, feature_set) %>% 
  inner_join(logit_coefs, by = join_by(feature_set, feature_name)) %>% 
  mutate(
    raw_weight = sign(estimate) * importance,
    norm_weight = raw_weight / max(abs(raw_weight), na.rm = TRUE)
  ) %>% 
  select(feature_name, norm_weight) %>% 
  deframe()

# Extract regression weights
weight_vec_reg <- rf_importance_reg %>% 
  filter(feature_set == select_model) %>% 
  select(feature_name, importance, feature_set) %>% 
  inner_join(lm_coefs, by = join_by(feature_set, feature_name)) %>% 
  mutate(
    # Invert sign to penalize features increasing delta Log2FC
    raw_weight = sign(estimate) * importance,
    norm_weight = raw_weight / max(abs(raw_weight), na.rm = TRUE)
  ) %>% 
  select(feature_name, norm_weight) %>% 
  deframe()

# Lookup functions
get_w_class <- function(name) {
  w <- unname(weight_vec_class[name])
  if (is.null(w) || is.na(w)) return(0) else return(w)
}

get_w_reg <- function(name) {
  w <- unname(weight_vec_reg[name])
  if (is.null(w) || is.na(w)) return(0) else return(w)
}

# Robust Min-Max scaling function handling zero-variance groups
min_max <- function(x) {
  if (all(is.na(x))) return(rep(0, length(x)))
  rng <- max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
  if (is.na(rng) || rng == 0) return(rep(0, length(x)))
  (x - min(x, na.rm = TRUE)) / rng
}

# =========================================================================
# 1. Compute Dual Composite Scores (Pathogen-Level Min-Max)
# =========================================================================
legs_notlegs_layers_score <- legs_notlegs_layers %>% 
  group_by(treatment, pathogen, timepoint_comparison) %>% 
  mutate(
    # Min-Max Normalization per pathogen (Preserves global variance contrast)
    f_fc            = min_max(abs(mean_log2fc_mouse)),
    f_rank          = min_max(abs(abs_rank_log2fc_mouse)),
    f_inv_se        = min_max(log1p(inverse_se_mouse)),
    f_kimura        = min_max(dist_k80),
    f_aa_id         = min_max(identity_human2mouse),
    f_tf_total      = min_max(n_tf_total),
    f_tf_shared     = min_max(pct_tf_shared),
    f_dels_ncres    = min_max(n_type_dELS),
    f_dels_match    = min_max(`pct_match_type_dELS`),
    f_dels_ctcf     = min_max(`pct_match_ctcf_dELS_CTCF-bound`),
    f_pels_ncres    = min_max(n_type_pELS),
    f_pels_match    = min_max(`pct_match_type_pELS`),
    f_pels_ctcf     = min_max(`pct_match_ctcf_pELS_CTCF-bound`),
    f_pls_ncres     = min_max(n_type_PLS),
    f_pls_match     = min_max(`pct_match_type_PLS`),
    f_pls_ctcf      = min_max(`pct_match_ctcf_PLS_CTCF-bound`),
    
    # Classification Raw Score
    score_raw_class = 
      (f_fc           * get_w_class("Log2FC (Mouse)")) + 
      (f_rank           * get_w_class("|Rank (Mouse)|")) + 
      (f_inv_se       * get_w_class("1/SE")) +
      (f_kimura       * get_w_class("Kimura distance")) + 
      (f_aa_id        * get_w_class("%Protein AA identity")) +
      (f_tf_total     * get_w_class("nTotal TFs")) + 
      (f_tf_shared    * get_w_class("%Shared TFs")) +
      (f_dels_ncres   * get_w_class("dELS nCREs")) + 
      (f_dels_match   * get_w_class("dELS %Match Type")) +
      (f_dels_ctcf    * get_w_class("dELS CTCF %Match CTCF")) + 
      (f_pels_ncres   * get_w_class("pELS nCREs")) +
      (f_pels_match   * get_w_class("pELS %Match Type")) + 
      (f_pels_ctcf    * get_w_class("pELS CTCF %Match CTCF")) +
      (f_pls_ncres    * get_w_class("PLS nCREs")) + 
      (f_pls_match    * get_w_class("PLS %Match Type")) +
      (f_pls_ctcf     * get_w_class("PLS CTCF %Match CTCF")),
      
    # Regression Raw Score
    score_raw_reg = 
      (f_fc           * get_w_reg("Log2FC (Mouse)")) + 
      (f_rank           * get_w_reg("|Rank (Mouse)|")) + 
      (f_inv_se       * get_w_reg("1/SE")) +
      (f_kimura       * get_w_reg("Kimura distance")) + 
      (f_aa_id        * get_w_reg("%Protein AA identity")) +
      (f_tf_total     * get_w_reg("nTotal TFs")) + 
      (f_tf_shared    * get_w_reg("%Shared TFs")) +
      (f_dels_ncres   * get_w_reg("dELS nCREs")) + 
      (f_dels_match   * get_w_reg("dELS %Match Type")) +
      (f_dels_ctcf    * get_w_reg("dELS CTCF %Match CTCF")) + 
      (f_pels_ncres   * get_w_reg("pELS nCREs")) +
      (f_pels_match   * get_w_reg("pELS %Match Type")) + 
      (f_pels_ctcf    * get_w_reg("pELS CTCF %Match CTCF")) +
      (f_pls_ncres    * get_w_reg("PLS nCREs")) + 
      (f_pls_match    * get_w_reg("PLS %Match Type")) +
      (f_pls_ctcf     * get_w_reg("PLS CTCF %Match CTCF")),
    
    # Rescale Composite Scores (0-100) and Pathogen Rank
    score_shared = min_max(score_raw_class) * 100,
    score_reg = (1 - min_max(score_raw_reg)) * 100
  ) %>% 
  select(-starts_with("f_"), -starts_with("score_raw_")) %>% 
  ungroup() %>% 
  
  # Process-level Rank calculated separately
  group_by(treatment, pathogen, process) %>% 
  mutate(rank_log2fc_process = percent_rank(abs(mean_log2fc_mouse) * -log10(pmax(p_adj_mouse, 1e-300))) * 100) %>% 
  ungroup()

legs_notlegs_layers_score_plot <- legs_notlegs_layers_score %>% 
  bind_rows(
    legs_notlegs_layers_score %>% 
      mutate(treatment = "Pooled", pathogen = "", timepoint_comparison = "")
  )

# =========================================================================
# 2. Dual Evaluation Pipeline (Statistical Summary Table)
# =========================================================================
calc_dual_metrics <- function(df) {
  df %>% 
    nest() %>% 
    mutate(
      # --- CLASSIFICATION METRICS (score_shared) ---
      auc_roc = map_dbl(data, function(d) {
        if (!"score_shared" %in% names(d) || !"status_original" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(score_shared), !is.na(status_original))
        if (length(unique(d_clean$status_original == "Shared")) < 2) return(NA_real_)
        tryCatch({
          as.numeric(auc(roc(response = d_clean$status_original == "Shared", 
                             predictor = d_clean$score_shared, quiet = TRUE)))
        }, error = function(e) NA_real_)
      }),
      
      p_val_roc = map_dbl(data, function(d) {
        if (!"score_shared" %in% names(d) || !"status_original" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(score_shared), !is.na(status_original))
        if (length(unique(d_clean$status_original == "Shared")) < 2) return(NA_real_)
        tryCatch({
          wilcox.test(score_shared ~ (status_original == "Shared"), data = d_clean)$p.value
        }, error = function(e) NA_real_)
      }),
      
      auc_pr = map_dbl(data, function(d) {
        if (!"score_shared" %in% names(d) || !"status_original" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(score_shared), !is.na(status_original))
        is_sh <- d_clean$status_original == "Shared"
        if (sum(is_sh) == 0 || sum(!is_sh) == 0) return(NA_real_)
        tryCatch({
          pr.curve(
            scores.class0 = d_clean$score_shared[is_sh],
            scores.class1 = d_clean$score_shared[!is_sh]
          )$auc.integral
        }, error = function(e) NA_real_)
      }),
      
      # --- SPEARMAN: rank_log2fc_process vs score_shared ---
      spearman_rank_rho = map_dbl(data, function(d) {
        if (!"rank_log2fc_process" %in% names(d) || !"score_shared" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(rank_log2fc_process), !is.na(score_shared))
        if (nrow(d_clean) < 3 || sd(d_clean$rank_log2fc_process) == 0 || sd(d_clean$score_shared) == 0) return(NA_real_)
        tryCatch({
          cor(d_clean$rank_log2fc_process, d_clean$score_shared, method = "spearman")
        }, error = function(e) NA_real_)
      }),
      
      p_val_spearman_rank = map_dbl(data, function(d) {
        if (!"rank_log2fc_process" %in% names(d) || !"score_shared" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(rank_log2fc_process), !is.na(score_shared))
        if (nrow(d_clean) < 3 || sd(d_clean$rank_log2fc_process) == 0 || sd(d_clean$score_shared) == 0) return(NA_real_)
        tryCatch({
          cor.test(d_clean$rank_log2fc_process, d_clean$score_shared, method = "spearman", exact = FALSE)$p.value
        }, error = function(e) NA_real_)
      }),

      # --- PEARSON: inverse_se_mouse vs score_reg ---
      pearson_se_rho = map_dbl(data, function(d) {
        if (!"inverse_se_mouse" %in% names(d) || !"score_reg" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(inverse_se_mouse), !is.na(score_reg))
        if (nrow(d_clean) < 3 || sd(d_clean$inverse_se_mouse) == 0 || sd(d_clean$score_reg) == 0) return(NA_real_)
        tryCatch({
          cor(d_clean$inverse_se_mouse, d_clean$score_reg, method = "pearson")
        }, error = function(e) NA_real_)
      }),
      
      p_val_pearson_se = map_dbl(data, function(d) {
        if (!"inverse_se_mouse" %in% names(d) || !"score_reg" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(inverse_se_mouse), !is.na(score_reg))
        if (nrow(d_clean) < 3 || sd(d_clean$inverse_se_mouse) == 0 || sd(d_clean$score_reg) == 0) return(NA_real_)
        tryCatch({
          cor.test(d_clean$inverse_se_mouse, d_clean$score_reg, method = "pearson")$p.value
        }, error = function(e) NA_real_)
      }),
      
      # --- SPEARMAN: score_reg vs. |abs_log2fc_diff| ---
      spearman_rho = map_dbl(data, function(d) {
        if (!"score_reg" %in% names(d) || !"abs_log2fc_diff" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(score_reg), !is.na(abs_log2fc_diff))
        if (nrow(d_clean) < 3 || sd(d_clean$score_reg) == 0 || sd(d_clean$abs_log2fc_diff) == 0) return(NA_real_)
        tryCatch({
          cor(d_clean$score_reg, abs(d_clean$abs_log2fc_diff), method = "spearman")
        }, error = function(e) NA_real_)
      }),

      p_val_spearman = map_dbl(data, function(d) {
        if (!"score_reg" %in% names(d) || !"abs_log2fc_diff" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(score_reg), !is.na(abs_log2fc_diff))
        if (nrow(d_clean) < 3 || sd(d_clean$score_reg) == 0 || sd(d_clean$abs_log2fc_diff) == 0) return(NA_real_)
        tryCatch({
          cor.test(d_clean$score_reg, abs(d_clean$abs_log2fc_diff), method = "spearman", exact = FALSE)$p.value
        }, error = function(e) NA_real_)
      }),
      
      # --- REGRESSION METRICS (X: score_reg, Y: |abs_log2fc_diff|) ---
      pearson_r = map_dbl(data, function(d) {
        if (!"score_reg" %in% names(d) || !"abs_log2fc_diff" %in% names(d)) return(NA_real_)
        tryCatch({
          cor(d$score_reg, abs(d$abs_log2fc_diff), method = "pearson", use = "complete.obs")
        }, error = function(e) NA_real_)
      }),
      
      r_squared = map_dbl(data, function(d) {
        if (!"score_reg" %in% names(d) || !"abs_log2fc_diff" %in% names(d)) return(NA_real_)
        tryCatch({
          fit <- lm(abs(abs_log2fc_diff) ~ score_reg, data = d) # Y ~ X
          summary(fit)$r.squared
        }, error = function(e) NA_real_)
      }),
      
      p_val_lm = map_dbl(data, function(d) {
        if (!"score_reg" %in% names(d) || !"abs_log2fc_diff" %in% names(d)) return(NA_real_)
        tryCatch({
          fit <- lm(abs(abs_log2fc_diff) ~ score_reg, data = d) # Y ~ X
          summary(fit)$coefficients[2, 4]
        }, error = function(e) NA_real_)
      })
    ) %>% 
    select(-data)
}

# A. Evaluation per Process / Condition (Pathogen-specific)
eval_condition_process <- calc_dual_metrics(
  legs_notlegs_layers_score_plot %>% 
    filter(!is.na(score_shared), !is.na(score_reg), !is.na(status_original)) %>% 
    group_by(treatment, pathogen, process)
)

# B. Evaluation per Process Across Pooled Conditions (Pooled Process Level)
eval_process_pooled <- calc_dual_metrics(
  legs_notlegs_layers_score_plot %>%
    filter(!is.na(score_shared), !is.na(score_reg), !is.na(status_original)) %>%
    group_by(treatment, pathogen)
) %>%
  mutate( process = "All BTMs", group = "All BTMs")

# C. Overall Global Evaluation (Pooled Overall)
eval_pooled_class <- calc_dual_metrics(
  legs_notlegs_layers_score_plot %>%
    filter(!is.na(score_shared), !is.na(status_original)) %>%
    select(-c(process, group)) %>%
    group_by(treatment, pathogen)
) %>%
  select(
    treatment,
    pathogen,
    spearman_rho,
    spearman_rank_rho,
    p_val_spearman_rank,
    p_val_spearman,
    pearson_r,
    r_squared,
    p_val_lm
  )

eval_pooled_reg <- legs_notlegs_layers_score_plot %>%
  filter(!is.na(score_reg)) %>% 
  select(
    treatment, 
    pathogen, 
    hgnc_symbol, 
    score_reg, 
    abs_log2fc_diff,       
    inverse_se_mouse
  ) %>% 
  distinct() %>% 
  group_by(treatment, pathogen) %>% 
  calc_dual_metrics() %>% 
  select(
    treatment, 
    pathogen, 
    spearman_rho, 
    p_val_spearman, 
    pearson_r, 
    r_squared, 
    p_val_lm
  )

eval_pooled <- inner_join(eval_pooled_class, 
                          eval_pooled_reg, by = c("treatment", "pathogen"), suffix = c("_class", "_reg")) %>% 
  mutate(process = "All genes", group = "All genes")

# Consolidation & FDR Correction
eval_metrics_all_summary <- bind_rows(
  eval_pooled %>% mutate(analysis_set = "Pooled"),
  eval_process_pooled %>% mutate(analysis_set = "Process_Pooled"),
  eval_condition_process %>% mutate(analysis_set = "Process")
) %>% 
  group_by(analysis_set) %>% 
  mutate(
    p_adj_roc           = p.adjust(p_val_roc, method = "BH"),
    p_adj_spearman      = p.adjust(p_val_spearman, method = "BH"),
    p_adj_spearman_rank = p.adjust(p_val_spearman_rank, method = "BH"),
    p_adj_pearson_se    = p.adjust(p_val_pearson_se, method = "BH"),
    p_adj_lm            = p.adjust(p_val_lm, method = "BH")
  ) %>% 
  ungroup() %>% 
  select(
    analysis_set, treatment, pathogen, process, group,
    auc_roc, auc_pr, p_val_roc, p_adj_roc,
    spearman_rho, p_val_spearman, p_adj_spearman,
    spearman_rank_rho, p_val_spearman_rank, p_adj_spearman_rank,
    pearson_se_rho, p_val_pearson_se, p_adj_pearson_se,
    pearson_r, r_squared, p_val_lm, p_adj_lm, everything()
  )

eval_metrics_all_summary
```



## Visualize by process
### Classification


#### LEGs shared prediction
```{r fig.height=4, fig.width=4}
legs_notlegs_layers_visualized_allprocesses = legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, abs_rank_log2fc_mouse, score_shared, p_adj_mouse, status, status_original) %>% 
  distinct() %>% 
  inner_join(
    eval_metrics_all_summary %>% 
      filter(analysis_set %in% c("Process_Pooled")),
    by = join_by(treatment, pathogen)
  ) %>% 
  filter(treatment == "Pooled") %>% 
  ggplot(aes(x = score_shared, y = abs_rank_log2fc_mouse, fill = status_original)) +
  
  facet_nested_wrap(
  ~ process + treatment + pathogen + 
    paste0("ROC-AUC=", round(auc_roc, 2), 
           " \nPR-AUC=", round(auc_pr, 2), 
           "\nrho=", round(spearman_rank_rho, 2), 
           ", padj=", format(p_val_spearman_rank, scientific = TRUE, digits = 2)), 
  nrow = 1, 
  strip = strip_nested(size = "variable")
) +
  
  # Layered scatter plot highlighting Shared vs. Human-only candidates
  geom_point(
    data = . %>% filter(status_original != "Shared"),
    shape = 21, alpha = 0.1, stroke = 0, size = 2
  ) +
  geom_point(
    data = . %>% filter(status_original == "Shared"),
    shape = 21, alpha = 1, stroke = 0.2, size = 2,
  ) +

  scale_fill_manual(values = colors$comparison) +
  theme_vaxgo() +
  labs(
    x = "Score",
    y = "Rank, Mouse",
    fill = "LEG Status",
  ) +
  
  # Right Marginals (X-axis position = 110)
  geom_violin(
    aes(y = abs_rank_log2fc_mouse, x = 110, fill = status_original),
    width = 15, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(y = abs_rank_log2fc_mouse, x = 110),
    width = 15, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  # Top Marginals 
  geom_violin(
    aes(x = score_shared, y = 110, fill = status_original),
    width = 15, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(x = score_shared, y = 110),
    width = 15, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  theme(legend.position = "top", legend.direction = "horizontal")

# Display Plot
legs_notlegs_layers_visualized_allprocesses

```


#### Score vs AbsDiff
```{r fig.height=4, fig.width=4}
######## Score vs AbsDiff ==============
reg_layers_mouse_human_absdiff <-  
  legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, abs_rank_log2fc_mouse, score_reg, p_adj_mouse) %>% 
  distinct() %>% 
    filter(p_adj_mouse <= 0.05) %>% 
 inner_join(
    eval_metrics_all_summary %>%
      filter(analysis_set == "Pooled"),
    by = join_by(treatment, pathogen)
  ) %>%
  distinct() %>% 
  filter(treatment == "Pooled") %>% 
  ggplot(aes(x = score_reg, y = abs_log2fc_diff, 
             fill = score_reg,
             )) +
  
  facet_nested_wrap(
    ~ treatment + pathogen + 
      paste0("rho=", round(spearman_rho_reg, 2), " (p=", sprintf("%.2e", p_val_spearman_reg), ")"),
    nrow = 1, 
    strip = strip_nested(size = "variable")
  ) +
  
  geom_point(
    aes(size = score_reg),
    shape = 21, 
    size = 2,
    alpha = 1,
    # color = NA,
    stroke = 0.5
  ) +
   # Right Marginals 
  geom_violin(
    aes(y = abs_log2fc_diff, x = 110,),
    fill = "#4361ee",
    width = 5, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(y = abs_log2fc_diff, x = 110),
    width = 5, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  # Top Marginals 
  geom_violin(
    aes(x = score_reg, y = 6),
    fill = "#4361ee",
    width = 0.5, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(x = score_reg, y = 6),
    fill = "#4361ee",
    width = 0.5, alpha = 0, color = "black", outlier.shape = NA
  ) +

  scale_fill_continuous(palette = c("white", "#4cc9f0", "#4361ee")) +
  theme_vaxgo() +
  labs(
    x = "Score",
    y = "|ΔLog2fc|",
    fill = "Score",
    size = "Score",
  ) +
  theme(legend.position = "top", legend.direction = "horizontal")

# Display
reg_layers_mouse_human_absdiff 
   
```


#### Mouse Log2FC vs 1/SE
```{r fig.height=4, fig.width=4}
######## Mouse Log2FC vs 1/SE
reg_layers_mouse_human_se <- legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, abs_rank_log2fc_mouse, score_reg, p_adj_mouse) %>% 
  filter(p_adj_mouse <= 0.05,
         treatment == "Pooled") %>%
  distinct() %>% 
 inner_join(
    eval_metrics_all_summary %>%
      filter(analysis_set == "Pooled"),
    by = join_by(treatment, pathogen)
  ) %>%
  distinct() %>% 
  ggplot(aes(x = inverse_se_mouse, 
             y = abs_log2fc_diff, 
             fill = score_reg,
             )) +
  
  # Facet display showing regression metrics (R² and Spearman ρ) with formatted p-values
  facet_nested_wrap(
    ~ treatment + pathogen,
      # paste0("ρ=", round(spearman_rho, 2), " (p=", sprintf("%.2e", p_adj_spearman), ")"),
    nrow = 1, scales = "free_x",
    strip = strip_nested(size = "variable")
  ) +
  
  geom_point(
    # aes(size = score_reg * abs(abs_log2fc_diff)),
    aes(size = score_reg),
    shape = 21, 
    size = 2,
    alpha = 1,
    # color = NA,
    stroke = 0.5
  ) +
  
   # Marginals 
  geom_violin(
    aes(x = inverse_se_mouse, y = 6, group = pathogen),
    width = 0.5, alpha = 0.5, color = NA,
    fill = "#4361ee"
  ) +
  geom_boxplot(
    aes(x = inverse_se_mouse, y = 6, group = pathogen),
    width = 0.5, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
    geom_violin(
    aes(y = abs_log2fc_diff, x = 15,),
    fill = "#4361ee",
    width = 1, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(y = abs_log2fc_diff, x = 15),
    width = 1, alpha = 0, color = "black", outlier.shape = NA
  ) +

  scale_fill_continuous(palette = c("white", "#4cc9f0", "#4361ee")) +
  theme_vaxgo() +
  labs(
    x = "1/SE mouse",
    y = "|ΔLog2fc|"
  ) +
  theme(legend.position = "top", legend.direction = "horizontal")

# Display
reg_layers_mouse_human_se 
```















#### Score vs Rank mouse
```{r fig.height=4, fig.width=4}
######## Score vs Rank mouse
reg_layers_mouse_score_rank <-  
  legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, abs_rank_log2fc_mouse, score_reg, p_adj_mouse) %>% 
  distinct() %>% 
 inner_join(
    eval_metrics_all_summary %>%
      filter(analysis_set == "Pooled"),
    by = join_by(treatment, pathogen)
  ) %>%
  distinct() %>% 
  filter(treatment == "Pooled") %>% 
  ggplot(aes(x = score_reg, y = abs_rank_log2fc_mouse
             )) +
  facet_nested_wrap(
    ~ treatment + pathogen,
    nrow = 1, 
    strip = strip_nested(size = "variable")
  ) +
  
  geom_point(
    aes(size = score_reg,
        fill = score_reg),
    shape = 21, 
    size = 2,
    alpha = 1,
    stroke = 0.5
  ) +
  
     # Right Marginals 
  geom_violin(
    aes(y = abs_rank_log2fc_mouse, x = 110,),
    fill = "#4361ee",
    width = 5, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(y = abs_rank_log2fc_mouse, x = 110),
    width = 5, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  # Top Marginals 
  geom_violin(
    aes(x = score_reg, y = -10),
    fill = "#4361ee",
    width = 5, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(x = score_reg, y = -10),
    fill = "#4361ee",
    width = 5, alpha = 0, color = "black", outlier.shape = NA
  ) +

  scale_fill_continuous(palette = c("white", "#4cc9f0", "#4361ee")) +
  scale_y_reverse() +
  theme_vaxgo() +
  labs(
    x = "Score",
    y = "Rank",
    fill = "Score",
    size = "Score",
  ) +
  theme(legend.position = "top", legend.direction = "horizontal") 

# Display
reg_layers_mouse_score_rank 


```


#### Score vs Rank mouse vs Abs Log2FC diff
```{r fig.height=4, fig.width=4}
######## Score vs Rank mouse vs Abs Log2FC diff
reg_layers_mouse_score_rank_diff_log2fc <-  
  legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, abs_rank_log2fc_mouse, score_reg, p_adj_mouse) %>% 
  distinct() %>% 
  filter(treatment == "Pooled") %>% 
 inner_join(
    eval_metrics_all_summary %>%
      filter(analysis_set == "Pooled"),
    by = join_by(treatment, pathogen)
  ) %>%
  distinct() %>% 
  ggplot(aes(x = abs_rank_log2fc_mouse, y = abs_log2fc_diff
             )) +
  facet_nested_wrap(
    ~ treatment + pathogen,
    nrow = 1, 
    strip = strip_nested(size = "variable")
  ) +
  
  geom_point(
    aes(fill = score_reg),
    data = . %>% filter(score_reg <90),
    shape = 21, 
    size = 2,
    alpha = 0.2,
    stroke = 0
  ) +
  
  geom_point(
    aes(fill = score_reg),
    data = . %>% filter(score_reg >=90),
    shape = 21, 
    size = 2,
    alpha = 1,
    stroke = 0.5
  ) +
  
     # Right Marginals 
  geom_violin(
    aes(y = abs_log2fc_diff, x = -10,),
    fill = "#4361ee",
    width = 5, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(y = abs_log2fc_diff, x = -10),
    width = 5, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  # Top Marginals 
  geom_violin(
    aes(x = abs_rank_log2fc_mouse, y = 10),
    fill = "#4361ee",
    width = 1, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(x = abs_rank_log2fc_mouse, y = 10),
    fill = "#4361ee",
    width = 1, alpha = 0, color = "black", outlier.shape = NA
  ) +

  scale_fill_continuous(palette = c("white", "#4cc9f0", "#4361ee")) +
  scale_x_reverse() +
  theme_vaxgo() +
  labs(
    x = "|Rank Mouse|",
    y = "|ΔLog2fc|",
    fill = "Score"
  ) +
  theme(legend.position = "top", legend.direction = "horizontal") 

# Display
reg_layers_mouse_score_rank_diff_log2fc 

#Save
# reg_layers_mouse_score_rank %>%
#   ggsave(filename = here("Figures", paste0("statistical_modelling_regression_scores_visualized_vsMouseRank.png")),
#          width = 12,
#          height = 3)
```


#### Rank human vs Rank mouse vs Abs Log2FC diff vs score
```{r fig.height=4, fig.width=4}
######## Rank human vs Rank mouse vs Abs Log2FC diff vs score
reg_layers_mouse_score_rank_human_mouse_diff_log2fc <-  
  legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, abs_rank_log2fc_mouse, abs_rank_log2fc_human,rank_log2fc_human, rank_log2fc_mouse, score_reg, p_adj_mouse) %>% 
  distinct() %>% 
 inner_join(
    eval_metrics_all_summary %>%
      filter(analysis_set == "Pooled"),
    by = join_by(treatment, pathogen)
  ) %>%
  filter(treatment == "Pooled") %>% 
  distinct() %>% 
  ggplot(aes(x = abs_rank_log2fc_mouse, y = abs_rank_log2fc_human
             )) +
  facet_nested_wrap(
    ~ treatment + pathogen,
    nrow = 1, 
    strip = strip_nested(size = "variable")
  ) +
  
  geom_point(
       aes(fill = score_reg
        ),
    shape = 21, 
    size = 2,
    alpha = 0.2,
    stroke = 0
  ) +
  
  geom_point(
    aes(fill = score_reg),
    data = . %>% filter(score_reg >=90),
    shape = 21,
    size = 2,
    alpha = 1,
    stroke = 0.5
  ) +

  scale_fill_continuous(palette = c("white", "#4cc9f0", "#4361ee")) +
  theme_vaxgo() +
  labs(
    x = "|Rank Mouse|",
    y = "|Rank Human|",
    fill = "Score",
    size = "AbsDiffLog2fc"
  ) +
  theme(legend.position = "top", legend.direction = "horizontal") 

# Display
reg_layers_mouse_score_rank_human_mouse_diff_log2fc 
```










### Combine plots
All processes, pooled condiitons
```{r fig.height=4, fig.width=12}

######## Shared Ranks vs. Score
legs_notlegs_layers_visualized_allprocesses_pooled = legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, abs_rank_log2fc_mouse, score_shared, p_adj_mouse, status, status_original) %>% 
  distinct() %>% 
  inner_join(
    eval_metrics_all_summary %>% 
      filter(analysis_set %in% c("Process_Pooled")),
    by = join_by(treatment, pathogen)
  ) %>% 
  filter(treatment == "Pooled") %>% 
  ggplot(aes(x = score_shared, y = abs_rank_log2fc_mouse, fill = status_original)) +
  
    facet_nested_wrap(
  ~ process + treatment + pathogen + 
    paste0("ROC-AUC=", round(auc_roc, 2), 
           " \nPR-AUC=", round(auc_pr, 2), 
           "\nrho=", round(spearman_rank_rho, 2), 
           ", padj=", format(p_val_spearman_rank, scientific = TRUE, digits = 2)), 
  nrow = 1, 
  strip = strip_nested(size = "variable")
) +
  
  geom_point(
    data = . %>% filter(status_original != "Shared"),
    shape = 21, alpha = 0.1, stroke = 0, size = 2
  ) +
  geom_point(
    data = . %>% filter(status_original == "Shared"),
    shape = 21, alpha = 1, stroke = 0.5, size = 2,
  ) +

  scale_fill_manual(values = colors$comparison) +
  theme_vaxgo() +
  labs(
    x = "Score",
    y = "Rank, Mouse",
    fill = "LEG Status",
  ) +
  
  # Right Marginals
  geom_violin(
    aes(y = abs_rank_log2fc_mouse, x = 110, fill = status_original),
    width = 15, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(y = abs_rank_log2fc_mouse, x = 110),
    width = 15, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  # Top Marginals
  geom_violin(
    aes(x = score_shared, y = 110, fill = status_original),
    width = 15, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(x = score_shared, y = 110),
    width = 15, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  theme(legend.position = "top")

# Display Plot
legs_notlegs_layers_visualized_allprocesses_pooled

######## Score vs AbsDiff
reg_layers_mouse_human_absdiff_pooled <-  
  legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, abs_rank_log2fc_mouse, score_reg, p_adj_mouse) %>% 
  distinct() %>% 
    filter(p_adj_mouse <= 0.05) %>%
  filter(treatment == "Pooled") %>% 
 inner_join(
    eval_metrics_all_summary %>%
      filter(analysis_set == "Pooled"),
    by = join_by(treatment, pathogen)
  ) %>%
  distinct() %>% 
  ggplot(aes(x = score_reg, y = abs_log2fc_diff, 
             # fill = status_original,
             fill = score_reg,
             )) +
  
  # Facet display showing regression metrics (R² and Spearman ρ) with formatted p-values
  facet_nested_wrap(
    ~ treatment + pathogen,
    nrow = 1, 
    strip = strip_nested(size = "variable")
  ) +
  
  # Layered scatter plot highlighting Shared vs. Human Only candidates
  geom_point(
    shape = 21, 
    size = 2,
    alpha = 1,
    stroke = 0.5
  ) +
   # Right Marginals 
  geom_violin(
    aes(y = abs_log2fc_diff, x = 110,),
    fill = "#4361ee",
    width = 5, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(y = abs_log2fc_diff, x = 110),
    width = 5, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  # Top Marginals 
  geom_violin(
    aes(x = score_reg, y = 6),
    fill = "#4361ee",
    width = 0.5, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(x = score_reg, y = 6),
    fill = "#4361ee",
    width = 0.5, alpha = 0, color = "black", outlier.shape = NA
  ) +
  scale_fill_continuous(palette = c("white", "#4cc9f0", "#4361ee")) +
  theme_vaxgo() +
  labs(
    x = "Score",
    y = "|ΔLog2fc|",
    fill = "Score",
  ) +
  theme(legend.position = "top")

# Display
reg_layers_mouse_human_absdiff_pooled

######## Mouse |Rank| vs Score =======
reg_layers_mouse_score_rank_pooled <-  
  legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, abs_rank_log2fc_mouse, score_reg, p_adj_mouse) %>% 
  distinct() %>% 
  
  filter(treatment == "Pooled") %>%
 inner_join(
    eval_metrics_all_summary %>%
      filter(analysis_set == "Pooled"),
    by = join_by(treatment, pathogen)
  ) %>%
  distinct() %>% 
  ggplot(aes(x = score_reg, y = abs_rank_log2fc_mouse, 
             fill = score_reg,
             )) +
  facet_nested_wrap(
    ~ treatment + pathogen,
    nrow = 1, 
    strip = strip_nested(size = "variable")
  ) +
  
  geom_point(
    aes(size = score_reg),
    shape = 21, 
    size = 2,
    alpha = 1,
    stroke = 0.5
  ) +
      # Right Marginals 
  geom_violin(
    aes(y = abs_rank_log2fc_mouse, x = 110,),
    fill = "#4361ee",
    width = 5, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(y = abs_rank_log2fc_mouse, x = 110),
    width = 5, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  # Top Marginals 
  geom_violin(
    aes(x = score_reg, y = -10),
    fill = "#4361ee",
    width = 5, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(x = score_reg, y = -10),
    fill = "#4361ee",
    width = 5, alpha = 0, color = "black", outlier.shape = NA
  ) +
  scale_fill_continuous(palette = c("white", "#4cc9f0", "#4361ee")) +
  theme_vaxgo() +
  labs(
    x = "Score",
    y = "Rank",
    fill = "Score",
    size = "Score",
  ) +
  theme(legend.position = "top", legend.direction = "horizontal") 

# Display
reg_layers_mouse_score_rank_pooled


#Merge
score_visualized_plots = (legs_notlegs_layers_visualized_allprocesses_pooled +theme(legend.position = "right", legend.direction = "vertical") |
reg_layers_mouse_human_absdiff_pooled + theme(legend.position = "right", legend.direction = "vertical") |
reg_layers_mouse_score_rank_pooled +theme(legend.position = "right", legend.direction = "vertical")) +
  plot_layout(guides = "collect") +
  plot_annotation(theme = theme(legend.position = "right"))

#Display
score_visualized_plots

#Save
score_visualized_plots %>%
  ggsave(filename = here("Figures", "statistical_modelling_regression_scores_visualized.png"),
         width = 12,
         height = 4)

```













