---
title: "6_Statistical_Modelling"
output: html_document
---
# Setup

```{r message=FALSE, warning=FALSE}
# Load core packages for the analysis session
library(here)
source(here("scripts_notebooks", "required.R"))

library(tidyverse)
library(tidymodels)
```


# Prepare features table
```{r}
# Gene expression
dge_btm_process_genes_diff_bygene_clean_filtered =  readRDS(file = here("tables", "dge_btm_process_genes_diff_bygene_clean_filtered.rds"))


#Protein identity
alignments_human2mouse_all = readRDS(file = here("tables", "alignments_human2mouse_all.rds"))

#Distance
human_mouse_cds_distance = readRDS(file = here::here("tables", "human_mouse_cds_distance.rds"))

#Transcription factors
gene_tf_shared_unique = readRDS(file = here("tables", "gene_tf_shared_unique.rds"))

#BTMs annotation
btm_annotation_genes <- read_csv(here("tables", "btm_annotation_genes.csv")) %>% 
  select(hgnc_symbol = symbol, process, group, subgroup) %>% 
  distinct()

btm_annotation_genes_others = bind_rows(
  btm_annotation_genes,
  #Add non-BTM genes
  dge_btm_process_genes_diff_bygene_clean_filtered %>% 
    select(hgnc_symbol) %>% 
    anti_join(btm_annotation_genes %>% 
                select(hgnc_symbol) %>% 
                distinct(),
              by = "hgnc_symbol") %>% 
    mutate(process = "Not-BTM",
           group = "Not-BTM",
           subgroup = "Not-BTM")
  )


btm_annotation_genes_wide = btm_annotation_genes_others %>% 
  select(hgnc_symbol, group) %>% 
  group_by(hgnc_symbol, group) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  pivot_wider(names_from = group,
              values_from = n) %>% 
  column_to_rownames("hgnc_symbol") %>% 
  as.matrix() %>% 
  replace(is.na(.), 0) %>% 
  as.data.frame() %>% 
  rownames_to_column("hgnc_symbol")


# CREs ===============
cres_type_homology_comparison_wide = 
  readRDS(file = here("tables", "cres_type_homology_comparison_wide.rds"))

#Merge features ==========
#By genes only
gene_annotated_layers = 
  #Annotate with CDS distance (Kimura)
  human_mouse_cds_distance %>% select(hgnc_symbol, dist_k80) %>% 
  #Annotate with AA identity %
  inner_join(alignments_human2mouse_all,
             by = join_by("hgnc_symbol" == "gene")) %>% 
  #Annotate with TFs
  inner_join(gene_tf_shared_unique,
             by = join_by("hgnc_symbol" == "gene")) %>% 
  #Annotate with CRES
  inner_join(cres_type_homology_comparison_wide,
             by = join_by(hgnc_symbol)) %>% 
     #Annotate with BTMs 
  inner_join(btm_annotation_genes_wide,
             by = "hgnc_symbol")

#Save
# gene_annotated_layers %>% 
#   saveRDS(file = here("tables", "human_mouse_statsmodelling_gene_annotated_layers.rds"))

#All parameters
human_mouse_parameters_values = 
  gene_annotated_layers %>% 
  #Annotate with gene expression
  inner_join(
    dge_btm_process_genes_diff_bygene_clean_filtered %>% 
               select(!c(p_adj_mouse, status_original, process, p_adjust_mouse_gsea, mean_log2fc_human, group, status_original, nes_human, nes_mouse)) %>% 
               distinct(),
             by = join_by(hgnc_symbol)) 

human_mouse_parameters_values

#Save
# human_mouse_parameters_values %>% 
#   saveRDS(file = here("tables", "human_mouse_statsmodelling_parameters_values.rds"))

human_mouse_parameters_values %>% 
  glimpse()



# #Create features for regression and classification
# human_mouse_features_regression = human_mouse_parameters_values %>%
#   select(-status) %>%
#   distinct() %>%
#   select(treatment:abs_log2fc_diff,inverse_se_mouse, hgnc_symbol, everything())
# human_mouse_features_regression
# 
# human_mouse_features_classification_legs = human_mouse_parameters_values %>%
#   drop_na(status) %>%
#   filter(status %in% c("Shared", "Not-shared")) %>%
#   mutate(status = fct_relevel(status, "Not-shared")) %>%
#   select(treatment:inverse_se_mouse, inverse_se_mouse, hgnc_symbol, status, everything()) %>%
#   select(!c(abs_log2fc_diff))
# human_mouse_features_classification_legs

```



# Statistical modelling

## Features dictionary
```{r}
#Import
human_mouse_parameters_values =
  readRDS(file = here("tables", "human_mouse_statsmodelling_parameters_values.rds"))

feature_dictionary <- tibble(
  feature_name_original = names(human_mouse_parameters_values)
) %>%
  filter(!feature_name_original %in% c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison", "status")) %>%
  
  # 1. Regex parsing for cCRE feature sets
  extract(
    col     = feature_name_original,
    into    = c("cre_metric", "cre_type", "cre_ctcf", "cre_homology"),
    regex   = "^(n_type|pct_match_type|n_ctcf|pct_match_ctcf|n_homology|pct_homology)_(PLS|pELS|dELS)(?:_(CTCF-bound|not-CTCF))?(?:_([A-Za-z0-9_]+))?$",
    remove  = FALSE
  ) %>%
  mutate(cre_ctcf = str_replace(cre_ctcf, "-bound", "")) %>% 
  
  # 2. Consolidated classification and label formatting
  mutate(
    # --- Category Assignment ---
    feature_type = case_when(
      feature_name_original %in% c("inverse_se_mouse", "abs_log2fc_diff", "mean_log2fc_mouse") ~ "Gene expression",
      feature_name_original %in% c("dist_k80", "identity_human2mouse") ~ "Sequence evolution",
      str_detect(feature_name_original, "^(n_tf|pct_tf)") ~ "Transcription Factor",
      feature_name_original %in% c(
        "Not-BTM", "INNATE RESPONSE", "B CELLS", "T CELLS", "PLATELETS", 
        "ECM AND MIGRATION", "ENERGY METABOLISM", "MONOCYTES", 
        "CELL CYCLE", "IFN", "SIGNAL TRANSDUCTION", "NK CELLS", "NEUTROPHILS"
      ) ~ "BTMs",
      feature_name_original == "n_total_cres_gene" ~ "CREs | Summary",
      
      !is.na(cre_type) & !is.na(cre_ctcf) ~ str_c(cre_type, " ", cre_ctcf),
      !is.na(cre_type)                    ~ cre_type,
      
      str_detect(feature_name_original, "PLS")  ~ "PLS",
      str_detect(feature_name_original, "pELS") ~ "pELS",
      str_detect(feature_name_original, "dELS") ~ "dELS",
      TRUE ~ "Other"
    ),
    
    # Clean whitespace and trailing commas
    feature_type = str_trim(feature_type),
    
    # --- Formatted Clean Feature Labels ---
    feature_name = case_when(
      feature_name_original == "mean_log2fc_mouse"      ~ "Mean Log2FC (Mouse)",
      feature_name_original == "abs_log2fc_diff"        ~ "|ΔMean Log2FC|",
      feature_name_original == "inverse_se_mouse"        ~ "1/SE",
      feature_name_original == "identity_human2mouse"   ~ "%Protein AA identity",
      feature_name_original == "dist_k80"                ~ "Kimura distance",
      
      feature_name_original == "n_tf_unique"             ~ "nUnique TFs",
      feature_name_original == "pct_tf_unique"           ~ "%Unique TFs",
      feature_name_original == "n_tf_shared"             ~ "nShared TFs",
      feature_name_original == "pct_tf_shared"           ~ "%Shared TFs",
      feature_name_original == "n_tf_total"              ~ "nTotal TFs",
      
      feature_name_original == "n_total_cres_gene"      ~ "nTotal cCREs",
      
      cre_metric == "n_type"                             ~ "nCREs",
      cre_metric == "pct_match_type"                    ~ "%Match Type",
      cre_metric == "n_ctcf"                             ~ "nCTCF",
      cre_metric == "pct_match_ctcf"                    ~ "%Match CTCF",
      
      cre_metric == "n_homology"                         ~ str_c("n", str_replace_all(cre_homology, "_", " ")),
      cre_metric == "pct_homology"                       ~ str_c("%", str_replace_all(cre_homology, "_", " ")),
      
      TRUE ~ feature_name_original
    )
  ) %>%
  
  # 3. Final selection, factor releveling, and ordering
  select(feature_name_original, feature_type, feature_name) %>% 
  mutate(
    feature_type = fct_relevel(
      feature_type, 
      "Gene expression", 
      "Sequence evolution", 
      "Transcription Factor", 
      "CREs | Summary",
      "PLS",              
      "PLS CTCF", 
      "PLS not-CTCF",
      "pELS",             
      "pELS CTCF", 
      "pELS not-CTCF",
      "dELS",             
      "dELS CTCF", 
      "dELS not-CTCF",
      "Other",
      "BTMs"
    )
  ) %>% 
  arrange(feature_type)

feature_dictionary

```


```{r}
# ==============================================================================
# 1. BUILDING BLOCKS (CATEGORY-SPECIFIC FEATURE DEFINITIONS)
# ==============================================================================

# Experimental metadata and target outcome
feat_base_experiment <- c("treatment", "pathogen", "timepoint_comparison", "hgnc_symbol")
feat_base_outcome    <- c("inverse_se_mouse", "abs_log2fc_diff", "status", "mean_log2fc_mouse")

# Biological Modules
feat_sequence_evolution <- c("dist_k80", "identity_human2mouse")

feat_btms <- c(
  "Not-BTM", "INNATE RESPONSE", "B CELLS", "T CELLS", "PLATELETS", 
  "ECM AND MIGRATION", "ENERGY METABOLISM", "MONOCYTES", "CELL CYCLE", 
  "IFN", "SIGNAL TRANSDUCTION", "NK CELLS", "NEUTROPHILS"
)

feat_tfs_selected    <- c("n_tf_total", "pct_tf_shared")


# 2. BIOLOGICAL ABLATION CONTROLS (NEGATIVE CONTROLS)

# [Ablation 1] Experimental metadata and DGE statistical precision only
features_ablation_dge_only <- c(feat_base_experiment, feat_base_outcome)

# [Ablation 2] DGE + Protein sequence conservation
features_ablation_dge_seq <- c(features_ablation_dge_only, feat_sequence_evolution)


# [Ablation 3] DGE + Sequence + Transcriptional Networks (TFs & BTMs) - NO epigenomics/cCREs
features_ablation_no_ccre <- c(features_ablation_dge_seq, feat_tfs_selected)


# ==============================================================================
# 3. REGULATORY RESOLUTION SETS (cCREs / ENHANCERS / PROMOTERS)
# ==============================================================================
features_ccre_baseline_minimal <- c(
  features_ablation_no_ccre,
  "n_total_cres_gene", 
  "n_type_PLS", "n_type_pELS", "n_type_dELS"
)


# [Resolution 4] High: Includes subtype matching rates (% match)
features_ccre_ctcf_match <- c(
  features_ablation_no_ccre,
  "n_total_cres_gene",
  ## PLS
  "n_type_PLS",                               # Count of PLS
  "pct_match_type_PLS",                       # % Matching PLS subtype
  "pct_match_ctcf_PLS_CTCF-bound",            # % Matching subtype
  ## pELS
  "n_type_pELS",                               # Count of pELS
  "pct_match_type_pELS",                       # % Matching pELS subtype
  "pct_match_ctcf_pELS_CTCF-bound",            # % Matching subtype
  ## dELS
  "n_type_dELS",                               # Count of dELS
  "pct_match_type_dELS",                       # % Matching dELS subtype
  "pct_match_ctcf_dELS_CTCF-bound"            # % Matching subtype
)

# [Resolution 5] Ultra-High: Adds complete homology and complex homology metrics
features_ccre_full_homology <- c(
  features_ablation_no_ccre,
  "n_total_cres_gene",
  
  # --- cCREs: Promoter-like Sequences (PLS) ---
  ## General PLS
  "n_type_PLS",                               # Count of PLS
  "pct_match_type_PLS",                       # % Matching PLS subtype
  "pct_match_ctcf_PLS_CTCF-bound",            # % Matching subtype
  "pct_homology_PLS_CTCF-bound_Homologous",   # % Homologous

  # --- cCREs: Proximal Enhancer-like Sequences (pELS) ---
  ## General pELS
  "n_type_pELS",                               # Count of pELS
  "pct_match_type_pELS",                       # % Matching pELS subtype
  "pct_match_ctcf_pELS_CTCF-bound",            # % Matching subtype
  "pct_homology_pELS_CTCF-bound_Homologous",   # % Homologous

  # --- cCREs: Distal Enhancer-like Sequences (dELS) ---
  ## General dELS
  "n_type_dELS",                               # Count of dELS
  "pct_match_type_dELS",                       # % Matching dELS subtype
  "pct_match_ctcf_dELS_CTCF-bound",            # % Matching subtype
  "pct_homology_dELS_CTCF-bound_Homologous"   # % Homologous
)



features_dge_seq_btms <- c(
  features_ccre_ctcf_match,
  feat_btms
)

# ==============================================================================
# 4. MASTER LIST FOR PIPELINE AND ROC EVALUATION
# ==============================================================================

feature_sets_list <- list(
  # --- Biological Ablation Controls ---
  "DGE Baseline"      = features_ablation_dge_only,
  "+ Sequence"        = features_ablation_dge_seq,
  "+ Sequence + TFs"             = features_ablation_no_ccre,
  
  # --- Regulatory Models (cCRE Resolution Gradient) ---
  "+ Sequence + TFs + CRE type"      = features_ccre_baseline_minimal,
  "+ Sequence + TFs + CRE type + CTCF"        = features_ccre_ctcf_match,
  # "+ Sequence + TFs + CRE type + CTCF + Homology"        = features_ccre_full_homology,
  # --- Functional Models ---
  "Full + BTM"            = features_dge_seq_btms
)

features_names_vector = c(
      "DGE Baseline",
      "+ Sequence",
      "+ Sequence + TFs",
      "+ Sequence + TFs + CRE type",
      "+ Sequence + TFs + CRE type + CTCF",
      "Full + BTM" 
    )

```

## Inputs
```{r}
#Create features tables for each model
input_features_selection = features_ccre_full_homology
input_features_name = "features_ccre_full_homology"

input_features = human_mouse_parameters_values %>% 
  select(all_of(input_features_selection))

input_features

#Create features for regression and classification
human_mouse_features_regression = input_features %>% 
  select(-status) %>% 
  distinct() %>% 
  select(treatment:abs_log2fc_diff,inverse_se_mouse, hgnc_symbol, everything())
human_mouse_features_regression

human_mouse_features_classification_legs = input_features %>% 
  drop_na(status) %>% 
  filter(status %in% c("Shared", "Not-shared")) %>% 
  mutate(status = fct_relevel(status, "Not-shared")) %>% 
  select(treatment:inverse_se_mouse, inverse_se_mouse, hgnc_symbol, status, everything()) %>% 
  select(!c(abs_log2fc_diff)) 
human_mouse_features_classification_legs
```

# Data exploration

## Correlation
```{r fig.height=8, fig.width=10}
library(Hmisc) # For fast pairwise correlation & p-value matrix computation

# 1. Prepare numeric data matrix
numeric_matrix <- input_features %>%
  select(where(is.numeric)) %>%
  select(where(~ var(.x, na.rm = TRUE) > 0)) %>%
  as.matrix()

# 2. Compute Spearman correlations and p-values simultaneously
cor_results <- Hmisc::rcorr(numeric_matrix, type = "spearman")

# Extract correlation (r) and p-value (P) matrices
cor_r <- cor_results$r
cor_p <- cor_results$P

# 3. Reorder variables using Hierarchical Clustering
ord <- hclust(as.dist(1 - cor_r))$order
var_names <- colnames(cor_r)[ord]

# 4. Convert matrices to long format and apply FDR (BH) p-value adjustment
cor_long <- cor_r %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  pivot_longer(cols = -var1, names_to = "var2", values_to = "spearman_rho") %>%
  inner_join(
    cor_p %>%
      as.data.frame() %>%
      mutate(var1 = rownames(.)) %>%
      pivot_longer(cols = -var1, names_to = "var2", values_to = "p_value"),
    by = c("var1", "var2")
  ) %>%
  # Handle self-correlations (diagonal p-values are NA in rcorr)
  mutate(p_value = if_else(var1 == var2, 0, p_value)) %>%
  # Perform FDR correction across all pairwise tests
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  # Add significance stars
  mutate(
    p_signif = case_when(
      p_adj <= 0.001 ~ "***",
      p_adj <= 0.01  ~ "**",
      p_adj <= 0.05  ~ "*",
      TRUE           ~ "" # Leave non-significant tiles blank (or use "ns")
    ),
    var1 = factor(var1, levels = var_names),
    var2 = factor(var2, levels = var_names)
  )

# 5. Join with feature_dictionary to attach category annotations
cor_long_annotated <- cor_long %>%
  inner_join(feature_dictionary, by = c("var1" = "feature_name_original")) %>%
  inner_join(feature_dictionary, by = c("var2" = "feature_name_original"), suffix = c("_var1", "_var2")) %>%
  mutate(
    feature_name_var1 = factor(feature_name_var1),
    feature_name_var2 = factor(feature_name_var2)
  )

# 6. Generate Plot with Significance Stars Layer
cor_long_annotated_matrixplot <- ggplot(
  cor_long_annotated, 
  aes(x = feature_name_var1, y = feature_name_var2, fill = spearman_rho)
) +
  geom_tile(color = "white", linewidth = 0.1) +
  
  # ANNOTATION LAYER: Overlay significance stars
  geom_text(
    aes(label = p_signif),
    color = "black",
    size = 2.5,
    vjust = 0.75 # Centers stars neatly within each tile
  ) +
  
  # Grid facets grouped by feature category
  facet_grid(
    feature_type_var2 ~ feature_type_var1, 
    scales = "free", 
    space = "free"
  ) +
  
  scale_fill_gradient2(
    low = "#4cc9f0",
    mid = "white",
    high = "#4361ee",
    midpoint = 0,
    limit = c(-1, 1),
    name = "Spearman\nRho (ρ)"
  ) +
  labs(
    x = "",
    y = "",
    title = "Spearman Correlation Matrix with FDR-Adjusted Significance (* p<0.05, ** p<0.01, *** p<0.001)"
  ) +
  theme_vaxgo() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    strip.text = element_text(size = 8, face = "bold"),
    strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5),
    strip.text.y = element_text(size = 8, hjust = 0, angle = 0),
    panel.grid = element_blank(),
    panel.spacing = unit(0.15, "lines")
  )

cor_long_annotated_matrixplot

cor_long_annotated_matrixplot %>%
  ggsave(filename = here("Figures", paste0( "statistical_modelling_features_matrixplot_", input_features_name, ".png")),
         width = 12, height = 10)

```









# Modelling

## Classify Shared vs Not-Shared LEGs

### One by one
```{r fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
# # Step 1: Define Classification Recipe with Near-Zero Variance Filter
# classification_recipe <- recipes::recipe(status ~ ., data = human_mouse_features_classification_legs) %>%
#   recipes::update_role(hgnc_symbol, treatment, pathogen, timepoint_comparison, new_role = "id") %>%
#   recipes::step_dummy(recipes::all_nominal_predictors()) %>%
#   recipes::step_nzv(recipes::all_predictors()) %>%
#   recipes::step_corr(recipes::all_numeric_predictors(), threshold = 0.9) %>%
#   recipes::step_normalize(recipes::all_numeric_predictors())
# 
# # Step 2: Fit Penalized Logistic Regression Model (glmnet)
# logistic_spec <- logistic_reg() %>%
#   parsnip::set_engine("glm") %>%
#   parsnip::set_mode("classification")
# 
# logistic_workflow <- workflows::workflow() %>%
#   workflows::add_recipe(classification_recipe) %>%
#   workflows::add_model(logistic_spec)
# 
# logistic_fit <- workflows::fit(logistic_workflow, data = human_mouse_features_classification_legs)
# 
# # Step 3: Extract Tidy Odds Ratios (OR) and Join Dictionary
# logistic_coefficients <- broom::tidy(logistic_fit, penalty = 0.01) %>%
#   dplyr::filter(term != "(Intercept)") %>%
#   dplyr::rename(feature_name_original = term) %>%
#   dplyr::mutate(
#     feature_name_original = stringr::str_replace_all(feature_name_original, "`", ""),
#     estimate = exp(estimate)
#   ) %>%
#   dplyr::left_join(
#     feature_dictionary,
#     by = dplyr::join_by("feature_name_original")
#   ) 
# 
# # Step 4: Compute Area Under the ROC Curve (AUC) Safely
# prob_df <- stats::predict(logistic_fit, new_data = human_mouse_features_classification_legs, type = "prob") %>%
#   dplyr::bind_cols(human_mouse_features_classification_legs %>% dplyr::select(status))
# 
# roc_obj_logistic <- pROC::roc(
#   response  = prob_df$status,
#   predictor = prob_df$.pred_Shared,
#   levels    = c("Not-shared", "Shared")
# )
# 
# cat("Logistic AUC:", round(pROC::auc(roc_obj_logistic), 3), "\n")
# 
# # APPROACH 2: RANDOM FOREST CLASSIFIER (GINI IMPORTANCE)
# rf_cls_spec <- rand_forest(
#   trees = 1000,
#   mtry  = 8,       # Fixed number of predictors sampled at each split
#   min_n = 5        # Fixed minimum node size
# ) %>%
#   parsnip::set_engine("ranger", importance = "impurity") %>%
#   parsnip::set_mode("classification")
# 
# # Assemble workflow with classification recipe
# rf_cls_workflow <- workflows::workflow() %>%
#   workflows::add_recipe(classification_recipe) %>%
#   workflows::add_model(rf_cls_spec)
# 
# # Fit Random Forest Classifier directly on the full classification dataset
# final_rf_cls_fit <- workflows::fit(
#   rf_cls_workflow, 
#   data = human_mouse_features_classification_legs
# )
# 
# # Extract native ranger engine
# rf_cls_engine <- workflows::extract_fit_engine(final_rf_cls_fit)
# 
# # Extract Gini Impurity feature importance metric
# rf_cls_importance <- tibble::tibble(
#   feature_name_original = names(rf_cls_engine$variable.importance),
#   importance            = as.numeric(rf_cls_engine$variable.importance)
# ) %>%
#   dplyr::inner_join(
#     feature_dictionary,
#     by = dplyr::join_by("feature_name_original")
#   )
# 
# # Step 4: Compute Area Under the ROC Curve (AUC)
# prob_df_rf <- stats::predict(final_rf_cls_fit, new_data = human_mouse_features_classification_legs, type = "prob") %>%
#   dplyr::bind_cols(human_mouse_features_classification_legs %>% dplyr::select(status))
# 
# roc_obj_rf <- pROC::roc(
#   response  = prob_df_rf$status,
#   predictor = prob_df_rf$.pred_Shared,
#   levels    = c("Not-shared", "Shared")
# )
# 
# cat("Logistic AUC:", round(pROC::auc(roc_obj_rf), 3), "\n")
# 
# # Reorder feature names
# feature_names_order = logistic_coefficients %>% 
#   arrange(estimate) %>% 
#   pull(feature_name) %>% 
#   unique()
# 
# # Step 5: Filter OR data and render Forest Plot
# p_odds_ratio_forest <- logistic_coefficients %>%
#   mutate(feature_name = fct_relevel(feature_name, c(feature_names_order))) %>% 
#   ggplot() +
#   aes(
#     x     = estimate,
#     y     = feature_name
#   ) +
#   facet_grid(feature_type~., scales = "free_y", space = "free_y") +
#   geom_vline(xintercept = 1, linetype = "dashed", color = "gray50", linewidth = 0.6) +
#     geom_point(aes(fill = feature_type),
#                         size = 2.5, shape = 21, 
#                       color = "black") +
#   scale_fill_manual(values = colors_all$blues) +
#   labs(
#     x     = "Odds Ratio",
#     y     = "Feature Predictor",
#     fill  = "Feature Category",
#     title = paste0("Logistic Regression:\nShared LEGs (AUC = ", round(pROC::auc(roc_obj_logistic), 3), ")")
#   ) +
#   theme_vaxgo() +
#   theme(
#     legend.position = "none",
#     legend.location = "plot",
#     axis.text.y     = element_text(size = 10), 
#     strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)
#   ) 
# 
# # Render Variable Importance Plot for Random Forest Classifier
# p_rf_cls_importance <- rf_cls_importance %>%
#   mutate(feature_name = fct_relevel(feature_name, c(feature_names_order))) %>% 
#   ggplot() +
#   aes(x = importance, y = feature_name, fill = feature_type) +
#   facet_grid(feature_type~., scales = "free_y", space = "free_y") +
#   geom_col(width = 0.6) +
#   scale_fill_manual(values = colors_all$blues) +
#   labs(
#     x     = "Gini Impurity",
#     y     = "Feature Predictor",
#     fill  = "Feature Category",
#     title = paste0("Random Forest Classifier:\nShared LEGs (AUC = ", round(pROC::auc(roc_obj_rf), 3), ")"
#   )) +
#   theme_vaxgo() +
#   theme(
#     legend.position = "none",
#     legend.location = "plot",
#     axis.text.y     = element_text(size = 10), 
#     strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)
#   ) 
# 
# 
# # 4. Merge plots (LM vs RF)
# coefficients_classifier_patch_plot = (p_odds_ratio_forest + theme(strip.text.y = element_blank())) +
#   p_rf_cls_importance +
#   plot_layout(widths = c(0.5, 0.5), guides = "collect", axes = "collect") +
#   plot_annotation(title = input_features_name)
# 
# coefficients_classifier_patch_plot 
# 
# coefficients_classifier_patch_plot %>%
#   ggsave(filename = here("Figures", paste0( "statistical_modelling_coefficients_classifier_", input_features_name, ".png")),
#          width = 8,
#          height = 1.5)

```


ROC curve
```{r fig.height=4, fig.width=4}
# # ROC curve
# df_roc_logistic <- pROC::ggroc(roc_obj_logistic)$data %>%
#   dplyr::mutate(
#     model = paste0("Logistic Regression (AUC = ", round(pROC::auc(roc_obj_logistic), 3), ")")
#   )
# 
# df_roc_rf <- pROC::ggroc(roc_obj_rf)$data %>%
#   dplyr::mutate(
#     model = paste0("Random Forest (AUC = ", round(pROC::auc(roc_obj_rf), 3), ")")
#   )
# 
# df_roc_combined <- dplyr::bind_rows(df_roc_logistic, df_roc_rf)
# 
# p_roc_comparison <- ggplot(df_roc_combined) +
#   aes(
#     x = 1 - specificity, # Convert to False Positive Rate
#     y = sensitivity, 
#     color = model
#   ) +
#   
#   # Diagonal reference line for random classifier (AUC = 0.5)
#   geom_segment(
#     aes(x = 0, y = 0, xend = 1, yend = 1),
#     linetype = "dashed", 
#     color = "gray60", 
#     linewidth = 0.6
#   ) +
#   
#   # Model ROC curves
#   geom_line(linewidth = 1.2) +
#   
#   # Color palette and scale adjustments
#   scale_color_manual(values = c(
#     "Logistic Regression" = "#001219", 
#     "Random Forest"       = "#4DBBD5FF"
#   ) %>% setNames(unique(df_roc_combined$model))) +
#   
#   scale_x_continuous(expand = c(0, 0.01), limits = c(0, 1)) +
#   scale_y_continuous(expand = c(0, 0.01), limits = c(0, 1)) +
#   
#   coord_fixed() + # Force 1:1 aspect ratio on the plot
#   
#   labs(
#     x = "1 - Specificity (False Positive Rate)",
#     y = "Sensitivity (True Positive Rate)",
#     color = "Classifier Model",
#     title = paste0("ROC Curves: LEGs Classification\n",
#                    input_features_name)
#   ) +
#   
#   theme_vaxgo() +
#   theme(
#     legend.position = c(0.65, 0.2),
#     legend.text = element_text(size = 9, face = "bold"),
#     plot.title.position = "plot"
#   )
# 
# #Display
# p_roc_comparison
# 
# #Save
# p_roc_comparison %>%
#   ggsave(filename = here("Figures", paste0( "statistical_modelling_coefficients_classifier_roccurve_", input_features_name, ".png")),
#          width = 4,
#          height = 4)

```





### Iterate over all features

```{r fig.height=12, fig.width=10}
# Create a directory to store the lightweight RDS files if it doesn't exist
dir.create(here::here("Models"), showWarnings = FALSE)

# Execute the iterative classification pipeline
all_classification_results <- imap(feature_sets_list, function(current_selection, set_name) {
  
  cat("\n=======================================================\n")
  cat(" FITTING CLASSIFICATION MODELS:", set_name, "\n")
  cat("=======================================================\n")
  
  clean_name <- str_replace_all(set_name, "[ +]+", "_") %>% 
    str_replace_all("^_|_$", "")
  
  # ==============================================================================
  # 1. PREPARE DATASET
  # ==============================================================================
  df_classification <- human_mouse_parameters_values %>%
    select(all_of(current_selection)) %>%
    drop_na(status) %>%
    filter(status %in% c("Shared", "Not-shared")) %>%
    mutate(status = factor(status, levels = c("Not-shared", "Shared"))) %>%
    select(!any_of("abs_log2fc_diff"))
  
  # ==============================================================================
  # 2. RECIPE DEFINITION
  # ==============================================================================
  cls_recipe <- recipe(status ~ ., data = df_classification) %>%
    update_role(any_of(c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison")), new_role = "id") %>%
    step_dummy(all_nominal_predictors()) %>%
    step_nzv(all_predictors()) %>%
    step_corr(all_numeric_predictors(), threshold = 0.9) %>%
    step_normalize(all_numeric_predictors())
  
  prep_rec <- prep(cls_recipe, training = df_classification)
  baked_predictors <- bake(prep_rec, new_data = NULL) %>% 
    select(-status, -has_role("id"))
  
  num_cols <- ncol(baked_predictors)
  
  # ==============================================================================
  # 3. MODEL FITTING
  # ==============================================================================
  
  # --- Logistic Regression (Native GLM) ---
  logistic_spec <- logistic_reg() %>% 
    set_engine("glm")
  
  fit_logistic <- workflow() %>%
    add_recipe(cls_recipe) %>%
    add_model(logistic_spec) %>%
    fit(data = df_classification)
  
  # --- Random Forest ---
  rf_mtry <- min(8, max(1, num_cols))
  
  rf_spec <- rand_forest(trees = 1000, mtry = rf_mtry, min_n = 5) %>% 
    set_engine("ranger", importance = "impurity") %>% 
    set_mode("classification")
  
  fit_rf <- workflow() %>%
    add_recipe(cls_recipe) %>%
    add_model(rf_spec) %>%
    fit(data = df_classification)
  
  # ==============================================================================
  # 4. EXTRACT COEFFICIENTS & IMPORTANCE
  # ==============================================================================
  
  log_coefs <- tidy(extract_fit_engine(fit_logistic), conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    rename(feature_name_original = term) %>%
    mutate(
      feature_name_original = str_replace_all(feature_name_original, "`", ""),
      feature_set = set_name
    ) %>% 
    inner_join(feature_dictionary, by = join_by("feature_name_original"))
  
  rf_engine <- extract_fit_engine(fit_rf)
  rf_imp <- tibble(
    feature_name_original = names(rf_engine$variable.importance),
    importance            = as.numeric(rf_engine$variable.importance),
    feature_set           = set_name
  ) %>%
    inner_join(feature_dictionary, by = join_by("feature_name_original"))
  
  # ==============================================================================
  # 5. EVALUATION METRICS (ROC)
  # ==============================================================================
  
  pred_logistic <- stats::predict(fit_logistic, new_data = df_classification, type = "prob")$.pred_Shared
  pred_rf       <- stats::predict(fit_rf, new_data = df_classification, type = "prob")$.pred_Shared
  
  roc_logistic  <- roc(response = df_classification$status, predictor = pred_logistic, levels = c("Not-shared", "Shared"))
  roc_rf        <- roc(response = df_classification$status, predictor = pred_rf, levels = c("Not-shared", "Shared"))
  
  auc_log <- round(auc(roc_logistic), 3)
  auc_rf  <- round(auc(roc_rf), 3)
  
  metrics_table <- tibble(
    feature_set = set_name,
    model = c("Logistic Regression", "Random Forest"),
    auc = c(auc_log, auc_rf)
  )
  
  roc_data <- bind_rows(
    ggroc(roc_logistic)$data %>% mutate(model = "Logistic Regression", auc = auc_log),
    ggroc(roc_rf)$data %>% mutate(model = "Random Forest", auc = auc_rf)
  ) %>% 
    mutate(
      feature_set = set_name,
      curve_label = paste0(model, " - ", set_name, " (AUC = ", auc, ")")
    )
  
  # ==============================================================================
  # 6. COMPILE AND SAVE LIGHTWEIGHT RDS
  # ==============================================================================
  
  iteration_results <- list(
    set_name     = set_name,
    metrics      = metrics_table,
    coefficients = log_coefs,
    importance   = rf_imp,
    roc_data     = roc_data
  )
  
  rds_path <- here::here("Models", paste0("classification_extracted_data_", clean_name, ".rds"))
  saveRDS(iteration_results, file = rds_path)
  
  return(iteration_results)
})


# ==============================================================================
# GENERATE MASTER TABLES
# ==============================================================================

master_class_metrics <- map_dfr(all_classification_results, "metrics") %>% 
  mutate(
    feature_set = fct_relevel(feature_set, features_names_vector)
  )
saveRDS(master_class_metrics, file = here::here("Models", "master_class_metrics.rds"))

master_class_coefs <- map_dfr(all_classification_results, "coefficients") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))
saveRDS(master_class_coefs, file = here::here("Models", "master_class_coefs.rds"))

master_class_var_imp <- map_dfr(all_classification_results, "importance")  %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

saveRDS(master_class_var_imp, file = here::here("Models", "master_class_var_imp.rds"))

master_class_roc_data <- map_dfr(all_classification_results, "roc_data")  %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

saveRDS(master_class_roc_data, file = here::here("Models", "master_class_roc_data.rds"))

# Generate wide summary table of AUC values across models and feature sets
classification_auc_table <- master_class_metrics %>%
  pivot_wider(names_from = model, values_from = auc)  %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector)) %>% 
  arrange(feature_set)

classification_auc_table

# ==============================================================================
# 5. VISUALIZATIONS
# ==============================================================================

master_class_metrics = readRDS(file = here::here("Models", "master_class_metrics.rds"))
master_class_coefs = readRDS(file = here::here("Models", "master_class_coefs.rds"))
master_class_var_imp = readRDS(file = here::here("Models", "master_class_var_imp.rds"))
master_class_roc_data = readRDS(file = here::here("Models", "master_class_roc_data.rds"))


# Logistic Regression Coefficients Plot================
log_coefs_plot_df = master_class_coefs %>%
  inner_join(
    master_class_metrics %>% filter(model == "Logistic Regression"),
    by = "feature_set"
  ) %>%
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  mutate(
    feature_set_label = paste0(feature_set, "\n(AUC = ", auc, ")"),
    feature_set       = fct_relevel(feature_set, features_names_vector),
    feature_name = if_else(
      str_detect(feature_type, "PLS|pELS|dELS"),
      paste(feature_type, feature_name),
      feature_name
    )
  ) %>%
  arrange(feature_set) %>%
  mutate(feature_name = fct_rev(fct_inorder(feature_name)))


log_coefs_plot = log_coefs_plot_df %>% 
  ggplot() +
  aes(x = estimate, 
      y = feature_set %>% fct_rev(), 
      xmin = conf.low, 
      xmax = conf.high, 
      fill = feature_set, 
      group = feature_set) +
  facet_grid(feature_name %>% fct_rev() ~., scales="free_y", space="free", switch = "y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.6) +
  geom_errorbarh(height = 0.2, linewidth = 0.6, width = 0,
                 alpha = 0.5,
                 position = position_dodge(width = 1), 
                 show.legend = FALSE) +
  geom_point(size = 2.2, pch = 21, position = position_dodge(width = 1)) +
  scale_fill_manual(values = colors$feature_sets) +
  labs(
    x = "Log-Odds",
    y = "Feature",
    fill = "Feature Set"
  ) +
  theme_vaxgo() +
  theme(legend.position = "top", 
    legend.location = "plot",
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1, vjust = 0.5),
    axis.text.y = element_blank()) +
  guides(fill = guide_legend(override.aes = list(pch = 22),
                             ncol = 1))

# 2. Random Forest Importance Plot================
# Classification RF Importance Data Preparation
rf_imp_plot_class_df <- master_class_var_imp %>%
  inner_join(
    master_class_metrics %>% filter(model == "Random Forest"),
    by = "feature_set"
  ) %>%
  mutate(
    feature_set_label = paste0(feature_set, "\n(AUC = ", auc, ")"),
    feature_set       = fct_relevel(feature_set, features_names_vector),
    feature_name = if_else(
      str_detect(feature_type, "PLS|pELS|dELS"),
      paste(feature_type, feature_name),
      feature_name
    )
  ) %>%
  arrange(feature_set) %>%
  mutate(feature_name = fct_rev(fct_inorder(feature_name)))

# Classification RF Importance Visualization
rf_imp_plot_class <- rf_imp_plot_class_df %>%
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  ggplot() +
  aes(
    x    = importance + 1, 
    y    = feature_set %>% fct_rev(), 
    fill = feature_set
  ) +
  facet_grid(feature_name %>% fct_rev() ~ ., scales = "free_y", space = "free", switch = "y") +
  geom_col(
    width    = 0.6, 
    position = position_dodge(1)
  ) +
  scale_fill_manual(values = colors$feature_sets) +
  scale_x_log10(
    breaks = c(1, 10, 50, 100, 250, 1000), 
    labels = c("0", "10", "50", "100", "250", "1000"), 
    expand = c(0, 0)
  ) +
  labs(
    x    = "Gini",
    y    = "Feature",
    fill = "Feature Set"
  ) +
  theme_vaxgo() +
  theme(
    legend.position   = "top", 
    legend.location   = "plot",
    strip.placement   = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1, vjust = 0.5),
    axis.text.y       = element_blank()
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(pch = 22),
      byrow        = TRUE,
      ncol         = 1
    )
  )

rf_imp_plot_class

# AUC ROC Bar Chart Comparison ========================
p_auc_bars <- master_class_metrics  %>% 
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector)) %>% 
  ggplot() +
  aes(x = auc, y = feature_set, fill = model) +
  facet_wrap(.~feature_set,
             scales = "free_y",
             ncol = 1) +
  # Draw side-by-side bars
  geom_col(position = position_dodge(width = 0.8), width = 0.7, color = "black", linewidth = 0.3) +
  # Add exact AUC values on top of the bars
  geom_text(
    aes(label = sprintf("%.3f", auc)), 
    position = position_dodge(width = 0.8), 
    vjust = 0.5,
    hjust = 1.5,
    size = 3,
    color = "black",
    fontface = "bold",
  ) +
  # Match the color scheme used in previous algorithm comparisons
  scale_fill_manual(values = c(
    "Random Forest" = "#4DBBD5FF", 
    "Logistic Regression" = "gray70" 
  )) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    x = "AUC ROC",
    y = "",
    fill = "Algorithm"
  ) +
  theme_vaxgo() +
  theme(
    legend.position = "top", 
    legend.title.position = "top",
    axis.text.y = element_blank(),
    plot.title.position = "plot",
    strip.text.x = element_text(hjust = 0)
  ) +
   guides(fill = guide_legend(byrow = T, nrow = 2))

# Merge plots
class_patch_plot_1 <- (
  (log_coefs_plot + theme(strip.text.y = element_blank(), legend.position = "none")) |
  (rf_imp_plot_class + guides(fill = guide_legend(byrow = TRUE, nrow = 2)) + 
     theme(legend.title.position = "top", legend.location = "plot", legend.justification = "left",
           strip.text.y.left = element_blank()))
) +
  plot_layout(guides = "collect", axes = "collect") +
  plot_annotation(theme = theme(legend.position = "top", legend.location = "plot", legend.justification = "left"))

# Final merge
class_patch_plot <- (class_patch_plot_1 |p_auc_bars) 

#Display
class_patch_plot

#Save
class_patch_plot %>% 
ggsave(filename = here("Figures", "statistical_modelling_classification_coefficients_rocauc_allmodels.png"),
         width = 10,
         height = 12)
```








## Predict Log2FC Divergence Magnitude

### One by one
```{r fig.height=12, fig.width=12}
# library(mgcv)     # For GAM engine
# library(vip)      # For Permutation Importance
# 
# # ==============================================================================
# # 1. RECIPE DEFINITION
# # ==============================================================================
# 
# regression_recipe <- recipes::recipe(abs_log2fc_diff ~ ., data = human_mouse_features_regression) %>%
#   recipes::update_role(hgnc_symbol, treatment, pathogen, timepoint_comparison, new_role = "id") %>%
#   recipes::step_dummy(recipes::all_nominal_predictors()) %>%
#   recipes::step_zv(recipes::all_predictors()) %>%
#   recipes::step_normalize(recipes::all_numeric_predictors())
# 
# 
# # ==============================================================================
# # 2. MODEL SPECIFICATIONS & FITTING
# # ==============================================================================
# 
# # MODEL 1: Linear Regression (LM) ----------------------------------------------
# linear_spec <- linear_reg() %>%
#   set_engine("lm") %>%
#   set_mode("regression")
# 
# linear_fit <- workflow() %>%
#   add_recipe(regression_recipe) %>%
#   add_model(linear_spec) %>%
#   fit(data = human_mouse_features_regression)
# 
# linear_lm_engine <- extract_fit_engine(linear_fit)
# 
# linear_coefficients <- broom::tidy(linear_lm_engine, conf.int = TRUE) %>%
#   filter(term != "(Intercept)") %>%
#   rename(feature_name_original = term) %>%
#   mutate(feature_name_original = str_replace_all(feature_name_original, "`", "")) %>% 
#   inner_join(feature_dictionary, by = join_by("feature_name_original"))
# 
# 
# # MODEL 2: Generalized Linear Model (GLM - Gamma / Log Link for skewed |Log2FC|)
# # Note: For non-negative continuous outcomes like |ΔLog2FC|, Gamma GLM is ideal.
# glm_spec <- linear_reg() %>%
#   set_engine("glm", family = stats::Gamma(link = "log")) %>%
#   set_mode("regression")
# 
# glm_fit <- workflow() %>%
#   add_recipe(regression_recipe) %>%
#   add_model(glm_spec) %>%
#   fit(data = human_mouse_features_regression)
# 
# glm_engine <- extract_fit_engine(glm_fit)
# 
# glm_coefficients <- broom::tidy(glm_engine, conf.int = TRUE) %>%
#   filter(term != "(Intercept)") %>%
#   rename(feature_name_original = term) %>%
#   mutate(feature_name_original = str_replace_all(feature_name_original, "`", "")) %>% 
#   inner_join(feature_dictionary, by = join_by("feature_name_original"))
# 
# 
# # MODEL : Random Forest (RF) --------------------------------------------------
# rf_spec <- rand_forest(
#   trees = 1000,
#   mtry  = 8,       
#   min_n = 5        
# ) %>%
#   set_engine("ranger", importance = "impurity") %>%
#   set_mode("regression")
# 
# rf_workflow <- workflow() %>%
#   add_recipe(regression_recipe) %>%
#   add_model(rf_spec)
# 
# final_rf_fit <- fit(rf_workflow, data = human_mouse_features_regression)
# rf_engine <- extract_fit_engine(final_rf_fit)
# 
# rf_importance <- tibble(
#   feature_name_original = names(rf_engine$variable.importance),
#   importance            = as.numeric(rf_engine$variable.importance)
# ) %>%
#   inner_join(feature_dictionary, by = join_by("feature_name_original"))
# 
# # ==============================================================================
# # 4. EVALUATION & MODEL COMPARISON METRICS
# # ==============================================================================
# 
# eval_table <- human_mouse_features_regression %>%
#   select(abs_log2fc_diff) %>%
#   bind_cols(
#     predict(linear_fit, new_data = human_mouse_features_regression) %>% rename(pred_linear = .pred),
#     predict(final_rf_fit, new_data = human_mouse_features_regression) %>% rename(pred_rf = .pred)
#   )
# 
# # Calculate RMSE and R2 across all 4 models
# metrics_lm  <- metric_set(rmse, rsq)(eval_table, truth = abs_log2fc_diff, estimate = pred_linear) %>% mutate(model = "Linear (LM)")
# metrics_rf  <- metric_set(rmse, rsq)(eval_table, truth = abs_log2fc_diff, estimate = pred_rf)     %>% mutate(model = "Random Forest")
# 
# model_comparison_metrics <- bind_rows(metrics_lm, metrics_rf)
# print(model_comparison_metrics)
# 
# # Preserve your exact factor level ordering logic
# name_order <- rf_importance %>%
#   arrange(importance) %>% 
#   mutate(feature_name = fct_inorder(feature_name))
# 
# # ==============================================================================
# # 5. VISUALIZATIONS
# # ==============================================================================
# 
# # 1. LM Forest Plot
# linear_coefficients_forest_plot <- linear_coefficients %>%
#   mutate(feature_name = factor(feature_name, levels = levels(name_order$feature_name))) %>%
#   ggplot(aes(x = estimate, y = feature_name, xmin = conf.low, xmax = conf.high)) +
#   facet_grid(feature_type ~ ., space = "free", scales = "free_y") +
#   geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.6) +
#   geom_errorbarh(height = 0.2, color = "black", linewidth = 0.6) +
#   geom_point(aes(color = p.value <= 0.05), size = 2.5) +
#   scale_color_manual(
#     values = c("TRUE" = "#4DBBD5FF", "FALSE" = "gray60"),
#     labels = c("TRUE" = "p <= 0.05", "FALSE" = "p > 0.05")
#   ) +
#   labs(
#     x     = "Standardized Effect Size",
#     y     = "Feature Predictor",
#     color = "Significance",
#     title = paste0("Linear Regression (R² = ", round(metrics_lm$ .estimate[metrics_lm$.metric == "rsq"], 3), ")")
#   ) +
#   theme_vaxgo() +
#   theme(legend.position = "right", strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5))
# 
# # 2. GLM Forest Plot
# glm_coefficients_forest_plot <- glm_coefficients %>%
#   mutate(feature_name = factor(feature_name, levels = levels(name_order$feature_name))) %>%
#   ggplot(aes(x = estimate, y = feature_name, xmin = conf.low, xmax = conf.high)) +
#   facet_grid(feature_type ~ ., space = "free", scales = "free_y") +
#   geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.6) +
#   geom_errorbarh(height = 0.2, color = "black", linewidth = 0.6) +
#   geom_point(aes(color = p.value <= 0.05), size = 2.5) +
#   scale_color_manual(
#     values = c("TRUE" = "#4DBBD5FF", "FALSE" = "gray60"),
#     labels = c("TRUE" = "p <= 0.05", "FALSE" = "p > 0.05")
#   ) +
#   labs(
#     x     = "Standardized Effect Size",
#     y     = "Feature Predictor",
#     color = "Significance",
#     title = paste0("GLM (R² = ", round(metrics_glm$.estimate[metrics_glm$.metric == "rsq"], 3), ")")
#   ) +
#   theme_vaxgo() +
#   theme(legend.position = "right", strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5))
# 
# # 3. Random Forest Gini Importance Plot
# p_rf_importance <- rf_importance %>%
#   mutate(feature_name = factor(feature_name, levels = levels(name_order$feature_name))) %>%
#   ggplot(aes(x = importance + 1, y = feature_name, fill = feature_type)) +
#   facet_grid(feature_type ~ ., space = "free", scales = "free_y") +
#   geom_col(width = 0.6) +
#   scale_fill_manual(values = colors_all$blues) +
#   scale_x_log10(
#     breaks = c(1, 10, 50, 100, 250, 1000),
#     labels = c("0", "10", "50", "100", "250", "1000")
#   ) +
#   labs(
#     x     = "Variable Importance (Gini)",
#     y     = "Feature Predictor",
#     fill  = "Feature Category",
#     title = paste0("Random Forest (R² = ", round(metrics_rf$.estimate[metrics_rf$.metric == "rsq"], 3), ")")
#   ) +
#   theme_vaxgo() +
#   theme(legend.position = "none", strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5))
# 
# 
# # 4. Merge plots (LM vs RF)
# coeffiecients_patch_plot = (linear_coefficients_forest_plot + theme(strip.text.y = element_blank())) +
#   p_rf_importance +
#   plot_layout(widths = c(0.5, 0.5), guides = "collect", axes = "collect")
# 
# coeffiecients_patch_plot
# 
# 
# coeffiecients_patch_plot %>%
#   ggsave(filename = here("Figures", paste0("statistical_modelling_regression_coefficients_patch_plot_",input_features_name,  ".png")),
#          width = 8,
#          height = 8)
# 
# #### Plot Observed vs Predicted values across models ============
# 
# # 1. Extract R² values dynamically for accurate facet labels
# r2_lm  <- round(model_comparison_metrics$.estimate[model_comparison_metrics$model == "Linear (LM)" & model_comparison_metrics$.metric == "rsq"], 3)
# r2_rf  <- round(model_comparison_metrics$.estimate[model_comparison_metrics$model == "Random Forest" & model_comparison_metrics$.metric == "rsq"], 3)
# 
# # 2. Reshape predictions and plot Observed vs. Predicted across all 4 models
# p_obs_vs_pred <- eval_table %>%
#   pivot_longer(
#     cols      = c(pred_linear, pred_rf),
#     names_to  = "model_key",
#     values_to = "predicted"
#   ) %>%
#   mutate(
#     model = case_when(
#       model_key == "pred_linear" ~ paste0("Linear Model (R² = ", r2_lm, ")"),
#       model_key == "pred_rf"     ~ paste0("Random Forest (R² = ", r2_rf, ")"),
#       TRUE ~ NA_character_
#     ),
#     # Relevel factors to preserve structural layout from linear to complex models
#     model = factor(model, levels = c(
#       paste0("Linear Model (R² = ", r2_lm, ")"),
#       paste0("Random Forest (R² = ", r2_rf, ")")
#     ))
#   ) %>%
#   ggplot(aes(x = abs_log2fc_diff, y = predicted)) +
#   
#   # Identity reference line (1:1 perfect prediction)
#   geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.6) +
#   
#   # Scatter points
#   geom_point(alpha = 0.15, size = 1, color = "black") +
#   
#   # Linear calibration fit line per model
#   geom_smooth(method = "lm", se = TRUE, linewidth = 1, color = "#4DBBD5FF") +
#   
#   # 2x2 Facet grid layout
#   facet_wrap(~ model, nrow = 1) +
#   
#   scale_x_continuous(expand = c(0, 0)) +
#   scale_y_continuous(expand = c(0, 0)) +
#   
#   # Force 1:1 aspect ratio across axes
#   tune::coord_obs_pred() +
#   
#   labs(
#     x = "Observed Divergence",
#     y = "Predicted",
#     title = paste0("Observed vs. Predicted\nLog2FC Divergence\n", input_features_name)
#   ) +
#   theme_vaxgo() +
#   theme(
#     legend.position = "none",
#     plot.title.position = "plot",
#     strip.text = element_text(size = 10, face = "bold")
#   )
# 
# # Display Plot
# print(p_obs_vs_pred)
# 
# #Save
# p_obs_vs_pred %>%
#   ggsave(filename = here("Figures", paste0("statistical_regression_modelling_p_obs_vs_pred_",input_features_name,  ".png")),
#          width = 5,
#          height = 3)
```






### Iterate over multiple models

```{r fig.height=12, fig.width=10}
# Create a directory to store the lightweight RDS files if it doesn't exist
dir.create(here::here("Models"), showWarnings = FALSE)

# Execute the iterative regression pipeline and store extracted data in a list
all_regression_results <- imap(feature_sets_list, function(current_selection, set_name) {
  
  cat("\n=======================================================\n")
  cat(" FITTING REGRESSION MODELS:", set_name, "\n")
  cat("=======================================================\n")
  
  # Clean the set name for file saving 
  clean_name <- str_replace_all(set_name, "[ +]+", "_") %>% 
    str_replace_all("^_|_$", "")
  
  # ==============================================================================
  # 1. PREPARE DATASET 
  # ==============================================================================
  reg_cols <- setdiff(current_selection, "status")
  if (!"abs_log2fc_diff" %in% reg_cols) {
    reg_cols <- c(reg_cols, "abs_log2fc_diff")
  }
  
  df_regression <- human_mouse_parameters_values %>%
    select(all_of(reg_cols)) %>%
    drop_na(abs_log2fc_diff) %>%
    distinct()
  
  # ==============================================================================
  # 2. RECIPE DEFINITION 
  # ==============================================================================
  reg_recipe <- recipe(abs_log2fc_diff ~ ., data = df_regression) %>%
    update_role(any_of(c("hgnc_symbol", "treatment", "pathogen", "timepoint_comparison")), new_role = "id") %>%
    step_dummy(all_nominal_predictors()) %>%
    step_zv(all_predictors()) %>%
    step_normalize(all_numeric_predictors())
  
  # Determine number of valid predictors
  prep_rec <- prep(reg_recipe, training = df_regression)
  baked_pred <- bake(prep_rec, new_data = NULL) %>% 
    select(-abs_log2fc_diff, -has_role("id"))
  
  num_cols <- ncol(baked_pred)
  
  # ==============================================================================
  # 3. MODEL FITTING 
  # ==============================================================================
  
  # --- Linear Regression (LM) ---
  lm_spec <- parsnip::linear_reg() %>%
    parsnip::set_engine("lm") %>%
    parsnip::set_mode("regression")
  
  lm_fit <- workflow() %>%
    add_recipe(reg_recipe) %>%
    add_model(lm_spec) %>%
    fit(data = df_regression)
  
  # --- Random Forest (RF) ---
  rf_mtry <- min(8, max(1, num_cols))
  
  rf_spec <- parsnip::rand_forest(trees = 1000, mtry = rf_mtry, min_n = 5) %>%
    parsnip::set_engine("ranger", importance = "impurity") %>%
    parsnip::set_mode("regression")
  
  rf_fit <- workflow() %>%
    add_recipe(reg_recipe) %>%
    add_model(rf_spec) %>%
    fit(data = df_regression)
  
  # ==============================================================================
  # 4. EXTRACT COEFFICIENTS & IMPORTANCE
  # ==============================================================================
  
  lm_coefs <- tidy(extract_fit_engine(lm_fit), conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    rename(feature_name_original = term) %>%
    mutate(
      feature_name_original = str_replace_all(feature_name_original, "`", ""),
      feature_set = set_name # Add identifier for the master table
    ) %>% 
    inner_join(feature_dictionary, by = join_by("feature_name_original"))
  
  rf_engine <- extract_fit_engine(rf_fit)
  rf_imp <- tibble(
    feature_name_original = names(rf_engine$variable.importance),
    importance            = as.numeric(rf_engine$variable.importance),
    feature_set           = set_name # Add identifier for the master table
  ) %>%
    inner_join(feature_dictionary, by = join_by("feature_name_original"))
  
  # ==============================================================================
  # 5. EVALUATION METRICS & PREDICTIONS
  # ==============================================================================
  
  eval_table <- df_regression %>%
    select(abs_log2fc_diff) %>%
    bind_cols(
      stats::predict(lm_fit, new_data = df_regression) %>% rename(pred_linear = .pred),
      stats::predict(rf_fit, new_data = df_regression) %>% rename(pred_rf = .pred)
    ) %>%
    mutate(feature_set = set_name)
  
  metrics_lm <- metric_set(rmse, rsq)(eval_table, truth = abs_log2fc_diff, estimate = pred_linear) %>% 
    mutate(model = "Linear (LM)", feature_set = set_name)
  
  metrics_rf <- metric_set(rmse, rsq)(eval_table, truth = abs_log2fc_diff, estimate = pred_rf) %>% 
    mutate(model = "Random Forest", feature_set = set_name)
  
  current_metrics <- bind_rows(metrics_lm, metrics_rf)
  
  # ==============================================================================
  # 6. COMPILE AND SAVE LIGHTWEIGHT RDS
  # ==============================================================================
  
  # Bundle ONLY the extracted data frames (excluding the heavy model fits)
  iteration_results <- list(
    set_name     = set_name,
    metrics      = current_metrics,
    coefficients = lm_coefs,
    importance   = rf_imp,
    predictions  = eval_table
  )
  
  # Save the lightweight bundle as an RDS file
  rds_path <- here::here("Models", paste0("regression_extracted_data_", clean_name, ".rds"))
  saveRDS(iteration_results, file = rds_path)
  
  cat("  -> Saved extracted data to:", rds_path, "\n")
  
  return(iteration_results)
})

# ==============================================================================
# GENERATE MASTER TABLES
# ==============================================================================

# 1. Master Metrics Table (RMSE, R2 for all models and sets)
master_regression_metrics <- map_dfr(all_regression_results, "metrics") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

master_regression_metrics %>% 
  saveRDS(file = here("Models", "master_regression_metrics.rds"))

# 2. Master Coefficients Table (Linear Model terms across all sets)
master_regression_coefs <- map_dfr(all_regression_results, "coefficients") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

master_regression_coefs %>% 
  saveRDS(file = here("Models", "master_regression_coefs.rds"))

# 3. Master Variable Importance Table (Random Forest Gini across all sets)
master_regression_var_imp <- map_dfr(all_regression_results, "importance") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

master_regression_var_imp %>% 
  saveRDS(file = here("Models", "master_regression_var_imp.rds"))

# 4. Master Predictions Table (Observed vs Predicted data across all sets)
master_regression_predictions <- map_dfr(all_regression_results, "predictions") %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector))

master_regression_predictions %>% 
  saveRDS(file = here("Models", "master_regression_predictions.rds"))

# Display the summary metrics
print(master_regression_metrics %>% arrange(model, feature_set))

master_regression_metrics %>% 
  filter(.metric == "rsq") %>% 
  arrange(model)

```


```{r fig.height=12, fig.width=10}
# ==============================================================================
# 5. VISUALIZATIONS
# ==============================================================================
master_regression_metrics = 
  readRDS(file = here("Models", "master_regression_metrics.rds"))
master_regression_coefs =
  readRDS(file = here("Models", "master_regression_coefs.rds"))
master_regression_var_imp = 
  readRDS(file = here("Models", "master_regression_var_imp.rds"))
master_regression_predictions =
  readRDS(file = here("Models", "master_regression_predictions.rds"))
```


```{r fig.height=12, fig.width=10}
# 1. Generate wide summary table of R² values across models and feature sets
regression_rsq_table <- master_regression_metrics %>%
  filter(.metric == "rsq") %>%
  mutate(.estimate = round(.estimate, 3)) %>%
  select(feature_set, model, .estimate) %>%
  pivot_wider(
    names_from = model,
    values_from = .estimate
  ) %>% 
   mutate(feature_set = fct_relevel(feature_set, features_names_vector)) %>% 
  arrange(feature_set)

# Display the summary table
print(regression_rsq_table)

# 1. LM Forest Plot Data Preparation
linear_coefficients_df <- master_regression_coefs %>%
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  inner_join(
    master_regression_metrics %>% 
      filter(model == "Linear (LM)", .metric == "rsq") %>% 
      select(model, .estimate, feature_set),
    by = "feature_set"
  ) %>%
  mutate(
    feature_set_label = paste0(feature_set, "\n(R² = ", round(.estimate, 2), ")"),
    feature_set       = fct_relevel(feature_set, features_names_vector),
    feature_name = if_else(
      str_detect(feature_type, "PLS|pELS|dELS"),
      paste(feature_type, feature_name),
      feature_name
    )
  ) %>%
  arrange(feature_set) %>%
  mutate(feature_name = fct_rev(fct_inorder(feature_name)))

# 2. LM Forest Plot Visualization
linear_coefficients_forest_plot <- linear_coefficients_df %>%
  ggplot() +
  aes(
    x     = estimate, 
    y     = feature_set %>% fct_rev(), 
    xmin  = conf.low, 
    xmax  = conf.high, 
    fill  = feature_set, 
    group = feature_set
  ) +
  facet_grid(feature_name %>% fct_rev() ~ ., scales = "free_y", space = "free", switch = "y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.6) +
  geom_errorbarh(
    height      = 0.2, 
    linewidth   = 0.6, 
    width       = 0,
    alpha       = 0.5,
    position    = position_dodge(width = 1), 
    show.legend = FALSE
  ) +
  geom_point(size = 2.2, pch = 21, position = position_dodge(width = 1)) +
  scale_fill_manual(values = colors$feature_sets) +
  labs(
    x    = "Std. Effect Size",
    y    = "Feature",
    fill = "Feature Set"
  ) +
  theme_vaxgo() +
  theme(
    legend.title.position = "top", 
    legend.position   = "top", 
    legend.location   = "plot",
    strip.placement   = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1, vjust = 0.5),
    axis.text.y       = element_blank()
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(pch = 22),
      ncol         = 1
    )
  )

linear_coefficients_forest_plot

# 3. Random Forest Gini Importance Plot
# 1. RF Importance Data Preparation
rf_importance_df <- master_regression_var_imp %>%
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  inner_join(
    master_regression_metrics %>% 
      filter(model == "Random Forest", .metric == "rsq") %>% 
      select(model, .estimate, feature_set),
    by = "feature_set"
  ) %>%
  mutate(
    feature_set_label = paste0(feature_set, "\n(R² = ", round(.estimate, 2), ")"),
    feature_set       = fct_relevel(feature_set, features_names_vector),
    feature_name = if_else(
      str_detect(feature_type, "PLS|pELS|dELS"),
      paste(feature_type, feature_name),
      feature_name
    )
  ) %>%
  arrange(feature_set) %>%
  mutate(feature_name = fct_rev(fct_inorder(feature_name)))

# 2. RF Importance Visualization
p_rf_importance <- rf_importance_df %>% 
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  ggplot() +
  aes(
    x    = importance + 1, 
    y    = feature_set %>% fct_rev(), 
    fill = feature_set
  ) +
  facet_grid(feature_name %>% fct_rev() ~ ., scales = "free_y", space = "free", switch = "y") +
  geom_col(
    width    = 0.6,
    position = position_dodge(1)
  ) +
  scale_fill_manual(values = colors$feature_sets) +
  scale_x_log10(
    breaks = c(1, 10, 50, 100, 250, 1000),
    labels = c("0", "10", "50", "100", "250", "1000"),
    expand = c(0, 0)
  ) +
  labs(
    x    = "Importance",
    y    = "Feature",
    fill = "Feature Set"
  ) +
  theme_vaxgo() +
  theme(
    legend.title.position = "top", 
    legend.position   = "top", 
    legend.location   = "plot",
    legend.justification = "left",
    strip.placement   = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1, vjust = 0.5),
    axis.text.y       = element_blank()
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(pch = 22),
      nrow        = 2
    )
  )

p_rf_importance

# 4. Merge plots (LM vs RF)
coeffiecients_patch_plot = (linear_coefficients_forest_plot + 
                              theme(strip.text.y = element_blank(),
                                    legend.position = "none")) +
  (p_rf_importance + theme(strip.text.y.left = element_blank()))+
  plot_layout(widths = c(0.5, 0.5), 
              guides = "collect", axes = "collect") +
  plot_annotation(theme = theme(legend.position = "top", legend.location = "plot"))

coeffiecients_patch_plot

#### Plot Observed vs Predicted values across models ============

pobs_vs_pred_df = master_regression_predictions %>% 
  filter(feature_set != "+ Sequence + TFs + CRE type + CTCF + Homology") %>% 
  inner_join(regression_rsq_table,
             by = "feature_set") %>% 
  mutate(feature_set_label = paste0(feature_set, "\nLinear model: R² = ", `Linear (LM)`, ",\nRandom forest: R² = ", `Random Forest`)) %>% 
  select(feature_set, feature_set_label, abs_log2fc_diff, pred_linear, pred_rf) %>% 
  pivot_longer(cols = c(pred_linear, pred_rf),
               names_to = "algorithm",
               values_to = "predicted") %>% 
  mutate(algorithm = case_when(algorithm == "pred_linear" ~ "Linear model",
                               algorithm == "pred_rf" ~ "Random forest",
                               T ~ algorithm),
         feature_set = fct_relevel(feature_set,features_names_vector)
    ) %>% 
  arrange(feature_set) %>% 
  mutate(feature_set_label = fct_inorder(feature_set_label))

p_obs_vs_pred = pobs_vs_pred_df %>%
  ggplot(aes(x = abs_log2fc_diff, 
             y = predicted,
             color = algorithm)) +
  facet_wrap(~ feature_set_label, ncol = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.6) +
  geom_point(alpha = 0.1, size = 1, pch = 21, stroke = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1, show.legend = F) +
  scale_color_manual(values = c("Random forest" = "#4DBBD5FF",
                                "Linear model" = "gray70")) +
  guides(
    color = guide_legend(override.aes = list(alpha = 1,
                                             size = 4,
                                             pch = 21,
                                             fill = c("gray70", "#4DBBD5FF")),
                         nrow = 2))  +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  
  labs(
    x = "Observed |ΔLog2FC|",
    y = "Predicted |ΔLog2FC|",
    fill = "",
    color = ""
  ) +
  theme_vaxgo() +
  theme(
    legend.position = "top",
    plot.title.position = "plot", 
    legend.justification = "left",
    legend.title = element_blank(),
  )

#Merge plots
coeffiecients_pobsvspred_patch_plot = (coeffiecients_patch_plot | p_obs_vs_pred ) +
  plot_layout(widths = c(0.6, 0.4)) 

coeffiecients_pobsvspred_patch_plot

coeffiecients_pobsvspred_patch_plot %>% 
  ggsave(filename = here("Figures", "statistical_modelling_regression_coefficients_predicted_vs_observed_patch_plot_allmodes.png"),
         width = 10,
         height = 12)
```

## Combine plots for figures

```{r fig.height=12, fig.width=6}
performance_plots = p_auc_bars + p_obs_vs_pred

performance_plots

#Save plots
performance_plots %>% 
      ggsave(filename = here("Figures", "statistical_modelling_classification_regression_aucroc_obs_pred.png"),
         width = 6,
         height = 12)
```


```{r fig.height=12, fig.width=12}
features_plots = wrap_plots(
  class_patch_plot_1 + theme(legend.position = "bottom", legend.direction = "vertical"),
  coeffiecients_patch_plot + theme(legend.position = "bottom"),
  ncol = 2,
  widths = c(1, 1)
) + 
  plot_layout(
    guides = "collect",
    axes = "collect",
    axis_titles = "collect"
  ) &
  plot_annotation(theme = theme(legend.position = "bottom"),
                  title = "Features")

features_plots %>% 
  ggsave(filename = here("Figures", "statistical_modelling_classification_regression_features.png"),
         width = 12,
         height = 12)

```










# Create Scores

```{r}
#Import
#Multilayer table
gene_annotated_layers = 
  readRDS(file = here("tables", "human_mouse_statsmodelling_gene_annotated_layers.rds"))

#Gene expression + BTMs results
dge_btm_process_genes_diff_bygene_clean_filtered =  readRDS(file = here("tables", "dge_btm_process_genes_diff_bygene_clean_filtered.rds"))

#Classification
logit_coefs = readRDS(file = here::here("Models", "master_class_coefs.rds"))#Logistic coefficients
rf_importance_class <- readRDS(file = here::here("Models", "master_class_var_imp.rds"))#Random forest importance

#Regression 
lm_coefs = readRDS(file = here("Models", "master_regression_coefs.rds")) #Logistic coefficients
rf_importance_reg = readRDS(file = here("Models", "master_regression_var_imp.rds")) #Random forest importance

#Annotate with status by organism ==========
legs_notlegs_layers = dge_btm_process_genes_diff_bygene_clean_filtered %>% 
  filter(treatment %in% c("Infection", "Injury")) %>% 
  #Annotate with layers
  inner_join(gene_annotated_layers,
             by = "hgnc_symbol")  %>%
  filter(status != "All genes")

#Select model to use corresponding weights =========
select_model = "+ Sequence + TFs + CRE type + CTCF"

# Extract classification weights (Shared vs. Not-Shared)
weight_vec_class <- rf_importance_class %>% 
  filter(feature_set == select_model) %>% 
  select(feature_name, importance, feature_set) %>% 
  inner_join(logit_coefs, by = join_by(feature_set, feature_name)) %>% 
  mutate(
    raw_weight = sign(estimate) * importance,
    norm_weight = raw_weight / max(abs(raw_weight), na.rm = TRUE)
  ) %>% 
  select(feature_name, norm_weight) %>% 
  deframe()

# Extract regression weights
weight_vec_reg <- rf_importance_reg %>% 
  filter(feature_set == select_model) %>% 
  select(feature_name, importance, feature_set) %>% 
  inner_join(lm_coefs, by = join_by(feature_set, feature_name)) %>% 
  mutate(
    # Invert sign to penalize features increasing delta Log2FC
    raw_weight = sign(estimate) * importance,
    norm_weight = raw_weight / max(abs(raw_weight), na.rm = TRUE)
  ) %>% 
  select(feature_name, norm_weight) %>% 
  deframe()

# Lookup functions
get_w_class <- function(name) {
  w <- unname(weight_vec_class[name])
  if (is.null(w) || is.na(w)) return(0) else return(w)
}

get_w_reg <- function(name) {
  w <- unname(weight_vec_reg[name])
  if (is.null(w) || is.na(w)) return(0) else return(w)
}

# Robust Min-Max scaling function handling zero-variance groups
min_max <- function(x) {
  if (all(is.na(x))) return(rep(0, length(x)))
  rng <- max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
  if (is.na(rng) || rng == 0) return(rep(0, length(x)))
  (x - min(x, na.rm = TRUE)) / rng
}

# =========================================================================
# 1. Compute Dual Composite Scores (Pathogen-Level Min-Max)
# =========================================================================
legs_notlegs_layers_score <- legs_notlegs_layers %>% 
  group_by(treatment, pathogen, timepoint_comparison) %>% 
  mutate(
    # Min-Max Normalization per pathogen (Preserves global variance contrast)
    f_fc            = min_max(abs(mean_log2fc_mouse)),
    f_inv_se        = min_max(log1p(inverse_se_mouse)),
    f_kimura        = min_max(dist_k80),
    f_aa_id         = min_max(identity_human2mouse),
    f_tf_total      = min_max(n_tf_total),
    f_tf_shared     = min_max(pct_tf_shared),
    f_dels_ncres    = min_max(n_type_dELS),
    f_dels_match    = min_max(`pct_match_type_dELS`),
    f_dels_ctcf     = min_max(`pct_match_ctcf_dELS_CTCF-bound`),
    f_pels_ncres    = min_max(n_type_pELS),
    f_pels_match    = min_max(`pct_match_type_pELS`),
    f_pels_ctcf     = min_max(`pct_match_ctcf_pELS_CTCF-bound`),
    f_pls_ncres     = min_max(n_type_PLS),
    f_pls_match     = min_max(`pct_match_type_PLS`),
    f_pls_ctcf      = min_max(`pct_match_ctcf_PLS_CTCF-bound`),
    
    # Classification Raw Score
    score_raw_class = 
      (f_fc           * get_w_class("Mean Log2FC (Mouse)")) + 
      (f_inv_se       * get_w_class("1/SE")) +
      (f_kimura       * get_w_class("Kimura distance")) + 
      (f_aa_id        * get_w_class("%Protein AA identity")) +
      (f_tf_total     * get_w_class("nTotal TFs")) + 
      (f_tf_shared    * get_w_class("%Shared TFs")) +
      (f_dels_ncres   * get_w_class("dELS nCREs")) + 
      (f_dels_match   * get_w_class("dELS %Match Type")) +
      (f_dels_ctcf    * get_w_class("dELS CTCF %Match CTCF")) + 
      (f_pels_ncres   * get_w_class("pELS nCREs")) +
      (f_pels_match   * get_w_class("pELS %Match Type")) + 
      (f_pels_ctcf    * get_w_class("pELS CTCF %Match CTCF")) +
      (f_pls_ncres    * get_w_class("PLS nCREs")) + 
      (f_pls_match    * get_w_class("PLS %Match Type")) +
      (f_pls_ctcf     * get_w_class("PLS CTCF %Match CTCF")),
      
    # Regression Raw Score
    score_raw_reg = 
      (f_fc           * get_w_reg("Mean Log2FC (Mouse)")) + 
      (f_inv_se       * get_w_reg("1/SE")) +
      (f_kimura       * get_w_reg("Kimura distance")) + 
      (f_aa_id        * get_w_reg("%Protein AA identity")) +
      (f_tf_total     * get_w_reg("nTotal TFs")) + 
      (f_tf_shared    * get_w_reg("%Shared TFs")) +
      (f_dels_ncres   * get_w_reg("dELS nCREs")) + 
      (f_dels_match   * get_w_reg("dELS %Match Type")) +
      (f_dels_ctcf    * get_w_reg("dELS CTCF %Match CTCF")) + 
      (f_pels_ncres   * get_w_reg("pELS nCREs")) +
      (f_pels_match   * get_w_reg("pELS %Match Type")) + 
      (f_pels_ctcf    * get_w_reg("pELS CTCF %Match CTCF")) +
      (f_pls_ncres    * get_w_reg("PLS nCREs")) + 
      (f_pls_match    * get_w_reg("PLS %Match Type")) +
      (f_pls_ctcf     * get_w_reg("PLS CTCF %Match CTCF")),
    
    # Rescale Composite Scores (0-100) and Pathogen Rank
    score_shared = min_max(score_raw_class) * 100,
    score_reg = (1 - min_max(score_raw_reg)) * 100,
    rank_log2fc  = percent_rank(abs(mean_log2fc_mouse) * -log10(pmax(p_adj_mouse, 1e-300))) * 100
  ) %>% 
  select(-starts_with("f_"), -starts_with("score_raw_")) %>% 
  ungroup() %>% 
  
  # Process-level Rank calculated separately
  group_by(treatment, pathogen, process) %>% 
  mutate(rank_log2fc_process = percent_rank(abs(mean_log2fc_mouse) * -log10(pmax(p_adj_mouse, 1e-300))) * 100) %>% 
  ungroup()

legs_notlegs_layers_score_plot <- legs_notlegs_layers_score %>% 
  bind_rows(
    legs_notlegs_layers_score %>% 
      mutate(treatment = "Pooled", pathogen = "", timepoint_comparison = "")
  )

# =========================================================================
# 2. Dual Evaluation Pipeline (Statistical Summary Table)
# =========================================================================
calc_dual_metrics <- function(df) {
  df %>% 
    nest() %>% 
    mutate(
      # --- CLASSIFICATION METRICS (score_shared) ---
      auc_roc = map_dbl(data, function(d) {
        if (!"score_shared" %in% names(d) || !"status_original" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(score_shared), !is.na(status_original))
        if (length(unique(d_clean$status_original == "Shared")) < 2) return(NA_real_)
        tryCatch({
          as.numeric(auc(roc(response = d_clean$status_original == "Shared", 
                             predictor = d_clean$score_shared, quiet = TRUE)))
        }, error = function(e) NA_real_)
      }),
      
      p_val_roc = map_dbl(data, function(d) {
        if (!"score_shared" %in% names(d) || !"status_original" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(score_shared), !is.na(status_original))
        if (length(unique(d_clean$status_original == "Shared")) < 2) return(NA_real_)
        tryCatch({
          wilcox.test(score_shared ~ (status_original == "Shared"), data = d_clean)$p.value
        }, error = function(e) NA_real_)
      }),
      
      auc_pr = map_dbl(data, function(d) {
        if (!"score_shared" %in% names(d) || !"status_original" %in% names(d)) return(NA_real_)
        d_clean <- d %>% filter(!is.na(score_shared), !is.na(status_original))
        is_sh <- d_clean$status_original == "Shared"
        if (sum(is_sh) == 0 || sum(!is_sh) == 0) return(NA_real_)
        tryCatch({
          pr.curve(
            scores.class0 = d_clean$score_shared[is_sh],
            scores.class1 = d_clean$score_shared[!is_sh]
          )$auc.integral
        }, error = function(e) NA_real_)
      }),
      
      # --- REGRESSION METRICS (score_reg vs. |mean_log2fc_mouse|) ---
      spearman_rho = map_dbl(data, function(d) {
        if (!"score_reg" %in% names(d) || !"mean_log2fc_mouse" %in% names(d)) return(NA_real_)
        tryCatch({
          cor(d$score_reg, abs(d$mean_log2fc_mouse), method = "spearman", use = "complete.obs")
        }, error = function(e) NA_real_)
      }),
      
      p_val_spearman = map_dbl(data, function(d) {
        if (!"score_reg" %in% names(d) || !"mean_log2fc_mouse" %in% names(d)) return(NA_real_)
        tryCatch({
          cor.test(d$score_reg, abs(d$mean_log2fc_mouse), method = "spearman", exact = FALSE)$p.value
        }, error = function(e) NA_real_)
      }),
      
      pearson_r = map_dbl(data, function(d) {
        if (!"score_reg" %in% names(d) || !"mean_log2fc_mouse" %in% names(d)) return(NA_real_)
        tryCatch({
          cor(d$score_reg, abs(d$mean_log2fc_mouse), method = "pearson", use = "complete.obs")
        }, error = function(e) NA_real_)
      }),
      
      r_squared = map_dbl(data, function(d) {
        if (!"score_reg" %in% names(d) || !"mean_log2fc_mouse" %in% names(d)) return(NA_real_)
        tryCatch({
          fit <- lm(abs(mean_log2fc_mouse) ~ score_reg, data = d)
          summary(fit)$r.squared
        }, error = function(e) NA_real_)
      }),
      
      p_val_lm = map_dbl(data, function(d) {
        if (!"score_reg" %in% names(d) || !"mean_log2fc_mouse" %in% names(d)) return(NA_real_)
        tryCatch({
          fit <- lm(abs(mean_log2fc_mouse) ~ score_reg, data = d)
          summary(fit)$coefficients[2, 4]
        }, error = function(e) NA_real_)
      })
    ) %>% 
    select(-data)
}

# A. Evaluation per Process / Condition (Pathogen-specific)
eval_condition_process <- calc_dual_metrics(
  legs_notlegs_layers_score_plot %>% 
    filter(!is.na(score_shared), !is.na(score_reg), !is.na(status_original)) %>% 
    group_by(treatment, pathogen, process)
)

# B. Evaluation per Process Across Pooled Conditions (Pooled Process Level)
eval_process_pooled <- calc_dual_metrics(
  legs_notlegs_layers_score_plot %>%
    filter(!is.na(score_shared), !is.na(score_reg), !is.na(status_original)) %>%
    group_by(treatment, pathogen)
) %>%
  mutate( process = "All BTMs", group = "All BTMs")

# C. Overall Global Evaluation (Pooled Overall)
eval_pooled_class <- calc_dual_metrics(
  legs_notlegs_layers_score_plot %>%
    filter(!is.na(score_shared), !is.na(status_original)) %>% 
    select(-c(process, group)) %>% 
    group_by(treatment, pathogen)
) %>% 
  select(treatment, pathogen, auc_roc, p_val_roc, auc_pr)

eval_pooled_reg <- calc_dual_metrics(
  legs_notlegs_layers_score_plot %>%
    filter(!is.na(score_reg)) %>% 
    select(treatment, pathogen, hgnc_symbol, score_reg, mean_log2fc_mouse) %>% 
    distinct() %>% 
    group_by(treatment, pathogen)
) %>% 
  select(treatment, pathogen, spearman_rho, p_val_spearman, pearson_r, r_squared, p_val_lm)

eval_pooled <- inner_join(eval_pooled_class, 
                          eval_pooled_reg, by = c("treatment", "pathogen")) %>% 
  mutate(process = "All genes", group = "All genes")

# Consolidation & FDR Correction
eval_metrics_all_summary <- bind_rows(
  eval_pooled %>% mutate(analysis_set = "Pooled"),
  eval_process_pooled %>% mutate(analysis_set = "Process_Pooled"),
  eval_condition_process %>% mutate(analysis_set = "Process")
) %>% 
  group_by(analysis_set) %>% 
  mutate(
    p_adj_roc      = p.adjust(p_val_roc, method = "BH"),
    p_adj_spearman = p.adjust(p_val_spearman, method = "BH"),
    p_adj_lm       = p.adjust(p_val_lm, method = "BH")
  ) %>% 
  ungroup() %>% 
  select(
    analysis_set, treatment, pathogen, process, group,
    auc_roc, auc_pr, p_val_roc, p_adj_roc,
    spearman_rho, p_val_spearman, p_adj_spearman,
    pearson_r, r_squared, p_val_lm, p_adj_lm
  )

eval_metrics_all_summary
```



## Visualize by process
### Classification
Selected process
```{r fig.height=4, fig.width=12}
# Append Pooled rows cleanly for plotting purposes only
# legs_notlegs_layers_score_plot <- legs_notlegs_layers_score %>% 
#   bind_rows(
#     legs_notlegs_layers_score %>% 
#       mutate(treatment = "Pooled", pathogen = "")
#   )

# Select process for visualization
select_process <- "immune activation - generic cluster"
select_process <- "cell cycle and transcription"

legs_notlegs_layers_visualized <- legs_notlegs_layers_score_plot %>% 
  filter(process == select_process) %>% 
  inner_join(
    eval_metrics_all_summary %>% 
      filter(analysis_set %in% c("Process")),
    by = join_by(treatment, pathogen, process)
  ) %>% 
  ggplot(aes(x = score_shared, y = rank_log2fc_process, fill = status_original)) +
  
  # Nested faceting displaying conditions and ROC/PR performance
  facet_nested_wrap(
    ~ process + treatment + pathogen + 
      paste0("ROC-AUC=", round(auc_roc, 2), ",\nPR-AUC=", round(auc_pr, 2)),
    nrow = 1, 
    strip = strip_nested(size = "variable")
  ) +
  
  # Layered scatter plot highlighting Shared vs. Human-only candidates
  geom_point(
    data = . %>% filter(status_original == "Human only"),
    aes(size = score_shared * abs(mean_log2fc_mouse)),
    shape = 21, alpha = 0.2
  ) +
  geom_point(
    data = . %>% filter(status_original != "Human only"),
    aes(size = score_shared * abs(mean_log2fc_mouse)),
    shape = 21, alpha = 1
  ) +

  scale_fill_manual(values = colors$comparison) +
  theme_vaxgo() +
  labs(
    x = "Shared LEG Composite Score",
    y = "Log2FC Rank Percentile (Mouse)",
    fill = "LEG Status",
    size = "|Log2FC|*Score",
    title = "Shared LEG Score vs. Expression Rank"
  ) +
  
  # Right Marginals (X-axis position = 110)
  geom_violin(
    aes(y = rank_log2fc_process, x = 110, fill = status_original),
    width = 15, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(y = rank_log2fc_process, x = 110),
    width = 15, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  # Top Marginals (Y-axis position = 110)
  geom_violin(
    aes(x = score_shared, y = 110, fill = status_original),
    width = 15, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(x = score_shared, y = 110),
    width = 15, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  theme(legend.position = "top", legend.direction = "horizontal")

# Display Plot
legs_notlegs_layers_visualized

#Save
legs_notlegs_layers_visualized %>%
  ggsave(filename = here("Figures", paste0("statistical_modelling_classification_scores_visualized_", select_process, ".png")),
         width = 12,
         height = 4)
```



```{r fig.height=3, fig.width=12}
legs_notlegs_layers_visualized_allprocesses = legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, rank_log2fc, score_shared, p_adj_mouse, status, status_original) %>% 
  distinct() %>% 
  inner_join(
    eval_metrics_all_summary %>% 
      filter(analysis_set %in% c("Process_Pooled")),
    by = join_by(treatment, pathogen)
  ) %>% 
  ggplot(aes(x = score_shared, y = rank_log2fc, fill = status_original)) +
  
  # Nested faceting displaying conditions and ROC/PR performance
  facet_nested_wrap(
    ~ process + treatment + pathogen + 
      paste0("ROC-AUC=", round(auc_roc, 2), ",\nPR-AUC=", round(auc_pr, 2)),
    nrow = 1, 
    strip = strip_nested(size = "variable")
  ) +
  
  # Layered scatter plot highlighting Shared vs. Human-only candidates
  geom_point(
    data = . %>% filter(status_original != "Shared"),
    shape = 21, alpha = 0.1, stroke = 0, size = 2
  ) +
  geom_point(
    data = . %>% filter(status_original == "Shared"),
    shape = 21, alpha = 0.8, stroke = 0.2, size = 2,
  ) +

  scale_fill_manual(values = colors$comparison) +
  theme_vaxgo() +
  labs(
    x = "Shared LEG Composite Score",
    y = "Rank, Mouse",
    fill = "LEG Status",
  ) +
  
  # Right Marginals (X-axis position = 110)
  geom_violin(
    aes(y = rank_log2fc, x = 110, fill = status_original),
    width = 15, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(y = rank_log2fc, x = 110),
    width = 15, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  # Top Marginals (Y-axis position = 110)
  geom_violin(
    aes(x = score_shared, y = 110, fill = status_original),
    width = 15, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(x = score_shared, y = 110),
    width = 15, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  theme(legend.position = "top", legend.direction = "horizontal")

# Display Plot
legs_notlegs_layers_visualized_allprocesses

# #Save
legs_notlegs_layers_visualized_allprocesses %>%
  ggsave(filename = here("Figures", paste0("statistical_modelling_classification_scores_visualized_allprocesses.png")),
         width = 12,
         height = 3)
```





Pooled, all processes
```{r fig.height=4, fig.width=12}

reg_layers_visualized_all <- legs_notlegs_layers_score %>% 
  inner_join(
    eval_metrics_all_summary %>% 
      filter(analysis_set == "Pooled"),
    by = join_by(treatment, pathogen, process)
  ) %>% 
  ggplot(aes(x = score_reg, y = mean_log2fc_mouse, fill = status_original)) +
  
    # Facet informando métricas de regressão (R² e Spearman)
  facet_nested_wrap(
    ~ process + treatment + pathogen + 
      paste0("R²=", round(r_squared, 2), " (p=", sprintf("%.2e", p_adj_lm), ")\n",
             "ρ=", round(spearman_rho, 2), " (p=", sprintf("%.2e", p_adj_spearman), ")"),
    nrow = 1, 
    strip = strip_nested(size = "variable")
  ) +
  
  
  # Scatter plot com camadas para destacar Shared vs Human Only
  geom_point(
    data = . %>% filter(status_original == "Human only"),
    aes(size = score_reg * abs(mean_log2fc_mouse)),
    shape = 21, alpha = 0.2
  ) +
  geom_point(
    data = . %>% filter(status_original != "Human only"),
    aes(size = score_reg * abs(mean_log2fc_mouse)),
    shape = 21, alpha = 0.8
  ) +

  scale_fill_manual(values = colors$comparison) +
  theme_vaxgo() +
  labs(
    x = "Shared LEG Regression Score (0-100)",
    y = "Log2 Fold Change (Mouse)",
    fill = "LEG Status",
    size = "Score × |Log2FC|",
    title = "Shared LEG Regression Score vs. Mouse Expression Signal"
  ) +
  scale_y_continuous(limits = c(0, 10)) +
  # Marginais Direita (Eixo Y: mean_log2fc_mouse, Posição X = 110)
  geom_violin(
    aes(y = mean_log2fc_mouse, x = 110, fill = status_original),
    width = 20, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(y = mean_log2fc_mouse, x = 110),
    width = 20, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  # Marginais Superior (Eixo X: score_reg, Posição Y = max(mean_log2fc_mouse)*1.1)
  geom_violin(
    aes(x = score_reg, y = 8, fill = status_original),
    width = 1, alpha = 0.5, color = NA
  ) +
  geom_boxplot(
    aes(x = score_reg, y = 8),
    width = 1, alpha = 0, color = "black", outlier.shape = NA
  ) +
  
  theme(legend.position = "top", legend.direction = "horizontal")

# Display
reg_layers_visualized

#Save
# legs_notlegs_layers_visualized %>% 
#   ggsave(filename = here("Figures", paste0("statistical_modelling_classification_scores_visualized_", select_process, ".png")),
#          width = 12,
#          height = 4)


```



By group
```{r fig.height=6, fig.width=16}
# Select process to get LEGs
legs_notlegs_layers_visualized_pooled_group_all = legs_notlegs_layers_score %>%
  filter(treatment == "Pooled") %>% 
  inner_join(eval_metrics_all_summary %>% 
               filter(analysis_set == "Group"),
             by = join_by(treatment, pathogen, group)) %>% 
  ggplot(aes(x = score_shared,
             y = rank_log2fc,
             fill = status_original,
             color = status_original)) +
  facet_nested_wrap(~treatment~paste0(group, ",\nROC-AUC=", round(auc_roc, 2), ", PR-AUC=", round(auc_pr, 2)),
             nrow = 2,
             strip = strip_nested(size = "variable"))+
  geom_point(alpha = 0.1, show.legend = F, stroke = 0) +
  geom_smooth(method = "lm", show.legend = F) +
  # Right
  geom_violin(mapping = aes(y = rank_log2fc,
                             x = 110,
                             fill = status_original),
               width = 20,
               alpha = 0.5, color = NA) +
  geom_boxplot(mapping = aes(x = 110,
                             y = rank_log2fc),
               width = 20,
               alpha = 0,
               color = "black",
               show.legend = F) +
  # Top
  geom_violin(mapping = aes(y = 110,
                             x = score_shared,
                             fill = status_original),
               width = 20,
               alpha = 0.5, color = NA) +
  geom_boxplot(mapping = aes(y = 110,
                             x = score_shared),
               width = 20,
               alpha = 0,
               color = "black",
               show.legend = F) +
  theme(legend.position = "top", legend.direction = "horizontal") +
  scale_color_manual(values = colors$comparison) +
  scale_fill_manual(values = colors$comparison) +
  theme_vaxgo() +
  theme(legend.position = "top") +
  labs(
    x = "Shared LEG Composite Score",
    y = "Log2FC Rank (Mouse)",
    fill = "LEG Status",
    size = "|Log2FC| (Mouse)",
    title = "Shared LEG Score vs. Expression Rank"
  ) 
 
legs_notlegs_layers_visualized_pooled_group_all 
  

```


```{r}
library(dplyr)
library(ggplot2)

# 1. Define os quadrantes e calcula as proporções
quadrant_summary <- legs_notlegs_layers_score %>%
  filter(treatment == "Pooled") %>% 
  mutate(
    quadrant_x = if_else(score_shared >= 50, "High Score (>=50)", "Low Score (<50)"),
    quadrant_y = if_else(rank_log2fc >= 50, "High Rank (>=50)", "Low Rank (<50)"),
    
    # Define a ordem dos quadrantes para exibição correta na matriz
    quadrant_label = case_when(
      score_shared >= 50 & rank_log2fc >= 50 ~ "Q1: Top Candidates\n(High Rank / High Score)",
      score_shared < 50  & rank_log2fc >= 50 ~ "Q2 (High Rank / Low Score)",
      score_shared < 50  & rank_log2fc < 50  ~ "Q3 (Low Rank / Low Score)",
      score_shared >= 50 & rank_log2fc < 50  ~ "Q4 (Low Rank / High Score)"
    )
  ) %>% 
  group_by(quadrant_x, quadrant_y, quadrant_label, status_original) %>% 
  summarise(count = n(), .groups = "drop_last") %>% 
  mutate(
    total_in_quadrant = sum(count),
    pct_in_quadrant = (count / total_in_quadrant) * 100
  ) %>% 
  ungroup()

quadrant_summary %>% 
  drop_na(quadrant_y) %>% 
  ggplot(aes(x = count, y = quadrant_label, fill = status_original)) +
  geom_col(position = "fill", width = 0.6, color = "black") +
  scale_x_continuous(labels = scales::percent) +
  scale_fill_manual(values = colors$comparison) +
  theme_vaxgo() +
  theme(
    # axis.text.x = element_text(angle = 15, hjust = 1),
        legend.position = "top") +
  labs(
    x = "Proportion of Genes",
    y = "Quadrant",
    fill = "LEG Status",
    title = "Class Distribution Across Score/Rank Quadrants"
  )

```

### Regression


```{r fig.height=3, fig.width=12}
######## Mouse Log2FC vs AbsDiff
reg_layers_mouse_human_absdiff <-  
  legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, rank_log2fc, score_reg, p_adj_mouse) %>% 
  distinct() %>% 
    filter(p_adj_mouse <= 0.05) %>% 
 inner_join(
    eval_metrics_all_summary %>%
      filter(analysis_set == "Pooled"),
    by = join_by(treatment, pathogen)
  ) %>%
  distinct() %>% 
  ggplot(aes(x = score_reg, y = abs_log2fc_diff, 
             # fill = status_original,
             fill = score_reg,
             )) +
  
  # Facet display showing regression metrics (R² and Spearman ρ) with formatted p-values
  facet_nested_wrap(
    ~ treatment + pathogen + 
      paste0("ρ=", round(spearman_rho, 2), " (p=", sprintf("%.2e", p_adj_spearman), ")"),
    nrow = 1, 
    strip = strip_nested(size = "variable")
  ) +
  
  # Layered scatter plot highlighting Shared vs. Human Only candidates
  geom_point(
    # aes(size = score_reg * abs(abs_log2fc_diff)),
    aes(size = score_reg),
    shape = 21, 
    size = 2,
    alpha = 1,
    # color = NA,
    stroke = 0.5
  ) +

  scale_fill_continuous(palette = c("white", "#4cc9f0", "#4361ee")) +
  theme_vaxgo() +
  labs(
    x = "Score",
    y = "|ΔLog2fc|",
    fill = "Score",
    size = "Score",
    title = "Shared LEG Regression Score vs. |ΔLog2fc|"
  ) +
  theme(legend.position = "top", legend.direction = "horizontal")

# Display
reg_layers_mouse_human_absdiff

#Save
reg_layers_mouse_human_absdiff %>%
  ggsave(filename = here("Figures", paste0("statistical_modelling_regression_scores_visualized_vsAbsLog2FC.png")),
         width = 12,
         height = 3)
```



```{r fig.height=3, fig.width=12}
######## Mouse Log2FC vs Score
reg_layers_mouse_human <- legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, rank_log2fc, score_reg,p_adj_mouse) %>%
  filter(p_adj_mouse <= 0.05) %>% 
  distinct() %>% 
 inner_join(
    eval_metrics_all_summary %>%
      filter(analysis_set == "Pooled"),
    by = join_by(treatment, pathogen)
  ) %>%
  distinct() %>% 
  ggplot(aes(x = mean_log2fc_mouse, 
             y = mean_log2fc_human, 
             # fill = status_original,
             fill = score_reg,
             )) +
  
  # Facet display showing regression metrics (R² and Spearman ρ) with formatted p-values
  facet_nested_wrap(
    ~ treatment + pathogen + 
      paste0("ρ=", round(spearman_rho, 2), " (p=", sprintf("%.2e", p_adj_spearman), ")"),
    nrow = 1, 
    scales = "free",
    strip = strip_nested(size = "variable")
  ) +
  
  # Layered scatter plot highlighting Shared vs. Human Only candidates
  geom_point(
    aes(
        # size = score_reg,
        alpha = score_reg),
    # aes(size = score_reg),
    shape = 21, 
    # alpha = 0.5, 
    stroke = 0.5,
  ) +

  scale_fill_continuous(palette = c("white", "#4cc9f0", "#4361ee")) +
  theme_vaxgo() +
  labs(
    x = "Log2FC (Mouse)",
    y = "Log2FC (Human)",
    size = "Score",
    fill = "Score"
  ) +
  theme(legend.position = "top", legend.direction = "horizontal") +
  geom_vline(xintercept = 0, color = "black", linetype = 2) +
  geom_hline(yintercept = 0, color = "black", linetype = 2) 

# Display
reg_layers_mouse_human 

#Save
reg_layers_mouse_human %>%
  ggsave(filename = here("Figures", paste0("statistical_modelling_regression_scores_visualized_L2fcMousevsLog2FCHuman.png")),
         width = 12,
         height = 3)
```



```{r fig.height=3, fig.width=12}
######## Mouse Log2FC vs 1/SE
reg_layers_mouse_human_se <- legs_notlegs_layers_score_plot %>% 
  select(treatment, pathogen, hgnc_symbol, mean_log2fc_human, mean_log2fc_mouse, inverse_se_mouse, abs_log2fc_diff, rank_log2fc, score_reg, p_adj_mouse) %>% 
  filter(p_adj_mouse <= 0.05) %>%
  distinct() %>% 
 inner_join(
    eval_metrics_all_summary %>%
      filter(analysis_set == "Pooled"),
    by = join_by(treatment, pathogen)
  ) %>%
  distinct() %>% 
  ggplot(aes(x = inverse_se_mouse, 
             y = abs_log2fc_diff, 
             fill = score_reg,
             )) +
  
  # Facet display showing regression metrics (R² and Spearman ρ) with formatted p-values
  facet_nested_wrap(
    ~ treatment + pathogen + 
      paste0("ρ=", round(spearman_rho, 2), " (p=", sprintf("%.2e", p_adj_spearman), ")"),
    nrow = 1, scales = "free_x",
    strip = strip_nested(size = "variable")
  ) +
  
  # Layered scatter plot highlighting Shared vs. Human Only candidates
  geom_point(
    # aes(size = score_reg * abs(abs_log2fc_diff)),
    aes(size = score_reg),
    shape = 21, 
    size = 2,
    alpha = 1,
    # color = NA,
    stroke = 0.5
  ) +
  
  # Top Marginals (Y-axis position = 110)
  geom_violin(
    aes(x = inverse_se_mouse, y = 6, group = pathogen),
    width = 0.5, alpha = 0.5, color = NA,
    fill = "#4361ee"
  ) +
  geom_boxplot(
    aes(x = inverse_se_mouse, y = 6, group = pathogen),
    width = 0.5, alpha = 0, color = "black", outlier.shape = NA
  ) +

  scale_fill_continuous(palette = c("white", "#4cc9f0", "#4361ee")) +
  theme_vaxgo() +
  labs(
    x = "1/SE mouse",
    y = "|ΔLog2fc|"
  ) +
  theme(legend.position = "top", legend.direction = "horizontal")

# Display
reg_layers_mouse_human_se 
  

#Save
reg_layers_mouse_human_se %>%
  ggsave(filename = here("Figures", paste0("statistical_modelling_regression_scores_visualized_AbsLog2FC_InverseSE.png")),
         width = 12,
         height = 3)
```



















