---
title: "E.coli infection"
author: "Wasim Aluísio Prates-Syed"
date: "2025-07-23"
output: 
  html_document: 
    toc: TRUE
    toc_depth: 4
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE, cache = TRUE)
```

# Exploring other datasets: A general analysis

## Import required files

```{r}
# install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))

#Import data
fit_training_datasets_metadata_clean = read_rds(file = here("FIT_Mouse2man", "fit_training_datasets_metadata_clean.rds"))
AllData_V2.0_blood = read_rds(here("new_models", "AllData_FIT_training_blood.rds"))
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))

```

# E. coli infection

### GSE33341

```{r}
#E. coli: GSE33341 -----
filename = "GSE33341"
GSE33341 <- getGEO("GSE33341", GSEMatrix = TRUE) 
eset_mouse <- GSE33341[[1]]
eset_human <- GSE33341[[2]]

#Metadata ------
#Mouse
GSE33341_ecoli_mouse_metadata = eset_mouse %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    #Host
    subject_title = "title",
    subject_num = "replicate number:ch1",
    batch = "experimental batch:ch1",
    host_strain = "host strain:ch1",
    organism = "organism_ch1",
    age = "host age (yr):ch1",
    sex = "host gender:ch1", 
    #Pathogen
    pathogen = "pathogen:ch1",
    status = "infection status:ch1",
    site_infection = "anatomic site of infection:ch1",
    route_infection = "inoculum route:ch1",
    dosage = "pathogen   dosage cfu/g:ch1",
    timepoint = "time after infection (h):ch1",
    bac_strain = "bacterial strain:ch1"
    ) %>% 
  filter(pathogen %in% c("Escherichia  coli", "-")) %>%
  mutate(pathogen = if_else(pathogen == "Escherichia  coli", "escherichia coli", "none"),
         subject_id = paste0("Mouse_", host_strain, "_Animal", subject_num, "_Batch", batch),
         status = str_replace(status, "-", "Healthy"),
         condition = paste("Mouse", host_strain, timepoint, sep = "_"),
         comparison = paste0("Mouse_", host_strain, "_Batch", batch)) %>% 
  select(geo_sample, comparison, condition, subject_id, everything())

GSE33341_ecoli_mouse_metadata
```

```{r}
#Human
GSE33341_ecoli_human_metadata = eset_human %>% 
  pData() %>% 
  filter(`pathogen:ch1` %in% c("Escherichia coli", "-")) %>%
  dplyr::select( 
    geo_sample = "geo_accession", 
    #Host
    subject_title = "title",
    # subject_num = "replicate number:ch1",
    batch = "experimental batch:ch1",
    # host_strain = "host strain:ch1",
    organism = "organism_ch1",
    age = "host age (yr):ch1",
    sex = "host gender:ch1", 
    ethnicity = "ethnicity:ch1",
    #Pathogen
    pathogen = "pathogen:ch1",
    # status = "infection status:ch1",
    site_infection = "anatomic site of infection:ch1"
    # route_infection = "inoculum route:ch1",
    # dosage = "pathogen   dosage cfu/g:ch1",
    # timepoint = "time after infection (h):ch1",
    # bac_strain = "bacterial strain:ch1"
    ) %>% 
  mutate(site_infection = str_to_lower(site_infection),
         organism = "Human",
         pathogen = if_else(pathogen == "-", "Healthy", pathogen),
         status = if_else(pathogen == "Healthy", pathogen, "infected"),
         timepoint = "0",
         subject_id = if_else(pathogen == "Escherichia coli", 
                              paste0("Ecoli_human_", site_infection,"_",geo_sample), 
                              paste0("Healthy_human_", geo_sample)),
         site_infection = if_else(site_infection == "-", "none", site_infection),
         condition = subject_id) %>% 
  select(geo_sample, condition, subject_id, everything())

GSE33341_ecoli_human_metadata

#Unite metadata
GSE33341_ecoli_mouse_human_metadata = GSE33341_ecoli_human_metadata %>% 
  bind_rows(GSE33341_ecoli_mouse_metadata)

GSE33341_ecoli_mouse_human_metadata %>% 
  saveRDS(file = here("tables", "GSE33341_ecoli_mouse_human_metadata.rds"))
```

```{r}
#Import
GSE33341_ecoli_mouse_human_metadata =
  readRDS(file = here("tables", "GSE33341_ecoli_mouse_human_metadata.rds"))

#Get expression data -----
#Mouse
GSE33341_mouse_eset <- exprs(eset_mouse) %>% 
  as.data.frame() %>% 
  select(GSE33341_ecoli_mouse_human_metadata %>% 
           filter(organism == "Mus musculus") %>% 
           pull(geo_sample)) %>% 
  rownames_to_column("probe_id") 

#Human
GSE33341_human_eset <- exprs(eset_human) %>% 
  as.data.frame() %>% 
  select(GSE33341_ecoli_mouse_human_metadata %>% 
           filter(organism == "Human") %>% 
           pull(geo_sample)) %>% 
  rownames_to_column("probe_id")
```

#### Mouse

##### Annotate genes

This procedure was done using only one batch (8)

```{r}
#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
GSE33341_mouse_expression_metadata  = GSE33341_mouse_eset %>% 
  pivot_longer(cols = 2:125,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(GSE33341_ecoli_mouse_human_metadata, 
             by = "geo_sample") 

#Get value from day 0 and calculate Log2FC
GSE33341_log2fc_sample = GSE33341_mouse_expression_metadata %>% 
  #Filter batch
  filter(organism == "Mus musculus",
         status == "Healthy",
         host_strain == "C57BL/6J",
         batch == "8") %>% 
  select(probe_id, comparison, batch, subject_num, day0_value_raw = value_raw) %>% 
  #Get the mean expression value for healthy
  group_by(probe_id, comparison, batch) %>% 
  summarize(n = n(),
            #Mean of healthy
            day0_value_raw = sum(day0_value_raw)/n,
            batch) %>% 
  select(probe_id, day0_value_raw) %>% 
  #Filter batch
  inner_join(GSE33341_mouse_expression_metadata %>% 
               filter(timepoint != "0",
                      host_strain == "C57BL/6J",
                      batch == "8"),
             by = join_by("probe_id", "comparison", "batch")) %>% 
  mutate(log2fc = value_raw - day0_value_raw) %>% 
  distinct() 

#Check duplicates
GSE33341_log2fc_sample %>% 
  filter(subject_id == "Mouse_C57BL/6J_Animal1_Batch8",
         probe_id == "1415670_at")

```

```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- GSE33341_log2fc_sample$probe_id %>% unique()
# listAttributes(ensembl_mart)
#Annotate
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2",
  values = gene_id,
  mart = ensembl_mart)

#Gene lookup table
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  rename(human_symbol = HGNC_symbol,
         mouse_symbol =MGI.symbol,
         HS.Entrez = Human.Entrez,
         MM.Entrez = Mouse.Ortholog) %>% 
  mutate(HS.Entrez = as.character(HS.Entrez),
         MM.Entrez = as.character(MM.Entrez))

#Filter probe_ids with highest median expression for the same entrezgene_id
GSE33341_mouse_counts_log2_summary_annotated_probes = GSE33341_log2fc_sample %>% 
  group_by(probe_id) %>% 
  summarize(mean = mean(value_raw, na.rm = TRUE),
            median = median(value_raw, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe_id = affy_mouse430_2), 
            by = "probe_id") %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>% 
  select(probe_id, entrezgene_id, mgi_symbol) %>% 
  distinct() %>%
  
  #Filter probes from the original table
  inner_join(GSE33341_log2fc_sample, 
             by = "probe_id") %>% 
  
  #Annotate hgnc symbols
  inner_join(HS_MM_Symbol_Entrez %>% 
               clean_names() %>% 
               rename(mgi_symbol = mouse_symbol) %>% 
               select(mgi_symbol, human_symbol), 
             by = join_by("mgi_symbol")) 

#Filter entrez ids with the highest median for the same symbol ------

#Get one entrez_id by human symbol
entrez_ids_highest_median = GSE33341_mouse_counts_log2_summary_annotated_probes %>%
               select(probe_id, entrezgene_id, human_symbol, mgi_symbol, value_raw) %>%
               group_by(human_symbol, mgi_symbol, entrezgene_id) %>%
               #Get entrezids with highest median
               summarize(median = median(value_raw, na.rm = T)) %>% 
               slice_max(median, n = 1) %>%
               ungroup() %>% 
               select(human_symbol, mgi_symbol, entrezgene_id, median) %>% 
               distinct() %>% 
    summarise(n = n(), .by = c(human_symbol), mgi_symbol, entrezgene_id, median)

#Get not duplicated genes
entrez_ids_highest_median_notduplicated = entrez_ids_highest_median %>% 
  filter(n == 1L) %>% 
  distinct()

#Filter randomly entrezids that have the same value for the same gene (human and mouse)
entrez_ids_highest_median_duplicated = entrez_ids_highest_median %>%
  filter(n > 1L) %>%
  distinct() %>%
  group_by(human_symbol) %>%
  #Filter irrespective to the mgi_symbol
  slice_sample(n = 1) %>%
  select(human_symbol, mgi_symbol, entrezgene_id, median)

#Merge tables
entrez_ids_selected = entrez_ids_highest_median_notduplicated %>% 
               bind_rows(entrez_ids_highest_median_duplicated) %>% 
               select(human_symbol, mgi_symbol, entrezgene_id)

#Annotate expression table
GSE33341_mouse_counts_log2_summary_annotated_genes = GSE33341_mouse_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  inner_join(entrez_ids_selected %>% distinct(),
             by = join_by("human_symbol", "mgi_symbol", "entrezgene_id"))%>% 
  distinct()

#Check if the duplicated entrez ids were filtered out
GSE33341_mouse_counts_log2_summary_annotated_genes %>%          
        filter(human_symbol == "APOC4-APOC2",
               timepoint == "2",
               subject_id == "Mouse_C57BL/6J_Animal1_Batch8")

#Save
GSE33341_mouse_counts_log2_summary_annotated_genes %>% 
  saveRDS(file = here("tables", "GSE33341_mouse_counts_log2_summary_annotated_probes.rds"))
```

Import

```{r}
#Import
GSE33341_mouse_counts_log2_summary_annotated_probes =
  readRDS(file = here("tables", "GSE33341_mouse_counts_log2_summary_annotated_probes.rds"))

#Pivot wider
GSE33341_log2fc_sample_wide = GSE33341_mouse_counts_log2_summary_annotated_probes %>% 
  select(human_symbol, subject_id, timepoint, log2fc) %>%
  group_by(human_symbol, timepoint) %>% 
  pivot_wider(names_from = subject_id,
              values_from = log2fc) %>% 
  ungroup()  %>% 
  mutate(timepoint = fct_reorder(timepoint, timepoint %>% as.numeric()))
  
#Print
GSE33341_log2fc_sample_wide
```

##### Correlation plots, pairwise

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(GSE33341_log2fc_sample_wide,
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, GSE33341, C57BL/6J, E. coli infection", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee"))

mouse_correlation_ggpairs_alldays
#Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_alldays_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()



#Timepoint 2
timepoint_select = 2
mouse_correlation_ggpairs_day2 = ggpairs(GSE33341_log2fc_sample_wide %>% 
                                              filter(timepoint == 2) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE33341, C57BL/6J, E.coli"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee"))

mouse_correlation_ggpairs_day2
#Save
mouse_correlation_ggpairs_day2 %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_day2_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 6
timepoint_select = 6
mouse_correlation_ggpairs_day6 = ggpairs(GSE33341_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE33341, C57BL/6J, E.coli"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee"))

mouse_correlation_ggpairs_day6
#Save
mouse_correlation_ggpairs_day6 %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_day6_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 12
timepoint_select = 12
mouse_correlation_ggpairs_day12 = ggpairs(GSE33341_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE33341, C57BL/6J, E.coli"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee"))

mouse_correlation_ggpairs_day12
#Save
mouse_correlation_ggpairs_day12 %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_day12_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 24
timepoint_select = 24
mouse_correlation_ggpairs_day24 = ggpairs(GSE33341_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE33341, C57BL/6J, E.coli"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee"))

mouse_correlation_ggpairs_day24
#Save
mouse_correlation_ggpairs_day24 %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_day24_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()
```

##### PCA

```{r fig.height=4, fig.width=5}
#Import metadata
annotation_samples_pca = GSE33341_mouse_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(subject_id, "_Day_", timepoint)) %>% 
  select(subject_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = GSE33341_mouse_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(subject_id, "_Day_", timepoint)) %>% 
  select(sample, entrezgene_id, log2fc) %>% 
  filter(!entrezgene_id == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("entrezgene_id") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = subject_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  # guides(fill = "none") +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  labs(title = "Mouse, E. coli, GSE33341",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

```

##### BTMs

```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)

#Annotate genes by BTM
GSE33341_genes_btm = GSE33341_mouse_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  rename(mm_entrez = entrezgene_id) %>% 
  select(mgi_symbol, symbol = human_symbol, participant_id = subject_id, geo_sample, timepoint, status, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = GSE33341_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process
```

```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(3:7) %>% 
  colnames()

timepoint_selected = 24
plot_data <-
  #Pivot longer, 1st participant
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == timepoint_selected) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == 2) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))

max_lim = max(plot_data$participant_1_value, plot_data$participant_2_value) + 0.5
min_lim = min(plot_data$participant_1_value, plot_data$participant_2_value) - 0.5

#Plot
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(min_lim, max_lim) +
  xlim(min_lim, max_lim) +
  theme(strip.text = element_text(size = 8),
        legend.position = "right") +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste("GSE33341, Mouse, E. coli infection, Day", timepoint_selected))

```

#### Human

##### Annotate genes

There is a batch effect

```{r}
#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
GSE33341_human_expression_metadata  = GSE33341_human_eset %>% 
  pivot_longer(cols = 2:63,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(GSE33341_ecoli_mouse_human_metadata, 
             by = "geo_sample") 

#There is a batch effect
#   batch   status  n
# 1  2034 infected  7
# 2  2287  Healthy 43
# 3  2287 infected  7
# 4  2739 infected  5

#Get value from day 0 and calculate Log2FC
GSE33341_human_log2fc_sample = GSE33341_human_expression_metadata %>% 
  #Filter batch
  filter(organism == "Human",
         status == "Healthy") %>% 
  select(probe_id, participant_id = subject_id, status, day0_value_raw = value_raw) %>% 
  group_by(probe_id, status) %>% 
  summarize(n = n(),
            #Mean of healthy
            day0_value_raw = sum(day0_value_raw)/n) %>% 
  select(probe_id, day0_value_raw) %>% 
  #Join by probe id to infected participants
  inner_join(GSE33341_human_expression_metadata %>% 
               filter(site_infection == "urinary tract"), #1 lung, 19 urinary tract),
             by = join_by("probe_id")) %>% 
  mutate(log2fc = value_raw - day0_value_raw) %>% 
  select(probe_id:subject_id, batch:status, log2fc)

#Check duplicates
GSE33341_human_log2fc_sample %>% 
  filter(subject_id == "Ecoli_human_urinary tract_GSM824707",
         probe_id == "1007_s_at") #OK

```

```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- GSE33341_human_log2fc_sample$probe_id %>% unique()
listAttributes(ensembl_mart)%>% View()
#Annotate
#[HG-U133A_2] Affymetrix Human Genome U133A 2.0 Array
annot <- getBM(
  attributes = c("affy_hg_u133a_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133a_2",
  values = gene_id,
  mart = ensembl_mart)

#Gene lookup table
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  rename(human_symbol = HGNC_symbol,
         mouse_symbol =MGI.symbol,
         HS.Entrez = Human.Entrez,
         MM.Entrez = Mouse.Ortholog) %>% 
  mutate(HS.Entrez = as.character(HS.Entrez),
         MM.Entrez = as.character(MM.Entrez))
```

```{r}
#Filter probe_ids with highest median expression for the same entrezgene_id
GSE33341_human_counts_log2_summary_annotated_probes = GSE33341_human_log2fc_sample %>% 
  group_by(probe_id) %>% 
  summarize(mean = mean(value_raw, na.rm = TRUE),
            median = median(value_raw, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe_id = affy_hg_u133a_2,
                     human_symbol = hgnc_symbol), 
            by = "probe_id") %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>% 
  select(probe_id, entrezgene_id, human_symbol) %>% 
  distinct() %>%
  
  #Filter probes from the original table
  inner_join(GSE33341_human_log2fc_sample, 
             by = "probe_id") %>% 
  
  #Annotate hgnc symbols
  inner_join(HS_MM_Symbol_Entrez %>% 
               clean_names() %>% 
               select(human_symbol), 
             by = join_by("human_symbol")) 

#Filter entrez ids with the highest median for the same symbol ------

#Get one entrez_id by human symbol
entrez_ids_highest_median = GSE33341_human_counts_log2_summary_annotated_probes %>%
               select(probe_id, entrezgene_id, human_symbol, value_raw) %>%
               group_by(human_symbol, entrezgene_id) %>%
               #Get entrezids with highest median
               summarize(median = median(value_raw, na.rm = T)) %>% 
               slice_max(median, n = 1) %>%
               ungroup() %>% 
               select(human_symbol, entrezgene_id, median) %>% 
               distinct() %>% 
    summarise(n = n(), .by = c(human_symbol), entrezgene_id, median)

#Get not duplicated genes
entrez_ids_highest_median_notduplicated = entrez_ids_highest_median %>% 
  filter(n == 1L) %>% 
  distinct()

#Filter randomly entrezids that have the same value for the same gene (human and mouse)
entrez_ids_highest_median_duplicated = entrez_ids_highest_median %>%
  filter(n > 1L) %>%
  distinct() %>%
  group_by(human_symbol) %>%
  #Filter entrez_ids randomly
  slice_sample(n = 1) %>%
  select(human_symbol, entrezgene_id, median)

#Merge tables
entrez_ids_selected = entrez_ids_highest_median_notduplicated %>% 
               bind_rows(entrez_ids_highest_median_duplicated) %>% 
               select(human_symbol, entrezgene_id)

#Annotate expression table
GSE33341_human_counts_log2_summary_annotated_genes = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  inner_join(entrez_ids_selected %>% distinct(),
             by = join_by("human_symbol","entrezgene_id"))%>% 
  ungroup() %>% 
  distinct()

#Check if the duplicated entrez ids were filtered out
GSE33341_human_counts_log2_summary_annotated_genes %>%          
        filter(human_symbol == "MAGOH",
               subject_id == "Ecoli_human_urinary tract_GSM824707") %>% 
  
  distinct()

#Save
GSE33341_human_counts_log2_summary_annotated_genes %>% 
  saveRDS(file = here("tables", "GSE33341_human_counts_log2_summary_annotated_probes.rds"))
```

##### Import

```{r}
#Import
GSE33341_human_counts_log2_summary_annotated_probes =
  readRDS(file = here("tables", "GSE33341_human_counts_log2_summary_annotated_probes.rds"))

#Check duplicates
GSE33341_human_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
dplyr::summarise(n = dplyr::n(), .by = c(entrezgene_id, human_symbol, subject_id)) |>
  dplyr::filter(n > 1L) %>% 
  select(human_symbol, entrezgene_id) %>% 
  distinct()

#Pivot wider
GSE33341_human_log2fc_sample_wide = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  select(human_symbol, status, subject_id, log2fc) %>%
  group_by(human_symbol) %>% 
  pivot_wider(names_from = subject_id,
              values_from = log2fc) %>% 
  ungroup()
  
#Print
GSE33341_human_log2fc_sample_wide
```

##### Correlation plots, pairwise

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}


# Plot correlations for human -------
participant_id_selected = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  select(subject_id) %>% 
  distinct() %>% 
  slice_sample(n = 5) %>% 
  pull()

#All timepoints
human_correlation_ggpairs_alldays = ggpairs(GSE33341_human_log2fc_sample_wide %>% 
                                              select(human_symbol, status, participant_id_selected),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = status,
                                                           fill = status),
                                    title = "GSE33341, Human, E. coli infection", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point), 
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = "#219ebc") +
  scale_fill_manual(values = "#219ebc") 

human_correlation_ggpairs_alldays

#Save
human_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "human_correlation_ggpairs_alldays_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()
```

##### PCA

```{r fig.height=4, fig.width=5}
#Import metadata
annotation_samples_pca = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  select(sample = subject_id, subject_id, status, organism, batch) %>% 
  distinct()

# PCA with raw values
clean_matrix_raw = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  select(sample = subject_id, entrezgene_id, log2fc) %>% 
  filter(!entrezgene_id == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("entrezgene_id") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = batch,
      fill = batch,
      shape = organism,
      label = sample) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  # guides(fill = "none") +
 # scale_color_manual(values = "#219ebc") +
 #  scale_fill_manual(values = "#219ebc") +
  labs(title = "Human, E. coli, GSE33341",
       shape = "Organism",
       fill = "Batch") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

```

##### BTMs

```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)

#Annotate genes by BTM
GSE33341_human_genes_btm = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  rename(mm_entrez = entrezgene_id) %>% 
  select(symbol = human_symbol, participant_id = subject_id, geo_sample, status, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_human_btm_mean_process = GSE33341_human_genes_btm %>%
  mutate(condition = "infected") %>% 
  group_by(participant_id, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_human_btm_mean_process
```

```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_human_btm_mean_process = dge_human_btm_mean_process %>%  
  ungroup()

sample_participants_5 = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  select(subject_id) %>% 
  distinct() %>% 
  slice_sample(n = 5) %>% 
  pull()

plot_data <-
  #Pivot longer, 1st participant
  ggpairs_data_dge_human_btm_mean_process %>%
  # filter(timepoint == timepoint_selected) %>% 
  select(group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_human_btm_mean_process %>%
      # filter(timepoint == timepoint_selected) %>% 
      select(group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))

max_lim = max(plot_data$participant_1_value, plot_data$participant_2_value) + 0.5
min_lim = min(plot_data$participant_1_value, plot_data$participant_2_value) - 0.5

#Plot
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(min_lim, max_lim) +
  xlim(min_lim, max_lim) +
  theme(strip.text = element_text(size = 8),
        legend.position = "right") +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste("GSE33341, Human, E. coli infection"))

```

### Compare Human vs Mouse

```{r}
#Unite Mouse and Human datasets. Samples.
human_mouse_ecoli_log2fc_samples = 
#Mouse
GSE33341_mouse_counts_log2_summary_annotated_probes %>%
  ungroup() %>% 
  select(human_symbol, log2fc, participant_id = subject_id, 
         pathogen, 
         value_raw_ref = day0_value_raw, 
         value_raw,
         timepoint) %>% 
  mutate(organism = "Mouse") %>% 
  
  #Human
  bind_rows(GSE33341_human_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  select(human_symbol, log2fc,
         value_raw_ref = day0_value_raw, 
         value_raw,
         participant_id = subject_id, pathogen) %>% 
  mutate(organism = "Human",
         timepoint = "Hospitalization"))
```

```{r}
#Calculate the mean log2fc
human_mouse_ecoli_log2fc_avg = human_mouse_ecoli_log2fc_samples %>% 
  group_by(organism, human_symbol, timepoint) %>% 
  summarize(
    mean_log2fc = mean(log2fc, na.rm = TRUE),
    sd_log2fc   = sd(log2fc, na.rm = TRUE),
    se_log2fc   = sd(log2fc, na.rm = TRUE) / sqrt(n()),   # erro padrão
    
    # Test t paired
    ttest = list(tryCatch(
      t.test(value_raw, value_raw_ref, paired = TRUE),
      error = function(e) NULL
    )),
    .groups = "drop"
  ) %>%
  mutate(
    p_value = map_dbl(ttest, ~ if (!is.null(.x)) .x$p.value else NA_real_),
    ci_low  = map_dbl(ttest, ~ if (!is.null(.x)) .x$conf.int[1] else NA_real_),
    ci_high = map_dbl(ttest, ~ if (!is.null(.x)) .x$conf.int[2] else NA_real_),
    t_stat  = map_dbl(ttest, ~ if (!is.null(.x)) .x$statistic else NA_real_),
    df      = map_dbl(ttest, ~ if (!is.null(.x)) .x$parameter else NA_real_),
    p_adj   = p.adjust(p_value, method = "BH")
  ) %>%
  select(-ttest) %>% 
  filter(human_symbol != "") %>%
  ungroup()

human_mouse_ecoli_log2fc_avg

```

```{r}
#Pivot wider to compare human to different mouse times
#Human
human_mouse_ecoli_log2fc_avg_wide = 
  
  #Human
  human_mouse_ecoli_log2fc_avg %>% 
  filter(organism == "Human") %>% 
  select(human_symbol, 
         timepoint_human = timepoint, 
         mean_log2fc_human = mean_log2fc,
         p_value_human = p_value,
         p_adj_human = p_adj) %>% 
  inner_join(
  #Mouse
  human_mouse_ecoli_log2fc_avg %>% 
  filter(organism == "Mouse") %>% 
  select(human_symbol, 
         timepoint_mouse = timepoint, 
         mean_log2fc_mouse = mean_log2fc,
         p_value_mouse = p_value,
         p_adj_mouse = p_adj) %>% 
    mutate(timepoint_mouse = fct_reorder(paste0(timepoint_mouse, "h"), c(timepoint_mouse %>% as.numeric()))),
  by = "human_symbol") 

human_mouse_ecoli_log2fc_avg_wide


#Get immune and non immune genes
human_mouse_ecoli_log2fc_avg_wide_degs_immune_nonimmune = human_mouse_ecoli_log2fc_avg_wide %>%
  mutate(gene_type = "all") %>% 
  bind_rows(
  #Classify as immune or non_immune 
    human_mouse_ecoli_log2fc_avg_wide %>% 
  mutate(gene_type = if_else(human_symbol %in% ImmuneGO_BTM_grouped_genes$genes,
                             "immune",
                             "non_immune"))) %>%
  #Filter DEGs in both organisms, same sign
  filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 &
      sign(mean_log2fc_human) == sign(mean_log2fc_mouse)) %>%
  group_by(timepoint_human, timepoint_mouse, gene_type)

#Summary
degs_immune_nonimmune_summary = human_mouse_ecoli_log2fc_avg_wide_degs_immune_nonimmune %>% 
  summarize(n_genes = n(), .groups = "drop")

# Calculate the correlation
cor_data <- human_mouse_ecoli_log2fc_avg_wide_degs_immune_nonimmune %>%
  # filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>%
  group_by(timepoint_human, timepoint_mouse, gene_type) %>%
  summarize(
    cor_test = list(cor.test(mean_log2fc_mouse, mean_log2fc_human,
                             method = "pearson", use = "complete.obs")),
    .groups = "drop"
  ) %>%
  inner_join(degs_immune_nonimmune_summary, by = join_by("timepoint_human", "timepoint_mouse", "gene_type")) %>% 
  mutate(
    cor_value = map_dbl(cor_test, ~ .x$estimate),
    p_value   = map_dbl(cor_test, ~ .x$p.value),
    label_pos_x = map_dbl(cor_test, ~ max(human_mouse_ecoli_log2fc_avg_wide$mean_log2fc_mouse, na.rm = TRUE)),
    label_pos_y = map_dbl(cor_test, ~ max(human_mouse_ecoli_log2fc_avg_wide$mean_log2fc_human + 1, na.rm = TRUE)),
    label = paste0("r = ", round(cor_value, 2), ", p = ", signif(p_value, 2), ", n = ", n_genes))

  cor_data
  
  
```

```{r fig.height=3, fig.width=10}

human_mouse_ecoli_log2fc_avg_wide_scatterplot = human_mouse_ecoli_log2fc_avg_wide %>% 
  filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>% 
  
  #Plot
  ggplot() +
  aes(x = mean_log2fc_mouse,
      y = mean_log2fc_human) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #Facet
  facet_grid(~timepoint_human~timepoint_mouse) +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_ecoli_log2fc_avg_wide %>%
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_ecoli_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  stat_smooth(method = "lm", geom = "line", color = "#4DBBD5FF", alpha = 0.5, se = F, size = 1,
              data = human_mouse_ecoli_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
    #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Add correlation metrics (Immune)
  geom_label(
    data = cor_data %>% filter(gene_type == "all"),
    aes(x = 0.1, y = 9, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 0,
    vjust = 1,
    size = 2.5) +
  geom_label(
    data = cor_data %>% filter(gene_type == "immune"),
    aes(x = 0.1, y = 8, label = label),
    inherit.aes = FALSE,
    fill = "#4DBBD5FF",
    hjust = 0,
    vjust = 1,
    size = 2.5) +
  geom_label(
    data = cor_data %>% filter(gene_type == "non_immune"),
    aes(x = 0.1, y = 7, label = label),
    inherit.aes = FALSE,
    fill = "#4361ee",
    color = "white",
    hjust = 0,
    vjust = 1,
    size = 2.5) +
  
  # #Adjust the axes range
  # ylim(-4, 7) +
  
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "top",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title = "Correlation plot, E. coli infection, Log2-Fold Change, GSE33341 ")
human_mouse_ecoli_log2fc_avg_wide_scatterplot

```

#### Plot correlation between genes (log2fc for every sample)

```{r}
#Calculate correlation ------
correlation_long_all_mouse_human_timepoints <- human_mouse_ecoli_log2fc_samples %>%
  select(timepoint, human_symbol, participant_id, log2fc) %>%
  group_by(timepoint) %>%
  
  # Pivot wider and calculate correlation
  group_modify(~ {
    .x %>%
      pivot_wider(names_from = participant_id, values_from = log2fc) %>%
      select(-human_symbol) %>%
      cor(use = "pairwise.complete.obs", method = "pearson") %>%
      as.data.frame() %>%
      rownames_to_column("participant1") %>%
      pivot_longer(-participant1, names_to = "participant2", values_to = "correlation") %>%
      filter(participant1 != participant2)
  }) %>%
  ungroup() %>% 
  
  #Filter by organism
  inner_join(GSE33341_ecoli_mouse_human_metadata %>% 
               select(participant_id = subject_id, participant_1_organism = organism, sex_p_1 = sex) %>% 
               distinct(),
             by = join_by("participant1" == "participant_id")) %>% 
  inner_join(GSE33341_ecoli_mouse_human_metadata %>% 
               select(participant_id = subject_id, participant_2_organism = organism, sex_p_2 = sex) %>% 
               distinct(),
             by = join_by("participant2" == "participant_id")) 


# Remove duplicated lines only if the organism is the same
correlation_long_all_timepoints_same_organisms = correlation_long_all_mouse_human_timepoints %>%
  filter(participant_1_organism == participant_2_organism) %>%
  mutate(
    participant_min = pmin(participant1, participant2),
    participant_max = pmax(participant1, participant2)
  ) %>%
  # Remove duplicates (ex: A vs B e B vs A)
  distinct(timepoint, participant_min, participant_max, .keep_all = TRUE) %>%
  select(timepoint, participant1, participant2, correlation, participant_1_organism, participant_2_organism, sex_p_1, sex_p_2) %>%
  group_by(timepoint) %>%
  arrange(-correlation)


#Unite table with different organisms with table with deduplicated rows with the same organism
correlation_long_all_mouse_human_timepoints = correlation_long_all_mouse_human_timepoints %>% 
  filter(participant_1_organism != participant_2_organism) %>% 
  bind_rows(correlation_long_all_timepoints_same_organisms)

#Summarize data and save table
correlation_long_all_mouse_human_timepoints_summary = correlation_long_all_mouse_human_timepoints %>% 
  
  #Filter only female and set Mouse as reference
  filter(sex_p_1 == "Female", 
         sex_p_2 == "Female") %>% 
  mutate(sex_comparison = paste(sex_p_1, "vs.", sex_p_2),
         comparison = paste(participant_1_organism, "vs.", participant_2_organism)) %>%
  group_by(timepoint, comparison) %>% 
  arrange(timepoint, -correlation) %>% 
  mutate(correlation = round(correlation, 2),
         timepoint = paste("Day", timepoint)) %>% 
  summarize(mean = mean(correlation, na.rm = T) %>% round(2),
            median = median(correlation, na.rm = T)%>% round(2),
            min = min(correlation, na.rm = T)%>% round(2),
            max = max(correlation, na.rm = T) %>% round(2),
            sd = sd(correlation),
            n = sum(!is.na(correlation)),
            se = sd/sqrt(n),
            t_critical = qt(0.975, df = n - 1),
            ci_lower = mean - t_critical * se,
            ci_upper = mean + t_critical * se,
            .groups = "drop") %>% 
  filter(comparison != "Human vs. Mouse")

correlation_long_all_mouse_human_timepoints_summary

#Create GT table
correlation_long_all_mouse_human_timepoints_gt = correlation_long_all_mouse_human_timepoints_summary %>% 
  gt::gt() 
#Save gt
correlation_long_all_mouse_human_timepoints_gt
gtsave(correlation_long_all_mouse_human_timepoints_gt, filename = here("tables", "correlation_long_all_ecoli_mouse_human_timepoints_gt.docx"))

```

#### Alternative

```{r}

# --- 1) Mouse vs Human no mesmo timepoint ---
#Get timepoints
timepoints_human <- unique(human_mouse_ecoli_log2fc_samples %>% filter(organism == "Human") %>% pull(timepoint))
timepoints_mouse <- unique(human_mouse_ecoli_log2fc_samples %>% filter(organism == "Mouse") %>% pull(timepoint))

#Get combinations
timepoint_combinations <- expand.grid(timepoint_human = timepoints_human,
                                      timepoint_mouse = timepoints_mouse,
                                      stringsAsFactors = FALSE)

# Loop
cor_human_mouse_timepoint <- timepoint_combinations %>%
  pmap_dfr(function(timepoint_human, timepoint_mouse) {
    
    # pegar os participantes do timepoint específico
    human_df <- human_mouse_ecoli_log2fc_samples %>%
      filter(organism == "Human", timepoint == timepoint_human) %>%
      select(participant_id, human_symbol, log2fc)
    
    mouse_df <- human_mouse_ecoli_log2fc_samples %>%
      filter(organism == "Mouse", timepoint == timepoint_mouse) %>%
      select(participant_id, human_symbol, log2fc)
    
    # manter apenas genes em comum
    common_genes <- intersect(human_df$human_symbol, mouse_df$human_symbol)
    human_df <- human_df %>% filter(human_symbol %in% common_genes)
    mouse_df <- mouse_df %>% filter(human_symbol %in% common_genes)
    
    # pivot para wide
    human_wide <- human_df %>% pivot_wider(names_from = participant_id, values_from = log2fc)
    mouse_wide <- mouse_df %>% pivot_wider(names_from = participant_id, values_from = log2fc)
    
    # IDs de participantes
    human_ids <- setdiff(colnames(human_wide), "human_symbol")
    mouse_ids <- setdiff(colnames(mouse_wide), "human_symbol")
    
    # calcular correlação para todos pares humano x mouse
    expand.grid(human = human_ids, mouse = mouse_ids, stringsAsFactors = FALSE) %>%
      mutate(
        cor_test = map2(human, mouse, ~{
          x <- human_wide[[.x]]
          y <- mouse_wide[[.y]]
          # remover NAs correspondentes
          keep <- complete.cases(x, y)
          cor.test(x[keep], y[keep], method = "pearson")
        }),
        r       = map_dbl(cor_test, ~.x$estimate),
        p_value = map_dbl(cor_test, ~.x$p.value),
        ci_low  = map_dbl(cor_test, ~.x$conf.int[1]),
        ci_high = map_dbl(cor_test, ~.x$conf.int[2])
      ) %>%
      select(-cor_test) %>%
      mutate(timepoint_human = timepoint_human,
             timepoint_mouse = timepoint_mouse)
  }) %>% 
  group_by(timepoint_human, timepoint_mouse) %>%
  summarize(
    comparison = "Human vs. Mouse",
    mean = mean(r, na.rm = TRUE),
    sd   = sd(r, na.rm = TRUE),
    n      = sum(!is.na(r)),
    se     = sd / sqrt(n),
    tval   = qt(0.975, df = n - 1),
    ci_low = mean - tval * se,
    ci_high = mean + tval * se,
    .groups = "drop"
  )

# --- 2) Human vs Human (entre timepoints) ---
cor_human_human <- human_mouse_ecoli_log2fc_samples %>%
  filter(organism == "Human") %>%
  select(participant_id, human_symbol, timepoint, log2fc) %>%
  pivot_wider(
    names_from = participant_id,
    values_from = log2fc
  ) %>%
  select(-human_symbol, -timepoint) %>%   # só colunas de participantes
  cor(use = "pairwise.complete.obs", method = "pearson") %>% 
  as.data.frame() %>%
  rownames_to_column("participant_1") %>%
  melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>% 
  filter(participant_1 != participant_2) %>%   # tira a diagonal (=1)
  # garantir combinações únicas
  rowwise() %>%
  mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
  ungroup() %>%
  distinct(pair, .keep_all = TRUE) %>%
  #Summarize
  summarize(
    comparison = "Human vs. Human",
    mean = mean(correlation, na.rm = TRUE),
    sd   = sd(correlation, na.rm = TRUE),
    n    = sum(!is.na(correlation)),
    se   = sd / sqrt(n),
    tval = qt(0.975, df = n - 1),
    ci_low  = mean - tval * se,
    ci_high = mean + tval * se
  ) 
  

# --- 3) Mouse vs Mouse ---
cor_mouse_mouse <- human_mouse_ecoli_log2fc_samples %>%
  filter(organism == "Mouse") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, human_symbol, log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = log2fc) %>%
      select(-human_symbol)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Mouse vs. Mouse",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup()


#Unite
correlation_long_all_mouse_human_timepoints_summary = cor_human_mouse_timepoint %>% 
  mutate(timepoint = timepoint_mouse) %>% 
  bind_rows(
    cor_human_human %>% mutate(timepoint = "24"),
    cor_mouse_mouse)

correlation_long_all_mouse_human_timepoints_summary

```

```{r fig.height=3, fig.width=5}
correlation_long_all_mouse_human_timepoints_summary %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  # geom_textline(aes(label = comparison), 
  #               size = 3, 
  #               vjust = -0.5,
  #               hjust = 0.8,
  #               linewidth = 1) +
  
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 2, 6, 12, 24), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_color_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = "Correlation between same and different organisms, by subject",
       x = "",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "right") +
  guides(fill = "none") 
```

#### BTMs

```{r}
#Annotate genes by BTM
dge_btm_mean_process = GSE33341_genes_btm %>%
  mutate(organism = "Mouse") %>% 
  bind_rows(GSE33341_human_genes_btm %>% 
              mutate(organism = "Human")) %>%
  # filter(timepoint %in% c(1, 2)) %>% 
  mutate(condition = if_else(organism == "Mouse", paste0(organism, "_", timepoint, "h"), organism)) %>% 
  group_by(condition, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  select(-timepoint) %>% 
  pivot_wider(names_from = "condition", values_from = "mean_log2fc") 

dge_btm_mean_process
```

##### Scatterplot

```{r fig.height=3, fig.width=10}

plot_data <-
  #Pivot longer, 1st participatn
  dge_btm_mean_process %>%
  pivot_longer(cols = c(Mouse_12h:Mouse_6h), 
               names_to = "mouse_subject", 
               values_to = "mouse_subject_value")


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(mouse_subject) %>%
  summarise(cor_value = cor(Human, mouse_subject_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(comparison = paste0("Human vs. ", mouse_subject),
         r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = mouse_subject_value, y = Human)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  facet_grid("Human" ~ mouse_subject,  drop = T) +
  stat_smooth(aes(x = mouse_subject_value, y = Human),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  theme_vaxgo() +
  # ylim(-3, 3) +
  xlim(-1.5, 2) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "Mouse Log2FC", 
       y = "Human Log2FC", 
       color = "",
       title = "Mouse vs. Human")
```
