---
title: "Exploring other datasets"
author: "Wasim Aluísio Prates-Syed"
date: "2025-07-23"
output: 
  html_document: 
    toc: TRUE
    toc_depth: 4
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE, cache = TRUE)
```


# Exploring other datasets: A general analysis

## Import required codes
```{r}
# install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```

## Found in translation
Ref: Normand, R., Du, W., Briller, M. et al. Found In Translation: a machine learning model for mouse-to-human inference. Nat Methods 15, 1067–1073 (2018). https://doi.org/10.1038/s41592-018-0214-9
Github: https://github.com/shenorrLabTRDF/FIT.mouse2man 

- Standard (ST): which include human and mouse datasets that were conducted in separate experiments;
- Reference (RF), in which the human and mouse datasets were directly contrasted in a publication authored by researchers who had generated at least one of the datasets.
- CSP IDs can be enumarated by different comparisons with different mice strains, for example A/J and CD1 mice, and timpoints (not clearly stated).
```{r}
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)

#Import files
pca_rotations = FIT.mouse2man::pca_rotations
best_models = FIT.mouse2man::best_models
slopes_per_gene_V2.0 = FIT.mouse2man::slopes_per_gene_V2.0

#Gene lookup table
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  rename(human_symbol = HGNC_symbol,
         mouse_symbol =MGI.symbol,
         HS.Entrez = Human.Entrez,
         MM.Entrez = Mouse.Ortholog) %>% 
  mutate(HS.Entrez = as.character(HS.Entrez),
         MM.Entrez = as.character(MM.Entrez))

# MM_Entrez_symbol_desc = FIT.mouse2man::MM_Entrez_symbol_desc
# Psoriasis_classifier_neg = FIT.mouse2man::Psoriasis_classifier_neg
# StaphInfection_classifier_pos = FIT.mouse2man::StaphInfection_classifier_pos

#Training data
AllData_V2.0 = FIT.mouse2man::AllData_V2.0 %>% 
  mutate(csp_id_short = str_remove(CSP_ID, "Microarrays_RF_"),
         csp_id_short = str_remove(csp_id_short, "Microarrays_ST_"),
         csp_id_short = str_remove(csp_id_short, "RNAseq_ST_"),
         csp_id_short = str_remove(csp_id_short, "RNAseq_RF_")) %>% 
  inner_join(HS_MM_Symbol_Entrez, 
             by = join_by("HS.Entrez", "MM.Entrez")) %>% 
  separate(csp_id_short, into = c("del", "csp_id_index")) %>% 
  select(-del)
  
AllData_V2.0$csp_id_index %>% unique()

```


## Filter datasets
Get metadata for datasets on Supplementary Files of FIT paper and filter datasets with blood-derived samples
```{r}
#Get metadata for datasets on Supplementary Files of FIT paper and filter datasets with blood-derived samples
#Import
fit_training_datasets_metadata <- read_excel(here("FIT_Mouse2man", "41592_2018_214_MOESM3_ESM.xlsx"))

#Clean grouped header
fit_training_datasets_metadata_newcols = fit_training_datasets_metadata %>% 
  slice_head(n = 2) %>% 
  t() %>% 
  as.data.frame() %>% 
  
  #Fill sub columns
  rownames_to_column("V0") %>% 
  
  mutate(V1 = if_else(is.na(V1), V0, V1),
         V0 = ifelse(str_detect(V0, "^\\.\\.\\.\\d+$"), NA, V0),
         V1 = ifelse(str_detect(V1, "^\\.\\.\\.\\d+$"), NA, V1)) %>% 
  
  #Fill columns
  fill(V0, .direction = "down") %>% 
  fill(V1, .direction = "down") %>% 
  
  #Fill columns manually
  mutate(V2 = if_else(V0 == V1, V0, V2),
         V3 = if_else(V0 == V1, V0, NA),
         V3 = if_else(is.na(V2), paste0(V0,"_", V1), V2),
         V2 = if_else(is.na(V2), V3, V2),
         V3 = if_else(V2 == "Disease" & V0 != "Disease", 
                      paste0(V0, "_", V1, "_", V2), 
                      V3),
         V3 = if_else(V2 == "Control" & V0 != "Control", 
                      paste0(V0, "_", V1, "_", V2), 
                      V3),
         #Clean table further
         V3 = str_remove(V3, " Data"),
         V3 = make_clean_names(V3)) %>% 
  select(V3)

#Set new names
fit_training_datasets_metadata_clean = fit_training_datasets_metadata %>% 
  filter(!row_number() %in% c(1:2)) %>% 
  setNames(fit_training_datasets_metadata_newcols %>% pull()) %>% 
  fill(disease, .direction = "down")

fit_training_datasets_metadata_clean %>% 
  saveRDS(file = here("FIT_Mouse2man", "fit_training_datasets_metadata_clean.rds"))
  
#Get datasets with blood-derived samples only, in both organisms
fit_training_datasets_blood = fit_training_datasets_metadata_clean %>% 
  filter(str_detect(human_tissue_control, "Blood"),
         str_detect(mouse_tissue_disease, "Blood"))

AllData_V2.0_blood = AllData_V2.0 %>% 
  filter(CSP_ID %in% c(fit_training_datasets_blood$comparison_id %>% unique())) 

AllData_V2.0_blood %>% 
  saveRDS(here("new_models", "AllData_FIT_training_blood.rds"))

```


##Import
```{r}
#Import data
fit_training_datasets_metadata_clean = read_rds(file = here("FIT_Mouse2man", "fit_training_datasets_metadata_clean.rds"))
AllData_V2.0_blood = read_rds(here("new_models", "AllData_FIT_training_blood.rds"))
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))

```

## Plot distributions
```{r fig.height=8, fig.width=10}
AllData_V2.0_distributions_plot = AllData_V2.0 %>% 
  clean_names() %>% 
  pivot_longer(cols = c(fc_hs:qval_mm),
               names_to = "metric",
               values_to = "value") %>% 
  mutate(organism = if_else(str_detect(metric, "mm"),"Mouse", "Human")) %>% 
  filter(metric %in% c("fc_hs", "fc_mm")) %>% 
  mutate(metric = if_else(str_detect(metric, "fc"), "fc", metric)) %>% 
  ggplot() +
  aes(x = value, fill = organism) +
  geom_histogram( position = position_identity(), alpha = 0.5, bins = 50) +
  facet_wrap(~disease, 
             scale = "free_y", axes = "all" 
             # strip.position="right"
             ) +
  scale_y_continuous(expand = c(0.01, 0.01), n.breaks = 3) +
  scale_x_continuous(limits = c(-2, 2)) +
    scale_fill_manual(
  values = c(
    FIT = "#CE1800FF", 
    Mouse = "#546E7AFF",
    Human = "#4DBBD5FF"))+
    scale_color_manual(
  values = c(
    FIT = "#CE1800FF", 
    Mouse = "#546E7AFF",
    Human = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0, hjust = 0), 
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA)) +
  labs(fill = "Organism",
       color = "Organism",
       x = "Log2-fold change",
       y = "Frequency",
       title = "All samples")

AllData_V2.0_blood_distributions_plot = AllData_V2.0_blood %>% 
  clean_names() %>% 
  pivot_longer(cols = c(fc_hs:qval_mm),
               names_to = "metric",
               values_to = "value") %>% 
  mutate(organism = if_else(str_detect(metric, "mm"),"Mouse", "Human")) %>% 
  filter(metric %in% c("fc_hs", "fc_mm")) %>% 
  mutate(metric = if_else(str_detect(metric, "fc"), "fc", metric)) %>% 
  ggplot() +
  aes(x = value, fill = organism) +
  geom_histogram( position = position_identity(), alpha = 0.5, bins = 50) +
  facet_wrap(~disease, 
             scale = "free_y", axes = "all",
             ncol = 1
             # strip.position="right"
             ) +
  scale_y_continuous(expand = c(0.01, 0.01), n.breaks = 3) +
  scale_x_continuous(limits = c(-2, 2)) +
    scale_fill_manual(
  values = c(
    Mouse = "#546E7AFF",
    Human = "#4DBBD5FF"))+
    scale_color_manual(
  values = c(
    Mouse = "#546E7AFF",
    Human = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0, hjust = 0), 
        panel.spacing.y = unit(0.8, "lines"),
        plot.margin = margin(0.5, 1, 0.2, 1, "cm"),
        strip.background.y = element_rect(fill = NA, color = NA)) +
  labs(fill = "Organism",
       color = "Organism",
       x = "Log2-fold change",
       y = "Frequency",
       title = "Blood samples")


#Plot all
AllData_V2.0_distributions_all = AllData_V2.0_distributions_plot + AllData_V2.0_blood_distributions_plot +
  plot_layout(widths = c(1, 0.15),
              guides = "collect",
              axes = "collect") +
  plot_annotation(theme = theme(legend.position = "top"))

#Save
AllData_V2.0_distributions_all %>% 
  ggsave(filename = here("Figures", "FIT_Training data_distributions_all.png"),
         width = 10,
         height = 10)

```



##Correlation plots
```{r fig.height=12, fig.width=10}
# Calculate the correlation
cor_data <- AllData_V2.0_blood %>%
  clean_names() %>%
  filter(qval_hs <= 0.05 | qval_mm <= 0.05) %>%
  mutate(csp_id_index = fct_relevel(csp_id_index, as.character(1:30))) %>%
  group_by(disease, csp_id_index) %>%
  summarize(
    cor_value = cor(fc_mm, fc_hs, method = "pearson", use = "complete.obs"),
    label_pos_x = max(fc_mm, na.rm = TRUE),
    label_pos_y = max(fc_hs, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("r = ", round(cor_value, 2)))


#Log2FC scatterplot
blood_l2fc_scatterplot = AllData_V2.0_blood %>%
  clean_names() %>%
  mutate(csp_id_index = fct_relevel(csp_id_index, as.character(1:30))) %>% 
  filter(qval_hs <= 0.05 | qval_mm <= 0.05) %>% 
  
  #Plot
  ggplot() +
  aes(x = fc_mm,
      y = fc_hs) +
  
  facet_nested_wrap(~disease+paste("Comparison", csp_id_index), scales = "free", nest_line = T) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             
             #Get only DEGs with the same sign (positive correlation)
             data = AllData_V2.0_blood %>%
               clean_names() %>%
               #Filter DEGs in both organisms, same sign
               filter(qval_hs <= 0.05 & qval_mm <= 0.05 & sign(fc_hs) == sign(fc_mm))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = AllData_V2.0_blood %>%
               clean_names() %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               filter(qval_hs <= 0.05 & qval_mm <= 0.05 & sign(fc_hs) == sign(fc_mm))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  
  #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  scale_x_continuous(breaks = c(0, 1, 2, 5, 7, 10, -1, -2, -5, -7, -10)) +
  scale_y_continuous(breaks = c(0, 1, 2, 5, 7, 10, -1, -2, -5, -7, -10)) +
  #Add correlation metrics
  geom_label(
    data = cor_data,
    aes(x = label_pos_x, y = label_pos_y, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 1,
    vjust = 1,
    size = 3
  ) +
  
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =8),
        axis.text.y = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "none",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title= "Correlation plot, Log2-Fold Change")

blood_l2fc_scatterplot

#Save
blood_l2fc_scatterplot %>% 
  ggsave(filename = here("Figures", "FIT_TrainingData_blood_l2fc_scatterplot.png"),
         height = 12,
         width = 10)

```





### Annotate genes
```{r fig.height=8, fig.width=8}

## Plot correlation-----
# cov_genes_FIT_training = AllData_V2.0 %>% 
#   clean_names() %>% 
#   select(hs_entrez, mean_human = fc_hs, mean_mouse = fc_mm) %>% 
#   inner_join(HS_MM_Symbol_Entrez %>%
#                clean_names() %>%
#                select(human_symbol = hgnc_symbol,
#                       hs_entrez = human_entrez) %>% 
#                mutate(hs_entrez = as.character(hs_entrez)),
#              by = "hs_entrez") %>% 
#   arrange(human_symbol) %>% 
#   group_by(human_symbol) %>% 
#   summarize(sd_human = sd(mean_human),
#             cov = cov(mean_human, mean_mouse),
#             mean_human_time = mean(mean_human),
#             mean_mouse_time = mean(mean_mouse)) %>% 
#   arrange(-cov)
# 
# cov_genes_FIT_training %>% filter(abs(cov) > 0.001) #15891 genes from 6,373 (92.9%)
# cov_genes_FIT_training %>% filter(abs(cov) > 0.01) #11320 genes from 6,373 (50.3%)
# cov_genes_FIT_training %>% filter(abs(cov) > 0.1) #1151 genes
# cov_genes_FIT_training %>% filter(abs(cov) > 0.2) #356 genes
# cov_genes_FIT_training %>% filter(abs(cov) > 0.3) #186 genes
# 
# 
# genes_vary_human_all_FIT = cov_genes_FIT_training %>% 
#   filter(abs(mean_human_time) >= 0.1,
#         abs(mean_mouse_time) <= 0.1) %>% 
#   arrange(-abs(cov))
# 
# genes_vary_mouse_all_FIT = cov_genes_FIT_training %>% 
#   filter(abs(mean_mouse_time) >= 0.1,
#                 abs(mean_human_time) <= 0.1) %>% 
#   arrange(-abs(cov))
# 
# genes_vary_human_mouse_all_FIT = cov_genes_FIT_training %>% 
#   filter(abs(mean_human_time) >= 0.1,
#          abs(mean_mouse_time) >= 0.1) %>% 
#   arrange(-abs(cov))
# 
# genes_not_vary_human_mouse_all_FIT = cov_genes_FIT_training %>% 
#    filter(abs(mean_human_time) <= 0.1, 
#           abs(mean_mouse_time) <= 0.1) %>%  
#   arrange(-abs(cov))
# 
# cov_genes_FIT_training %>% filter(abs(mean_human_time) > 0.1) #3,962
# cov_genes_FIT_training %>% filter(abs(mean_human_time) > 0.2) #2191
# 
# cov_genes_FIT_training %>% filter(abs(mean_mouse_time) > 0.1) #3363
# cov_genes_FIT_training %>% filter(abs(mean_mouse_time) > 0.2) #1631
# 
# cov_genes_FIT_training %>% filter(abs(mean_mouse_time) < 0.1) #3,010
# cov_genes_FIT_training %>% filter(abs(mean_mouse_time) < 0.2) #4742
# 
# 
# 
# genes_vary_human_all_FIT #5,787
# genes_vary_mouse_all_FIT #1,807
# genes_not_vary_human_mouse_all_FIT #6,476
# genes_vary_human_mouse_all_FIT #2,355

# #Plot
# genes_vary_human = cov_genes_FIT_training %>% 
#   ggplot() +
#   aes(x = mean_human_time,
#       y = mean_mouse_time) +
#     geom_point(color = "#3182bd", fill = NA, pch = 21) +
#   scale_x_continuous(limits = c(-1.5, 1.5)) +
#   scale_y_continuous(limits = c(-1.5, 1.5)) +
#     gghighlight(abs(mean_human_time) >= 0.1,
#                 abs(mean_mouse_time) <= 0.1,
#               use_direct_label = FALSE, 
#               unhighlighted_params = list(alpha = 0.2, fill = NA)) +
#   geom_label_repel(aes(label = human_symbol), max.overlaps = 20, force_pull = 4, size = 3,
#                   data = cov_genes_FIT_training %>% 
#                     filter(abs(mean_human_time) > 0.1,
#                 abs(mean_mouse_time) <= 0.1) %>% 
#                   arrange(-abs(cov)) %>% 
#                   slice_head(n = 10)) +
#    theme_vaxgo() +
#   labs(title = "Which genes vary only in human?",
#        subtitle = "Top 10 genes with highest covariance. Mean cutoff > 0.1.")
# 
# genes_vary_mouse = cov_genes_FIT_training %>% 
#   ggplot() +
#   aes(x = mean_human_time,
#       y = mean_mouse_time) +
#     geom_point(color = "#90A4AEFF", fill = NA, pch = 21) +
#   scale_x_continuous(limits = c(-1.5, 1.5)) +
#   scale_y_continuous(limits = c(-1.5, 1.5)) +
#   gghighlight(abs(mean_mouse_time) >= 0.1,
#                 abs(mean_human_time) <= 0.1,
#               use_direct_label = FALSE,
#               unhighlighted_params = list(alpha = 0.2, fill = NA)) +
#   geom_label_repel(aes(label = human_symbol), max.overlaps = 20, force_pull = 4, size = 3,
#                   data = cov_genes_FIT_training %>% 
#                     filter(abs(mean_mouse_time) >= 0.1,
#                 abs(mean_human_time) <= 0.1) %>% 
#                   arrange(-abs(cov)) %>% 
#                   slice_head(n = 10)) +
# 
#   theme_vaxgo() +
#   labs(title = "Which genes vary only in mouse?",
#        subtitle = "Top 10 genes with highest covariance. Mean cutoff > 0.1") 
# 
# 
# genes_vary_mouse_human = cov_genes_FIT_training %>% 
#   ggplot() +
#   aes(x = mean_human_time,
#       y = mean_mouse_time) +
#   geom_point(color = "black", fill = NA, pch = 21) +
#   scale_x_continuous(limits = c(-1.5, 1.5)) +
#   scale_y_continuous(limits = c(-1.5, 1.5)) +
#   gghighlight(abs(mean_human_time) >= 0.1,
#                 abs(mean_mouse_time) >= 0.1,
#                             unhighlighted_params = list(alpha = 0.2, fill = NA)) +
#   geom_label_repel(aes(label = human_symbol), max.overlaps = 20, force_pull = 4, size = 3,
#                   data = cov_genes_FIT_training %>% 
#                     filter(abs(mean_human_time) >= 0.1,
#                 abs(mean_mouse_time) >= 0.1) %>% 
#                   arrange(-abs(cov)) %>% 
#                   slice_head(n = 10)) +
#   theme_vaxgo() +
#   labs(title = "Which genes vary in both human and mouse?",
#        subtitle = "Top 10 genes with highest covariance. Mean cutoff > 0.1")
# 
# genes_notvary_mouse_human = cov_genes_FIT_training %>% 
#   ggplot() +
#   aes(x = mean_human_time,
#       y = mean_mouse_time) +
#   geom_point(color = "black", fill = NA, pch = 21) +
#   scale_x_continuous(limits = c(-1.5, 1.5)) +
#   scale_y_continuous(limits = c(-1.5, 1.5)) +
#   gghighlight(abs(mean_human_time) <= 0.1,
#                 abs(mean_mouse_time) <= 0.1,
#               unhighlighted_params = list(alpha = 0.2, fill = NA)) +
#   geom_label_repel(aes(label = human_symbol), 
#                   # color = "red",
#                   max.overlaps = 20, force_pull = 4, size = 3, 
#                   data = cov_genes_FIT_training %>% 
#                     filter(abs(mean_human_time) <= 0.1, 
#                              abs(mean_mouse_time) <= 0.1) %>%  
#                     arrange(-abs(cov)) %>% 
#                     slice_head(n = 10)) +
#   theme_vaxgo() +
#   labs(title = "Which genes do not vary in both human and mouse?",
#        subtitle = "Top 10 genes with highest covariance. Mean cutoff < 0.1")
#   
# 
# #Plot all
# genes_vary_mouse + genes_vary_human + genes_vary_mouse_human + genes_notvary_mouse_human +
#   plot_layout(guides = "collect", axes = "collect", axis_titles =  "collect",
#               ncol =  2)
  
```


