---
title: "Exploring other datasets"
author: "Wasim Aluísio Prates-Syed"
date: "2025-07-23"
output: 
  html_document: 
    toc: TRUE
    toc_depth: 4
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE, cache = TRUE)
```


# Exploring other datasets: A general analysis

## Import required codes
```{r}
# install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```

## Found in translation
Ref: Normand, R., Du, W., Briller, M. et al. Found In Translation: a machine learning model for mouse-to-human inference. Nat Methods 15, 1067–1073 (2018). https://doi.org/10.1038/s41592-018-0214-9
Github: https://github.com/shenorrLabTRDF/FIT.mouse2man 

- Standard (ST): which include human and mouse datasets that were conducted in separate experiments;
- Reference (RF), in which the human and mouse datasets were directly contrasted in a publication authored by researchers who had generated at least one of the datasets.
- CSP IDs can be enumarated by different comparisons with different mice strains, for example A/J and CD1 mice, and timpoints (not clearly stated).
```{r}
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)

#Import files
pca_rotations = FIT.mouse2man::pca_rotations
best_models = FIT.mouse2man::best_models
slopes_per_gene_V2.0 = FIT.mouse2man::slopes_per_gene_V2.0

#Gene lookup table
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  rename(human_symbol = HGNC_symbol,
         mouse_symbol =MGI.symbol,
         HS.Entrez = Human.Entrez,
         MM.Entrez = Mouse.Ortholog) %>% 
  mutate(HS.Entrez = as.character(HS.Entrez),
         MM.Entrez = as.character(MM.Entrez))

# MM_Entrez_symbol_desc = FIT.mouse2man::MM_Entrez_symbol_desc
# Psoriasis_classifier_neg = FIT.mouse2man::Psoriasis_classifier_neg
# StaphInfection_classifier_pos = FIT.mouse2man::StaphInfection_classifier_pos

#Training data
AllData_V2.0 = FIT.mouse2man::AllData_V2.0 %>% 
  mutate(csp_id_short = str_remove(CSP_ID, "Microarrays_RF_"),
         csp_id_short = str_remove(csp_id_short, "Microarrays_ST_"),
         csp_id_short = str_remove(csp_id_short, "RNAseq_ST_"),
         csp_id_short = str_remove(csp_id_short, "RNAseq_RF_")) %>% 
  inner_join(HS_MM_Symbol_Entrez, 
             by = join_by("HS.Entrez", "MM.Entrez")) %>% 
  separate(csp_id_short, into = c("del", "csp_id_index")) %>% 
  select(-del)
  
AllData_V2.0$csp_id_index %>% unique()

```


## Filter datasets
Get metadata for datasets on Supplementary Files of FIT paper and filter datasets with blood-derived samples
```{r}
#Get metadata for datasets on Supplementary Files of FIT paper and filter datasets with blood-derived samples
#Import
fit_training_datasets_metadata <- read_excel(here("FIT_Mouse2man", "41592_2018_214_MOESM3_ESM.xlsx"))

#Clean grouped header
fit_training_datasets_metadata_newcols = fit_training_datasets_metadata %>% 
  slice_head(n = 2) %>% 
  t() %>% 
  as.data.frame() %>% 
  
  #Fill sub columns
  rownames_to_column("V0") %>% 
  
  mutate(V1 = if_else(is.na(V1), V0, V1),
         V0 = ifelse(str_detect(V0, "^\\.\\.\\.\\d+$"), NA, V0),
         V1 = ifelse(str_detect(V1, "^\\.\\.\\.\\d+$"), NA, V1)) %>% 
  
  #Fill columns
  fill(V0, .direction = "down") %>% 
  fill(V1, .direction = "down") %>% 
  
  #Fill columns manually
  mutate(V2 = if_else(V0 == V1, V0, V2),
         V3 = if_else(V0 == V1, V0, NA),
         V3 = if_else(is.na(V2), paste0(V0,"_", V1), V2),
         V2 = if_else(is.na(V2), V3, V2),
         V3 = if_else(V2 == "Disease" & V0 != "Disease", 
                      paste0(V0, "_", V1, "_", V2), 
                      V3),
         V3 = if_else(V2 == "Control" & V0 != "Control", 
                      paste0(V0, "_", V1, "_", V2), 
                      V3),
         #Clean table further
         V3 = str_remove(V3, " Data"),
         V3 = make_clean_names(V3)) %>% 
  select(V3)

#Set new names
fit_training_datasets_metadata_clean = fit_training_datasets_metadata %>% 
  filter(!row_number() %in% c(1:2)) %>% 
  setNames(fit_training_datasets_metadata_newcols %>% pull()) %>% 
  fill(disease, .direction = "down")

fit_training_datasets_metadata_clean %>% 
  saveRDS(file = here("FIT_Mouse2man", "fit_training_datasets_metadata_clean.rds"))
  
#Get datasets with blood-derived samples only, in both organisms
fit_training_datasets_blood = fit_training_datasets_metadata_clean %>% 
  filter(str_detect(human_tissue_control, "Blood"),
         str_detect(mouse_tissue_disease, "Blood"))

AllData_V2.0_blood = AllData_V2.0 %>% 
  filter(CSP_ID %in% c(fit_training_datasets_blood$comparison_id %>% unique())) 

AllData_V2.0_blood %>% 
  saveRDS(here("new_models", "AllData_FIT_training_blood.rds"))

```


##Import
```{r}
#Import data
fit_training_datasets_metadata_clean = read_rds(file = here("FIT_Mouse2man", "fit_training_datasets_metadata_clean.rds"))

fit_training_datasets_metadata_clean = fit_training_datasets_metadata_clean %>% 
  mutate(
  #Shorten tissue names
  human_tissue_disease_short = case_when(
    human_tissue_disease %in% c("Lung", "Lung epithelium") ~ "Lung",
    human_tissue_disease == "Mammary Epithelium" ~ "Mammary",
    human_tissue_disease %in% c("Blood", "White Blood Cells") ~ "Blood",
    human_tissue_disease %in% c("Peripheral Blood Monocytes", "Monocytes") ~ "Monocytes",
    human_tissue_disease == "Peripheral Blood Mononuclear Cell" ~ "PBMC",
    human_tissue_disease %in% c("Neuron") ~ "Neuron",
    human_tissue_disease %in% c("Brain", "Cerebrum") ~ "Brain",
    human_tissue_disease == "Cerebral Cortex" ~ "Cortex",
    human_tissue_disease == "Cerebellum" ~ "Cerebellum",
    human_tissue_disease == "Caudate Nucleus" ~ "Caudate",
    human_tissue_disease == "Renal Tubule" ~ "Kidney",
    human_tissue_disease == "Ovary" ~ "Ovary",
    human_tissue_disease == "Colon" ~ "Colon",
    human_tissue_disease == "Skin" ~ "Skin",
    human_tissue_disease == "Heart" ~ "Heart",
    human_tissue_disease %in% c("MUSCLE, QUADRICEPS FEMORIS", "Quadriceps") ~ "Quadriceps",
    human_tissue_disease %in% c("Tibialis", "Calf", "skeletal muscle") ~ "Muscle",
    human_tissue_disease == "Fibroblast Culture" ~ "Fibroblast",
    human_tissue_disease == "Nasal Epithelial Cells" ~ "Nasal Epith.",
    human_tissue_disease == "Biopsy Sample" ~ "Biopsy",
    human_tissue_disease == "Glioblastoma Cell Line" ~ "Glioblastoma",
    human_tissue_disease %in% c("Human Cell Line", "LNCaP cells", "lymphoblast") ~ "Cell line",
    human_tissue_disease == "T-cell" ~ "T cell",
    TRUE ~ human_tissue_disease
  ),
  #Group by organ
  human_organ = case_when(
  human_tissue_disease %in% c(
    "Lung",
    "Lung epithelium"
  ) ~ "Lung",

  human_tissue_disease %in% c(
    "Brain",
    "Cerebrum",
    "Cerebral Cortex",
    "Cerebellum",
    "Caudate Nucleus",
    "Neuron"
  ) ~ "Brain",

  human_tissue_disease %in% c(
    "Blood",
    "White Blood Cells",
    "Peripheral Blood Monocytes",
    "Monocytes",
    "Peripheral Blood Mononuclear Cell",
    "T-cell"
  ) ~ "Blood/Immune",

  human_tissue_disease == "Renal Tubule" ~ "Kidney",

  human_tissue_disease == "Mammary Epithelium" ~ "Mammary",

  human_tissue_disease == "Ovary" ~ "Ovary",

  human_tissue_disease == "Colon" ~ "Colon",

  human_tissue_disease == "Skin" ~ "Skin",

  human_tissue_disease == "Heart" ~ "Heart",

  human_tissue_disease %in% c(
    "MUSCLE, QUADRICEPS FEMORIS",
    "Quadriceps",
    "Tibialis",
    "Calf",
    "skeletal muscle"
  ) ~ "Muscle",

  human_tissue_disease == "Nasal Epithelial Cells" ~ "Nasal",

  human_tissue_disease %in% c(
    "Fibroblast Culture",
    "Biopsy Sample"
  ) ~ "Other tissue",

  human_tissue_disease %in% c(
    "Glioblastoma Cell Line",
    "Human Cell Line",
    "LNCaP cells",
    "lymphoblast"
  ) ~ "Cell line",

  TRUE ~ "Other"
),
  mouse_tissue_short = case_when(
    mouse_tissue_disease == "Lung" ~ "Lung",
    mouse_tissue_disease == "Embryonic Tissue" ~ "Embryo",
    mouse_tissue_disease %in% c("Blood", "White Blood Cells") ~ "Blood",
    mouse_tissue_disease == "Brain" ~ "Brain",
    mouse_tissue_disease == "Corpus Striatum" ~ "Striatum",
    mouse_tissue_disease == "BRAIN, HIPPOCAMPUS" ~ "Hippocampus",
    mouse_tissue_disease == "Cerebral Cortex" ~ "Cortex",
    mouse_tissue_disease == "Brain Olig2+ Cells" ~ "Olig2+ cells",
    mouse_tissue_disease %in% c("Kidney", "immortal kidney epithelium") ~ "Kidney",
    mouse_tissue_disease == "Liver" ~ "Liver",
    mouse_tissue_disease == "Ovary" ~ "Ovary",
    mouse_tissue_disease == "Colon" ~ "Colon",
    mouse_tissue_disease == "Skin" ~ "Skin",
    mouse_tissue_disease == "Spleen" ~ "Spleen",
    mouse_tissue_disease == "Heart" ~ "Heart",
    mouse_tissue_disease == "Diaphragm" ~ "Diaphragm",
    mouse_tissue_disease %in% c("MUSCLE, QUADRICEPS FEMORIS", "skeletal muscle", "Calf") ~ "Muscle",
    mouse_tissue_disease == "Peritoneal Macrophage" ~ "Macrophage",
    mouse_tissue_disease == "Bone Marrow Sca+Lin- Cells" ~ "HSC prog.",
    mouse_tissue_disease == "Hematopoietic Stem Cell" ~ "HSC",
    mouse_tissue_disease == "CD4+ T cells" ~ "CD4 T",
    mouse_tissue_disease %in% c("Peritoneal Cavity B Cells", "Splenic B Cells") ~ "B cell",
    mouse_tissue_disease == "Organoid" ~ "Organoid",
    TRUE ~ mouse_tissue_disease
  ),
  #Group by organ
  mouse_organ = case_when(
    mouse_tissue_disease == "Lung" ~ "Lung",
    mouse_tissue_disease == "Embryonic Tissue" ~ "Embryo",
    mouse_tissue_disease %in% c(
      "Brain",
      "Corpus Striatum",
      "Cerebral Cortex",
      "BRAIN, HIPPOCAMPUS",
      "Brain Olig2+ Cells"
    ) ~ "Brain",
    mouse_tissue_disease %in% c(
      "Blood",
      "White Blood Cells",
      "Bone Marrow Sca+Lin- Cells",
      "Hematopoietic Stem Cell",
      "Peritoneal Macrophage",
      "Peritoneal Cavity B Cells",
      "Splenic B Cells",
      "CD4+ T cells"
    ) ~ "Blood/Immune",
    mouse_tissue_disease %in% c(
      "Kidney",
      "immortal kidney epithelium"
    ) ~ "Kidney",
    mouse_tissue_disease == "Liver" ~ "Liver",
    mouse_tissue_disease == "Spleen" ~ "Spleen",
    mouse_tissue_disease == "Ovary" ~ "Ovary",
    mouse_tissue_disease == "Colon" ~ "Colon",
    mouse_tissue_disease == "Skin" ~ "Skin",
    mouse_tissue_disease == "Heart" ~ "Heart",
    mouse_tissue_disease == "Diaphragm" ~ "Diaphragm",
    mouse_tissue_disease %in% c(
      "MUSCLE, QUADRICEPS FEMORIS",
      "skeletal muscle",
      "Calf"
    ) ~ "Muscle",
    mouse_tissue_disease == "Organoid" ~ "Organoid",
    TRUE ~ "Other"
  ),
 #Group diseases
disease_group = case_when(
  disease %in% c(
    "Cancer",
    "Adenocarcinoma",
    "Ovarian Cancer",
    "Prostate Cancer",
    "Brain Tumor",
    "Alveolar Soft Part Sarcoma (ASPS)",
    "Mucosa-Associated Lymphoid Tissue Lymphoma (MALT lymphoma)"
  ) ~ "Cancer",

  disease %in% c(
    "E.coli Infection",
    "Influenza",
    "Leptospiral Infection",
    "Staphylococcal Infection (Staph Infection)",
    "Sepsis"
  ) ~ "Infectious",

  disease %in% c(
    "Lupus",
    "Rheumatoid Arthritis (RA)",
    "Atopic Dermatitis",
    "Inflammatory Bowel Diseases (IBD)"
  ) ~ "Autoimmune/Inflammatory",

  disease %in% c(
    "Autism",
    "Epilepsy-Autism",
    "Down Syndrome",
    "Huntington"
  ) ~ "Neurological",

  disease %in% c(
    "Duchenne Muscular Dystrophy (DMD)",
    "Myotonic Dystrophy"
  ) ~ "Muscular Dystrophy",

  disease %in% c(
    "Cystic Fibrosis",
    "Pulmonary Fibrosis"
  ) ~ "Pulmonary Disease",

  disease %in% c(
    "Alpha-1 Antitrypsin Deficiancy",
    "Gaucher's Disease (Gaucher)"
  ) ~ "Genetic / Metabolic",

  disease %in% c(
    "Burns",
    "Injury"
  ) ~ "Injury / Trauma",

  TRUE ~ "Other"
),
disease_short = case_when(
  disease == "Adenocarcinoma" ~ "Adenoca.",
  disease == "Cancer" ~ "Cancer",
  disease == "Ovarian Cancer" ~ "Ov. cancer",
  disease == "Prostate Cancer" ~ "Pros. cancer",
  disease == "Brain Tumor" ~ "Brain tumor",
  disease == "Alveolar Soft Part Sarcoma (ASPS)" ~ "ASPS",
  disease == "Mucosa-Associated Lymphoid Tissue Lymphoma (MALT lymphoma)" ~ "MALT lymphoma",

  disease == "E.coli Infection" ~ "E. coli inf.",
  disease == "Influenza" ~ "Influenza",
  disease == "Leptospiral Infection" ~ "Leptospiral inf.",
  disease == "Staphylococcal Infection (Staph Infection)" ~ "Staph. inf.",
  disease == "Sepsis" ~ "Sepsis",

  disease == "Lupus" ~ "Lupus",
  disease == "Rheumatoid Arthritis (RA)" ~ "RA",
  disease == "Atopic Dermatitis" ~ "Atopic derm.",
  disease == "Inflammatory Bowel Diseases (IBD)" ~ "IBD",

  disease == "Autism" ~ "Autism",
  disease == "Epilepsy-Autism" ~ "Epilepsy-Autism",
  disease == "Down Syndrome" ~ "Down syn.",
  disease == "Huntington" ~ "Huntington",

  disease == "Duchenne Muscular Dystrophy (DMD)" ~ "DMD",
  disease == "Myotonic Dystrophy" ~ "Myotonic dyst.",

  disease == "Cystic Fibrosis" ~ "Cystic Fibrosis",
  disease == "Pulmonary Fibrosis" ~ "Pulm. fibrosis",

  disease == "Alpha-1 Antitrypsin Deficiancy" ~ "A1AT def.",
  disease == "Gaucher's Disease (Gaucher)" ~ "Gaucher",

  disease == "Burns" ~ "Burns",
  disease == "Injury" ~ "Injury",

  TRUE ~ disease
)
  )

AllData_V2.0_blood = read_rds(here("new_models", "AllData_FIT_training_blood.rds"))
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))


AllData_V2.0_manual = AllData_V2.0 %>% 
  select(CSP_ID:qval.MM) %>% 
  inner_join(fit_training_datasets_metadata_clean %>% 
               rename(CSP_ID = comparison_id),
             by = "CSP_ID") %>% 
  clean_names() 


```

## Plot distributions
```{r fig.height=12, fig.width=20}
AllData_V2.0_distributions_plot = AllData_V2.0_manual %>% 
  clean_names() %>% 
  pivot_longer(cols = c(fc_hs:qval_mm),
               names_to = "metric",
               values_to = "value") %>% 
  mutate(organism = if_else(str_detect(metric, "mm"),"Mouse", "Human")) %>% 
  filter(metric %in% c("fc_hs", "fc_mm")) %>% 
  mutate(metric = if_else(str_detect(metric, "fc"), "fc", metric)) %>% 
  ggplot() +
  aes(x = value, fill = organism) +
  geom_histogram( position = position_identity(), alpha = 0.5, bins = 50) +
  #Facet
  facet_nested_wrap(vars(disease_group, disease_short, paste0(mouse_organ, ", ", human_organ)),
                    scales = "free_y",
                    # axes = "all",
                    strip = strip_nested(size = "variable")) +
  scale_y_continuous(expand = c(0.01, 0.01), n.breaks = 3) +
  scale_x_continuous(limits = c(-2, 2)) +
  scale_fill_manual(values = c(colors_immune$organism))+
  scale_color_manual(values = c(colors_immune$organism)) +
  geom_vline(linetype = 2, xintercept = 0) + 
    theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0, hjust = 0), 
        panel.spacing.y = unit(0, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) +
  labs(fill = "Organism",
       color = "Organism",
       x = "Log2-fold change",
       y = "Frequency",
       title = "All samples")

AllData_V2.0_distributions_plot


#Save
AllData_V2.0_distributions_plot %>% 
  ggsave(filename = here("Figures", "FIT_Trainingdata_distributions_all.png"),
         width = 20,
         height = 12)

```


```{r fig.height=8, fig.width=10}
AllData_V2.0_blood_distributions_plot = AllData_V2.0_manual %>% 
  clean_names() %>% 
  pivot_longer(cols = c(fc_hs:qval_mm),
               names_to = "metric",
               values_to = "value") %>% 
  mutate(organism = if_else(str_detect(metric, "mm"),"Mouse", "Human")) %>% 
  filter(metric %in% c("fc_hs", "fc_mm")) %>% 
  mutate(metric = if_else(str_detect(metric, "fc"), "fc", metric)) %>% 
  ggplot() +
  aes(x = value, fill = organism) +
  geom_histogram( position = position_identity(), alpha = 0.5, bins = 50) +
  facet_wrap(~disease, 
             scale = "free_y", axes = "all",
             ncol = 1
             # strip.position="right"
             ) +
  scale_y_continuous(expand = c(0.01, 0.01), n.breaks = 3) +
  scale_x_continuous(limits = c(-2, 2)) +
    scale_fill_manual(
  values = c(
    Mouse = "#546E7AFF",
    Human = "#4DBBD5FF"))+
    scale_color_manual(
  values = c(
    Mouse = "#546E7AFF",
    Human = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0, hjust = 0), 
        panel.spacing.y = unit(0.8, "lines"),
        plot.margin = margin(0.5, 1, 0.2, 1, "cm"),
        strip.background.y = element_rect(fill = NA, color = NA)) +
  labs(fill = "Organism",
       color = "Organism",
       x = "Log2-fold change",
       y = "Frequency",
       title = "Blood samples")


#Plot all
AllData_V2.0_distributions_all = AllData_V2.0_distributions_plot + AllData_V2.0_blood_distributions_plot +
  plot_layout(widths = c(1, 0.15),
              guides = "collect",
              axes = "collect") +
  plot_annotation(theme = theme(legend.position = "top"))

#Save
AllData_V2.0_distributions_all %>% 
  ggsave(filename = here("Figures", "FIT_Trainingdata_distributions_all.png"),
         width = 10,
         height = 10)

```



##Correlation plots
```{r fig.height=12, fig.width=10}
# Calculate the correlation

AllData_V2.0_manual_Blood = AllData_V2.0_manual %>%
  clean_names() %>%
  # filter(qval_hs <= 0.05 | qval_mm <= 0.05) %>%
  filter(str_detect(human_tissue_control, "Blood"),
         str_detect(mouse_tissue_disease, "Blood")) %>% 
  mutate(
    csp_id_index = parse_number(csp_id)
  ) 

cor_data <- AllData_V2.0_manual_Blood %>% 
  group_by(disease, csp_id_index) %>%
  dplyr::summarise(
    test = list(
      cor.test(
        fc_mm,
        fc_hs,
        method = "spearman",
        use = "pairwise.complete.obs"
      )
    ),
    label_pos_x = max(fc_mm, na.rm = TRUE),
    label_pos_y = max(fc_hs, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    cor_value = map_dbl(test, ~ .x$estimate),
    p_value   = map_dbl(test, ~ .x$p.value),
    p_sig = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE ~ "ns"
    ),
    label_all = paste0("r = ", round(cor_value, 2),p_sig)
  )  %>%
  select(-test)


cor_data_degs <- AllData_V2.0_manual_Blood %>% 
  filter(qval_hs <= 0.05 & qval_mm <= 0.05) %>% 
  drop_na(fc_mm, fc_hs) %>% 
  group_by(disease, csp_id_index) %>%
  filter(n() >= 3) %>%      
  dplyr::summarise(
    test = list(
      cor.test(
        fc_mm,
        fc_hs,
        method = "spearman",
        use = "pairwise.complete.obs"
      )
    ),
    label_pos_x = max(fc_mm, na.rm = TRUE),
    label_pos_y = max(fc_hs, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    cor_value = purrr::map_dbl(test, ~ as.numeric(.x$estimate)),
    p_value   = purrr::map_dbl(test, ~ .x$p.value),
    p_sig = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE ~ "ns"
    ),
    label_degs = paste0("r = ", round(cor_value, 2), " ", p_sig)
  ) %>%
  select(-test)



cor_data
cor_data_degs

AllData_V2.0_manual_Blood_cor = AllData_V2.0_manual_Blood %>% 
  full_join(cor_data %>% select(disease, csp_id_index, label_all),
             by = join_by("disease", "csp_id_index")) %>% 
  full_join(cor_data_degs %>% select(disease, csp_id_index, label_degs),
             by = join_by("disease", "csp_id_index")) 
```


```{r fig.height=10, fig.width=16}
#Log2FC scatterplot
blood_l2fc_scatterplot = AllData_V2.0_manual_Blood_cor %>%
  filter(qval_hs <= 0.05 | qval_mm <= 0.05) %>%
  mutate(
    csp_id_index = parse_number(csp_id),
    comparison = paste("Comparison", csp_id_index)
  ) %>%

  #Plot
  ggplot() +
  aes(x = fc_mm,
      y = fc_hs) +
  
  facet_nested_wrap(~disease+paste("Comparison", csp_id_index, "\n All: ", label_all, "\n DEGs: ", label_degs), 
                    scales = "free", nest_line = T,
                    nrow = 4,
                    strip = strip_nested(size = "variable")) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             
             #Get only shared DEGs
             data = AllData_V2.0_manual_Blood_cor %>% 
               filter(qval_hs <= 0.05 & qval_mm <= 0.05)
             ) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  
  #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =8),
        axis.text.y = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "none",
        panel.spacing.y = unit(0, "lines"),
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title= "Correlation plot, Log2-Fold Change")

blood_l2fc_scatterplot

#Save
blood_l2fc_scatterplot %>%
  ggsave(filename = here("Figures", "FIT_TrainingData_blood_l2fc_scatterplot.png"),
         height = 10,
         width = 16)

```





### Annotate genes
```{r fig.height=8, fig.width=8}

## Plot correlation-----
# cov_genes_FIT_training = AllData_V2.0 %>% 
#   clean_names() %>% 
#   select(hs_entrez, mean_human = fc_hs, mean_mouse = fc_mm) %>% 
#   inner_join(HS_MM_Symbol_Entrez %>%
#                clean_names() %>%
#                select(human_symbol = hgnc_symbol,
#                       hs_entrez = human_entrez) %>% 
#                mutate(hs_entrez = as.character(hs_entrez)),
#              by = "hs_entrez") %>% 
#   arrange(human_symbol) %>% 
#   group_by(human_symbol) %>% 
#   summarize(sd_human = sd(mean_human),
#             cov = cov(mean_human, mean_mouse),
#             mean_human_time = mean(mean_human),
#             mean_mouse_time = mean(mean_mouse)) %>% 
#   arrange(-cov)
# 
# cov_genes_FIT_training %>% filter(abs(cov) > 0.001) #15891 genes from 6,373 (92.9%)
# cov_genes_FIT_training %>% filter(abs(cov) > 0.01) #11320 genes from 6,373 (50.3%)
# cov_genes_FIT_training %>% filter(abs(cov) > 0.1) #1151 genes
# cov_genes_FIT_training %>% filter(abs(cov) > 0.2) #356 genes
# cov_genes_FIT_training %>% filter(abs(cov) > 0.3) #186 genes
# 
# 
# genes_vary_human_all_FIT = cov_genes_FIT_training %>% 
#   filter(abs(mean_human_time) >= 0.1,
#         abs(mean_mouse_time) <= 0.1) %>% 
#   arrange(-abs(cov))
# 
# genes_vary_mouse_all_FIT = cov_genes_FIT_training %>% 
#   filter(abs(mean_mouse_time) >= 0.1,
#                 abs(mean_human_time) <= 0.1) %>% 
#   arrange(-abs(cov))
# 
# genes_vary_human_mouse_all_FIT = cov_genes_FIT_training %>% 
#   filter(abs(mean_human_time) >= 0.1,
#          abs(mean_mouse_time) >= 0.1) %>% 
#   arrange(-abs(cov))
# 
# genes_not_vary_human_mouse_all_FIT = cov_genes_FIT_training %>% 
#    filter(abs(mean_human_time) <= 0.1, 
#           abs(mean_mouse_time) <= 0.1) %>%  
#   arrange(-abs(cov))
# 
# cov_genes_FIT_training %>% filter(abs(mean_human_time) > 0.1) #3,962
# cov_genes_FIT_training %>% filter(abs(mean_human_time) > 0.2) #2191
# 
# cov_genes_FIT_training %>% filter(abs(mean_mouse_time) > 0.1) #3363
# cov_genes_FIT_training %>% filter(abs(mean_mouse_time) > 0.2) #1631
# 
# cov_genes_FIT_training %>% filter(abs(mean_mouse_time) < 0.1) #3,010
# cov_genes_FIT_training %>% filter(abs(mean_mouse_time) < 0.2) #4742
# 
# 
# 
# genes_vary_human_all_FIT #5,787
# genes_vary_mouse_all_FIT #1,807
# genes_not_vary_human_mouse_all_FIT #6,476
# genes_vary_human_mouse_all_FIT #2,355

# #Plot
# genes_vary_human = cov_genes_FIT_training %>% 
#   ggplot() +
#   aes(x = mean_human_time,
#       y = mean_mouse_time) +
#     geom_point(color = "#3182bd", fill = NA, pch = 21) +
#   scale_x_continuous(limits = c(-1.5, 1.5)) +
#   scale_y_continuous(limits = c(-1.5, 1.5)) +
#     gghighlight(abs(mean_human_time) >= 0.1,
#                 abs(mean_mouse_time) <= 0.1,
#               use_direct_label = FALSE, 
#               unhighlighted_params = list(alpha = 0.2, fill = NA)) +
#   geom_label_repel(aes(label = human_symbol), max.overlaps = 20, force_pull = 4, size = 3,
#                   data = cov_genes_FIT_training %>% 
#                     filter(abs(mean_human_time) > 0.1,
#                 abs(mean_mouse_time) <= 0.1) %>% 
#                   arrange(-abs(cov)) %>% 
#                   slice_head(n = 10)) +
#    theme_vaxgo() +
#   labs(title = "Which genes vary only in human?",
#        subtitle = "Top 10 genes with highest covariance. Mean cutoff > 0.1.")
# 
# genes_vary_mouse = cov_genes_FIT_training %>% 
#   ggplot() +
#   aes(x = mean_human_time,
#       y = mean_mouse_time) +
#     geom_point(color = "#90A4AEFF", fill = NA, pch = 21) +
#   scale_x_continuous(limits = c(-1.5, 1.5)) +
#   scale_y_continuous(limits = c(-1.5, 1.5)) +
#   gghighlight(abs(mean_mouse_time) >= 0.1,
#                 abs(mean_human_time) <= 0.1,
#               use_direct_label = FALSE,
#               unhighlighted_params = list(alpha = 0.2, fill = NA)) +
#   geom_label_repel(aes(label = human_symbol), max.overlaps = 20, force_pull = 4, size = 3,
#                   data = cov_genes_FIT_training %>% 
#                     filter(abs(mean_mouse_time) >= 0.1,
#                 abs(mean_human_time) <= 0.1) %>% 
#                   arrange(-abs(cov)) %>% 
#                   slice_head(n = 10)) +
# 
#   theme_vaxgo() +
#   labs(title = "Which genes vary only in mouse?",
#        subtitle = "Top 10 genes with highest covariance. Mean cutoff > 0.1") 
# 
# 
# genes_vary_mouse_human = cov_genes_FIT_training %>% 
#   ggplot() +
#   aes(x = mean_human_time,
#       y = mean_mouse_time) +
#   geom_point(color = "black", fill = NA, pch = 21) +
#   scale_x_continuous(limits = c(-1.5, 1.5)) +
#   scale_y_continuous(limits = c(-1.5, 1.5)) +
#   gghighlight(abs(mean_human_time) >= 0.1,
#                 abs(mean_mouse_time) >= 0.1,
#                             unhighlighted_params = list(alpha = 0.2, fill = NA)) +
#   geom_label_repel(aes(label = human_symbol), max.overlaps = 20, force_pull = 4, size = 3,
#                   data = cov_genes_FIT_training %>% 
#                     filter(abs(mean_human_time) >= 0.1,
#                 abs(mean_mouse_time) >= 0.1) %>% 
#                   arrange(-abs(cov)) %>% 
#                   slice_head(n = 10)) +
#   theme_vaxgo() +
#   labs(title = "Which genes vary in both human and mouse?",
#        subtitle = "Top 10 genes with highest covariance. Mean cutoff > 0.1")
# 
# genes_notvary_mouse_human = cov_genes_FIT_training %>% 
#   ggplot() +
#   aes(x = mean_human_time,
#       y = mean_mouse_time) +
#   geom_point(color = "black", fill = NA, pch = 21) +
#   scale_x_continuous(limits = c(-1.5, 1.5)) +
#   scale_y_continuous(limits = c(-1.5, 1.5)) +
#   gghighlight(abs(mean_human_time) <= 0.1,
#                 abs(mean_mouse_time) <= 0.1,
#               unhighlighted_params = list(alpha = 0.2, fill = NA)) +
#   geom_label_repel(aes(label = human_symbol), 
#                   # color = "red",
#                   max.overlaps = 20, force_pull = 4, size = 3, 
#                   data = cov_genes_FIT_training %>% 
#                     filter(abs(mean_human_time) <= 0.1, 
#                              abs(mean_mouse_time) <= 0.1) %>%  
#                     arrange(-abs(cov)) %>% 
#                     slice_head(n = 10)) +
#   theme_vaxgo() +
#   labs(title = "Which genes do not vary in both human and mouse?",
#        subtitle = "Top 10 genes with highest covariance. Mean cutoff < 0.1")
#   
# 
# #Plot all
# genes_vary_mouse + genes_vary_human + genes_vary_mouse_human + genes_notvary_mouse_human +
#   plot_layout(guides = "collect", axes = "collect", axis_titles =  "collect",
#               ncol =  2)
  
```


