---
title: "Frameworks"
output: html_document
---

#Download datasets
```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}

#Download datasets
GSEXXXX <- getGEO("GSEXXXX", GSEMatrix = TRUE) 

#Get Expression Set
eset <- GSEXXXX[[1]] #Some series may have two or more eSets listed

#Get metadata -------
GSEXXXX_metadata = eset %>% 
  pData()

#Select, clean, and standardize attributes
GSEXXXX_metadata_annotated = GSEXXXX_metadata %>% 
  select(geo_sample = "geo_accession") %>% 
  mutate(treatment = fct_relevel(treatment, "reference"))

#Save metadata
write_csv(GSEXXXX_metadata_annotated, 
          file = here("tables", "GSEXXXX_metadata_annotated.csv"))

#Get counts matrix ------
eset <- exprs(eset)

#Save as RDS
eset %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "GSE120661_exprs.rds"))

```


```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Import
GSE120661_eset = readRDS(here("tables", "GSE120661_exprs.rds"))
GSE120661_counts_not_normalized_log2 = readRDS(here("tables", "GSE120661_counts_not_normalized_log2.rds"))

#Check distribution
GSE120661_exprs_distribution = GSE120661_eset %>% 
  rownames_to_column("probes") %>% 
  pivot_longer(cols = -probes,
               names_to = "sample_id",
               values_to = "value") 

#Plot
GSE120661_exprs_distribution %>% 
  filter(sample_id == "GSM3407136") %>% 
  ggplot() +
  aes(x = value) +
  geom_histogram()
```


#### Get metadata
```{r}
#Import
GSE120661_metadata = read_csv(here("tables", "GSE120661_metadata.csv")) 

#Clean
GSE120661_metadata = GSE120661_metadata %>% 
  mutate(timepoint = as.integer(timepoint) %>% 
           as.character() %>% 
           as.factor() %>% 
           fct_relevel(., "0"),
         treatment = fct_relevel(treatment, "saline"),
         condition = paste(treatment, timepoint, tissue, sep = "_"))

#Save
GSE120661_metadata %>% saveRDS(here("tables", "GSE120661_metadata.rds"))
```

Import
```{r}
#Import
GSE120661_metadata = readRDS(here("tables", "GSE120661_metadata.rds"))
GSE120661_counts_not_normalized_log2 = read_rds(here("tables", "GSE120661_counts_not_normalized_log2.rds"))

#Check length
GSE120661_counts_not_normalized_log2 %>% rownames() %>% length()

#Select and filter
#Filter PBMC samples
GSE120661_samples_pbmc_vector = GSE120661_metadata %>% 
  filter(tissue == "PBMC") %>% 
  select(geo_sample) %>% 
  pull(geo_sample)

GSE120661_samples_pbmc = GSE120661_metadata %>% 
  filter(tissue == "PBMC")

#Select only PBMC samples counts
GSE120661_counts_not_normalized_log2 = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  select(c(GSE120661_samples_pbmc_vector)) %>% 
  as.matrix()
```

### Annotate genes
```{r}
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- GSE120661_counts_not_normalized_log2 %>% rownames()

#Get other annotations
listAttributes(ensembl_mart)

#Annotate
annot <- getBM(
  attributes = c("agilent_sureprint_g3_ge_8x60k", #Agilent microarray data (design ID 028005, retrieved from GEO)
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "agilent_sureprint_g3_ge_8x60k",
  values = gene_id,
  mart = ensembl_mart)

#Get annotations for each probe 
### Filter genes with highest median expression
GSE120661_counts_not_normalized_log2_summary = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = agilent_sureprint_g3_ge_8x60k), 
            by = "probe") 

#Print out the annotated table
GSE120661_counts_not_normalized_log2_summary
```
Stats
- 39,430 probes in the microarray,
- 39,392 that were annotated in BiomaRt (~99.90%)
- 37,983 probes had a correspondent mgi symbol (96.33% of probes in microarray)
- 31,140 probes had a correspondent entrez id (78.97% of probes in microarray)
- 30,787 probes with entrez id also had a correspondent mgi_symbol (78.08% of probes in microarray)
- 22,435 probes have an entrez gene id. As this is the input in FIT, we used this to filter, not mgi_symbol (56.89% of probes in microarray)
- 21,744 genes were filtered by the highest median (55.14% were kept from the microarray)
- 86 Dupes. The duplication is due to different mgi_symbols. Same Entrez IDs can correspond to different mgi_symbol
- 20849 probes were used (52.87%)

```{r}
# Stats
## 24009 probes in the microarray
GSE120661_counts_not_normalized_log2_summary %>% dim()
## 24009 probes with mgi_symbol
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(mgi_symbol) %>% 
  filter(mgi_symbol != "")
## 22363 probes with Entrez gene id
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(entrezgene_id) %>% 
  dim()
## 22082 of these probes with mgi_symbol also had a correspondent entrez gene id
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(mgi_symbol) %>% 
  filter(mgi_symbol != "") %>% 
  drop_na(entrezgene_id) %>% 
  dim()
## 22082 probes with entrez id also had a correspondent mgi_symbol
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(entrezgene_id) %>% 
  filter(mgi_symbol != "") %>% 
  dim()

#Get the probes with the highest median across samples for each entrez_id
GSE120661_expression_highest_median_probes = GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(entrezgene_id) %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median) %>% 
  ungroup() %>% 
  arrange(entrezgene_id, probe)  %>% 
  select(probe, entrezgene_id, mgi_symbol, median) %>% 
  distinct()

#84 Dupes. The duplication is due to different mgi_symbols. Same Entrez IDs can correspond to different mgi_symbol
GSE120661_expression_highest_median_probes %>% 
  get_dupes(entrezgene_id)


#Modify dataset
GSE120661_counts_not_normalized_log2_annotations = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(GSE120661_expression_highest_median_probes, by = "probe") %>% 
  select(colnames(GSE120661_expression_highest_median_probes), -median, everything())

#Modify expression matrix
GSE120661_counts_not_normalized_log2 = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  filter(probe %in% c(GSE120661_expression_highest_median_probes$probe)) %>% 
  column_to_rownames("probe") %>% 
  as.matrix()

#20845
GSE120661_counts_not_normalized_log2 %>% dim()


#---------------

#Save 
GSE120661_counts_not_normalized_log2 %>% 
  saveRDS(here("tables", "GSE120661_counts_not_normalized_log2.rds"))

GSE120661_counts_not_normalized_log2_annotations %>% 
  saveRDS(here("tables", "GSE120661_counts_not_normalized_log2_annotations_not_normalized_log2.rds"))

GSE120661_samples_pbmc %>% 
  mutate(sample_id = subject_id,
         subject_id = str_remove(subject_id, "_\\d+h")) %>% 
  saveRDS(here("tables", "GSE120661_samples_pbmc.rds"))
```
