---
title: "Frameworks"
output: html_document
---

#Download datasets
```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}

#Download datasets
GSEXXXX <- getGEO("GSEXXXX", GSEMatrix = TRUE) 

#Get Expression Set
eset <- GSEXXXX[[1]] #Some series may have two or more eSets listed

#Get metadata -------
GSEXXXX_metadata = eset %>% 
  pData()

#Select, clean, and standardize attributes
GSEXXXX_metadata_annotated = GSEXXXX_metadata %>% 
  select(geo_sample = "geo_accession") %>% 
  mutate(treatment = fct_relevel(treatment, "reference"))

#Save metadata
write_csv(GSEXXXX_metadata_annotated, 
          file = here("tables", "GSEXXXX_metadata_annotated.csv"))

#Get counts matrix ------
eset <- exprs(eset)

#Save as RDS
eset %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "GSE120661_exprs.rds"))

```


```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Import
GSE120661_eset = readRDS(here("tables", "GSE120661_exprs.rds"))
GSE120661_counts_not_normalized_log2 = readRDS(here("tables", "GSE120661_counts_not_normalized_log2.rds"))

#Check distribution
GSE120661_exprs_distribution = GSE120661_eset %>% 
  rownames_to_column("probes") %>% 
  pivot_longer(cols = -probes,
               names_to = "sample_id",
               values_to = "value") 

#Plot
GSE120661_exprs_distribution %>% 
  filter(sample_id == "GSM3407136") %>% 
  ggplot() +
  aes(x = value) +
  geom_histogram()
```


#### Get metadata
```{r}
#Import
GSE120661_metadata = read_csv(here("tables", "GSE120661_metadata.csv")) 

#Clean
GSE120661_metadata = GSE120661_metadata %>% 
  mutate(timepoint = as.integer(timepoint) %>% 
           as.character() %>% 
           as.factor() %>% 
           fct_relevel(., "0"),
         treatment = fct_relevel(treatment, "saline"),
         condition = paste(treatment, timepoint, tissue, sep = "_"))

#Save
GSE120661_metadata %>% saveRDS(here("tables", "GSE120661_metadata.rds"))
```

Import
```{r}
#Import
GSE120661_metadata = readRDS(here("tables", "GSE120661_metadata.rds"))
GSE120661_counts_not_normalized_log2 = read_rds(here("tables", "GSE120661_counts_not_normalized_log2.rds"))

#Check length
GSE120661_counts_not_normalized_log2 %>% rownames() %>% length()

#Select and filter
#Filter PBMC samples
GSE120661_samples_pbmc_vector = GSE120661_metadata %>% 
  filter(tissue == "PBMC") %>% 
  select(geo_sample) %>% 
  pull(geo_sample)

GSE120661_samples_pbmc = GSE120661_metadata %>% 
  filter(tissue == "PBMC")

#Select only PBMC samples counts
GSE120661_counts_not_normalized_log2 = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  select(c(GSE120661_samples_pbmc_vector)) %>% 
  as.matrix()
```

### Annotate genes
```{r}
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- GSE120661_counts_not_normalized_log2 %>% rownames()

#Get other annotations
listAttributes(ensembl_mart)

#Annotate
annot <- getBM(
  attributes = c("agilent_sureprint_g3_ge_8x60k", #Agilent microarray data (design ID 028005, retrieved from GEO)
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "agilent_sureprint_g3_ge_8x60k",
  values = gene_id,
  mart = ensembl_mart)

#Get annotations for each probe 
### Filter genes with highest median expression
GSE120661_counts_not_normalized_log2_summary = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = agilent_sureprint_g3_ge_8x60k), 
            by = "probe") 

#Print out the annotated table
GSE120661_counts_not_normalized_log2_summary
```
Stats
- 39,430 probes in the microarray,
- 39,392 that were annotated in BiomaRt (~99.90%)
- 37,983 probes had a correspondent mgi symbol (96.33% of probes in microarray)
- 31,140 probes had a correspondent entrez id (78.97% of probes in microarray)
- 30,787 probes with entrez id also had a correspondent mgi_symbol (78.08% of probes in microarray)
- 22,435 probes have an entrez gene id. As this is the input in FIT, we used this to filter, not mgi_symbol (56.89% of probes in microarray)
- 21,744 genes were filtered by the highest median (55.14% were kept from the microarray)
- 86 Dupes. The duplication is due to different mgi_symbols. Same Entrez IDs can correspond to different mgi_symbol
- 20849 probes were used (52.87%)

```{r}
# Stats
## 24009 probes in the microarray
GSE120661_counts_not_normalized_log2_summary %>% dim()
## 24009 probes with mgi_symbol
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(mgi_symbol) %>% 
  filter(mgi_symbol != "")
## 22363 probes with Entrez gene id
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(entrezgene_id) %>% 
  dim()
## 22082 of these probes with mgi_symbol also had a correspondent entrez gene id
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(mgi_symbol) %>% 
  filter(mgi_symbol != "") %>% 
  drop_na(entrezgene_id) %>% 
  dim()
## 22082 probes with entrez id also had a correspondent mgi_symbol
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(entrezgene_id) %>% 
  filter(mgi_symbol != "") %>% 
  dim()

#Get the probes with the highest median across samples for each entrez_id
GSE120661_expression_highest_median_probes = GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(entrezgene_id) %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median) %>% 
  ungroup() %>% 
  arrange(entrezgene_id, probe)  %>% 
  select(probe, entrezgene_id, mgi_symbol, median) %>% 
  distinct()

#84 Dupes. The duplication is due to different mgi_symbols. Same Entrez IDs can correspond to different mgi_symbol
GSE120661_expression_highest_median_probes %>% 
  get_dupes(entrezgene_id)


#Modify dataset
GSE120661_counts_not_normalized_log2_annotations = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(GSE120661_expression_highest_median_probes, by = "probe") %>% 
  select(colnames(GSE120661_expression_highest_median_probes), -median, everything())

#Modify expression matrix
GSE120661_counts_not_normalized_log2 = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  filter(probe %in% c(GSE120661_expression_highest_median_probes$probe)) %>% 
  column_to_rownames("probe") %>% 
  as.matrix()

#20845
GSE120661_counts_not_normalized_log2 %>% dim()


#---------------

#Save 
GSE120661_counts_not_normalized_log2 %>% 
  saveRDS(here("tables", "GSE120661_counts_not_normalized_log2.rds"))

GSE120661_counts_not_normalized_log2_annotations %>% 
  saveRDS(here("tables", "GSE120661_counts_not_normalized_log2_annotations_not_normalized_log2.rds"))

GSE120661_samples_pbmc %>% 
  mutate(sample_id = subject_id,
         subject_id = str_remove(subject_id, "_\\d+h")) %>% 
  saveRDS(here("tables", "GSE120661_samples_pbmc.rds"))
```






#BTMs correlation, intra- and inter-species

#### Samples' means
Annotate genes log2fc table with BTMs
```{r}
#Import BTMs annotation table
btm_annotation_genes =
  read_csv(file = here("tables", "btm_annotation_genes.csv"))

#Example input
results_log2_summary_annotated_probes =
  readRDS(file = here("tables", "GSE33341_mouse_counts_log2_summary_annotated_probes.rds"))

#Annotate genes by BTM
results_genes_btm_df = results_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  rename(mm_entrez = entrezgene_id) %>% 
  select(mgi_symbol, symbol = human_symbol, participant_id = subject_id, geo_sample, timepoint, status, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")

#Pivot wider
dge_btm_mean_process = results_genes_btm_df %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

#Print
dge_btm_mean_process

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("Tables", "example_dge_btm_mean_process.rds"))

```


#### Import data from different organisms
```{r}
#Genes L2FC table annotated with BTMs
example_dge_btm_mean_process_samples =
  readRDS(file = here("tables", "dge_Ecoli_btm_mean_process_samples.rds")) %>% 
   mutate(timepoint = if_else(organism == "Human", "24", timepoint)) 

#Import metadata
example_mouse_human_metadata =
  readRDS(file = here("tables", "GSE33341_ecoli_mouse_human_metadata.rds"))

#Calculate the mean log2fc by module
example_dge_btm_mean_process_sample_long = example_dge_btm_mean_process_samples %>%
  group_by(organism, participant_id, timepoint, process) %>% 
  summarize(group,
            subgroup, 
            mean_log2fc = mean(log2fc, na.rm = T)) %>% 
  distinct() %>% 
  ungroup()

example_dge_btm_mean_process_sample_long %>% 
  saveRDS(file = here("tables", "example_dge_btm_mean_process_sample_long.rds"))

#Pivot wider
example_dge_btm_mean_process_sample_wide = example_dge_btm_mean_process_sample_long %>% 
  group_by(organism, timepoint, process) %>% 
  pivot_wider(names_from = "participant_id", 
              values_from = "mean_log2fc") %>% 
  ungroup()

example_dge_btm_mean_process_sample_wide
  
```

#### Summarize interindividual correlation

```{r}
# --- 1) Mouse vs Human, same timepoint ---
#Get timepoints
timepoints_human <- unique(example_dge_btm_mean_process_sample_long %>% filter(organism == "Human") %>% pull(timepoint))
timepoints_mouse <- unique(example_dge_btm_mean_process_sample_long %>% filter(organism == "Mouse") %>% pull(timepoint))

#Get combinations
timepoint_combinations <- expand.grid(timepoint_human = timepoints_human,
                                      timepoint_mouse = timepoints_mouse,
                                      stringsAsFactors = FALSE)

timepoint_combinations

# Pearson Correlation
# --- 1) Human vs Mouse (between timepoints) ---
cor_human_mouse_timepoint <- timepoint_combinations %>%
  pmap_dfr(function(timepoint_human, timepoint_mouse) {
    
    # get subjects from the given timepoint
    human_df <- example_dge_btm_mean_process_sample_long %>%
      filter(organism == "Human", timepoint == timepoint_human) %>%
      select(participant_id, process, mean_log2fc)
    
    mouse_df <- example_dge_btm_mean_process_sample_long %>%
      filter(organism == "Mouse", timepoint == timepoint_mouse) %>%
      select(participant_id, process, mean_log2fc)
    
    # keep the common genes
    common_process <- intersect(human_df$process, mouse_df$process)
    human_df <- human_df %>% filter(process %in% common_process)
    mouse_df <- mouse_df %>% filter(process %in% common_process)
    
    # pivot wider
    human_wide <- human_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    mouse_wide <- mouse_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    
    # Get subjects id
    human_ids <- setdiff(colnames(human_wide), "process")
    mouse_ids <- setdiff(colnames(mouse_wide), "process")
    
    # Calculate correlation for every pair of organism 1 vs organism 2
    expand.grid(human = human_ids, mouse = mouse_ids, stringsAsFactors = FALSE) %>%
      mutate(
        cor_test = map2(human, mouse, ~{
          x <- human_wide[[.x]]
          y <- mouse_wide[[.y]]
          # remove NAs
          keep <- complete.cases(x, y)
          cor.test(x[keep], y[keep], method = "pearson")
        }),
        r       = map_dbl(cor_test, ~.x$estimate),
        p_value = map_dbl(cor_test, ~.x$p.value),
        ci_low  = map_dbl(cor_test, ~.x$conf.int[1]),
        ci_high = map_dbl(cor_test, ~.x$conf.int[2])
      ) %>%
      select(-cor_test) %>%
      mutate(timepoint_human = timepoint_human %>% as.numeric(),
             timepoint_mouse = timepoint_mouse %>% as.numeric())
  }) %>% 
  #Summarize
  group_by(timepoint_human, timepoint_mouse) %>%
  summarize(
    comparison = "Human vs. Mouse",
    mean = mean(r, na.rm = TRUE),
    sd   = sd(r, na.rm = TRUE),
    n      = sum(!is.na(r)),
    se     = sd / sqrt(n),
    tval   = qt(0.975, df = n - 1),
    ci_low = mean - tval * se,
    ci_high = mean + tval * se,
    .groups = "drop"
  )
#Print
cor_human_mouse_timepoint

# --- 2) Human vs Human
cor_human_human <- example_dge_btm_mean_process_sample_long %>%
  mutate(timepoint = as.numeric(timepoint)) %>% 
  filter(organism == "Human") %>%
  select(participant_id, process, timepoint, mean_log2fc) %>%
  group_by(timepoint) %>% 
  pivot_wider(
    names_from = participant_id,
    values_from = mean_log2fc
  ) %>%
  select(-process, -timepoint) %>%
  cor(use = "pairwise.complete.obs", method = "pearson") %>% 
  as.data.frame() %>%
  rownames_to_column("participant_1") %>%
  melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>% 
  filter(participant_1 != participant_2) %>%
  
  # keep unique combinations
  rowwise() %>%
  mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
  ungroup() %>%
  distinct(pair, .keep_all = TRUE) %>%
  #Summarize
  summarize(
    comparison = "Human vs. Human",
    mean = mean(correlation, na.rm = TRUE),
    sd   = sd(correlation, na.rm = TRUE),
    n    = sum(!is.na(correlation)),
    se   = sd / sqrt(n),
    tval = qt(0.975, df = n - 1),
    ci_low  = mean - tval * se,
    ci_high = mean + tval * se
  ) %>% 
  mutate(timepoint = 24,
         timepoint_human = timepoint)
  

# --- 3) Mouse vs Mouse ---
cor_mouse_mouse <- example_dge_btm_mean_process_sample_long %>%
  filter(organism == "Mouse") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, process, mean_log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = mean_log2fc) %>%
      select(-process)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Mouse vs. Mouse",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup() %>% 
  mutate(timepoint = as.numeric(timepoint), 
         timepoint_mouse = timepoint)


#Merge tables
example_samples_btm_correlation_all_timepoints_summary = cor_human_mouse_timepoint %>% 
  mutate(timepoint = timepoint_mouse) %>% 
  bind_rows(cor_human_human, cor_mouse_mouse)

#Print
example_samples_btm_correlation_all_timepoints_summary

example_samples_btm_correlation_all_timepoints_summary %>% 
  saveRDS(file = here("tables", "example_samples_btm_correlation_all_timepoints_summary.rds"))
```

Plot scatterplot
```{r fig.height=3, fig.width=3}
#Scatterplot
example_samples_btm_correlation_all_timepoints_summary_plot = example_samples_btm_correlation_all_timepoints_summary %>% 
  mutate(comparison = str_replace_all(comparison, "Mus musculus", "Mouse")) %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1,
                data = example_samples_btm_correlation_all_timepoints_summary %>% 
                  filter(comparison != "Human vs. Human")) +
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.5,
                angle = 90,
                linewidth = 1,
                data = example_samples_btm_correlation_all_timepoints_summary %>% 
                  filter(comparison == "Human vs. Human")) +
  
  
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 2, 6, 12, 24), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(0.4, 1)) +
  theme_vaxgo() +
  scale_color_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = "Correlation between organisms,\nBTMs, Example",
       x = "Hour",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 

example_samples_btm_correlation_all_timepoints_summary_plot

example_samples_btm_correlation_all_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "example_samples_btm_correlation_all_timepoints_summary_plot.png"),
         width = 3,
         height = 5)

example_samples_btm_correlation_all_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "example_samples_btm_correlation_all_timepoints_summary_plot.png"),
         width = 3,
         height = 3)

correlation_long_all_btm_mouse_human_timepoints_summary_plot
  
```



