---
title: "Frameworks"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Download datasets

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}

#Download datasets
GSEXXXX <- getGEO("GSEXXXX", GSEMatrix = TRUE) 

#Get Expression Set
eset <- GSEXXXX[[1]] #Some series may have two or more eSets listed

#Get metadata -------
GSEXXXX_metadata = eset %>% 
  pData()

#Select, clean, and standardize attributes
GSEXXXX_metadata_annotated = GSEXXXX_metadata %>% 
  select(geo_sample = "geo_accession") %>% 
  mutate(treatment = fct_relevel(treatment, "reference"))

#Save metadata
write_csv(GSEXXXX_metadata_annotated, 
          file = here("tables", "GSEXXXX_metadata_annotated.csv"))

#Get counts matrix ------
eset <- exprs(eset)

#Save as RDS
eset %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "GSE120661_exprs.rds"))

```

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Import
GSE120661_eset = readRDS(here("tables", "GSE120661_exprs.rds"))
GSE120661_counts_not_normalized_log2 = readRDS(here("tables", "GSE120661_counts_not_normalized_log2.rds"))

#Check distribution
GSE120661_exprs_distribution = GSE120661_eset %>% 
  rownames_to_column("probes") %>% 
  pivot_longer(cols = -probes,
               names_to = "sample_id",
               values_to = "value") 

#Plot
GSE120661_exprs_distribution %>% 
  filter(sample_id == "GSM3407136") %>% 
  ggplot() +
  aes(x = value) +
  geom_histogram()
```

#### Get metadata

```{r}
#Import
GSE120661_metadata = read_csv(here("tables", "GSE120661_metadata.csv")) 

#Clean
GSE120661_metadata = GSE120661_metadata %>% 
  mutate(timepoint = as.integer(timepoint) %>% 
           as.character() %>% 
           as.factor() %>% 
           fct_relevel(., "0"),
         treatment = fct_relevel(treatment, "saline"),
         condition = paste(treatment, timepoint, tissue, sep = "_"))

#Save
GSE120661_metadata %>% saveRDS(here("tables", "GSE120661_metadata.rds"))
```

Import

```{r}
#Import
GSE120661_metadata = readRDS(here("tables", "GSE120661_metadata.rds"))
GSE120661_counts_not_normalized_log2 = read_rds(here("tables", "GSE120661_counts_not_normalized_log2.rds"))

#Check length
GSE120661_counts_not_normalized_log2 %>% rownames() %>% length()

#Select and filter
#Filter PBMC samples
GSE120661_samples_pbmc_vector = GSE120661_metadata %>% 
  filter(tissue == "PBMC") %>% 
  select(geo_sample) %>% 
  pull(geo_sample)

GSE120661_samples_pbmc = GSE120661_metadata %>% 
  filter(tissue == "PBMC")

#Select only PBMC samples counts
GSE120661_counts_not_normalized_log2 = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  select(c(GSE120661_samples_pbmc_vector)) %>% 
  as.matrix()
```

### Annotate genes

```{r}
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- GSE120661_counts_not_normalized_log2 %>% rownames()

#Get other annotations
listAttributes(ensembl_mart)

#Annotate
annot <- getBM(
  attributes = c("agilent_sureprint_g3_ge_8x60k", #Agilent microarray data (design ID 028005, retrieved from GEO)
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "agilent_sureprint_g3_ge_8x60k",
  values = gene_id,
  mart = ensembl_mart)

#Get annotations for each probe 
### Filter genes with highest median expression
GSE120661_counts_not_normalized_log2_summary = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = agilent_sureprint_g3_ge_8x60k), 
            by = "probe") 

#Print out the annotated table
GSE120661_counts_not_normalized_log2_summary
```

Stats - 39,430 probes in the microarray, - 39,392 that were annotated in
BiomaRt (\~99.90%) - 37,983 probes had a correspondent mgi symbol
(96.33% of probes in microarray) - 31,140 probes had a correspondent
entrez id (78.97% of probes in microarray) - 30,787 probes with entrez
id also had a correspondent mgi_symbol (78.08% of probes in
microarray) - 22,435 probes have an entrez gene id. As this is the input
in FIT, we used this to filter, not mgi_symbol (56.89% of probes in
microarray) - 21,744 genes were filtered by the highest median (55.14%
were kept from the microarray) - 86 Dupes. The duplication is due to
different mgi_symbols. Same Entrez IDs can correspond to different
mgi_symbol - 20849 probes were used (52.87%)

```{r}
# Stats
## 24009 probes in the microarray
GSE120661_counts_not_normalized_log2_summary %>% dim()
## 24009 probes with mgi_symbol
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(mgi_symbol) %>% 
  filter(mgi_symbol != "")
## 22363 probes with Entrez gene id
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(entrezgene_id) %>% 
  dim()
## 22082 of these probes with mgi_symbol also had a correspondent entrez gene id
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(mgi_symbol) %>% 
  filter(mgi_symbol != "") %>% 
  drop_na(entrezgene_id) %>% 
  dim()
## 22082 probes with entrez id also had a correspondent mgi_symbol
GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(entrezgene_id) %>% 
  filter(mgi_symbol != "") %>% 
  dim()

#Get the probes with the highest median across samples for each entrez_id
GSE120661_expression_highest_median_probes = GSE120661_counts_not_normalized_log2_summary %>% 
  drop_na(entrezgene_id) %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median) %>% 
  ungroup() %>% 
  arrange(entrezgene_id, probe)  %>% 
  select(probe, entrezgene_id, mgi_symbol, median) %>% 
  distinct()

#84 Dupes. The duplication is due to different mgi_symbols. Same Entrez IDs can correspond to different mgi_symbol
GSE120661_expression_highest_median_probes %>% 
  get_dupes(entrezgene_id)


#Modify dataset
GSE120661_counts_not_normalized_log2_annotations = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(GSE120661_expression_highest_median_probes, by = "probe") %>% 
  select(colnames(GSE120661_expression_highest_median_probes), -median, everything())

#Modify expression matrix
GSE120661_counts_not_normalized_log2 = GSE120661_counts_not_normalized_log2 %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  filter(probe %in% c(GSE120661_expression_highest_median_probes$probe)) %>% 
  column_to_rownames("probe") %>% 
  as.matrix()

#20845
GSE120661_counts_not_normalized_log2 %>% dim()


#---------------

#Save 
GSE120661_counts_not_normalized_log2 %>% 
  saveRDS(here("tables", "GSE120661_counts_not_normalized_log2.rds"))

GSE120661_counts_not_normalized_log2_annotations %>% 
  saveRDS(here("tables", "GSE120661_counts_not_normalized_log2_annotations_not_normalized_log2.rds"))

GSE120661_samples_pbmc %>% 
  mutate(sample_id = subject_id,
         subject_id = str_remove(subject_id, "_\\d+h")) %>% 
  saveRDS(here("tables", "GSE120661_samples_pbmc.rds"))
```


# Differential gene expression

Create expression set
```{r}
#Differential gene expression ----
#Matrix
matrix_genes = human_mouse_fluad_matrix
#Filter samples
human_mouse_matrix = matrix_genes %>% 
  as.data.frame() %>% 
  select(c(human_mouse_metadata %>% 
           filter(organism == "Mouse") %>% 
             pull(sample_id))) %>% 
  as.matrix()

#Create Expression Set
eset_mouse <- ExpressionSet(
  assayData = human_mouse_matrix,
  phenoData = AnnotatedDataFrame(human_mouse_metadata %>% 
                                   column_to_rownames("sample_id") %>% 
                                   filter(organism == "Mouse")))

#Save
eset_mouse %>% saveRDS(file = here("tables", "Example_eset_mouse.rds"))
```

Import
```{r}
#Import
eset_mouse = readRDS(file = here("tables", "Example_eset_mouse.rds"))
human_mouse_fluad_matrix = eset_mouse %>% 
  exprs()

```



##Limma
```{r}
#Create model matrix
design <- model.matrix(~ condition + 0, # 0 defines the way we compare the samples
                       eset_mouse) 
colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

# Estimate the correlation between measurements made on the same subject (paired-samples)
corfit <- duplicateCorrelation(eset_mouse, 
                               design = design, 
                               block = eset_mouse$participant_id)
corfit$consensus

#Fit Limma model
fit <- lmFit(eset_mouse, design, 
             block = eset_mouse$participant_id, 
             correlation = corfit$consensus) #Use inter-subject correlation as input into the linear model fit

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Test for Fluad
cont.matrix <- makeContrasts(conditionFluad_4_PBMC - conditionFluad_0_PBMC,
                             conditionFluad_8_PBMC - conditionFluad_0_PBMC,
                             conditionFluad_24_PBMC - conditionFluad_0_PBMC,
                             conditionFluad_48_PBMC - conditionFluad_0_PBMC,
                             conditionFluad_72_PBMC - conditionFluad_0_PBMC,
                             conditionFluad_168_PBMC - conditionFluad_0_PBMC,
                             levels=design)
#Get contrasts
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- treat(fit2, lfc = 0.0)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_mouse = all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  mutate(condition = case_when(comparisonA == "conditionFluad_4_PBMC"	~ "Fluad_4_PBMC",
                               comparisonA == "conditionFluad_8_PBMC"	~ "Fluad_8_PBMC",
                               comparisonA == "conditionFluad_24_PBMC"	~ "Fluad_24_PBMC",
                               comparisonA == "conditionFluad_48_PBMC"	~ "Fluad_48_PBMC",
                               comparisonA == "conditionFluad_72_PBMC"	~ "Fluad_72_PBMC",
                               comparisonA == "conditionFluad_168_PBMC"	~ "Fluad_168_PBMC",
                               T~comparisonA)) %>% 
  select(condition, everything())
	
# #Save
all_degs_mouse %>% saveRDS(file = here("tables", "Example_GSE120661_degs_mouse.rds"))
```


## T test
```{r}
#Import table
human_mouse_example_raw = readRDS(file = here("tables", "human_mouse_fluad_raw_qn.rds"))

# Calculate Log2-Fold Change ------
human_mouse_example_raw_log2fc <- human_mouse_example_raw %>%
  filter(measure %in% c("value_raw"),
         timepoint %in% c("0", "1", "3", "7")) %>%
  select(participant_id, symbol, timepoint, value, measure, organism) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>%
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, symbol, measure) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")

#Tests
example_log2fc_avg_paired_ttest_wilcoxon <- human_mouse_example_raw_log2fc %>% 
  # Group by gene and experimental variables
  group_by(symbol, organism, timepoint, measure) %>% 
  summarize(
    # --- Basic descriptive statistics ---
    avg_log2fc = mean(log2fc, na.rm = TRUE),  # mean log2 fold change
    n_subjects = n(),                         # number of samples
    sd = sd(value, na.rm = TRUE),             # standard deviation
    se = sd / sqrt(n_subjects),               # standard error
    effect_size = avg_log2fc / sd,            # Cohen’s d-like effect size

    # --- Paired t-test ---
    pval_ttest_paired = tryCatch(
      t.test(value, day0, paired = TRUE, var.equal = FALSE)$p.value,
      error = function(e) NA_real_
    ),

    # --- Paired Wilcoxon signed-rank test ---
    pval_wilcoxon_paired = tryCatch(
      wilcox.test(value, day0, paired = TRUE)$p.value,
      error = function(e) NA_real_
    ),

    # --- Unpaired (independent) t-test ---
    pval_ttest_unpaired = tryCatch(
      t.test(value, day0, paired = FALSE, var.equal = FALSE)$p.value,
      error = function(e) NA_real_
    ),

    # --- Unpaired (independent) Wilcoxon rank-sum test ---
    pval_wilcoxon_unpaired = tryCatch(
      wilcox.test(value, day0, paired = FALSE)$p.value,
      error = function(e) NA_real_
    )
  ) %>% 
  ungroup() %>% 
  mutate(
    # --- Confidence intervals for the mean log2FC ---
    ci_low = avg_log2fc - 1.96 * se,
    ci_high = avg_log2fc + 1.96 * se,

    # --- Multiple testing correction (FDR) ---
    pval_adj_ttest_paired = p.adjust(pval_ttest_paired, method = "fdr"),
    pval_adj_wilcoxon_paired = p.adjust(pval_wilcoxon_paired, method = "fdr"),
    pval_adj_ttest_unpaired = p.adjust(pval_ttest_unpaired, method = "fdr"),
    pval_adj_wilcoxon_unpaired = p.adjust(pval_wilcoxon_unpaired, method = "fdr"),

    # --- Significance classification for paired t-test ---
    significant_ttest_paired = case_when(
      pval_adj_ttest_paired <= 0.01 ~ "***",
      pval_adj_ttest_paired <= 0.05 ~ "**",
      pval_adj_ttest_paired <= 0.10 ~ "*",
      TRUE ~ "ns"
    ),

    # --- Significance classification for paired Wilcoxon ---
    significant_wilcoxon_paired = case_when(
      pval_adj_wilcoxon_paired <= 0.01 ~ "***",
      pval_adj_wilcoxon_paired <= 0.05 ~ "**",
      pval_adj_wilcoxon_paired <= 0.10 ~ "*",
      TRUE ~ "ns"
    ),

    # --- Significance classification for unpaired t-test ---
    significant_ttest_unpaired = case_when(
      pval_adj_ttest_unpaired <= 0.01 ~ "***",
      pval_adj_ttest_unpaired <= 0.05 ~ "**",
      pval_adj_ttest_unpaired <= 0.10 ~ "*",
      TRUE ~ "ns"
    ),

    # --- Significance classification for unpaired Wilcoxon ---
    significant_wilcoxon_unpaired = case_when(
      pval_adj_wilcoxon_unpaired <= 0.01 ~ "***",
      pval_adj_wilcoxon_unpaired <= 0.05 ~ "**",
      pval_adj_wilcoxon_unpaired <= 0.10 ~ "*",
      TRUE ~ "ns"
    )
  )

#Count DEGs
example_log2fc_avg_paired_ttest_summary = example_log2fc_avg_paired_ttest_wilcoxon %>% 
    group_by(organism, timepoint, significant_ttest, significant_wilcoxon) %>% 
  select(organism, timepoint, starts_with("significant_")) %>%
  pivot_longer(
    cols = starts_with("significant_"),
    names_to = "test",
    values_to = "significance"
  ) %>%
  group_by(organism, timepoint, test, significance) %>%
  summarise(n = n(), .groups = "drop")

example_log2fc_avg_paired_ttest_summary %>% 
  filter(significance == "0.05")


```


```{r}
# Pacotes necessários
library(coin)        # Permutation test
library(boot)        # Bootstrap

# Função auxiliar: bootstrap mean difference
boot_mean_diff <- function(data, indices) {
  d <- data[indices, ]
  return(mean(d$value - d$day0, na.rm = TRUE))
}

# Tests
example_log2fc_all_tests <- human_mouse_example_raw_log2fc %>% 
  group_by(symbol, organism, timepoint, measure) %>% 
  summarize(
    # Descriptive statistics
    avg_log2fc = mean(log2fc, na.rm = TRUE),
    n_subjects = n(),
    sd = sd(value, na.rm = TRUE),
    se = sd / sqrt(n_subjects),
    effect_size_cohen_d = avg_log2fc / sd,

    # --- PAIRED TESTS ---
    pval_ttest_paired = tryCatch(
      t.test(value, day0, paired = TRUE)$p.value, error = function(e) NA_real_),

    pval_wilcoxon_paired = tryCatch(
      wilcox.test(value, day0, paired = TRUE)$p.value, error = function(e) NA_real_),

    pval_perm_paired = tryCatch(
      coin::pvalue(coin::oneway_test(value - day0 ~ 1, distribution = coin::approximate(B = 10000))),
      error = function(e) NA_real_),

    pval_boot_paired = tryCatch({
      boot_res <- boot(data.frame(value, day0), boot_mean_diff, R = 1000)
      mean(boot_res$t >= 0) # aproximação de p-valor
    }, error = function(e) NA_real_),

    # --- INDEPENDENT ---
    pval_ttest_unpaired = tryCatch(
      t.test(value, day0, paired = FALSE, var.equal = FALSE)$p.value,
      error = function(e) NA_real_),

    pval_wilcoxon_unpaired = tryCatch(
      wilcox.test(value, day0, paired = FALSE)$p.value,
      error = function(e) NA_real_),

    pval_perm_unpaired = tryCatch(
      coin::pvalue(coin::independence_test(value ~ day0, distribution = coin::approximate(B = 10000))),
      error = function(e) NA_real_),

    pval_boot_unpaired = tryCatch({
      boot_diff <- boot(data.frame(value, day0),
                        function(d, i) mean(d$value[i], na.rm = TRUE) - mean(d$day0[i], na.rm = TRUE),
                        R = 1000)
      mean(boot_diff$t >= 0)
    }, error = function(e) NA_real_),

    .groups = "drop"
  ) %>%
  ungroup() %>%
  mutate(
    ci_low = avg_log2fc - 1.96 * se,
    ci_high = avg_log2fc + 1.96 * se,
    
    # FDR adjustment
    across(starts_with("pval_"), ~ p.adjust(.x, method = "fdr"), 
           .names = "{.col}_adj")
  )


#Save
example_log2fc_avg_paired_ttest %>% 
  saveRDS(file = here("tables", "example_log2fc_avg_paired_ttest.rds"))

```






#Genes correlation, intra- and inter-species

Import

```{r}
#Import
human_mouse_fluad_raw_qn_log2fc = 
  readRDS(file = here("tables", "human_mouse_fluad_raw_qn_log2fc.rds"))
human_mouse_fluad_raw_qn_log2fc_avg =  
  readRDS(file = here("tables", "human_mouse_fluad_raw_qn_log2fc_avg.rds"))

human_mouse_fluad_metadata = readRDS(file = here("tables", "human_mouse_fluad_metadata.rds"))
human_mouse_fluad_metadata_grouped = human_mouse_fluad_metadata %>% 
  group_by(organism, vaccine, timepoint) %>% 
  summarize(n = n()) %>% 
  ungroup()


#Input -------
results_log2fc = human_mouse_fluad_raw_qn_log2fc %>% 
  filter(measure == "value_raw",
         timepoint %in% c(1, 3, 7))

results_log2fc_avg_ttest = human_mouse_fluad_raw_qn_log2fc_avg %>% 
  filter(measure == "value_raw",
         timepoint %in% c(1, 3, 7))

results_log2fc_avg = read_rds(file= here("tables", "dge_fit_mouse_human_long.rds")) %>% 
  select(symbol, vaccine, organism, timepoint, avg_log2fc = log2fc, pval = p_value, pval_adj = adj_p_val, ci_low, ci_high) %>% 
  mutate(timepoint = as.character(timepoint)) %>% 
  inner_join(human_mouse_fluad_metadata_grouped, 
             by = join_by("organism", "vaccine", "timepoint")) %>% 
  filter(timepoint %in% c(1, 3, 7))

#Metadata
results_metadata = human_mouse_fluad_metadata
```

Line plot of DEGs by timepoint

```{r fig.height=5, fig.width=3}
#Get DEGs from either organisms
human_or_mouse_degs = results_log2fc_avg %>% 
  select(symbol, organism, timepoint, pval_adj) %>% 
  pivot_wider(names_from = "organism",
              values_from = "pval_adj") %>% #14150 genes
  filter(Mouse <= 0.05) %>%  
  select(symbol, timepoint) %>% 
  distinct()

human_and_mouse_degs_summary = human_or_mouse_degs %>% 
  group_by(timepoint) %>% 
  summarize(n_genes = n(),
            test = "Limma")


#Compare with t test
human_or_mouse_degs_ttest = results_log2fc_avg_ttest %>% 
  select(symbol, organism, timepoint, pval_adj) %>% 
  pivot_wider(names_from = "organism",
              values_from = "pval_adj") %>% #14150 genes
  filter(Mouse <= 0.05) %>%  
  select(symbol, timepoint) %>% 
  distinct()

human_and_mouse_degs_summary_ttest = human_or_mouse_degs_ttest %>% 
  group_by(timepoint) %>% 
  summarize(n_genes = n(),
            test = "t-Test")

human_and_mouse_degs_summary_plot = human_and_mouse_degs_summary %>% 
  bind_rows(human_and_mouse_degs_summary_ttest) %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = n_genes,
      fill = test) +
  geom_col(width = 0.6, position = position_dodge2()) +
  scale_fill_manual(values = c("t-Test" = "gray",
                               "Limma" = "#4DBBD5FF")) +
  scale_x_continuous(breaks = c(1, 3, 7), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0,0)) +
  theme_vaxgo() +
  labs(title = "Total shared DEGs, by day",
       x = "Day",
       y = "Total genes",
       fill = "Test") +
  theme(legend.position = "top",
        axis.line.y = element_blank(),
        panel.grid.major.y = element_line(color = "gray75", linetype = 3),
        axis.ticks.y = element_line(color = "gray75")) 

human_and_mouse_degs_summary_plot

human_and_mouse_degs_summary_plot %>% 
  ggsave(filename = here("Figures", "human_and_mouse_degs_summary_plot_ttest_VS_Limma.png"),
         width = 5,
         height = 5)
```

## 1 individual vs 1 individual

```{r}
#Calculate correlation ------

#Compute correlation
correlation_long_all_mouse_human_timepoints <- results_log2fc %>%
  select(timepoint, symbol, participant_id, log2fc) %>%
  inner_join(human_or_mouse_degs,
             by = join_by("timepoint", "symbol")) %>%
  group_by(timepoint) %>%
  
  # Pivot wider and calculate correlation
  group_modify(~ {
    .x %>%
      pivot_wider(names_from = participant_id, values_from = log2fc) %>%
      select(-symbol) %>%
      cor(use = "pairwise.complete.obs", method = "spearman") %>%
      as.data.frame() %>%
      rownames_to_column("participant1") %>%
      pivot_longer(-participant1, names_to = "participant2", values_to = "correlation") %>%
      filter(participant1 != participant2)
  }) %>%
  ungroup() %>% 
  
  #Filter by organism
  inner_join(results_metadata %>% 
               select(participant_id, participant_1_organism = organism, sex_p_1 = sex) %>% 
               distinct(),
             by = join_by("participant1" == "participant_id")) %>% 
  inner_join(results_metadata %>% 
               select(participant_id, participant_2_organism = organism, sex_p_2 = sex) %>% 
               distinct(),
             by = join_by("participant2" == "participant_id")) 


# Remove duplicated lines only if the organism is the same
correlation_long_all_timepoints_same_organisms = correlation_long_all_mouse_human_timepoints %>%
  filter(participant_1_organism == participant_2_organism) %>%
  mutate(
    participant_min = pmin(participant1, participant2),
    participant_max = pmax(participant1, participant2)
  ) %>%
  # Remove duplicates (ex: A vs B e B vs A)
  distinct(timepoint, participant_min, participant_max, .keep_all = TRUE) %>%
  select(timepoint, participant1, participant2, correlation, participant_1_organism, participant_2_organism, sex_p_1, sex_p_2) %>%
  group_by(timepoint) %>%
  arrange(-correlation)


#Unite table with different organisms with table with deduplicated rows with the same organism
correlation_long_all_mouse_human_timepoints = correlation_long_all_mouse_human_timepoints %>% 
  filter(participant_1_organism != participant_2_organism) %>% 
  bind_rows(correlation_long_all_timepoints_same_organisms)

#Summarize data and save table
correlation_long_all_mouse_human_timepoints_summary = correlation_long_all_mouse_human_timepoints %>% 
  
  #Filter only female and set Mouse as reference
  filter(sex_p_1 == "Female", 
         sex_p_2 == "Female") %>% 
  mutate(sex_comparison = paste(sex_p_1, "vs.", sex_p_2),
         comparison = paste(participant_1_organism, "vs.", participant_2_organism)) %>%
  group_by(timepoint, comparison) %>% 
  arrange(timepoint, -correlation) %>% 
  mutate(correlation = round(correlation, 2),
         timepoint = paste("Day", timepoint)) %>% 
  summarize(mean = mean(correlation, na.rm = T) %>% round(2),
            median = median(correlation, na.rm = T)%>% round(2),
            min = min(correlation, na.rm = T)%>% round(2),
            max = max(correlation, na.rm = T) %>% round(2),
            sd = sd(correlation),
            n = sum(!is.na(correlation)),
            se = sd/sqrt(n),
            t_critical = qt(0.975, df = n - 1),
            ci_lower = mean - t_critical * se,
            ci_upper = mean + t_critical * se,
            .groups = "drop") %>% 
  filter(comparison != "Human vs. Mouse")

correlation_long_all_mouse_human_timepoints_summary

#Save
correlation_long_all_mouse_human_timepoints_summary %>% 
  saveRDS(file = here("tables", "example_correlation_genes_long_all_mouse_human_timepoints_summary.rds"))

```

## Bootstrap

```{r}
# This script performs cross-species correlation (Human vs. Mouse)
# using the bootstrap method (1000 iterations), sampling 5 female participants
# from each organism in each iteration, as requested.


# --- 1. BOOTSTRAP CONFIGURATION ---
N_BOOTSTRAPS = 1000   # Number of bootstrap iterations
SAMPLE_SIZE = 5       # 5 humans vs 5 mice in each iteration
TIMEPOINTS = c(1, 3, 7)

# --- 2. DATA PREPARATION AND FILTERING (Female vs. Female) ---

#Get DEGs from either organisms


# Filters raw log2fc data only for the relevant timepoints
# and joins metadata to filter only 'Female' participants.
data_filtered <- results_log2fc %>%
  filter(timepoint %in% TIMEPOINTS) %>%
  inner_join(human_or_mouse_degs, 
             by = join_by("symbol", "timepoint")) %>% 
  inner_join(results_metadata %>%
               filter(sex == "Female") %>% # Filters only Female participants
               select(participant_id), # Need to keep 'organism' for filtering in the next steps
             by = "participant_id")

human_data <- data_filtered %>% filter(organism == "Human")
mouse_data <- data_filtered %>% filter(organism == "Mouse")

# --- 3. FUNCTION FOR A SINGLE BOOTSTRAP ITERATION ---

# This function performs a single bootstrap iteration for a timepoint.
# 1. Samples 5 Human and 5 Mouse IDs (with replacement).
# 2. Filters data for sampled IDs and common symbols (genes).
# 3. Calculates the cross-correlation matrix (5x5).
# 4. Returns all 25 correlation values.
get_bootstrap_cors <- function(timepoint_val, all_human_data, all_mouse_data, n_sample) {

  # Filters data for the timepoint and identifies common symbols (genes)
  h_tp_data <- all_human_data %>% filter(timepoint == timepoint_val)
  m_tp_data <- all_mouse_data %>% filter(timepoint == timepoint_val)
  common_symbols <- intersect(h_tp_data$symbol, m_tp_data$symbol)

  # If there is no data or common symbols, returns an empty tibble
  if (length(unique(h_tp_data$participant_id)) < n_sample || 
      length(unique(m_tp_data$participant_id)) < n_sample ||
      length(common_symbols) == 0) {
    message(paste("Insufficient data for Timepoint:", timepoint_val))
    return(tibble(timepoint = timepoint_val, correlation = NA_real_))
  }

  # Samples N participant IDs (with replacement - bootstrap)
  sampled_human_pids <- sample(unique(h_tp_data$participant_id), n_sample, replace = TRUE)
  sampled_mouse_pids <- sample(unique(m_tp_data$participant_id), n_sample, replace = TRUE)

  # Prepares Human matrix (rows=symbols, columns=participants)
  h_matrix <- h_tp_data %>%
    filter(participant_id %in% sampled_human_pids, symbol %in% common_symbols) %>%
    # FIX: Ensure uniqueness of symbol/participant_id before pivoting to prevent 'duplicate row names' error
    distinct(symbol, participant_id, log2fc) %>%
    pivot_wider(names_from = participant_id, values_from = log2fc) %>%
    column_to_rownames("symbol")

  # Prepares Mouse matrix (rows=symbols, columns=participants)
  m_matrix <- m_tp_data %>%
    filter(participant_id %in% sampled_mouse_pids, symbol %in% common_symbols) %>%
    # FIX: Ensure uniqueness of symbol/participant_id before pivoting to prevent 'duplicate row names' error
    distinct(symbol, participant_id, log2fc) %>%
    pivot_wider(names_from = participant_id, values_from = log2fc) %>%
    column_to_rownames("symbol")

  # Ensures columns have unique names for correlation (necessary due to replace=TRUE)
  colnames(h_matrix) <- paste0("Human_", 1:ncol(h_matrix))
  colnames(m_matrix) <- paste0("Mouse_", 1:ncol(m_matrix))

  # Calculates the cross-correlation matrix (Spearman)
  cross_cors <- cor(h_matrix, m_matrix, use = "pairwise.complete.obs", method = "spearman")

  # Extracts the N*N correlation values and adds the timepoint
  tibble(
    timepoint = timepoint_val,
    correlation = as.vector(cross_cors)
  )
}

# --- 4. BOOTSTRAP EXECUTION ---

# Creates a dataframe of bootstrap IDs and applies the function for all timepoints
bootstrapped_human_mouse_cors <- tibble(
    bootstrap_id = 1:N_BOOTSTRAPS
) %>%
rowwise() %>%
mutate(
    # Executes the bootstrap function for each timepoint (1, 3, 7)
    cors_data = list(
        bind_rows(
            get_bootstrap_cors(1, human_data, mouse_data, SAMPLE_SIZE),
            get_bootstrap_cors(3, human_data, mouse_data, SAMPLE_SIZE),
            get_bootstrap_cors(7, human_data, mouse_data, SAMPLE_SIZE)
        ))) %>%
ungroup() %>%
unnest(cors_data)

# --- 5. BOOTSTRAP RESULTS SUMMARIZATION ---

# Calculates descriptive statistics on the distribution of correlations
# obtained by the 1000 bootstrap iterations (N_total = 1000 * 25 = 25000 observations)
correlation_bootstrap_summary <- bootstrapped_human_mouse_cors %>%
 group_by(timepoint) %>%
  summarize(
    comparison = "Bootstrap Mouse vs. Human",
    mean = mean(correlation, na.rm = T) %>% round(4),
    median = median(correlation, na.rm = T) %>% round(4),
    min = min(correlation, na.rm = T) %>% round(4),
    max = max(correlation, na.rm = T) %>% round(4),
    sd = sd(correlation, na.rm = T) %>% round(4),
    se = sd,
    ci_lower = quantile(correlation, 0.025, na.rm = TRUE) %>% round(4),
    ci_upper = quantile(correlation, 0.975, na.rm = TRUE) %>% round(4),
    .groups = "drop") %>% 
  mutate(timepoint = as.character(timepoint),
         unit = "Day")
```

```{r fig.height=5, fig.width=3}
#Unite results
correlation_long_all_mouse_human_timepoints_summary_df = correlation_long_all_mouse_human_timepoints_summary %>% 
  separate(timepoint, into = c("unit", "timepoint"), remove = T) %>% 
  filter(timepoint %in% c("1", "3", "7")) %>% 
  bind_rows(correlation_bootstrap_summary %>% 
              mutate(comparison = "Bootstrap"))

# Prints the summarized result
correlation_bootstrap_summary_plot = correlation_long_all_mouse_human_timepoints_summary_df %>% 
  filter(comparison %in% c("Bootstrap", "Mouse vs. Human")) %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_textline(aes(label = comparison), 
                data = . %>% filter(comparison == "Bootstrap"),
                size = 3, 
                vjust = -0.5,
                hjust = 0,
                linewidth = 1) +
  
  geom_textline(aes(label = comparison), 
                data = . %>% filter(comparison == "Mouse vs. Human"),
                size = 3, 
                vjust = 1.3,
                hjust = 0.8,
                linewidth = 1) +
  
    scale_color_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Bootstrap" = "#06d6a0",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  
  geom_errorbar(aes(ymin = ci_lower, 
                  ymax = ci_upper), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 1, 3, 7), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  labs(title = "Correlation between organisms,\ngenes, Fluad vaccine",
       x = "",
       y = " rho, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 

correlation_bootstrap_summary_plot %>% 
  ggsave(filename = here("Figures", "correlation_bootstrap_summary_plot.png"),
         width = 3,
         height = 5)

correlation_bootstrap_summary_plot

```

Line plot of correlation

```{r fig.height=5, fig.width=3}
#Import
correlation_long_all_mouse_human_timepoints_summary = 
  readRDS(file = here("tables", "example_correlation_genes_long_all_mouse_human_timepoints_summary.rds"))


correlation_long_all_mouse_human_timepoints_summary_df = correlation_long_all_mouse_human_timepoints_summary %>% 
  separate(timepoint, into = c("unit", "timepoint"), remove = T) %>% 
  filter(timepoint %in% c("1", "3", "7"))
  # bind_rows(correlation_bootstrap_summary %>% 
  #             mutate(comparison = "Bootstrap"))

#Plot
correlation_long_all_mouse_human_timepoints_summary_plot = correlation_long_all_mouse_human_timepoints_summary_df %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.2,
                hjust = 0.8,
                linewidth = 1,
                data = correlation_long_all_mouse_human_timepoints_summary_df %>% 
                  filter(comparison == "Human vs. Human")) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.2,
                hjust = 0.8,
                linewidth = 1,
                data = correlation_long_all_mouse_human_timepoints_summary_df %>% 
                  filter(comparison == "Mouse vs. Mouse")) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = 1,
                hjust = 0.8,
                linewidth = 1,
                data = correlation_long_all_mouse_human_timepoints_summary_df %>% 
                  filter(comparison == "Mouse vs. Human")) +
  
  # geom_textline(aes(label = comparison), 
  #               size = 3, 
  #               vjust = -0.2,
  #               hjust = 0.05,
  #               linewidth = 1,
  #               data = correlation_long_all_mouse_human_timepoints_summary_df %>% 
  #                 filter(comparison == "Bootstrap")) +
  
  geom_errorbar(aes(ymin = ci_lower, 
                  ymax = ci_upper), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 1, 3, 7), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01), 
                     # limits = c(-0.2, 0.5)
                     ) +
  theme_vaxgo() +
  scale_color_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Bootstrap" = "#06d6a0",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Mouse vs. Human" = "#4361ee",
                               "Bootstrap" = "#06d6a0",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = "Correlation between organisms,\ngenes, Fluad vaccine",
       x = "Day",
       y = " rho, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 


(human_and_mouse_degs_summary_plot + 
    theme(plot.title = element_blank())) /
(correlation_long_all_mouse_human_timepoints_summary_plot + 
   theme(plot.title = element_blank()))  +
  plot_layout(heights = c(0.3, 0.7),
              axis_titles = "collect")

correlation_long_all_mouse_human_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "correlation_long_all_mouse_human_Example_timepoints_genes_summary_plot.png"),
                         width =3,
                         height = 5)

correlation_long_all_mouse_human_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "correlation_long_all_mouse_human_Example_timepoints_genes_summary_plot_square.png"),
                         width =3,
                         height = 3)
  
```

#Immune functional analysis

BTMs correlation, intra- and inter-species

## Mean Log2FC and GSEA
Compare means and NES
Import
```{r}
#Import datasets ------
#Import DGE output
pairwise_human_fit_measures = read_rds(here("FIT_Mouse2man", "pairwise_human_fit_measures.rds"))
dge_long_genes = read_rds(file= here("tables", "dge_fit_mouse_human_long.rds"))

#Import BTMs annotation table
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))


#Clean metadata 
annotation_conditions = dge_long_genes %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct()

# Gene set enrichment analysis (GSEA) --------
#Get calculate means -----
#Annotate genes by BTM 

dge_btm_process = dge_long_genes %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(immune_set == "BTM by Modules") %>%
  filter(organism %in% c("Human", "Mouse")) %>% 
  group_by(condition, timepoint, process)

dge_btm_mean_process = dge_btm_process %>% 
  summarize(mean_log2fc = mean(log2fc),
            median_log2fc = median(log2fc),
            sd_log2fc = sd(log2fc),
            var_log2fc = var(log2fc),
            organism,
            group, subgroup) %>%
  distinct() %>% 
  ungroup()

#Print
dge_btm_mean_process
```

```{r}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = ImmuneGO_BTM_grouped_genes %>% 
  select(process, genes)

#Clean table for GSEA standard
df_genes2GSEA_human = dge_long_genes %>% 
  distinct() %>% 
  select(condition, genes = symbol, log2fold_change = log2fc)  

#Run multiple GSEA and unite means
gsea_results_immuneGO_Manual_human = autoGSEA(df = df_genes2GSEA_human,
         TERM2GENE = term2gene,
         geneset_name = "Human_ImmuneGO") %>% 
  clean_names() %>% 
  rename(process = id) %>% 
  #Unite means and GSEA
  inner_join(dge_btm_mean_process, by = join_by("process", "condition"))
  
  
#Save
gsea_results_immuneGO_Manual_human %>% 
    saveRDS(here("GSEA_ORA", "gsea_results_immuneGO_BTM_Human_Manual.rds"))
```

#### Plot distributions

```{r fig.height=12, fig.width=10}
#Pivot to wide GSEA + mean
dge_btm_gsea_mean_wide = gsea_results_immuneGO_Manual_human %>% 
  mutate(organism = case_when(str_detect(condition, "Mouse") ~ "Mouse",
                              str_detect(condition, "Human") ~ "Human",
                              T~condition)) %>% 
  select(organism, timepoint, process, group, subgroup, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process, subgroup, group, set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

#Arrange process by log2fc
process_order_log2fc = dge_btm_mean_process %>% 
  select(process, timepoint, organism, mean_log2fc, group) %>%
  filter(timepoint == 1,
         organism == "Human") %>% 
  group_by(organism, timepoint, group) %>% 
  arrange(-mean_log2fc) %>%
  pull(process) %>% 
  unique() 

dge_btm_process_ordered = dge_btm_process %>% 
  ungroup() %>% 
    mutate(group = fct_relevel(as.factor(group), c(immune_order)),
           process = fct_relevel(as.factor(process), c(process_order_log2fc)) %>% fct_rev()) 

labels_process <- dge_btm_gsea_mean_wide %>%
  ungroup() %>% 
  distinct(process, p_adjust_Human, p_adjust_Mouse) %>%
  mutate(label = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
                         "bold", "plain"),
         label_color = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
                         "#4DBBD5FF", "gray50"))  %>% 
  ungroup() %>% 
  distinct() %>% 
  mutate(process = fct_relevel(process, process_order_log2fc))
  

#Plot
dge_btm_process_distribution = dge_btm_process_ordered %>% 
  ggplot() +
  aes(y = process,
      x = log2fc,
      color = organism) +
  facet_grid(~group~paste("Day", timepoint), scales = "free", space = "free") +
  geom_vline(xintercept = 0, color = "black") +
  geom_point(pch = "|", alpha = 0.2, 
             show.legend = F,
             size = 3) +
  stat_summary(
  aes(color = organism),
  fun.data = mean_se,
  geom = "errorbar",
  width = 0,
  size = 1
) +
  stat_summary(
    aes(fill = organism), 
    fun = mean,             
    color = "black",
    geom = "point",             
    size = 2,    
    pch = 21,  
    alpha = 1,
    show.legend = T 
  ) +
  xlim(-2, 2) +
  theme_vaxgo() +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), 
        axis.ticks.y = element_line(color = "gray", alpha = 0.5),
        legend.position = "top",
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8, face = c(labels_process$label), 
                                   color = c(labels_process$label_color)),
        axis.line.y = element_blank(),
        panel.spacing.y = unit(0.3, "lines"),
        panel.grid.major.y = element_line(alpha = 0.5),
        panel.grid.major.x = element_line(alpha = 0.5)) +
  scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
  scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
    labs(color = "Organism",
       y = "",
       x = "Log2FC",
       fill = "Organism",
       title = "Mouse vs Human, BTMs genes, mean Log2FC") 

dge_btm_process_distribution 

dge_btm_process_distribution %>% 
  ggsave(filename = here("Figures", "Example_dge_btm_process_distribution.png"),
         height = 12, 
         width = 10)
  
```





#### Correlation plot (BTM, immune)
Import
```{r}
#Import results -------
gsea_results_immuneGO_Manual_human = readRDS(here("GSEA_ORA", "gsea_results_immuneGO_BTM_Human_Manual.rds"))
dge_long_genes = read_rds(file= here("tables", "dge_fit_mouse_human_long.rds"))

```


Plot correlation
```{r fig.height=5, fig.width=7}
#Mouse vs Human
#Pivot to wide GSEA + mean
dge_btm_gsea_mean_wide = gsea_results_immuneGO_Manual_human %>% 
  mutate(organism = case_when(str_detect(condition, "Mouse") ~ "Mouse",
                              str_detect(condition, "Human") ~ "Human",
                              T~condition)) %>% 
  select(organism, timepoint, process, group, subgroup, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process, subgroup, group, set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")


#Calculate correlation -----

# Means vs means
cor_labels <- dge_btm_gsea_mean_wide %>%
  select(process, timepoint, group, subgroup, Mouse = mean_log2fc_Mouse, Human = mean_log2fc_Human) %>% 
  group_by(timepoint) %>% 
  summarise(test_result = list(cor.test(x = Human, 
                                        y = Mouse, 
                                        use = "pairwise.complete.obs", 
                                        method = "spearman")), 
            .groups = "drop") %>%
  mutate(cor_value = purrr::map_dbl(test_result, "estimate"),
         p_value = purrr::map_dbl(test_result, "p.value")) %>%
  select(-test_result) %>%
  # Labels
  mutate(
    r_squared = cor_value^2,
    p_label = ifelse(p_value < 0.01,
                     "p = <0.01",
                     paste0("p = ", round(p_value, 2))),
    cor_label = paste0(
      "r = ", round(cor_value, 2),
      ", R² = ", round(r_squared, 2),
      ", ", p_label))

# NES vs NES
cor_labels_nes <- dge_btm_gsea_mean_wide %>%
  select(process, timepoint, group, subgroup, Mouse = nes_Mouse, Human = nes_Human) %>% 
  group_by(timepoint) %>% 
  summarise(test_result = list(cor.test(x = Human, 
                                        y = Mouse, 
                                        use = "pairwise.complete.obs", 
                                        method = "spearman")), 
            .groups = "drop") %>%
  mutate(cor_value = purrr::map_dbl(test_result, "estimate"),
         p_value = purrr::map_dbl(test_result, "p.value")) %>%
  select(-test_result) %>%
  # Labels
  mutate(
    r_squared = cor_value^2,
    p_label = ifelse(p_value < 0.01,
                     "p = <0.01",
                     paste0("p = ", round(p_value, 2))),
    cor_label = paste0(
      "r = ", round(cor_value, 2),
      ", R² = ", round(r_squared, 2),
      ", ", p_label))

```

Plot means
```{r fig.height=3, fig.width=9}
#Plot correlation plot (means)
dge_btm_mean_group_correlation_plot = dge_btm_gsea_mean_wide %>% 
  mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>% 
  ggplot() +
  aes(x = mean_log2fc_Mouse,
      y = mean_log2fc_Human,
      labels = process) +
    #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  facet_wrap(~timepoint, 
             # scales = "free"
             ) +
  #Plot all points
  geom_point(aes(fill = group,
                 color = group),
             size = 2,
             alpha = 0.2,
             pch = 21) +
  #Plot only significant in both
  geom_point(aes(fill = group),
             data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <= 0.25), 
             color = "black",
             size = 2,
             alpha = 1,
             pch = 21) +

  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  guides(color = F) +
  labs(fill = "Immune group",
       y = "Human (Mean Log2FC)",
       x = "Model (Mean Log2FC)") +
  ylim(-2, 2) +
  xlim(-2, 2) +
  
  #Add correlation metrics
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            fill= "white",
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  
  geom_text_repel(aes(label = process), 
                  min.segment.length = 0,
                  data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <=0.25), 
                  size = 1.5, 
                  segment.size = 0.2,  
                  lineheight = 0.8, 
                  max.overlaps = 200000) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "right",
          legend.direction = "vertical",
          legend.text = element_text(size = 8, hjust = 0)
          )

#Print
dge_btm_mean_group_correlation_plot

#Save results
dge_btm_mean_group_correlation_plot %>%
  ggsave(filename = here("Figures", "Example_dge_btm_mean_group_correlation_plot.png"),
         width = 9,
         height = 3)
```


Plot NES
```{r fig.height=3, fig.width=9}
#Plot correlation plot (NES)
dge_btm_nes_group_correlation_plot = dge_btm_gsea_mean_wide %>% 
  mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>% 
  ggplot() +
  aes(x = nes_Mouse,
      y = nes_Human,
      labels = process) +
    #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  facet_wrap(~timepoint, 
             # scales = "free"
             ) +
  #Plot all points
  geom_point(aes(fill = group,
                 color = group),
             size = 2,
             alpha = 0.2,
             pch = 21) +
  #Plot only significant in both
  geom_point(aes(fill = group),
             data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <= 0.25), 
             color = "black",
             size = 2,
             alpha = 1,
             pch = 21) +

  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  guides(color = F) +
  labs(fill = "Immune group",
       y = "Human (NES)",
       x = "Model (NES)") +
  ylim(-4, 4) +
  xlim(-4, 4) +
  
  #Add correlation metrics
  geom_label(data = cor_labels_nes,
            aes(x = 0, y = Inf, label = cor_label), 
            fill= "white",
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  
  geom_text_repel(aes(label = process), 
                  min.segment.length = 0,
                  data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <=0.25), 
                  size = 1.5, 
                  segment.size = 0.2,  
                  lineheight = 0.8, 
                  max.overlaps = 200000) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "right",
          legend.direction = "vertical",
          legend.text = element_text(size = 8, hjust = 0)
          )

#Print
dge_btm_nes_group_correlation_plot

#Save results
dge_btm_nes_group_correlation_plot %>%
  ggsave(filename = here("Figures", "Example_dge_btm_nes_group_correlation_plot.png"),
         width = 9,
         height = 3)
```

## Sample-wise analysis
Annotate genes log2fc table with BTMs
```{r}
#Import BTMs annotation table
btm_annotation_genes =
  read_csv(file = here("tables", "btm_annotation_genes.csv"))

#Example input
results_log2_summary_annotated_probes =
  readRDS(file = here("tables", "GSE33341_mouse_counts_log2_summary_annotated_probes.rds"))

#Annotate genes by BTM 
results_genes_btm_df = results_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  rename(mm_entrez = entrezgene_id) %>% 
  select(mgi_symbol, symbol = human_symbol, participant_id = subject_id, geo_sample, timepoint, status, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")

#Pivot wider
dge_btm_mean_process = results_genes_btm_df %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

#Print
dge_btm_mean_process

#Save
dge_btm_mean_process %>% 
  saveRDS(file = here("Tables", "example_dge_btm_mean_process.rds"))

```

Import data from different organisms

```{r}
#Genes L2FC table annotated with BTMs
example_dge_btm_mean_process_samples =
  readRDS(file = here("tables", "dge_Ecoli_btm_mean_process_samples.rds")) %>% 
   mutate(timepoint = if_else(organism == "Human", "24", timepoint)) 

#Import metadata
example_mouse_human_metadata =
  readRDS(file = here("tables", "GSE33341_ecoli_mouse_human_metadata.rds"))

#Calculate the mean log2fc by module
example_dge_btm_mean_process_sample_long = example_dge_btm_mean_process_samples %>%
  group_by(organism, participant_id, timepoint, process) %>% 
  summarize(group,
            subgroup, 
            mean_log2fc = mean(log2fc, na.rm = T)) %>% 
  distinct() %>% 
  ungroup()

example_dge_btm_mean_process_sample_long %>% 
  saveRDS(file = here("tables", "example_dge_btm_mean_process_sample_long.rds"))

#Pivot wider
example_dge_btm_mean_process_sample_wide = example_dge_btm_mean_process_sample_long %>% 
  group_by(organism, timepoint, process) %>% 
  pivot_wider(names_from = "participant_id", 
              values_from = "mean_log2fc") %>% 
  ungroup()

example_dge_btm_mean_process_sample_wide
  
```

### Summarize inter-individual correlation

#### 1 individual vs 1 individual

```{r}
# --- 1) Mouse vs Human, same timepoint ---
#Get timepoints
timepoints_human <- unique(example_dge_btm_mean_process_sample_long %>% filter(organism == "Human") %>% pull(timepoint))
timepoints_mouse <- unique(example_dge_btm_mean_process_sample_long %>% filter(organism == "Mouse") %>% pull(timepoint))

#Get combinations
timepoint_combinations <- expand.grid(timepoint_human = timepoints_human,
                                      timepoint_mouse = timepoints_mouse,
                                      stringsAsFactors = FALSE) %>% 
  filter(timepoint_human <= timepoint_mouse) #filter symetric pairs

timepoint_combinations

# Spearman Correlation
# --- 1) Human vs Mouse (between timepoints) ---
cor_human_mouse_timepoint <- timepoint_combinations %>%
  pmap_dfr(function(timepoint_human, timepoint_mouse) {
    
    # get subjects from the given timepoint
    human_df <- example_dge_btm_mean_process_sample_long %>%
      filter(organism == "Human", timepoint == timepoint_human) %>%
      select(participant_id, process, mean_log2fc)
    
    mouse_df <- example_dge_btm_mean_process_sample_long %>%
      filter(organism == "Mouse", timepoint == timepoint_mouse) %>%
      select(participant_id, process, mean_log2fc)
    
    # keep the common genes
    common_process <- intersect(human_df$process, mouse_df$process)
    human_df <- human_df %>% filter(process %in% common_process)
    mouse_df <- mouse_df %>% filter(process %in% common_process)
    
    # pivot wider
    human_wide <- human_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    mouse_wide <- mouse_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    
    # Get subjects id
    human_ids <- setdiff(colnames(human_wide), "process")
    mouse_ids <- setdiff(colnames(mouse_wide), "process")
    
    # Calculate correlation for every pair of organism 1 vs organism 2
    expand.grid(human = human_ids, mouse = mouse_ids, stringsAsFactors = FALSE) %>%
      mutate(
        cor_test = map2(human, mouse, ~{
          x <- human_wide[[.x]]
          y <- mouse_wide[[.y]]
          # remove NAs
          keep <- complete.cases(x, y)
          cor.test(x[keep], y[keep], method = "spearman")
        }),
        r       = map_dbl(cor_test, ~.x$estimate),
        p_value = map_dbl(cor_test, ~.x$p.value),
        ci_low  = map_dbl(cor_test, ~.x$conf.int[1]),
        ci_high = map_dbl(cor_test, ~.x$conf.int[2])
      ) %>%
      select(-cor_test) %>%
      mutate(timepoint_human = timepoint_human %>% as.numeric(),
             timepoint_mouse = timepoint_mouse %>% as.numeric())
  }) %>% 
  #Summarize
  group_by(timepoint_human, timepoint_mouse) %>%
  summarize(
    comparison = "Human vs. Mouse",
    mean = mean(r, na.rm = TRUE),
    sd   = sd(r, na.rm = TRUE),
    n      = sum(!is.na(r)),
    se     = sd / sqrt(n),
    tval   = qt(0.975, df = n - 1),
    ci_low = mean - tval * se,
    ci_high = mean + tval * se,
    .groups = "drop"
  )
#Print
cor_human_mouse_timepoint

# --- 2) Human vs Human
cor_human_human <- example_dge_btm_mean_process_sample_long %>%
  mutate(timepoint = as.numeric(timepoint)) %>% 
  filter(organism == "Human") %>%
  select(participant_id, process, timepoint, mean_log2fc) %>%
  group_by(timepoint) %>% 
  pivot_wider(
    names_from = participant_id,
    values_from = mean_log2fc
  ) %>%
  select(-process, -timepoint) %>%
  cor(use = "pairwise.complete.obs", method = "spearman") %>% 
  as.data.frame() %>%
  rownames_to_column("participant_1") %>%
  melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>% 
  filter(participant_1 != participant_2) %>%
  
  # keep unique combinations
  rowwise() %>%
  mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
  ungroup() %>%
  distinct(pair, .keep_all = TRUE) %>%
  #Summarize
  summarize(
    comparison = "Human vs. Human",
    mean = mean(correlation, na.rm = TRUE),
    sd   = sd(correlation, na.rm = TRUE),
    n    = sum(!is.na(correlation)),
    se   = sd / sqrt(n),
    tval = qt(0.975, df = n - 1),
    ci_low  = mean - tval * se,
    ci_high = mean + tval * se
  ) %>% 
  mutate(timepoint = 24,
         timepoint_human = timepoint)
  

# --- 3) Mouse vs Mouse ---
cor_mouse_mouse <- example_dge_btm_mean_process_sample_long %>%
  filter(organism == "Mouse") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, process, mean_log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = mean_log2fc) %>%
      select(-process)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "spearman")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Mouse vs. Mouse",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup() %>% 
  mutate(timepoint = as.numeric(timepoint), 
         timepoint_mouse = timepoint)


#Merge tables
example_samples_btm_correlation_all_timepoints_summary = cor_human_mouse_timepoint %>% 
  mutate(timepoint = timepoint_mouse) %>% 
  bind_rows(cor_human_human, cor_mouse_mouse)

#Print
example_samples_btm_correlation_all_timepoints_summary

example_samples_btm_correlation_all_timepoints_summary %>% 
  saveRDS(file = here("tables", "example_samples_btm_correlation_all_timepoints_summary.rds"))
```


##### Plot scatterplot

```{r fig.height=3, fig.width=3}
#Import
example_samples_btm_correlation_all_timepoints_summary = 
  readRDS(file = here("tables", "example_samples_btm_correlation_all_timepoints_summary.rds"))

#Scatterplot
example_samples_btm_correlation_all_timepoints_summary_plot = example_samples_btm_correlation_all_timepoints_summary %>% 
  mutate(comparison = str_replace_all(comparison, "Mus musculus", "Mouse")) %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1,
                data = example_samples_btm_correlation_all_timepoints_summary %>% 
                  filter(comparison != "Human vs. Human")) +
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.5,
                angle = 90,
                linewidth = 1,
                data = example_samples_btm_correlation_all_timepoints_summary %>% 
                  filter(comparison == "Human vs. Human")) +
  
  
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 2, 6, 12, 24), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(0.4, 1)) +
  theme_vaxgo() +
  scale_color_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = "Correlation between organisms,\nBTMs, Example",
       x = "Hour",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 

example_samples_btm_correlation_all_timepoints_summary_plot

example_samples_btm_correlation_all_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "example_samples_btm_correlation_all_timepoints_summary_plot.png"),
         width = 3,
         height = 5)

example_samples_btm_correlation_all_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "example_samples_btm_correlation_all_timepoints_summary_plot.png"),
         width = 3,
         height = 3)

correlation_long_all_btm_mouse_human_timepoints_summary_plot
  
```



### ssGSEA
Import data
```{r}
#Import data ---------
#Raw data
ssgsea_pre_data = 
  readRDS(file = here("tables", "human_mouse_fluad_raw_qn_log2fc.rds")) %>% 
  filter(measure == "value_raw") %>% 
  distinct() %>% 
  mutate(condition = paste0(participant_id, "_Day", timepoint))

#Get metadata
ssgsea_metadata = ssgsea_pre_data %>% 
  select(condition, participant_id, organism, timepoint) %>% distinct()

#BTMs
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))
```


Run ssGSEA 
```{r}
#ssGSEA-----
#Clean table for GSEA standard
term2gene = btm_annotation_genes %>% 
  filter(immune_set == "BTM by Modules") %>% 
  select(process, genes = symbol)

#Run GSEA for samples
df_genes2ssGSEA_human_mice = ssgsea_pre_data %>%
  select(condition, genes = symbol, log2fold_change = log2fc) %>%
  autoGSEA(df = .,
         TERM2GENE = term2gene,
         geneset_name = "Human_ImmuneGO")

df_genes2ssGSEA_human_mice %>% 
  saveRDS(file = here("GSEA_ORA", "Example_df_genes_2_ssGSEA_human_mice.rds"))

beep()
```


Import and annotate results
```{r}
#Import results
df_genes2ssGSEA_human_mice = readRDS(file = here("GSEA_ORA", "df_genes2ssGSEA_human_mice.rds"))
df_genes2ssGSEA_human_mice = df_genes2ssGSEA_human_mice$gsea %>% 
  clean_names() %>% 
  rownames_to_column("del") %>% 
  select(-del) %>% 
  rename(process = id)

#Annotate results
ssgsea_results_BTM_annotated = df_genes2ssGSEA_human_mice %>% 
  inner_join(btm_annotation %>% rename(process = id), by = "process") %>% 
  select(process, nes, p_adjust, group, condition) %>% 
  inner_join(ssgsea_metadata, by = "condition") %>% 
  distinct() %>% 
  arrange(process)
```

#### Plot ssGSEA correlations for mice
```{r fig.height=8, fig.width=8, message=FALSE, warning=FALSE}
#Plot inputs ----------
#Plot correlations for mice
ssgsea_results_BTM_annotated_wide = ssgsea_results_BTM_annotated %>% 
  filter(organism == "Mouse") %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "nes") %>%
  group_by(process, timepoint) %>% 
  fill(`TriFLU-MF59_BL_1`:`TriFLU-MF59_BL_5`, .direction = "updown") %>% 
  filter(process %in% c(labels_process %>% 
                          filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <= 0.25) %>% 
                          pull(process)))

#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Plot ------------
#All days
mouse_correlation_ggpairs_btms_gsea_alldays = ggpairs(ssgsea_results_BTM_annotated_wide ,
                                    columns = 7:11,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("All days"), 
                                    upper = list(continuous = wrap(ggally_cor, method = "spearman")),
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("0" = "black",
                                "1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc")) +
  scale_fill_manual(values = c("0" = "black",
                               "1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc"))

mouse_correlation_ggpairs_btms_gsea_alldays

mouse_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "Example_mouse_correlation_ggpairs_btms_gsea_alldays.png"),
         height = 10,
         width = 10)




#Day 1
timepoint_selected = 1

mouse_correlation_ggpairs_btms_gsea_day1 = ggpairs(ssgsea_results_BTM_annotated_wide %>% 
                                              filter(timepoint == timepoint_selected),
                                    columns = 7:11,
                                    mapping = ggplot2::aes(color = group,
                                                           fill = group),
                                    title = paste("Day", timepoint_selected), 
                                    upper = list(continuous = wrap(ggally_cor, method = "spearman", size = 2)),
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors)

mouse_correlation_ggpairs_btms_gsea_day1

mouse_correlation_ggpairs_btms_gsea_day1 %>%
  ggsave(filename = here("Figures", "Example_mouse_correlation_ggpairs_btms_gsea_day1.png"),
         height = 10,
         width = 10)


```














# Performance tests 

## ROC Curve 

Do for one direction (up and then down) Do one sided t test for mouse,
fit, and human Positive control: Human Neg control: Random permutation

Import data

```{r}
#Import DEGs table
dge_long_genes = read_rds(file= here("tables", "dge_fit_mouse_human_long.rds")) 
#ImmuneGO ontologies
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))


#All genes
genes_roc_allgenes = dge_long_genes
#Filter immune genees
genes_roc_allgenes_immunegenes = dge_long_genes %>% 
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))

  
#Only DEGs
degs_genes_human = dge_long_genes %>% 
  filter(adj_p_val <= 0.05, 
         organism == "Human") %>% 
  select(symbol, timepoint)
#All DEGs
genes_roc_DEGs_immune = dge_long_genes %>% 
  inner_join(degs_genes_human,
             by = join_by("symbol", "timepoint")) %>% 
  #Filter immune genes only
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))
#Immune DEGs
genes_roc_DEGs_immune = dge_long_genes %>% 
  inner_join(degs_genes_human,
             by = join_by("symbol", "timepoint")) %>% 
  #Filter immune genes only
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))
```

### All genes

Calculate ROC curve

```{r fig.height=4, fig.width=10}
set.seed(123)
######### Input -----
genes_df_input = genes_roc_allgenes
filename = "Fluad_All"
# filename = "Fluad_All_Immune"
# filename = "Fluad_AllDEGs_HumanDEGs"
# filename = "Fluad_AllDEGs_HumanDEGs_Immune"
```


```{r fig.height=5, fig.width=8}
######### UP -----------
#Do one sided t-test for each model
dge_long_genes_up = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = F), #onesided, p(gene) > 0 (right tail)
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj < 0.05, "up", "not_up")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj)

#Assign human log2fc direction as truth
truth_human = dge_long_genes_up %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism",
              values_from = "direction") %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#Generate comparable distributions
perfect_prediction = truth_human %>%
  mutate(prob = adj_p_val) %>%
  rename(pred = truth)

uniform_prediction = perfect_prediction %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()),
         organism = "Uniform")

#Get only up and not up genes from models
models_up = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = F), #one sided, higher
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj < 0.05, "up", "not_up")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj) %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         prob = adj_p_val) %>% 
  #Add controls
  bind_rows(
    perfect_prediction,
    uniform_prediction
    )

#Get probabilities (UP)
classified_df_probs_up = truth_human %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(symbol, timepoint, pred, prob, model = organism),
             by = join_by("symbol", "timepoint")) %>% 
  mutate(not_prob = 1 - prob,
         sum = prob + not_prob,
         truth = as.factor(truth)) %>% 
  distinct()


#Get ROC
roc_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Human", "Mouse", "Uniform", "FIT")) %>%
  select(symbol, timepoint, truth, pred, model, prob, not_prob, sum) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Up vs Non-Up")

######### DOWN -----------
#Do one sided t-test for each model
dge_long_genes_down = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = T), #onesided, p(gene) < 0 (left tail)
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj < 0.05, "down", "not_down")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj)

#Assign human log2fc direction as truth
truth_human = dge_long_genes_down %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism",
              values_from = "direction") %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#Generate comparable distributions
perfect_prediction = truth_human %>%
  #Invert probabilities
  mutate(prob = 1 - adj_p_val) %>%
  rename(pred = truth)
uniform_prediction = perfect_prediction %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()),
         organism = "Uniform")


#Get only up and not up genes from models
models_down = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = T), #one sided, less
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    #Invert probabilities
    pval_one_sided_adj = 1 - pval_one_sided_adj,
    direction = if_else(pval_one_sided_adj < 0.05, "down", "not_down")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj) %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         prob = adj_p_val) %>% 
  #Add controls
  bind_rows(perfect_prediction, uniform_prediction)

#Get probabilities (DOWN)
classified_df_probs_down = truth_human %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(symbol, timepoint, pred, prob, model = organism),
             by = join_by("symbol", "timepoint")) %>% 
  mutate(not_prob = 1 - prob,
         sum = prob + not_prob,
         truth = as.factor(truth)) %>% 
  distinct()


#Get ROC
roc_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Human", "Mouse", "Uniform", "FIT")) %>%
  select(symbol, timepoint, truth, pred, model, prob, not_prob, sum) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Down vs Non-Down")


#Plot curves ------------------
#PLOT ROC Curve
roc_curve_plot = roc_curve_up_notup %>% 
  bind_rows(roc_curve_down_notdown) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(size = 1, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_grid(~fct_rev(comparison)~paste("Day", timepoint)) + 
  scale_color_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "FIT"   = "#4361ee",
    "Human" = "#90A4AEFF",
    "Uniform" = "gray80"
    )) +
  labs(
    title = filename,
  x = "FPR",
  y = "TPR",
  color = "Model") + 
  theme(
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3)

roc_curve_plot

roc_curve_plot %>%
  ggsave(filename = here("Figures", paste0("roc_curve_plot_", filename, ".png")),
         width = 8,
         height = 5)

```






