---
title: "Genomic CRE Annotation"
subtitle: "Comparative annotation of human and mouse cis-regulatory elements"
author: "Wasim Aluísio Prates-Syed"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = FALSE
)
```

# Setup

```{r packages}
source(here::here("scripts_notebooks", "required.R"))

library(BSgenome.Hsapiens.UCSC.hg38)
library(BSgenome.Mmusculus.UCSC.mm10)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(ChIPseeker)
library(GenomeInfoDb)
library(GenomicRanges)
library(IRanges)
library(rtracklayer)
library(org.Hs.eg.db)
library(pwalign)
library(biomaRt)
```

------------------------------------------------------------------------

# Import

```{r import-data}
#Import Human and Mouse CREs ===========
#In contrast to the Human dataset, there are no CTCF-bound CREs in the main, full CRE mouse dataset. So, we imported CREs separately and merged them.

# Human CREs (GRCh38) =============

#PLS
cres_human_pls <- read_tsv(
  here("Genomic", "genomic_human_cCREs_GRCh38_PLS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_original_human = type_human,
         subtype_human = "not-CTCF") 

#pELS-dELS
cres_human_pelsdels <- read_tsv(
  here("Genomic", "genomic_human_cCREs_GRCh38_pELS-dELS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_original_human = type_human,
         subtype_human = "not-CTCF") 

#CTCF BOUND
cres_human_ctcfbound <- read_tsv(
  here("Genomic", "genomic_human_cCREs_GRCh38_CTCF-bound.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type_original"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_human")) %>% 
  mutate(type_human = str_remove(type_original_human, ",CTCF-bound"),
         subtype_human = "CTCF")

#Check overlaps and remove CTCF-bound cCREs that are not annotated accordingly =====
#PLS
cres_human_ctcfbound_pls_overlap = cres_human_ctcfbound %>% 
  inner_join(cres_human_pls,
             by = "id_ccre_human") %>% 
  select(id_ccre_human) %>% 
  distinct() %>% 
  arrange(id_ccre_human) #36,430 IDs

#47,532 -> 11,102 IDs
cres_human_pls_clean = cres_human_pls %>% 
  filter(!id_ccre_human %in% cres_human_ctcfbound_pls_overlap$id_ccre_human) 

#pELS-dELS
cres_human_ctcfbound_pelsdels_overlap = cres_human_ctcfbound %>% 
  inner_join(cres_human_pelsdels,
             by = "id_ccre_human")%>% 
  select(id_ccre_human) %>% 
  distinct() #720,306 IDs

#1,718,669 -> 998,363
cres_human_pelsdels_clean = cres_human_pelsdels %>% 
  filter(!id_ccre_human %in% cres_human_ctcfbound_pelsdels_overlap$id_ccre_human)

#Merge
cres_human = bind_rows(cres_human_pls_clean,
          cres_human_pelsdels_clean,
          cres_human_ctcfbound) S

```

```{r import-data}
# Mouse CREs (mm10) ============
#PLS
cres_mouse_pls <- read_tsv(
  here("Genomic", "genomic_mouse_cCREs_mm10_PLS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_original_mouse = type_mouse,
         subtype_mouse = "not-CTCF") 

#pELS-dELS
cres_mouse_pelsdels <- read_tsv(
  here("Genomic", "genomic_mouse_cCREs_mm10_pELS-dELS.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_original_mouse = type_mouse,
         subtype_mouse = "not-CTCF")

#CTCF BOUND
cres_mouse_ctcfbound <- read_tsv(
  here("Genomic", "genomic_mouse_cCREs_mm10_CTCF-bound.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type_original"),
  show_col_types = FALSE
) %>%
  rename_with(~paste0(., "_mouse")) %>% 
  mutate(type_mouse = str_remove(type_original_mouse, ",CTCF-bound"),
         subtype_mouse = "CTCF")

#Check overlaps and remove CTCF-bound cCREs that are not annotated accordingly =====
#PLS
cres_mouse_ctcfbound_pls_overlap = cres_mouse_ctcfbound %>% 
  inner_join(cres_mouse_pls,
             by = "id_ccre_mouse") %>% 
  select(id_ccre_mouse) %>% 
  distinct() %>% 
  arrange(id_ccre_mouse) #11.884 IDs

#27.332 -> 15,448 IDs
cres_mouse_pls_clean = cres_mouse_pls %>% 
  filter(!id_ccre_mouse %in% cres_mouse_ctcfbound_pls_overlap$id_ccre_mouse) 

#pELS-dELS
cres_mouse_ctcfbound_pelsdels_overlap = cres_mouse_ctcfbound %>% 
  inner_join(cres_mouse_pelsdels,
             by = "id_ccre_mouse")%>% 
  select(id_ccre_mouse) %>% 
  distinct() #66.186 IDs

#512.001 -> 445,815 
cres_mouse_pelsdels_clean = cres_mouse_pelsdels %>% 
  filter(!id_ccre_mouse %in% cres_mouse_ctcfbound_pelsdels_overlap$id_ccre_mouse)

#Merge
cres_mouse = bind_rows(cres_mouse_pls_clean,
          cres_mouse_pelsdels_clean,
          cres_mouse_ctcfbound)

```

Import homologous CREs

```{r}
# Homologous CRE pairs from ENCODE (hg38–mm10) Nearest gene, protein coding only
homologous_cres <- read_tsv(
  here("Genomic", "genomic_human_mouse_cCREs_homologous_GRCh38_Mm10.tsv"),
  col_names = c("human_enhancer", "mouse_enhancer"),
  show_col_types = FALSE
) #537,822

# Load ENCODE comprehensive gene-CRE association table
genes_cres_human <- read_tsv(
  here("Genomic", "genomic_human_cCREs_GRCh38_Closest-Genes-PC.tsv"),
  col_names = c("chrom_cre", "start_cre", "end_cre", "id_dhs", "id_ccre", "type_cre",
                "chrom_gene", "start_gene", "end_gene", "transcript_id", "score_gene", "strand_gene", "gene_id"),
  show_col_types = FALSE
) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) #2,651,787
```
Annotate with GeneHancer
```{r}
########### GENEHANCER

#Gene interactions with enhancers
genehancer_interactions_doubleelite = read_tsv(here("Genomic", "GeneHancer_Chr1-22_ChrX_Interactions_DoubleElite_hgTables.txt")) %>% 
  separate(name, into = c("hgnc_symbol", "genehancer_id")) %>%
  filter(chromEnd >= chromStart) %>% 
  rename(geneScore = score,
         geneValue = value,)
                                               
#Enhancers                                              
genehancer_enhancers_doubleelite = read_tsv(here("Genomic", "GeneHancer_Chr1-22_ChrX_Enhancers_DoubleElite_hgTables.txt")) %>%
  filter(chromEnd >= chromStart)


########## Get overlap between ENCODE4 and Genehancer
library(GenomicRanges)
#Convert to GRanges
genehancer_enhancers_doubleelite_ranges = makeGRangesFromDataFrame(genehancer_enhancers_doubleelite, 
                                  keep.extra.columns = TRUE,
                                  seqnames.field = "chrom",
                                  start.field = "chromStart",
                                  end.field = "chromEnd", 
                                  na.rm = T)

#Convert to GRanges
cres_human_ranges = makeGRangesFromDataFrame(cres_human, 
                                  keep.extra.columns = TRUE,
                                  seqnames.field = "chrom_human",
                                  start.field = "start_human",
                                  end.field = "end_human", 
                                  na.rm = T)
# Find overlaps
distances_ranges <- distanceToNearest(genehancer_enhancers_doubleelite_ranges, 
                     cres_human_ranges) %>% 
  as.data.frame()

#Annotate Genehancer ids (ELITE) with ENCODE4 ids
cres_human_encode_genehancer <- data.frame(
  genehancer_enhancers_doubleelite_ranges[distances_ranges$queryHits], 
  cres_human_ranges[distances_ranges$subjectHits],
  distance = distances_ranges$distance
) %>% 
  rename_with(
    ~ str_replace(., "\\.1$", "_encode4"), 
    ends_with(".1")                        
  ) %>% 
  rename(chrom = seqnames,
         chromStart = start,
         chromEnd = end,
         genehancer_id = name) %>%
  
  #Merge gene interaction scores
  inner_join(genehancer_interactions_doubleelite,
             by = join_by("genehancer_id", "chrom", "chromStart", "chromEnd")
             )


cres_human_encode_genehancer %>% 
  glimpse()
```




Summarize data Count how many mouse CREs each human CRE maps to (and vice versa)

```{r classify-match-type}

#Summarize classes
cres_human_encode_genehancer %>% 
  dplyr::count(type_human, subtype_human) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_human, desc(n))

cres_mouse %>% 
  dplyr::count(type_mouse, subtype_mouse) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(type_mouse, desc(n))

# Count how many mouse CREs each human CRE maps to (and vice versa)
human_counts <- homologous_cres %>%
  dplyr::count(human_enhancer, name = "n_mouse_per_human")

mouse_counts <- homologous_cres %>%
  dplyr::count(mouse_enhancer, name = "n_human_per_mouse")

# Full join: preserves orphan CREs on both sides
cres_annotated <- homologous_cres %>%
  left_join(human_counts, by = "human_enhancer") %>%
  left_join(mouse_counts, by = "mouse_enhancer") %>%
  full_join(cres_human_encode_genehancer, by = join_by(human_enhancer == id_ccre_human)) %>%
  full_join(cres_mouse, by = join_by(mouse_enhancer == id_ccre_mouse)) %>%
  mutate(
    is_homologous = !is.na(human_enhancer) & !is.na(mouse_enhancer),
    match_type = case_when(
      is.na(human_enhancer)                            ~ "mouse_only",
      is.na(mouse_enhancer)                            ~ "human_only",
      n_mouse_per_human == 1 & n_human_per_mouse == 1  ~ "1:1",
      n_mouse_per_human >  1 & n_human_per_mouse == 1  ~ "N:1",
      n_mouse_per_human == 1 & n_human_per_mouse >  1  ~ "1:N",
      TRUE                                             ~ "N:M"
    )
  ) %>% 
  rename(chrom_human = chrom,
         start_human = chromStart,
         end_human = chromEnd,
         )

#Filter 1 to 1
cres_homologous_1to1 <- cres_annotated %>% #856,415
  dplyr::filter(match_type == "1:1") %>% #277,898
  #Filter enhancers with complete annotation in both species
  dplyr::filter(!is.na(chrom_human), !is.na(chrom_mouse))#1,567

# Separate tables
cres_homologous <- cres_annotated %>% dplyr::filter(is_homologous)
cres_human_only <- cres_annotated %>% dplyr::filter(match_type == "human_only")
cres_mouse_only <- cres_annotated %>% dplyr::filter(match_type == "mouse_only")
```

Visualize

```{r classify-match-type, fig.height=5, fig.width=6}
#Summarize
cres_homologous_1to1$type_human %>% unique()
#Compare classes of CREs between species
cres_homologous_1to1_comparison = cres_homologous_1to1 %>% 
  
  #Filter PLS, pELS, dELS
    filter(type_human %in% c("pELS", "dELS", "PLS"),
         type_mouse %in% c("pELS", "dELS", "PLS")) %>% 
  
  #Summarize
  group_by(type_human, subtype_human) %>% 
  dplyr::count(type_human, subtype_human, type_mouse, subtype_mouse) %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(desc(n)) %>%
  mutate(type_human = fct_relevel(type_human, "PLS", "pELS", "dELS"),
         type_mouse = fct_relevel(type_mouse, "PLS", "pELS", "dELS")
         )
```

```{r classify-match-type, fig.height=5, fig.width=5}
#Heatmap

#Elite CREs
cres_homologous %>% 
    dplyr::filter(!is.na(hgnc_symbol)) %>% #1,706
  #Filter ELITE enhancers
  dplyr::filter(eliteness == "Elite") %>%  ##1,567
  #Plot
  ggplot() +
    aes(x = subtype_mouse %>% str_replace(., "-", "\n"),
         y = subtype_human %>% str_replace(., "-", "\n")) +
  facet_grid(~type_human~type_mouse) +
  geom_tile(aes(fill = pct), color = "black", linewidth = 0.4) +
  geom_text(aes(label = paste0(pct, "%")),
            size = 3) +
  theme_vaxgo() +
  scale_fill_gradient2(low = "white",
                       high = "#4361ee",
                       guide = guide_colorbar(barwidth = unit(2, "cm"))) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  theme(axis.text.x = element_text(angle = 0,
                                   # vjust = 0,
                                   hjust = 0.5),
        legend.position = "top") +
  labs(x = "Mouse",
       y = "Human",
       fill = "%Shared, from human")


# All CREs
cres_homologous %>% 
  #Plot
  ggplot() +
    aes(x = subtype_mouse %>% str_replace(., "-", "\n"),
         y = subtype_human %>% str_replace(., "-", "\n")) +
  facet_grid(~type_human~type_mouse) +
  geom_tile(aes(fill = pct), color = "black", linewidth = 0.4) +
  geom_text(aes(label = paste0(pct, "%")),
            size = 3) +
  theme_vaxgo() +
  scale_fill_gradient2(low = "white",
                       high = "#4361ee",
                       guide = guide_colorbar(barwidth = unit(2, "cm"))) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  theme(axis.text.x = element_text(angle = 0,
                                   # vjust = 0,
                                   hjust = 0.5),
        legend.position = "top") +
  labs(x = "Mouse",
       y = "Human",
       fill = "%Shared, from human")


```

------------------------------------------------------------------------

# Gene annotation

```{r import-gene-annotation}

###############
# ADAPT
###############

# Import gene expression correlation
# Mock
expression_correlation <- tibble(
  hgnc_symbol = c("ENTPD1", "ASGR1", "MARCKS", "IL7R"),
  mgi_symbol  = c("Entpd1", "Asgr1", "Marcks", "Il7r"),
  rho         = c(0.85, 0.79, -0.42, -0.61)
)

all_human_mouse_dge_btm_process_genes_correlation_bygene <- readRDS( here("tables", "all_human_mouse_dge_btm_process_genes_correlation_bygene.rds"))

############## 
# Query biomaRt to map human Ensembl IDs to HGNC and MGI symbols
human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

hgnc_mgi_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = unique(cres_human_encode_genehancer$gene_id_clean),
  mart       = human_mart
)

# Join HGNC symbols to gene table
genes_cres_human <- genes_cres_human %>%
  left_join(hgnc_mgi_map, by = c("gene_id_clean" = "ensembl_gene_id"))


# Compute multi-CRE metrics per homologous gene pair
gene_level_cre_metrics <- genes_cres_human %>%
  group_by(hgnc_symbol, type_cre) %>%
  summarise(
    n_cres_human       = n(),
    n_cres_orthologous = sum(id_ccre %in% cres_homologous_1to1$human_enhancer),
    pct_cres_conserved = n_cres_orthologous / n_cres_human,
    .groups = "drop"
  ) 
  # inner_join(all_human_mouse_dge_btm_process_genes_correlation_bygene, by = c("hgnc_symbol" = "hgnc_symbol"))


# Annotate cres_annotated with human gene info (one gene per CRE)
cres_annotated <- cres_annotated %>%
  left_join(
    genes_cres_human %>%
      dplyr::select(id_ccre, gene_id, hgnc_symbol,                     transcript_id, chrom_gene, start_gene, end_gene, strand_gene) %>%
      distinct(id_ccre, .keep_all = TRUE),
    by = c("human_enhancer" = "id_ccre")
  )

```

How many CREs per gene? And how many genes per enhancer?
```{r fig.height=5, fig.width=10}
#Distribution
gene_level_cre_metrics %>% 
  filter(hgnc_symbol != "") %>% 
  ggplot() +
  aes(x = n_cres_human,
      y = "Human",
      label = hgnc_symbol) +
  geom_violin(fill = "gray90",
              color = NA) +
  geom_boxplot(width = 0.2,
               color = "#4DBBD5FF",
               alpha = 0) +
  scale_x_continuous(transform = "log10") +
  geom_quasirandom(groupOnX = FALSE, # Permite a distribuição vertical em um X contínuo
                 varwidth = TRUE,  # Ajusta a largura de acordo com a densidade
                 pch = 21,
                 mapping = aes(fill = pct_cres_conserved)) +
  scale_color_continuous() +
  geom_label_repel(size = 2.5,
                  data = . %>% slice_max(n = 5, 
                                         order_by = n_cres_human)) +
  theme_vaxgo()

```


```{r fig.height=4, fig.width=10}

library(ggpubr)
#Is there any correlation between CREs metrics and types of CREs?
gene_level_cre_metrics %>% 
  inner_join(all_human_mouse_dge_btm_process_genes_correlation_bygene %>% 
               rename(hgnc_symbol = symbol), 
             by = join_by(hgnc_symbol)) %>% 
  filter(hgnc_symbol != "",
         # p_value <= 0.10,
         n_pathogens > 10,
         type_cre %in% c("PLS", "pELS", "dELS", "TF")) %>% 
  ggplot() +
  aes(
      x = n_cres_orthologous,
      # x = n_cres_human,
      # x = pct_cres_conserved,
      y = cor_value, 
      label = hgnc_symbol,
      color = type_cre,
      # size = n_pathogens
      ) +
  theme_vaxgo() +
  facet_wrap(~type_cre, nrow = 1, 
             scales = "free_x")+
  geom_point(alpha = 0.2,
             ) +
  geom_smooth(method = "lm", color = "black", se = TRUE, size = 0.8) +
  
  stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top", color = "black") +
  theme(legend.position = "top") 
  
  # geom_label_repel(color = "black",
  #                 size = 2.5,
  #                 data = . %>% 
  #                   slice_max(pct_cres_conserved, n = 10))


```



------------------------------------------------------------------------

# Sequence extraction

The HOXD55 matrix (originally derived by Chiaromonte et al., 2002, and deeply analyzed in Blanchette et al., 2004) defines nucleotide matches and mismatches using large integers to maximize sensitivity when calculating neutral substitution rates in mammalian alignments:Match: $+91$Mismatch: $-114$Because a single mismatch penalizes the score by $-114$, the gap penalties must be scaled proportionally. If you used a standard gap opening penalty of $-10$, the alignment algorithm would erroneously introduce hundreds of micro-gaps instead of accepting a single nucleotide mismatch, completely ruining the biological alignment of your cis-regulatory elements (CREs).

To maintain perfect parity with the alignments you see on the UCSC Genome Browser for human (hg38) and mouse (mm10), you use their default parameters:Gap Opening Penalty ($O$): $-400$Gap Extension Penalty ($E$): $-30$

Chiaromonte, F., et al. (2002). Scoring pairwise genomic alignments. Pacific Symposium on Biocomputing. This paper established the ancestral mammalian repeat substitution matrices (including the HOXD matrices). Blanchette, M., et al. (2004). Blastz: A tool for aligning a large number of long genomic sequences. Genome Research. This paper formally documents the open-source BLASTZ/LASTZ defaults ($O = 400, E = 30$, match $= 91$, mismatch $= 114$) used by UCSC. Kent, W. J., et al. (2003). Evolution's Cauldron: Duplication, Deletion, and Rearrangement in the Mouse and Human Genomes. PNAS. Documents how UCSC constructs the chain and net alignments utilizing these exact parameters.

Align homolog enhancer sequences

```{r extract-sequences}
#Filter homologs with mapped coordinates
cres_to_align_homologous <- cres_homologous_1to1 %>%
  filter(human_enhancer %in% genes_cres_human$id_ccre) %>%
  filter(!is.na(chrom_human), !is.na(chrom_mouse)) %>%
  #Annotate gene symbol
  inner_join(cres_annotated %>% 
            select(human_enhancer, mouse_enhancer, hgnc_symbol),
        by = join_by(human_enhancer, mouse_enhancer))

#Convert to GRanges objects
gr_human <- GRanges(seqnames = cres_to_align_homologous$chrom_human, ranges = IRanges(start = cres_to_align_homologous$start_human, end = cres_to_align_homologous$end_human))
gr_mouse <- GRanges(seqnames = cres_to_align_homologous$chrom_mouse, ranges = IRanges(start = cres_to_align_homologous$start_mouse, end = cres_to_align_homologous$end_mouse))

#Get sequences
cres_to_align_homologous$sequence_human <- as.character(getSeq(BSgenome.Hsapiens.UCSC.hg38, gr_human))
cres_to_align_homologous$sequence_mouse <- as.character(getSeq(BSgenome.Mmusculus.UCSC.mm10, gr_mouse))

#Clean sequences from undefined characters
cres_to_align_homologous_clean = cres_to_align_homologous %>% 
  mutate(sequence_human = str_remove_all(sequence_human, "N"),
         sequence_mouse = str_remove_all(sequence_mouse, "N"))

# Parameterize a nucleotide substitution matrix mirroring the standard HOXD55 values
matriz_hoxd55 <- nucleotideSubstitutionMatrix(match = 91, mismatch = -114, baseOnly = TRUE)

#Align sequences
aln_cres <- pairwiseAlignment(
  DNAStringSet(cres_to_align_homologous_clean$sequence_human),
  DNAStringSet(cres_to_align_homologous_clean$sequence_mouse),
  type         = "local",
  substitutionMatrix = matriz_hoxd55,
  gapOpening   = -400, # Score scale adjusted for standard UCSC genomic alignment scoring
  gapExtension = -30
)

#Calculate percent identity
cres_to_align_homologous_identity <- cres_to_align_homologous_clean %>%
  mutate(
    cre_identity_pct = pid(aln_cres, type = "PID1"),
    cre_align_score  = score(aln_cres),
    cre_align_length = nchar(alignedPattern(aln_cres)),
    cre_coverage_human = cre_align_length / nchar(sequence_human)
  )



#Save results
cres_to_align_homologous_identity %>% 
  saveRDS(file = here("tables", "cres_homologous_1to1_annotated_aligned.rds"))

#Summarize alignment results
summary(cres_to_align$cre_identity_pct)

cat("\n% identity > 80%:", round(mean(cres_to_align$cre_identity_pct > 80) * 100, 1), "%\n")
cat("% identity > 70%:", round(mean(cres_to_align$cre_identity_pct > 70) * 100, 1), "%\n")
cat("% identity > 60%:", round(mean(cres_to_align$cre_identity_pct > 60) * 100, 1), "%\n")

#Which genes have the most conserved enhancers?
cres_to_align %>% 
  select(hgnc_symbol, human_enhancer, mouse_enhancer, type_human, type_mouse, cre_identity_pct:cre_coverage_human) %>% 
  group_by(hgnc_symbol, type_human, type_mouse) %>% 
  summarise(median_pct_id = median(cre_identity_pct)) %>%
  arrange(-median_pct_id) 

```

# Alignment of proximal TSS windows

```{r}
tx_human <- transcripts(TxDb.Hsapiens.UCSC.hg38.knownGene)

# Multi-scale biological window definitions centered on the TSS
windows_definition <- list(
  w3000 = list(up = 3000, down = 300),  # Full proximal regulatory landscape (pELS/dELS capture)
  w1500 = list(up = 1500, down = 300),  # Core promoter + immediate proximal enhancers
  w500  = list(up = 500,  down = 100),  # Core promoter locus only
  w200  = list(up = 200,  down = 50)    # Minimal transcription initiation footprint
)

# Exemplary extraction and global alignment (Needleman-Wunsch) loop initialization for w3000
tss_human_w3000 <- promoters(TxDb.Hsapiens.UCSC.hg38.knownGene, upstream = 3000, downstream = 300)
tss_mouse_w3000 <- promoters(TxDb.Mmusculus.UCSC.mm10.knownGene, upstream = 3000, downstream = 300)

# Implementation Note: Filter your GRanges objects by matched homologous pairs prior to running global alignment:
aln_global <- pairwiseAlignment(seq_human_gene, seq_mouse_gene, type = "global", substitutionMatrix = matriz_hoxd55)
```

# Annotate conservation and TEs

```{r}
ah <- AnnotationHub()

# 1. Fetch the most recent RepeatMasker records for hg38
rmsk_query <- query(ah, c("RepeatMasker", "Homo sapiens", "hg38"))
rmsk_hg38  <- ah[[names(rmsk_query)[1]]]

# 2. Compute the exact fraction of the CRE masked by repetitive components
gr_cres_human_all <- GRanges(
  seqnames = cres_to_align$chrom_human, 
  ranges = IRanges(start = cres_to_align$start_human, end = cres_to_align$end_human)
)

hits_rmsk <- findOverlaps(gr_cres_human_all, rmsk_hg38)

# Calculate coverage to map base-pair level masking within the ranges
repeat_cov <- GenomicRanges::coverage(rmsk_hg38)
seqlevels(gr_cres_human_all, pruning.mode = "coercion") <- names(repeat_cov)
masked_views <- Views(repeat_cov, ranges(gr_cres_human_all))

cres_to_align$repeat_masked_pct <- sapply(masked_views, function(x) sum(as.numeric(x) > 0)) / width(gr_cres_human_all) * 100
```

#Integrative Statistical Modeling (Linear Mixed-Effects Models)

Because individual homologous genes map to multiple independent flanking CREs throughout the genome, we employ a Linear Mixed-Effects Model (LMM). Treating the gene pair as a random effect handles the nested data dependency and protects against degrees-of-freedom inflation.

```{r}
# Consolidate feature matrix across all annotation scales
df_integrated <- cres_to_align %>%
  select(human_enhancer, mouse_enhancer, cre_identity_pct, repeat_masked_pct) %>%
  left_join(genes_cres_human, by = join_by(human_enhancer == id_ccre)) %>%
  left_join(gene_level_cre_metrics, by = "hgnc_symbol") %>%
  mutate(gene_pair_id = paste0(hgnc_symbol, "_", mmusculus_homolog_associated_gene_name)) %>%
  filter(!is.na(rho))

# Fit Linear Mixed-Effects Model
lmm_fit <- lmer(
  rho ~ cre_identity_pct + pct_cres_conserved + repeat_masked_pct + (1 | gene_pair_id),
  data = df_integrated
)

# Extract clean model coefficients and variance components
tidy(lmm_fit)
```

# Export

```{r export}
# Merge alignment results back into cres_annotated
cres_annotated <- cres_annotated %>%
  left_join(
    cres_to_align %>%
      dplyr::select(human_enhancer, mouse_enhancer,
                    sequence_human, sequence_mouse,
                    cre_identity_pct, cre_align_score, cre_align_length),
    by = c("human_enhancer", "mouse_enhancer")
  )

# Update split tables
# cres_homologous <- cres_annotated %>% filter(is_homologous)
# cres_human_only <- cres_annotated %>% filter(match_type == "human_only")
# cres_mouse_only <- cres_annotated %>% filter(match_type == "mouse_only")

# Save
saveRDS(cres_annotated,  here("tables", "cres_annotated_full.rds"))
# saveRDS(cres_homologous, here("tables", "cres_homologous_annotated.rds"))
# saveRDS(cres_human_only, here("tables", "cres_human_only_annotated.rds"))
# saveRDS(cres_mouse_only, here("tables", "cres_mouse_only_annotated.rds"))

message("All outputs saved to tables/")
```
