---
title: "E.coli infection"
author: "Wasim Aluísio Prates-Syed"
date: "2025-07-23"
output: 
  html_document: 
    toc: TRUE
    toc_depth: 4
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE, cache = TRUE)
```

# Exploring other datasets: A general analysis

## Import required files

```{r}
# install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))

#Import data
fit_training_datasets_metadata_clean = read_rds(file = here("FIT_Mouse2man", "fit_training_datasets_metadata_clean.rds"))
AllData_V2.0_blood = read_rds(here("new_models", "AllData_FIT_training_blood.rds"))
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))

```

# E. coli infection

### GSE33341

```{r}
#E. coli: GSE33341 -----
filename = "GSE33341"
GSE33341 <- getGEO("GSE33341", GSEMatrix = TRUE) 
eset_mouse <- GSE33341[[1]]
eset_human <- GSE33341[[2]]

#Metadata ------
#Mouse
GSE33341_ecoli_mouse_metadata = eset_mouse %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    #Host
    subject_title = "title",
    subject_num = "replicate number:ch1",
    batch = "experimental batch:ch1",
    host_strain = "host strain:ch1",
    organism = "organism_ch1",
    age = "host age (yr):ch1",
    sex = "host gender:ch1", 
    #Pathogen
    pathogen = "pathogen:ch1",
    status = "infection status:ch1",
    site_infection = "anatomic site of infection:ch1",
    route_infection = "inoculum route:ch1",
    dosage = "pathogen   dosage cfu/g:ch1",
    timepoint = "time after infection (h):ch1",
    bac_strain = "bacterial strain:ch1"
    ) %>% 
  filter(pathogen %in% c("Escherichia  coli", "-")) %>%
  mutate(pathogen = if_else(pathogen == "Escherichia  coli", "escherichia coli", "none"),
         subject_id = paste0("Mouse_", host_strain, "_Animal", subject_num, "_Batch", batch),
         status = str_replace(status, "-", "Healthy"),
         condition = paste("Mouse", host_strain, timepoint, sep = "_"),
         comparison = paste0("Mouse_", host_strain, "_Batch", batch)) %>% 
  select(geo_sample, comparison, condition, subject_id, everything())


#Human
GSE33341_ecoli_human_metadata = eset_human %>% 
  pData() %>% 
  filter(`pathogen:ch1` %in% c("Escherichia coli", "-")) %>%
  dplyr::select( 
    geo_sample = "geo_accession", 
    #Host
    subject_title = "title",
    # subject_num = "replicate number:ch1",
    batch = "experimental batch:ch1",
    # host_strain = "host strain:ch1",
    organism = "organism_ch1",
    age = "host age (yr):ch1",
    sex = "host gender:ch1", 
    ethnicity = "ethnicity:ch1",
    #Pathogen
    pathogen = "pathogen:ch1",
    # status = "infection status:ch1",
    site_infection = "anatomic site of infection:ch1"
    # route_infection = "inoculum route:ch1",
    # dosage = "pathogen   dosage cfu/g:ch1",
    # timepoint = "time after infection (h):ch1",
    # bac_strain = "bacterial strain:ch1"
    ) %>% 
  mutate(site_infection = str_to_lower(site_infection),
         organism = "Human",
         pathogen = if_else(pathogen == "-", "Healthy", pathogen),
         status = if_else(pathogen == "Healthy", pathogen, "infected"),
         timepoint = "0",
         subject_id = if_else(pathogen == "Escherichia coli", 
                              paste0("Ecoli_human_", site_infection,"_",geo_sample), 
                              paste0("Healthy_human_", geo_sample)),
         site_infection = if_else(site_infection == "-", "none", site_infection),
         condition = subject_id) %>% 
  select(geo_sample, condition, subject_id, everything())

GSE33341_ecoli_human_metadata

#Unite metadata
GSE33341_ecoli_mouse_human_metadata = GSE33341_ecoli_human_metadata %>% 
  bind_rows(GSE33341_ecoli_mouse_metadata)

GSE33341_ecoli_mouse_human_metadata %>% 
  saveRDS(file = here("tables", "GSE33341_ecoli_mouse_human_metadata.rds"))
```

```{r}
#Import
GSE33341_ecoli_mouse_human_metadata =
  readRDS(file = here("tables", "GSE33341_ecoli_mouse_human_metadata.rds"))

#Get expression data -----
#Mouse
GSE33341_mouse_exprs <- exprs(eset_mouse) %>% 
  as.data.frame() %>% 
  select(GSE33341_ecoli_mouse_human_metadata %>% 
           filter(organism == "Mus musculus") %>% 
           pull(geo_sample)) %>% 
  rownames_to_column("probe_id") 

GSE33341_mouse_exprs %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "Ecoli_GSE33341_mouse_exprs.rds"))

#Human
GSE33341_human_exprs <- exprs(eset_human) %>% 
  as.data.frame() %>% 
  select(GSE33341_ecoli_mouse_human_metadata %>% 
           filter(organism == "Human") %>% 
           pull(geo_sample)) %>% 
  rownames_to_column("probe_id")

GSE33341_human_exprs %>% 
  as.data.frame() %>% 
  saveRDS(here("tables", "Ecoli_GSE33341_human_exprs.rds"))

```

#### Mouse

##### Annotate genes

This procedure was done using only one batch (8)

```{r}

ecoli_human_exprs = readRDS(here("tables", "Ecoli_GSE33341_human_exprs.rds"))
ecoli_mouse_exprs = readRDS(here("tables", "Ecoli_GSE33341_mouse_exprs.rds"))
ecoli_mouse_human_metadata = readRDS(file = here("tables", "GSE33341_ecoli_mouse_human_metadata.rds"))

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
GSE33341_mouse_expression_metadata  = ecoli_mouse_exprs %>% 
  pivot_longer(cols = 2:125,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(GSE33341_ecoli_mouse_human_metadata, 
             by = "geo_sample") 

mouse_metadata =  GSE33341_ecoli_mouse_human_metadata %>% 
               filter(host_strain == "C57BL/6J",
                      batch == "8")

mouse_metadata %>% 
  saveRDS(file = here("tables", "ecoli_mouse_metadata.rds"))

#Get value from day 0 and calculate Log2FC
GSE33341_log2fc_sample = GSE33341_mouse_expression_metadata %>% 
  #Filter batch
  filter(organism == "Mus musculus",
         status == "Healthy",
         host_strain == "C57BL/6J",
         batch == "8") %>% 
  select(probe_id, batch, subject_num, day0_value_raw = value_raw) %>% 
  #Get the mean expression value for healthy
  group_by(probe_id, batch) %>% 
  summarize(n = n(),
            #Mean of healthy
            day0_value_raw = sum(day0_value_raw)/n,
            batch) %>% 
  select(probe_id, day0_value_raw) %>% 
  #Filter batch
  inner_join(GSE33341_mouse_expression_metadata %>% 
               filter(timepoint != "0",
                      host_strain == "C57BL/6J",
                      batch == "8"),
             by = join_by("probe_id", "batch")) %>% 
  mutate(log2fc = value_raw - day0_value_raw) %>% 
  distinct() 

#Check duplicates
GSE33341_log2fc_sample %>% 
  filter(subject_id == "Mouse_C57BL/6J_Animal1_Batch8",
         probe_id == "1415670_at")

```

```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- GSE33341_log2fc_sample$probe_id %>% unique()
# listAttributes(ensembl_mart)
#Annotate
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2",
  values = gene_id,
  mart = ensembl_mart)

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_log2summary = ecoli_mouse_exprs %>% #20,845 probes
  as.data.frame() %>% 
  rename(probe = probe_id) %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_mouse430_2), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = mouse_log2summary %>% 
  #filter ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

mouse_log2summary = mouse_log2summary %>% 
  select(probe, hgnc_symbol) %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
mouse_log2_filtered_probes = ecoli_mouse_exprs %>% 
  as.data.frame() %>% 
  rename(probe = probe_id) %>% 
  inner_join(mouse_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol, 
         c(mouse_metadata %>% 
             pull(geo_sample)))
#14395
mouse_log2_filtered_probes %>% dim()

mouse_log2_hgnc_symbol = mouse_log2_filtered_probes 

mouse_log2_hgnc_symbol %>% 
  get_dupes(human_symbol)

#Save
mouse_log2_hgnc_symbol %>% 
  saveRDS(file = here("tables", "GSE33341_mouse_counts_log2_summary_annotated_probes.rds"))
```

Import

```{r}
#Import
GSE33341_mouse_counts_log2_summary_annotated_genes = 
  readRDS(file = here("tables", "GSE33341_mouse_counts_log2_summary_annotated_probes.rds"))

#Pivot wider
GSE33341_log2fc_sample_wide = GSE33341_mouse_counts_log2_summary_annotated_probes %>% 
  select(probe = probe_id, entrezgene_id, batch:human_symbol) %>% 
  distinct() %>% 
  select(human_symbol, subject_id, timepoint, log2fc) %>%
  group_by(human_symbol, timepoint) %>% 
  pivot_wider(names_from = subject_id,
              values_from = log2fc) %>% 
  ungroup()  %>% 
  mutate(timepoint = fct_reorder(timepoint, timepoint %>% as.numeric()))
  
#Print
GSE33341_log2fc_sample_wide
```


##### Differential expression

#### Create ExpressionSet

```{r}
#Import
ecoli_mouse_log2_hgnc_symbol = readRDS(file = here("tables", "GSE33341_mouse_counts_log2_summary_annotated_probes.rds"))
ecoli_mouse_metadata = readRDS(file = here("tables", "ecoli_mouse_metadata.rds"))

#Create Expression Set
eset_mouse_ecoli <- ExpressionSet(
  assayData = ecoli_mouse_log2_hgnc_symbol %>% 
    column_to_rownames("human_symbol") %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(ecoli_mouse_metadata))

#Save
eset_mouse_ecoli %>% saveRDS(file = here("tables", "eset_mouse_ecoli.rds"))
```



#### Differential expression (Limma)
Paired samples design
```{r}

#Import
ecoli_mouse_log2_hgnc_symbol = readRDS(file = here("tables", "GSE33341_mouse_counts_log2_summary_annotated_probes.rds"))
ecoli_mouse_metadata = readRDS(file = here("tables", "ecoli_mouse_metadata.rds"))
eset_mouse_ecoli %>% saveRDS(file = here("tables", "eset_mouse_ecoli.rds"))

#Create model matrix
design <- model.matrix(~ condition + 0, # 0 defines the way we compare the samples
                       eset_mouse_ecoli) 

colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

# Estimate the correlation between measurements made on the same subject (paired-samples)
corfit <- duplicateCorrelation(eset_mouse_ecoli, 
                               design = design, 
                               block = eset_mouse_ecoli$subject_id)
corfit$consensus #-0.06

#Fit Limma model
fit <- lmFit(eset_mouse_ecoli, 
             design, 
             block = eset_mouse_ecoli$subject_id, 
             correlation = corfit$consensus) #Use inter-subject correlation as input into the linear model fit

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(conditionMouse_C57BL.6J_2	 - conditionMouse_C57BL.6J_0,
                             conditionMouse_C57BL.6J_6	 - conditionMouse_C57BL.6J_0,
                             conditionMouse_C57BL.6J_12	 - conditionMouse_C57BL.6J_0,
                             conditionMouse_C57BL.6J_24	 - conditionMouse_C57BL.6J_0,
                             levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_mouse_ecoli= all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  mutate(condition = case_when(comparisonA == "conditionMouse_C57BL.6J_2"	~ "Ecoli_2h",
                               comparisonA == "conditionMouse_C57BL.6J_6"	~ "Ecoli_6h",
                               comparisonA == "conditionMouse_C57BL.6J_12"	~ "Ecoli_12h",
                               comparisonA == "conditionMouse_C57BL.6J_24"	~ "Ecoli_24h",
                               T~comparisonA),
         organism = "Mouse") %>% 
  select(condition, everything()) %>% 
  clean_names() %>% 
  rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs_mouse_ecoli %>% saveRDS(file = here("tables", "Ecoli_GGSE33341_degs_mouse.rds"))

all_degs_mouse_ecoli %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarize(n_degs = n())

```












##### Correlation plots, pairwise

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(GSE33341_log2fc_sample_wide,
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, GSE33341, C57BL/6J, E. coli infection", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee"))

mouse_correlation_ggpairs_alldays
#Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_alldays_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()



#Timepoint 2
timepoint_select = 2
mouse_correlation_ggpairs_day2 = ggpairs(GSE33341_log2fc_sample_wide %>% 
                                              filter(timepoint == 2) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE33341, C57BL/6J, E.coli"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee"))

mouse_correlation_ggpairs_day2
#Save
mouse_correlation_ggpairs_day2 %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_day2_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 6
timepoint_select = 6
mouse_correlation_ggpairs_day6 = ggpairs(GSE33341_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE33341, C57BL/6J, E.coli"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee"))

mouse_correlation_ggpairs_day6
#Save
mouse_correlation_ggpairs_day6 %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_day6_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 12
timepoint_select = 12
mouse_correlation_ggpairs_day12 = ggpairs(GSE33341_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE33341, C57BL/6J, E.coli"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee"))

mouse_correlation_ggpairs_day12
#Save
mouse_correlation_ggpairs_day12 %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_day12_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 24
timepoint_select = 24
mouse_correlation_ggpairs_day24 = ggpairs(GSE33341_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE33341, C57BL/6J, E.coli"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee"))

mouse_correlation_ggpairs_day24
#Save
mouse_correlation_ggpairs_day24 %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_day24_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()
```

##### PCA

```{r fig.height=4, fig.width=5}
#Import metadata
annotation_samples_pca = GSE33341_mouse_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(subject_id, "_Day_", timepoint)) %>% 
  select(subject_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = GSE33341_mouse_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(subject_id, "_Day_", timepoint)) %>% 
  select(sample, entrezgene_id, log2fc) %>% 
  filter(!entrezgene_id == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("entrezgene_id") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = subject_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  # guides(fill = "none") +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "6" = "#219ebc",
                                "12" = "#669bbc",
                                "24" = "#4361ee")) +
  labs(title = "Mouse, E. coli, GSE33341",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

```

##### BTMs

```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)

#Annotate genes by BTM
GSE33341_genes_btm = GSE33341_mouse_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  rename(mm_entrez = entrezgene_id) %>% 
  select(mgi_symbol, symbol = human_symbol, participant_id = subject_id, geo_sample, timepoint, status, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = GSE33341_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process
```

```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(3:7) %>% 
  colnames()

timepoint_selected = 24
plot_data <-
  #Pivot longer, 1st participant
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == timepoint_selected) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == 2) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))

max_lim = max(plot_data$participant_1_value, plot_data$participant_2_value) + 0.5
min_lim = min(plot_data$participant_1_value, plot_data$participant_2_value) - 0.5

#Plot
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(min_lim, max_lim) +
  xlim(min_lim, max_lim) +
  theme(strip.text = element_text(size = 8),
        legend.position = "right") +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste("GSE33341, Mouse, E. coli infection, Day", timepoint_selected))

```

#### Human

##### Annotate genes

There is a batch effect

```{r}
#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
GSE33341_human_expression_metadata  = GSE33341_human_exprs %>% 
  pivot_longer(cols = 2:63,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(GSE33341_ecoli_mouse_human_metadata, 
             by = "geo_sample") 

GSE33341_human_expression_metadata %>% 
  saveRDS(file = here("tables", "GSE33341_human_expression_metadata.rds"))

human_metadata =  GSE33341_ecoli_mouse_human_metadata %>% 
  filter(organism == "Human")

human_metadata %>% 
  saveRDS(file = here("tables", "ecoli_human_metadata.rds"))

#There is a batch effect
#   batch   status  n
# 1  2034 infected  7
# 2  2287  Healthy 43
# 3  2287 infected  7
# 4  2739 infected  5

#Get value from day 0 and calculate Log2FC
GSE33341_human_log2fc_sample = GSE33341_human_expression_metadata %>% 
  #Filter batch
  filter(organism == "Human",
         status == "Healthy") %>% 
  select(probe_id, participant_id = subject_id, status, day0_value_raw = value_raw) %>% 
  group_by(probe_id, status) %>% 
  summarize(n = n(),
            #Mean of healthy
            day0_value_raw = sum(day0_value_raw)/n) %>% 
  select(probe_id, day0_value_raw) %>% 
  #Join by probe id to infected participants
  inner_join(GSE33341_human_expression_metadata %>% 
               filter(site_infection == "urinary tract"), #1 lung, 19 urinary tract),
             by = join_by("probe_id")) %>% 
  mutate(log2fc = value_raw - day0_value_raw) %>% 
  select(probe_id:subject_id, batch:status, log2fc)

#Check duplicates
GSE33341_human_log2fc_sample %>% 
  filter(subject_id == "Ecoli_human_urinary tract_GSM824707",
         probe_id == "1007_s_at") #OK

```

```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- GSE33341_human_log2fc_sample$probe_id %>% unique()
listAttributes(ensembl_mart)%>% View()
#Annotate
#[HG-U133A_2] Affymetrix Human Genome U133A 2.0 Array
annot <- getBM(
  attributes = c("affy_hg_u133a_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133a_2",
  values = gene_id,
  mart = ensembl_mart)

#Get annotations for each probe 
### Filter probes with highest median expression
#Get annotations for each probe 
### Filter probes with highest median expression
human_ecoli_log2summary = GSE33341_human_exprs %>% 
  rename(probe = probe_id) %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_hg_u133a_2), 
            by = "probe") %>% 
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(human_symbol = hgnc_symbol) %>% 
  filter(human_symbol != "") %>%
  group_by(entrezgene_id) %>%
  slice_sample(n = 1) %>%  # one line per entrezgene_id
  ungroup() %>%
  distinct(probe, .keep_all = TRUE)  # remove dup probes

human_ecoli_log2summary_entrez %>% 
  get_dupes(probe)



#Filtered probes
human_ecoli_log2_filtered_probes = GSE33341_human_exprs %>%
  rename(probe = probe_id) %>% 
  inner_join(human_ecoli_log2summary_entrez, by = "probe") %>%   #214,396
  select(human_symbol,
         c(human_metadata$geo_sample))
#14395
human_ecoli_log2_filtered_probes %>% dim()

human_ecoli_log2_filtered_symbol = human_ecoli_log2_filtered_probes 

human_ecoli_log2_filtered_symbol %>% 
  get_dupes(human_symbol)

#Save 
human_ecoli_log2_filtered_symbol %>% 
  saveRDS(here("tables", "human_ecoli_log2_filtered_symbol.rds"))
```

##### Import

```{r}
#Import
GSE33341_human_counts_log2_summary_annotated_probes =
  readRDS(file = here("tables", "human_ecoli_log2_filtered_symbol.rds"))

#Check duplicates
GSE33341_human_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
dplyr::summarise(n = dplyr::n(), .by = c(entrezgene_id, human_symbol, subject_id)) |>
  dplyr::filter(n > 1L) %>% 
  select(human_symbol, entrezgene_id) %>% 
  distinct()

#Pivot wider
GSE33341_human_log2fc_sample_wide = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  select(human_symbol, status, subject_id, log2fc) %>%
  group_by(human_symbol) %>% 
  pivot_wider(names_from = subject_id,
              values_from = log2fc) %>% 
  ungroup()
  
#Print
GSE33341_human_log2fc_sample_wide
```



##### Differential expression

#### Create ExpressionSet

```{r}
#Import
human_ecoli_log2_filtered_symbol = readRDS(file = here("tables", "human_ecoli_log2_filtered_symbol.rds"))

dim(human_ecoli_log2_filtered_symbol)
  
ecoli_human_metadata = readRDS(file = here("tables", "ecoli_human_metadata.rds")) %>% 
  mutate(timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         age = scale(as.numeric(age)),
         ethnicity = as.factor(ethnicity),
         status = as.factor(status),
         status = fct_relevel(status, "Healthy", "infected"),
         site_infection = as.factor(site_infection)) %>% 
  dplyr::filter(!is.na(age))

dim(ecoli_human_metadata)

#Create Expression Set
eset_human_ecoli <- ExpressionSet(
  assayData = human_ecoli_log2_filtered_symbol %>% 
    column_to_rownames("human_symbol") %>% 
    dplyr::select(ecoli_human_metadata$geo_sample) %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(ecoli_human_metadata)) 

#metadata = 62 rows
#exprs data = 62 cols

#Save
eset_human_ecoli %>% saveRDS(file = here("tables", "eset_human_ecoli.rds"))


```



#### Differential expression (Limma)
Paired samples design
```{r}

#Import
human_ecoli_log2_filtered_symbol = readRDS(file = here("tables", "human_ecoli_log2_filtered_symbol.rds"))
ecoli_human_metadata = readRDS(file = here("tables", "ecoli_human_metadata.rds"))
eset_human_ecoli = readRDS(file = here("tables", "eset_human_ecoli.rds"))


#Create model matrix
design <- model.matrix(~ age + sex + ethnicity + status,
                       data = pData(eset_human_ecoli))

colnames(design) <- make.names(colnames(design))  # Avoids blank spaces
rownames(design) <- rownames(pData(eset_human_ecoli))

#Fit Limma model
fit <- lmFit(eset_human_ecoli, 
             design) 

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(statusinfected, levels = design)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- limma::eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_human_ecoli= all_contrasts_degs %>% 
  dplyr::select(comparison, everything()) %>% 
  dplyr::mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  dplyr::mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  dplyr::mutate(condition = case_when(comparisonA == "statusinfected"	~ "Ecoli_24h",
                               T~comparisonA),
         organism = "Human") %>% 
  dplyr::select(condition, everything()) %>% 
  clean_names() %>% 
  dplyr::rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs_human_ecoli %>% saveRDS(file = here("tables", "Ecoli_GGSE33341_degs_human.rds"))

all_degs_human_ecoli %>% 
  dplyr::group_by(condition) %>%
  dplyr::filter(adj_p_val <= 0.05) %>% 
  dplyr::summarize(n_degs = n())

```














##### Correlation plots, pairwise

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}


# Plot correlations for human -------
participant_id_selected = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  select(subject_id) %>% 
  distinct() %>% 
  slice_sample(n = 5) %>% 
  pull()

#All timepoints
human_correlation_ggpairs_alldays = ggpairs(GSE33341_human_log2fc_sample_wide %>% 
                                              select(human_symbol, status, participant_id_selected),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = status,
                                                           fill = status),
                                    title = "GSE33341, Human, E. coli infection", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point), 
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = "#219ebc") +
  scale_fill_manual(values = "#219ebc") 

human_correlation_ggpairs_alldays

#Save
human_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "human_correlation_ggpairs_alldays_Ecoli_GSE33341.png"),
         height = 10,
         width = 10)
beep()
```

##### PCA

```{r fig.height=4, fig.width=5}
#Import metadata
annotation_samples_pca = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  select(sample = subject_id, subject_id, status, organism, batch) %>% 
  distinct()

# PCA with raw values
clean_matrix_raw = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  select(sample = subject_id, entrezgene_id, log2fc) %>% 
  filter(!entrezgene_id == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("entrezgene_id") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = batch,
      fill = batch,
      shape = organism,
      label = sample) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  # guides(fill = "none") +
 # scale_color_manual(values = "#219ebc") +
 #  scale_fill_manual(values = "#219ebc") +
  labs(title = "Human, E. coli, GSE33341",
       shape = "Organism",
       fill = "Batch") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

```

##### BTMs

```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)

#Annotate genes by BTM
GSE33341_human_genes_btm = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  rename(mm_entrez = entrezgene_id) %>% 
  select(symbol = human_symbol, participant_id = subject_id, geo_sample, status, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_human_btm_mean_process = GSE33341_human_genes_btm %>%
  mutate(condition = "infected") %>% 
  group_by(participant_id, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_human_btm_mean_process
```

```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_human_btm_mean_process = dge_human_btm_mean_process %>%  
  ungroup()

sample_participants_5 = GSE33341_human_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  select(subject_id) %>% 
  distinct() %>% 
  slice_sample(n = 5) %>% 
  pull()

plot_data <-
  #Pivot longer, 1st participant
  ggpairs_data_dge_human_btm_mean_process %>%
  # filter(timepoint == timepoint_selected) %>% 
  select(group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_human_btm_mean_process %>%
      # filter(timepoint == timepoint_selected) %>% 
      select(group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))

max_lim = max(plot_data$participant_1_value, plot_data$participant_2_value) + 0.5
min_lim = min(plot_data$participant_1_value, plot_data$participant_2_value) - 0.5

#Plot
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(min_lim, max_lim) +
  xlim(min_lim, max_lim) +
  theme(strip.text = element_text(size = 8),
        legend.position = "right") +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste("GSE33341, Human, E. coli infection"))

```

### Compare Human vs Mouse

```{r}
GSE33341_human_counts_log2_summary_annotated_probes =
  readRDS(file = here("tables", "GSE33341_human_counts_log2_summary_annotated_probes.rds"))

GSE33341_mouse_counts_log2_summary_annotated_probes =
  readRDS(file = here("tables", "GSE33341_mouse_counts_log2_summary_annotated_probes.rds"))


#Unite Mouse and Human datasets. Samples.
human_mouse_ecoli_log2fc_samples = 
#Mouse
GSE33341_mouse_counts_log2_summary_annotated_probes %>%
  ungroup() %>% 
  select(human_symbol, log2fc, participant_id = subject_id, 
         pathogen, 
         value_raw_ref = day0_value_raw, 
         value_raw,
         timepoint) %>% 
  mutate(organism = "Mouse") %>% 
  
  #Human
  bind_rows(GSE33341_human_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  select(human_symbol, log2fc,
         value_raw_ref = day0_value_raw, 
         value_raw,
         participant_id = subject_id, pathogen) %>% 
  mutate(organism = "Human",
         timepoint = "Hospitalization"))

#Save
human_mouse_ecoli_log2fc_samples %>% 
  saveRDS(file = here("tables", "human_mouse_ecoli_log2fc_samples.rds"))
```

```{r}
#Calculate the mean log2fc
human_mouse_ecoli_log2fc_avg = human_mouse_ecoli_log2fc_samples %>% 
  group_by(organism, human_symbol, timepoint) %>% 
  summarize(
    mean_log2fc = mean(log2fc, na.rm = TRUE),
    sd_log2fc   = sd(log2fc, na.rm = TRUE),
    se_log2fc   = sd(log2fc, na.rm = TRUE) / sqrt(n()),   # erro padrão
    
    # Test t paired
    ttest = list(tryCatch(
      t.test(value_raw, value_raw_ref, paired = TRUE),
      error = function(e) NULL
    )),
    .groups = "drop"
  ) %>%
  mutate(
    p_value = map_dbl(ttest, ~ if (!is.null(.x)) .x$p.value else NA_real_),
    ci_low  = map_dbl(ttest, ~ if (!is.null(.x)) .x$conf.int[1] else NA_real_),
    ci_high = map_dbl(ttest, ~ if (!is.null(.x)) .x$conf.int[2] else NA_real_),
    t_stat  = map_dbl(ttest, ~ if (!is.null(.x)) .x$statistic else NA_real_),
    df      = map_dbl(ttest, ~ if (!is.null(.x)) .x$parameter else NA_real_),
    p_adj   = p.adjust(p_value, method = "BH")
  ) %>%
  select(-ttest) %>% 
  filter(human_symbol != "") %>%
  ungroup()

human_mouse_ecoli_log2fc_avg


human_mouse_ecoli_log2fc_avg %>% 
  saveRDS(file = here("tables", "human_mouse_ecoli_log2fc_avg.rds"))

```

```{r}
#Pivot wider to compare human to different mouse times
#Human
human_mouse_ecoli_log2fc_avg_wide = 
  
  #Human
  human_mouse_ecoli_log2fc_avg %>% 
  filter(organism == "Human") %>% 
  select(human_symbol, 
         timepoint_human = timepoint, 
         mean_log2fc_human = mean_log2fc,
         p_value_human = p_value,
         p_adj_human = p_adj) %>% 
  inner_join(
  #Mouse
  human_mouse_ecoli_log2fc_avg %>% 
  filter(organism == "Mouse") %>% 
  select(human_symbol, 
         timepoint_mouse = timepoint, 
         mean_log2fc_mouse = mean_log2fc,
         p_value_mouse = p_value,
         p_adj_mouse = p_adj) %>% 
    mutate(timepoint_mouse = fct_reorder(paste0(timepoint_mouse, "h"), c(timepoint_mouse %>% as.numeric()))),
  by = "human_symbol") 

human_mouse_ecoli_log2fc_avg_wide


#Get immune and non immune genes
human_mouse_ecoli_log2fc_avg_wide_degs_immune_nonimmune = human_mouse_ecoli_log2fc_avg_wide %>%
  mutate(gene_type = "all") %>% 
  bind_rows(
  #Classify as immune or non_immune 
    human_mouse_ecoli_log2fc_avg_wide %>% 
  mutate(gene_type = if_else(human_symbol %in% ImmuneGO_BTM_grouped_genes$genes,
                             "immune",
                             "non_immune"))) %>%
  #Filter DEGs in both organisms, same sign
  filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 &
      sign(mean_log2fc_human) == sign(mean_log2fc_mouse)) %>%
  group_by(timepoint_human, timepoint_mouse, gene_type)

#Summary
degs_immune_nonimmune_summary = human_mouse_ecoli_log2fc_avg_wide_degs_immune_nonimmune %>% 
  summarize(n_genes = n(), .groups = "drop")

# Calculate the correlation
cor_data <- human_mouse_ecoli_log2fc_avg_wide_degs_immune_nonimmune %>%
  # filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>%
  group_by(timepoint_human, timepoint_mouse, gene_type) %>%
  summarize(
    cor_test = list(cor.test(mean_log2fc_mouse, mean_log2fc_human,
                             method = "pearson", use = "complete.obs")),
    .groups = "drop"
  ) %>%
  inner_join(degs_immune_nonimmune_summary, by = join_by("timepoint_human", "timepoint_mouse", "gene_type")) %>% 
  mutate(
    cor_value = map_dbl(cor_test, ~ .x$estimate),
    p_value   = map_dbl(cor_test, ~ .x$p.value),
    label_pos_x = map_dbl(cor_test, ~ max(human_mouse_ecoli_log2fc_avg_wide$mean_log2fc_mouse, na.rm = TRUE)),
    label_pos_y = map_dbl(cor_test, ~ max(human_mouse_ecoli_log2fc_avg_wide$mean_log2fc_human + 1, na.rm = TRUE)),
    label = paste0("r = ", round(cor_value, 2), ", p = ", signif(p_value, 2), ", n = ", n_genes))

  cor_data
  
  
```

```{r fig.height=3, fig.width=10}

human_mouse_ecoli_log2fc_avg_wide_scatterplot = human_mouse_ecoli_log2fc_avg_wide %>% 
  filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>% 
  
  #Plot
  ggplot() +
  aes(x = mean_log2fc_mouse,
      y = mean_log2fc_human) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #Facet
  facet_grid(~timepoint_human~timepoint_mouse) +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_ecoli_log2fc_avg_wide %>%
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_ecoli_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  stat_smooth(method = "lm", geom = "line", color = "#4DBBD5FF", alpha = 0.5, se = F, size = 1,
              data = human_mouse_ecoli_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
    #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Add correlation metrics (Immune)
  geom_label(
    data = cor_data %>% filter(gene_type == "all"),
    aes(x = 0.1, y = 9, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 0,
    vjust = 1,
    size = 2.5) +
  geom_label(
    data = cor_data %>% filter(gene_type == "immune"),
    aes(x = 0.1, y = 8, label = label),
    inherit.aes = FALSE,
    fill = "#4DBBD5FF",
    hjust = 0,
    vjust = 1,
    size = 2.5) +
  geom_label(
    data = cor_data %>% filter(gene_type == "non_immune"),
    aes(x = 0.1, y = 7, label = label),
    inherit.aes = FALSE,
    fill = "#4361ee",
    color = "white",
    hjust = 0,
    vjust = 1,
    size = 2.5) +
  
  # #Adjust the axes range
  # ylim(-4, 7) +
  
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "top",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title = "Correlation plot, E. coli infection, Log2-Fold Change, GSE33341 ")

human_mouse_ecoli_log2fc_avg_wide_scatterplot

human_mouse_ecoli_log2fc_avg_wide_scatterplot %>% 
  ggsave(filename = here("Figures", "human_mouse_Ecoli_log2fc_avg_wide_scatterplot.png"),
         height = 3,
         width = 10)

```


Single timepoint
```{r fig.height=3, fig.width=3}
human_mouse_ecoli_log2fc_avg_wide_scatterplot_singletimepoint = human_mouse_ecoli_log2fc_avg_wide %>% 
  filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>% 
  filter(timepoint_mouse == "2h") %>% 
  
  #Plot
  ggplot() +
  aes(x = mean_log2fc_mouse,
      y = mean_log2fc_human) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #Facet
  facet_grid(~timepoint_human~timepoint_mouse) +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_ecoli_log2fc_avg_wide %>%
               filter(timepoint_mouse == "2h") %>%
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_ecoli_log2fc_avg_wide %>%
               filter(timepoint_mouse == "2h") %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  stat_smooth(method = "lm", geom = "line", color = "#4DBBD5FF", alpha = 0.5, se = F, size = 1,
              data = human_mouse_ecoli_log2fc_avg_wide %>%
                filter(timepoint_mouse == "2h") %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
    #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Add correlation metrics (Immune)
  geom_label(
    data = cor_data %>% 
      filter(timepoint_mouse == "2h") %>%
      filter(gene_type == "all"),
    aes(x = 0.1, y = 9, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 0,
    vjust = 1,
    size = 3) +
  geom_label(
    data = cor_data %>% 
      filter(timepoint_mouse == "2h") %>%
      filter(gene_type == "immune"),
    aes(x = 0.1, y = 8, label = label),
    inherit.aes = FALSE,
    fill = "#4DBBD5FF",
    hjust = 0,
    vjust = 1,
    size = 3) +
  geom_label(
    data = cor_data %>% 
      filter(timepoint_mouse == "2h") %>%
      filter(gene_type == "non_immune"),
    aes(x = 0.1, y = 7, label = label),
    inherit.aes = FALSE,
    fill = "#4361ee",
    color = "white",
    hjust = 0,
    vjust = 1,
    size = 3) +
  
  # #Adjust the axes range
  # ylim(-4, 7) +
  
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_blank(),
        legend.position = "top",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title = "Mouse vs. Human, gene,\nE. coli infection")

human_mouse_ecoli_log2fc_avg_wide_scatterplot_singletimepoint

human_mouse_ecoli_log2fc_avg_wide_scatterplot_singletimepoint %>% 
  ggsave(filename = here("Figures", "human_mouse_ecoli_log2fc_avg_wide_scatterplot_2h.png"),
         height = 3,
         width = 3)
```


#### Plot correlation between genes (log2fc for every sample)

```{r}

# --- 1) Mouse vs Human no mesmo timepoint ---
#Get timepoints
timepoints_human <- unique(human_mouse_ecoli_log2fc_samples %>% filter(organism == "Human") %>% pull(timepoint))
timepoints_mouse <- unique(human_mouse_ecoli_log2fc_samples %>% filter(organism == "Mouse") %>% pull(timepoint))

#Get combinations
timepoint_combinations <- expand.grid(timepoint_human = timepoints_human,
                                      timepoint_mouse = timepoints_mouse,
                                      stringsAsFactors = FALSE)

# Loop
cor_human_mouse_timepoint <- timepoint_combinations %>%
  pmap_dfr(function(timepoint_human, timepoint_mouse) {
    
    # pegar os participantes do timepoint específico
    human_df <- human_mouse_ecoli_log2fc_samples %>%
      filter(organism == "Human", timepoint == timepoint_human) %>%
      select(participant_id, human_symbol, log2fc)
    
    mouse_df <- human_mouse_ecoli_log2fc_samples %>%
      filter(organism == "Mouse", timepoint == timepoint_mouse) %>%
      select(participant_id, human_symbol, log2fc)
    
    # manter apenas genes em comum
    common_genes <- intersect(human_df$human_symbol, mouse_df$human_symbol)
    human_df <- human_df %>% filter(human_symbol %in% common_genes)
    mouse_df <- mouse_df %>% filter(human_symbol %in% common_genes)
    
    # pivot para wide
    human_wide <- human_df %>% pivot_wider(names_from = participant_id, values_from = log2fc)
    mouse_wide <- mouse_df %>% pivot_wider(names_from = participant_id, values_from = log2fc)
    
    # IDs de participantes
    human_ids <- setdiff(colnames(human_wide), "human_symbol")
    mouse_ids <- setdiff(colnames(mouse_wide), "human_symbol")
    
    # calcular correlação para todos pares humano x mouse
    expand.grid(human = human_ids, mouse = mouse_ids, stringsAsFactors = FALSE) %>%
      mutate(
        cor_test = map2(human, mouse, ~{
          x <- human_wide[[.x]]
          y <- mouse_wide[[.y]]
          # remover NAs correspondentes
          keep <- complete.cases(x, y)
          cor.test(x[keep], y[keep], method = "pearson")
        }),
        r       = map_dbl(cor_test, ~.x$estimate),
        p_value = map_dbl(cor_test, ~.x$p.value),
        ci_low  = map_dbl(cor_test, ~.x$conf.int[1]),
        ci_high = map_dbl(cor_test, ~.x$conf.int[2])
      ) %>%
      select(-cor_test) %>%
      mutate(timepoint_human = timepoint_human,
             timepoint_mouse = timepoint_mouse)
  }) %>% 
  group_by(timepoint_human, timepoint_mouse) %>%
  summarize(
    comparison = "Human vs. Mouse",
    mean = mean(r, na.rm = TRUE),
    sd   = sd(r, na.rm = TRUE),
    n      = sum(!is.na(r)),
    se     = sd / sqrt(n),
    tval   = qt(0.975, df = n - 1),
    ci_low = mean - tval * se,
    ci_high = mean + tval * se,
    .groups = "drop"
  )

# --- 2) Human vs Human (entre timepoints) ---
cor_human_human <- human_mouse_ecoli_log2fc_samples %>%
  filter(organism == "Human") %>%
  select(participant_id, human_symbol, timepoint, log2fc) %>%
  pivot_wider(
    names_from = participant_id,
    values_from = log2fc
  ) %>%
  select(-human_symbol, -timepoint) %>%   # só colunas de participantes
  cor(use = "pairwise.complete.obs", method = "pearson") %>% 
  as.data.frame() %>%
  rownames_to_column("participant_1") %>%
  melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>% 
  filter(participant_1 != participant_2) %>%   # tira a diagonal (=1)
  # garantir combinações únicas
  rowwise() %>%
  mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
  ungroup() %>%
  distinct(pair, .keep_all = TRUE) %>%
  #Summarize
  summarize(
    comparison = "Human vs. Human",
    mean = mean(correlation, na.rm = TRUE),
    sd   = sd(correlation, na.rm = TRUE),
    n    = sum(!is.na(correlation)),
    se   = sd / sqrt(n),
    tval = qt(0.975, df = n - 1),
    ci_low  = mean - tval * se,
    ci_high = mean + tval * se
  ) 
  

# --- 3) Mouse vs Mouse ---
cor_mouse_mouse <- human_mouse_ecoli_log2fc_samples %>%
  filter(organism == "Mouse") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, human_symbol, log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = log2fc) %>%
      select(-human_symbol)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Mouse vs. Mouse",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup()


#Unite
correlation_long_all_mouse_human_Ecoli_timepoints_summary = cor_human_mouse_timepoint %>% 
  mutate(timepoint = timepoint_mouse) %>% 
  bind_rows(
    cor_human_human %>% mutate(timepoint = "24"),
    cor_mouse_mouse)

correlation_long_all_mouse_human_Ecoli_timepoints_summary

```

```{r fig.height=5, fig.width=3}
correlation_long_all_mouse_human_Ecoli_timepoints_summary_plot = correlation_long_all_mouse_human_Ecoli_timepoints_summary %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1,
                data = correlation_long_all_mouse_human_timepoints_summary %>% 
                  filter(comparison != "Human vs. Human")) +
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 1,
                linewidth = 1,
                data = correlation_long_all_mouse_human_timepoints_summary %>% 
                  filter(comparison == "Human vs. Human")) +
  scale_x_continuous(breaks = c(0, 2, 6, 12, 24), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_color_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = "Correlation between organisms,\ngenes, E. coli infection",
       x = "",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 

correlation_long_all_mouse_human_Ecoli_timepoints_summary_plot

correlation_long_all_mouse_human_Ecoli_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "correlation_long_all_mouse_human_Ecoli_timepoints_summary_plot.png"),
         width = 3,
         height = 5)


correlation_long_all_mouse_human_Ecoli_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "correlation_long_all_mouse_human_Ecoli_timepoints_summary_plot_square.png"),
         width = 3,
         height = 3)
```

#### BTMs

```{r}
#Annotate genes by BTM
dge_btm_mean_process_samples = GSE33341_genes_btm %>%
  mutate(organism = "Mouse") %>% 
  bind_rows(GSE33341_human_genes_btm %>% 
              mutate(organism = "Human"))

dge_btm_mean_process_samples %>% 
  saveRDS(file = here("tables", "dge_Ecoli_btm_mean_process_samples.rds"))

#Import
dge_btm_mean_process_samples = 
  readRDS(file = here("tables", "dge_Ecoli_btm_mean_process_samples.rds"))

dge_btm_mean_process = dge_btm_mean_process_samples %>% 
  # filter(timepoint %in% c(1, 2)) %>% 
  mutate(condition = if_else(organism == "Mouse", paste0(organism, "_", timepoint, "h"), organism)) %>% 
  group_by(condition, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  select(-timepoint) %>% 
  pivot_wider(names_from = "condition", values_from = "mean_log2fc") 

dge_btm_mean_process %>% 
  saveRDS(file = here("tables", "dge_Ecoli_btm_mean_process.rds"))
dge_btm_mean_process
```

##### Import
```{r}
dge_btm_mean_process =
  readRDS(file = here("tables", "dge_Ecoli_btm_mean_process.rds"))
```


##### Scatterplot

```{r fig.height=3, fig.width=10}

plot_data <-
  #Pivot longer, 1st participatn
  dge_btm_mean_process %>%
  pivot_longer(cols = c(Mouse_12h:Mouse_6h), 
               names_to = "mouse_subject", 
               values_to = "mouse_subject_value")


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(mouse_subject) %>%
  summarise(cor_value = cor(Human, mouse_subject_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(comparison = paste0("Human vs. ", mouse_subject),
         r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
ecoli_btm_mean_plot = plot_data %>% 
ggplot(aes(x = mouse_subject_value, y = Human)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  facet_grid("Human" ~ mouse_subject,  drop = T) +
  stat_smooth(aes(x = mouse_subject_value, y = Human),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  theme_vaxgo() +
  # ylim(-3, 3) +
  xlim(-1.5, 2) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "Mouse Log2FC", 
       y = "Human Log2FC", 
       color = "",
       title = "Mouse vs. Human")

ecoli_btm_mean_plot

ecoli_btm_mean_plot %>% 
  ggsave(filename = here("Figures", "ecoli_btm_mean_plot.png"),
         width = 10, height = 3)
```

One timepoint
```{r fig.height=4, fig.width=3}
#Plot 
ecoli_btm_mean_plot = plot_data %>% 
  filter(str_detect(mouse_subject, "_2h")) %>% 
  
  ggplot(aes(x = mouse_subject_value, y = Human)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  facet_grid("Human" ~ mouse_subject,  drop = T) +
  stat_smooth(aes(x = mouse_subject_value, y = Human),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  
  geom_label(data = cor_labels %>% filter(str_detect(mouse_subject, "_2h")),
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  theme_vaxgo() +
  ylim(-2, 2) +
  xlim(-1, 2) +
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "Mouse Log2FC", 
       y = "Human Log2FC", 
       color = "",
       fill = "",
       title = "Mouse vs. Human, BTMs,\nE. coli infection")

ecoli_btm_mean_plot

ecoli_btm_mean_plot %>% 
  ggsave(filename = here("Figures", "ecoli_btm_mean_plot_2h.png"),
         width = 3, height = 3)

```



##### By samples

####Samples' means

```{r}
#Import
dge_btm_mean_process_samples =
  readRDS(file = here("tables", "dge_Ecoli_btm_mean_process_samples.rds")) %>% 
   mutate(timepoint = if_else(organism == "Human", "24", timepoint)) 
GSE33341_ecoli_mouse_human_metadata =
  readRDS(file = here("tables", "GSE33341_ecoli_mouse_human_metadata.rds"))

#Calculate the mean log2fc by module
dge_btm_mean_process_sample_long = dge_btm_mean_process_samples %>% 
  mutate(timepoint = if_else(organism == "Human", "24", timepoint)) %>% 
  group_by(organism, participant_id, timepoint, process) %>% 
  summarize(group,
            subgroup, 
            mean_log2fc = mean(log2fc, na.rm = T)) %>% 
  distinct() %>% 
  ungroup()

dge_btm_mean_process_sample_long %>% 
  saveRDS(file = here("GSEA_ORA", "dge_Ecoli_btm_mean_process_sample_long.rds"))

#Pivot wider
dge_btm_mean_process_sample_wide = dge_btm_mean_process_sample_long %>% 
  group_by(organism, timepoint, process) %>% 
  pivot_wider(names_from = "participant_id", 
              values_from = "mean_log2fc")
  
```

###### Summarize interindividual correlation


Alternative
```{r}
# --- 1) Mouse vs Human no mesmo timepoint ---
#Get timepoints
timepoints_human <- unique(dge_btm_mean_process_sample_long %>% filter(organism == "Human") %>% pull(timepoint))
timepoints_mouse <- unique(dge_btm_mean_process_sample_long %>% filter(organism == "Mouse") %>% pull(timepoint))

#Get combinations
timepoint_combinations <- expand.grid(timepoint_human = timepoints_human,
                                      timepoint_mouse = timepoints_mouse,
                                      stringsAsFactors = FALSE)

# Loop
cor_human_mouse_timepoint <- timepoint_combinations %>%
  pmap_dfr(function(timepoint_human, timepoint_mouse) {
    
    # pegar os participantes do timepoint específico
    human_df <- dge_btm_mean_process_sample_long %>%
      filter(organism == "Human", timepoint == timepoint_human) %>%
      select(participant_id, process, mean_log2fc)
    
    mouse_df <- dge_btm_mean_process_sample_long %>%
      filter(organism == "Mouse", timepoint == timepoint_mouse) %>%
      select(participant_id, process, mean_log2fc)
    
    # manter apenas genes em comum
    common_process <- intersect(human_df$process, mouse_df$process)
    human_df <- human_df %>% filter(process %in% common_process)
    mouse_df <- mouse_df %>% filter(process %in% common_process)
    
    # pivot para wide
    human_wide <- human_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    mouse_wide <- mouse_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    
    # IDs de participantes
    human_ids <- setdiff(colnames(human_wide), "process")
    mouse_ids <- setdiff(colnames(mouse_wide), "process")
    
    # calcular correlação para todos pares humano x mouse
    expand.grid(human = human_ids, mouse = mouse_ids, stringsAsFactors = FALSE) %>%
      mutate(
        cor_test = map2(human, mouse, ~{
          x <- human_wide[[.x]]
          y <- mouse_wide[[.y]]
          # remover NAs correspondentes
          keep <- complete.cases(x, y)
          cor.test(x[keep], y[keep], method = "pearson")
        }),
        r       = map_dbl(cor_test, ~.x$estimate),
        p_value = map_dbl(cor_test, ~.x$p.value),
        ci_low  = map_dbl(cor_test, ~.x$conf.int[1]),
        ci_high = map_dbl(cor_test, ~.x$conf.int[2])
      ) %>%
      select(-cor_test) %>%
      mutate(timepoint_human = timepoint_human %>% as.numeric(),
             timepoint_mouse = timepoint_mouse %>% as.numeric())
  }) %>% 
  group_by(timepoint_human, timepoint_mouse) %>%
  summarize(
    comparison = "Human vs. Mouse",
    mean = mean(r, na.rm = TRUE),
    sd   = sd(r, na.rm = TRUE),
    n      = sum(!is.na(r)),
    se     = sd / sqrt(n),
    tval   = qt(0.975, df = n - 1),
    ci_low = mean - tval * se,
    ci_high = mean + tval * se,
    .groups = "drop"
  )

cor_human_mouse_timepoint

# --- 2) Human vs Human (entre timepoints) ---
cor_human_human <- dge_btm_mean_process_sample_long %>%
  mutate(timepoint = as.numeric(timepoint)) %>% 
  filter(organism == "Human") %>%
  select(participant_id, process, timepoint, mean_log2fc) %>%
  group_by(timepoint) %>% 
  pivot_wider(
    names_from = participant_id,
    values_from = mean_log2fc
  ) %>%
  select(-process, -timepoint) %>%   # só colunas de participantes
  cor(use = "pairwise.complete.obs", method = "pearson") %>% 
  as.data.frame() %>%
  rownames_to_column("participant_1") %>%
  melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>% 
  filter(participant_1 != participant_2) %>%   # tira a diagonal (=1)
  # garantir combinações únicas
  rowwise() %>%
  mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
  ungroup() %>%
  distinct(pair, .keep_all = TRUE) %>%
  #Summarize
  summarize(
    comparison = "Human vs. Human",
    mean = mean(correlation, na.rm = TRUE),
    sd   = sd(correlation, na.rm = TRUE),
    n    = sum(!is.na(correlation)),
    se   = sd / sqrt(n),
    tval = qt(0.975, df = n - 1),
    ci_low  = mean - tval * se,
    ci_high = mean + tval * se
  ) %>% 
  mutate(timepoint = 24,
         timepoint_human = timepoint)
  

# --- 3) Mouse vs Mouse ---
cor_mouse_mouse <- dge_btm_mean_process_sample_long %>%
  filter(organism == "Mouse") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, process, mean_log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = mean_log2fc) %>%
      select(-process)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Mouse vs. Mouse",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup() %>% 
  mutate(timepoint = as.numeric(timepoint), 
         timepoint_mouse = timepoint)


#Unite
correlation_long_all_btm_mouse_human_timepoints_summary = cor_human_mouse_timepoint %>% 
  mutate(timepoint = timepoint_mouse) %>% 
  bind_rows(
    cor_human_human,
    cor_mouse_mouse)

correlation_long_all_btm_mouse_human_timepoints_summary
```

Plot scatterplot
```{r fig.height=3, fig.width=3}
correlation_long_all_btm_mouse_human_timepoints_summary_plot = correlation_long_all_btm_mouse_human_timepoints_summary %>% 
  mutate(comparison = str_replace_all(comparison, "Mus musculus", "Mouse")) %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1,
                data = correlation_long_all_btm_mouse_human_timepoints_summary %>% 
                  filter(comparison != "Human vs. Human")) +
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.5,
                angle = 90,
                linewidth = 1,
                data = correlation_long_all_btm_mouse_human_timepoints_summary %>% 
                  filter(comparison == "Human vs. Human")) +
  
  
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 2, 6, 12, 24), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(0.4, 1)) +
  theme_vaxgo() +
  scale_color_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = "Correlation between organisms,\nBTMs, E. coli infection",
       x = "Hour",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 

correlation_long_all_btm_mouse_human_timepoints_summary_plot

correlation_long_all_btm_mouse_human_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "correlation_long_all_mouse_human_Ecoli_BTMs_timepoints_summary_plot.png"),
         width = 3,
         height = 5)

correlation_long_all_btm_mouse_human_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "correlation_long_all_mouse_human_Ecoli_BTMs_timepoints_summary_plot_square.png"),
         width = 3,
         height = 3)

correlation_long_all_btm_mouse_human_timepoints_summary_plot
  
```









#Performance tests
## ROC Curve
Do for one direction (up and then down)
Do one sided t test for mouse, fit, and human
Positive control: Human
Neg control: Random permutation


Import data
```{r}
#ImmuneGO ontologies
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))

#Import DEGs table
human_mouse_ecoli_log2fc_avg =
  readRDS(file = here("tables", "human_mouse_ecoli_log2fc_avg.rds"))

dge_long_genes = human_mouse_ecoli_log2fc_avg  %>% 
  rename(log2fc = mean_log2fc, se = se_log2fc, sd = sd_log2fc, adj_p_val = p_adj, symbol = human_symbol) %>% 
  mutate(timepoint = if_else(timepoint == "Hospitalization", "24", timepoint)) %>% 
  filter(timepoint == "24")

```

Calculate ROC curve
```{r fig.height=5, fig.width=4}
set.seed(123)

######### Input -----

#All genes
# genes_df_input = dge_long_genes 
# filename = "EcoliInfection_Mouse24h_HumanHospitalization_AllGenes"
# filename_title ="Ecoli Infection,\nMouse (24h) to Human (Hospitalization),\nAll"

#Immune
genes_df_input = dge_long_genes 
  #Filter immune genes only
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))
filename = "EcoliInfection_Mouse24h_HumanHospitalization_Immune"
filename_title ="Ecoli Infection,\nMouse (24h) to Human (Hospitalization),\nImmune genes"

######### UP -----------
#Do one sided t-test for each model
dge_long_genes_up = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = F), #onesided, p(gene) > 0 (right tail)
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj < 0.05, "up", "not_up")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj)

#Assign human log2fc direction as truth
truth_human = dge_long_genes_up %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism",
              values_from = "direction") %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#Generate comparable distributions
perfect_prediction = truth_human %>%
  mutate(prob = adj_p_val) %>%
  rename(pred = truth)

uniform_prediction = perfect_prediction %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()),
         organism = "Uniform")

#Get only up and not up genes from models
models_up = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = F), #one sided, higher
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj < 0.05, "up", "not_up")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj) %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         prob = adj_p_val) %>% 
  #Add controls
  bind_rows(
    perfect_prediction,
    uniform_prediction
    )

#Get probabilities (UP)
classified_df_probs_up = truth_human %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(symbol, timepoint, pred, prob, model = organism),
             by = join_by("symbol", "timepoint")) %>% 
  mutate(not_prob = 1 - prob,
         sum = prob + not_prob,
         truth = as.factor(truth)) %>% 
  distinct()


#Get ROC
roc_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Human", "Mouse", "Uniform", "FIT")) %>%
  select(symbol, timepoint, truth, pred, model, prob, not_prob, sum) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Up vs Non-Up")

classified_df_probs_up %>% 
  dplyr::count(model, timepoint)

######### DOWN -----------
#Do one sided t-test for each model
dge_long_genes_down = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = T), #onesided, p(gene) < 0 (left tail)
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj < 0.05, "down", "not_down")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj)

#Assign human log2fc direction as truth
truth_human = dge_long_genes_down %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism",
              values_from = "direction") %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#Generate comparable distributions
perfect_prediction = truth_human %>%
  #Invert probabilities
  mutate(prob = 1 - adj_p_val) %>%
  rename(pred = truth)
uniform_prediction = perfect_prediction %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()),
         organism = "Uniform")


#Get only up and not up genes from models
models_down = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = T), #one sided, less
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    #Invert probabilities
    pval_one_sided_adj = 1 - pval_one_sided_adj,
    direction = if_else(pval_one_sided_adj < 0.05, "down", "not_down")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj) %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         prob = adj_p_val) %>% 
  #Add controls
  bind_rows(perfect_prediction, uniform_prediction)

#Get probabilities (DOWN)
classified_df_probs_down = truth_human %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(symbol, timepoint, pred, prob, model = organism),
             by = join_by("symbol", "timepoint")) %>% 
  mutate(not_prob = 1 - prob,
         sum = prob + not_prob,
         truth = as.factor(truth)) %>% 
  distinct()


#Get ROC
roc_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Human", "Mouse", "Uniform", "FIT")) %>%
  select(symbol, timepoint, truth, pred, model, prob, not_prob, sum) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Down vs Non-Down")

#Calculate AUC --------
#Get AUC
auc_results = classified_df_probs_down %>%
  group_by(model, timepoint) %>%
  yardstick::roc_auc(truth, prob) %>% 
  mutate(comparison = "Down vs Non-Down") %>% 
  bind_rows(
    classified_df_probs_up %>%
      group_by(model, timepoint) %>%
      yardstick::roc_auc(truth, prob) %>%
      mutate(comparison = "Up vs Non-Up")) %>% 
  mutate(.estimate = round(.estimate, 2)) %>% 
  rename(roc_auc = .estimate)

#Merge data -------
roc_curve_data = roc_curve_up_notup %>% 
  bind_rows(roc_curve_down_notdown) %>% 
  left_join(auc_results, 
            by = c("model", "comparison", "timepoint")) %>% 
  mutate(condition = "E. coli infection",
         filename = filename)

roc_curve_data %>% 
  saveRDS(file = here("tables", paste0("roc_curve_data_", filename, ".rds")))

#PLOT ROC Curve
roc_curve_plot = roc_curve_data %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(size = 1, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_grid(~fct_rev(comparison)~paste0(timepoint, "h")) + 
  scale_color_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "FIT"   = "#4361ee",
    "Human" = "#90A4AEFF",
    "Uniform" = "gray80"
    )) +
  labs(
    title = filename_title, 
  x = "FPR",
  y = "TPR",
  color = "Model") + 
  theme(
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10,),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3)

roc_curve_plot

roc_curve_plot %>%
  ggsave(filename = here("Figures", paste0("roc_curve_plot_", filename, ".png")),
         width = 4,
         height = 5)

```



# Performance tests 

## ROC Curve 

Do for one direction (up and then down) Do one sided t test for mouse,
fit, and human Positive control: Human Neg control: Random permutation

Import data

```{r}
#Import DEGs table
dge_long_genes = read_rds(file= here("tables", "dge_fit_mouse_human_long.rds")) 
#Import metadata
human_mouse_fluad_metadata = readRDS(file = here("tables", "human_mouse_fluad_metadata.rds"))
#ImmuneGO ontologies
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))


dge_long_genes = dge_long_genes %>% 
  filter(organism != "FIT") %>% 
  inner_join(human_mouse_fluad_metadata %>%
               group_by(organism, timepoint) %>%
               summarize(n = n()) %>% 
               mutate(timepoint = as.numeric(timepoint)),
               by = join_by("organism", "timepoint"))
#All genes
genes_roc_allgenes = dge_long_genes
#Filter immune genes
genes_roc_allgenes_immunegenes = dge_long_genes %>% 
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))

#Only DEGs
degs_genes_human = dge_long_genes %>%
  filter(adj_p_val <= 0.05,
         organism == "Human") %>%
  select(symbol, timepoint)

#Only immune DEGs
genes_roc_DEGs_all = dge_long_genes %>%
  inner_join(degs_genes_human,
             by = join_by("symbol", "timepoint"))

#Only immune DEGs
genes_roc_DEGs_immune = dge_long_genes %>%
  inner_join(degs_genes_human,
             by = join_by("symbol", "timepoint")) %>%
  #Filter immune genes only
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))
```

Calculate ROC curve

Inputs
```{r fig.height=4, fig.width=10}
set.seed(123)
######### Input -----
# filename = "Fluad_All_OnlyImmune"
# genes_df_input = bind_rows(
#   genes_roc_allgenes %>% mutate(model_type = "Mouse"),
#   genes_roc_allgenes_immunegenes %>% mutate(model_type = "Immune")
# )


filename = "Fluad_AllDEGs_HumanDEGs_OnlyImmune"
genes_df_input = bind_rows(
  genes_roc_DEGs_all %>% mutate(model_type = "Mouse"),
  genes_roc_DEGs_immune %>% mutate(model_type = "Immune")
)

```



### Using t statistic to compute probabilities
```{r fig.height=5, fig.width=8}

#=====================================================
# Compute probabilities
#=====================================================

dge_long_genes_up = genes_df_input %>% 
  mutate(
    df = n - 1,
    pval_one_sided = pt(t, df, lower.tail = FALSE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t > 0, "up", "not_up"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  select(-adj_p_val) %>%
  rename(adj_p_val = pval_one_sided_adj)

dge_long_genes_down = genes_df_input %>% 
  mutate(
    df = n - 1,
    pval_one_sided = pt(t, df, lower.tail = TRUE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t < 0, "down", "not_down"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  select(-adj_p_val) %>%
  rename(adj_p_val = pval_one_sided_adj)

#=====================================================
# Assign human truth
#=====================================================
truth_human_up = dge_long_genes_up %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

truth_human_down = dge_long_genes_down %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#=====================================================
#Generate reference distributions
#=====================================================
perfect_prediction_up = truth_human_up %>%
  mutate(prob = if_else(truth == "up", 1, 0),
         model = "Human") %>%
  rename(pred = truth)

uniform_prediction_up = perfect_prediction_up %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()), organism = "Uniform",
         model = "Uniform")

perfect_prediction_down = truth_human_down %>%
  mutate(prob = if_else(truth == "down", 1, 0),
          model = "Human") %>%
  rename(pred = truth)

uniform_prediction_down = perfect_prediction_down %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()), organism = "Uniform",
         model = "Uniform")

#=====================================================
# Combine models (Mouse + Immune + Uniform + Perfect)
#=====================================================
models_up = dge_long_genes_up %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = if_else(model_type == "Immune", "Immune", organism)) %>%
  bind_rows(perfect_prediction_up, uniform_prediction_up)

models_down = dge_long_genes_down %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = if_else(model_type == "Immune", "Immune", organism)) %>%
  bind_rows(perfect_prediction_down, uniform_prediction_down)

#=====================================================
# Classify probabilities 
#=====================================================
classified_df_probs_up = truth_human_up %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(symbol, timepoint, pred, prob, model),
             by = join_by(symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct()

classified_df_probs_down = truth_human_down %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(symbol, timepoint, pred, prob, model),
             by = join_by(symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct()


#=====================================================
#Compute ROC 
#=====================================================

roc_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Up")

roc_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Down")

#=====================================================
#Compute AUC 
#=====================================================
auc_results_up = classified_df_probs_up %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(model, timepoint, auc, comparison)

auc_results_down = classified_df_probs_down %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(model, timepoint, auc, comparison)

#Plot curves ------------------
#PLOT ROC Curve
roc_curve_plot = roc_curve_up_notup %>% 
  bind_rows(roc_curve_down_notdown) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(size = 1, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_grid(~fct_rev(comparison)~paste("Day", timepoint)) + 
  scale_color_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "Human" = "#90A4AEFF",
    "Immune" = "#4361ee",
    "Uniform" = "gray80"
    )) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model") + 
  theme(
    legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3)


# Visualize AUC
auc_plot <- bind_rows(auc_results_up, 
                      auc_results_down) %>% 
  mutate(model = factor(model, levels = c("Perfect", "Mouse", "Immune", "Uniform")),
         comparison = factor(comparison, levels = c("Up", "Down"))) %>%
  
  ggplot(aes(x = auc, y = factor(timepoint) %>% fct_rev(), fill = model)) +
  geom_col(width = 0.8,
           position = position_dodge()) +
  
  # Add a reference line for random prediction (AUC = 0.5)
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black", linewidth = 0.5) +
  
  geom_text(aes(label = round(auc, 2)),
            position = position_dodge2(0.9),
            hjust = 1.2, size = 2) +
  
  facet_grid(comparison ~ ., scales = "free_y") +
  
  labs(
    x = "AUC",
    y = "Day",
    fill = "Model"
  ) +
    scale_fill_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "Immune" = "#4361ee",
    "Human" = "#90A4AEFF",
    "Uniform" = "gray80"
    )) +
  theme_vaxgo()  +
  theme(
    legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(expand = c(0,0), limits = c(0,1))

roc_auc_curve_plot = ((roc_curve_plot) |
  auc_plot) +
  plot_layout(guides = "collect", widths = c(1, 0.4))+
  plot_annotation(theme = theme(legend.position = "top", plot.title = element_text(hjust = 0.5)),
                  title = filename)

roc_auc_curve_plot %>%
  ggsave(filename = here("Figures", paste0("roc_curve_plot_", filename, ".png")),
         width = 8,
         height = 5)

roc_auc_curve_plot
```



