---
title: "S. aureus"
author: "Wasim Aluísio Prates-Syed"
date: "2025-07-23"
output: 
  html_document: 
    toc: TRUE
    toc_depth: 4
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE, cache = TRUE)
```

Import required codes
```{r message=FALSE, warning=FALSE}
# install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```



# S. aureus
## Mouse: GSE19668

### Retrieve dataset
Retrieve Exprs and Metadata from GEO
```{r}
#S aureus: GSE19668 -----
filename = "saureus_infection_GSE19668"
GSE19668 <- getGEO("GSE19668", GSEMatrix = TRUE) 
eset <- GSE19668[[1]]

#Metadata ------
GSE19668_metadata = eset %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    subject_id = "title",
    # sex = "gender:ch1", 
    treatment = "infection:ch1", 
    # tissue = "tissue:ch1",
    timepoint = "time point:ch1",
    # animal_replicate = "animal replicate:ch1",
    strain = "strain:ch1") %>% 
  mutate(treatment = fct_relevel(treatment, "none"),
         timepoint = fct_relevel(timepoint, "0"),
         condition = paste(treatment, strain, timepoint, sep = "_"))

#Save
GSE19668_metadata %>% saveRDS(file = here("tables", paste0(filename, "_metadata.rds")))


#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% as.data.frame() %>% saveRDS(here("tables", paste0(filename, "_exprs.rds")))
```

Clean metadata
```{r}
#Import
GSE19668_metadata = readRDS(file = here("tables", "saureus_infection_GSE19668_metadata.rds"))

GSE19668_metadata %>% 
  select(strain)

#Filter metadata
GSE19668_metadata_filtered = GSE19668_metadata %>%
  filter(strain == "C57BL/6J") %>% 
  separate(subject_id, into = c("tissue", "strain1", "timepoint1", "participant_id"), sep = " ") %>% 
  select(!c(tissue:timepoint1)) %>% 
  arrange(participant_id, timepoint) %>% 
  separate(timepoint, into = c("timepoint_number", "time_unit")) %>% 
  mutate(time_unit = "hours",
         timepoint = timepoint_number %>% as.numeric(),
         organism = "Mouse") %>% 
  arrange(timepoint) %>% 
  select(-timepoint_number) %>% 
  select(geo_sample, participant_id, condition, timepoint, time_unit, everything())

#Save
GSE19668_metadata_filtered %>% saveRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

```

## Annotate dataset
```{r}
#Import
saureus_mouse_exprs = readRDS(here("tables","saureus_infection_GSE19668_exprs.rds"))
saureus_mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

#Inputs
filename = "saureus_infection_mouse"
mouse_exprs = saureus_mouse_exprs 
mouse_metadata = saureus_mouse_metadata
```


```{r}

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
mouse_exprs_metadata  = mouse_exprs %>% 
  select(c(mouse_metadata$geo_sample)) %>%
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = c(mouse_metadata$geo_sample),
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(mouse_metadata, 
             by = "geo_sample") 


#Get value from day 0 and calculate Log2FC
mouse_exprs_log2fc_sample = mouse_exprs_metadata %>% 
  filter(timepoint == 0) %>% 
  select(probe_id, participant_id, day0_value_raw = value_raw) %>% 
  #Filter batch
  inner_join(mouse_exprs_metadata %>% 
               filter(timepoint != "0"),
             by = join_by("probe_id", "participant_id")) %>% 
  mutate(log2fc = value_raw - day0_value_raw) %>% 
  distinct() 

#Check duplicates
mouse_exprs_log2fc_sample %>% 
  filter(participant_id == "rep1",
         probe_id == "1415670_at")


mouse_exprs_log2fc_sample %>% 
  saveRDS(here("tables", "saureus_infection_mouse_log2fc_sample_clean_long.rds"))
```


```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pull(probe_id) %>% unique()

# listAttributes(ensembl_mart)
#Annotate
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2", #Probes
  values = gene_id,
  mart = ensembl_mart)

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_log2summary = mouse_exprs %>% #20,845 probes
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_mouse430_2), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = mouse_log2summary %>% 
  #filter ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

#Filter out probes with multiple mgi and hgnc symbols
probe_dupes = mouse_log2summary %>% get_dupes(probe)

mouse_log2summary = mouse_log2summary %>% 
  select(probe, hgnc_symbol) %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique()),
         !probe %in% c(probe_dupes$probe %>% unique())) 

#Modify expression matrix

#Filtered probes
mouse_log2_filtered_probes = mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
  inner_join(mouse_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol,
         probe,
         c(mouse_metadata %>% 
             pull(geo_sample)))
#Check
mouse_log2_filtered_probes %>% dim()
mouse_log2_hgnc_symbol = mouse_log2_filtered_probes 
mouse_log2_hgnc_symbol %>% get_dupes(probe)

#Save
mouse_log2_hgnc_symbol %>% saveRDS(file = here("tables","saureus_infection_mouse_exprs_hgnc_symbol.rds"))
```

Pivot data and create expression set
```{r}
#Import
mouse_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_mouse_exprs_hgnc_symbol.rds"))
mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

#Pivot to long and wide format, and annotate
mouse_log2_samples_log2fc_long = mouse_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(mouse_metadata,
             by = "geo_sample")


mouse_log2_samples_log2fc_wide = mouse_log2_samples_log2fc_long %>% 
  select(human_symbol, probe, participant_id, timepoint, log2fc) %>%
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc") %>% 
  mutate(timepoint = as.character(timepoint) %>% fct_reorder(timepoint %>% as.numeric()))


#Create Expression Set
eset_mouse <- ExpressionSet(
  assayData = mouse_log2_hgnc_symbol %>% 
    select(-probe) %>% 
    column_to_rownames("human_symbol") %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(mouse_metadata))

#Save
eset_mouse %>% saveRDS(file = here("tables", "saureus_infection_mouse_eset.rds"))

```


### Differential expression 

#### Limma
Paired samples design
```{r}

#Import
mouse_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_mouse_exprs_hgnc_symbol.rds"))
mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))
eset_mouse = readRDS(file = here("tables", "saureus_infection_mouse_eset.rds"))

#Create model matrix
design <- model.matrix(~ condition + 0, # 0 defines the way we compare the samples
                       eset_mouse) 

colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

# Estimate the correlation between measurements made on the same subject (paired-samples)
corfit <- duplicateCorrelation(eset_mouse, 
                               design = design, 
                               block = eset_mouse$participant_id)
corfit$consensus #0.001357108

#Fit Limma model
fit <- lmFit(eset_mouse, 
             design, 
             block = eset_mouse$participant_id, 
             correlation = corfit$consensus) #Use inter-subject correlation as input into the linear model fit

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(conditionStaphylococcus.aureus_C57BL.6J_2.hours - conditionnone_C57BL.6J_0,
                             conditionStaphylococcus.aureus_C57BL.6J_4.hours - conditionnone_C57BL.6J_0,
                             conditionStaphylococcus.aureus_C57BL.6J_6.hours - conditionnone_C57BL.6J_0,
                             conditionStaphylococcus.aureus_C57BL.6J_12.hours - conditionnone_C57BL.6J_0,
                             levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs = all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "conditionnone_C57BL.6J_0", "0")) %>% 
  mutate(condition = case_when(comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_2.hours"	~ "Saureus_2h",
                               comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_4.hours"	~ "Saureus_4h",
                               comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_6.hours"	~ "Saureus_6h",
                               comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_12.hours"	~ "Saureus_12h",
                               T~comparisonA),
         organism = "Mouse",
         timepoint = str_remove(condition, "Saureus_") %>% str_remove("h")) %>% 
  select(condition, everything()) %>% 
  clean_names() %>% 
  rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs %>% saveRDS(file = here("tables", "saureus_infection_mouse_dge_limma_degs.rds"))

all_degs %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarize(n_degs = n())

```



##Import


```{r}
mouse_degs = readRDS(file = here("tables", "saureus_infection_mouse_dge_limma_degs.rds"))
```



#### Correlation plots, pairwise
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, GSE19668, C57BL/6J", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_alldays
#Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)
beep()



#Timepoint 2
timepoint_select = 2
mouse_correlation_ggpairs_day2 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == 2) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day2
#Save
mouse_correlation_ggpairs_day2 %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_day2.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 6
timepoint_select = 6
mouse_correlation_ggpairs_day6 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day6
#Save
mouse_correlation_ggpairs_day6 %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_day6.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 12
timepoint_select = 12
mouse_correlation_ggpairs_day12 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day12
#Save
mouse_correlation_ggpairs_day12 %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_day12.png"),
         height = 10,
         width = 10)
beep()

```




#### PCA

```{r fig.height=4, fig.width=5}
#Import metadata
annotation_samples_pca = GSE19668_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = GSE19668_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, entrezgene_id, log2fc) %>% 
  filter(!entrezgene_id == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("entrezgene_id") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Mouse, S. aureus, GSE19668",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

```

#### BTMs
```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)

#Annotate genes by BTM
GSE19668_genes_btm = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  rename(mm_entrez = entrezgene_id) %>% 
  select(mgi_symbol, symbol = human_symbol, participant_id, geo_sample, timepoint, treatment, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = GSE19668_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process
```



##### GGPairs
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

mouse_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) 

mouse_correlation_ggpairs_btms_gsea_alldays

mouse_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_gsea_btm_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)


#All days, colored by group
mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor = ggpairs_data_dge_btm_mean_process %>%
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = group,
                                 fill = group),
          title = paste("Day", timepoint_selected),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors)

mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor

mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_gsea_btm_correlation_ggpairs_alldays_groupcolor.png"),
         height = 10,
         width = 10)

beep()

```




##### Ggplot manual
```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(3:7) %>% 
  colnames()


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == 2) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == 2) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = "GSE19668, Mouse, SAureus infeciton, random sample")
```













# Human: GSE33341

### Prepare data
```{r}
#S. aureus: GSE33341 -----
filename = "GSE33341"
GSE <- getGEO("GSE33341", GSEMatrix = TRUE) 
eset <- GSE[[2]] #The matrix #1 the mouse annotated

#Metadata ------
GSE_metadata = eset %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    participant_id = "title",
    sex = "host gender:ch1",
    condition = "source_name_ch1",
    pathogen = "pathogen:ch1",
    description = "source_name_ch1", 
    ethnicity = "ethnicity:ch1",
    tissue_infection = "anatomic site of infection:ch1",
    batch_experimental = "experimental batch:ch1",
    age = "host age (yr):ch1") %>% 
  mutate(condition = pathogen,
         organism = "Human",
         timepoint = 1,
         time_unit = "Day",
         tissue = "Blood",
         participant_id = str_extract(participant_id, "\\[.*?\\]") %>% 
           str_remove("\\[") %>% 
           str_remove("\\]"),
         condition = case_when(description == "blood patient with E. coli bacteremia prior antibiotic treatment (Emergency Department)" ~ "E. coli prior antibiotic",
                               description == "blood patient with S. aureus bacteremia prior antibiotic treatment (Emergency Department)" ~ "S aureus prior antibiotic",
                               description == "blood human NOT infected" ~ "No infection",
                               T ~ description),
         pathogen = if_else(pathogen == "-", "Healthy", pathogen))
#Save
GSE_metadata %>% saveRDS(file = here("tables", "saureus_infection_human_metadata.rds"))

#Filter metadata
GSE_metadata_filtered = GSE_metadata %>%
  filter(pathogen %in% c("Staphylococcus aureus", "Healthy")) %>% 
  select(geo_sample, participant_id, condition, timepoint, everything()) %>% 
  mutate(ethnicity = str_to_title(ethnicity)) %>% 
  filter(batch_experimental != "2739",
         ethnicity != "Unknown",
         age != "unknown")

GSE_metadata_filtered %>% 
  saveRDS(file = here("tables", "saureus_infection_human_metadata.rds"))



#Expression data ------------
GSE_exprs <- exprs(eset) %>% as.data.frame()

#Save
GSE_exprs %>% saveRDS(file = here("tables", "saureus_infection_human_exprs.rds"))
```

Prepare data
```{r}
#Import
saureus_human_exprs = readRDS(file = here("tables", "saureus_infection_human_exprs.rds"))
saureus_human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))


#Inputs
filename = "saureus_infection_human"
human_exprs = saureus_human_exprs 
human_metadata = saureus_human_metadata
```


```{r}
#Annotate genes ------
#Human -----
#Annotate data and calculate log2fc
human_exprs_metadata  = human_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = -probe_id,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(human_metadata, 
             by = "geo_sample") 

#Get value from healthy and calculate Log2FC
human_exprs_log2fc_sample = human_exprs_metadata %>% 
  filter(pathogen == "Healthy") %>% 
  select(probe_id, value_raw_ref = value_raw) %>% 
  group_by(probe_id) %>% 
  summarize(value_raw_ref = mean(value_raw_ref)) %>% 
  inner_join(human_exprs_metadata %>% 
               filter(pathogen != "Healthy"),
             by = join_by("probe_id")) %>% 
  filter(condition != "No infection") %>% 
  mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

#Check duplicates
human_exprs_log2fc_sample %>% 
  filter(participant_id == "CSD4620.CEL",
         probe_id == "1007_s_at")

human_exprs_log2fc_sample %>% 
  saveRDS(here("tables", "saureus_infection_human_log2fc_sample_clean_long.rds"))
```


```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
# listDatasets(ensembl_mart)
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- human_exprs_log2fc_sample$probe_id %>% unique()

#Annotate
annot <- getBM(
  attributes = c("affy_hg_u133a_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133a_2",
  values = gene_id,
  mart = ensembl_mart) %>% 
  #Standardize
  rename(probe = affy_hg_u133a_2,
                     entrezgene_id = entrezgene_id,
                     human_symbol = hgnc_symbol,
                     ensembl_id = ensembl_gene_id)

#Filter genes with highest median expression
human_log2summary = human_exprs %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot, 
            by = "probe") %>% 
  #Filter ensembl id
  group_by(ensembl_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup()  


#Filter out genes with multiple hgnc_symbols
hgnc_symbol_dupes = human_log2summary %>% 
  #filter ensembl id with multiple hgnc_symbols
  get_dupes(human_symbol)

human_log2summary = human_log2summary %>% 
  filter(!human_symbol %in% c(hgnc_symbol_dupes$human_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
human_log2_filtered_probes = human_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(human_log2summary, by = "probe") %>%   #214,396
  select(human_symbol, probe,
         c(human_metadata$geo_sample))
#14395
human_log2_filtered_probes %>% dim()

human_log2_filtered_probes = human_log2_filtered_probes 

human_log2_filtered_probes %>% 
  get_dupes(human_symbol)

#Save 
human_log2_filtered_probes %>% 
  saveRDS(here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
```


```{r}
#Import
human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))

#Pivot to long and wide format, and annotate
human_log2_samples_log2fc_long = human_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(human_metadata,
             by = "geo_sample")


human_log2_samples_log2fc_wide = human_log2_samples_log2fc_long %>% 
  select(human_symbol, probe, participant_id, timepoint, log2fc) %>%
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc") %>% 
  mutate(timepoint = as.character(timepoint) %>% fct_reorder(timepoint %>% as.numeric()))


#Create Expression Set
human_metadata_eset = human_metadata %>% 
  mutate(age = as.numeric(age) %>% scale(),
         timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         pathogen = fct_relevel(pathogen, "Healthy"),
         batch_experimental = as.factor(batch_experimental),
         ethnicity = as.factor(ethnicity)) 

eset_human <- ExpressionSet(
  assayData = human_log2_hgnc_symbol %>% 
    select(-probe) %>% 
    column_to_rownames("human_symbol") %>% 
    select(c(human_metadata_eset$geo_sample)) %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(human_metadata_eset))

#Save
eset_human %>% saveRDS(file = here("tables", "saureus_infection_human_eset.rds"))

```






### Differential expression

#### Limma

```{r}
#Import
human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))
eset_human = readRDS(file = here("tables", "saureus_infection_human_eset.rds"))

#Create model matrix
design <- model.matrix(~ age + sex + ethnicity + batch_experimental + pathogen,
                       data = pData(eset_human))

colnames(design) <- make.names(colnames(design))  
rownames(design) <- rownames(pData(eset_human))

#Fit Limma model
fit <- lmFit(eset_human, 
             design) 

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(pathogenStaphylococcus.aureus, levels = design)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- limma::eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_human = all_contrasts_degs %>% 
  dplyr::select(comparison, everything()) %>% 
  dplyr::mutate(condition = case_when(comparison == "pathogenStaphylococcus.aureus"	~ "Saureus_Day1",
                               T~comparison),
         organism = "Human",
         timepoint = "1") %>% 
  dplyr::select(condition, everything()) %>% 
  clean_names() %>% 
  dplyr::rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs_human %>% saveRDS(file = here("tables", "saureus_infection_human_dge_limma_degs.rds"))

all_degs_human %>% 
  dplyr::group_by(condition) %>%
  dplyr::filter(adj_p_val <= 0.05) %>% 
  dplyr::summarize(n_degs = n())

```


#### t-Test
```{r}
#Import ================
human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))

#Wrangle ==============
#Pivot to long and wide format, and annotate
human_log2_samples_log2fc_long = human_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(human_metadata,
             by = "geo_sample")

human_log2_wide = human_log2_hgnc_symbol %>%
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(human_metadata %>% 
               select(geo_sample, participant_id, pathogen),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_")

human_exprs_log2_infected_healthy = human_exprs_metadata %>% 
  filter(pathogen == "Healthy") %>% 
  select(probe_id, value_raw_ref = value_raw) %>% 
  group_by(probe_id) %>% 
  summarize(value_raw_ref = mean(value_raw_ref)) %>% 
  inner_join(human_exprs_metadata %>% 
               filter(pathogen != "Healthy"),
             by = join_by("probe_id")) %>% 
  filter(pathogen != "Healthy") %>% 
  inner_join(human_log2_samples_log2fc_long %>% 
               select(probe_id = probe, human_symbol),
             by = join_by("probe_id")) %>% 
  mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

# Unpaired t-test ----------
human_log2fc_unpaired_ttest <- human_exprs_log2_infected_healthy %>%
  group_by(human_symbol) %>%
  
  summarise(
    pval = tryCatch(
      t.test(value_raw, value_raw_ref,
             paired = FALSE, var.equal = FALSE, 
             alternative = "two.sided")$p.value,
      error = function(e) NA_real_
    ),
    .groups = 'drop'
  ) %>%
  mutate(
    padj = p.adjust(pval, method = "fdr")
    ) %>% 
  mutate(
    condition = "Saureus infection",
    timepoint = "1",
    test = "t-Test",
    paired = "Unpaired")


human_log2fc_unpaired_ttest %>% saveRDS(file = here("tables", "saureus_infection_human_dge_unpaired_ttest.rds"))

```







#### Import
```{r}
#Import
GSE33341_counts_log2_summary_annotated_probes = 
  readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol_long.rds"))

human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))



#Pivot to long and wide format, and annotate
human_log2_samples_log2fc_long = human_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(human_metadata,
             by = "geo_sample")


human_log2_samples_log2fc_wide = human_log2_samples_log2fc_long %>% 
  select(human_symbol, probe, participant_id, timepoint, log2fc) %>%
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc") %>% 
  mutate(timepoint = as.character(timepoint) %>% fct_reorder(timepoint %>% as.numeric()))

#Pivot wider
GSE33341_log2fc_sample_wide = GSE33341_counts_log2_summary_annotated_probes %>%
  select(human_symbol, participant_id, timepoint, log2fc) %>% 
  pivot_wider(names_from = participant_id,
              values_from = log2fc) %>% 
  ungroup()
```





# Human vs Mouse

## Combine datasets

```{r}
#Import results =======
all_degs_mouse = read_rds(here("tables", "saureus_infection_mouse_dge_limma_degs.rds"))
all_degs_human = read_rds(here("tables", "saureus_infection_human_dge_limma_degs.rds"))
mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))
mouse_log2fc_sample = readRDS(here("tables", "saureus_infection_mouse_log2fc_sample_clean_long.rds")) 
mouse_exprs = readRDS(file = here("tables", "saureus_infection_mouse_exprs_hgnc_symbol_long.rds"))
human_log2fc_sample = readRDS(here("tables", "saureus_infection_human_log2fc_sample_clean_long.rds")) 
human_exprs = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))



#DGE results ===========
all_degs_human_mouse = all_degs_mouse %>% 
  bind_rows(all_degs_human) %>% 
  select(condition, timepoint, human_symbol, mean_l2fc:organism)
all_degs_human_mouse %>% saveRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds"))


#Metadata ===========
human_mouse_metadata = mouse_metadata %>% bind_rows(human_metadata)
human_mouse_metadata %>% saveRDS(file = here("tables", "saureus_infection_human_mouse_metadata.rds"))

#Merge sample log2fc =============
mouse_log2fc_sample_symbol = mouse_log2fc_sample %>% 
  inner_join(mouse_exprs %>% 
               ungroup() %>% 
               select(human_symbol, probe_id) %>% 
               distinct(),
             by = "probe_id") %>% 
  mutate(organism = "Mouse",
         subject_id = participant_id)

human_symbol_dupes = mouse_log2fc_sample_symbol %>% 
select(organism, timepoint, participant_id, human_symbol, log2fc) %>% 
distinct() |>
dplyr::summarise(n = dplyr::n(), .by = c(human_symbol, participant_id, timepoint)) |>
dplyr::filter(n > 1L)

mouse_log2fc_sample_symbol = mouse_log2fc_sample_symbol %>% 
  filter(!human_symbol %in% c(human_symbol_dupes$human_symbol))

human_log2fc_sample_symbol = human_log2fc_sample %>% 
  inner_join(human_exprs %>% 
               select(human_symbol, probe_id = probe) %>% 
               distinct(),
             by = "probe_id") %>% 
  mutate(organism = "Human",
         timepoint = 24,
         subject_id = participant_id) 

human_mouse_log2fc_samples = bind_rows(mouse_log2fc_sample_symbol, human_log2fc_sample_symbol) %>% 
  ungroup()

#Check duplicates and remove
human_symbol_dupes = human_mouse_log2fc_samples %>% 
select(organism, timepoint, participant_id, geo_sample, human_symbol, log2fc) %>% 
distinct() |>
dplyr::summarise(n = dplyr::n(), .by = c(human_symbol, participant_id, timepoint)) |>
dplyr::filter(n > 1L)

human_mouse_log2fc_samples$organism %>% unique()

human_mouse_log2fc_samples %>% 
  saveRDS(here("tables", "saureus_infection_human_mouse_samples_log2fc.rds"))


```



## Get overlap

Input

```{r}
# Import datasets
dge_long_genes = read_rds(file = here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds"))
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

# DEGs table
degs_all_updown = dge_long_genes %>% 
  dplyr::filter(organism %in% c("Human", "Mouse")) %>% 
  dplyr::filter(adj_p_val <= 0.05) %>% 
  dplyr::mutate(condition = paste0(organism, "_", timepoint)) %>% 
  dplyr::select(process = condition, genes = human_symbol)

# Metadata
annotation_metadata = dge_long_genes %>% 
  #Arrange columns in the desired order to be displayed on the heatmap
  dplyr::select(condition, 
                organism, 
                # type, 
                # vaccine, 
                timepoint) %>% 
  mutate(condition = paste0(organism, "_", timepoint))

annotation_metadata$condition %>% unique()
#Inputs
filename = "saureus_infection"
```

Perform calculations

```{r}
conflicted::conflicts_prefer(base::intersect)
conflicted::conflicts_prefer(base::setdiff)

# Perform calculations
unique_cond <- unique(degs_all_updown$process) # List sets
shared_genes_df <- data.frame() # Store data

for (i in 1:(length(unique_cond) - 1)) {
  for (j in (i + 1):length(unique_cond)) {
    results_temp <- overlap_genes(unique_cond[i], unique_cond[j], degs_all_updown)
    shared_genes_df <- bind_rows(shared_genes_df, results_temp)
  }
}

# Fisher's exact test
fisher_exact_test <- function(shared, notshared1, notshared2) {
  cont_table <- matrix(c(shared, notshared1, notshared2, 0), nrow = 2)
  results_fisher <- stats::fisher.test(cont_table)
  return(results_fisher$p.value)
}
shared_genes_df = shared_genes_df %>%
  mutate(pvalue = pmap_dbl(list(Shared, NotShared_cond1, NotShared_cond2), fisher_exact_test),
         p_adj = p.adjust(pvalue))


#Mirror table
shared_genes_df2 = shared_genes_df %>% 
  dplyr::rename(Cond1_temp = Cond1, Cond2_temp = Cond2) %>%
  dplyr::rename(Cond1 = Cond2_temp, Cond2 = Cond1_temp)

shared_genes_df_mirror = shared_genes_df %>% 
  dplyr::bind_rows(shared_genes_df2) %>% 
  dplyr::mutate(Percentage_Shared_Cond1 = round(Percentage_Shared_Cond1, 2),
         Percentage_Shared_Cond2 = round(Percentage_Shared_Cond2, 2))
  
#Jaccard distance
jaccard.matrix <- degs_all_updown %>%
  dplyr::mutate(present = 1) %>% 
  dplyr::distinct() %>% 
  pivot_wider(names_from = genes, values_from = present, values_fill = 0) %>%
  column_to_rownames(var = "process") %>%
  vegan::vegdist(binary = TRUE, method = "jaccard", diag = TRUE, upper = TRUE) %>%
  as.matrix() %>%
  {1 - .}
shared_genes_df_mirror = jaccard.matrix %>% 
  as.data.frame() %>% 
  rownames_to_column("Cond1") %>% 
  pivot_longer(cols = -"Cond1",
               names_to = "Cond2",
               values_to = "jaccard_distance") %>% 
  inner_join(shared_genes_df_mirror, by = join_by(Cond1, Cond2)) %>% 
  mutate(jaccard_distance = round(jaccard_distance, 3))

#Save
write_csv(shared_genes_df_mirror, 
          file = here("Tables", paste0(filename, "_sharedgenes_Fisher_Jaccard.csv")))
shared_genes_df_mirror
```

### Heatmap

#### Shared genes

```{r}
shared_genes_df_mirror = read_csv(file = here("Tables", "ecoli_infection_sharedgenes_Fisher_Jaccard.csv"))

matrix_data = shared_genes_df_mirror %>% 
  dcast(Cond1 ~ Cond2, 
        value.var = "Shared", 
        fun.aggregate = mean) %>% 
  column_to_rownames("Cond1") %>%
  mutate_if(is.character, as.numeric) %>% 
  as.matrix() 

# Create a matrix of p-values
pvalue_matrix <- shared_genes_df_mirror %>%
  dcast(Cond1 ~ Cond2,
        value.var = "p_adj",
        fun.aggregate = mean) %>%
  column_to_rownames("Cond1") %>%
  as.matrix()

dim(matrix_data)

annotation_metadata_complex = matrix_data %>% 
  as.data.frame() %>%
  rownames_to_column("condition") %>% 
  dplyr::select(condition) %>% 
  inner_join(annotation_metadata , by = "condition") %>%
  distinct() 

# merge
annotation_metadata_heatmap = matrix_data %>% 
  colnames() %>% 
  data.frame() %>% 
  dplyr::filter(. != "process") %>% 
  dplyr::rename(condition = '.') %>% 
  dplyr::inner_join(annotation_metadata_complex, by = "condition") %>% 
  dplyr::arrange(condition) %>% 
  column_to_rownames(var= "condition")


# Reorder
matrix_data_ordered = matrix_data %>% 
  as.data.frame() %>% 
  dplyr::select(c(annotation_metadata_complex %>% pull(condition))) %>% 
  rownames_to_column("rownames") %>% 
  dplyr::filter(rownames %in% c(annotation_metadata_complex %>% pull(condition))) %>% 
  column_to_rownames(var= "rownames") %>% 
  as.matrix()

#Check
dim(matrix_data_ordered) 
dim(annotation_metadata_complex)

col_fun <- colorRamp2(c(0, max(matrix_data_ordered, na.rm = T)), c("white", "#00A087FF"))


######## HEATMAP ANNOTATION
annotation_metadata_heatmap = annotation_metadata_heatmap %>% 
  mutate(timepoint = as_factor(timepoint),
         timepoint = fct_reorder(timepoint, as.numeric(timepoint)))

ha = HeatmapAnnotation(df = annotation_metadata_heatmap, 
                       col = colors_immune, 
                       annotation_name_side = "left",
                       show_legend = F)

row_ha = rowAnnotation(df = annotation_metadata_heatmap, 
                       col = colors_immune, 
                       show_annotation_name = F)



mat = matrix_data_ordered
width_image = 15
height_image = 15
width = unit(5, "cm")
height = unit(5, "cm")
column_names_gp = gpar(fontsize = 10)
row_names_gp = gpar(fontsize = 10)
gp = gpar(fontsize = 10)
rect_gp = gpar(col = "gray1", lwd = 0)



fileimageheatmap_grouped = paste0(filename, sep ="_", "Shared_Heatmap.png")

png(
  file = here("Figures", fileimageheatmap_grouped),
  width = width_image,
  height = height_image,
  units = "cm",
  res = 900
)

heatmap_plot_all = Heatmap(mat,
                        top_annotation = ha,
                        left_annotation = row_ha,
                        show_row_names = TRUE,
                        show_heatmap_legend = TRUE,
                        heatmap_legend_param = list(direction = "vertical"),
                        name = "Shared", #Legend
                        col = col_fun, #color
                         cell_fun = function(j, i, x, y, width, height, fill) {
  if (!is.na(mat[i, j]) && mat[i, j] > 0) {
    grid.text(sprintf("%.f", mat[i, j]), x, y, gp = gpar(fontsize = 8, color = "white"))
  }
                           i# Add significance stars based on p-value thresholds
  if (!is.na(pvalue_matrix[i, j])) {
    pval <- pvalue_matrix[i, j]
    sig <- ""
    if (pval < 0.01) {
      sig <- "***"
    } else if (pval < 0.05) {
      sig <- "**"
    } else if (pval < 0.10) {
      sig <- "*"
    }
    
    # Draw the symbol slightly above the cell center
    if (sig != "") {
      grid.text(sig, x, y + unit(2, "mm"), gp = gpar(fontsize = 6, fontface = "plain", color = "white"))
    }
  }
},
                        column_title = fileimageheatmap_grouped,
                        column_title_gp = gpar(fontsize = 15, fontface = "bold"),
                        column_names_rot = 45,
row_names_rot = 0,
                        column_names_gp = column_names_gp,
                        cluster_rows = TRUE,
                        rect_gp = rect_gp,
                        row_names_gp = row_names_gp,
                        # row_split = 2, #Split row clusters
                        row_gap = unit(2, "mm"),
                        cluster_columns = TRUE,
                        na_col = "black",
                        column_split = NULL, #Split group clusters
                        #column_km = 3,
                        column_gap = unit(8, "mm"),
                        show_column_dend = F,
                        show_row_dend = F,
                        row_dend_side = "left",
                        column_dend_side = "top",
                        clustering_distance_rows = "euclidean",
                        border = TRUE,
                        column_dend_reorder = TRUE,
                        width = width, 
                        height = height
)

draw(heatmap_plot_all,  heatmap_legend_side = "left", annotation_legend_side = "left")

dev.off()
```


### Stacked barplot of overlapped genes
```{r fig.width=10, fig.height=4}
#Import datasets
all_degs_human_mouse = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds"))  
ImmuneGO_Mouse_Manual_genes <- read_excel(here("VaxGO", "ImmuneGO_Mouse.xlsx")) %>% 
  separate_rows(genes)


#Compare every mouse timepoint to a human reference
timepoint_reference = "24"
human_data_reference <- all_degs_human_mouse %>% filter(organism == "Human")
new_timepoints <- c(all_degs_human_mouse %>% filter(organism != "Human") %>% pull(timepoint) %>% unique() %>% as.character())
duplicated_human_data <- map_dfr(new_timepoints, ~ {
  df_copy <- human_data_reference
  df_copy <- df_copy %>%
    mutate(timepoint = .x)

  return(df_copy)
})

dge_long_genes_matched = all_degs_human_mouse %>% 
  ungroup() %>% 
  filter(organism != "Human") %>% 
  bind_rows(duplicated_human_data)  %>% 
    mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) 


#Get total genes by each condition
degs_mouse = dge_long_genes_matched %>% 
  filter(adj_p_val <= 0.05,
         organism == "Mouse")
degs_human = dge_long_genes_matched %>% 
  filter(adj_p_val <= 0.05,
         organism == "Human")

degs_mouse_summarize = dge_long_genes_matched %>% 
  filter(adj_p_val <= 0.05) %>% 
  group_by(organism, timepoint) %>% 
  summarize(n_genes = n()) 

#Print
degs_mouse_summarize

#Get overlap -----
#Get symbols
degs_mouse_genes = degs_mouse %>% 
  select(human_symbol, timepoint) %>% 
  distinct()

degs_human_genes = degs_human %>% 
  select(human_symbol, timepoint) %>% 
  distinct()

#Inner join, Mouse and Human ------

#Mouse
degs_human_and_mouse_genes = degs_human_genes %>% 
  inner_join(degs_mouse_genes, 
             by = join_by("human_symbol", "timepoint")) %>% 
  mutate(condition = "Mouse AND Human",
         comparison = "Mouse vs Human",
         organism = "Mouse",
         group = "Common")

#Duplicate rows for human
degs_human_and_mouse_genes_human = degs_human_and_mouse_genes %>% 
  mutate(organism = "Human")


#Anti join, Human vs Mouse -------
#Human
#Human vs Mouse
degs_human_Not_mouse_genes = degs_human_genes %>% 
  anti_join(degs_mouse_genes, 
             by = join_by("human_symbol", "timepoint")) %>% 
  mutate(condition = "Human NOT Mouse", 
         comparison = "Mouse vs Human",
         organism = "Human",
         group = "Not common")

#Models
#Mouse vs Human
degs_mouse_Not_human_genes = degs_mouse_genes %>% 
  anti_join(degs_human_genes, 
             by = join_by("human_symbol", "timepoint")) %>% 
  mutate(condition = "Mouse NOT Human",
         organism = "Mouse",
         comparison = "Mouse vs Human",
         group = "Not common")


#Merge data ------
degs_human_mouse_overlap_genes = 
  bind_rows(degs_human_and_mouse_genes,
            degs_human_and_mouse_genes_human,
            degs_human_Not_mouse_genes,
            degs_mouse_Not_human_genes 
            )

#Classify by same and exclusive DEGs -----
degs_human_mouse_overlap_genes_summary = degs_human_mouse_overlap_genes %>% 
  group_by(condition, timepoint) %>% 
  summarize(n = n(),
            organism,
            comparison,
            group) %>% 
  distinct()

#Print
degs_human_mouse_overlap_genes_summary
```


### Plot overlap (bars)

```{r fig.height=5, fig.width=4}
#Plot overlap
degs_human_mouse_overlap_common_plot = degs_human_mouse_overlap_genes_summary %>%
  ggplot() +
  aes(x = organism,
      y = n,
      fill = fct_rev(group)) + 
  geom_col(alpha = 1) +
  facet_nested(rows = vars(fct_rev(comparison)), 
               cols = vars(timepoint), 
               scales = "free", 
               nest_line = element_line(linetype = 1, color = "black", size = 2),
               independent = "x") +
  scale_y_continuous(n.breaks = 5,
                     expand = c(0.01, 0.01)) +
  scale_fill_manual(values = c("Common" = "#4DBBD5FF",
                               "Not common" = "gray90"))    +
  scale_color_manual(values = c("Common" = "#4DBBD5FF",
                               "Not common" = "gray90"))    +
  theme_vaxgo() +
  theme(
    strip.text = element_text(size = 10),
    strip.text.y = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0, "lines"), 
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "top",
    legend.position.inside = c(0.5, 0.95),
    legend.direction = "horizontal", 
    legend.key.width = unit(0.4, 'cm'),
    legend.key.height = unit(0.4, 'cm'),
    legend.title = element_blank(),
    legend.text = element_text(size = 7, vjust = 0.5, hjust = -0.2)
  ) +
  labs(
    x = "",
    y = "Total",
    color = ""
    ) 
#Print
degs_human_mouse_overlap_common_plot

#Save
degs_human_mouse_overlap_common_plot %>%
  ggsave(filename = here("Figures", "saureus_infection_human_mouse_dge_degs_overlap_barplot.png"),
         width = 4,
         height = 5)
```





## Correlation analysis (TERMINAR)
```{r}
all_degs_human_mouse = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds"))  

#Pivot wider to compare human to different mouse times
#Human
human_mouse_log2fc_avg_wide = 
  
  #Human
  all_degs_human_mouse %>% 
  filter(organism == "Human") %>% 
  mutate(timepoint = "24h") %>% 
  select(human_symbol, 
         timepoint_human = timepoint, 
         mean_log2fc_human = mean_l2fc,
         p_value_human = p_value,
         p_adj_human = adj_p_val) %>% 
  inner_join(
  #Mouse
  all_degs_human_mouse%>% 
  filter(organism == "Mouse") %>% 
  select(human_symbol, 
         timepoint_mouse = timepoint, 
         mean_log2fc_mouse = mean_l2fc,
         p_value_mouse = p_value,
         p_adj_mouse = adj_p_val) %>% 
    mutate(timepoint_mouse = fct_reorder(timepoint_mouse, c(timepoint_mouse %>% as.numeric()))),
  by = "human_symbol") 

human_mouse_log2fc_avg_wide
```


```{r}
#Get immune and non immune genes
human_mouse_log2fc_avg_wide_degs_immune_nonimmune = human_mouse_log2fc_avg_wide %>%
  mutate(gene_type = "all") %>% 
  bind_rows(
  #Classify as immune or non_immune 
    human_mouse_log2fc_avg_wide %>% 
  mutate(gene_type = if_else(human_symbol %in% ImmuneGO_BTM_grouped_genes$genes,
                             "immune",
                             "non_immune"))) %>%
  #Filter DEGs in both organisms, same sign
  filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 &
      sign(mean_log2fc_human) == sign(mean_log2fc_mouse)) %>%
  group_by(timepoint_human, timepoint_mouse, gene_type)

#Summary
degs_immune_nonimmune_summary = human_mouse_log2fc_avg_wide_degs_immune_nonimmune %>% 
  summarize(n_genes = n(), .groups = "drop")

# Calculate the correlation
cor_data <- human_mouse_log2fc_avg_wide_degs_immune_nonimmune %>%
  # filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>%
  group_by(timepoint_human, timepoint_mouse, gene_type) %>%
  summarize(
    cor_test = list(cor.test(mean_log2fc_mouse, mean_log2fc_human,
                             method = "pearson", use = "complete.obs")),
    .groups = "drop"
  ) %>%
  inner_join(degs_immune_nonimmune_summary, by = join_by("timepoint_human", "timepoint_mouse", "gene_type")) %>% 
  mutate(
    cor_value = map_dbl(cor_test, ~ .x$estimate),
    p_value   = map_dbl(cor_test, ~ .x$p.value),
    label_pos_x = map_dbl(cor_test, ~ max(human_mouse_ecoli_log2fc_avg_wide$mean_log2fc_mouse, na.rm = TRUE)),
    label_pos_y = map_dbl(cor_test, ~ max(human_mouse_ecoli_log2fc_avg_wide$mean_log2fc_human + 1, na.rm = TRUE)),
    label = paste0("r = ", round(cor_value, 2), ", p = ", signif(p_value, 2), ", n = ", n_genes))

  cor_data
  
  
```

```{r fig.height=3, fig.width=10}

human_mouse_log2fc_avg_wide_scatterplot = human_mouse_log2fc_avg_wide %>% 
  filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>% 
  
  #Plot
  ggplot() +
  aes(x = mean_log2fc_mouse,
      y = mean_log2fc_human) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #Facet
  facet_grid(~timepoint_human~timepoint_mouse) +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_log2fc_avg_wide %>%
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  stat_smooth(method = "lm", geom = "line", color = "#4DBBD5FF", alpha = 0.5, se = F, size = 1,
              data = human_mouse_ecoli_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05 & sign(mean_log2fc_human) == sign(mean_log2fc_mouse))) +
  
    #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Add correlation metrics (Immune)
  geom_label(
    data = cor_data %>% filter(gene_type == "all"),
    aes(x = 0.1, y = 9, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 0,
    vjust = 1,
    size = 2.5) +
  geom_label(
    data = cor_data %>% filter(gene_type == "immune"),
    aes(x = 0.1, y = 8, label = label),
    inherit.aes = FALSE,
    fill = "#4361ee", 
    hjust = 0,
    vjust = 1,
    size = 2.5) +
  geom_label(
    data = cor_data %>% filter(gene_type == "non_immune"),
    aes(x = 0.1, y = 7, label = label),
    inherit.aes = FALSE,
    fill = "#4DBBD5FF",
    color = "white",
    hjust = 0,
    vjust = 1,
    size = 2.5) +
  
  # #Adjust the axes range
  # ylim(-4, 7) +
  
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "top",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title = "Correlation plot, S. aureus infection, Log2-Fold Change")

human_mouse_ecoli_log2fc_avg_wide_scatterplot

human_mouse_ecoli_log2fc_avg_wide_scatterplot %>% 
  ggsave(filename = here("Figures", "saureus_infection_human_mouse_correlation_mean_scatterplot.png"),
         height = 3,
         width = 10)

```



#### Plot correlation between genes

```{r}
#Import
human_mouse_saureus_log2fc_samples = readRDS(here("tables", "saureus_infection_human_mouse_samples_log2fc.rds"))
all_degs_human_mouse = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds"))
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))
human_mouse_metadata = readRDS(file = here("tables", "saureus_infection_human_mouse_metadata.rds"))

human_mouse_saureus_log2fc_samples = human_mouse_saureus_log2fc_samples %>% 
  mutate(timepoint = as.character(timepoint)) %>% 
  distinct()


human_mouse_log2fc_samples = human_mouse_saureus_log2fc_samples

human_mouse_saureus_log2fc_samples$organism %>% unique()
  
```



```{r}
# #Input
# filename = "saureus_infection_human_mouse"
# gene_set_name = "AllGenes"
# human_mouse_log2fc_samples = human_mouse_saureus_log2fc_samples

#Input
# filename = "saureus_infection_human_mouse"
# gene_set_name = "Immune"
# human_mouse_log2fc_samples = human_mouse_saureus_log2fc_samples  %>%
#   filter(human_symbol %in% c(btm_annotation_genes$symbol %>% unique()))


# #Input
# filename = "saureus_infection_human_mouse"
# gene_set_name = "DEGs"
# human_degs = all_degs_human_mouse %>%
#   filter(organism == "Human", adj_p_val <= 0.05) %>%
#   select(human_symbol)
# 
# human_mouse_log2fc_samples = human_mouse_saureus_log2fc_samples  %>%
#   inner_join(human_degs, by = join_by("human_symbol"))


filename = "saureus_infection_human_mouse"
gene_set_name = "DEGs_Immune"
human_mouse_log2fc_samples = human_mouse_saureus_log2fc_samples  %>%
  inner_join(human_degs, by = join_by("human_symbol")) %>%
  filter(human_symbol %in% c(btm_annotation_genes$symbol %>% unique()))
```



#### Plot correlation between genes (log2fc for every sample)


```{r fig.height=3, fig.width=3}
# --- 1) Mouse vs Human no mesmo timepoint ---
#Get timepoints
timepoints_human <- unique(human_mouse_log2fc_samples %>% filter(organism == "Human") %>% pull(timepoint))
timepoints_mouse <- unique(human_mouse_log2fc_samples %>% filter(organism == "Mouse") %>% pull(timepoint))

#Get combinations
timepoint_combinations <- expand.grid(timepoint_human = timepoints_human,
                                      timepoint_mouse = timepoints_mouse,
                                      stringsAsFactors = FALSE)

# Loop
cor_human_mouse_timepoint <- timepoint_combinations %>%
  pmap_dfr(function(timepoint_human, timepoint_mouse) {
    
    # pegar os participantes do timepoint específico
    human_df <- human_mouse_log2fc_samples %>%
      filter(organism == "Human", timepoint == timepoint_human) %>%
      select(participant_id, human_symbol, log2fc)
    
    mouse_df <- human_mouse_log2fc_samples %>%
      filter(organism == "Mouse", timepoint == timepoint_mouse) %>%
      select(participant_id, human_symbol, log2fc)
    
    # manter apenas genes em comum
    common_genes <- intersect(human_df$human_symbol, mouse_df$human_symbol)
    human_df <- human_df %>% filter(human_symbol %in% common_genes)
    mouse_df <- mouse_df %>% filter(human_symbol %in% common_genes)
    
    # pivot para wide
    human_wide <- human_df %>% pivot_wider(names_from = participant_id, values_from = log2fc)
    mouse_wide <- mouse_df %>% pivot_wider(names_from = participant_id, values_from = log2fc)
    
    # IDs de participantes
    human_ids <- setdiff(colnames(human_wide), "human_symbol")
    mouse_ids <- setdiff(colnames(mouse_wide), "human_symbol")
    
    # calcular correlação para todos pares humano x mouse
    expand.grid(human = human_ids, mouse = mouse_ids, stringsAsFactors = FALSE) %>%
      mutate(
        cor_test = map2(human, mouse, ~{
          x <- human_wide[[.x]]
          y <- mouse_wide[[.y]]
          # remover NAs correspondentes
          keep <- complete.cases(x, y)
          cor.test(x[keep], y[keep], method = "pearson")
        }),
        r       = map_dbl(cor_test, ~.x$estimate),
        p_value = map_dbl(cor_test, ~.x$p.value),
        ci_low  = map_dbl(cor_test, ~.x$conf.int[1]),
        ci_high = map_dbl(cor_test, ~.x$conf.int[2])
      ) %>%
      select(-cor_test) %>%
      mutate(timepoint_human = timepoint_human,
             timepoint_mouse = timepoint_mouse)
  }) %>% 
  group_by(timepoint_human, timepoint_mouse) %>%
  summarize(
    comparison = "Human vs. Mouse",
    mean = mean(r, na.rm = TRUE),
    sd   = sd(r, na.rm = TRUE),
    n      = sum(!is.na(r)),
    se     = sd / sqrt(n),
    tval   = qt(0.975, df = n - 1),
    ci_low = mean - tval * se,
    ci_high = mean + tval * se,
    .groups = "drop"
  )

# --- 2) Human vs Human (entre timepoints) ---
cor_human_human <- human_mouse_log2fc_samples %>%
  filter(organism == "Human") %>%
  select(participant_id, human_symbol, timepoint, log2fc) %>%
  pivot_wider(
    names_from = participant_id,
    values_from = log2fc
  ) %>%
  select(-human_symbol, -timepoint) %>%   # só colunas de participantes
  cor(use = "pairwise.complete.obs", method = "pearson") %>% 
  as.data.frame() %>%
  rownames_to_column("participant_1") %>%
  melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>% 
  filter(participant_1 != participant_2) %>%   # tira a diagonal (=1)
  # garantir combinações únicas
  rowwise() %>%
  mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
  ungroup() %>%
  distinct(pair, .keep_all = TRUE) %>%
  #Summarize
  summarize(
    comparison = "Human vs. Human",
    mean = mean(correlation, na.rm = TRUE),
    sd   = sd(correlation, na.rm = TRUE),
    n    = sum(!is.na(correlation)),
    se   = sd / sqrt(n),
    tval = qt(0.975, df = n - 1),
    ci_low  = mean - tval * se,
    ci_high = mean + tval * se
  ) 
  

# --- 3) Mouse vs Mouse ---
cor_mouse_mouse <- human_mouse_log2fc_samples %>%
  filter(organism == "Mouse") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, human_symbol, log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = log2fc) %>%
      select(-human_symbol)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Mouse vs. Mouse",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup()


#Unite
correlation_long_all_mouse_human_timepoints_summary = cor_human_mouse_timepoint %>% 
  mutate(timepoint = timepoint_mouse) %>% 
  bind_rows(
    cor_human_human %>% mutate(timepoint = "24"),
    cor_mouse_mouse)

correlation_long_all_mouse_human_timepoints_summary


correlation_long_all_mouse_human_timepoints_summary_plot = correlation_long_all_mouse_human_timepoints_summary %>% 
    mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(x = timepoint,
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1,
                data = . %>% 
                  filter(comparison != "Human vs. Human")) +
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                angle = 45,
                hjust = 1,
                linewidth = 1,
                data = . %>% 
                  filter(comparison == "Human vs. Human")) +
  scale_x_discrete(expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01), limits = c(-0.2, 1)) +
  theme_vaxgo() +
  scale_color_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = paste0("Correlation between organisms\n", gene_set_name, ", ", filename),
       x = "",
       y = "Pearson p, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 

correlation_long_all_mouse_human_timepoints_summary_plot


correlation_long_all_mouse_human_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", paste(filename, gene_set_name, "correlation_timepoints_summary_plot.png", sep = "_")),
         width = 3,
         height = 3)

correlation_long_all_mouse_human_timepoints_summary_plot
```


# BOOTSTRAP (ATUALIZAR)
```{r message=FALSE, warning=FALSE}
# This script performs cross-species correlation (Human vs. Mouse)
# using the bootstrap method (1000 iterations), sampling 5 female participants
# from each organism in each iteration, as requested.

# --- 1. BOOTSTRAP CONFIGURATION ---
N_BOOTSTRAPS = 1000   # Number of bootstrap iterations
SAMPLE_SIZE = 5       # 5 humans vs 5 mice in each iteration

TIMEPOINTS = unique(human_mouse_log2fc_samples %>% filter(organism == "Mouse") %>% pull(timepoint)) %>% as.numeric()

# --- 2. DATA PREPARATION AND FILTERING (Female vs. Female) ---

#Get DEGs from either organisms


# Filters raw log2fc data only for the relevant timepoints
# and joins metadata to filter only 'Female' participants.
data_filtered <- human_mouse_log2fc_samples %>%
  filter(timepoint %in% TIMEPOINTS) 

human_data <- data_filtered %>% filter(organism == "Human")
mouse_data <- data_filtered %>% filter(organism == "Mouse")

# --- 3. FUNCTION FOR A SINGLE BOOTSTRAP ITERATION ---

# This function performs a single bootstrap iteration for a timepoint.
# 1. Samples 5 Human and 5 Mouse IDs (with replacement).
# 2. Filters data for sampled IDs and common symbols (genes).
# 3. Calculates the cross-correlation matrix (5x5).
# 4. Returns all 25 correlation values.
get_bootstrap_cors <- function(timepoint_val, all_human_data, all_mouse_data, n_sample) {

  # Filters data for the timepoint and identifies common symbols (genes)
  h_tp_data <- all_human_data %>% filter(timepoint == timepoint_val)
  m_tp_data <- all_mouse_data %>% filter(timepoint == timepoint_val)
  common_symbols <- intersect(h_tp_data$symbol, m_tp_data$symbol)

  # If there is no data or common symbols, returns an empty tibble
  if (length(unique(h_tp_data$participant_id)) < n_sample || 
      length(unique(m_tp_data$participant_id)) < n_sample ||
      length(common_symbols) == 0) {
    message(paste("Insufficient data for Timepoint:", timepoint_val))
    return(tibble(timepoint = timepoint_val, correlation = NA_real_))
  }

  # Samples N participant IDs (with replacement - bootstrap)
  sampled_human_pids <- sample(unique(h_tp_data$participant_id), n_sample, replace = TRUE)
  sampled_mouse_pids <- sample(unique(m_tp_data$participant_id), n_sample, replace = TRUE)

  # Prepares Human matrix (rows=symbols, columns=participants)
  h_matrix <- h_tp_data %>%
    filter(participant_id %in% sampled_human_pids, symbol %in% common_symbols) %>%
    # FIX: Ensure uniqueness of symbol/participant_id before pivoting to prevent 'duplicate row names' error
    distinct(symbol, participant_id, log2fc) %>%
    pivot_wider(names_from = participant_id, values_from = log2fc) %>%
    column_to_rownames("symbol")

  # Prepares Mouse matrix (rows=symbols, columns=participants)
  m_matrix <- m_tp_data %>%
    filter(participant_id %in% sampled_mouse_pids, symbol %in% common_symbols) %>%
    # FIX: Ensure uniqueness of symbol/participant_id before pivoting to prevent 'duplicate row names' error
    distinct(symbol, participant_id, log2fc) %>%
    pivot_wider(names_from = participant_id, values_from = log2fc) %>%
    column_to_rownames("symbol")

  # Ensures columns have unique names for correlation (necessary due to replace=TRUE)
  colnames(h_matrix) <- paste0("Human_", 1:ncol(h_matrix))
  colnames(m_matrix) <- paste0("Mouse_", 1:ncol(m_matrix))

  # Calculates the cross-correlation matrix (Spearman)
  cross_cors <- cor(h_matrix, m_matrix, use = "pairwise.complete.obs", method = "pearson")

  # Extracts the N*N correlation values and adds the timepoint
  tibble(
    timepoint = timepoint_val,
    correlation = as.vector(cross_cors)
  )
}

# --- 4. BOOTSTRAP EXECUTION ---

# Creates a dataframe of bootstrap IDs and applies the function for all timepoints
bootstrapped_human_mouse_cors <- tibble(
    bootstrap_id = 1:N_BOOTSTRAPS
) %>%
rowwise() %>%
mutate(
    # Executes the bootstrap function for each timepoint 
    cors_data = list(
        bind_rows(
            get_bootstrap_cors(2, human_data, mouse_data, SAMPLE_SIZE),
            get_bootstrap_cors(6, human_data, mouse_data, SAMPLE_SIZE),
            get_bootstrap_cors(12, human_data, mouse_data, SAMPLE_SIZE),
            get_bootstrap_cors(24, human_data, mouse_data, SAMPLE_SIZE)
        ))) %>%
ungroup() %>%
unnest(cors_data)

# --- 5. BOOTSTRAP RESULTS SUMMARIZATION ---

# Calculates descriptive statistics on the distribution of correlations
# obtained by the 1000 bootstrap iterations (N_total = 1000 * 25 = 25000 observations)
correlation_bootstrap_summary <- bootstrapped_human_mouse_cors %>%
 group_by(timepoint) %>%
  summarize(
    comparison = "Bootstrap Mouse vs. Human",
    mean = mean(correlation, na.rm = T) %>% round(4),
    median = median(correlation, na.rm = T) %>% round(4),
    min = min(correlation, na.rm = T) %>% round(4),
    max = max(correlation, na.rm = T) %>% round(4),
    sd = sd(correlation, na.rm = T) %>% round(4),
    se = sd,
    ci_lower = quantile(correlation, 0.025, na.rm = TRUE) %>% round(4),
    ci_upper = quantile(correlation, 0.975, na.rm = TRUE) %>% round(4),
    .groups = "drop") %>% 
  mutate(timepoint = as.character(timepoint),
         unit = "Day")
```








#### BTMs


```{r}
#Annotate genes by BTM
dge_btm_mean_process = GSE33341_genes_btm %>%
  mutate(organism = "Human") %>% 
  bind_rows(GSE19668_genes_btm %>% 
              mutate(organism = "Mouse")) %>%
  filter(timepoint %in% c(1, 2)) %>% 
  mutate(condition = paste0(organism, "_Day", timepoint)) %>% 
  group_by(condition, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  select(-timepoint) %>% 
  pivot_wider(names_from = "condition", values_from = "mean_log2fc") 

dge_btm_mean_process
```
##### Scatterplot
```{r fig.height=5, fig.width=7}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(4:5) %>% 
  colnames()


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  select(group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      select(group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = participant_2_value, y = participant_1_value)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  stat_smooth(aes(x = participant_2_value, y = participant_1_value),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "Mouse Log2FC", 
       y = "Human Log2FC", 
       color = "",
       title = "Mouse vs. Human")



```



##### Heatmap

```{r}
#Matrix data
matrix_data = dge_btm_mean_process %>% 
  select(process, Human_Day1, Mouse_Day2) %>% 
  drop_na() %>% 
  column_to_rownames("process") %>% 
  as.matrix()

# Column and row annotation
######## Annotations
annotation_heatmap = tribble(~condition, ~timepoint, ~organism,
        "Human_Day1", 1, "Human",
        "Mouse_Day2", 2, "Mouse")

#Arrange columns in the same order as the matrix
ann_cols_heatmap = matrix_data %>% 
  colnames() %>% 
  data.frame() %>% 
  rename(condition = '.') %>% 
  inner_join(annotation_heatmap, by = "condition") %>% 
  distinct() %>% 
  arrange(condition)

#Arrange matrix columns 
mat = matrix_data %>% 
  as.data.frame() %>% 
  select(c(ann_cols_heatmap$condition %>% unique())) %>% 
  as.matrix() 

#Column to rownames
ann_cols_heatmap = ann_cols_heatmap %>% 
  column_to_rownames("condition")

#Annotate rows
ann_rows_heatmap = mat %>% 
  rownames() %>% 
  as.data.frame() %>% 
  rename("process" = ".") %>% 
  inner_join(dge_btm_mean_process %>% 
               select(process, group, subgroup), 
             by = "process") %>% 
  column_to_rownames("process")
  

#Check dimensions
dim(mat)
dim(ann_cols_heatmap)

#Colors from required.R
btm_immune_groups
colors_immune

```


```{r fig.width= 6, fig.height=20}
#Heatmap annotation
ha = HeatmapAnnotation(df = ann_cols_heatmap, 
                       col = colors_immune,
                       gp = gpar(col = "black"), 
                       simple_anno_size = unit(0.3, "cm"),
                       annotation_name_side = "right")

row_ha = HeatmapAnnotation(df = ann_rows_heatmap,
                       col = btm_immune_groups,
                       which= "row",
                       simple_anno_size = unit(0.2, "cm"),
                       show_annotation_name = T,
                       show_legend = T)

# Values colors
col_fun <- colorRamp2(c(-2,0,2), c("#45efc6", "white", "#165289"))

file = "BTM_Modules_Saureus_Mouse_Human"

width_image = 25
height_image = 30
width = unit(1, "cm")
height = unit(25, "cm")
column_names_gp = gpar(fontsize = 8)
row_names_gp = gpar(fontsize = 8)
rect_gp = gpar(col = "black", lwd = 0.5)

fileimageheatmap_grouped = here("Figures", paste0(file, sep ="_", "Complexheatmap.png"))

png(file = fileimageheatmap_grouped, 
    width = width_image, 
    height = height_image, 
    units="cm", 
    res=900)

heatmap_plot = Heatmap(mat,
                        top_annotation = ha, 
                       left_annotation = row_ha,
                        show_row_names = TRUE,
                        row_names_side = "right",
                        show_heatmap_legend = TRUE,
                        heatmap_legend_param = list(direction = "horizontal"),
                        col = col_fun,
                        show_column_names = TRUE,
                        column_title = file,
                        column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                        column_names_rot = 45,
                        column_names_gp = column_names_gp,
                        rect_gp = rect_gp,
                        na_col = "black",
                        name = "Mean Log2FC",
                        cluster_rows = TRUE,
                        row_names_gp = row_names_gp, row_km = 2,
                        row_dend_width = unit(3, "cm"),
                        row_split = NULL,
                        column_split = NULL,
                        row_gap = unit(2, "mm"),
                        cluster_columns = TRUE,
                        column_gap = unit(8, "mm"),
                        show_column_dend = F,
                        show_row_dend = T, 
                        row_dend_side = "left",
                        column_dend_side = "top",
                        clustering_distance_rows = "euclidean",
                        border = TRUE,
                        width = width,
                        height = height)

heatmap_plot = draw(heatmap_plot, 
     annotation_legend_side = "left", 
     heatmap_legend_side = "left")

heatmap_plot

dev.off()


```














#5. Functional analysis

## Immune
BTMs correlation, intra- and inter-species

### Mean Log2FC and GSEA
Compare means and NES
Import
```{r}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Saureus")
#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "saureus_infection_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Gene set enrichment analysis (GSEA) --------
#Get calculate means -----
#Annotate genes by BTM 

dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(immune_set == "BTM by Modules") %>%
  filter(organism %in% c("Human", "Mouse")) 

dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarize(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            n_genes = n(),
            organism, pathogen,
            group, subgroup) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

#Print
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
```

```{r message=FALSE, warning=FALSE}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) %>% 
  select(condition, genes, log2fold_change)


#Run multiple GSEA and unite means
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "ImmuneGO_Human")


gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  inner_join(dge_btm_mean_process, by = join_by("condition", "process"))
  
  
#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```

#### Plot distributions

```{r fig.height=12, fig.width=10}
#Import ========================
gsea_btm_results = readRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
dge_btm_mean_process = readRDS(file = here("tables", paste0(filename, "_dge_btm_mean_process.rds")))
dge_btm_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))


# #If timepoints match between ref and target, use this
# dge_btm_process_genes_matched = dge_btm_process_genes

#Duplicate reference, if timepoints do not match =================

timepoint_reference = "1"
human_data_reference <- dge_btm_process_genes %>% filter(organism == "Human")
new_timepoints <- c(dge_btm_process_genes %>% filter(organism != "Human") %>% pull(timepoint) %>% unique() %>% as.character())
duplicated_human_data <- map_dfr(new_timepoints, ~ {
  df_copy <- human_data_reference
  df_copy <- df_copy %>%
    mutate(timepoint = .x)

  return(df_copy)
})

dge_btm_process_genes_matched = dge_btm_process_genes %>% 
  filter(organism != "Human") %>% 
  bind_rows(duplicated_human_data) 

# ================================


#Pivot to wide GSEA + mean
dge_btm_gsea_mean_wide = gsea_btm_results %>% 
  mutate(organism = case_when(str_detect(condition, "Mouse") ~ "Mouse",
                              str_detect(condition, "Human") ~ "Human",
                              T~condition)) %>% 
  select(organism, timepoint, process, group, subgroup, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process, subgroup, group, set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

#Arrange process by log2fc
process_order_log2fc = dge_btm_process_genes_matched %>% 
  select(process, timepoint, organism, mean_l2fc, group) %>%
  filter(timepoint == 12,
         organism == "Human") %>% 
  group_by(organism, timepoint, group) %>% 
  arrange(-mean_l2fc) %>%
  pull(process) %>% 
  unique() 

dge_btm_process_ordered = dge_btm_process_genes_matched %>% 
  ungroup() %>% 
    mutate(group = fct_relevel(as.factor(group), c(immune_order)),
           process = fct_relevel(as.factor(process), c(process_order_log2fc)) %>% fct_rev()) 

labels_process <- dge_btm_gsea_mean_wide %>%
  ungroup() %>% 
  distinct(process, p_adjust_Human, p_adjust_Mouse) %>%
  mutate(label = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
                         "bold", "plain"),
         label_color = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
                         "#4DBBD5FF", "gray50"))  %>% 
  ungroup() %>% 
  distinct() %>% 
  mutate(process = fct_relevel(process, process_order_log2fc)) %>% 
  filter(!is.na(p_adjust_Human))
```


```{r fig.height=12, fig.width=10}
#Plot
dge_btm_process_distribution = dge_btm_process_genes_matched %>% 
    mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(y = process,
      x = mean_l2fc,
      color = organism) +
  facet_grid(~group~timepoint, scales = "free", space = "free") +
  geom_vline(xintercept = 0, color = "black") +
  geom_point(pch = "|", alpha = 1, 
             show.legend = F,
             size = 3) +
  stat_summary(
  aes(color = organism),
  fun.data = mean_se,
  geom = "errorbar",
  width = 0.5,
  size = 1) +
  
  stat_summary(
    aes(fill = organism), 
    fun = mean,             
    color = "black",
    geom = "point",             
    size = 2,    
    pch = 21,  
    alpha = 1,
    show.legend = T 
  ) +
  
  xlim(-2, 2) +
  theme_vaxgo() +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), 
        axis.ticks.y = element_line(color = "gray", alpha = 0.5),
        legend.position = "top",
        axis.text.x = element_text(size = 8),
        # axis.text.y = element_text(size = 8, face = c(labels_process$label), 
        #                            color = c(labels_process$label_color)),
        axis.line.y = element_blank(),
        panel.spacing.y = unit(0.3, "lines"),
        panel.grid.major.y = element_line(alpha = 0.5),
        panel.grid.major.x = element_line(alpha = 0.5)) +
  scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
  scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
    labs(color = "Organism",
       y = "",
       x = "Log2FC",
       fill = "Organism",
       title = "Mouse vs Human, BTMs genes, mean Log2FC") 

dge_btm_process_distribution 

dge_btm_process_distribution %>% 
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_process_distribution.png")),
         height = 12, 
         width = 10)


dge_btm_process_ordered %>% filter(organism == "Human") %>% distinct()

```


#### Correlation plot (Gene-level, BTMs, immune)

##### Plot genes by btm group
```{r fig.height=16, fig.width=8}
#Convert to wide
dge_btm_process_genes_wide = dge_btm_process_genes_matched  %>% 
    ungroup() %>%
  select(symbol, mean_l2fc, adj_p_val, organism, timepoint, process, group) %>% 
  distinct() %>% 
  group_by(organism, timepoint, process) %>% 
  pivot_wider(names_from = organism,
              values_from = c(mean_l2fc, adj_p_val)) %>% 
  ungroup() %>% 
  mutate(group = fct_relevel(group, c(immune_order)))  %>% 
  #Remove duplicates
  select(-process) %>% 
  distinct()

#Plot
dge_btm_process_genes_correlation_plot = dge_btm_process_genes_wide %>% 
      mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(x = mean_l2fc_Mouse,
      y = mean_l2fc_Human,
      labels = symbol) +
    #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  facet_grid(~group~timepoint, scales = "free") +
  #Plot all points
  geom_point(aes(fill = group,
                 color = group),
             size = 2,
             alpha = 0.2,
             pch = 21) +
  #Plot only significant in both
  geom_point(aes(fill = group),
             data = . %>% filter(adj_p_val_Human <= 0.05 & adj_p_val_Mouse <= 0.05), 
             color = "black",
             size = 2,
             alpha = 1,
             pch = 21) +

  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  guides(color = F) +
  labs(fill = "Immune group",
       y = "Human (Mean Log2FC)",
       x = "Model (Mean Log2FC)",
       title = "Genes by BTM",
       subtitle = "Genes with adjusted p-value <= 0.05 in both organisms.") +
  ylim(-2, 2) +
  xlim(-2, 2) +
  
  geom_text_repel(aes(label = symbol), 
                  min.segment.length = 0,
                  data = . %>% filter(adj_p_val_Human <= 0.05 & adj_p_val_Mouse <=0.05) %>% 
                    distinct(), 
                  size = 2, 
                  segment.size = 0.2,  
                  lineheight = 0.8, 
                  max.overlaps = 10) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.text = element_text(size = 8, hjust = 0)
          )

#Save results
dge_btm_process_genes_correlation_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_process_genes_correlation_plot_facetted.png")),
         width = 8,
         height = 16)

#Print
dge_btm_process_genes_correlation_plot
```


#### Correlation plot (Module-level, BTMs, immune)
Import
```{r}
#Import results -------
gsea_results = readRDS(here("tables", "saureus_infection_human_mouse_gsea_btm_results.rds"))
dge_long_genes = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Saureus")

#Duplicate reference, if timepoints do not match =================

timepoint_reference = "1"
human_data_reference <- dge_btm_process_genes %>% filter(organism == "Human")
new_timepoints <- c(dge_btm_process_genes %>% filter(organism != "Human") %>% pull(timepoint) %>% unique() %>% as.character())
duplicated_human_data <- map_dfr(new_timepoints, ~ {
  df_copy <- human_data_reference
  df_copy <- df_copy %>%
    mutate(timepoint = .x)

  return(df_copy)
})

dge_btm_process_genes_matched = dge_btm_process_genes %>% 
  filter(organism != "Human") %>% 
  bind_rows(duplicated_human_data) 


#GSEA results
gsea_results
timepoint_reference = "1"
human_data_reference <- gsea_results %>% filter(organism == "Human")
new_timepoints <- c(gsea_results %>% filter(organism != "Human") %>% pull(timepoint) %>% unique() %>% as.character())
duplicated_human_data <- map_dfr(new_timepoints, ~ {
  df_copy <- human_data_reference
  df_copy <- df_copy %>%
    mutate(timepoint = .x)

  return(df_copy)
})

gsea_results_matched = gsea_results %>%
  filter(organism != "Human") %>%
  bind_rows(duplicated_human_data)


# ================================


gsea_results %>% 
  filter(timepoint == 12) %>% 
  select(organism, timepoint, process, group, subgroup, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process, subgroup, group, set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_") %>% 
  ungroup()

```


Plot correlation
```{r fig.height=5, fig.width=7}
#Mouse vs Human
#Pivot to wide GSEA + mean
dge_btm_gsea_mean_wide = gsea_results_matched %>% 
  select(organism, timepoint, process, group, subgroup, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process, subgroup, group, set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_") %>% 
  ungroup()


dge_btm_process_genes_wide = dge_btm_process_genes_matched  %>% 
    ungroup() %>%
  select(symbol, mean_l2fc, adj_p_val, organism, timepoint, process, group) %>% 
  distinct() %>% 
  group_by(organism, timepoint, process) %>% 
  pivot_wider(names_from = organism,
              values_from = c(mean_l2fc, adj_p_val)) %>% 
  ungroup() %>% 
  mutate(group = fct_relevel(group, c(immune_order))) %>% 
  drop_na(mean_l2fc_Human)


```


```{r fig.height=5, fig.width=7}
#Calculate correlation -----

# Means vs means
cor_labels <- dge_btm_process_genes_wide %>%
  select(process, timepoint, group, Mouse = mean_l2fc_Mouse, Human = mean_l2fc_Human) %>% 
  group_by(timepoint) %>% 
  summarise(test_result = list(cor.test(x = Human, 
                                        y = Mouse, 
                                        use = "pairwise.complete.obs", 
                                        method = "pearson")), 
            .groups = "drop") %>%
  mutate(cor_value = purrr::map_dbl(test_result, "estimate"),
         p_value = purrr::map_dbl(test_result, "p.value")) %>%
  select(-test_result) %>%
  # Labels
  mutate(r_squared = cor_value^2,
    p_label = ifelse(p_value < 0.01,
                     "p = <0.01",
                     paste0("p = ", round(p_value, 2))),
    cor_label = paste0(
      "r = ", round(cor_value, 2),
      ", R² = ", round(r_squared, 2),
      ", ", p_label)) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) 

# NES vs NES
cor_labels_nes <- dge_btm_gsea_mean_wide %>%
  ungroup() %>% 
  select(process, timepoint, group, subgroup, Mouse = nes_Mouse, Human = nes_Human) %>% 
  group_by(timepoint) %>% 
  summarise(test_result = list(cor.test(x = Human, 
                                        y = Mouse, 
                                        use = "pairwise.complete.obs", 
                                        method = "pearson")), 
            .groups = "drop") %>%
  mutate(cor_value = purrr::map_dbl(test_result, "estimate"),
         p_value = purrr::map_dbl(test_result, "p.value")) %>%
  select(-test_result) %>%
  # Labels
  mutate(r_squared = cor_value^2,
    p_label = ifelse(p_value < 0.01,
                     "p = <0.01",
                     paste0("p = ", round(p_value, 2))),
    cor_label = paste0(
      "r = ", round(cor_value, 2),
      ", R² = ", round(r_squared, 2),
      ", ", p_label)) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) 

```

Plot means
```{r fig.height=3, fig.width=9}
dge_btm_mean_group_correlation_df = dge_btm_gsea_mean_wide %>% 
  ungroup() %>% 
  mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  drop_na(mean_log2fc_Mouse)

#Plot correlation plot (means)
dge_btm_mean_group_correlation_plot =  dge_btm_mean_group_correlation_df %>% 
  ggplot() +
  aes(x = mean_log2fc_Mouse,
      y = mean_log2fc_Human,
      labels = process) +
    #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  facet_wrap(~timepoint, nrow = 1) +
  #Plot all points
  geom_point(aes(fill = group,
                 color = group),
             size = 2,
             alpha = 0.2,
             pch = 21) +
  #Plot only significant in both
  geom_point(aes(fill = group),
             data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <= 0.25), 
             color = "black",
             size = 2,
             alpha = 1,
             pch = 21) +

  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  guides(color = F) +
  labs(fill = "Immune group",
       y = "Human (Mean Log2FC)",
       x = "Model (Mean Log2FC)") +
  ylim(-2, 2) +
  xlim(-2, 2) +
  
  #Add correlation metrics
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label),
            fill= "white",
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  #Label only the top 10 processes with least difference in log2fc
  geom_text_repel(aes(label = process),
                  min.segment.length = 0,
                  data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <= 0.25) %>%
                    mutate(diff = mean_log2fc_Human - mean_log2fc_Mouse) %>%
                    slice_min(n = 20, abs(diff)), 
                  size = 1.5,
                  segment.size = 0.2,
                  lineheight = 0.8,
                  max.overlaps = 20) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "right",
          legend.direction = "vertical",
          legend.text = element_text(size = 8, hjust = 0)
          )

#Print
dge_btm_mean_group_correlation_plot

#Save results
dge_btm_mean_group_correlation_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_mean_group_correlation_plot.png")),
         width = 9,
         height = 3)
```


Plot NES
```{r fig.height=3, fig.width=9}
#Plot correlation plot (NES)
dge_btm_nes_group_correlation_plot = dge_btm_gsea_mean_wide %>% 
  mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(x = nes_Mouse,
      y = nes_Human,
      labels = process) +
    #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  facet_wrap(~timepoint, nrow = 1
             # scales = "free"
             ) +
  #Plot all points
  geom_point(aes(fill = group,
                 color = group),
             size = 2,
             alpha = 0.2,
             pch = 21) +
  #Plot only significant in both
  geom_point(aes(fill = group),
             data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <= 0.25), 
             color = "black",
             size = 2,
             alpha = 1,
             pch = 21) +

  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  guides(color = F) +
  labs(fill = "Immune group",
       y = "Human (NES)",
       x = "Model (NES)") +
  ylim(-4, 4) +
  xlim(-4, 4) +
  
  #Add correlation metrics
  geom_label(data = cor_labels_nes,
            aes(x = 0, y = Inf, label = cor_label), 
            fill= "white",
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  
  geom_text_repel(aes(label = process), 
                  min.segment.length = 0,
                  data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <= 0.25) %>%
                    mutate(diff = nes_Human - nes_Mouse) %>%
                    slice_min(n = 20, abs(diff)), 
                  size = 1.5, 
                  segment.size = 0.2,  
                  lineheight = 0.8, 
                  max.overlaps = 200000) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "right",
          legend.direction = "vertical",
          legend.text = element_text(size = 8, hjust = 0)
          )

#Print
dge_btm_nes_group_correlation_plot

#Save results
dge_btm_nes_group_correlation_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_nes_group_correlation_plot.png")),
         width = 9,
         height = 3)
```


##### Facet by group

Mean log2fc
```{r fig.height=10, fig.width=8}
dge_btm_mean_group_correlation_plot_facetted = dge_btm_gsea_mean_wide %>% 
  mutate(timepoint = as.character(timepoint)) %>% 
    mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(x = mean_log2fc_Mouse,
      y = mean_log2fc_Human,
      color = group) +
  facet_grid(~group~timepoint, scales = "free") +
  #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
    geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
    geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
    geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  
    #Plot all
    geom_point(aes(fill = group),
               color = "black",
               size = 2,
               alpha = 0.2,
               pch = 21) +
    #Plot only significant in both
    geom_point(aes(fill = group),
               data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <= 0.25), 
               color = "black",
               size = 2,
               alpha = 1,
               pch = 21) +
    geom_text_repel(aes(label = process),
                    color = "black",
                  min.segment.length = 0,
                  data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <= 0.25) %>%
                    mutate(diff = nes_Human - nes_Mouse) %>%
                    group_by(timepoint, group) %>% 
                    slice_min(n = 5, abs(diff)) %>% 
                    ungroup(), 
                  size = 2, 
                  segment.size = 0.2,  
                  lineheight = 0.8, 
                  max.overlaps = 20) +
    scale_fill_manual(values = immune_colors) +
    scale_color_manual(values = immune_colors) +
    guides(color = F) +
    labs(fill = "Immune group",
         y = "Human (Log2FC)",
         x = "Model (Log2FC)") +
    #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.text = element_text(size = 8, hjust = 0))

dge_btm_mean_group_correlation_plot_facetted

#Save results
dge_btm_mean_group_correlation_plot_facetted %>%
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_mean_group_correlation_plot_facetted.png")),
         width = 8,
         height = 11)
  
```


NES
```{r fig.height=10, fig.width=8}
dge_btm_nes_group_correlation_plot_facetted = dge_btm_gsea_mean_wide %>% 
  mutate(timepoint = as.character(timepoint)) %>% 
    mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(x = nes_Mouse,
      y = nes_Human,
      color = group) +
  facet_grid(~group~timepoint, scales = "free") +
  #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
    geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
    geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
    geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  
    #Plot all
    geom_point(aes(fill = group),
               color = "black",
               size = 2,
               alpha = 0.2,
               pch = 21) +
    #Plot only significant in both
    geom_point(aes(fill = group),
               data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <= 0.25), 
               color = "black",
               size = 2,
               alpha = 1,
               pch = 21) +
    geom_text_repel(aes(label = process),
                    color = "black",
                  min.segment.length = 0,
                  data = . %>% filter(p_adjust_Human <= 0.25 & p_adjust_Mouse <= 0.25) %>%
                    mutate(diff = nes_Human - nes_Mouse) %>%
                    group_by(timepoint, group) %>% 
                    slice_min(n = 5, abs(diff)) %>% 
                    ungroup(), 
                  size = 2, 
                  segment.size = 0.2,  
                  lineheight = 0.8, 
                  max.overlaps = 200000) +
    scale_fill_manual(values = immune_colors) +
    scale_color_manual(values = immune_colors) +
    guides(color = F) +
    labs(fill = "Immune group",
         y = "Human (NES)",
         x = "Model (NES)") +
    #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.text = element_text(size = 8, hjust = 0))

dge_btm_mean_group_correlation_plot_facetted

#Save results
dge_btm_nes_group_correlation_plot_facetted %>%
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_nes_group_correlation_plot_facetted.png")),
         width = 8,
         height = 11)
  
```






## Sample-wise analysis
Annotate genes log2fc table with BTMs
```{r}
#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

#Example input
samples_log2fc = readRDS(file = here("tables", "saureus_infection_human_mouse_samples_log2fc.rds"))

#Annotate genes by BTM 
results_genes_btm_df = samples_log2fc %>% 
  ungroup() %>% 
  select(symbol = human_symbol, participant_id = subject_id, geo_sample, timepoint, log2fc, organism) %>% 
  inner_join(btm_annotation_genes, by = "symbol")

#Pivot wider
dge_btm_mean_process = results_genes_btm_df %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

#Print
dge_btm_mean_process

#Save
dge_btm_mean_process %>% saveRDS(file = here("Tables", paste0(filename, "_dge_btm_samples_log2fc.rds")))

```

Import data 

```{r}
#Genes L2FC table annotated with BTMs
dge_btm_mean_process_samples = readRDS(file = here("tables", "saureus_infection_human_mouse_dge_btm_samples_log2fc.rds")) 

#Import metadata
mouse_human_metadata = readRDS(file = here("tables", "saureus_infection_human_mouse_metadata.rds"))


#Duplicate reference, if timepoints do not match =================
human_data_reference <- results_genes_btm_df %>% filter(organism == "Human")
new_timepoints <- c(results_genes_btm_df %>% filter(organism != "Human") %>% pull(timepoint) %>% unique() %>% as.character())
duplicated_human_data <- map_dfr(new_timepoints, ~ {
  df_copy <- human_data_reference
  df_copy <- df_copy %>%
    mutate(timepoint = .x)

  return(df_copy)
})

results_genes_btm_df_matched = results_genes_btm_df %>% 
  mutate(timepoint = as.character(timepoint)) %>% 
  filter(organism != "Human") %>% 
  bind_rows(duplicated_human_data) 


# ======================

#Calculate the mean log2fc by module
dge_btm_mean_process_sample_long = results_genes_btm_df_matched %>% 
  # mutate(timepoint = if_else(organism == "Human", 12, timepoint)) %>%
  group_by(organism, participant_id, timepoint, process) %>% 
  summarize(group,
            mean_log2fc = mean(log2fc, na.rm = T)) %>% 
  distinct() %>% 
  ungroup()

#Pivot wider
dge_btm_mean_process_sample_wide = dge_btm_mean_process_sample_long %>% 
  select(-organism) %>% 
  group_by(timepoint, process, group) %>% 
  pivot_wider(names_from = "participant_id", 
              values_from = "mean_log2fc") 
  
```

### Summarize inter-individual correlation

#### 1 individual vs 1 individual

```{r}
# --- 1) Mouse vs Human no mesmo timepoint ---
#Get timepoints
timepoints_human <- unique(dge_btm_mean_process_sample_long %>% filter(organism == "Human") %>% pull(timepoint))
timepoints_mouse <- unique(dge_btm_mean_process_sample_long %>% filter(organism == "Mouse") %>% pull(timepoint))

#Get combinations
timepoint_combinations <- expand.grid(timepoint_human = timepoints_human,
                                      timepoint_mouse = timepoints_mouse,
                                      stringsAsFactors = FALSE)

# Loop
cor_human_mouse_timepoint <- timepoint_combinations %>%
  pmap_dfr(function(timepoint_human, timepoint_mouse) {
    
    # pegar os participantes do timepoint específico
    human_df <- dge_btm_mean_process_sample_long %>%
      filter(organism == "Human", timepoint == timepoint_human) %>%
      select(participant_id, process, mean_log2fc)
    
    mouse_df <- dge_btm_mean_process_sample_long %>%
      filter(organism == "Mouse", timepoint == timepoint_mouse) %>%
      select(participant_id, process, mean_log2fc)
    
    # manter apenas genes em comum
    common_process <- intersect(human_df$process, mouse_df$process)
    human_df <- human_df %>% filter(process %in% common_process)
    mouse_df <- mouse_df %>% filter(process %in% common_process)
    
    # pivot para wide
    human_wide <- human_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    mouse_wide <- mouse_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    
    # IDs de participantes
    human_ids <- setdiff(colnames(human_wide), "process")
    mouse_ids <- setdiff(colnames(mouse_wide), "process")
    
    # calcular correlação para todos pares humano x mouse
    expand.grid(human = human_ids, mouse = mouse_ids, stringsAsFactors = FALSE) %>%
      mutate(
        cor_test = map2(human, mouse, ~{
          x <- human_wide[[.x]]
          y <- mouse_wide[[.y]]
          # remover NAs correspondentes
          keep <- complete.cases(x, y)
          cor.test(x[keep], y[keep], method = "pearson")
        }),
        r       = map_dbl(cor_test, ~.x$estimate),
        p_value = map_dbl(cor_test, ~.x$p.value),
        ci_low  = map_dbl(cor_test, ~.x$conf.int[1]),
        ci_high = map_dbl(cor_test, ~.x$conf.int[2])
      ) %>%
      select(-cor_test) %>%
      mutate(timepoint_human = timepoint_human %>% as.numeric(),
             timepoint_mouse = timepoint_mouse %>% as.numeric())
  }) %>% 
  group_by(timepoint_human, timepoint_mouse) %>%
  summarize(
    comparison = "Human vs. Mouse",
    mean = mean(r, na.rm = TRUE),
    sd   = sd(r, na.rm = TRUE),
    n      = sum(!is.na(r)),
    se     = sd / sqrt(n),
    tval   = qt(0.975, df = n - 1),
    ci_low = mean - tval * se,
    ci_high = mean + tval * se,
    .groups = "drop"
  )

cor_human_mouse_timepoint

# --- 2) Human vs Human (entre timepoints) ---
cor_human_human <- dge_btm_mean_process_sample_long %>%
  mutate(timepoint = as.numeric(timepoint)) %>% 
  filter(organism == "Human") %>%
  select(participant_id, process, timepoint, mean_log2fc) %>%
  group_by(timepoint) %>% 
  pivot_wider(
    names_from = participant_id,
    values_from = mean_log2fc
  ) %>%
  select(-process, -timepoint) %>%   # só colunas de participantes
  cor(use = "pairwise.complete.obs", method = "pearson") %>% 
  as.data.frame() %>%
  rownames_to_column("participant_1") %>%
  melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>% 
  filter(participant_1 != participant_2) %>%   # tira a diagonal (=1)
  # garantir combinações únicas
  rowwise() %>%
  mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
  ungroup() %>%
  distinct(pair, .keep_all = TRUE) %>%
  #Summarize
  summarize(
    comparison = "Human vs. Human",
    mean = mean(correlation, na.rm = TRUE),
    sd   = sd(correlation, na.rm = TRUE),
    n    = sum(!is.na(correlation)),
    se   = sd / sqrt(n),
    tval = qt(0.975, df = n - 1),
    ci_low  = mean - tval * se,
    ci_high = mean + tval * se
  ) %>% 
  mutate(timepoint = 24,
         timepoint_human = timepoint)
  

# --- 3) Mouse vs Mouse ---
cor_mouse_mouse <- dge_btm_mean_process_sample_long %>%
  filter(organism == "Mouse") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, process, mean_log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = mean_log2fc) %>%
      select(-process)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Mouse vs. Mouse",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup() %>% 
  mutate(timepoint = as.numeric(timepoint), 
         timepoint_mouse = timepoint)


#Unite
correlation_long_all_btm_mouse_human_timepoints_summary = cor_human_mouse_timepoint %>% 
  mutate(timepoint = timepoint_mouse) %>% 
  bind_rows(
    cor_human_human,
    cor_mouse_mouse)

#Print
correlation_long_all_btm_mouse_human_timepoints_summary

correlation_long_all_btm_mouse_human_timepoints_summary %>% 
  saveRDS(file = here("tables", paste0(filename, "_samples_btm_correlation_all_timepoints_summary.rds")))
```


##### Plot scatterplot

```{r fig.height=3, fig.width=3}
#Import

#Scatterplot
correlation_long_all_btm_mouse_human_timepoints_summary_plot = correlation_long_all_btm_mouse_human_timepoints_summary %>% 
  mutate(timepoint_num = timepoint,
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(x = timepoint,
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1,
                data = . %>% 
                  filter(comparison != "Human vs. Human")) +
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.5,
                angle = 90,
                linewidth = 1,
                data = . %>% 
                  filter(comparison == "Human vs. Human")) +
  
  
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  # scale_x_continuous(breaks = c(0, 2, 6, 12, 24), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     limits = c(0.4, 1)) +
  theme_vaxgo() +
  scale_color_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = paste0("Correlation between organisms,\nBTMs,", filename),
       x = "Hour",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 

correlation_long_all_btm_mouse_human_timepoints_summary_plot

correlation_long_all_btm_mouse_human_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", paste0(filename, "_samples_btm_correlation_all_timepoints_summary_plot.png")),
         width = 3,
         height = 3)

  
```







# Compare with FIT prediction

```{r}
#Get log2fc mean
GSE19668_input2FIT = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
    filter(timepoint == 2) %>% 
  select(participant_id, MM.Entrez = entrezgene_id, EffectSize = log2fc) %>% 
  group_by(MM.Entrez) %>% 
  summarize(mean_log2fc = mean(EffectSize))

#Save
GSE19668_input2FIT %>% 
  write_csv(file = here("FIT_Mouse2man", "saureus_infection_mouse_fit_prediction_dge_degs_day2.csv"))

#Run FIT
GSE19668_input2FIT_day2 = FIT.mouse2man::FIT(MouseFile = here("FIT_Mouse2man", "saureus_infection_mouse_fit_prediction_dge_degs_day2.csv"), DataType = "rnaseq")
```



### Unite log2FC values from mouse, human, and FIT
```{r fig.height=5, fig.width=8}
#Bind rows
mouse_human_fit_log2fc = bind_rows(GSE19668_input2FIT_day2 %>%
                                     #Get DEGs for FIT
                                     filter((CI_low != 0 | CI_high != 0) & !is.na(CI_low) & !is.na(CI_high)) %>%  
                                   select(human_symbol = Human.symbol,
                                          log2fc = FIT_prediction) %>%
                                     mutate(organism = "FIT"),
                                   
                                   #Get DEGs for Human
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_human <= 0.05) %>% 
                                     select(human_symbol, log2fc = Human) %>%
                                     mutate(organism = "Human"),
                                   
                                   #Get DEGs for Mouse
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_mouse <= 0.05) %>% 
                                     select(human_symbol, log2fc = Mouse) %>%
                                     mutate(organism = "Mouse"))


#Pivot wider
mouse_human_fit_log2fc_wide = GSE19668_input2FIT_day2 %>%
            select(human_symbol = Human.symbol,
                   FIT_prediction) %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Human),
            by = "human_symbol") %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Mouse),
            by = "human_symbol")

#Save
mouse_human_fit_log2fc %>% 
  saveRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))

mouse_human_fit_log2fc_wide %>% 
  saveRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))
```

## Distribution

### Import tables
```{r fig.height=5, fig.width=8}
#Import
mouse_human_fit_log2fc = readRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))
mouse_human_fit_log2fc_wide = readRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))

GSE33341_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))
GSE19668_metadata = readRDS(file = here("tables", "saureus_infection_GSE19668_metadata.rds"))
```

### Plot histogram

```{r fig.height=5, fig.width=8}
#Plot histogram
mouse_human_fit_log2fc_histogram = mouse_human_fit_log2fc %>% 
  ggplot() +
  aes(x = log2fc,
      fill = organism) +
  geom_density(aes(y = after_stat(count),
                   color = organism),
                  position = position_identity(), 
               alpha = 0.1,
               size = 1,
              show.legend = T) +
  scale_fill_manual(values = c(FIT = "#4361ee",
                               Mouse = "#4DBBD5FF",
                               Human = "#546E7AFF"))+
  scale_color_manual(values = c(FIT = "#4361ee",
                                Mouse = "#4DBBD5FF",
                                Human = "#546E7AFF"))+
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        legend.key = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0),
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(limits = c(-4, 4)) +
  labs(color = "",
       fill = "")

mouse_human_fit_log2fc_histogram

#Faceted
mouse_human_fit_log2fc_histogram_faceted = mouse_human_fit_log2fc_histogram +
  facet_wrap(~organism,
             scales = "free_y")


#Plot scatter plot
mouse_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = Mouse,
      y = Human) +
#All DEGs
geom_point(color = "#4DBBD5FF",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "Mouse L2FC",
         y = "Human L2FC")

mouse_human_scatterplot


fit_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = FIT_prediction,
      y = Human) +
#All DEGs
geom_point(color = "#4361ee",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "FIT L2FC",
         y = "Human L2FC")

#Histograms
(((mouse_human_fit_log2fc_histogram + mouse_human_fit_log2fc_histogram_faceted) + 
    plot_layout(widths = c(0.5, 1), guides = "collect") +
    plot_annotation(theme = theme(legend.position = "top")))  /
    #Scatter plots
    (mouse_human_scatterplot + fit_human_scatterplot)) +
  plot_layout(heights = c(0.5, 1))

```



## BTMs
Import
```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)
```

```{r}
#Annotate genes by BTM
dge_btm_mean_process = mouse_human_fit_log2fc %>%
  filter(organism != "Mouse") %>% 
  rename(symbol = human_symbol) %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  group_by(organism, process) %>% 
  summarize(mean_log2fc = mean(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism", values_from = "mean_log2fc") 

dge_btm_mean_process
```
##### Scatterplot
```{r fig.height=5, fig.width=7}

#Calculate correlation
cor_labels <- dge_btm_mean_process %>%
  summarise(cor_value = cor(Human, FIT, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
dge_btm_mean_process %>% 
ggplot(aes(x = FIT, y = Human)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  stat_smooth(aes(x = FIT, y = Human),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "FIT Log2FC", 
       y = "Human Log2FC", 
       color = "",
       title = "Mouse vs. Human")



```









# Performance tests
## ROC Curve: Genes
Comparison must be done between human at day 1 and multiple mice timepoints 
```{r}
#Import
all_degs_human_mouse = readRDS(file = here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds"))
human_mouse_metadata = readRDS(file = here("tables", "saureus_infection_human_mouse_metadata.rds"))
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))


#Prepare
all_degs_human_mouse = all_degs_human_mouse %>% 
  inner_join(human_mouse_metadata %>%
               group_by(organism, timepoint) %>%
               summarize(n = n()) %>% 
               mutate(timepoint = as.character(timepoint)),
               by = join_by("organism", "timepoint")) %>% 
  mutate(timepoint = if_else(organism == "Human", "24", timepoint))


timepoint_reference = "24"
human_data_reference <- all_degs_human_mouse %>% filter(organism == "Human")
new_timepoints <- c(all_degs_human_mouse %>% filter(organism != "Human") %>% pull(timepoint) %>% unique() %>% as.character())
duplicated_human_data <- map_dfr(new_timepoints, ~ {
  df_copy <- human_data_reference
  df_copy <- df_copy %>%
    mutate(timepoint = .x)

  return(df_copy)
})

dge_long_genes_matched = all_degs_human_mouse %>% 
  filter(organism != "Human") %>% 
  bind_rows(duplicated_human_data) 

#Input
dge_long_genes = dge_long_genes_matched

#All genes
genes_roc_allgenes = dge_long_genes %>% 
  rename(symbol = human_symbol)
#Filter immune genes
genes_roc_allgenes_immunegenes = dge_long_genes %>% 
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))

#Only DEGs
degs_genes_human = dge_long_genes %>%
  filter(adj_p_val <= 0.05,
         organism == "Human") %>%
  select(symbol, timepoint)

#Only immune DEGs
genes_roc_DEGs_all = dge_long_genes %>%
  inner_join(degs_genes_human,
             by = join_by("symbol", "timepoint"))

#Only immune DEGs
genes_roc_DEGs_immune = dge_long_genes %>%
  inner_join(degs_genes_human,
             by = join_by("symbol", "timepoint")) %>%
  #Filter immune genes only
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))
```



Inputs
```{r fig.height=4, fig.width=10}
set.seed(123)
######### Input -----
# filename = "saureus_infection_human_mouse"
# gene_set_filename = "All_OnlyImmune"
# genes_df_input = bind_rows(
#   genes_roc_allgenes %>% mutate(model_type = "Mouse"),
#   genes_roc_allgenes_immunegenes %>% mutate(model_type = "Immune")
# )


filename = "saureus_infection_human_mouse"
gene_set_filename = "HumanDEGs_OnlyImmune"
genes_df_input = bind_rows(
  genes_roc_DEGs_all %>% mutate(model_type = "Mouse"),
  genes_roc_DEGs_immune %>% mutate(model_type = "Immune")
)

```




Calculate ROC curve
```{r fig.height=5, fig.width=10}
#=====================================================
# Compute probabilities
#=====================================================

dge_long_genes_up = genes_df_input %>% 
  mutate(
    df = n - 1,
    pval_one_sided = pt(t, df, lower.tail = FALSE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t > 0, "up", "not_up"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  select(-adj_p_val) %>%
  rename(adj_p_val = pval_one_sided_adj)

dge_long_genes_down = genes_df_input %>% 
  mutate(
    df = n - 1,
    pval_one_sided = pt(t, df, lower.tail = TRUE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t < 0, "down", "not_down"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  select(-adj_p_val) %>%
  rename(adj_p_val = pval_one_sided_adj)

#=====================================================
# Assign human truth
#=====================================================
truth_human_up = dge_long_genes_up %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

truth_human_down = dge_long_genes_down %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#=====================================================
#Generate reference distributions
#=====================================================
perfect_prediction_up = truth_human_up %>%
  mutate(prob = if_else(truth == "up", 1, 0),
         model = "Human") %>%
  rename(pred = truth)

uniform_prediction_up = perfect_prediction_up %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()), organism = "Uniform",
         model = "Uniform")

perfect_prediction_down = truth_human_down %>%
  mutate(prob = if_else(truth == "down", 1, 0),
          model = "Human") %>%
  rename(pred = truth)

uniform_prediction_down = perfect_prediction_down %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()), organism = "Uniform",
         model = "Uniform")

#=====================================================
# Combine models (Mouse + Immune + Uniform + Perfect)
#=====================================================
models_up = dge_long_genes_up %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = if_else(model_type == "Immune", "Immune", organism)) %>%
  bind_rows(perfect_prediction_up, uniform_prediction_up)

models_down = dge_long_genes_down %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = if_else(model_type == "Immune", "Immune", organism)) %>%
  bind_rows(perfect_prediction_down, uniform_prediction_down)

#=====================================================
# Classify probabilities 
#=====================================================
classified_df_probs_up = truth_human_up %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(symbol, timepoint, pred, prob, model),
             by = join_by(symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct()

classified_df_probs_down = truth_human_down %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(symbol, timepoint, pred, prob, model),
             by = join_by(symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct()


#=====================================================
#Compute ROC 
#=====================================================

roc_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Up") %>% 
  ungroup()


roc_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Down") %>% 
  ungroup()

#=====================================================
#Compute AUC 
#=====================================================
auc_results_up = classified_df_probs_up %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(model, timepoint, auc, comparison)

auc_results_down = classified_df_probs_down %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(model, timepoint, auc, comparison)

#Plot curves ------------------
#PLOT ROC Curve
roc_curve_plot = roc_curve_up_notup %>% 
  bind_rows(roc_curve_down_notdown) %>%
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot(aes(x = 1 - specificity, 
             y = sensitivity, 
             color = model)) +
  geom_path(size = 1, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_grid(~fct_rev(comparison)~timepoint) +
  scale_color_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "Human" = "#90A4AEFF",
    "Immune" = "#4361ee",
    "Uniform" = "gray80"
    )) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model") + 
  theme(
    legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3)

roc_curve_plot

# Visualize AUC
auc_plot <- bind_rows(auc_results_up, 
                      auc_results_down) %>% 
  mutate(model = factor(model, levels = c("Perfect", "Mouse", "Immune", "Uniform")),
         comparison = factor(comparison, levels = c("Up", "Down"))) %>%
  
  ggplot(aes(x = auc, 
             y = fct_reorder(timepoint, as.numeric(timepoint)) %>% fct_rev(), 
             fill = model)) +
  geom_col(width = 1,
           position = position_dodge(1)) +
  
  # Add a reference line for random prediction (AUC = 0.5)
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black", linewidth = 0.5) +
  
  geom_text(aes(label = round(auc, 2)),
            position = position_dodge2(1),
            hjust = -0.2, size = 2, vjust = 0.5) +
  
  facet_grid(comparison ~ ., scales = "free_y") +
  
  labs(
    x = "AUC",
    y = "Day",
    fill = "Model"
  ) +
    scale_fill_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "Immune" = "#4361ee",
    "Human" = "#90A4AEFF",
    "Uniform" = "gray80"
    )) +
  theme_vaxgo()  +
  theme(
    legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(expand = c(0,0), limits = c(0,1))

roc_auc_curve_plot = ((roc_curve_plot) |
  (auc_plot + theme(legend.position = "none"))) +
  plot_layout(guides = "collect", widths = c(1, 0.4))+
  plot_annotation(theme = theme(legend.position = "top", plot.title = element_text(hjust = 0.5)),
                  title = paste(filename, gene_set_filename))

roc_auc_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_", gene_set_filename, ".png")),
         width = 10,
         height = 5)

roc_auc_curve_plot
```


## ROC Curve: BTMs

```{r}
#Import ============
#Import DEGs table
dge_btm_mean_process = readRDS(file = here("tables", "saureus_infection_human_mouse_dge_btm_mean_process.rds"))

#Import metadata
human_mouse_metadata = readRDS(file = here("tables", "saureus_infection_human_mouse_metadata.rds")) %>% 
  mutate(timepoint = if_else(organism == "Human", 24, timepoint),
         time_unit = "hours")



#Pre-process data =============
all_btms_human_mouse = dge_btm_mean_process 

human_mouse_metadata %>% select(timepoint, organism) %>% distinct()
all_btms_human_mouse %>% select(timepoint, organism) %>% distinct()

#Timpoints differ from human and model
human_data_reference <- all_btms_human_mouse %>% filter(organism == "Human")
new_timepoints <- c(all_btms_human_mouse %>% filter(organism != "Human") %>% pull(timepoint) %>% unique() %>% as.character())
duplicated_human_data <- map_dfr(new_timepoints, ~ {
  df_copy <- human_data_reference
  df_copy <- df_copy %>%
    mutate(timepoint = .x)

  return(df_copy)
})

all_btms_human_mouse_matched = all_btms_human_mouse %>% 
  filter(organism != "Human") %>% 
  bind_rows(duplicated_human_data) 
#Check
all_btms_human_mouse_matched %>% select(timepoint, organism) %>% distinct()

#Input ======================
set.seed(123)
filename = "saureus_infection"
gene_set_filename = "BTMs"

btms_df_input = all_btms_human_mouse_matched %>% 
  rename(symbol = process)

```

Calculate ROC curve


```{r fig.height=5, fig.width=10}
#=====================================================
# Compute probabilities
#=====================================================

dge_long_genes_up = btms_df_input %>% 
  mutate(
    df = n_genes - 1,
    t = mean_log2fc/(sd_log2fc/sqrt(n_genes)), 
    pval_one_sided = pt(t, df, lower.tail = FALSE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t > 0, "up", "not_up"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  rename(adj_p_val = pval_one_sided_adj)

dge_long_genes_down = btms_df_input %>% 
  mutate(
    df = n_genes - 1,
    t = mean_log2fc/(sd_log2fc/sqrt(n_genes)), 
    pval_one_sided = pt(t, df, lower.tail = TRUE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t < 0, "down", "not_down"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  rename(adj_p_val = pval_one_sided_adj)

#=====================================================
# Assign human truth
#=====================================================
truth_human_up = dge_long_genes_up %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

truth_human_down = dge_long_genes_down %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#=====================================================
#Generate reference distributions
#=====================================================
perfect_prediction_up = truth_human_up %>%
  mutate(prob = if_else(truth == "up", 1, 0),
         model = "Human") %>%
  rename(pred = truth)

uniform_prediction_up = perfect_prediction_up %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()), organism = "Uniform",
         model = "Uniform")

perfect_prediction_down = truth_human_down %>%
  mutate(prob = if_else(truth == "down", 1, 0),
          model = "Human") %>%
  rename(pred = truth)

uniform_prediction_down = perfect_prediction_down %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()), organism = "Uniform",
         model = "Uniform")

#=====================================================
# Combine models (Mouse + Immune + Uniform + Perfect)
#=====================================================
models_up = dge_long_genes_up %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_up, uniform_prediction_up)

models_down = dge_long_genes_down %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = organism) %>%
  bind_rows(perfect_prediction_down, uniform_prediction_down)

#=====================================================
# Classify probabilities 
#=====================================================
classified_df_probs_up = truth_human_up %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(symbol, timepoint, pred, prob, model),
             by = join_by(symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct()

classified_df_probs_down = truth_human_down %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(symbol, timepoint, pred, prob, model),
             by = join_by(symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct()


#=====================================================
#Compute ROC 
#=====================================================

roc_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Up")

roc_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Down")

#=====================================================
#Compute AUC 
#=====================================================
auc_results_up = classified_df_probs_up %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(model, timepoint, auc, comparison)

auc_results_down = classified_df_probs_down %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(model, timepoint, auc, comparison)

#Plot curves ------------------
#PLOT ROC Curve
roc_curve_plot = roc_curve_up_notup %>% 
  bind_rows(roc_curve_down_notdown) %>%
  ungroup() %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot(aes(x = 1 - specificity, 
             y = sensitivity, 
             color = model)) +
  geom_path(size = 1, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_grid(~fct_rev(comparison)~timepoint) +
  scale_color_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "Human" = "#90A4AEFF",
    "Immune" = "#4361ee",
    "Uniform" = "gray80"
    )) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model") + 
  theme(
    legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3)

roc_curve_plot


# Visualize AUC
auc_plot <- bind_rows(auc_results_up, 
                      auc_results_down) %>% 
  mutate(model = factor(model, levels = c("Perfect", "Mouse", "Immune", "Uniform")),
         comparison = factor(comparison, levels = c("Up", "Down"))) %>%
  
  ggplot(aes(x = auc, 
             y = fct_reorder(timepoint, as.numeric(timepoint)) %>% fct_rev(), 
             fill = model)) +
  geom_col(width = 1,
           position = position_dodge(1)) +
  
  # Add a reference line for random prediction (AUC = 0.5)
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black", linewidth = 0.5) +
  
  geom_text(aes(label = round(auc, 2)),
            position = position_dodge2(1),
            color = "black", 
            hjust = 1.2, size = 3, vjust = 0.5) +
  
  facet_grid(comparison ~ ., scales = "free_y") +
  
  labs(
    x = "AUC",
    y = "Day",
    fill = "Model"
  ) +
    scale_fill_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "Immune" = "#4361ee",
    "Human" = "#90A4AEFF",
    "Uniform" = "gray80"
    )) +
  theme_vaxgo()  +
  theme(
    legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(expand = c(0,0), limits = c(0,1))

roc_auc_curve_plot = ((roc_curve_plot) |
  (auc_plot + theme(legend.position = "none"))) +
  plot_layout(guides = "collect", widths = c(1, 0.3))+
  plot_annotation(theme = theme(legend.position = "top", plot.title = element_text(hjust = 0.5)),
                  title = paste(filename, gene_set_filename))

roc_auc_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_", gene_set_filename, ".png")),
         width = 10,
         height = 5)

roc_auc_curve_plot
```

