---
title: "S. aureus"
author: "Wasim Aluísio Prates-Syed"
date: "2025-07-23"
output: 
  html_document: 
    toc: TRUE
    toc_depth: 4
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE, cache = TRUE)
```


## Import required codes
```{r message=FALSE, warning=FALSE}
# install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```


# S. aureus infection

## GSE19668 (Mouse)

### Retrieve dataset
Retrieve Exprs and Metadata from GEO
```{r}
#S aureus: GSE19668 -----
filename = "saureus_infection_GSE19668"
GSE19668 <- getGEO("GSE19668", GSEMatrix = TRUE) 
eset <- GSE19668[[1]]

#Metadata ------
GSE19668_metadata = eset %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    subject_id = "title",
    # sex = "gender:ch1", 
    treatment = "infection:ch1", 
    # tissue = "tissue:ch1",
    timepoint = "time point:ch1",
    # animal_replicate = "animal replicate:ch1",
    strain = "strain:ch1") %>% 
  mutate(treatment = fct_relevel(treatment, "none"),
         timepoint = fct_relevel(timepoint, "0"),
         condition = paste(treatment, strain, timepoint, sep = "_"))

#Save
GSE19668_metadata %>% saveRDS(file = here("tables", paste0(filename, "_metadata.rds")))


#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% as.data.frame() %>% saveRDS(here("tables", paste0(filename, "_exprs.rds")))
```

Clean metadata
```{r}
#Import
GSE19668_metadata = readRDS(file = here("tables", "saureus_infection_GSE19668_metadata.rds"))

#Filter metadata
GSE19668_metadata_filtered = GSE19668_metadata %>%
  filter(strain == "C57BL/6J") %>% 
  separate(subject_id, into = c("tissue", "strain1", "timepoint1", "participant_id"), sep = " ") %>% 
  select(!c(tissue:timepoint1)) %>% 
  arrange(participant_id, timepoint) %>% 
  separate(timepoint, into = c("timepoint_number", "time_unit")) %>% 
  mutate(time_unit = "hours",
         timepoint = timepoint_number %>% as.numeric(),
         organism = "Mouse") %>% 
  arrange(timepoint) %>% 
  select(-timepoint_number) %>% 
  select(geo_sample, participant_id, condition, timepoint, time_unit, everything())

#Save
GSE19668_metadata_filtered %>% saveRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

GSE19668_metadata %>% filter(treatment == "Staphylococcus aureus")
```


```{r}
#Import
saureus_mouse_exprs = readRDS(here("tables","saureus_infection_GSE19668_exprs.rds"))
saureus_mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

#Inputs
filename = "saureus_infection_mouse"
mouse_exprs = saureus_mouse_exprs 
mouse_metadata = saureus_mouse_metadata
```


```{r}

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
mouse_exprs_metadata  = mouse_exprs %>% 
  select(c(mouse_metadata$geo_sample)) %>%
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = c(mouse_metadata$geo_sample),
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(mouse_metadata, 
             by = "geo_sample") 


#Get value from day 0 and calculate Log2FC
mouse_exprs_log2fc_sample = mouse_exprs_metadata %>% 
  filter(timepoint == 0) %>% 
  select(probe_id, participant_id, day0_value_raw = value_raw) %>% 
  #Filter batch
  inner_join(mouse_exprs_metadata %>% 
               filter(timepoint != "0"),
             by = join_by("probe_id", "participant_id")) %>% 
  mutate(log2fc = value_raw - day0_value_raw) %>% 
  distinct() 

#Check duplicates
mouse_exprs_log2fc_sample %>% 
  filter(participant_id == "rep1",
         probe_id == "1415670_at")
```


```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pull(probe_id) %>% unique()

# listAttributes(ensembl_mart)
#Annotate
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2", #Probes
  values = gene_id,
  mart = ensembl_mart)

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_log2summary = mouse_exprs %>% #20,845 probes
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_mouse430_2), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = mouse_log2summary %>% 
  #filter ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

mouse_log2summary = mouse_log2summary %>% 
  select(probe, hgnc_symbol) %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
mouse_log2_filtered_probes = mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
  inner_join(mouse_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol,
         probe,
         c(mouse_metadata %>% 
             pull(geo_sample)))
#Check
mouse_log2_filtered_probes %>% dim()
mouse_log2_hgnc_symbol = mouse_log2_filtered_probes 
mouse_log2_hgnc_symbol %>% get_dupes(human_symbol)

#Save
mouse_log2_hgnc_symbol %>% saveRDS(file = here("tables", paste0(filename, "_exprs_hgnc_symbol.rds")))
```

Pivot data and create expression set
```{r}
#Import
mouse_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_mouse_exprs_hgnc_symbol.rds"))
mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))

#Pivot to long and wide format, and annotate
mouse_log2_samples_log2fc_long = mouse_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(mouse_metadata,
             by = "geo_sample")


mouse_log2_samples_log2fc_wide = mouse_log2_samples_log2fc_long %>% 
  select(human_symbol, probe, participant_id, timepoint, log2fc) %>%
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc") %>% 
  mutate(timepoint = as.character(timepoint) %>% fct_reorder(timepoint %>% as.numeric()))


#Create Expression Set
eset_mouse <- ExpressionSet(
  assayData = mouse_log2_hgnc_symbol %>% 
    select(-probe) %>% 
    column_to_rownames("human_symbol") %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(mouse_metadata))

#Save
eset_mouse %>% saveRDS(file = here("tables", "saureus_infection_mouse_eset.rds"))

```


### Differential expression 

#### Limma
Paired samples design
```{r}

#Import
mouse_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_mouse_exprs_hgnc_symbol.rds"))
mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))
eset_mouse = readRDS(file = here("tables", "saureus_infection_mouse_eset.rds"))

#Create model matrix
design <- model.matrix(~ condition + 0, # 0 defines the way we compare the samples
                       eset_mouse) 

colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

# Estimate the correlation between measurements made on the same subject (paired-samples)
corfit <- duplicateCorrelation(eset_mouse, 
                               design = design, 
                               block = eset_mouse$participant_id)
corfit$consensus #0.001357108

#Fit Limma model
fit <- lmFit(eset_mouse, 
             design, 
             block = eset_mouse$participant_id, 
             correlation = corfit$consensus) #Use inter-subject correlation as input into the linear model fit

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(conditionStaphylococcus.aureus_C57BL.6J_2.hours - conditionnone_C57BL.6J_0,
                             conditionStaphylococcus.aureus_C57BL.6J_4.hours - conditionnone_C57BL.6J_0,
                             conditionStaphylococcus.aureus_C57BL.6J_6.hours - conditionnone_C57BL.6J_0,
                             conditionStaphylococcus.aureus_C57BL.6J_12.hours - conditionnone_C57BL.6J_0,
                             levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs = all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "conditionnone_C57BL.6J_0", "0")) %>% 
  mutate(condition = case_when(comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_2.hours"	~ "Saureus_2h",
                               comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_4.hours"	~ "Saureus_4h",
                               comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_6.hours"	~ "Saureus_6h",
                               comparisonA == "conditionStaphylococcus.aureus_C57BL.6J_12.hours"	~ "Saureus_12h",
                               T~comparisonA),
         organism = "Mouse",
         timepoint = str_remove(condition, "Saureus_") %>% str_remove("h")) %>% 
  select(condition, everything()) %>% 
  clean_names() %>% 
  rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs %>% saveRDS(file = here("tables", "saureus_infection_mouse_dge_limma_degs.rds"))

all_degs %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarize(n_degs = n())

```



##Import


```{r}
mouse_degs = readRDS(file = here("tables", "saureus_infection_mouse_dge_limma_degs.rds"))
```



#### Correlation plots, pairwise
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, GSE19668, C57BL/6J", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_alldays
#Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)
beep()



#Timepoint 2
timepoint_select = 2
mouse_correlation_ggpairs_day2 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == 2) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day2
#Save
mouse_correlation_ggpairs_day2 %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_day2.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 6
timepoint_select = 6
mouse_correlation_ggpairs_day6 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day6
#Save
mouse_correlation_ggpairs_day6 %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_day6.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 12
timepoint_select = 12
mouse_correlation_ggpairs_day12 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day12
#Save
mouse_correlation_ggpairs_day12 %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_correlation_ggpairs_day12.png"),
         height = 10,
         width = 10)
beep()

```




#### PCA

```{r fig.height=4, fig.width=5}
#Import metadata
annotation_samples_pca = GSE19668_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = GSE19668_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, entrezgene_id, log2fc) %>% 
  filter(!entrezgene_id == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("entrezgene_id") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Mouse, S. aureus, GSE19668",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

```

#### BTMs
```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)

#Annotate genes by BTM
GSE19668_genes_btm = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  rename(mm_entrez = entrezgene_id) %>% 
  select(mgi_symbol, symbol = human_symbol, participant_id, geo_sample, timepoint, treatment, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = GSE19668_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process
```



##### GGPairs
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

mouse_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) 

mouse_correlation_ggpairs_btms_gsea_alldays

mouse_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_gsea_btm_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)


#All days, colored by group
mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor = ggpairs_data_dge_btm_mean_process %>%
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = group,
                                 fill = group),
          title = paste("Day", timepoint_selected),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors)

mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor

mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor %>%
  ggsave(filename = here("Figures", "saureus_infection_mouse_gsea_btm_correlation_ggpairs_alldays_groupcolor.png"),
         height = 10,
         width = 10)

beep()

```




##### Ggplot manual
```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(3:7) %>% 
  colnames()


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == 2) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == 2) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = "GSE19668, Mouse, SAureus infeciton, random sample")
```













## GSE33341 (Human)

### Prepare data
```{r}
#S. aureus: GSE33341 -----
filename = "GSE33341"
GSE <- getGEO("GSE33341", GSEMatrix = TRUE) 
eset <- GSE[[2]] #The matrix #1 the mouse annotated

#Metadata ------
GSE_metadata = eset %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    participant_id = "title",
    sex = "host gender:ch1",
    condition = "source_name_ch1",
    pathogen = "pathogen:ch1",
    description = "source_name_ch1", 
    ethnicity = "ethnicity:ch1",
    tissue_infection = "anatomic site of infection:ch1",
    batch_experimental = "experimental batch:ch1",
    age = "host age (yr):ch1") %>% 
  mutate(condition = pathogen,
         organism = "Human",
         timepoint = 1,
         time_unit = "Day",
         tissue = "Blood",
         participant_id = str_extract(participant_id, "\\[.*?\\]") %>% 
           str_remove("\\[") %>% 
           str_remove("\\]"),
         condition = case_when(description == "blood patient with E. coli bacteremia prior antibiotic treatment (Emergency Department)" ~ "E. coli prior antibiotic",
                               description == "blood patient with S. aureus bacteremia prior antibiotic treatment (Emergency Department)" ~ "S aureus prior antibiotic",
                               description == "blood human NOT infected" ~ "No infection",
                               T ~ description),
         pathogen = if_else(pathogen == "-", "Healthy", pathogen))
#Save
GSE_metadata %>% saveRDS(file = here("tables", "saureus_infection_human_metadata.rds"))

#Filter metadata
GSE_metadata_filtered = GSE_metadata %>%
  filter(pathogen %in% c("Staphylococcus aureus", "Healthy")) %>% 
  select(geo_sample, participant_id, condition, timepoint, everything()) %>% 
  mutate(ethnicity = str_to_title(ethnicity)) %>% 
  filter(batch_experimental != "2739",
         ethnicity != "Unknown",
         age != "unknown")

GSE_metadata_filtered %>% 
  saveRDS(file = here("tables", "saureus_infection_human_metadata.rds"))



#Expression data ------------
GSE_exprs <- exprs(eset) %>% as.data.frame()

#Save
GSE_exprs %>% saveRDS(file = here("tables", "saureus_infection_human_exprs.rds"))
```

Prepare data
```{r}
#Import
saureus_human_exprs = readRDS(file = here("tables", "saureus_infection_human_exprs.rds"))
saureus_human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))


#Inputs
filename = "saureus_infection_human"
human_exprs = saureus_human_exprs 
human_metadata = saureus_human_metadata
```


```{r}
#Annotate genes ------
#Human -----
#Annotate data and calculate log2fc
human_exprs_metadata  = human_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = -probe_id,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(human_metadata, 
             by = "geo_sample") 

#Get value from healthy and calculate Log2FC
human_exprs_log2fc_sample = human_exprs_metadata %>% 
  filter(pathogen == "Healthy") %>% 
  select(probe_id, value_raw_ref = value_raw) %>% 
  group_by(probe_id) %>% 
  summarize(value_raw_ref = mean(value_raw_ref)) %>% 
  inner_join(human_exprs_metadata %>% 
               filter(pathogen != "Healthy"),
             by = join_by("probe_id")) %>% 
  filter(condition != "No infection") %>% 
  mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

#Check duplicates
human_exprs_log2fc_sample %>% 
  filter(participant_id == "CSD4620.CEL",
         probe_id == "1007_s_at")
```


```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
# listDatasets(ensembl_mart)
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- human_exprs_log2fc_sample$probe_id %>% unique()

#Annotate
annot <- getBM(
  attributes = c("affy_hg_u133a_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133a_2",
  values = gene_id,
  mart = ensembl_mart) %>% 
  #Standardize
  rename(probe = affy_hg_u133a_2,
                     entrezgene_id = entrezgene_id,
                     human_symbol = hgnc_symbol,
                     ensembl_id = ensembl_gene_id)

#Filter genes with highest median expression
human_log2summary = human_exprs %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot, 
            by = "probe") %>% 
  #Filter ensembl id
  group_by(ensembl_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup()  


#Filter out genes with multiple hgnc_symbols
hgnc_symbol_dupes = human_log2summary %>% 
  #filter ensembl id with multiple hgnc_symbols
  get_dupes(human_symbol)

human_log2summary = human_log2summary %>% 
  filter(!human_symbol %in% c(hgnc_symbol_dupes$human_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
human_log2_filtered_probes = human_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(human_log2summary, by = "probe") %>%   #214,396
  select(human_symbol, probe,
         c(human_metadata$geo_sample))
#14395
human_log2_filtered_probes %>% dim()

human_log2_filtered_probes = human_log2_filtered_probes 

human_log2_filtered_probes %>% 
  get_dupes(human_symbol)

#Save 
human_log2_filtered_probes %>% 
  saveRDS(here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
```


```{r}
#Import
human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))

#Pivot to long and wide format, and annotate
human_log2_samples_log2fc_long = human_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(human_metadata,
             by = "geo_sample")


human_log2_samples_log2fc_wide = human_log2_samples_log2fc_long %>% 
  select(human_symbol, probe, participant_id, timepoint, log2fc) %>%
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc") %>% 
  mutate(timepoint = as.character(timepoint) %>% fct_reorder(timepoint %>% as.numeric()))


#Create Expression Set
human_metadata_eset = human_metadata %>% 
  mutate(age = as.numeric(age) %>% scale(),
         timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         pathogen = fct_relevel(pathogen, "Healthy"),
         batch_experimental = as.factor(batch_experimental),
         ethnicity = as.factor(ethnicity)) 

eset_human <- ExpressionSet(
  assayData = human_log2_hgnc_symbol %>% 
    select(-probe) %>% 
    column_to_rownames("human_symbol") %>% 
    select(c(human_metadata_eset$geo_sample)) %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(human_metadata_eset))

#Save
eset_human %>% saveRDS(file = here("tables", "saureus_infection_human_eset.rds"))

```






### Differential expression

#### Limma

```{r}
#Import
human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))
eset_human = readRDS(file = here("tables", "saureus_infection_human_eset.rds"))

#Create model matrix
design <- model.matrix(~ age + sex + ethnicity + batch_experimental + pathogen,
                       data = pData(eset_human))

colnames(design) <- make.names(colnames(design))  
rownames(design) <- rownames(pData(eset_human))

#Fit Limma model
fit <- lmFit(eset_human, 
             design) 

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(pathogenStaphylococcus.aureus, levels = design)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- limma::eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_human = all_contrasts_degs %>% 
  dplyr::select(comparison, everything()) %>% 
  dplyr::mutate(condition = case_when(comparison == "pathogenStaphylococcus.aureus"	~ "Saureus_Day1",
                               T~comparison),
         organism = "Human",
         timepoint = "1") %>% 
  dplyr::select(condition, everything()) %>% 
  clean_names() %>% 
  dplyr::rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs_human %>% saveRDS(file = here("tables", "saureus_infection_human_dge_limma_degs.rds"))

all_degs_human %>% 
  dplyr::group_by(condition) %>%
  dplyr::filter(adj_p_val <= 0.05) %>% 
  dplyr::summarize(n_degs = n())

```


#### t-Test
```{r}
#Import ================
human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))

#Wrangle ==============
#Pivot to long and wide format, and annotate
human_log2_samples_log2fc_long = human_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(human_metadata,
             by = "geo_sample")

human_log2_wide = human_log2_hgnc_symbol %>%
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(human_metadata %>% 
               select(geo_sample, participant_id, pathogen),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_")

human_exprs_log2_infected_healthy = human_exprs_metadata %>% 
  filter(pathogen == "Healthy") %>% 
  select(probe_id, value_raw_ref = value_raw) %>% 
  group_by(probe_id) %>% 
  summarize(value_raw_ref = mean(value_raw_ref)) %>% 
  inner_join(human_exprs_metadata %>% 
               filter(pathogen != "Healthy"),
             by = join_by("probe_id")) %>% 
  filter(pathogen != "Healthy") %>% 
  inner_join(human_log2_samples_log2fc_long %>% 
               select(probe_id = probe, human_symbol),
             by = join_by("probe_id")) %>% 
  mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

# Unpaired t-test ----------
human_log2fc_unpaired_ttest <- human_exprs_log2_infected_healthy %>%
  group_by(human_symbol) %>%
  
  summarise(
    pval = tryCatch(
      t.test(value_raw, value_raw_ref,
             paired = FALSE, var.equal = FALSE, 
             alternative = "two.sided")$p.value,
      error = function(e) NA_real_
    ),
    .groups = 'drop'
  ) %>%
  mutate(
    padj = p.adjust(pval, method = "fdr")
    ) %>% 
  mutate(
    condition = "Saureus infection",
    timepoint = "1",
    test = "t-Test",
    paired = "Unpaired")


human_log2fc_unpaired_ttest %>% saveRDS(file = here("tables", "saureus_infection_human_dge_unpaired_ttest.rds"))

```







#### Import
```{r}
#Import
GSE33341_counts_log2_summary_annotated_probes = 
  readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol_long.rds"))

human_log2_hgnc_symbol = readRDS(file = here("tables", "saureus_infection_human_exprs_hgnc_symbol.rds"))



#Pivot to long and wide format, and annotate
human_log2_samples_log2fc_long = human_log2_hgnc_symbol %>% 
  pivot_longer(cols = -c(human_symbol, probe),
               names_to = "geo_sample",
               values_to = "log2fc") %>% 
  inner_join(human_metadata,
             by = "geo_sample")


human_log2_samples_log2fc_wide = human_log2_samples_log2fc_long %>% 
  select(human_symbol, probe, participant_id, timepoint, log2fc) %>%
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc") %>% 
  mutate(timepoint = as.character(timepoint) %>% fct_reorder(timepoint %>% as.numeric()))

#Pivot wider
GSE33341_log2fc_sample_wide = GSE33341_counts_log2_summary_annotated_probes %>%
  select(human_symbol, participant_id, timepoint, log2fc) %>% 
  pivot_wider(names_from = participant_id,
              values_from = log2fc) %>% 
  ungroup()
```





# Human vs Mouse

## Combine datasets

```{r}
all_degs_mouse = read_rds(here("tables", "saureus_infection_mouse_dge_limma_degs.rds"))
all_degs_human = read_rds(here("tables", "saureus_infection_human_dge_limma_degs.rds"))

#Combine Limma DEGs from mice and humans
all_degs_human_mouse = all_degs_mouse %>% 
  bind_rows(all_degs_human) %>% 
  select(condition, timepoint, human_symbol, mean_l2fc:organism)
#Save
all_degs_human_mouse %>% saveRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds"))

#Combine metadata from mice and humans
mouse_metadata = readRDS(file = here("tables", "saureus_infection_mouse_metadata_filtered.rds"))
human_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))

human_mouse_metadata = mouse_metadata %>% bind_rows(human_metadata)
human_mouse_metadata %>% saveRDS(file = here("tables", "saureus_infection_human_mouse_metadata.rds"))


```



## Correlation
```{r fig.height=5, fig.width=5}

# Calculate the correlation
cor_data <- human_mouse_saureus_log2fc_avg_wide %>%
  filter(padj_human <= 0.05 | padj_mouse <= 0.05) %>%
  summarize(
    cor_value = cor(Mouse, Human, method = "pearson", use = "complete.obs"),
    label_pos_x = max(Mouse, na.rm = TRUE),
    label_pos_y = max(Human, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("r = ", round(cor_value, 2)))



ylim_range = c(-5, 5)
xlim_range = c(-5, 5)
human_mouse_saureus_log2fc_avg_wide_scatterplot = human_mouse_saureus_log2fc_avg_wide %>% 
  filter(padj_human <= 0.05 | padj_mouse <= 0.05) %>% 
  
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_saureus_log2fc_avg_wide %>%
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_saureus_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  stat_smooth(method = "lm", geom = "line", color = "#4DBBD5FF", alpha = 0.5, se = F, size = 1,
              data = human_mouse_saureus_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add correlation metrics (Pearson)
  geom_label(
    data = cor_data,
    aes(x = label_pos_x, y = label_pos_y, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 1,
    vjust = 1,
    size = 3
  ) +
  
  #Adjust the axes range
  ylim(ylim_range) +
  xlim(xlim_range) +
  
  #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "none",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title= "Correlation plot, Log2-Fold Change, GSE19668 and GSE33341 ")

human_mouse_saureus_log2fc_avg_wide_scatterplot

```


#### Plot correlation between genes (log2fc for every sample)

```{r}
#Calculate correlation ------
correlation_long_all_mouse_human_timepoints <- human_mouse_saureus_log2fc_samples %>%
  group_by(timepoint, symbol = human_symbol, participant_id) %>%
  summarise(log2fc = max(log2fc, na.rm = TRUE), .groups = "drop") %>% 
  group_by(timepoint) %>% 
  distinct() %>% 
  group_modify(~ {
    .x %>%
      pivot_wider(
        id_cols = symbol,
        names_from = participant_id,
        values_from = log2fc
      ) %>%
      select(-symbol) %>%
      cor(use = "pairwise.complete.obs", method = "pearson") %>%
      as.data.frame() %>%
      rownames_to_column("participant1") %>%
      pivot_longer(-participant1, names_to = "participant2", values_to = "correlation") %>%
      filter(participant1 != participant2)
  }) %>%
  ungroup() %>% 
  inner_join(GSE19668_metadata_filtered %>% 
               select(participant_id, participant_1_organism = organism, sex_p_1 = sex) %>% 
               distinct(),
             by = join_by("participant1" == "participant_id")) %>% 
  inner_join(GSE19668_metadata_filtered %>% 
               select(participant_id, participant_2_organism = organism, sex_p_2 = sex) %>% 
               distinct(),
             by = join_by("participant2" == "participant_id"))


# Remove duplicated lines only if the organism is the same
correlation_long_all_timepoints_same_organisms = correlation_long_all_mouse_human_timepoints %>%
  filter(participant_1_organism == participant_2_organism) %>%
  mutate(
    participant_min = pmin(participant1, participant2),
    participant_max = pmax(participant1, participant2)
  ) %>%
  # Remove duplicates (ex: A vs B e B vs A)
  distinct(timepoint, participant_min, participant_max, .keep_all = TRUE) %>%
  select(timepoint, participant1, participant2, correlation, participant_1_organism, participant_2_organism, sex_p_1, sex_p_2) %>%
  group_by(timepoint) %>%
  arrange(-correlation)


#Unite table with different organisms with table with deduplicated rows with the same organism
correlation_long_all_mouse_human_timepoints = correlation_long_all_mouse_human_timepoints %>% 
  filter(participant_1_organism != participant_2_organism) %>% 
  bind_rows(correlation_long_all_timepoints_same_organisms)

#Summarize data and save table
correlation_long_all_mouse_human_timepoints_summary = correlation_long_all_mouse_human_timepoints %>% 
  
  #Filter only female and set Mouse as reference
  filter(sex_p_1 == "Female", 
         sex_p_2 == "Female") %>% 
  mutate(sex_comparison = paste(sex_p_1, "vs.", sex_p_2),
         comparison = paste(participant_1_organism, "vs.", participant_2_organism)) %>%
  group_by(timepoint, comparison) %>% 
  arrange(timepoint, -correlation) %>% 
  mutate(correlation = round(correlation, 2),
         timepoint = paste("Day", timepoint)) %>% 
  summarize(mean = mean(correlation, na.rm = T) %>% round(2),
            median = median(correlation, na.rm = T)%>% round(2),
            min = min(correlation, na.rm = T)%>% round(2),
            max = max(correlation, na.rm = T) %>% round(2),
            sd = sd(correlation),
            n = sum(!is.na(correlation)),
            se = sd/sqrt(n),
            t_critical = qt(0.975, df = n - 1),
            ci_lower = mean - t_critical * se,
            ci_upper = mean + t_critical * se,
            .groups = "drop") %>% 
  filter(comparison != "Human vs. Mouse")

correlation_long_all_mouse_human_timepoints_summary

#Create GT table
correlation_long_all_mouse_human_timepoints_gt = correlation_long_all_mouse_human_timepoints_summary %>% 
  gt::gt() 
#Save gt
correlation_long_all_mouse_human_timepoints_gt
gtsave(correlation_long_all_mouse_human_timepoints_gt, filename = here("tables", "correlation_long_all_mouse_human_Saureus_timepoints_gt.docx"))

```


```{r fig.height=5, fig.width=5}
correlation_long_all_mouse_human_timepoints_summary_plot = correlation_long_all_mouse_human_timepoints_summary %>% 
  separate(timepoint, into = c("unit", "timepoint"), remove = T) %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1) +
  
  geom_errorbar(aes(ymin = ci_lower, 
                  ymax = ci_upper), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 1, 3, 7), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_color_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = "Correlation between same and different organisms, by subject",
       x = "",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 
  
```




#### BTMs


```{r}
#Annotate genes by BTM
dge_btm_mean_process = GSE33341_genes_btm %>%
  mutate(organism = "Human") %>% 
  bind_rows(GSE19668_genes_btm %>% 
              mutate(organism = "Mouse")) %>%
  filter(timepoint %in% c(1, 2)) %>% 
  mutate(condition = paste0(organism, "_Day", timepoint)) %>% 
  group_by(condition, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  select(-timepoint) %>% 
  pivot_wider(names_from = "condition", values_from = "mean_log2fc") 

dge_btm_mean_process
```
##### Scatterplot
```{r fig.height=5, fig.width=7}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(4:5) %>% 
  colnames()


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  select(group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      select(group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = participant_2_value, y = participant_1_value)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  stat_smooth(aes(x = participant_2_value, y = participant_1_value),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "Mouse Log2FC", 
       y = "Human Log2FC", 
       color = "",
       title = "Mouse vs. Human")



```



##### Heatmap

```{r}
#Matrix data
matrix_data = dge_btm_mean_process %>% 
  select(process, Human_Day1, Mouse_Day2) %>% 
  drop_na() %>% 
  column_to_rownames("process") %>% 
  as.matrix()

# Column and row annotation
######## Annotations
annotation_heatmap = tribble(~condition, ~timepoint, ~organism,
        "Human_Day1", 1, "Human",
        "Mouse_Day2", 2, "Mouse")

#Arrange columns in the same order as the matrix
ann_cols_heatmap = matrix_data %>% 
  colnames() %>% 
  data.frame() %>% 
  rename(condition = '.') %>% 
  inner_join(annotation_heatmap, by = "condition") %>% 
  distinct() %>% 
  arrange(condition)

#Arrange matrix columns 
mat = matrix_data %>% 
  as.data.frame() %>% 
  select(c(ann_cols_heatmap$condition %>% unique())) %>% 
  as.matrix() 

#Column to rownames
ann_cols_heatmap = ann_cols_heatmap %>% 
  column_to_rownames("condition")

#Annotate rows
ann_rows_heatmap = mat %>% 
  rownames() %>% 
  as.data.frame() %>% 
  rename("process" = ".") %>% 
  inner_join(dge_btm_mean_process %>% 
               select(process, group, subgroup), 
             by = "process") %>% 
  column_to_rownames("process")
  

#Check dimensions
dim(mat)
dim(ann_cols_heatmap)

#Colors from required.R
btm_immune_groups
colors_immune

```


```{r fig.width= 6, fig.height=20}
#Heatmap annotation
ha = HeatmapAnnotation(df = ann_cols_heatmap, 
                       col = colors_immune,
                       gp = gpar(col = "black"), 
                       simple_anno_size = unit(0.3, "cm"),
                       annotation_name_side = "right")

row_ha = HeatmapAnnotation(df = ann_rows_heatmap,
                       col = btm_immune_groups,
                       which= "row",
                       simple_anno_size = unit(0.2, "cm"),
                       show_annotation_name = T,
                       show_legend = T)

# Values colors
col_fun <- colorRamp2(c(-2,0,2), c("#45efc6", "white", "#165289"))

file = "BTM_Modules_Saureus_Mouse_Human"

width_image = 25
height_image = 30
width = unit(1, "cm")
height = unit(25, "cm")
column_names_gp = gpar(fontsize = 8)
row_names_gp = gpar(fontsize = 8)
rect_gp = gpar(col = "black", lwd = 0.5)

fileimageheatmap_grouped = here("Figures", paste0(file, sep ="_", "Complexheatmap.png"))

png(file = fileimageheatmap_grouped, 
    width = width_image, 
    height = height_image, 
    units="cm", 
    res=900)

heatmap_plot = Heatmap(mat,
                        top_annotation = ha, 
                       left_annotation = row_ha,
                        show_row_names = TRUE,
                        row_names_side = "right",
                        show_heatmap_legend = TRUE,
                        heatmap_legend_param = list(direction = "horizontal"),
                        col = col_fun,
                        show_column_names = TRUE,
                        column_title = file,
                        column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                        column_names_rot = 45,
                        column_names_gp = column_names_gp,
                        rect_gp = rect_gp,
                        na_col = "black",
                        name = "Mean Log2FC",
                        cluster_rows = TRUE,
                        row_names_gp = row_names_gp, row_km = 2,
                        row_dend_width = unit(3, "cm"),
                        row_split = NULL,
                        column_split = NULL,
                        row_gap = unit(2, "mm"),
                        cluster_columns = TRUE,
                        column_gap = unit(8, "mm"),
                        show_column_dend = F,
                        show_row_dend = T, 
                        row_dend_side = "left",
                        column_dend_side = "top",
                        clustering_distance_rows = "euclidean",
                        border = TRUE,
                        width = width,
                        height = height)

heatmap_plot = draw(heatmap_plot, 
     annotation_legend_side = "left", 
     heatmap_legend_side = "left")

heatmap_plot

dev.off()


```











# Compare with FIT prediction

```{r}
#Get log2fc mean
GSE19668_input2FIT = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
    filter(timepoint == 2) %>% 
  select(participant_id, MM.Entrez = entrezgene_id, EffectSize = log2fc) %>% 
  group_by(MM.Entrez) %>% 
  summarize(mean_log2fc = mean(EffectSize))

#Save
GSE19668_input2FIT %>% 
  write_csv(file = here("FIT_Mouse2man", "saureus_infection_mouse_fit_prediction_dge_degs_day2.csv"))

#Run FIT
GSE19668_input2FIT_day2 = FIT.mouse2man::FIT(MouseFile = here("FIT_Mouse2man", "saureus_infection_mouse_fit_prediction_dge_degs_day2.csv"), DataType = "rnaseq")
```



### Unite log2FC values from mouse, human, and FIT
```{r fig.height=5, fig.width=8}
#Bind rows
mouse_human_fit_log2fc = bind_rows(GSE19668_input2FIT_day2 %>%
                                     #Get DEGs for FIT
                                     filter((CI_low != 0 | CI_high != 0) & !is.na(CI_low) & !is.na(CI_high)) %>%  
                                   select(human_symbol = Human.symbol,
                                          log2fc = FIT_prediction) %>%
                                     mutate(organism = "FIT"),
                                   
                                   #Get DEGs for Human
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_human <= 0.05) %>% 
                                     select(human_symbol, log2fc = Human) %>%
                                     mutate(organism = "Human"),
                                   
                                   #Get DEGs for Mouse
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_mouse <= 0.05) %>% 
                                     select(human_symbol, log2fc = Mouse) %>%
                                     mutate(organism = "Mouse"))


#Pivot wider
mouse_human_fit_log2fc_wide = GSE19668_input2FIT_day2 %>%
            select(human_symbol = Human.symbol,
                   FIT_prediction) %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Human),
            by = "human_symbol") %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Mouse),
            by = "human_symbol")

#Save
mouse_human_fit_log2fc %>% 
  saveRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))

mouse_human_fit_log2fc_wide %>% 
  saveRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))
```

## Distribution

### Import tables
```{r fig.height=5, fig.width=8}
#Import
mouse_human_fit_log2fc = readRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))
mouse_human_fit_log2fc_wide = readRDS(file = here("tables", "saureus_infection_human_mouse_fit_dge_long.rds"))

GSE33341_metadata = readRDS(file = here("tables", "saureus_infection_human_metadata.rds"))
GSE19668_metadata = readRDS(file = here("tables", "saureus_infection_GSE19668_metadata.rds"))
```

### Plot histogram

```{r fig.height=5, fig.width=8}
#Plot histogram
mouse_human_fit_log2fc_histogram = mouse_human_fit_log2fc %>% 
  ggplot() +
  aes(x = log2fc,
      fill = organism) +
  geom_density(aes(y = after_stat(count),
                   color = organism),
                  position = position_identity(), 
               alpha = 0.1,
               size = 1,
              show.legend = T) +
  scale_fill_manual(values = c(FIT = "#4361ee",
                               Mouse = "#4DBBD5FF",
                               Human = "#546E7AFF"))+
  scale_color_manual(values = c(FIT = "#4361ee",
                                Mouse = "#4DBBD5FF",
                                Human = "#546E7AFF"))+
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        legend.key = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0),
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(limits = c(-4, 4)) +
  labs(color = "",
       fill = "")

mouse_human_fit_log2fc_histogram

#Faceted
mouse_human_fit_log2fc_histogram_faceted = mouse_human_fit_log2fc_histogram +
  facet_wrap(~organism,
             scales = "free_y")


#Plot scatter plot
mouse_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = Mouse,
      y = Human) +
#All DEGs
geom_point(color = "#4DBBD5FF",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "Mouse L2FC",
         y = "Human L2FC")

mouse_human_scatterplot


fit_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = FIT_prediction,
      y = Human) +
#All DEGs
geom_point(color = "#4361ee",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "FIT L2FC",
         y = "Human L2FC")

#Histograms
(((mouse_human_fit_log2fc_histogram + mouse_human_fit_log2fc_histogram_faceted) + 
    plot_layout(widths = c(0.5, 1), guides = "collect") +
    plot_annotation(theme = theme(legend.position = "top")))  /
    #Scatter plots
    (mouse_human_scatterplot + fit_human_scatterplot)) +
  plot_layout(heights = c(0.5, 1))

```



## BTMs
Import
```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)
```

```{r}
#Annotate genes by BTM
dge_btm_mean_process = mouse_human_fit_log2fc %>%
  filter(organism != "Mouse") %>% 
  rename(symbol = human_symbol) %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  group_by(organism, process) %>% 
  summarize(mean_log2fc = mean(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism", values_from = "mean_log2fc") 

dge_btm_mean_process
```
##### Scatterplot
```{r fig.height=5, fig.width=7}

#Calculate correlation
cor_labels <- dge_btm_mean_process %>%
  summarise(cor_value = cor(Human, FIT, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
dge_btm_mean_process %>% 
ggplot(aes(x = FIT, y = Human)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  stat_smooth(aes(x = FIT, y = Human),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "FIT Log2FC", 
       y = "Human Log2FC", 
       color = "",
       title = "Mouse vs. Human")



```









#Performance tests
## ROC Curve
Comparison must be done between human at day 1 and multiple mice timepoints 
```{r}
#Import
all_degs_human_mouse = readRDS(file = here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds"))
human_mouse_metadata = readRDS(file = here("tables", "saureus_infection_human_mouse_metadata.rds"))
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))


#Prepare
all_degs_human_mouse = all_degs_human_mouse %>% 
  inner_join(human_mouse_metadata %>%
               group_by(organism, timepoint) %>%
               summarize(n = n()) %>% 
               mutate(timepoint = as.character(timepoint)),
               by = join_by("organism", "timepoint")) %>% 
  mutate(symbol = human_symbol) %>% 
  mutate(timepoint = if_else(organism == "Human", "24", timepoint))


# Duplicate the human reference data to match with the mice timepoints
human_data_day1 <- all_degs_human_mouse %>% filter(organism == "Human")
new_timepoints <- c("2", "4", "6", "12")
duplicated_human_data <- map_dfr(new_timepoints, ~ {
  df_copy <- human_data_day1
  df_copy <- df_copy %>%
    mutate(timepoint = .x)

  return(df_copy)
})

all_degs_human_mouse = all_degs_human_mouse %>% 
  filter(organism != "Human") %>% 
  bind_rows(duplicated_human_data) 

all_degs_human_mouse %>% 
  select(organism, timepoint) %>% 
  distinct()

#Input
dge_long_genes = all_degs_human_mouse

#All genes
genes_roc_allgenes = dge_long_genes
#Filter immune genes
genes_roc_allgenes_immunegenes = dge_long_genes %>% 
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))

#Only DEGs
degs_genes_human = dge_long_genes %>%
  filter(adj_p_val <= 0.05,
         organism == "Human") %>%
  select(symbol, timepoint)

#Only immune DEGs
genes_roc_DEGs_all = dge_long_genes %>%
  inner_join(degs_genes_human,
             by = join_by("symbol", "timepoint"))

#Only immune DEGs
genes_roc_DEGs_immune = dge_long_genes %>%
  inner_join(degs_genes_human,
             by = join_by("symbol", "timepoint")) %>%
  #Filter immune genes only
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))
```



Inputs
```{r fig.height=4, fig.width=10}
set.seed(123)
######### Input -----
filename = "saureus_infection_human_mouse"
gene_set_filename = "All_OnlyImmune"
genes_df_input = bind_rows(
  genes_roc_allgenes %>% mutate(model_type = "Mouse"),
  genes_roc_allgenes_immunegenes %>% mutate(model_type = "Immune")
)



# gene_set_filename = "HumanDEGs_OnlyImmune"
# genes_df_input = bind_rows(
#   genes_roc_DEGs_all %>% mutate(model_type = "Mouse"),
#   genes_roc_DEGs_immune %>% mutate(model_type = "Immune")
# )

```




Calculate ROC curve
```{r fig.height=5, fig.width=10}
#=====================================================
# Compute probabilities
#=====================================================

dge_long_genes_up = genes_df_input %>% 
  mutate(
    df = n - 1,
    pval_one_sided = pt(t, df, lower.tail = FALSE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t > 0, "up", "not_up"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  select(-adj_p_val) %>%
  rename(adj_p_val = pval_one_sided_adj)

dge_long_genes_down = genes_df_input %>% 
  mutate(
    df = n - 1,
    pval_one_sided = pt(t, df, lower.tail = TRUE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t < 0, "down", "not_down"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  select(-adj_p_val) %>%
  rename(adj_p_val = pval_one_sided_adj)

#=====================================================
# Assign human truth
#=====================================================
truth_human_up = dge_long_genes_up %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

truth_human_down = dge_long_genes_down %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#=====================================================
#Generate reference distributions
#=====================================================
perfect_prediction_up = truth_human_up %>%
  mutate(prob = if_else(truth == "up", 1, 0),
         model = "Human") %>%
  rename(pred = truth)

uniform_prediction_up = perfect_prediction_up %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()), organism = "Uniform",
         model = "Uniform")

perfect_prediction_down = truth_human_down %>%
  mutate(prob = if_else(truth == "down", 1, 0),
          model = "Human") %>%
  rename(pred = truth)

uniform_prediction_down = perfect_prediction_down %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()), organism = "Uniform",
         model = "Uniform")

#=====================================================
# Combine models (Mouse + Immune + Uniform + Perfect)
#=====================================================
models_up = dge_long_genes_up %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = if_else(model_type == "Immune", "Immune", organism)) %>%
  bind_rows(perfect_prediction_up, uniform_prediction_up)

models_down = dge_long_genes_down %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = if_else(model_type == "Immune", "Immune", organism)) %>%
  bind_rows(perfect_prediction_down, uniform_prediction_down)

#=====================================================
# Classify probabilities 
#=====================================================
classified_df_probs_up = truth_human_up %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(symbol, timepoint, pred, prob, model),
             by = join_by(symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct()

classified_df_probs_down = truth_human_down %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(symbol, timepoint, pred, prob, model),
             by = join_by(symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct()


#=====================================================
#Compute ROC 
#=====================================================

roc_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Up") %>% 
  ungroup()


roc_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Down") %>% 
  ungroup()

#=====================================================
#Compute AUC 
#=====================================================
auc_results_up = classified_df_probs_up %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(model, timepoint, auc, comparison)

auc_results_down = classified_df_probs_down %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(model, timepoint, auc, comparison)

#Plot curves ------------------
#PLOT ROC Curve
roc_curve_plot = roc_curve_up_notup %>% 
  bind_rows(roc_curve_down_notdown) %>%
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot(aes(x = 1 - specificity, 
             y = sensitivity, 
             color = model)) +
  geom_path(size = 1, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_grid(~fct_rev(comparison)~timepoint) +
  scale_color_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "Human" = "#90A4AEFF",
    "Immune" = "#4361ee",
    "Uniform" = "gray80"
    )) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model") + 
  theme(
    legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3)

roc_curve_plot

# Visualize AUC
auc_plot <- bind_rows(auc_results_up, 
                      auc_results_down) %>% 
  mutate(model = factor(model, levels = c("Perfect", "Mouse", "Immune", "Uniform")),
         comparison = factor(comparison, levels = c("Up", "Down"))) %>%
  
  ggplot(aes(x = auc, 
             y = fct_reorder(timepoint, as.numeric(timepoint)) %>% fct_rev(), 
             fill = model)) +
  geom_col(width = 0.8,
           position = position_dodge()) +
  
  # Add a reference line for random prediction (AUC = 0.5)
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black", linewidth = 0.5) +
  
  geom_text(aes(label = round(auc, 2)),
            position = position_dodge2(0.9),
            hjust = 1.2, size = 2) +
  
  facet_grid(comparison ~ ., scales = "free_y") +
  
  labs(
    x = "AUC",
    y = "Day",
    fill = "Model"
  ) +
    scale_fill_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "Immune" = "#4361ee",
    "Human" = "#90A4AEFF",
    "Uniform" = "gray80"
    )) +
  theme_vaxgo()  +
  theme(
    legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(expand = c(0,0), limits = c(0,1))

roc_auc_curve_plot = ((roc_curve_plot) |
  auc_plot) +
  plot_layout(guides = "collect", widths = c(1, 0.4))+
  plot_annotation(theme = theme(legend.position = "top", plot.title = element_text(hjust = 0.5)),
                  title = filename)

roc_auc_curve_plot %>%
  ggsave(filename = here("Figures", paste0("roc_curve_plot_", filename, ".png")),
         width = 10,
         height = 5)

roc_auc_curve_plot
```


```{r fig.height=5, fig.width=10}
roc_auc_curve_plot
```


