---
title: "Exploring other datasets"
author: "Wasim Aluísio Prates-Syed"
date: "2025-07-23"
output: 
  html_document: 
    toc: TRUE
    toc_depth: 4
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE, cache = TRUE)
```


# Exploring other datasets: A general analysis

## Import required codes
```{r}
# install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```

## Found in translation
Ref: Normand, R., Du, W., Briller, M. et al. Found In Translation: a machine learning model for mouse-to-human inference. Nat Methods 15, 1067–1073 (2018). https://doi.org/10.1038/s41592-018-0214-9
Github: https://github.com/shenorrLabTRDF/FIT.mouse2man 

- Standard (ST): which include human and mouse datasets that were conducted in separate experiments;
- Reference (RF), in which the human and mouse datasets were directly contrasted in a publication authored by researchers who had generated at least one of the datasets.
- CSP IDs can be enumarated by different comparisons with different mice strains, for example A/J and CD1 mice, and timpoints (not clearly stated).
```{r}
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)

#Import files
# pca_rotations = FIT.mouse2man::pca_rotations
# best_models = FIT.mouse2man::best_models
# slopes_per_gene_V2.0 = FIT.mouse2man::slopes_per_gene_V2.0

#Gene lookup table
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  rename(human_symbol = HGNC_symbol,
         mouse_symbol =MGI.symbol,
         HS.Entrez = Human.Entrez,
         MM.Entrez = Mouse.Ortholog) %>% 
  mutate(HS.Entrez = as.character(HS.Entrez),
         MM.Entrez = as.character(MM.Entrez))

#Training data
AllData_V2.0 = FIT.mouse2man::AllData_V2.0 %>% 
  mutate(csp_id_short = str_remove(CSP_ID, "Microarrays_RF_"),
         csp_id_short = str_remove(csp_id_short, "Microarrays_ST_"),
         csp_id_short = str_remove(csp_id_short, "RNAseq_ST_"),
         csp_id_short = str_remove(csp_id_short, "RNAseq_RF_")) %>% 
  inner_join(HS_MM_Symbol_Entrez, 
             by = join_by("HS.Entrez", "MM.Entrez")) %>% 
  separate(csp_id_short, into = c("del", "csp_id_index")) %>% 
  select(-del)

```



##Import
```{r}
#Import data
fit_training_datasets_metadata_clean = read_rds(file = here("FIT_Mouse2man", "fit_training_datasets_metadata_clean.rds"))
AllData_V2.0_blood = read_rds(here("new_models", "AllData_FIT_training_blood.rds"))
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))

```

# S. aureus infection
### GSE19668 (Mouse)
```{r}
#S aureus: GSE19668 -----
filename = "GSE19668"
GSE19668 <- getGEO("GSE19668", GSEMatrix = TRUE) 
eset <- GSE19668[[1]]

#Metadata ------
GSE19668_metadata = eset %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    subject_id = "title",
    # sex = "gender:ch1", 
    treatment = "infection:ch1", 
    # tissue = "tissue:ch1",
    timepoint = "time point:ch1",
    # animal_replicate = "animal replicate:ch1",
    strain = "strain:ch1") %>% 
  mutate(treatment = fct_relevel(treatment, "none"),
         timepoint = fct_relevel(timepoint, "0"),
         condition = paste(treatment, strain, timepoint, sep = "_"))

GSE19668_metadata %>% 
  saveRDS(file = here("tables", "GSE19668_metadata.rds"))
```


```{r}
#Import
GSE19668_metadata = 
  readRDS(file = here("tables", "GSE19668_metadata.rds"))

#Filter metadata
GSE19668_metadata_filtered = GSE19668_metadata %>%
  filter(strain == "C57BL/6J") %>% 
  separate(subject_id, into = c("tissue", "strain1", "timepoint1", "participant_id"), sep = " ") %>% 
  select(!c(tissue:timepoint1)) %>% 
  arrange(participant_id, timepoint) %>% 
  separate(timepoint, into = c("timepoint_number", "time_unit")) %>% 
  mutate(time_unit = "hours",
         timepoint = timepoint_number %>% as.numeric()) %>% 
  arrange(timepoint) %>% 
  select(-timepoint_number) %>% 
  select(geo_sample, participant_id, condition, timepoint, time_unit, everything())

GSE19668_metadata_filtered %>% 
  saveRDS(file = here("tables", "GSE19668_metadata_filtered.rds"))

#Counts matrix
GSE19668_eset <- exprs(eset) %>% as.data.frame()

#Annotate data and calculate log2fc
GSE19668_expression_metadata  = GSE19668_eset %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = 2:51,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(GSE19668_metadata_filtered, 
             by = "geo_sample") 

#Get value from day 0 and calculate Log2FC
GSE19668_log2fc_sample = GSE19668_expression_metadata %>% 
  filter(timepoint == 0) %>% 
  select(probe_id, participant_id, day0_value_raw = value_raw) %>% 
  inner_join(GSE19668_expression_metadata,
             by = join_by("probe_id", "participant_id")) %>% 
  filter(timepoint != 0) %>% 
  mutate(log2fc = value_raw - day0_value_raw)

########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- GSE19668_log2fc_sample$probe_id %>% unique()
# listAttributes(ensembl_mart)
#Annotate
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2",
  values = gene_id,
  mart = ensembl_mart)

#Filter genes with highest median expression
GSE19668_counts_log2_summary_annotated_probes = GSE19668_log2fc_sample %>% 
  group_by(probe_id) %>% 
  summarize(mean = mean(value_raw, na.rm = TRUE),
            median = median(value_raw, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe_id = affy_mouse430_2), 
            by = "probe_id") %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>% #19.393 probes left
  select(probe_id, entrezgene_id, mgi_symbol) %>% 
  distinct() %>%
  
  #Filter probes from the original table
  inner_join(GSE19668_log2fc_sample, 
             by = "probe_id") %>% 
  
  #Annotate hgnc symbols
  inner_join(HS_MM_Symbol_Entrez %>% 
               clean_names() %>% 
               rename(mgi_symbol = mouse_symbol) %>% 
               select(mgi_symbol, human_symbol), 
             by = join_by("mgi_symbol")) 

#Filter human symbol by entrez
GSE19668_counts_log2_summary_annotated_probes = GSE19668_counts_log2_summary_annotated_probes %>% 
  inner_join(GSE19668_counts_log2_summary_annotated_probes %>%
               select(probe_id, entrezgene_id, human_symbol, value_raw) %>%
               group_by(human_symbol, entrezgene_id) %>%
               summarize(median = median(value_raw, na.rm = T)) %>% 
               ungroup() %>% 
               select(entrezgene_id) %>% 
               distinct(),
             by = "entrezgene_id")

 

#Save
GSE19668_counts_log2_summary_annotated_probes %>% 
  saveRDS(file = here("tables", "GSE19668_counts_log2_summary_annotated_probes.rds"))
```


####Import
```{r}
#Import
GSE19668_counts_log2_summary_annotated_probes = readRDS(file = here("tables", "GSE19668_counts_log2_summary_annotated_probes.rds"))

#Pivot wider
GSE19668_log2fc_sample_wide = GSE19668_counts_log2_summary_annotated_probes %>%
  select(human_symbol, participant_id, timepoint, log2fc) %>% 
  distinct() %>% 
  pivot_wider(names_from = participant_id,
              values_from = log2fc)
  
#Print
GSE19668_log2fc_sample_wide
```




#### Correlation plots, pairwise
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, GSE19668, C57BL/6J", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_alldays
#Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_alldays_SAureus_GSE19668.png"),
         height = 10,
         width = 10)
beep()



#Timepoint 2
timepoint_select = 2
mouse_correlation_ggpairs_day2 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == 2) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day2
#Save
mouse_correlation_ggpairs_day2 %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_day2_SAureus_GSE19668.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 6
timepoint_select = 6
mouse_correlation_ggpairs_day6 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day6
#Save
mouse_correlation_ggpairs_day6 %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_day6_SAureus_GSE19668.png"),
         height = 10,
         width = 10)
beep()

#Timepoint 12
timepoint_select = 12
mouse_correlation_ggpairs_day12 = ggpairs(GSE19668_log2fc_sample_wide %>% 
                                              filter(timepoint == timepoint_select) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 4:8,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = paste("Day", timepoint_select, "GSE19668, C57BL/6J"), 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "#4361ee"))

mouse_correlation_ggpairs_day12
#Save
mouse_correlation_ggpairs_day12 %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_day12_SAureus_GSE19668.png"),
         height = 10,
         width = 10)
beep()

```




#### PCA

```{r fig.height=4, fig.width=5}
#Import metadata
annotation_samples_pca = GSE19668_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = GSE19668_counts_log2_summary_annotated_probes %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, entrezgene_id, log2fc) %>% 
  filter(!entrezgene_id == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("entrezgene_id") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Mouse, S. aureus, GSE19668",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

```

#### BTMs
```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)

#Annotate genes by BTM
GSE19668_genes_btm = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  rename(mm_entrez = entrezgene_id) %>% 
  select(mgi_symbol, symbol = human_symbol, participant_id, geo_sample, timepoint, treatment, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = GSE19668_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process
```


```{r}
#Define colors
btm_annotation %>% pull(group) %>% unique()

immune_colors = c("SIGNAL TRANSDUCTION" = "#708d81",
                  "CELL CYCLE" = "#06d6a0",
                  "ECM AND MIGRATION" = "#52b788",
                  "ENERGY METABOLISM" = "#95d5b2", 
                  "INNATE RESPONSE" = "#184e77",
                  "NEUTROPHILS" = "#219ebc",
                  "NK CELLS" = "#a2d2ff",
                  "IFN"= "#4361ee",
                  "MONOCYTES" = "#4cc9f0",
                  "PLATELETS" = "#7209b7",
                  "B CELLS" = "#e5383b",
                  "T CELLS" = "#f72585")
immune_order = c("SIGNAL TRANSDUCTION",
                  "CELL CYCLE",
                  "ECM AND MIGRATION",
                  "ENERGY METABOLISM", 
                  "INNATE RESPONSE",
                  "NEUTROPHILS",
                  "NK CELLS",
                  "IFN",
                  "MONOCYTES",
                  "PLATELETS",
                  "B CELLS",
                  "T CELLS")

```

##### GGPairs
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

mouse_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("2" = "#caf0f8",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "12" = "purple")) 

mouse_correlation_ggpairs_btms_gsea_alldays

mouse_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_btms_gsea_alldays_Saureus.png"),
         height = 10,
         width = 10)


#All days, colored by group
mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor = ggpairs_data_dge_btm_mean_process %>%
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = group,
                                 fill = group),
          title = paste("Day", timepoint_selected),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors)

mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor

mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_btms_gsea_alldays_groupcolor_Saureus.png"),
         height = 10,
         width = 10)

beep()

```




##### Ggplot manual
```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(3:7) %>% 
  colnames()


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == 2) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == 2) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = "GSE19668, Mouse, SAureus infeciton, random sample")
```













### GSE33341 (Human)

#### Prepare data
```{r}
#S. aureus: GSE33341 -----
filename = "GSE33341"
GSE33341 <- getGEO("GSE33341", GSEMatrix = TRUE) 
eset <- GSE33341[[2]] #The matrix #1 the mouse annotated

#Metadata ------
GSE33341_metadata = eset %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    participant_id = "title",
    sex = "host gender:ch1",
    condition = "source_name_ch1",
    pathogen = "pathogen:ch1",
    description = "source_name_ch1", 
    ethnicity = "ethnicity:ch1",
    tissue_infection = "anatomic site of infection:ch1",
    batch_experimental = "experimental batch:ch1",
    age = "host age (yr):ch1") %>% 
  mutate(condition = pathogen,
         organism = "Human",
         timepoint = 1,
         time_unit = "Day",
         tissue = "Blood",
         participant_id = str_extract(participant_id, "\\[.*?\\]") %>% 
           str_remove("\\[") %>% 
           str_remove("\\]"),
         condition = case_when(description == "blood patient with E. coli bacteremia prior antibiotic treatment (Emergency Department)" ~ "E. coli prior antibiotic",
                               description == "blood patient with S. aureus bacteremia prior antibiotic treatment (Emergency Department)" ~ "S aureus prior antibiotic",
                               description == "blood human NOT infected" ~ "No infection",
                               T ~ description),
         pathogen = if_else(pathogen == "-", "Healthy", pathogen))
#Save
GSE33341_metadata %>% 
  saveRDS(file = here("tables", "GSE33341_metadata.rds"))

#Filter metadata
GSE33341_metadata_filtered = GSE33341_metadata %>%
  filter(pathogen %in% c("Staphylococcus aureus", "Healthy")) %>% 
  select(geo_sample, participant_id, condition, timepoint, everything())

#Counts matrix
GSE33341_eset <- exprs(eset) %>% as.data.frame()

#Annotate data and calculate log2fc
GSE33341_expression_metadata  = GSE33341_eset %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = -probe_id,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(GSE33341_metadata_filtered, 
             by = "geo_sample") 

#Get value from healthy and calculate Log2FC
GSE33341_log2fc_sample = GSE33341_expression_metadata %>% 
  filter(pathogen == "Healthy") %>% 
  select(probe_id, value_raw_ref = value_raw) %>% 
  group_by(probe_id) %>% 
  summarize(value_raw_ref = mean(value_raw_ref)) %>% 
  inner_join(GSE33341_expression_metadata %>% 
               filter(pathogen != "Healthy"),
             by = join_by("probe_id")) %>% 
  filter(condition != "No infection") %>% 
  mutate(log2fc = value_raw - value_raw_ref)

########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
# listDatasets(ensembl_mart)
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- GSE33341_log2fc_sample$probe_id %>% unique()
# listAttributes(ensembl_mart) %>% View()
#Annotate
annot <- getBM(
  attributes = c("affy_hg_u133a_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133a_2",
  values = gene_id,
  mart = ensembl_mart)

#Filter genes with highest median expression
GSE33341_counts_log2_summary_annotated_probes = GSE33341_log2fc_sample %>% 
  group_by(probe_id) %>% 
  summarize(mean = mean(value_raw, na.rm = TRUE),
            median = median(value_raw, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe_id = affy_hg_u133a_2), 
            by = "probe_id") %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>% 
  select(probe_id, entrezgene_id, hgnc_symbol) %>% 
  distinct() %>%  ##41.398 -> 19.394 probes left
  
  #Filter probes from the original table
  inner_join(GSE33341_log2fc_sample, 
             by = "probe_id")
#Save
GSE33341_counts_log2_summary_annotated_probes %>% 
  saveRDS(file = here("tables", "GSE33341_counts_log2_summary_annotated_probes.rds"))
  
```


#### Import
```{r}
#Import
GSE33341_counts_log2_summary_annotated_probes = 
  readRDS(file = here("tables", "GSE33341_counts_log2_summary_annotated_probes.rds"))

#Pivot wider
GSE33341_log2fc_sample_wide = GSE33341_counts_log2_summary_annotated_probes %>%
  select(hgnc_symbol, participant_id, timepoint, log2fc) %>% 
  pivot_wider(names_from = participant_id,
              values_from = log2fc) %>% 
  ungroup()
```

#### Correlation plots, pairwise
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

# Plot correlations for human -------
participant_id_selected = GSE33341_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  select(participant_id) %>% 
  distinct() %>% 
  slice_sample(n = 5) %>% 
  pull()

#All timepoints
human_correlation_ggpairs_alldays = ggpairs(GSE33341_log2fc_sample_wide %>% 
                                              select(hgnc_symbol, timepoint, participant_id_selected) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, GSE33341, Human", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#219ebc")) +
  scale_fill_manual(values = c("1" = "#219ebc"))

human_correlation_ggpairs_alldays
#Save
human_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "human_correlation_ggpairs_alldays_SAureus_GSE33341.png"),
         height = 10,
         width = 10)
beep()


```


#### BTMs
```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)

#Annotate genes by BTM
GSE33341_genes_btm = GSE33341_counts_log2_summary_annotated_probes %>% 
  rename(hs_entrez = entrezgene_id,
         human_symbol = hgnc_symbol) %>% 
  inner_join(HS_MM_Symbol_Entrez %>% 
               clean_names() %>% 
               rename(mgi_symbol = mouse_symbol) %>% 
               select(mgi_symbol, human_symbol), 
             by = join_by("human_symbol")) %>% 
  select(mgi_symbol, symbol = human_symbol, participant_id, geo_sample, timepoint, pathogen, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = GSE33341_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process")

dge_btm_mean_process
```


```{r}
#Define colors
btm_annotation %>% pull(group) %>% unique()

immune_colors = c("SIGNAL TRANSDUCTION" = "#708d81",
                  "CELL CYCLE" = "#06d6a0",
                  "ECM AND MIGRATION" = "#52b788",
                  "ENERGY METABOLISM" = "#95d5b2", 
                  "INNATE RESPONSE" = "#184e77",
                  "NEUTROPHILS" = "#219ebc",
                  "NK CELLS" = "#a2d2ff",
                  "IFN"= "#4361ee",
                  "MONOCYTES" = "#4cc9f0",
                  "PLATELETS" = "#7209b7",
                  "B CELLS" = "#e5383b",
                  "T CELLS" = "#f72585")
immune_order = c("SIGNAL TRANSDUCTION",
                  "CELL CYCLE",
                  "ECM AND MIGRATION",
                  "ENERGY METABOLISM", 
                  "INNATE RESPONSE",
                  "NEUTROPHILS",
                  "NK CELLS",
                  "IFN",
                  "MONOCYTES",
                  "PLATELETS",
                  "B CELLS",
                  "T CELLS")

```




```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)
sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(3:35) %>% 
  colnames() %>% 
  sample(5)


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  select(group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      select(group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(
    r_squared = cor_value^2,
    cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2))
  )


#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = "GSE33341, Human, SAureus infeciton, random sample")

```



### Compare Human vs Mouse

```{r}
#Unite Mouse and Human datasets. Samples.
human_mouse_saureus_log2fc_samples = 
#Mouse
GSE19668_counts_log2_summary_annotated_probes %>%
  ungroup() %>% 
  select(human_symbol, log2fc, participant_id, value_raw, 
         timepoint, pathogen = treatment,
         value_raw_ref = day0_value_raw) %>% 
  mutate(organism = "Mouse") %>% 
  
  #Human
  bind_rows(GSE33341_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
  select(human_symbol = hgnc_symbol, log2fc, value_raw, value_raw_ref, 
         participant_id, timepoint, pathogen) %>% 
  mutate(organism = "Human"))

human_mouse_saureus_log2fc_samples %>% 
  saveRDS(file = here("tables", "human_mouse_saureus_log2fc_samples.rds"))
```

####Import
```{r}

#Import
human_mouse_saureus_log2fc_samples =
  readRDS(file = here("tables", "human_mouse_saureus_log2fc_samples.rds"))

GSE19668_metadata_filtered =
  readRDS(file = here("tables", "GSE19668_metadata_filtered.rds"))

 #Calculate the mean log2fc
human_mouse_saureus_log2fc_avg = human_mouse_saureus_log2fc_samples %>% 
  group_by(organism, human_symbol, timepoint) %>% 
  summarize(
            #Calculate the mean
            mean_log2fc = mean(log2fc),
            
            #Calculate pvalue, t test
            p_value = tryCatch(
      t.test(value_raw, value_raw_ref, paired = TRUE, )$p.value,
      error = function(e) NA),
    .groups = "drop") %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>% 
  filter(human_symbol != "") %>% 
  ungroup() %>% 
  filter(timepoint %in% c(1, 2)) %>% 
  mutate(timepoint = "1-2")

#Pivot wider
human_mouse_saureus_log2fc_avg_wide = human_mouse_saureus_log2fc_avg %>% 
  select(organism, human_symbol, timepoint, mean_log2fc) %>% 
  pivot_wider(names_from = "organism",
              values_from = "mean_log2fc") %>% 
  #Add padj columns
  inner_join(human_mouse_saureus_log2fc_avg %>%
               select(organism, human_symbol, timepoint, p_adj) %>%
               pivot_wider(names_from = "organism",
                           values_from = "p_adj") %>%
               rename(padj_human = Human, padj_mouse = Mouse))

human_mouse_saureus_log2fc_avg_wide

```


```{r fig.height=5, fig.width=5}

# Calculate the correlation
cor_data <- human_mouse_saureus_log2fc_avg_wide %>%
  filter(padj_human <= 0.05 | padj_mouse <= 0.05) %>%
  summarize(
    cor_value = cor(Mouse, Human, method = "pearson", use = "complete.obs"),
    label_pos_x = max(Mouse, na.rm = TRUE),
    label_pos_y = max(Human, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("r = ", round(cor_value, 2)))



ylim_range = c(-5, 5)
xlim_range = c(-5, 5)
human_mouse_saureus_log2fc_avg_wide_scatterplot = human_mouse_saureus_log2fc_avg_wide %>% 
  filter(padj_human <= 0.05 | padj_mouse <= 0.05) %>% 
  
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_saureus_log2fc_avg_wide %>%
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_saureus_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  stat_smooth(method = "lm", geom = "line", color = "#4DBBD5FF", alpha = 0.5, se = F, size = 1,
              data = human_mouse_saureus_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add correlation metrics (Pearson)
  geom_label(
    data = cor_data,
    aes(x = label_pos_x, y = label_pos_y, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 1,
    vjust = 1,
    size = 3
  ) +
  
  #Adjust the axes range
  ylim(ylim_range) +
  xlim(xlim_range) +
  
  #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "none",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title= "Correlation plot, Log2-Fold Change, GSE19668 and GSE33341 ")

human_mouse_saureus_log2fc_avg_wide_scatterplot

```


#### Plot correlation between genes (log2fc for every sample)

```{r}
#Calculate correlation ------
correlation_long_all_mouse_human_timepoints <- human_mouse_saureus_log2fc_samples %>%
  group_by(timepoint, symbol = human_symbol, participant_id) %>%
  summarise(log2fc = max(log2fc, na.rm = TRUE), .groups = "drop") %>% 
  group_by(timepoint) %>% 
  distinct() %>% 
  group_modify(~ {
    .x %>%
      pivot_wider(
        id_cols = symbol,
        names_from = participant_id,
        values_from = log2fc
      ) %>%
      select(-symbol) %>%
      cor(use = "pairwise.complete.obs", method = "pearson") %>%
      as.data.frame() %>%
      rownames_to_column("participant1") %>%
      pivot_longer(-participant1, names_to = "participant2", values_to = "correlation") %>%
      filter(participant1 != participant2)
  }) %>%
  ungroup() %>% 
  inner_join(GSE19668_metadata_filtered %>% 
               select(participant_id, participant_1_organism = organism, sex_p_1 = sex) %>% 
               distinct(),
             by = join_by("participant1" == "participant_id")) %>% 
  inner_join(GSE19668_metadata_filtered %>% 
               select(participant_id, participant_2_organism = organism, sex_p_2 = sex) %>% 
               distinct(),
             by = join_by("participant2" == "participant_id"))


# Remove duplicated lines only if the organism is the same
correlation_long_all_timepoints_same_organisms = correlation_long_all_mouse_human_timepoints %>%
  filter(participant_1_organism == participant_2_organism) %>%
  mutate(
    participant_min = pmin(participant1, participant2),
    participant_max = pmax(participant1, participant2)
  ) %>%
  # Remove duplicates (ex: A vs B e B vs A)
  distinct(timepoint, participant_min, participant_max, .keep_all = TRUE) %>%
  select(timepoint, participant1, participant2, correlation, participant_1_organism, participant_2_organism, sex_p_1, sex_p_2) %>%
  group_by(timepoint) %>%
  arrange(-correlation)


#Unite table with different organisms with table with deduplicated rows with the same organism
correlation_long_all_mouse_human_timepoints = correlation_long_all_mouse_human_timepoints %>% 
  filter(participant_1_organism != participant_2_organism) %>% 
  bind_rows(correlation_long_all_timepoints_same_organisms)

#Summarize data and save table
correlation_long_all_mouse_human_timepoints_summary = correlation_long_all_mouse_human_timepoints %>% 
  
  #Filter only female and set Mouse as reference
  filter(sex_p_1 == "Female", 
         sex_p_2 == "Female") %>% 
  mutate(sex_comparison = paste(sex_p_1, "vs.", sex_p_2),
         comparison = paste(participant_1_organism, "vs.", participant_2_organism)) %>%
  group_by(timepoint, comparison) %>% 
  arrange(timepoint, -correlation) %>% 
  mutate(correlation = round(correlation, 2),
         timepoint = paste("Day", timepoint)) %>% 
  summarize(mean = mean(correlation, na.rm = T) %>% round(2),
            median = median(correlation, na.rm = T)%>% round(2),
            min = min(correlation, na.rm = T)%>% round(2),
            max = max(correlation, na.rm = T) %>% round(2),
            sd = sd(correlation),
            n = sum(!is.na(correlation)),
            se = sd/sqrt(n),
            t_critical = qt(0.975, df = n - 1),
            ci_lower = mean - t_critical * se,
            ci_upper = mean + t_critical * se,
            .groups = "drop") %>% 
  filter(comparison != "Human vs. Mouse")

correlation_long_all_mouse_human_timepoints_summary

#Create GT table
correlation_long_all_mouse_human_timepoints_gt = correlation_long_all_mouse_human_timepoints_summary %>% 
  gt::gt() 
#Save gt
correlation_long_all_mouse_human_timepoints_gt
gtsave(correlation_long_all_mouse_human_timepoints_gt, filename = here("tables", "correlation_long_all_mouse_human_Saureus_timepoints_gt.docx"))

```


```{r fig.height=5, fig.width=5}
correlation_long_all_mouse_human_timepoints_summary_plot = correlation_long_all_mouse_human_timepoints_summary %>% 
  separate(timepoint, into = c("unit", "timepoint"), remove = T) %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1) +
  
  geom_errorbar(aes(ymin = ci_lower, 
                  ymax = ci_upper), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 1, 3, 7), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_color_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = "Correlation between same and different organisms, by subject",
       x = "",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 
  
```




#### BTMs


```{r}
#Annotate genes by BTM
dge_btm_mean_process = GSE33341_genes_btm %>%
  mutate(organism = "Human") %>% 
  bind_rows(GSE19668_genes_btm %>% 
              mutate(organism = "Mouse")) %>%
  filter(timepoint %in% c(1, 2)) %>% 
  mutate(condition = paste0(organism, "_Day", timepoint)) %>% 
  group_by(condition, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  select(-timepoint) %>% 
  pivot_wider(names_from = "condition", values_from = "mean_log2fc") 

dge_btm_mean_process
```
##### Scatterplot
```{r fig.height=5, fig.width=7}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(4:5) %>% 
  colnames()


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  select(group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      select(group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = participant_2_value, y = participant_1_value)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  stat_smooth(aes(x = participant_2_value, y = participant_1_value),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "Mouse Log2FC", 
       y = "Human Log2FC", 
       color = "",
       title = "Mouse vs. Human")



```



##### Heatmap

```{r}
#Matrix data
matrix_data = dge_btm_mean_process %>% 
  select(process, Human_Day1, Mouse_Day2) %>% 
  drop_na() %>% 
  column_to_rownames("process") %>% 
  as.matrix()

# Column and row annotation
######## Annotations
annotation_heatmap = tribble(~condition, ~timepoint, ~organism,
        "Human_Day1", 1, "Human",
        "Mouse_Day2", 2, "Mouse")

#Arrange columns in the same order as the matrix
ann_cols_heatmap = matrix_data %>% 
  colnames() %>% 
  data.frame() %>% 
  rename(condition = '.') %>% 
  inner_join(annotation_heatmap, by = "condition") %>% 
  distinct() %>% 
  arrange(condition)

#Arrange matrix columns 
mat = matrix_data %>% 
  as.data.frame() %>% 
  select(c(ann_cols_heatmap$condition %>% unique())) %>% 
  as.matrix() 

#Column to rownames
ann_cols_heatmap = ann_cols_heatmap %>% 
  column_to_rownames("condition")

#Annotate rows
ann_rows_heatmap = mat %>% 
  rownames() %>% 
  as.data.frame() %>% 
  rename("process" = ".") %>% 
  inner_join(dge_btm_mean_process %>% 
               select(process, group, subgroup), 
             by = "process") %>% 
  column_to_rownames("process")
  

#Check dimensions
dim(mat)
dim(ann_cols_heatmap)

#Colors from required.R
btm_immune_groups
colors_immune

```


```{r fig.width= 6, fig.height=20}
#Heatmap annotation
ha = HeatmapAnnotation(df = ann_cols_heatmap, 
                       col = colors_immune,
                       gp = gpar(col = "black"), 
                       simple_anno_size = unit(0.3, "cm"),
                       annotation_name_side = "right")

row_ha = HeatmapAnnotation(df = ann_rows_heatmap,
                       col = btm_immune_groups,
                       which= "row",
                       simple_anno_size = unit(0.2, "cm"),
                       show_annotation_name = T,
                       show_legend = T)

# Values colors
col_fun <- colorRamp2(c(-2,0,2), c("#45efc6", "white", "#165289"))

file = "BTM_Modules_Saureus_Mouse_Human"

width_image = 25
height_image = 30
width = unit(1, "cm")
height = unit(25, "cm")
column_names_gp = gpar(fontsize = 8)
row_names_gp = gpar(fontsize = 8)
rect_gp = gpar(col = "black", lwd = 0.5)

fileimageheatmap_grouped = here("Figures", paste0(file, sep ="_", "Complexheatmap.png"))

png(file = fileimageheatmap_grouped, 
    width = width_image, 
    height = height_image, 
    units="cm", 
    res=900)

heatmap_plot = Heatmap(mat,
                        top_annotation = ha, 
                       left_annotation = row_ha,
                        show_row_names = TRUE,
                        row_names_side = "right",
                        show_heatmap_legend = TRUE,
                        heatmap_legend_param = list(direction = "horizontal"),
                        col = col_fun,
                        show_column_names = TRUE,
                        column_title = file,
                        column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                        column_names_rot = 45,
                        column_names_gp = column_names_gp,
                        rect_gp = rect_gp,
                        na_col = "black",
                        name = "Mean Log2FC",
                        cluster_rows = TRUE,
                        row_names_gp = row_names_gp, row_km = 2,
                        row_dend_width = unit(3, "cm"),
                        row_split = NULL,
                        column_split = NULL,
                        row_gap = unit(2, "mm"),
                        cluster_columns = TRUE,
                        column_gap = unit(8, "mm"),
                        show_column_dend = F,
                        show_row_dend = T, 
                        row_dend_side = "left",
                        column_dend_side = "top",
                        clustering_distance_rows = "euclidean",
                        border = TRUE,
                        width = width,
                        height = height)

heatmap_plot = draw(heatmap_plot, 
     annotation_legend_side = "left", 
     heatmap_legend_side = "left")

heatmap_plot

dev.off()


```











# Compare with FIT prediction

## Run FIT on a training dataset
```{r}
#Get log2fc mean
GSE19668_input2FIT = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
    filter(timepoint == 2) %>% 
  select(participant_id, MM.Entrez = entrezgene_id, EffectSize = log2fc) %>% 
  group_by(MM.Entrez) %>% 
  summarize(mean_log2fc = mean(EffectSize))

#Save
GSE19668_input2FIT %>% 
  write_csv(file = here("FIT_Mouse2man", "GSE19668_day2_FIT_mouse.csv"))

#Run FIT
GSE19668_input2FIT_day2 = FIT.mouse2man::FIT(MouseFile = here("FIT_Mouse2man", "GSE19668_day2_FIT_mouse.csv"), DataType = "rnaseq")
```



### Unite log2FC values from mouse, human, and FIT
```{r fig.height=5, fig.width=8}
#Bind rows
mouse_human_fit_log2fc = bind_rows(GSE19668_input2FIT_day2 %>%
                                     #Get DEGs for FIT
                                     filter((CI_low != 0 | CI_high != 0) & !is.na(CI_low) & !is.na(CI_high)) %>%  
                                   select(human_symbol = Human.symbol,
                                          log2fc = FIT_prediction) %>%
                                     mutate(organism = "FIT"),
                                   
                                   #Get DEGs for Human
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_human <= 0.05) %>% 
                                     select(human_symbol, log2fc = Human) %>%
                                     mutate(organism = "Human"),
                                   
                                   #Get DEGs for Mouse
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_mouse <= 0.05) %>% 
                                     select(human_symbol, log2fc = Mouse) %>%
                                     mutate(organism = "Mouse"))


#Pivot wider
mouse_human_fit_log2fc_wide = GSE19668_input2FIT_day2 %>%
            select(human_symbol = Human.symbol,
                   FIT_prediction) %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Human),
            by = "human_symbol") %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Mouse),
            by = "human_symbol")

#Save
mouse_human_fit_log2fc %>% 
  saveRDS(file = here("tables", "mouse_human_fit_log2fc_long_SAureus.rds"))

mouse_human_fit_log2fc_wide %>% 
  saveRDS(file = here("tables", "mouse_human_fit_log2fc_wide_SAureus.rds"))
```

## Distribution

### Import tables
```{r fig.height=5, fig.width=8}
#Import
mouse_human_fit_log2fc = readRDS(file = here("tables", "mouse_human_fit_log2fc_long_SAureus.rds"))
mouse_human_fit_log2fc_wide = readRDS(file = here("tables", "mouse_human_fit_log2fc_wide_SAureus.rds"))

GSE33341_metadata = readRDS(file = here("tables", "GSE33341_metadata.rds"))
GSE19668_metadata = readRDS(file = here("tables", "GSE19668_metadata.rds"))
```

### Plot histogram

```{r fig.height=5, fig.width=8}
#Plot histogram
mouse_human_fit_log2fc_histogram = mouse_human_fit_log2fc %>% 
  ggplot() +
  aes(x = log2fc,
      fill = organism) +
  geom_density(aes(y = after_stat(count),
                   color = organism),
                  position = position_identity(), 
               alpha = 0.1,
               size = 1,
              show.legend = T) +
  scale_fill_manual(values = c(FIT = "#4361ee",
                               Mouse = "#4DBBD5FF",
                               Human = "#546E7AFF"))+
  scale_color_manual(values = c(FIT = "#4361ee",
                                Mouse = "#4DBBD5FF",
                                Human = "#546E7AFF"))+
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        legend.key = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0),
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(limits = c(-4, 4)) +
  labs(color = "",
       fill = "")

mouse_human_fit_log2fc_histogram

#Faceted
mouse_human_fit_log2fc_histogram_faceted = mouse_human_fit_log2fc_histogram +
  facet_wrap(~organism,
             scales = "free_y")


#Plot scatter plot
mouse_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = Mouse,
      y = Human) +
#All DEGs
geom_point(color = "#4DBBD5FF",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "Mouse L2FC",
         y = "Human L2FC")

mouse_human_scatterplot


fit_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = FIT_prediction,
      y = Human) +
#All DEGs
geom_point(color = "#4361ee",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "FIT L2FC",
         y = "Human L2FC")

#Histograms
(((mouse_human_fit_log2fc_histogram + mouse_human_fit_log2fc_histogram_faceted) + 
    plot_layout(widths = c(0.5, 1), guides = "collect") +
    plot_annotation(theme = theme(legend.position = "top")))  /
    #Scatter plots
    (mouse_human_scatterplot + fit_human_scatterplot)) +
  plot_layout(heights = c(0.5, 1))

```



## BTMs
Import
```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)
```

```{r}
#Annotate genes by BTM
dge_btm_mean_process = mouse_human_fit_log2fc %>%
  filter(organism != "Mouse") %>% 
  rename(symbol = human_symbol) %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  group_by(organism, process) %>% 
  summarize(mean_log2fc = mean(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism", values_from = "mean_log2fc") 

dge_btm_mean_process
```
##### Scatterplot
```{r fig.height=5, fig.width=7}

#Calculate correlation
cor_labels <- dge_btm_mean_process %>%
  summarise(cor_value = cor(Human, FIT, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
dge_btm_mean_process %>% 
ggplot(aes(x = FIT, y = Human)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  stat_smooth(aes(x = FIT, y = Human),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "FIT Log2FC", 
       y = "Human Log2FC", 
       color = "",
       title = "Mouse vs. Human")



```

