---
title: "Seok_datasets"
output: html_document
---

# Required

```{r}
# install.packages("here") if needed
library(here)
source(here("scripts_notebooks", "required.R"))
```


# 1. Download datasets

## Human

### Trauma, Human (GSE36809)
Clinical metadata were curated and standardized to ensure consistency across the dataset. The time elapsed since injury, originally recorded in hours, was converted to days and rounded to one decimal place. To account for the irregular temporal distribution of samples, timepoints were discretized into seven categorical bins: <1 day (Early Hours), 1–2 days, 2–5 days, 5–10 days, 10–18 days, 18–25 days, and >25 days. Samples were stratified into 'Trauma' and 'Control' groups based on subject identifiers. Rigorous quality control was applied by excluding samples explicitly flagged in the metadata for low RNA quality or incomplete temporal data. Finally, temporal bins were treated as ordered factors to facilitate downstream differential expression modeling.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Download datasets
filename = "trauma_human"
GSE_matrix <- getGEO(filename = here("tables", "GSE36809_series_matrix.txt.gz"), 
                     GSEMatrix = T)

#Get Expression Set
eset <- GSE_matrix

#Get metadata -------
GSE_metadata = eset %>% Biobase::pData()

#Select, clean, and standardize attributes
GSE_metadata_annotated = GSE_metadata %>% 
  select(geo_sample = "geo_accession",
         sample_id = "source_name_ch1",
         description = title,
         tissue = `tissue:ch1`,
         sex = `Sex:ch1`,
         timepoint = `hours_since_injury:ch1`,
         platform_id
         ) %>% 
  separate(sample_id, sep = ", ", into = c("subject_id", "delete")) %>% 
  select(-delete) %>% 
    mutate(timepoint = if_else(timepoint == "N/A", "0", timepoint),
         timepoint = (as.numeric(timepoint)/24) %>% round(x = ., digits = 1),
         subject_id = str_remove(subject_id, "Subject "),
         geo = "GSE36809",
         organism = "Human",
         treatment = if_else(subject_id != "Control", "Trauma", "Control")) %>% 
  filter(!description %in% c("Please notice that this sample with low RNA quality.", 
                            "Sample from patients with incomplete time points.")) %>% 
  mutate(
    
    # Binned timepoints
    timepoints_new = case_when(
      timepoint < 1            ~ "Early_Hours",
      timepoint >= 1 & timepoint < 2  ~ "Day_1",
      timepoint >= 2 & timepoint < 5  ~ "Day_3",
      timepoint >= 5 & timepoint < 10 ~ "Day_7",
      timepoint >= 10 & timepoint < 18 ~ "Day_14",
      timepoint >= 18 & timepoint < 25 ~ "Day_21",
      timepoint >= 25          ~ "Day_28",
      TRUE                         ~ "Other"
    ),
    
    #Explicit bins
    timepoints_bin = case_when(
      timepoint < 1            ~ "<Day_1",
      timepoint >= 1 & timepoint < 2  ~ "Day_1_>=1 to <2",
      timepoint >= 2 & timepoint < 5  ~ "Day_3_>=2 to <5",
      timepoint >= 5 & timepoint < 10 ~ "Day_7_>=5 to <10",
      timepoint >= 10 & timepoint < 18 ~ "Day_14_>=10 to <18",
      timepoint >= 18 & timepoint < 25 ~ "Day_21_>=18 to <25",
      timepoint >= 25          ~ "Day_28_>25",
      TRUE                         ~ "Other"
    ),
    
    timepoints_new = factor(timepoints_new, levels = c("Early_Hours", "Day_1", "Day_3", "Day_7", "Day_14", "Day_21", "Day_28"))
  )



GSE_metadata_annotated %>% 
  group_by(subject_id)%>%
  arrange(subject_id, timepoint) %>%
  View()

GSE_metadata_annotated %>% 
  filter(subject_id != "Control") %>% 
  group_by(subject_id)%>%
  summarise(n_timepoints = n()) %>% 
  arrange(subject_id, n_timepoints) %>% 
  filter(n_timepoints > 1) # -10 participants


GSE_metadata_annotated_filtered %>% 
  group_by(timepoints_new, treatment)%>%
  summarise(n_sample = n()) %>% 
  arrange(timepoints_new) %>% 
  ggplot() +
  aes(x = timepoints_new,
      size = n_sample,
      y = treatment,
      color = treatment) +
  geom_point()

GSE_metadata_annotated_filtered = GSE_metadata_annotated %>% 
  filter(timepoints_new %in% c("Early_Hours", "Day_1", "Day_3", "Day_7"))


#Save metadata
write_csv(GSE_metadata_annotated_filtered, file = here("tables", paste0(filename, "_metadata.csv")))

#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% as.data.frame() %>% 
  select(GSE_metadata_annotated_filtered$geo_sample) %>% 
  saveRDS(here("tables", paste0(filename, "_exprs.rds")))

```












### Burn, Human (GSE37069)

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Download datasets
filename = "burn_human"
GSE_matrix <- getGEO(filename = here("tables", "GSE37069_series_matrix.txt.gz"), 
                     GSEMatrix = T)

#Get Expression Set
eset <- GSE_matrix

#Get metadata -------
GSE_metadata = eset %>% Biobase::pData()

#Select, clean, and standardize attributes
GSE_metadata_annotated = GSE_metadata %>% 
  select(geo_sample = "geo_accession",
         treatment = "source_name_ch1",
         organism = organism_ch1,
         age = `age:ch1`,
         tissue = `tissue:ch1`,
         sex = `Sex:ch1`,
         timepoint = `hours_since_injury:ch1`,
         platform_id,
         ) %>% 
  separate(treatment, sep = ", ", into = c("subject_id", "delete")) %>% 
  select(-delete) %>% 
    mutate(subject_id = str_remove(subject_id, "Subject "),
           timepoint = if_else(timepoint == "N/A", "0", timepoint),
         timepoint = (as.numeric(timepoint)/24) %>% round(x = ., digits = 1),
         geo = "GSE37069",
         treatment = if_else(subject_id != "Control", "Burn", "Control")) %>% 
  mutate(
    
    # Binned timepoints
    timepoints_new = case_when(
      timepoint < 1            ~ "Early_Hours",
      timepoint >= 1 & timepoint < 2  ~ "Day_1",
      timepoint >= 2 & timepoint < 5  ~ "Day_3",
      timepoint >= 5 & timepoint < 10 ~ "Day_7",
      timepoint >= 10 & timepoint < 18 ~ "Day_14",
      timepoint >= 18 & timepoint < 25 ~ "Day_21",
      timepoint >= 25          ~ "Day_28",
      TRUE                         ~ "Other"
    ),
    
    #Explicit bins
    timepoints_bin = case_when(
      timepoint < 1            ~ "<Day_1",
      timepoint >= 1 & timepoint < 2  ~ "Day_1_>=1 to <2",
      timepoint >= 2 & timepoint < 5  ~ "Day_3_>=2 to <5",
      timepoint >= 5 & timepoint < 10 ~ "Day_7_>=5 to <10",
      timepoint >= 10 & timepoint < 18 ~ "Day_14_>=10 to <18",
      timepoint >= 18 & timepoint < 25 ~ "Day_21_>=18 to <25",
      timepoint >= 25          ~ "Day_28_>25",
      TRUE                         ~ "Other"
    ),
    
    timepoints_new = factor(timepoints_new, levels = c("Early_Hours", "Day_1", "Day_3", "Day_7", "Day_14", "Day_21", "Day_28"))
  )


GSE_metadata_annotated %>% 
  group_by(timepoints_new, treatment)%>%
  summarise(n_sample = n()) %>% 
  arrange(timepoints_new) %>% 
  ggplot() +
  aes(x = timepoints_new,
      size = n_sample,
      y = treatment,
      color = treatment) +
  geom_point()

GSE_metadata_annotated %>%
  group_by(timepoint, treatment)%>%
  summarise(n_sample = n()) %>%
  arrange(timepoint) %>%
  ggplot() +
  aes(x = timepoint,
      color = treatment) +
  geom_histogram()

GSE_metadata_annotated_filtered = GSE_metadata_annotated %>% 
  filter(timepoints_new %in% c("Early_Hours", "Day_1", "Day_3", "Day_7"))
  

#Save metadata
write_csv(GSE_metadata_annotated_filtered, file = here("tables", paste0(filename, "_metadata.csv")))

#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% 
  as.data.frame() %>% 
  select(GSE_metadata_annotated_filtered$geo_sample) %>% 
  saveRDS(here("tables", paste0(filename, "_exprs.rds")))

```

## Mouse
### Trauma and Burn, Mouse (GSE7404)

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Download datasets
filename = "trauma_burn_mouse"
GSE_matrix <- getGEO(filename = here("tables", "GSE7404_series_matrix.txt.gz"), 
                     GSEMatrix = T)

#Get Expression Set
eset <- GSE_matrix

#Get metadata -------
GSE_metadata = eset %>% Biobase::pData()

#Select, clean, and standardize attributes
GSE_metadata_annotated = GSE_metadata %>% 
  select(geo_sample = "geo_accession",
         sample_id = "source_name_ch1",
         participant_description = title,
         characteristics = characteristics_ch1,
         timepoint = `time:ch1`,
         genetic = `genetic background:ch1`,
         tissue = `source_name_ch1`,
         organism = organism_ch1,
         description = description,
         # sex = `Sex:ch1`,
         platform_id
         ) %>% 
  filter(tissue == "mouse leukocytes") %>% 
  fill(genetic, .direction = "up") %>% 
  mutate(treatment = case_when(str_detect(description, "burn") ~ "Burn", 
                               str_detect(description, "trauma") ~ "Trauma", 
                               str_detect(description, "LPS") ~ "LPS",
                               T~description),
         treatment = if_else(str_detect(participant_description, "Sham"), paste0(treatment, "_", "Control"), treatment),
         tissue = str_remove(tissue, "mouse "), 
         sex = "male",
         geo = "GSE7404") %>% 
  extract(
    col = participant_description, 
    into = c("timepoint", "participant_id"), 
    regex = "(\\d+\\s?(?:day|Day|hr))\\s+(rep\\s?\\d+)", 
    remove = T 
  ) %>% 
  mutate(timepoint = str_replace(timepoint, "hr", "Hour") %>% str_replace("day", "Day"),
         timepoint = fct_reorder(timepoint, parse_number(timepoint))) %>% 
  select(geo_sample, participant_id, timepoint, sex, 
         strain = genetic, tissue, organism, treatment, platform_id) 


GSE_metadata_annotated %>% 
  group_by(timepoint, treatment)%>%
  summarise(n_sample = n()) %>% 
  arrange(timepoint) %>% 
  ggplot() +
  aes(x = timepoint,
      size = n_sample,
      y = treatment,
      color = treatment) +
  geom_point()
  

#Save metadata
write_csv(GSE_metadata_annotated, file = here("tables", paste0(filename, "_metadata.csv")))

#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% as.data.frame() %>% 
  saveRDS(here("tables", paste0(filename, "_exprs.rds")))

```


## Annotate genes

### Mouse datasets
#### Trauman and Burn

```{r}
#Import datasets
mouse_exprs = readRDS(here("tables", "trauma_burn_mouse_exprs.rds"))
mouse_metadata = read_csv(file = here("tables", "trauma_burn_mouse_metadata.csv")) %>% 
  mutate(status = if_else(str_detect(treatment, "_Control"), "Control", "Treated"),
         treatment = str_remove(treatment, "_Control"))

#Inputs
filename = "trauma_burn_mouse"

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
mouse_exprs_metadata  = mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = 2:145,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(mouse_metadata, 
             by = "geo_sample") 

#Get value from day 0 and calculate Log2FC (non-paired)
mouse_exprs_log2fc_sample = mouse_exprs_metadata %>% 
  filter(status == "Control") %>% 
  select(probe_id, treatment, timepoint, subject_num = participant_id, value_raw_ref = value_raw) %>% 
  distinct() %>% 
  group_by(timepoint, treatment, probe_id) %>% 
  summarise(value_raw_ref = mean(value_raw_ref)) %>% 
  ungroup() %>% 
  inner_join(mouse_exprs_metadata %>% 
               filter(status != "Control"),
             by = join_by("probe_id", "treatment", "timepoint")) %>% 
  mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

#Check duplicates
mouse_exprs_log2fc_sample %>% 
  filter(participant_id == "rep 1",
         treatment == "Burn",
         probe_id == "1415670_at")


########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- rownames(mouse_exprs) %>% unique()

# listAttributes(ensembl_mart) %>% View()
#Annotate
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2", #Probes
  values = gene_id,
  mart = ensembl_mart)

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_log2summary = mouse_exprs %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarise(median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_mouse430_2), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = mouse_log2summary %>% 
  #filter out ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

mouse_log2summary = mouse_log2summary %>% 
  select(probe, hgnc_symbol) %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
mouse_log2_filtered_probes = mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  as.data.frame() %>% 
  rename(probe = probe_id) %>% 
  inner_join(mouse_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol,
         probe,
         c(mouse_metadata %>% 
             pull(geo_sample)))
#Check
mouse_log2_filtered_probes %>% dim()
mouse_log2_hgnc_symbol = mouse_log2_filtered_probes 
mouse_log2_hgnc_symbol %>% get_dupes(human_symbol)

#Save
mouse_log2_hgnc_symbol %>% saveRDS(file = here("tables", paste0(filename, "_exprs_hgnc_symbol.rds")))

# #Pivot to long and wide format, and annotate
# mouse_log2_samples_log2fc_long = mouse_log2_filtered_probes %>% 
#   pivot_longer(cols = -c(human_symbol, probe),
#                names_to = "geo_sample",
#                values_to = "log2fc") %>% 
#   inner_join(mouse_metadata,
#              by = "geo_sample")
# 
# 
# mouse_log2_samples_log2fc_wide = mouse_log2_samples_log2fc_long %>% 
#   select(human_symbol, probe, subject_id = participant_id, timepoint, log2fc) %>%
#   pivot_wider(names_from = "subject_id",
#               values_from = "log2fc") %>% 
#   mutate(timepoint = fct_reorder(timepoint, timepoint %>% as.numeric()))

```




###Human datasets
#### Trauma

```{r}
#Import datasets
human_exprs = readRDS(here("tables", "trauma_human_exprs.rds"))
human_metadata = read_csv(file = here("tables", "trauma_human_metadata.csv"))

#Inputs
filename = "trauma_human"

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
human_exprs_metadata  = human_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = 2:681,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(human_metadata, 
             by = "geo_sample")
```


```{r}
#Get value from day 0 and calculate Log2FC (non-paired)
human_exprs_log2fc_sample = human_exprs_metadata %>% 
  filter(treatment == "Control") %>% 
  select(probe_id, treatment, value_raw_ref = value_raw) %>% 
  distinct() %>% 
  group_by(treatment, probe_id) %>% 
  summarise(value_raw_ref = mean(value_raw_ref)) %>% 
  ungroup() %>% 
  inner_join(human_exprs_metadata %>% 
               filter(treatment != "Control"),
             by = join_by("probe_id", "treatment")) %>% 
  mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

#Check duplicates
human_exprs_log2fc_sample %>% 
  filter(participant_id == "rep 1",
         treatment == "Burn",
         probe_id == "1415670_at")
```


```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- rownames(human_exprs) %>% unique()

# listAttributes(ensembl_mart) %>% View()
#Annotate
annot <- getBM(
  attributes = c("affy_human430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_human430_2", #Probes
  values = gene_id,
  mart = ensembl_mart)

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.human2man')
library(FIT.human2man)
HS_MM_Symbol_Entrez = FIT.human2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(human_entrez = human_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
human_log2summary = human_exprs %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarise(median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_human430_2), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(human_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("human_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = human_log2summary %>% 
  #filter out ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

human_log2summary = human_log2summary %>% 
  select(probe, hgnc_symbol) %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
human_log2_filtered_probes = human_exprs %>% 
  rownames_to_column("probe_id") %>% 
  as.data.frame() %>% 
  rename(probe = probe_id) %>% 
  inner_join(human_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol,
         probe,
         c(human_metadata %>% 
             pull(geo_sample)))
#Check
human_log2_filtered_probes %>% dim()
human_log2_hgnc_symbol = human_log2_filtered_probes 
human_log2_hgnc_symbol %>% get_dupes(human_symbol)

#Save
human_log2_hgnc_symbol %>% saveRDS(file = here("tables", paste0(filename, "_exprs_hgnc_symbol.rds")))
```










