---
title: "Seok_datasets"
output: html_document
---

# Required

```{r}
# install.packages("here") if needed
library(here)
source(here("scripts_notebooks", "required.R"))
```


# 1. Download datasets

## Human

### Trauma, Human (GSE36809)
Clinical metadata were curated and standardized to ensure consistency across the dataset. The time elapsed since injury, originally recorded in hours, was converted to days and rounded to one decimal place. To account for the irregular temporal distribution of samples, timepoints were discretized into seven categorical bins: <1 day (Early Hours), 1–2 days, 2–5 days, 5–10 days, 10–18 days, 18–25 days, and >25 days. Samples were stratified into 'Trauma' and 'Control' groups based on subject identifiers. Rigorous quality control was applied by excluding samples explicitly flagged in the metadata for low RNA quality or incomplete temporal data. Finally, temporal bins were treated as ordered factors to facilitate downstream differential expression modeling.

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Download datasets
filename = "trauma_human"
GSE_matrix <- getGEO(filename = here("tables", "GSE36809_series_matrix.txt.gz"), 
                     GSEMatrix = T)

#Get Expression Set
eset <- GSE_matrix

#Get metadata -------
GSE_metadata = eset %>% Biobase::pData()

#Select, clean, and standardize attributes
GSE_metadata_annotated = GSE_metadata %>% 
  select(geo_sample = "geo_accession",
         sample_id = "source_name_ch1",
         description = title,
         tissue = `tissue:ch1`,
         sex = `Sex:ch1`,
         timepoint = `hours_since_injury:ch1`,
         platform_id
         ) %>% 
  separate(sample_id, sep = ", ", into = c("subject_id", "delete")) %>% 
  select(-delete) %>% 
    mutate(timepoint = if_else(timepoint == "N/A", "0", timepoint),
         timepoint = (as.numeric(timepoint)/24) %>% round(x = ., digits = 1),
         subject_id = str_remove(subject_id, "Subject "),
         geo = "GSE36809",
         organism = "Human",
         treatment = if_else(subject_id != "Control", "Trauma", "Control")) %>% 
  filter(!description %in% c("Please notice that this sample with low RNA quality.", 
                            "Sample from patients with incomplete time points.")) %>% 
  mutate(
    
    # Binned timepoints
    timepoints_new = case_when(
      timepoint < 1            ~ "Early_Hours",
      timepoint >= 1 & timepoint < 2  ~ "Day_1",
      timepoint >= 2 & timepoint < 5  ~ "Day_3",
      timepoint >= 5 & timepoint < 10 ~ "Day_7",
      timepoint >= 10 & timepoint < 18 ~ "Day_14",
      timepoint >= 18 & timepoint < 25 ~ "Day_21",
      timepoint >= 25          ~ "Day_28",
      TRUE                         ~ "Other"
    ),
    
    #Explicit bins
    timepoints_bin = case_when(
      timepoint < 1            ~ "<Day_1",
      timepoint >= 1 & timepoint < 2  ~ "Day_1_>=1 to <2",
      timepoint >= 2 & timepoint < 5  ~ "Day_3_>=2 to <5",
      timepoint >= 5 & timepoint < 10 ~ "Day_7_>=5 to <10",
      timepoint >= 10 & timepoint < 18 ~ "Day_14_>=10 to <18",
      timepoint >= 18 & timepoint < 25 ~ "Day_21_>=18 to <25",
      timepoint >= 25          ~ "Day_28_>25",
      TRUE                         ~ "Other"
    ),
    
    timepoints_new = factor(timepoints_new, levels = c("Early_Hours", "Day_1", "Day_3", "Day_7", "Day_14", "Day_21", "Day_28"))
  )



GSE_metadata_annotated %>% 
  group_by(subject_id)%>%
  arrange(subject_id, timepoint) %>%
  View()

GSE_metadata_annotated %>% 
  filter(subject_id != "Control") %>% 
  group_by(subject_id)%>%
  summarise(n_timepoints = n()) %>% 
  arrange(subject_id, n_timepoints) %>% 
  filter(n_timepoints > 1) # -10 participants


GSE_metadata_annotated_filtered %>% 
  group_by(timepoints_new, treatment)%>%
  summarise(n_sample = n()) %>% 
  arrange(timepoints_new) %>% 
  ggplot() +
  aes(x = timepoints_new,
      size = n_sample,
      y = treatment,
      color = treatment) +
  geom_point()

GSE_metadata_annotated_filtered = GSE_metadata_annotated %>% 
  filter(timepoints_new %in% c("Early_Hours", "Day_1", "Day_3", "Day_7"))


#Save metadata
write_csv(GSE_metadata_annotated_filtered, file = here("tables", paste0(filename, "_metadata.csv")))

#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% as.data.frame() %>% 
  select(GSE_metadata_annotated_filtered$geo_sample) %>% 
  saveRDS(here("tables", paste0(filename, "_exprs.rds")))

```












### Burn, Human (GSE37069)

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Download datasets
filename = "burn_human"
GSE_matrix <- getGEO(filename = here("tables", "GSE37069_series_matrix.txt.gz"), 
                     GSEMatrix = T)

#Get Expression Set
eset <- GSE_matrix

#Get metadata -------
GSE_metadata = eset %>% Biobase::pData()

#Select, clean, and standardize attributes
GSE_metadata_annotated = GSE_metadata %>% 
  select(geo_sample = "geo_accession",
         treatment = "source_name_ch1",
         organism = organism_ch1,
         age = `age:ch1`,
         tissue = `tissue:ch1`,
         sex = `Sex:ch1`,
         timepoint = `hours_since_injury:ch1`,
         platform_id,
         ) %>% 
  separate(treatment, sep = ", ", into = c("subject_id", "delete")) %>% 
  select(-delete) %>% 
    mutate(subject_id = str_remove(subject_id, "Subject "),
           timepoint = if_else(timepoint == "N/A", "0", timepoint),
         timepoint = (as.numeric(timepoint)/24) %>% round(x = ., digits = 1),
         geo = "GSE37069",
         treatment = if_else(subject_id != "Control", "Burn", "Control")) %>% 
  mutate(
    
    # Binned timepoints
    timepoints_new = case_when(
      timepoint < 1            ~ "Early_Hours",
      timepoint >= 1 & timepoint < 2  ~ "Day_1",
      timepoint >= 2 & timepoint < 5  ~ "Day_3",
      timepoint >= 5 & timepoint < 10 ~ "Day_7",
      timepoint >= 10 & timepoint < 18 ~ "Day_14",
      timepoint >= 18 & timepoint < 25 ~ "Day_21",
      timepoint >= 25          ~ "Day_28",
      TRUE                         ~ "Other"
    ),
    
    #Explicit bins
    timepoints_bin = case_when(
      timepoint < 1            ~ "<Day_1",
      timepoint >= 1 & timepoint < 2  ~ "Day_1_>=1 to <2",
      timepoint >= 2 & timepoint < 5  ~ "Day_3_>=2 to <5",
      timepoint >= 5 & timepoint < 10 ~ "Day_7_>=5 to <10",
      timepoint >= 10 & timepoint < 18 ~ "Day_14_>=10 to <18",
      timepoint >= 18 & timepoint < 25 ~ "Day_21_>=18 to <25",
      timepoint >= 25          ~ "Day_28_>25",
      TRUE                         ~ "Other"
    ),
    
    timepoints_new = factor(timepoints_new, levels = c("Early_Hours", "Day_1", "Day_3", "Day_7", "Day_14", "Day_21", "Day_28"))
  )


GSE_metadata_annotated %>% 
  group_by(timepoints_new, treatment)%>%
  summarise(n_sample = n()) %>% 
  arrange(timepoints_new) %>% 
  ggplot() +
  aes(x = timepoints_new,
      size = n_sample,
      y = treatment,
      color = treatment) +
  geom_point()

GSE_metadata_annotated %>%
  group_by(timepoint, treatment)%>%
  summarise(n_sample = n()) %>%
  arrange(timepoint) %>%
  ggplot() +
  aes(x = timepoint,
      color = treatment) +
  geom_histogram()

GSE_metadata_annotated_filtered = GSE_metadata_annotated %>% 
  filter(timepoints_new %in% c("Early_Hours", "Day_1", "Day_3", "Day_7"))
  

#Save metadata
write_csv(GSE_metadata_annotated_filtered, file = here("tables", paste0(filename, "_metadata.csv")))

#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% 
  as.data.frame() %>% 
  select(GSE_metadata_annotated_filtered$geo_sample) %>% 
  saveRDS(here("tables", paste0(filename, "_exprs.rds")))

```

## Mouse
### Trauma and Burn, Mouse (GSE7404)

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE, cache = TRUE}
#Download datasets
filename = "trauma_burn_mouse"
GSE_matrix <- getGEO(filename = here("tables", "GSE7404_series_matrix.txt.gz"), 
                     GSEMatrix = T)

#Get Expression Set
eset <- GSE_matrix

#Get metadata -------
GSE_metadata = eset %>% Biobase::pData()

#Select, clean, and standardize attributes
GSE_metadata_annotated = GSE_metadata %>% 
  select(geo_sample = "geo_accession",
         sample_id = "source_name_ch1",
         participant_description = title,
         characteristics = characteristics_ch1,
         timepoint = `time:ch1`,
         genetic = `genetic background:ch1`,
         tissue = `source_name_ch1`,
         organism = organism_ch1,
         description = description,
         # sex = `Sex:ch1`,
         platform_id
         ) %>% 
  filter(tissue == "mouse leukocytes") %>% 
  fill(genetic, .direction = "up") %>% 
  mutate(treatment = case_when(str_detect(description, "burn") ~ "Burn", 
                               str_detect(description, "trauma") ~ "Trauma", 
                               str_detect(description, "LPS") ~ "LPS",
                               T~description),
         treatment = if_else(str_detect(participant_description, "Sham"), paste0(treatment, "_", "Control"), treatment),
         tissue = str_remove(tissue, "mouse "), 
         sex = "male",
         geo = "GSE7404") %>% 
  extract(
    col = participant_description, 
    into = c("timepoint", "participant_id"), 
    regex = "(\\d+\\s?(?:day|Day|hr))\\s+(rep\\s?\\d+)", 
    remove = T 
  ) %>% 
  mutate(timepoint = str_replace(timepoint, "hr", "Hour") %>% str_replace("day", "Day"),
         timepoint = fct_reorder(timepoint, parse_number(timepoint))) %>% 
  select(geo_sample, participant_id, timepoint, sex, 
         strain = genetic, tissue, organism, treatment, platform_id) 


GSE_metadata_annotated %>% 
  group_by(timepoint, treatment)%>%
  summarise(n_sample = n()) %>% 
  arrange(timepoint) %>% 
  ggplot() +
  aes(x = timepoint,
      size = n_sample,
      y = treatment,
      color = treatment) +
  geom_point()
  

#Save metadata
write_csv(GSE_metadata_annotated, file = here("tables", paste0(filename, "_metadata.csv")))

#Get counts matrix ------
GSE_exprs <- eset %>% exprs()

#Save as RDS
GSE_exprs %>% as.data.frame() %>% 
  saveRDS(here("tables", paste0(filename, "_exprs.rds")))

```


#2.Annotate genes

### Mouse datasets

#### Trauman and Burn

```{r}
#Import datasets
mouse_exprs = readRDS(here("tables", "trauma_burn_mouse_exprs.rds"))
mouse_metadata = read_csv(file = here("tables", "trauma_burn_mouse_metadata.csv")) %>% 
  mutate(status = if_else(str_detect(treatment, "_Control"), "Control", "Treated"),
         treatment = str_remove(treatment, "_Control"))

#Inputs
filename = "trauma_burn_mouse"

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
mouse_exprs_metadata  = mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = 2:145,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(mouse_metadata, 
             by = "geo_sample") 

#Get value from day 0 and calculate Log2FC (non-paired)
mouse_exprs_log2fc_sample = mouse_exprs_metadata %>% 
  filter(status == "Control") %>% 
  select(probe_id, treatment, timepoint, subject_num = participant_id, value_raw_ref = value_raw) %>% 
  distinct() %>% 
  group_by(timepoint, treatment, probe_id) %>% 
  summarise(value_raw_ref = mean(value_raw_ref)) %>% 
  ungroup() %>% 
  inner_join(mouse_exprs_metadata %>% 
               filter(status != "Control"),
             by = join_by("probe_id", "treatment", "timepoint")) %>% 
  mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

mouse_exprs_log2fc_sample %>% 
  saveRDS(here("tables", "trauma_burn_mouse_log2fc_sample_clean_long.rds"))

#Check duplicates
mouse_exprs_log2fc_sample %>% 
  filter(participant_id == "rep 1",
         treatment == "Burn",
         probe_id == "1415670_at")


########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- rownames(mouse_exprs) %>% unique()

# listAttributes(ensembl_mart) %>% View()
#Annotate
annot <- getBM(
  attributes = c("affy_mouse430_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "affy_mouse430_2", #Probes
  values = gene_id,
  mart = ensembl_mart)

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_log2summary = mouse_exprs %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe_id") %>% 
  rename(probe = probe_id) %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarise(median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = affy_mouse430_2), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = mouse_log2summary %>% 
  #filter out ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

mouse_log2summary = mouse_log2summary %>% 
  select(probe, hgnc_symbol) %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
mouse_log2_filtered_probes = mouse_exprs %>% 
  rownames_to_column("probe_id") %>% 
  as.data.frame() %>% 
  rename(probe = probe_id) %>% 
  inner_join(mouse_log2summary, by = "probe") %>%   #214,396
  select(human_symbol = hgnc_symbol,
         probe,
         c(mouse_metadata %>% 
             pull(geo_sample)))
#Check
mouse_log2_filtered_probes %>% dim()
mouse_log2_hgnc_symbol = mouse_log2_filtered_probes 
mouse_log2_hgnc_symbol %>% get_dupes(human_symbol)

#Save
mouse_log2_hgnc_symbol %>% saveRDS(file = here("tables", paste0(filename, "_exprs_hgnc_symbol.rds")))

```




###Human datasets
#### Trauma

```{r}
#Import datasets
human_exprs = readRDS(here("tables", "trauma_human_exprs.rds"))
human_metadata = read_csv(file = here("tables", "trauma_human_metadata.csv"))

#Inputs
filename = "trauma_human"

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
human_exprs_metadata  = human_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = 2:681,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(human_metadata, 
             by = "geo_sample")
```


```{r}
#Get value from day 0 and calculate Log2FC (non-paired)
human_exprs_log2fc_sample = human_exprs_metadata %>% 
  filter(treatment == "Control") %>% 
  select(probe_id, treatment, value_raw_ref = value_raw) %>% 
  distinct() %>% 
  group_by(probe_id) %>% 
  summarise(value_raw_ref = mean(value_raw_ref)) %>% 
  ungroup() %>% 
  inner_join(human_exprs_metadata %>% 
               filter(treatment != "Control"),
             by = join_by("probe_id")) %>% 
  mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

#Check for duplicates
human_exprs_log2fc_sample %>% 
  filter(description == "Blood, subject 15429976, hours since injury 5.8",
         probe_id == "1007_s_at")
```


```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- rownames(human_exprs) %>% unique()

# listAttributes(ensembl_mart) %>% View()
#Annotate
annot <- getBM(
  attributes = c("affy_hg_u133_plus_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133_plus_2", #Probes
  values = gene_id,
  mart = ensembl_mart) %>% #Standardize
  rename(probe = affy_hg_u133_plus_2,
                     entrezgene_id = entrezgene_id,
                     human_symbol = hgnc_symbol,
                     ensembl_id = ensembl_gene_id)

#Filter genes with highest median expression
human_log2summary = human_exprs %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarise(median = median(expression, na.rm = T)) %>%  
  inner_join(annot, 
            by = "probe") %>% 
  
  #Filter ensembl id
  group_by(ensembl_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup()  


#Filter out genes with multiple hgnc_symbols
hgnc_symbol_dupes = human_log2summary %>% 
  #filter ensembl id with multiple hgnc_symbols
  get_dupes(human_symbol)

human_log2summary = human_log2summary %>% 
  filter(!human_symbol %in% c(hgnc_symbol_dupes$human_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
human_log2_filtered_probes = human_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(human_log2summary, by = "probe") %>%   #214,396
  select(human_symbol, probe,
         c(human_metadata$geo_sample))
#14395
human_log2_filtered_probes %>% dim()

human_log2_filtered_probes = human_log2_filtered_probes 

human_log2_filtered_probes %>% 
  get_dupes(human_symbol)

#Save 
human_log2_filtered_probes %>% 
  saveRDS(here("tables", "trauma_human_exprs_hgnc_symbol.rds"))
```



#### Burn

```{r}
#Import datasets
human_exprs = readRDS(here("tables", "burn_human_exprs.rds"))
human_metadata = read_csv(file = here("tables", "burn_human_metadata.csv"))

#Inputs
filename = "burn_human"

#Annotate genes ------
#Mouse -----
#Annotate data and calculate log2fc
human_exprs_metadata  = human_exprs %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = 2:329,
               names_to = "geo_sample",
               values_to = "value_raw") %>% 
  inner_join(human_metadata, 
             by = "geo_sample")
```


```{r}
#Get value from day 0 and calculate Log2FC (non-paired)
human_exprs_log2fc_sample = human_exprs_metadata %>% 
  filter(treatment == "Control") %>% 
  select(probe_id, treatment, value_raw_ref = value_raw) %>% 
  distinct() %>% 
  group_by(probe_id) %>% 
  summarise(value_raw_ref = mean(value_raw_ref)) %>% 
  ungroup() %>% 
  inner_join(human_exprs_metadata %>% 
               filter(treatment != "Control"),
             by = join_by("probe_id")) %>% 
  mutate(log2fc = value_raw - value_raw_ref) %>% 
  distinct() 

#Check for duplicates
human_exprs_log2fc_sample %>% 
  filter(subject_id == "11710452",
         probe_id == "1007_s_at")
```


```{r}
########### Filter and annotate probes -----
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- rownames(human_exprs) %>% unique()

# listAttributes(ensembl_mart) %>% View()
#Annotate
annot <- getBM(
  attributes = c("affy_hg_u133_plus_2",
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "hgnc_symbol",
                 "description"),
  filters = "affy_hg_u133_plus_2", #Probes
  values = gene_id,
  mart = ensembl_mart) %>% #Standardize
  rename(probe = affy_hg_u133_plus_2,
                     entrezgene_id = entrezgene_id,
                     human_symbol = hgnc_symbol,
                     ensembl_id = ensembl_gene_id)

#Filter genes with highest median expression
human_log2summary = human_exprs %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarise(median = median(expression, na.rm = T)) %>%  
  inner_join(annot, 
            by = "probe") %>% 
  
  #Filter ensembl id
  group_by(ensembl_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup()  


#Filter out genes with multiple hgnc_symbols
hgnc_symbol_dupes = human_log2summary %>% 
  #filter ensembl id with multiple hgnc_symbols
  get_dupes(human_symbol)

human_log2summary = human_log2summary %>% 
  filter(!human_symbol %in% c(hgnc_symbol_dupes$human_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
human_log2_filtered_probes = human_exprs %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(human_log2summary, by = "probe") %>%   #214,396
  select(human_symbol, probe,
         c(human_metadata$geo_sample))
#18196
human_log2_filtered_probes %>% dim()

human_log2_filtered_probes = human_log2_filtered_probes 

human_log2_filtered_probes %>% 
  get_dupes(human_symbol)

#Save 
human_log2_filtered_probes %>% 
  saveRDS(here("tables", "burn_human_exprs_hgnc_symbol.rds"))
```


















#3. Differential expression

## Mouse
### Burn
###### Limma
Paired samples design
```{r}
#Import
#Import
mouse_log2_hgnc_symbol = readRDS(file = here("tables", "trauma_burn_mouse_exprs_hgnc_symbol.rds"))
mouse_metadata = read_csv(file = here("tables", "trauma_burn_mouse_metadata.csv")) %>% 
  filter(str_detect(treatment, "Burn"))

exprs(eset_mouse_log) <- log2(exprs(eset_mouse) + 1)


#Create Expression Set
eset_mouse <- ExpressionSet(
  assayData = mouse_log2_hgnc_symbol %>% 
    column_to_rownames("human_symbol") %>% 
    select(all_of(mouse_metadata$geo_sample)) %>%
    as.matrix() %>% 
    { log2(. + 1) },
  phenoData = AnnotatedDataFrame(mouse_metadata %>% 
                                   column_to_rownames("geo_sample") %>% 
                                   mutate(condition = paste0(treatment, "_", timepoint))))

#Create model matrix
design <- model.matrix(~ condition + 0, # 0 defines the way we compare the samples
                       eset_mouse) 

colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

aw <- arrayWeights(eset_mouse, design)
barplot(aw, 
        names.arg = eset_mouse$condition, # Define os nomes no eixo X
        main = "Sample Weights Diagnostic", 
        ylab = "Weight", 
        las = 2,           # Rotaciona os nomes em 90 graus
        cex.names = 0.7,   # Diminui o tamanho da fonte se houver muitos nomes
        col = "skyblue")

#Fit Limma model
fit <- lmFit(eset_mouse, 
             design, weights = aw) 

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
colnames(design) <- make.names(colnames(design))

cont_matrix <- makeContrasts(
  # --- Burn Comparisons ---
  Burn_2h = conditionBurn_2.Hour - conditionBurn_Control_2.Hour,
  Burn_D1 = conditionBurn_1.Day  - conditionBurn_Control_1.Day,
  Burn_D3 = conditionBurn_3.Day  - conditionBurn_Control_3.Day,
  Burn_D7 = conditionBurn_7.Day  - conditionBurn_Control_7.Day,
  levels = design
)

fit2 <- contrasts.fit(fit, cont_matrix)
fit2 <- eBayes(fit2, trend = T, robust = T)

# Veja se os genes com esses pesos altos estão distorcendo a curva
plotSA(fit2, main = "Sigma vs Abundance (Weights Applied)")

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_mouse_burn = all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(organism = "Mouse") %>% 
  select(condition = comparison, everything()) %>% 
  clean_names() %>% 
  rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)

all_degs_mouse_burn$condition %>% unique()
	
# #Save
all_degs_mouse_burn %>% saveRDS(file = here("tables", "burn_mouse_dge_limma_degs.rds"))

all_degs_mouse_burn %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarise(n_degs = n())

```



###### T test
```{r}
#Import
mouse_exprs = readRDS(file = here("tables", "trauma_burn_mouse_exprs_hgnc_symbol.rds"))
mouse_metadata = read_csv(file = here("tables", "trauma_burn_mouse_metadata.csv"))
log2fc_sample = readRDS(here("tables", "trauma_burn_mouse_log2fc_sample_clean_long.rds"))

log2fc_sample_symbol = log2fc_sample %>% 
  inner_join(mouse_exprs %>% 
               select(human_symbol, probe_id = probe),
             by = "probe_id") %>% 
  select(-probe_id) %>% 
  distinct() %>% 
  filter(treatment == "Burn")


```


```{r}
#Compute summary statistics
log2fc_summary = log2fc_sample_symbol %>% 
  group_by(human_symbol, timepoint) %>%
  summarise(
    avg_log2fc = mean(log2fc, na.rm = TRUE),  # mean log2 fold change
    n_subjects = n(),                         # number of samples
    sd = sd(log2fc, na.rm = TRUE),             # standard deviation
    se = sd / sqrt(n_subjects),               # standard error
    effect_size = avg_log2fc / sd)            # Cohen’s d-like effect size

log2_hgnc_symbol_wide 

log2fc_sample_symbol %>%
  ungroup() %>% 
  inner_join(mouse_metadata %>% 
               select(geo_sample, timepoint),
             by = join_by("geo_sample", "timepoint")) %>% 
  select(human_symbol, timepoint, ref = value_raw_ref, value = value_raw) %>% 
  group_by(human_symbol)
  pivot_wider(names_from = timepoint, values_from = c(ref, value))
```


```{r}
#Unpaired
log2fc_unpaired_ttest = log2_hgnc_symbol_wide %>%
  group_by(human_symbol) %>%
  
  summarise(
    pval_hour2 = tryCatch(t.test(hour_2, hour_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_hour6 = tryCatch(t.test(hour_6, hour_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_hour12 = tryCatch(t.test(hour_12, hour_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    pval_hour24 = tryCatch(t.test(hour_24, hour_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    .groups = 'drop'
  ) %>%
  
  mutate(
    padj_hour2 = p.adjust(pval_hour2, method = "fdr"),
    padj_hour6 = p.adjust(pval_hour6, method = "fdr"),
    padj_hour12 = p.adjust(pval_hour12, method = "fdr"),
    padj_hour24 = p.adjust(pval_hour24, method = "fdr"),
  ) %>% 
  pivot_longer(
    cols = starts_with(c("pval_", "padj_")),
    names_to = c(".value", "timepoint"),
    names_pattern = "(pval|padj)_(hour\\d+)"
  ) %>% 
  mutate(timepoint = str_remove(timepoint, "hour"),
         test = "t-Test",
         paired = "Unpaired") 

log2fc_paired_unpaired_ttest = log2fc_unpaired_ttest %>% 
  bind_rows(log2fc_paired_ttest) 

log2fc_paired_unpaired_ttest %>% 
  saveRDS(file = here("tables", "ecoli_infection_mouse_log2fc_paired_unpaired_ttest.rds"))

```







## Human
Create ExpressionSet

```{r}
#Import
human_ecoli_log2_filtered_symbol = readRDS(file = here("tables", "ecoli_infection_human_exprs_hgnc_symbol.rds"))

dim(human_ecoli_log2_filtered_symbol)
  
ecoli_human_metadata = readRDS(file = here("tables", "ecoli_infection_human_metadata.rds")) %>% 
  mutate(timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         age = scale(as.numeric(age)),
         ethnicity = as.factor(ethnicity),
         status = as.factor(status),
         status = fct_relevel(status, "Healthy", "infected"),
         site_infection = as.factor(site_infection)) %>% 
  dplyr::filter(!is.na(age))

dim(ecoli_human_metadata)

#Create Expression Set
eset_human_ecoli <- ExpressionSet(
  assayData = human_ecoli_log2_filtered_symbol %>% 
    column_to_rownames("human_symbol") %>% 
    dplyr::select(ecoli_human_metadata$geo_sample) %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(ecoli_human_metadata)) 

#metadata = 62 rows
#exprs data = 62 cols

#Save
eset_human_ecoli %>% saveRDS(file = here("tables", "ecoli_infection_human_eset.rds"))


```



###### Limma
Paired samples design
```{r}

#Import
human_ecoli_log2_filtered_symbol = readRDS(file = here("tables", "ecoli_infection_human_exprs_hgnc_symbol.rds"))
ecoli_human_metadata = readRDS(file = here("tables", "ecoli_infection_human_metadata.rds"))
eset_human_ecoli = readRDS(file = here("tables", "ecoli_infection_human_eset.rds"))


#Create model matrix
design <- model.matrix(~ age + sex + ethnicity + status,
                       data = pData(eset_human_ecoli))

colnames(design) <- make.names(colnames(design))  # Avoids blank spaces
rownames(design) <- rownames(pData(eset_human_ecoli))

#Fit Limma model
fit <- lmFit(eset_human_ecoli, 
             design) 

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(statusinfected, levels = design)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- limma::eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_human_ecoli= all_contrasts_degs %>% 
  dplyr::select(comparison, everything()) %>% 
  dplyr::mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  dplyr::mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  dplyr::mutate(condition = case_when(comparisonA == "statusinfected"	~ "Ecoli_24h",
                               T~comparisonA),
         organism = "Human",
        timepoint = str_remove(condition, "Ecoli_") %>% str_remove("h")) %>% 
  dplyr::select(condition, everything()) %>% 
  clean_names() %>% 
  dplyr::rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs_human_ecoli %>% saveRDS(file = here("tables", "ecoli_infection_human_dge_limma_degs.rds"))

all_degs_human_ecoli %>% 
  dplyr::group_by(condition) %>%
  dplyr::filter(adj_p_val <= 0.05) %>% 
  dplyr::summarise(n_degs = n())

```




###### T test
```{r}
#Import
human_exprs = readRDS(file = here("tables", "ecoli_infection_human_exprs_hgnc_symbol.rds"))
human_metadata = readRDS(file = here("tables", "ecoli_infection_human_metadata.rds"))
log2fc_sample = readRDS(here("tables", "ecoli_infection_human_log2fc_sample_clean_long.rds"))
log2fc_sample_symbol = log2fc_sample %>% 
  inner_join(human_exprs %>% 
               select(human_symbol, probe_id = probe),
             by = "probe_id")


#Compute summary statistics
log2fc_summary = log2fc_sample_symbol %>% 
  group_by(human_symbol) %>%
  summarise(
    avg_log2fc = mean(log2fc, na.rm = TRUE),  # mean log2 fold change
    n_subjects = n(),                         # number of samples
    sd = sd(log2fc, na.rm = TRUE),             # standard deviation
    se = sd / sqrt(n_subjects),               # standard error
    effect_size = avg_log2fc / sd)            # Cohen’s d-like effect size

log2_hgnc_symbol_wide = log2fc_sample_symbol %>%
  ungroup() %>% 
  inner_join(human_metadata %>% 
               select(geo_sample, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, subject_id, timepoint, day0_value_raw, value_raw) %>% 
  distinct() %>% 
  select(subject_id, human_symbol, timepoint, hour_0 = day0_value_raw, value = value_raw) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "hour_")

#Unpaired
log2fc_unpaired_ttest = log2_hgnc_symbol_wide %>%
  group_by(human_symbol) %>%
  
  summarise(pval_hour24 = tryCatch(t.test(hour_24, hour_0, paired = F, var.equal = FALSE, alternative = "two.sided")$p.value,
                         error = function(e) NA_real_),
    .groups = 'drop'
  ) %>%
  
  mutate(padj_hour24 = p.adjust(pval_hour24, method = "fdr")) %>% 
  pivot_longer(
    cols = starts_with(c("pval_", "padj_")),
    names_to = c(".value", "timepoint"),
    names_pattern = "(pval|padj)_(hour\\d+)"
  ) %>% 
  mutate(timepoint = str_remove(timepoint, "hour"),
         test = "t-Test",
         paired = "Unpaired") 


log2fc_unpaired_ttest %>% 
  saveRDS(file = here("tables", "ecoli_infection_human_log2fc_unpaired_ttest.rds"))

```






## Combine datasets

```{r}

#Merge DGE results
all_degs_mouse = read_rds(here("tables", "ecoli_infection_mouse_dge_limma_degs.rds"))
all_degs_human = read_rds(here("tables", "ecoli_infection_human_dge_limma_degs.rds"))
mouse_log2fc_sample = readRDS(here("tables", "ecoli_infection_mouse_log2fc_sample_clean_long.rds")) 
mouse_exprs = readRDS(file = here("tables", "ecoli_infection_mouse_exprs_hgnc_symbol_long.rds"))
human_log2fc_sample = readRDS(here("tables", "ecoli_infection_human_log2fc_sample_clean_long.rds")) 
human_exprs = readRDS(file = here("tables", "ecoli_infection_human_exprs_hgnc_symbol.rds"))

#Merge DGE results ===========
all_degs_human_mouse = all_degs_mouse %>% 
  bind_rows(all_degs_human) %>% 
  select(condition, timepoint, human_symbol, mean_l2fc:organism)

#Save
all_degs_human_mouse %>% saveRDS(here("tables", "ecoli_infection_human_mouse_dge_limma_degs.rds"))

#Merge sample log2fc =============
mouse_log2fc_sample_symbol = mouse_log2fc_sample %>% 
  inner_join(mouse_exprs %>% 
               select(human_symbol, probe_id = probe),
             by = "probe_id") %>% 
  mutate(organism = "Mouse")

human_log2fc_sample_symbol = human_log2fc_sample %>% 
  inner_join(human_exprs %>% 
               select(human_symbol, probe_id = probe),
             by = "probe_id") %>% 
  mutate(organism = "Human",
         timepoint = "24")

human_mouse_ecoli_log2fc_samples = bind_rows(mouse_log2fc_sample_symbol, human_log2fc_sample_symbol) %>% 
  ungroup()

human_mouse_ecoli_log2fc_samples %>% 
  saveRDS(here("tables", "ecoli_infection_human_mouse_samples_log2fc.rds"))


# Metadata
human_mouse_metadata_ecoli = human_mouse_ecoli_log2fc_samples_ecoli %>% 
  select(subject_id, batch, geo_sample, subject_title:bac_strain) %>% 
  distinct()

human_mouse_metadata_ecoli %>% 
  saveRDS(here("tables", "ecoli_infection_human_mouse_samples_metadata.rds"))
```

#### Compare DGE methods
```{r}
#Import ------
#t-Test
human_log2fc_paired_unpaired_ttest =
  readRDS(file = here("tables", "ecoli_infection_human_log2fc_unpaired_ttest.rds")) %>% 
  mutate(organism = "Human")
mouse_log2fc_paired_unpaired_ttest =
  readRDS(file = here("tables", "ecoli_infection_mouse_log2fc_paired_unpaired_ttest.rds")) %>% 
  mutate(organism = "Mouse")

#Limma
human_degs = readRDS(file = here("tables", "ecoli_infection_human_dge_limma_degs.rds")) %>% 
  clean_names() %>% 
  rename(padj = adj_p_val) %>% 
  mutate(organism = "Human",
         test = "Limma",
         paired = "Unpaired") 

mouse_degs = readRDS(file = here("tables", "ecoli_infection_mouse_dge_limma_degs.rds")) %>% 
  clean_names() %>% 
  rename(padj = adj_p_val) %>% 
  mutate(organism = "Mouse",
         test = "Limma",
         paired = "Paired") 

#Merge datasets
degs_ttest_limma =
  bind_rows(
    human_log2fc_paired_unpaired_ttest,
    human_degs,
    mouse_log2fc_paired_unpaired_ttest,
    mouse_degs
    )

#Summarize  
degs_ttest_limma_summary = degs_ttest_limma %>% 
  group_by(organism, timepoint, test, paired) %>% 
  summarise(
    n_degs = sum(padj <= 0.05, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) 

degs_ttest_limma_summary

```


```{r fig.height=5, fig.width=5}
# Plot number of DEGs per test and significance level
degs_ttest_limma_summary_plot_gg <- degs_ttest_limma_summary %>%
  ggplot(aes(
    x = timepoint,
    y = n_degs,
    fill = test,
    alpha = paired
  )) +
  geom_col(width = 0.8, position = position_dodge2()) +
  
  # Manual colors per test
  scale_fill_manual(values = c(
    "t-Test" = "gray25",
    "Limma" = "#4DBBD5FF"
  )) +
  
  # Alpha: paired = 1, unpaired = 0.5
  scale_alpha_manual(values = c("Paired" = 1, "Unpaired" = 0.2)) +
  
  # Continuous x-axis for numeric timepoints
  scale_x_discrete(
    breaks = sort(unique(degs_ttest_limma_summary$timepoint)),
    expand = c(0.01, 0.01)
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  
  theme_vaxgo() +
  labs(
    title = "Total DEGs by test and timepoint",
    subtitle = "Comparison across tests (FDR ≤ 0.05)",
    x = "Day",
    y = "Number of DEGs",
    fill = "Test",
    alpha = "Pairing"
  ) +
  facet_wrap(~organism, scales = "free_x", space = "free_x") +
  theme(
    legend.position = "top",
    axis.line.y = element_blank(),
    panel.grid.major.y = element_line(color = "gray75", linetype = 3),
    axis.ticks.y = element_line(color = "gray75")
  )

# Render plot
degs_ttest_limma_summary_plot_gg

degs_ttest_limma_summary_plot_gg %>% 
  ggsave(filename = here("Figures", "ecoli_infection_human_mouse_dge_ttest_limma_summary.png"),
         width = 5,
         height = 5)
```





#===================

# Combine all datasets

## Metadata
```{r}
#Merge results =====================
#Metadata
human_mouse_metadata_saureus = readRDS(file = here("tables", "saureus_infection_human_mouse_metadata.rds")) %>% 
  select(geo_sample, participant_id, organism, timepoint, time_unit, pathogen, treatment, age, sex, ethnicity, strain, tissue, tissue_infection, batch = batch_experimental) %>% 
  mutate(pathogen = "Staphylococcus aureus",
         tissue = "Blood",
         timepoint = as.numeric(timepoint),
         age = as.character(age)) %>% 
  mutate(timepoint = if_else(timepoint == 1, 24, timepoint),
         time_unit = "Hour") 

human_mouse_metadata_ecoli = readRDS(here("tables", "ecoli_infection_human_mouse_metadata.rds")) %>% 
  filter(host_strain == "C57BL/6J" | organism == "Human",
         !batch %in% c("1", "2")) %>% 
  mutate(timepoint = as.numeric(timepoint), 
         time_unit = "Hour", 
         pathogen = "Escherichia coli", 
         tissue = "Blood", tissue_infection = site_infection, 
         age = str_remove(age, "0.50-0.66"),
         organism = if_else(organism == "Mus musculus", "Mouse", organism)) %>% 
  select(geo_sample, participant_id = subject_id, organism, timepoint, time_unit, pathogen, treatment = status, age, sex, ethnicity, strain = host_strain,
         tissue, tissue_infection, batch) %>% 
  arrange(organism, timepoint)

human_mouse_engerix_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_metadata.rds")) %>% 
  mutate(pathogen = "Hepatitis B",
         vaccine = "Engerix B",
       treatment = "Engerix B", 
       type = "VLP",
       timepoint = as.numeric(timepoint),
       age = as.character(age)) %>% 
  select(geo_sample, participant_id, organism, timepoint, time_unit, pathogen, vaccine, type, treatment, age, sex, ethnicity = race, strain, tissue)

fluad_human_mouse_metadata = readRDS(here("tables", "fluad_human_mouse_metadata.rds"))  %>% 
  mutate(time_unit = "Day", timepoint = as.numeric(timepoint), age = as.character(age)) %>% 
  select(geo_sample, participant_id, organism, timepoint, time_unit, pathogen, vaccine, type, treatment, age, sex, ethnicity, strain, tissue)

all_human_mouse_metadata = bind_rows(
  human_mouse_metadata_saureus,
  human_mouse_metadata_ecoli,
  human_mouse_engerix_metadata,
  fluad_human_mouse_metadata
)
all_human_mouse_metadata %>% 
  saveRDS(here("tables", "all_human_mouse_metadata.rds"))
```

## Sample log2FC
```{r}
# Sample log2fc
#Import
human_mouse_log2fc_samples_ecoli = readRDS(here("tables", "ecoli_infection_human_mouse_samples_log2fc.rds"))  %>% 
  select(human_symbol, participant_id = subject_id, geo_sample, log2fc, value_raw, day0_value_raw, timepoint) %>% 
  mutate(timepoint = as.character(timepoint),
         pathogen = "Escherichia coli")
human_mouse_log2fc_samples_saureus = readRDS(here("tables", "saureus_infection_human_mouse_samples_log2fc.rds"))  %>% 
  select(human_symbol, participant_id, geo_sample, log2fc, value_raw, day0_value_raw, timepoint) %>% 
  mutate(timepoint = as.character(timepoint),
         pathogen = "Staphylococcus aureus")

human_mouse_log2fc_samples_engerixb = readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_exprs_samples_log2fc.rds"))  %>% 
   mutate(timepoint = as.numeric(timepoint)) %>% 
  inner_join(human_mouse_engerix_metadata,
             by = join_by("participant_id", "timepoint", "organism")) %>% 
  select(human_symbol, participant_id, geo_sample, log2fc, value_raw = value, day0_value_raw = day0, timepoint) %>% 
  mutate(timepoint = as.character(timepoint))

human_mouse_log2fc_samples_fluad = readRDS(file = here("tables", "influenza_fluad_human_mouse_log2fc_samples.rds")) %>% 
  mutate(timepoint = as.numeric(timepoint)) %>% 
  inner_join(fluad_human_mouse_metadata,
             by = join_by("participant_id", "timepoint", "organism")) %>% 
  select(human_symbol, participant_id, geo_sample, log2fc, value_raw = value, day0_value_raw = day0, timepoint) %>% 
  mutate(timepoint = as.character(timepoint))

#Unite
all_human_mouse_samples_log2fc = bind_rows(
  human_mouse_log2fc_samples_saureus,
  human_mouse_log2fc_samples_ecoli,
  human_mouse_log2fc_samples_engerixb,
  human_mouse_log2fc_samples_fluad
) %>% 
  select(human_symbol, geo_sample, timepoint, log2fc, value_raw, day0_value_raw) 

#Save
all_human_mouse_samples_log2fc %>% 
  saveRDS(here("tables", "all_human_mouse_samples_log2fc.rds"))

all_human_mouse_samples_log2fc %>%
  inner_join(all_human_mouse_metadata %>% 
               mutate(timepoint = as.character(timepoint)), 
             by = join_by("geo_sample", "timepoint")) %>% 
  filter(organism == "Human",
         pathogen == "Staphylococcus aureus")
```

## DGE results
```{r}
# DGE results
#Import
all_degs_human_mouse_engerixb = readRDS(here("tables", "hepatitisb_engerixb_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Hepatitis B", treatment = "Engerix B",
         timepoint = as.character(timepoint))
all_degs_human_mouse_saureus = readRDS(here("tables", "saureus_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Staphylococcus aureus", treatment = "Infection")
all_degs_human_mouse_ecoli = readRDS(here("tables", "ecoli_infection_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Escherichia coli", treatment = "Infection")
all_degs_human_mouse_fluad = readRDS(here("tables", "influenza_fluad_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "Influenza", treatment = "Fluad") %>% 
  rename(mean_l2fc = mean, ci_lower = lower, ci_upper = upper)

#Unite
all_degs_human_mouse = bind_rows(all_degs_human_mouse_saureus, 
                                 all_degs_human_mouse_ecoli, 
                                 all_degs_human_mouse_fluad, 
                                 all_degs_human_mouse_engerixb
                                 ) %>% 
  select(organism, pathogen, treatment, timepoint, condition, human_symbol, everything()) %>% 
  mutate(timepoint = if_else(pathogen == "Staphylococcus aureus" & timepoint == "1", "24", timepoint))
#Save
all_degs_human_mouse %>% 
  saveRDS(here("tables", "all_human_mouse_dge_limma_degs.rds"))

```

