```{r}
pairwise_human_fit_test =  read_rds(here("FIT_Mouse2man", "pairwise_human_fit_test.rds"))
```


```{r fig.height=4, fig.width=8}

quantile_diff_genes = pairwise_human_fit_test 

abs_diff_correlation_plot_data_split_qn = quantile_diff_genes %>% 
  mutate(
    quartile_label = cut(abs(human_2_mouse_qn),
                         breaks = quantile(abs(human_2_mouse_qn),
                                           probs = c(0, 0.10, 0.25, 0.5, 0.75, 1),
                                           na.rm = TRUE),
                         include.lowest = TRUE,
                         labels = c("Q10", "Q25", "Q50", "Q75", "Q100")),
    timepoint_day = paste0("Day ", timepoint)) %>% 
  mutate(pval_mouse_adj = if_else(pval_mouse_adj <= 0.05, "DEG", "Not DEG"),
         pval_human_adj = if_else(pval_human_adj <= 0.05, "DEG", "Not DEG"))

abs_diff_correlation_plot_data_split_qn %>% dplyr::count(timepoint)

#Get all genes
quantiles <- quantile(abs(quantile_diff_genes$human_2_mouse_qn), 
                    probs = c(0.10, 0.25, 0.5, 0.75, 1), 
                    na.rm = TRUE)

diff_correlation_plot_data_qn = quantile_diff_genes %>% 
  arrange(timepoint)  %>% 
  ungroup() %>% 
  #Split by quantile
    mutate(
    Q10 = abs(human_2_mouse_qn) <= quantiles[1],
    Q25 = abs(human_2_mouse_qn) <= quantiles[2],
    Q50 = abs(human_2_mouse_qn) <= quantiles[3],
    Q75 = abs(human_2_mouse_qn) <= quantiles[4],
    Q100 = abs(human_2_mouse_qn) <= quantiles[5]
  ) %>%
  #Cumulative
  pivot_longer(cols = starts_with("Q"),
               names_to = "quartile_label",
               values_to = "in_quantile") %>%
  mutate(quartile_label = fct_relevel(quartile_label, "Q10", "Q25", "Q50", "Q75", "Q100")) %>% 
  filter(in_quantile) %>%
  select(-in_quantile)
```


```{r fig.height=4, fig.width=8}
abs_diff_correlation_plot_data_split_qn %>% 
  ungroup() %>% 
  ggplot() +
  aes(x = mean_mouse_qn,
      y = mean_human_qn) +
  facet_wrap(~paste("Day", timepoint)) +
  #All points, colored by differences
  geom_point(
             aes(color = quartile_label,
                 # alpha = quartile_label,
                 fill = quartile_label),
             # fill = NA,
             # alpha = 1,
             pch = 21) +

  #Same DEGs (WORST)
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1,
             size = 0.5,
             pch = 21,
             data = abs_diff_correlation_plot_data_split_qn %>%
               filter(pval_mouse_adj == pval_human_adj,
                      pval_mouse_adj == "DEG") %>%
               filter(quartile_label %in% c("Q50", "Q75", "Q100"))) +
  #Same DEGs (BEST)
  geom_point(color = "black",
             fill = "#4DBBD5FF",
             alpha = 1,
             size = 2,
             pch = 21,
             data = abs_diff_correlation_plot_data_split_qn %>%
               filter(pval_mouse_adj == pval_human_adj,
                      pval_mouse_adj == "DEG") %>%
               filter(quartile_label %in% c(
                 "Q10",
                 "Q25"
                 ))) +
   #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(slope = -1, alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  scale_color_manual(values = c("Q10" = "black",
                               "Q25" = "gray25",
                               "Q50" = "gray50",
                               "Q75" = "gray75",
                               "Q100" = "gray90")) +
  scale_fill_manual(values = c("Q10" = "black",
                               "Q25" = "gray25",
                               "Q50" = "gray50",
                               "Q75" = "gray75",
                               "Q100" = "gray90")) +
  scale_x_continuous(limits = c(-1.5, 1.5)) +
  scale_y_continuous(limits = c(-1.5, 1.5)) +
  
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.position = "right",
        legend.direction = "vertical",
        legend.key.width = unit(0.4, 'cm'),
        legend.key.height = unit(0.4, 'cm'),
        legend.key = element_blank(),
        legend.title = element_text(size = 10, vjust = 0.9),
        legend.text = element_text(size = 7, vjust = 0.5, hjust = -0.2)) +
  labs(title = "Which genes and DEGs vary in both human and mouse?",
       subtitle = "Differences in mouse to human, by timepoint.In colors, DEGs in human AND mouse, p_adjust < 0.05, |diff| < 0.2. \nPurple = Negatively correlated, Blue = Positively correlated",
       x = "Mouse L2FC",
       y = "Human L2FC",
       alpha = "Quantiles",
       fill = "Quantiles",
       color = "|Z|")


```






### plot btms with CIs

Engerix
```{r fig.height=4, fig.width=8}

dge_btm_mean_process_samples = 
  readRDS(file = here("tables", "engerixb_dge_btm_mean_process_samples.rds"))


dge_btm_mean_process = dge_btm_mean_process_samples %>% 
  group_by(organism, timepoint, process) %>% 
  summarize(
    mean = mean(log2fc, na.rm = TRUE) %>% round(2),
    median = median(log2fc, na.rm = TRUE) %>% round(2),
    min = min(log2fc, na.rm = TRUE) %>% round(2),
    max = max(log2fc, na.rm = TRUE) %>% round(2),
    sd = sd(log2fc, na.rm = T),
    n = sum(!is.na(log2fc)),
    se = sd/sqrt(n),
    t_critical = qt(0.975, df = n - 1),
    ci_lower = mean - t_critical * se,
    ci_upper = mean + t_critical * se,
    p_value = t.test(log2fc, mu = 0)$p.value,
    .groups = "drop",
    group,
    subgroup
  ) %>%
  distinct() %>% 
  mutate(p_adj = p.adjust(p_value, method = "BH"), 
         sig_cat = case_when(p_adj < 0.01 ~ "< 0.01", 
                             p_adj < 0.05 ~ "< 0.05",
                             p_adj < 0.10 ~ "< 0.10", 
                             p_adj < 0.25 ~ "< 0.25", 
                             TRUE ~ "≥ 0.25"),
         sig_cat = factor(sig_cat,
                   levels = c("< 0.01", "< 0.05", "< 0.10", "< 0.25", "≥ 0.25"))) %>% 
  select(organism, timepoint, process, mean, median, p_adj, p_value, sig_cat, ci_lower, ci_upper, group, subgroup) %>% 
  pivot_wider(names_from = organism, 
              values_from = c(mean, median, p_value, p_adj, ci_lower, ci_upper, sig_cat), 
              names_sep = "_", names_sort = TRUE) 



dge_btm_mean_process  %>% 
  ggplot() +
  aes(x = mean_Mouse,
      y = mean_Human) +
  facet_wrap(~timepoint, scales = "free") +
  geom_point(aes(fill = group), alpha = 1, size = 3, 
             color = "black", pch = 21,
             data = . %>% filter(p_value_Mouse <= 0.25| p_value_Mouse <= 0.25)) +
  geom_point(aes(fill = group), alpha = 0.2, size = 3, 
             color = "black", pch = 21,
             data = . %>% filter(p_value_Mouse > 0.25| p_value_Mouse > 0.25)) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "Mouse Log2FC", 
       y = "Human Log2FC", 
       color = "",
       fill = "",
       title = "Mouse vs. Human, BTMs,\nEngerixB vaccination")

```





```{r}
#Import
dge_btm_mean_process_samples = 
  readRDS(file = here("tables", "dge_Ecoli_btm_mean_process_samples.rds")) %>% 
  mutate(timepoint = if_else(is.na(timepoint), "24", timepoint)) %>% 
  filter(timepoint == 24)


dge_btm_mean_process = dge_btm_mean_process_samples %>% 
  group_by(organism, timepoint, process) %>% 
  summarize(
    mean = mean(log2fc, na.rm = TRUE) %>% round(2),
    median = median(log2fc, na.rm = TRUE) %>% round(2),
    min = min(log2fc, na.rm = TRUE) %>% round(2),
    max = max(log2fc, na.rm = TRUE) %>% round(2),
    sd = sd(log2fc, na.rm = T),
    n = sum(!is.na(log2fc)),
    se = sd/sqrt(n),
    t_critical = qt(0.975, df = n - 1),
    ci_lower = mean - t_critical * se,
    ci_upper = mean + t_critical * se,
    p_value = t.test(log2fc, mu = 0)$p.value,
    .groups = "drop",
    group,
    subgroup
  ) %>%
  distinct() %>% 
  mutate(p_adj = p.adjust(p_value, method = "BH"), 
         sig_cat = case_when(p_adj < 0.01 ~ "< 0.01", 
                             p_adj < 0.05 ~ "< 0.05",
                             p_adj < 0.10 ~ "< 0.10", 
                             p_adj < 0.25 ~ "< 0.25", 
                             TRUE ~ "≥ 0.25"),
         sig_cat = factor(sig_cat,
                   levels = c("< 0.01", "< 0.05", "< 0.10", "< 0.25", "≥ 0.25"))) %>% 
  select(organism, timepoint, process, mean, median, p_adj, p_value, sig_cat, ci_lower, ci_upper, group, subgroup) %>% 
  pivot_wider(names_from = organism, 
              values_from = c(mean, median, p_value, p_adj, ci_lower, ci_upper, sig_cat), 
              names_sep = "_", names_sort = TRUE) 



dge_btm_mean_process %>%
  mutate(
    significance = case_when(
      p_adj_Mouse <= 0.25 & p_adj_Human <= 0.25 ~ "Significant in both",
      p_adj_Mouse > 0.25 & p_adj_Human > 0.25 ~ "Non-significant in both",
      TRUE ~ "Mixed"
    ),
    significance = factor(
      significance,
      levels = c("Significant in both", "Mixed", "Non-significant in both")
    )
  ) %>%
  ggplot(aes(x = mean_Mouse, y = mean_Human)) +
  facet_wrap(~timepoint, scales = "free") +
  
  geom_point(
  aes(fill = group, color = group, shape = significance),
  size = 3,
  stroke = 0.7) +
scale_fill_manual(values = immune_colors, name = "Group") +
scale_color_manual(values = immune_colors, guide = "none") + # esconde legenda duplicada
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  
  theme_vaxgo() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(size = 10),
    strip.text.y = element_blank()
  ) +
  
#   scale_fill_manual(values = immune_colors, name = "Group") +
#   guides(
#   fill = guide_legend(override.aes = list(color = "black"))
# ) +
  scale_shape_manual(
    values = c(
      "Significant in both" = 21,  # círculo cheio
      "Mixed" = 24,                # triângulo
      "Non-significant in both" = 22  # quadrado
    ),
    name = "Adjusted p-value match"
  ) +
  
  labs(
    x = "Mouse Log2FC",
    y = "Human Log2FC",
    title = "Mouse vs. Human, BTMs,\nEngerixB vaccination"
  )
```



#Scatterplot of correlation (arrows, day)
```{r fig.height=3, fig.width=10}

zscore_genes_plot_timepoints = zscore_genes %>% 
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = mean_human,
      y = mean_mouse) +
  #All points, colored by zscore
  geom_point(color ="gray90",
             fill = NA,
             pch = 21,
             alpha = 0.4) +
  #Select
  geom_point(aes(shape = as.character(timepoint)),
             # pch = 21, 
             alpha = 0.5,
             # stroke = 0.5, 
             size = 2, 
             color = "#4DBBD5FF",
            data =degs_pos_neg_mouse %>% 
              filter(quantile_label %in% c("Q10", "Q25", "Q50"))) +
  geom_path(aes(group = human_symbol),
            arrow=arrow(length=unit(0.05,"in")), 
            show.legend= FALSE,
            alpha = 0.5,
            data = degs_pos_neg_mouse %>% 
              filter(quantile_label %in% c("Q10", "Q25", "Q50")),
            # color = "black"
            ) +
  # scale_color_manual(values = c("1" = "#4DBBD5",
  #                               "3" = "#A6DDEA",
  #                               "7" = "white")) +
    scale_x_continuous(limits = c(-0.6, 0.6)) +
  scale_y_continuous(limits = c(-0.6, 0.6)) +
  geom_abline(alpha = 0.5, linetype = 2) +
  geom_abline(slope = -1, alpha = 0.5, linetype = 2) +
  facet_wrap(~quantile_label) + 
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
         legend.title = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.key = element_rect(colour = NA)) +
  labs(title = "Are these genes well predicted in most timepoints?",
       subtitle = "Differences in mouse to human, using DEGs in human AND mouse, p_adjust < 0.05.\nAll timepoints. Only Q50 genes.",
       x = "Human L2FC",
       y = "Mouse L2FC",
       fill = "Timepoint",
       color = "Timepoint",
       shape = "Timepoint")  

zscore_genes_plot_timepoints$layers[[2]]

```



```{r}

degs_segments <- degs_pos_neg_mouse %>%
  filter(quantile_label %in% c("Q25")) %>%
  mutate(quantile_label = if_else(quantile_label == "Q25", "<Q25", quantile_label)) %>% 
  bind_rows(degs_pos_neg_mouse_split %>% 
    mutate(quantile_label = if_else(quartiles_diff == "Q100" | 
                                      quartiles_diff == "Q75", "Q75>", 
                                    quartiles_diff)) %>% 
      filter(quantile_label == "Q75>")) %>% 
  arrange(human_symbol, timepoint) %>%
  group_by(human_symbol, quantile_label) %>%
  mutate(
    mean_human_next = lead(mean_human),
    mean_mouse_next = lead(mean_mouse),
    timepoint_next = lead(timepoint),
    direction = paste0("D", timepoint, "→D", timepoint_next)
  ) %>%
  filter(!is.na(mean_human_next), !is.na(mean_mouse_next)) %>%
  ungroup() %>% 
  filter(!(mean_human == mean_human_next & mean_mouse == mean_mouse_next))
  
  

degs_segments
```


```{r}
# Plot com caminhos coloridos por direção
zscore_genes_plot_timepoints <- zscore_genes %>% 
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = mean_human, y = mean_mouse) +
  # Curva com seta no fim
  geom_curve(data = degs_segments,
             aes(x = mean_human, y = mean_mouse,
                 xend = mean_human_next, yend = mean_mouse_next,
                 color = direction,
                 group = human_symbol),
             curvature = 0,     # Ajuste a curvatura conforme desejado
             size = 1,
             arrow = arrow(length = unit(0.1, "in")),
             alpha = 0.5,
             lineend = c('round'),
              linejoin = c('mitre'),
             show.legend = TRUE) +
    scale_x_continuous(limits = c(-0.6, 0.6)) +
  scale_y_continuous(limits = c(-0.6, 0.6)) +
  geom_abline(alpha = 0.5, linetype = 2) +
  geom_abline(slope = -1, alpha = 0.5, linetype = 2) +
  facet_wrap(~quantile_label) + 
   scale_color_manual(values = c(
  "D1→D3" = "#7400b8",
  "D3→D7" = "#4dbbd5",
  "D1→D7" = "#3c5488"
)) + 
  scale_fill_manual(values = c(
  "D1→D3" = "#7400b8",
  "D3→D7" = "#4dbbd5",
  "D1→D7" = "#3c5488"
)) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.key.width = unit(0.4, 'cm'),
        legend.key.height = unit(0.4, 'cm'),
        legend.title = element_text(size = 10, vjust = 0.9),
        legend.text = element_text(size = 7, vjust = 0.5, hjust = -0.2),
        legend.key = element_rect(colour = NA)) +
  labs(title = "Are these genes well predicted in most timepoints?",
       subtitle = "Differences in mouse to human, using DEGs in human AND mouse, p_adjust < 0.05.\nAll timepoints.",
       x = "Human L2FC",
       y = "Mouse L2FC",
       fill = "Timepoint",
       color = "Timepoint",
       shape = "Timepoint")  

zscore_genes_plot_timepoints
```


```{r}
zscore_genes_plot_timepoints_bar = degs_segments %>% 
  group_by(quantile_label, direction) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  ggplot() +
  aes(x = direction, y = n, fill = direction) +
  geom_col(position = "dodge2", alpha = 0.5) +
  geom_text(aes(label = n),
            position = "dodge2", 
            color = "black",
            hjust = 0.5,
            vjust = -0.5
            ) +
  scale_color_manual(values = c(
  "D1→D3" = "#7400b8",
  "D3→D7" = "#4dbbd5",
  "D1→D7" = "#3c5488"
)) + 
  scale_fill_manual(values = c(
  "D1→D3" = "#7400b8",
  "D3→D7" = "#4dbbd5",
  "D1→D7" = "#3c5488"
)) +
  # scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 110))+
  facet_wrap(~quantile_label) + 
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.position = "none") +
  labs(
    # title = "Are these genes well predicted in most timepoints?",
       # subtitle = "Differences in mouse to human, using DEGs in human AND mouse, p_adjust < 0.05.\nAll timepoints. Only Q50 genes.",
       x = "Direction",
       y = "Total",
       fill = "Direction")  

  zscore_genes_plot_timepoints_bar
```






## Correlation unweighted



### Log2FC differences

```{r fig.height=6, fig.width=6}
#Plot weighted correlation scatterplot
human_mouse_log2fc_avg_wide_degs_all_immune_nonimmune %>%
  inner_join(correlation_filtered_inter_weighted_filtered %>% 
               select(pathogen_human, pathogen_mouse, 
                          treatment_human, treatment_mouse,
                          timepoint_human, timepoint_mouse,
                      label),
             by = join_by(pathogen_human, pathogen_mouse, 
                          treatment_human, treatment_mouse,
                          timepoint_human, timepoint_mouse)) %>% 
  mutate(diff_log2FC_cat = case_when(
    mean_log2fc_diff < -2 ~ "<-2",
    mean_log2fc_diff > 2  ~ ">2",
    TRUE        ~ as.character(round(mean_log2fc_diff, 0))),
  diff_log2FC_cat = fct_relevel(diff_log2FC_cat, "<-2", "-2", "-1", "0", "1", "2", ">2")) %>% 
  ggplot() +
  aes(x = mean_log2fc_mouse, y = mean_log2fc_human) +
  facet_wrap(~treatment_human+paste0(pathogen_human, "-", pathogen_mouse) + timepoint_comparison + label, scales = "free") +
  geom_point(aes(size = weight_se_inverse,
                 fill = diff_log2FC_cat),
             alpha = 0.5,
             stroke = 0,
             pch = 21) +
  
  scale_fill_manual(
    values = c("<-2" = "darkred", "-2" = "red", "-1" = "orange", 
               "0" = "gray", "1" = "orange", "2" = "red", ">2" = "darkred")
  ) +
  labs(
    x = "Human Median Log2FC",
    y = "Mouse Median Log2FC"
  ) +
  theme_vaxgo()+
  theme(legend.position = "top")

```

##### Log2FC vs Log2fc

```{r}
#Correlation ------

#Log2fc vs Log2fc
cor_data <- human_mouse_log2fc_avg_wide_degs_immune_nonimmune %>%
  group_by(pathogen, timepoint_comparison, gene_set) %>%
  summarise(
    cor_test_spearman = list(safe_cor_test(mean_log2fc_mouse, mean_log2fc_human, method = "spearman")),
    cor_test_pearson = list(safe_cor_test(mean_log2fc_mouse, mean_log2fc_human, method = "pearson")),,
    .groups = "drop"
  ) %>%
  inner_join(degs_immune_nonimmune_summary, by = join_by("pathogen", "timepoint_comparison", "gene_set")) %>% 
  mutate(
    #Spearman
    cor_value_spearman = map_dbl(cor_test_spearman, ~ if (is.null(.x)) NA_real_ else .x$estimate),
    p_spearman   = map_dbl(cor_test_spearman, ~ if (is.null(.x)) NA_real_ else .x$p.value),
    p_label_spearman = case_when(
      is.na(p_spearman)      ~ "ns",
      p_spearman <= 0.01     ~ "***",
      p_spearman <= 0.05     ~ "**",
      p_spearman <= 0.10     ~ "*",
      TRUE                ~ "ns"
    ),
    label_spearman = paste0("rho=", round(cor_value_spearman, 2), ", n=", n_genes, ", ", p_label_spearman),
    #Pearson
    cor_value_pearson = map_dbl(cor_test_pearson, ~ if (is.null(.x)) NA_real_ else .x$estimate),
    p_pearson   = map_dbl(cor_test_pearson, ~ if (is.null(.x)) NA_real_ else .x$p.value),
    p_label_pearson = case_when(
      is.na(p_pearson)      ~ "ns",
      p_pearson <= 0.01     ~ "***",
      p_pearson <= 0.05     ~ "**",
      p_pearson <= 0.10     ~ "*",
      TRUE                ~ "ns"
    ),
    label_pearson = paste0("r=", round(cor_value_pearson, 2), ", n=", n_genes, ", ", p_label_pearson)
  ) %>% 
  drop_na(cor_value_spearman,
          cor_value_pearson) %>% 
  mutate(
    gene_set = factor(gene_set, levels = c("All genes", "DEGs", "Not-immune", "Immune")),
    facet_label = paste0(gene_set, ": ", label_spearman)) %>% 
    #Reorder timepoints
  separate(timepoint_comparison, into = c("timepoint_mouse", "timepoint_human"), remove = F) %>% 
  mutate(timepoint_mouse = fct_relevel(timepoint_mouse, c("2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7"))) %>% 
  arrange(timepoint_mouse) %>%
  mutate(timepoint_comparison = fct_inorder(timepoint_comparison))

cor_data_facetted = cor_data %>% 
  arrange(gene_set) %>% 
   group_by(pathogen, timepoint_comparison) %>%
  summarise(
    facet_label = paste(facet_label, collapse = "\n"),
    .groups = "drop"
  )  %>% 
  mutate(facet_label = paste0(timepoint_comparison, "\n", facet_label)) %>% 
  arrange(timepoint_comparison) %>% 
  mutate(facet_label = fct_inorder(facet_label))

cor_data
cor_data_facetted


human_mouse_log2fc_avg_wide_labels = human_mouse_log2fc_avg_wide %>% 
  inner_join(cor_data_facetted, 
             by = join_by("pathogen","timepoint_comparison")) %>% 
  mutate(pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma")) %>% 
  distinct() 

#Unite gene data and correlation/facets
human_mouse_log2fc_avg_wide_labels_filters = 
  bind_rows(
    #All genes
    human_mouse_log2fc_avg_wide_labels %>% 
  mutate(color_group = "All genes"),

    #All DEGs
   human_mouse_log2fc_avg_wide_labels %>% 
  filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>% 
  mutate(color_group = "All DEGs"),

   #Shared DEGs
   human_mouse_log2fc_avg_wide_labels %>% 
  filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05) %>% 
  mutate(color_group = "Shared DEGs"),

  #Immune DEGs
    human_mouse_log2fc_avg_wide_labels %>% 
  filter(human_symbol %in% c(btm_annotation_genes %>% 
                               filter(module_category == "immune") %>% 
                               pull(symbol)),
         p_adj_human <= 0.05 & p_adj_mouse <= 0.05) %>% 
  mutate(color_group = "Immune")
  ) %>% 
  mutate(color_group = fct_relevel(color_group, "All genes", "All DEGs", "Shared DEGs", "Immune"))



```


```{r fig.height=12, fig.width=16}
# Log2FC vs Log2fc
human_mouse_log2fc_log2fc_avg_wide_scatterplot = human_mouse_log2fc_avg_wide_labels_filters %>% 
  
  #Plot
  ggplot() +
  aes(x = mean_log2fc_mouse,
      y = mean_log2fc_human) +
  #Facet
  facet_nested_wrap(vars(pathogen, facet_label),
                    scales = "free",
                    nrow = 4,
                    strip = strip_nested(size = "variable")) +
  
  geom_point(aes(fill = color_group),
             alpha = 1,
             color = "gray95",
             pch = 21,
             size = 1.5,
             data = . %>% filter(color_group == "All genes")) +
  
  geom_point(aes(fill = color_group),
             alpha = 1,
             color = "gray85",
             pch = 21,
             size = 1.5,
             data = . %>% filter(color_group == "All DEGs")) +
  
  geom_point(aes(fill = color_group,
                 color = color_group),
             alpha = 1,
             pch = 21,
             size = 1.5,
             data = . %>% filter(!color_group %in% c("All genes", "All DEGs"))) +
  
  #Regression line: Linear model, Pearson
  stat_smooth(method = "lm", geom = "line", alpha = 1, se = F, size = 1,
              mapping = aes(group = color_group,
                        color = color_group)) +
  guides(fill = guide_legend(override.aes = list(size = 4)),
         color = guide_legend(override.aes = list(linewidth = 4, shape = 1))
         ) +
  scale_fill_manual(values = c("All genes" = "gray95",
                               "All DEGs" = "gray85",
                               "Shared DEGs" = "#175289",
                               "Immune" ="#46eec6")) +
  scale_color_manual(values = c("All genes" = "black",
                               "All DEGs" = "#4361ee",
                               "Shared DEGs" = "#175289",
                               "Immune" ="#46eec6")) +
  #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "top",
        legend.direction = "horizontal",
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       fill = "Gene type",
       color = "Linear regression",
       title = "Correlation plot,\nHuman vs Mouse")


human_mouse_log2fc_log2fc_avg_wide_scatterplot 

#Save
# human_mouse_log2fc_log2fc_avg_wide_scatterplot %>%
#   ggsave(filename = here("Figures", "all_human_mouse_dge_degs_scatterplot.png"),
#          width = 16, height = 12)

```


##### Ranks vs Ranks

```{r fig.height=12, fig.width=16}
# 
# #Create ranks
# human_mouse_log2fc_avg_wide_labels_ranked <-
#   human_mouse_log2fc_avg_wide_labels %>%
#   group_by(pathogen, facet_label) %>%
#   mutate(
#     rank_log2fc_mouse = dense_rank(mean_log2fc_mouse),
#     rank_log2fc_human = dense_rank(mean_log2fc_human)
#   ) %>%
#   ungroup()
# 
# 
# human_mouse_rank_rank_avg_wide_scatterplot = human_mouse_log2fc_avg_wide_labels_ranked %>% 
#   
#   #Plot
#   ggplot() +
#   aes(x = rank_log2fc_mouse,
#       y = rank_log2fc_human) +
#   #Facet
#   facet_nested_wrap(vars(pathogen, facet_label),
#                     scales = "free",
#                     nrow = 4,
#                     strip = strip_nested(size = "variable")) +
#   
#   geom_point(aes(color = color_group,
#                  fill = color_group),
#              alpha = 1,
#              size = 1.5,
#              data = human_mouse_log2fc_avg_wide_labels_ranked %>% 
#                mutate(color_group = "All genes")) +
#   
#   #All DEGs
#   geom_point(aes(color = color_group,
#                  fill = color_group), 
#              alpha = 1,
#              size = 1.5,
#              data = human_mouse_log2fc_avg_wide_labels_ranked %>% 
#                filter(p_adj_human <= 0.05 | p_adj_mouse <= 0.05) %>% 
#                mutate(color_group = "All DEGs")) +
#   
#   #DEGs in both organisms
#   geom_point(aes(fill = color_group,
#                  color = color_group),
#              alpha = 1,
#              size = 1.5,
#              pch = 21,
#              data = human_mouse_log2fc_avg_wide_labels_ranked %>%
#                #Filter DEGs in both organisms, same sign
#                filter(p_adj_human <= 0.05 & p_adj_mouse <= 0.05) %>% 
#              mutate(color_group = "Shared DEGs")) +
# 
#   #Immune DEGs
#   geom_point(aes(fill = color_group,
#                  color = color_group),
#              alpha = 1,
#              size = 1.5,
#              pch = 21,
#              data = human_mouse_log2fc_avg_wide_labels_ranked %>%
#                filter(human_symbol %in% c(btm_annotation_genes %>%                                                
#                                             filter(module_category == "immune") %>%
#                                             pull(symbol)),
#                       p_adj_human <= 0.05 & p_adj_mouse <= 0.05) %>% 
#                mutate(color_group = "Immune")) +
#   guides(color = "none",
#          fill = guide_legend(override.aes = list(size = 4))) +
#   scale_fill_manual(values = colors$shared) +
#   scale_color_manual(values = colors$shared) +
#   #Add lines
#   geom_abline(alpha = 0.2, linetype = 2) +
#   #Theme
#   theme_vaxgo() +
#   theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
#         axis.text.y = element_text(size = 10),
#         strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
#         legend.position = "top",
#         legend.direction = "horizontal",
#         ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)) +
#   labs(x = "Mouse L2FC",
#        y = "Human L2FC",
#        fill = "Gene type",
#        title = "Correlation plot,\nHuman vs Mouse")
# 
# 
# human_mouse_rank_rank_avg_wide_scatterplot 
# human_mouse_rank_rank_avg_wide_scatterplot %>%
#   ggsave(filename = here("Figures", "all_human_mouse_dge_degs_scatterplot_ranks.png"),
#          width = 16, height = 12)

```







#Glassplot

```{r}
library(ggplot2)
library(ggfx)
library(dplyr)
library(MASS)
library(here) # Para salvar

# ---------------------------------------------------------
# 1. GERAR DADOS
# ---------------------------------------------------------
set.seed(101)
n_genes <- 1000

# Correlação alvo (r = 0.9)
target_r <- 0.9
sigma_matrix <- matrix(c(1, target_r, target_r, 1), nrow = 2)

dados_brutos <- mvrnorm(n = n_genes, mu = c(0, 0), Sigma = sigma_matrix)

df_log2fc <- data.frame(
  human_logfc = dados_brutos[, 1] * 2.5, 
  mouse_logfc = dados_brutos[, 2] * 2.5 
) %>%
  mutate(
    # --- LÓGICA DE RANKING (Para as CAMADAS do Blur) ---
    distancia = abs(human_logfc) + abs(mouse_logfc),
    fake_pval = 1 / (distancia + runif(n(), 0, 2)), 
    pval_rank = rank(fake_pval), # 1 a 1000
    
    # --- LÓGICA DE COR (Para o FILL/COLOR) ---
    # Aqui criamos grupos que NÃO dependem do ranking.
    # Simula vias biológicas diferentes misturadas nos dados.
    pathway_group = sample(c("Inflammation", "Cell Cycle"), 
                           n(), replace = TRUE)
  )

# Calcular R² para o texto
r_squared <- round(cor(df_log2fc$human_logfc, df_log2fc$mouse_logfc)^2, 2)
label_text <- paste0("R² ≈ ", r_squared)

# ---------------------------------------------------------
# 2. DEFINIÇÃO DE CORES
# ---------------------------------------------------------
# Cores para os grupos (e não para o rank)
my_colors <- c(
  "Inflammation" = "#607d8b", # Vermelho
  "Cell Cycle"   = "#455a64"  # Azul
)

# ---------------------------------------------------------
# 3. PLOT COM GGFX
# ---------------------------------------------------------
glass_plot <- ggplot(df_log2fc, aes(x = human_logfc, y = mouse_logfc)) +
  
  # Guias de fundo
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey80") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey80") +
  
  # --- CAMADA 1: BACKGROUND (Todos os 1000 pontos) ---
  # Blur forte. Usamos shape 16 (sem borda) para um desfoque mais suave.
  with_blur(
    geom_point(
      data = . %>% filter(pval_rank > 500),
      aes(color = pathway_group), # Colore pelo grupo
      size = 3,
      alpha = 0.5,
      shape = 16
    ),
    sigma = 5 # Blur alto
  ) +
  
  # --- CAMADA 2: MIDGROUND (Top 500 Rank) ---
  # Blur médio. Filtrado pelo Rank, mas colorido pelo Grupo.
  with_blur(
    geom_point(
      data = . %>% filter(pval_rank <= 500),
      aes(color = pathway_group),
      size = 3,
      alpha = 0.8,
      shape = 16
    ),
    sigma = 2 # Blur médio
  ) +
  
  # --- CAMADA 3: FOREGROUND (Top 100 Rank - Foco) ---
  # Nítido. Usamos shape 21 para ter borda preta e preenchimento (fill) colorido.
  with_shadow(
    geom_point(
    data = . %>% filter(pval_rank <= 300),
    aes(fill = pathway_group,
        color = pathway_group), # Note que aqui usamos FILL, não color
    size = 3,
    # color = "black",
    shape = 21,                # Círculo preenchido
    stroke = 0.2,                # Espessura da borda preta
    alpha = 1
  ),
  x_offset = 2,
  y_offset = 2,
  sigma = 3
  ) +
  
  geom_smooth(method = "lm", se = FALSE, color = "#c59d45", size = 1.5) +
  
  # --- TEMA E ESCALAS ---
  scale_color_manual(values = my_colors) + # Para as camadas 1 e 2 (blur)
  scale_fill_manual(values = my_colors) +  # Para a camada 3 (foco)
  coord_cartesian(xlim = c(-7, 7), ylim = c(-7, 7)) +
  theme_classic() +
  labs(
    x = "Human Log2FC", 
    y = "Mouse Log2FC", 
    fill = "Pathway", 
    color = "Pathway"
  ) +
  theme(
    axis.line = element_line(color = "black", linewidth = 1),
    axis.text = element_text(size = 12, color = "black"),
    axis.title = element_text(size = 14, color = "black"),
    legend.position = "top"
  )

# Exibir o plot
print(glass_plot)

# Salvar
ggsave(filename = here("Figures", "glassplot_test.png"), plot = glass_plot, width = 6, height = 6)


```


#===============================
#Not used

## Functional analysis Sample-wise analysis

#### Immune: Summarize inter-individual correlation

Annotate genes log2fc table with BTMs

```{r}
# Define output filename root for the combined dataset --------
filename <- "all_human_mouse"

# Load sample-level log2FC values for all genes across all conditions --------
samples_log2fc <- readRDS(file = here("tables", "all_human_mouse_samples_log2fc.rds"))

# Quick inspection: how many unique participants are in the dataset --------
samples_log2fc %>% select(participant_id) %>% distinct() %>% View()

# Load sample metadata (pathogen label, treatment, organism, timepoint) --------
samples_annotation <- readRDS(file = here("tables", "all_human_mouse_metadata.rds"))

# Check which pathogens are represented in the annotation table --------
samples_annotation$pathogen %>% unique()

# Annotate each gene's log2FC with its BTM module membership --------
# Join sample-level expression to metadata, then join to BTM annotation by gene symbol
dge_btm_mean_process_samples <- samples_log2fc %>%
  inner_join(samples_annotation, by = "geo_sample") %>%
  ungroup() %>%
  select(symbol = human_symbol, participant_id, geo_sample, timepoint, log2fc, organism, pathogen, treatment) %>%
  inner_join(btm_annotation_genes, by = "symbol")

# Verify which pathogens are present in the BTM-annotated data --------
dge_btm_mean_process_samples$pathogen %>% unique()

# Match human infection timepoints to mouse timepoints --------
# For S. aureus and E. coli, human samples have different time labels than mouse.
# Duplicate human samples across all available mouse timepoints for that pathogen
# so that the correlation can be computed at matched temporal windows.
dge_btm_mean_process_samples_matched <-
  bind_rows(
    # Keep non-infection conditions as-is (vaccination and injury already match between species)
    dge_btm_mean_process_samples %>%
      filter(organism == "Mouse" | !(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))),

    # For human infection samples, duplicate across all mouse timepoints of the same pathogen
    dge_btm_mean_process_samples %>%
      filter(organism == "Human",
             pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
      select(-timepoint) %>%
      inner_join(
        dge_btm_mean_process_samples %>%
          filter(organism == "Mouse",
                 pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
          select(pathogen, timepoint) %>%
          distinct(),
        by = "pathogen",
        relationship = "many-to-many"
      )
  )

# Verify the matching: check a specific human infection sample at a mouse timepoint --------
dge_btm_mean_process_samples_matched %>%
  filter(pathogen == "Staphylococcus aureus",
         timepoint == "2h",
         organism == "Human")

# Pivot BTM-annotated data to wide format with one column per participant_id --------
# Compute the mean log2FC per participant, timepoint, and BTM process,
# then spread participants to columns for cross-sample correlation.
dge_btm_mean_process_wide <- dge_btm_mean_process_samples_matched %>%
  mutate(condition_label = paste0(participant_id, "_Day", timepoint),
         timepoint = as.character(timepoint)) %>%
  group_by(pathogen, participant_id, timepoint, process) %>%
  summarise(mean_log2fc = mean(log2fc, na.rm = TRUE)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>%
  inner_join(btm_annotation_genes %>%
               select(process, group, subgroup) %>%
               distinct(),
             by = "process") %>%
  ungroup()

# Print sample counts per condition, used to verify the matching step --------
dge_btm_mean_process_samples_matched %>%
  dplyr::count(pathogen, organism, timepoint)
```

Import data

```{r}
# Compute mean log2FC per BTM module for each sample --------
# Group by pathogen, organism, participant, timepoint, and BTM process,
# then average the log2FC across all genes belonging to that module.
dge_btm_mean_process_sample_long <- dge_btm_mean_process_samples_matched %>%
  group_by(pathogen, organism, participant_id, timepoint, process) %>%
  summarise(mean_log2fc = mean(log2fc, na.rm = TRUE)) %>%
  distinct() %>%
  ungroup()

# Inspect: check S. aureus samples to verify that matching produced reasonable timepoint combinations --------
dge_btm_mean_process_sample_long %>%
  filter(pathogen == "Staphylococcus aureus")

# Count samples per condition to verify balanced representation --------
dge_btm_mean_process_sample_long %>%
  dplyr::count(pathogen, organism, timepoint) 
  
```

1 individual vs 1 individual

```{r}
# Prepare long-format table of mean log2FC per sample and BTM process --------
# Each row is a (sample, process) pair; we create a unique condition identifier for wide pivoting
btm_samples_long <- dge_btm_mean_process_sample_long %>%
  select(pathogen, organism, process, participant_id, timepoint, mean_log2fc) %>%
  distinct() %>%
  mutate(condition = paste0(pathogen, "_", participant_id, "_", timepoint)) %>%
  drop_na()

# Create annotation table that maps each condition back to its pathogen, organism, and timepoint --------
btm_samples_annotation <- btm_samples_long %>%
  select(pathogen, organism, participant_id, timepoint, condition) %>%
  distinct()

# Pivot to wide format: rows = BTM processes, columns = individual samples (conditions) --------
btm_samples_wide_all <- btm_samples_long %>%
  select(-pathogen, -timepoint, -organism, -participant_id) %>%
  pivot_wider(names_from = "condition",
              values_from = "mean_log2fc") %>%
  column_to_rownames("process")

# Compute Pearson correlation matrix and p-values between all pairs of samples --------
cor_btm_samples_wide_all <- btm_samples_wide_all %>%
  as.matrix() %>%
  rcorr(type = "pearson")

# Convert correlation matrix to long format: each row is a pair (condition1, condition2) --------
cor_btm_samples_wide_all_r <- cor_btm_samples_wide_all$r %>%
  as.data.frame() %>%
  rownames_to_column("condition1") %>%
  pivot_longer(cols = -condition1,
               names_to = "condition2",
               values_to = "correlation")

# Convert p-value matrix to long format: each row is a pair (condition1, condition2) --------
cor_btm_samples_wide_all_p <- cor_btm_samples_wide_all$P %>%
  as.data.frame() %>%
  rownames_to_column("condition1") %>%
  pivot_longer(cols = -condition1,
               names_to = "condition2",
               values_to = "p_value")

# Join correlations, p-values, and sample annotations into a single table --------
# Annotate both condition1 and condition2 with their pathogen, organism, timepoint metadata
correlation_long_all_btm_mouse_human_timepoints_summary <- cor_btm_samples_wide_all_r %>%
  inner_join(cor_btm_samples_wide_all_p,
             by = join_by("condition1", "condition2")) %>%
  # Annotate condition1 with its sample metadata
  inner_join(btm_samples_annotation %>%
               rename_with(~ paste0(.x, "_condition1")) %>%
               rename(condition1 = condition_condition1),
             by = "condition1") %>%
  # Annotate condition2 with its sample metadata
  inner_join(btm_samples_annotation %>%
               rename_with(~ paste0(.x, "_condition2")) %>%
               rename(condition2 = condition_condition2),
             by = "condition2") %>%
  # Extract pathogen names from the condition string (pathogen is the first underscore-delimited token)
  mutate(pathogen1 = str_split_i(condition1, "_", 1),
         pathogen2 = str_split_i(condition2, "_", 1)) %>%
  # Remove symmetric duplicates: keep only one direction (condition1 < condition2) --------
  # and only compare samples within the same pathogen
  filter(condition1 < condition2,
         pathogen1 == pathogen2) %>%
  # Keep Human as the reference: drop rows where Mouse would appear as condition1 when Human is condition2 --------
  filter(!(organism_condition1 == "Mouse" & organism_condition2 == "Human")) %>%
  # Annotate comparison type and create unified timepoint and pathogen fields --------
  mutate(comparison = paste0(organism_condition1, " vs. ", organism_condition2),
         timepoint_comparison = paste0(timepoint_condition1, "-", timepoint_condition2),
         pathogen = pathogen1) %>%
  # Compute summary statistics per comparison and timepoint --------
  group_by(pathogen, comparison, timepoint_comparison) %>%
  mutate(
    mean = mean(correlation, na.rm = TRUE),
    sd   = sd(correlation, na.rm = TRUE),
    n = n(),
    se   = sd / sqrt(n),
    tval = qt(0.975, df = n - 1),
    ci_low  = mean - tval * se,
    ci_high = mean + tval * se
  ) %>%
  ungroup() %>%
  select(comparison, pathogen, timepoint_condition1, timepoint_condition2, mean:ci_high) %>%
  distinct() %>%
  arrange(comparison) %>%
  # Keep only comparisons where both conditions share the same timepoint label --------
  filter(timepoint_condition1 == timepoint_condition2) %>%
  mutate(timepoint = timepoint_condition1)

# Save the summarised correlation table for downstream plotting --------
saveRDS(correlation_long_all_btm_mouse_human_timepoints_summary,
        file = here("tables", paste0(filename, "_samples_btm_correlation_all_timepoints_summary.rds")))
correlation_long_all_btm_mouse_human_timepoints_summary

```

Plot scatterplot

```{r fig.height=3, fig.width=12}
# Data preparation: reload correlation summary and clean labels for publication-quality plotting --------
correlation_long_all_btm_mouse_human_timepoints_summary <-
  readRDS(file = here("tables", "all_human_mouse_samples_btm_correlation_all_timepoints_summary.rds"))

# Data preparation: rename pathogens, recode timepoints, and assign treatment categories --------
correlation_long_all_btm_mouse_human_timepoints_summary_plot <- correlation_long_all_btm_mouse_human_timepoints_summary %>%
  mutate(pathogen = case_when(pathogen == "Staphylococcus aureus" ~ "S. aureus",
                              pathogen == "Escherichia coli" ~ "E. coli",
                              TRUE ~ pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "S. aureus", "E. coli", "Burn", "Trauma"),
         timepoint = case_when(timepoint == "Early hours" ~ "Early\nhr",
                               TRUE ~ timepoint),
         timepoint = str_replace(timepoint, " ", "\n"),
         timepoint = fct_relevel(timepoint, c("Early hr", "2h", "4h", "6h", "12h", "24h", "Day\n1", "Day\n3", "Day\n7")),
         treatment = case_when(pathogen %in% c("Hepatitis B", "Influenza") ~ "Vaccination",
                               pathogen %in% c("S. aureus", "E. coli") ~ "Infection",
                               pathogen %in% c("Burn", "Trauma") ~ "Injury",
                               TRUE ~ pathogen),
         treatment = fct_relevel(treatment, "Vaccination", "Infection", "Injury"))

# Initialise ggplot with aesthetic mappings: x = timepoint, y = mean correlation, grouped and colored by comparison type --------
correlation_long_all_btm_mouse_human_timepoints_summary_plot <- correlation_long_all_btm_mouse_human_timepoints_summary_plot %>%
  ggplot() +
  aes(x = timepoint,
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +

  # Geom layers: line + error bars + points for the correlation-over-time trend --------
  geom_line(size = 1) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2, size = 0.5) +
  geom_point(pch = 21, fill = "white", size = 1.5, stroke = 1) +
  geom_hline(yintercept = 0, linetype = 2, color = "black") +

  # Facetting and scales: nested facets (treatment → pathogen) with free x-scales --------
  facet_nested_wrap(~ treatment + pathogen,
                  scales = "free_x",
                  nrow = 1) +
  scale_x_discrete(expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01), limits = c(-0.25, 1)) +
  scale_color_manual(values = colors$comparison_versus) +
  scale_fill_manual(values = colors$comparison_versus) +

  # Theme customization: project theme with nested facet line, top legend, compact spacing --------
  theme_vaxgo() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0, size = 10),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 10),
    strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
    panel.spacing = unit(2, "lines"),
    plot.margin = margin(0, 0.5, 0, 0.5, unit = "cm"),
    legend.position = "top",
    legend.direction = "horizontal",
    legend.margin = margin(0, 0, 0, 0),
    legend.box.margin = margin(0, 0, -10, 0),
    legend.text = element_text(size = 10, hjust = 0),
    legend.title = element_text(size = 10),
    ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
  ) +
  guides(fill = "none") +

  # Labels --------
  labs(x = "Time",
       y = "Correlation",
       color = "Comparison")

# Render plot --------
correlation_long_all_btm_mouse_human_timepoints_summary_plot

# Save --------
correlation_long_all_btm_mouse_human_timepoints_summary_plot %>%
  ggsave(filename = here("Figures", paste0("all_human_mouse_samples_btm_correlation_all_timepoints_summary_plot.png")),
         width = 10, height = 3)

```

#### Hallmarks: Summarize inter-individual correlation

Load the Hallmark gene annotation table and the per-sample log2FC
data, then compute intra-module mean log2FC for each Hallmark process.
Duplicate human infection timepoints to match mouse timepoints, then
pivot to wide format for cross-sample correlation analysis.

```{r}
# Import the Hallmark annotation table with gene-to-process mappings --------
msigdbr_grouped <- read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))

# Define the file prefix used for both input and output paths --------
filename <- "all_human_mouse"

# Load the per-sample log2FC matrix from the merged human-mouse DGE analysis --------
samples_log2fc <- readRDS(file = here("tables", "all_human_mouse_samples_log2fc.rds"))

# Load the sample metadata (organism, pathogen, timepoint, participant) --------
samples_annotation <- readRDS(file = here("tables", "all_human_mouse_metadata.rds"))

# Join log2FC values with sample metadata, select relevant columns, then join with Hallmarks --------
dge_btm_mean_process_samples <- samples_log2fc %>%
  # Attach sample-level metadata via geo_sample identifier --------
  inner_join(samples_annotation, by = "geo_sample") %>%
  ungroup() %>%
  # Keep only columns needed for downstream module-level aggregation --------
  select(symbol = human_symbol, participant_id, geo_sample, timepoint, log2fc, organism, pathogen, treatment) %>%
  # Map each gene to its Hallmark process via the annotation table --------
  inner_join(msigdbr_grouped, by = "symbol")

# Duplicate human infection timepoints to align with mouse timepoints --------
dge_btm_mean_process_samples_matched <-
  bind_rows(
  # Keep mouse rows as-is and human rows from non-infection conditions (vaccination/injury) --------
  dge_btm_mean_process_samples %>%
    filter(organism == "Mouse" | !(pathogen %in% c("Staphylococcus aureus", "Escherichia coli"))),
  # For human infection samples, create all combinations with available mouse timepoints --------
  dge_btm_mean_process_samples %>%
    filter(organism == "Human",
           pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
    # Remove human-specific timepoint to allow cross-product with mouse timepoints --------
    select(-timepoint) %>%
    inner_join(
      # Extract the set of mouse timepoints for each infection pathogen --------
      dge_btm_mean_process_samples %>%
        filter(organism == "Mouse",
               pathogen %in% c("Staphylococcus aureus", "Escherichia coli")) %>%
        select(pathogen, timepoint) %>%
        distinct(),
      by = "pathogen",
      relationship = "many-to-many"
    )
)

# Inspect the matched table: verify that human S. aureus 2h samples exist --------
dge_btm_mean_process_samples_matched %>%
  filter(pathogen == "Staphylococcus aureus",
         timepoint == "2h",
         organism == "Human")

#Save
# dge_btm_mean_process_samples %>% saveRDS(file = here("Tables", paste0(filename, "_dge_btm_samples_log2fc_long.rds")))

# Pivot wider: collapse gene-level log2FC to participant-level mean per Hallmark process --------
dge_btm_mean_process <- dge_btm_mean_process_samples_matched %>%
  # Create a unique condition identifier combining participant and timepoint --------
  mutate(condition = paste0(participant_id, "_Day", timepoint),
         timepoint = as.character(timepoint)) %>%
  # Group by pathogen, participant, timepoint, and Hallmark process --------
  group_by(pathogen, participant_id, timepoint, process) %>%
  # Summarise gene-level log2FC to the mean within each Hallmark process --------
  summarise(mean_log2fc = mean(log2fc, na.rm = TRUE)) %>%
  # Spread participant IDs as columns for wide-format correlation --------
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>%
  # Attach module group and subgroup metadata from the BTM annotation table --------
  inner_join(btm_annotation_genes %>%
               select(process, group, subgroup) %>%
               distinct(),
             by = "process") %>%
  ungroup()

# Inspect: count rows per pathogen, organism, and timepoint in the matched table --------
dge_btm_mean_process_samples_matched %>%
  dplyr::count(pathogen, organism, timepoint)

# Compute the mean log2FC per Hallmark process for each sample --------
dge_btm_mean_process_sample_long <- dge_btm_mean_process_samples_matched %>%
  # Group by pathogen, organism, participant, timepoint, and process --------
  group_by(pathogen, organism, participant_id, timepoint, process) %>%
  # Retain the group label and compute the mean log2FC per process per sample --------
  summarise(group,
            mean_log2fc = mean(log2fc, na.rm = TRUE)) %>%
  distinct() %>%
  ungroup()

# Verify sample counts per condition in the long-format table --------
dge_btm_mean_process_sample_long %>%
  dplyr::count(pathogen, organism, timepoint)

```

Compute pairwise Pearson correlations between all human and mouse
samples using Hallmark process-level mean log2FC. Reshape the correlation
matrix to long format, annotate comparison types, and summarise
correlation distributions per (pathogen, comparison, timepoint) group.

```{r}
# Build a long-format table of mean log2FC per sample and Hallmark process --------
btm_samples_long <- dge_btm_mean_process_sample_long %>%
  # Keep only the columns needed for the correlation matrix --------
  select(pathogen, organism, process, participant_id, timepoint, mean_log2fc) %>%
  distinct() %>%
  # Create a unique condition identifier embedding pathogen, participant, and timepoint --------
  mutate(condition = paste0(pathogen, "_", participant_id, "_", timepoint)) %>%
  drop_na()

# Build a distinct annotation table mapping each condition to its metadata --------
btm_samples_annotation <- btm_samples_long %>%
  select(pathogen, organism, participant_id, timepoint, condition) %>%
  distinct()

# Pivot the long table to wide format: rows = Hallmark processes, columns = conditions --------
btm_samples_wide_all <- btm_samples_long %>%
  # Remove metadata columns — only process and condition remain for the matrix --------
  select(-pathogen, -timepoint, -organism, -participant_id) %>%
  pivot_wider(names_from = "condition",
              values_from = "mean_log2fc") %>%
  column_to_rownames("process")

# Compute the full Pearson correlation matrix and associated p-values using Hmisc --------
cor_btm_samples_wide_all <- btm_samples_wide_all %>%
  as.matrix() %>%
  rcorr(type = "pearson")

# Convert the correlation coefficient matrix from wide to long format --------
cor_btm_samples_wide_all_r <- cor_btm_samples_wide_all$r %>%
  as.data.frame() %>%
  rownames_to_column("condition1") %>%
  pivot_longer(cols = -condition1,
               names_to = "condition2",
               values_to = "correlation")

# Convert the p-value matrix from wide to long format --------
cor_btm_samples_wide_all_p <- cor_btm_samples_wide_all$P %>%
  as.data.frame() %>%
  rownames_to_column("condition1") %>%
  pivot_longer(cols = -condition1,
               names_to = "condition2",
               values_to = "p_value")

# Join correlations with p-values and annotate both condition sides with sample metadata --------
correlation_long_all_btm_mouse_human_timepoints_summary <- cor_btm_samples_wide_all_r %>%
  # Attach the matching p-value for each condition pair --------
  inner_join(cor_btm_samples_wide_all_p,
             by = join_by("condition1", "condition2")) %>%
  # Attach sample metadata for condition1 (pathogen, organism, timepoint) --------
  inner_join(btm_samples_annotation %>%
               rename_with(~ paste0(.x, "_condition1")) %>%
               rename(condition1 = condition_condition1),
             by = "condition1") %>%
  # Attach sample metadata for condition2 --------
  inner_join(btm_samples_annotation %>%
               rename_with(~ paste0(.x, "_condition2")) %>%
               rename(condition2 = condition_condition2),
             by = "condition2") %>%
  # Extract the pathogen label from the condition string (first field before underscore) --------
  mutate(pathogen1 = str_split_i(condition1, "_", 1),
         pathogen2 = str_split_i(condition2, "_", 1)) %>%
  # Keep each unique pair only once (condition1 < condition2) and only within the same pathogen --------
  filter(condition1 < condition2,
         pathogen1 == pathogen2) %>%
  # Keep Human as the reference: drop rows where Mouse would appear as condition1 when Human is condition2 --------
  filter(!(organism_condition1 == "Mouse" & organism_condition2 == "Human")) %>%
  # Annotate comparison type and create unified timepoint and pathogen fields --------
  mutate(comparison = paste0(organism_condition1, " vs. ", organism_condition2),
         timepoint_comparison = paste0(timepoint_condition1, "-", timepoint_condition2),
         pathogen = pathogen1) %>%
  # Compute summary statistics (mean, sd, se, 95% CI) per comparison and timepoint --------
  group_by(pathogen, comparison, timepoint_comparison) %>%
  mutate(
    mean = mean(correlation, na.rm = TRUE),
    sd   = sd(correlation, na.rm = TRUE),
    n = n(),
    se   = sd / sqrt(n),
    tval = qt(0.975, df = n - 1),
    ci_low  = mean - tval * se,
    ci_high = mean + tval * se
  ) %>%
  ungroup() %>%
  select(comparison, pathogen, timepoint_condition1, timepoint_condition2, mean:ci_high) %>%
  distinct() %>%
  arrange(comparison) %>%
  # Keep only comparisons where both conditions share the same timepoint label --------
  filter(timepoint_condition1 == timepoint_condition2) %>%
  mutate(timepoint = timepoint_condition1)

# Save the summarised correlation table for downstream plotting --------
saveRDS(correlation_long_all_btm_mouse_human_timepoints_summary,
        file = here("tables", paste0(filename, "_samples_btm_correlation_all_timepoints_summary.rds")))
correlation_long_all_btm_mouse_human_timepoints_summary

```

Plot the Hallmark correlation summary as a multi-panel scatterplot
showing mean inter-individual correlation over time, with error bars
and faceted by pathogen.

```{r fig.height=2, fig.width=10}
# Data preparation: relevel pathogen order and recode timepoint labels for consistent plotting --------
correlation_long_all_btm_mouse_human_timepoints_summary_plot <- correlation_long_all_btm_mouse_human_timepoints_summary %>%
  mutate(pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma"),
         timepoint = case_when(timepoint == "Early hours" ~ "Early hr",
                               TRUE ~ timepoint),
         timepoint = fct_relevel(timepoint, c("Early hr", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")))

# Initialise ggplot with aesthetic mappings: x = timepoint, y = mean correlation, grouped and colored by comparison type --------
correlation_long_all_btm_mouse_human_timepoints_summary_plot <- correlation_long_all_btm_mouse_human_timepoints_summary_plot %>%
  ggplot() +
  aes(x = timepoint,
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +

  # Geom layers: line + error bars + points for the correlation-over-time trend --------
  geom_line(size = 1) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2, size = 0.5) +
  geom_point(pch = 21, fill = "white", size = 1.5, stroke = 1) +
  geom_hline(yintercept = 0, linetype = 2, color = "black") +

  # Facetting: one panel per pathogen with independent x-axes --------
  facet_wrap(~pathogen, scales = "free_x", space = "free_x", nrow = 1) +

  # Scales: discrete x, continuous y, manual color and fill palettes --------
  scale_x_discrete(expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01), limits = c(-0.2, 1)) +
  scale_color_manual(values = colors$comparison_versus) +
  scale_fill_manual(values = colors$comparison_versus) +

  # Labels --------
  labs(x = "Time",
       y = "Correlation",
       color = "Comparison") +

  # Theme customization: project theme with compact spacing, top legend, nested line --------
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0, size = 8),
        axis.text.y = element_text(size = 8),
        strip.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 8),
        panel.spacing = unit(2, "lines"),
        plot.margin = margin(0, 0.5, 0, 0.5, unit = "cm"),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, -10, 0),
        legend.text = element_text(size = 8, hjust = 0),
        legend.title = element_text(size = 8)) +
  guides(fill = "none")

# Render plot --------
correlation_long_all_btm_mouse_human_timepoints_summary_plot

# Save --------
correlation_long_all_btm_mouse_human_timepoints_summary_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_samples_hallmarks_correlation_all_timepoints_summary_plot.png")),
         width = 10, height = 2)

```

##-----------------------

## Correlation analysis 
### Hallmarks

Hallmarks correlation, intra- and inter-species Compare means and NES

Import

```{r}
#Import datasets ------
#Import DGE output
dge_long_genes_matched = readRDS(here("tables", "all_human_mouse_dge_limma_degs_matched.rds")) 
#Import Hallmarks annotation table
msigdbr_grouped = read_csv(file = here("tables", "msigdb_hallmarks_grouped_genes.csv"))


filename = "all_human_mouse"

#Clean metadata 
annotation_conditions = dge_long_genes_matched %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

```

Import means and NES

```{r fig.height=12, fig.width=10}
#Import ========================
all_human_mouse_metadata = readRDS(here("tables", "all_human_mouse_metadata.rds"))
#Unite Hallmarks means and NES
gsea_hallmarks_results_all = readRDS(here("tables", paste0("all_human_mouse_gsea_hallmarks_results.rds")))

gsea_hallmarks_results_all = gsea_hallmarks_results_all %>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint == "2h", "Early hours", timepoint)) %>% 
  inner_join(msigdbr_grouped %>% 
               select(process, group) %>% distinct(),
             by = "process") %>% 
  #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Influenza") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Ecoli", "Saureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_comparison = if_else(pathogen %in% c("Ecoli", "Saureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early"))



#Unite Hallmarks means
dge_hallmarks_mean_process_all = readRDS(file = here("tables", paste0("all_human_mouse_dge_hallmarks_mean_process.rds")))

dge_hallmarks_mean_process_all = dge_hallmarks_mean_process_all %>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint == "2h", "Early hours", timepoint)) %>% 
  inner_join(msigdbr_grouped %>% 
               select(process, group) %>% distinct(),
             by = "process")

#Unite DGE results annotated with Hallmarks
dge_hallmarks_process_genes_all = readRDS(file = here("tables", paste0("all_human_mouse_dge_hallmarks_process_genes.rds")))

dge_hallmarks_process_genes_all = dge_hallmarks_process_genes_all %>% 
  mutate(timepoint = if_else(pathogen %in% c("Burn", "Trauma") & timepoint == "2h", "Early hours", timepoint)) %>% 
   #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Influenza") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Ecoli", "Saureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_comparison = if_else(pathogen %in% c("Ecoli", "Saureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early"))

dge_hallmarks_process_genes_all

```

Import

```{r}
#Import results -------

filename = "all_human_mouse"

#Duplicate reference, if timepoints do not match =================
gsea_hallmarks_results_all_matched = gsea_hallmarks_results_all %>% 
  select(pathogen, organism, process, group, timepoint, nes, p_adjust, p_adjust_mean, mean_log2fc, core_enrichment, set_size) %>% 
  bind_rows(
  #Duplicate Human
  gsea_hallmarks_results_all %>% 
  filter(organism == "Human",
         pathogen %in% c("Saureus", "Ecoli")) %>% 
  select(pathogen, organism, process, group, nes, p_adjust, p_adjust_mean, mean_log2fc, core_enrichment, set_size) %>% 
  inner_join(gsea_hallmarks_results_all %>% 
               filter(organism == "Mouse",
                      pathogen %in% c("Saureus", "Ecoli")) %>% 
               select(pathogen, timepoint) %>% 
               distinct(),
             by = "pathogen")
  ) %>% 
  distinct()

dge_hallmarks_process_genes_matched = dge_hallmarks_process_genes_all %>% 
  select(pathogen, organism, symbol, process, group, timepoint, adj_p_val, mean_l2fc) %>% 
  bind_rows(
  #Duplicate Human
  dge_hallmarks_process_genes_all %>% 
  filter(organism == "Human",
         pathogen %in% c("Saureus", "Ecoli")) %>% 
  select(pathogen, symbol, organism, process, group, adj_p_val,mean_l2fc) %>% 
  inner_join(dge_hallmarks_process_genes_all %>% 
               filter(organism == "Mouse",
                      pathogen %in% c("Saureus", "Ecoli")) %>% 
               select(pathogen, timepoint) %>% 
               distinct(),
             by = "pathogen")
  ) %>% 
  distinct() 

gsea_hallmarks_results_all_matched %>% 
  ungroup() %>% 
  dplyr::count(pathogen, organism, timepoint)
 
dge_hallmarks_process_genes_matched %>% 
  ungroup() %>% 
  dplyr::count(pathogen, organism, timepoint)
```

```{r}
#Pivot to wide GSEA + mean
dge_hallmarks_gsea_mean_wide = gsea_hallmarks_results_all_matched %>% 
  select(organism, pathogen, timepoint, process, group, set_size, nes, mean_log2fc, p_adjust, p_adjust_mean) %>% 
  group_by(pathogen, process, group, set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, p_adjust_mean, set_size),
    names_sep = "_")

dge_hallmarks_gsea_mean_wide %>% 
  ungroup() %>% 
  dplyr::count(pathogen,timepoint)

```

#### Gene-level, Hallmarks

Plot genes by hallmarks group

```{r fig.height=16, fig.width=16}
#Convert to wide
dge_hallmarks_process_genes_wide = dge_hallmarks_process_genes_matched  %>% 
    ungroup() %>%
  select(pathogen, symbol, mean_l2fc, adj_p_val, organism, timepoint, process, group) %>% 
  distinct() %>% 
  group_by(pathogen, organism, timepoint, process) %>% 
  pivot_wider(names_from = organism,
              values_from = c(mean_l2fc, adj_p_val)) %>% 
  ungroup() %>% 
  mutate(group = fct_relevel(group, c(immune_order)))  %>% 
  #Remove duplicates
  select(-process) %>% 
  distinct()

#Convert to wide
dge_hallmarks_process_genes_wide =
  dge_hallmarks_process_genes_matched  %>% 
    ungroup() %>%
    select(symbol, mean_l2fc, adj_p_val, pathogen, organism, timepoint, process, group) %>% 
    distinct() %>% 
    group_by(pathogen, organism, timepoint) %>% 
    pivot_wider(
      names_from = organism,
      values_from = c(mean_l2fc, adj_p_val)
    ) %>% 
    ungroup() %>% 
    mutate(group = fct_relevel(group, c(immune_order))) %>% 
    distinct() %>%
    select(
      symbol,
      pathogen,
      process, 
      timepoint,
      Mouse = mean_l2fc_Mouse,
      Human = mean_l2fc_Human,
      adj_p_val_Mouse,
      adj_p_val_Human,
      group
    ) %>% 
  drop_na(Human, Mouse) 
  


safe_cor_test <- purrr::possibly(
  ~cor.test(.x, .y),
  otherwise = NULL
)

dge_hallmarks_process_genes_wide %>% 
  dplyr::count(pathogen, timepoint)

dge_hallmarks_process_genes_correlation = dge_hallmarks_process_genes_wide %>%
  group_by(pathogen, timepoint, process) %>%
  summarise(
    test_result = list(safe_cor_test(
      x = Human, y = Mouse, method = "pearson"
    )),
    .groups = "drop",
    n_genes = n(),
    group
  ) %>%

  mutate(test_result = purrr::map(
    test_result,
    ~ if (is.null(.x))
      list(estimate = NA_real_, p.value = NA_real_)
    else
      .x 
    )) %>%
  mutate(
    cor_value = purrr::map_dbl(test_result, ~ .x$estimate),
    p_value   = purrr::map_dbl(test_result, ~ .x$p.value)
  ) %>%
  select(-test_result) %>%
  # Labels
  mutate(
    r_squared = cor_value^2,
    p_label = case_when(
      p_value <= 0.01 ~ "***",
      p_value <= 0.05 ~ "**",
      p_value <= 0.10 ~ "*",
      TRUE            ~ "ns"
    ),
    cor_label = paste0(
      "r=",
      round(cor_value, 2),
      ", R²=",
      round(r_squared, 2),
      ", ",
      p_label
    )
  ) %>%
  arrange(timepoint) %>% 
  distinct()

dge_hallmarks_process_genes_correlation %>% 
  dplyr::count(pathogen, timepoint)

filter_1 = dge_hallmarks_process_genes_correlation %>% 
  group_by(pathogen, timepoint) %>% 
  arrange(process, 
          -abs(cor_value)) %>% 
  distinct() %>% 
  ungroup() %>% 
  select(-cor_label,  -p_value)

filter_1 %>% 
  group_by(pathogen, timepoint)  %>% 
  arrange() %>% 
  summarise(mean_cor = mean(cor_value, na.rm = T),
            median_cor = median(cor_value, na.rm = T),
            mean_rsq = mean(r_squared, na.rm = T))
```

Does the choice of using DEGs or all genes affect the correlation?

```{r fig.height=20, fig.width=20}
#Does the choice of using DEGs or all genes affect the correlation?
hallmarks_correlation_process_plot = filter_1 %>% 
  arrange(-cor_value) %>% 
  drop_na(cor_value) %>% 
  mutate(process = fct_reorder(process, cor_value)) %>% 
  ggplot() +
  aes(y = process, 
      x = cor_value,
      fill = group,
      group = process) +
  facet_nested_wrap(vars(pathogen, timepoint),
                    nrow = 1,
                    strip = strip_nested(size = "variable")) +
  geom_point(aes(size = n_genes),
             pch = 21,
             color = "black") +
  scale_fill_manual(values = hallmarks_colors) +
  scale_color_manual(values = hallmarks_colors) +
  geom_line() +
  geom_vline(xintercept = 0) +
  labs(fill = "Group",
       y = "",
       x = "Correlation",
       title = paste0("Hallmarks immune, ", filename)) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.text = element_text(size = 8, hjust = 0),
           ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
          )

hallmarks_correlation_process_plot

# hallmarks_correlation_process_plot %>% 
#   ggsave(filename = here("Figures", "all_human_mouse_gsea_hallmarks_results_intramodule_correlation_processes.png"),
#          width = 20,
#          height = 20)
```

Heatmap of intra-module correlation per Hallmark process across conditions.
Rows are Hallmark processes (ordered by mean correlation), columns are
timepoints, faceted by pathogen and functional group.

```{r fig.height=15, fig.width=10}
# Build heatmap of intra-module correlation for Hallmarks
hallmarks_correlation_process_heatmap <-
  filter_1 %>%
  drop_na(cor_value) %>%
  mutate(
    process = fct_reorder(process, cor_value, .fun = mean),
    timepoint = as.factor(timepoint)
  ) %>%
  ggplot(aes(x = timepoint, y = process, fill = cor_value)) +
  facet_grid(
    rows = vars(group),
    cols = vars(pathogen),
    scales = "free",
    space = "free"
  ) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "#669bbc", mid = "white", high = "#06d6a0",
    midpoint = 0, name = "Correlation"
  ) +
  labs(x = "", y = "",
       title = paste0("Hallmarks immune — intramodule correlation (", filename, ")")) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.text = element_text(size = 8),
        ggh4x.facet.nestline = element_line(colour = "black", linetype = 1))

# Render heatmap
hallmarks_correlation_process_heatmap

# Save heatmap
ggsave(hallmarks_correlation_process_heatmap,
       filename = here("Figures", "all_human_mouse_gsea_hallmarks_results_intramodule_correlation_processes_heatmap.png"),
       width = 10, height = 15)
```

Are the leading edges in mice the same in humans?

```{r}
gsea_hallmarks_results_core = gsea_hallmarks_results_all_matched %>% 
  separate_rows(core_enrichment, sep = "/") 

gsea_hallmarks_results_core_joined =
  gsea_hallmarks_results_core %>% 
  filter(organism == "Human") %>% 
  select(pathogen, process, timepoint,
         gene = core_enrichment,
         set_size_human = set_size) %>% 
  mutate(human_core = TRUE) %>% 
  full_join(
    gsea_hallmarks_results_core %>% 
      filter(organism == "Mouse") %>% 
      select(pathogen, process, timepoint,
             gene = core_enrichment,
             set_size_mouse = set_size) %>% 
      mutate(mouse_core = TRUE),
    by = c("pathogen", "process", "timepoint", "gene")
  ) %>% 
  mutate(
    status = case_when(
      human_core == TRUE & mouse_core == TRUE ~ "Shared",
      human_core == TRUE & is.na(mouse_core) ~ "Human only",
      mouse_core == TRUE & is.na(human_core) ~ "Mouse only",
      TRUE ~ "Other"
    )
  ) %>% 
  arrange(pathogen, timepoint, process, status)

gsea_hallmarks_results_core_joined %>% 
  dplyr::count(pathogen, timepoint)
```

```{r}
shared_perc = gsea_hallmarks_results_core_joined %>% 
  group_by(pathogen, timepoint, process, status) %>% 
  summarise(n = n(), .groups = "drop_last") %>% 
  mutate(percent = (n / sum(n)) * 100,
         percent = percent %>% round(0)) %>% 
  ungroup() 

#Which are the processes that share most of the core genes?
shared_perc %>% 
  filter(status == "Shared") %>% 
  arrange(-n, -percent)

#What is the mean percentage?
mean_percentage_shared = shared_perc %>% 
  group_by(pathogen, timepoint, status) %>% 
  summarise(mean_perc = mean(percent),
            median_perc = median(percent)) %>% 
  filter(status == "Shared")

mean_percentage_shared

mean_percentage_shared %>% 
  ungroup() %>% 
  summarise(mean_perc = mean(mean_perc))


shared_perc_df_plot = shared_perc %>% 
  inner_join(gsea_hallmarks_results_core %>% select(pathogen, timepoint, process, group),
             by = join_by("pathogen", "timepoint", "process")) %>% 
  filter(status == "Shared") %>% 
  inner_join(mean_percentage_shared %>% 
               filter(status == "Shared"), by = join_by(pathogen, timepoint, status)) %>% 
  mutate(process = fct_reorder(process, percent),
         facet_label = paste0(timepoint, ",\n Mean Shared = ", round(mean_perc, 1), "%")) %>% 
  arrange(-n, -percent) 

shared_perc_df_plot %>% 
  dplyr::count(pathogen, timepoint)
```

Bubble plot

```{r fig.height=6, fig.width=16}
#What does the choice of using DEGs or all genes affect the correlation and the number of genes?

hallmarks_correlation_setsize_plot = shared_perc_df_plot %>% 
  inner_join(filter_1 %>% select(pathogen, timepoint, n_genes, process, cor_value, r_squared, p_label),
             by = join_by(pathogen, timepoint, process)) %>% 
  distinct() %>% 
  ggplot() +
  aes(y = cor_value, 
      x = percent,
      fill = group,
      group = process) +
  facet_nested_wrap(vars(pathogen, timepoint),
                    scales = "free",
                    nrow = 2,
                    strip = strip_nested(size = "variable")) +
  geom_point(aes(size = n_genes),
             pch = 21,
             alpha = 1,
             # color = "black"
             ) +
  # geom_point(aes(size = n_genes),
  #            data = . %>% filter(process == "immune activation - generic cluster"),
  #            pch = 21,
  #            alpha = 1,
  #            color = "black") +
  
  geom_hline(yintercept = 0, linetype = 2) +
  scale_fill_manual(values = hallmarks_colors) +
  scale_color_manual(values = hallmarks_colors) +
  theme_vaxgo() +
  guides(color = F) +
  labs(fill = "Group",
       y = "Correlation",
       x = "Shared (%)",
       size = "Set size",
       title = paste0("Hallmarks, ", filename)) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.text = element_text(size = 8, hjust = 0),
           ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
          )

hallmarks_correlation_setsize_plot

hallmarks_correlation_setsize_plot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_gsea_hallmarks_results_intramodule_correlation_setsize_shared.png"),
         width = 16,
         height = 6)
  
```

```{r fig.height=20, fig.width=20}
shared_perc_plot =  shared_perc_df_plot %>% 
  
  ggplot() +
  aes(x = percent, 
      y = process,
      fill = group,
      size = n) +
  facet_nested_wrap(~pathogen + facet_label,
                    nrow = 1,
                    strip = strip_nested(size = "variable")) +
  geom_point(pch = 21)+
  theme_vaxgo() +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.text = element_text(size = 8, hjust = 0),
           ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
          ) +
  scale_fill_manual(values = hallmarks_colors) +
  scale_color_manual(values = hallmarks_colors)


shared_perc_plot

shared_perc_plot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_gsea_hallmarks_results_leadingedge_core_shared_perc.png"),
         width = 20,
         height = 20)
  
  
```

```{r fig.height=15, fig.width=10}
process_order = shared_perc_df_plot %>% 
  group_by(group, process) %>% 
  summarise(median = median(percent*n, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(process = as.character(process)) %>% 
  arrange(group, median)

shared_perc_plot =  shared_perc_df_plot %>% 
  mutate(process = fct_relevel(process, process_order$process),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Saureus", "Ecoli", "Burn", "Trauma"),
         ) %>% 
  ggplot() +
  aes(x = timepoint, 
      y = process,
      fill = percent,
      size = n) +
  facet_grid(
  rows = vars(group),
  cols = vars(pathogen),
  scales = "free",
  space  = "free"
) +
  geom_point(pch = 21)+
  scale_fill_stepsn(
  colours = c(
    "white",      # 0–25
    "#06d6a0",    # 25–50
    "#669bbc",    # 50–75
    "black"     # 75–100
  ),
  breaks = c(0, 25, 50, 75, 100),
  limits = c(0, 100),
  name = "Shared %"
)+
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          axis.text.y = element_text(size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.text = element_text(size = 8, hjust = 0),
           ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
          ) +
  labs(x = "Time",
       y = "")


shared_perc_plot

shared_perc_plot %>% 
  ggsave(filename = here("Figures", "all_human_mouse_gsea_hallmarks_results_leadingedge_core_shared_perc_heatmap.png"),
         width = 10,
         height = 15)
```

Do these leading edges have the same log2fc?

```{r}
#Do these leading edges have the same log2fc?

process_selected = "HEME_METABOLISM"

df_legs =
  dge_hallmarks_process_genes_wide %>% 
  filter(str_detect(process, process_selected)) %>% 
  
  # Join com os genes core
  full_join(
    gsea_hallmarks_results_core_joined %>% 
      filter(str_detect(process, process_selected)),
    by = join_by("pathogen", "process", "timepoint", symbol == gene)
  ) %>% 
  
  # join com shared %
  inner_join(
    shared_perc %>% 
      filter(str_detect(process, process_selected),
             status == "Shared") %>%
      rename(percent = percent,
             status_percent = status),
    by = join_by("pathogen", "process", "timepoint")
  ) %>%
  
  # join com NES
  inner_join(
    dge_hallmarks_gsea_mean_wide %>% 
      ungroup() %>% 
      filter(str_detect(process, process_selected)) %>% 
      select(process, pathogen, timepoint, nes_Mouse, nes_Human),
    by = join_by("pathogen", "process", "timepoint")
  ) %>% 
  
  mutate(
    nes_Mouse = round(nes_Mouse, 2),
    nes_Human = round(nes_Human, 2)
  ) %>% 
  mutate(status = if_else(is.na(status), "Not core", status)) %>% 
  inner_join(dge_hallmarks_process_genes_correlation %>% 
               select(timepoint, pathogen, process, cor_value, r_squared, p_label, cor_label),
             by = join_by("pathogen", "process", "timepoint")) %>% 
  mutate(
    facet_label = paste0(timepoint, ",\nShared = ", percent, "%,\n",
      "NES Mouse = ", nes_Mouse, ",\nNES Human = ", nes_Human)) %>% 
  arrange(timepoint) %>% 
  mutate(facet_label = fct_inorder(facet_label),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Saureus", "Ecoli", "Burn", "Trauma"),)

df_legs



```

Leading-edge visualisation: scatterplot, log2FC line plot, and rank line plot for
core genes vs non-core genes across Hallmark processes. The three panels
show the same data from complementary perspectives, followed by a
patchwork combining all views.

```{r fig.height=16, fig.width=20}
#Scatterplot: human vs mouse log2FC for each LEG, coloured by core status
df_test_scatter = df_legs %>% 
  ggplot() +
  aes(x = Mouse,
      y = Human,
      fill = status) +
  facet_nested_wrap(~process+pathogen + facet_label,
                    scales = "free",
                    nrow = 2,
                    strip = strip_nested(size = "variable")) +
  geom_point(size = 2, alpha = 0.5, fill = "white", pch = 21, stroke = 0.5, data = . %>% filter(status == "Not core")) +
  geom_point(size = 2, alpha = 1, pch = 21, color = "black", data = . %>% filter(status != "Not core")) +
  theme_vaxgo() +
    geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  scale_fill_manual(values = colors$comparison) +
  labs(fill = "Core",
       x = "Mouse (Log2FC)",
       y = "Human (Log2FC)") +
   guides(fill = guide_legend(override.aes = list(size = 4))) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          legend.position = "top",
          legend.direction = "horizontal",
          panel.spacing = unit(0, "lines"),
          legend.text = element_text(size = 8, hjust = 0),
           ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
          )

# df_test_scatter

# df_test_scatter %>% 
#   ggsave(filename = here("Figures", paste0("all_human_mouse_gsea_hallmarks_results_leadingedge_core_shared_scatterplot_", process_selected, ".png"),
#          width = 20,
#          height = 4)

#Positions using log2fc
df_test_log2fc = df_legs %>% 
  pivot_longer(names_to = "organism",
               values_to = "log2fc",
               cols = c("Mouse", "Human")) %>% 
  mutate(
    facet_label = paste0(timepoint, ",\n", cor_label, ",\nShared = ", percent, "%,\n",
      "NES Mouse = ", nes_Mouse, ",\nNES Human = ", nes_Human)) %>% 
   arrange(timepoint) %>% 
  mutate(facet_label = fct_inorder(facet_label),
    organism = if_else(organism == "Human", "H", "M")) %>% 
  ggplot() +
  aes(x = organism,
      y = log2fc,
      fill = status,
      group = symbol,
      color = status) +
  facet_nested_wrap(~process + pathogen + facet_label,
                    scales = "free",
                    nrow = 2,
                    strip = strip_nested(size = "variable")) +
  geom_line(size = 0.5, alpha = 0.5, data = . %>% filter(status == "Not core")) +
  geom_line(size = 0.5, alpha = 1, data = . %>% filter(status != "Not core")) +
  geom_point(size = 1.5, alpha = 0.5, pch = "l", data = . %>% filter(status == "Not core")) +
  geom_point(size = 1.5, alpha = 1, pch = "l", data = . %>% filter(status != "Not core")) +
  theme_vaxgo() +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2, size = 1, color = "black") +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(fill = "Core",
       color = "Core",
       y = "Log2fc",
       x = "") +
  scale_x_discrete(expand = c(0.01, 0.01)) +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          panel.spacing = unit(0, "lines"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.text = element_text(size = 8, hjust = 0),
           ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
          ) 


# df_test_log2fc

# df_test_log2fc %>% 
#   ggsave(filename = here("Figures", paste0("all_human_mouse_gsea_hallmarks_results_leadingedge_core_log2fc_lineplot_", process_selected, ".png"),
#          width = 20,
#          height = 4)

#Positions using ranks
df_legs_ranked <- df_legs %>%
  pivot_longer(
    names_to = "organism",
    values_to = "log2fc",
    cols = c("Mouse", "Human")
  ) %>%
  group_by(pathogen, organism) %>%
  mutate(rank = rank(-log2fc, ties.method = "first")) %>%   
  ungroup() %>% 
  arrange(rank) %>% 
  drop_na(log2fc) %>% 
  mutate(
    facet_label = paste0(timepoint, ",\n", cor_label, ",\nShared = ", percent, "%,\n",
      "NES Mouse = ", nes_Mouse, ",\nNES Human = ", nes_Human
    )) %>% 
    arrange(timepoint) %>% 
  mutate(facet_label = fct_inorder(facet_label),
    organism = if_else(organism == "Human", "H", "M")
    )

df_legs_ranked_plot = df_legs_ranked %>% 
  ggplot() +
  aes(x = organism,
      y = rank,
      fill = status,
      group = symbol,
      color = status) +
  facet_nested_wrap(~process + pathogen + facet_label,
                    scales = "free",
                    nrow = 2,
                    strip = strip_nested(size = "variable")) +
  geom_line(size = 0.5, alpha = 0.5, data = . %>% filter(status == "Not core")) +
  geom_line(size = 0.5,alpha = 1,  data = . %>% filter(status != "Not core")) +
  geom_point(size = 3, alpha = 0.5, pch = "l", data = . %>% filter(status == "Not core")) +
  geom_point(size = 3, alpha = 1, pch = "l", data = . %>% filter(status != "Not core")) +
  theme_vaxgo() +
  scale_fill_manual(values = colors$comparison) +
  scale_color_manual(values = colors$comparison) +
  labs(fill = "Core",
       color = "Core",
       y = "Rank",
       x = "") +
  scale_x_discrete(expand = c(0.01, 0.01)) +
  scale_y_reverse() +
  guides(color = guide_legend(override.aes = list(linewidth = 3))) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0,  hjust = 0.5, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.x = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5, size = 10),
          panel.spacing = unit(0, "lines"),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.text = element_text(size = 8, hjust = 0),
           ggh4x.facet.nestline = element_line(colour = "black", linetype = 1)
          ) 

# df_legs_ranked_plot

# df_legs_ranked_plot %>% 
# ggsave(filename = here("Figures", paste0("all_human_mouse_gsea_hallmarks_results_leadingedge_core_ranks_lineplot.png", process_selected)),
#          width = 20,
#          height = 4)


leading_edges_plots = (
  (df_test_scatter + theme(legend.position = "top")) / 
  plot_spacer() / 
  (df_test_log2fc + theme(legend.position = "none")) / 
  plot_spacer() / 
  (df_legs_ranked_plot + theme(legend.position = "none"))
) +
plot_layout(
  guides = "collect",
  heights = c(0.8, 0, 1, 0, 1) 
) +
plot_annotation(
  theme = theme(
    legend.position = "top",
    plot.margin = margin(1, 1,1,1, "pt")
  )
) & 
theme(
  panel.spacing = unit(0, "lines"),
  plot.margin = margin(0, 2, 0, 2, "pt") 
)
leading_edges_plots

leading_edges_plots %>%
  ggsave(filename = here("Figures", paste0("all_human_mouse_gsea_hallmarks_results_leadingedge_core_", process_selected, ".png")),
         width = 20,
         height = 16)
```

```{r fig.height=12, fig.width=20}
#Shortened version
leading_edges_plots_short = (
  (df_test_scatter + theme(legend.position = "top")) / 
  plot_spacer() / 
  (df_legs_ranked_plot + theme(legend.position = "none"))
) +
plot_layout(
  guides = "collect",
  heights = c(1, -0.2, 1) 
) +
plot_annotation(
  theme = theme(
    legend.position = "top",
    plot.margin = margin(1, 1,1,1, "pt")
  )
) & 
theme(
  panel.spacing = unit(0, "lines"),
  # plot.margin = margin(0, 2, 0, 2, "pt"),
   legend.margin=margin(0,0,0,0),
   legend.box.margin=margin(1,0,-10,0),
   plot.margin = margin(0.5, 0.5, 0.5, 0.5, unit = "cm")
)
leading_edges_plots_short

leading_edges_plots_short %>%
  ggsave(filename = here("Figures", paste0("all_human_mouse_gsea_hallmarks_results_leadingedge_core_short_", process_selected, ".png")),
         width = 20,
         height = 12)


```

####Module-level, Hallmarks

Plot correlation

```{r fig.height=5, fig.width=7}
#Mouse vs Human
#Pivot to wide GSEA + mean
dge_hallmarks_gsea_mean_wide = gsea_hallmarks_results_all_matched %>% 
  select(pathogen, organism, timepoint, process, group, nes, mean_log2fc, p_adjust, p_adjust_mean) %>% 
  group_by(pathogen, timepoint, process, group) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, p_adjust_mean),
    names_sep = "_") %>% 
  ungroup() %>% 
  # mutate(timepoint = as.numeric(timepoint)) %>% 
  drop_na(mean_log2fc_Mouse)

dge_hallmarks_gsea_mean_wide %>% 
  dplyr::count(pathogen, timepoint)


dge_hallmarks_process_genes_wide = dge_hallmarks_process_genes_matched  %>% 
    ungroup() %>%
  select(symbol, mean_l2fc, adj_p_val, organism, timepoint, process, group) %>% 
  distinct() %>% 
  group_by(organism, timepoint, process) %>% 
  pivot_wider(names_from = organism,
              values_from = c(mean_l2fc, adj_p_val)) %>% 
  ungroup() %>% 
  drop_na(mean_l2fc_Human) %>% 
  mutate(timepoint = as.numeric(timepoint))


```

```{r fig.height=5, fig.width=7}
#Calculate correlation -----
#Safe correlation 
safe_cor_test <- function(x, y, method) {
  # Remove NAs and infinite values first
  ok <- is.finite(x) & is.finite(y)
  x <- x[ok]
  y <- y[ok]
  
  # cor.test requires n >= 3 to run without error
  if (length(x) < 3 || sd(x) == 0 || sd(y) == 0) {
    return(list(estimate = NA_real_, p.value = NA_real_))
  } else {
    tryCatch({
      ct <- cor.test(x, y, method = method)
      return(list(estimate = unname(ct$estimate), p.value = ct$p.value))
    }, error = function(e) {
      return(list(estimate = NA_real_, p.value = NA_real_))
    })
  }
}
# Means vs means
cor_labels <- dge_hallmarks_gsea_mean_wide %>%
  select(pathogen, process, timepoint, group,
         Mouse = mean_log2fc_Mouse,
         Human = mean_log2fc_Human) %>%
  group_by(pathogen, timepoint) %>%
  summarise(
    test_result = list(safe_cor_test(Human, Mouse, method =  "pearson")),
    .groups = "drop"
  ) %>%
  mutate(
    cor_value = purrr::map_dbl(test_result, "estimate"),
    p_value   = purrr::map_dbl(test_result, "p.value")
  ) %>%
  select(-test_result) %>%
  mutate(
    filter = "all",
    r_squared = cor_value^2,
    p_label = case_when(
      p_value <= 0.01 ~ "***",
      p_value <= 0.05 ~ "**",
      p_value <= 0.10 ~ "*",
      TRUE            ~ "ns"
    ),
    cor_label = paste0(
      "r=", round(cor_value, 2),
      ", R²=", round(r_squared, 2),
      p_label
    )
  )

cor_labels_filter_Mouse <- dge_hallmarks_gsea_mean_wide %>%
  filter(p_adjust_mean_Mouse <= 0.25) %>% 
  select(pathogen, process, timepoint, group,
         Mouse = mean_log2fc_Mouse,
         Human = mean_log2fc_Human) %>%
  group_by(pathogen, timepoint) %>%
  summarise(
    test_result = list(safe_cor_test(Human, Mouse, method = "pearson")),
    .groups = "drop"
  ) %>%
  mutate(
    cor_value_mouse = purrr::map_dbl(test_result, "estimate"),
    p_value_mouse   = purrr::map_dbl(test_result, "p.value")
  ) %>%
  select(-test_result) %>%
  mutate(
    r_squared_mouse = cor_value_mouse^2,
    p_label_mouse = case_when(
      p_value_mouse <= 0.01 ~ "***",
      p_value_mouse <= 0.05 ~ "**",
      p_value_mouse <= 0.10 ~ "*",
      TRUE            ~ "ns"
    ),
    cor_label_mouse = paste0(
      "r=", round(cor_value_mouse, 2),
      ", R²=", round(r_squared_mouse, 2),
      p_label_mouse
    )
  )

cor_labels_filter_Mouse_Human <- dge_hallmarks_gsea_mean_wide %>%
  filter(p_adjust_mean_Mouse <= 0.25, p_adjust_mean_Human <= 0.25) %>% 
  select(pathogen, process, timepoint, group,
         Mouse = mean_log2fc_Mouse,
         Human = mean_log2fc_Human) %>%
  group_by(pathogen, timepoint) %>%
  summarise(
    test_result = list(safe_cor_test(Human, Mouse, method = "pearson")),
    .groups = "drop"
  ) %>%
  mutate(
    cor_value_mouse_human = purrr::map_dbl(test_result, "estimate"),
    p_value_mouse_human   = purrr::map_dbl(test_result, "p.value")
  ) %>%
  select(-test_result) %>%
  mutate(
    r_squared_mouse_human = cor_value_mouse_human^2,
    p_label_mouse_human = case_when(
      p_value_mouse_human <= 0.01 ~ "***",
      p_value_mouse_human <= 0.05 ~ "**",
      p_value_mouse_human <= 0.10 ~ "*",
      TRUE            ~ "ns"
    ),
    cor_label_mouse_human = paste0(
      "r=", round(cor_value_mouse_human, 2),
      ", R²=", round(r_squared_mouse_human, 2),
      p_label_mouse_human
    )
  )


# NES vs NES
# Means vs means
cor_labels_nes <- dge_hallmarks_gsea_mean_wide %>%
  select(pathogen, process, timepoint, group,
         Mouse = nes_Mouse,
         Human = nes_Human) %>%
  group_by(pathogen, timepoint) %>%
  summarise(
    test_result = list(safe_cor_test(Human, Mouse, method =  "pearson")),
    .groups = "drop"
  ) %>%
  mutate(
    cor_value = purrr::map_dbl(test_result, "estimate"),
    p_value   = purrr::map_dbl(test_result, "p.value")
  ) %>%
  select(-test_result) %>%
  mutate(
    filter = "all",
    r_squared = cor_value^2,
    p_label = case_when(
      p_value <= 0.01 ~ "***",
      p_value <= 0.05 ~ "**",
      p_value <= 0.10 ~ "*",
      TRUE            ~ "ns"
    ),
    cor_label = paste0(
      "r=", round(cor_value, 2),
      ", R²=", round(r_squared, 2),
      p_label
    )
  )

cor_labels_nes_filter_Mouse <- dge_hallmarks_gsea_mean_wide %>%
  filter(p_adjust_Mouse <= 0.25) %>% 
  select(pathogen, process, timepoint, group,
         Mouse = nes_Mouse,
         Human = nes_Human) %>%
  group_by(pathogen, timepoint) %>%
  summarise(
    test_result = list(safe_cor_test(Human, Mouse, method = "pearson")),
    .groups = "drop"
  ) %>%
  mutate(
    cor_value_mouse = purrr::map_dbl(test_result, "estimate"),
    p_value_mouse   = purrr::map_dbl(test_result, "p.value")
  ) %>%
  select(-test_result) %>%
  mutate(
    r_squared_mouse = cor_value_mouse^2,
    p_label_mouse = case_when(
      p_value_mouse <= 0.01 ~ "***",
      p_value_mouse <= 0.05 ~ "**",
      p_value_mouse <= 0.10 ~ "*",
      TRUE            ~ "ns"
    ),
    cor_label_mouse = paste0(
      "r=", round(cor_value_mouse, 2),
      ", R²=", round(r_squared_mouse, 2),
      p_label_mouse
    )
  )

cor_labels_nes_filter_Mouse_Human <- dge_hallmarks_gsea_mean_wide %>%
  filter(p_adjust_Mouse <= 0.25, p_adjust_Human <= 0.25) %>% 
  select(pathogen, process, timepoint, group,
         Mouse = nes_Mouse,
         Human = nes_Human) %>%
  group_by(pathogen, timepoint) %>%
  summarise(
    test_result = list(safe_cor_test(Human, Mouse, method = "pearson")),
    .groups = "drop"
  ) %>%
  mutate(
    cor_value_mouse_human = purrr::map_dbl(test_result, "estimate"),
    p_value_mouse_human   = purrr::map_dbl(test_result, "p.value")
  ) %>%
  select(-test_result) %>%
  mutate(
    r_squared_mouse_human = cor_value_mouse_human^2,
    p_label_mouse_human = case_when(
      p_value_mouse_human <= 0.01 ~ "***",
      p_value_mouse_human <= 0.05 ~ "**",
      p_value_mouse_human <= 0.10 ~ "*",
      TRUE            ~ "ns"
    ),
    cor_label_mouse_human = paste0(
      "r=", round(cor_value_mouse_human, 2),
      ", R²=", round(r_squared_mouse_human, 2),
      p_label_mouse_human
    )
  )



dge_hallmarks_mean_group_correlation_df = dge_hallmarks_gsea_mean_wide %>% 
  ungroup() %>% 
  mutate(process = str_wrap(process, width = 15)) %>% 
  drop_na(mean_log2fc_Mouse) %>% 
  inner_join(cor_labels, by = join_by(pathogen, timepoint)) %>% 
  inner_join(cor_labels_filter_Mouse, by = join_by(pathogen, timepoint)) %>% 
  inner_join(cor_labels_filter_Mouse_Human, by = join_by(pathogen, timepoint)) %>% 
  mutate(cor_labels_all_filtered = paste0("All: ", cor_label, ",\nM: ", cor_label_mouse, ",\nM&H: ", cor_label_mouse_human)) %>% 
  #Annotate timepoints
  mutate( timepoint_comparison = if_else(pathogen %in% c("Ecoli", "Saureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early"),
         
         #Create facet names
         facet_name = paste0(timepoint_comparison, "\n", cor_labels_all_filtered)) %>% 
  arrange(timepoint) %>% 
  mutate(
         facet_name = fct_inorder(facet_name),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Saureus", "Ecoli", "Burn", "Trauma"))


dge_hallmarks_mean_group_correlation_df %>% 
  dplyr::count(pathogen, timepoint, timepoint_comparison)

```

##### Plot means

Faceted correlation plot of Hallmark module means showing Human vs Mouse,
Human vs FIT, and Mouse vs FIT comparisons across all conditions.

```{r fig.height=9, fig.width=10, message=FALSE, warning=FALSE}

#Plot correlation plot (means)
dge_hallmarks_mean_group_correlation_plot = 
  
  bind_rows(
    dge_hallmarks_mean_group_correlation_df %>% 
      mutate(filter = "All"),
    
    dge_hallmarks_mean_group_correlation_df %>% 
      filter(p_adjust_mean_Mouse <= 0.25, p_adjust_mean_Human >0.25) %>% 
      mutate(filter = "Mouse"),
    
    dge_hallmarks_mean_group_correlation_df %>% 
      filter(p_adjust_mean_Mouse <= 0.25 & p_adjust_mean_Human <= 0.25) %>% 
      mutate(filter = "Mouse and Human"),
  ) %>% 
  ggplot() +
  aes(x = mean_log2fc_Mouse,
      y = mean_log2fc_Human) +
  facet_nested_wrap(vars(pathogen, facet_name),
                    scales = "free",
                    nrow = 4,
                    strip = strip_nested(size = "variable")) +
  
    #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  
#Plot all points
  with_blur(
    geom_point(aes(fill = group,
                 color = group,
                 shape = filter),
                 data = . %>% filter(filter == c("All")),
             size = 1.5,
             alpha = 0.2,
    sigma = 2)
    ) +
  #Filtered
  geom_point(aes(fill = group,
                 shape = filter,
                 color = group),
             data = . %>% filter(filter == c("Mouse")),
             size = 1.5,
             alpha = 1
             ) +
  geom_point(aes(fill = group,
                 shape = filter),
             color = "black",
             data = . %>% filter(filter == c("Mouse and Human")),
             size = 1.5,
             stroke = 0.5,
             alpha = 1,
             ) +
  scale_shape_manual(values = c(21, 22, 24)) +
  scale_fill_manual(values = hallmarks_colors) +
  scale_color_manual(values = hallmarks_colors) +
  guides(color = F,
         fill = guide_legend(byrow = TRUE,  nrow = 3,
                           keywidth = unit(0.3, "cm"),
                           keyheight = unit(0.3, "cm"),
                           override.aes = list(size = 3, shape = 21, color = "black"),
                           default.unit="cm"),
         shape = guide_legend(byrow = TRUE,  
                              nrow = 3,
                             keywidth = unit(0.3, "cm"),
                           keyheight = unit(0.3, "cm"),
                          override.aes = list(size = 3),
                           default.unit="cm")) +
  labs(fill = "Group",
       shape = "Significance",
       y = "Human (Mean Log2FC)",
       x = "Model (Mean Log2FC)") +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 8),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 8, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing = unit(0, "lines")
          )

dge_hallmarks_mean_group_correlation_plot

#Save results
dge_hallmarks_mean_group_correlation_plot %>%
  ggsave(filename = here("Figures",  "all_human_mouse_dge_hallmarks_mean_group_correlation_plot.png"),
         width = 10,
         height = 9)
```

##### Plot NES

```{r fig.height=8, fig.width=8}
dge_hallmarks_nes_group_correlation_df = dge_hallmarks_gsea_mean_wide %>% 
  ungroup() %>% 
  mutate(process = str_wrap(process, width = 15)) %>% 
  drop_na(mean_log2fc_Mouse) %>% 
  inner_join(cor_labels_nes, by = join_by(pathogen, timepoint)) %>% 
  inner_join(cor_labels_nes_filter_Mouse, by = join_by(pathogen, timepoint)) %>% 
  inner_join(cor_labels_nes_filter_Mouse_Human, by = join_by(pathogen, timepoint)) %>% 
  mutate(cor_labels_all_filtered = paste0("All: ", cor_label, ",\nM: ", cor_label_mouse, ",\nM&H: ", cor_label_mouse_human)) %>% 
  #Annotate timepoints
  mutate(timepoint = case_when(pathogen %in% c("Burn", "Trauma") & timepoint %in% c("2h", "2", "Early hours")~"Early hours", 
                               pathogen %in% c("Influenza", "Hepatitis B") ~ paste0("Day ", timepoint),
                               pathogen %in% c("Ecoli", "Saureus") ~ paste0(timepoint, "h"),
                               T ~timepoint),
         timepoint = fct_relevel(timepoint, c("Early hours", "2h", "4h", "6h", "12h", "24h", "Day 1", "Day 3", "Day 7")),
         timepoint_comparison = if_else(pathogen %in% c("Escherichia coli", "Staphylococcus aureus"),
                                        paste0(timepoint, "-", "24h"), 
                                        paste0(timepoint, "-", timepoint)) %>% 
                                str_replace(., "Early hours-Early hours", "2h-Early"),
         
         #Create facet names
         facet_name = paste0(timepoint_comparison, "\n", cor_labels_all_filtered)) %>% 
  arrange(timepoint) %>% 
  mutate(
         facet_name = fct_inorder(facet_name),
         pathogen = case_when(pathogen == "Saureus" ~ "Staphylococcus aureus",
                              pathogen == "Ecoli" ~ "Escherichia coli",
                              T ~pathogen),
         pathogen = fct_relevel(pathogen, "Influenza", "Hepatitis B", "Staphylococcus aureus", "Escherichia coli", "Burn", "Trauma"))

#Plot correlation plot (NES)
dge_hallmarks_nes_group_correlation_plot =  
  bind_rows(
    dge_hallmarks_nes_group_correlation_df %>% 
      mutate(filter = "All"),
    
    dge_hallmarks_nes_group_correlation_df %>% 
      filter(p_adjust_Mouse <= 0.25, p_adjust_Human >0.25) %>% 
      mutate(filter = "Mouse"),
    
    dge_hallmarks_nes_group_correlation_df %>% 
      filter(p_adjust_Mouse <= 0.25 & p_adjust_Human <= 0.25) %>% 
      mutate(filter = "Mouse and Human"),
  ) %>% 
  ggplot() +
  aes(x = nes_Mouse,
      y = nes_Human,
      labels = process) +
  facet_nested_wrap(vars(pathogen, facet_name),
                    scales = "free",
                    nrow = 4,
                    strip = strip_nested(size = "variable")) +
    #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  
#Plot all points
  with_blur(
    geom_point(aes(fill = group,
                 color = group,
                 shape = filter),
                 data = . %>% filter(filter == c("All")),
             size = 1.5,
             alpha = 0.2,
    sigma = 2)
    ) +
  #Filtered
  geom_point(aes(fill = group,
                 shape = filter,
                 color = group),
             data = . %>% filter(filter == c("Mouse")),
             size = 1.5,
             alpha = 1
             ) +
  geom_point(aes(fill = group,
                 shape = filter),
             color = "black",
             data = . %>% filter(filter == c("Mouse and Human")),
             size = 1.5,
             stroke = 0.5,
             alpha = 1,
             ) +
  scale_shape_manual(values = c(21, 22, 24)) +
  scale_fill_manual(values = hallmarks_colors) +
  scale_color_manual(values = hallmarks_colors) +
  guides(color = F,
         fill = guide_legend(byrow = TRUE,  nrow = 3,
                           keywidth = unit(0.3, "cm"),
                           keyheight = unit(0.3, "cm"),
                           override.aes = list(size = 3, shape = 21, color = "black"),
                           default.unit="cm"),
         shape = guide_legend(byrow = TRUE,  
                              nrow = 3,
                             keywidth = unit(0.3, "cm"),
                           keyheight = unit(0.3, "cm"),
                          override.aes = list(size = 3),
                           default.unit="cm")) +
  labs(fill = "Group",
       shape = "Significance",
       y = "Human (NES)",
       x = "Model (NES)") +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 8),
          axis.text.y = element_text(size = 8),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          strip.text.x = element_text(angle = 0, hjust = 0., vjust = 0.5, size = 8),
          legend.position = "top",
          legend.direction = "horizontal", 
          legend.text = element_text(size = 6, hjust = 0),
          legend.title = element_text(size = 8),
          ggh4x.facet.nestline = element_line(colour = "black", linetype = 1),
          legend.box.spacing = unit(0, "cm"),
          panel.spacing = unit(0, "lines")
          )

dge_hallmarks_nes_group_correlation_plot

#Save results
dge_hallmarks_nes_group_correlation_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_dge_hallmarks_nes_group_correlation_plot.png")),
         width = 10,
         height = 9)
```


#=================== 
#TO DO

## ORA of best and worst quartiles

```{r fig.width=10, fig.height= 8}
# Enrichment analysis -------
organism = "org.Hs.eg.db"
#Hallmarks
#Retrieve human H (hallmark) gene set
msigdbr_raw <- msigdbr(species = "human", category = "H") 
#Clean
msigdbr_df = msigdbr_raw %>% 
  select(process = gs_name, genes = gene_symbol) %>% 
  mutate(process = str_remove(process, "HALLMARK_"))

#Discretize by quartile
df_genes_best_worst_fit_mouse = zscore_genes %>% 
  ungroup() %>% 
  mutate(quartiles_grouped = case_when(
    quartile_label %in% c("Q10", "Q25") ~ "<Q25",
    quartile_label %in% c("Q75", "Q100") ~ ">Q75",
    TRUE ~ quartile_label)) %>% 
    mutate(condition = paste0(comparison, "", quartiles_grouped, " (", timepoint_day,")")) %>% 
  filter(quartiles_grouped != "Q50") %>% 
  select(genes = human_symbol,
         condition)

#Run ORA Hallmarks
ora_all_diff_hallmark <- enricher_condition(
  df = df_genes_best_worst_fit_mouse,
  term2gene_df = msigdbr_df) %>% 
  mutate(log_p_adjust = -log(p_adjust)) %>% 
  mutate(description = str_replace_all(description, "_", " ")) %>% 
  mutate(timepoint = str_extract(condition, "(?<=\\(Day )\\d+")) %>% 
  mutate(comparison = case_when(str_detect(condition, "FIT") ~ "FIT",
                                str_detect(condition, "Mouse") ~ "Mouse"))
#Save
ora_all_diff_hallmark %>% 
  saveRDS(file = here("tables", "ora_all_diff_hallmark.rds"))

#Print
ora_all_diff_hallmark
```


```{r fig.height=8, fig.width=8}
##### Plot ORA Hallmark ------
#Import
ora_all_diff_hallmark = readRDS(file = here("tables", "ora_all_diff_hallmark.rds"))
#Cluster rows
row_clustered = ora_all_diff_hallmark %>% 
    dplyr::select(condition, description, log_p_adjust) %>% 
    distinct() %>% 
    pivot_wider(names_from = "condition",
                values_from= "log_p_adjust") %>%
    column_to_rownames("description") %>% 
    as.matrix() %>% 
    replace(is.na(.), 0) %>% 
    dist(.) %>% 
    hclust()

#Plot
ora_all_diff_hallmark_heatmap = ora_all_diff_hallmark %>% 
  clean_names() %>% 
  filter(description %in% c(row_clustered$labels)) %>% 
  ungroup() %>% 
  select(condition, description, log_p_adjust, p_adjust, everything()) %>% 
    mutate(signific = case_when(p_adjust <= 0.10 ~ "*",
                              p_adjust <= 0.25 ~ "-",
                              p_adjust <= 0.50 ~ "/",
                              p_adjust <= 0.05 ~ "**",
                              TRUE ~ ""),
           quantile = str_extract(condition, "Q\\d{2,3}"),
        condition = paste(quantile)) %>% 
  ggplot(aes(x=condition, y = str_to_upper(description), fill = log_p_adjust)) +
  geom_tile(color = "black", 
            size = 0.2, na.rm = F, stat = "identity") +
  geom_text(aes(label = signific), size = 4, vjust = 0.5, color = "black") +
  scale_fill_gradient2(
        low = "white",     
        high = "#4DBBD5FF",   
        n.breaks = 3,
        oob = scales::squish,
        limits = c(0, 4),      # fixa o intervalo
        name = "-Log(FDR)") +
    theme_vaxgo() +
    theme(axis.text = element_text(color = "black", size = 8) ,
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 8),
          legend.key.width = unit(0.2, 'cm'),
          legend.key.height = unit(0.4, 'cm'),
          axis.line.y  = element_blank(),
          axis.text.x = element_text(angle = 0, vjust = 1, hjust=0.5),
          axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1),
          axis.ticks = element_blank(),
          strip.background.x = element_rect(fill = "black"),
          strip.text.x = element_text(color = "white"),
          panel.grid = element_blank(),
          plot.margin = margin(0, 0.5, 0, 1, unit = "cm"),
          legend.direction = "horizontal",
          legend.position = "top") +
    scale_y_discrete(position = "right") +
     labs(x = "",
         y = "",
         title = "ORA, MSigDB Hallmarks") +
  facet_nested(~ paste("Day", timepoint) + fct_rev(comparison), scales = "free", space = "free") 

#Print
ora_all_diff_hallmark_heatmap

#Save
ora_all_diff_hallmark_heatmap %>%
  ggsave(filename = here("Figures", "ORA_Best_Worst_heatmap.png"),
         width = 8,
         height = 8)
```

#### Top best and worst

```{r}
#Import data
pairwise_human_fit_measures = readRDS(here("tables", "influenza_fluad_human_mouse_fit_dge_pairwise2human.rds")) 
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
# Top best and worst predicted genes in FIT and Mouse -----
pairwise_human_fit_genes_long =  pairwise_human_fit_measures %>% 
  pivot_longer(cols = c(mean_human, mean_mouse, mean_fit),
               values_to = "l2fc",
               names_to = "organism") %>% 
  rename(diff_h2m = human_2_mouse,
         diff_h2fit = human_2_fit) %>% 
  mutate(organism = if_else(organism == "FIT Prediction", "FIT", organism)) %>% 
  
  pivot_longer(cols = -c(symbol, timepoint, organism),
               values_to = "value",
               names_to = "metric") %>% 
  mutate(
    organism = case_when(
      str_detect(metric, "human") ~ "Human",
      str_detect(metric, "fit") ~ "FIT",
      str_detect(metric, "mouse") ~ "Mouse",
      metric == "Human" ~ "Human",
      metric == "FIT Prediction" ~ "FIT",
      TRUE ~ organism),
    metric = str_remove_all(metric, "_?(human|mouse|fit)$"),
    metric = str_remove_all(metric, "_?(human|mouse|fit)")
  ) %>% 
  mutate(organism = case_when(str_detect(metric, "h2m") ~ "Mouse",
                              str_detect(metric, "h2") ~ "FIT",
                              T~organism))  %>% 
  mutate(organism = case_when(organism == "mean_fit" ~ "FIT",
                              organism == "mean_human" ~ "Human",
                              organism == "mean_mouse" ~ "Mouse",
                              T ~ organism))

#Comparisons
pairwise_human_fit_measures_comparisons = pairwise_human_fit_measures %>%
    select(timepoint, symbol, human_2_mouse:effect_size_h2m) 

#Pivot table and join the comparisons
pairwise_human_fit_genes_wide = pairwise_human_fit_genes_long %>% 
  pivot_wider(names_from = "metric",
              values_from = "value", 
              values_fn = mean) %>% 
  select(!c(pval:effect_size_h2m)) %>% 
  inner_join(pairwise_human_fit_measures_comparisons,
                by = join_by("timepoint", "symbol")) 


#Get top genes in each model
top_genes_human = pairwise_human_fit_genes_wide %>% 
  filter(organism == "Human") %>% 
  relocate(ci_width,  .after = organism) %>% 
  group_by(timepoint) %>% 
  arrange(ci_width) 
  
 top_genes_mouse = pairwise_human_fit_genes_wide %>% 
  filter(organism == "Mouse",
         symbol %in% c(top_genes_human$symbol)) %>% 
   arrange(ci_width, desc(p_value_h2m_adj), human_2_mouse)
 
 top_genes_fit = pairwise_human_fit_genes_wide %>% 
  filter(organism == "FIT",
         symbol %in% c(top_genes_human$symbol)) %>% 
   arrange(ci_width, desc(p_value_h2fit_adj), human_2_fit)
```

##### Plot confidence intervals for each top gene

```{r fig.height=7, fig.width=8}
#Plot -----
#Mouse to human
fit_example_human_plot_mouse = 
  pairwise_human_fit_genes_wide %>% 
  # rename(l2fc = value) %>% 
  filter(symbol %in% c(top_genes_mouse %>% 
                               select(symbol) %>% 
                               distinct() %>% 
                               slice_head(n = 5) %>% 
                               pull(symbol))) %>% 
  arrange(symbol, timepoint) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  facet_wrap(~symbol, nrow = 1) +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1), 
                show.legend = F
                ) +
    geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_color_manual(values = colors$organism)+
  scale_fill_manual(values = colors$organism)+
    theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        strip.text.x = element_text(color = "black"),
        strip.background = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "Time (days)",
       y = "LogFold-Change",
       title = "Prediction - Influenza, Fluad Vaccine\nTop 5 MOUSE genes with lowest CI widths and l2fc differences to human l2fc, with highest padj")


#FIT to human
fit_example_human_plot_fit= 
  pairwise_human_fit_genes_wide %>% 
  filter(symbol %in% c(top_genes_fit %>% 
                               select(symbol) %>% 
                               distinct() %>% 
                               slice_head(n = 5) %>% 
                               pull(symbol))) %>% 
  arrange(symbol, timepoint) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  facet_wrap(~symbol, nrow = 1) +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1), show.legend = F
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
    scale_color_manual(values = colors$organism)+
  scale_fill_manual(values = colors$organism)+
    theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        strip.text.x = element_text(color = "black"),
        strip.background = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "Time (days)",
       y = "LogFold-Change",
       title = "Top 5 FIT genes with lowest CI widths and l2fc differences to human l2fc, with highest padj")

#Print
fit_example_human_plot_mouse/
fit_example_human_plot_fit
```

How many genes are not significantly different from human gene expression?

```{r fig.height=8, fig.width=8}
# #How many genes are not significantly different from human gene expression? ------
# pairwise_human_fit_genes_size_plot = pairwise_human_fit_genes_wide %>% 
#   # filter(!is.na(p_value_diff_human)) %>% 
#   mutate(
#     # p_value_diff_human = if_else(organism == "Human", pval_adj, p_value_diff_human),
#     `<0.01` = abs(p_value_h2fit_adj) <= 0.01,
#     `<0.05` = abs(p_value_h2fit_adj) <= 0.05,
#     `<0.10` = abs(p_value_h2fit_adj) <= 0.10,
#     `<0.25` = abs(p_value_h2fit_adj) <= 0.25,
#     # `<1.00` = abs(p_value_diff_human) <= 1.00
#   ) %>%
#   pivot_longer(cols = starts_with("<"),
#                names_to = "threshold_label",
#                values_to = "in_threshold") %>%
#   mutate(threshold_label = forcats::fct_relevel(
#     threshold_label, "<0.01", "<0.05", "<0.10", "<0.25", "<1.00"
#   )) %>%
#   filter(in_threshold) %>%
#   select(-in_threshold) %>% 
#   group_by(organism, timepoint, threshold_label) %>% 
#   summarise(n = n()) %>% 
#   ungroup() %>% 
#   mutate(timepoint = as.character(timepoint) %>% 
#            fct_relevel(., "1", "3", "7")) %>% 
#   ggplot() + 
#   aes(x = timepoint,
#       y = n, 
#       fill = organism) +
#   geom_col(position = position_dodge2(1), show.legend = F,
#            alpha = 0.2) +
#   geom_line(aes(group = organism,
#                 color = organism),
#             position = position_dodge2(1), size =1) +
#   geom_point(position = position_dodge2(1),
#              pch = 21,
#              color = "black") +
#   scale_y_continuous(expand = c(0.01, 0.01)) +
#   facet_grid(~threshold_label, scales = "free_y") +
#     scale_fill_manual(
#   values = c(
#     FIT = "#4361ee", 
#     Mouse = "#4DBBD5FF",
#     Human = "#546E7AFF"))+
#     scale_color_manual(
#   values = c(
#     FIT = "#4361ee", 
#     Mouse = "#4DBBD5FF",
#     Human = "#546E7AFF"))+
#     theme_vaxgo() +
#   theme(axis.text.x = element_text(angle = 0),
#         legend.key = element_blank(),
#         strip.text.x = element_text(color = "black"),
#         # strip.background = element_rect(fill = "gray95", color = NA)
#         ) +
#   labs(fill = "Organism",
#        color = "Organism",
#        x = "Time (days)",
#        y = "Total",
#        title = "Total of genes with high similarity by p value cutoff")
# 
# pairwise_human_fit_genes_size_plot
```

#####Plot histogram of genes by cutoff

```{r fig.height=5, fig.width=3}
# #Plot histogram of genes by cutoff
# pairwise_human_fit_genes_wide_distribution_data = pairwise_human_fit_genes_wide %>% 
#   # filter(!is.na(p_value_diff_human)) %>% 
#   mutate(
#     # p_value_diff_human = if_else(organism == "Human", pval_adj, p_value_diff_human),
#     `<0.01` = abs(pval_adj) <= 0.01,
#     `<0.05` = abs(pval_adj) <= 0.05,
#     `<0.10` = abs(pval_adj) <= 0.10,
#     `<0.25` = abs(pval_adj) <= 0.25,
#     `<1.00` = abs(pval_adj) <= 1.00
#   ) %>%
#   pivot_longer(cols = starts_with("<"),
#                names_to = "threshold_label",
#                values_to = "in_threshold") %>%
#   mutate(threshold_label = forcats::fct_relevel(
#     threshold_label, "<0.01", "<0.05", "<0.10", "<0.25", "<1.00"
#   )) %>%
#   filter(in_threshold) %>%
#   select(-in_threshold) 
#   
# options(scipen = 999)
# pairwise_human_fit_genes_wide_distribution = pairwise_human_fit_genes_wide_distribution_data %>% 
#   filter(!organism == "Human") %>% 
#   ggplot() +
#   aes(x = mean) +
#   geom_density(aes(y = after_stat(count),
#                    fill = organism,
#                    # fill = organism,
#                    ), 
#                   position = position_identity(), 
#                alpha = 0.5,
#                  show.legend = F) +
#   guides(fill = "none") +
#   geom_density(aes(y = after_stat(count),
#                    color = organism,
#                    # fill = organism,
#                    ), 
#                   position = position_identity(), alpha = 0.5,
#                  show.legend = F) +
#   geom_density(aes(y = after_stat(count),
#                    color = organism), 
#                position = position_identity(),
#                data = pairwise_human_fit_genes_wide_distribution_data %>% 
#                     filter(organism == "Human"),
#                  fill = NA,
#                show.legend = T,
#                alpha = 1) +
#   
#   facet_wrap(~threshold_label, 
#              scale = "free_y",
#              ncol = 1, 
#              strip.position="right") +
#   scale_y_continuous(expand = c(0.01, 0.01), n.breaks = 3, 
#                      # labels = scales::comma
#                      ) +
#     scale_fill_manual(
#   values = c(
#     FIT = "#4361ee", 
#     Mouse = "#4DBBD5FF",
#     Human = "#546E7AFF"))+
#     scale_color_manual(
#   values = c(
#     FIT = "#4361ee", 
#     Mouse = "#4DBBD5FF",
#     Human = "#546E7AFF"))+
#     theme_vaxgo() +
#   theme(axis.text.x = element_text(angle = 0),
#         legend.position = "top",
#         legend.key = element_blank(),
#         strip.text.x = element_text(color = "black", size = 10),
#         strip.text.y = element_text(color = "black", angle = 0), 
#         panel.spacing.y = unit(0.8, "lines"),
#         strip.background.y = element_rect(fill = NA, color = NA)) +
#   labs(fill = "",
#        color = "",
#        x = "Log2-fold change",
#        y = "Total",
#        title = "Frequency of genes by p value cutoff")
# 
# pairwise_human_fit_genes_wide_distribution
```

Unite plots

```{r fig.height=8, fig.width=12}

#Unite plots ------
a = fit_example_human_plot_mouse + theme(legend.position = "none")  + theme(plot.title = element_blank())
b = fit_example_human_plot_fit + guides(fill = "none") + theme(plot.title = element_blank())
c = pairwise_human_fit_genes_size_plot + theme(legend.position = "none") + guides(fill = "none") + theme(plot.title = element_blank())
d = pairwise_human_fit_genes_wide_distribution + theme(legend.position = "top",
                                                       legend.title = element_text(size = 10),
                                                       legend.text = element_text(size = 8),
                                                       axis.text.y = element_text(size = 8),
                                                       plot.title = element_blank())


plot_patchwork_1 = d + (c/a/b) +
  plot_layout(widths = c(0.2, 1), 
              axis_titles = "collect",
              axes = "collect")

plot_patchwork_1

#Save
plot_patchwork_1 %>%
  ggsave(filename = here("Figures", "influenza_fluad_mouse_fit_prediction_model2human_top5_fit_and_mouse.png"),
         width = 12,
         height = 8)



```

##### Get best and worst genes predicted

Get best and worst genes predicted by mouse and FIT, and see what can be better Difference in ranks

```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
# Get best and worst predicted genes ------
pairwise_genes_wide_mouse_fit = pairwise_human_fit_genes_wide %>% 
  filter(organism != "Human",
         symbol %in% c(top_genes_human$symbol)) 

#Get comparisons
z_scores_best_worst_models_comparisons = pairwise_human_fit_measures_comparisons %>% 
  mutate(diff_z_fit_mouse = abs(z_score_h2fit) - abs(z_score_h2m),
         diff_z_mouse_fit = abs(z_score_h2m) - abs(z_score_h2fit)) 

#Best and worst FIT
best_fit = z_scores_best_worst_models_comparisons %>% 
  arrange(abs(z_score_h2fit))

worst_fit = z_scores_best_worst_models_comparisons %>% 
  arrange(-abs(z_score_h2fit))

#Best and worst Mouse
best_mouse = z_scores_best_worst_models_comparisons %>% 
  arrange(abs(z_score_h2m))

worst_mouse = z_scores_best_worst_models_comparisons %>% 
  arrange(-abs(z_score_h2m))

#Merge FIT vs. Mouse
worst_fit_mouse = z_scores_best_worst_models_comparisons %>% 
  arrange(-diff_z_fit_mouse)

#Merge Mouse vs. FIT
worst_mouse_fit = z_scores_best_worst_models_comparisons %>% 
  arrange(-diff_z_mouse_fit)
```

Plots

```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
#Plot best vs worst -----
#FIT ------
#Best FIT
best_fit_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(best_fit %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~human_symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Top genes in FIT prediction. Difference in Z-score to human")

#Worst FIT
worst_fit_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(worst_fit %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Worst predicted genes in Mouse. Difference in Z-score to human")

#Mouse -----
#Best mouse
best_mouse_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(best_mouse %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~human_symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Top genes in Mouse prediction. Difference in Z-score to human")

#Worst Mouse
worst_mouse_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(worst_mouse %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~human_symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Worst predicted genes in Mouse. Difference in Z-score to human")


#FIT and Mouse -------
#Worst FIT to best Mouse
worst_fit_mouse_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(worst_fit_mouse %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~human_symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Genes that were most different in FIT prediction from mouse and human. Difference in Z-score")

#Worst mouse to best FIT
worst_mouse_fit_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(worst_mouse_fit %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~human_symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Genes that were most different in Mouse prediction from FIT and human. Difference in Z-score")


worst_mouse_fit_plot
```

##### Get ranks

```{r}
#Get ranks -----
#Mouse
ranks_mouse = pairwise_human_fit_genes_wide %>%
  filter(organism == "Mouse") %>%
  select(symbol, timepoint, z_score_h2m, organism) %>%
  group_by(timepoint) %>% 
  arrange(timepoint, abs(z_score_h2m)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2m)),
         percentile = round(percent_rank * 100, 2)) %>% 
  distinct() %>% 
  rename(zscore_h2model = z_score_h2m)
#FIT
ranks_fit = pairwise_human_fit_genes_wide %>%
  filter(organism == "FIT") %>%
  select(symbol, timepoint, z_score_h2fit, organism) %>%
  group_by(timepoint) %>% 
  arrange(timepoint, abs(z_score_h2fit)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2fit)),
         percentile = round(percent_rank * 100, 2)) %>% 
  distinct() %>% 
  rename(zscore_h2model = z_score_h2fit)

#Fit vs Mouse
ranks_fit_vs_mouse = ranks_fit %>% 
  select(symbol, fit = percentile, timepoint_ref = timepoint, zscore_h2fit = zscore_h2model) %>% 
  inner_join(ranks_mouse %>% 
               select(symbol, mouse = percentile,
                      timepoint_other = timepoint,
                      zscore_h2mouse = zscore_h2model),
             by = "symbol") %>% 
  mutate(timepoint_ref = as.character(timepoint_ref),
         timepoint_other = as.character(timepoint_other),
         condition = "Top FIT vs Mouse")

#Mouse vs FIT
ranks_mouse_vs_fit = ranks_mouse %>% 
  select(symbol, mouse = percentile, timepoint_ref = timepoint, zscore_h2mouse = zscore_h2model) %>% 
  inner_join(ranks_fit %>% 
               select(symbol, fit = percentile,
                      timepoint_other = timepoint,
                      zscore_h2fit = zscore_h2model),
             by = "symbol") %>% 
  mutate(timepoint_ref = as.character(timepoint_ref),
         timepoint_other = as.character(timepoint_other),
         condition = "Top Mouse vs FIT")


#Get genes that are in the 10th percentile in one model, but are in the 90th percentile in the other
ranks_best_fit_and_mouse = ranks_fit_vs_mouse %>%
  filter(fit < 10, mouse > 90) %>%
  bind_rows(ranks_mouse_vs_fit %>%
              filter(mouse < 10, fit > 90)) %>%
  select(condition,
         symbol,
         timepoint_ref,
         timepoint_other, 
         fit, zscore_h2fit,
         mouse, zscore_h2mouse)

ranks_best_fit_and_mouse_long <- ranks_best_fit_and_mouse %>%
  pivot_longer(
    cols = c(fit, mouse),
    names_to = "model",
    values_to = "percentile"
  )


#Get genes with least zscores in one model, but have the highest zscore in the other
ranks_best_fit_and_mouse_zscore = ranks_fit_vs_mouse %>%

  # FILTER BY N
  filter(timepoint_ref == timepoint_other) %>%
  arrange(abs(zscore_h2fit), desc(abs(zscore_h2mouse))) %>%
  slice_head(n = 20) %>% 
  bind_rows(ranks_mouse_vs_fit %>%
              filter(timepoint_ref == timepoint_other) %>%
  arrange(abs(zscore_h2mouse), desc(abs(zscore_h2fit))) %>%
  slice_head(n = 20)) %>%
  select(condition,
         symbol,
         timepoint_ref,
         timepoint_other, 
         fit = zscore_h2fit,
         mouse = zscore_h2mouse)
#Pivot longer
ranks_best_fit_and_mouse_long_zscore <- ranks_best_fit_and_mouse_zscore %>%
  pivot_longer(
    cols = c(fit, mouse),
    names_to = "model",
    values_to = "zscore"
  )

#Print
ranks_best_fit_and_mouse_long_zscore

```

##### Plot percentiles

```{r fig.height=4, fig.width=8}
#Plot percentiles -------
#Select genes
selected_genes = ranks_best_fit_and_mouse %>% 
  group_by(condition, timepoint_ref, timepoint_other) %>% 
  filter(timepoint_ref == timepoint_other) %>% 
  # filter(fit < 1 | mouse <1) %>% 
  pivot_longer(
    cols = c(fit, mouse),
    names_to = "model",
    values_to = "percentile"
  ) 

#Plot
ranks_percentile_best_fit_and_mouse_long_plot = ranks_best_fit_and_mouse_long %>% 
  filter(timepoint_ref == timepoint_other) %>% 
  ggplot() +
  aes(x = model, y = percentile, group = interaction(condition, symbol), color = condition) +
  geom_line(alpha = 0.5) +
  guides(color = guide_legend(override.aes = list(linewidth = 2))) +
  geom_text_repel(data = selected_genes %>%
                    inner_join(ImmuneGO_BTM_grouped_genes %>% 
                                               select(symbol = genes) %>% 
                                               distinct() %>% 
                                               mutate(gene_function = "Immune"),
                  by = "symbol") %>% 
                    mutate(gene_function = "Other") %>% 
                    filter(condition == "Top FIT vs Mouse" & model == "mouse") %>% 
                    arrange(-percentile) %>% 
                    slice_max(percentile, n = 3),
             mapping = aes(label = symbol,
                           x = model,
                           y = percentile),
             color = "black",
             segment.color = "#4361ee",
             direction = c("y"),
             size = 2.5, 
            min.segment.length = 0.5,
             max.overlaps = 10000, 
             force = 2,
             nudge_x = 1, 
            hjust = "left") +
  
geom_text_repel(data = selected_genes %>%
                    filter(condition == "Top Mouse vs FIT" & model == "fit") %>% 
                  arrange(-percentile) %>% 
                  slice_max(percentile, n = 3),
                color = "black",
             segment.color = "#4DBBD5FF",
             direction = c("y"),
             mapping = aes(label = symbol,
                           x = model,
                           y = percentile),
             size = 2.5, 
             min.segment.length = 0.5,
             max.overlaps = 10000, 
             nudge_x = -1, 
             force = 2, hjust = "right") +
  facet_grid(~paste("Day", timepoint_ref))  +
  labs(x = "Model",
       y = "Percentile",
       color = ""
       # title = "Best 10% versus worst 100%"
       ) +
    scale_color_manual(values = c(
    "Top FIT vs Mouse" = "#4361ee", 
    "Top Mouse vs FIT" = "#4DBBD5FF")) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01), limits = c(0, 110)) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "right",
        legend.key = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0), 
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA))
  
#Print
ranks_percentile_best_fit_and_mouse_long_plot 
 
#Save
ranks_percentile_best_fit_and_mouse_long_plot %>% 
  ggsave(filename = here("Figures", "influenza_fluad_mouse_fit_prediction_ranks_best_plot.png"),
         width = 8,
         height = 5)
```

##### Plot ranks by zscores

```{r fig.height=4, fig.width=8}
#Plot ranks by zscores ------
#Select genes by zscore
selected_genes_zscore = ranks_best_fit_and_mouse_zscore %>%
  group_by(condition, timepoint_ref, timepoint_other) %>%
  pivot_longer(cols = c(fit, mouse),
               names_to = "model",
               values_to = "zscore") 

#Get the genes with the highest zscores in Mouse
worst_zscores_Mouse = selected_genes_zscore %>%
  filter(condition == "Top FIT vs Mouse" &
           model == "mouse") %>%
  arrange(-abs(zscore)) %>%
  slice_head(n = 3) %>%
  ungroup()

#Get the genes with the lowest zscores in Mouse
best_zscores_Mouse = selected_genes_zscore %>%
  filter(condition == "Top FIT vs Mouse" &
           model == "mouse") %>%
  arrange(abs(zscore)) %>%
  slice_head(n = 3) %>%
  ungroup()

#Get the genes with the highest zscores in FIT
worst_zscores_FIT = selected_genes_zscore %>%
  filter(condition == "Top Mouse vs FIT" &
           model == "fit") %>%
  arrange(-abs(zscore)) %>%
  slice_head(n = 3) %>% 
  ungroup()

#Get the genes with the lowest zscores in FIT
best_zscores_FIT = selected_genes_zscore %>%
  filter(condition == "Top Mouse vs FIT" &
           model == "fit") %>%
  arrange(abs(zscore)) %>%
  slice_head(n = 3) %>% 
  ungroup()

```

```{r fig.height=3, fig.width=8}
#Plot
ranks_best_worst_fit_and_mouse_long_zscore_plot = ranks_best_fit_and_mouse_long_zscore %>% 
  ungroup() %>% 
  ggplot() +
  aes(x = model, y = zscore, 
      group = interaction(condition, symbol), 
      color = condition) +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.3) +
   geom_line(alpha = 0.5) +
  
  #Get the worst zscores for mouse and see where they are in the fit
  geom_text_repel(data = worst_zscores_Mouse,
             mapping = aes(label = symbol,
                           x = model,
                           y = zscore),
             size = 2.5, 
             direction = c("y"),
             color = "black",
             segment.color = "black", 
             max.overlaps = 10000, 
             # force = 2,
             force_pull = 5,
             nudge_x = 1, 
             min.segment.length = 0.5, 
             hjust = "left") +
  
  #Get the worst zscores fit and see if they improved in mouse
  geom_text_repel(data = worst_zscores_FIT,
                  direction = c("y"),
                  mapping = aes(label = symbol,
                                x = model,
                                y = zscore),
                  color = "black",
                  segment.color = "black",
                  size = 2.5,
                  max.overlaps = 10000,
                  force = 2,
                  nudge_x = -1, 
                  min.segment.length = 0.5, 
                  hjust = "right")+


  facet_wrap(~paste("Day", timepoint_ref), scales = "free")  +
  labs(x = "Model",
       y = "Zscore",
       color = "") +
    scale_color_manual(values = c(
    "Top FIT vs Mouse" = "#4361ee", 
    "Top Mouse vs FIT" = "#4DBBD5FF")) +
  guides(color = guide_legend(override.aes = list(linewidth = 2))) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "right",
        legend.key = element_blank(),
        # legend.key.height = unit(0.4,"cm"),
        # legend.key.width = unit(0.4,"cm"),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0), 
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA)) 

#Print
ranks_best_worst_fit_and_mouse_long_zscore_plot
  
#Save
# ranks_best_worst_fit_and_mouse_long_zscore_plot %>% 
#   ggsave(filename = here("Figures", "ranks_best_worst_fit_and_mouse_long_zscore_plot.png"),
#          width = 8,
#          height = 5)
```

Combine

```{r fig.height=5, fig.width=8}
#Combine plots
percentile_zscore_ranks = (ranks_best_worst_fit_and_mouse_long_zscore_plot/ranks_percentile_best_fit_and_mouse_long_plot) +
  plot_layout(guides = "collect") 

#Save
ggsave(percentile_zscore_ranks,
       filename = here("Figures", "influenza_fluad_mouse_fit_prediction_ranks_zscore_patch.png"),
         width = 8,
         height = 5)
  
```

Plot all

```{r fig.height=8, fig.width=12}
#Combine plots -------
a = pairwise_human_fit_genes_wide_distribution + theme(legend.position = "top",
                                                       legend.direction = "horizontal",
                                                       plot.title = element_blank(),
                                                       legend.title = element_text(size = 10),
                                                       legend.text = element_text(size = 8),
                                                       axis.text.y = element_text(size = 10))
b = pairwise_human_fit_genes_size_plot + theme(legend.position = "right",
                                               legend.direction = "vertical",
                                               plot.title = element_blank(),
                                               legend.title = element_text(size = 10),
                                               legend.text = element_text(size = 8),
                                               axis.text.y = element_text(size = 10))
c = ranks_best_worst_fit_and_mouse_long_zscore_plot + theme(legend.position = "right", 
                                               legend.direction = "vertical",
                                               plot.title = element_blank(),
                                               legend.title = element_text(size = 10),
                                               legend.text = element_text(size = 8),
                                               axis.text.y = element_text(size = 10)) + 
  theme(title = element_blank())
d = ranks_percentile_best_fit_and_mouse_long_plot + theme(legend.position = "right", 
                                               legend.direction = "vertical",
                                               plot.title = element_blank(),
                                               legend.title = element_text(size = 10),
                                               legend.text = element_text(size = 8),
                                               axis.text.y = element_text(size = 10))

mouse_vs_fit_ranks_zscores_patch = a + (b/c/d) +
  plot_layout(widths = c(0.2, 1), 
              axis_titles = "collect"
              ) 


mouse_vs_fit_ranks_zscores_patch

#Save
mouse_vs_fit_ranks_zscores_patch %>% 
  ggsave(filename = here("Figures", "influenza_fluad_mouse_fit_prediction_ranks_zscore_patch.png"),
         width = 12,
         height = 8)
```

##### Heatmap: Best 10 vs Worst 90

Get best 10 and worst 90

```{r}
#Plot heatmap -----
#Get best 10 and worst 90
#Mouse
mouse_best_10 = best_mouse %>%
  inner_join(pairwise_genes_wide_mouse_fit %>% 
               select(timepoint:l2fc) %>% 
               filter(organism == "Mouse"),
             by = join_by("symbol", "timepoint")) %>% 
  # filter(organism == "Mouse") %>%
  select(symbol, z_score_h2m, l2fc, timepoint, organism) %>%
  group_by(organism, timepoint) %>% 
  arrange(abs(z_score_h2m)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2m)),
         percentile = round(percent_rank * 100, 2)) %>%
  filter(percentile <= 10) %>% 
  mutate(condition = paste("Mouse Best Day", timepoint))

mouse_worst_90 = best_mouse %>%
  inner_join(pairwise_genes_wide_mouse_fit %>% 
               select(timepoint:l2fc) %>% 
               filter(organism == "Mouse"),
             by = join_by("symbol", "timepoint")) %>% 
  select(symbol, z_score_h2m, l2fc, timepoint, organism) %>%
    group_by(organism, timepoint) %>% 
  arrange(abs(z_score_h2m)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2m)),
         percentile = round(percent_rank * 100, 2)) %>%
  filter(percentile >= 90) %>% 
  mutate(condition = paste("Mouse Worst Day", timepoint))

#FIT
fit_best_10 = best_fit %>%
  inner_join(pairwise_genes_wide_mouse_fit %>% 
               select(timepoint:l2fc) %>% 
               filter(organism == "FIT"),
             by = join_by("symbol", "timepoint")) %>% 
  select(symbol, z_score_h2fit, l2fc, timepoint, organism) %>%
  group_by(organism, timepoint) %>% 
  arrange(abs(z_score_h2fit)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2fit)),
         percentile = round(percent_rank * 100, 2)) %>%
  filter(percentile <= 10) %>% 
  mutate(condition = paste("FIT Best Day", timepoint))

fit_worst_90 = best_fit %>%
  inner_join(pairwise_genes_wide_mouse_fit %>% 
               select(timepoint:l2fc) %>% 
               filter(organism == "FIT"),
             by = join_by("symbol", "timepoint")) %>% 
  select(symbol, z_score_h2fit, l2fc, timepoint, organism) %>%
   group_by(organism, timepoint) %>% 
  arrange(abs(z_score_h2fit)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2fit)),
         percentile = round(percent_rank * 100, 2)) %>%
  filter(percentile >= 90) %>% 
  mutate(condition = paste("FIT Worst Day", timepoint))

#Bind tables
mouse_fit_best_worst = bind_rows(mouse_best_10, 
                                 fit_best_10,
                                 mouse_worst_90, 
                                 fit_worst_90) 

#Get overlaps
overlap_best_worst = mouse_fit_best_worst %>%
  select(symbol, l2fc, condition, organism, timepoint)  %>% 
  ungroup()
  
```

Plot heatmap

```{r fig.height=8, fig.width=3}

#Plot heatmap
row_clustered = overlap_best_worst %>% 
    dplyr::select(condition, symbol, l2fc) %>% 
    distinct() %>% 
    pivot_wider(names_from = "condition",
                values_from= "l2fc") %>%
    column_to_rownames("symbol") %>% 
    as.matrix() %>% 
    replace(is.na(.), 0) %>% 
    dist(.) %>% 
    hclust()

df_genes_mouse_fit_best_worst_perc_hallmark_heatmap = overlap_best_worst %>% 
  clean_names() %>% 
  ungroup() %>% 
  select(condition, symbol, everything()) %>% 
    mutate(model = case_when(str_detect(condition, "Mouse") ~ "Mouse", 
                          str_detect(condition, "FIT") ~ "FIT",
                          T ~ NA),
           symbol = fct_relevel(symbol, c(row_clustered$labels)), 
           condition = if_else(str_detect(condition, "Best"), "Best", "Worst"))%>% 
  ggplot(aes(x=model, y = symbol, fill = l2fc, color= l2fc)) +
  geom_tile(size = 0.2, na.rm = F, stat = "identity") +
  scale_fill_gradient2(
  low = "#45efc6",
  mid = "white",  
  high = "#165289",
        n.breaks = 3,
        oob = scales::squish,
        limits = c(-1, 1),
        name = "L2FC") +
  scale_color_gradient2(
  low = "#45efc6",
  mid = "white",  
  high = "#165289",
        n.breaks = 3,
        oob = scales::squish,
        limits = c(-1, 1),      # fixa o intervalo
        name = "L2FC") +
  scale_y_discrete(position = "right") +
    theme_vaxgo() +
    theme(axis.text = element_text(color = "black", size = 8) ,
          # axis.line.x  = element_blank(),
          axis.line.y  = element_blank(),
          axis.text.x = element_text(angle = 90, vjust = 1, hjust=1),
          axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1, size = 0),
          axis.ticks.y = element_blank(), 
          panel.spacing.x = unit(0.1, "cm"),
          strip.background.x = element_rect(fill = "black"),
          strip.text.x = element_text(color = "white", size = 10, 
                                      margin = margin(0.2,1,0.2,1, "cm")),
          panel.grid = element_blank(),
          plot.margin = margin(0, 0, 0, 0, unit = "cm"),
          legend.direction = "horizontal",
          legend.position = "bottom") +
     labs(x = "",
         y = "",
         title = "Best and Worst genes, Overlap") +
  facet_nested(~ paste("Day", timepoint) + condition, scales = "free", space = "free")

df_genes_mouse_fit_best_worst_perc_hallmark_heatmap

df_genes_mouse_fit_best_worst_perc_hallmark_heatmap %>% 
  ggsave(filename = here("Figures", "df_genes_mouse_fit_best_worst_perc_hallmark_heatmap.png"),
         width = 3,
         height = 8)

```

#### ORA: Which processes are better or worse predicted?

```{r fig.width=10, fig.height= 8}
#ORA Hallmarks DIFF
#HALLMARK
#Retrieve human H (hallmark) gene set
msigdbr_df <- msigdbr(species = "human", collection = "H") %>% 
  select(process = gs_name, genes = gene_symbol) %>% 
  mutate(process = str_remove(process, "HALLMARK_"))

ora_mouse_fit_best_worst_perc_hallmark <- enricher_condition(
  df = overlap_best_worst %>% 
    mutate(genes = symbol),
  term2gene_df = msigdbr_df) %>% 
  mutate(log_q = -log(qvalue)) %>% 
  mutate(description = str_replace_all(description, "_", " "),
         timepoint = str_extract(condition, "Day \\d+"))


ora_mouse_fit_best_worst_perc_hallmark
```

ORA Hallmark

```{r fig.height=6, fig.width=6}
#Plot heatmap for ORA Hallmark ------
row_clustered = ora_mouse_fit_best_worst_perc_hallmark %>%  
    dplyr::select(condition, description, log_q) %>% 
    distinct() %>% 
    pivot_wider(names_from = "condition",
                values_from= "log_q") %>%
    column_to_rownames("description") %>% 
    as.matrix() %>% 
    replace(is.na(.), 0) %>% 
    dist(.) %>% 
    hclust()

df_ora_mouse_fit_best_worst_perc_hallmark_heatmap = ora_mouse_fit_best_worst_perc_hallmark %>% 
  clean_names() %>% 
  filter(description %in% c(row_clustered$labels)) %>% 
  ungroup() %>% 
  select(condition, description, log_q, qvalue, everything()) %>% 
    mutate(signific = case_when(qvalue <= 0.10 ~ "*",
                              qvalue <= 0.25 ~ "-",
                              qvalue <= 0.50 ~ "/",
                              qvalue <= 0.05 ~ "**",
                              TRUE ~ ""),
        model = case_when(str_detect(condition, "Mouse") ~ "Mouse", 
                          str_detect(condition, "FIT") ~ "FIT",
                          T ~ NA),
        best_worst = case_when(str_detect(condition, "Best") ~ "10%", 
                          str_detect(condition, "Worst") ~ "90",
                          T ~ NA)) %>% 
  ggplot(aes(x=fct_rev(model), y = str_to_upper(description), fill = log_q)) +
  geom_tile(color = "black", 
            size = 0.2, na.rm = F, stat = "identity") +
  geom_text(aes(label = signific), size = 4, vjust = 0.5, color = "black") +
  scale_fill_gradient2(
        low = "white",     
        high = "#165289",
        n.breaks = 3,
        oob = scales::squish,
        limits = c(0, 4),      # fixa o intervalo
        name = "-Log(FDR)") +
  scale_y_discrete(position = "right") +
    theme_vaxgo() +
    theme(axis.text = element_text(color = "black", size = 8) ,
          # axis.line.x  = element_blank(),
          axis.line.y  = element_blank(),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1),
          axis.ticks = element_blank(),
          strip.background.x = element_rect(fill = "black"),
          strip.text.x = element_text(color = "white", size = 10, 
                                      margin = margin(0.2,1,0.2,1, "cm")),
          panel.grid = element_blank(), 
          panel.spacing.x = unit(0.1, "cm"),
          plot.margin = margin(0, 0, 0, 1, unit = "cm"),
          legend.direction = "horizontal",
          legend.position = "bottom") +
     labs(x = "",
         y = "",
         title = "ORA, Best and Worst genes, Percentile") +
  facet_nested(~ timepoint + best_worst, scales = "free", space = "free")


df_ora_mouse_fit_best_worst_perc_hallmark_heatmap

df_ora_mouse_fit_best_worst_perc_hallmark_heatmap %>% 
  ggsave(filename = here("Figures", "influenza_fluad_mouse_fit_prediction_ora_hallmark_bet_worst_percentile_heatmap_2.png"),
         width = 6,
         height = 6)
```



#====================
# Align sequences

```{r}


#Import homologous CREs from ENCODE 
homologous_cres = read_tsv(here("Genomic", "hg38-mm10-Homologous.tsv"),
  col_names = c("human_enhancer", "mouse_enhancer")
) 

#Import human CREs
cres_human <- read_tsv(
  here("Genomic", "GRCh38-cCREs.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>% 
   rename_with(~paste0(., "_human"))
              
              
#Import mouse CREs
cres_mouse <- read_tsv(
  here("Genomic", "mm10-cCREs.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>% rename_with(~paste0(., "_mouse"))


#Unite info
cres_annotated = homologous_cres %>% 
  full_join(cres_human, by = join_by(human_enhancer == id_ccre_human)) %>% 
  full_join(cres_mouse, by = join_by(mouse_enhancer == id_ccre_mouse)) 
  
```

```{r}
# Instalar ChIPseeker e TxDb se necessário
if (!require("ChIPseeker", quietly = TRUE)) {
  BiocManager::install("ChIPseeker")
}

library(ChIPseeker)
library(rtracklayer)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)

# Look up data types and keytypes
columns(org.Hs.eg.db)
keytypes(org.Hs.eg.db)

# Convert to a GRanges object
cres_human_gr <- GRanges(
  seqnames = cres_human$chrom_human,
  ranges = IRanges(start = cres_human$start_human, end = cres_human$end_human)
)

# Get levels and seqinfo
seqlevels(cres_human_gr) <- seqlevels(BSgenome.Hsapiens.UCSC.hg38)
seqinfo(cres_human_gr) <- seqinfo(BSgenome.Hsapiens.UCSC.hg38)

# Annotate peak
cres_anno <- annotatePeak(
  cres_human_gr,
  tssRegion = c(-3000, 3000),
  TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene,
  annoDb = "org.Hs.eg.db"
)

# Get sequences
sequences_human <- getSeq(BSgenome.Hsapiens.UCSC.hg38, cres_human_gr)

# Convert to dataframe
cres_anno_df <- as.data.frame(cres_anno)

cres_human_seq <- cres_human %>%
  mutate(
    sequence_human = as.character(sequences_human),
    gene_name = cres_anno_df$SYMBOL,
    gene_id = cres_anno_df$GENEID,
    annotation = cres_anno_df$annotation,
    distance_to_tss = cres_anno_df$distanceToTSS
  )
```

Annotate genes
```{r}
library(AnnotationHub)

# 1. Obter as coordenadas dos genes humanos (Ex: hg38)
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
genes_gr <- genes(txdb)

#Get sequence in a range
promoters_custom <- promoters(genes_gr, upstream = 2500, downstream = 300)

# Inicializar o Hub
ah <- AnnotationHub()

# ---- A. ELEMENTOS TRANSPONÍVEIS (RepeatMasker) ----
rmsk_query <- query(ah, c("RepeatMasker", "Homo sapiens", "hg38"))
# Inspecione rmsk_query para escolher o ID correto (ex: AH61234 ou similar)
# Supondo que o ID retornado seja AH99341:
rmsk_gr <- ah[["AH99341"]] 

# ---- B. CONSERVAÇÃO FILOGENÉTICA (PhastCons) ----
# Buscar scores de PhastCons (ex: conservação entre 100 vertebrados)
phast_query <- query(ah, c("phastCons", "Homo sapiens", "hg38", "BigWig"))
# BigWigFiles retornam uma referência ao arquivo para consulta rápida por coordenada
phast_bw <- ah[["AH75265"]] # Substitua pelo ID real da sua busca


# Encontrar intersecções entre as janelas dos genes e os elementos repetitivos
overlaps_rmsk <- findOverlaps(promoters_custom, rmsk_gr)

# Criar um data.frame com o mapeamento
gene_te_mapping <- data.frame(
  gene_id = names(promoters_custom)[queryHits(overlaps_rmsk)],
  te_name = rmsk_gr$repName[subjectHits(overlaps_rmsk)],
  te_family = rmsk_gr$repClass[subjectHits(overlaps_rmsk)] # Ex: SINE/Alu
)

# Exemplo para calcular a densidade de elementos Alu por gene:
library(dplyr)
te_density_per_gene <- gene_te_mapping %>%
  group_by(gene_id) %>%
  summarise(
    total_TEs = n(),
    total_Alu = sum(te_family == "SINE/Alu")
  )


# Extrair os scores de conservação apenas para as regiões das nossas janelas promotoras
phast_scores <- import.bw(phast_bw, selection = BigWigSelection(promoters_custom))

# Calcular o score médio de conservação para a janela de cada gene
# Usamos aggregate para calcular a média do score sobrepondo as janelas
score_per_promoter <- aggregate(phast_scores, promoters_custom, score = mean)

# Adicionar o score ao objeto original
promoters_custom$mean_phastCons <- score_per_promoter$score


# Intersecção espacial
overlaps_cres <- findOverlaps(promoters_custom, cres_human_seq)


```









