```{r}
pairwise_human_fit_test =  read_rds(here("FIT_Mouse2man", "pairwise_human_fit_test.rds"))
```


```{r fig.height=4, fig.width=8}

quantile_diff_genes = pairwise_human_fit_test 

abs_diff_correlation_plot_data_split_qn = quantile_diff_genes %>% 
  mutate(
    quartile_label = cut(abs(human_2_mouse_qn),
                         breaks = quantile(abs(human_2_mouse_qn),
                                           probs = c(0, 0.10, 0.25, 0.5, 0.75, 1),
                                           na.rm = TRUE),
                         include.lowest = TRUE,
                         labels = c("Q10", "Q25", "Q50", "Q75", "Q100")),
    timepoint_day = paste0("Day ", timepoint)) %>% 
  mutate(pval_mouse_adj = if_else(pval_mouse_adj <= 0.05, "DEG", "Not DEG"),
         pval_human_adj = if_else(pval_human_adj <= 0.05, "DEG", "Not DEG"))

abs_diff_correlation_plot_data_split_qn %>% dplyr::count(timepoint)

#Get all genes
quantiles <- quantile(abs(quantile_diff_genes$human_2_mouse_qn), 
                    probs = c(0.10, 0.25, 0.5, 0.75, 1), 
                    na.rm = TRUE)

diff_correlation_plot_data_qn = quantile_diff_genes %>% 
  arrange(timepoint)  %>% 
  ungroup() %>% 
  #Split by quantile
    mutate(
    Q10 = abs(human_2_mouse_qn) <= quantiles[1],
    Q25 = abs(human_2_mouse_qn) <= quantiles[2],
    Q50 = abs(human_2_mouse_qn) <= quantiles[3],
    Q75 = abs(human_2_mouse_qn) <= quantiles[4],
    Q100 = abs(human_2_mouse_qn) <= quantiles[5]
  ) %>%
  #Cumulative
  pivot_longer(cols = starts_with("Q"),
               names_to = "quartile_label",
               values_to = "in_quantile") %>%
  mutate(quartile_label = fct_relevel(quartile_label, "Q10", "Q25", "Q50", "Q75", "Q100")) %>% 
  filter(in_quantile) %>%
  select(-in_quantile)
```


```{r fig.height=4, fig.width=8}
abs_diff_correlation_plot_data_split_qn %>% 
  ungroup() %>% 
  ggplot() +
  aes(x = mean_mouse_qn,
      y = mean_human_qn) +
  facet_wrap(~paste("Day", timepoint)) +
  #All points, colored by differences
  geom_point(
             aes(color = quartile_label,
                 # alpha = quartile_label,
                 fill = quartile_label),
             # fill = NA,
             # alpha = 1,
             pch = 21) +

  #Same DEGs (WORST)
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1,
             size = 0.5,
             pch = 21,
             data = abs_diff_correlation_plot_data_split_qn %>%
               filter(pval_mouse_adj == pval_human_adj,
                      pval_mouse_adj == "DEG") %>%
               filter(quartile_label %in% c("Q50", "Q75", "Q100"))) +
  #Same DEGs (BEST)
  geom_point(color = "black",
             fill = "#4DBBD5FF",
             alpha = 1,
             size = 2,
             pch = 21,
             data = abs_diff_correlation_plot_data_split_qn %>%
               filter(pval_mouse_adj == pval_human_adj,
                      pval_mouse_adj == "DEG") %>%
               filter(quartile_label %in% c(
                 "Q10",
                 "Q25"
                 ))) +
   #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(slope = -1, alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  scale_color_manual(values = c("Q10" = "black",
                               "Q25" = "gray25",
                               "Q50" = "gray50",
                               "Q75" = "gray75",
                               "Q100" = "gray90")) +
  scale_fill_manual(values = c("Q10" = "black",
                               "Q25" = "gray25",
                               "Q50" = "gray50",
                               "Q75" = "gray75",
                               "Q100" = "gray90")) +
  scale_x_continuous(limits = c(-1.5, 1.5)) +
  scale_y_continuous(limits = c(-1.5, 1.5)) +
  
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.position = "right",
        legend.direction = "vertical",
        legend.key.width = unit(0.4, 'cm'),
        legend.key.height = unit(0.4, 'cm'),
        legend.key = element_blank(),
        legend.title = element_text(size = 10, vjust = 0.9),
        legend.text = element_text(size = 7, vjust = 0.5, hjust = -0.2)) +
  labs(title = "Which genes and DEGs vary in both human and mouse?",
       subtitle = "Differences in mouse to human, by timepoint.In colors, DEGs in human AND mouse, p_adjust < 0.05, |diff| < 0.2. \nPurple = Negatively correlated, Blue = Positively correlated",
       x = "Mouse L2FC",
       y = "Human L2FC",
       alpha = "Quantiles",
       fill = "Quantiles",
       color = "|Z|")


```






### plot btms with CIs

Engerix
```{r fig.height=4, fig.width=8}

dge_btm_mean_process_samples = 
  readRDS(file = here("tables", "engerixb_dge_btm_mean_process_samples.rds"))


dge_btm_mean_process = dge_btm_mean_process_samples %>% 
  group_by(organism, timepoint, process) %>% 
  summarize(
    mean = mean(log2fc, na.rm = TRUE) %>% round(2),
    median = median(log2fc, na.rm = TRUE) %>% round(2),
    min = min(log2fc, na.rm = TRUE) %>% round(2),
    max = max(log2fc, na.rm = TRUE) %>% round(2),
    sd = sd(log2fc, na.rm = T),
    n = sum(!is.na(log2fc)),
    se = sd/sqrt(n),
    t_critical = qt(0.975, df = n - 1),
    ci_lower = mean - t_critical * se,
    ci_upper = mean + t_critical * se,
    p_value = t.test(log2fc, mu = 0)$p.value,
    .groups = "drop",
    group,
    subgroup
  ) %>%
  distinct() %>% 
  mutate(p_adj = p.adjust(p_value, method = "BH"), 
         sig_cat = case_when(p_adj < 0.01 ~ "< 0.01", 
                             p_adj < 0.05 ~ "< 0.05",
                             p_adj < 0.10 ~ "< 0.10", 
                             p_adj < 0.25 ~ "< 0.25", 
                             TRUE ~ "≥ 0.25"),
         sig_cat = factor(sig_cat,
                   levels = c("< 0.01", "< 0.05", "< 0.10", "< 0.25", "≥ 0.25"))) %>% 
  select(organism, timepoint, process, mean, median, p_adj, p_value, sig_cat, ci_lower, ci_upper, group, subgroup) %>% 
  pivot_wider(names_from = organism, 
              values_from = c(mean, median, p_value, p_adj, ci_lower, ci_upper, sig_cat), 
              names_sep = "_", names_sort = TRUE) 



dge_btm_mean_process  %>% 
  ggplot() +
  aes(x = mean_Mouse,
      y = mean_Human) +
  facet_wrap(~timepoint, scales = "free") +
  geom_point(aes(fill = group), alpha = 1, size = 3, 
             color = "black", pch = 21,
             data = . %>% filter(p_value_Mouse <= 0.25| p_value_Mouse <= 0.25)) +
  geom_point(aes(fill = group), alpha = 0.2, size = 3, 
             color = "black", pch = 21,
             data = . %>% filter(p_value_Mouse > 0.25| p_value_Mouse > 0.25)) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_blank(),
        legend.position = "none") +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "Mouse Log2FC", 
       y = "Human Log2FC", 
       color = "",
       fill = "",
       title = "Mouse vs. Human, BTMs,\nEngerixB vaccination")

```





```{r}
#Import
dge_btm_mean_process_samples = 
  readRDS(file = here("tables", "dge_Ecoli_btm_mean_process_samples.rds")) %>% 
  mutate(timepoint = if_else(is.na(timepoint), "24", timepoint)) %>% 
  filter(timepoint == 24)


dge_btm_mean_process = dge_btm_mean_process_samples %>% 
  group_by(organism, timepoint, process) %>% 
  summarize(
    mean = mean(log2fc, na.rm = TRUE) %>% round(2),
    median = median(log2fc, na.rm = TRUE) %>% round(2),
    min = min(log2fc, na.rm = TRUE) %>% round(2),
    max = max(log2fc, na.rm = TRUE) %>% round(2),
    sd = sd(log2fc, na.rm = T),
    n = sum(!is.na(log2fc)),
    se = sd/sqrt(n),
    t_critical = qt(0.975, df = n - 1),
    ci_lower = mean - t_critical * se,
    ci_upper = mean + t_critical * se,
    p_value = t.test(log2fc, mu = 0)$p.value,
    .groups = "drop",
    group,
    subgroup
  ) %>%
  distinct() %>% 
  mutate(p_adj = p.adjust(p_value, method = "BH"), 
         sig_cat = case_when(p_adj < 0.01 ~ "< 0.01", 
                             p_adj < 0.05 ~ "< 0.05",
                             p_adj < 0.10 ~ "< 0.10", 
                             p_adj < 0.25 ~ "< 0.25", 
                             TRUE ~ "≥ 0.25"),
         sig_cat = factor(sig_cat,
                   levels = c("< 0.01", "< 0.05", "< 0.10", "< 0.25", "≥ 0.25"))) %>% 
  select(organism, timepoint, process, mean, median, p_adj, p_value, sig_cat, ci_lower, ci_upper, group, subgroup) %>% 
  pivot_wider(names_from = organism, 
              values_from = c(mean, median, p_value, p_adj, ci_lower, ci_upper, sig_cat), 
              names_sep = "_", names_sort = TRUE) 



dge_btm_mean_process %>%
  mutate(
    significance = case_when(
      p_adj_Mouse <= 0.25 & p_adj_Human <= 0.25 ~ "Significant in both",
      p_adj_Mouse > 0.25 & p_adj_Human > 0.25 ~ "Non-significant in both",
      TRUE ~ "Mixed"
    ),
    significance = factor(
      significance,
      levels = c("Significant in both", "Mixed", "Non-significant in both")
    )
  ) %>%
  ggplot(aes(x = mean_Mouse, y = mean_Human)) +
  facet_wrap(~timepoint, scales = "free") +
  
  geom_point(
  aes(fill = group, color = group, shape = significance),
  size = 3,
  stroke = 0.7) +
scale_fill_manual(values = immune_colors, name = "Group") +
scale_color_manual(values = immune_colors, guide = "none") + # esconde legenda duplicada
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  
  theme_vaxgo() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    strip.text.x = element_text(size = 10),
    strip.text.y = element_blank()
  ) +
  
#   scale_fill_manual(values = immune_colors, name = "Group") +
#   guides(
#   fill = guide_legend(override.aes = list(color = "black"))
# ) +
  scale_shape_manual(
    values = c(
      "Significant in both" = 21,  # círculo cheio
      "Mixed" = 24,                # triângulo
      "Non-significant in both" = 22  # quadrado
    ),
    name = "Adjusted p-value match"
  ) +
  
  labs(
    x = "Mouse Log2FC",
    y = "Human Log2FC",
    title = "Mouse vs. Human, BTMs,\nEngerixB vaccination"
  )
```



#Scatterplot of correlation (arrows, day)
```{r fig.height=3, fig.width=10}

zscore_genes_plot_timepoints = zscore_genes %>% 
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = mean_human,
      y = mean_mouse) +
  #All points, colored by zscore
  geom_point(color ="gray90",
             fill = NA,
             pch = 21,
             alpha = 0.4) +
  #Select
  geom_point(aes(shape = as.character(timepoint)),
             # pch = 21, 
             alpha = 0.5,
             # stroke = 0.5, 
             size = 2, 
             color = "#4DBBD5FF",
            data =degs_pos_neg_mouse %>% 
              filter(quantile_label %in% c("Q10", "Q25", "Q50"))) +
  geom_path(aes(group = human_symbol),
            arrow=arrow(length=unit(0.05,"in")), 
            show.legend= FALSE,
            alpha = 0.5,
            data = degs_pos_neg_mouse %>% 
              filter(quantile_label %in% c("Q10", "Q25", "Q50")),
            # color = "black"
            ) +
  # scale_color_manual(values = c("1" = "#4DBBD5",
  #                               "3" = "#A6DDEA",
  #                               "7" = "white")) +
    scale_x_continuous(limits = c(-0.6, 0.6)) +
  scale_y_continuous(limits = c(-0.6, 0.6)) +
  geom_abline(alpha = 0.5, linetype = 2) +
  geom_abline(slope = -1, alpha = 0.5, linetype = 2) +
  facet_wrap(~quantile_label) + 
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
         legend.title = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.key = element_rect(colour = NA)) +
  labs(title = "Are these genes well predicted in most timepoints?",
       subtitle = "Differences in mouse to human, using DEGs in human AND mouse, p_adjust < 0.05.\nAll timepoints. Only Q50 genes.",
       x = "Human L2FC",
       y = "Mouse L2FC",
       fill = "Timepoint",
       color = "Timepoint",
       shape = "Timepoint")  

zscore_genes_plot_timepoints$layers[[2]]

```



```{r}

degs_segments <- degs_pos_neg_mouse %>%
  filter(quantile_label %in% c("Q25")) %>%
  mutate(quantile_label = if_else(quantile_label == "Q25", "<Q25", quantile_label)) %>% 
  bind_rows(degs_pos_neg_mouse_split %>% 
    mutate(quantile_label = if_else(quartiles_diff == "Q100" | 
                                      quartiles_diff == "Q75", "Q75>", 
                                    quartiles_diff)) %>% 
      filter(quantile_label == "Q75>")) %>% 
  arrange(human_symbol, timepoint) %>%
  group_by(human_symbol, quantile_label) %>%
  mutate(
    mean_human_next = lead(mean_human),
    mean_mouse_next = lead(mean_mouse),
    timepoint_next = lead(timepoint),
    direction = paste0("D", timepoint, "→D", timepoint_next)
  ) %>%
  filter(!is.na(mean_human_next), !is.na(mean_mouse_next)) %>%
  ungroup() %>% 
  filter(!(mean_human == mean_human_next & mean_mouse == mean_mouse_next))
  
  

degs_segments
```


```{r}
# Plot com caminhos coloridos por direção
zscore_genes_plot_timepoints <- zscore_genes %>% 
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = mean_human, y = mean_mouse) +
  # Curva com seta no fim
  geom_curve(data = degs_segments,
             aes(x = mean_human, y = mean_mouse,
                 xend = mean_human_next, yend = mean_mouse_next,
                 color = direction,
                 group = human_symbol),
             curvature = 0,     # Ajuste a curvatura conforme desejado
             size = 1,
             arrow = arrow(length = unit(0.1, "in")),
             alpha = 0.5,
             lineend = c('round'),
              linejoin = c('mitre'),
             show.legend = TRUE) +
    scale_x_continuous(limits = c(-0.6, 0.6)) +
  scale_y_continuous(limits = c(-0.6, 0.6)) +
  geom_abline(alpha = 0.5, linetype = 2) +
  geom_abline(slope = -1, alpha = 0.5, linetype = 2) +
  facet_wrap(~quantile_label) + 
   scale_color_manual(values = c(
  "D1→D3" = "#7400b8",
  "D3→D7" = "#4dbbd5",
  "D1→D7" = "#3c5488"
)) + 
  scale_fill_manual(values = c(
  "D1→D3" = "#7400b8",
  "D3→D7" = "#4dbbd5",
  "D1→D7" = "#3c5488"
)) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.key.width = unit(0.4, 'cm'),
        legend.key.height = unit(0.4, 'cm'),
        legend.title = element_text(size = 10, vjust = 0.9),
        legend.text = element_text(size = 7, vjust = 0.5, hjust = -0.2),
        legend.key = element_rect(colour = NA)) +
  labs(title = "Are these genes well predicted in most timepoints?",
       subtitle = "Differences in mouse to human, using DEGs in human AND mouse, p_adjust < 0.05.\nAll timepoints.",
       x = "Human L2FC",
       y = "Mouse L2FC",
       fill = "Timepoint",
       color = "Timepoint",
       shape = "Timepoint")  

zscore_genes_plot_timepoints
```


```{r}
zscore_genes_plot_timepoints_bar = degs_segments %>% 
  group_by(quantile_label, direction) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  ggplot() +
  aes(x = direction, y = n, fill = direction) +
  geom_col(position = "dodge2", alpha = 0.5) +
  geom_text(aes(label = n),
            position = "dodge2", 
            color = "black",
            hjust = 0.5,
            vjust = -0.5
            ) +
  scale_color_manual(values = c(
  "D1→D3" = "#7400b8",
  "D3→D7" = "#4dbbd5",
  "D1→D7" = "#3c5488"
)) + 
  scale_fill_manual(values = c(
  "D1→D3" = "#7400b8",
  "D3→D7" = "#4dbbd5",
  "D1→D7" = "#3c5488"
)) +
  # scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 110))+
  facet_wrap(~quantile_label) + 
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.position = "none") +
  labs(
    # title = "Are these genes well predicted in most timepoints?",
       # subtitle = "Differences in mouse to human, using DEGs in human AND mouse, p_adjust < 0.05.\nAll timepoints. Only Q50 genes.",
       x = "Direction",
       y = "Total",
       fill = "Direction")  

  zscore_genes_plot_timepoints_bar
```


#Glassplot

```{r}
library(ggplot2)
library(ggfx)
library(dplyr)
library(MASS)
library(here) # Para salvar

# ---------------------------------------------------------
# 1. GERAR DADOS
# ---------------------------------------------------------
set.seed(101)
n_genes <- 1000

# Correlação alvo (r = 0.9)
target_r <- 0.9
sigma_matrix <- matrix(c(1, target_r, target_r, 1), nrow = 2)

dados_brutos <- mvrnorm(n = n_genes, mu = c(0, 0), Sigma = sigma_matrix)

df_log2fc <- data.frame(
  human_logfc = dados_brutos[, 1] * 2.5, 
  mouse_logfc = dados_brutos[, 2] * 2.5 
) %>%
  mutate(
    # --- LÓGICA DE RANKING (Para as CAMADAS do Blur) ---
    distancia = abs(human_logfc) + abs(mouse_logfc),
    fake_pval = 1 / (distancia + runif(n(), 0, 2)), 
    pval_rank = rank(fake_pval), # 1 a 1000
    
    # --- LÓGICA DE COR (Para o FILL/COLOR) ---
    # Aqui criamos grupos que NÃO dependem do ranking.
    # Simula vias biológicas diferentes misturadas nos dados.
    pathway_group = sample(c("Inflammation", "Cell Cycle"), 
                           n(), replace = TRUE)
  )

# Calcular R² para o texto
r_squared <- round(cor(df_log2fc$human_logfc, df_log2fc$mouse_logfc)^2, 2)
label_text <- paste0("R² ≈ ", r_squared)

# ---------------------------------------------------------
# 2. DEFINIÇÃO DE CORES
# ---------------------------------------------------------
# Cores para os grupos (e não para o rank)
my_colors <- c(
  "Inflammation" = "#607d8b", # Vermelho
  "Cell Cycle"   = "#455a64"  # Azul
)

# ---------------------------------------------------------
# 3. PLOT COM GGFX
# ---------------------------------------------------------
glass_plot <- ggplot(df_log2fc, aes(x = human_logfc, y = mouse_logfc)) +
  
  # Guias de fundo
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey80") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey80") +
  
  # --- CAMADA 1: BACKGROUND (Todos os 1000 pontos) ---
  # Blur forte. Usamos shape 16 (sem borda) para um desfoque mais suave.
  with_blur(
    geom_point(
      data = . %>% filter(pval_rank > 500),
      aes(color = pathway_group), # Colore pelo grupo
      size = 3,
      alpha = 0.5,
      shape = 16
    ),
    sigma = 5 # Blur alto
  ) +
  
  # --- CAMADA 2: MIDGROUND (Top 500 Rank) ---
  # Blur médio. Filtrado pelo Rank, mas colorido pelo Grupo.
  with_blur(
    geom_point(
      data = . %>% filter(pval_rank <= 500),
      aes(color = pathway_group),
      size = 3,
      alpha = 0.8,
      shape = 16
    ),
    sigma = 2 # Blur médio
  ) +
  
  # --- CAMADA 3: FOREGROUND (Top 100 Rank - Foco) ---
  # Nítido. Usamos shape 21 para ter borda preta e preenchimento (fill) colorido.
  with_shadow(
    geom_point(
    data = . %>% filter(pval_rank <= 300),
    aes(fill = pathway_group,
        color = pathway_group), # Note que aqui usamos FILL, não color
    size = 3,
    # color = "black",
    shape = 21,                # Círculo preenchido
    stroke = 0.2,                # Espessura da borda preta
    alpha = 1
  ),
  x_offset = 2,
  y_offset = 2,
  sigma = 3
  ) +
  
  geom_smooth(method = "lm", se = FALSE, color = "#c59d45", size = 1.5) +
  
  # --- TEMA E ESCALAS ---
  scale_color_manual(values = my_colors) + # Para as camadas 1 e 2 (blur)
  scale_fill_manual(values = my_colors) +  # Para a camada 3 (foco)
  coord_cartesian(xlim = c(-7, 7), ylim = c(-7, 7)) +
  theme_classic() +
  labs(
    x = "Human Log2FC", 
    y = "Mouse Log2FC", 
    fill = "Pathway", 
    color = "Pathway"
  ) +
  theme(
    axis.line = element_line(color = "black", linewidth = 1),
    axis.text = element_text(size = 12, color = "black"),
    axis.title = element_text(size = 14, color = "black"),
    legend.position = "top"
  )

# Exibir o plot
print(glass_plot)

# Salvar
ggsave(filename = here("Figures", "glassplot_test.png"), plot = glass_plot, width = 6, height = 6)


```



#=================== 
#TO DO

## ORA of best and worst quartiles

```{r fig.width=10, fig.height= 8}
# Enrichment analysis -------
organism = "org.Hs.eg.db"
#Hallmarks
#Retrieve human H (hallmark) gene set
msigdbr_raw <- msigdbr(species = "human", category = "H") 
#Clean
msigdbr_df = msigdbr_raw %>% 
  select(process = gs_name, genes = gene_symbol) %>% 
  mutate(process = str_remove(process, "HALLMARK_"))

#Discretize by quartile
df_genes_best_worst_fit_mouse = zscore_genes %>% 
  ungroup() %>% 
  mutate(quartiles_grouped = case_when(
    quartile_label %in% c("Q10", "Q25") ~ "<Q25",
    quartile_label %in% c("Q75", "Q100") ~ ">Q75",
    TRUE ~ quartile_label)) %>% 
    mutate(condition = paste0(comparison, "", quartiles_grouped, " (", timepoint_day,")")) %>% 
  filter(quartiles_grouped != "Q50") %>% 
  select(genes = human_symbol,
         condition)

#Run ORA Hallmarks
ora_all_diff_hallmark <- enricher_condition(
  df = df_genes_best_worst_fit_mouse,
  term2gene_df = msigdbr_df) %>% 
  mutate(log_p_adjust = -log(p_adjust)) %>% 
  mutate(description = str_replace_all(description, "_", " ")) %>% 
  mutate(timepoint = str_extract(condition, "(?<=\\(Day )\\d+")) %>% 
  mutate(comparison = case_when(str_detect(condition, "FIT") ~ "FIT",
                                str_detect(condition, "Mouse") ~ "Mouse"))
#Save
ora_all_diff_hallmark %>% 
  saveRDS(file = here("tables", "ora_all_diff_hallmark.rds"))

#Print
ora_all_diff_hallmark
```


```{r fig.height=8, fig.width=8}
##### Plot ORA Hallmark ------
#Import
ora_all_diff_hallmark = readRDS(file = here("tables", "ora_all_diff_hallmark.rds"))
#Cluster rows
row_clustered = ora_all_diff_hallmark %>% 
    dplyr::select(condition, description, log_p_adjust) %>% 
    distinct() %>% 
    pivot_wider(names_from = "condition",
                values_from= "log_p_adjust") %>%
    column_to_rownames("description") %>% 
    as.matrix() %>% 
    replace(is.na(.), 0) %>% 
    dist(.) %>% 
    hclust()

#Plot
ora_all_diff_hallmark_heatmap = ora_all_diff_hallmark %>% 
  clean_names() %>% 
  filter(description %in% c(row_clustered$labels)) %>% 
  ungroup() %>% 
  select(condition, description, log_p_adjust, p_adjust, everything()) %>% 
    mutate(signific = case_when(p_adjust <= 0.10 ~ "*",
                              p_adjust <= 0.25 ~ "-",
                              p_adjust <= 0.50 ~ "/",
                              p_adjust <= 0.05 ~ "**",
                              TRUE ~ ""),
           quantile = str_extract(condition, "Q\\d{2,3}"),
        condition = paste(quantile)) %>% 
  ggplot(aes(x=condition, y = str_to_upper(description), fill = log_p_adjust)) +
  geom_tile(color = "black", 
            size = 0.2, na.rm = F, stat = "identity") +
  geom_text(aes(label = signific), size = 4, vjust = 0.5, color = "black") +
  scale_fill_gradient2(
        low = "white",     
        high = "#4DBBD5FF",   
        n.breaks = 3,
        oob = scales::squish,
        limits = c(0, 4),      # fixa o intervalo
        name = "-Log(FDR)") +
    theme_vaxgo() +
    theme(axis.text = element_text(color = "black", size = 8) ,
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 8),
          legend.key.width = unit(0.2, 'cm'),
          legend.key.height = unit(0.4, 'cm'),
          axis.line.y  = element_blank(),
          axis.text.x = element_text(angle = 0, vjust = 1, hjust=0.5),
          axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1),
          axis.ticks = element_blank(),
          strip.background.x = element_rect(fill = "black"),
          strip.text.x = element_text(color = "white"),
          panel.grid = element_blank(),
          plot.margin = margin(0, 0.5, 0, 1, unit = "cm"),
          legend.direction = "horizontal",
          legend.position = "top") +
    scale_y_discrete(position = "right") +
     labs(x = "",
         y = "",
         title = "ORA, MSigDB Hallmarks") +
  facet_nested(~ paste("Day", timepoint) + fct_rev(comparison), scales = "free", space = "free") 

#Print
ora_all_diff_hallmark_heatmap

#Save
ora_all_diff_hallmark_heatmap %>%
  ggsave(filename = here("Figures", "ORA_Best_Worst_heatmap.png"),
         width = 8,
         height = 8)
```

#### Top best and worst

```{r}
#Import data
pairwise_human_fit_measures = readRDS(here("tables", "influenza_fluad_human_mouse_fit_dge_pairwise2human.rds")) 
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
# Top best and worst predicted genes in FIT and Mouse -----
pairwise_human_fit_genes_long =  pairwise_human_fit_measures %>% 
  pivot_longer(cols = c(mean_human, mean_mouse, mean_fit),
               values_to = "l2fc",
               names_to = "organism") %>% 
  rename(diff_h2m = human_2_mouse,
         diff_h2fit = human_2_fit) %>% 
  mutate(organism = if_else(organism == "FIT Prediction", "FIT", organism)) %>% 
  
  pivot_longer(cols = -c(symbol, timepoint, organism),
               values_to = "value",
               names_to = "metric") %>% 
  mutate(
    organism = case_when(
      str_detect(metric, "human") ~ "Human",
      str_detect(metric, "fit") ~ "FIT",
      str_detect(metric, "mouse") ~ "Mouse",
      metric == "Human" ~ "Human",
      metric == "FIT Prediction" ~ "FIT",
      TRUE ~ organism),
    metric = str_remove_all(metric, "_?(human|mouse|fit)$"),
    metric = str_remove_all(metric, "_?(human|mouse|fit)")
  ) %>% 
  mutate(organism = case_when(str_detect(metric, "h2m") ~ "Mouse",
                              str_detect(metric, "h2") ~ "FIT",
                              T~organism))  %>% 
  mutate(organism = case_when(organism == "mean_fit" ~ "FIT",
                              organism == "mean_human" ~ "Human",
                              organism == "mean_mouse" ~ "Mouse",
                              T ~ organism))

#Comparisons
pairwise_human_fit_measures_comparisons = pairwise_human_fit_measures %>%
    select(timepoint, symbol, human_2_mouse:effect_size_h2m) 

#Pivot table and join the comparisons
pairwise_human_fit_genes_wide = pairwise_human_fit_genes_long %>% 
  pivot_wider(names_from = "metric",
              values_from = "value", 
              values_fn = mean) %>% 
  select(!c(pval:effect_size_h2m)) %>% 
  inner_join(pairwise_human_fit_measures_comparisons,
                by = join_by("timepoint", "symbol")) 


#Get top genes in each model
top_genes_human = pairwise_human_fit_genes_wide %>% 
  filter(organism == "Human") %>% 
  relocate(ci_width,  .after = organism) %>% 
  group_by(timepoint) %>% 
  arrange(ci_width) 
  
 top_genes_mouse = pairwise_human_fit_genes_wide %>% 
  filter(organism == "Mouse",
         symbol %in% c(top_genes_human$symbol)) %>% 
   arrange(ci_width, desc(p_value_h2m_adj), human_2_mouse)
 
 top_genes_fit = pairwise_human_fit_genes_wide %>% 
  filter(organism == "FIT",
         symbol %in% c(top_genes_human$symbol)) %>% 
   arrange(ci_width, desc(p_value_h2fit_adj), human_2_fit)
```

##### Plot confidence intervals for each top gene

```{r fig.height=7, fig.width=8}
#Plot -----
#Mouse to human
fit_example_human_plot_mouse = 
  pairwise_human_fit_genes_wide %>% 
  # rename(l2fc = value) %>% 
  filter(symbol %in% c(top_genes_mouse %>% 
                               select(symbol) %>% 
                               distinct() %>% 
                               slice_head(n = 5) %>% 
                               pull(symbol))) %>% 
  arrange(symbol, timepoint) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  facet_wrap(~symbol, nrow = 1) +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1), 
                show.legend = F
                ) +
    geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_color_manual(values = colors$organism)+
  scale_fill_manual(values = colors$organism)+
    theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        strip.text.x = element_text(color = "black"),
        strip.background = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "Time (days)",
       y = "LogFold-Change",
       title = "Prediction - Influenza, Fluad Vaccine\nTop 5 MOUSE genes with lowest CI widths and l2fc differences to human l2fc, with highest padj")


#FIT to human
fit_example_human_plot_fit= 
  pairwise_human_fit_genes_wide %>% 
  filter(symbol %in% c(top_genes_fit %>% 
                               select(symbol) %>% 
                               distinct() %>% 
                               slice_head(n = 5) %>% 
                               pull(symbol))) %>% 
  arrange(symbol, timepoint) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  facet_wrap(~symbol, nrow = 1) +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1), show.legend = F
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
    scale_color_manual(values = colors$organism)+
  scale_fill_manual(values = colors$organism)+
    theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        strip.text.x = element_text(color = "black"),
        strip.background = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "Time (days)",
       y = "LogFold-Change",
       title = "Top 5 FIT genes with lowest CI widths and l2fc differences to human l2fc, with highest padj")

#Print
fit_example_human_plot_mouse/
fit_example_human_plot_fit
```

How many genes are not significantly different from human gene expression?

```{r fig.height=8, fig.width=8}
# #How many genes are not significantly different from human gene expression? ------
# pairwise_human_fit_genes_size_plot = pairwise_human_fit_genes_wide %>% 
#   # filter(!is.na(p_value_diff_human)) %>% 
#   mutate(
#     # p_value_diff_human = if_else(organism == "Human", pval_adj, p_value_diff_human),
#     `<0.01` = abs(p_value_h2fit_adj) <= 0.01,
#     `<0.05` = abs(p_value_h2fit_adj) <= 0.05,
#     `<0.10` = abs(p_value_h2fit_adj) <= 0.10,
#     `<0.25` = abs(p_value_h2fit_adj) <= 0.25,
#     # `<1.00` = abs(p_value_diff_human) <= 1.00
#   ) %>%
#   pivot_longer(cols = starts_with("<"),
#                names_to = "threshold_label",
#                values_to = "in_threshold") %>%
#   mutate(threshold_label = forcats::fct_relevel(
#     threshold_label, "<0.01", "<0.05", "<0.10", "<0.25", "<1.00"
#   )) %>%
#   filter(in_threshold) %>%
#   select(-in_threshold) %>% 
#   group_by(organism, timepoint, threshold_label) %>% 
#   summarise(n = n()) %>% 
#   ungroup() %>% 
#   mutate(timepoint = as.character(timepoint) %>% 
#            fct_relevel(., "1", "3", "7")) %>% 
#   ggplot() + 
#   aes(x = timepoint,
#       y = n, 
#       fill = organism) +
#   geom_col(position = position_dodge2(1), show.legend = F,
#            alpha = 0.2) +
#   geom_line(aes(group = organism,
#                 color = organism),
#             position = position_dodge2(1), size =1) +
#   geom_point(position = position_dodge2(1),
#              pch = 21,
#              color = "black") +
#   scale_y_continuous(expand = c(0.01, 0.01)) +
#   facet_grid(~threshold_label, scales = "free_y") +
#     scale_fill_manual(
#   values = c(
#     FIT = "#4361ee", 
#     Mouse = "#4DBBD5FF",
#     Human = "#546E7AFF"))+
#     scale_color_manual(
#   values = c(
#     FIT = "#4361ee", 
#     Mouse = "#4DBBD5FF",
#     Human = "#546E7AFF"))+
#     theme_vaxgo() +
#   theme(axis.text.x = element_text(angle = 0),
#         legend.key = element_blank(),
#         strip.text.x = element_text(color = "black"),
#         # strip.background = element_rect(fill = "gray95", color = NA)
#         ) +
#   labs(fill = "Organism",
#        color = "Organism",
#        x = "Time (days)",
#        y = "Total",
#        title = "Total of genes with high similarity by p value cutoff")
# 
# pairwise_human_fit_genes_size_plot
```

#####Plot histogram of genes by cutoff

```{r fig.height=5, fig.width=3}
# #Plot histogram of genes by cutoff
# pairwise_human_fit_genes_wide_distribution_data = pairwise_human_fit_genes_wide %>% 
#   # filter(!is.na(p_value_diff_human)) %>% 
#   mutate(
#     # p_value_diff_human = if_else(organism == "Human", pval_adj, p_value_diff_human),
#     `<0.01` = abs(pval_adj) <= 0.01,
#     `<0.05` = abs(pval_adj) <= 0.05,
#     `<0.10` = abs(pval_adj) <= 0.10,
#     `<0.25` = abs(pval_adj) <= 0.25,
#     `<1.00` = abs(pval_adj) <= 1.00
#   ) %>%
#   pivot_longer(cols = starts_with("<"),
#                names_to = "threshold_label",
#                values_to = "in_threshold") %>%
#   mutate(threshold_label = forcats::fct_relevel(
#     threshold_label, "<0.01", "<0.05", "<0.10", "<0.25", "<1.00"
#   )) %>%
#   filter(in_threshold) %>%
#   select(-in_threshold) 
#   
# options(scipen = 999)
# pairwise_human_fit_genes_wide_distribution = pairwise_human_fit_genes_wide_distribution_data %>% 
#   filter(!organism == "Human") %>% 
#   ggplot() +
#   aes(x = mean) +
#   geom_density(aes(y = after_stat(count),
#                    fill = organism,
#                    # fill = organism,
#                    ), 
#                   position = position_identity(), 
#                alpha = 0.5,
#                  show.legend = F) +
#   guides(fill = "none") +
#   geom_density(aes(y = after_stat(count),
#                    color = organism,
#                    # fill = organism,
#                    ), 
#                   position = position_identity(), alpha = 0.5,
#                  show.legend = F) +
#   geom_density(aes(y = after_stat(count),
#                    color = organism), 
#                position = position_identity(),
#                data = pairwise_human_fit_genes_wide_distribution_data %>% 
#                     filter(organism == "Human"),
#                  fill = NA,
#                show.legend = T,
#                alpha = 1) +
#   
#   facet_wrap(~threshold_label, 
#              scale = "free_y",
#              ncol = 1, 
#              strip.position="right") +
#   scale_y_continuous(expand = c(0.01, 0.01), n.breaks = 3, 
#                      # labels = scales::comma
#                      ) +
#     scale_fill_manual(
#   values = c(
#     FIT = "#4361ee", 
#     Mouse = "#4DBBD5FF",
#     Human = "#546E7AFF"))+
#     scale_color_manual(
#   values = c(
#     FIT = "#4361ee", 
#     Mouse = "#4DBBD5FF",
#     Human = "#546E7AFF"))+
#     theme_vaxgo() +
#   theme(axis.text.x = element_text(angle = 0),
#         legend.position = "top",
#         legend.key = element_blank(),
#         strip.text.x = element_text(color = "black", size = 10),
#         strip.text.y = element_text(color = "black", angle = 0), 
#         panel.spacing.y = unit(0.8, "lines"),
#         strip.background.y = element_rect(fill = NA, color = NA)) +
#   labs(fill = "",
#        color = "",
#        x = "Log2-fold change",
#        y = "Total",
#        title = "Frequency of genes by p value cutoff")
# 
# pairwise_human_fit_genes_wide_distribution
```

Unite plots

```{r fig.height=8, fig.width=12}

#Unite plots ------
a = fit_example_human_plot_mouse + theme(legend.position = "none")  + theme(plot.title = element_blank())
b = fit_example_human_plot_fit + guides(fill = "none") + theme(plot.title = element_blank())
c = pairwise_human_fit_genes_size_plot + theme(legend.position = "none") + guides(fill = "none") + theme(plot.title = element_blank())
d = pairwise_human_fit_genes_wide_distribution + theme(legend.position = "top",
                                                       legend.title = element_text(size = 10),
                                                       legend.text = element_text(size = 8),
                                                       axis.text.y = element_text(size = 8),
                                                       plot.title = element_blank())


plot_patchwork_1 = d + (c/a/b) +
  plot_layout(widths = c(0.2, 1), 
              axis_titles = "collect",
              axes = "collect")

plot_patchwork_1

#Save
plot_patchwork_1 %>%
  ggsave(filename = here("Figures", "influenza_fluad_mouse_fit_prediction_model2human_top5_fit_and_mouse.png"),
         width = 12,
         height = 8)



```

##### Get best and worst genes predicted

Get best and worst genes predicted by mouse and FIT, and see what can be better Difference in ranks

```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
# Get best and worst predicted genes ------
pairwise_genes_wide_mouse_fit = pairwise_human_fit_genes_wide %>% 
  filter(organism != "Human",
         symbol %in% c(top_genes_human$symbol)) 

#Get comparisons
z_scores_best_worst_models_comparisons = pairwise_human_fit_measures_comparisons %>% 
  mutate(diff_z_fit_mouse = abs(z_score_h2fit) - abs(z_score_h2m),
         diff_z_mouse_fit = abs(z_score_h2m) - abs(z_score_h2fit)) 

#Best and worst FIT
best_fit = z_scores_best_worst_models_comparisons %>% 
  arrange(abs(z_score_h2fit))

worst_fit = z_scores_best_worst_models_comparisons %>% 
  arrange(-abs(z_score_h2fit))

#Best and worst Mouse
best_mouse = z_scores_best_worst_models_comparisons %>% 
  arrange(abs(z_score_h2m))

worst_mouse = z_scores_best_worst_models_comparisons %>% 
  arrange(-abs(z_score_h2m))

#Merge FIT vs. Mouse
worst_fit_mouse = z_scores_best_worst_models_comparisons %>% 
  arrange(-diff_z_fit_mouse)

#Merge Mouse vs. FIT
worst_mouse_fit = z_scores_best_worst_models_comparisons %>% 
  arrange(-diff_z_mouse_fit)
```

Plots

```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
#Plot best vs worst -----
#FIT ------
#Best FIT
best_fit_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(best_fit %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~human_symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Top genes in FIT prediction. Difference in Z-score to human")

#Worst FIT
worst_fit_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(worst_fit %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Worst predicted genes in Mouse. Difference in Z-score to human")

#Mouse -----
#Best mouse
best_mouse_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(best_mouse %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~human_symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Top genes in Mouse prediction. Difference in Z-score to human")

#Worst Mouse
worst_mouse_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(worst_mouse %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~human_symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Worst predicted genes in Mouse. Difference in Z-score to human")


#FIT and Mouse -------
#Worst FIT to best Mouse
worst_fit_mouse_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(worst_fit_mouse %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~human_symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Genes that were most different in FIT prediction from mouse and human. Difference in Z-score")

#Worst mouse to best FIT
worst_mouse_fit_plot = pairwise_human_fit_genes_wide %>% 
  arrange(symbol, timepoint) %>% 
  inner_join(worst_mouse_fit %>% 
              select(symbol, timepoint) %>%
               group_by(timepoint) %>% 
              mutate(day = paste0("Day ", timepoint)) %>% 
              distinct() %>% 
              slice_head(n = 5),
            by = join_by("symbol", "timepoint")) %>% 
  ggplot() +
  aes(x = timepoint,
      y = l2fc,
      color = organism,
      fill = organism) +
  geom_hline(yintercept = 0, color = "gray75", alpha = 1, linetype = 2) +
  geom_line(aes(group = organism), position = position_dodge2(1)) +
  # facet_wrap(~day~human_symbol, nrow = 3, scales = "free") +
  facet_grid(~day~symbol, scales = "free") +
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high),
                position =  position_dodge(1)
                ) +
      geom_point( position = position_dodge(1),
             color = "black",
             pch = 21) +
  scale_x_continuous(breaks = c(1, 3, 7),minor_breaks = F) +
  scale_y_continuous(n.breaks = 3) +
  scale_color_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
  scale_fill_manual(
  values = c(
    "FIT" = "#4361ee", 
    "Mouse" = "#546E7AFF",
    "Human" = "#4DBBD5FF"))+
    theme_vaxgo() +
  theme(axis.text.x = element_blank(),
        strip.text.x = element_text(color = "black"),
        strip.text.y = element_text(color = "black", angle = 0),
        strip.background.x = element_rect(fill = "gray95", color = NA), 
        panel.grid.major.y = element_blank()) +
  labs(color = "Organism",
       x = "",
       y = "LogFold-Change",
       title = "Genes that were most different in Mouse prediction from FIT and human. Difference in Z-score")


worst_mouse_fit_plot
```

##### Get ranks

```{r}
#Get ranks -----
#Mouse
ranks_mouse = pairwise_human_fit_genes_wide %>%
  filter(organism == "Mouse") %>%
  select(symbol, timepoint, z_score_h2m, organism) %>%
  group_by(timepoint) %>% 
  arrange(timepoint, abs(z_score_h2m)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2m)),
         percentile = round(percent_rank * 100, 2)) %>% 
  distinct() %>% 
  rename(zscore_h2model = z_score_h2m)
#FIT
ranks_fit = pairwise_human_fit_genes_wide %>%
  filter(organism == "FIT") %>%
  select(symbol, timepoint, z_score_h2fit, organism) %>%
  group_by(timepoint) %>% 
  arrange(timepoint, abs(z_score_h2fit)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2fit)),
         percentile = round(percent_rank * 100, 2)) %>% 
  distinct() %>% 
  rename(zscore_h2model = z_score_h2fit)

#Fit vs Mouse
ranks_fit_vs_mouse = ranks_fit %>% 
  select(symbol, fit = percentile, timepoint_ref = timepoint, zscore_h2fit = zscore_h2model) %>% 
  inner_join(ranks_mouse %>% 
               select(symbol, mouse = percentile,
                      timepoint_other = timepoint,
                      zscore_h2mouse = zscore_h2model),
             by = "symbol") %>% 
  mutate(timepoint_ref = as.character(timepoint_ref),
         timepoint_other = as.character(timepoint_other),
         condition = "Top FIT vs Mouse")

#Mouse vs FIT
ranks_mouse_vs_fit = ranks_mouse %>% 
  select(symbol, mouse = percentile, timepoint_ref = timepoint, zscore_h2mouse = zscore_h2model) %>% 
  inner_join(ranks_fit %>% 
               select(symbol, fit = percentile,
                      timepoint_other = timepoint,
                      zscore_h2fit = zscore_h2model),
             by = "symbol") %>% 
  mutate(timepoint_ref = as.character(timepoint_ref),
         timepoint_other = as.character(timepoint_other),
         condition = "Top Mouse vs FIT")


#Get genes that are in the 10th percentile in one model, but are in the 90th percentile in the other
ranks_best_fit_and_mouse = ranks_fit_vs_mouse %>%
  filter(fit < 10, mouse > 90) %>%
  bind_rows(ranks_mouse_vs_fit %>%
              filter(mouse < 10, fit > 90)) %>%
  select(condition,
         symbol,
         timepoint_ref,
         timepoint_other, 
         fit, zscore_h2fit,
         mouse, zscore_h2mouse)

ranks_best_fit_and_mouse_long <- ranks_best_fit_and_mouse %>%
  pivot_longer(
    cols = c(fit, mouse),
    names_to = "model",
    values_to = "percentile"
  )


#Get genes with least zscores in one model, but have the highest zscore in the other
ranks_best_fit_and_mouse_zscore = ranks_fit_vs_mouse %>%

  # FILTER BY N
  filter(timepoint_ref == timepoint_other) %>%
  arrange(abs(zscore_h2fit), desc(abs(zscore_h2mouse))) %>%
  slice_head(n = 20) %>% 
  bind_rows(ranks_mouse_vs_fit %>%
              filter(timepoint_ref == timepoint_other) %>%
  arrange(abs(zscore_h2mouse), desc(abs(zscore_h2fit))) %>%
  slice_head(n = 20)) %>%
  select(condition,
         symbol,
         timepoint_ref,
         timepoint_other, 
         fit = zscore_h2fit,
         mouse = zscore_h2mouse)
#Pivot longer
ranks_best_fit_and_mouse_long_zscore <- ranks_best_fit_and_mouse_zscore %>%
  pivot_longer(
    cols = c(fit, mouse),
    names_to = "model",
    values_to = "zscore"
  )

#Print
ranks_best_fit_and_mouse_long_zscore

```

##### Plot percentiles

```{r fig.height=4, fig.width=8}
#Plot percentiles -------
#Select genes
selected_genes = ranks_best_fit_and_mouse %>% 
  group_by(condition, timepoint_ref, timepoint_other) %>% 
  filter(timepoint_ref == timepoint_other) %>% 
  # filter(fit < 1 | mouse <1) %>% 
  pivot_longer(
    cols = c(fit, mouse),
    names_to = "model",
    values_to = "percentile"
  ) 

#Plot
ranks_percentile_best_fit_and_mouse_long_plot = ranks_best_fit_and_mouse_long %>% 
  filter(timepoint_ref == timepoint_other) %>% 
  ggplot() +
  aes(x = model, y = percentile, group = interaction(condition, symbol), color = condition) +
  geom_line(alpha = 0.5) +
  guides(color = guide_legend(override.aes = list(linewidth = 2))) +
  geom_text_repel(data = selected_genes %>%
                    inner_join(ImmuneGO_BTM_grouped_genes %>% 
                                               select(symbol = genes) %>% 
                                               distinct() %>% 
                                               mutate(gene_function = "Immune"),
                  by = "symbol") %>% 
                    mutate(gene_function = "Other") %>% 
                    filter(condition == "Top FIT vs Mouse" & model == "mouse") %>% 
                    arrange(-percentile) %>% 
                    slice_max(percentile, n = 3),
             mapping = aes(label = symbol,
                           x = model,
                           y = percentile),
             color = "black",
             segment.color = "#4361ee",
             direction = c("y"),
             size = 2.5, 
            min.segment.length = 0.5,
             max.overlaps = 10000, 
             force = 2,
             nudge_x = 1, 
            hjust = "left") +
  
geom_text_repel(data = selected_genes %>%
                    filter(condition == "Top Mouse vs FIT" & model == "fit") %>% 
                  arrange(-percentile) %>% 
                  slice_max(percentile, n = 3),
                color = "black",
             segment.color = "#4DBBD5FF",
             direction = c("y"),
             mapping = aes(label = symbol,
                           x = model,
                           y = percentile),
             size = 2.5, 
             min.segment.length = 0.5,
             max.overlaps = 10000, 
             nudge_x = -1, 
             force = 2, hjust = "right") +
  facet_grid(~paste("Day", timepoint_ref))  +
  labs(x = "Model",
       y = "Percentile",
       color = ""
       # title = "Best 10% versus worst 100%"
       ) +
    scale_color_manual(values = c(
    "Top FIT vs Mouse" = "#4361ee", 
    "Top Mouse vs FIT" = "#4DBBD5FF")) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01), limits = c(0, 110)) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "right",
        legend.key = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0), 
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA))
  
#Print
ranks_percentile_best_fit_and_mouse_long_plot 
 
#Save
ranks_percentile_best_fit_and_mouse_long_plot %>% 
  ggsave(filename = here("Figures", "influenza_fluad_mouse_fit_prediction_ranks_best_plot.png"),
         width = 8,
         height = 5)
```

##### Plot ranks by zscores

```{r fig.height=4, fig.width=8}
#Plot ranks by zscores ------
#Select genes by zscore
selected_genes_zscore = ranks_best_fit_and_mouse_zscore %>%
  group_by(condition, timepoint_ref, timepoint_other) %>%
  pivot_longer(cols = c(fit, mouse),
               names_to = "model",
               values_to = "zscore") 

#Get the genes with the highest zscores in Mouse
worst_zscores_Mouse = selected_genes_zscore %>%
  filter(condition == "Top FIT vs Mouse" &
           model == "mouse") %>%
  arrange(-abs(zscore)) %>%
  slice_head(n = 3) %>%
  ungroup()

#Get the genes with the lowest zscores in Mouse
best_zscores_Mouse = selected_genes_zscore %>%
  filter(condition == "Top FIT vs Mouse" &
           model == "mouse") %>%
  arrange(abs(zscore)) %>%
  slice_head(n = 3) %>%
  ungroup()

#Get the genes with the highest zscores in FIT
worst_zscores_FIT = selected_genes_zscore %>%
  filter(condition == "Top Mouse vs FIT" &
           model == "fit") %>%
  arrange(-abs(zscore)) %>%
  slice_head(n = 3) %>% 
  ungroup()

#Get the genes with the lowest zscores in FIT
best_zscores_FIT = selected_genes_zscore %>%
  filter(condition == "Top Mouse vs FIT" &
           model == "fit") %>%
  arrange(abs(zscore)) %>%
  slice_head(n = 3) %>% 
  ungroup()

```

```{r fig.height=3, fig.width=8}
#Plot
ranks_best_worst_fit_and_mouse_long_zscore_plot = ranks_best_fit_and_mouse_long_zscore %>% 
  ungroup() %>% 
  ggplot() +
  aes(x = model, y = zscore, 
      group = interaction(condition, symbol), 
      color = condition) +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.3) +
   geom_line(alpha = 0.5) +
  
  #Get the worst zscores for mouse and see where they are in the fit
  geom_text_repel(data = worst_zscores_Mouse,
             mapping = aes(label = symbol,
                           x = model,
                           y = zscore),
             size = 2.5, 
             direction = c("y"),
             color = "black",
             segment.color = "black", 
             max.overlaps = 10000, 
             # force = 2,
             force_pull = 5,
             nudge_x = 1, 
             min.segment.length = 0.5, 
             hjust = "left") +
  
  #Get the worst zscores fit and see if they improved in mouse
  geom_text_repel(data = worst_zscores_FIT,
                  direction = c("y"),
                  mapping = aes(label = symbol,
                                x = model,
                                y = zscore),
                  color = "black",
                  segment.color = "black",
                  size = 2.5,
                  max.overlaps = 10000,
                  force = 2,
                  nudge_x = -1, 
                  min.segment.length = 0.5, 
                  hjust = "right")+


  facet_wrap(~paste("Day", timepoint_ref), scales = "free")  +
  labs(x = "Model",
       y = "Zscore",
       color = "") +
    scale_color_manual(values = c(
    "Top FIT vs Mouse" = "#4361ee", 
    "Top Mouse vs FIT" = "#4DBBD5FF")) +
  guides(color = guide_legend(override.aes = list(linewidth = 2))) +
  scale_x_discrete(expand = c(0.05, 0.05)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "right",
        legend.key = element_blank(),
        # legend.key.height = unit(0.4,"cm"),
        # legend.key.width = unit(0.4,"cm"),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0), 
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA)) 

#Print
ranks_best_worst_fit_and_mouse_long_zscore_plot
  
#Save
# ranks_best_worst_fit_and_mouse_long_zscore_plot %>% 
#   ggsave(filename = here("Figures", "ranks_best_worst_fit_and_mouse_long_zscore_plot.png"),
#          width = 8,
#          height = 5)
```

Combine

```{r fig.height=5, fig.width=8}
#Combine plots
percentile_zscore_ranks = (ranks_best_worst_fit_and_mouse_long_zscore_plot/ranks_percentile_best_fit_and_mouse_long_plot) +
  plot_layout(guides = "collect") 

#Save
ggsave(percentile_zscore_ranks,
       filename = here("Figures", "influenza_fluad_mouse_fit_prediction_ranks_zscore_patch.png"),
         width = 8,
         height = 5)
  
```

Plot all

```{r fig.height=8, fig.width=12}
#Combine plots -------
a = pairwise_human_fit_genes_wide_distribution + theme(legend.position = "top",
                                                       legend.direction = "horizontal",
                                                       plot.title = element_blank(),
                                                       legend.title = element_text(size = 10),
                                                       legend.text = element_text(size = 8),
                                                       axis.text.y = element_text(size = 10))
b = pairwise_human_fit_genes_size_plot + theme(legend.position = "right",
                                               legend.direction = "vertical",
                                               plot.title = element_blank(),
                                               legend.title = element_text(size = 10),
                                               legend.text = element_text(size = 8),
                                               axis.text.y = element_text(size = 10))
c = ranks_best_worst_fit_and_mouse_long_zscore_plot + theme(legend.position = "right", 
                                               legend.direction = "vertical",
                                               plot.title = element_blank(),
                                               legend.title = element_text(size = 10),
                                               legend.text = element_text(size = 8),
                                               axis.text.y = element_text(size = 10)) + 
  theme(title = element_blank())
d = ranks_percentile_best_fit_and_mouse_long_plot + theme(legend.position = "right", 
                                               legend.direction = "vertical",
                                               plot.title = element_blank(),
                                               legend.title = element_text(size = 10),
                                               legend.text = element_text(size = 8),
                                               axis.text.y = element_text(size = 10))

mouse_vs_fit_ranks_zscores_patch = a + (b/c/d) +
  plot_layout(widths = c(0.2, 1), 
              axis_titles = "collect"
              ) 


mouse_vs_fit_ranks_zscores_patch

#Save
mouse_vs_fit_ranks_zscores_patch %>% 
  ggsave(filename = here("Figures", "influenza_fluad_mouse_fit_prediction_ranks_zscore_patch.png"),
         width = 12,
         height = 8)
```

##### Heatmap: Best 10 vs Worst 90

Get best 10 and worst 90

```{r}
#Plot heatmap -----
#Get best 10 and worst 90
#Mouse
mouse_best_10 = best_mouse %>%
  inner_join(pairwise_genes_wide_mouse_fit %>% 
               select(timepoint:l2fc) %>% 
               filter(organism == "Mouse"),
             by = join_by("symbol", "timepoint")) %>% 
  # filter(organism == "Mouse") %>%
  select(symbol, z_score_h2m, l2fc, timepoint, organism) %>%
  group_by(organism, timepoint) %>% 
  arrange(abs(z_score_h2m)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2m)),
         percentile = round(percent_rank * 100, 2)) %>%
  filter(percentile <= 10) %>% 
  mutate(condition = paste("Mouse Best Day", timepoint))

mouse_worst_90 = best_mouse %>%
  inner_join(pairwise_genes_wide_mouse_fit %>% 
               select(timepoint:l2fc) %>% 
               filter(organism == "Mouse"),
             by = join_by("symbol", "timepoint")) %>% 
  select(symbol, z_score_h2m, l2fc, timepoint, organism) %>%
    group_by(organism, timepoint) %>% 
  arrange(abs(z_score_h2m)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2m)),
         percentile = round(percent_rank * 100, 2)) %>%
  filter(percentile >= 90) %>% 
  mutate(condition = paste("Mouse Worst Day", timepoint))

#FIT
fit_best_10 = best_fit %>%
  inner_join(pairwise_genes_wide_mouse_fit %>% 
               select(timepoint:l2fc) %>% 
               filter(organism == "FIT"),
             by = join_by("symbol", "timepoint")) %>% 
  select(symbol, z_score_h2fit, l2fc, timepoint, organism) %>%
  group_by(organism, timepoint) %>% 
  arrange(abs(z_score_h2fit)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2fit)),
         percentile = round(percent_rank * 100, 2)) %>%
  filter(percentile <= 10) %>% 
  mutate(condition = paste("FIT Best Day", timepoint))

fit_worst_90 = best_fit %>%
  inner_join(pairwise_genes_wide_mouse_fit %>% 
               select(timepoint:l2fc) %>% 
               filter(organism == "FIT"),
             by = join_by("symbol", "timepoint")) %>% 
  select(symbol, z_score_h2fit, l2fc, timepoint, organism) %>%
   group_by(organism, timepoint) %>% 
  arrange(abs(z_score_h2fit)) %>%
  mutate(percent_rank = percent_rank(abs(z_score_h2fit)),
         percentile = round(percent_rank * 100, 2)) %>%
  filter(percentile >= 90) %>% 
  mutate(condition = paste("FIT Worst Day", timepoint))

#Bind tables
mouse_fit_best_worst = bind_rows(mouse_best_10, 
                                 fit_best_10,
                                 mouse_worst_90, 
                                 fit_worst_90) 

#Get overlaps
overlap_best_worst = mouse_fit_best_worst %>%
  select(symbol, l2fc, condition, organism, timepoint)  %>% 
  ungroup()
  
```

Plot heatmap

```{r fig.height=8, fig.width=3}

#Plot heatmap
row_clustered = overlap_best_worst %>% 
    dplyr::select(condition, symbol, l2fc) %>% 
    distinct() %>% 
    pivot_wider(names_from = "condition",
                values_from= "l2fc") %>%
    column_to_rownames("symbol") %>% 
    as.matrix() %>% 
    replace(is.na(.), 0) %>% 
    dist(.) %>% 
    hclust()

df_genes_mouse_fit_best_worst_perc_hallmark_heatmap = overlap_best_worst %>% 
  clean_names() %>% 
  ungroup() %>% 
  select(condition, symbol, everything()) %>% 
    mutate(model = case_when(str_detect(condition, "Mouse") ~ "Mouse", 
                          str_detect(condition, "FIT") ~ "FIT",
                          T ~ NA),
           symbol = fct_relevel(symbol, c(row_clustered$labels)), 
           condition = if_else(str_detect(condition, "Best"), "Best", "Worst"))%>% 
  ggplot(aes(x=model, y = symbol, fill = l2fc, color= l2fc)) +
  geom_tile(size = 0.2, na.rm = F, stat = "identity") +
  scale_fill_gradient2(
  low = "#45efc6",
  mid = "white",  
  high = "#165289",
        n.breaks = 3,
        oob = scales::squish,
        limits = c(-1, 1),
        name = "L2FC") +
  scale_color_gradient2(
  low = "#45efc6",
  mid = "white",  
  high = "#165289",
        n.breaks = 3,
        oob = scales::squish,
        limits = c(-1, 1),      # fixa o intervalo
        name = "L2FC") +
  scale_y_discrete(position = "right") +
    theme_vaxgo() +
    theme(axis.text = element_text(color = "black", size = 8) ,
          # axis.line.x  = element_blank(),
          axis.line.y  = element_blank(),
          axis.text.x = element_text(angle = 90, vjust = 1, hjust=1),
          axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1, size = 0),
          axis.ticks.y = element_blank(), 
          panel.spacing.x = unit(0.1, "cm"),
          strip.background.x = element_rect(fill = "black"),
          strip.text.x = element_text(color = "white", size = 10, 
                                      margin = margin(0.2,1,0.2,1, "cm")),
          panel.grid = element_blank(),
          plot.margin = margin(0, 0, 0, 0, unit = "cm"),
          legend.direction = "horizontal",
          legend.position = "bottom") +
     labs(x = "",
         y = "",
         title = "Best and Worst genes, Overlap") +
  facet_nested(~ paste("Day", timepoint) + condition, scales = "free", space = "free")

df_genes_mouse_fit_best_worst_perc_hallmark_heatmap

df_genes_mouse_fit_best_worst_perc_hallmark_heatmap %>% 
  ggsave(filename = here("Figures", "df_genes_mouse_fit_best_worst_perc_hallmark_heatmap.png"),
         width = 3,
         height = 8)

```

#### ORA: Which processes are better or worse predicted?

```{r fig.width=10, fig.height= 8}
#ORA Hallmarks DIFF
#HALLMARK
#Retrieve human H (hallmark) gene set
msigdbr_df <- msigdbr(species = "human", collection = "H") %>% 
  select(process = gs_name, genes = gene_symbol) %>% 
  mutate(process = str_remove(process, "HALLMARK_"))

ora_mouse_fit_best_worst_perc_hallmark <- enricher_condition(
  df = overlap_best_worst %>% 
    mutate(genes = symbol),
  term2gene_df = msigdbr_df) %>% 
  mutate(log_q = -log(qvalue)) %>% 
  mutate(description = str_replace_all(description, "_", " "),
         timepoint = str_extract(condition, "Day \\d+"))


ora_mouse_fit_best_worst_perc_hallmark
```

ORA Hallmark

```{r fig.height=6, fig.width=6}
#Plot heatmap for ORA Hallmark ------
row_clustered = ora_mouse_fit_best_worst_perc_hallmark %>%  
    dplyr::select(condition, description, log_q) %>% 
    distinct() %>% 
    pivot_wider(names_from = "condition",
                values_from= "log_q") %>%
    column_to_rownames("description") %>% 
    as.matrix() %>% 
    replace(is.na(.), 0) %>% 
    dist(.) %>% 
    hclust()

df_ora_mouse_fit_best_worst_perc_hallmark_heatmap = ora_mouse_fit_best_worst_perc_hallmark %>% 
  clean_names() %>% 
  filter(description %in% c(row_clustered$labels)) %>% 
  ungroup() %>% 
  select(condition, description, log_q, qvalue, everything()) %>% 
    mutate(signific = case_when(qvalue <= 0.10 ~ "*",
                              qvalue <= 0.25 ~ "-",
                              qvalue <= 0.50 ~ "/",
                              qvalue <= 0.05 ~ "**",
                              TRUE ~ ""),
        model = case_when(str_detect(condition, "Mouse") ~ "Mouse", 
                          str_detect(condition, "FIT") ~ "FIT",
                          T ~ NA),
        best_worst = case_when(str_detect(condition, "Best") ~ "10%", 
                          str_detect(condition, "Worst") ~ "90",
                          T ~ NA)) %>% 
  ggplot(aes(x=fct_rev(model), y = str_to_upper(description), fill = log_q)) +
  geom_tile(color = "black", 
            size = 0.2, na.rm = F, stat = "identity") +
  geom_text(aes(label = signific), size = 4, vjust = 0.5, color = "black") +
  scale_fill_gradient2(
        low = "white",     
        high = "#165289",
        n.breaks = 3,
        oob = scales::squish,
        limits = c(0, 4),      # fixa o intervalo
        name = "-Log(FDR)") +
  scale_y_discrete(position = "right") +
    theme_vaxgo() +
    theme(axis.text = element_text(color = "black", size = 8) ,
          # axis.line.x  = element_blank(),
          axis.line.y  = element_blank(),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1),
          axis.ticks = element_blank(),
          strip.background.x = element_rect(fill = "black"),
          strip.text.x = element_text(color = "white", size = 10, 
                                      margin = margin(0.2,1,0.2,1, "cm")),
          panel.grid = element_blank(), 
          panel.spacing.x = unit(0.1, "cm"),
          plot.margin = margin(0, 0, 0, 1, unit = "cm"),
          legend.direction = "horizontal",
          legend.position = "bottom") +
     labs(x = "",
         y = "",
         title = "ORA, Best and Worst genes, Percentile") +
  facet_nested(~ timepoint + best_worst, scales = "free", space = "free")


df_ora_mouse_fit_best_worst_perc_hallmark_heatmap

df_ora_mouse_fit_best_worst_perc_hallmark_heatmap %>% 
  ggsave(filename = here("Figures", "influenza_fluad_mouse_fit_prediction_ora_hallmark_bet_worst_percentile_heatmap_2.png"),
         width = 6,
         height = 6)
```



#====================
# Align sequences

```{r}


#Import homologous CREs from ENCODE 
homologous_cres = read_tsv(here("Genomic", "hg38-mm10-Homologous.tsv"),
  col_names = c("human_enhancer", "mouse_enhancer")
) 

#Import human CREs
cres_human <- read_tsv(
  here("Genomic", "GRCh38-cCREs.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>% 
   rename_with(~paste0(., "_human"))
              
              
#Import mouse CREs
cres_mouse <- read_tsv(
  here("Genomic", "mm10-cCREs.bed"),
  col_names = c("chrom", "start", "end", "id_dhs", "id_ccre", "type"),
  show_col_types = FALSE
) %>% rename_with(~paste0(., "_mouse"))


#Unite info
cres_annotated = homologous_cres %>% 
  full_join(cres_human, by = join_by(human_enhancer == id_ccre_human)) %>% 
  full_join(cres_mouse, by = join_by(mouse_enhancer == id_ccre_mouse)) 
  
```

```{r}

# Instalar ChIPseeker e TxDb se necessário
if (!require("ChIPseeker", quietly = TRUE)) {
  BiocManager::install("ChIPseeker")
}

library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)

# 1. Converter para GRanges
cres_human_gr <- GRanges(
  seqnames = cres_human$chrom_human,
  ranges = IRanges(start = cres_human$start_human, end = cres_human$end_human)
)

# 2. Propagar seqlevels e seqinfo
seqlevels(cres_human_gr) <- seqlevels(BSgenome.Hsapiens.UCSC.hg38)
seqinfo(cres_human_gr) <- seqinfo(BSgenome.Hsapiens.UCSC.hg38)

# 3. Anotar com genes mais próximos
cres_anno <- annotatePeak(
  cres_human_gr,
  tssRegion = c(-3000, 3000),
  TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene,
  annoDb = "org.Hs.eg.db"
)

# 4. Extrair sequências
sequences_human <- getSeq(BSgenome.Hsapiens.UCSC.hg38, cres_human_gr)

# 5. Converter anotações para dataframe e adicionar ao dados originais
cres_anno_df <- as.data.frame(cres_anno)

cres_human_seq <- cres_human %>%
  mutate(
    sequence_human = as.character(sequences_human),
    gene_name = cres_anno_df$SYMBOL,
    gene_id = cres_anno_df$GENEID,
    annotation = cres_anno_df$annotation,
    distance_to_tss = cres_anno_df$distanceToTSS
  )

# Ver resultado
cres_human_seq |> 
  select(id_ccre_human, chrom_human, start_human, end_human, 
         gene_name, gene_id, distance_to_tss, annotation) |>
  head(10)





```









