---
title: "Exploring other datasets"
author: "Wasim Aluísio Prates-Syed"
date: "2025-07-23"
output: 
  html_document: 
    toc: TRUE
    toc_depth: 4
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE, cache = TRUE)
```


# Exploring other datasets: A general analysis

## Import required codes
```{r}
# install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```

#Hepatitis B
## EngerixB
### GSE19668 (Mouse)
```{r}
#Import
GSE120661_counts_not_normalized_log2 = readRDS(here("tables", "GSE120661_counts_not_normalized_log2.rds")) %>% 
  as.data.frame()
GSE120661_metadata = read_csv(file = here("tables", "GSE120661_metadata.csv"))


#Filter data
mouse_engerixb_metadata = GSE120661_metadata %>% 
  filter(treatment == "Engerix B",
         tissue == "PBMC") %>% 
  mutate(timepoint = (timepoint / 24) %>% round(1),
         participant_id = paste0("ENG_BL_", animal_replicate)) %>% 
  filter(timepoint %in% c(0, 1, 3, 7))

#Get log2 values
mouse_engerixb_log2 = GSE120661_counts_not_normalized_log2 %>% 
  select(mouse_engerixb_metadata %>% pull(geo_sample))

```


### Annotate genes
```{r}
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- mouse_engerixb_log2 %>% rownames()

#Get other annotations
listAttributes(ensembl_mart)

#Annotate
annot <- getBM(
  attributes = c("agilent_sureprint_g3_ge_8x60k", #Agilent microarray data (design ID 028005, retrieved from GEO)
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "agilent_sureprint_g3_ge_8x60k",
  values = gene_id,
  mart = ensembl_mart)

#Get lookup table from FIT
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_engerixb_log2summary = mouse_engerixb_log2 %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = agilent_sureprint_g3_ge_8x60k), 
            by = "probe") %>%  #24.009 rows
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1) #14.396


#Modify expression matrix
mouse_engerixb_log2_filtered = mouse_engerixb_log2 %>% 
  rownames_to_column("probe") %>% 
  inner_join(mouse_engerixb_log2summary, by = "probe") %>%   #214,396
  select(probe, mouse_entrez, human_symbol = hgnc_symbol, c(mouse_engerixb_metadata %>% pull(geo_sample)))
#14395
mouse_engerixb_log2_filtered %>% dim()

#Save 
mouse_engerixb_log2_filtered %>% 
  saveRDS(here("tables", "GSE120661_mouse_engerixb_log2.rds"))
```


####Import
```{r}
#Import
mouse_engerixb_log2_filtered = readRDS(here("tables", "GSE120661_mouse_engerixb_log2.rds"))

#Annotate
mouse_engerixb_log2fc_filtered_sample_long = mouse_engerixb_log2_filtered %>%
  pivot_longer(cols = -(probe:human_symbol),
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(mouse_engerixb_metadata %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>%
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")


#Pivot wider
mouse_engerixb_log2fc_filtered_sample_wide = mouse_engerixb_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")

beep()
```




#### Correlation plots, pairwise
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(mouse_engerixb_log2fc_filtered_sample_wide %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, HepB/EngerixB, GSE120661, C57BL/6J", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) 

mouse_correlation_ggpairs_alldays

# Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_alldays_HepB_GSE120661.png"),
         height = 10,
         width = 10)
beep()
```




#### PCA

```{r fig.height=4, fig.width=5}
#Input
samples_metadata_pca_input = mouse_engerixb_metadata
samples_log2fc_values_pca_input = mouse_engerixb_log2fc_filtered_sample_long


#Import metadata
annotation_samples_pca = samples_metadata_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = samples_log2fc_values_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, human_symbol, log2fc) %>% 
  filter(!human_symbol == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("human_symbol") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Mouse, HepB, EngerixB, GSE120661",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

beep()

```

#### BTMs
```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)

#Annotate genes by BTM
samples_genes_btm = samples_log2fc_values_pca_input %>% 
  ungroup() %>% 
  select(symbol = human_symbol, participant_id, timepoint, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = samples_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process

dge_btm_mean_process_mouse = dge_btm_mean_process
```


```{r}
#Define colors
btm_annotation %>% pull(group) %>% unique()

immune_colors = c("SIGNAL TRANSDUCTION" = "#708d81",
                  "CELL CYCLE" = "#06d6a0",
                  "ECM AND MIGRATION" = "#52b788",
                  "ENERGY METABOLISM" = "#95d5b2", 
                  "INNATE RESPONSE" = "#184e77",
                  "NEUTROPHILS" = "#219ebc",
                  "NK CELLS" = "#a2d2ff",
                  "IFN"= "#4361ee",
                  "MONOCYTES" = "#4cc9f0",
                  "PLATELETS" = "#7209b7",
                  "B CELLS" = "#e5383b",
                  "T CELLS" = "#f72585")
immune_order = c("SIGNAL TRANSDUCTION",
                  "CELL CYCLE",
                  "ECM AND MIGRATION",
                  "ENERGY METABOLISM", 
                  "INNATE RESPONSE",
                  "NEUTROPHILS",
                  "NK CELLS",
                  "IFN",
                  "MONOCYTES",
                  "PLATELETS",
                  "B CELLS",
                  "T CELLS")

```

##### GGPairs
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

mouse_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple"))

mouse_correlation_ggpairs_btms_gsea_alldays

mouse_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_btms_gsea_alldays_EngerixB_Mouse.png"),
         height = 10,
         width = 10)

```




##### Ggplot manual
```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(3:7) %>% 
  colnames()

#Inputs
timepoint_selected = 7


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == timepoint_selected) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == timepoint_selected) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-1, 1) +
  xlim(-1, 1) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste0("Mouse, Hepatitis B, EngerixB Mouse, Day", timepoint_selected))
```













### GSE124533 (Human)
Download directly from GEO. The data is too large.

#### Prepare data
```{r}
#EngerixB: GSE124533 -----
filename = "GSE124533"
GSE124533 <- getGEO(filename = here("data_studies", "GSE124533_series_matrix.txt"), GSEMatrix = F)

#Metadata ------
GSE124533_metadata = GSE124533@phenoData %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    participant_id = "participant:ch1",
    sex = "gender:ch1",
    treatment = "treatment:ch1",
    ethnicity = "ethnicity:ch1",
    race = "race:ch1",
    age = "age:ch1",
    timepoint = "day:ch1") %>% 
  mutate(condition = treatment,
         organism = "Human",
         time_unit = "Day")
#Save
GSE124533_metadata %>% saveRDS(file = here("tables", "GSE124533_metadata.rds"))

#Import
GSE124533_metadata = readRDS(file = here("tables", "GSE124533_metadata.rds"))


#Filter metadata
GSE124533_metadata_filtered = GSE124533_metadata %>%
  filter(treatment %in% c("ENGERIXB1"),
         timepoint %in% c("0", "1" , "3", "7"))

#Counts matrix
GSE124533_exprs = GSE124533@assayData[["exprs"]] %>% as.data.frame()
GSE124533_exprs_engerixb = GSE124533_exprs %>% 
  select(c(GSE124533_metadata_filtered$geo_sample))
#Save
GSE124533_exprs_engerixb %>% saveRDS(file = here("tables", "GSE124533_exprs_engerixb.rds"))
```


```{r}
#Import
GSE124533_exprs_engerixb = readRDS(file = here("tables", "GSE124533_exprs_engerixb.rds"))

########### Filter and annotate probes -----
probes_annotation <- read.delim(here("tables", "GPL21272-58809.txt"), skip = 19, header = TRUE, sep = "\t", quote = "") %>% 
  select(gene_id = ID, entrez_id = ENTREZ_GENE_ID, ensembl_id = ENSEMBL_ID, gene_symbol = GENE_SYMBOL) %>% 
  mutate(gene_id = as.character(gene_id))

#Filter genes with highest median expression
GSE124533_counts_log2_summary_annotated_probes = GSE124533_exprs_engerixb %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = -probe_id,
               names_to = "sample_id",
               values_to = "value_raw") %>% 
  group_by(probe_id) %>% 
  summarize(mean = mean(value_raw, na.rm = TRUE),
            median = median(value_raw, na.rm = T)) %>%  
  inner_join(probes_annotation %>% 
              rename(probe_id = gene_id,
                     entrezgene_id = entrez_id,
                     human_symbol = gene_symbol), 
            by = "probe_id") %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>% 
  select(probe_id, entrezgene_id, human_symbol) %>% 
  distinct() %>%  ##41.398 -> 19.394 probes left
  inner_join(GSE124533_exprs_engerixb %>% 
               rownames_to_column("probe_id"),
             by = "probe_id")
#Save
GSE124533_counts_log2_summary_annotated_probes %>% 
  saveRDS(file = here("tables", "GSE124533_counts_log2_summary_annotated_probes.rds"))

beep()
```


```{r}
#Get value from time 0 and calculate Log2FC
human_engerixb_log2fc_filtered_sample_long = GSE124533_counts_log2_summary_annotated_probes %>%
  ungroup() %>% 
  pivot_longer(cols = -(probe_id:human_symbol),
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(GSE124533_metadata_filtered %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) 

#Get over duplicated symbols
overduplicated = human_engerixb_log2fc_filtered_sample_long %>% 
  select(human_symbol) %>% 
  group_by(human_symbol) %>% 
  dplyr::count() %>% 
  arrange(-n) 

#Calculate log2fc
human_engerixb_log2fc_filtered_sample_long = human_engerixb_log2fc_filtered_sample_long %>%
  filter(!human_symbol %in% c(overduplicated %>% 
                                ungroup() %>% 
                                filter(n > median(n)) %>% 
                                pull(human_symbol))) %>% 
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>% 
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")


#Annotate sample and pivot longer
human_engerixb_log2fc_filtered_sample_long_annotated = human_engerixb_log2fc_filtered_sample_long %>% 
  select(-value, -day0) %>% 
  inner_join(GSE124533_metadata_filtered %>% 
               select(sample_id = geo_sample,
                      participant_id, timepoint), by = join_by("participant_id", "timepoint"))

beep()

#Pivot wider
human_engerixb_log2fc_filtered_sample_wide = human_engerixb_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")
  
```


#### Import
```{r}
#Import
human_engerixb_log2fc_filtered_sample_long = 
  readRDS(file = here("tables", "GSE124533_counts_log2_summary_annotated_probes.rds"))
```

#### Correlation plots, pairwise
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Sample 5 participants
set.seed(3141)
sample_participants_5 = GSE124533_metadata_filtered %>% 
  select(participant_id) %>% 
  distinct() %>% 
  slice_sample(n = 5) %>% 
  pull()

#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(human_engerixb_log2fc_filtered_sample_wide %>% 
                                              select(human_symbol, timepoint, c(sample_participants_5)) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, HepB/EngerixB, GSE124533, Human", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) 

mouse_correlation_ggpairs_alldays
# Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "human_correlation_ggpairs_alldays_HepB_GSE124533.png"),
         height = 10,
         width = 10)
beep()
```




#### PCA

```{r fig.height=4, fig.width=5}
#Input
samples_metadata_pca_input = GSE124533_metadata_filtered
samples_log2fc_values_pca_input = human_engerixb_log2fc_filtered_sample_long_annotated


#Import metadata
annotation_samples_pca = samples_metadata_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = samples_log2fc_values_pca_input %>% 
  ungroup() %>% 
  select(human_symbol, participant_id, timepoint, log2fc) %>%
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, human_symbol, log2fc) %>% 
  filter(!human_symbol == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("human_symbol") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Human, HepB, EngerixB, GSE124533",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

beep()

```

#### BTMs
```{r}

#Input
samples_metadata_pca_input = GSE124533_metadata_filtered
samples_log2fc_values_pca_input = human_engerixb_log2fc_filtered_sample_long_annotated



#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)

#Annotate genes by BTM
samples_genes_btm = samples_log2fc_values_pca_input %>% 
  ungroup() %>% 
  select(symbol = human_symbol, participant_id, timepoint, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = samples_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process_human = dge_btm_mean_process

```


```{r}
#Define colors
btm_annotation %>% pull(group) %>% unique()

immune_colors = c("SIGNAL TRANSDUCTION" = "#708d81",
                  "CELL CYCLE" = "#06d6a0",
                  "ECM AND MIGRATION" = "#52b788",
                  "ENERGY METABOLISM" = "#95d5b2", 
                  "INNATE RESPONSE" = "#184e77",
                  "NEUTROPHILS" = "#219ebc",
                  "NK CELLS" = "#a2d2ff",
                  "IFN"= "#4361ee",
                  "MONOCYTES" = "#4cc9f0",
                  "PLATELETS" = "#7209b7",
                  "B CELLS" = "#e5383b",
                  "T CELLS" = "#f72585")
immune_order = c("SIGNAL TRANSDUCTION",
                  "CELL CYCLE",
                  "ECM AND MIGRATION",
                  "ENERGY METABOLISM", 
                  "INNATE RESPONSE",
                  "NEUTROPHILS",
                  "NK CELLS",
                  "IFN",
                  "MONOCYTES",
                  "PLATELETS",
                  "B CELLS",
                  "T CELLS")

```

##### GGPairs
```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

human_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = sample_participants_5,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days, BTM"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple"))

human_correlation_ggpairs_btms_gsea_alldays

human_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "human_correlation_ggpairs_btms_gsea_alldays_EngerixB_Human.png"),
         height = 10,
         width = 10)

```




##### Ggplot manual
```{r fig.height=8, fig.width=10}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#Input
timepoint_selected = 7

#Plot table
plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == timepoint_selected) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == timepoint_selected) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove identical pairs
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))

#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-1, 1) +
  xlim(-1, 1) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste("Human, Hepatitis B, EngerixB Human, Day", timepoint_selected))
```







### Compare Human vs Mouse

```{r}
#Unite Mouse and Human datasets. Samples.
human_mouse_log2fc_samples = 
#Mouse
mouse_engerixb_log2fc_filtered_sample_long %>%
  ungroup() %>% 
  select(human_symbol, log2fc, participant_id,
         timepoint, value, day0) %>% 
  mutate(organism = "Mouse") %>% 
  
  #Human
  bind_rows(human_engerixb_log2fc_filtered_sample_long %>%
  ungroup() %>% 
  select(human_symbol, log2fc, participant_id,
         timepoint,  value, day0) %>% 
  mutate(organism = "Human"))

#Unite metadata
human_mouse_engerix_metadata = 
  #Human
  GSE124533_metadata_filtered %>% 
  select(geo_sample, participant_id, sex, organism, timepoint) %>% 
  mutate(timepoint = as.numeric(timepoint),
         sex = str_to_title(sex)) %>% 
  #Mouse
  bind_rows(
    mouse_engerixb_metadata %>%
      select(geo_sample, participant_id, sex, timepoint) %>%
      mutate(organism = "Mouse"))



```


```{r}
#Calculate the mean log2fc
human_mouse_log2fc_log2fc_avg = human_mouse_log2fc_samples %>% 
  group_by(organism, human_symbol, timepoint) %>% 
  summarize(
            #Calculate the mean
            mean_log2fc = mean(log2fc),
            
            #Calculate pvalue, t test
            p_value = tryCatch(
      t.test(value, day0, paired = TRUE, )$p.value,
      error = function(e) NA),
    .groups = "drop") %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>% 
  filter(human_symbol != "") %>% 
  ungroup()

#Pivot wider
human_mouse_log2fc_log2fc_avg_wide = human_mouse_log2fc_log2fc_avg %>% 
  select(organism, human_symbol, timepoint, mean_log2fc) %>% 
  pivot_wider(names_from = "organism",
              values_from = "mean_log2fc") %>% 
  #Add padj columns
  inner_join(human_mouse_log2fc_log2fc_avg %>%
               select(organism, human_symbol, timepoint, p_adj) %>%
               pivot_wider(names_from = "organism",
                           values_from = "p_adj") %>%
               rename(padj_human = Human, padj_mouse = Mouse)) %>% 
  #Drop genes that are absent in one of the organisms 
  drop_na()

```


```{r fig.height=5, fig.width=5}
# Calculate the correlation
cor_data <- human_mouse_log2fc_log2fc_avg_wide %>%
  group_by(timepoint) %>% 
  filter(padj_human <= 0.05 | padj_human <= 0.05) %>%
  summarize(
    cor_value = cor(Mouse, Human, method = "pearson", use = "complete.obs"),
    label_pos_x = max(Mouse, na.rm = TRUE),
    label_pos_y = max(Human, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("r = ", round(cor_value, 2)))
```


```{r fig.height=6, fig.width=3}
ylim_range = c(-2.5, 2.5)
xlim_range = c(-2.5, 2.5)
human_mouse_log2fc_log2fc_avg_wide_scatterplot = human_mouse_log2fc_log2fc_avg_wide %>% 
  
  #Filter DEGs only
  filter(padj_human <= 0.05 | padj_mouse <= 0.05) %>%
  
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human) +
  
    #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #Facet
  facet_wrap(~paste("Day", timepoint), ncol = 1) +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_log2fc_log2fc_avg_wide %>%
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_log2fc_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  stat_smooth(method = "lm", geom = "line", color = "#4DBBD5FF", alpha = 0.5, se = F, size = 1,
              data = human_mouse_log2fc_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add correlation metrics (Pearson)
  geom_label(
    data = cor_data,
    aes(x = label_pos_x, y = label_pos_y, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 1,
    vjust = 1,
    size = 3
  ) +
  
  #Adjust the axes range
  ylim(ylim_range) +
  xlim(xlim_range) +

  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "none",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title= "Correlation plot, EngerixB,\nHuman vs Mouse")

human_mouse_log2fc_log2fc_avg_wide_scatterplot

```


#### Plot correlation between genes (log2fc for every sample)

```{r}
#Calculate correlation ------
correlation_long_all_mouse_human_timepoints <- human_mouse_log2fc_samples %>%
  select(timepoint, human_symbol, participant_id, log2fc) %>%
  group_by(timepoint) %>%
  
  # Pivot wider and calculate correlation
  group_modify(~ {
    .x %>%
      pivot_wider(names_from = participant_id, values_from = log2fc) %>%
      select(-human_symbol) %>%
      cor(use = "pairwise.complete.obs", method = "pearson") %>%
      as.data.frame() %>%
      rownames_to_column("participant1") %>%
      pivot_longer(-participant1, names_to = "participant2", values_to = "correlation") %>%
      filter(participant1 != participant2)
  }) %>%
  ungroup() %>% 
  
  #Filter by organism
  inner_join(human_mouse_engerix_metadata %>% 
               select(participant_id, participant_1_organism = organism, sex_p_1 = sex) %>% 
               distinct(),
             by = join_by("participant1" == "participant_id")) %>% 
  inner_join(human_mouse_engerix_metadata %>% 
               select(participant_id, participant_2_organism = organism, sex_p_2 = sex) %>% 
               distinct(),
             by = join_by("participant2" == "participant_id")) 


# Remove duplicated lines only if the organism is the same
correlation_long_all_timepoints_same_organisms = correlation_long_all_mouse_human_timepoints %>%
  filter(participant_1_organism == participant_2_organism) %>%
  mutate(
    participant_min = pmin(participant1, participant2),
    participant_max = pmax(participant1, participant2)
  ) %>%
  # Remove duplicates (ex: A vs B e B vs A)
  distinct(timepoint, participant_min, participant_max, .keep_all = TRUE) %>%
  select(timepoint, participant1, participant2, correlation, participant_1_organism, participant_2_organism, sex_p_1, sex_p_2) %>%
  group_by(timepoint) %>%
  arrange(-correlation)


#Unite table with different organisms with table with deduplicated rows with the same organism
correlation_long_all_mouse_human_timepoints = correlation_long_all_mouse_human_timepoints %>% 
  filter(participant_1_organism != participant_2_organism) %>% 
  bind_rows(correlation_long_all_timepoints_same_organisms)

#Summarize data and save table
correlation_long_all_mouse_human_timepoints_summary = correlation_long_all_mouse_human_timepoints %>% 
  
  #Filter only female and set Mouse as reference
  filter(sex_p_1 == "Female",
         sex_p_2 == "Female") %>%
  mutate(sex_comparison = paste(sex_p_1, "vs.", sex_p_2),
         comparison = paste(participant_1_organism, "vs.", participant_2_organism)) %>%
  group_by(timepoint, comparison) %>% 
  arrange(timepoint, -correlation) %>% 
  mutate(correlation = round(correlation, 2),
         timepoint = paste("Day", timepoint)) %>% 
  summarize(mean = mean(correlation, na.rm = T) %>% round(2),
            median = median(correlation, na.rm = T)%>% round(2),
            min = min(correlation, na.rm = T)%>% round(2),
            max = max(correlation, na.rm = T) %>% round(2),
            sd = sd(correlation),
            n = sum(!is.na(correlation)),
            se = sd/sqrt(n),
            t_critical = qt(0.975, df = n - 1),
            ci_lower = mean - t_critical * se,
            ci_upper = mean + t_critical * se,
            .groups = "drop") %>% 
  filter(comparison != "Human vs. Mouse")

correlation_long_all_mouse_human_timepoints_summary

#Create GT table
correlation_long_all_mouse_human_timepoints_gt = correlation_long_all_mouse_human_timepoints_summary %>% 
  gt::gt() 
#Save gt
correlation_long_all_mouse_human_timepoints_gt
gtsave(correlation_long_all_mouse_human_timepoints_gt, filename = here("tables", "correlation_long_all_mouse_human_timepoints_EngerixB_gt.docx"))

```


```{r fig.height=5, fig.width=3}
correlation_long_all_mouse_human_timepoints_summary %>% 
  separate(timepoint, into = c("unit", "timepoint"), remove = T) %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1) +
  
  geom_errorbar(aes(ymin = ci_lower, 
                  ymax = ci_upper), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 1, 3, 7), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_color_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = "Correlation between same and \ndifferent organisms, by subject.\nHepatitis B, GSE124533, GSE19668",
       x = "Day",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 
  
```




#### BTMs


```{r}

```
##### Scatterplot



```{r fig.height=5, fig.width=5}
#Annotate genes by BTM
 
dge_btm_mean_process_human_long = dge_btm_mean_process_human %>% 
  pivot_longer(cols = -c(timepoint, process, group, subgroup), 
               names_to = "participant_id", 
               values_to = "log2fc") %>% 
  mutate(organism = "Human")

dge_btm_mean_process_mouse_long = dge_btm_mean_process_mouse %>% 
  pivot_longer(cols = -c(timepoint, process, group, subgroup), 
               names_to = "participant_id", 
               values_to = "log2fc") %>% 
  mutate(organism = "Mouse")



dge_btm_mean_process = dge_btm_mean_process_human_long %>%
  bind_rows(dge_btm_mean_process_mouse_long) %>%
  filter(timepoint %in% c(1, 3, 7)) %>% 
  group_by(organism, timepoint, process) %>% 
  summarize(mean_log2fc = median(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism",
              values_from = "mean_log2fc")

dge_btm_mean_process

# Calculate the correlation
cor_data <- dge_btm_mean_process %>%
  # Group the data by timepoint
  group_by(timepoint) %>% 
  
  # For each timepoint, run Pearson correlation test between Mouse and Human
  summarize(
    cor_test = list(cor.test(Mouse, Human, method = "pearson")), # store full test result as a list
    label_pos_x = max(Mouse, na.rm = TRUE),  # position for label in the plot (x = max Mouse value)
    label_pos_y = max(Human, na.rm = TRUE),  # position for label in the plot (y = max Human value)
    .groups = "drop" # remove grouping after summarizing
  ) %>%
  
  # Extract correlation coefficient (r) and p-value from cor_test, and create a label
  mutate(
    cor_value = map_dbl(cor_test, ~ .x$estimate),  # correlation coefficient
    p_value   = map_dbl(cor_test, ~ .x$p.value),   # p-value
    label = paste0(
      "r = ", round(cor_value, 2),                 # round r to 2 decimal places
      ", p = ", signif(p_value, 3)                 # show p with 3 significant digits
    )
  )
```


```{r fig.height=8, fig.width=5}

human_mouse_log2fc_BTMs_avg_wide_scatterplot = dge_btm_mean_process %>% 
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human,
      color = group) +
  
  #Add points
  geom_point() +
  
#Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Facet
  facet_wrap(~paste("Day", timepoint), 
             ncol = 1, 
             scales = "free", axes = "all") +
  
  #Color
  scale_color_manual(values = immune_colors) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  
  #Add correlation metrics (Pearson)
  geom_label(data = cor_data,
             aes(x = label_pos_x, y = label_pos_y, label = label),
             inherit.aes = FALSE,
             fill = "white",
             hjust = 1,
             vjust = 1,
             size = 3)+

  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "right",
        legend.direction = "vertical") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       color = "BTM group",
       title= "Correlation plot, EngerixB,\nHuman vs Mouse")


human_mouse_log2fc_BTMs_avg_wide_scatterplot
```




##### Heatmap

```{r}
#Matrix data
matrix_data = dge_btm_mean_process %>% 
  select(process, Human_Day1, Mouse_Day2) %>% 
  drop_na() %>% 
  column_to_rownames("process") %>% 
  as.matrix()

# Column and row annotation
######## Annotations
annotation_heatmap = tribble(~condition, ~timepoint, ~organism,
        "Human_Day1", 1, "Human",
        "Mouse_Day2", 2, "Mouse")

#Arrange columns in the same order as the matrix
ann_cols_heatmap = matrix_data %>% 
  colnames() %>% 
  data.frame() %>% 
  rename(condition = '.') %>% 
  inner_join(annotation_heatmap, by = "condition") %>% 
  distinct() %>% 
  arrange(condition)

#Arrange matrix columns 
mat = matrix_data %>% 
  as.data.frame() %>% 
  select(c(ann_cols_heatmap$condition %>% unique())) %>% 
  as.matrix() 

#Column to rownames
ann_cols_heatmap = ann_cols_heatmap %>% 
  column_to_rownames("condition")

#Annotate rows
ann_rows_heatmap = mat %>% 
  rownames() %>% 
  as.data.frame() %>% 
  rename("process" = ".") %>% 
  inner_join(dge_btm_mean_process %>% 
               select(process, group, subgroup), 
             by = "process") %>% 
  column_to_rownames("process")
  

#Check dimensions
dim(mat)
dim(ann_cols_heatmap)

#Colors from required.R
btm_immune_groups
colors_immune

```


```{r fig.width= 6, fig.height=20}
#Heatmap annotation
ha = HeatmapAnnotation(df = ann_cols_heatmap, 
                       col = colors_immune,
                       gp = gpar(col = "black"), 
                       simple_anno_size = unit(0.3, "cm"),
                       annotation_name_side = "right")

row_ha = HeatmapAnnotation(df = ann_rows_heatmap,
                       col = btm_immune_groups,
                       which= "row",
                       simple_anno_size = unit(0.2, "cm"),
                       show_annotation_name = T,
                       show_legend = T)

# Values colors
col_fun <- colorRamp2(c(-2,0,2), c("#45efc6", "white", "#165289"))

file = "BTM_Modules_Saureus_Mouse_Human"

width_image = 25
height_image = 30
width = unit(1, "cm")
height = unit(25, "cm")
column_names_gp = gpar(fontsize = 8)
row_names_gp = gpar(fontsize = 8)
rect_gp = gpar(col = "black", lwd = 0.5)

fileimageheatmap_grouped = here("Figures", paste0(file, sep ="_", "Complexheatmap.png"))

png(file = fileimageheatmap_grouped, 
    width = width_image, 
    height = height_image, 
    units="cm", 
    res=900)

heatmap_plot = Heatmap(mat,
                        top_annotation = ha, 
                       left_annotation = row_ha,
                        show_row_names = TRUE,
                        row_names_side = "right",
                        show_heatmap_legend = TRUE,
                        heatmap_legend_param = list(direction = "horizontal"),
                        col = col_fun,
                        show_column_names = TRUE,
                        column_title = file,
                        column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                        column_names_rot = 45,
                        column_names_gp = column_names_gp,
                        rect_gp = rect_gp,
                        na_col = "black",
                        name = "Mean Log2FC",
                        cluster_rows = TRUE,
                        row_names_gp = row_names_gp, row_km = 2,
                        row_dend_width = unit(3, "cm"),
                        row_split = NULL,
                        column_split = NULL,
                        row_gap = unit(2, "mm"),
                        cluster_columns = TRUE,
                        column_gap = unit(8, "mm"),
                        show_column_dend = F,
                        show_row_dend = T, 
                        row_dend_side = "left",
                        column_dend_side = "top",
                        clustering_distance_rows = "euclidean",
                        border = TRUE,
                        width = width,
                        height = height)

heatmap_plot = draw(heatmap_plot, 
     annotation_legend_side = "left", 
     heatmap_legend_side = "left")

heatmap_plot

dev.off()


```











# Compare with FIT prediction

## Run FIT on a training dataset
```{r}
#Get log2fc mean
GSE19668_input2FIT = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
    filter(timepoint == 2) %>% 
  select(participant_id, MM.Entrez = entrezgene_id, EffectSize = log2fc) %>% 
  group_by(MM.Entrez) %>% 
  summarize(mean_log2fc = mean(EffectSize))

#Save
GSE19668_input2FIT %>% 
  write_csv(file = here("FIT_Mouse2man", "GSE19668_day2_FIT_mouse.csv"))

#Run FIT
GSE19668_input2FIT_day2 = FIT.mouse2man::FIT(MouseFile = here("FIT_Mouse2man", "GSE19668_day2_FIT_mouse.csv"), DataType = "rnaseq")
```



### Unite log2FC values from mouse, human, and FIT
```{r fig.height=5, fig.width=8}
#Bind rows
mouse_human_fit_log2fc = bind_rows(GSE19668_input2FIT_day2 %>%
                                     #Get DEGs for FIT
                                     filter((CI_low != 0 | CI_high != 0) & !is.na(CI_low) & !is.na(CI_high)) %>%  
                                   select(human_symbol = Human.symbol,
                                          log2fc = FIT_prediction) %>%
                                     mutate(organism = "FIT"),
                                   
                                   #Get DEGs for Human
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_human <= 0.05) %>% 
                                     select(human_symbol, log2fc = Human) %>%
                                     mutate(organism = "Human"),
                                   
                                   #Get DEGs for Mouse
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_mouse <= 0.05) %>% 
                                     select(human_symbol, log2fc = Mouse) %>%
                                     mutate(organism = "Mouse"))


#Pivot wider
mouse_human_fit_log2fc_wide = GSE19668_input2FIT_day2 %>%
            select(human_symbol = Human.symbol,
                   FIT_prediction) %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Human),
            by = "human_symbol") %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Mouse),
            by = "human_symbol")

#Save
mouse_human_fit_log2fc %>% 
  saveRDS(file = here("tables", "mouse_human_fit_log2fc_long_SAureus.rds"))

mouse_human_fit_log2fc_wide %>% 
  saveRDS(file = here("tables", "mouse_human_fit_log2fc_wide_SAureus.rds"))
```

## Distribution

### Import tables
```{r fig.height=5, fig.width=8}
#Import
mouse_human_fit_log2fc = readRDS(file = here("tables", "mouse_human_fit_log2fc_long_SAureus.rds"))
mouse_human_fit_log2fc_wide = readRDS(file = here("tables", "mouse_human_fit_log2fc_wide_SAureus.rds"))

GSE33341_metadata = readRDS(file = here("tables", "GSE33341_metadata.rds"))
GSE19668_metadata = readRDS(file = here("tables", "GSE19668_metadata.rds"))
```

### Plot histogram

```{r fig.height=5, fig.width=8}
#Plot histogram
mouse_human_fit_log2fc_histogram = mouse_human_fit_log2fc %>% 
  ggplot() +
  aes(x = log2fc,
      fill = organism) +
  geom_density(aes(y = after_stat(count),
                   color = organism),
                  position = position_identity(), 
               alpha = 0.1,
               size = 1,
              show.legend = T) +
  scale_fill_manual(values = c(FIT = "#4361ee",
                               Mouse = "#4DBBD5FF",
                               Human = "#546E7AFF"))+
  scale_color_manual(values = c(FIT = "#4361ee",
                                Mouse = "#4DBBD5FF",
                                Human = "#546E7AFF"))+
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        legend.key = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0),
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(limits = c(-4, 4)) +
  labs(color = "",
       fill = "")

mouse_human_fit_log2fc_histogram

#Faceted
mouse_human_fit_log2fc_histogram_faceted = mouse_human_fit_log2fc_histogram +
  facet_wrap(~organism,
             scales = "free_y")


#Plot scatter plot
mouse_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = Mouse,
      y = Human) +
#All DEGs
geom_point(color = "#4DBBD5FF",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "Mouse L2FC",
         y = "Human L2FC")

mouse_human_scatterplot


fit_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = FIT_prediction,
      y = Human) +
#All DEGs
geom_point(color = "#4361ee",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "FIT L2FC",
         y = "Human L2FC")

#Histograms
(((mouse_human_fit_log2fc_histogram + mouse_human_fit_log2fc_histogram_faceted) + 
    plot_layout(widths = c(0.5, 1), guides = "collect") +
    plot_annotation(theme = theme(legend.position = "top")))  /
    #Scatter plots
    (mouse_human_scatterplot + fit_human_scatterplot)) +
  plot_layout(heights = c(0.5, 1))

```



## BTMs
Import
```{r}
#New BTM Annotation
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
btm_annotation = read.table(file = here("VaxGO",  "BTM functional groups V3.txt"), sep="\t", header=TRUE) %>% 
  clean_names() %>% 
  rename(id = name) %>% 
  filter(group != "")

btm_annotation_genes = ImmuneGO_BTM_grouped_genes %>% 
               inner_join(btm_annotation %>% 
                            rename(process = id), 
                          by = "process") %>% 
               rename(symbol = genes)
```

```{r}
#Annotate genes by BTM
dge_btm_mean_process = mouse_human_fit_log2fc %>%
  filter(organism != "Mouse") %>% 
  rename(symbol = human_symbol) %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  group_by(organism, process) %>% 
  summarize(mean_log2fc = mean(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism", values_from = "mean_log2fc") 

dge_btm_mean_process
```
##### Scatterplot
```{r fig.height=5, fig.width=7}

#Calculate correlation
cor_labels <- dge_btm_mean_process %>%
  summarise(cor_value = cor(Human, FIT, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
dge_btm_mean_process %>% 
ggplot(aes(x = FIT, y = Human)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  stat_smooth(aes(x = FIT, y = Human),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "FIT Log2FC", 
       y = "Human Log2FC", 
       color = "",
       title = "Mouse vs. Human")



```

