---
title: "Exploring other datasets"
author: "Wasim Aluísio Prates-Syed"
date: "2025-07-23"
output: 
  html_document: 
    toc: TRUE
    toc_depth: 4
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE, cache = TRUE)
```

# Hepatitis B

## Import required codes

```{r}
# install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```

## EngerixB

### GSE19668 (Mouse)

```{r}
#Import
GSE120661_counts_not_normalized_log2 = readRDS(here("tables", "GSE120661_counts_not_normalized_log2.rds")) %>% 
  as.data.frame()
GSE120661_metadata = read_csv(file = here("tables", "GSE120661_metadata.csv"))


#Filter data
mouse_engerixb_metadata = GSE120661_metadata %>% 
  filter(treatment == "Engerix B",
         tissue == "PBMC") %>% 
  mutate(timepoint = (timepoint / 24) %>% round(1),
         participant_id = paste0("ENG_BL_", animal_replicate)) %>% 
  filter(timepoint %in% c(0, 1, 3, 7))

mouse_engerixb_metadata %>% 
  saveRDS(file = here("tables", "mouse_engerixb_metadata.rds"))

#Get log2 values
mouse_engerixb_log2 = GSE120661_counts_not_normalized_log2 %>% 
  select(mouse_engerixb_metadata %>% pull(geo_sample))

```

### Annotate genes

Connect to the Ensembl BioMart database to annotate mouse microarray
probes. To do this, extract the probe identifiers from the expression
matrix and retrieve their Ensembl IDs, Entrez IDs, MGI symbols, and gene
descriptions.

Load the FIT reference table linking mouse genes to their human
orthologs, and reshape the expression data to calculate mean and median
values for each probe. For every gene, keep only the probe with the
highest median expression to remove redundancy, and repeat the same step
at the human ortholog level so that each gene is represented once.

The final product is a reduced expression matrix of about 14,000 human
ortholog genes derived from the mouse dataset.

```{r}
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- mouse_engerixb_log2 %>% rownames()

#Get other annotations
listAttributes(ensembl_mart)

#Annotate
annot <- getBM(
  attributes = c("agilent_sureprint_g3_ge_8x60k", #Agilent microarray data (design ID 028005, retrieved from GEO)
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "agilent_sureprint_g3_ge_8x60k",
  values = gene_id,
  mart = ensembl_mart)

beep()

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_engerixb_log2summary = mouse_engerixb_log2 %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = agilent_sureprint_g3_ge_8x60k), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = mouse_engerixb_log2summary %>% 
  #filter ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

mouse_engerixb_log2summary = mouse_engerixb_log2summary %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
mouse_engerixb_log2_filtered_probes = mouse_engerixb_log2 %>% 
  rownames_to_column("probe") %>% 
  inner_join(mouse_engerixb_log2summary, by = "probe") %>%   #214,396
  select(probe, mouse_entrez, human_symbol = hgnc_symbol, c(mouse_engerixb_metadata %>% pull(geo_sample)))
#14395
mouse_engerixb_log2_filtered %>% dim()



mouse_engerixb_log2_hgnc_symbol = mouse_engerixb_log2_filtered_probes %>% 
  select(human_symbol, c(mouse_engerixb_metadata$geo_sample)) %>% 
  column_to_rownames("human_symbol")


#Save 
mouse_engerixb_log2_hgnc_symbol %>% 
  saveRDS(here("tables", "GSE120661_mouse_engerixb_log2.rds"))
```

#### Import

```{r}
#Import
mouse_engerixb_log2_hgnc_symbol = readRDS(here("tables", "GSE120661_mouse_engerixb_log2.rds"))
mouse_engerixb_metadata = readRDS(file = here("tables", "mouse_engerixb_metadata.rds"))

#Create Expression Set
eset_mouse_engerixb <- ExpressionSet(
  assayData = mouse_engerixb_log2_hgnc_symbol %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(mouse_engerixb_metadata %>% 
                                   column_to_rownames("geo_sample")))

#Save
eset_mouse_engerixb %>% saveRDS(file = here("tables", "eset_mouse_engerixb.rds"))

#Annotate
mouse_engerixb_log2fc_filtered_sample_long = mouse_engerixb_log2_hgnc_symbol %>%
  rownames_to_column("human_symbol") %>% 
  pivot_longer(cols = -human_symbol,
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(mouse_engerixb_metadata %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>%
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")

#Save
mouse_engerixb_log2fc_filtered_sample_long %>% 
  saveRDS(here("tables", "mouse_engerixb_log2fc_filtered_sample_long.rds"))


#Pivot wider
mouse_engerixb_log2fc_filtered_sample_wide = mouse_engerixb_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")

beep()
```


#### Differential expression (Limma)
Paired samples design
```{r}

#Import
mouse_engerixb_log2_hgnc_symbol = readRDS(here("tables", "GSE120661_mouse_engerixb_log2.rds"))
mouse_engerixb_metadata = readRDS(file = here("tables", "mouse_engerixb_metadata.rds"))
eset_mouse_engerixb = readRDS(file = here("tables", "eset_mouse_engerixb.rds"))

#Create model matrix
design <- model.matrix(~ timepoint + 0, # 0 defines the way we compare the samples
                       eset_mouse) 
colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

# Estimate the correlation between measurements made on the same subject (paired-samples)
corfit <- duplicateCorrelation(eset_mouse, 
                               design = design, 
                               block = eset_mouse$participant_id)
corfit$consensus #-0.02688727

#Fit Limma model
fit <- lmFit(eset_mouse, 
             design, 
             block = eset_mouse$participant_id, 
             correlation = corfit$consensus) #Use inter-subject correlation as input into the linear model fit
fit <- eBayes(fit)
#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(conditionEngerix.B_24_PBMC	 - conditionEngerix.B_0_PBMC,
                             conditionEngerix.B_72_PBMC	 - conditionEngerix.B_0_PBMC,
                             conditionEngerix.B_168_PBMC	 - conditionEngerix.B_0_PBMC,
                             levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- treat(fit2, lfc = 0.0)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_mouse_engerixb = all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  mutate(condition = case_when(comparisonA == "conditionEngerix.B_24_PBMC"	~ "Engerix_24_PBMC",
                               comparisonA == "conditionEngerix.B_72_PBMC"	~ "Engerix_72_PBMC",
                               comparisonA == "conditionEngerix.B_168_PBMC"	~ "Engerix_168_PBMC",
                               T~comparisonA)) %>% 
  select(condition, everything())
	
# #Save
all_degs_mouse_engerixb %>% saveRDS(file = here("tables", "EngerixB_GSE120661_degs_mouse.rds"))
```



#### Correlation plots, pairwise

Plot correlations between mouse samples across all timepoints using
pairwise scatterplots and density plots from the GGally package.

To do this, define custom plotting functions for densities on the
diagonal and points on the lower panels, then apply ggpairs to the
filtered log2 fold change expression matrix. Reorder the timepoints as
factors, color and fill by timepoint, and apply a custom theme along
with manual color scales.

The result is a matrix of correlation plots titled “All days,
HepB/EngerixB, GSE120661, C57BL/6J”.

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(mouse_engerixb_log2fc_filtered_sample_wide %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, HepB/EngerixB, GSE120661, C57BL/6J", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) 

mouse_correlation_ggpairs_alldays

# Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_alldays_HepB_GSE120661.png"),
         height = 10,
         width = 10)
beep()
```

#### PCA

Perform a PCA on the **mouse** EngerixB dataset using log2 fold change
values from each **sample**.

Merge metadata with expression values, reshape the data into a wide
matrix with genes as rows and samples as columns, and replace missing or
non-finite values with zeros.

Run PCA on the transposed matrix to capture sample-level variation, then
extract the first two principal components and merge them back with
sample annotations.

Finally, plot PC1 versus PC2, coloring and filling by timepoint and
labeling by participant, with custom color scales and a themed layout.

```{r fig.height=4, fig.width=5}

#Import
mouse_engerixb_metadata = readRDS(file = here("tables", "mouse_engerixb_metadata.rds"))
mouse_engerixb_log2fc_filtered_sample_long = readRDS(here("tables", "mouse_engerixb_log2fc_filtered_sample_long.rds"))

#Input
samples_metadata_pca_input = mouse_engerixb_metadata
samples_log2fc_values_pca_input = mouse_engerixb_log2fc_filtered_sample_long


#Import metadata
annotation_samples_pca = samples_metadata_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = samples_log2fc_values_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, human_symbol, log2fc) %>% 
  filter(!human_symbol == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("human_symbol") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Mouse, HepB, EngerixB, GSE120661",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

beep()

```







#### BTMs

Annotate mouse genes with Blood Transcription Module (BTM) functional
groups and use them to assess correlations across timepoints. Load
pre-curated BTM annotations and clean the table of functional groups,
then merge it with an external list of immune-related genes. Map mouse
log2 fold change values to BTM processes by joining gene symbols with
their corresponding modules. For each participant and timepoint, compute
the mean log2 fold change per process, reshape the data into a wide
format, and attach module group information.

```{r}
#Import files
btm_annotation_genes =
  read_csv(file = here("tables", "btm_annotation_genes.csv"))
mouse_engerixb_metadata = readRDS(file = here("tables", "mouse_engerixb_metadata.rds"))
mouse_engerixb_log2fc_filtered_sample_long = readRDS(here("tables", "mouse_engerixb_log2fc_filtered_sample_long.rds"))

#Annotate genes by BTM
samples_genes_btm = mouse_engerixb_log2fc_filtered_sample_long %>% 
  ungroup() %>% 
  select(symbol = human_symbol, participant_id, timepoint, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = samples_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process

dge_btm_mean_process_mouse = dge_btm_mean_process

dge_btm_mean_process_mouse %>% 
  saveRDS(file = here("tables", "hepatitisb_dge_btm_mean_process_mouse.rds"))
```

```{r}
#Define colors
btm_annotation %>% pull(group) %>% unique()

immune_colors = c("SIGNAL TRANSDUCTION" = "#708d81",
                  "CELL CYCLE" = "#06d6a0",
                  "ECM AND MIGRATION" = "#52b788",
                  "ENERGY METABOLISM" = "#95d5b2", 
                  "INNATE RESPONSE" = "#184e77",
                  "NEUTROPHILS" = "#219ebc",
                  "NK CELLS" = "#a2d2ff",
                  "IFN"= "#4361ee",
                  "MONOCYTES" = "#4cc9f0",
                  "PLATELETS" = "#7209b7",
                  "B CELLS" = "#e5383b",
                  "T CELLS" = "#f72585")
immune_order = c("SIGNAL TRANSDUCTION",
                  "CELL CYCLE",
                  "ECM AND MIGRATION",
                  "ENERGY METABOLISM", 
                  "INNATE RESPONSE",
                  "NEUTROPHILS",
                  "NK CELLS",
                  "IFN",
                  "MONOCYTES",
                  "PLATELETS",
                  "B CELLS",
                  "T CELLS")

```

##### Using GGPairs

Using this summarized BTM-level expression, generate pairwise
correlation plots with ggpairs, showing scatterplots and density plots
across modules, colored by timepoint. Add reference lines at zero, apply
custom color scales, and use a themed layout.

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

mouse_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple"))

mouse_correlation_ggpairs_btms_gsea_alldays

mouse_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "mouse_correlation_ggpairs_btms_gsea_alldays_EngerixB_Mouse.png"),
         height = 10,
         width = 10)

```

##### Using ggplot

Compute pairwise correlations of BTM module expression for five mouse
participants at a selected timepoint. Create unique participant pairs,
calculate Pearson correlations and R², and plot scatterplots with
regression lines, colored by module group, faceted by participant pairs,
and labeled with correlation values.

```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(3:7) %>% 
  colnames()

#Inputs
timepoint_selected = 7


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == timepoint_selected) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == timepoint_selected) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-1, 1) +
  xlim(-1, 1) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste0("Mouse, Hepatitis B, EngerixB Mouse, Day", timepoint_selected))
```

### GSE124533 (Human)

Download directly from GEO. The data is too large.

#### Prepare data

```{r}
#EngerixB: GSE124533 -----
filename = "GSE124533"
GSE124533 <- getGEO(filename = here("data_studies", "GSE124533_series_matrix.txt"), GSEMatrix = F)

#Metadata ------
GSE124533_metadata = GSE124533@phenoData %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    participant_id = "participant:ch1",
    sex = "gender:ch1",
    treatment = "treatment:ch1",
    ethnicity = "ethnicity:ch1",
    race = "race:ch1",
    age = "age:ch1",
    timepoint = "day:ch1") %>% 
  mutate(condition = treatment,
         organism = "Human",
         time_unit = "Day")
#Save
GSE124533_metadata %>% saveRDS(file = here("tables", "GSE124533_metadata.rds"))

#Import
GSE124533_metadata = readRDS(file = here("tables", "GSE124533_metadata.rds"))


#Filter metadata
GSE124533_metadata_filtered = GSE124533_metadata %>%
  filter(treatment %in% c("ENGERIXB1"),
         timepoint %in% c("0", "1" , "3", "7"))

#Save
GSE124533_metadata_filtered %>% 
  saveRDS(file = here("tables", "GSE124533_metadata_filtered.rds"))


#Counts matrix
GSE124533_exprs = GSE124533@assayData[["exprs"]] %>% as.data.frame()
GSE124533_exprs_engerixb = GSE124533_exprs %>% 
  select(c(GSE124533_metadata_filtered$geo_sample))
#Save
GSE124533_exprs_engerixb %>% saveRDS(file = here("tables", "GSE124533_exprs_engerixb.rds"))
```




```{r}
#Import
GSE124533_exprs_engerixb = readRDS(file = here("tables", "GSE124533_exprs_engerixb.rds"))

########### Filter and annotate probes -----
probes_annotation <- read.delim(here("tables", "GPL21272-58809.txt"), skip = 19, header = TRUE, sep = "\t", quote = "") %>% 
  select(gene_id = ID, entrez_id = ENTREZ_GENE_ID, ensembl_id = ENSEMBL_ID, gene_symbol = GENE_SYMBOL) %>% 
  mutate(gene_id = as.character(gene_id))

#Filter genes with highest median expression
GSE124533_counts_log2_summary_annotated_probes = GSE124533_exprs_engerixb %>% 
  rownames_to_column("probe_id") %>% 
  pivot_longer(cols = -probe_id,
               names_to = "sample_id",
               values_to = "value_raw") %>% 
  group_by(probe_id) %>% 
  summarize(mean = mean(value_raw, na.rm = TRUE),
            median = median(value_raw, na.rm = T)) %>%  
  inner_join(probes_annotation %>% 
              rename(probe_id = gene_id,
                     entrezgene_id = entrez_id,
                     human_symbol = gene_symbol), 
            by = "probe_id") %>% 
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>% 
  select(probe_id, entrezgene_id, human_symbol) %>% 
  distinct() %>%  ##41.398 -> 19.394 probes left
  inner_join(GSE124533_exprs_engerixb %>% 
               rownames_to_column("probe_id"),
             by = "probe_id")
#Save
GSE124533_counts_log2_summary_annotated_probes %>% 
  saveRDS(file = here("tables", "GSE124533_counts_log2_summary_annotated_probes.rds"))

beep()
```

```{r}

#Import
GSE124533_counts_log2_summary_annotated_probes =
  readRDS(file = here("tables", "GSE124533_counts_log2_summary_annotated_probes.rds"))
GSE124533_metadata_filtered =
  readRDS(file = here("tables", "GSE124533_metadata_filtered.rds"))

#Get value from time 0 and calculate Log2FC
human_engerixb_log2fc_filtered_sample_long = GSE124533_counts_log2_summary_annotated_probes %>%
  ungroup() %>% 
  pivot_longer(cols = -(probe_id:human_symbol),
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(GSE124533_metadata_filtered %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) 

#Get over duplicated symbols
overduplicated = human_engerixb_log2fc_filtered_sample_long %>% 
  select(human_symbol) %>% 
  group_by(human_symbol) %>% 
  dplyr::count() %>% 
  arrange(-n) 

#Calculate log2fc
human_engerixb_log2fc_filtered_sample_long = human_engerixb_log2fc_filtered_sample_long %>%
  filter(!human_symbol %in% c(overduplicated %>% 
                                ungroup() %>% 
                                filter(n > median(n)) %>% 
                                pull(human_symbol))) %>% 
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>% 
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")


#Annotate sample and pivot longer
human_engerixb_log2fc_filtered_sample_long_annotated = human_engerixb_log2fc_filtered_sample_long %>% 
  select(-value, -day0) %>% 
  inner_join(GSE124533_metadata_filtered %>% 
               select(sample_id = geo_sample,
                      participant_id, timepoint), by = join_by("participant_id", "timepoint"))

#Save
human_engerixb_log2fc_filtered_sample_long_annotated %>% 
  saveRDS(file = here("tables", "human_engerixb_log2fc_filtered_sample_long_annotated.rds"))

beep()

#Pivot wider
human_engerixb_log2fc_filtered_sample_wide = human_engerixb_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")
  
```

#### Import

```{r}
#Import
human_engerixb_log2fc_filtered_sample_long = 
  readRDS(file = here("tables", "GSE124533_counts_log2_summary_annotated_probes.rds"))
```

#### Correlation plots, pairwise

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Sample 5 participants
set.seed(3141)
sample_participants_5 = GSE124533_metadata_filtered %>% 
  select(participant_id) %>% 
  distinct() %>% 
  slice_sample(n = 5) %>% 
  pull()

#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(human_engerixb_log2fc_filtered_sample_wide %>% 
                                              select(human_symbol, timepoint, c(sample_participants_5)) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, HepB/EngerixB, GSE124533, Human", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) 

mouse_correlation_ggpairs_alldays
# Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "human_correlation_ggpairs_alldays_HepB_GSE124533.png"),
         height = 10,
         width = 10)
beep()
```

#### PCA

```{r fig.height=4, fig.width=5}

GSE124533_counts_log2_summary_annotated_probes =
  readRDS(file = here("tables", "GSE124533_counts_log2_summary_annotated_probes.rds"))

#Input
samples_metadata_pca_input = GSE124533_metadata_filtered
samples_log2fc_values_pca_input = human_engerixb_log2fc_filtered_sample_long_annotated


#Import metadata
annotation_samples_pca = samples_metadata_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = samples_log2fc_values_pca_input %>% 
  ungroup() %>% 
  select(human_symbol, participant_id, timepoint, log2fc) %>%
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, human_symbol, log2fc) %>% 
  filter(!human_symbol == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("human_symbol") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Human, HepB, EngerixB, GSE124533",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

beep()

```

#### BTMs

```{r}

#Import files
btm_annotation_genes =
  read_csv(file = here("tables", "btm_annotation_genes.csv"))
human_engerixb_log2fc_filtered_sample_long_annotated = 
  readRDS(file = here("tables", "human_engerixb_log2fc_filtered_sample_long_annotated.rds"))
GSE124533_metadata_filtered =
  readRDS(file = here("tables", "GSE124533_metadata_filtered.rds"))


#Input
samples_metadata_pca_input = GSE124533_metadata_filtered
samples_log2fc_values_pca_input = human_engerixb_log2fc_filtered_sample_long_annotated

#Annotate genes by BTM
samples_genes_btm = samples_log2fc_values_pca_input %>% 
  ungroup() %>% 
  select(symbol = human_symbol, participant_id, timepoint, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = samples_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process_human = dge_btm_mean_process

dge_btm_mean_process_human %>% 
  saveRDS(file = here("tables", "hepatitisb_dge_btm_mean_process_human.rds"))

immune_order = c("SIGNAL TRANSDUCTION",
                  "CELL CYCLE",
                  "ECM AND MIGRATION",
                  "ENERGY METABOLISM", 
                  "INNATE RESPONSE",
                  "NEUTROPHILS",
                  "NK CELLS",
                  "IFN",
                  "MONOCYTES",
                  "PLATELETS",
                  "B CELLS",
                  "T CELLS")
```

##### GGPairs

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

human_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = sample_participants_5,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days, BTM"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple"))

human_correlation_ggpairs_btms_gsea_alldays

human_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "human_correlation_ggpairs_btms_gsea_alldays_EngerixB_Human.png"),
         height = 10,
         width = 10)

```

##### Ggplot manual

```{r fig.height=8, fig.width=10}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#Input
timepoint_selected = 7

#Plot table
plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == timepoint_selected) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == timepoint_selected) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove identical pairs
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))

#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-1, 1) +
  xlim(-1, 1) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste("Human, Hepatitis B, EngerixB Human, Day", timepoint_selected))
```

## Compare Human vs Mouse

```{r}
#Unite Mouse and Human datasets. Samples.
human_mouse_log2fc_samples = 
#Mouse
mouse_engerixb_log2fc_filtered_sample_long %>%
  ungroup() %>% 
  select(human_symbol, log2fc, participant_id,
         timepoint, value, day0) %>% 
  mutate(organism = "Mouse") %>% 
  
  #Human
  bind_rows(human_engerixb_log2fc_filtered_sample_long %>%
  ungroup() %>% 
  select(human_symbol, log2fc, participant_id,
         timepoint,  value, day0) %>% 
  mutate(organism = "Human"))

#Unite metadata
human_mouse_engerix_metadata = 
  #Human
  GSE124533_metadata_filtered %>% 
  select(geo_sample, participant_id, sex, organism, timepoint) %>% 
  mutate(timepoint = as.numeric(timepoint),
         sex = str_to_title(sex)) %>% 
  #Mouse
  bind_rows(
    mouse_engerixb_metadata %>%
      select(geo_sample, participant_id, sex, timepoint) %>%
      mutate(organism = "Mouse"))

human_mouse_log2fc_samples %>% 
  saveRDS(file = here("tables", "engerixb_human_mouse_log2fc_samples.rds"))

human_mouse_engerix_metadata %>% 
  saveRDS(file = here("tables", "engerixb_human_mouse_engerix_metadata.rds"))

```

```{r}
#Import
human_mouse_log2fc_samples =
  readRDS(file = here("tables", "engerixb_human_mouse_log2fc_samples.rds"))

human_mouse_engerix_metadata =
  readRDS(file = here("tables", "engerixb_human_mouse_engerix_metadata.rds"))

#Summarize
#Calculate the mean log2fc
human_mouse_engerixb_log2fc_avg = human_mouse_log2fc_samples %>% 
  group_by(organism, human_symbol, timepoint) %>% 
  summarize(
    mean_log2fc = mean(log2fc, na.rm = TRUE),
    sd_log2fc   = sd(log2fc, na.rm = TRUE),
    se_log2fc   = sd(log2fc, na.rm = TRUE) / sqrt(n()),   # erro padrão
    
    # Test t paired
    ttest = list(tryCatch(
      t.test(value, day0, paired = TRUE),
      error = function(e) NULL
    )),
    .groups = "drop"
  ) %>%
  mutate(
    p_value = map_dbl(ttest, ~ if (!is.null(.x)) .x$p.value else NA_real_),
    ci_low  = map_dbl(ttest, ~ if (!is.null(.x)) .x$conf.int[1] else NA_real_),
    ci_high = map_dbl(ttest, ~ if (!is.null(.x)) .x$conf.int[2] else NA_real_),
    t_stat  = map_dbl(ttest, ~ if (!is.null(.x)) .x$statistic else NA_real_),
    df      = map_dbl(ttest, ~ if (!is.null(.x)) .x$parameter else NA_real_),
    p_adj   = p.adjust(p_value, method = "BH")
  ) %>%
  select(-ttest) %>% 
  filter(human_symbol != "") %>%
  ungroup()


#Save
human_mouse_engerixb_log2fc_avg %>% 
  saveRDS(here("tables", "human_mouse_engerixb_log2fc_avg.rds"))

#Pivot wider
human_mouse_engerixb_log2fc_avg_wide = human_mouse_engerixb_log2fc_avg %>% 
  select(organism, human_symbol, timepoint, mean_log2fc) %>% 
  pivot_wider(names_from = "organism",
              values_from = "mean_log2fc") %>% 
  #Add padj columns
  inner_join(human_mouse_engerixb_log2fc_avg %>%
               select(organism, human_symbol, timepoint, p_adj) %>%
               pivot_wider(names_from = "organism",
                           values_from = "p_adj") %>%
               rename(padj_human = Human, padj_mouse = Mouse)) %>% 
  #Drop genes that are absent in one of the organisms 
  drop_na()

# Calculate the correlation
cor_data <- human_mouse_engerixb_log2fc_avg_wide %>%
  group_by(timepoint) %>% 
  filter(padj_human <= 0.05 | padj_human <= 0.05) %>%
  summarize(
    cor_value = cor(Mouse, Human, method = "pearson", use = "complete.obs"),
    label_pos_x = max(Mouse, na.rm = TRUE),
    label_pos_y = max(Human, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("r = ", round(cor_value, 2)))
```

```{r fig.height=6, fig.width=3}
ylim_range = c(-2.5, 2.5)
xlim_range = c(-2.5, 2.5)
human_mouse_log2fc_log2fc_avg_wide_scatterplot = human_mouse_engerixb_log2fc_avg_wide %>% 
  
  #Filter DEGs only
  filter(padj_human <= 0.05 | padj_mouse <= 0.05) %>%
  
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human) +
  
    #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #Facet
  facet_wrap(~paste("Day", timepoint), ncol = 1) +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_engerixb_log2fc_avg_wide %>%
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_engerixb_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  stat_smooth(method = "lm", geom = "line", color = "#4DBBD5FF", alpha = 0.5, se = F, size = 1,
              data = human_mouse_engerixb_log2fc_avg_wide %>%
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add correlation metrics (Pearson)
  geom_label(
    data = cor_data,
    aes(x = label_pos_x, y = label_pos_y, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 1,
    vjust = 1,
    size = 3
  ) +
  
  #Adjust the axes range
  ylim(ylim_range) +
  xlim(xlim_range) +

  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "none",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title= "Correlation plot, EngerixB,\nHuman vs Mouse")

human_mouse_log2fc_log2fc_avg_wide_scatterplot

human_mouse_log2fc_log2fc_avg_wide_scatterplot %>% 
  ggsave(filename = here("Figures", "EngerixB_human_mouse_degs_scatterplot.png"),
         width = 3, height = 6)
```



Single timepoint
```{r fig.height=3, fig.width=3}
human_mouse_log2fc_degs_scatterplot_singletimepoint = human_mouse_log2fc_log2fc_avg_wide %>% 
  filter(timepoint == 7) %>% 
  
  #Filter DEGs only
  filter(padj_human <= 0.05 | padj_mouse <= 0.05) %>%
  
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human) +
  
    #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #Facet
  facet_wrap(~paste("Day", timepoint), ncol = 1) +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_log2fc_log2fc_avg_wide %>%
               filter(timepoint == 7) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_log2fc_log2fc_avg_wide %>%
               filter(timepoint == 7) %>% 
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  stat_smooth(method = "lm", geom = "line", color = "#4DBBD5FF", alpha = 0.5, se = F, size = 1,
              data = human_mouse_log2fc_log2fc_avg_wide %>%
                filter(timepoint == 7) %>% 
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add correlation metrics (Pearson)
  geom_label(
    data = cor_data %>% 
      filter(timepoint == 7),
    aes(x = label_pos_x, y = label_pos_y, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 1,
    vjust = 1,
    size = 3
  ) +
  
  #Adjust the axes range
  ylim(ylim_range) +
  xlim(xlim_range) +

  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "none",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title= "Correlation plot, EngerixB,\nHuman vs Mouse")


human_mouse_log2fc_degs_scatterplot_singletimepoint %>% 
  ggsave(filename = here("Figures", "EngerixB_human_mouse_degs_scatterplot_singletimepoint.png"),
         width = 3, height = 3)

human_mouse_log2fc_degs_scatterplot_singletimepoint

```

#### Plot correlation between genes (log2fc for every sample)

```{r}
#Calculate correlation ------
correlation_long_all_mouse_human_timepoints <- human_mouse_log2fc_samples %>%
  select(timepoint, human_symbol, participant_id, log2fc) %>%
  group_by(timepoint) %>%
  
  # Pivot wider and calculate correlation
  group_modify(~ {
    .x %>%
      pivot_wider(names_from = participant_id, values_from = log2fc) %>%
      select(-human_symbol) %>%
      cor(use = "pairwise.complete.obs", method = "pearson") %>%
      as.data.frame() %>%
      rownames_to_column("participant1") %>%
      pivot_longer(-participant1, names_to = "participant2", values_to = "correlation") %>%
      filter(participant1 != participant2)
  }) %>%
  ungroup() %>% 
  
  #Filter by organism
  inner_join(human_mouse_engerix_metadata %>% 
               select(participant_id, participant_1_organism = organism, sex_p_1 = sex) %>% 
               distinct(),
             by = join_by("participant1" == "participant_id")) %>% 
  inner_join(human_mouse_engerix_metadata %>% 
               select(participant_id, participant_2_organism = organism, sex_p_2 = sex) %>% 
               distinct(),
             by = join_by("participant2" == "participant_id")) 


# Remove duplicated lines only if the organism is the same
correlation_long_all_timepoints_same_organisms = correlation_long_all_mouse_human_timepoints %>%
  filter(participant_1_organism == participant_2_organism) %>%
  mutate(
    participant_min = pmin(participant1, participant2),
    participant_max = pmax(participant1, participant2)
  ) %>%
  # Remove duplicates (ex: A vs B e B vs A)
  distinct(timepoint, participant_min, participant_max, .keep_all = TRUE) %>%
  select(timepoint, participant1, participant2, correlation, participant_1_organism, participant_2_organism, sex_p_1, sex_p_2) %>%
  group_by(timepoint) %>%
  arrange(-correlation)


#Unite table with different organisms with table with deduplicated rows with the same organism
correlation_long_all_mouse_human_timepoints = correlation_long_all_mouse_human_timepoints %>% 
  filter(participant_1_organism != participant_2_organism) %>% 
  bind_rows(correlation_long_all_timepoints_same_organisms)

#Summarize data and save table
correlation_long_all_mouse_human_timepoints_summary = correlation_long_all_mouse_human_timepoints %>% 
  
  #Filter only female and set Mouse as reference
  filter(sex_p_1 == "Female",
         sex_p_2 == "Female") %>%
  mutate(sex_comparison = paste(sex_p_1, "vs.", sex_p_2),
         comparison = paste(participant_1_organism, "vs.", participant_2_organism)) %>%
  group_by(timepoint, comparison) %>% 
  arrange(timepoint, -correlation) %>% 
  mutate(correlation = round(correlation, 2),
         timepoint = paste("Day", timepoint)) %>% 
  summarize(mean = mean(correlation, na.rm = T) %>% round(2),
            median = median(correlation, na.rm = T)%>% round(2),
            min = min(correlation, na.rm = T)%>% round(2),
            max = max(correlation, na.rm = T) %>% round(2),
            sd = sd(correlation),
            n = sum(!is.na(correlation)),
            se = sd/sqrt(n),
            t_critical = qt(0.975, df = n - 1),
            ci_lower = mean - t_critical * se,
            ci_upper = mean + t_critical * se,
            .groups = "drop") %>% 
  filter(comparison != "Human vs. Mouse")

correlation_long_all_mouse_human_timepoints_summary

#Create GT table
correlation_long_all_mouse_human_timepoints_gt = correlation_long_all_mouse_human_timepoints_summary %>% 
  gt::gt() 
#Save gt
correlation_long_all_mouse_human_timepoints_gt
gtsave(correlation_long_all_mouse_human_timepoints_gt, filename = here("tables", "EngerixB_correlation_long_all_mouse_human_timepoints_EngerixB_gt.docx"))

```

```{r fig.height=3, fig.width=3}
correlation_long_all_mouse_human_timepoints_summary_plot = correlation_long_all_mouse_human_timepoints_summary %>% 
  separate(timepoint, into = c("unit", "timepoint"), remove = T) %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1) +
  
  geom_errorbar(aes(ymin = ci_lower, 
                  ymax = ci_upper), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 1, 3, 7), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme_vaxgo() +
  scale_color_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = "Correlation between organisms, genes,\nHepatitis B, GSE124533, GSE19668",
       x = "Day",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 

correlation_long_all_mouse_human_timepoints_summary_plot %>% 
ggsave(filename = here("Figures", "EngerixB_samples_genes_correlation_all_timepoints_summary_plot.png"),
         width = 3,
         height = 3)

correlation_long_all_mouse_human_timepoints_summary_plot
  
```

#### BTMs

##### Scatterplot

```{r fig.height=5, fig.width=5}
#Import files
dge_btm_mean_process_human = 
  readRDS(file = here("tables", "hepatitisb_dge_btm_mean_process_human.rds"))
dge_btm_mean_process_mouse =
  readRDS(file = here("tables", "hepatitisb_dge_btm_mean_process_mouse.rds"))

#Merge tables
dge_btm_mean_process_human_long = dge_btm_mean_process_human %>% 
  pivot_longer(cols = -c(timepoint, process, group, subgroup), 
               names_to = "participant_id", 
               values_to = "log2fc") %>% 
  mutate(organism = "Human")

dge_btm_mean_process_mouse_long = dge_btm_mean_process_mouse %>% 
  pivot_longer(cols = -c(timepoint, process, group, subgroup), 
               names_to = "participant_id", 
               values_to = "log2fc") %>% 
  mutate(organism = "Mouse")

dge_btm_mean_process_samples = dge_btm_mean_process_human_long %>%
  bind_rows(dge_btm_mean_process_mouse_long) %>%
  filter(timepoint %in% c(1, 3, 7))

dge_btm_mean_process_samples %>% 
  saveRDS(file = here("tables", "engerixb_dge_btm_mean_process_samples.rds"))

dge_btm_mean_process_samples = 
  readRDS(file = here("tables", "engerixb_dge_btm_mean_process_samples.rds"))

dge_btm_mean_process = dge_btm_mean_process_samples %>% 
  group_by(organism, timepoint, process) %>% 
  summarize(mean_log2fc = median(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism",
              values_from = "mean_log2fc")

#Save
# dge_btm_mean_process %>% 
#   saveRDS(file = here("tables", "hepatitisb_dge_btm_mean_process_mouse_human.rds"))
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", "engerixb_dge_btm_mean_process_mouse_human.rds"))
```


```{r fig.height=5, fig.width=5}
#Import
dge_btm_mean_process =
  readRDS(file = here("tables", "hepatitisb_dge_btm_mean_process_mouse_human.rds"))

# Calculate the correlation
cor_data <- dge_btm_mean_process %>%
  # Group the data by timepoint
  group_by(timepoint) %>% 
  
  # For each timepoint, run Pearson correlation test between Mouse and Human
  summarize(
    cor_test = list(cor.test(Mouse, Human, method = "pearson")), # store full test result as a list
    label_pos_x = max(Mouse, na.rm = TRUE),  # position for label in the plot (x = max Mouse value)
    label_pos_y = max(Human, na.rm = TRUE),  # position for label in the plot (y = max Human value)
    .groups = "drop" # remove grouping after summarizing
  ) %>%
  
  # Extract correlation coefficient (r) and p-value from cor_test, and create a label
  mutate(
    cor_value = map_dbl(cor_test, ~ .x$estimate),  # correlation coefficient
    p_value   = map_dbl(cor_test, ~ .x$p.value),   # p-value
    label = paste0(
      "r = ", round(cor_value, 2),                 # round r to 2 decimal places
      ", p = ", signif(p_value, 3)                 # show p with 3 significant digits
    )
  )
```

```{r fig.height=8, fig.width=5}

human_mouse_log2fc_BTMs_avg_wide_scatterplot = dge_btm_mean_process %>% 
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human,
      color = group) +
  
  #Add points
  geom_point() +
  
#Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Facet
  facet_wrap(~paste("Day", timepoint), 
             ncol = 1, 
             scales = "free", axes = "all") +
  
  #Color
  scale_color_manual(values = immune_colors) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  
  #Add correlation metrics (Pearson)
  geom_label(data = cor_data,
             aes(x = label_pos_x, y = label_pos_y, label = label),
             inherit.aes = FALSE,
             fill = "white",
             hjust = 1,
             vjust = 1,
             size = 3)+

  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "right",
        legend.direction = "vertical") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       color = "BTM group",
       title= "Correlation plot, EngerixB,\nHuman vs Mouse")

#Save 
human_mouse_log2fc_BTMs_avg_wide_scatterplot %>% 
  ggsave(filename = here("Figures", "EngerixB_human_mouse_log2fc_BTMs_avg_scatterplot.png"),
         height = 8,
         width = 5)
human_mouse_log2fc_BTMs_avg_wide_scatterplot
```

Single timepoint

```{r fig.height=3, fig.width=3}
human_mouse_log2fc_BTMs_scatterplot_singletimepoint = dge_btm_mean_process %>% 
  filter(timepoint == 7) %>% 
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human) +
  
  #Add points
  geom_point(aes(fill = group),
             color = "black", size = 2, pch = 21) +
  
  ylim(-0.4, 0.4) +
  xlim(-0.4, 0.4) +
#Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Facet
  facet_wrap(~paste("Day", timepoint), 
             ncol = 1, 
             scales = "free", axes = "all") +
  
  #Color
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  
  #Add correlation metrics (Pearson)
  geom_label(data = cor_data %>% 
               filter(timepoint == 7),
             aes(x = 0.2, y = 0.3, label = label),
             inherit.aes = FALSE,
             fill = "white",
             hjust = 1,
             vjust = 1,
             size = 3)+

  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "none",
        legend.direction = "vertical") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       color = "BTM group",
       title= "Correlation plot, EngerixB,\nHuman vs Mouse")

human_mouse_log2fc_BTMs_scatterplot_singletimepoint

#Save 
human_mouse_log2fc_BTMs_scatterplot_singletimepoint %>% 
  ggsave(filename = here("Figures", "EngerixB_human_mouse_log2fc_BTMs_avg_scatterplot_singletimepoint.png"),
         height = 3,
         width = 3)
```

##### Heatmap

```{r}
#Matrix data
matrix_data = dge_btm_mean_process %>% 
  select(process, Human_Day1, Mouse_Day2) %>% 
  drop_na() %>% 
  column_to_rownames("process") %>% 
  as.matrix()

# Column and row annotation
######## Annotations
annotation_heatmap = tribble(~condition, ~timepoint, ~organism,
        "Human_Day1", 1, "Human",
        "Mouse_Day2", 2, "Mouse")

#Arrange columns in the same order as the matrix
ann_cols_heatmap = matrix_data %>% 
  colnames() %>% 
  data.frame() %>% 
  rename(condition = '.') %>% 
  inner_join(annotation_heatmap, by = "condition") %>% 
  distinct() %>% 
  arrange(condition)

#Arrange matrix columns 
mat = matrix_data %>% 
  as.data.frame() %>% 
  select(c(ann_cols_heatmap$condition %>% unique())) %>% 
  as.matrix() 

#Column to rownames
ann_cols_heatmap = ann_cols_heatmap %>% 
  column_to_rownames("condition")

#Annotate rows
ann_rows_heatmap = mat %>% 
  rownames() %>% 
  as.data.frame() %>% 
  rename("process" = ".") %>% 
  inner_join(dge_btm_mean_process %>% 
               select(process, group, subgroup), 
             by = "process") %>% 
  column_to_rownames("process")
  

#Check dimensions
dim(mat)
dim(ann_cols_heatmap)

#Colors from required.R
btm_immune_groups
colors_immune

```

```{r fig.width= 6, fig.height=20}
#Heatmap annotation
ha = HeatmapAnnotation(df = ann_cols_heatmap, 
                       col = colors_immune,
                       gp = gpar(col = "black"), 
                       simple_anno_size = unit(0.3, "cm"),
                       annotation_name_side = "right")

row_ha = HeatmapAnnotation(df = ann_rows_heatmap,
                       col = btm_immune_groups,
                       which= "row",
                       simple_anno_size = unit(0.2, "cm"),
                       show_annotation_name = T,
                       show_legend = T)

# Values colors
col_fun <- colorRamp2(c(-2,0,2), c("#45efc6", "white", "#165289"))

file = "BTM_Modules_Saureus_Mouse_Human"

width_image = 25
height_image = 30
width = unit(1, "cm")
height = unit(25, "cm")
column_names_gp = gpar(fontsize = 8)
row_names_gp = gpar(fontsize = 8)
rect_gp = gpar(col = "black", lwd = 0.5)

fileimageheatmap_grouped = here("Figures", paste0(file, sep ="_", "Complexheatmap.png"))

png(file = fileimageheatmap_grouped, 
    width = width_image, 
    height = height_image, 
    units="cm", 
    res=900)

heatmap_plot = Heatmap(mat,
                        top_annotation = ha, 
                       left_annotation = row_ha,
                        show_row_names = TRUE,
                        row_names_side = "right",
                        show_heatmap_legend = TRUE,
                        heatmap_legend_param = list(direction = "horizontal"),
                        col = col_fun,
                        show_column_names = TRUE,
                        column_title = file,
                        column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                        column_names_rot = 45,
                        column_names_gp = column_names_gp,
                        rect_gp = rect_gp,
                        na_col = "black",
                        name = "Mean Log2FC",
                        cluster_rows = TRUE,
                        row_names_gp = row_names_gp, row_km = 2,
                        row_dend_width = unit(3, "cm"),
                        row_split = NULL,
                        column_split = NULL,
                        row_gap = unit(2, "mm"),
                        cluster_columns = TRUE,
                        column_gap = unit(8, "mm"),
                        show_column_dend = F,
                        show_row_dend = T, 
                        row_dend_side = "left",
                        column_dend_side = "top",
                        clustering_distance_rows = "euclidean",
                        border = TRUE,
                        width = width,
                        height = height)

heatmap_plot = draw(heatmap_plot, 
     annotation_legend_side = "left", 
     heatmap_legend_side = "left")

heatmap_plot

dev.off()


```

# Compare with FIT prediction

## Run FIT on a training dataset

```{r}
#Get log2fc mean
GSE19668_input2FIT = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
    filter(timepoint == 2) %>% 
  select(participant_id, MM.Entrez = entrezgene_id, EffectSize = log2fc) %>% 
  group_by(MM.Entrez) %>% 
  summarize(mean_log2fc = mean(EffectSize))

#Save
GSE19668_input2FIT %>% 
  write_csv(file = here("FIT_Mouse2man", "GSE19668_day2_FIT_mouse.csv"))

#Run FIT
GSE19668_input2FIT_day2 = FIT.mouse2man::FIT(MouseFile = here("FIT_Mouse2man", "GSE19668_day2_FIT_mouse.csv"), DataType = "rnaseq")
```

### Unite log2FC values from mouse, human, and FIT

```{r fig.height=5, fig.width=8}
#Bind rows
mouse_human_fit_log2fc = bind_rows(GSE19668_input2FIT_day2 %>%
                                     #Get DEGs for FIT
                                     filter((CI_low != 0 | CI_high != 0) & !is.na(CI_low) & !is.na(CI_high)) %>%  
                                   select(human_symbol = Human.symbol,
                                          log2fc = FIT_prediction) %>%
                                     mutate(organism = "FIT"),
                                   
                                   #Get DEGs for Human
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_human <= 0.05) %>% 
                                     select(human_symbol, log2fc = Human) %>%
                                     mutate(organism = "Human"),
                                   
                                   #Get DEGs for Mouse
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_mouse <= 0.05) %>% 
                                     select(human_symbol, log2fc = Mouse) %>%
                                     mutate(organism = "Mouse"))


#Pivot wider
mouse_human_fit_log2fc_wide = GSE19668_input2FIT_day2 %>%
            select(human_symbol = Human.symbol,
                   FIT_prediction) %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Human),
            by = "human_symbol") %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Mouse),
            by = "human_symbol")

#Save
mouse_human_fit_log2fc %>% 
  saveRDS(file = here("tables", "mouse_human_fit_log2fc_long_SAureus.rds"))

mouse_human_fit_log2fc_wide %>% 
  saveRDS(file = here("tables", "mouse_human_fit_log2fc_wide_SAureus.rds"))
```

## Distribution

### Import tables

```{r fig.height=5, fig.width=8}
#Import
mouse_human_fit_log2fc = readRDS(file = here("tables", "mouse_human_fit_log2fc_long_SAureus.rds"))
mouse_human_fit_log2fc_wide = readRDS(file = here("tables", "mouse_human_fit_log2fc_wide_SAureus.rds"))

GSE33341_metadata = readRDS(file = here("tables", "GSE33341_metadata.rds"))
GSE19668_metadata = readRDS(file = here("tables", "GSE19668_metadata.rds"))
```

### Plot histogram

```{r fig.height=5, fig.width=8}
#Plot histogram
mouse_human_fit_log2fc_histogram = mouse_human_fit_log2fc %>% 
  ggplot() +
  aes(x = log2fc,
      fill = organism) +
  geom_density(aes(y = after_stat(count),
                   color = organism),
                  position = position_identity(), 
               alpha = 0.1,
               size = 1,
              show.legend = T) +
  scale_fill_manual(values = c(FIT = "#4361ee",
                               Mouse = "#4DBBD5FF",
                               Human = "#546E7AFF"))+
  scale_color_manual(values = c(FIT = "#4361ee",
                                Mouse = "#4DBBD5FF",
                                Human = "#546E7AFF"))+
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        legend.key = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0),
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(limits = c(-4, 4)) +
  labs(color = "",
       fill = "")

mouse_human_fit_log2fc_histogram

#Faceted
mouse_human_fit_log2fc_histogram_faceted = mouse_human_fit_log2fc_histogram +
  facet_wrap(~organism,
             scales = "free_y")


#Plot scatter plot
mouse_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = Mouse,
      y = Human) +
#All DEGs
geom_point(color = "#4DBBD5FF",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "Mouse L2FC",
         y = "Human L2FC")

mouse_human_scatterplot


fit_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = FIT_prediction,
      y = Human) +
#All DEGs
geom_point(color = "#4361ee",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "FIT L2FC",
         y = "Human L2FC")

#Histograms
(((mouse_human_fit_log2fc_histogram + mouse_human_fit_log2fc_histogram_faceted) + 
    plot_layout(widths = c(0.5, 1), guides = "collect") +
    plot_annotation(theme = theme(legend.position = "top")))  /
    #Scatter plots
    (mouse_human_scatterplot + fit_human_scatterplot)) +
  plot_layout(heights = c(0.5, 1))

```

## BTMs

### Means of means
```{r}
#Annotate genes by BTM
dge_btm_mean_process = mouse_human_fit_log2fc %>%
  filter(organism != "Mouse") %>% 
  rename(symbol = human_symbol) %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  group_by(organism, process) %>% 
  summarize(mean_log2fc = mean(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism", values_from = "mean_log2fc") 

dge_btm_mean_process
```

```{r fig.height=5, fig.width=7}

#Calculate correlation
cor_labels <- dge_btm_mean_process %>%
  summarise(cor_value = cor(Human, FIT, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
dge_btm_mean_process %>% 
ggplot(aes(x = FIT, y = Human)) +
  geom_point(aes(fill = group), alpha = 1, size = 2, color = "black", pch = 21) +
  stat_smooth(aes(x = FIT, y = Human),
              method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
    geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  theme_vaxgo() +
  ylim(-3, 3) +
  xlim(-3, 3) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  labs(x = "FIT Log2FC", 
       y = "Human Log2FC", 
       color = "",
       title = "Mouse vs. Human")



```



### Samples means

#### Import data from different organisms
```{r}

engerixb_dge_btm_mean_process_samples =
  readRDS(file = here("tables", "engerixb_dge_btm_mean_process_samples.rds"))

human_mouse_engerix_metadata %>% 
  saveRDS(file = here("tables", "engerixb_human_mouse_engerix_metadata.rds"))

#Calculate the mean log2fc by module
example_dge_btm_mean_process_sample_long = engerixb_dge_btm_mean_process_samples %>%
  group_by(organism, participant_id, timepoint, process) %>% 
  summarize(group,
            subgroup, 
            mean_log2fc = mean(log2fc, na.rm = T)) %>% 
  distinct() %>% 
  ungroup()

example_dge_btm_mean_process_sample_long %>% 
  saveRDS(file = here("tables", "engerixb_dge_btm_mean_process_sample_long.rds"))

#Pivot wider
example_dge_btm_mean_process_sample_wide = example_dge_btm_mean_process_sample_long %>% 
  group_by(organism, timepoint, process) %>% 
  pivot_wider(names_from = "participant_id", 
              values_from = "mean_log2fc") %>% 
  ungroup()

example_dge_btm_mean_process_sample_wide
  
```

#### Summarize interindividual correlation

```{r}
# --- 1) Mouse vs Human, same timepoint ---
#Get timepoints
timepoints_human <- unique(example_dge_btm_mean_process_sample_long %>% filter(organism == "Human") %>% pull(timepoint))
timepoints_mouse <- unique(example_dge_btm_mean_process_sample_long %>% filter(organism == "Mouse") %>% pull(timepoint))

#Get combinations
timepoint_combinations <- expand.grid(timepoint_human = timepoints_human,
                                      timepoint_mouse = timepoints_mouse,
                                      stringsAsFactors = FALSE) %>% 
  filter(timepoint_human <= timepoint_mouse) #filter symetric pairs

timepoint_combinations
```


```{r}
# Pearson Correlation
# --- 1) Human vs Mouse (between timepoints) ---
cor_human_mouse_timepoint <- timepoint_combinations %>%
  pmap_dfr(function(timepoint_human, timepoint_mouse) {
    
    # get subjects from the given timepoint
    human_df <- example_dge_btm_mean_process_sample_long %>%
      filter(organism == "Human", timepoint == timepoint_human) %>%
      select(participant_id, process, mean_log2fc)
    
    mouse_df <- example_dge_btm_mean_process_sample_long %>%
      filter(organism == "Mouse", timepoint == timepoint_mouse) %>%
      select(participant_id, process, mean_log2fc)
    
    # keep the common genes
    common_process <- intersect(human_df$process, mouse_df$process)
    human_df <- human_df %>% filter(process %in% common_process)
    mouse_df <- mouse_df %>% filter(process %in% common_process)
    
    # pivot wider
    human_wide <- human_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    mouse_wide <- mouse_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    
    # Get subjects id
    human_ids <- setdiff(colnames(human_wide), "process")
    mouse_ids <- setdiff(colnames(mouse_wide), "process")
    
    # Calculate correlation for every pair of organism 1 vs organism 2
    expand.grid(human = human_ids, mouse = mouse_ids, stringsAsFactors = FALSE) %>%
      mutate(
        cor_test = map2(human, mouse, ~{
          x <- human_wide[[.x]]
          y <- mouse_wide[[.y]]
          # remove NAs
          keep <- complete.cases(x, y)
          cor.test(x[keep], y[keep], method = "pearson")
        }),
        r       = map_dbl(cor_test, ~.x$estimate),
        p_value = map_dbl(cor_test, ~.x$p.value),
        ci_low  = map_dbl(cor_test, ~.x$conf.int[1]),
        ci_high = map_dbl(cor_test, ~.x$conf.int[2])
      ) %>%
      select(-cor_test) %>%
      mutate(timepoint_human = timepoint_human %>% as.numeric(),
             timepoint_mouse = timepoint_mouse %>% as.numeric())
  }) %>% 
  #Summarize
  group_by(timepoint_human, timepoint_mouse) %>%
  summarize(
    comparison = "Human vs. Mouse",
    mean = mean(r, na.rm = TRUE),
    sd   = sd(r, na.rm = TRUE),
    n      = sum(!is.na(r)),
    se     = sd / sqrt(n),
    tval   = qt(0.975, df = n - 1),
    ci_low = mean - tval * se,
    ci_high = mean + tval * se,
    .groups = "drop"
  )
#Print
cor_human_mouse_timepoint
```


```{r}
# --- 2) Human vs Human
cor_human_human <- example_dge_btm_mean_process_sample_long %>%
  filter(organism == "Human") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, process, mean_log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = mean_log2fc) %>%
      select(-process)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Human vs. Human",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup() %>% 
  mutate(timepoint = as.numeric(timepoint), 
         timepoint_human = timepoint)
  

# --- 3) Mouse vs Mouse ---
cor_mouse_mouse <- example_dge_btm_mean_process_sample_long %>%
  filter(organism == "Mouse") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, process, mean_log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = mean_log2fc) %>%
      select(-process)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Mouse vs. Mouse",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup() %>% 
  mutate(timepoint = as.numeric(timepoint), 
         timepoint_mouse = timepoint)


#Merge tables
example_samples_btm_correlation_all_timepoints_summary = cor_human_mouse_timepoint %>% 
  filter(timepoint_human == timepoint_mouse) %>% 
  mutate(timepoint = timepoint_mouse) %>% 
  select(!c(timepoint_mouse, timepoint_human)) %>% 
  select(timepoint, comparison, everything()) %>% 
  bind_rows(cor_human_human, cor_mouse_mouse)

#Print
example_samples_btm_correlation_all_timepoints_summary

example_samples_btm_correlation_all_timepoints_summary %>% 
  saveRDS(file = here("tables", "engerixb_samples_btm_correlation_all_timepoints_summary.rds"))
```

Plot scatterplot
```{r fig.height=3, fig.width=3}
#Scatterplot
example_samples_btm_correlation_all_timepoints_summary_plot = example_samples_btm_correlation_all_timepoints_summary %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1,
                data = example_samples_btm_correlation_all_timepoints_summary %>% 
                  filter(comparison != "Human vs. Human")) +
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.5,
                angle = 90,
                linewidth = 1,
                data = example_samples_btm_correlation_all_timepoints_summary %>% 
                  filter(comparison == "Human vs. Human")) +
  
  
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 1, 3, 7), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     # limits = c(0.4, 1)
                     ) +
  theme_vaxgo() +
  scale_color_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = "Correlation between organisms,\nBTMs, EngerixB",
       x = "Day",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 

example_samples_btm_correlation_all_timepoints_summary_plot

#Save image
example_samples_btm_correlation_all_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "EngerixB_samples_btm_correlation_all_timepoints_summary_plot.png"),
         width = 3,
         height = 5)

example_samples_btm_correlation_all_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", "EngerixB_samples_btm_correlation_all_timepoints_summary_plot.png"),
         width = 3,
         height = 3)

example_samples_btm_correlation_all_timepoints_summary_plot
  
```
























#Performance tests
## ROC Curve
Do for one direction (up and then down)
Do one sided t test for mouse, fit, and human
Positive control: Human
Neg control: Random permutation


Import data
```{r}
#ImmuneGO ontologies
ImmuneGO_Mouse_Manual_genes <- read_excel(here("VaxGO", "ImmuneGO_Mouse.xlsx")) %>% 
 separate_rows(genes)
immunego_btm_manual_grouped = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genesets.csv"))
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))

#Import DEGs table
human_mouse_engerixb_log2fc_avg =
  readRDS(file = here("tables", "human_mouse_engerixb_log2fc_avg.rds"))

dge_long_genes = human_mouse_engerixb_log2fc_avg  %>% 
  rename(log2fc = mean_log2fc, se = se_log2fc, sd = sd_log2fc, adj_p_val = p_adj, symbol = human_symbol) 

#Use only same genes
dge_long_genes = dge_long_genes %>% 
  filter(organism == "Mouse") %>% 
  select(symbol) %>% 
  distinct() %>% 
  inner_join(dge_long_genes %>%
               filter(organism == "Human") %>%
               select(symbol) %>%
               distinct(), 
             by = "symbol") %>% 
  inner_join(dge_long_genes,
             by = "symbol")

```

Calculate ROC curve
```{r fig.height=4, fig.width=6}
set.seed(123)

######### Input -----

filename = "EngerixB_AllGenes"
filename_title ="EngerixB,\nAll genes"

genes_df_input = dge_long_genes 
  # Filter immune genes only
  # filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))


######### UP -----------
#Do one sided t-test for each model
dge_long_genes_up = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = F), #onesided, p(gene) > 0 (right tail)
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj < 0.05, "up", "not_up")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj)

#Assign human log2fc direction as truth
truth_human = dge_long_genes_up %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism",
              values_from = "direction") %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#Generate comparable distributions
perfect_prediction = truth_human %>%
  mutate(prob = adj_p_val) %>%
  rename(pred = truth)

uniform_prediction = perfect_prediction %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()),
         organism = "Uniform")

#Get only up and not up genes from models
models_up = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = F), #one sided, higher
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj < 0.05, "up", "not_up")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj) %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         prob = adj_p_val) %>% 
  #Add controls
  bind_rows(
    perfect_prediction,
    permutation_prediction,
    uniform_prediction
    )

#Get probabilities (UP)
classified_df_probs_up = truth_human %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(symbol, timepoint, pred, prob, model = organism),
             by = join_by("symbol", "timepoint")) %>% 
  mutate(not_prob = 1 - prob,
         sum = prob + not_prob,
         truth = as.factor(truth)) %>% 
  distinct()


#Get ROC
roc_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Human", "Mouse", "Uniform", "FIT")) %>%
  select(symbol, timepoint, truth, pred, model, prob, not_prob, sum) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Up vs Non-Up")

######### DOWN -----------
#Do one sided t-test for each model
dge_long_genes_down = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = T), #onesided, p(gene) < 0 (left tail)
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj < 0.05, "down", "not_down")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj)

#Assign human log2fc direction as truth
truth_human = dge_long_genes_down %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism",
              values_from = "direction") %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#Generate comparable distributions
perfect_prediction = truth_human %>%
  #Invert probabilities
  mutate(prob = 1 - adj_p_val) %>%
  rename(pred = truth)

uniform_prediction = perfect_prediction %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()),
         organism = "Uniform")


#Get only up and not up genes from models
models_down = genes_df_input %>% 
  mutate(
    se = (ci_high - ci_low) / (2 * 1.96),
    z = log2fc / se,
    pval_one_sided = pnorm(z, lower.tail = T), #one sided, less
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    #Invert probabilities
    pval_one_sided_adj = 1 - pval_one_sided_adj,
    direction = if_else(pval_one_sided_adj < 0.05, "down", "not_down")) %>%
  select(-adj_p_val) %>% 
  rename(adj_p_val = pval_one_sided_adj) %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         prob = adj_p_val) %>% 
  #Add controls
  bind_rows(perfect_prediction, uniform_prediction)

#Get probabilities (DOWN)
classified_df_probs_down = truth_human %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(symbol, timepoint, pred, prob, model = organism),
             by = join_by("symbol", "timepoint")) %>% 
  mutate(not_prob = 1 - prob,
         sum = prob + not_prob,
         truth = as.factor(truth)) %>% 
  distinct()



#Get ROC
roc_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Human", "Mouse", "Uniform")) %>%
  select(symbol, timepoint, truth, pred, model, prob, not_prob, sum) %>%
  group_by(model, timepoint) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Down vs Non-Down")

#Calculate AUC --------
#Get AUC
auc_results = classified_df_probs_down %>%
  group_by(model, timepoint) %>%
  yardstick::roc_auc(truth, prob) %>% 
  mutate(comparison = "Down vs Non-Down") %>% 
  bind_rows(
    classified_df_probs_up %>%
      group_by(model, timepoint) %>%
      yardstick::roc_auc(truth, prob) %>%
      mutate(comparison = "Up vs Non-Up")) %>% 
  mutate(.estimate = round(.estimate, 2)) %>% 
  rename(roc_auc = .estimate)

#Merge data -------
roc_curve_data = roc_curve_up_notup %>% 
  bind_rows(roc_curve_down_notdown) %>% 
  left_join(auc_results, 
            by = c("model", "comparison", "timepoint")) %>% 
  mutate(condition = "EngerixB_Vaccination",
         filename = filename)

roc_curve_data %>% 
  saveRDS(file = here("tables", paste0("roc_curve_data_", filename, ".rds")))

auc_results
roc_curve_data 

```


```{r fig.height=4, fig.width=6}
##### PLOT ROC Curve ------
roc_curve_plot = roc_curve_up_notup %>% 
  bind_rows(roc_curve_down_notdown) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(size = 1, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_grid(~fct_rev(comparison)~paste0(timepoint, "h")) + 
  scale_color_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "FIT"   = "#4361ee",
    "Human" = "#90A4AEFF",
    "Uniform" = "gray80"
    )) +
  labs(
    title = filename_title, 
  x = "FPR",
  y = "TPR",
  color = "Model") + 
  theme(
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10,),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3)


roc_curve_plot

roc_curve_plot %>%
  ggsave(filename = here("Figures", paste0("roc_curve_plot_", filename, ".png")),
         width = 6,
         height = 4)

```



