---
title: "Hepatitis B"
author: "Wasim Aluísio Prates-Syed"
date: "2025-07-23"
output: 
  html_document: 
    toc: TRUE
    toc_depth: 4
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE, cache = TRUE)
```

# Hepatitis B

#Import required codes

```{r}
# install.packages("here")
library(here)
source(here("scripts_notebooks", "required.R"))
```


# 1. General analyses
## GSE19668 (Mouse)

```{r}
#Import
GSE120661_counts_not_normalized_log2 = readRDS(here("tables", "GSE120661_counts_not_normalized_log2.rds")) %>% 
  as.data.frame()
GSE120661_metadata = read_csv(file = here("tables", "GSE120661_metadata.csv"))


#Filter data
mouse_engerixb_metadata = GSE120661_metadata %>% 
  filter(treatment == "Engerix B",
         tissue == "PBMC") %>% 
  mutate(timepoint = (timepoint / 24) %>% round(1),
         participant_id = paste0("ENG_BL_", animal_replicate)) %>% 
  filter(timepoint %in% c(0, 1, 3, 7))

mouse_engerixb_metadata %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))

#Get log2 values
mouse_engerixb_log2 = GSE120661_counts_not_normalized_log2 %>% 
  select(mouse_engerixb_metadata %>% pull(geo_sample))

```

#### Annotate genes

```{r}
## Set up biomart
ensembl_mart <- useMart("ensembl")
ensembl_mart <- useDataset("mmusculus_gene_ensembl", mart = ensembl_mart)

## Get the MGI symbol
gene_id <- mouse_engerixb_log2 %>% rownames()

#Get other annotations
listAttributes(ensembl_mart)

#Annotate
annot <- getBM(
  attributes = c("agilent_sureprint_g3_ge_8x60k", #Agilent microarray data (design ID 028005, retrieved from GEO)
                 "ensembl_gene_id",
                 "entrezgene_id",
                 "mgi_symbol",
                 "description"),
  filters = "agilent_sureprint_g3_ge_8x60k",
  values = gene_id,
  mart = ensembl_mart)

beep()

#Get lookup table from FIT
#Install model package
# library(devtools)
# devtools::install_github('shenorrLabTRDF/FIT.mouse2man')
library(FIT.mouse2man)
HS_MM_Symbol_Entrez = FIT.mouse2man::HS_MM_Symbol_Entrez %>% 
  clean_names() %>% 
  rename(mouse_entrez = mouse_ortholog)

#Get annotations for each probe 
### Filter probes with highest median expression
mouse_engerixb_log2summary = mouse_engerixb_log2 %>% #20,845 probes
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(annot %>% 
              rename(probe = agilent_sureprint_g3_ge_8x60k), 
            by = "probe") %>%  #24.009 rows
  #Filter ensembl id
  group_by(ensembl_gene_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_gene_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() %>% 
  rename(mouse_entrez = entrezgene_id) %>% 
  
  #Filter hgnc symbol
  inner_join(HS_MM_Symbol_Entrez, by = c("mouse_entrez", "mgi_symbol")) %>% 
  group_by(hgnc_symbol) %>% 
  slice_max(median, n = 1)  %>% #14.396
  ungroup()

#Filter out genes with multiple mgi_symbols
hgnc_symbol_dupes = mouse_engerixb_log2summary %>% 
  #filter ensembl id with multiple mgi_symbols
  get_dupes(hgnc_symbol)

mouse_engerixb_log2summary = mouse_engerixb_log2summary %>% 
  filter(!hgnc_symbol %in% c(hgnc_symbol_dupes$hgnc_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
mouse_engerixb_log2_filtered_probes = mouse_engerixb_log2 %>% 
  rownames_to_column("probe") %>% 
  inner_join(mouse_engerixb_log2summary, by = "probe") %>%   #214,396
  select(probe, mouse_entrez, human_symbol = hgnc_symbol, c(mouse_engerixb_metadata %>% pull(geo_sample)))
#14395
mouse_engerixb_log2_filtered %>% dim()



mouse_engerixb_log2_hgnc_symbol = mouse_engerixb_log2_filtered_probes %>% 
  select(human_symbol, c(mouse_engerixb_metadata$geo_sample)) %>% 
  column_to_rownames("human_symbol")


#Save 
mouse_engerixb_log2_hgnc_symbol %>% 
  saveRDS(here("tables", "hepatitisb_engerixb_mouse_exprs.rds"))
```

#### Import

```{r}
#Import
mouse_engerixb_log2_hgnc_symbol = readRDS(here("tables", "hepatitisb_engerixb_mouse_exprs.rds"))
mouse_engerixb_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))

#Create Expression Set
eset_mouse_engerixb <- ExpressionSet(
  assayData = mouse_engerixb_log2_hgnc_symbol %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(mouse_engerixb_metadata %>% 
                                   column_to_rownames("geo_sample")))

#Save
eset_mouse_engerixb %>% saveRDS(file = here("tables", "hepatitisb_engerixb_mouse_eset.rds"))
```


```{r}
#Annotate
mouse_engerixb_log2fc_filtered_sample_long = mouse_engerixb_log2_hgnc_symbol %>%
  rownames_to_column("human_symbol") %>% 
  pivot_longer(cols = -human_symbol,
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(mouse_engerixb_metadata %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) %>%
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>%
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")

#Save
mouse_engerixb_log2fc_filtered_sample_long %>% 
  saveRDS(here("tables", "mouse_engerixb_log2fc_filtered_sample_long.rds"))


#Pivot wider
mouse_engerixb_log2fc_filtered_sample_wide = mouse_engerixb_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")

beep()
```


#### Differential expression (Limma)
Paired samples design
```{r}

#Import
mouse_engerixb_log2_hgnc_symbol = readRDS(here("tables", "hepatitisb_engerixb_mouse_exprs.rds"))
mouse_engerixb_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))
eset_mouse = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_eset.rds"))

#Create model matrix
design <- model.matrix(~ condition + 0, # 0 defines the way we compare the samples
                       eset_mouse) 
colnames(design) <- make.names(colnames(design))  # Avoids blank spaces

# Estimate the correlation between measurements made on the same subject (paired-samples)
corfit <- duplicateCorrelation(eset_mouse, 
                               design = design, 
                               block = eset_mouse$participant_id)
corfit$consensus #-0.02688727

#Fit Limma model
fit <- lmFit(eset_mouse, 
             design, 
             block = eset_mouse$participant_id, 
             correlation = corfit$consensus) #Use inter-subject correlation as input into the linear model fit

#Get conditions
colnames(design) %>% 
  as.data.frame() %>% 
  arrange(.)

#Get contrasts 
cont.matrix <- makeContrasts(conditionEngerix.B_24_PBMC	 - conditionEngerix.B_0_PBMC,
                             conditionEngerix.B_72_PBMC	 - conditionEngerix.B_0_PBMC,
                             conditionEngerix.B_168_PBMC	 - conditionEngerix.B_0_PBMC,
                             levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

#Get DEGs

## For every contrast
results_list <- lapply(1:ncol(fit2), function(i) {
  topTreat(fit2, coef = i, adjust = "fdr", p.value = 1, number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>% 
    rownames_to_column("symbol")
  })

# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs and clean dataset
all_degs_mouse_engerixb = all_contrasts_degs %>% 
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  mutate(condition = case_when(comparisonA == "conditionEngerix.B_24_PBMC"	~ "Engerix_24_PBMC",
                               comparisonA == "conditionEngerix.B_72_PBMC"	~ "Engerix_72_PBMC",
                               comparisonA == "conditionEngerix.B_168_PBMC"	~ "Engerix_168_PBMC",
                               T~comparisonA),
         organism = "Mouse") %>% 
  select(condition, everything()) %>% 
  clean_names() %>% 
  rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)
	
# #Save
all_degs_mouse_engerixb %>% saveRDS(file = here("tables", "hepatitisb_engerixb_mouse_degs_mouse.rds"))

all_degs_mouse_engerixb %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarize(n_degs = n())

```



#### Correlation plots, pairwise

Plot correlations between mouse samples across all timepoints using
pairwise scatterplots and density plots from the GGally package.

To do this, define custom plotting functions for densities on the
diagonal and points on the lower panels, then apply ggpairs to the
filtered log2 fold change expression matrix. Reorder the timepoints as
factors, color and fill by timepoint, and apply a custom theme along
with manual color scales.

The result is a matrix of correlation plots titled “All days,
HepB/EngerixB, GSE120661, C57BL/6J”.

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(mouse_engerixb_log2fc_filtered_sample_wide %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, HepB/EngerixB, GSE120661, C57BL/6J", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) 

mouse_correlation_ggpairs_alldays

# Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "hepatitisb_engerixb_mouse_samples_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)
beep()
```

#### PCA

Perform a PCA on the **mouse** EngerixB dataset using log2 fold change
values from each **sample**.

Merge metadata with expression values, reshape the data into a wide
matrix with genes as rows and samples as columns, and replace missing or
non-finite values with zeros.

Run PCA on the transposed matrix to capture sample-level variation, then
extract the first two principal components and merge them back with
sample annotations.

Finally, plot PC1 versus PC2, coloring and filling by timepoint and
labeling by participant, with custom color scales and a themed layout.

```{r fig.height=4, fig.width=5}

#Import
mouse_engerixb_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))
mouse_engerixb_log2fc_filtered_sample_long = readRDS(here("tables", "mouse_engerixb_log2fc_filtered_sample_long.rds"))

#Input
samples_metadata_pca_input = mouse_engerixb_metadata
samples_log2fc_values_pca_input = mouse_engerixb_log2fc_filtered_sample_long


#Import metadata
annotation_samples_pca = samples_metadata_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = samples_log2fc_values_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, human_symbol, log2fc) %>% 
  filter(!human_symbol == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("human_symbol") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Mouse, HepB, EngerixB, GSE120661",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

beep()

```







#### BTMs

Annotate mouse genes with Blood Transcription Module (BTM) functional
groups and use them to assess correlations across timepoints. Load
pre-curated BTM annotations and clean the table of functional groups,
then merge it with an external list of immune-related genes. Map mouse
log2 fold change values to BTM processes by joining gene symbols with
their corresponding modules. For each participant and timepoint, compute
the mean log2 fold change per process, reshape the data into a wide
format, and attach module group information.

```{r}
#Import files
btm_annotation_genes =
  read_csv(file = here("tables", "btm_annotation_genes.csv"))
mouse_engerixb_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_metadata.rds"))
mouse_engerixb_log2fc_filtered_sample_long = readRDS(here("tables", "mouse_engerixb_log2fc_filtered_sample_long.rds"))

#Annotate genes by BTM
samples_genes_btm = mouse_engerixb_log2fc_filtered_sample_long %>% 
  ungroup() %>% 
  select(symbol = human_symbol, participant_id, timepoint, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = samples_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process

dge_btm_mean_process_mouse = dge_btm_mean_process

dge_btm_mean_process_mouse %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_mouse_gsea_btm_mean.rds"))
```



##### Using GGPairs

Using this summarized BTM-level expression, generate pairwise
correlation plots with ggpairs, showing scatterplots and density plots
across modules, colored by timepoint. Add reference lines at zero, apply
custom color scales, and use a themed layout.

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

mouse_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = 3:7,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple"))

mouse_correlation_ggpairs_btms_gsea_alldays

mouse_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "hepatitisb_engerixb_mouse_gsea_samples_btm_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)

```

##### Using ggplot

Compute pairwise correlations of BTM module expression for five mouse
participants at a selected timepoint. Create unique participant pairs,
calculate Pearson correlations and R², and plot scatterplots with
regression lines, colored by module group, faceted by participant pairs,
and labeled with correlation values.

```{r fig.height=8, fig.width=10}
#Sample 5 participants
set.seed(31415)

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

sample_participants_5 = ggpairs_data_dge_btm_mean_process %>% 
  select(3:7) %>% 
  colnames()

#Inputs
timepoint_selected = 7


plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == timepoint_selected) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == timepoint_selected) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove pares idênticos (mesma variável)
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))


#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-1, 1) +
  xlim(-1, 1) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste0("Mouse, Hepatitis B, EngerixB Mouse, Day", timepoint_selected))
```

### GSE124533 (Human)

Download directly from GEO. The data is too large.

#### Prepare data

```{r}
#EngerixB: GSE124533 -----
filename = "GSE124533"
GSE124533 <- getGEO(filename = here("data_studies", "GSE124533_series_matrix.txt"), GSEMatrix = F)

#Metadata ------
GSE124533_metadata = GSE124533@phenoData %>% 
  pData() %>% 
  dplyr::select( 
    geo_sample = "geo_accession", 
    participant_id = "participant:ch1",
    sex = "gender:ch1",
    treatment = "treatment:ch1",
    ethnicity = "ethnicity:ch1",
    race = "race:ch1",
    age = "age:ch1",
    timepoint = "day:ch1") %>% 
  mutate(condition = treatment,
         organism = "Human",
         time_unit = "Day")
#Save
GSE124533_metadata %>% saveRDS(file = here("tables", "hepatitisb_engerixb_human_GSE124533_metadata.rds"))

#Import
GSE124533_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_human_GSE124533_metadata.rds"))


#Filter metadata
GSE124533_metadata_filtered = GSE124533_metadata %>%
  filter(treatment %in% c("ENGERIXB1"),
         timepoint %in% c("0", "1" , "3", "7"))

#Save
GSE124533_metadata_filtered %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_metadata.rds"))


#Counts matrix
GSE124533_exprs = GSE124533@assayData[["exprs"]] %>% as.data.frame()

GSE124533_exprs_engerixb = GSE124533_exprs %>% 
  select(c(GSE124533_metadata_filtered$geo_sample))
#Save
GSE124533_exprs_engerixb %>% saveRDS(file = here("tables", "GSE124533_exprs_engerixb.rds"))
```



Import
```{r}
#Import
human_exprs_engerixb = readRDS(file = here("tables", "GSE124533_exprs_engerixb.rds"))
human_engerixb_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_human_metadata.rds"))

########### Filter and annotate probes -----
probes_annotation <- read.delim(here("tables", "GPL21272-58809.txt"), skip = 19, header = TRUE, sep = "\t", quote = "") %>% 
  select(gene_id = ID, entrez_id = ENTREZ_GENE_ID, ensembl_id = ENSEMBL_ID, gene_symbol = GENE_SYMBOL) %>% 
  mutate(gene_id = as.character(gene_id))


#Get annotations for each probe 
### Filter probes with highest median expression
human_engerix_log2summary = human_exprs_engerixb %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  pivot_longer(cols = -probe,
               names_to = "sample",
               values_to = "expression") %>% 
  group_by(probe) %>% 
  summarize(mean = mean(expression,na.rm = TRUE),
            median = median(expression, na.rm = T)) %>%  
  inner_join(probes_annotation %>% 
              rename(probe = gene_id,
                     entrezgene_id = entrez_id,
                     human_symbol = gene_symbol), 
            by = "probe") %>% 
  #Filter ensembl id
  group_by(ensembl_id) %>% 
  slice_max(median, n = 1) %>%
  ungroup() %>% 
  select(-ensembl_id) %>% 
  distinct() %>% 
  
  #Filter entrez id
  group_by(entrezgene_id) %>% 
  slice_max(median, n = 1) %>%  #21.863 probes/entrez_ids
  ungroup() 

#Filter out genes with multiple hgnc_symbols
hgnc_symbol_dupes = human_engerix_log2summary %>% 
  #filter ensembl id with multiple hgnc_symbols
  get_dupes(human_symbol)

human_engerix_log2summary = human_engerix_log2summary %>% 
  filter(!human_symbol %in% c(hgnc_symbol_dupes$human_symbol %>% unique())) 

#Modify expression matrix

#Filtered probes
human_engerix_log2_filtered_probes = human_exprs_engerixb %>% 
  as.data.frame() %>% 
  rownames_to_column("probe") %>% 
  inner_join(human_engerix_log2summary, by = "probe") %>%   #214,396
  select(human_symbol,
         c(human_engerixb_metadata$geo_sample))
#14395
human_engerix_log2_filtered_probes %>% dim()

human_engerix_log2_hgnc_symbol = human_engerix_log2_filtered_probes 

human_engerix_log2_hgnc_symbol %>% 
  get_dupes(human_symbol)

#Save 
human_engerix_log2_hgnc_symbol %>% 
  saveRDS(here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol.rds"))
```




```{r}

#Import
human_engerix_log2_hgnc_symbol = readRDS(here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol.rds"))
human_metadata_filtered = readRDS(file = here("tables", "hepatitisb_engerixb_human_metadata.rds"))

#Get value from time 0 and calculate Log2FC
human_engerixb_log2fc_filtered_sample_long = human_engerix_log2_hgnc_symbol %>%
  ungroup() %>% 
  pivot_longer(cols = -(human_symbol),
               names_to = "geo_sample",
               values_to = "log2") %>% 
  inner_join(human_metadata_filtered %>% 
               select(geo_sample, participant_id, timepoint),
             by = "geo_sample") %>% 
  select(human_symbol, participant_id, timepoint, log2) %>% 
  distinct() %>% 
  select(participant_id, human_symbol, timepoint, value = log2) 

# #Get over duplicated symbols
# overduplicated = human_engerixb_log2fc_filtered_sample_long %>% 
#   select(human_symbol) %>% 
#   group_by(human_symbol) %>% 
#   dplyr::count() %>% 
#   arrange(-n) 

#Calculate log2fc
human_engerixb_log2fc_filtered_sample_long = human_engerixb_log2fc_filtered_sample_long %>%
  # filter(!human_symbol %in% c(overduplicated %>% 
  #                               ungroup() %>% 
  #                               filter(n > median(n)) %>% 
  #                               pull(human_symbol))) %>% 
  pivot_wider(names_from = timepoint, values_from = value, names_prefix = "day_") %>% 
  pivot_longer(cols = starts_with("day_"),
               names_to = "timepoint",
               names_prefix = "day_",
               values_to = "value") %>%
  group_by(participant_id, human_symbol) %>% 
  mutate(day0 = value[timepoint == "0"]) %>%
  ungroup() %>%
  mutate(log2fc = value - day0) %>%
  filter(timepoint != "0")


#Annotate sample and pivot longer
human_engerixb_log2fc_filtered_sample_long_annotated = human_engerixb_log2fc_filtered_sample_long %>% 
  select(-value, -day0) %>% 
  inner_join(GSE124533_metadata_filtered %>% 
               select(sample_id = geo_sample,
                      participant_id, timepoint), by = join_by("participant_id", "timepoint"))

#Save
human_engerixb_log2fc_filtered_sample_long_annotated %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol_long.rds"))

beep()

#Pivot wider
human_engerixb_log2fc_filtered_sample_wide = human_engerixb_log2fc_filtered_sample_long %>% 
  select(-c(value:day0)) %>% 
  pivot_wider(names_from = "participant_id",
              values_from = "log2fc")
  
```


#### Differential expression (Limma)


##### Create expression set
```{r}
#Import
human_engerix_log2_hgnc_symbol = readRDS(here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol.rds"))
human_metadata_filtered = readRDS(file = here("tables", "hepatitisb_engerixb_human_metadata.rds"))

human_metadata_filtered = human_metadata_filtered %>% 
  mutate(age = str_remove(age, "y") %>% as.numeric() %>% scale(),
         timepoint = as.factor(timepoint),
         sex = as.factor(sex),
         race = as.factor(race))

#Create Expression Set
eset_human <- ExpressionSet(
  assayData = human_engerix_log2_hgnc_symbol %>% 
    column_to_rownames("human_symbol") %>% 
    as.matrix(),
  phenoData = AnnotatedDataFrame(human_metadata_filtered))

#Save
eset_human %>% saveRDS(file = here("tables", "hepatitisb_engerixb_human_eset.rds"))
```



#####Limma
```{r}
#Import
human_engerix_log2_hgnc_symbol = readRDS(here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol.rds"))
human_metadata_filtered = readRDS(file = here("tables", "hepatitisb_engerixb_human_metadata.rds"))
eset_human = readRDS(file = here("tables", "hepatitisb_engerixb_human_eset.rds"))

#Create design matrix
design <- model.matrix(~ timepoint + sex + age + race + 0, eset_human) 

#Avoid blank spaces
colnames(design) <- make.names(colnames(design))

#Get intra-individual correlation
corfit <- duplicateCorrelation(eset_human, 
                               design = design, 
                               block = eset_human$participant_id)
corfit$consensus #0.22

# Fit
engerix_human_limma_fit <- lmFit(eset_human, 
                                 design, 
                                 block = eset_human$participant_id, 
                                 correlation = corfit$consensus)

# Contrasts
cont.matrix <- makeContrasts(
  timepoint1 - timepoint0,
  timepoint3 - timepoint0,
  timepoint7 - timepoint0,
  levels = design
)

fit2 <- contrasts.fit(engerix_human_limma_fit, cont.matrix)

# eBayes
fit2 <- eBayes(fit2)

# Extract
results_list <- lapply(1:ncol(fit2), function(i) {
  topTable(fit2, coef = i, adjust.method = "fdr", number = Inf, confint = TRUE) %>%
    mutate(comparison = colnames(fit2)[i]) %>%
    rownames_to_column("symbol")
})


# Combine all results
all_contrasts_degs <- bind_rows(results_list)

# Get all DEGs
all_degs_human = all_contrasts_degs %>% 
  #Clean table
  select(comparison, everything()) %>% 
  mutate(comparison = str_replace(comparison, "\\.\\.\\.", "_") %>% 
           str_replace(., "timepoint", "Time") %>% 
           str_replace(., "treatment", "")) %>% 
  separate(comparison, sep = " - ", into = c("comparisonA", "comparisonRef")) %>% 
  mutate(comparisonRef = str_replace(comparisonRef, "treatment", "Time0")) %>% 
  mutate(condition = case_when(comparisonA == "Time1"	~ "EngerixB_Day1",
                               comparisonA == "Time3"	~ "EngerixB_Day3",
                               comparisonA == "Time7"	~ "EngerixB_Day7",
                               T~comparisonA),
         organism = "Human") %>% 
  select(condition, everything()) %>% 
  clean_names() %>% 
  rename(human_symbol = symbol,
         mean_l2fc = log_fc,
         ci_upper = ci_r,
         ci_lower = ci_l)

all_degs_human %>% 
  group_by(condition) %>%
  filter(adj_p_val <= 0.05) %>% 
  summarize(n_degs = n())
	
#Save
all_degs_human %>% saveRDS(file = here("tables", "hepatitisb_engerixb_human_dge_limma_degs.rds"))
```




#### Correlation plots, pairwise

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Sample 5 participants
set.seed(3141)
sample_participants_5 = GSE124533_metadata_filtered %>% 
  select(participant_id) %>% 
  distinct() %>% 
  slice_sample(n = 5) %>% 
  pull()

#Plot correlation between animals ------
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.8, ...)  
}

#All timepoints
mouse_correlation_ggpairs_alldays = ggpairs(human_engerixb_log2fc_filtered_sample_wide %>% 
                                              select(human_symbol, timepoint, c(sample_participants_5)) %>% 
                                              mutate(timepoint = as.character(timepoint),
                                                     timepoint = fct_reorder(timepoint, as.numeric(timepoint))),
                                    columns = 3:7,
                                    mapping = ggplot2::aes(color = timepoint,
                                                           fill = timepoint),
                                    title = "All days, HepB/EngerixB, GSE124533, Human", 
                                    diag = list(continuous = my_dens),
                                    lower = list(continuous = my_point)
                                    ) +
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "3" = "#219ebc",
                                "7" = "#669bbc",
                                "12" = "#4361ee")) 

mouse_correlation_ggpairs_alldays
# Save
mouse_correlation_ggpairs_alldays %>%
  ggsave(filename = here("Figures", "hepatitisb_engerixb_human_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)
beep()
```

#### PCA

```{r fig.height=4, fig.width=5}

GSE124533_counts_log2_summary_annotated_probes =
  readRDS(file = here("tables", "GSE124533_counts_log2_summary_annotated_probes.rds"))

#Input
samples_metadata_pca_input = GSE124533_metadata_filtered
samples_log2fc_values_pca_input = human_engerixb_log2fc_filtered_sample_long_annotated


#Import metadata
annotation_samples_pca = samples_metadata_pca_input %>% 
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(participant_id, timepoint, sample) %>% 
  distinct()


# PCA with raw values
clean_matrix_raw = samples_log2fc_values_pca_input %>% 
  ungroup() %>% 
  select(human_symbol, participant_id, timepoint, log2fc) %>%
  mutate(sample = paste0(participant_id, "_Day_", timepoint)) %>% 
  select(sample, human_symbol, log2fc) %>% 
  filter(!human_symbol == "") %>% 
  distinct() %>%
  pivot_wider(names_from = "sample",
              values_from = "log2fc") %>% 
  column_to_rownames("human_symbol") %>% 
  as.matrix()

clean_matrix_raw[!is.finite(clean_matrix_raw)] <- 0

pca_raw <- prcomp(t(clean_matrix_raw), scale. = TRUE)
pca_raw_df <- as_tibble(pca_raw$x[, 1:2]) %>%
  mutate(sample = rownames(pca_raw$x)) %>% 
  left_join(annotation_samples_pca, by = "sample")

#Plot
pca_raw <- pca_raw_df %>%
  mutate(timepoint = as.character(timepoint)) %>% 
  ggplot() +
  aes(x = PC1,
      y = PC2,
      color = timepoint,
      fill = timepoint,
      shape = organism,
      label = participant_id) +
  geom_point(size = 3, 
             alpha = 1, 
             pch = 24, 
             color = "black") +
  theme_vaxgo() +
  guides(fill = "none") +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  labs(title = "Human, HepB, EngerixB, GSE124533",
       shape = "Organism",
       color = "Timepoint") +
  theme(axis.text.x = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill=NA, linewidth=1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank())
pca_raw 

beep()

```

#### BTMs

```{r}

#Import files
btm_annotation_genes =
  read_csv(file = here("tables", "btm_annotation_genes.csv"))
human_engerixb_log2fc_filtered_sample_long_annotated = 
  readRDS(file = here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol_long.rds"))
GSE124533_metadata_filtered =
  readRDS(file = here("tables", "hepatitisb_engerixb_human_exprs_hgnc_symbol.rds"))


#Input
samples_metadata_pca_input = GSE124533_metadata_filtered
samples_log2fc_values_pca_input = human_engerixb_log2fc_filtered_sample_long_annotated

#Annotate genes by BTM
samples_genes_btm = samples_log2fc_values_pca_input %>% 
  ungroup() %>% 
  select(symbol = human_symbol, participant_id, timepoint, log2fc) %>% 
  inner_join(btm_annotation_genes, by = "symbol")


dge_btm_mean_process = samples_genes_btm %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

dge_btm_mean_process_human = dge_btm_mean_process

dge_btm_mean_process_human %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_gsea_btm_mean.rds"))
```

##### GGPairs

```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
#Change params in geoms
#Density
my_dens <- function(data, mapping, ...) {
  ggplot(data = data, mapping=mapping) +
    geom_density(alpha = 0.2, color = "black")
}
#Points
my_point <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5)  
}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#All days

human_correlation_ggpairs_btms_gsea_alldays = ggpairs_data_dge_btm_mean_process %>% 
  ggpairs(columns = sample_participants_5,
          mapping = ggplot2::aes(color = timepoint,
                                 fill = timepoint),
          title = paste("All days, BTM"),
          diag = list(continuous = my_dens),
          lower = list(continuous = my_point)) +
  
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2)+
  theme_vaxgo() +
  scale_color_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple")) +
  scale_fill_manual(values = c("1" = "#caf0f8",
                                "2" = "#caf0f8",
                                "3" = "#219ebc",
                                "4" = "#219ebc",
                                "6" = "#669bbc",
                                "7" = "#669bbc",
                                "12" = "purple"))

human_correlation_ggpairs_btms_gsea_alldays

human_correlation_ggpairs_btms_gsea_alldays %>%
  ggsave(filename = here("Figures", "hepatitisb_engerixb_human_gsea_btm_correlation_ggpairs_alldays.png"),
         height = 10,
         width = 10)

```

##### Ggplot manual

```{r fig.height=8, fig.width=10}

#Create table
ggpairs_data_dge_btm_mean_process = dge_btm_mean_process %>% 
  mutate(timepoint = as.character(timepoint),
         timepoint = fct_reorder(timepoint, timepoint %>% as.numeric)) %>% 
  ungroup()

#Input
timepoint_selected = 7

#Plot table
plot_data <-
  #Pivot longer, 1st participatn
  ggpairs_data_dge_btm_mean_process %>%
  filter(timepoint == timepoint_selected) %>% 
  select(timepoint, group, process, sample_participants_5) %>%
  pivot_longer(cols = !c("group", "process", "timepoint"), 
               names_to = "participant_1", 
               values_to = "participant_1_value") %>%
  
  #Add a column to the 2nd participant
  inner_join(
    ggpairs_data_dge_btm_mean_process %>%
      filter(timepoint == timepoint_selected) %>% 
      select(timepoint, group, process, sample_participants_5) %>%
      pivot_longer(cols = !c("group", "process", "timepoint"), 
                   names_to = "participant_2",
                   values_to = "participant_2_value"),
    by = c("process", "group")) %>%
  #Remove identical pairs
  filter(participant_1 != participant_2) %>%
  
  #Remove dupes (ex: A-B e B-A → A-B)
  mutate(pair_min = pmin(participant_1, participant_2),
         pair_max = pmax(participant_1, participant_2)) %>%
  distinct(process, pair_min, pair_max, .keep_all = TRUE) %>% 
  
  # Arrange participants to create a pairwise matrix
  mutate(
    participant_1 = factor(participant_1, levels = sample_participants_5),
    participant_2 = factor(participant_2, levels = rev(sample_participants_5)))


#Calculate correlation
cor_labels <- plot_data %>%
  group_by(participant_1, participant_2) %>%
  summarise(cor_value = cor(participant_1_value, participant_2_value, use = "pairwise.complete.obs", method = "pearson"),
            .groups = "drop") %>%
  mutate(r_squared = cor_value ^ 2,
         cor_label = paste0("r = ", round(cor_value, 2), ", R² = ", round(r_squared, 2)))

#Plot 
plot_data %>% 
ggplot(aes(x = participant_1_value, y = participant_2_value, color = group)) +
  geom_point(alpha = 1) +
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label), 
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  facet_grid(participant_2 ~ participant_1, scales = "free", drop = T) +
  theme_vaxgo() +
  ylim(-1, 1) +
  xlim(-1, 1) +
  theme(strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0, hjust = 0)) +
  scale_color_manual(values = immune_colors) +
  labs(x = "Log2FC", 
       y = "Log2FC", 
       color = "",
       title = paste("Human, Hepatitis B, EngerixB Human, Day", timepoint_selected))
```

# 2. Compare Human vs Mouse

### Combine datasets
```{r}
#Unite Mouse and Human datasets. Samples.
human_mouse_log2fc_samples = 
#Mouse
mouse_engerixb_log2fc_filtered_sample_long %>%
  ungroup() %>% 
  select(human_symbol, log2fc, participant_id,
         timepoint, value, day0) %>% 
  mutate(organism = "Mouse") %>% 
  
  #Human
  bind_rows(human_engerixb_log2fc_filtered_sample_long %>%
  ungroup() %>% 
  select(human_symbol, log2fc, participant_id,
         timepoint,  value, day0) %>% 
  mutate(organism = "Human"))

#Unite metadata
human_mouse_engerix_metadata = 
  #Human
  GSE124533_metadata_filtered %>% 
  select(geo_sample, participant_id, sex, organism, timepoint) %>% 
  mutate(timepoint = as.numeric(timepoint),
         sex = str_to_title(sex)) %>% 
  #Mouse
  bind_rows(
    mouse_engerixb_metadata %>%
      select(geo_sample, participant_id, sex, timepoint) %>%
      mutate(organism = "Mouse"))

human_mouse_log2fc_samples %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_exprs_samples_log2fc.rds"))

human_mouse_engerix_metadata %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_metadata.rds"))


all_degs_human = readRDS(file = here("tables", "hepatitisb_engerixb_human_dge_limma_degs.rds")) %>% 
  mutate(timepoint = str_remove(comparison_a, "Time"))
all_degs_mouse = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_degs_mouse.rds")) %>% 
  mutate(timepoint = str_remove(condition, "Engerix_") %>% str_remove("_PBMC"))
all_degs_human_mouse = all_degs_mouse %>% 
  bind_rows(all_degs_human) %>% 
  select(condition, timepoint, human_symbol, mean_l2fc:organism)
all_degs_human_mouse %>% saveRDS(here("tables", "hepatitisb_engerixb_human_mouse_dge_limma_degs.rds"))
```




### Get overlap

Input

```{r}
# Import datasets
dge_long_genes = readRDS(here("tables", "hepatitisb_engerixb_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(timepoint = if_else(organism == "Mouse", (as.numeric(timepoint)/24) %>% round(0) %>% as.character(), timepoint) ,
         condition = paste0(organism, "_", timepoint)) %>% 
  distinct()
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

# DEGs table
degs_all_updown = dge_long_genes %>% 
  dplyr::filter(organism %in% c("Human", "Mouse")) %>% 
  dplyr::filter(adj_p_val <= 0.05) %>% 
  dplyr::mutate(condition = paste0(organism, "_", timepoint)) %>% 
  dplyr::select(process = condition, genes = human_symbol) 

# Metadata
annotation_metadata = dge_long_genes %>% 
  #Arrange columns in the desired order to be displayed on the heatmap
  dplyr::select(condition, 
                organism, 
                # type, 
                # vaccine, 
                timepoint) %>% 
  distinct()

annotation_metadata$condition %>% unique()
#Inputs
filename = "hepatitisb_engerixb"
```

Perform calculations

```{r}
conflicted::conflicts_prefer(base::intersect)
conflicted::conflicts_prefer(base::setdiff)

# Perform calculations
unique_cond <- unique(degs_all_updown$process) # List sets
shared_genes_df <- data.frame() # Store data

for (i in 1:(length(unique_cond) - 1)) {
  for (j in (i + 1):length(unique_cond)) {
    results_temp <- overlap_genes(unique_cond[i], unique_cond[j], degs_all_updown)
    shared_genes_df <- bind_rows(shared_genes_df, results_temp)
  }
}

# Fisher's exact test
fisher_exact_test <- function(shared, notshared1, notshared2) {
  cont_table <- matrix(c(shared, notshared1, notshared2, 0), nrow = 2)
  results_fisher <- stats::fisher.test(cont_table)
  return(results_fisher$p.value)
}
shared_genes_df = shared_genes_df %>%
  mutate(pvalue = pmap_dbl(list(Shared, NotShared_cond1, NotShared_cond2), fisher_exact_test),
         p_adj = p.adjust(pvalue))


#Mirror table
shared_genes_df2 = shared_genes_df %>% 
  dplyr::rename(Cond1_temp = Cond1, Cond2_temp = Cond2) %>%
  dplyr::rename(Cond1 = Cond2_temp, Cond2 = Cond1_temp)

shared_genes_df_mirror = shared_genes_df %>% 
  dplyr::bind_rows(shared_genes_df2) %>% 
  dplyr::mutate(Percentage_Shared_Cond1 = round(Percentage_Shared_Cond1, 2),
         Percentage_Shared_Cond2 = round(Percentage_Shared_Cond2, 2))
  
#Jaccard distance
jaccard.matrix <- degs_all_updown %>%
  dplyr::mutate(present = 1) %>% 
  dplyr::distinct() %>% 
  pivot_wider(names_from = genes, values_from = present, values_fill = 0) %>%
  column_to_rownames(var = "process") %>%
  vegan::vegdist(binary = TRUE, method = "jaccard", diag = TRUE, upper = TRUE) %>%
  as.matrix() %>%
  {1 - .}
shared_genes_df_mirror = jaccard.matrix %>% 
  as.data.frame() %>% 
  rownames_to_column("Cond1") %>% 
  pivot_longer(cols = -"Cond1",
               names_to = "Cond2",
               values_to = "jaccard_distance") %>% 
  inner_join(shared_genes_df_mirror, by = join_by(Cond1, Cond2)) %>% 
  mutate(jaccard_distance = round(jaccard_distance, 3))

#Save
write_csv(shared_genes_df_mirror, 
          file = here("Tables", paste0(filename, "_sharedgenes_Fisher_Jaccard.csv")))
shared_genes_df_mirror
```

### Heatmap

#### Shared genes

```{r}
shared_genes_df_mirror = read_csv(file = here("Tables", "hepatitisb_engerixb_sharedgenes_Fisher_Jaccard.csv"))

matrix_data = shared_genes_df_mirror %>% 
  dcast(Cond1 ~ Cond2, 
        value.var = "Shared", 
        fun.aggregate = mean) %>% 
  column_to_rownames("Cond1") %>%
  mutate_if(is.character, as.numeric) %>% 
  as.matrix() 

# Create a matrix of p-values
pvalue_matrix <- shared_genes_df_mirror %>%
  dcast(Cond1 ~ Cond2,
        value.var = "p_adj",
        fun.aggregate = mean) %>%
  column_to_rownames("Cond1") %>%
  as.matrix()

dim(matrix_data)

annotation_metadata_complex = matrix_data %>% 
  as.data.frame() %>%
  rownames_to_column("condition") %>% 
  dplyr::select(condition) %>% 
  inner_join(annotation_metadata , by = "condition") %>%
  distinct() 

# merge
annotation_metadata_heatmap = matrix_data %>% 
  colnames() %>% 
  data.frame() %>% 
  dplyr::filter(. != "process") %>% 
  dplyr::rename(condition = '.') %>% 
  dplyr::inner_join(annotation_metadata_complex, by = "condition") %>% 
  dplyr::arrange(condition) %>% 
  column_to_rownames(var= "condition")


# Reorder
matrix_data_ordered = matrix_data %>% 
  as.data.frame() %>% 
  dplyr::select(c(annotation_metadata_complex %>% pull(condition))) %>% 
  rownames_to_column("rownames") %>% 
  dplyr::filter(rownames %in% c(annotation_metadata_complex %>% pull(condition))) %>% 
  column_to_rownames(var= "rownames") %>% 
  as.matrix()

#Check
dim(matrix_data_ordered) 
dim(annotation_metadata_complex)

col_fun <- colorRamp2(c(0, max(matrix_data_ordered, na.rm = T)), c("white", "#00A087FF"))


######## HEATMAP ANNOTATION
annotation_metadata_heatmap = annotation_metadata_heatmap %>% 
  mutate(timepoint = as_factor(timepoint),
         timepoint = fct_reorder(timepoint, as.numeric(timepoint)))

ha = HeatmapAnnotation(df = annotation_metadata_heatmap, 
                       col = colors_immune, 
                       annotation_name_side = "left",
                       show_legend = F)

row_ha = rowAnnotation(df = annotation_metadata_heatmap, 
                       col = colors_immune, 
                       show_annotation_name = F)



mat = matrix_data_ordered
width_image = 15
height_image = 15
width = unit(5, "cm")
height = unit(5, "cm")
column_names_gp = gpar(fontsize = 10)
row_names_gp = gpar(fontsize = 10)
gp = gpar(fontsize = 10)
rect_gp = gpar(col = "gray1", lwd = 0)



fileimageheatmap_grouped = paste0(filename, sep ="_", "Shared_Heatmap.png")

png(
  file = here("Figures", fileimageheatmap_grouped),
  width = width_image,
  height = height_image,
  units = "cm",
  res = 900
)

heatmap_plot_all = Heatmap(mat,
                        top_annotation = ha,
                        left_annotation = row_ha,
                        show_row_names = TRUE,
                        show_heatmap_legend = TRUE,
                        heatmap_legend_param = list(direction = "vertical"),
                        name = "Shared", #Legend
                        col = col_fun, #color
                         cell_fun = function(j, i, x, y, width, height, fill) {
  if (!is.na(mat[i, j]) && mat[i, j] > 0) {
    grid.text(sprintf("%.f", mat[i, j]), x, y, gp = gpar(fontsize = 8, color = "white"))
  }
                           i# Add significance stars based on p-value thresholds
  if (!is.na(pvalue_matrix[i, j])) {
    pval <- pvalue_matrix[i, j]
    sig <- ""
    if (pval < 0.01) {
      sig <- "***"
    } else if (pval < 0.05) {
      sig <- "**"
    } else if (pval < 0.10) {
      sig <- "*"
    }
    
    # Draw the symbol slightly above the cell center
    if (sig != "") {
      grid.text(sig, x, y + unit(2, "mm"), gp = gpar(fontsize = 6, fontface = "plain", color = "white"))
    }
  }
},
                        column_title = fileimageheatmap_grouped,
                        column_title_gp = gpar(fontsize = 15, fontface = "bold"),
                        column_names_rot = 45,
row_names_rot = 0,
                        column_names_gp = column_names_gp,
                        cluster_rows = TRUE,
                        rect_gp = rect_gp,
                        row_names_gp = row_names_gp,
                        # row_split = 2, #Split row clusters
                        row_gap = unit(2, "mm"),
                        cluster_columns = TRUE,
                        na_col = "black",
                        column_split = NULL, #Split group clusters
                        #column_km = 3,
                        column_gap = unit(8, "mm"),
                        show_column_dend = F,
                        show_row_dend = F,
                        row_dend_side = "left",
                        column_dend_side = "top",
                        clustering_distance_rows = "euclidean",
                        border = TRUE,
                        column_dend_reorder = TRUE,
                        width = width, 
                        height = height
)

draw(heatmap_plot_all,  heatmap_legend_side = "left", annotation_legend_side = "left")

dev.off()
```


### Stacked barplot of overlapped genes
```{r fig.width=10, fig.height=4}
#Import datasets
all_degs_human_mouse = readRDS(here("tables", "hepatitisb_engerixb_human_mouse_dge_limma_degs.rds"))  
ImmuneGO_Mouse_Manual_genes <- read_excel(here("VaxGO", "ImmuneGO_Mouse.xlsx")) %>% 
  separate_rows(genes)


#Compare every mouse timepoint to a human reference
timepoint_reference = "7"
human_data_reference <- all_degs_human_mouse %>% filter(organism == "Human")
new_timepoints <- c(all_degs_human_mouse %>% filter(organism != "Human") %>% pull(timepoint) %>% unique() %>% as.character())
duplicated_human_data <- map_dfr(new_timepoints, ~ {
  df_copy <- human_data_reference
  df_copy <- df_copy %>%
    mutate(timepoint = .x)

  return(df_copy)
})

dge_long_genes_matched = all_degs_human_mouse %>% 
  ungroup() %>% 
  filter(organism != "Human") %>% 
  bind_rows(duplicated_human_data)  %>% 
    mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0(timepoint, "h"),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) 


#Get total genes by each condition
degs_mouse = dge_long_genes_matched %>% 
  filter(adj_p_val <= 0.05,
         organism == "Mouse")
degs_human = dge_long_genes_matched %>% 
  filter(adj_p_val <= 0.05,
         organism == "Human")

degs_mouse_summarize = dge_long_genes_matched %>% 
  filter(adj_p_val <= 0.05) %>% 
  group_by(organism, timepoint) %>% 
  summarize(n_genes = n()) 

#Print
degs_mouse_summarize

#Get overlap -----
#Get symbols
degs_mouse_genes = degs_mouse %>% 
  select(human_symbol, timepoint) %>% 
  distinct()

degs_human_genes = degs_human %>% 
  select(human_symbol, timepoint) %>% 
  distinct()

#Inner join, Mouse and Human ------

#Mouse
degs_human_and_mouse_genes = degs_human_genes %>% 
  inner_join(degs_mouse_genes, 
             by = join_by("human_symbol", "timepoint")) %>% 
  mutate(condition = "Mouse AND Human",
         comparison = "Mouse vs Human",
         organism = "Mouse",
         group = "Common")

#Duplicate rows for human
degs_human_and_mouse_genes_human = degs_human_and_mouse_genes %>% 
  mutate(organism = "Human")


#Anti join, Human vs Mouse -------
#Human
#Human vs Mouse
degs_human_Not_mouse_genes = degs_human_genes %>% 
  anti_join(degs_mouse_genes, 
             by = join_by("human_symbol", "timepoint")) %>% 
  mutate(condition = "Human NOT Mouse", 
         comparison = "Mouse vs Human",
         organism = "Human",
         group = "Not common")

#Models
#Mouse vs Human
degs_mouse_Not_human_genes = degs_mouse_genes %>% 
  anti_join(degs_human_genes, 
             by = join_by("human_symbol", "timepoint")) %>% 
  mutate(condition = "Mouse NOT Human",
         organism = "Mouse",
         comparison = "Mouse vs Human",
         group = "Not common")


#Merge data ------
degs_human_mouse_overlap_genes = 
  bind_rows(degs_human_and_mouse_genes,
            degs_human_and_mouse_genes_human,
            degs_human_Not_mouse_genes,
            degs_mouse_Not_human_genes 
            )

#Classify by same and exclusive DEGs -----
degs_human_mouse_overlap_genes_summary = degs_human_mouse_overlap_genes %>% 
  group_by(condition, timepoint) %>% 
  summarize(n = n(),
            organism,
            comparison,
            group) %>% 
  distinct()

#Print
degs_human_mouse_overlap_genes_summary
```


### Plot overlap (bars)

```{r fig.height=5, fig.width=4}
#Plot overlap
degs_human_mouse_overlap_common_plot = degs_human_mouse_overlap_genes_summary %>%
  ggplot() +
  aes(x = organism,
      y = n,
      fill = fct_rev(group)) + 
  geom_col(alpha = 1) +
  facet_nested(rows = vars(fct_rev(comparison)), 
               cols = vars(timepoint), 
               scales = "free", 
               nest_line = element_line(linetype = 1, color = "black", size = 2),
               independent = "x") +
  scale_y_continuous(n.breaks = 5,
                     expand = c(0.01, 0.01)) +
  scale_fill_manual(values = c("Common" = "#4DBBD5FF",
                               "Not common" = "gray90"))    +
  scale_color_manual(values = c("Common" = "#4DBBD5FF",
                               "Not common" = "gray90"))    +
  theme_vaxgo() +
  theme(
    strip.text = element_text(size = 10),
    strip.text.y = element_text(hjust = 0.5, size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0, "lines"), 
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "top",
    legend.position.inside = c(0.5, 0.95),
    legend.direction = "horizontal", 
    legend.key.width = unit(0.4, 'cm'),
    legend.key.height = unit(0.4, 'cm'),
    legend.title = element_blank(),
    legend.text = element_text(size = 7, vjust = 0.5, hjust = -0.2)
  ) +
  labs(
    x = "",
    y = "Total",
    color = ""
    ) 
#Print
degs_human_mouse_overlap_common_plot

#Save
degs_human_mouse_overlap_common_plot %>%
  ggsave(filename = here("Figures", "hepatitisb_engerixb_human_mouse_dge_degs_overlap_barplot.png"),
         width = 4,
         height = 5)
```




###Plot correlation
```{r}
#Import
human_mouse_log2fc_samples =
  readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_exprs_samples_log2fc.rds"))

human_mouse_engerix_metadata =
  readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_metadata.rds"))

all_degs_human = readRDS(file = here("tables", "hepatitisb_engerixb_human_dge_limma_degs.rds")) %>% 
  mutate(timepoint = str_remove(comparison_a, "Time"))
all_degs_mouse = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_degs_mouse.rds")) %>% 
  mutate(timepoint = str_remove(condition, "Engerix_") %>% str_remove("_PBMC") %>% as.numeric(),
         timepoint = timepoint /24,
         timepoint = timepoint %>% as.character())



#Pivot wider
human_mouse_engerixb_log2fc_avg_wide = all_degs_human %>% 
  bind_rows(all_degs_mouse) %>% 
  select(organism, human_symbol, timepoint, mean_log2fc = mean_l2fc) %>% 
  pivot_wider(names_from = "organism",
              values_from = "mean_log2fc") %>% 
  #Add padj columns
  inner_join(all_degs_human %>%
               bind_rows(all_degs_mouse) %>% 
               select(organism, human_symbol, timepoint, adj_p_val) %>%
               pivot_wider(names_from = "organism",
                           values_from = "adj_p_val") %>%
               rename(padj_human = Human, padj_mouse = Mouse)) %>% 
  #Drop genes that are absent in one of the organisms 
  drop_na()

# Calculate the correlation
cor_data <- human_mouse_engerixb_log2fc_avg_wide %>%
  group_by(timepoint) %>% 
  filter(padj_human <= 0.05 | padj_human <= 0.05) %>%
  summarize(
    cor_value = cor(Mouse, Human, method = "spearman", use = "complete.obs"),
    label_pos_x = max(Mouse, na.rm = TRUE),
    label_pos_y = max(Human, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("r = ", round(cor_value, 2)))
```

```{r fig.height=3, fig.width=7}
ylim_range = c(-2.5, 2.5)
xlim_range = c(-2.5, 2.5)
human_mouse_log2fc_log2fc_avg_wide_scatterplot = human_mouse_engerixb_log2fc_avg_wide %>% 
  
  #Filter DEGs only
  filter(padj_human <= 0.05 | padj_mouse <= 0.05) %>%
  
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human) +
  
    #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #Facet
  facet_wrap(~paste("Day", timepoint), nrow = 1) +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_engerixb_log2fc_avg_wide %>%
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_engerixb_log2fc_avg_wide %>%
               filter(human_symbol %in% c(btm_annotation_genes$symbol)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  stat_smooth(method = "lm", geom = "line", color = "#4DBBD5FF", alpha = 0.5, se = F, size = 1,
              data = human_mouse_engerixb_log2fc_avg_wide %>%
               filter(human_symbol %in% c(btm_annotation_genes$symbol)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add correlation metrics (Pearson)
  geom_label(
    data = cor_data,
    aes(x = label_pos_x, y = label_pos_y, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 1,
    vjust = 1,
    size = 3
  ) +
  
  #Adjust the axes range
  ylim(ylim_range) +
  xlim(xlim_range) +

  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "none",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title= "Correlation plot, EngerixB,\nHuman vs Mouse")

human_mouse_log2fc_log2fc_avg_wide_scatterplot

human_mouse_log2fc_log2fc_avg_wide_scatterplot %>% 
  ggsave(filename = here("Figures", "hepatitisb_engerixb_human_mouse_dge_degs_scatterplot.png"),
         width = 7, height = 3)
```



Single timepoint
```{r fig.height=3, fig.width=3}
human_mouse_log2fc_degs_scatterplot_singletimepoint = human_mouse_engerixb_log2fc_avg_wide %>% 
  filter(timepoint == 7) %>% 
  
  #Filter DEGs only
  filter(padj_human <= 0.05 | padj_mouse <= 0.05) %>%
  
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human) +
  
    #Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #All DEGs
  geom_point(color = "gray90") +
  
  #Facet
  facet_wrap(~paste("Day", timepoint), ncol = 1) +
  
  #DEGs in both organisms
  geom_point(color = "#4DBBD5FF",
             fill = "#4DBBD5FF",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_engerixb_log2fc_avg_wide %>%
               filter(timepoint == 7) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Immune DEGs
  geom_point(color = "#4361ee",
             fill = "#4361ee",
             alpha = 1.5,
             size = 1,
             pch = 21,
             data = human_mouse_engerixb_log2fc_avg_wide %>%
               filter(timepoint == 7) %>% 
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  stat_smooth(method = "lm", geom = "line", color = "#4DBBD5FF", alpha = 0.5, se = F, size = 1,
              data = human_mouse_engerixb_log2fc_avg_wide %>%
                filter(timepoint == 7) %>% 
               filter(human_symbol %in% c(ImmuneGO_BTM_grouped_genes$genes)) %>% 
               #Filter DEGs in both organisms, same sign
               filter(padj_human <= 0.05 & padj_mouse <= 0.05 & sign(Human) == sign(Mouse))) +
  
  #Add correlation metrics (Pearson)
  geom_label(
    data = cor_data %>% 
      filter(timepoint == 7),
    aes(x = label_pos_x, y = label_pos_y, label = label),
    inherit.aes = FALSE,
    fill = "white",
    hjust = 1,
    vjust = 1,
    size = 3
  ) +
  
  #Adjust the axes range
  ylim(ylim_range) +
  xlim(xlim_range) +

  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "none",
        legend.direction = "horizontal") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       title= "Correlation plot, EngerixB,\nHuman vs Mouse")


human_mouse_log2fc_degs_scatterplot_singletimepoint %>% 
  ggsave(filename = here("Figures", "hepatitisb_engerixb_human_mouse_dge_degs_correlation_scatterplot_singletimepoint.png"),
         width = 3, height = 3)

human_mouse_log2fc_degs_scatterplot_singletimepoint

```

#### Plot correlation between genes (log2fc for every sample)

```{r}
#Import
human_mouse_log2fc_samples =
  readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_exprs_samples_log2fc.rds"))

human_mouse_engerix_metadata =
  readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_metadata.rds"))

all_degs_human = readRDS(file = here("tables", "hepatitisb_engerixb_human_dge_limma_degs.rds")) %>% 
  mutate(timepoint = str_remove(comparison_a, "Time"))
all_degs_mouse = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_degs_mouse.rds")) %>% 
  mutate(timepoint = str_remove(condition, "Engerix_") %>% str_remove("_PBMC") %>% as.numeric(),
         timepoint = timepoint /24,
         timepoint = timepoint %>% as.character())

human_mouse_engerix_log2fc_samples = human_mouse_log2fc_samples %>% 
  mutate(timepoint = as.character(timepoint)) %>% 
  distinct()


human_mouse_log2fc_samples = human_mouse_engerix_log2fc_samples

human_mouse_engerix_log2fc_samples$organism %>% unique()
  
```



```{r}
# #Input
# filename = "hepatitisb_engerixb_human_mouse"
# gene_set_name = "AllGenes"
# human_mouse_log2fc_samples = human_mouse_engerix_log2fc_samples

#Input
# filename = "hepatitisb_engerixb_human_mouse"
# gene_set_name = "Immune"
# human_mouse_log2fc_samples = human_mouse_engerix_log2fc_samples  %>%
#   filter(human_symbol %in% c(btm_annotation_genes$symbol %>% unique()))


# #Input
# filename = "hepatitisb_engerixb_human_mouse"
# gene_set_name = "DEGs"
# human_degs = all_degs_human_mouse %>%
#   filter(organism == "Human", adj_p_val <= 0.05) %>%
#   select(human_symbol)
# human_mouse_log2fc_samples = human_mouse_engerix_log2fc_samples  %>%
#   inner_join(human_degs, by = join_by("human_symbol"))

# #Input
filename = "hepatitisb_engerixb_human_mouse"
gene_set_name = "DEGs_Immune"
human_mouse_log2fc_samples = human_mouse_engerix_log2fc_samples  %>%
  inner_join(human_degs, by = join_by("human_symbol")) %>%
  filter(human_symbol %in% c(btm_annotation_genes$symbol %>% unique()))
```



```{r fig.height=3, fig.width=3}
# --- 1) Mouse vs Human, matched timepoints ---
#Get timepoints
timepoints_human <- unique(human_mouse_log2fc_samples %>% filter(organism == "Human") %>% pull(timepoint))
timepoints_mouse <- unique(human_mouse_log2fc_samples %>% filter(organism == "Mouse") %>% pull(timepoint))

#Matched timepoints
timepoint_combinations <- tibble(timepoint_human = timepoints_human,
                                      timepoint_mouse = timepoints_mouse)

# Loop for matched timepoints
cor_human_mouse_timepoint <- timepoint_combinations %>%
  pmap_dfr(function(timepoint_human, timepoint_mouse) {
    
    # pegar os participantes do timepoint específico
    human_df <- human_mouse_log2fc_samples %>%
      filter(organism == "Human", timepoint == timepoint_human) %>%
      select(participant_id, human_symbol, log2fc)
    
    mouse_df <- human_mouse_log2fc_samples %>%
      filter(organism == "Mouse", timepoint == timepoint_mouse) %>%
      select(participant_id, human_symbol, log2fc)
    
    # manter apenas genes em comum
    common_genes <- intersect(human_df$human_symbol, mouse_df$human_symbol)
    human_df <- human_df %>% filter(human_symbol %in% common_genes)
    mouse_df <- mouse_df %>% filter(human_symbol %in% common_genes)
    
    # pivot para wide
    human_wide <- human_df %>% pivot_wider(names_from = participant_id, values_from = log2fc)
    mouse_wide <- mouse_df %>% pivot_wider(names_from = participant_id, values_from = log2fc)
    
    # IDs de participantes
    human_ids <- setdiff(colnames(human_wide), "human_symbol")
    mouse_ids <- setdiff(colnames(mouse_wide), "human_symbol")
    
    # calcular correlação para todos pares humano x mouse
    expand.grid(human = human_ids, mouse = mouse_ids, stringsAsFactors = FALSE) %>%
      mutate(
        cor_test = map2(human, mouse, ~{
          x <- human_wide[[.x]]
          y <- mouse_wide[[.y]]
          # remover NAs correspondentes
          keep <- complete.cases(x, y)
          cor.test(x[keep], y[keep], method = "pearson")
        }),
        r       = map_dbl(cor_test, ~.x$estimate),
        p_value = map_dbl(cor_test, ~.x$p.value),
        ci_low  = map_dbl(cor_test, ~.x$conf.int[1]),
        ci_high = map_dbl(cor_test, ~.x$conf.int[2])
      ) %>%
      select(-cor_test) %>%
      mutate(timepoint_human = timepoint_human,
             timepoint_mouse = timepoint_mouse)
  }) %>% 
  group_by(timepoint_human, timepoint_mouse) %>%
  summarize(
    comparison = "Human vs. Mouse",
    mean = mean(r, na.rm = TRUE),
    sd   = sd(r, na.rm = TRUE),
    n      = sum(!is.na(r)),
    se     = sd / sqrt(n),
    tval   = qt(0.975, df = n - 1),
    ci_low = mean - tval * se,
    ci_high = mean + tval * se,
    .groups = "drop"
  )

# --- 2) Human vs Human ---
cor_human_human <- human_mouse_log2fc_samples %>%
  filter(organism == "Human") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, human_symbol, log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = log2fc) %>%
      select(-human_symbol)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Human vs. Human",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup()
  

# --- 3) Mouse vs Mouse ---
cor_mouse_mouse <- human_mouse_log2fc_samples %>%
  filter(organism == "Mouse") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, human_symbol, log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = log2fc) %>%
      select(-human_symbol)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Mouse vs. Mouse",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup()


#Unite
correlation_long_all_mouse_human_timepoints_summary = cor_human_mouse_timepoint %>% 
  mutate(timepoint = timepoint_mouse) %>% 
  bind_rows(
    cor_human_human,
    cor_mouse_mouse)

correlation_long_all_mouse_human_timepoints_summary

#Plot
correlation_long_all_mouse_human_timepoints_summary_plot = correlation_long_all_mouse_human_timepoints_summary %>% 
  ggplot() +
  aes(x = as.numeric(timepoint),
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1,
                data = . %>% 
                  filter(comparison != "Human vs. Human")) +
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                # angle = 45,
                hjust = 1,
                linewidth = 1,
                data = . %>% 
                  filter(comparison == "Human vs. Human")) +
  scale_x_continuous(breaks = c(1, 3, 7), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01), 
                     # limits = c(-0.2, 1)
                     ) +
  theme_vaxgo() +
  scale_color_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Mouse vs. Human" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = paste0("Correlation between organisms\n", gene_set_name, ", ", filename),
       x = "",
       y = "Pearson p, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 

correlation_long_all_mouse_human_timepoints_summary_plot


correlation_long_all_mouse_human_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", paste(filename, gene_set_name, "correlation_timepoints_summary_plot_square.png", sep = "_")),
         width = 3,
         height = 3)

```

#### BTMs

##### Scatterplot

```{r fig.height=5, fig.width=5}
#Import files
dge_btm_mean_process_human = 
  readRDS(file = here("tables", "hepatitisb_engerixb_human_gsea_btm_mean.rds"))
dge_btm_mean_process_mouse =
  readRDS(file = here("tables", "hepatitisb_engerixb_mouse_gsea_btm_mean.rds"))

#Merge tables
dge_btm_mean_process_human_long = dge_btm_mean_process_human %>% 
  pivot_longer(cols = -c(timepoint, process, group, subgroup), 
               names_to = "participant_id", 
               values_to = "log2fc") %>% 
  mutate(organism = "Human")

dge_btm_mean_process_mouse_long = dge_btm_mean_process_mouse %>% 
  pivot_longer(cols = -c(timepoint, process, group, subgroup), 
               names_to = "participant_id", 
               values_to = "log2fc") %>% 
  mutate(organism = "Mouse")

dge_btm_mean_process_samples = dge_btm_mean_process_human_long %>%
  bind_rows(dge_btm_mean_process_mouse_long) %>%
  filter(timepoint %in% c(1, 3, 7))

dge_btm_mean_process_samples %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_gsea_btm_sample_mean.rds"))

dge_btm_mean_process_samples = 
  readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_gsea_btm_sample_mean.rds"))

dge_btm_mean_process = dge_btm_mean_process_samples %>% 
  group_by(organism, timepoint, process) %>% 
  summarize(mean_log2fc = median(log2fc),
            group,
            subgroup) %>%
  ungroup() %>% 
  distinct() %>% 
  pivot_wider(names_from = "organism",
              values_from = "mean_log2fc")

#Save
# dge_btm_mean_process %>% 
#   saveRDS(file = here("tables", "hepatitisb_dge_btm_mean_process_mouse_human.rds"))
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_gsea_btm_mean.rds"))
```


```{r fig.height=5, fig.width=5}
#Import
dge_btm_mean_process =
  readRDS(file = here("tables", "hepatitisb_dge_btm_mean_process_mouse_human.rds"))

# Calculate the correlation
cor_data <- dge_btm_mean_process %>%
  # Group the data by timepoint
  group_by(timepoint) %>% 
  
  # For each timepoint, run Pearson correlation test between Mouse and Human
  summarize(
    cor_test = list(cor.test(Mouse, Human, method = "spearman")), # store full test result as a list
    label_pos_x = max(Mouse, na.rm = TRUE),  # position for label in the plot (x = max Mouse value)
    label_pos_y = max(Human, na.rm = TRUE),  # position for label in the plot (y = max Human value)
    .groups = "drop"
  ) %>%
  
  # Extract correlation coefficient (r) and p-value from cor_test, and create a label
  mutate(
    cor_value = map_dbl(cor_test, ~ .x$estimate),  
    p_value   = map_dbl(cor_test, ~ .x$p.value),  
    label = paste0(
      "r = ", round(cor_value, 2),      
      ", p = ", signif(p_value, 3)   
    )
  )
```

```{r fig.height=8, fig.width=5}

human_mouse_log2fc_BTMs_avg_wide_scatterplot = dge_btm_mean_process %>% 
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human,
      color = group) +
  
  #Add points
  geom_point() +
  
#Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Facet
  facet_wrap(~paste("Day", timepoint), 
             ncol = 1, 
             scales = "free", axes = "all") +
  
  #Color
  scale_color_manual(values = immune_colors) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  
  #Add correlation metrics (Pearson)
  geom_label(data = cor_data,
             aes(x = label_pos_x, y = label_pos_y, label = label),
             inherit.aes = FALSE,
             fill = "white",
             hjust = 1,
             vjust = 1,
             size = 3)+

  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "right",
        legend.direction = "vertical") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       color = "BTM group",
       title= "Correlation plot, EngerixB,\nHuman vs Mouse")

#Save 
human_mouse_log2fc_BTMs_avg_wide_scatterplot %>% 
  ggsave(filename = here("Figures", "hepatitisb_engerixb_human_mouse_gsea_btm_mean_scatterplot.png"),
         height = 8,
         width = 5)
human_mouse_log2fc_BTMs_avg_wide_scatterplot
```

Single timepoint

```{r fig.height=3, fig.width=3}
human_mouse_log2fc_BTMs_scatterplot_singletimepoint = dge_btm_mean_process %>% 
  filter(timepoint == 7) %>% 
  #Plot
  ggplot() +
  aes(x = Mouse,
      y = Human) +
  
  #Add points
  geom_point(aes(fill = group),
             color = "black", size = 2, pch = 21) +
  
  ylim(-0.4, 0.4) +
  xlim(-0.4, 0.4) +
#Add lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_hline(yintercept = 0, alpha = 1, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 1, linetype = 2) +
  
  #Facet
  facet_wrap(~paste("Day", timepoint), 
             ncol = 1, 
             scales = "free", axes = "all") +
  
  #Color
  scale_color_manual(values = immune_colors) +
  scale_fill_manual(values = immune_colors) +
  
  #Add linear models
  stat_smooth(method = "lm", geom = "line", color = "black", alpha = 0.5, se = F, size = 1) +
  
  #Add correlation metrics (Pearson)
  geom_label(data = cor_data %>% 
               filter(timepoint == 7),
             aes(x = 0.2, y = 0.3, label = label),
             inherit.aes = FALSE,
             fill = "white",
             hjust = 1,
             vjust = 1,
             size = 3)+

  #Theme
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
        axis.text.y = element_text(size = 10),
        strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
  theme(legend.position = "none",
        legend.direction = "vertical") +
  labs(x = "Mouse L2FC",
       y = "Human L2FC",
       color = "BTM group",
       title= "Correlation plot, EngerixB,\nHuman vs Mouse")

human_mouse_log2fc_BTMs_scatterplot_singletimepoint

#Save 
human_mouse_log2fc_BTMs_scatterplot_singletimepoint %>% 
  ggsave(filename = here("Figures", "hepatitisb_engerixb_human_mouse_gsea_btm_mean_scatterplot_singletimepoint.png"),
         height = 3,
         width = 3)
```

##### Heatmap

```{r}
#Matrix data
matrix_data = dge_btm_mean_process %>% 
  select(process, Human_Day1, Mouse_Day2) %>% 
  drop_na() %>% 
  column_to_rownames("process") %>% 
  as.matrix()

# Column and row annotation
######## Annotations
annotation_heatmap = tribble(~condition, ~timepoint, ~organism,
        "Human_Day1", 1, "Human",
        "Mouse_Day2", 2, "Mouse")

#Arrange columns in the same order as the matrix
ann_cols_heatmap = matrix_data %>% 
  colnames() %>% 
  data.frame() %>% 
  rename(condition = '.') %>% 
  inner_join(annotation_heatmap, by = "condition") %>% 
  distinct() %>% 
  arrange(condition)

#Arrange matrix columns 
mat = matrix_data %>% 
  as.data.frame() %>% 
  select(c(ann_cols_heatmap$condition %>% unique())) %>% 
  as.matrix() 

#Column to rownames
ann_cols_heatmap = ann_cols_heatmap %>% 
  column_to_rownames("condition")

#Annotate rows
ann_rows_heatmap = mat %>% 
  rownames() %>% 
  as.data.frame() %>% 
  rename("process" = ".") %>% 
  inner_join(dge_btm_mean_process %>% 
               select(process, group, subgroup), 
             by = "process") %>% 
  column_to_rownames("process")
  

#Check dimensions
dim(mat)
dim(ann_cols_heatmap)

#Colors from required.R
btm_immune_groups
colors_immune

```

```{r fig.width= 6, fig.height=20}
#Heatmap annotation
ha = HeatmapAnnotation(df = ann_cols_heatmap, 
                       col = colors_immune,
                       gp = gpar(col = "black"), 
                       simple_anno_size = unit(0.3, "cm"),
                       annotation_name_side = "right")

row_ha = HeatmapAnnotation(df = ann_rows_heatmap,
                       col = btm_immune_groups,
                       which= "row",
                       simple_anno_size = unit(0.2, "cm"),
                       show_annotation_name = T,
                       show_legend = T)

# Values colors
col_fun <- colorRamp2(c(-2,0,2), c("#45efc6", "white", "#165289"))

file = "BTM_Modules_Saureus_Mouse_Human"

width_image = 25
height_image = 30
width = unit(1, "cm")
height = unit(25, "cm")
column_names_gp = gpar(fontsize = 8)
row_names_gp = gpar(fontsize = 8)
rect_gp = gpar(col = "black", lwd = 0.5)

fileimageheatmap_grouped = here("Figures", paste0(file, sep ="_", "Complexheatmap.png"))

png(file = fileimageheatmap_grouped, 
    width = width_image, 
    height = height_image, 
    units="cm", 
    res=900)

heatmap_plot = Heatmap(mat,
                        top_annotation = ha, 
                       left_annotation = row_ha,
                        show_row_names = TRUE,
                        row_names_side = "right",
                        show_heatmap_legend = TRUE,
                        heatmap_legend_param = list(direction = "horizontal"),
                        col = col_fun,
                        show_column_names = TRUE,
                        column_title = file,
                        column_title_gp = gpar(fontsize = 10, fontface = "bold"),
                        column_names_rot = 45,
                        column_names_gp = column_names_gp,
                        rect_gp = rect_gp,
                        na_col = "black",
                        name = "Mean Log2FC",
                        cluster_rows = TRUE,
                        row_names_gp = row_names_gp, row_km = 2,
                        row_dend_width = unit(3, "cm"),
                        row_split = NULL,
                        column_split = NULL,
                        row_gap = unit(2, "mm"),
                        cluster_columns = TRUE,
                        column_gap = unit(8, "mm"),
                        show_column_dend = F,
                        show_row_dend = T, 
                        row_dend_side = "left",
                        column_dend_side = "top",
                        clustering_distance_rows = "euclidean",
                        border = TRUE,
                        width = width,
                        height = height)

heatmap_plot = draw(heatmap_plot, 
     annotation_legend_side = "left", 
     heatmap_legend_side = "left")

heatmap_plot

dev.off()


```

# 3. FIT prediction

## Run FIT on a training dataset

```{r}
#Get log2fc mean
GSE19668_input2FIT = GSE19668_counts_log2_summary_annotated_probes %>% 
  ungroup() %>% 
    filter(timepoint == 2) %>% 
  select(participant_id, MM.Entrez = entrezgene_id, EffectSize = log2fc) %>% 
  group_by(MM.Entrez) %>% 
  summarize(mean_log2fc = mean(EffectSize))

#Save
GSE19668_input2FIT %>% 
  write_csv(file = here("FIT_Mouse2man", "GSE19668_day2_FIT_mouse.csv"))

#Run FIT
GSE19668_input2FIT_day2 = FIT.mouse2man::FIT(MouseFile = here("FIT_Mouse2man", "GSE19668_day2_FIT_mouse.csv"), DataType = "rnaseq")
```

### Unite log2FC values from mouse, human, and FIT

```{r fig.height=5, fig.width=8}
#Bind rows
mouse_human_fit_log2fc = bind_rows(GSE19668_input2FIT_day2 %>%
                                     #Get DEGs for FIT
                                     filter((CI_low != 0 | CI_high != 0) & !is.na(CI_low) & !is.na(CI_high)) %>%  
                                   select(human_symbol = Human.symbol,
                                          log2fc = FIT_prediction) %>%
                                     mutate(organism = "FIT"),
                                   
                                   #Get DEGs for Human
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_human <= 0.05) %>% 
                                     select(human_symbol, log2fc = Human) %>%
                                     mutate(organism = "Human"),
                                   
                                   #Get DEGs for Mouse
                                   human_mouse_saureus_log2fc_avg_wide %>%
                                     filter(padj_mouse <= 0.05) %>% 
                                     select(human_symbol, log2fc = Mouse) %>%
                                     mutate(organism = "Mouse"))


#Pivot wider
mouse_human_fit_log2fc_wide = GSE19668_input2FIT_day2 %>%
            select(human_symbol = Human.symbol,
                   FIT_prediction) %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Human),
            by = "human_symbol") %>% 
  inner_join(human_mouse_saureus_log2fc_avg_wide %>%
            select(human_symbol, Mouse),
            by = "human_symbol")

#Save
mouse_human_fit_log2fc %>% 
  saveRDS(file = here("tables", "mouse_human_fit_log2fc_long_SAureus.rds"))

mouse_human_fit_log2fc_wide %>% 
  saveRDS(file = here("tables", "mouse_human_fit_log2fc_wide_SAureus.rds"))
```

## Distribution

### Import tables

```{r fig.height=5, fig.width=8}
#Import
mouse_human_fit_log2fc = readRDS(file = here("tables", "mouse_human_fit_log2fc_long_SAureus.rds"))
mouse_human_fit_log2fc_wide = readRDS(file = here("tables", "mouse_human_fit_log2fc_wide_SAureus.rds"))

GSE33341_metadata = readRDS(file = here("tables", "GSE33341_metadata.rds"))
GSE19668_metadata = readRDS(file = here("tables", "GSE19668_metadata.rds"))
```

### Plot histogram

```{r fig.height=5, fig.width=8}
#Plot histogram
mouse_human_fit_log2fc_histogram = mouse_human_fit_log2fc %>% 
  ggplot() +
  aes(x = log2fc,
      fill = organism) +
  geom_density(aes(y = after_stat(count),
                   color = organism),
                  position = position_identity(), 
               alpha = 0.1,
               size = 1,
              show.legend = T) +
  scale_fill_manual(values = c(FIT = "#4361ee",
                               Mouse = "#4DBBD5FF",
                               Human = "#546E7AFF"))+
  scale_color_manual(values = c(FIT = "#4361ee",
                                Mouse = "#4DBBD5FF",
                                Human = "#546E7AFF"))+
  theme_vaxgo() +
  theme(axis.text.x = element_text(angle = 0),
        legend.position = "top",
        legend.key = element_blank(),
        strip.text.x = element_text(color = "black", size = 10),
        strip.text.y = element_text(color = "black", angle = 0),
        panel.spacing.y = unit(0.8, "lines"),
        strip.background.y = element_rect(fill = NA, color = NA)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(limits = c(-4, 4)) +
  labs(color = "",
       fill = "")

mouse_human_fit_log2fc_histogram

#Faceted
mouse_human_fit_log2fc_histogram_faceted = mouse_human_fit_log2fc_histogram +
  facet_wrap(~organism,
             scales = "free_y")


#Plot scatter plot
mouse_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = Mouse,
      y = Human) +
#All DEGs
geom_point(color = "#4DBBD5FF",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "Mouse L2FC",
         y = "Human L2FC")

mouse_human_scatterplot


fit_human_scatterplot = mouse_human_fit_log2fc_wide %>% 
  ggplot() + 
  aes(x = FIT_prediction,
      y = Human) +
#All DEGs
geom_point(color = "#4361ee",
           alpha = 1.5,
               size = 1,
               pch = 21) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 0, linetype = 2, linewidth = 0.2) +
  
  scale_x_continuous(limits = c(-8, 8)) +
#Theme
theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size =10),
          axis.text.y = element_text(size = 10),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5)) +
    theme(legend.position = "none",
          legend.direction = "horizontal") +
    labs(x = "FIT L2FC",
         y = "Human L2FC")

#Histograms
(((mouse_human_fit_log2fc_histogram + mouse_human_fit_log2fc_histogram_faceted) + 
    plot_layout(widths = c(0.5, 1), guides = "collect") +
    plot_annotation(theme = theme(legend.position = "top")))  /
    #Scatter plots
    (mouse_human_scatterplot + fit_human_scatterplot)) +
  plot_layout(heights = c(0.5, 1))

```



# 3. Functional analysis

## Immune
BTMs correlation, intra- and inter-species

### Mean Log2FC and GSEA
Compare means and NES
Import
```{r}
#Import datasets ------
#Import DGE output
all_degs_human_mouse = readRDS(here("tables", "hepatitisb_engerixb_human_mouse_dge_limma_degs.rds")) %>% 
  mutate(pathogen = "HepatitisB")
#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

filename = "hepatitisb_engerixb_human_mouse"

#Clean metadata 
annotation_conditions = all_degs_human_mouse %>% 
  select(condition, timepoint, model = organism) %>% 
  distinct() 

# Gene set enrichment analysis (GSEA) --------
#Get calculate means -----
#Annotate genes by BTM 

dge_btm_process_genes = all_degs_human_mouse %>% 
  rename(symbol = human_symbol) %>% 
  ungroup() %>% 
  inner_join(btm_annotation_genes, by = "symbol") %>% 
  filter(immune_set == "BTM by Modules") %>%
  filter(organism %in% c("Human", "Mouse")) 

dge_btm_process_genes %>% 
  saveRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))

dge_btm_mean_process = dge_btm_process_genes %>% 
  group_by(organism, timepoint, process) %>% 
  summarize(mean_log2fc = mean(mean_l2fc),
            median_log2fc = median(mean_l2fc),
            sd_log2fc = sd(mean_l2fc),
            var_log2fc = var(mean_l2fc),
            organism, pathogen,
            group, subgroup) %>%
  distinct() %>% 
  ungroup() %>% 
  mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) 

#Print
dge_btm_mean_process %>% 
  saveRDS(file = here("tables", paste0(filename, "dge_btm_mean_process.rds")))
```

```{r}
#Run GSEA for DGE outputs-----
#Human ImmuneGO/BTMs
term2gene = btm_annotation_genes %>% 
  select(process, genes = symbol)

#Clean table for GSEA standard
df_genes2GSEA <- all_degs_human_mouse %>% 
  distinct() %>% 
  select(organism, pathogen, timepoint, 
         genes = human_symbol, log2fold_change = mean_l2fc)  %>% 
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
   mutate(condition = paste0(organism,"_", pathogen, "_", timepoint)) %>% 
  select(condition, genes, log2fold_change)

#Run multiple GSEA and unite means
gsea_btm_output = autoGSEA(df = df_genes2GSEA,
         TERM2GENE = term2gene,
         geneset_name = "BTM")


gsea_btm_results = gsea_btm_output$gsea %>% 
  as.data.frame() %>% 
  clean_names() %>% 
  rename(process = id) %>%
  #Unite means and GSEA
  inner_join(dge_btm_mean_process, by = join_by("condition", "process"))

gsea_btm_output %>% 
  filter(condition == "Human_HepatitisB_7")
  
  
#Save
gsea_btm_results %>% 
    saveRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
```

#### Plot distributions

```{r fig.height=12, fig.width=10}
#Import ========================
gsea_btm_results = readRDS(here("tables", paste0(filename, "_gsea_btm_results.rds")))
dge_btm_mean_process = readRDS(file = here("tables", paste0(filename, "dge_btm_mean_process.rds")))
dge_btm_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))


#If timepoints match between ref and target, use this
dge_btm_process_genes_matched = dge_btm_process_genes


#Pivot to wide GSEA + mean
dge_btm_gsea_mean_wide = gsea_btm_results %>% 
  select(organism, timepoint, process, group, subgroup, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  group_by(timepoint, process, subgroup, group, set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_")

#Arrange process by log2fc, choose best timepoint for human as reference
process_order_log2fc = dge_btm_mean_process %>% 
  select(process, timepoint, organism, mean_log2fc, group) %>%
  filter(timepoint == 7,
         organism == "Human") %>% 
  group_by(organism, timepoint, group) %>% 
  arrange(-mean_log2fc) %>%
  pull(process) %>% 
  unique() 

dge_btm_process_ordered = dge_btm_mean_process %>% 
  ungroup() %>% 
    mutate(group = fct_relevel(as.factor(group), c(immune_order)),
           process = fct_relevel(as.factor(process), c(process_order_log2fc)) %>% fct_rev()) 

#Label axis labels with significant p
# labels_process <- dge_btm_gsea_mean_wide %>%
#   ungroup() %>% 
#   distinct(process,
#            # p_adjust_Human, 
#            p_adjust_Mouse) %>%
#   mutate(label = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
#                          "bold", "plain"),
#          label_color = ifelse(p_adjust_Human <= 0.10 & p_adjust_Mouse <= 0.10, 
#                          "#4DBBD5FF", "gray50"))  %>% 
#   ungroup() %>% 
#   distinct() %>% 
#   mutate(process = fct_relevel(process, process_order_log2fc)) %>% 
#   filter(!is.na(p_adjust_Human))
  

#Plot
dge_btm_process_distribution = dge_btm_process_genes_matched %>% 
  mutate(timepoint = if_else(organism == "Mouse", as.character(as.numeric(timepoint)/24), timepoint)) %>% 
    mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0("Day", timepoint),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(y = process,
      x = mean_l2fc,
      color = organism) +
  facet_grid(~group~timepoint, scales = "free", space = "free") +
  geom_vline(xintercept = 0, color = "black") +
  geom_point(pch = "|", alpha = 1, 
             show.legend = F,
             size = 3) +
  stat_summary(
  aes(color = organism),
  fun.data = mean_se,
  geom = "errorbar",
  width = 0.5,
  size = 1) +
  
  stat_summary(
    aes(fill = organism), 
    fun = mean,             
    color = "black",
    geom = "point",             
    size = 2,    
    pch = 21,  
    alpha = 1,
    show.legend = T 
  ) +
  
  xlim(-2, 2) +
  theme_vaxgo() +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), 
        axis.ticks.y = element_line(color = "gray", alpha = 0.5),
        legend.position = "top",
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8, 
                                   # face = c(labels_process$label), 
                                   # color = c(labels_process$label_color)
                                   ),
        axis.line.y = element_blank(),
        panel.spacing.y = unit(0.3, "lines"),
        panel.grid.major.y = element_line(alpha = 0.5),
        panel.grid.major.x = element_line(alpha = 0.5)) +
  scale_fill_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
  scale_color_manual(values = c("Mouse" = "#4DBBD5FF",
                               "Human" = "#90A4AEFF")) +
    labs(color = "Organism",
       y = "",
       x = "Log2FC",
       fill = "Organism",
       title = "Mouse vs Human, BTMs genes, mean Log2FC") 

dge_btm_process_distribution 

dge_btm_process_distribution %>% 
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_process_distribution.png")),
         height = 12, 
         width = 10)


```


#### Correlation plot (Gene-level, BTMs, immune)

##### Plot genes by btm group
```{r fig.height=16, fig.width=8}
#Convert to wide
dge_btm_process_genes_wide = dge_btm_process_genes_matched  %>% 
  mutate(timepoint = if_else(organism == "Mouse", as.character(as.numeric(timepoint)/24), timepoint)) %>% 
    ungroup() %>%
  select(symbol, mean_l2fc, adj_p_val, organism, timepoint, process, group) %>% 
  distinct() %>% 
  group_by(organism, timepoint, process) %>% 
  pivot_wider(names_from = organism,
              values_from = c(mean_l2fc, adj_p_val)) %>% 
  ungroup() %>% 
  mutate(group = fct_relevel(group, c(immune_order)))  %>% 
  #Remove duplicates
  select(-process) %>% 
  distinct()

#Plot
dge_btm_process_genes_correlation_plot = dge_btm_process_genes_wide %>% 
    mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0("Day", timepoint),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(x = mean_l2fc_Mouse,
      y = mean_l2fc_Human,
      labels = symbol) +
    #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  facet_grid(~group~timepoint, scales = "free") +
  #Plot all points
  geom_point(aes(fill = group,
                 color = group),
             size = 2,
             alpha = 0.2,
             pch = 21) +
  #Plot only significant in both
  geom_point(aes(fill = group),
             data = . %>% filter(adj_p_val_Human <= 0.05 & adj_p_val_Mouse <= 0.05), 
             color = "black",
             size = 2,
             alpha = 1,
             pch = 21) +

  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  guides(color = F) +
  labs(fill = "Immune group",
       y = "Human (Mean Log2FC)",
       x = "Model (Mean Log2FC)",
       title = "Genes by BTM",
       subtitle = "Genes with adjusted p-value <= 0.05 in both organisms.") +
  ylim(-2, 2) +
  xlim(-2, 2) +
  
  geom_text_repel(aes(label = symbol), 
                  min.segment.length = 0,
                  data = . %>% filter(adj_p_val_Human <= 0.05 & adj_p_val_Mouse <=0.05) %>% 
                    distinct(), 
                  size = 2, 
                  segment.size = 0.2,  
                  lineheight = 0.8, 
                  max.overlaps = 10) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.text = element_text(size = 8, hjust = 0)
          )

#Save results
dge_btm_process_genes_correlation_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_process_genes_correlation_plot_facetted.png")),
         width = 8,
         height = 16)

#Print
dge_btm_process_genes_correlation_plot
```


#### Correlation plot (Module-level, BTMs, immune)
Import
```{r}
#Import results -------
gsea_results = readRDS(here("tables", "hepatitisb_engerixb_human_mouse_gsea_btm_results.rds"))
dge_btm_process_genes = readRDS(file = here("tables", paste0(filename, "_dge_btm_process_genes.rds")))
dge_btm_mean_process = readRDS(file = here("tables", paste0(filename, "dge_btm_mean_process.rds")))

gsea_results_matched = gsea_results %>% mutate(timepoint = if_else(organism == "Mouse", as.character(as.numeric(timepoint)/24), timepoint))
dge_btm_process_genes_matched = dge_btm_process_genes  %>% mutate(timepoint = if_else(organism == "Mouse", as.character(as.numeric(timepoint)/24), timepoint))

```


Plot correlation
```{r fig.height=5, fig.width=7}
#Mouse vs Human
#Pivot to wide GSEA + mean
dge_btm_gsea_mean_wide = gsea_results_matched %>% 
  select(organism, timepoint, process, group, set_size, nes, mean_log2fc, p_adjust, qvalue) %>% 
  bind_rows(dge_btm_mean_process %>% 
              select(organism, process, timepoint, group, mean_log2fc) %>% 
               filter(organism == "Human")) %>% 
  group_by(timepoint, process, group, set_size) %>%
  pivot_wider(names_from = organism,
    values_from = c(mean_log2fc, nes, p_adjust, qvalue, set_size),
    names_sep = "_") %>% 
  ungroup()


dge_btm_process_genes_wide = dge_btm_process_genes_matched  %>% 
    ungroup() %>%
  select(symbol , mean_l2fc, adj_p_val, organism, timepoint, process, group) %>% 
  distinct() %>% 
  group_by(organism, timepoint, process) %>% 
  pivot_wider(names_from = organism,
              values_from = c(mean_l2fc, adj_p_val)) %>% 
  ungroup() %>% 
  mutate(group = fct_relevel(group, c(immune_order))) %>% 
  drop_na(mean_l2fc_Human)


```


```{r fig.height=5, fig.width=7}
#Calculate correlation -----

# Means vs means
cor_labels <- dge_btm_process_genes_wide %>%
  select(process, timepoint, group, Mouse = mean_l2fc_Mouse, Human = mean_l2fc_Human) %>% 
  group_by(timepoint) %>% 
  summarise(test_result = list(cor.test(x = Human, 
                                        y = Mouse, 
                                        use = "pairwise.complete.obs", 
                                        method = "pearson")), 
            .groups = "drop") %>%
  mutate(cor_value = purrr::map_dbl(test_result, "estimate"),
         p_value = purrr::map_dbl(test_result, "p.value")) %>%
  select(-test_result) %>%
  # Labels
  mutate(r_squared = cor_value^2,
    p_label = ifelse(p_value < 0.01,
                     "p = <0.01",
                     paste0("p = ", round(p_value, 2))),
    cor_label = paste0(
      "r = ", round(cor_value, 2),
      ", R² = ", round(r_squared, 2),
      ", ", p_label)) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0("Day", timepoint),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)])))

```

Plot means
```{r fig.height=3, fig.width=9}
dge_btm_mean_group_correlation_df = dge_btm_gsea_mean_wide %>% 
  ungroup() %>% 
  mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0("Day", timepoint),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  drop_na(mean_log2fc_Mouse)

#Plot correlation plot (means)
dge_btm_mean_group_correlation_plot =  dge_btm_mean_group_correlation_df %>% 
  ggplot() +
  aes(x = mean_log2fc_Mouse,
      y = mean_log2fc_Human) +
    #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
  geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
  geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
  geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  facet_wrap(~timepoint, nrow = 1) +
  #Plot all points
  geom_point(aes(fill = group,
                 color = group),
             size = 2,
             alpha = 0.2,
             pch = 21) +
  geom_smooth(color = "black", method = lm, se = F) +
  #Plot only significant in both
  geom_point(aes(fill = group),
             data = . %>% filter(p_adjust_Mouse <= 0.25) %>%
                    mutate(diff = mean_log2fc_Human - mean_log2fc_Mouse) %>%
                     group_by(timepoint) %>% 
                    slice_min(n = 20, abs(diff)), 
             color = "black",
             size = 2,
             alpha = 1,
             pch = 21) +

  scale_fill_manual(values = immune_colors) +
  scale_color_manual(values = immune_colors) +
  guides(color = F) +
  labs(fill = "Immune group",
       y = "Human (Mean Log2FC)",
       x = "Model (Mean Log2FC)",
       subtitle = "Labels: significantly enriched modules in mice only.") +
  ylim(-0.5, 0.5) +
  xlim(-0.5, 0.5 ) +
  
  #Add correlation metrics
  geom_label(data = cor_labels,
            aes(x = 0, y = Inf, label = cor_label),
            fill= "white",
            hjust = 0.5, vjust = 1, inherit.aes = FALSE, size = 3) +
  #Label only the top 10 processes with least difference in log2fc
  geom_text_repel(aes(label = process),
                  min.segment.length = 0,
                  data = . %>% filter(p_adjust_Mouse <= 0.25) %>%
                    mutate(diff = mean_log2fc_Human - mean_log2fc_Mouse) %>%
                    group_by(timepoint) %>% 
                    slice_min(n = 20, abs(diff)), 
                  size = 1.5,
                  segment.size = 0.2,
                  lineheight = 0.8,
                  max.overlaps = 20) +
  #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "right",
          legend.direction = "vertical",
          legend.text = element_text(size = 8, hjust = 0)
          )

#Print
dge_btm_mean_group_correlation_plot

#Save results
dge_btm_mean_group_correlation_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_mean_group_correlation_plot.png")),
         width = 9,
         height = 3)
```




##### Facet by group

Mean log2fc
```{r fig.height=10, fig.width=8}
dge_btm_mean_group_correlation_plot_facetted = dge_btm_gsea_mean_wide %>% 
  mutate(timepoint = as.character(timepoint)) %>% 
    mutate(group = fct_relevel(group, c(immune_order)),
         process = str_wrap(process, width = 15)) %>% 
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0("Day", timepoint),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot() +
  aes(x = mean_log2fc_Mouse,
      y = mean_log2fc_Human,
      color = group) +
  facet_grid(~group~timepoint, scales = "free") +
  #Guide lines
  geom_abline(alpha = 0.2, linetype = 2) +
    geom_abline(alpha = 0.2, linetype = 2, slope = -1) +
    geom_hline(yintercept = 0, alpha = 0.2, linetype = 2) +
    geom_vline(xintercept = 0, alpha = 0.2, linetype = 2) +
  
    #Plot all
    geom_point(aes(fill = group),
               color = "black",
               size = 2,
               alpha = 0.2,
               pch = 21) +
    #Plot only significant in both
    geom_point(aes(fill = group),
               data = . %>% filter(p_adjust_Mouse <= 0.25), 
               color = "black",
               size = 2,
               alpha = 1,
               pch = 21) +
    geom_text_repel(aes(label = process),
                    color = "black",
                  min.segment.length = 0,
                  data = . %>% filter( p_adjust_Mouse <= 0.25) %>%
                    mutate(diff = mean_log2fc_Human - mean_log2fc_Human) %>%
                    group_by(timepoint, group) %>% 
                    slice_min(n = 5, abs(diff)) %>% 
                    ungroup(), 
                  size = 2, 
                  segment.size = 0.2,  
                  lineheight = 0.8, 
                  max.overlaps = 20) +
    scale_fill_manual(values = immune_colors) +
    scale_color_manual(values = immune_colors) +
    guides(color = F) +
    labs(fill = "Immune group",
         y = "Human (Log2FC)",
         x = "Model (Log2FC)") +
    #Theme
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
          strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5),
          legend.position = "top",
          legend.direction = "horizontal",
          legend.text = element_text(size = 8, hjust = 0))

dge_btm_mean_group_correlation_plot_facetted

#Save results
dge_btm_mean_group_correlation_plot_facetted %>%
  ggsave(filename = here("Figures", paste0(filename, "_dge_btm_mean_group_correlation_plot_facetted.png")),
         width = 8,
         height = 11)
  
```

## Sample-wise analysis
Annotate genes log2fc table with BTMs
```{r}
#Import BTMs annotation table
btm_annotation_genes = read_csv(file = here("tables", "btm_annotation_genes.csv"))

#Example input
samples_log2fc = readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_exprs_samples_log2fc.rds"))

#Annotate genes by BTM 
results_genes_btm_df = samples_log2fc %>% 
  ungroup() %>% 
  select(symbol = human_symbol, participant_id, timepoint, log2fc, organism) %>% 
  inner_join(btm_annotation_genes, by = "symbol")

#Pivot wider
dge_btm_mean_process = results_genes_btm_df %>%
  mutate(condition = paste0(participant_id, "_Day", timepoint)) %>% 
  group_by(participant_id, timepoint, process) %>% 
  summarize(mean_log2fc = mean(log2fc)) %>%
  pivot_wider(names_from = "participant_id", values_from = "mean_log2fc") %>% 
  inner_join(btm_annotation_genes %>% 
               select(process, group, subgroup) %>% 
               distinct(),
             by = "process") %>% 
  ungroup()

#Print
dge_btm_mean_process

#Save
dge_btm_mean_process %>% saveRDS(file = here("Tables", paste0(filename, "_dge_btm_samples_log2fc.rds")))

```

Import data from different organisms

```{r}
#Genes L2FC table annotated with BTMs
dge_btm_mean_process_samples = readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_dge_btm_samples_log2fc.rds")) 

#Import metadata
mouse_human_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_metadata.rds"))

#Calculate the mean log2fc by module
dge_btm_mean_process_sample_long = results_genes_btm_df %>% 
  group_by(organism, participant_id, timepoint, process) %>% 
  summarize(group,
            subgroup, 
            mean_log2fc = mean(log2fc, na.rm = T)) %>% 
  distinct() %>% 
  ungroup()

#Pivot wider
dge_btm_mean_process_sample_wide = dge_btm_mean_process_sample_long %>% 
  group_by(organism, timepoint, process) %>% 
  pivot_wider(names_from = "participant_id", 
              values_from = "mean_log2fc")
  
```

### Summarize inter-individual correlation

#### 1 individual vs 1 individual

```{r}
# --- 1) Mouse vs Human, matched timepoints ---
#Get timepoints
timepoints_human <- unique(dge_btm_mean_process_sample_long %>% filter(organism == "Human") %>% pull(timepoint))
timepoints_mouse <- unique(dge_btm_mean_process_sample_long %>% filter(organism == "Mouse") %>% pull(timepoint))

#Matched timepoints
timepoint_combinations <- tibble(timepoint_human = timepoints_human,
                                      timepoint_mouse = timepoints_mouse)

# Loop
cor_human_mouse_timepoint <- timepoint_combinations %>%
  pmap_dfr(function(timepoint_human, timepoint_mouse) {
    
    # pegar os participantes do timepoint específico
    human_df <- dge_btm_mean_process_sample_long %>%
      filter(organism == "Human", timepoint == timepoint_human) %>%
      select(participant_id, process, mean_log2fc)
    
    mouse_df <- dge_btm_mean_process_sample_long %>%
      filter(organism == "Mouse", timepoint == timepoint_mouse) %>%
      select(participant_id, process, mean_log2fc)
    
    # manter apenas genes em comum
    common_process <- intersect(human_df$process, mouse_df$process)
    human_df <- human_df %>% filter(process %in% common_process)
    mouse_df <- mouse_df %>% filter(process %in% common_process)
    
    # pivot para wide
    human_wide <- human_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    mouse_wide <- mouse_df %>% pivot_wider(names_from = participant_id, values_from = mean_log2fc)
    
    # IDs de participantes
    human_ids <- setdiff(colnames(human_wide), "process")
    mouse_ids <- setdiff(colnames(mouse_wide), "process")
    
    # calcular correlação para todos pares humano x mouse
    expand.grid(human = human_ids, mouse = mouse_ids, stringsAsFactors = FALSE) %>%
      mutate(
        cor_test = map2(human, mouse, ~{
          x <- human_wide[[.x]]
          y <- mouse_wide[[.y]]
          # remover NAs correspondentes
          keep <- complete.cases(x, y)
          cor.test(x[keep], y[keep], method = "pearson")
        }),
        r       = map_dbl(cor_test, ~.x$estimate),
        p_value = map_dbl(cor_test, ~.x$p.value),
        ci_low  = map_dbl(cor_test, ~.x$conf.int[1]),
        ci_high = map_dbl(cor_test, ~.x$conf.int[2])
      ) %>%
      select(-cor_test) %>%
      mutate(timepoint_human = timepoint_human %>% as.numeric(),
             timepoint_mouse = timepoint_mouse %>% as.numeric())
  }) %>% 
  group_by(timepoint_human, timepoint_mouse) %>%
  summarize(
    comparison = "Human vs. Mouse",
    mean = mean(r, na.rm = TRUE),
    sd   = sd(r, na.rm = TRUE),
    n      = sum(!is.na(r)),
    se     = sd / sqrt(n),
    tval   = qt(0.975, df = n - 1),
    ci_low = mean - tval * se,
    ci_high = mean + tval * se,
    .groups = "drop"
  )

cor_human_mouse_timepoint

# --- 2) Human vs Human (entre timepoints) ---
cor_human_human <- dge_btm_mean_process_sample_long %>%
  filter(organism == "Human") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, process, mean_log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = mean_log2fc) %>%
      select(-process)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Human vs. Human",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup() %>% 
  mutate(timepoint = as.numeric(timepoint), 
         timepoint_human = timepoint)
  

# --- 3) Mouse vs Mouse ---
cor_mouse_mouse <- dge_btm_mean_process_sample_long %>%
  filter(organism == "Mouse") %>%
  group_by(timepoint) %>%
  group_modify(~{
    df <- .x %>%
      select(participant_id, process, mean_log2fc) %>%
      pivot_wider(names_from = participant_id, values_from = mean_log2fc) %>%
      select(-process)
    
    # Calculate correlation
    cor_mat <- cor(df, use = "pairwise.complete.obs", method = "pearson")
    
    # Long format
    cor_long <- as.data.frame(cor_mat) %>%
      rownames_to_column("participant_1") %>%
      melt(id.vars = "participant_1", variable.name = "participant_2", value.name = "correlation") %>%
      filter(participant_1 != participant_2) %>%
      rowwise() %>%
      mutate(pair = paste(sort(c(participant_1, participant_2)), collapse = "_")) %>%
      ungroup() %>%
      distinct(pair, .keep_all = TRUE)
    
    # Summary
    tibble(
      comparison = "Mouse vs. Mouse",
      mean = mean(cor_long$correlation, na.rm = TRUE),
      sd   = sd(cor_long$correlation, na.rm = TRUE),
      n    = nrow(cor_long),
      se   = sd / sqrt(n),
      tval = qt(0.975, df = n - 1),
      ci_low  = mean - tval * se,
      ci_high = mean + tval * se
    )
  }) %>%
  ungroup() %>% 
  mutate(timepoint = as.numeric(timepoint), 
         timepoint_mouse = timepoint)


#Unite
correlation_long_all_btm_mouse_human_timepoints_summary = cor_human_mouse_timepoint %>% 
  mutate(timepoint = timepoint_mouse) %>% 
  bind_rows(
    cor_human_human,
    cor_mouse_mouse)

#Print
correlation_long_all_btm_mouse_human_timepoints_summary

correlation_long_all_btm_mouse_human_timepoints_summary %>% 
  saveRDS(file = here("tables", paste0(filename, "_samples_btm_correlation_all_timepoints_summary.rds")))


#Scatterplot
correlation_long_all_btm_mouse_human_timepoints_summary_plot = correlation_long_all_btm_mouse_human_timepoints_summary %>% 
  ggplot() +
  aes(x = timepoint,
      y = mean,
      group = comparison,
      color = comparison,
      fill = comparison) +
  geom_line(size = 1) +
  
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.8,
                linewidth = 1,
                data = . %>% 
                  filter(comparison != "Human vs. Human")) +
  geom_textline(aes(label = comparison), 
                size = 3, 
                vjust = -0.5,
                hjust = 0.5,
                angle = 90,
                linewidth = 1,
                data = . %>% 
                  filter(comparison == "Human vs. Human")) +
  
  
  geom_errorbar(aes(ymin = ci_low, 
                  ymax = ci_high), width = 0.2,
                 size = 0.5) +
  geom_point(pch = 21, color = "black", size = 2) +
  scale_x_continuous(breaks = c(0, 1, 3, 7), expand = c(0.01, 0.01)) +
  scale_y_continuous(expand = c(0.01, 0.01),
                     # limits = c(0.4, 1)
                     ) +
  theme_vaxgo() +
  scale_color_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  scale_fill_manual(values = c("Human vs. Mouse" = "#4361ee",
                                "Mouse vs. Mouse" = "#4DBBD5FF",
                                "Human vs. Human" = "#90A4AEFF")) +
  labs(title = paste0("Correlation between organisms,\nBTMs, ", filename),
       x = "Day",
       y = "Pearson r, mean",
       color = "Comparison") +
  theme(legend.position = "none") +
  guides(fill = "none") 

correlation_long_all_btm_mouse_human_timepoints_summary_plot

correlation_long_all_btm_mouse_human_timepoints_summary_plot %>% 
  ggsave(filename = here("Figures", paste0(filename, "_samples_btm_correlation_all_timepoints_summary_plot.png")),
         width = 3,
         height = 3)

  
```




# 4. Performance tests
## ROC Curve
Do for one direction (up and then down)
Do one sided t test for mouse, fit, and human
Positive control: Human
Neg control: Random permutation


Import data
```{r}
#Import =============
#Import ImmuneGO ontologies
ImmuneGO_BTM_grouped_genes = read_csv(file = here("VaxGO", "ImmuneGO_BTM_grouped_genes.csv"))
#Import metadata
human_mouse_metadata = readRDS(file = here("tables", "hepatitisb_engerixb_human_mouse_metadata.rds"))

#Import DEGs table
all_degs_human = readRDS(file = here("tables", "hepatitisb_engerixb_human_dge_limma_degs.rds")) %>% 
  mutate(timepoint = str_remove(comparison_a, "Time"))
all_degs_mouse = readRDS(file = here("tables", "hepatitisb_engerixb_mouse_degs_mouse.rds")) %>% 
  mutate(timepoint = str_remove(condition, "Engerix_") %>% str_remove("_PBMC") %>% as.numeric(),
         timepoint = timepoint /24,
         timepoint = timepoint %>% as.character())


#Pre-process data
dge_long_genes_matched = all_degs_human  %>% 
  bind_rows(all_degs_mouse) %>% 
  rename(symbol = human_symbol, log2fc = mean_l2fc, adj_p_val = adj_p_val)  %>% 
  inner_join(human_mouse_metadata %>%
               group_by(organism, timepoint) %>%
               summarize(n = n()) %>% 
               mutate(timepoint = as.character(timepoint)),
               by = join_by("organism", "timepoint")) %>% 
  filter(timepoint == 7)




#Input ======================
dge_long_genes = dge_long_genes_matched

#All genes
genes_roc_allgenes = dge_long_genes
#Filter immune genes
genes_roc_allgenes_immunegenes = dge_long_genes %>% 
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))

#Only DEGs
degs_genes_human = dge_long_genes %>%
  filter(adj_p_val <= 0.05,
         organism == "Human") %>%
  select(symbol, timepoint)

#Only immune DEGs
genes_roc_DEGs_all = dge_long_genes %>%
  inner_join(degs_genes_human,
             by = join_by("symbol", "timepoint"))

#Only immune DEGs
genes_roc_DEGs_immune = dge_long_genes %>%
  inner_join(degs_genes_human,
             by = join_by("symbol", "timepoint")) %>%
  #Filter immune genes only
  filter(symbol %in% c(ImmuneGO_BTM_grouped_genes$genes %>% unique()))

```


Inputs
```{r fig.height=4, fig.width=10}
set.seed(123)
######### Input -----
# filename = "hepatitisb_engerixb"
# gene_set_filename = "All_OnlyImmune"
# genes_df_input = bind_rows(
#   genes_roc_allgenes %>% mutate(model_type = "Mouse"),
#   genes_roc_allgenes_immunegenes %>% mutate(model_type = "Immune")
# )

filename = "hepatitisb_engerixb"
gene_set_filename = "HumanDEGs_OnlyImmune"
genes_df_input = bind_rows(
  genes_roc_DEGs_all %>% mutate(model_type = "Mouse"),
  genes_roc_DEGs_immune %>% mutate(model_type = "Immune")
)

```

Calculate ROC curve
```{r fig.height=5, fig.width=8}
#=====================================================
# Compute probabilities
#=====================================================

dge_long_genes_up = genes_df_input %>% 
  mutate(
    df = n - 1,
    pval_one_sided = pt(t, df, lower.tail = FALSE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t > 0, "up", "not_up"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  select(-adj_p_val) %>%
  rename(adj_p_val = pval_one_sided_adj)

dge_long_genes_up %>% 
  group_by(organism, timepoint, direction) %>%
  summarise(n = n(), .groups = "drop")

dge_long_genes_down = genes_df_input %>% 
  mutate(
    df = n - 1,
    pval_one_sided = pt(t, df, lower.tail = TRUE),
    pval_one_sided_adj = p.adjust(pval_one_sided, method = "BH"),
    direction = if_else(pval_one_sided_adj <= 0.05 & t < 0, "down", "not_down"),
    prob = 1 - pval_one_sided_adj
  ) %>%
  select(-adj_p_val) %>%
  rename(adj_p_val = pval_one_sided_adj)

#=====================================================
# Assign human truth
#=====================================================
truth_human_up = dge_long_genes_up %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

truth_human_down = dge_long_genes_down %>% 
  select(symbol, organism, timepoint, adj_p_val, direction) %>% 
  filter(organism == "Human") %>% 
  distinct() %>% 
  pivot_wider(names_from = organism, values_from = direction) %>% 
  rename(truth = Human) %>% 
  mutate(organism = "Human")

#=====================================================
#Generate reference distributions
#=====================================================
perfect_prediction_up = truth_human_up %>%
  mutate(prob = if_else(truth == "up", 1, 0),
         model = "Human") %>%
  rename(pred = truth)

uniform_prediction_up = perfect_prediction_up %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()), organism = "Uniform",
         model = "Uniform")

perfect_prediction_down = truth_human_down %>%
  mutate(prob = if_else(truth == "down", 1, 0),
          model = "Human") %>%
  rename(pred = truth)

uniform_prediction_down = perfect_prediction_down %>%
  group_by(timepoint) %>% 
  mutate(prob = runif(n()), organism = "Uniform",
         model = "Uniform")

#=====================================================
# Combine models (Mouse + Immune + Uniform + Perfect)
#=====================================================
models_up = dge_long_genes_up %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = if_else(model_type == "Immune", "Immune", organism)) %>%
  bind_rows(perfect_prediction_up, uniform_prediction_up)

models_down = dge_long_genes_down %>% 
  filter(organism != "Human") %>% 
  mutate(pred = direction,
         model = if_else(model_type == "Immune", "Immune", organism)) %>%
  bind_rows(perfect_prediction_down, uniform_prediction_down)

#=====================================================
# Classify probabilities 
#=====================================================
classified_df_probs_up = truth_human_up %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_up %>% 
               select(symbol, timepoint, pred, prob, model),
             by = join_by(symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("up", "not_up"))
  ) %>% 
  distinct()

classified_df_probs_down = truth_human_down %>% 
  select(symbol, timepoint, truth) %>%
  inner_join(models_down %>% 
               select(symbol, timepoint, pred, prob, model),
             by = join_by(symbol, timepoint)) %>% 
  mutate(
    truth = factor(truth, levels = c("down", "not_down"))
  ) %>% 
  distinct()


#=====================================================
#Compute ROC 
#=====================================================

roc_curve_up_notup = classified_df_probs_up %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  filter(any(truth == "up")) %>%
  roc_curve(truth, prob) %>%
  mutate(comparison = "Up") %>% 
  ungroup()


roc_curve_down_notdown = classified_df_probs_down %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  filter(any(truth == "down")) %>%    
  roc_curve(truth, prob) %>%
  mutate(comparison = "Down") %>% 
  ungroup()

#=====================================================
#Compute AUC 
#=====================================================
auc_results_up = classified_df_probs_up %>% 
  mutate(truth = fct_relevel(truth, "up", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  filter(any(truth == "up")) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Up") %>% 
  ungroup() %>% 
  select(model, timepoint, auc, comparison)

auc_results_down = classified_df_probs_down %>% 
  mutate(truth = fct_relevel(truth, "down", after = 0)) %>% 
  filter(model %in% c("Mouse", "Immune", "Uniform", "Human", "FIT")) %>%
  group_by(model, timepoint) %>%
  filter(any(truth == "down")) %>%
  roc_auc(truth, prob) %>%
  rename(auc = .estimate) %>%
  mutate(comparison = "Down") %>% 
  ungroup() %>% 
  select(model, timepoint, auc, comparison)

#Plot curves ------------------
#PLOT ROC Curve
roc_curve_plot = roc_curve_up_notup %>% 
  bind_rows(roc_curve_down_notdown) %>%
  mutate(timepoint_num = readr::parse_number(timepoint),
    timepoint = paste0("Day" ,timepoint),
    timepoint = factor(timepoint, levels = unique(timepoint[order(timepoint_num)]))) %>% 
  ggplot(aes(x = 1 - specificity, 
             y = sensitivity, 
             color = model)) +
  geom_path(size = 1, alpha = 1) +
  geom_abline(lty = 3) +
  theme_vaxgo() +
  facet_grid(~fct_rev(comparison)~timepoint) +
  scale_color_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "Human" = "#90A4AEFF",
    "Immune" = "#4361ee",
    "Uniform" = "gray80"
    )) +
  labs(
  x = "FPR",
  y = "TPR",
  color = "Model") + 
  theme(
    legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(n.breaks = 3) +
  scale_y_continuous(n.break = 3)

roc_curve_plot

# Visualize AUC
auc_plot <- bind_rows(auc_results_up, 
                      auc_results_down) %>% 
  mutate(model = factor(model, levels = c("Perfect", "Mouse", "Immune", "Uniform")),
         comparison = factor(comparison, levels = c("Up", "Down"))) %>%
  
  ggplot(aes(x = auc, 
             y = fct_reorder(timepoint, as.numeric(timepoint)) %>% fct_rev(), 
             fill = model)) +
  geom_col(width = 1,
           position = position_dodge(1)) +
  
  # Add a reference line for random prediction (AUC = 0.5)
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black", linewidth = 0.5) +
  
  geom_text(aes(label = round(auc, 2)),
            position = position_dodge2(1),
            hjust = -0.2, size = 2, vjust = 0.5) +
  
  facet_grid(comparison ~ ., scales = "free_y") +
  
  labs(
    x = "AUC",
    y = "Day",
    fill = "Model"
  ) +
    scale_fill_manual(values = c(
    "Mouse" = "#4DBBD5FF",
    "Immune" = "#4361ee",
    "Human" = "#90A4AEFF",
    "Uniform" = "gray80"
    )) +
  theme_vaxgo()  +
  theme(
    legend.position = "top",
    strip.text = element_text(size = 10),
    strip.placement = "outside",
    # strip.background.x = element_rect(fill = "gray95", color = NA), 
    strip.text.y = element_text(size = 10, angle = 0, hjust = 0),
    axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5, vjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.spacing.x = unit(0.5, "lines"),
    legend.key = element_blank()
  ) +
  scale_x_continuous(expand = c(0,0), limits = c(0,1))

roc_auc_curve_plot = ((roc_curve_plot) |
  (auc_plot + theme(legend.position = "none"))) +
  plot_layout(guides = "collect", widths = c(1, 0.8))+
  plot_annotation(theme = theme(legend.position = "top", plot.title = element_text(hjust = 0.5)),
                  title = paste(filename, gene_set_filename))

roc_auc_curve_plot %>%
  ggsave(filename = here("Figures", paste0(filename, "_performance_roc_", gene_set_filename, ".png")),
         width = 5,
         height = 5)

roc_auc_curve_plot
```


5